### 2017-10-03 manuscript.Rmd
initial commit: add files; add outline;

---
title: | 
       | Containers and R:
       | the Rockerverse and beyond
author:
  - name: Daniel Nüst
    affiliation: Institute for Geoinformatics, University of Münster 
    address:
    - Heisenbergstr. 2, 48149 Münster, Germany
    email: daniel.nuest@uni-muenster.de
abstract: >
  The Rocker project provides widely-used Docker images for R across different application scenarios. In this article we describe the downstream projects building upon Rocker and look beyond towards other container solutions and libraries to interact with them from R.
preamble: >
   \usepackage{longtable}
output: rticles::rjournal_article
---

## Introduction

\citep{R}

### Containerization

### Rocker

## R images and containers

### Rocker

### Bioconductor

### MRO

### Renjin

### pqR

### Singularity

### OCI

## Packages

### Automation and packaging

- `containerit`
- `dockertest`
- `liftr`

### Control and provisioning

- `harbor`
- `docker`
- `RSelenium`
- `googleComputeEngineR`
- `analogsea`

## Downstream Applications

### Deployment and processing

- `plumber`
- `batchtools`

### Cross-R testing

Show how you can run the same script in multiple R variants in containers (new contribution).
Ideally the images should be created using `containerit` and controlled with `harbor`/`docker`, i.e. via an R script.

### Workflows and reproducibility

...

### Scalable deployments

- Shiny talks @ useR!2017
- ...

\bibliography{RJreferences}

### 2017-10-03 manuscript.tex
initial commit: add files; add outline;

% !TeX root = RJwrapper.tex
\title{Containers and R:\\
the Rockerverse and beyond}
\author{by Daniel Nüst}

\maketitle

\abstract{%
The Rocker project provides widely-used Docker images for R across
different application scenarios. In this article we describe the
downstream projects building upon Rocker and look beyond towards other
container solutions and libraries to interact with them from R.
}

\subsection{Introduction}\label{introduction}

\citep{R}

\subsubsection{Containerization}\label{containerization}

\subsubsection{Rocker}\label{rocker}

\subsection{R images and containers}\label{r-images-and-containers}

\subsubsection{Rocker}\label{rocker-1}

\subsubsection{Bioconductor}\label{bioconductor}

\subsubsection{MRO}\label{mro}

\subsubsection{Renjin}\label{renjin}

\subsubsection{pqR}\label{pqr}

\subsubsection{Singularity}\label{singularity}

\subsubsection{OCI}\label{oci}

\subsection{Packages}\label{packages}

\subsubsection{Automation and packaging}\label{automation-and-packaging}

\begin{itemize}
\tightlist
\item
  \texttt{containerit}
\item
  \texttt{dockertest}
\item
  \texttt{liftr}
\end{itemize}

\subsubsection{Control and provisioning}\label{control-and-provisioning}

\begin{itemize}
\tightlist
\item
  \texttt{harbor}
\item
  \texttt{docker}
\item
  \texttt{RSelenium}
\item
  \texttt{googleComputeEngineR}
\item
  \texttt{analogsea}
\end{itemize}

\subsection{Downstream Applications}\label{downstream-applications}

\subsubsection{Deployment and
processing}\label{deployment-and-processing}

\begin{itemize}
\tightlist
\item
  \texttt{plumber}
\item
  \texttt{batchtools}
\end{itemize}

\subsubsection{Cross-R testing}\label{cross-r-testing}

Show how you can run the same script in multiple R variants in
containers (new contribution). Ideally the images should be created
using \texttt{containerit} and controlled with
\texttt{harbor}/\texttt{docker}, i.e.~via an R script.

\subsubsection{Workflows and
reproducibility}\label{workflows-and-reproducibility}

\ldots{}

\subsubsection{Scalable deployments}\label{scalable-deployments}

\begin{itemize}
\tightlist
\item
  Shiny talks @ useR!2017
\item
  \ldots{}
\end{itemize}

\bibliography{RJreferences}

\address{%
Daniel Nüst\\
Institute for Geoinformatics, University of Münster\\
Heisenbergstr. 2, 48149 Münster, Germany\\
}
\href{mailto:daniel.nuest@uni-muenster.de}{\nolinkurl{daniel.nuest@uni-muenster.de}}

### 2019-07-28 manuscript.Rmd
Update manuscript with suggested contents and responsibilities from #3

@@
 ---
-title: | 
-       | Containers and R:
-       | the Rockerverse and beyond
+title: "Containers and R: the Rockerverse and beyond"
 author:
@@
     email: daniel.nuest@uni-muenster.de
+  - name: Carl Boettiger
+    #affiliation: -
+    #email: -
+  - name: Robrecht Cannoodt
+    #affiliatin: -
+    #email: -
+  - name: Dirk Eddelbuettel
+    #affiliation: -
+    #email: -
+  - name: Mark Edmondson
+    #affiliation: -
+    #email: -
+  - name: Colin Fay
+    #affiliation: -
+    #email: -
+  - name: Ben Marwick
+    #affiliation: -
+    #email: -
+  - name: Noam Ross
+    #affiliation: -
+    #email: -
+  - name: Nan Xiao
+    #affiliation: -
+    #email: - 
 abstract: >
   The Rocker project provides widely-used Docker images for R across different application scenarios.
-In this article we describe the downstream projects building upon Rocker and look beyond towards other container solutions and libraries to interact with them from R.
+  This articles surveys describe downstream projects building upon Rocker.
+  We also look beyond Rocker to other projects connecting containerisation with R.
+  These use cases and the diversity of applications demonstrate the power of Rocker and containerisation for collaboration, effectivity, scalability, and transparency.
 preamble: >
@@
 
-\citep{R}
+The \citep{R} community keeps growing: the number of new packages on CRAN keeps on rising, meetups, conferences, online courses and unconferences prosper, and the uptake in education and industry increases [REFs].
+All this cements `R`'s role as the lingua franca of statistics, data visualisation, and computational research.
+Part of this success can be traced back to the _Rocker Project_ ([https://www.rocker-project.org/](https://www.rocker-project.org/)).
+The community project was introduced in 2014 and provides a number of [Docker](https://en.wikipedia.org/wiki/Docker_(software)) images for various use cases \citet{RJ-2017-065}.
+The considerable uptake and continued evolution of _Rocker_'s suite has lead to numerous projects extending or building upon _Rocker_ images ranging from reproducible research to production deployments.
+This article presents this _Rockerverse_ of projects.
+It also introduces related activities connecting the R language and environment with other containerisation solutions.
+The main contributions is a coherent picture of the current lay of the land of containers in, with, and for `R`.
+This diversity includes demonstrators, early prototypes, and mature projects.
 
-### Containerization
+## Containerization and Rocker
 
-### Rocker
+_Do we need a generic intro here?_
 
-## R images and containers
+## Images for alternative R distributions [@nuest]
 
-### Rocker
+- [MRO](https://github.com/nuest/mro-docker)
+- [Renjin](https://github.com/nuest/renjin-docker)
+- [pqR](https://github.com/nuest/pqr-docker)
+- [FastR](https://github.com/nuest/fastr-docker)
 
-### Bioconductor
+## Other containerisation platforms [NN]
 
-### MRO
+- R images for Singularity
+- Running Rocker images with
+  - Singularity
+  - CoreOS rkt?
+- nix?
 
-### Renjin
+## Non-Debian images [NN]
 
-### pqR
+- Alpine images
+- Images used by R-Hub (overlap with CI?)
+- [Windows](https://github.com/nuest/rocker-win) [@nuest]
+- https://github.com/jlisic/R-docker-centos
 
-### Singularity
+## Further Use-case Docker images [NN]
 
-### OCI
+- part of Rocker (if not mentioned in \citet{RJ-2017-065}) ?
+  - Binder
+  - GPU [@noamross, @cboettig]
+- [Bioconductor](http://bioconductor.org/help/docker/), https://github.com/Bioconductor/bioc_docker
+- Gigantum stack
+- ...
 
-## Packages
+## Rocker for Research Compendia [@benmarwick]
 
-### Automation and packaging
+...
+
+## Packages for automation and packaging [@nuest]
 
@@
 - `dockertest`
-- `liftr`
+- `liftr` for OS-level reproducibility [@nanx]
+- [`rize` for Shiny](https://github.com/cole-brokamp/rize)
+- [`dockerfiler`](https://github.com/ColinFay/dockerfiler)
 
-### Control and provisioning
+## Packages for control and provisioning [@nuest]
 
-- `harbor`
-- `docker`
+- [`stevedore`](https://github.com/richfitz/stevedore)
+- [`babelwhale`](https://cran.r-project.org/web/packages/babelwhale/index.html): Running a Docker from R with Singularity or Docker as back-end.
+This is really useful in HPC environments where a user might not have root access but is able to install Singularity instead of Docker.
+[@rcannood]
+- [`docker`](https://bhaskarvk.github.io/docker/)
 - `RSelenium`
-- `googleComputeEngineR`
-- `analogsea`
+- [`googleComputeEngineR`](https://cloudyr.github.io/googleComputeEngineR/) (function `docker_run`)
+- [`analogsea`](https://github.com/sckott/analogsea)
+- [`harbor`](https://github.com/wch/harbor/)
+- [`dockermachine`](https://github.com/cboettig/dockermachine)
 
-## Downstream Applications
+## Continuous integration and continuous delivery [@noamross, @ColinFay]
 
-### Deployment and processing
+- R-Hub
+- DevOps
+  - https://www.opencpu.org/posts/opencpu-with-docker/
+- `r-ci`: https://github.com/ColinFay/r-ci
+- [dynwrap](https://github.com/dynverse/dynwrap_containers/blob/master/.travis.yml) [@rcannood]
+  - For this project, we use travis-ci to build rocker-derived containers, test them, and only push them to docker hub (from travis-ci.org) if the integration tests succeed.
+- ...
 
-- `plumber`
-- `batchtools`
+## Common work environments [@noamross]
+
+- ...
+- in education
+  - ready to use learning environments
+  - [`r-db`](https://github.com/ColinFay/r-db) [@ColinFay]
+- [RCloud](https://github.com/att/rcloud/tree/master/docker) ?
+
+## Deployment, processing, cloud
+
+- Docker images for cloud services [@MarkEdmondson1234]
+- [`plumber`](https://www.rplumber.io/docs/hosting.html#docker)
+- `batchtools` \citep{Lang2017batchtools} can [schedule jobs with Docker Swarm](https://mllg.github.io/batchtools/reference/makeClusterFunctionsDocker.html)
+- scalable deployments, e.g. start with numerous Shiny talks mentioning Rocker at useR!2017
+- [dynmethods](https://github.com/dynverse/dynmethods) [@rcannood]: In order to evaluate ±50 computational methods which all used different environments (R, Python, C++, ...), we wrapped each of them in a docker container and can execute these methods from R.
+Again, all of these containers are being built on travis-ci, and will only be pushed to docker hub if the integration test succeeds.
+- [ShinyProxy](https://www.shinyproxy.io/) [creates a container](https://github.com/openanalytics/shinyproxy/blob/master/src/main/java/eu/openanalytics/services/DockerService.java#L388) for each user
+
+## R versions [@ColinFay]
+
+- run code with Versions of R: https://srv.colinfay.me/r-online
+- old R versions: https://github.com/rocker-org/rocker-versioned/issues/138
+- [semantic version tags for Rocker](https://github.com/rocker-org/rocker-versioned/#version-tags)
+- debugging memory problems across multiple versions: https://github.com/wch/r-debug
 
@@
 
-### Workflows and reproducibility
+## Discussion/outlook [all, please add bullet points]
 
-...
-
-### Scalable deployments
-
-- Shiny talks @ useR!2017
+- Missing pieces?
+- consolidation (e.g. via packages using `dockerfiler` and `stevedore`)
+- Common themes
+  - reproducibility
+- will knowledge about containers continue to spread?
+- what is needed for even more containers with R?
 - ...

### 2019-07-30 manuscript.Rmd
distribute content from "further images" section, add @karthik with holepunch
also add some more project links

@@
   - name: Ben Marwick
+    #affiliation: -
+    #email: -
+  - name: Karthik Ram
     #affiliation: -
@@
 - [FastR](https://github.com/nuest/fastr-docker)
+- [Bioconductor](http://bioconductor.org/help/docker/), https://github.com/Bioconductor/bioc_docker
 
@@
 
-## Further Use-case Docker images [NN]
-
-- part of Rocker (if not mentioned in \citet{RJ-2017-065}) ?
-  - Binder
-  - GPU [@noamross, @cboettig]
-- [Bioconductor](http://bioconductor.org/help/docker/), https://github.com/Bioconductor/bioc_docker
-- Gigantum stack
-- ...
-
 ## Rocker for Research Compendia [@benmarwick]
@@
 
-- `containerit`
-- `dockertest`
-- `liftr` for OS-level reproducibility [@nanx]
+- [`containerit`](https://github.com/o2r-project/containerit/)
+- [`dockertest`](https://github.com/traitecoevo/dockertest/)
+- [`liftr`](https://liftr.me/) for OS-level reproducibility [@nanx]
 - [`rize` for Shiny](https://github.com/cole-brokamp/rize)
@@
   - For this project, we use travis-ci to build rocker-derived containers, test them, and only push them to docker hub (from travis-ci.org) if the integration tests succeed.
-- ...
 
-## Common work environments [@noamross]
+## Common or public work environments [div]
 
-- ...
-- in education
-  - ready to use learning environments
+- Binder
+  - [`holepunch`](https://github.com/karthik/holepunch) streamlines making an R research compendium Binder-ready [@karthik]
+  - [`repo2docker`](https://repo2docker.readthedocs.io/en/latest/config_files.html) (`install.R`, `DESCRPTION`, `runtime.txt`) based on the Rocker images [@nuest]
+- GPU [@noamross, @cboettig]
+- Gigantum stack
+- in education (ready to use teaching/learning environments)
   - [`r-db`](https://github.com/ColinFay/r-db) [@ColinFay]
@@
 
-## Deployment, processing, cloud
+## Deployment, processing, cloud [div]
 
@@
 
-## R versions [@ColinFay]
+## R versions [@ColinFay, @nuest]
 
@@
 - debugging memory problems across multiple versions: https://github.com/wch/r-debug
-
-### Cross-R testing
-
-Show how you can run the same script in multiple R variants in containers (new contribution).
-Ideally the images should be created using `containerit` and controlled with `harbor`/`docker`, i.e. via an R script.
+- cross-R testing: Show how you can run the same script in multiple R variants in containers (new contribution).
+Ideally the images should be created using `containerit` and controlled with `stevedore`/`docker`, i.e.
+via an R script.
 

### 2019-08-12 manuscript.Rmd
Rough bioc brainstroming session for rockerverse paper.
- This session included @lshep and @nturaga

- Jot down all the points which might be relevant.

- Next sessions include, arrange , clean up and write

@@
 - [FastR](https://github.com/nuest/fastr-docker)
+
 - [Bioconductor](http://bioconductor.org/help/docker/), https://github.com/Bioconductor/bioc_docker
+
+- Most system dependencies come from rocker/r-ver and rocker/rstudio
+  as needed.
+
+- Bioconductor works on two release cycles every year.
+
+- Docker images based on rocker/rstudio for release and devel.
+
+	- Naming and tagging of images.
+
+- Community maintained docker images like proteomics and metabolomics
+  and cytometry. (Include community names L gatto, sneumann, m.jiang)
+
+- Linux nightly build replica (bioconductor_full).
+
+	- testing environment for users and developers/maintainers.
+
+	- Aim is to install MOST (if not all) Bioconductor packages.
+
+	- Better alternative than distributing specific packages in a
+      image.
+
+	- These images will be used for cloud based solutions in the near
+      future like AnVIL project.
+
+	- distributed on singularity as well.
+
+- Improve reproducibility of Bioconductor versions and R versions
+  which are closely tied together.
+
+- Ruby and a rakefile along with a templating engine to generate all
+  of our images once every 6 months. (link to src)
+
+	- Takes away the pain of hand crafting dockerfiles each release cycl
+
+	- recent work: improve src Dockerfiles, so that images are
+      smaller. Improvements like cleaning cache etc. Not installing
+      system dependencies once they have already been installed
+      rocker - images.
+
+- Docker hub allows us to build and distribute images to the
+  community.
+
+- Future:
+
+	- Teaching workshops, and they don't have to worry about AWS
+      working vs not.
+
+	- Docker images used for distributing Bioconductor workshop
+      material post conference.
 

### 2019-08-12 manuscript.Rmd
Lori's rough draft paragraph style
  - Overview of Bioconductor
  - Bioconductor release version closely correlated with R release version
  - Bioconductor images build on top of rocker images
  - Tag Bioconductor images with R_ver_Bioc_ver for current/past accessibility
  - Built and distribute through DockerHub
  - Community contributed images

@@
     #email: - 
+  - name: Lori Shepherd
+    affiliation: Roswell Park Cancer Institute
+    address:
+    - Elm & Carlton Streets, Buffalo, New York, 14263 USA
+    email: lori.shepherd@roswellpark.org
 abstract: >
@@
 - [Bioconductor](http://bioconductor.org/help/docker/), https://github.com/Bioconductor/bioc_docker
+
+#
+# Draft bullet point notes (brainstorming session)
+#
 
@@
       material post conference.
+
+#
+# Rough draft - they asked one sentance per line
+#
+
+Bioconductor is a free, open source and open development project aimed at the analysis and visualization of genetic and genomic data utilizing the R programming language [1]. 
+The project currently consists of approximately 1741 R packages and has a semiannual release subsequent to the Comprehensive R Archive Networks release of R [2]. 
+Biocondctor release versions are tightly correlated with a version of R. 
+Docker images allow availability of current and past versions of R/Bioconductor for convenience and reproducibility.
+
+The Bioconductor docker images are built on top of rocker/r-ver and rocker/rstudio.
+By utilizing rocker as a base image, the R version and majority of package system dependencies are already included in a Bioconductor base image.
+Bioconductor will install packages based on the selected R version and therefore leverage the rocker/rstudio and rocker/r-ver use of R version tagging. 
+Bioconductor selects the desired version of R from rocker, installs BiocManager which will default to the appropriate Bioconductor version, and creates a Bioconductor docker image with a informative tag (R_version_Bioc_version).
+The images are built and distributed to the community through DockerHub [3]. 
+All past and current combinations of R and Bioconductor are therefore accessible via a specific docker tag.
+
+The Bioconductor base image contains the minimial system depenedencies and R from rocker, and the BiocManager package for installation of packages configured for the proper version of Bioconductor given the installed R version.
+Bioconductor has several other docker images specific for various areas of research. 
+The Bioconductor core image installs the most downloaded Bioconductor packages based on collected statistics. 
+The other Bioconductor images are community maintained: Laurent Gatto maintains proteomics images, Steffen Neumann maintains metabolomics images, and the RGLab maintains cytometry images. 
+All community maintained images build on top of the Bioconductor base image and therefore indirectly the rocker images [4].
+
+#
+# Need to move these to bib file and properly format. 
+#
+
+[1]  Gentleman, Robert C.; Carey, Vincent J.; Bates, Douglas M.; Bolstad, Ben;
+Dettling, Marcel; Dudoit, Sandrine; Ellis, Byron; Gautier, Laurent; Ge,
+Yongchao; Gentry, Jeff; Hornik, Kurt; Hothorn, Torsten; Huber, Wolfgang; Iacus,
+Stefano; Irizarry, Rafael; Leisch, Friedrich; Li, Cheng; Maechler, Martin;
+Rossini, Anthony J.; Sawitzki, Gunther; Smith, Colin; Smyth, Gordon; Tierney,
+Luke; Yang, Jean Y. H.; Zhang, Jianhua (2004). "Bioconductor: open software
+development for computational biology and bioinformatics". Genome Biology. 5
+(10): R80. doi:10.1186/gb-2004-5-10-r80. PMC 545600
+
+[2] R Core Team (2019). R: A language and environment for statistical
+  computing. R Foundation for Statistical Computing, Vienna, Austria.
+  URL https://www.R-project.org/
+
+[3] ??DockerHub citation?
+
+[4] ??Reference our docker help page/github?
 

### 2019-08-14 manuscript.Rmd
Proper Rowsell affiliation

@@
   - name: Lori Shepherd
-    affiliation: Roswell Park Cancer Institute
+    affiliation: Roswell Park Comprehensive Cancer Center
     address:

### 2019-08-15 manuscript.Rmd
whitespace cleanup
light edits
- mention bioc website and github repository
- don't mention community maintainers by name, to avoid beuracracy of seeking
  approval from them; credit is on github and dockerhub
- italicize _Bioconductor_
- minor word-smithing

@@
 
-Bioconductor is a free, open source and open development project aimed at the analysis and visualization of genetic and genomic data utilizing the R programming language [1]. 
-The project currently consists of approximately 1741 R packages and has a semiannual release subsequent to the Comprehensive R Archive Networks release of R [2]. 
-Biocondctor release versions are tightly correlated with a version of R. 
-Docker images allow availability of current and past versions of R/Bioconductor for convenience and reproducibility.
+_Bioconductor_ is a free, open source and open development project for the analysis and comprehension of genomic data [1].
+The project consists of 1741 R software packages, as well as packages containing annotation or experiment data.
+_Bioconductor_ has a semiannual release cycle, each release is associated with a particular version of R.
+Docker images allow availability of current and past versions of _Bioconductor_ for convenience and reproducibility.
 
-The Bioconductor docker images are built on top of rocker/r-ver and rocker/rstudio.
-By utilizing rocker as a base image, the R version and majority of package system dependencies are already included in a Bioconductor base image.
-Bioconductor will install packages based on the selected R version and therefore leverage the rocker/rstudio and rocker/r-ver use of R version tagging. 
-Bioconductor selects the desired version of R from rocker, installs BiocManager which will default to the appropriate Bioconductor version, and creates a Bioconductor docker image with a informative tag (R_version_Bioc_version).
-The images are built and distributed to the community through DockerHub [3]. 
-All past and current combinations of R and Bioconductor are therefore accessible via a specific docker tag.
+_Bioconductor_ docker images are built on top of rocker/r-ver and rocker/rstudio.
+By using rocker as a base image, the R version and most package and system dependencies are included in a _Bioconductor_ base image.
+_Bioconductor_ installs packages based on the R version, and therefore leverage the rocker/rstudio and rocker/r-ver use of R version tagging.
+_Bioconductor_ selects the desired version of R from rocker, adds the BiocManager CRAN package for installing appropriate versions of _Bioconductor_ packages, and creates a _Bioconductor_ docker image with a informative tag (R\_version\_Bioc\_version).
+The images are summarized on the _Bioconductor_ web site [5], maintained in github [6], and available to the community through DockerHub [3].
+All past and current combinations of R and _Bioconductor_ are therefore accessible via a specific docker tag.
 
-The Bioconductor base image contains the minimial system depenedencies and R from rocker, and the BiocManager package for installation of packages configured for the proper version of Bioconductor given the installed R version.
-Bioconductor has several other docker images specific for various areas of research. 
-The Bioconductor core image installs the most downloaded Bioconductor packages based on collected statistics. 
-The other Bioconductor images are community maintained: Laurent Gatto maintains proteomics images, Steffen Neumann maintains metabolomics images, and the RGLab maintains cytometry images. 
-All community maintained images build on top of the Bioconductor base image and therefore indirectly the rocker images [4].
+_Bioconductor_ has several other docker images specific to various areas of research.
+The 'core' image installs the most commonly used _Bioconductor_ packages.
+Other _Bioconductor_ images for proteomics, metabolomics, and flow cytometry are community maintained.
+All community maintained images build on top of the _Bioconductor_ base image and therefore indirectly the rocker images [4].
 
@@
 [4] ??Reference our docker help page/github?
+
+[5] ?? https://bioconductor.org/help/docker/
+
+[6] ?? https://github.com/Bioconductor/bioc_docker
 

### 2019-08-19 manuscript.Rmd
Adding sections about
- Future

- bioconductor_full

- Cloud activities

- recent improvements to bioc_docker (audit and optimization)

@@
 
+	- kubernetes orchestrated
+
 - Improve reproducibility of Bioconductor versions and R versions
@@
 
-_Bioconductor_ is a free, open source and open development project for the analysis and comprehension of genomic data [1].
-The project consists of 1741 R software packages, as well as packages containing annotation or experiment data.
-_Bioconductor_ has a semiannual release cycle, each release is associated with a particular version of R.
-Docker images allow availability of current and past versions of _Bioconductor_ for convenience and reproducibility.
+_Bioconductor_ is a free, open source and open development project for
+the analysis and comprehension of genomic data [1]. The project
+consists of 1741 R software packages as of August 15th 2019, as well
+as packages containing annotation or experiment data. _Bioconductor_
+has a semiannual release cycle, each release is associated with a
+particular version of R. Docker images allow availability of current
+and past versions of _Bioconductor_ for convenience and
+reproducibility.
 
-_Bioconductor_ docker images are built on top of rocker/r-ver and rocker/rstudio.
-By using rocker as a base image, the R version and most package and system dependencies are included in a _Bioconductor_ base image.
-_Bioconductor_ installs packages based on the R version, and therefore leverage the rocker/rstudio and rocker/r-ver use of R version tagging.
-_Bioconductor_ selects the desired version of R from rocker, adds the BiocManager CRAN package for installing appropriate versions of _Bioconductor_ packages, and creates a _Bioconductor_ docker image with a informative tag (R\_version\_Bioc\_version).
-The images are summarized on the _Bioconductor_ web site [5], maintained in github [6], and available to the community through DockerHub [3].
-All past and current combinations of R and _Bioconductor_ are therefore accessible via a specific docker tag.
+_Bioconductor_ docker images are built on top of rocker/r-ver and
+rocker/rstudio. By using rocker as a base image, the R version and
+most package and system dependencies are included in a _Bioconductor_
+base image.  _Bioconductor_ installs packages based on the R version,
+and therefore leverage the rocker/rstudio and rocker/r-ver use of R
+version tagging. _Bioconductor_ selects the desired version of R from
+rocker, adds the BiocManager CRAN package for installing appropriate
+versions of _Bioconductor_ packages, and creates a _Bioconductor_
+docker image with an informative tag (R\_version\_Bioc\_version). The
+images are summarized on the _Bioconductor_ web site [5], maintained
+in github [6], and available to the community through DockerHub [3].
+All past and current combinations of R and _Bioconductor_ are
+therefore accessible via a specific docker tag.
 
-_Bioconductor_ has several other docker images specific to various areas of research.
-The 'core' image installs the most commonly used _Bioconductor_ packages.
-Other _Bioconductor_ images for proteomics, metabolomics, and flow cytometry are community maintained.
-All community maintained images build on top of the _Bioconductor_ base image and therefore indirectly the rocker images [4].
+_Bioconductor_ has several other docker images specific to various
+areas of research.  The 'core' image installs the most commonly used
+_Bioconductor_ packages.  Other _Bioconductor_ images for proteomics,
+metabolomics, and flow cytometry are community maintained.  All
+community maintained images build on top of the _Bioconductor_ base
+image and therefore indirectly the rocker images [4].
+
+Bioconductor's _bioc_docker_ software works using a templating engine
+based in ruby, has a source directory `src` and an output directory
+`out`.[*] This takes away in large part the repeated handcrafting of
+each Dockerfile produced by Bioconductor at each release cycle. Edits
+to the Dockerfile need to be made in the source directory, based on a
+config file, the output directory is populated with the appropriate
+Dockerfile for each specific to the various areas of research. A
+recent systematic "audit" of our docker images following best
+practices (as issued by the Docker website) led to a reduction in the
+number of layers and also size of the overall images being produced by
+Bioconductor. Some of the optimizations include removing cache,
+cleaning up after each layer produced by a `RUN` statement, and most
+importantly, not re-installing any dependencies already provided by
+rocker.
+
+Bioconductor now produces a Docker image to mimic the _Bioconductor_
+nightly build system on Linux. This image is called the
+`bioconductor_full` (based on rocker/rstudio). The goal of the Docker
+image is to install _all the system dependencies_ needed to install
+all of the Bioconductor packages. We achieve this goal to a large
+extent, and are able to greater than _99%_ of packages (1730 out of
+1741). This image allows developers to test their packages extensively
+by replicating errors on the nightly build system, and users to
+install most of the packages they need without dealing with system
+dependencies. There is the added bonus that it comes with an RStudio
+interface to ease users into using R with Docker, thanks to rocker.
+
+We found this to be a easier alternative to distributing installed
+_Bioconductor_ packages within Docker images as this is not entirely
+consistent with the way R/Bioconductor is used. Users usually install
+and add packages to their scripts in real time. It also keeps the
+Docker image smaller in size. Using R is almost never like a
+'workflow', which is defined by the pre-installed packages. We
+encourage users to mount a volume to their Bioconductor docker image,
+where all the installed (compiled) packages are stored in the external
+mountable volume. The library tree `.libPaths()` has
+`/usr/local/lib/R/host-site-library` as the first location searched
+for installed packages, which allows this functionality to work. This
+enables the docker image to be updated by using a `docker pull`
+command without the user having to worry about any packages saved on
+the image.
+
+The `bioconductor_full` image is forming the base for our cloud based
+solutions in upcoming future projects. We are currently able to host a
+Kubernetes orchestrated cluster on the Google Cloud Platform using the
+bioconductor_full image and the rocker/r-ver images. The distribution
+of computation is done through an R package called `RedisParam`
+(github: mtmorgan/RedisParam), where the `bioconductor_full` image is
+the master node and the `rocker/r-ver` images are the workers.
+
+We envision these Docker images being the primary way of teaching in
+the future. The goal is to let workshop participants pull a pre-built
+docker image for the specific worksop and run them on their local
+machines. As of right now, the same task is accomplished by creating
+and spinning up multiple AWS images (one per participant). The docker
+images can work in more fault tolerant, lasting, cost-effective
+manner, and be used by participants throughout their learning
+journey. The users can choose to scale their workshop skills using
+their cloud accounts to scale up with this provided docker image
+(using either the AWS Elastic Container service or Google Kubernetes
+Engine etc.)
 
@@
 [6] ?? https://github.com/Bioconductor/bioc_docker
+
+[*] The original work on `bioc_docker` was done by Dan Tanenbaum
 

### 2019-08-19 manuscript.Rmd
Nitesh: Add name to author list

@@
     email: lori.shepherd@roswellpark.org
+  - name: Nitesh Turaga
+    affiliation: Roswell Park Comprehensive Cancer Center
+    address:
+    - Elm & Carlton Streets, Buffalo, New York, 14263 USA
+    email: nitesh.turaga@roswellpark.org
 abstract: >

### 2019-08-19 manuscript.Rmd
remove bullet points

@@
 - [Bioconductor](http://bioconductor.org/help/docker/), https://github.com/Bioconductor/bioc_docker
-
-#
-# Draft bullet point notes (brainstorming session)
-#
-
-- Most system dependencies come from rocker/r-ver and rocker/rstudio
-  as needed.
-
-- Bioconductor works on two release cycles every year.
-
-- Docker images based on rocker/rstudio for release and devel.
-
-    - Naming and tagging of images.
-
-- Community maintained docker images like proteomics and metabolomics
-  and cytometry. (Include community names L gatto, sneumann, m.jiang)
-
-- Linux nightly build replica (bioconductor_full).
-
-    - testing environment for users and developers/maintainers.
-
-    - Aim is to install MOST (if not all) Bioconductor packages.
-
-    - Better alternative than distributing specific packages in a
-      image.
-
-    - These images will be used for cloud based solutions in the near
-      future like AnVIL project.
-
-    - distributed on singularity as well.
-
-	- kubernetes orchestrated
-
-- Improve reproducibility of Bioconductor versions and R versions
-  which are closely tied together.
-
-- Ruby and a rakefile along with a templating engine to generate all
-  of our images once every 6 months. (link to src)
-
-    - Takes away the pain of hand crafting dockerfiles each release cycl
-
-    - recent work: improve src Dockerfiles, so that images are
-      smaller. Improvements like cleaning cache etc. Not installing
-      system dependencies once they have already been installed
-      rocker - images.
-
-- Docker hub allows us to build and distribute images to the
-  community.
-
-- Future:
-
-    - Teaching workshops, and they don't have to worry about AWS
-      working vs not.
-
-    - Docker images used for distributing Bioconductor workshop
-      material post conference.
 

### 2019-08-19 manuscript.Rmd
One sentance per line

@@
 
-_Bioconductor_ is a free, open source and open development project for
-the analysis and comprehension of genomic data [1]. The project
-consists of 1741 R software packages as of August 15th 2019, as well
-as packages containing annotation or experiment data. _Bioconductor_
-has a semiannual release cycle, each release is associated with a
-particular version of R. Docker images allow availability of current
-and past versions of _Bioconductor_ for convenience and
-reproducibility.
+_Bioconductor_ is a free, open source and open development project for the analysis and comprehension of genomic data [1].
+The project consists of 1741 R software packages as of August 15th 2019, as well as packages containing annotation or experiment data.
+_Bioconductor_has a semiannual release cycle, each release is associated with a particular version of R.
+Docker images allow availability of current and past versions of _Bioconductor_ for convenience and reproducibility.
 
-_Bioconductor_ docker images are built on top of rocker/r-ver and
-rocker/rstudio. By using rocker as a base image, the R version and
-most package and system dependencies are included in a _Bioconductor_
-base image.  _Bioconductor_ installs packages based on the R version,
-and therefore leverage the rocker/rstudio and rocker/r-ver use of R
-version tagging. _Bioconductor_ selects the desired version of R from
-rocker, adds the BiocManager CRAN package for installing appropriate
-versions of _Bioconductor_ packages, and creates a _Bioconductor_
-docker image with an informative tag (R\_version\_Bioc\_version). The
-images are summarized on the _Bioconductor_ web site [5], maintained
-in github [6], and available to the community through DockerHub [3].
-All past and current combinations of R and _Bioconductor_ are
-therefore accessible via a specific docker tag.
+_Bioconductor_ docker images are built on top of rocker/r-ver and rocker/rstudio.
+By using rocker as a base image, the R version and most package and system dependencies are included in a _Bioconductor_ base image.
+_Bioconductor_ installs packages based on the R version, and therefore leverage the rocker/rstudio and rocker/r-ver use of R version tagging.
+_Bioconductor_ selects the desired version of R from rocker, adds the BiocManager CRAN package for installing appropriate
+versions of _Bioconductor_ packages, and creates a _Bioconductor_ docker image with an informative tag (R\_version\_Bioc\_version).
+The images are summarized on the _Bioconductor_ web site [5], maintained in github [6], and available to the community through DockerHub [3].
+All past and current combinations of R and _Bioconductor_ are therefore accessible via a specific docker tag.
 
-_Bioconductor_ has several other docker images specific to various
-areas of research.  The 'core' image installs the most commonly used
-_Bioconductor_ packages.  Other _Bioconductor_ images for proteomics,
-metabolomics, and flow cytometry are community maintained.  All
-community maintained images build on top of the _Bioconductor_ base
-image and therefore indirectly the rocker images [4].
+_Bioconductor_ has several other docker images specific to various areas of research.
+The 'core' image installs the most commonly used _Bioconductor_ packages.  Other _Bioconductor_ images for proteomics, metabolomics, and flow cytometry are community maintained.
+All community maintained images build on top of the _Bioconductor_ base image and therefore indirectly the rocker images [4].
 
-Bioconductor's _bioc_docker_ software works using a templating engine
-based in ruby, has a source directory `src` and an output directory
-`out`.[*] This takes away in large part the repeated handcrafting of
-each Dockerfile produced by Bioconductor at each release cycle. Edits
-to the Dockerfile need to be made in the source directory, based on a
-config file, the output directory is populated with the appropriate
-Dockerfile for each specific to the various areas of research. A
-recent systematic "audit" of our docker images following best
-practices (as issued by the Docker website) led to a reduction in the
-number of layers and also size of the overall images being produced by
-Bioconductor. Some of the optimizations include removing cache,
-cleaning up after each layer produced by a `RUN` statement, and most
-importantly, not re-installing any dependencies already provided by
-rocker.
+Bioconductor's _bioc_docker_ software works using a templating engine based in ruby, has a source directory `src` and an output directory `out`[*].
+This takes away in large part the repeated handcrafting of each Dockerfile produced by Bioconductor at each release cycle.
+Edits to the Dockerfile need to be made in the source directory, based on a config file, the output directory is populated with the appropriate Dockerfile for each specific to the various areas of research.
+A recent systematic "audit" of our docker images following best practices (as issued by the Docker website) led to a reduction in the number of layers and also size of the overall images being produced by Bioconductor.
+Some of the optimizations include removing cache, cleaning up after each layer produced by a `RUN` statement, and most importantly, not re-installing any dependencies already provided by rocker.
 
-Bioconductor now produces a Docker image to mimic the _Bioconductor_
-nightly build system on Linux. This image is called the
-`bioconductor_full` (based on rocker/rstudio). The goal of the Docker
-image is to install _all the system dependencies_ needed to install
-all of the Bioconductor packages. We achieve this goal to a large
-extent, and are able to greater than _99%_ of packages (1730 out of
-1741). This image allows developers to test their packages extensively
-by replicating errors on the nightly build system, and users to
-install most of the packages they need without dealing with system
-dependencies. There is the added bonus that it comes with an RStudio
-interface to ease users into using R with Docker, thanks to rocker.
+Bioconductor now produces a Docker image to mimic the _Bioconductor_ nightly build system on Linux.
+This image is called the `bioconductor_full` (based on rocker/rstudio).
+The goal of the Docker image is to install _all the system dependencies_ needed to install all of the Bioconductor packages.
+We achieve this goal to a large extent, and are able to greater than _99%_ of packages (1730 out of 1741).
+This image allows developers to test their packages extensively by replicating errors on the nightly build system, and users to install most of the packages they need without dealing with system dependencies.
+There is the added bonus that it comes with an RStudio interface to ease users into using R with Docker, thanks to rocker.
 
-We found this to be a easier alternative to distributing installed
-_Bioconductor_ packages within Docker images as this is not entirely
-consistent with the way R/Bioconductor is used. Users usually install
-and add packages to their scripts in real time. It also keeps the
-Docker image smaller in size. Using R is almost never like a
-'workflow', which is defined by the pre-installed packages. We
-encourage users to mount a volume to their Bioconductor docker image,
-where all the installed (compiled) packages are stored in the external
-mountable volume. The library tree `.libPaths()` has
-`/usr/local/lib/R/host-site-library` as the first location searched
-for installed packages, which allows this functionality to work. This
-enables the docker image to be updated by using a `docker pull`
-command without the user having to worry about any packages saved on
-the image.
+We found this to be a easier alternative to distributing installed _Bioconductor_ packages within Docker images as this is not entirely consistent with the way R/Bioconductor is used.
+Users usually install and add packages to their scripts in real time.
+It also keeps the Docker image smaller in size.
+Using R is almost never like a 'workflow', which is defined by the pre-installed packages.
+We encourage users to mount a volume to their Bioconductor docker image, where all the installed (compiled) packages are stored in the external mountable volume.
+The library tree `.libPaths()` has `/usr/local/lib/R/host-site-library` as the first location searched for installed packages, which allows this functionality to work.
+This enables the docker image to be updated by using a `docker pull` command without the user having to worry about any packages saved on the image.
 
-The `bioconductor_full` image is forming the base for our cloud based
-solutions in upcoming future projects. We are currently able to host a
-Kubernetes orchestrated cluster on the Google Cloud Platform using the
-bioconductor_full image and the rocker/r-ver images. The distribution
-of computation is done through an R package called `RedisParam`
-(github: mtmorgan/RedisParam), where the `bioconductor_full` image is
-the master node and the `rocker/r-ver` images are the workers.
+The `bioconductor_full` image is forming the base for our cloud based solutions in upcoming future projects.
+We are currently able to host a Kubernetes orchestrated cluster on the Google Cloud Platform using the bioconductor_full image and the rocker/r-ver images.
+The distribution of computation is done through an R package called `RedisParam` (github: mtmorgan/RedisParam), where the `bioconductor_full` image is the master node and the `rocker/r-ver` images are the workers.
 
-We envision these Docker images being the primary way of teaching in
-the future. The goal is to let workshop participants pull a pre-built
-docker image for the specific worksop and run them on their local
-machines. As of right now, the same task is accomplished by creating
-and spinning up multiple AWS images (one per participant). The docker
-images can work in more fault tolerant, lasting, cost-effective
-manner, and be used by participants throughout their learning
-journey. The users can choose to scale their workshop skills using
-their cloud accounts to scale up with this provided docker image
-(using either the AWS Elastic Container service or Google Kubernetes
-Engine etc.)
+We envision these Docker images being the primary way of teaching in the future.
+The goal is to let workshop participants pull a pre-built docker image for the specific worksop and run them on their local machines.
+As of right now, the same task is accomplished by creating and spinning up multiple AWS images (one per participant).
+The docker images can work in more fault tolerant, lasting, cost-effective manner, and be used by participants throughout their learning journey.
+The users can choose to scale their workshop skills using their cloud accounts to scale up with this provided docker image (using either the AWS Elastic Container service or Google Kubernetes Engine etc.)
 

### 2019-08-23 manuscript.Rmd
added some bits from me

@@
   - name: Dirk Eddelbuettel
-    #affiliation: -
+    #affiliation:
     #email: -
   - name: Mark Edmondson
-    #affiliation: -
-    #email: -
+    affiliation: IIH Nordic A/S, Google Developer Expert for GCP
+    email: mark@markedmondson.me
   - name: Colin Fay
@@
     #email: - 
-  - name: Lori Shepherd
-    affiliation: Roswell Park Comprehensive Cancer Center
-    address:
-    - Elm & Carlton Streets, Buffalo, New York, 14263 USA
-    email: lori.shepherd@roswellpark.org
-  - name: Nitesh Turaga
-    affiliation: Roswell Park Comprehensive Cancer Center
-    address:
-    - Elm & Carlton Streets, Buffalo, New York, 14263 USA
-    email: nitesh.turaga@roswellpark.org
 abstract: >
@@
 - [FastR](https://github.com/nuest/fastr-docker)
-
 - [Bioconductor](http://bioconductor.org/help/docker/), https://github.com/Bioconductor/bioc_docker
-
-#
-# Rough draft - they asked one sentance per line
-#
-
-_Bioconductor_ is a free, open source and open development project for the analysis and comprehension of genomic data [1].
-The project consists of 1741 R software packages as of August 15th 2019, as well as packages containing annotation or experiment data.
-_Bioconductor_has a semiannual release cycle, each release is associated with a particular version of R.
-Docker images allow availability of current and past versions of _Bioconductor_ for convenience and reproducibility.
-
-_Bioconductor_ docker images are built on top of rocker/r-ver and rocker/rstudio.
-By using rocker as a base image, the R version and most package and system dependencies are included in a _Bioconductor_ base image.
-_Bioconductor_ installs packages based on the R version, and therefore leverage the rocker/rstudio and rocker/r-ver use of R version tagging.
-_Bioconductor_ selects the desired version of R from rocker, adds the BiocManager CRAN package for installing appropriate
-versions of _Bioconductor_ packages, and creates a _Bioconductor_ docker image with an informative tag (R\_version\_Bioc\_version).
-The images are summarized on the _Bioconductor_ web site [5], maintained in github [6], and available to the community through DockerHub [3].
-All past and current combinations of R and _Bioconductor_ are therefore accessible via a specific docker tag.
-
-_Bioconductor_ has several other docker images specific to various areas of research.
-The 'core' image installs the most commonly used _Bioconductor_ packages.  Other _Bioconductor_ images for proteomics, metabolomics, and flow cytometry are community maintained.
-All community maintained images build on top of the _Bioconductor_ base image and therefore indirectly the rocker images [4].
-
-Bioconductor's _bioc_docker_ software works using a templating engine based in ruby, has a source directory `src` and an output directory `out`[*].
-This takes away in large part the repeated handcrafting of each Dockerfile produced by Bioconductor at each release cycle.
-Edits to the Dockerfile need to be made in the source directory, based on a config file, the output directory is populated with the appropriate Dockerfile for each specific to the various areas of research.
-A recent systematic "audit" of our docker images following best practices (as issued by the Docker website) led to a reduction in the number of layers and also size of the overall images being produced by Bioconductor.
-Some of the optimizations include removing cache, cleaning up after each layer produced by a `RUN` statement, and most importantly, not re-installing any dependencies already provided by rocker.
-
-Bioconductor now produces a Docker image to mimic the _Bioconductor_ nightly build system on Linux.
-This image is called the `bioconductor_full` (based on rocker/rstudio).
-The goal of the Docker image is to install _all the system dependencies_ needed to install all of the Bioconductor packages.
-We achieve this goal to a large extent, and are able to greater than _99%_ of packages (1730 out of 1741).
-This image allows developers to test their packages extensively by replicating errors on the nightly build system, and users to install most of the packages they need without dealing with system dependencies.
-There is the added bonus that it comes with an RStudio interface to ease users into using R with Docker, thanks to rocker.
-
-We found this to be a easier alternative to distributing installed _Bioconductor_ packages within Docker images as this is not entirely consistent with the way R/Bioconductor is used.
-Users usually install and add packages to their scripts in real time.
-It also keeps the Docker image smaller in size.
-Using R is almost never like a 'workflow', which is defined by the pre-installed packages.
-We encourage users to mount a volume to their Bioconductor docker image, where all the installed (compiled) packages are stored in the external mountable volume.
-The library tree `.libPaths()` has `/usr/local/lib/R/host-site-library` as the first location searched for installed packages, which allows this functionality to work.
-This enables the docker image to be updated by using a `docker pull` command without the user having to worry about any packages saved on the image.
-
-The `bioconductor_full` image is forming the base for our cloud based solutions in upcoming future projects.
-We are currently able to host a Kubernetes orchestrated cluster on the Google Cloud Platform using the bioconductor_full image and the rocker/r-ver images.
-The distribution of computation is done through an R package called `RedisParam` (github: mtmorgan/RedisParam), where the `bioconductor_full` image is the master node and the `rocker/r-ver` images are the workers.
-
-We envision these Docker images being the primary way of teaching in the future.
-The goal is to let workshop participants pull a pre-built docker image for the specific worksop and run them on their local machines.
-As of right now, the same task is accomplished by creating and spinning up multiple AWS images (one per participant).
-The docker images can work in more fault tolerant, lasting, cost-effective manner, and be used by participants throughout their learning journey.
-The users can choose to scale their workshop skills using their cloud accounts to scale up with this provided docker image (using either the AWS Elastic Container service or Google Kubernetes Engine etc.)
-
-#
-# Need to move these to bib file and properly format.
-#
-
-[1]  Gentleman, Robert C.; Carey, Vincent J.; Bates, Douglas M.; Bolstad, Ben;
-Dettling, Marcel; Dudoit, Sandrine; Ellis, Byron; Gautier, Laurent; Ge,
-Yongchao; Gentry, Jeff; Hornik, Kurt; Hothorn, Torsten; Huber, Wolfgang; Iacus,
-Stefano; Irizarry, Rafael; Leisch, Friedrich; Li, Cheng; Maechler, Martin;
-Rossini, Anthony J.; Sawitzki, Gunther; Smith, Colin; Smyth, Gordon; Tierney,
-Luke; Yang, Jean Y. H.; Zhang, Jianhua (2004). "Bioconductor: open software
-development for computational biology and bioinformatics". Genome Biology. 5
-(10): R80. doi:10.1186/gb-2004-5-10-r80. PMC 545600
-
-[2] R Core Team (2019). R: A language and environment for statistical
-  computing. R Foundation for Statistical Computing, Vienna, Austria.
-  URL https://www.R-project.org/
-
-[3] ??DockerHub citation?
-
-[4] ??Reference our docker help page/github?
-
-[5] ?? https://bioconductor.org/help/docker/
-
-[6] ?? https://github.com/Bioconductor/bioc_docker
-
-[*] The original work on `bioc_docker` was done by Dan Tanenbaum
 
@@
 - [`googleComputeEngineR`](https://cloudyr.github.io/googleComputeEngineR/) (function `docker_run`)
+  - I don't think `docker_run` is main focus for containers for this package, its more how it uses Docker to launch custom environments that can be templated (e.g.
+rstudio, shiny etc.) and for parrallisation e.g.
+`library(future)`  using GCE VMs as endpoints.
+[@MarkEdmondson1234]
 - [`analogsea`](https://github.com/sckott/analogsea)
@@
   - For this project, we use travis-ci to build rocker-derived containers, test them, and only push them to docker hub (from travis-ci.org) if the integration tests succeed.
+- Google Cloud Build [@MarkEdmondson1234]
+  - https://cloud.google.com/cloud-build/
+  - For me this is what makes Docker containers viable, as it builds the Dockerfiles on each GitHub commit.
+It couples with Google Container Registry to build private Docker images, for downstream applications.
 
@@
 - Docker images for cloud services [@MarkEdmondson1234]
+  - the most popular functionality for `googleComputeEngineR` is its use of Docker to enable parallel processing across VMS - [some demos here](https://cloudyr.github.io/googleComputeEngineR/articles/massive-parallel.html)
+- Google Cloud Run - CaaS (Containers as a Service) that lets you launch a Docker container without worrying about underlying infrastructure.
+An R implementation is shown here at [cloudRunR](https://github.com/MarkEdmondson1234/cloudRunR) which uses it to create a scalable R plumber API.
 - [`plumber`](https://www.rplumber.io/docs/hosting.html#docker)
@@
 - what is needed for even more containers with R?
+- the ability to move processing between services easily (e.g locally, one cloud providers VM, another cloud provider's Container-as-a-Service)
 - ...

### 2019-08-26 manuscript.Rmd
manuscript edits
- wording, with minimal changes to conten

--- 2019-08-19 manuscript.Rmd
@@
 
-_Bioconductor_ is a free, open source and open development project for the analysis and comprehension of genomic data [1].
+_Bioconductor_ is an open source, open development project for the analysis and comprehension of genomic data [1].
 The project consists of 1741 R software packages as of August 15th 2019, as well as packages containing annotation or experiment data.
@@
 
-_Bioconductor_ docker images are built on top of rocker/r-ver and rocker/rstudio.
-By using rocker as a base image, the R version and most package and system dependencies are included in a _Bioconductor_ base image.
-_Bioconductor_ installs packages based on the R version, and therefore leverage the rocker/rstudio and rocker/r-ver use of R version tagging.
-_Bioconductor_ selects the desired version of R from rocker, adds the BiocManager CRAN package for installing appropriate
-versions of _Bioconductor_ packages, and creates a _Bioconductor_ docker image with an informative tag (R\_version\_Bioc\_version).
+_Bioconductor_ 'base' docker images are built on top of rocker/r-ver and rocker/rstudio.
+_Bioconductor_ installs packages based on the R version, and therefore uses rocker/rstudio and rocker/r-ver version tagging.
+_Bioconductor_ selects the desired version of R from rocker, adds the BiocManager CRAN package for installing appropriate versions of _Bioconductor_ packages, and creates a _Bioconductor_ docker image with an informative tag (R\_version\_Bioc\_version).
 The images are summarized on the _Bioconductor_ web site [5], maintained in github [6], and available to the community through DockerHub [3].
-All past and current combinations of R and _Bioconductor_ are therefore accessible via a specific docker tag.
+Past and current combinations of R and _Bioconductor_ are therefore accessible via a specific docker tag.
 
-_Bioconductor_ has several other docker images specific to various areas of research.
-The 'core' image installs the most commonly used _Bioconductor_ packages.  Other _Bioconductor_ images for proteomics, metabolomics, and flow cytometry are community maintained.
+_Bioconductor_ has several images in addition to 'base', specific to various areas of research.
+The 'core' image installs the most commonly used _Bioconductor_ packages.
+_Bioconductor_ images for proteomics, metabolomics, and flow cytometry are community maintained.
 All community maintained images build on top of the _Bioconductor_ base image and therefore indirectly the rocker images [4].
+To simplify building and maintaining _Bioconductor_ images, we use a Ruby templating engine.
 
-Bioconductor's _bioc_docker_ software works using a templating engine based in ruby, has a source directory `src` and an output directory `out`[*].
-This takes away in large part the repeated handcrafting of each Dockerfile produced by Bioconductor at each release cycle.
-Edits to the Dockerfile need to be made in the source directory, based on a config file, the output directory is populated with the appropriate Dockerfile for each specific to the various areas of research.
-A recent systematic "audit" of our docker images following best practices (as issued by the Docker website) led to a reduction in the number of layers and also size of the overall images being produced by Bioconductor.
-Some of the optimizations include removing cache, cleaning up after each layer produced by a `RUN` statement, and most importantly, not re-installing any dependencies already provided by rocker.
+A recent audit of our Dockerfiles, following best practices from the Docker website, led to a reduction in the size and number of layers in images we produce.
+The most important insights involve the 'union' file system used by Docker [REF:FIXME].
+In this file system, once a layer (e.g., `RUN` statement) writes to a file path, the file path is never altered.
+A subsequent layer that might appear to remove or overwrite the path actually masks, rather than alters, the original.
+It is therefore important to clean up (e.g., cache removal) within each layer, and to avoid re-installing existing dependencies.
 
-Bioconductor now produces a Docker image to mimic the _Bioconductor_ nightly build system on Linux.
-This image is called the `bioconductor_full` (based on rocker/rstudio).
-The goal of the Docker image is to install _all the system dependencies_ needed to install all of the Bioconductor packages.
-We achieve this goal to a large extent, and are able to greater than _99%_ of packages (1730 out of 1741).
-This image allows developers to test their packages extensively by replicating errors on the nightly build system, and users to install most of the packages they need without dealing with system dependencies.
-There is the added bonus that it comes with an RStudio interface to ease users into using R with Docker, thanks to rocker.
+A recent innovation is to produce a `bioconductor_full` image to emulate the _Bioconductor_ nightly Linux build machine.
+The image contains the _system dependencies_ needed to install and check almost all (1730 of 1741) _Bioconductor_ software packages.
+Users no longer have to manage complciated system dependencies.
+The image is configured so that `.libPaths()` has `/usr/local/lib/R/host-site-library` as the first location.
+Users mounting a location on the host file system to this location then persist installed packages across docker sessions or updates.
+Many _R_ users pursue flexible work flows tailored to particular analysis needs, rather than standardized work flows.
+The `bioconductor_full` image is well-suited to this pattern.
+`bioconductor_full` provides developers with a test environment like _Bioconductor_'s build system.
 
-We found this to be a easier alternative to distributing installed _Bioconductor_ packages within Docker images as this is not entirely consistent with the way R/Bioconductor is used.
-Users usually install and add packages to their scripts in real time.
-It also keeps the Docker image smaller in size.
-Using R is almost never like a 'workflow', which is defined by the pre-installed packages.
-We encourage users to mount a volume to their Bioconductor docker image, where all the installed (compiled) packages are stored in the external mountable volume.
-The library tree `.libPaths()` has `/usr/local/lib/R/host-site-library` as the first location searched for installed packages, which allows this functionality to work.
-This enables the docker image to be updated by using a `docker pull` command without the user having to worry about any packages saved on the image.
-
-The `bioconductor_full` image is forming the base for our cloud based solutions in upcoming future projects.
-We are currently able to host a Kubernetes orchestrated cluster on the Google Cloud Platform using the bioconductor_full image and the rocker/r-ver images.
-The distribution of computation is done through an R package called `RedisParam` (github: mtmorgan/RedisParam), where the `bioconductor_full` image is the master node and the `rocker/r-ver` images are the workers.
-
-We envision these Docker images being the primary way of teaching in the future.
-The goal is to let workshop participants pull a pre-built docker image for the specific worksop and run them on their local machines.
-As of right now, the same task is accomplished by creating and spinning up multiple AWS images (one per participant).
-The docker images can work in more fault tolerant, lasting, cost-effective manner, and be used by participants throughout their learning journey.
-The users can choose to scale their workshop skills using their cloud accounts to scale up with this provided docker image (using either the AWS Elastic Container service or Google Kubernetes Engine etc.)
+Use of images suggests several interesting possibilities for the _Bioconductor_ project.
+Images may be valuable in teaching, where participants pull pre-built images to avoid complicated configuration of their own computing environemnts.
+An appeal of this over our current approach (providing Amazon Machine Instances for the duration of the course) is the utility of the image to participant after the course is over.
+`bioconductor_full` introduces a common system configuration, so it becomes increasingly sensible for _Bioconductor_ to distribute convenient _binary_ packages.
+Images also suggest approaches to more advanced computational models.
+For instance, we are exploring use of images for Helm [REF:FIXME] orchestrated Kubernetes [REF:FIXME] clusters on the Google Cloud Platform.
+The user interacts with a manager image based on `bioconductor_full`, configured to perform map-reduce style computations via the BiocParallel package communicating with minimally-configured worker images.
+A strength of this approach is that the responsibility for complex software configuration (including customized development) is shifted from the user to the experienced _Bioconductor_ core team.
 
@@
 Rossini, Anthony J.; Sawitzki, Gunther; Smith, Colin; Smyth, Gordon; Tierney,
-Luke; Yang, Jean Y. H.; Zhang, Jianhua (2004). "Bioconductor: open software
+Luke; Yang, Jean Y. H.; Zhang, Jianhua (2004). "_Bioconductor_: open software
 development for computational biology and bioinformatics". Genome Biology. 5

### 2019-08-26 manuscript.Rmd
Add references

@@
 A recent audit of our Dockerfiles, following best practices from the Docker website, led to a reduction in the size and number of layers in images we produce.
-The most important insights involve the 'union' file system used by Docker [REF:FIXME].
+The most important insights involve the 'union' file system used by Docker [7].
 In this file system, once a layer (e.g., `RUN` statement) writes to a file path, the file path is never altered.
@@
 Images also suggest approaches to more advanced computational models.
-For instance, we are exploring use of images for Helm [REF:FIXME] orchestrated Kubernetes [REF:FIXME] clusters on the Google Cloud Platform.
+For instance, we are exploring use of images for Helm [8] orchestrated Kubernetes [9] clusters on the Google Cloud Platform.
 The user interacts with a manager image based on `bioconductor_full`, configured to perform map-reduce style computations via the BiocParallel package communicating with minimally-configured worker images.
@@
 
-[*] The original work on `bioc_docker` was done by Dan Tanenbaum
+[7] https://docs.docker.com/storage/storagedriver/overlayfs-driver/#how-container-reads-and-writes-work-with-overlay-or-overlay2
+
+[8] https://helm.sh/
+
+[9] https://kubernetes.io/
 

### 2019-08-27 manuscript.Rmd
Update manuscript.Rmd
Added info about my affiliation

--- 2019-08-26 manuscript.Rmd
@@
   - name: Colin Fay
-    #affiliation: -
-    #email: -
+    affiliation: ThinkR
+    email: contact@colinfay.me
   - name: Ben Marwick
@@
 - [FastR](https://github.com/nuest/fastr-docker)
-
 - [Bioconductor](http://bioconductor.org/help/docker/), https://github.com/Bioconductor/bioc_docker
-
-#
-# Rough draft - they asked one sentance per line
-#
-
-_Bioconductor_ is an open source, open development project for the analysis and comprehension of genomic data [1].
-The project consists of 1741 R software packages as of August 15th 2019, as well as packages containing annotation or experiment data.
-_Bioconductor_has a semiannual release cycle, each release is associated with a particular version of R.
-Docker images allow availability of current and past versions of _Bioconductor_ for convenience and reproducibility.
-
-_Bioconductor_ 'base' docker images are built on top of rocker/r-ver and rocker/rstudio.
-_Bioconductor_ installs packages based on the R version, and therefore uses rocker/rstudio and rocker/r-ver version tagging.
-_Bioconductor_ selects the desired version of R from rocker, adds the BiocManager CRAN package for installing appropriate versions of _Bioconductor_ packages, and creates a _Bioconductor_ docker image with an informative tag (R\_version\_Bioc\_version).
-The images are summarized on the _Bioconductor_ web site [5], maintained in github [6], and available to the community through DockerHub [3].
-Past and current combinations of R and _Bioconductor_ are therefore accessible via a specific docker tag.
-
-_Bioconductor_ has several images in addition to 'base', specific to various areas of research.
-The 'core' image installs the most commonly used _Bioconductor_ packages.
-_Bioconductor_ images for proteomics, metabolomics, and flow cytometry are community maintained.
-All community maintained images build on top of the _Bioconductor_ base image and therefore indirectly the rocker images [4].
-To simplify building and maintaining _Bioconductor_ images, we use a Ruby templating engine.
-
-A recent audit of our Dockerfiles, following best practices from the Docker website, led to a reduction in the size and number of layers in images we produce.
-The most important insights involve the 'union' file system used by Docker [REF:FIXME].
-In this file system, once a layer (e.g., `RUN` statement) writes to a file path, the file path is never altered.
-A subsequent layer that might appear to remove or overwrite the path actually masks, rather than alters, the original.
-It is therefore important to clean up (e.g., cache removal) within each layer, and to avoid re-installing existing dependencies.
-
-A recent innovation is to produce a `bioconductor_full` image to emulate the _Bioconductor_ nightly Linux build machine.
-The image contains the _system dependencies_ needed to install and check almost all (1730 of 1741) _Bioconductor_ software packages.
-Users no longer have to manage complciated system dependencies.
-The image is configured so that `.libPaths()` has `/usr/local/lib/R/host-site-library` as the first location.
-Users mounting a location on the host file system to this location then persist installed packages across docker sessions or updates.
-Many _R_ users pursue flexible work flows tailored to particular analysis needs, rather than standardized work flows.
-The `bioconductor_full` image is well-suited to this pattern.
-`bioconductor_full` provides developers with a test environment like _Bioconductor_'s build system.
-
-Use of images suggests several interesting possibilities for the _Bioconductor_ project.
-Images may be valuable in teaching, where participants pull pre-built images to avoid complicated configuration of their own computing environemnts.
-An appeal of this over our current approach (providing Amazon Machine Instances for the duration of the course) is the utility of the image to participant after the course is over.
-`bioconductor_full` introduces a common system configuration, so it becomes increasingly sensible for _Bioconductor_ to distribute convenient _binary_ packages.
-Images also suggest approaches to more advanced computational models.
-For instance, we are exploring use of images for Helm [REF:FIXME] orchestrated Kubernetes [REF:FIXME] clusters on the Google Cloud Platform.
-The user interacts with a manager image based on `bioconductor_full`, configured to perform map-reduce style computations via the BiocParallel package communicating with minimally-configured worker images.
-A strength of this approach is that the responsibility for complex software configuration (including customized development) is shifted from the user to the experienced _Bioconductor_ core team.
-
-#
-# Need to move these to bib file and properly format.
-#
-
-[1]  Gentleman, Robert C.; Carey, Vincent J.; Bates, Douglas M.; Bolstad, Ben;
-Dettling, Marcel; Dudoit, Sandrine; Ellis, Byron; Gautier, Laurent; Ge,
-Yongchao; Gentry, Jeff; Hornik, Kurt; Hothorn, Torsten; Huber, Wolfgang; Iacus,
-Stefano; Irizarry, Rafael; Leisch, Friedrich; Li, Cheng; Maechler, Martin;
-Rossini, Anthony J.; Sawitzki, Gunther; Smith, Colin; Smyth, Gordon; Tierney,
-Luke; Yang, Jean Y. H.; Zhang, Jianhua (2004). "_Bioconductor_: open software
-development for computational biology and bioinformatics". Genome Biology. 5
-(10): R80. doi:10.1186/gb-2004-5-10-r80. PMC 545600
-
-[2] R Core Team (2019). R: A language and environment for statistical
-  computing. R Foundation for Statistical Computing, Vienna, Austria.
-  URL https://www.R-project.org/
-
-[3] ??DockerHub citation?
-
-[4] ??Reference our docker help page/github?
-
-[5] ?? https://bioconductor.org/help/docker/
-
-[6] ?? https://github.com/Bioconductor/bioc_docker
-
-[*] The original work on `bioc_docker` was done by Dan Tanenbaum
 

### 2019-08-29 manuscript.Rmd
Remove comments
update references and links for Bioconductor

@@
 
-- [Bioconductor](http://bioconductor.org/help/docker/), https://github.com/Bioconductor/bioc_docker
+### Bioconductor
 
-#
-# Rough draft - they asked one sentance per line
-#
-
-_Bioconductor_ is an open source, open development project for the analysis and comprehension of genomic data [1].
+[_Bioconductor_](http://bioconductor.org/help/docker/) is an open source, open development project for the analysis and comprehension of genomic data [@gentleman_bioconductor_2004].
 The project consists of 1741 R software packages as of August 15th 2019, as well as packages containing annotation or experiment data.
@@
 
-_Bioconductor_ 'base' docker images are built on top of rocker/r-ver and rocker/rstudio.
-_Bioconductor_ installs packages based on the R version, and therefore uses rocker/rstudio and rocker/r-ver version tagging.
-_Bioconductor_ selects the desired version of R from rocker, adds the BiocManager CRAN package for installing appropriate versions of _Bioconductor_ packages, and creates a _Bioconductor_ docker image with an informative tag (R\_version\_Bioc\_version).
-The images are summarized on the _Bioconductor_ web site [5], maintained in github [6], and available to the community through DockerHub [3].
+_Bioconductor_ 'base' docker images are built on top of `rocker/r-ver` and `rocker/rstudio`.
+_Bioconductor_ installs packages based on the R version, and therefore uses `rocker/rstudio` and `rocker/r-ver` version tagging.
+_Bioconductor_ selects the desired version of R from _Rocker_, adds the BiocManager CRAN package for installing appropriate versions of _Bioconductor_ packages, and creates a _Bioconductor_ docker image with an informative tag (R\_version\_Bioc\_version).
+The images are summarized on the _Bioconductor_ web site (https://bioconductor.org/help/docker/), maintained on GitHub ([https://github.com/Bioconductor/bioc_docker](https://github.com/Bioconductor/bioc_docker)), and available to the community through [DockerHub](https://hub.docker.com/).
 Past and current combinations of R and _Bioconductor_ are therefore accessible via a specific docker tag.
@@
 _Bioconductor_ images for proteomics, metabolomics, and flow cytometry are community maintained.
-All community maintained images build on top of the _Bioconductor_ base image and therefore indirectly the rocker images [4].
+All community maintained images build on top of the _Bioconductor_ base image and therefore indirectly the _Rocker_ images.
 To simplify building and maintaining _Bioconductor_ images, we use a Ruby templating engine.
@@
 A recent audit of our Dockerfiles, following best practices from the Docker website, led to a reduction in the size and number of layers in images we produce.
-The most important insights involve the 'union' file system used by Docker [7].
+The most important insights involve the ['union' file system used by Docker](https://docs.docker.com/storage/storagedriver/overlayfs-driver/#how-container-reads-and-writes-work-with-overlay-or-overlay2).
 In this file system, once a layer (e.g., `RUN` statement) writes to a file path, the file path is never altered.
@@
 Images also suggest approaches to more advanced computational models.
-For instance, we are exploring use of images for Helm [8] orchestrated Kubernetes [9] clusters on the Google Cloud Platform.
+For instance, we are exploring use of images for [Helm](https://helm.sh/)-orchestrated [Kubernetes](https://kubernetes.io/) clusters on the Google Cloud Platform.
 The user interacts with a manager image based on `bioconductor_full`, configured to perform map-reduce style computations via the BiocParallel package communicating with minimally-configured worker images.
 A strength of this approach is that the responsibility for complex software configuration (including customized development) is shifted from the user to the experienced _Bioconductor_ core team.
-
-#
-# Need to move these to bib file and properly format.
-#
-
-[1]  Gentleman, Robert C.; Carey, Vincent J.; Bates, Douglas M.; Bolstad, Ben;
-Dettling, Marcel; Dudoit, Sandrine; Ellis, Byron; Gautier, Laurent; Ge,
-Yongchao; Gentry, Jeff; Hornik, Kurt; Hothorn, Torsten; Huber, Wolfgang; Iacus,
-Stefano; Irizarry, Rafael; Leisch, Friedrich; Li, Cheng; Maechler, Martin;
-Rossini, Anthony J.; Sawitzki, Gunther; Smith, Colin; Smyth, Gordon; Tierney,
-Luke; Yang, Jean Y. H.; Zhang, Jianhua (2004). "_Bioconductor_: open software
-development for computational biology and bioinformatics". Genome Biology. 5
-(10): R80. doi:10.1186/gb-2004-5-10-r80. PMC 545600
-
-[2] R Core Team (2019). R: A language and environment for statistical
-  computing. R Foundation for Statistical Computing, Vienna, Austria.
-  URL https://www.R-project.org/
-
-[3] ??DockerHub citation?
-
-[4] ??Reference our docker help page/github?
-
-[5] ?? https://bioconductor.org/help/docker/
-
-[6] ?? https://github.com/Bioconductor/bioc_docker
-
-[7] https://docs.docker.com/storage/storagedriver/overlayfs-driver/#how-container-reads-and-writes-work-with-overlay-or-overlay2
-
-[8] https://helm.sh/
-
-[9] https://kubernetes.io/
 

### 2019-09-03 manuscript.Rmd
add first  draft of distributions section

@@
   \usepackage{longtable}
+  \let\svthefootnote\thefootnote
+  \textheight 1in
+  \newcommand\blankfootnote[1]{%
+    \let\thefootnote\relax\footnotetext{#1}%
+    \let\thefootnote\svthefootnote%
+  }
 output: rticles::rjournal_article
@@
 ## Introduction
+
+\blankfootnote{This articles was collaboratively written at \href{https://github.com/nuest/rockerverse-paper/}{https://github.com/nuest/rockerverse-paper/}.}
 
@@
 
-## Images for alternative R distributions [@nuest]
+## Images for alternative R distributions
 
-- [MRO](https://github.com/nuest/mro-docker)
-- [Renjin](https://github.com/nuest/renjin-docker)
-- [pqR](https://github.com/nuest/pqr-docker)
-- [FastR](https://github.com/nuest/fastr-docker)
+As outlined above, `R` is widespread language with a large community.
+The large number of extension packages provides access to an unrivaled variety of established and upcoming features.
+Nevertheless, special use cases and experimental projects exist to test approaches or provide features different to what _"base R"_ provides.
+These projects stem both form academia and industry.
+_Base R_, sometimes `GNU-R`, is the R distribution maintained by the [R Core Team](https://www.r-project.org/contributors.html) and and provided via [CRAN](https://cran.r-project.org/).
+
+_Microsoft R Open_ ([MRO](https://github.com/nuest/mro-docker)) is an R distribution formerly known as Revolution R Open (RRO) before Revolution Analytics was acquired by Microsoft.
+MRO is compatible with main R and it's packages.sss
+"It includes additional capabilities for improved performance, reproducibility, and platform support." [@microsoft_mro]
+Most notably these capabilities are the MRAN repository a.k.a.
+CRAN Time Machine, which is enabled by default, and the (optional) integration with [Intel® Math Kernel Library](https://software.intel.com/en-us/mkl) (MKL) for multi-threading in linear algebra operations [@microsoft_multithread].
+<!-- MRAN is also used by versioned Rocker images. -->
+MRO does not provide official Docker images, but a set of community-maintained `Dockerfile`s and Docker images `nuest/mro` are available on Docker Hub at [https://github.com/nuest/mro-docker](https://github.com/nuest/mro-docker) and on GitHub at [https://github.com/nuest/mro-docker](https://github.com/nuest/mro-docker) respectively.
+The images are inspired by the Rocker images and can be used much in the same fashion, effectively a drop-in replacement allowing users to quickly evaluate if the benefits of MRO + Intel® MKL apply to their use case.
+Version-tagged images are provided for the latest bugfix release of recent R versions.
+Extended license information about MKL is printed at every startup.
+
+_Renjin_ is an interpreter for the R language running on top of the [Java Virtual Machine](https://en.wikipedia.org/wiki/Java_virtual_machine) (JVM) providing full two-way access between Java and R code [@wikipedia_renjin_2018].
+It was developed to combine the benefits of R, such scripting and extension packages, with the JVM's advantages in the areas of security, cross-platform availability, and established position in enterprise settings.
+R extension packages need to be specially compiled and are distributed via the Java package manager [Apache Maven](https://en.wikipedia.org/wiki/Apache_Maven), cf.
+[http://packages.renjin.org/packages](http://packages.renjin.org/packages) for available packages.
+Packages are loaded on demand, i.e. at the first call to `library()`.
+Not all R packages, especially one linking to binary libraries, are available, e.g. `rgdal` [^1].
+[^1]: [http://packages.renjin.org/package/org.renjin.cran/rgdal/1.4-4/build/1](http://packages.renjin.org/package/org.renjin.cran/rgdal/1.4-4/build/1)
+There are no offical Docker images for Renjin, but community-maintained images for selected releases only are available under `nuest/renjin` on Docker Hub and GitHub at [https://hub.docker.com/r/nuest/renjin](https://hub.docker.com/r/nuest/renjin) and [https://github.com/nuest/renjin-docker](https://github.com/nuest/renjin-docker) respectively.
+These images expose the command line interface of Renjin in a similar fashion as Rocker images and allow an easy evaluation of Renjin's suitability, but are not intended for production use.
+
+[_pqR_](http://www.pqr-project.org/) is a "a pretty quick version of R".
+_pqR_ fixes some opinionated issues in the R language and is the basis for experimental features, e.g. automatic diffentiation [^2].
+The source code development [on GitHub](https://github.com/radfordneal/pqR/) is a one man project and it does not provide any Docker images.
+But especially disruptive approaches may contribute to the development of the R ecosystem, so the `nuest/pqr` images on Docker Hub and GitHub at [https://hub.docker.com/r/nuest/pqr/](https://hub.docker.com/r/nuest/pqr/) and [https://github.com/nuest/pqr-docker](https://github.com/nuest/pqr-docker) respectively.
+[^2]: [https://riotworkshop.github.io/abstracts/riot-2019-pqr.txt](https://riotworkshop.github.io/abstracts/riot-2019-pqr.txt)
+
+_FastR_ is "A high-performance implementation of the R programming language, built on GraalVM" (https://github.com/oracle/fastr).
+It is developed by Oracle, connects R to the [GraalVM ecosystem](), and also claims superior performance but also targets full compatibility with base R[^3].
+<!-- https://riotworkshop.github.io/abstracts/riot-2019-fastr.txt -->
+[^3]: [https://github.com/oracle/fastr#current-status](https://github.com/oracle/fastr#current-status)
+There are no offical Docker images provided, but `nuest/fastr-docker` provides image and `Dockerfile` on Docker Hub and GitHub at []() and [https://github.com/nuest/fastr-docker](https://github.com/nuest/fastr-docker) respectively.
+
+While the images presented in this section are far from being as vetted, stable, and widely used as any of the _Rocker_ images, they demonstrate an important advantage of containerisation technology, namely the abiliy to transparently build portable stacks of open source software and make them easily accessible to users.
+All different distributions are published under GPL licenses.
+Since all of the different R distributions claim better performance as a core motivation, a comparision based on Docker images, potentially leveraging the [resource restriction mechanisms](https://docs.docker.com/config/containers/resource_constraints/) of Docker to level the playing field, seems useful future work.
 

### 2019-09-09 manuscript.Rmd
fix footnotes and citations
see #4

@@
   These use cases and the diversity of applications demonstrate the power of Rocker and containerisation for collaboration, effectivity, scalability, and transparency.
+# blank footnote? https://tex.stackexchange.com/questions/170511/footnotes-without-numbering
 preamble: >
   \usepackage{longtable}
-  \let\svthefootnote\thefootnote
-  \textheight 1in
-  \newcommand\blankfootnote[1]{%
-    \let\thefootnote\relax\footnotetext{#1}%
-    \let\thefootnote\svthefootnote%
-  }
+  \usepackage{hyperref}
 output: rticles::rjournal_article
+#  includes:
+#    in_header: preamble.tex
 ---
@@
 ## Introduction
-
-\blankfootnote{This articles was collaboratively written at \href{https://github.com/nuest/rockerverse-paper/}{https://github.com/nuest/rockerverse-paper/}.}
 
@@
 MRO is compatible with main R and it's packages.sss
-"It includes additional capabilities for improved performance, reproducibility, and platform support." [@microsoft_mro]
+"It includes additional capabilities for improved performance, reproducibility, and platform support." \citep{@microsoft_mro}
 Most notably these capabilities are the MRAN repository a.k.a.
-CRAN Time Machine, which is enabled by default, and the (optional) integration with [Intel® Math Kernel Library](https://software.intel.com/en-us/mkl) (MKL) for multi-threading in linear algebra operations [@microsoft_multithread].
+CRAN Time Machine, which is enabled by default, and the (optional) integration with [Intel® Math Kernel Library](https://software.intel.com/en-us/mkl) (MKL) for multi-threading in linear algebra operations \citep{microsoft_multithread}.
 <!-- MRAN is also used by versioned Rocker images. -->
@@
 
-_Renjin_ is an interpreter for the R language running on top of the [Java Virtual Machine](https://en.wikipedia.org/wiki/Java_virtual_machine) (JVM) providing full two-way access between Java and R code [@wikipedia_renjin_2018].
+_Renjin_ is an interpreter for the R language running on top of the [Java Virtual Machine](https://en.wikipedia.org/wiki/Java_virtual_machine) (JVM) providing full two-way access between Java and R code \citep{wikipedia_renjin_2018}.
 It was developed to combine the benefits of R, such scripting and extension packages, with the JVM's advantages in the areas of security, cross-platform availability, and established position in enterprise settings.
@@
 Packages are loaded on demand, i.e. at the first call to `library()`.
-Not all R packages, especially one linking to binary libraries, are available, e.g. `rgdal` [^1].
-[^1]: [http://packages.renjin.org/package/org.renjin.cran/rgdal/1.4-4/build/1](http://packages.renjin.org/package/org.renjin.cran/rgdal/1.4-4/build/1)
+Not all R packages, especially one linking to binary libraries, are available, e.g.
+`rgdal` \footnote{\href{http://packages.renjin.org/package/org.renjin.cran/rgdal/1.4-4/build/1}{http://packages.renjin.org/package/org.renjin.cran/rgdal/1.4-4/build/1}}.
+
 There are no offical Docker images for Renjin, but community-maintained images for selected releases only are available under `nuest/renjin` on Docker Hub and GitHub at [https://hub.docker.com/r/nuest/renjin](https://hub.docker.com/r/nuest/renjin) and [https://github.com/nuest/renjin-docker](https://github.com/nuest/renjin-docker) respectively.
@@
 [_pqR_](http://www.pqr-project.org/) is a "a pretty quick version of R".
-_pqR_ fixes some opinionated issues in the R language and is the basis for experimental features, e.g. automatic diffentiation [^2].
+_pqR_ fixes some opinionated issues in the R language and is the basis for experimental features, e.g.
+automatic diffentiation\footnote{ \href{https://riotworkshop.github.io/abstracts/riot-2019-pqr.txt}{https://riotworkshop.github.io/abstracts/riot-2019-pqr.txt}}.
 The source code development [on GitHub](https://github.com/radfordneal/pqR/) is a one man project and it does not provide any Docker images.
 But especially disruptive approaches may contribute to the development of the R ecosystem, so the `nuest/pqr` images on Docker Hub and GitHub at [https://hub.docker.com/r/nuest/pqr/](https://hub.docker.com/r/nuest/pqr/) and [https://github.com/nuest/pqr-docker](https://github.com/nuest/pqr-docker) respectively.
-[^2]: [https://riotworkshop.github.io/abstracts/riot-2019-pqr.txt](https://riotworkshop.github.io/abstracts/riot-2019-pqr.txt)
 
 _FastR_ is "A high-performance implementation of the R programming language, built on GraalVM" (https://github.com/oracle/fastr).
-It is developed by Oracle, connects R to the [GraalVM ecosystem](), and also claims superior performance but also targets full compatibility with base R[^3].
-<!-- https://riotworkshop.github.io/abstracts/riot-2019-fastr.txt -->
-[^3]: [https://github.com/oracle/fastr#current-status](https://github.com/oracle/fastr#current-status)
+It is developed by Oracle, connects R to the GraalVM ecosystem \citep{wikipedia_graal_vm}, and also claims superior performance but also targets full compatibility with base R \footnote{ \href{https://github.com/oracle/fastr}{https://github.com/oracle/fastr}}.
+
 There are no offical Docker images provided, but `nuest/fastr-docker` provides image and `Dockerfile` on Docker Hub and GitHub at []() and [https://github.com/nuest/fastr-docker](https://github.com/nuest/fastr-docker) respectively.
@@
 
-[_Bioconductor_](http://bioconductor.org/help/docker/) is an open source, open development project for the analysis and comprehension of genomic data [@gentleman_bioconductor_2004].
+[_Bioconductor_](http://bioconductor.org/help/docker/) is an open source, open development project for the analysis and comprehension of genomic data \citep{gentleman_bioconductor_2004}.
 The project consists of 1741 R software packages as of August 15th 2019, as well as packages containing annotation or experiment data.

### 2019-09-09 manuscript.Rmd
add outline for author contribution section

@@
 
+## Author contributions
+
+DN conceived of the presented idea and \href{https://github.com/nuest/rockerverse-paper/issues/3}{initialised the formation of the writing team}.
+CB ..
+RC ..
+DE ..
+ME ..
+CF ..
+BW ..
+KR ..
+NR ..
+NX ..
+LS \& NT wrote the section on Bioconductor.
+All authors contributed to the discussion and outlook section and approved the final version.
+This articles was collaboratively written at \href{https://github.com/nuest/rockerverse-paper/}{https://github.com/nuest/rockerverse-paper/}.
+The \href{https://github.com/nuest/rockerverse-paper/graphs/contributors}{contributors page} and \href{https://github.com/nuest/rockerverse-paper/commits/master}{commit history} provide a detailed view on the respective contributions.
+
 \bibliography{RJreferences}

### 2019-09-09 manuscript.Rmd

@@
 
-## Introduction
+# Introduction
 
@@
 
-## Containerization and Rocker
+# Containerization and Rocker
 
 _Do we need a generic intro here?_
+
+# Container Images
 
@@
 
-### Bioconductor
+## Bioconductor
 
@@
 
-## Other containerisation platforms [NN]
+## Images for (historic) R versions [@ColinFay, @nuest]
 
-- R images for Singularity
-- Running Rocker images with
-  - Singularity
-  - CoreOS rkt?
-- nix?
+- run code with Versions of R: https://srv.colinfay.me/r-online
+- old R versions: https://github.com/rocker-org/rocker-versioned/issues/138
+- [semantic version tags for Rocker](https://github.com/rocker-org/rocker-versioned/#version-tags)
+- debugging memory problems across multiple versions: https://github.com/wch/r-debug
+- cross-R testing: Show how you can run the same script in multiple R variants in containers (new contribution).
+Ideally the images should be created using `containerit` and controlled with `stevedore`/`docker`, i.e.
+via an R script.
 
-## Non-Debian images [NN]
+## Windows Images [@nuest]
+
+- [rocker-win](https://github.com/nuest/rocker-win)
+- is possible, only relevant in organisations with an existing Windows Server-based infrastructure, can meet policies then
+
+## Non-Debian Linux images [NN]
 
 - Alpine images
-- Images used by R-Hub (overlap with CI?)
-- [Windows](https://github.com/nuest/rocker-win) [@nuest]
+- [Images used by R-Hub](https://github.com/r-hub/rhub-linux-builders) (overlap with CI?)
 - https://github.com/jlisic/R-docker-centos
 
-## Rocker for Research Compendia [@benmarwick]
-
-...
+# R Packages
 
@@
 - [`dockermachine`](https://github.com/cboettig/dockermachine)
+
+# Use cases and applications
 
@@
 
-## R versions [@ColinFay, @nuest]
+## Research Compendia [@benmarwick]
 
-- run code with Versions of R: https://srv.colinfay.me/r-online
-- old R versions: https://github.com/rocker-org/rocker-versioned/issues/138
-- [semantic version tags for Rocker](https://github.com/rocker-org/rocker-versioned/#version-tags)
-- debugging memory problems across multiple versions: https://github.com/wch/r-debug
-- cross-R testing: Show how you can run the same script in multiple R variants in containers (new contribution).
-Ideally the images should be created using `containerit` and controlled with `stevedore`/`docker`, i.e.
-via an R script.
+...
 
-## Discussion/outlook [all, please add bullet points]
+# Other containerisation platforms [NN]
+
+- R images for Singularity
+- Running Rocker images with
+  - Singularity
+  - CoreOS rkt?
+- nix?
+
+# Discussion/outlook/conclusion [@all, please add bullet points]
 
@@
 
-## Author contributions
+# Author contributions
 

### 2019-09-09 manuscript.Rmd

@@
   - name: Nan Xiao
-    #affiliation: -
-    #email: -
+    affiliation: Seven Bridges Genomics
+    address: "529 Main St, Suite 6610 \\newline Charlestown, MA 02129, USA \\newline ORCiD: 0000-0002-0250-5673"
+    email: me@nanx.me
   - name: Lori Shepherd
@@
 - [`dockertest`](https://github.com/traitecoevo/dockertest/)
-- [`liftr`](https://liftr.me/) for OS-level reproducibility [@nanx]
+
+### liftr
+
+[_liftr_](https://nanx.me/liftr/) \citep{liftr2019} aims to solve the problem of persistent reproducible reporting in statistical computing.
+Currently, the R Markdown format and its backend compilation engine _knitr_ offer a _de facto_ standard for creating dynamic documents \citep{xie2018}.
+However, the reproducibility of such content authoring environments is often limited to individual machines --- it is not easy to replicate the system environment (libraries, R versions, R packages) where the document was compiled.
+This issue becomes even more serious when it comes to collaborative document authoring and creating large-scale document building services.
+_liftr_ solves this reproducibility problem by bringing Docker to the game.
+In essence, _liftr_ helps R Markdown users create and manage Docker containers for rendering the documents, thus make the computations utterly reproducible across machines and systems.
+
+On implementation, with no side effects, _liftr_ extended and introduced new metadata fields to R Markdown, allowing users to declare the dependencies for rendering the document.
+_liftr_ parses such fields and generates a `Dockerfile` for creating Docker containers.
+_liftr_ then helps render the document inside the created Docker container.
+This workflow is summarized in Figure \ref{figure:liftr}.
+
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[width=\textwidth]{liftr-workflow}
+  \caption{The liftr workflow for rendering containerized R Markdown documents.}
+  \label{figure:liftr}
+\end{figure}
+
+_knitr_ and R Markdown are used as the template engine to generate the `Dockerfile`.
+Features such as caching container layers for saving image build time, automatic housekeeping for fault-tolerant builds, and Docker status check are supported by _liftr_.
+Four RStudio addins are also offered by _liftr_ to allow push-button compilation of documents and provide better IDE integrations.
+
+Three basic principles are followed to design the _liftr_ package since its inception.
+
+1. Continuous reproducibility.
+Continuous integration, continuous delivery, and continuous deployment are well-accepted practices in software engineering.
+Similarly, it is believed by the authors of _liftr_ that ensuring computational reproducibility means a continuous process instead of creating static data/code archives or a one-time deal.
+Specifically, the software packages used in data analysis should be upgraded regularly in a manageable way.
+Therefore, _liftr_ supports specifying particular versions of package dependencies, while users are encouraged to always use the latest version of packages (without a version number) by default.
+
+2. Document first.
+Many data analysis workflows could be wrapped as either R packages or dynamic documents.
+In _liftr_, the endpoint of dynamic report creation is the focus of containerization, because this offers more possibilities for organizing both computations and documentation.
+Users are encouraged to start thinking from the visible research output from the first day.
+
+3. Minimal footprint.
+R Markdown and Docker are already complex software systems.
+Making them work together seamlessly can be complicated.
+Therefore, API designs such as function arguments are simplified while being kept as expressive and flexible as possible.
+
+In summary, _liftr_ tries to redefine the meaning of computational reproducibility by offering system-level reproducibility for data analysis.
+It provided a practical way for achieving it --- a new perspective on how reproducible research could be done in reality.
+Further, sharing system environments for data analysis also becomes extremely easy, since users only need to share the R Markdown document (with a few extra metadata fields), and compile them with _liftr_.
+As an example, _liftr_ demonstrated its advantage for R Markdown-based computational workflow orchestration, by effortlessly containerizing 18 complex Bioconductor workflows in the DockFlow project (https://dockflow.org) in 2017.
+
 - [`rize` for Shiny](https://github.com/cole-brokamp/rize)

### 2019-09-09 manuscript.Rmd
add notes from #5
minor fix for liftr

@@
 Further, sharing system environments for data analysis also becomes extremely easy, since users only need to share the R Markdown document (with a few extra metadata fields), and compile them with _liftr_.
-As an example, _liftr_ demonstrated its advantage for R Markdown-based computational workflow orchestration, by effortlessly containerizing 18 complex Bioconductor workflows in the DockFlow project (https://dockflow.org) in 2017.
+As an example, _liftr_ demonstrated its advantage for R Markdown-based computational workflow orchestration, by effortlessly containerizing 18 complex _Bioconductor_ workflows in the DockFlow project (https://dockflow.org) in 2017.
 
@@
 NR ..
-NX ..
+NX wrote the section on liftr.
 LS \& NT wrote the section on Bioconductor.

### 2019-09-15 manuscript.tex
small edits
adds my affiliation
remove citation for R -- R Journal house style prohibits its
replaces `R` with R -- I like \proglang{} from JSS but R Journal does neither (I think)
very minor rewording and sentence flow in two or three places
fixes two references

@@
 % !TeX root = RJwrapper.tex
-\title{Containers and R:\\
-the Rockerverse and beyond}
-\author{by Daniel Nüst}
+\title{Containers and R: the Rockerverse and beyond}
+\author{by Daniel Nüst, Carl Boettiger, Robrecht Cannoodt, Dirk Eddelbuettel, Mark Edmondson, Colin Fay, Ben Marwick, Karthik Ram, Noam Ross, Nan Xiao, Lori Shepherd, Nitesh Turaga}
 
@@
 The Rocker project provides widely-used Docker images for R across
-different application scenarios. In this article we describe the
-downstream projects building upon Rocker and look beyond towards other
-container solutions and libraries to interact with them from R.
+different application scenarios. This articles surveys describe
+downstream projects building upon Rocker. We also look beyond Rocker to
+other projects connecting containerisation with R. These use cases and
+the diversity of applications demonstrate the power of Rocker and
+containerisation for collaboration, effectivity, scalability, and
+transparency.
 }
 
-\subsection{Introduction}\label{introduction}
+\hypertarget{introduction}{%
+\section{Introduction}\label{introduction}}
 
-\citep{R}
+The R community keeps growing: the number of new packages on CRAN keeps
+on rising, meetups, conferences, online courses and unconferences
+prosper, and the uptake in education and industry increases {[}REFs{]}.
+All this cements the role of R as the lingua franca of statistics, data
+visualisation, and computational research. Coinciding with the rise of R
+was the advent of Docker {[}CITATION?{]} as a general tool for
+development, distribution, and deployment of software. Combining both
+these topics, the \emph{Rocker
+Project}(\url{https://www.rocker-project.org/}) was introduced in 2014
+and provides a number of
+\href{https://en.wikipedia.org/wiki/Docker_(software)}{Docker} images
+for various use cases as described in \citet{RJ-2017-065}. The
+considerable uptake and continued evolution of the \emph{Rocker} suite
+of containers has lead to numerous projects extending or building upon
+\emph{Rocker} images ranging from reproducible research to production
+deployments. This article presents this \emph{Rockerverse} of projects.
+It also introduces related activities connecting the R language and
+environment with other containerisation solutions. The main
+contributions is a coherent picture of the current lay of the land of
+containers in, with, and for \texttt{R}. This diversity includes
+demonstrators, early prototypes, and mature projects.
 
-\subsubsection{Containerization}\label{containerization}
+\hypertarget{containerization-and-rocker}{%
+\section{Containerization and
+Rocker}\label{containerization-and-rocker}}
 
-\subsubsection{Rocker}\label{rocker}
+\emph{Do we need a generic intro here?} DE: Yes, and happy to take a
+stab.
 
-\subsection{R images and containers}\label{r-images-and-containers}
+\hypertarget{container-images}{%
+\section{Container Images}\label{container-images}}
 
-\subsubsection{Rocker}\label{rocker-1}
+\hypertarget{images-for-alternative-r-distributions}{%
+\subsection{Images for alternative R
+distributions}\label{images-for-alternative-r-distributions}}
 
-\subsubsection{Bioconductor}\label{bioconductor}
+As outlined above, R is a widely-used language with a large community.
+The large number of extension packages provides access to an unrivaled
+variety of established and upcoming features. Nevertheless, special use
+cases and experimental projects exist to test approaches or provide
+features different to what \emph{``base R''} provides. These projects
+stem both form academia and industry. \emph{Base R}, sometimes called
+\texttt{GNU-R}, is the R distribution maintained by the
+\href{https://www.r-project.org/contributors.html}{R Core Team} and and
+provided via \href{https://cran.r-project.org/}{CRAN}.
 
-\subsubsection{MRO}\label{mro}
+\emph{Microsoft R Open}
+(\href{https://github.com/nuest/mro-docker}{MRO}) is an R distribution
+formerly known as Revolution R Open (RRO) before Revolution Analytics
+was acquired by Microsoft. MRO is compatible with main R and it's
+packages.sss ``It includes additional capabilities for improved
+performance, reproducibility, and platform support.''
+\citep{microsoft_mro} Most notably these capabilities are the MRAN
+repository a.k.a. CRAN Time Machine, which is enabled by default, and
+the (optional) integration with
+\href{https://software.intel.com/en-us/mkl}{Intel® Math Kernel Library}
+(MKL) for multi-threading in linear algebra operations
+\citep{microsoft_multithread}. MRO does not provide official Docker
+images, but a set of community-maintained \texttt{Dockerfile}s and
+Docker images \texttt{nuest/mro} are available on Docker Hub at
+\url{https://github.com/nuest/mro-docker} and on GitHub at
+\url{https://github.com/nuest/mro-docker} respectively. The images are
+inspired by the Rocker images and can be used much in the same fashion,
+effectively a drop-in replacement allowing users to quickly evaluate if
+the benefits of MRO + Intel® MKL apply to their use case. Version-tagged
+images are provided for the latest bugfix release of recent R versions.
+Extended license information about MKL is printed at every startup.
 
-\subsubsection{Renjin}\label{renjin}
+\emph{Renjin} is an interpreter for the R language running on top of the
+\href{https://en.wikipedia.org/wiki/Java_virtual_machine}{Java Virtual
+Machine} (JVM) providing full two-way access between Java and R code
+\citep{wikipedia_renjin_2018}. It was developed to combine the benefits
+of R, such scripting and extension packages, with the JVM's advantages
+in the areas of security, cross-platform availability, and established
+position in enterprise settings. R extension packages need to be
+specially compiled and are distributed via the Java package manager
+\href{https://en.wikipedia.org/wiki/Apache_Maven}{Apache Maven}, cf.
+\url{http://packages.renjin.org/packages} for available packages.
+Packages are loaded on demand, i.e.~at the first call to
+\texttt{library()}. Not all R packages, especially one linking to binary
+libraries, are available, e.g. \texttt{rgdal}
+\footnote{\href{http://packages.renjin.org/package/org.renjin.cran/rgdal/1.4-4/build/1}{http://packages.renjin.org/package/org.renjin.cran/rgdal/1.4-4/build/1}}.
 
-\subsubsection{pqR}\label{pqr}
+There are no offical Docker images for Renjin, but community-maintained
+images for selected releases only are available under
+\texttt{nuest/renjin} on Docker Hub and GitHub at
+\url{https://hub.docker.com/r/nuest/renjin} and
+\url{https://github.com/nuest/renjin-docker} respectively. These images
+expose the command line interface of Renjin in a similar fashion as
+Rocker images and allow an easy evaluation of Renjin's suitability, but
+are not intended for production use.
 
-\subsubsection{Singularity}\label{singularity}
+\href{http://www.pqr-project.org/}{\emph{pqR}} is a ``a pretty quick
+version of R''. \emph{pqR} fixes some opinionated issues in the R
+language and is the basis for experimental features, e.g.~automatic
+diffentiation\footnote{ \href{https://riotworkshop.github.io/abstracts/riot-2019-pqr.txt}{https://riotworkshop.github.io/abstracts/riot-2019-pqr.txt}}.
+The source code development
+\href{https://github.com/radfordneal/pqR/}{on GitHub} is a one man
+project and it does not provide any Docker images. But especially
+disruptive approaches may contribute to the development of the R
+ecosystem, so the \texttt{nuest/pqr} images on Docker Hub and GitHub at
+\url{https://hub.docker.com/r/nuest/pqr/} and
+\url{https://github.com/nuest/pqr-docker} respectively.
 
-\subsubsection{OCI}\label{oci}
+\emph{FastR} is ``A high-performance implementation of the R programming
+language, built on GraalVM'' (\url{https://github.com/oracle/fastr}). It
+is developed by Oracle, connects R to the GraalVM ecosystem
+\citep{wikipedia_graalvm_2019}, and also claims superior performance but
+also targets full compatibility with base R
+\footnote{ \href{https://github.com/oracle/fastr}{https://github.com/oracle/fastr}}.
 
-\subsection{Packages}\label{packages}
+There are no offical Docker images provided, but
+\texttt{nuest/fastr-docker} provides image and \texttt{Dockerfile} on
+Docker Hub and GitHub at \href{}{} and
+\url{https://github.com/nuest/fastr-docker} respectively.
 
-\subsubsection{Automation and packaging}\label{automation-and-packaging}
+While the images presented in this section are far from being as vetted,
+stable, and widely used as any of the \emph{Rocker} images, they
+demonstrate an important advantage of containerisation technology,
+namely the abiliy to transparently build portable stacks of open source
+software and make them easily accessible to users. All different
+distributions are published under GPL licenses. Since all of the
+different R distributions claim better performance as a core motivation,
+a comparision based on Docker images, potentially leveraging the
+\href{https://docs.docker.com/config/containers/resource_constraints/}{resource
+restriction mechanisms} of Docker to level the playing field, seems
+useful future work.
+
+\hypertarget{bioconductor}{%
+\subsection{Bioconductor}\label{bioconductor}}
+
+\href{http://bioconductor.org/help/docker/}{\emph{Bioconductor}} is an
+open source, open development project for the analysis and comprehension
+of genomic data \citep{gentleman_bioconductor_2004}. The project
+consists of 1741 R software packages as of August 15th 2019, as well as
+packages containing annotation or experiment data. \_Bioconductor\_has a
+semiannual release cycle, each release is associated with a particular
+version of R. Docker images allow availability of current and past
+versions of \emph{Bioconductor} for convenience and reproducibility.
+
+\emph{Bioconductor} `base' docker images are built on top of
+\texttt{rocker/r-ver} and \texttt{rocker/rstudio}. \emph{Bioconductor}
+installs packages based on the R version, and therefore uses
+\texttt{rocker/rstudio} and \texttt{rocker/r-ver} version tagging.
+\emph{Bioconductor} selects the desired version of R from \emph{Rocker},
+adds the BiocManager CRAN package for installing appropriate versions of
+\emph{Bioconductor} packages, and creates a \emph{Bioconductor} docker
+image with an informative tag (R\_version\_Bioc\_version). The images
+are summarized on the \emph{Bioconductor} web site
+(\url{https://bioconductor.org/help/docker/}), maintained on GitHub
+(\url{https://github.com/Bioconductor/bioc_docker}), and available to
+the community through \href{https://hub.docker.com/}{DockerHub}. Past
+and current combinations of R and \emph{Bioconductor} are therefore
+accessible via a specific docker tag.
+
+\emph{Bioconductor} has several images in addition to `base', specific
+to various areas of research. The `core' image installs the most
+commonly used \emph{Bioconductor} packages. \emph{Bioconductor} images
+for proteomics, metabolomics, and flow cytometry are community
+maintained. All community maintained images build on top of the
+\emph{Bioconductor} base image and therefore indirectly the
+\emph{Rocker} images. To simplify building and maintaining
+\emph{Bioconductor} images, we use a Ruby templating engine.
+
+A recent audit of our Dockerfiles, following best practices from the
+Docker website, led to a reduction in the size and number of layers in
+images we produce. The most important insights involve the
+\href{https://docs.docker.com/storage/storagedriver/overlayfs-driver/\#how-container-reads-and-writes-work-with-overlay-or-overlay2}{`union'
+file system used by Docker}. In this file system, once a layer (e.g.,
+\texttt{RUN} statement) writes to a file path, the file path is never
+altered. A subsequent layer that might appear to remove or overwrite the
+path actually masks, rather than alters, the original. It is therefore
+important to clean up (e.g., cache removal) within each layer, and to
+avoid re-installing existing dependencies.
+
+A recent innovation is to produce a \texttt{bioconductor\_full} image to
+emulate the \emph{Bioconductor} nightly Linux build machine. The image
+contains the \emph{system dependencies} needed to install and check
+almost all (1730 of 1741) \emph{Bioconductor} software packages. Users
+no longer have to manage complciated system dependencies. The image is
+configured so that \texttt{.libPaths()} has
+\texttt{/usr/local/lib/R/host-site-library} as the first location. Users
+mounting a location on the host file system to this location then
+persist installed packages across docker sessions or updates. Many
+\emph{R} users pursue flexible work flows tailored to particular
+analysis needs, rather than standardized work flows. The
+\texttt{bioconductor\_full} image is well-suited to this pattern.
+\texttt{bioconductor\_full} provides developers with a test environment
+like \emph{Bioconductor}'s build system.
+
+Use of images suggests several interesting possibilities for the
+\emph{Bioconductor} project. Images may be valuable in teaching, where
+participants pull pre-built images to avoid complicated configuration of
+their own computing environemnts. An appeal of this over our current
+approach (providing Amazon Machine Instances for the duration of the
+course) is the utility of the image to participant after the course is
+over. \texttt{bioconductor\_full} introduces a common system
+configuration, so it becomes increasingly sensible for
+\emph{Bioconductor} to distribute convenient \emph{binary} packages.
+Images also suggest approaches to more advanced computational models.
+For instance, we are exploring use of images for
+\href{https://helm.sh/}{Helm}-orchestrated
+\href{https://kubernetes.io/}{Kubernetes} clusters on the Google Cloud
+Platform. The user interacts with a manager image based on
+\texttt{bioconductor\_full}, configured to perform map-reduce style
+computations via the BiocParallel package communicating with
+minimally-configured worker images. A strength of this approach is that
+the responsibility for complex software configuration (including
+customized development) is shifted from the user to the experienced
+\emph{Bioconductor} core team.
+
+\hypertarget{images-for-historic-r-versions-nuestcolinfay-nuest}{%
+\subsection{\texorpdfstring{Images for (historic) R versions
+{[}@ColinFay,
+@nuest{]}}{Images for (historic) R versions , @nuest{[}@ColinFay, @nuest{]}}}\label{images-for-historic-r-versions-nuestcolinfay-nuest}}
 
@@
 \item
-  \texttt{containerit}
+  run code with Versions of R: \url{https://srv.colinfay.me/r-online}
 \item
-  \texttt{dockertest}
+  old R versions:
+  \url{https://github.com/rocker-org/rocker-versioned/issues/138}
 \item
-  \texttt{liftr}
+  \href{https://github.com/rocker-org/rocker-versioned/\#version-tags}{semantic
+  version tags for Rocker}
+\item
+  debugging memory problems across multiple versions:
+  \url{https://github.com/wch/r-debug}
+\item
+  cross-R testing: Show how you can run the same script in multiple R
+  variants in containers (new contribution). Ideally the images should
+  be created using \texttt{containerit} and controlled with
+  \texttt{stevedore}/\texttt{docker}, i.e.~via an R script.
 \end{itemize}
 
-\subsubsection{Control and provisioning}\label{control-and-provisioning}
+\hypertarget{windows-images-nuest}{%
+\subsection{Windows Images {[}@nuest{]}}\label{windows-images-nuest}}
 
@@
 \item
-  \texttt{harbor}
+  \href{https://github.com/nuest/rocker-win}{rocker-win}
 \item
-  \texttt{docker}
-\item
-  \texttt{RSelenium}
-\item
-  \texttt{googleComputeEngineR}
-\item
-  \texttt{analogsea}
+  is possible, only relevant in organisations with an existing Windows
+  Server-based infrastructure, can meet policies then
 \end{itemize}
 
-\subsection{Downstream Applications}\label{downstream-applications}
-
-\subsubsection{Deployment and
-processing}\label{deployment-and-processing}
+\hypertarget{non-debian-linux-images-nn}{%
+\subsection{Non-Debian Linux images
+{[}NN{]}}\label{non-debian-linux-images-nn}}
 
@@
 \item
-  \texttt{plumber}
+  Alpine images
 \item
-  \texttt{batchtools}
+  \href{https://github.com/r-hub/rhub-linux-builders}{Images used by
+  R-Hub} (overlap with CI?)
+\item
+  \url{https://github.com/jlisic/R-docker-centos}
 \end{itemize}
 
-\subsubsection{Cross-R testing}\label{cross-r-testing}
+\hypertarget{r-packages}{%
+\section{R Packages}\label{r-packages}}
 
-Show how you can run the same script in multiple R variants in
-containers (new contribution). Ideally the images should be created
-using \texttt{containerit} and controlled with
-\texttt{harbor}/\texttt{docker}, i.e.~via an R script.
-
-\subsubsection{Workflows and
-reproducibility}\label{workflows-and-reproducibility}
-
-\ldots{}
-
-\subsubsection{Scalable deployments}\label{scalable-deployments}
+\hypertarget{packages-for-automation-and-packaging-nuest}{%
+\subsection{Packages for automation and packaging
+{[}@nuest{]}}\label{packages-for-automation-and-packaging-nuest}}
 
@@
 \item
-  Shiny talks @ useR!2017
+  \href{https://github.com/o2r-project/containerit/}{\texttt{containerit}}
+\item
+  \href{https://github.com/traitecoevo/dockertest/}{\texttt{dockertest}}
+\end{itemize}
+
+\hypertarget{liftr}{%
+\subsubsection{liftr}\label{liftr}}
+
+\href{https://nanx.me/liftr/}{\emph{liftr}} \citep{liftr2019} aims to
+solve the problem of persistent reproducible reporting in statistical
+computing. Currently, the R Markdown format and its backend compilation
+engine \emph{knitr} offer a \emph{de facto} standard for creating
+dynamic documents \citep{xie2018}. However, the reproducibility of such
+content authoring environments is often limited to individual machines
+--- it is not easy to replicate the system environment (libraries, R
+versions, R packages) where the document was compiled. This issue
+becomes even more serious when it comes to collaborative document
+authoring and creating large-scale document building services.
+\emph{liftr} solves this reproducibility problem by bringing Docker to
+the game. In essence, \emph{liftr} helps R Markdown users create and
+manage Docker containers for rendering the documents, thus make the
+computations utterly reproducible across machines and systems.
+
+On implementation, with no side effects, \emph{liftr} extended and
+introduced new metadata fields to R Markdown, allowing users to declare
+the dependencies for rendering the document. \emph{liftr} parses such
+fields and generates a \texttt{Dockerfile} for creating Docker
+containers. \emph{liftr} then helps render the document inside the
+created Docker container. This workflow is summarized in Figure
+\ref{figure:liftr}.
+
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[width=\textwidth]{liftr-workflow}
+  \caption{The liftr workflow for rendering containerized R Markdown documents.}
+  \label{figure:liftr}
+\end{figure}
+
+\emph{knitr} and R Markdown are used as the template engine to generate
+the \texttt{Dockerfile}. Features such as caching container layers for
+saving image build time, automatic housekeeping for fault-tolerant
+builds, and Docker status check are supported by \emph{liftr}. Four
+RStudio addins are also offered by \emph{liftr} to allow push-button
+compilation of documents and provide better IDE integrations.
+
+Three basic principles are followed to design the \emph{liftr} package
+since its inception.
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\item
+  Continuous reproducibility. Continuous integration, continuous
+  delivery, and continuous deployment are well-accepted practices in
+  software engineering. Similarly, it is believed by the authors of
+  \emph{liftr} that ensuring computational reproducibility means a
+  continuous process instead of creating static data/code archives or a
+  one-time deal. Specifically, the software packages used in data
+  analysis should be upgraded regularly in a manageable way. Therefore,
+  \emph{liftr} supports specifying particular versions of package
+  dependencies, while users are encouraged to always use the latest
+  version of packages (without a version number) by default.
+\item
+  Document first. Many data analysis workflows could be wrapped as
+  either R packages or dynamic documents. In \emph{liftr}, the endpoint
+  of dynamic report creation is the focus of containerization, because
+  this offers more possibilities for organizing both computations and
+  documentation. Users are encouraged to start thinking from the visible
+  research output from the first day.
+\item
+  Minimal footprint. R Markdown and Docker are already complex software
+  systems. Making them work together seamlessly can be complicated.
+  Therefore, API designs such as function arguments are simplified while
+  being kept as expressive and flexible as possible.
+\end{enumerate}
+
+In summary, \emph{liftr} tries to redefine the meaning of computational
+reproducibility by offering system-level reproducibility for data
+analysis. It provided a practical way for achieving it --- a new
+perspective on how reproducible research could be done in reality.
+Further, sharing system environments for data analysis also becomes
+extremely easy, since users only need to share the R Markdown document
+(with a few extra metadata fields), and compile them with \emph{liftr}.
+As an example, \emph{liftr} demonstrated its advantage for R
+Markdown-based computational workflow orchestration, by effortlessly
+containerizing 18 complex \emph{Bioconductor} workflows in the DockFlow
+project (\url{https://dockflow.org}) in 2017.
+
+\begin{itemize}
+\tightlist
+\item
+  \href{https://github.com/cole-brokamp/rize}{\texttt{rize} for Shiny}
+\item
+  \href{https://github.com/ColinFay/dockerfiler}{\texttt{dockerfiler}}
+\end{itemize}
+
+\hypertarget{packages-for-control-and-provisioning-richfitz-rcannoodnuest-richfitz-rcannood}{%
+\subsection{\texorpdfstring{Packages for control and provisioning
+{[}@nuest, @richfitz,
+@rcannood{]}}{Packages for control and provisioning , @richfitz, @rcannood{[}@nuest, @richfitz, @rcannood{]}}}\label{packages-for-control-and-provisioning-richfitz-rcannoodnuest-richfitz-rcannood}}
+
+\begin{itemize}
+\tightlist
+\item
+  \href{https://github.com/richfitz/stevedore}{\texttt{stevedore}}
+\item
+  \href{https://cran.r-project.org/web/packages/babelwhale/index.html}{\texttt{babelwhale}}:
+  Running a Docker from R with Singularity or Docker as back-end. This
+  is really useful in HPC environments where a user might not have root
+  access but is able to install Singularity instead of Docker.
+  {[}@rcannood{]}
+\item
+  \href{https://bhaskarvk.github.io/docker/}{\texttt{docker}}
+\item
+  \texttt{RSelenium}
+\item
+  \href{https://cloudyr.github.io/googleComputeEngineR/}{\texttt{googleComputeEngineR}}
+  (function \texttt{docker\_run})
+
+  \begin{itemize}
+  \tightlist
+  \item
+    I don't think \texttt{docker\_run} is main focus for containers for
+    this package, its more how it uses Docker to launch custom
+    environments that can be templated (e.g.~rstudio, shiny etc.) and
+    for parrallisation e.g. \texttt{library(future)} using GCE VMs as
+    endpoints. {[}@MarkEdmondson1234{]}
+  \end{itemize}
+\item
+  \href{https://github.com/sckott/analogsea}{\texttt{analogsea}}
+\item
+  \href{https://github.com/wch/harbor/}{\texttt{harbor}}
+\item
+  \href{https://github.com/cboettig/dockermachine}{\texttt{dockermachine}}
+\end{itemize}
+
+\hypertarget{use-cases-and-applications}{%
+\section{Use cases and applications}\label{use-cases-and-applications}}
+
+\hypertarget{continuous-integration-and-continuous-delivery-colinfaynoamross-colinfay}{%
+\subsection{\texorpdfstring{Continuous integration and continuous
+delivery {[}@noamross,
+@ColinFay{]}}{Continuous integration and continuous delivery , @ColinFay{[}@noamross, @ColinFay{]}}}\label{continuous-integration-and-continuous-delivery-colinfaynoamross-colinfay}}
+
+\begin{itemize}
+\tightlist
+\item
+  R-Hub
+\item
+  DevOps
+
+  \begin{itemize}
+  \tightlist
+  \item
+    \url{https://www.opencpu.org/posts/opencpu-with-docker/}
+  \end{itemize}
+\item
+  \texttt{r-ci}: \url{https://github.com/ColinFay/r-ci}
+\item
+  \href{https://github.com/dynverse/dynwrap_containers/blob/master/.travis.yml}{dynwrap}
+  {[}@rcannood{]}
+
+  \begin{itemize}
+  \tightlist
+  \item
+    For this project, we use travis-ci to build rocker-derived
+    containers, test them, and only push them to docker hub (from
+    travis-ci.org) if the integration tests succeed.
+  \end{itemize}
+\item
+  Google Cloud Build {[}@MarkEdmondson1234{]}
+
+  \begin{itemize}
+  \tightlist
+  \item
+    \url{https://cloud.google.com/cloud-build/}
+  \item
+    For me this is what makes Docker containers viable, as it builds the
+    Dockerfiles on each GitHub commit. It couples with Google Container
+    Registry to build private Docker images, for downstream
+    applications.
+  \end{itemize}
+\end{itemize}
+
+\hypertarget{common-or-public-work-environments-div}{%
+\subsection{Common or public work environments
+{[}div{]}}\label{common-or-public-work-environments-div}}
+
+\begin{itemize}
+\tightlist
+\item
+  Binder
+
+  \begin{itemize}
+  \tightlist
+  \item
+    \href{https://github.com/karthik/holepunch}{\texttt{holepunch}}
+    streamlines making an R research compendium Binder-ready
+    {[}@karthik{]}
+  \item
+    \href{https://repo2docker.readthedocs.io/en/latest/config_files.html}{\texttt{repo2docker}}
+    (\texttt{install.R}, \texttt{DESCRPTION}, \texttt{runtime.txt})
+    based on the Rocker images {[}@nuest{]}
+  \end{itemize}
+\item
+  GPU {[}@noamross, @cboettig{]}
+\item
+  Gigantum stack
+\item
+  in education (ready to use teaching/learning environments)
+
+  \begin{itemize}
+  \tightlist
+  \item
+    \href{https://github.com/ColinFay/r-db}{\texttt{r-db}}
+    {[}@ColinFay{]}
+  \end{itemize}
+\item
+  \href{https://github.com/att/rcloud/tree/master/docker}{RCloud} ?
+\end{itemize}
+
+\hypertarget{deployment-processing-cloud-div}{%
+\subsection{Deployment, processing, cloud
+{[}div{]}}\label{deployment-processing-cloud-div}}
+
+\begin{itemize}
+\tightlist
+\item
+  Docker images for cloud services {[}@MarkEdmondson1234{]}
+
+  \begin{itemize}
+  \tightlist
+  \item
+    the most popular functionality for \texttt{googleComputeEngineR} is
+    its use of Docker to enable parallel processing across VMS -
+    \href{https://cloudyr.github.io/googleComputeEngineR/articles/massive-parallel.html}{some
+    demos here}
+  \end{itemize}
+\item
+  Google Cloud Run - CaaS (Containers as a Service) that lets you launch
+  a Docker container without worrying about underlying infrastructure.
+  An R implementation is shown here at
+  \href{https://github.com/MarkEdmondson1234/cloudRunR}{cloudRunR} which
+  uses it to create a scalable R plumber API.
+\item
+  \href{https://www.rplumber.io/docs/hosting.html\#docker}{\texttt{plumber}}
+\item
+  \texttt{batchtools} \citep{Lang2017batchtools} can
+  \href{https://mllg.github.io/batchtools/reference/makeClusterFunctionsDocker.html}{schedule
+  jobs with Docker Swarm}
+\item
+  scalable deployments, e.g.~start with numerous Shiny talks mentioning
+  Rocker at useR!2017
+\item
+  \href{https://github.com/dynverse/dynmethods}{dynmethods}
+  {[}@rcannood{]}: In order to evaluate ±50 computational methods which
+  all used different environments (R, Python, C++, \ldots{}), we wrapped
+  each of them in a docker container and can execute these methods from
+  R. Again, all of these containers are being built on travis-ci, and
+  will only be pushed to docker hub if the integration test succeeds.
+\item
+  \href{https://www.shinyproxy.io/}{ShinyProxy}
+  \href{https://github.com/openanalytics/shinyproxy/blob/master/src/main/java/eu/openanalytics/services/DockerService.java\#L388}{creates
+  a container} for each user
+\end{itemize}
+
+\hypertarget{research-compendia-benmarwick}{%
+\subsection{Research Compendia
+{[}@benmarwick{]}}\label{research-compendia-benmarwick}}
+
+\ldots{}
+
+\hypertarget{other-containerisation-platforms-nn}{%
+\section{Other containerisation platforms
+{[}NN{]}}\label{other-containerisation-platforms-nn}}
+
+\begin{itemize}
+\tightlist
+\item
+  R images for Singularity
+\item
+  Running Rocker images with
+
+  \begin{itemize}
+  \tightlist
+  \item
+    Singularity
+  \item
+    CoreOS rkt?
+  \end{itemize}
+\item
+  nix?
+\end{itemize}
+
+\hypertarget{discussionoutlookconclusion-please-add-bullet-pointsall-please-add-bullet-points}{%
+\section{\texorpdfstring{Discussion/outlook/conclusion {[}@all, please
+add bullet
+points{]}}{Discussion/outlook/conclusion , please add bullet points{[}@all, please add bullet points{]}}}\label{discussionoutlookconclusion-please-add-bullet-pointsall-please-add-bullet-points}}
+
+\begin{itemize}
+\tightlist
+\item
+  Missing pieces?
+\item
+  consolidation (e.g.~via packages using \texttt{dockerfiler} and
+  \texttt{stevedore})
+\item
+  Common themes
+
+  \begin{itemize}
+  \tightlist
+  \item
+    reproducibility
+  \end{itemize}
+\item
+  will knowledge about containers continue to spread?
+\item
+  what is needed for even more containers with R?
+\item
+  the ability to move processing between services easily (e.g locally,
+  one cloud providers VM, another cloud provider's
+  Container-as-a-Service)
 \item
@@
 \end{itemize}
+
+\hypertarget{author-contributions}{%
+\section{Author contributions}\label{author-contributions}}
+
+DN conceived of the presented idea and
+\href{https://github.com/nuest/rockerverse-paper/issues/3}{initialised the formation of the writing team}.
+CB .. RC .. DE .. ME .. CF .. BW .. KR .. NR .. NX wrote the section on
+liftr. LS \& NT wrote the section on Bioconductor. All authors
+contributed to the discussion and outlook section and approved the final
+version. This articles was collaboratively written at
+\href{https://github.com/nuest/rockerverse-paper/}{https://github.com/nuest/rockerverse-paper/}.
+The
+\href{https://github.com/nuest/rockerverse-paper/graphs/contributors}{contributors page}
+and
+\href{https://github.com/nuest/rockerverse-paper/commits/master}{commit history}
+provide a detailed view on the respective contributions.
 
@@
 \href{mailto:daniel.nuest@uni-muenster.de}{\nolinkurl{daniel.nuest@uni-muenster.de}}
+
+\address{%
+Carl Boettiger\\
+\\
+\\
+}
+
+\address{%
+Robrecht Cannoodt\\
+\\
+\\
+}
+
+\address{%
+Dirk Eddelbuettel\\
+University of Illinois at Urbana-Champaign\\
+\\
+}
+\href{mailto:dirkd@eddelbuettel.com}{\nolinkurl{dirkd@eddelbuettel.com}}
+
+\address{%
+Mark Edmondson\\
+IIH Nordic A/S, Google Developer Expert for GCP\\
+\\
+}
+\href{mailto:mark@markedmondson.me}{\nolinkurl{mark@markedmondson.me}}
+
+\address{%
+Colin Fay\\
+ThinkR\\
+\\
+}
+\href{mailto:contact@colinfay.me}{\nolinkurl{contact@colinfay.me}}
+
+\address{%
+Ben Marwick\\
+\\
+\\
+}
+
+\address{%
+Karthik Ram\\
+\\
+\\
+}
+
+\address{%
+Noam Ross\\
+\\
+\\
+}
+
+\address{%
+Nan Xiao\\
+Seven Bridges Genomics\\
+529 Main St, Suite 6610 \newline Charlestown, MA 02129, USA
+\newline ORCiD: 0000-0002-0250-5673\\
+}
+\href{mailto:me@nanx.me}{\nolinkurl{me@nanx.me}}
+
+\address{%
+Lori Shepherd\\
+Roswell Park Comprehensive Cancer Center\\
+Elm \& Carlton Streets, Buffalo, New York, 14263 USA\\
+}
+\href{mailto:lori.shepherd@roswellpark.org}{\nolinkurl{lori.shepherd@roswellpark.org}}
+
+\address{%
+Nitesh Turaga\\
+Roswell Park Comprehensive Cancer Center\\
+Elm \& Carlton Streets, Buffalo, New York, 14263 USA\\
+}
+\href{mailto:nitesh.turaga@roswellpark.org}{\nolinkurl{nitesh.turaga@roswellpark.org}}

### 2019-09-16 manuscript.Rmd

@@
   - name: Dirk Eddelbuettel
-    #affiliation:
-    #email: -
+    affiliation: University of Illinois at Urbana-Champaign
+    email: dirkd@eddelbuettel.com
   - name: Mark Edmondson
@@
 
-The \citep{R} community keeps growing: the number of new packages on CRAN keeps on rising, meetups, conferences, online courses and unconferences prosper, and the uptake in education and industry increases [REFs].
-All this cements `R`'s role as the lingua franca of statistics, data visualisation, and computational research.
-Part of this success can be traced back to the _Rocker Project_ ([https://www.rocker-project.org/](https://www.rocker-project.org/)).
-The community project was introduced in 2014 and provides a number of [Docker](https://en.wikipedia.org/wiki/Docker_(software)) images for various use cases \citet{RJ-2017-065}.
-The considerable uptake and continued evolution of _Rocker_'s suite has lead to numerous projects extending or building upon _Rocker_ images ranging from reproducible research to production deployments.
+The R community keeps growing: the number of new packages on CRAN keeps on rising, meetups, conferences, online courses and unconferences prosper, and the uptake in education and industry increases [REFs].
+All this cements the role of R as the lingua franca of statistics, data visualisation, and computational research.
+Coinciding with the rise of R was the advent of Docker [CITATION?] as a general tool for development, distribution, and deployment of software.
+Combining both these topics, the _Rocker Project_([https://www.rocker-project.org/](https://www.rocker-project.org/)) was introduced in 2014 and provides a number of [Docker](https://en.wikipedia.org/wiki/Docker_(software)) images for various use cases as described in \citet{RJ-2017-065}.
+The considerable uptake and continued evolution of the  _Rocker_ suite of containers has lead to numerous projects extending or building upon _Rocker_ images ranging from reproducible research to production deployments.
 This article presents this _Rockerverse_ of projects.
@@
 
-_Do we need a generic intro here?_
+_Do we need a generic intro here?_ DE: Yes, and happy to take a stab.
 
@@
 
-As outlined above, `R` is widespread language with a large community.
+As outlined above, R is a widely-used language with a large community.
 The large number of extension packages provides access to an unrivaled variety of established and upcoming features.
@@
 These projects stem both form academia and industry.
-_Base R_, sometimes `GNU-R`, is the R distribution maintained by the [R Core Team](https://www.r-project.org/contributors.html) and and provided via [CRAN](https://cran.r-project.org/).
+_Base R_, sometimes called `GNU-R`, is the R distribution maintained by the [R Core Team](https://www.r-project.org/contributors.html) and and provided via [CRAN](https://cran.r-project.org/).
 
@@
 MRO is compatible with main R and it's packages.sss
-"It includes additional capabilities for improved performance, reproducibility, and platform support." \citep{@microsoft_mro}
+"It includes additional capabilities for improved performance, reproducibility, and platform support." \citep{microsoft_mro}
 Most notably these capabilities are the MRAN repository a.k.a.
@@
 _FastR_ is "A high-performance implementation of the R programming language, built on GraalVM" (https://github.com/oracle/fastr).
-It is developed by Oracle, connects R to the GraalVM ecosystem \citep{wikipedia_graal_vm}, and also claims superior performance but also targets full compatibility with base R \footnote{ \href{https://github.com/oracle/fastr}{https://github.com/oracle/fastr}}.
+It is developed by Oracle, connects R to the GraalVM ecosystem \citep{wikipedia_graalvm_2019}, and also claims superior performance but also targets full compatibility with base R \footnote{ \href{https://github.com/oracle/fastr}{https://github.com/oracle/fastr}}.
 
@@
 
-## Packages for control and provisioning [@nuest]
+## Packages for control and provisioning [@nuest, @richfitz, @rcannood]
 

### 2019-09-18 manuscript.Rmd
restructure after telco 2

--- 2019-09-09 manuscript.Rmd
@@
 
-# Containerization and Rocker
+# Containerization and Rocker [NN]
 
@@
 # R Packages
+<!-- from packages closer to Docker to second level packages to more complex ones -->
 
-## Packages for automation and packaging [@nuest]
+## Clients
+<!-- first order packages, talking directyl to Docker -->
 
+- [`stevedore`](https://github.com/richfitz/stevedore)
+- [`docker`](https://bhaskarvk.github.io/docker/)
+- [`harbor`](https://github.com/wch/harbor/)
+- [`dockermachine`](https://github.com/cboettig/dockermachine)
+
+## Capture and create environments [@nuest]
+<!-- second order packages -->
+
+- [`dockerfiler`](https://github.com/ColinFay/dockerfiler/)
 - [`containerit`](https://github.com/o2r-project/containerit/)
 - [`dockertest`](https://github.com/traitecoevo/dockertest/)
-
-### liftr
+- [`rize` for Shiny](https://github.com/cole-brokamp/rize/)
 
@@
 
-- [`rize` for Shiny](https://github.com/cole-brokamp/rize)
-- [`dockerfiler`](https://github.com/ColinFay/dockerfiler)
+# Use cases and applications
 
-## Packages for control and provisioning [@nuest]
+## Deployment to cloud services [@nuest, @richfitz, @rcannood]
 
-- [`stevedore`](https://github.com/richfitz/stevedore)
 - [`babelwhale`](https://cran.r-project.org/web/packages/babelwhale/index.html): Running a Docker from R with Singularity or Docker as back-end.
@@
 [@rcannood]
-- [`docker`](https://bhaskarvk.github.io/docker/)
 - `RSelenium`
@@
 [@MarkEdmondson1234]
-- [`analogsea`](https://github.com/sckott/analogsea)
-- [`harbor`](https://github.com/wch/harbor/)
-- [`dockermachine`](https://github.com/cboettig/dockermachine)
-
-# Use cases and applications
+- [`analogsea`](https://github.com/sckott/analogsea) (digital ocean R client)
 
@@
 
+- next steps
+  - consolidation?
+- support more complex set-ups with ready-to-use stacks, e.g. GPU, AI, ML
 - Missing pieces?
@@
 - the ability to move processing between services easily (e.g locally, one cloud providers VM, another cloud provider's Container-as-a-Service)
+- running RStudio in a container is a common approach for reproducibility - _is there a need to run/control containers from a UI integrated in RStudio?_
+- moving away from Docker focus towards containers?
+  - is there a need for a wrapper for [podman's Varlink-API](https://github.com/containers/libpod/blob/master/API.md) (similar to stevedore) and buildah to manage containers and images respectively with alternative software stacks?
+  - will there be control packages for Singularity support?
 - ...

### 2019-10-07 manuscript.Rmd
Piece about r-online

--- 2019-09-16 manuscript.Rmd
@@
 
-- run code with Versions of R: https://srv.colinfay.me/r-online
+As with any other software, each new version of R comes with its share of changes. 
+Some are breaking changes, some are not, but the fact is that running a piece of code in a given version might give different s from another version ; which, in some case, can be an issue when it comes to reproducibility.
+For example, think about R random seed: starting with `R 3.6.0`, running `set.seed(2811); sample(1:1000, 2)` will not give the same result as if it was run inside an older version of R.
+That might seem trivial, but all the code using the random number generator will not be exactly reproducible after this breaking change. 
+Detecting this kind of behavior is the reason why the [r-online](https://github.com/ColinFay/ronline) app has been developped: available as a standalone NodeJS app or [online](https://srv.colinfay.me/r-online), this app was designed for historic exploration of R: with it, the user can compare a piece of code run in two separate versions of R.
+Internally, this app takes the piece of code written by the user, opens one or two docker instances with the given version of R, and returns the result to the user.
+
 - old R versions: https://github.com/rocker-org/rocker-versioned/issues/138

### 2019-10-07 manuscript.Rmd
part about {dockerfiler}

--- 2019-09-16 manuscript.Rmd
@@
 - [`rize` for Shiny](https://github.com/cole-brokamp/rize)
-- [`dockerfiler`](https://github.com/ColinFay/dockerfiler)
+
+One of the key tool of Docker are dockerfiles, which can be thought of as "recipes" used to build a specific image. 
+But building these files can feel like a workflow break as it demands to open a separate file, and / or to write a small wrapper around R's `write()` function to add elements to the file. 
+At the same time, doing it by hand prevents from using programming language for what they are good for: iteration, and automation. 
+Iteration and automation for Dockerfiles is the very reason behind [`dockerfiler`](https://github.com/ColinFay/dockerfiler), an R package designed for building Dockerfiles straigth from R.
+
+For example, the [`golem`](https://github.com/ThinkR-open/golem) package (an R package for building production-grade Shiny application) makes an heavy use of `dockerfiler` when it comes to creating the `Dockerfile` for deploying the Shiny application to production.
+The reason behind this is that `dockerfiler`, on top of being scriptable from R, can leverage all the tools available in R to parse a `DESCRIPTION` file, to get system requirements, to list dependencies, versions, etc.
 

### 2019-10-07 manuscript.Rmd
part about r-ci

@@
 - [`rize` for Shiny](https://github.com/cole-brokamp/rize)
-
-One of the key tool of Docker are dockerfiles, which can be thought of as "recipes" used to build a specific image. 
-But building these files can feel like a workflow break as it demands to open a separate file, and / or to write a small wrapper around R's `write()` function to add elements to the file. 
-At the same time, doing it by hand prevents from using programming language for what they are good for: iteration, and automation. 
-Iteration and automation for Dockerfiles is the very reason behind [`dockerfiler`](https://github.com/ColinFay/dockerfiler), an R package designed for building Dockerfiles straigth from R.
-
-For example, the [`golem`](https://github.com/ThinkR-open/golem) package (an R package for building production-grade Shiny application) makes an heavy use of `dockerfiler` when it comes to creating the `Dockerfile` for deploying the Shiny application to production.
-The reason behind this is that `dockerfiler`, on top of being scriptable from R, can leverage all the tools available in R to parse a `DESCRIPTION` file, to get system requirements, to list dependencies, versions, etc.
+- [`dockerfiler`](https://github.com/ColinFay/dockerfiler)
 
@@
   - https://www.opencpu.org/posts/opencpu-with-docker/
-- `r-ci`: https://github.com/ColinFay/r-ci
+
+When doing continuous integration and continuous delivery, it's crucial to test in an environment that matches the production environment.
+Tools like `Gitlab-CI` are built on top of Docker images: the user specifies a base Docker image, and the whole tests are run inside this environment. 
+But, as we just said, this environment has to be fixed, but it also have to contain the necessary toolkit. 
+In order to achieve that, [`r-ci`](https://github.com/ColinFay/r-ci) combines `rocker` versioning and a series of tools specifically designed for testing. 
+That way, package builders can use this image as a base image for there testing environment, without having to install the necessary packages every time they need to run a new test.
+
 - [dynwrap](https://github.com/dynverse/dynwrap_containers/blob/master/.travis.yml) [@rcannood]

### 2019-10-07 manuscript.Rmd
part about r-db

--- 2019-10-07 manuscript.Rmd
@@
 - Gigantum stack
-- in education (ready to use teaching/learning environments)
-  - [`r-db`](https://github.com/ColinFay/r-db) [@ColinFay]
+
+R is a fantastic tool when it comes to interfacing with databases: almost every open source and proprietary database system has an R package that allows users to connect and interact with the it 
+This flexibility is even broader now that we have tools like `DBI`, that allows to create a common API for interfacing these databases, or like `dbplyr`, which are designed to run `dplyr` code straight against the database.
+But learning and teaching these tools comes with a cost: the cost of deploying or having access to an environment with the softwares and drivers installed. 
+For people teaching R, it can become a barrier if they need to install local versions of the drivers, or to connect to remote instances which might or might not be made available by the IT. 
+Giving access to a sandbox for the most common database environments is the idea behind [`r-db`](https://github.com/ColinFay/r-db) [@ColinFay], a docker image that contains everything needed to connect, from R, to a database.
+Notably, with `r-db`, the users don't have to install complex drivers or to configure their machine in a specific way. 
+On top of that, `r-db` comes with a comprehensive [online guide](http://colinfay.me/r-db/), explaining how to spin a docker instance of a specific database, and how to use `r-db` to connect to it. 
+Each package contained inside this image also has a series examples, so that the user can get started with the database right away.
+
 - [RCloud](https://github.com/att/rcloud/tree/master/docker) ?

### 2019-10-07 manuscript.Rmd
piece written by CF

--- 2019-10-07 manuscript.Rmd
@@
 ME ..
-CF ..
+CF wrote the section about `r-online`, `dockerfiler`, `r-ci` and `r-db`.
 BW ..

### 2019-10-08 manuscript.tex
draft addition of docker and rocker intro

@@
 
-\emph{Do we need a generic intro here?} DE: Yes, and happy to take a
-stab.
+Docker, an application and service provide by the eponymous company, has
+in just a few short years risen to prominence for development, testing,
+deployment and distribution of computer software. While there are
+related approaches such as LXC or Singularity, Docker has become
+synomymous with ``containerization''---the method of taking software
+artefacts and bundling them in such a way that use becomes standardized
+and portable across operationg systems. In doing so, Docker had
+recognised and validated the importance of one very important thread
+that had been emerging, namely virtualization. By allowing multiple
+concurrent applications or services to run concurrently on one host
+machine without any fear of interference between them, an important
+scalability feature had been realized. But Docker improved on
+virtualization by accessing the host system---generally Linux---through
+a much thinner and smaller shim than a full operating system emulation
+or virtualization. This makes for more efficient use of system resources
+and allowed another order of magnitude in terms of scalability of
+deployment. {[}Reference?{]} While Docker makes use of Linux kernel
+features, its importance was large enough so that some required aspects
+of running Docker have been added to other operating systems to support
+Docker there more efficiently too. {[}Win10; reference?{]}
+
+The key accomplishment of Docker is to make a ``bundled'' aggregation of
+software available to any system running Docker, without requiring much
+else from the host. This is a rather attractive, and novel, proposition
+which has lead to widespread adoption and use of Docker in a variety of
+domains. It provided also a natural match for ``cloud deployment'' which
+runs (or at least appears to run) ``seamlessly'' without much explicit
+reference to the underlying machine, architecture or operating system.
+Using containers on cloud hosted servers is a natural fit also because
+these containers can be deployed with very littler in terms of
+dependencies on the host system.
+
+For statistical computing and analysis centered around R, the Rocker
+Project has provided a variety of Docker containers since its start in
+2014 \citep{RJ-2017-065}. The Rocker Project provides several lines of
+containeres spanning to from building blocks with R-release or R-devel
+to containers with RStudio Server and Shiny Server. Also of note is a
+series of ``versioned'' containers which match the R release they
+contain with the \emph{then-current} set of packages via the MRAN
+Snapshot views of CRAN. {[}Maybe at a later point add forward reference
+to other topics building on Rocker described here.{]}
 
@@
 University of Illinois at Urbana-Champaign\\
-\\
+Department of Statistics\\ Illini Hall, 725 S Wright St\\ Champaign, IL 61820\\ ORCiD \href{https://orcid.org/0000-0001-6419-907X}{0000-0001-6419-907X}\\
 }

### 2019-10-18 manuscript.Rmd
Added draft of holepunch section

@@
   - name: Karthik Ram
-    #affiliation: -
-    #email: -
+    affiliation: Berkeley Institute for Data Science, University of California, Berkeley, CA 94720 USA
+    email: karthik.ram@berkeley.edu
   - name: Noam Ross
@@
 - Binder
-  - [`holepunch`](https://github.com/karthik/holepunch) streamlines making an R research compendium Binder-ready [@karthik]
+
+The [Binder project](https://mybinder.readthedocs.io/en/latest/), maintained by the team behind Jupyter, makes it possible for users to create and share their computing environments with others.
+A deployed version of this service called binder hub allows anyone with access to a web browser and an internet connection to launch a temporary instance of these custom environments and execute any notebooks contained within.
+From a reproducibility standpoint, Binder makes it exceedingly easy to compile a paper, visualize data, and run small examples from papers or tutorials without the need for any local installation.
+
+To set up Binder for a project, a user typically starts at an instance of a Binderhub (examples include mybinder.org and binder.pangeo.io) and passes the location of a hosted Git repository.
+Using internal tools such as [repo2docker](https://repo2docker.readthedocs.io/en/latest/config_files.html), the binderhub builds a Dockerfile by parsing the contents of all code contained within.
+While this approach works well for most run of the mill Python projects, it is not so seamless for R projects.
+For any R projects that use the Tidyverse suite [citation in review], the time and resources required to build all dependencies from source can often time out before completion, making it frustrating for the average R user.
+
+[`holepunch`](https://github.com/karthik/holepunch) is a R package that was designed to remove some of these limitations and make Binder more accessible to novice R users.
+`holepunch` achieves this by leveraging Rocker images that contains the Tidyverse along special Jupyter dependencies, and only installs additional packages from CRAN and Bioconductor that are not already part of these images.
+Starting with the Binder/Tidyverse base images eliminates a large part of the build time and in most cases results in a binder instance launching within a minute.
+`holepunch` as a side effect also creates a DESCRIPTION file which then turns any project into a research compendium [Marwick et al].
+The Dockerfile included with the project can also be used to launch a RStudio server locally independent of binder which is especially useful when more computational resources are required.
+
   - [`repo2docker`](https://repo2docker.readthedocs.io/en/latest/config_files.html) (`install.R`, `DESCRPTION`, `runtime.txt`) based on the Rocker images [@nuest]
@@
 ME ..
-CF wrote the section about `r-online`, `dockerfiler`, `r-ci` and `r-db`.
+CF ..
 BW ..

### 2019-10-18 manuscript.tex
Added draft of holepunch section

--- 2019-09-15 manuscript.tex
@@
 specially compiled and are distributed via the Java package manager
-\href{https://en.wikipedia.org/wiki/Apache_Maven}{Apache Maven}, cf.
-\url{http://packages.renjin.org/packages} for available packages.
+\href{https://en.wikipedia.org/wiki/Apache_Maven}{Apache Maven},
+cf.~\url{http://packages.renjin.org/packages} for available packages.
 Packages are loaded on demand, i.e.~at the first call to
 \texttt{library()}. Not all R packages, especially one linking to binary
-libraries, are available, e.g. \texttt{rgdal}
+libraries, are available, e.g.~\texttt{rgdal}
 \footnote{\href{http://packages.renjin.org/package/org.renjin.cran/rgdal/1.4-4/build/1}{http://packages.renjin.org/package/org.renjin.cran/rgdal/1.4-4/build/1}}.
@@
     environments that can be templated (e.g.~rstudio, shiny etc.) and
-    for parrallisation e.g. \texttt{library(future)} using GCE VMs as
+    for parrallisation e.g.~\texttt{library(future)} using GCE VMs as
     endpoints. {[}@MarkEdmondson1234{]}
@@
   {[}@rcannood{]}: In order to evaluate ±50 computational methods which
-  all used different environments (R, Python, C++, \ldots{}), we wrapped
+  all used different environments (R, Python, C++, \ldots), we wrapped
   each of them in a docker container and can execute these methods from
@@
 Karthik Ram\\
-\\
+Berkeley Institute for Data Science, University of California, Berkeley,
+CA 94720 USA\\
 \\
 }
+\href{mailto:karthik.ram@berkeley.edu}{\nolinkurl{karthik.ram@berkeley.edu}}
 

### 2019-10-21 manuscript.Rmd
Added Marwick paper citation

@@
 While this approach works well for most run of the mill Python projects, it is not so seamless for R projects.
-For any R projects that use the Tidyverse suite [citation in review], the time and resources required to build all dependencies from source can often time out before completion, making it frustrating for the average R user.
+For any R projects that use the Tidyverse suite [citation in review.
+Will add by the time this gets submitted], the time and resources required to build all dependencies from source can often time out before completion, making it frustrating for the average R user.
 
@@
 Starting with the Binder/Tidyverse base images eliminates a large part of the build time and in most cases results in a binder instance launching within a minute.
-`holepunch` as a side effect also creates a DESCRIPTION file which then turns any project into a research compendium [Marwick et al].
+`holepunch` as a side effect also creates a DESCRIPTION file which then turns any project into a research compendium [doi:10.1080/00031305.2017.1375986].
 The Dockerfile included with the project can also be used to launch a RStudio server locally independent of binder which is especially useful when more computational resources are required.

### 2019-10-21 manuscript.tex
Added Marwick paper citation

@@
   Binder
+\end{itemize}
+
+The \href{https://mybinder.readthedocs.io/en/latest/}{Binder project},
+maintained by the team behind Jupyter, makes it possible for users to
+create and share their computing environments with others. A deployed
+version of this service called binder hub allows anyone with access to a
+web browser and an internet connection to launch a temporary instance of
+these custom environments and execute any notebooks contained within.
+From a reproducibility standpoint, Binder makes it exceedingly easy to
+compile a paper, visualize data, and run small examples from papers or
+tutorials without the need for any local installation.
+
+To set up Binder for a project, a user typically starts at an instance
+of a Binderhub (examples include mybinder.org and binder.pangeo.io) and
+passes the location of a hosted Git repository. Using internal tools
+such as
+\href{https://repo2docker.readthedocs.io/en/latest/config_files.html}{repo2docker},
+the binderhub builds a Dockerfile by parsing the contents of all code
+contained within. While this approach works well for most run of the
+mill Python projects, it is not so seamless for R projects. For any R
+projects that use the Tidyverse suite {[}citation in review{]}, the time
+and resources required to build all dependencies from source can often
+time out before completion, making it frustrating for the average R
+user.
+
+\href{https://github.com/karthik/holepunch}{\texttt{holepunch}} is a R
+package that was designed to remove some of these limitations and make
+Binder more accessible to novice R users. \texttt{holepunch} achieves
+this by leveraging Rocker images that contains the Tidyverse along
+special Jupyter dependencies, and only installs additional packages from
+CRAN and Bioconductor that are not already part of these images.
+Starting with the Binder/Tidyverse base images eliminates a large part
+of the build time and in most cases results in a binder instance
+launching within a minute. \texttt{holepunch} as a side effect also
+creates a DESCRIPTION file which then turns any project into a research
+compendium {[}\url{doi:10.1080/00031305.2017.1375986}{]}. The Dockerfile
+included with the project can also be used to launch a RStudio server
+locally independent of binder which is especially useful when more
+computational resources are required.
 
@@
 \item
-    \href{https://github.com/karthik/holepunch}{\texttt{holepunch}}
-    streamlines making an R research compendium Binder-ready
-    {[}@karthik{]}
-  \item
   \href{https://repo2docker.readthedocs.io/en/latest/config_files.html}{\texttt{repo2docker}}
-    (\texttt{install.R}, \texttt{DESCRPTION}, \texttt{runtime.txt})
-    based on the Rocker images {[}@nuest{]}
-  \end{itemize}
+  (\texttt{install.R}, \texttt{DESCRPTION}, \texttt{runtime.txt}) based
+  on the Rocker images {[}@nuest{]}
 \item

### 2019-10-25 manuscript.Rmd
Add section on shared work environments and GPUs

--- 2019-10-07 manuscript.Rmd
@@
 - GPU [@noamross, @cboettig]
+
+One use for containers is to run on shared local hardware where teams manage
+their own high-performance servers  This can follow one of several
+design patterns: users may deploy containers to hardware as a work environment
+for a specific project, conatiners may provide per-user persistent environments, or a single container can act as
+a common multi-user environment for a server.  The former models provide modularity,
+while the latter approach is most similar to a simple shared server. In all cases,
+though, the containerized approach provides several advantages: First, users may
+use the same image and thus work environment on desktop and laptop computers, as well.
+Second, software updates can be achieved by updating and redeploying the container,
+rather than tracking local installs on each server.  Third, the containerized
+environment can be quickly deployed to other hardware, cloud or local, if 
+more resources are neccessary or in case of server destruction or failure. In any of
+these cases, users need a method to interact with the containers, be it and
+IDE, or command-like access and tools such as SSH, which is usually not part
+of standard container recipes and must be added.  The Rocker project provides
+containers pre-installed with the RStudio IDE.
+
+In cases where users store nontrivial amounts of data for their projects, data
+needs to persist beyond the life of the container. This may be via in shared disks, attached
+network volumes, or in separate storage where it is uploaded between sessions. In
+the case of shared disks or network-attached volumes, care must be taken to
+persist user permissions, and of course backups are still neccessary.
+When working with multiple servers, an automation
+framework such as [Ansible](https://www.ansible.com) may be useful for managing
+users, permisions, and disks along with containers.
+
+In some cases, containers are not completely portable between hardware environments.
+This is the case for software using graphical processing units (GPUs)
+which are increasingly popular for compute-intensive machine learning tasks.
+Containers running GPU software  require drivers and libraries specific
+to GPU models and versions, and containers require a specialized runtime to connect
+to the underlying GPU hardware. For NVIDIA GPUs, the [NVIDIA Container Toolkit](https://github.com/NVIDIA/nvidia-docker)
+includes a specialized runtime plugn for docker and set of base images with
+appropriate drivers and libraries, and the Rocker project has (beta) images
+based on these that include GPU-enabled versions of machine-learning R packages.
+
 - Gigantum stack

### 2019-11-04 manuscript.Rmd

--- 2019-10-18 manuscript.Rmd
@@
     affiliation: University of Illinois at Urbana-Champaign
+    address:
+      - Department of Statistics
+      - Illini Hall, 725 S Wright St
+      - Champaign, IL 61820
+      - ORCiD \href{https://orcid.org/0000-0001-6419-907X}{0000-0001-6419-907X}
     email: dirkd@eddelbuettel.com
@@
   - name: Karthik Ram
-    affiliation: Berkeley Institute for Data Science, University of California, Berkeley, CA 94720 USA
-    email: karthik.ram@berkeley.edu
+    #affiliation: -
+    #email: -
   - name: Noam Ross
@@
 
-_Do we need a generic intro here?_ DE: Yes, and happy to take a stab.
+<!--_Do we need a generic intro here?_ DE: Yes, and happy to take a stab.-->
+
+Docker, an application and service provide by the eponymous company, has in just a few short years risen to prominence for development, testing, deployment and distribution of computer software.
+While there are related approaches such as LXC or Singularity, Docker has become synomymous with "containerization"---the method of taking software artefacts and bundling them in such a way that use becomes standardized and portable across operationg systems.
+In doing so, Docker had recognised and validated the importance of one very important thread that had been emerging, namely virtualization.
+By allowing multiple concurrent applications or services to run concurrently on one host machine without any fear of interference between them, an important scalability feature had been realized.
+But Docker improved on virtualization by accessing the host system---generally Linux---through a much thinner and smaller shim than a full operating system emulation or virtualization.
+This makes for more efficient use of system resources and allowed another order of magnitude in terms of scalability of deployment.
+[Reference?]   While Docker makes use of Linux kernel features, its importance was large enough so that some required aspects of running Docker have been added to other operating systems to support Docker there more efficiently too.
+[Win10; reference?]
+
+The key accomplishment of Docker is to make a "bundled" aggregation of software available to any system running Docker, without requiring much else from the host.
+This is a rather attractive, and novel, proposition which has lead to widespread adoption and use of Docker in a variety of domains.
+It provided also a natural match for "cloud deployment" which runs (or at least appears to run) "seamlessly" without much explicit reference to the underlying machine, architecture or operating system.
+Using containers on cloud hosted servers is a natural fit also because these containers can be deployed with very littler in terms of dependencies on the host system.
+
+For statistical computing and analysis centered around R, the Rocker Project has provided a variety of Docker containers since its start in 2014 \citep{RJ-2017-065}.
+The Rocker Project provides several lines of containeres spanning to from building blocks with R-release or R-devel to containers with RStudio Server and Shiny Server.
+Also of note is a series of "versioned" containers which match the R release they contain with the _then-current_ set of packages via the MRAN Snapshot views of CRAN.
+[Maybe at a later point add forward reference to other topics building on Rocker described here.]
 
@@
 - Binder
-
-The [Binder project](https://mybinder.readthedocs.io/en/latest/), maintained by the team behind Jupyter, makes it possible for users to create and share their computing environments with others.
-A deployed version of this service called binder hub allows anyone with access to a web browser and an internet connection to launch a temporary instance of these custom environments and execute any notebooks contained within.
-From a reproducibility standpoint, Binder makes it exceedingly easy to compile a paper, visualize data, and run small examples from papers or tutorials without the need for any local installation.
-
-To set up Binder for a project, a user typically starts at an instance of a Binderhub (examples include mybinder.org and binder.pangeo.io) and passes the location of a hosted Git repository.
-Using internal tools such as [repo2docker](https://repo2docker.readthedocs.io/en/latest/config_files.html), the binderhub builds a Dockerfile by parsing the contents of all code contained within.
-While this approach works well for most run of the mill Python projects, it is not so seamless for R projects.
-For any R projects that use the Tidyverse suite [citation in review], the time and resources required to build all dependencies from source can often time out before completion, making it frustrating for the average R user.
-
-[`holepunch`](https://github.com/karthik/holepunch) is a R package that was designed to remove some of these limitations and make Binder more accessible to novice R users.
-`holepunch` achieves this by leveraging Rocker images that contains the Tidyverse along special Jupyter dependencies, and only installs additional packages from CRAN and Bioconductor that are not already part of these images.
-Starting with the Binder/Tidyverse base images eliminates a large part of the build time and in most cases results in a binder instance launching within a minute.
-`holepunch` as a side effect also creates a DESCRIPTION file which then turns any project into a research compendium [Marwick et al].
-The Dockerfile included with the project can also be used to launch a RStudio server locally independent of binder which is especially useful when more computational resources are required.
-
+  - [`holepunch`](https://github.com/karthik/holepunch) streamlines making an R research compendium Binder-ready [@karthik]
   - [`repo2docker`](https://repo2docker.readthedocs.io/en/latest/config_files.html) (`install.R`, `DESCRPTION`, `runtime.txt`) based on the Rocker images [@nuest]

### 2019-11-05 manuscript.Rmd
Add ORCID rendering, add ids for current authors

@@
 ---
-title: "Containers and R: the Rockerverse and beyond"
+title: "Containers and R---the Rockerverse and beyond"
 author:
@@
     address:
-    - Heisenbergstr. 2, 48149 Münster, Germany
+    - Heisenbergstr. 2
+    - 48149 Münster
+    - Germany
+    - \orcid{0000-0002-0024-5046}
     email: daniel.nuest@uni-muenster.de
@@
       - Champaign, IL 61820
-      - ORCiD \href{https://orcid.org/0000-0001-6419-907X}{0000-0001-6419-907X}
+      - \orcid{0000-0001-6419-907X}
     email: dirkd@eddelbuettel.com
@@
     affiliation: Seven Bridges Genomics
-    address: "529 Main St, Suite 6610 \\newline Charlestown, MA 02129, USA \\newline ORCiD: 0000-0002-0250-5673"
+    address:
+      - 529 Main St, Suite 6610
+      - Charlestown, MA 02129, USA
+      - \orcid{0000-0002-0250-5673}
     email: me@nanx.me
@@
     address:
-    - Elm & Carlton Streets, Buffalo, New York, 14263 USA
+      - Elm & Carlton Streets
+      - Buffalo, New York, 14263 USA
+      - \orcid{0000-0002-5910-4010}
     email: lori.shepherd@roswellpark.org
@@
     address:
-    - Elm & Carlton Streets, Buffalo, New York, 14263 USA
+    - Elm & Carlton Streets
+    - Buffalo, New York, 14263 USA
+    - \orcid{0000-0002-0224-9817}
     email: nitesh.turaga@roswellpark.org
@@
 # blank footnote? https://tex.stackexchange.com/questions/170511/footnotes-without-numbering
-preamble: >
-  \usepackage{longtable}
-  \usepackage{hyperref}
-output: rticles::rjournal_article
-#  includes:
-#    in_header: preamble.tex
+output:
+  rticles::rjournal_article:
+    keep_tex: true
 ---

### 2019-11-05 manuscript.Rmd
chapter 1: add some references and put each sentence on a line

@@
 
-<!--_Do we need a generic intro here?_ DE: Yes, and happy to take a stab.-->
-
 Docker, an application and service provide by the eponymous company, has in just a few short years risen to prominence for development, testing, deployment and distribution of computer software.
-While there are related approaches such as LXC or Singularity, Docker has become synomymous with "containerization"---the method of taking software artefacts and bundling them in such a way that use becomes standardized and portable across operationg systems.
+While there are related approaches such as LXC\footnote{\href{https://en.wikipedia.org/wiki/LXC}{https://en.wikipedia.org/wiki/LXC}} or Singularity \citep{kurtzer_singularity_2017}, Docker has become synomymous with "containerization"---the method of taking software artefacts and bundling them in such a way that use becomes standardized and portable across operationg systems.
 In doing so, Docker had recognised and validated the importance of one very important thread that had been emerging, namely virtualization.
@@
 But Docker improved on virtualization by accessing the host system---generally Linux---through a much thinner and smaller shim than a full operating system emulation or virtualization.
-This makes for more efficient use of system resources and allowed another order of magnitude in terms of scalability of deployment.
-[Reference?]   While Docker makes use of Linux kernel features, its importance was large enough so that some required aspects of running Docker have been added to other operating systems to support Docker there more efficiently too.
-[Win10; reference?]
+This makes for more efficient use of system resources and allowed another order of magnitude in terms of scalability of deployment [REFERENCES!].   
+While Docker makes use of Linux kernel features, its importance was large enough so that some required aspects of running Docker have been added to other operating systems to support Docker there more efficiently too \citep{microsoft_linux_2019}.
+The success even lead to standardisation and industry collaboration \citep{oci_open_2019}.
 
 The key accomplishment of Docker is to make a "bundled" aggregation of software available to any system running Docker, without requiring much else from the host.
-This is a rather attractive, and novel, proposition which has lead to widespread adoption and use of Docker in a variety of domains.
-It provided also a natural match for "cloud deployment" which runs (or at least appears to run) "seamlessly" without much explicit reference to the underlying machine, architecture or operating system.
-Using containers on cloud hosted servers is a natural fit also because these containers can be deployed with very littler in terms of dependencies on the host system.
+This is a rather attractive, and novel, proposition which has lead to widespread adoption and use of Docker in a variety of domains, e.g., cloud computing infrastructure \citep{Bernstein2014}, data science \citep{boettiger_introduction_2015}, and edge computing \citep{alam_orchestration_2018} [more examples!].
+It provided to be a natural match for "cloud deployment" which runs (or at least appears to run) "seamlessly" without much explicit reference to the underlying machine, architecture or operating system: containers can be deployed with very little in terms of dependencies on the host system---only the container runtime is required.
 
 For statistical computing and analysis centered around R, the Rocker Project has provided a variety of Docker containers since its start in 2014 \citep{RJ-2017-065}.
-The Rocker Project provides several lines of containeres spanning to from building blocks with R-release or R-devel to containers with RStudio Server and Shiny Server.
-Also of note is a series of "versioned" containers which match the R release they contain with the _then-current_ set of packages via the MRAN Snapshot views of CRAN.
+The Rocker Project provides several lines of containeres spanning to from building blocks with R-release or R-devel to containers with RStudio~Server and Shiny~Server.
+Also of note is a series of "versioned" containers which match the R release they contain with the _then-current_ set of packages via the MRAN Snapshot views of CRAN \citep{microsoft_cran_2019}.
 [Maybe at a later point add forward reference to other topics building on Rocker described here.]
@@
 "It includes additional capabilities for improved performance, reproducibility, and platform support." \citep{microsoft_mro}
-Most notably these capabilities are the MRAN repository a.k.a.
-CRAN Time Machine, which is enabled by default, and the (optional) integration with [Intel® Math Kernel Library](https://software.intel.com/en-us/mkl) (MKL) for multi-threading in linear algebra operations \citep{microsoft_multithread}.
-<!-- MRAN is also used by versioned Rocker images. -->
+Most notably these capabilities are the MRAN repository, which is enabled by default, and the (optional) integration with [Intel® Math Kernel Library](https://software.intel.com/en-us/mkl) (MKL) for multi-threading in linear algebra operations \citep{microsoft_multithread}.
 MRO does not provide official Docker images, but a set of community-maintained `Dockerfile`s and Docker images `nuest/mro` are available on Docker Hub at [https://github.com/nuest/mro-docker](https://github.com/nuest/mro-docker) and on GitHub at [https://github.com/nuest/mro-docker](https://github.com/nuest/mro-docker) respectively.

### 2019-11-05 manuscript.Rmd
update introduction

@@
 All this cements the role of R as the lingua franca of statistics, data visualisation, and computational research.
-Coinciding with the rise of R was the advent of Docker [CITATION?] as a general tool for development, distribution, and deployment of software.
-Combining both these topics, the _Rocker Project_([https://www.rocker-project.org/](https://www.rocker-project.org/)) was introduced in 2014 and provides a number of [Docker](https://en.wikipedia.org/wiki/Docker_(software)) images for various use cases as described in \citet{RJ-2017-065}.
-The considerable uptake and continued evolution of the  _Rocker_ suite of containers has lead to numerous projects extending or building upon _Rocker_ images ranging from reproducible research to production deployments.
-This article presents this _Rockerverse_ of projects.
-It also introduces related activities connecting the R language and environment with other containerisation solutions.
-The main contributions is a coherent picture of the current lay of the land of containers in, with, and for `R`.
-This diversity includes demonstrators, early prototypes, and mature projects.
+Coinciding with the rise of R was the advent of [Docker](https://en.wikipedia.org/wiki/Docker_(software)) as a general tool for distribution and deployment of software.
+Combining both these topics, the Rocker Project ([https://www.rocker-project.org/](https://www.rocker-project.org/)) provides images with R (see details next Section).
+The considerable uptake and continued evolution of Rocker images has lead to numerous projects extending or building upon Rocker images, ranging from reproducible research to production deployments.
+This article presents this _Rockerverse_ of projects across all development stages: demonstrators, prototypes, and mature products.
+We also introduce related activities connecting the R language and environment with other containerisation solutions.
+The main contributions is a coherent picture of the current lay of the land of using containers in, with, and for R.
 
@@
 
-Docker, an application and service provide by the eponymous company, has in just a few short years risen to prominence for development, testing, deployment and distribution of computer software.
+Docker, an application and service provide by the eponymous company, has in just a few short years risen to prominence for development, testing, deployment and distribution of computer software \citep[cf.][]{datadog_8_2018,munoz_history_2019}.
 While there are related approaches such as LXC\footnote{\href{https://en.wikipedia.org/wiki/LXC}{https://en.wikipedia.org/wiki/LXC}} or Singularity \citep{kurtzer_singularity_2017}, Docker has become synomymous with "containerization"---the method of taking software artefacts and bundling them in such a way that use becomes standardized and portable across operationg systems.
@@
 But Docker improved on virtualization by accessing the host system---generally Linux---through a much thinner and smaller shim than a full operating system emulation or virtualization.
-This makes for more efficient use of system resources and allowed another order of magnitude in terms of scalability of deployment [REFERENCES!].   
+This makes for more efficient use of system resources and allowed another order of magnitude in terms of scalability of deployment [REFERENCE!].   
 While Docker makes use of Linux kernel features, its importance was large enough so that some required aspects of running Docker have been added to other operating systems to support Docker there more efficiently too \citep{microsoft_linux_2019}.
@@
 For statistical computing and analysis centered around R, the Rocker Project has provided a variety of Docker containers since its start in 2014 \citep{RJ-2017-065}.
-The Rocker Project provides several lines of containeres spanning to from building blocks with R-release or R-devel to containers with RStudio~Server and Shiny~Server.
+The Rocker Project provides several lines of containers spanning to from building blocks with `R-release` or `R-devel`, via containers with RStudio~Server and Shiny~Server, to domain-specific containers such as `geospatial`.
 Also of note is a series of "versioned" containers which match the R release they contain with the _then-current_ set of packages via the MRAN Snapshot views of CRAN \citep{microsoft_cran_2019}.

### 2019-11-06 manuscript.Rmd
Update style file and don't add extra packages for ORCiD while still following ORCID display guidelines
See also https://orcid.org/trademark-and-id-display-guidelines and https://journal.r-project.org/share/author-guide.pdf

@@
     - Germany
-    - \orcid{0000-0002-0024-5046}
+    - \href{https://orcid.org/0000-0002-0024-5046}{\includegraphics[height=\fontcharht\font`\0]{ORCID-iD_icon-vector.pdf} https://orcid.org/0000-0002-0024-5046}
     email: daniel.nuest@uni-muenster.de
@@
       - Champaign, IL 61820
-      - \orcid{0000-0001-6419-907X}
+      - \includegraphics[height=\fontcharht\font`\0]{ORCID-iD_icon-vector.pdf} \href{https://orcid.org/0000-0001-6419-907X}{https://orcid.org/0000-0001-6419-907X}
     email: dirkd@eddelbuettel.com
@@
       - Charlestown, MA 02129, USA
-      - \orcid{0000-0002-0250-5673}
+      - \includegraphics[height=\fontcharht\font`\0]{ORCID-iD_icon-vector.pdf} \href{https://orcid.org/0000-0002-0250-5673}{https://orcid.org/0000-0002-0250-5673}
     email: me@nanx.me
@@
       - Buffalo, New York, 14263 USA
-      - \orcid{0000-0002-5910-4010}
+      - \includegraphics[height=\fontcharht\font`\0]{ORCID-iD_icon-vector.pdf} \href{https://orcid.org/0000-0002-5910-4010}{https://orcid.org/0000-0002-5910-4010}
     email: lori.shepherd@roswellpark.org
@@
     - Buffalo, New York, 14263 USA
-    - \orcid{0000-0002-0224-9817}
+    - \includegraphics[height=\fontcharht\font`\0]{ORCID-iD_icon-vector.pdf} \href{https://orcid.org/0000-0002-0224-9817}{https://orcid.org/0000-0002-0224-9817}
     email: nitesh.turaga@roswellpark.org

### 2019-11-06 manuscript.Rmd
add ORCID for @ColinFay

--- 2019-11-06 manuscript.Rmd
@@
     affiliation: ThinkR
+    address:
+        - \includegraphics[height=\fontcharht\font`\0]{ORCID-iD_icon-vector.pdf} \href{https://orcid.org/0000-0001-7343-1846}{https://orcid.org/0000-0001-7343-1846}
     email: contact@colinfay.me

### 2019-11-06 manuscript.Rmd

@@
 # R Packages
+<!-- from packages closer to Docker to second level packages to more complex ones -->
 
-## Packages for automation and packaging [@nuest]
+## Clients
+<!-- first order packages, talking directyl to Docker -->
 
+- [`stevedore`](https://github.com/richfitz/stevedore)
+- [`docker`](https://bhaskarvk.github.io/docker/)
+- [`harbor`](https://github.com/wch/harbor/)
+- [`dockermachine`](https://github.com/cboettig/dockermachine)
+
+## Capture and create environments [@nuest]
+<!-- second order packages -->
+
+- [`dockerfiler`](https://github.com/ColinFay/dockerfiler/)
 - [`containerit`](https://github.com/o2r-project/containerit/)
 - [`dockertest`](https://github.com/traitecoevo/dockertest/)
-
-### liftr
 
@@
 
-## Packages for control and provisioning [@nuest, @richfitz, @rcannood]
+# Use cases and applications
 
-- [`stevedore`](https://github.com/richfitz/stevedore)
+## Deployment to cloud services [@nuest, @richfitz, @rcannood]
+
 - [`babelwhale`](https://cran.r-project.org/web/packages/babelwhale/index.html): Running a Docker from R with Singularity or Docker as back-end.

### 2019-11-06 manuscript.Rmd
update chapter on versions

@@
 
-## Images for (historic) R versions [@ColinFay, @nuest]
+## Images for (historic) R versions
 
 As with any other software, each new version of R comes with its share of changes.
-Some are breaking changes, some are not, but the fact is that running a piece of code in a given version might give different s from another version ; which, in some case, can be an issue when it comes to reproducibility.
+Some are breaking changes, some are not, but the fact is that running a piece of code in a given version might give different results from another version can be an issue when it comes to reproducibility of workflows and stability of applications.
 For example, think about R random seed: starting with `R 3.6.0`, running `set.seed(2811); sample(1:1000, 2)` will not give the same result as if it was run inside an older version of R.
 That might seem trivial, but all the code using the random number generator will not be exactly reproducible after this breaking change.
-Detecting this kind of behavior is the reason why the [r-online](https://github.com/ColinFay/ronline) app has been developped: available as a standalone NodeJS app or [online](https://srv.colinfay.me/r-online), this app was designed for historic exploration of R: with it, the user can compare a piece of code run in two separate versions of R.
-Internally, this app takes the piece of code written by the user, opens one or two docker instances with the given version of R, and returns the result to the user.
+Controlling the version of software is crucial for reproducible research [REF].
 
-- old R versions: https://github.com/rocker-org/rocker-versioned/issues/138
-- [semantic version tags for Rocker](https://github.com/rocker-org/rocker-versioned/#version-tags)
-- debugging memory problems across multiple versions: https://github.com/wch/r-debug
-- cross-R testing: Show how you can run the same script in multiple R variants in containers (new contribution).
-Ideally the images should be created using `containerit` and controlled with `stevedore`/`docker`, i.e.
-via an R script.
+Containers are perfectly suited to capture a specific configuration of a computing environment to prohibit such problems.
+The Rocker images provides the _versioned stack_ for different the version of R since the project inception, which use the _then-current_ stable Debian base image \citep[cf.][]{RJ-2017-065}.
+Based on custom build phase hooks\footnote{\href{https://docs.docker.com/docker-hub/builds/advanced/}{https://docs.docker.com/docker-hub/builds/advanced/}}, i.e.
+small shell scripts executed at different phases in the automated build of Rocker images, there are also semantic version tags for the most recent freezed (i.e.
+using MRAN) versions with only the major and minor version\footnote{See \href{https://github.com/rocker-org/rocker-versioned/blob/master/VERSIONS.md}{https://github.com/rocker-org/rocker-versioned/blob/master/VERSIONS.md}, FIXME: \href{https://github.com/rocker-org/rocker-versioned/issues/42#issuecomment-550266782}{https://github.com/rocker-org/rocker-versioned/issues/42\#issuecomment-550266782}}.
+So, at the time of writing this article, `rocker/r-ver:3` and `3.6` are aliases for `3.6.0`, because `3.6.1` is the latest release.
+With the release of `3.6.2` and pinning of the MRAN version in `3.6.1`, the tags `3` and `3.6` will deliver the same image as `3.6.1`.
+These tags allow users to update the base image to retrieve bugfixes while reducing the risk of also introducing breaking changes.
+
+[r-online](https://github.com/ColinFay/ronline) is an app for helping users to detect breaking changes between different R versions, and for historic exploration of R.
+With a standalone NodeJS app or [online](https://srv.colinfay.me/r-online), the user can compare a piece of code run in two separate versions of R.
+Internally, r-online opens one or two Docker instances with the given version of R based on Rocker images, executes a given piece of code, and returns the result to the user.
+
+A container also provides an isolated sandbox environment suitable for testing and evaluation, without interfering with the "main" working environment.
+This enables cross version testing and bugfixing.
+To take this even further, the Rocker contributors are discussing the provision of R versions reaching back futher than the project's own inception, reaching back as far as R `2.x` and `1.x` \footnote{\href{https://github.com/rocker-org/rocker-versioned/issues/138}{https://github.com/rocker-org/rocker-versioned/issues/138}}.
+The main challenges are finding a suitable base image with a matching OS version, or making the adjustments to compile R on more recent OS releases.
 

### 2019-11-06 manuscript.Rmd
use consistent writing of Rocker, add content:
- add section on debugging
- add dockyard
- add production use case by @jnolis and @nolistic

--- 2019-11-06 manuscript.Rmd
@@
 Coinciding with the rise of R was the advent of Docker [CITATION?] as a general tool for development, distribution, and deployment of software.
-Combining both these topics, the _Rocker Project_([https://www.rocker-project.org/](https://www.rocker-project.org/)) was introduced in 2014 and provides a number of [Docker](https://en.wikipedia.org/wiki/Docker_(software)) images for various use cases as described in \citet{RJ-2017-065}.
-The considerable uptake and continued evolution of the  _Rocker_ suite of containers has lead to numerous projects extending or building upon _Rocker_ images ranging from reproducible research to production deployments.
+Combining both these topics, the _Rocker Project_ ([https://www.rocker-project.org/](https://www.rocker-project.org/)) was introduced in 2014 and provides a number of [Docker](https://en.wikipedia.org/wiki/Docker_(software)) images for various use cases as described in \citet{RJ-2017-065}.
+The considerable uptake and continued evolution of the Rocker suite of containers has lead to numerous projects extending or building upon Rocker images ranging from reproducible research to production deployments.
 This article presents this _Rockerverse_ of projects.
@@
 `rgdal` \footnote{\href{http://packages.renjin.org/package/org.renjin.cran/rgdal/1.4-4/build/1}{http://packages.renjin.org/package/org.renjin.cran/rgdal/1.4-4/build/1}}.
-
 There are no offical Docker images for Renjin, but community-maintained images for selected releases only are available under `nuest/renjin` on Docker Hub and GitHub at [https://hub.docker.com/r/nuest/renjin](https://hub.docker.com/r/nuest/renjin) and [https://github.com/nuest/renjin-docker](https://github.com/nuest/renjin-docker) respectively.
@@
 It is developed by Oracle, connects R to the GraalVM ecosystem \citep{wikipedia_graalvm_2019}, and also claims superior performance but also targets full compatibility with base R \footnote{ \href{https://github.com/oracle/fastr}{https://github.com/oracle/fastr}}.
-
 There are no offical Docker images provided, but `nuest/fastr-docker` provides image and `Dockerfile` on Docker Hub and GitHub at []() and [https://github.com/nuest/fastr-docker](https://github.com/nuest/fastr-docker) respectively.
 
-While the images presented in this section are far from being as vetted, stable, and widely used as any of the _Rocker_ images, they demonstrate an important advantage of containerisation technology, namely the abiliy to transparently build portable stacks of open source software and make them easily accessible to users.
+While the images presented in this section are far from being as vetted, stable, and widely used as any of the Rocker images, they demonstrate an important advantage of containerisation technology, namely the abiliy to transparently build portable stacks of open source software and make them easily accessible to users.
 All different distributions are published under GPL licenses.
@@
 _Bioconductor_ installs packages based on the R version, and therefore uses `rocker/rstudio` and `rocker/r-ver` version tagging.
-_Bioconductor_ selects the desired version of R from _Rocker_, adds the BiocManager CRAN package for installing appropriate versions of _Bioconductor_ packages, and creates a _Bioconductor_ docker image with an informative tag (R\_version\_Bioc\_version).
+_Bioconductor_ selects the desired version of R from Rocker, adds the BiocManager CRAN package for installing appropriate versions of _Bioconductor_ packages, and creates a _Bioconductor_ docker image with an informative tag (R\_version\_Bioc\_version).
 The images are summarized on the _Bioconductor_ web site (https://bioconductor.org/help/docker/), maintained on GitHub ([https://github.com/Bioconductor/bioc_docker](https://github.com/Bioconductor/bioc_docker)), and available to the community through [DockerHub](https://hub.docker.com/).
@@
 _Bioconductor_ images for proteomics, metabolomics, and flow cytometry are community maintained.
-All community maintained images build on top of the _Bioconductor_ base image and therefore indirectly the _Rocker_ images.
+All community maintained images build on top of the _Bioconductor_ base image and therefore indirectly the Rocker images.
 To simplify building and maintaining _Bioconductor_ images, we use a Ruby templating engine.
@@
 
-- old R versions: https://github.com/rocker-org/rocker-versioned/issues/138
-- [semantic version tags for Rocker](https://github.com/rocker-org/rocker-versioned/#version-tags)
-- debugging memory problems across multiple versions: https://github.com/wch/r-debug
-- cross-R testing: Show how you can run the same script in multiple R variants in containers (new contribution).
-Ideally the images should be created using `containerit` and controlled with `stevedore`/`docker`, i.e.
-via an R script.
+Containers are perfectly suited to capture a specific configuration of a computing environment to prohibit such problems.
+The Rocker images provides the _versioned stack_ for different the version of R since the project inception, which use the _then-current_ stable Debian base image \citep[cf.][]{RJ-2017-065}.
+Based on custom build phase hooks\footnote{\href{https://docs.docker.com/docker-hub/builds/advanced/}{https://docs.docker.com/docker-hub/builds/advanced/}}, i.e.
+small shell scripts executed at different phases in the automated build of Rocker images, there are also semantic version tags for the most recent freezed (i.e.
+using MRAN) versions with only the major and minor version\footnote{See \href{https://github.com/rocker-org/rocker-versioned/blob/master/VERSIONS.md}{https://github.com/rocker-org/rocker-versioned/blob/master/VERSIONS.md}, FIXME: \href{https://github.com/rocker-org/rocker-versioned/issues/42}{https://github.com/rocker-org/rocker-versioned/issues/42}.}.
+So, at the time of writing this article, `rocker/r-ver:3` and `3.6` are aliases for `3.6.0`, because `3.6.1` is the latest release.
+With the release of `3.6.2` and pinning of the MRAN version in `3.6.1`, the tags `3` and `3.6` will deliver the same image as `3.6.1`.
+These tags allow users to update the base image to retrieve bugfixes while reducing the risk of also introducing breaking changes.
 
@@
 - [`dockermachine`](https://github.com/cboettig/dockermachine)
+- [`dockyard`](https://github.com/thebioengineer/dockyard)
 
@@
 - [ShinyProxy](https://www.shinyproxy.io/) [creates a container](https://github.com/openanalytics/shinyproxy/blob/master/src/main/java/eu/openanalytics/services/DockerService.java#L388) for each user
+- [production deployment of neural networks](https://github.com/tmobile/r-tensorflow-api) [@jnolis, @nolistic]
 
@@
 ...
+
+## Development and debugging
+
+Containers can also serve as useful playgrounds to create environments ad-hoc or to provide very specific environments that are not often needed.
+Developers can readily start a local container based on a Rocker image to investigate a bug report, which might require them to install a specific combination of package versions or use a specific R version (see also versioned images above).
+Unlike recreating the reporter's environment on their regular machine, a container can simply be discarded after use, and there is no need to tediously roll back explorative changes to a previous state.
+Using the Rocker images with RStudio, the disposable environments lack no development comfort (see also the usage in research compendia).
+
+[`r-debug`](https://github.com/wch/r-debug) is a purpose built Docker image for debugging R memory problems.
+\citet{eddelbuettel_debugging_2019} describes how a Docker container was used to debug an issue with a package only occuring with a particular version of Fortran, and using tools which are not readily available on all platforms (e.g., not on macOS).
 

### 2019-11-06 manuscript.Rmd

@@
 
-## Images for (historic) R versions [@ColinFay, @nuest]
+## Images for (historic) R versions
 
 As with any other software, each new version of R comes with its share of changes.
-Some are breaking changes, some are not, but the fact is that running a piece of code in a given version might give different s from another version ; which, in some case, can be an issue when it comes to reproducibility.
+Some are breaking changes, some are not, but the fact is that running a piece of code in a given version might give different results from another version can be an issue when it comes to reproducibility of workflows and stability of applications.
 For example, think about R random seed: starting with `R 3.6.0`, running `set.seed(2811); sample(1:1000, 2)` will not give the same result as if it was run inside an older version of R.
 That might seem trivial, but all the code using the random number generator will not be exactly reproducible after this breaking change.
-Detecting this kind of behavior is the reason why the [r-online](https://github.com/ColinFay/ronline) app has been developped: available as a standalone NodeJS app or [online](https://srv.colinfay.me/r-online), this app was designed for historic exploration of R: with it, the user can compare a piece of code run in two separate versions of R.
-Internally, this app takes the piece of code written by the user, opens one or two docker instances with the given version of R, and returns the result to the user.
+Controlling the version of software is crucial for reproducible research [REF].
 
@@
 These tags allow users to update the base image to retrieve bugfixes while reducing the risk of also introducing breaking changes.
+
+[r-online](https://github.com/ColinFay/ronline) is an app for helping users to detect breaking changes between different R versions, and for historic exploration of R.
+With a standalone NodeJS app or [online](https://srv.colinfay.me/r-online), the user can compare a piece of code run in two separate versions of R.
+Internally, r-online opens one or two Docker instances with the given version of R based on Rocker images, executes a given piece of code, and returns the result to the user.
+
+A container also provides an isolated sandbox environment suitable for testing and evaluation, without interfering with the "main" working environment.
+This enables cross version testing and bugfixing.
+To take this even further, the Rocker contributors are discussing the provision of R versions reaching back futher than the project's own inception, reaching back as far as R `2.x` and `1.x` \footnote{\href{https://github.com/rocker-org/rocker-versioned/issues/138}{https://github.com/rocker-org/rocker-versioned/issues/138}}.
+The main challenges are finding a suitable base image with a matching OS version, or making the adjustments to compile R on more recent OS releases.
 

### 2019-11-06 manuscript.Rmd

--- 2019-11-06 manuscript.Rmd
@@
 - Binder
-  - [`holepunch`](https://github.com/karthik/holepunch) streamlines making an R research compendium Binder-ready [@karthik]
+
+The [Binder project](https://mybinder.readthedocs.io/en/latest/), maintained by the team behind Jupyter, makes it possible for users to create and share their computing environments with others.
+A deployed version of this service called binder hub allows anyone with access to a web browser and an internet connection to launch a temporary instance of these custom environments and execute any notebooks contained within.
+From a reproducibility standpoint, Binder makes it exceedingly easy to compile a paper, visualize data, and run small examples from papers or tutorials without the need for any local installation.
+
+To set up Binder for a project, a user typically starts at an instance of a Binderhub (examples include mybinder.org and binder.pangeo.io) and passes the location of a hosted Git repository.
+Using internal tools such as [repo2docker](https://repo2docker.readthedocs.io/en/latest/config_files.html), the binderhub builds a Dockerfile by parsing the contents of all code contained within.
+While this approach works well for most run of the mill Python projects, it is not so seamless for R projects.
+For any R projects that use the Tidyverse suite [citation in review.
+Will add by the time this gets submitted], the time and resources required to build all dependencies from source can often time out before completion, making it frustrating for the average R user.
+
+[`holepunch`](https://github.com/karthik/holepunch) is a R package that was designed to remove some of these limitations and make Binder more accessible to novice R users.
+`holepunch` achieves this by leveraging Rocker images that contains the Tidyverse along special Jupyter dependencies, and only installs additional packages from CRAN and Bioconductor that are not already part of these images.
+Starting with the Binder/Tidyverse base images eliminates a large part of the build time and in most cases results in a binder instance launching within a minute.
+`holepunch` as a side effect also creates a DESCRIPTION file which then turns any project into a research compendium [doi:10.1080/00031305.2017.1375986].
+The Dockerfile included with the project can also be used to launch a RStudio server locally independent of binder which is especially useful when more computational resources are required.
+
   - [`repo2docker`](https://repo2docker.readthedocs.io/en/latest/config_files.html) (`install.R`, `DESCRPTION`, `runtime.txt`) based on the Rocker images [@nuest]

### 2019-11-06 manuscript.Rmd
one sentence per line
restructure contribution by @karthik, minor changes, add reference

@@
 
-## Common or public work environments [div]
+## Common or public work environments
 
-- Binder
+The fact that Docker images are portable and well defined make them useful when more than one person needs access to the same computing environment.
+This is even more useful when some of the users do not have the expertise to create such an environment themselves, and when these environments can be run in public or shared infrastructure.
 
-The [Binder project](https://mybinder.readthedocs.io/en/latest/), maintained by the team behind Jupyter, makes it possible for users to create and share their computing environments with others.
-A deployed version of this service called binder hub allows anyone with access to a web browser and an internet connection to launch a temporary instance of these custom environments and execute any notebooks contained within.
+[`holepunch`](https://github.com/karthik/holepunch) is an R package that was designed to make sharing work environments accessible to novice R users based on _Binder_.
+The [Binder project](https://mybinder.readthedocs.io/en/latest/), maintained by the team behind Jupyter, makes it possible for users to create and share their computing environments with others \citep{jupyter_binder_2018}.
+A _BinderHub_ allows anyone with access to a web browser and an internet connection to launch a temporary instance of these custom environments and execute any workflows contained within.
 From a reproducibility standpoint, Binder makes it exceedingly easy to compile a paper, visualize data, and run small examples from papers or tutorials without the need for any local installation.
-
-To set up Binder for a project, a user typically starts at an instance of a Binderhub (examples include mybinder.org and binder.pangeo.io) and passes the location of a hosted Git repository.
-Using internal tools such as [repo2docker](https://repo2docker.readthedocs.io/en/latest/config_files.html), the binderhub builds a Dockerfile by parsing the contents of all code contained within.
+To set up Binder for a project, a user typically starts at an instance of a BinderHub and passes the location of a repository with a workspace, e.g., a hosted Git repository, or a data repository like Zenodo.
+Binder's core internal tool is [repo2docker](https://repo2docker.readthedocs.io/en/latest/config_files.html).
+It deterministically builds a Docker image by parsing the contents of a repository, e.g. project dependency configurations or simple configuration files.
+In the most powerful case, `repo2docker` builds a given Dockerfile.
 While this approach works well for most run of the mill Python projects, it is not so seamless for R projects.
-For any R projects that use the Tidyverse suite [citation in review.
-Will add by the time this gets submitted], the time and resources required to build all dependencies from source can often time out before completion, making it frustrating for the average R user.
-
-[`holepunch`](https://github.com/karthik/holepunch) is a R package that was designed to remove some of these limitations and make Binder more accessible to novice R users.
-`holepunch` achieves this by leveraging Rocker images that contains the Tidyverse along special Jupyter dependencies, and only installs additional packages from CRAN and Bioconductor that are not already part of these images.
-Starting with the Binder/Tidyverse base images eliminates a large part of the build time and in most cases results in a binder instance launching within a minute.
-`holepunch` as a side effect also creates a DESCRIPTION file which then turns any project into a research compendium [doi:10.1080/00031305.2017.1375986].
-The Dockerfile included with the project can also be used to launch a RStudio server locally independent of binder which is especially useful when more computational resources are required.
-
-  - [`repo2docker`](https://repo2docker.readthedocs.io/en/latest/config_files.html) (`install.R`, `DESCRPTION`, `runtime.txt`) based on the Rocker images [@nuest]
-- GPU [@noamross, @cboettig]
-- Gigantum stack
+For any R projects that use the Tidyverse suite [citation in review, will add by the time this gets submitted], the time and resources required to build all dependencies from source can often time out before completion, making it frustrating for the average R user.
+`holepunch` removes some of these limitations by leveraging Rocker images that contain the Tidyverse along special Jupyter dependencies, and only installs additional packages from CRAN and Bioconductor that are not already part of these images.
+It short cicuits the configuration file parsing in `repo2docker` and starts with the Binder/Tidyverse base images, which eliminates a large part of the build time and in most cases results in a binder instance launching within a minute.
+`holepunch` as a side effect also creates a `DESCRIPTION` file which then turns any project into a research compendium \citep{marwick_packaging_2018}.
+The Dockerfile included with the project can also be used to launch a RStudio server locally independent of Binder which is especially useful when more or special computational resources are required.
 
@@
 
+- GPU environments [@noamross, @cboettig]
 - [RCloud](https://github.com/att/rcloud/tree/master/docker) ?

### 2019-11-06 manuscript.Rmd

--- 2019-11-06 manuscript.Rmd
@@
 
-## Common or public work environments [div]
-
-- Binder
+## Common or public work environments
 

### 2019-11-06 manuscript.Rmd
one sentence per line, fix remaining bullet points in common envs section

@@
 
-  - [`repo2docker`](https://repo2docker.readthedocs.io/en/latest/config_files.html) (`install.R`, `DESCRPTION`, `runtime.txt`) based on the Rocker images [@nuest]
-- GPU [@noamross, @cboettig]
+One use for containers is to run on shared local hardware where teams manage their own high-performance servers.
+This can follow one of several design patterns: users may deploy containers to hardware as a work environment for a specific project, conatiners may provide per-user persistent environments, or a single container can act as a common multi-user environment for a server.
+The former models provide modularity, while the latter approach is most similar to a simple shared server.
+In all cases, though, the containerized approach provides several advantages: First, users may use the same image and thus work environment on desktop and laptop computers, as well.
+Second, software updates can be achieved by updating and redeploying the container, rather than tracking local installs on each server.
+Third, the containerized environment can be quickly deployed to other hardware, cloud or local, if more resources are neccessary or in case of server destruction or failure.
+In any of these cases, users need a method to interact with the containers, be it and IDE, or command-like access and tools such as SSH, which is usually not part of standard container recipes and must be added.
+The Rocker project provides containers pre-installed with the RStudio IDE.
 
-One use for containers is to run on shared local hardware where teams manage
-their own high-performance servers  This can follow one of several
-design patterns: users may deploy containers to hardware as a work environment
-for a specific project, conatiners may provide per-user persistent environments, or a single container can act as
-a common multi-user environment for a server.  The former models provide modularity,
-while the latter approach is most similar to a simple shared server. In all cases,
-though, the containerized approach provides several advantages: First, users may
-use the same image and thus work environment on desktop and laptop computers, as well.
-Second, software updates can be achieved by updating and redeploying the container,
-rather than tracking local installs on each server.  Third, the containerized
-environment can be quickly deployed to other hardware, cloud or local, if 
-more resources are neccessary or in case of server destruction or failure. In any of
-these cases, users need a method to interact with the containers, be it and
-IDE, or command-like access and tools such as SSH, which is usually not part
-of standard container recipes and must be added.  The Rocker project provides
-containers pre-installed with the RStudio IDE.
-
-In cases where users store nontrivial amounts of data for their projects, data
-needs to persist beyond the life of the container. This may be via in shared disks, attached
-network volumes, or in separate storage where it is uploaded between sessions. In
-the case of shared disks or network-attached volumes, care must be taken to
-persist user permissions, and of course backups are still neccessary.
-When working with multiple servers, an automation
-framework such as [Ansible](https://www.ansible.com) may be useful for managing
-users, permisions, and disks along with containers.
+In cases where users store nontrivial amounts of data for their projects, data needs to persist beyond the life of the container.
+This may be via in shared disks, attached network volumes, or in separate storage where it is uploaded between sessions.
+In the case of shared disks or network-attached volumes, care must be taken to persist user permissions, and of course backups are still neccessary.
+When working with multiple servers, an automation framework such as [Ansible](https://www.ansible.com) may be useful for managing users, permisions, and disks along with containers.
 
 In some cases, containers are not completely portable between hardware environments.
-This is the case for software using graphical processing units (GPUs)
-which are increasingly popular for compute-intensive machine learning tasks.
-Containers running GPU software  require drivers and libraries specific
-to GPU models and versions, and containers require a specialized runtime to connect
-to the underlying GPU hardware. For NVIDIA GPUs, the [NVIDIA Container Toolkit](https://github.com/NVIDIA/nvidia-docker)
-includes a specialized runtime plugn for docker and set of base images with
-appropriate drivers and libraries, and the Rocker project has (beta) images
-based on these that include GPU-enabled versions of machine-learning R packages.
-
-- Gigantum stack
+This is the case for software using graphical processing units (GPUs) which are increasingly popular for compute-intensive machine learning tasks.
+Containers running GPU software  require drivers and libraries specific to GPU models and versions, and containers require a specialized runtime to connect to the underlying GPU hardware.
+For NVIDIA GPUs, the [NVIDIA Container Toolkit](https://github.com/NVIDIA/nvidia-docker) includes a specialized runtime plugn for docker and set of base images with appropriate drivers and libraries, and the Rocker project has (beta) images based on these that include GPU-enabled versions of machine-learning R packages.
 
@@
 
-- [RCloud](https://github.com/att/rcloud/tree/master/docker) ?
+[RCloud](https://github.com/att/rcloud/tree/master/docker) uses a `rocker/drd` base image for creation of collaborative data analysis and visualisation environments.
 

### 2019-11-06 manuscript.Rmd

--- 2019-11-06 manuscript.Rmd
@@
 
+One use for containers is to run on shared local hardware where teams manage their own high-performance servers.
+This can follow one of several design patterns: users may deploy containers to hardware as a work environment for a specific project, conatiners may provide per-user persistent environments, or a single container can act as a common multi-user environment for a server.
+In all cases, though, the containerized approach provides several advantages: First, users may use the same image and thus work environment on desktop and laptop computers, as well.
+The former models provide modularity, while the latter approach is most similar to a simple shared server.
+Second, software updates can be achieved by updating and redeploying the container, rather than tracking local installs on each server.
+Third, the containerized environment can be quickly deployed to other hardware, cloud or local, if more resources are neccessary or in case of server destruction or failure.
+In any of these cases, users need a method to interact with the containers, be it and IDE, or command-like access and tools such as SSH, which is usually not part of standard container recipes and must be added.
+The Rocker project provides containers pre-installed with the RStudio IDE.
+
+In cases where users store nontrivial amounts of data for their projects, data needs to persist beyond the life of the container.
+This may be via in shared disks, attached network volumes, or in separate storage where it is uploaded between sessions.
+In the case of shared disks or network-attached volumes, care must be taken to persist user permissions, and of course backups are still neccessary.
+When working with multiple servers, an automation framework such as [Ansible](https://www.ansible.com) may be useful for managing users, permisions, and disks along with containers.
+
+In some cases, containers are not completely portable between hardware environments.
+This is the case for software using graphical processing units (GPUs) which are increasingly popular for compute-intensive machine learning tasks.
+Containers running GPU software  require drivers and libraries specific to GPU models and versions, and containers require a specialized runtime to connect to the underlying GPU hardware.
+For NVIDIA GPUs, the [NVIDIA Container Toolkit](https://github.com/NVIDIA/nvidia-docker) includes a specialized runtime plugn for docker and set of base images with appropriate drivers and libraries, and the Rocker project has (beta) images based on these that include GPU-enabled versions of machine-learning R packages.
+
 R is a fantastic tool when it comes to interfacing with databases: almost every open source and proprietary database system has an R package that allows users to connect and interact with the it 
@@
 
-- GPU environments [@noamross, @cboettig]
-- [RCloud](https://github.com/att/rcloud/tree/master/docker) ?
+[RCloud](https://github.com/att/rcloud/tree/master/docker) uses a `rocker/drd` base image for creation of collaborative data analysis and visualisation environments.
 

### 2019-11-06 manuscript.Rmd
restructure common envs section

@@
 
-One use for containers is to run on shared local hardware where teams manage their own high-performance servers.
+In **high-performance computing**, one use for containers is to run workflows on shared local hardware where teams manage their own high-performance servers.
 This can follow one of several design patterns: users may deploy containers to hardware as a work environment for a specific project, conatiners may provide per-user persistent environments, or a single container can act as a common multi-user environment for a server.
@@
 The Rocker project provides containers pre-installed with the RStudio IDE.
-
 In cases where users store nontrivial amounts of data for their projects, data needs to persist beyond the life of the container.
@@
 
-In some cases, containers are not completely portable between hardware environments.
-This is the case for software using graphical processing units (GPUs) which are increasingly popular for compute-intensive machine learning tasks.
+Using **GPUs** (graphical processing units) as a specialised hardware from containerized common work environments is also possible [REF].
+GPUs are increasingly popular for compute-intensive machine learning tasks [REF].
+Though in this case, containers are not completely portable between hardware environments.
 Containers running GPU software require drivers and libraries specific to GPU models and versions, and containers require a specialized runtime to connect to the underlying GPU hardware.
-For NVIDIA GPUs, the [NVIDIA Container Toolkit](https://github.com/NVIDIA/nvidia-docker) includes a specialized runtime plugn for docker and set of base images with appropriate drivers and libraries, and the Rocker project has (beta) images based on these that include GPU-enabled versions of machine-learning R packages.
+For NVIDIA GPUs, the [NVIDIA Container Toolkit](https://github.com/NVIDIA/nvidia-docker) includes a specialized runtime plugin for Docker and a set of base images with appropriate drivers and libraries.
+The Rocker project has (beta) images based on these that include GPU-enabled versions of machine-learning R packages: https://github.com/rocker-org/ml.
 
+**Teaching** is a further example where sharing a prepared computing environment can greatly improve the process, especially for courses that require access to a relatively complex setup of software tools, e.g., as in the case of database systems.
 R is a fantastic tool when it comes to interfacing with databases: almost every open source and proprietary database system has an R package that allows users to connect and interact with the it 
@@
 But learning and teaching these tools comes with a cost: the cost of deploying or having access to an environment with the softwares and drivers installed. 
-For people teaching R, it can become a barrier if they need to install local versions of the drivers, or to connect to remote instances which might or might not be made available by the IT. 
-Giving access to a sandbox for the most common database environments is the idea behind [`r-db`](https://github.com/ColinFay/r-db) [@ColinFay], a docker image that contains everything needed to connect, from R, to a database.
+For people teaching R, it can become a barrier if they need to install local versions of the drivers, or to connect to remote instances which might or might not be made available by IT services.
+Giving access to a sandbox for the most common database environments is the idea behind **[`r-db`](https://github.com/ColinFay/r-db)** [@ColinFay], a Docker image that contains everything needed to connect, from R, to a database.
 Notably, with `r-db`, the users don't have to install complex drivers or to configure their machine in a specific way. 
-On top of that, `r-db` comes with a comprehensive [online guide](http://colinfay.me/r-db/), explaining how to spin a docker instance of a specific database, and how to use `r-db` to connect to it. 
+On top of that, `r-db` comes with a comprehensive [online guide](http://colinfay.me/r-db/), explaining how to spin a Docker instance of a specific database, and how to use `r-db` to connect to it.
 Each package contained inside this image also has a series examples, so that the user can get started with the database right away.
+The `rocker/tidyverse` base image ensures that users can also readily use packages for analysis, display, and reporting.
 
-[RCloud](https://github.com/att/rcloud/tree/master/docker) uses a `rocker/drd` base image for creation of collaborative data analysis and visualisation environments.
+**[RCloud](https://github.com/att/rcloud/tree/master/docker)** uses a `rocker/drd` base image for creation of collaborative data analysis and visualisation environments.
 

### 2019-11-06 manuscript.Rmd
highlight main section content in bold

--- 2019-11-06 manuscript.Rmd
@@
 
-_Microsoft R Open_ ([MRO](https://github.com/nuest/mro-docker)) is an R distribution formerly known as Revolution R Open (RRO) before Revolution Analytics was acquired by Microsoft.
+**Microsoft R Open** ([MRO](https://github.com/nuest/mro-docker)) is an R distribution formerly known as Revolution R Open (RRO) before Revolution Analytics was acquired by Microsoft.
 MRO is compatible with main R and it's packages.sss
@@
 
-_Renjin_ is an interpreter for the R language running on top of the [Java Virtual Machine](https://en.wikipedia.org/wiki/Java_virtual_machine) (JVM) providing full two-way access between Java and R code \citep{wikipedia_renjin_2018}.
+**Renjin** is an interpreter for the R language running on top of the [Java Virtual Machine](https://en.wikipedia.org/wiki/Java_virtual_machine) (JVM) providing full two-way access between Java and R code \citep{wikipedia_renjin_2018}.
 It was developed to combine the benefits of R, such scripting and extension packages, with the JVM's advantages in the areas of security, cross-platform availability, and established position in enterprise settings.
@@
 
-[_pqR_](http://www.pqr-project.org/) is a "a pretty quick version of R".
+**[pqR](http://www.pqr-project.org/)** is a "a pretty quick version of R".
 _pqR_ fixes some opinionated issues in the R language and is the basis for experimental features, e.g.
@@
 
-_FastR_ is "A high-performance implementation of the R programming language, built on GraalVM" (https://github.com/oracle/fastr).
+**FastR** is "A high-performance implementation of the R programming language, built on GraalVM" (https://github.com/oracle/fastr).
 It is developed by Oracle, connects R to the GraalVM ecosystem \citep{wikipedia_graalvm_2019}, and also claims superior performance but also targets full compatibility with base R \footnote{ \href{https://github.com/oracle/fastr}{https://github.com/oracle/fastr}}.
-There are no offical Docker images provided, but `nuest/fastr-docker` provides image and `Dockerfile` on Docker Hub and GitHub at []() and [https://github.com/nuest/fastr-docker](https://github.com/nuest/fastr-docker) respectively.
+There are no offical Docker images provided, but `nuest/fastr` images and `Dockerfile` are independently maintained on [Docker Hub](https://cloud.docker.com/repository/docker/nuest/fastr/) and [GitHub](https://github.com/nuest/fastr-docker) respectively.
 
@@
 
-[_Bioconductor_](http://bioconductor.org/help/docker/) is an open source, open development project for the analysis and comprehension of genomic data \citep{gentleman_bioconductor_2004}.
+[_Bioconductor_](http://bioconductor.org/) is an open source, open development project for the analysis and comprehension of genomic data \citep{gentleman_bioconductor_2004}.
 The project consists of 1741 R software packages as of August 15th 2019, as well as packages containing annotation or experiment data.
@@
 Docker images allow availability of current and past versions of _Bioconductor_ for convenience and reproducibility.
-
 _Bioconductor_ 'base' docker images are built on top of `rocker/r-ver` and `rocker/rstudio`.
@@
 To simplify building and maintaining _Bioconductor_ images, we use a Ruby templating engine.
-
-A recent audit of our Dockerfiles, following best practices from the Docker website, led to a reduction in the size and number of layers in images we produce.
+A recent audit of the _Bioconductor_ Dockerfiles, following best practices from the Docker website, led to a reduction in the size and number of layers.
 The most important insights involve the ['union' file system used by Docker](https://docs.docker.com/storage/storagedriver/overlayfs-driver/#how-container-reads-and-writes-work-with-overlay-or-overlay2).
@@
 
-[_liftr_](https://nanx.me/liftr/) \citep{liftr2019} aims to solve the problem of persistent reproducible reporting in statistical computing.
+**[_liftr_](https://nanx.me/liftr/)** \citep{liftr2019} aims to solve the problem of persistent reproducible reporting in statistical computing.
 Currently, the R Markdown format and its backend compilation engine _knitr_ offer a _de facto_ standard for creating dynamic documents \citep{xie2018}.
@@
 
-For example, the [`golem`](https://github.com/ThinkR-open/golem) package (an R package for building production-grade Shiny application) makes an heavy use of `dockerfiler` when it comes to creating the `Dockerfile` for deploying the Shiny application to production.
+For example, the **[`golem`](https://github.com/ThinkR-open/golem)** package makes an heavy use of `dockerfiler` when it comes to creating the `Dockerfile` for building production-grade Shiny applications and deploying them.
 The reason behind this is that `dockerfiler`, on top of being scriptable from R, can leverage all the tools available in R to parse a `DESCRIPTION` file, to get system requirements, to list dependencies, versions, etc.
@@
 
-[`r-debug`](https://github.com/wch/r-debug) is a purpose built Docker image for debugging R memory problems.
+**[`r-debug`](https://github.com/wch/r-debug)** is a purpose built Docker image for debugging R memory problems.
 \citet{eddelbuettel_debugging_2019} describes how a Docker container was used to debug an issue with a package only occuring with a particular version of Fortran, and using tools which are not readily available on all platforms (e.g., not on macOS).
@@
 
-DN conceived of the presented idea and \href{https://github.com/nuest/rockerverse-paper/issues/3}{initialised the formation of the writing team}.
+DN conceived of the presented idea and \href{https://github.com/nuest/rockerverse-paper/issues/3}{initialised the formation of the writing team} and wrote the section about images for R distributions.
 CB ..

### 2019-11-06 manuscript.Rmd
add some intro sentences, fix docker example command

@@
 ## Clients
-<!-- first order packages, talking directyl to Docker -->
+
+A number first order R packages exist that talk directly to Docker.
+They can be used to control the Docker daemon, e.g. to build images or run containers, from R without explicit knowledge of the Docker command line interface.
 
@@
 ## Capture and create environments [@nuest]
-<!-- second order packages -->
+
+Several second order R packages attempt to make the process of creating Docker images and using containers for specific tasks, such as running tests or rendering reproducible reports, easier.
 
@@
 
-## Deployment to cloud services [@nuest, @richfitz, @rcannood]
+## Deployment to cloud services [div]
+
+The cloud is the natural environment of containers, and becomes the go-to mechanism to expose applications written in R to users.
 
@@
 - [`analogsea`](https://github.com/sckott/analogsea) (digital ocean R client)
+- [`plumber`](https://www.rplumber.io/docs/hosting.html#docker)
+- [production deployment of neural networks](https://github.com/tmobile/r-tensorflow-api) [@jnolis, @nolistic]
 
@@
 
-- R-Hub
+The controlled nature of containers, i.e.
+it is possible to define the software environment very well, even on remote machines, make them also useful for continuous integration (CI) and continuous delivery of applications.
+
 - DevOps
@@
 
-[RCloud](https://github.com/att/rcloud/tree/master/docker) uses a `rocker/drd` base image for creation of collaborative data analysis and visualisation environments.
+- [RCloud](https://github.com/att/rcloud/tree/master/docker) uses a `rocker/drd` base image for creation of collaborative data analysis and visualisation environments.
+- [ShinyProxy](https://www.shinyproxy.io/) [creates a container](https://github.com/openanalytics/shinyproxy/blob/master/src/main/java/eu/openanalytics/services/DockerService.java#L388) for each user
 
-## Deployment, processing, cloud [div]
+## Processing [div]
+
+The portability of containers becomes particularly useful when complex processing tasks shall be offloaded to a server.
 
@@
 An R implementation is shown here at [cloudRunR](https://github.com/MarkEdmondson1234/cloudRunR) which uses it to create a scalable R plumber API.
-- [`plumber`](https://www.rplumber.io/docs/hosting.html#docker)
 - `batchtools` \citep{Lang2017batchtools} can [schedule jobs with Docker Swarm](https://mllg.github.io/batchtools/reference/makeClusterFunctionsDocker.html)
@@
 Again, all of these containers are being built on travis-ci, and will only be pushed to docker hub if the integration test succeeds.
-- [ShinyProxy](https://www.shinyproxy.io/) [creates a container](https://github.com/openanalytics/shinyproxy/blob/master/src/main/java/eu/openanalytics/services/DockerService.java#L388) for each user
-- [production deployment of neural networks](https://github.com/tmobile/r-tensorflow-api) [@jnolis, @nolistic]
 
@@
 
-# Other containerisation platforms [NN]
+- **[R-Hub](https://github.com/r-hub/)**
 
-- R images for Singularity
-- Running Rocker images with
-  - Singularity
-  - CoreOS rkt?
-- nix?
+# Other containerisation platforms
+
+Docker is not the only containerisation software.
+An alternative stemming from the domain of high-perfomance computing is Singularity \citep{kurtzer_singularity_2017}.
+Singularity can run Docker images, and in the case of Rocker works out of the box if the main process is R, e.g., in `rocker/r-base`, but does not succeed in running images where there is an init script, e.g.
+in containers that by default run RStudio.
+In the latter case, a `Singularity` file, a set of instructions akin to a `Dockerfile` needs to be used.
 

### 2019-11-07 manuscript.Rmd
update affiliations and addresses

@@
   - name: Daniel Nüst
-    affiliation: Institute for Geoinformatics, University of Münster
+    affiliation: University of Münster
     address:
+      - Institute for Geoinformatics
       - Heisenbergstr. 2
-    - 48149 Münster
-    - Germany
+      - 48149 Münster, Germany
       - \href{https://orcid.org/0000-0002-0024-5046}{\includegraphics[height=\fontcharht\font`\0]{ORCID-iD_icon-vector.pdf} https://orcid.org/0000-0002-0024-5046}
@@
       - Illini Hall, 725 S Wright St
-      - Champaign, IL 61820
+      - Champaign, IL 61820, USA
       - \includegraphics[height=\fontcharht\font`\0]{ORCID-iD_icon-vector.pdf} \href{https://orcid.org/0000-0001-6419-907X}{https://orcid.org/0000-0001-6419-907X}
@@
     address:
+      - 5O rue Arthur Rimbaud
+      - 93300 Aubervilliers, France
       - \includegraphics[height=\fontcharht\font`\0]{ORCID-iD_icon-vector.pdf} \href{https://orcid.org/0000-0001-7343-1846}{https://orcid.org/0000-0001-7343-1846}
@@
   - name: Karthik Ram
-    affiliation: Berkeley Institute for Data Science, University of California, Berkeley, CA 94720 USA
+    affiliation: Berkeley Institute for Data Science
+    address:
+      - University of California
+      - Berkeley, CA 94720, USA
+      - \includegraphics[height=\fontcharht\font`\0]{ORCID-iD_icon-vector.pdf} \href{https://orcid.org/0000-0002-0233-1757}{https://orcid.org/0000-0002-0233-1757}
     email: karthik.ram@berkeley.edu
   - name: Noam Ross
-    #affiliation: -
-    #email: -
+    affiliation: EcoHealth Alliance
+    address:
+      - 460 W 34th St., Ste. 1701
+      - New York, NY 10001, USA
+      - \includegraphics[height=\fontcharht\font`\0]{ORCID-iD_icon-vector.pdf} \href{https://orcid.org/0000-0002-0233-1757}{https://orcid.org/0000-0002-0233-1757}
+    email: ross@ecohealthalliance.org
   - name: Nan Xiao
@@
       - Elm & Carlton Streets
-      - Buffalo, New York, 14263 USA
+      - Buffalo, New York, 14263, USA
       - \includegraphics[height=\fontcharht\font`\0]{ORCID-iD_icon-vector.pdf} \href{https://orcid.org/0000-0002-5910-4010}{https://orcid.org/0000-0002-5910-4010}
@@
     - Elm & Carlton Streets
-    - Buffalo, New York, 14263 USA
+    - Buffalo, New York, 14263, USA
     - \includegraphics[height=\fontcharht\font`\0]{ORCID-iD_icon-vector.pdf} \href{https://orcid.org/0000-0002-0224-9817}{https://orcid.org/0000-0002-0224-9817}

### 2019-11-08 manuscript.Rmd
add details on googleComputeEngineR features and Cloud Run

@@
 At the same time, doing it by hand prevents from using programming language for what they are good for: iteration, and automation. 
-Iteration and automation for Dockerfiles is the very reason behind [`dockerfiler`](https://github.com/ColinFay/dockerfiler), an R package designed for building Dockerfiles straigth from R.
+Iteration and automation for Dockerfiles is the very reason behind [`dockerfiler`](https://github.com/ColinFay/dockerfiler), an R package designed for building Dockerfiles straight from R.
 
@@
 - `RSelenium`
-- [`googleComputeEngineR`](https://cloudyr.github.io/googleComputeEngineR/) (function `docker_run`)
-  - I don't think `docker_run` is main focus for containers for this package, its more how it uses Docker to launch custom environments that can be templated (e.g.
-rstudio, shiny etc.) and for parrallisation e.g.
-`library(future)`  using GCE VMs as endpoints.
-[@MarkEdmondson1234]
+- [`googleComputeEngineR`](https://cloudyr.github.io/googleComputeEngineR/) (function `gce_vm_template()`)
+  - To enable quick deployments of key R services such as RStudio and Shiny onto cloud virtual machines (VMs), this package utilises Dockerfiles to move the labour of setting up those services from the user to a premade Docker image.
+For example, by specifying the template `template="rstudio"` in `gce_vm_template()/gce_vm()` an up to date RStudio Service image is launched.
+Specifying `template="rstudio-gpu"` will launch an RStudio Server image with a GPU attached, etc.
 - [`analogsea`](https://github.com/sckott/analogsea) (digital ocean R client)
@@
   - For this project, we use travis-ci to build rocker-derived containers, test them, and only push them to docker hub (from travis-ci.org) if the integration tests succeed.
-- Google Cloud Build [@MarkEdmondson1234]
+- Google Cloud Build
   - https://cloud.google.com/cloud-build/
-  - For me this is what makes Docker containers viable, as it builds the Dockerfiles on each GitHub commit.
-It couples with Google Container Registry to build private Docker images, for downstream applications.
+  - Cloud Build is another continuous intergation service that helps move the workload of building the Docker images to an online service.
+Cloud Build can be set up to build the Dockerfiles on each GitHub commit or release.
+This means you do not need to build the Docker images locally, which can tie up resources since Docker images can be several GBs and take a long time to compile.
+Google Cloud Build works alongside Google Container Registry to allow you to build private and public Docker images, which allows you to build up your own dependency graphs for downstream applications.
 
@@
 
-- Docker images for cloud services [@MarkEdmondson1234]
-  - the most popular functionality for `googleComputeEngineR` is its use of Docker to enable parallel processing across VMS - [some demos here](https://cloudyr.github.io/googleComputeEngineR/articles/massive-parallel.html)
-- Google Cloud Run - CaaS (Containers as a Service) that lets you launch a Docker container without worrying about underlying infrastructure.
-An R implementation is shown here at [cloudRunR](https://github.com/MarkEdmondson1234/cloudRunR) which uses it to create a scalable R plumber API.
+- Docker images for cloud services
+  - A popular application of using Docker is to create the environments that can run in parallel for speeding up R code.
+For example, [`googleComputeEngineR`](https://CRAN.R-project.org/package=googleComputeEngineR)'s `gce_vm_cluster()` function can create clusters of 2 or more virtual machines, running multi-CPU architectures.
+Instead of running a local R script with the local CPU and RAM restrictions, the same code can be processed on all CPU threads of the cluster of machines in the cloud, all running in a Docker container with the same R environments.
+This is achieved through `googleComputeEngineR`'s integration with the R parralisation library [`library(future)` by Henrik Bengtsson]( https://CRAN.R-project.org/package=future).
+Local R computation can be thrown up to a multi-CPU and VM environment to achieve parrell computation in a few lines of R code in your local session.
+- [some demonstrations are available here](https://cloudyr.github.io/googleComputeEngineR/articles/massive-parallel.html).
+- Google Cloud Run - This is a CaaS (Containers as a Service) that lets you launch a Docker container without worrying about underlying infrastructure.
+This dispenses with the developer creating a cloud server to run the Docker image on, by abstracting away those servers to a more serverless configuration.
+Cloud Run lets you run your code on top of a managed or your own Kubernetes cluster running Knative, and can accept any Docker image.
+The service takes care of network ingress, scaling machines up and down to zero, authentication and authorisation, all features which are non-trivial for a developer to create on their own.
+This can be used to scale up R code to millions of instances if they need to, with little or no changes to existing code.
+An R implementation is shown here at [cloudRunR](https://github.com/MarkEdmondson1234/cloudRunR) which uses Cloud Run to create a scalable R plumber API.
 - `batchtools` \citep{Lang2017batchtools} can [schedule jobs with Docker Swarm](https://mllg.github.io/batchtools/reference/makeClusterFunctionsDocker.html)

### 2019-11-08 manuscript.Rmd
cloud run link

@@
 - [some demonstrations are available here](https://cloudyr.github.io/googleComputeEngineR/articles/massive-parallel.html).
-- Google Cloud Run - This is a CaaS (Containers as a Service) that lets you launch a Docker container without worrying about underlying infrastructure.
+- [Google Cloud Run](https://cloud.run) - This is a CaaS (Containers as a Service) that lets you launch a Docker container without worrying about underlying infrastructure.
 This dispenses with the developer creating a cloud server to run the Docker image on, by abstracting away those servers to a more serverless configuration.

### 2019-11-08 manuscript.Rmd
Initial work on deploying R to production

--- 2019-11-07 manuscript.Rmd
@@
 - [`analogsea`](https://github.com/sckott/analogsea) (digital ocean R client)
-- [`plumber`](https://www.rplumber.io/docs/hosting.html#docker)
-- [production deployment of neural networks](https://github.com/tmobile/r-tensorflow-api) [@jnolis, @nolistic]
+
+## Using R to power enterprise software in production environments [@jnolis, @nolistic]
+
+R has been historically viewed as a tool for analysis and scientific research--not for creating software that corporations can rely on to continously run in real time.
+However, thanks to advancements in R running as a web service, along with with the ability to deploy R in docker containers, modern enterprises are now capable of having real-time machine learning powered by R.
+
+The backbone of this work is the _[plumber](https://www.rplumber.io/docs/hosting.html#docker)_ package, created by Jeff Allen.
+The plumber package allows R to be used to host web services through RESTful APIs.
+When R is running on a server, a different server can ask R to make a computation through an HTTP request and get an immediate response.
+For instance, if a data scientist at a company creates a machine learning model to tell which customers are likely to make a repeat purchase, if they serve the model through a plumber API then someone else on the marketing team could write software that sends different emails depending on that real-time prediction.
+Plumber syntax is simple to read and understand: the API is created by commenting above function names in an R script.
+
+```
+#* Return the sum of two numbers
+#* @get /random
+function(){
+  runif()
+}
+```
+_Example of a plumber API to generated a random number by visiting `http://127.0.0.1/random`_
+
+Plumber becomes far more powerful with the addition of Docker.
+Docker can be used to create images that [include both R and plumber](https://www.rplumber.io/docs/hosting.html#docker).
+Then when the image is deployed, that creates the web endpoints that other services can hit and call the R code.
+This pattern of deploying code mimics those used by software engineering services created in Java, Python, and other modern languages.
+By using Docker to deploy plumber APIs, R can be used alongside those langages as a first class member of a software engineering technical stack.
+
+At T-Mobile, R is used to power real-time services that are called over a million times a day.
+T-Mobile created a set of neural network machine learning natural language processing models to help customer care agents manage text-based messages for customers.
+The models are convolutional neural networks that use the RStudio Keras package and are built on top of a Rocker Docker image.
+Since the models power tools for agents and customers, they need to have extremely high uptime and reliability--and they found that R is able to perform.
+The AI @ T-Mobile team open sourced [their docker images](https://github.com/tmobile/r-tensorflow-api  which are available on GitHub.
 

### 2019-11-09 manuscript.Rmd
add kubernetes paragraph

--- 2019-11-08 manuscript.Rmd
@@
 - [production deployment of neural networks](https://github.com/tmobile/r-tensorflow-api) [@jnolis, @nolistic]
+- Kubernetes - A popular platform for managing Docker containers is [Kubernetes](https://kubernetes.io/), which is used for a wide variety of containerized applications.
+It may be your organisation has a Kubernetes cluster already for other applications.
+Docker containers are used within Kubernetes clusters to hold native code, for which Kubernetes creates a framework around network connections and scaling of resources up and down.
+An introduction on using [Kubernetes with R is published at this blog post](https://code.markedmondson.me/r-on-kubernetes-serverless-shiny-r-apis-and-scheduled-scripts/).
 

### 2019-11-28 manuscript.Rmd
add azure+kubernetes section

--- 2019-11-08 manuscript.Rmd
@@
 **Microsoft R Open** ([MRO](https://github.com/nuest/mro-docker)) is an R distribution formerly known as Revolution R Open (RRO) before Revolution Analytics was acquired by Microsoft.
-MRO is compatible with main R and it's packages.sss
+MRO is compatible with main R and its packages.
 "It includes additional capabilities for improved performance, reproducibility, and platform support." \citep{microsoft_mro}
@@
 - [`analogsea`](https://github.com/sckott/analogsea) (digital ocean R client)
+- [`AzureContainers`](https://cran.r-project.org/package=AzureContainers): Umbrella package for working with containers in Microsoft's Azure cloud.
+Provides interfaces to three Azure services: Container Registry, Container Instances (for running individual containers), and Kubernetes Services (for orchestrated deployments)
 
@@
 The AI @ T-Mobile team open sourced [their docker images](https://github.com/tmobile/r-tensorflow-api  which are available on GitHub.
+
+## Deployment to the cloud [@hongooi]
+
+While Plumber provides the low-level infrastructure for turning a statistical model into a predictive service, for production purposes it is usually necessary to take into account considerations such as scalability, reliability and ease of management.
+A single container is limited in the volume of requests it can service.
+Furthermore, if the container goes down, or is occupied with processing a computationally intensive task, the service becomes unavailable until it is restarted.
+
+These issues can be mitigated by deploying not a single container, but a _cluster_ of containers, orchestrated as a single deployment.
+The deployment system of choice at the time of writing is [Kubernetes](https://kubernetes.io).
+This allows you to deploy a single image to a pool of servers, which can then serve requests in parallel.
+Client-side requests are made to the cluster ingress endpoint, which then redircts them to individual servers.
+The orchestrator can identify nodes that have failed, and restart them automatically; it can also detect when there is no capacity to service more requests, and start additional containers as needed.
+
+While these load-balancing and autoscaling properties of a Kubernetes cluster are invaluable, they are still limited by the hardware resources that it has access to.
+A cluster cannot scale out if it has run out of servers.
+This issue in turn can be addressed by deploying not to on-premise servers but to the cloud, where effectively unlimited computing resources are available (as long as one can pay for them).
+From a management perspective, cloud deployments have the additional advantage that hardware maintenance becomes the responsibility of the cloud provider, eliminating another pain point.
+Orchestrator services are available from all the major cloud service providers, including Amazon Web Services (AWS), Google Cloud Platform, and Microsoft Azure.
+
+An example of an R package for working with containers in the cloud is AzureContainers, which interacts with three Azure services: Container Instances (for running a single container), Container Registry (a private docker registry service), and Kubernetes Service (for Kubernetes on Azure).
+AzureContainers provides a powerful interface that builds on the framework provided by AzureRMR, which is an overarching package for working with Resource Manager, the framwork for deploying and managing arbitrary resource in Azure.
+On the client side, it provides simple shells to the Docker, Kubectl and Helm tools that are commonly used for managing containers and Kubernetes clusters.
+
+Here is some simple code that pushes a Docker image to an Azure Container Registry, and then creates a deployment using Azure Kubernetes Service.
+This is a slightly modified version of the code from the "Deploying a prediction service with Plumber" AzureContainers vignette.
+The code makes use of a yaml file to define the deployment and predictive service, which is standard practice with Kubernetes; the yaml can be seen in the aforementioned vignette.
+
+```r
+library(AzureContainers)
+# create a resource group for our deployments
+deployresgrp <- AzureRMR::get_azure_login()$
+    get_subscription("subscription_id")$
+    create_resource_group("deployresgrp", location="australiaeast")
+
+### create a container registry
+deployreg_svc <- deployresgrp$create_acr("deployreg")
+
+# build our image (a random forest on the Boston housing data, available in the MASS package)
+call_docker("build -t bos-rf .")
+
+# upload the image to Azure
+deployreg <- deployreg_svc$get_docker_registry()
+deployreg$push("bos-rf")
+
+### create a Kubernetes cluster with 2 nodes
+deployclus_svc <- deployresgrp$create_aks("deployclus", agent_pools=aks_pools("pool1", 2))
+
+# grant the cluster pull access to the registry
+gr <- AzureGraph::get_graph_login()
+aks_app_id <- deployclus$properties$servicePrincipalProfile$clientID
+reg$add_role_assignment(gr$get_app(aks_app_id), "Acrpull")
+
+# get the cluster endpoint
+deployclus <- deployclus_svc$get_cluster()
+
+# create and start the service
+deployclus$create("bos-rf.yaml")
+```
+
+Once the service has been created, we can check on its status and obtain predictions:
+
+```r
+deployclus$get("deployment bos-rf")
+#> Kubernetes operation: get deployment bos-rf  --kubeconfig=".../kubeconfigxxxx"
+#> NAME      DESIRED   CURRENT   UP-TO-DATE   AVAILABLE   AGE
+#> bos-rf    1         1         1            1           5m
+
+deployclus$get("service bos-rf-svc")
+#> Kubernetes operation: get service bos-rf-svc  --kubeconfig=".../kubeconfigxxxx"
+#> NAME         TYPE           CLUSTER-IP   EXTERNAL-IP     PORT(S)          AGE
+#> bos-rf-svc   LoadBalancer   10.0.8.189   xxx.xxx.xxx.xxx 8000:32276/TCP   5m
+
+response <- httr::POST("http://xxx.xxx.xxx.xxx:8000/score",
+    body=list(df=MASS::Boston[1:10,]), encode="json")
+httr::content(response, simplifyVector=TRUE)
+#> [1] 25.9269 22.0636 34.1876 33.7737 34.8081 27.6394 21.8007 22.3577 16.7812 18.9785
+```
+
+Note that Plumber, by itself, provides no authentication or security features; anybody who knows the cluster's IP address can access the predictive service.
+At minimum, a basic authentication layer should be added to any Kubernetes deployment.
+Standard Kubernetes procedure involves creating an ingress controller, for example with the nginx or traefik reverse proxy software, and one can then add application-specific authentication layers on top of that (for example, to authenticate with a user's Azure Active Directory or LDAP credentials).
+More details can be found from most Kubernetes learning resources, and are outside the scope of this paper.
 

### 2019-11-28 manuscript.Rmd
some cleanup

@@
 This allows you to deploy a single image to a pool of servers, which can then serve requests in parallel.
-Client-side requests are made to the cluster ingress endpoint, which then redircts them to individual servers.
-The orchestrator can identify nodes that have failed, and restart them automatically; it can also detect when there is no capacity to service more requests, and start additional containers as needed.
+Client-side requests are made to the cluster ingress endpoint, which then redirects them to individual servers.
+The orchestrator can identify cluster nodes that have failed, and restart them automatically; it can also detect when there is no capacity to service more requests, and start additional servers and containers as needed.
 
@@
 An example of an R package for working with containers in the cloud is AzureContainers, which interacts with three Azure services: Container Instances (for running a single container), Container Registry (a private docker registry service), and Kubernetes Service (for Kubernetes on Azure).
-AzureContainers provides a powerful interface that builds on the framework provided by AzureRMR, which is an overarching package for working with Resource Manager, the framwork for deploying and managing arbitrary resource in Azure.
+AzureContainers provides a lightweight yet powerful interface for working with Resource Manager, the framework for deploying and managing arbitrary resources in Azure.
 On the client side, it provides simple shells to the Docker, Kubectl and Helm tools that are commonly used for managing containers and Kubernetes clusters.

### 2019-11-28 manuscript.Rmd
add me

@@
     email: nitesh.turaga@roswellpark.org
+  - name: Hong Ooi
+    affiliation: Microsoft
+    address:
+    - Level 5, 4 Freshwater Place
+    - Southbank, VIC 3006, Australia
+    email: hongooi@microsoft.com
 abstract: >

### 2019-11-30 manuscript.Rmd
some more edits

--- 2019-11-28 manuscript.Rmd
@@
   - name: Daniel Nüst
-    affiliation: University of Münster
+    affiliation: Institute for Geoinformatics, University of Münster
     address:
-      - Institute for Geoinformatics
     - Heisenbergstr. 2
-      - 48149 Münster, Germany
-      - \href{https://orcid.org/0000-0002-0024-5046}{\includegraphics[height=\fontcharht\font`\0]{ORCID-iD_icon-vector.pdf} https://orcid.org/0000-0002-0024-5046}
+    - 48149 Münster
+    - Germany
+    - \orcid{0000-0002-0024-5046}
     email: daniel.nuest@uni-muenster.de
@@
       - Illini Hall, 725 S Wright St
-      - Champaign, IL 61820, USA
-      - \includegraphics[height=\fontcharht\font`\0]{ORCID-iD_icon-vector.pdf} \href{https://orcid.org/0000-0001-6419-907X}{https://orcid.org/0000-0001-6419-907X}
+      - Champaign, IL 61820
+      - \orcid{0000-0001-6419-907X}
     email: dirkd@eddelbuettel.com
@@
     affiliation: ThinkR
-    address:
-      - 5O rue Arthur Rimbaud
-      - 93300 Aubervilliers, France
-      - \includegraphics[height=\fontcharht\font`\0]{ORCID-iD_icon-vector.pdf} \href{https://orcid.org/0000-0001-7343-1846}{https://orcid.org/0000-0001-7343-1846}
     email: contact@colinfay.me
@@
   - name: Karthik Ram
-    affiliation: Berkeley Institute for Data Science
-    address:
-      - University of California
-      - Berkeley, CA 94720, USA
-      - \includegraphics[height=\fontcharht\font`\0]{ORCID-iD_icon-vector.pdf} \href{https://orcid.org/0000-0002-0233-1757}{https://orcid.org/0000-0002-0233-1757}
-    email: karthik.ram@berkeley.edu
+    #affiliation: -
+    #email: -
   - name: Noam Ross
-    affiliation: EcoHealth Alliance
-    address:
-      - 460 W 34th St., Ste. 1701
-      - New York, NY 10001, USA
-      - \includegraphics[height=\fontcharht\font`\0]{ORCID-iD_icon-vector.pdf} \href{https://orcid.org/0000-0002-0233-1757}{https://orcid.org/0000-0002-0233-1757}
-    email: ross@ecohealthalliance.org
+    #affiliation: -
+    #email: -
   - name: Nan Xiao
@@
       - Charlestown, MA 02129, USA
-      - \includegraphics[height=\fontcharht\font`\0]{ORCID-iD_icon-vector.pdf} \href{https://orcid.org/0000-0002-0250-5673}{https://orcid.org/0000-0002-0250-5673}
+      - \orcid{0000-0002-0250-5673}
     email: me@nanx.me
@@
       - Elm & Carlton Streets
-      - Buffalo, New York, 14263, USA
-      - \includegraphics[height=\fontcharht\font`\0]{ORCID-iD_icon-vector.pdf} \href{https://orcid.org/0000-0002-5910-4010}{https://orcid.org/0000-0002-5910-4010}
+      - Buffalo, New York, 14263 USA
+      - \orcid{0000-0002-5910-4010}
     email: lori.shepherd@roswellpark.org
@@
     - Elm & Carlton Streets
-    - Buffalo, New York, 14263, USA
-    - \includegraphics[height=\fontcharht\font`\0]{ORCID-iD_icon-vector.pdf} \href{https://orcid.org/0000-0002-0224-9817}{https://orcid.org/0000-0002-0224-9817}
+    - Buffalo, New York, 14263 USA
+    - \orcid{0000-0002-0224-9817}
     email: nitesh.turaga@roswellpark.org
@@
 
-The R community keeps growing: the number of new packages on CRAN keeps on rising, meetups, conferences, online courses and unconferences prosper, and the uptake in education and industry increases [REFs].
-All this cements the role of R as the lingua franca of statistics, data visualisation, and computational research.
-Coinciding with the rise of R was the advent of Docker [CITATION?] as a general tool for development, distribution, and deployment of software.
-Combining both these topics, the _Rocker Project_ ([https://www.rocker-project.org/](https://www.rocker-project.org/)) was introduced in 2014 and provides a number of [Docker](https://en.wikipedia.org/wiki/Docker_(software)) images for various use cases as described in \citet{RJ-2017-065}.
-The considerable uptake and continued evolution of the Rocker suite of containers has lead to numerous projects extending or building upon Rocker images ranging from reproducible research to production deployments.
-This article presents this _Rockerverse_ of projects.
-It also introduces related activities connecting the R language and environment with other containerisation solutions.
-The main contributions is a coherent picture of the current lay of the land of containers in, with, and for `R`.
-This diversity includes demonstrators, early prototypes, and mature projects.
+The R community keeps growing.
+This can be seen in the number of new packages on CRAN which keeps on rising, but also in the increased number of conferences, online courses, meetups, and unconferences, as well as in the uptake in education and industry increases [REFs].
+All this cements the role of R as the _lingua franca_ of statistics, data visualisation, and computational research.
+Coinciding with the rise of R was the advent of [Docker](https://en.wikipedia.org/wiki/Docker_(software)) as a general tool for distribution and deployment of software.
+Combining both these topics, the Rocker Project ([https://www.rocker-project.org/](https://www.rocker-project.org/)) provides images with R (see the next Section for more details).
+The considerable uptake and continued evolution of the Rocker Project has lead to numerous projects extending or building upon Rocker images, ranging from reproducible research to production deployments.
+This article presents this _Rockerverse_ of projects across all development stages: demonstrations, prototypes, and mature products.
+We also introduce related activities connecting the R language and environment with other containerisation solutions.
+The main contributions is a coherent picture of the current lay of the land of using containers in, with, and for R.
 
-# Containerization and Rocker [NN]
+# Containerization and Rocker
 
-<!--_Do we need a generic intro here?_ DE: Yes, and happy to take a stab.-->
+Docker, an application and service provide by the eponymous company, has in just a few short years risen to prominence for development, testing, deployment and distribution of computer software \citep[cf.][]{datadog_8_2018,munoz_history_2019}.
+While there are related approaches such as LXC\footnote{\href{https://en.wikipedia.org/wiki/LXC}{https://en.wikipedia.org/wiki/LXC}} or Singularity \citep{kurtzer_singularity_2017}, Docker has become synomymous with "containerization"---the method of taking software artefacts and bundling them in such a way that use becomes standardized and portable across operationg systems.
+In doing so, Docker had recognised and validated the importance of one very important thread that had been emerging, namely virtualization.
+By allowing (one or possibly) multiple applications or services to run concurrently on one host machine without any fear of interference between them, an important scalability opportunity is being provided.
+But Docker improved on virtualization by accessing the host system---generally Linux---through a much thinner and smaller shim than a full operating system emulation or virtualization.
+This makes for more efficient use of system resources and allowed another order of magnitude in terms of scalability of deployment [REFERENCE!].   
+While Docker makes use of Linux kernel features, its importance was large enough so that some required aspects of running Docker have been added to other operating systems to support Docker there more efficiently too \citep{microsoft_linux_2019}.
+The success even lead to standardisation and industry collaboration \citep{oci_open_2019}.
 
-Docker, an application and service provide by the eponymous company, has in just a few short years risen to prominence for development, testing, deployment and distribution of computer software.
-While there are related approaches such as LXC or Singularity, Docker has become synomymous with "containerization"---the method of taking software artefacts and bundling them in such a way that use becomes standardized and portable across operationg systems.
-In doing so, Docker had recognised and validated the importance of one very important thread that had been emerging, namely virtualization.
-By allowing multiple concurrent applications or services to run concurrently on one host machine without any fear of interference between them, an important scalability feature had been realized.
-But Docker improved on virtualization by accessing the host system---generally Linux---through a much thinner and smaller shim than a full operating system emulation or virtualization.
-This makes for more efficient use of system resources and allowed another order of magnitude in terms of scalability of deployment.
-[Reference?]   While Docker makes use of Linux kernel features, its importance was large enough so that some required aspects of running Docker have been added to other operating systems to support Docker there more efficiently too.
-[Win10; reference?]
-
-The key accomplishment of Docker is to make a "bundled" aggregation of software available to any system running Docker, without requiring much else from the host.
-This is a rather attractive, and novel, proposition which has lead to widespread adoption and use of Docker in a variety of domains.
-It provided also a natural match for "cloud deployment" which runs (or at least appears to run) "seamlessly" without much explicit reference to the underlying machine, architecture or operating system.
-Using containers on cloud hosted servers is a natural fit also because these containers can be deployed with very littler in terms of dependencies on the host system.
+The key accomplishment of Docker (as an "application") is to make a "bundled" aggregation of software (the so-called "image") available to any system equipped to run Docker, without requiring much else from the host besides the actual Docker application installation.
+This is a rather attractive, and novel, proposition which has lead to widespread adoption and use of Docker in a variety of domains, _e.g._, cloud computing infrastructure \citep{Bernstein2014}, data science \citep{boettiger_introduction_2015}, and edge computing \citep{alam_orchestration_2018} [more examples!].
+It provided to be a natural match for "cloud deployment" which runs (or at least appears to run) "seamlessly" without much explicit reference to the underlying machine, architecture or operating system: containers can be deployed with very little in terms of dependencies on the host system---only the container runtime is required.
 
 For statistical computing and analysis centered around R, the Rocker Project has provided a variety of Docker containers since its start in 2014 \citep{RJ-2017-065}.
-The Rocker Project provides several lines of containeres spanning to from building blocks with R-release or R-devel to containers with RStudio Server and Shiny Server.
-Also of note is a series of "versioned" containers which match the R release they contain with the _then-current_ set of packages via the MRAN Snapshot views of CRAN.
+The Rocker Project provides several lines of containers spanning to from building blocks with `R-release` or `R-devel`, via containers with RStudio~Server and Shiny~Server, to domain-specific containers such as `geospatial`.
+Also of note is a series of "versioned" containers which match the R release they contain with the _then-current_ set of packages via the MRAN Snapshot views of CRAN \citep{microsoft_cran_2019}.
 [Maybe at a later point add forward reference to other topics building on Rocker described here.]
@@
 
-**Microsoft R Open** ([MRO](https://github.com/nuest/mro-docker)) is an R distribution formerly known as Revolution R Open (RRO) before Revolution Analytics was acquired by Microsoft.
-MRO is compatible with main R and its packages.
+_Microsoft R Open_ ([MRO](https://github.com/nuest/mro-docker)) is an R distribution formerly known as Revolution R Open (RRO) before Revolution Analytics was acquired by Microsoft.
+MRO is compatible with main R and it's packages.sss
 "It includes additional capabilities for improved performance, reproducibility, and platform support." \citep{microsoft_mro}
-Most notably these capabilities are the MRAN repository a.k.a.
-CRAN Time Machine, which is enabled by default, and the (optional) integration with [Intel® Math Kernel Library](https://software.intel.com/en-us/mkl) (MKL) for multi-threading in linear algebra operations \citep{microsoft_multithread}.
-<!-- MRAN is also used by versioned Rocker images. -->
+Most notably these capabilities are the MRAN repository, which is enabled by default, and the (optional) integration with [Intel® Math Kernel Library](https://software.intel.com/en-us/mkl) (MKL) for multi-threading in linear algebra operations \citep{microsoft_multithread}.
 MRO does not provide official Docker images, but a set of community-maintained `Dockerfile`s and Docker images `nuest/mro` are available on Docker Hub at [https://github.com/nuest/mro-docker](https://github.com/nuest/mro-docker) and on GitHub at [https://github.com/nuest/mro-docker](https://github.com/nuest/mro-docker) respectively.
@@
 
-**Renjin** is an interpreter for the R language running on top of the [Java Virtual Machine](https://en.wikipedia.org/wiki/Java_virtual_machine) (JVM) providing full two-way access between Java and R code \citep{wikipedia_renjin_2018}.
+_Renjin_ is an interpreter for the R language running on top of the [Java Virtual Machine](https://en.wikipedia.org/wiki/Java_virtual_machine) (JVM) providing full two-way access between Java and R code \citep{wikipedia_renjin_2018}.
 It was developed to combine the benefits of R, such scripting and extension packages, with the JVM's advantages in the areas of security, cross-platform availability, and established position in enterprise settings.
@@
 `rgdal` \footnote{\href{http://packages.renjin.org/package/org.renjin.cran/rgdal/1.4-4/build/1}{http://packages.renjin.org/package/org.renjin.cran/rgdal/1.4-4/build/1}}.
+
 There are no offical Docker images for Renjin, but community-maintained images for selected releases only are available under `nuest/renjin` on Docker Hub and GitHub at [https://hub.docker.com/r/nuest/renjin](https://hub.docker.com/r/nuest/renjin) and [https://github.com/nuest/renjin-docker](https://github.com/nuest/renjin-docker) respectively.
@@
 
-**[pqR](http://www.pqr-project.org/)** is a "a pretty quick version of R".
+[_pqR_](http://www.pqr-project.org/) is a "a pretty quick version of R".
 _pqR_ fixes some opinionated issues in the R language and is the basis for experimental features, e.g.
@@
 
-**FastR** is "A high-performance implementation of the R programming language, built on GraalVM" (https://github.com/oracle/fastr).
+_FastR_ is "A high-performance implementation of the R programming language, built on GraalVM" (https://github.com/oracle/fastr).
 It is developed by Oracle, connects R to the GraalVM ecosystem \citep{wikipedia_graalvm_2019}, and also claims superior performance but also targets full compatibility with base R \footnote{ \href{https://github.com/oracle/fastr}{https://github.com/oracle/fastr}}.
-There are no offical Docker images provided, but `nuest/fastr` images and `Dockerfile` are independently maintained on [Docker Hub](https://cloud.docker.com/repository/docker/nuest/fastr/) and [GitHub](https://github.com/nuest/fastr-docker) respectively.
 
-While the images presented in this section are far from being as vetted, stable, and widely used as any of the Rocker images, they demonstrate an important advantage of containerisation technology, namely the abiliy to transparently build portable stacks of open source software and make them easily accessible to users.
+There are no offical Docker images provided, but `nuest/fastr-docker` provides image and `Dockerfile` on Docker Hub and GitHub at []() and [https://github.com/nuest/fastr-docker](https://github.com/nuest/fastr-docker) respectively.
+
+While the images presented in this section are far from being as vetted, stable, and widely used as any of the _Rocker_ images, they demonstrate an important advantage of containerisation technology, namely the abiliy to transparently build portable stacks of open source software and make them easily accessible to users.
 All different distributions are published under GPL licenses.
@@
 
-[_Bioconductor_](http://bioconductor.org/) is an open source, open development project for the analysis and comprehension of genomic data \citep{gentleman_bioconductor_2004}.
+[_Bioconductor_](http://bioconductor.org/help/docker/) is an open source, open development project for the analysis and comprehension of genomic data \citep{gentleman_bioconductor_2004}.
 The project consists of 1741 R software packages as of August 15th 2019, as well as packages containing annotation or experiment data.
@@
 Docker images allow availability of current and past versions of _Bioconductor_ for convenience and reproducibility.
+
 _Bioconductor_ 'base' docker images are built on top of `rocker/r-ver` and `rocker/rstudio`.
 _Bioconductor_ installs packages based on the R version, and therefore uses `rocker/rstudio` and `rocker/r-ver` version tagging.
-_Bioconductor_ selects the desired version of R from Rocker, adds the BiocManager CRAN package for installing appropriate versions of _Bioconductor_ packages, and creates a _Bioconductor_ docker image with an informative tag (R\_version\_Bioc\_version).
+_Bioconductor_ selects the desired version of R from _Rocker_, adds the BiocManager CRAN package for installing appropriate versions of _Bioconductor_ packages, and creates a _Bioconductor_ docker image with an informative tag (R\_version\_Bioc\_version).
 The images are summarized on the _Bioconductor_ web site (https://bioconductor.org/help/docker/), maintained on GitHub ([https://github.com/Bioconductor/bioc_docker](https://github.com/Bioconductor/bioc_docker)), and available to the community through [DockerHub](https://hub.docker.com/).
@@
 _Bioconductor_ images for proteomics, metabolomics, and flow cytometry are community maintained.
-All community maintained images build on top of the _Bioconductor_ base image and therefore indirectly the Rocker images.
+All community maintained images build on top of the _Bioconductor_ base image and therefore indirectly the _Rocker_ images.
 To simplify building and maintaining _Bioconductor_ images, we use a Ruby templating engine.
-A recent audit of the _Bioconductor_ Dockerfiles, following best practices from the Docker website, led to a reduction in the size and number of layers.
+
+A recent audit of our Dockerfiles, following best practices from the Docker website, led to a reduction in the size and number of layers in images we produce.
 The most important insights involve the ['union' file system used by Docker](https://docs.docker.com/storage/storagedriver/overlayfs-driver/#how-container-reads-and-writes-work-with-overlay-or-overlay2).
@@
 
-As with any other software, each new version of R comes with its share of changes. 
-Some are breaking changes, some are not, but the fact is that running a piece of code in a given version might give different s from another version ; which, in some case, can be an issue when it comes to reproducibility.
-For example, think about R random seed: starting with `R 3.6.0`, running `set.seed(2811); sample(1:1000, 2)` will not give the same result as if it was run inside an older version of R.
-That might seem trivial, but all the code using the random number generator will not be exactly reproducible after this breaking change. 
-Detecting this kind of behavior is the reason why the [r-online](https://github.com/ColinFay/ronline) app has been developped: available as a standalone NodeJS app or [online](https://srv.colinfay.me/r-online), this app was designed for historic exploration of R: with it, the user can compare a piece of code run in two separate versions of R.
-Internally, this app takes the piece of code written by the user, opens one or two docker instances with the given version of R, and returns the result to the user.
-
-Containers are perfectly suited to capture a specific configuration of a computing environment to prohibit such problems.
-The Rocker images provides the _versioned stack_ for different the version of R since the project inception, which use the _then-current_ stable Debian base image \citep[cf.][]{RJ-2017-065}.
-Based on custom build phase hooks\footnote{\href{https://docs.docker.com/docker-hub/builds/advanced/}{https://docs.docker.com/docker-hub/builds/advanced/}}, i.e.
-small shell scripts executed at different phases in the automated build of Rocker images, there are also semantic version tags for the most recent freezed (i.e.
-using MRAN) versions with only the major and minor version\footnote{See \href{https://github.com/rocker-org/rocker-versioned/blob/master/VERSIONS.md}{https://github.com/rocker-org/rocker-versioned/blob/master/VERSIONS.md}, FIXME: \href{https://github.com/rocker-org/rocker-versioned/issues/42}{https://github.com/rocker-org/rocker-versioned/issues/42}.}.
-So, at the time of writing this article, `rocker/r-ver:3` and `3.6` are aliases for `3.6.0`, because `3.6.1` is the latest release.
-With the release of `3.6.2` and pinning of the MRAN version in `3.6.1`, the tags `3` and `3.6` will deliver the same image as `3.6.1`.
-These tags allow users to update the base image to retrieve bugfixes while reducing the risk of also introducing breaking changes.
+- run code with Versions of R: https://srv.colinfay.me/r-online
+- old R versions: https://github.com/rocker-org/rocker-versioned/issues/138
+- [semantic version tags for Rocker](https://github.com/rocker-org/rocker-versioned/#version-tags)
+- debugging memory problems across multiple versions: https://github.com/wch/r-debug
+- cross-R testing: Show how you can run the same script in multiple R variants in containers (new contribution).
+Ideally the images should be created using `containerit` and controlled with `stevedore`/`docker`, i.e.
+via an R script.
 
@@
 # R Packages
-<!-- from packages closer to Docker to second level packages to more complex ones -->
 
-## Clients
+## Packages for automation and packaging [@nuest]
 
-A number first order R packages exist that talk directly to Docker.
-They can be used to control the Docker daemon, e.g. to build images or run containers, from R without explicit knowledge of the Docker command line interface.
-
-- [`stevedore`](https://github.com/richfitz/stevedore)
-- [`docker`](https://bhaskarvk.github.io/docker/)
-- [`harbor`](https://github.com/wch/harbor/)
-- [`dockermachine`](https://github.com/cboettig/dockermachine)
-- [`dockyard`](https://github.com/thebioengineer/dockyard)
-
-## Capture and create environments [@nuest]
-
-Several second order R packages attempt to make the process of creating Docker images and using containers for specific tasks, such as running tests or rendering reproducible reports, easier.
-
-- [`dockerfiler`](https://github.com/ColinFay/dockerfiler/)
 - [`containerit`](https://github.com/o2r-project/containerit/)
@@
 
-**[_liftr_](https://nanx.me/liftr/)** \citep{liftr2019} aims to solve the problem of persistent reproducible reporting in statistical computing.
+### liftr
+
+[_liftr_](https://nanx.me/liftr/) \citep{liftr2019} aims to solve the problem of persistent reproducible reporting in statistical computing.
 Currently, the R Markdown format and its backend compilation engine _knitr_ offer a _de facto_ standard for creating dynamic documents \citep{xie2018}.
@@
 - [`rize` for Shiny](https://github.com/cole-brokamp/rize)
+- [`dockerfiler`](https://github.com/ColinFay/dockerfiler)
 
-One of the key tool of Docker are dockerfiles, which can be thought of as "recipes" used to build a specific image. 
-But building these files can feel like a workflow break as it demands to open a separate file, and / or to write a small wrapper around R's `write()` function to add elements to the file. 
-At the same time, doing it by hand prevents from using programming language for what they are good for: iteration, and automation. 
-Iteration and automation for Dockerfiles is the very reason behind [`dockerfiler`](https://github.com/ColinFay/dockerfiler), an R package designed for building Dockerfiles straigth from R.
+## Packages for control and provisioning [@nuest, @richfitz, @rcannood]
 
-For example, the **[`golem`](https://github.com/ThinkR-open/golem)** package makes an heavy use of `dockerfiler` when it comes to creating the `Dockerfile` for building production-grade Shiny applications and deploying them.
-The reason behind this is that `dockerfiler`, on top of being scriptable from R, can leverage all the tools available in R to parse a `DESCRIPTION` file, to get system requirements, to list dependencies, versions, etc.
-
-# Use cases and applications
-
-## Deployment to cloud services [div]
-
-The cloud is the natural environment of containers, and becomes the go-to mechanism to expose applications written in R to users.
-
+- [`stevedore`](https://github.com/richfitz/stevedore)
 - [`babelwhale`](https://cran.r-project.org/web/packages/babelwhale/index.html): Running a Docker from R with Singularity or Docker as back-end.
@@
 [@rcannood]
+- [`docker`](https://bhaskarvk.github.io/docker/)
 - `RSelenium`
@@
 [@MarkEdmondson1234]
-- [`analogsea`](https://github.com/sckott/analogsea) (digital ocean R client)
-- [`AzureContainers`](https://cran.r-project.org/package=AzureContainers): Umbrella package for working with containers in Microsoft's Azure cloud.
-Provides interfaces to three Azure services: Container Registry, Container Instances (for running individual containers), and Kubernetes Services (for orchestrated deployments)
+- [`analogsea`](https://github.com/sckott/analogsea)
+- [`harbor`](https://github.com/wch/harbor/)
+- [`dockermachine`](https://github.com/cboettig/dockermachine)
 
-## Using R to power enterprise software in production environments [@jnolis, @nolistic]
-
-R has been historically viewed as a tool for analysis and scientific research--not for creating software that corporations can rely on to continously run in real time.
-However, thanks to advancements in R running as a web service, along with with the ability to deploy R in docker containers, modern enterprises are now capable of having real-time machine learning powered by R.
-
-The backbone of this work is the _[plumber](https://www.rplumber.io/docs/hosting.html#docker)_ package, created by Jeff Allen.
-The plumber package allows R to be used to host web services through RESTful APIs.
-When R is running on a server, a different server can ask R to make a computation through an HTTP request and get an immediate response.
-For instance, if a data scientist at a company creates a machine learning model to tell which customers are likely to make a repeat purchase, if they serve the model through a plumber API then someone else on the marketing team could write software that sends different emails depending on that real-time prediction.
-Plumber syntax is simple to read and understand: the API is created by commenting above function names in an R script.
-
-```
-#* Return the sum of two numbers
-#* @get /random
-function(){
-  runif()
-}
-```
-_Example of a plumber API to generated a random number by visiting `http://127.0.0.1/random`_
-
-Plumber becomes far more powerful with the addition of Docker.
-Docker can be used to create images that [include both R and plumber](https://www.rplumber.io/docs/hosting.html#docker).
-Then when the image is deployed, that creates the web endpoints that other services can hit and call the R code.
-This pattern of deploying code mimics those used by software engineering services created in Java, Python, and other modern languages.
-By using Docker to deploy plumber APIs, R can be used alongside those langages as a first class member of a software engineering technical stack.
-
-At T-Mobile, R is used to power real-time services that are called over a million times a day.
-T-Mobile created a set of neural network machine learning natural language processing models to help customer care agents manage text-based messages for customers.
-The models are convolutional neural networks that use the RStudio Keras package and are built on top of a Rocker Docker image.
-Since the models power tools for agents and customers, they need to have extremely high uptime and reliability--and they found that R is able to perform.
-The AI @ T-Mobile team open sourced [their docker images](https://github.com/tmobile/r-tensorflow-api  which are available on GitHub.
-
-## Deployment to the cloud [@hongooi]
-
-While Plumber provides the low-level infrastructure for turning a statistical model into a predictive service, for production purposes it is usually necessary to take into account considerations such as scalability, reliability and ease of management.
-A single container is limited in the volume of requests it can service.
-Furthermore, if the container goes down, or is occupied with processing a computationally intensive task, the service becomes unavailable until it is restarted.
-
-These issues can be mitigated by deploying not a single container, but a _cluster_ of containers, orchestrated as a single deployment.
-The deployment system of choice at the time of writing is [Kubernetes](https://kubernetes.io).
-This allows you to deploy a single image to a pool of servers, which can then serve requests in parallel.
-Client-side requests are made to the cluster ingress endpoint, which then redirects them to individual servers.
-The orchestrator can identify cluster nodes that have failed, and restart them automatically; it can also detect when there is no capacity to service more requests, and start additional servers and containers as needed.
-
-While these load-balancing and autoscaling properties of a Kubernetes cluster are invaluable, they are still limited by the hardware resources that it has access to.
-A cluster cannot scale out if it has run out of servers.
-This issue in turn can be addressed by deploying not to on-premise servers but to the cloud, where effectively unlimited computing resources are available (as long as one can pay for them).
-From a management perspective, cloud deployments have the additional advantage that hardware maintenance becomes the responsibility of the cloud provider, eliminating another pain point.
-Orchestrator services are available from all the major cloud service providers, including Amazon Web Services (AWS), Google Cloud Platform, and Microsoft Azure.
-
-An example of an R package for working with containers in the cloud is AzureContainers, which interacts with three Azure services: Container Instances (for running a single container), Container Registry (a private docker registry service), and Kubernetes Service (for Kubernetes on Azure).
-AzureContainers provides a lightweight yet powerful interface for working with Resource Manager, the framework for deploying and managing arbitrary resources in Azure.
-On the client side, it provides simple shells to the Docker, Kubectl and Helm tools that are commonly used for managing containers and Kubernetes clusters.
-
-Here is some simple code that pushes a Docker image to an Azure Container Registry, and then creates a deployment using Azure Kubernetes Service.
-This is a slightly modified version of the code from the "Deploying a prediction service with Plumber" AzureContainers vignette.
-The code makes use of a yaml file to define the deployment and predictive service, which is standard practice with Kubernetes; the yaml can be seen in the aforementioned vignette.
-
-```r
-library(AzureContainers)
-# create a resource group for our deployments
-deployresgrp <- AzureRMR::get_azure_login()$
-    get_subscription("subscription_id")$
-    create_resource_group("deployresgrp", location="australiaeast")
-
-### create a container registry
-deployreg_svc <- deployresgrp$create_acr("deployreg")
-
-# build our image (a random forest on the Boston housing data, available in the MASS package)
-call_docker("build -t bos-rf .")
-
-# upload the image to Azure
-deployreg <- deployreg_svc$get_docker_registry()
-deployreg$push("bos-rf")
-
-### create a Kubernetes cluster with 2 nodes
-deployclus_svc <- deployresgrp$create_aks("deployclus", agent_pools=aks_pools("pool1", 2))
-
-# grant the cluster pull access to the registry
-gr <- AzureGraph::get_graph_login()
-aks_app_id <- deployclus$properties$servicePrincipalProfile$clientID
-reg$add_role_assignment(gr$get_app(aks_app_id), "Acrpull")
-
-# get the cluster endpoint
-deployclus <- deployclus_svc$get_cluster()
-
-# create and start the service
-deployclus$create("bos-rf.yaml")
-```
-
-Once the service has been created, we can check on its status and obtain predictions:
-
-```r
-deployclus$get("deployment bos-rf")
-#> Kubernetes operation: get deployment bos-rf  --kubeconfig=".../kubeconfigxxxx"
-#> NAME      DESIRED   CURRENT   UP-TO-DATE   AVAILABLE   AGE
-#> bos-rf    1         1         1            1           5m
-
-deployclus$get("service bos-rf-svc")
-#> Kubernetes operation: get service bos-rf-svc  --kubeconfig=".../kubeconfigxxxx"
-#> NAME         TYPE           CLUSTER-IP   EXTERNAL-IP     PORT(S)          AGE
-#> bos-rf-svc   LoadBalancer   10.0.8.189   xxx.xxx.xxx.xxx 8000:32276/TCP   5m
-
-response <- httr::POST("http://xxx.xxx.xxx.xxx:8000/score",
-    body=list(df=MASS::Boston[1:10,]), encode="json")
-httr::content(response, simplifyVector=TRUE)
-#> [1] 25.9269 22.0636 34.1876 33.7737 34.8081 27.6394 21.8007 22.3577 16.7812 18.9785
-```
-
-Note that Plumber, by itself, provides no authentication or security features; anybody who knows the cluster's IP address can access the predictive service.
-At minimum, a basic authentication layer should be added to any Kubernetes deployment.
-Standard Kubernetes procedure involves creating an ingress controller, for example with the nginx or traefik reverse proxy software, and one can then add application-specific authentication layers on top of that (for example, to authenticate with a user's Azure Active Directory or LDAP credentials).
-More details can be found from most Kubernetes learning resources, and are outside the scope of this paper.
+# Use cases and applications
 
@@
 
-The controlled nature of containers, i.e.
-it is possible to define the software environment very well, even on remote machines, make them also useful for continuous integration (CI) and continuous delivery of applications.
-
+- R-Hub
 - DevOps
   - https://www.opencpu.org/posts/opencpu-with-docker/
-
-When doing continuous integration and continuous delivery, it's crucial to test in an environment that matches the production environment.
-Tools like `Gitlab-CI` are built on top of Docker images: the user specifies a base Docker image, and the whole tests are run inside this environment. 
-But, as we just said, this environment has to be fixed, but it also have to contain the necessary toolkit. 
-In order to achieve that, [`r-ci`](https://github.com/ColinFay/r-ci) combines `rocker` versioning and a series of tools specifically designed for testing. 
-That way, package builders can use this image as a base image for there testing environment, without having to install the necessary packages every time they need to run a new test.
-
+- `r-ci`: https://github.com/ColinFay/r-ci
 - [dynwrap](https://github.com/dynverse/dynwrap_containers/blob/master/.travis.yml) [@rcannood]
@@
 
-## Common or public work environments
+## Common or public work environments [div]
 
-The [Binder project](https://mybinder.readthedocs.io/en/latest/), maintained by the team behind Jupyter, makes it possible for users to create and share their computing environments with others.
-A deployed version of this service called binder hub allows anyone with access to a web browser and an internet connection to launch a temporary instance of these custom environments and execute any notebooks contained within.
-From a reproducibility standpoint, Binder makes it exceedingly easy to compile a paper, visualize data, and run small examples from papers or tutorials without the need for any local installation.
+- Binder
+  - [`holepunch`](https://github.com/karthik/holepunch) streamlines making an R research compendium Binder-ready [@karthik]
+  - [`repo2docker`](https://repo2docker.readthedocs.io/en/latest/config_files.html) (`install.R`, `DESCRPTION`, `runtime.txt`) based on the Rocker images [@nuest]
+- GPU [@noamross, @cboettig]
+- Gigantum stack
+- in education (ready to use teaching/learning environments)
+  - [`r-db`](https://github.com/ColinFay/r-db) [@ColinFay]
+- [RCloud](https://github.com/att/rcloud/tree/master/docker) ?
 
-To set up Binder for a project, a user typically starts at an instance of a Binderhub (examples include mybinder.org and binder.pangeo.io) and passes the location of a hosted Git repository.
-Using internal tools such as [repo2docker](https://repo2docker.readthedocs.io/en/latest/config_files.html), the binderhub builds a Dockerfile by parsing the contents of all code contained within.
-While this approach works well for most run of the mill Python projects, it is not so seamless for R projects.
-For any R projects that use the Tidyverse suite [citation in review.
-Will add by the time this gets submitted], the time and resources required to build all dependencies from source can often time out before completion, making it frustrating for the average R user.
-
-[`holepunch`](https://github.com/karthik/holepunch) is a R package that was designed to remove some of these limitations and make Binder more accessible to novice R users.
-`holepunch` achieves this by leveraging Rocker images that contains the Tidyverse along special Jupyter dependencies, and only installs additional packages from CRAN and Bioconductor that are not already part of these images.
-Starting with the Binder/Tidyverse base images eliminates a large part of the build time and in most cases results in a binder instance launching within a minute.
-`holepunch` as a side effect also creates a DESCRIPTION file which then turns any project into a research compendium [doi:10.1080/00031305.2017.1375986].
-The Dockerfile included with the project can also be used to launch a RStudio server locally independent of binder which is especially useful when more computational resources are required.
-
-One use for containers is to run on shared local hardware where teams manage their own high-performance servers.
-This can follow one of several design patterns: users may deploy containers to hardware as a work environment for a specific project, conatiners may provide per-user persistent environments, or a single container can act as a common multi-user environment for a server.
-The former models provide modularity, while the latter approach is most similar to a simple shared server.
-In all cases, though, the containerized approach provides several advantages: First, users may use the same image and thus work environment on desktop and laptop computers, as well.
-Second, software updates can be achieved by updating and redeploying the container, rather than tracking local installs on each server.
-Third, the containerized environment can be quickly deployed to other hardware, cloud or local, if more resources are neccessary or in case of server destruction or failure.
-In any of these cases, users need a method to interact with the containers, be it and IDE, or command-like access and tools such as SSH, which is usually not part of standard container recipes and must be added.
-The Rocker project provides containers pre-installed with the RStudio IDE.
-
-In cases where users store nontrivial amounts of data for their projects, data needs to persist beyond the life of the container.
-This may be via in shared disks, attached network volumes, or in separate storage where it is uploaded between sessions.
-In the case of shared disks or network-attached volumes, care must be taken to persist user permissions, and of course backups are still neccessary.
-When working with multiple servers, an automation framework such as [Ansible](https://www.ansible.com) may be useful for managing users, permisions, and disks along with containers.
-
-In some cases, containers are not completely portable between hardware environments.
-This is the case for software using graphical processing units (GPUs) which are increasingly popular for compute-intensive machine learning tasks.
-Containers running GPU software  require drivers and libraries specific to GPU models and versions, and containers require a specialized runtime to connect to the underlying GPU hardware.
-For NVIDIA GPUs, the [NVIDIA Container Toolkit](https://github.com/NVIDIA/nvidia-docker) includes a specialized runtime plugn for docker and set of base images with appropriate drivers and libraries, and the Rocker project has (beta) images based on these that include GPU-enabled versions of machine-learning R packages.
-
-R is a fantastic tool when it comes to interfacing with databases: almost every open source and proprietary database system has an R package that allows users to connect and interact with the it 
-This flexibility is even broader now that we have tools like `DBI`, that allows to create a common API for interfacing these databases, or like `dbplyr`, which are designed to run `dplyr` code straight against the database.
-But learning and teaching these tools comes with a cost: the cost of deploying or having access to an environment with the softwares and drivers installed. 
-For people teaching R, it can become a barrier if they need to install local versions of the drivers, or to connect to remote instances which might or might not be made available by the IT. 
-Giving access to a sandbox for the most common database environments is the idea behind [`r-db`](https://github.com/ColinFay/r-db) [@ColinFay], a docker image that contains everything needed to connect, from R, to a database.
-Notably, with `r-db`, the users don't have to install complex drivers or to configure their machine in a specific way. 
-On top of that, `r-db` comes with a comprehensive [online guide](http://colinfay.me/r-db/), explaining how to spin a docker instance of a specific database, and how to use `r-db` to connect to it. 
-Each package contained inside this image also has a series examples, so that the user can get started with the database right away.
-
-- [RCloud](https://github.com/att/rcloud/tree/master/docker) uses a `rocker/drd` base image for creation of collaborative data analysis and visualisation environments.
-- [ShinyProxy](https://www.shinyproxy.io/) [creates a container](https://github.com/openanalytics/shinyproxy/blob/master/src/main/java/eu/openanalytics/services/DockerService.java#L388) for each user
-
-## Processing [div]
-
-The portability of containers becomes particularly useful when complex processing tasks shall be offloaded to a server.
+## Deployment, processing, cloud [div]
 
@@
 An R implementation is shown here at [cloudRunR](https://github.com/MarkEdmondson1234/cloudRunR) which uses it to create a scalable R plumber API.
+- [`plumber`](https://www.rplumber.io/docs/hosting.html#docker)
 - `batchtools` \citep{Lang2017batchtools} can [schedule jobs with Docker Swarm](https://mllg.github.io/batchtools/reference/makeClusterFunctionsDocker.html)
@@
 Again, all of these containers are being built on travis-ci, and will only be pushed to docker hub if the integration test succeeds.
+- [ShinyProxy](https://www.shinyproxy.io/) [creates a container](https://github.com/openanalytics/shinyproxy/blob/master/src/main/java/eu/openanalytics/services/DockerService.java#L388) for each user
 
@@
 
-## Development and debugging
+# Other containerisation platforms [NN]
 
-Containers can also serve as useful playgrounds to create environments ad-hoc or to provide very specific environments that are not often needed.
-Developers can readily start a local container based on a Rocker image to investigate a bug report, which might require them to install a specific combination of package versions or use a specific R version (see also versioned images above).
-Unlike recreating the reporter's environment on their regular machine, a container can simply be discarded after use, and there is no need to tediously roll back explorative changes to a previous state.
-Using the Rocker images with RStudio, the disposable environments lack no development comfort (see also the usage in research compendia).
-
-**[`r-debug`](https://github.com/wch/r-debug)** is a purpose built Docker image for debugging R memory problems.
-\citet{eddelbuettel_debugging_2019} describes how a Docker container was used to debug an issue with a package only occuring with a particular version of Fortran, and using tools which are not readily available on all platforms (e.g., not on macOS).
-
-- **[R-Hub](https://github.com/r-hub/)**
-
-# Other containerisation platforms
-
-Docker is not the only containerisation software.
-An alternative stemming from the domain of high-perfomance computing is Singularity \citep{kurtzer_singularity_2017}.
-Singularity can run Docker images, and in the case of Rocker works out of the box if the main process is R, e.g., in `rocker/r-base`, but does not succeed in running images where there is an init script, e.g.
-in containers that by default run RStudio.
-In the latter case, a `Singularity` file, a set of instructions akin to a `Dockerfile` needs to be used.
+- R images for Singularity
+- Running Rocker images with
+  - Singularity
+  - CoreOS rkt?
+- nix?
 
@@
 
-- next steps
-  - consolidation?
-- support more complex set-ups with ready-to-use stacks, e.g. GPU, AI, ML
 - Missing pieces?
@@
 - the ability to move processing between services easily (e.g locally, one cloud providers VM, another cloud provider's Container-as-a-Service)
-- running RStudio in a container is a common approach for reproducibility - _is there a need to run/control containers from a UI integrated in RStudio?_
-- moving away from Docker focus towards containers?
-  - is there a need for a wrapper for [podman's Varlink-API](https://github.com/containers/libpod/blob/master/API.md) (similar to stevedore) and buildah to manage containers and images respectively with alternative software stacks?
-  - will there be control packages for Singularity support?
 - ...
@@
 
-DN conceived of the presented idea and \href{https://github.com/nuest/rockerverse-paper/issues/3}{initialised the formation of the writing team} and wrote the section about images for R distributions.
+DN conceived of the presented idea and \href{https://github.com/nuest/rockerverse-paper/issues/3}{initialised the formation of the writing team}.
 CB ..

### 2019-11-30 manuscript.Rmd
Initial addition of details around R docker clients.
Added some details on `dockyard` and its purpose in the docker life cycle.

--- 2019-11-28 manuscript.Rmd
@@
 **Microsoft R Open** ([MRO](https://github.com/nuest/mro-docker)) is an R distribution formerly known as Revolution R Open (RRO) before Revolution Analytics was acquired by Microsoft.
-MRO is compatible with main R and its packages.
+MRO is compatible with main R and it's packages.sss
 "It includes additional capabilities for improved performance, reproducibility, and platform support." \citep{microsoft_mro}
@@
 
-## Clients
+## Clients [Ellis Hughes]
 
-A number first order R packages exist that talk directly to Docker.
-They can be used to control the Docker daemon, e.g. to build images or run containers, from R without explicit knowledge of the Docker command line interface.
+Interfacing with the Docker daemon is done through the Docker Command Line Interface (CLI).
+Moving back and forth between an R session and the command line can create friction in workflows and reduce reproducibility.
+To solve this problem there are a number first order R packages exist that provide an interface to the Docker CLI.
+This then means that R users are able to take advantage of the benefits of Docker without explicitly working with the CLI.
 
@@
 - [`dockyard`](https://github.com/thebioengineer/dockyard)
+
+At the time of writing this paper, these packages can work in concert to provide interfaces for installing Docker, building images, and running/controlling the life cycle of docker containers.
+
+Each of these packages provide functional wrappers to the Docker CLI at different stages of the docker lifecycle.
+
+_dockyard_ has the goal of lowering barrier to creating dockerfiles and docker images, and deploying docker containers. 
+The way docker images are made is by reading and following a set of instructions found in dockerfiles/
+This requires stepping out of the R ecosystem and into an infrastructure mindset where the user needs to specify the environment they require.
+Because dockerfiles can be read as a series of steps, this lends itself well to the `tidyverse` style of programming where each step is piped into the next.
+_dockyard_ has a 'dockerfile' object that either starts out as an empty dockerfile, or can use existing dockerfile to use as a template.
+This combined with a series of functions that are analogous to the steps commonly found in a dockerfile, it makes building docker images much more approachable to R users.
+There are also wrappers for commonly used steps for R programmers like adding libraries and copying files into the R docker containers.
+
+Once the dockerfile is created, _dockyard_ has the functionality to build the image and deploy it.
+
+These clients are important for bringing accessibility to the Docker ecosystem for the wide variety of R programmers.
 
@@
 - [`analogsea`](https://github.com/sckott/analogsea) (digital ocean R client)
-- [`AzureContainers`](https://cran.r-project.org/package=AzureContainers): Umbrella package for working with containers in Microsoft's Azure cloud.
-Provides interfaces to three Azure services: Container Registry, Container Instances (for running individual containers), and Kubernetes Services (for orchestrated deployments)
-
-## Using R to power enterprise software in production environments [@jnolis, @nolistic]
-
-R has been historically viewed as a tool for analysis and scientific research--not for creating software that corporations can rely on to continously run in real time.
-However, thanks to advancements in R running as a web service, along with with the ability to deploy R in docker containers, modern enterprises are now capable of having real-time machine learning powered by R.
-
-The backbone of this work is the _[plumber](https://www.rplumber.io/docs/hosting.html#docker)_ package, created by Jeff Allen.
-The plumber package allows R to be used to host web services through RESTful APIs.
-When R is running on a server, a different server can ask R to make a computation through an HTTP request and get an immediate response.
-For instance, if a data scientist at a company creates a machine learning model to tell which customers are likely to make a repeat purchase, if they serve the model through a plumber API then someone else on the marketing team could write software that sends different emails depending on that real-time prediction.
-Plumber syntax is simple to read and understand: the API is created by commenting above function names in an R script.
-
-```
-#* Return the sum of two numbers
-#* @get /random
-function(){
-  runif()
-}
-```
-_Example of a plumber API to generated a random number by visiting `http://127.0.0.1/random`_
-
-Plumber becomes far more powerful with the addition of Docker.
-Docker can be used to create images that [include both R and plumber](https://www.rplumber.io/docs/hosting.html#docker).
-Then when the image is deployed, that creates the web endpoints that other services can hit and call the R code.
-This pattern of deploying code mimics those used by software engineering services created in Java, Python, and other modern languages.
-By using Docker to deploy plumber APIs, R can be used alongside those langages as a first class member of a software engineering technical stack.
-
-At T-Mobile, R is used to power real-time services that are called over a million times a day.
-T-Mobile created a set of neural network machine learning natural language processing models to help customer care agents manage text-based messages for customers.
-The models are convolutional neural networks that use the RStudio Keras package and are built on top of a Rocker Docker image.
-Since the models power tools for agents and customers, they need to have extremely high uptime and reliability--and they found that R is able to perform.
-The AI @ T-Mobile team open sourced [their docker images](https://github.com/tmobile/r-tensorflow-api  which are available on GitHub.
-
-## Deployment to the cloud [@hongooi]
-
-While Plumber provides the low-level infrastructure for turning a statistical model into a predictive service, for production purposes it is usually necessary to take into account considerations such as scalability, reliability and ease of management.
-A single container is limited in the volume of requests it can service.
-Furthermore, if the container goes down, or is occupied with processing a computationally intensive task, the service becomes unavailable until it is restarted.
-
-These issues can be mitigated by deploying not a single container, but a _cluster_ of containers, orchestrated as a single deployment.
-The deployment system of choice at the time of writing is [Kubernetes](https://kubernetes.io).
-This allows you to deploy a single image to a pool of servers, which can then serve requests in parallel.
-Client-side requests are made to the cluster ingress endpoint, which then redirects them to individual servers.
-The orchestrator can identify cluster nodes that have failed, and restart them automatically; it can also detect when there is no capacity to service more requests, and start additional servers and containers as needed.
-
-While these load-balancing and autoscaling properties of a Kubernetes cluster are invaluable, they are still limited by the hardware resources that it has access to.
-A cluster cannot scale out if it has run out of servers.
-This issue in turn can be addressed by deploying not to on-premise servers but to the cloud, where effectively unlimited computing resources are available (as long as one can pay for them).
-From a management perspective, cloud deployments have the additional advantage that hardware maintenance becomes the responsibility of the cloud provider, eliminating another pain point.
-Orchestrator services are available from all the major cloud service providers, including Amazon Web Services (AWS), Google Cloud Platform, and Microsoft Azure.
-
-An example of an R package for working with containers in the cloud is AzureContainers, which interacts with three Azure services: Container Instances (for running a single container), Container Registry (a private docker registry service), and Kubernetes Service (for Kubernetes on Azure).
-AzureContainers provides a lightweight yet powerful interface for working with Resource Manager, the framework for deploying and managing arbitrary resources in Azure.
-On the client side, it provides simple shells to the Docker, Kubectl and Helm tools that are commonly used for managing containers and Kubernetes clusters.
-
-Here is some simple code that pushes a Docker image to an Azure Container Registry, and then creates a deployment using Azure Kubernetes Service.
-This is a slightly modified version of the code from the "Deploying a prediction service with Plumber" AzureContainers vignette.
-The code makes use of a yaml file to define the deployment and predictive service, which is standard practice with Kubernetes; the yaml can be seen in the aforementioned vignette.
-
-```r
-library(AzureContainers)
-# create a resource group for our deployments
-deployresgrp <- AzureRMR::get_azure_login()$
-    get_subscription("subscription_id")$
-    create_resource_group("deployresgrp", location="australiaeast")
-
-### create a container registry
-deployreg_svc <- deployresgrp$create_acr("deployreg")
-
-# build our image (a random forest on the Boston housing data, available in the MASS package)
-call_docker("build -t bos-rf .")
-
-# upload the image to Azure
-deployreg <- deployreg_svc$get_docker_registry()
-deployreg$push("bos-rf")
-
-### create a Kubernetes cluster with 2 nodes
-deployclus_svc <- deployresgrp$create_aks("deployclus", agent_pools=aks_pools("pool1", 2))
-
-# grant the cluster pull access to the registry
-gr <- AzureGraph::get_graph_login()
-aks_app_id <- deployclus$properties$servicePrincipalProfile$clientID
-reg$add_role_assignment(gr$get_app(aks_app_id), "Acrpull")
-
-# get the cluster endpoint
-deployclus <- deployclus_svc$get_cluster()
-
-# create and start the service
-deployclus$create("bos-rf.yaml")
-```
-
-Once the service has been created, we can check on its status and obtain predictions:
-
-```r
-deployclus$get("deployment bos-rf")
-#> Kubernetes operation: get deployment bos-rf  --kubeconfig=".../kubeconfigxxxx"
-#> NAME      DESIRED   CURRENT   UP-TO-DATE   AVAILABLE   AGE
-#> bos-rf    1         1         1            1           5m
-
-deployclus$get("service bos-rf-svc")
-#> Kubernetes operation: get service bos-rf-svc  --kubeconfig=".../kubeconfigxxxx"
-#> NAME         TYPE           CLUSTER-IP   EXTERNAL-IP     PORT(S)          AGE
-#> bos-rf-svc   LoadBalancer   10.0.8.189   xxx.xxx.xxx.xxx 8000:32276/TCP   5m
-
-response <- httr::POST("http://xxx.xxx.xxx.xxx:8000/score",
-    body=list(df=MASS::Boston[1:10,]), encode="json")
-httr::content(response, simplifyVector=TRUE)
-#> [1] 25.9269 22.0636 34.1876 33.7737 34.8081 27.6394 21.8007 22.3577 16.7812 18.9785
-```
-
-Note that Plumber, by itself, provides no authentication or security features; anybody who knows the cluster's IP address can access the predictive service.
-At minimum, a basic authentication layer should be added to any Kubernetes deployment.
-Standard Kubernetes procedure involves creating an ingress controller, for example with the nginx or traefik reverse proxy software, and one can then add application-specific authentication layers on top of that (for example, to authenticate with a user's Azure Active Directory or LDAP credentials).
-More details can be found from most Kubernetes learning resources, and are outside the scope of this paper.
+- [`plumber`](https://www.rplumber.io/docs/hosting.html#docker)
+- [production deployment of neural networks](https://github.com/tmobile/r-tensorflow-api) [@jnolis, @nolistic]
 

### 2019-12-07 manuscript.Rmd
rework section on clients and add overview table

@@
   - name: Robrecht Cannoodt
-    #affiliatin: -
+    affiliation: Ghent University
+    address:
+      - Data Mining and Modelling for Biomedicine group
+      - VIB Center for Inflammation Research
+      - Technologiepark 71
+      - 9052 Ghent, Belgium
+      - \includegraphics[height=\fontcharht\font`\0]{ORCID-iD_icon-vector.pdf} \href{https://orcid.org/0000-0003-3641-729X}{https://orcid.org/0000-0003-3641-729X}
     #email: -
@@
 ---
+
+```{r setup, include=FALSE}
+library(tidyverse)
+library(kableExtra)
+```
 
@@
 
-## Clients [Ellis Hughes]
+## Clients [Ellis Hughes, Robrecht Cannoodt @rcannood]
 
-Interfacing with the Docker daemon is done through the Docker Command Line Interface (CLI).
-Moving back and forth between an R session and the command line can create friction in workflows and reduce reproducibility.
-To solve this problem there are a number first order R packages exist that provide an interface to the Docker CLI.
-This then means that R users are able to take advantage of the benefits of Docker without explicitly working with the CLI.
+Interfacing with the Docker daemon is typically done through the Docker Command Line Interface (CLI).
+However, moving back and forth between an R console and the command line can create friction in workflows and reduce reproducibility.
+A number of first-order R packages provide a interface to the Docker CLI, allowing to automate interaction with the Docker CLI from an R console.
 
-- [`stevedore`](https://github.com/richfitz/stevedore)
-- [`docker`](https://bhaskarvk.github.io/docker/)
-- [`harbor`](https://github.com/wch/harbor/)
-- [`dockermachine`](https://github.com/cboettig/dockermachine)
-- [`dockyard`](https://github.com/thebioengineer/dockyard)
+Each of these packages has particular advantages as they provide function wrappers for interacting with the Docker CLI at different stages of a container's life cycle.
+Examples of such interactions are installing the Docker software, creating Dockerfiles, building an image, and launching a container on a local machine or on the cloud.
+As such, the choice of which package is most useful depends on the use-case at hand.
+<!-- ^ in the examples listed above, the name of the packages which provide such functionality could be mentioned, this would highlight the variety in functionality already. -->
 
-At the time of writing this paper, these packages can work in concert to provide interfaces for installing Docker, building images, and running/controlling the life cycle of docker containers.
+```{r clientfunctionalities, echo=FALSE, results='asis', warning=FALSE, message=FALSE,}
 
-Each of these packages provide functional wrappers to the Docker CLI at different stages of the docker lifecycle.
+library(kableExtra)
 
-_dockyard_ has the goal of lowering barrier to creating dockerfiles and docker images, and deploying docker containers. 
+functionalities <- tribble(
+  ~id, ~Functionality,
+  "dockerfile", "Generate a Dockerfile",
+  "build", "Build an image",
+  "execute", "Execute a container locally or remotely",
+  "cloud", "Deploy or manage an instances in the cloud",
+  "interact", "Interact with an instance (e.g. file transfer)",
+  "images", "Manage storage of images",
+  "othercontainers", "Supports Docker and Singularity",
+  "api", "Direct access to Docker API instead of using the CLI",
+  "install", "Installing Docker software"
+)
+
+# some functionalities might be missing. please add them if you find that any are missing!
+package_functionalities <- tribble(
+  ~package, ~id,
+  "stevedore", c("api", "execute", "cloud", "images", "interact"), # build?
+  "dockyard", c("dockerfile", "build", "execute"), # cloud?
+  "dockermachine", c("install", "execute", "cloud", "interact"),
+  "AzureContainers", c("execute", "cloud"),
+  "babelwhale", c("execute", "interact", "othercontainers"),  
+  "harbor", c("execute", "cloud", "images")
+)
+
+tab <- 
+  full_join(
+  functionalities %>% crossing(package = package_functionalities$package), 
+  package_functionalities %>% unnest(id) %>% mutate(check = "\\checkmark"),#\u2714 # \\checkmark
+  by = c("id", "package")
+) %>%
+  mutate(check = ifelse(is.na(check), "", check)) %>% 
+  spread(package, check) %>% 
+  slice(match(functionalities$id, id)) %>% 
+  select(-id)
+
+knitr::kable(tab, escape = FALSE) %>% 
+  # kable_styling(latex_options = "striped") %>%
+  row_spec(0, angle = -90)
+# don't print this table just yet, wait for this PR to be merged
+# https://github.com/rstudio/rticles/pull/261
+```
+
+**stevedore** ([https://cran.r-project.org/package=stevedore](https://cran.r-project.org/package=stevedore)) 
+<!-- from the CRAN description:
+Work with containers over the Docker API.
+Rather than using system calls to interact with a docker client, using the API directly means that we can receive richer information from docker.
+The interface in the package is automatically generated using the 'OpenAPI' (a.k.a., 'swagger') specification, and all return values are checked in order to make them type stable.
+-->
+
+<!--
+**docker** ([https://bhaskarvk.github.io/docker](https://bhaskarvk.github.io/docker)
+docker has been removed from CRAN, seems unmaintained by authors. Maybe it should be removed from this list?
+-->
+
+**dockyard** ([https://github.com/thebioengineer/dockyard](https://github.com/thebioengineer/dockyard)) has the goal of lowering barrier to creating dockerfiles and docker images, and deploying docker containers.
 The way docker images are made is by reading and following a set of instructions found in dockerfiles/
@@
 Because dockerfiles can be read as a series of steps, this lends itself well to the `tidyverse` style of programming where each step is piped into the next.
-_dockyard_ has a 'dockerfile' object that either starts out as an empty dockerfile, or can use existing dockerfile to use as a template.
+dockyard has a 'dockerfile' object that either starts out as an empty dockerfile, or can use existing dockerfile to use as a template.
 This combined with a series of functions that are analogous to the steps commonly found in a dockerfile, it makes building docker images much more approachable to R users.
 There are also wrappers for commonly used steps for R programmers like adding libraries and copying files into the R docker containers.
+Once the dockerfile is created, dockyard has the functionality to build the image and deploy it.
 
-Once the dockerfile is created, _dockyard_ has the functionality to build the image and deploy it.
+<!-- From the github description:
+The goal of **dockermachine** ([https://github.com/cboettig/dockermachine](https://github.com/cboettig/dockermachine)) is to provide a convenient interface to docker-machine from the R command line.
+This will allow R users to easily create, control, and destroy cloud instances across a wide variety of providers.
+-->
 
-These clients are important for bringing accessibility to the Docker ecosystem for the wide variety of R programmers.
+<!-- From the cran description
+**AzureContainers** ([https://cran.r-project.org/package=AzureContainers](https://cran.r-project.org/package=AzureContainers)) lets you build and deploy containerised services in R, using Docker and Kubernetes.
+For full functionality, you should have Docker installed, as well as the kubectl and helm commandline tools.
+Otherwise it is relatively lightweight, requiring neither Powershell nor Python.
+-->
+
+**babelwhale** ([https://cran.r-project.org/package=babelwhale](https://cran.r-project.org/package=babelwhale)) 
+allows executing and interacting with containers, which can use either Docker or Singularity as a backend.
+provides a unified interface to interact with Docker and Singularity containers. You can execute a command inside a container, mount a volume or copy a file.
+
+<!-- from the github description
+**harbor** ([https://github.com/wch/harbor](https://github.com/wch/harbor) is an R package for controlling docker containers on local and remote hosts.
+-->
 
@@
 
-- [`babelwhale`](https://cran.r-project.org/web/packages/babelwhale/index.html): Running a Docker from R with Singularity or Docker as back-end.
-This is really useful in HPC environments where a user might not have root access but is able to install Singularity instead of Docker.
-[@rcannood]
 - `RSelenium`

### 2019-12-07 manuscript.Rmd
Rename 'clients' to 'Interfaces for Docker in R' and some other minor changes.

@@
       - \includegraphics[height=\fontcharht\font`\0]{ORCID-iD_icon-vector.pdf} \href{https://orcid.org/0000-0003-3641-729X}{https://orcid.org/0000-0003-3641-729X}
-    #email: -
+    email: robrecht@cannoodt.dev
+  - name: Ellis Hughes
   - name: Dirk Eddelbuettel
@@
 
-## Clients [Ellis Hughes, Robrecht Cannoodt @rcannood]
+## Interfaces for Docker in R [Ellis Hughes, Robrecht Cannoodt]
 
@@
   "dockermachine", c("install", "execute", "cloud", "interact"),
-  "AzureContainers", c("execute", "cloud"),
+  "AzureContainers", c("build", "execute", "cloud"),
   "babelwhale", c("execute", "interact", "othercontainers"),  
@@
   row_spec(0, angle = -90)
-# don't print this table just yet, wait for this PR to be merged
-# https://github.com/rstudio/rticles/pull/261
 ```
@@
 
+**dockermachine** ([https://github.com/cboettig/dockermachine](https://github.com/cboettig/dockermachine))
 <!-- From the github description:
-The goal of **dockermachine** ([https://github.com/cboettig/dockermachine](https://github.com/cboettig/dockermachine)) is to provide a convenient interface to docker-machine from the R command line.
+**dockermachine** ([https://github.com/cboettig/dockermachine](https://github.com/cboettig/dockermachine)) provides a convenient interface to docker-machine from the R command line.
 This will allow R users to easily create, control, and destroy cloud instances across a wide variety of providers.
@@
 
+**AzureContainers** ([https://cran.r-project.org/package=AzureContainers](https://cran.r-project.org/package=AzureContainers))
 <!-- From the cran description
-**AzureContainers** ([https://cran.r-project.org/package=AzureContainers](https://cran.r-project.org/package=AzureContainers)) lets you build and deploy containerised services in R, using Docker and Kubernetes.
+ lets you build and deploy containerised services in R, using Docker and Kubernetes.
 For full functionality, you should have Docker installed, as well as the kubectl and helm commandline tools.
@@
 
+**harbor** ([https://github.com/wch/harbor](https://github.com/wch/harbor)
 <!-- from the github description
-**harbor** ([https://github.com/wch/harbor](https://github.com/wch/harbor) is an R package for controlling docker containers on local and remote hosts.
+is an R package for controlling docker containers on local and remote hosts.
 -->
@@
 CB ..
-RC ..
+RC and RC wrote the section on interfaces for Docker in R.
 DE ..

### 2019-12-07 manuscript.Rmd
allow the manuscript to build with the CRAN version of rticles

@@
 library(tidyverse)
-library(kableExtra)
 ```
@@
 
-```{r clientfunctionalities, echo=FALSE, results='asis', warning=FALSE, message=FALSE,}
-
-library(kableExtra)
-
+```{r clientfunctionalities, echo=FALSE, results='asis', warning=FALSE, message=FALSE}
 functionalities <- tribble(
@@
 
+# inserting a table in the R Journal currently fails
+# waiting for this PR to merge
+# https://github.com/rstudio/rticles/pull/261
+if (packageVersion("rticles") >= "0.12.3") {
+  requireNamespace("kableExtra")
   knitr::kable(tab, escape = FALSE) %>%
-  # kable_styling(latex_options = "striped") %>%
-  row_spec(0, angle = -90)
+    kableExtra::row_spec(0, angle = -90)
+} else {
+  cat("\n\n[Functionality overview table omitted due to a small bug in rticles]\n\n")
+}
 ```

### 2019-12-07 manuscript.Rmd
move figures
simplify orcid specification

@@
       - 48149 Münster, Germany
-      - \href{https://orcid.org/0000-0002-0024-5046}{\includegraphics[height=\fontcharht\font`\0]{ORCID-iD_icon-vector.pdf} https://orcid.org/0000-0002-0024-5046}
+      - \orcid{0000-0002-0024-5046}
     email: daniel.nuest@uni-muenster.de
@@
       - 9052 Ghent, Belgium
-      - \includegraphics[height=\fontcharht\font`\0]{ORCID-iD_icon-vector.pdf} \href{https://orcid.org/0000-0003-3641-729X}{https://orcid.org/0000-0003-3641-729X}
+      - \orcid{0000-0003-3641-729X}
     email: robrecht@cannoodt.dev
@@
       - Champaign, IL 61820, USA
-      - \includegraphics[height=\fontcharht\font`\0]{ORCID-iD_icon-vector.pdf} \href{https://orcid.org/0000-0001-6419-907X}{https://orcid.org/0000-0001-6419-907X}
+      - \orcid{0000-0001-6419-907X}
     email: dirkd@eddelbuettel.com
@@
       - 93300 Aubervilliers, France
-      - \includegraphics[height=\fontcharht\font`\0]{ORCID-iD_icon-vector.pdf} \href{https://orcid.org/0000-0001-7343-1846}{https://orcid.org/0000-0001-7343-1846}
+      - \orcid{0000-0001-7343-1846}
     email: contact@colinfay.me
@@
       - Berkeley, CA 94720, USA
-      - \includegraphics[height=\fontcharht\font`\0]{ORCID-iD_icon-vector.pdf} \href{https://orcid.org/0000-0002-0233-1757}{https://orcid.org/0000-0002-0233-1757}
+      - \orcid{0000-0002-0233-1757}
     email: karthik.ram@berkeley.edu
@@
       - New York, NY 10001, USA
-      - \includegraphics[height=\fontcharht\font`\0]{ORCID-iD_icon-vector.pdf} \href{https://orcid.org/0000-0002-0233-1757}{https://orcid.org/0000-0002-0233-1757}
+      - \orcid{0000-0002-0233-1757}
     email: ross@ecohealthalliance.org
@@
       - Buffalo, New York, 14263, USA
-      - \includegraphics[height=\fontcharht\font`\0]{ORCID-iD_icon-vector.pdf} \href{https://orcid.org/0000-0002-5910-4010}{https://orcid.org/0000-0002-5910-4010}
+      - \orcid{0000-0002-5910-4010}
     email: lori.shepherd@roswellpark.org
@@
     - Buffalo, New York, 14263, USA
-    - \includegraphics[height=\fontcharht\font`\0]{ORCID-iD_icon-vector.pdf} \href{https://orcid.org/0000-0002-0224-9817}{https://orcid.org/0000-0002-0224-9817}
+    - \orcid{0000-0002-0224-9817}
     email: nitesh.turaga@roswellpark.org
@@
     keep_tex: true
+    includes:
+      in_header: header.tex
 ---
@@
   \centering
-  \includegraphics[width=\textwidth]{liftr-workflow}
+  \includegraphics[width=\textwidth]{figures/liftr-workflow}
   \caption{The liftr workflow for rendering containerized R Markdown documents.}

### 2019-12-09 manuscript.Rmd
Use vectorised logo instead of rasterised logo if the \titlepage command would be used
update AzureContainers description

@@
 - is possible, only relevant in organisations with an existing Windows Server-based infrastructure, can meet policies then
+<!-- can also be relevant for standalone Win 10 Professional users who have Docker for Windows installed -- HO -->
 
@@
 
-**AzureContainers** ([https://cran.r-project.org/package=AzureContainers](https://cran.r-project.org/package=AzureContainers))
-<!-- From the cran description
- lets you build and deploy containerised services in R, using Docker and Kubernetes.
-For full functionality, you should have Docker installed, as well as the kubectl and helm commandline tools.
-Otherwise it is relatively lightweight, requiring neither Powershell nor Python.
--->
+**AzureContainers** ([https://cran.r-project.org/package=AzureContainers](https://cran.r-project.org/package=AzureContainers)) is an interface to a number of container-related services in Microsoft's Azure cloud, namely Container Instances, Container Registry and Kubernetes Service.
+While it is mainly intended for working with Azure, as a convenience feature it includes lightweight, cross-platform shells to Docker, kubectl and helm.
+These can be used to create and manage arbitrary Docker images and containers, as well as Kubernetes clusters on any platform or cloud service.
 
@@
   \centering
-  \includegraphics[width=\textwidth]{figures/liftr-workflow}
+  \includegraphics[width=\textwidth]{liftr-workflow}
   \caption{The liftr workflow for rendering containerized R Markdown documents.}

### 2019-12-12 manuscript.Rmd

@@
 
-The R community keeps growing: the number of new packages on CRAN keeps on rising, meetups, conferences, online courses and unconferences prosper, and the uptake in education and industry increases [REFs].
-All this cements the role of R as the lingua franca of statistics, data visualisation, and computational research.
-Coinciding with the rise of R was the advent of Docker [CITATION?] as a general tool for development, distribution, and deployment of software.
-Combining both these topics, the _Rocker Project_ ([https://www.rocker-project.org/](https://www.rocker-project.org/)) was introduced in 2014 and provides a number of [Docker](https://en.wikipedia.org/wiki/Docker_(software)) images for various use cases as described in \citet{RJ-2017-065}.
-The considerable uptake and continued evolution of the Rocker suite of containers has lead to numerous projects extending or building upon Rocker images ranging from reproducible research to production deployments.
-This article presents this _Rockerverse_ of projects.
-It also introduces related activities connecting the R language and environment with other containerisation solutions.
-The main contributions is a coherent picture of the current lay of the land of containers in, with, and for `R`.
-This diversity includes demonstrators, early prototypes, and mature projects.
+The R community keeps growing.
+This can be seen in the number of new packages on CRAN which keeps on rising, but also in the increased number of conferences, online courses, meetups, and unconferences, as well as in the uptake in education and industry increases [REFs].
+All this cements the role of R as the _lingua franca_ of statistics, data visualisation, and computational research.
+Coinciding with the rise of R was the advent of [Docker](https://en.wikipedia.org/wiki/Docker_(software)) as a general tool for distribution and deployment of software.
+Combining both these topics, the Rocker Project ([https://www.rocker-project.org/](https://www.rocker-project.org/)) provides images with R (see the next Section for more details).
+The considerable uptake and continued evolution of the Rocker Project has lead to numerous projects extending or building upon Rocker images, ranging from reproducible research to production deployments.
+This article presents this _Rockerverse_ of projects across all development stages: demonstrations, prototypes, and mature products.
+We also introduce related activities connecting the R language and environment with other containerisation solutions.
+The main contributions is a coherent picture of the current lay of the land of using containers in, with, and for R.
 
@@
 
-<!--_Do we need a generic intro here?_ DE: Yes, and happy to take a stab.-->
+Docker, an application and service provide by the eponymous company, has in just a few short years risen to prominence for development, testing, deployment and distribution of computer software \citep[cf.][]{datadog_8_2018,munoz_history_2019}.
+While there are related approaches such as LXC\footnote{\href{https://en.wikipedia.org/wiki/LXC}{https://en.wikipedia.org/wiki/LXC}} or Singularity \citep{kurtzer_singularity_2017}, Docker has become synomymous with "containerization"---the method of taking software artefacts and bundling them in such a way that use becomes standardized and portable across operationg systems.
+In doing so, Docker had recognised and validated the importance of one very important thread that had been emerging, namely virtualization.
+By allowing (one or possibly) multiple applications or services to run concurrently on one host machine without any fear of interference between them, an important scalability opportunity is being provided.
+But Docker improved on virtualization by accessing the host system---generally Linux---through a much thinner and smaller shim than a full operating system emulation or virtualization.
+This makes for more efficient use of system resources and allowed another order of magnitude in terms of scalability of deployment [REFERENCE!].   
+While Docker makes use of Linux kernel features, its importance was large enough so that some required aspects of running Docker have been added to other operating systems to support Docker there more efficiently too \citep{microsoft_linux_2019}.
+The success even lead to standardisation and industry collaboration \citep{oci_open_2019}.
 
-Docker, an application and service provide by the eponymous company, has in just a few short years risen to prominence for development, testing, deployment and distribution of computer software.
-While there are related approaches such as LXC or Singularity, Docker has become synomymous with "containerization"---the method of taking software artefacts and bundling them in such a way that use becomes standardized and portable across operationg systems.
-In doing so, Docker had recognised and validated the importance of one very important thread that had been emerging, namely virtualization.
-By allowing multiple concurrent applications or services to run concurrently on one host machine without any fear of interference between them, an important scalability feature had been realized.
-But Docker improved on virtualization by accessing the host system---generally Linux---through a much thinner and smaller shim than a full operating system emulation or virtualization.
-This makes for more efficient use of system resources and allowed another order of magnitude in terms of scalability of deployment.
-[Reference?]   While Docker makes use of Linux kernel features, its importance was large enough so that some required aspects of running Docker have been added to other operating systems to support Docker there more efficiently too.
-[Win10; reference?]
-
-The key accomplishment of Docker is to make a "bundled" aggregation of software available to any system running Docker, without requiring much else from the host.
-This is a rather attractive, and novel, proposition which has lead to widespread adoption and use of Docker in a variety of domains.
-It provided also a natural match for "cloud deployment" which runs (or at least appears to run) "seamlessly" without much explicit reference to the underlying machine, architecture or operating system.
-Using containers on cloud hosted servers is a natural fit also because these containers can be deployed with very littler in terms of dependencies on the host system.
+The key accomplishment of Docker (as an "application") is to make a "bundled" aggregation of software (the so-called "image") available to any system equipped to run Docker, without requiring much else from the host besides the actual Docker application installation.
+This is a rather attractive, and novel, proposition which has lead to widespread adoption and use of Docker in a variety of domains, _e.g._, cloud computing infrastructure \citep{Bernstein2014}, data science \citep{boettiger_introduction_2015}, and edge computing \citep{alam_orchestration_2018} [more examples!].
+It provided to be a natural match for "cloud deployment" which runs (or at least appears to run) "seamlessly" without much explicit reference to the underlying machine, architecture or operating system: containers can be deployed with very little in terms of dependencies on the host system---only the container runtime is required.
 
 For statistical computing and analysis centered around R, the Rocker Project has provided a variety of Docker containers since its start in 2014 \citep{RJ-2017-065}.
-The Rocker Project provides several lines of containeres spanning to from building blocks with R-release or R-devel to containers with RStudio Server and Shiny Server.
-Also of note is a series of "versioned" containers which match the R release they contain with the _then-current_ set of packages via the MRAN Snapshot views of CRAN.
+The Rocker Project provides several lines of containers spanning to from building blocks with `R-release` or `R-devel`, via containers with RStudio~Server and Shiny~Server, to domain-specific containers such as `geospatial`.
+Also of note is a series of "versioned" containers which match the R release they contain with the _then-current_ set of packages via the MRAN Snapshot views of CRAN \citep{microsoft_cran_2019}.
 [Maybe at a later point add forward reference to other topics building on Rocker described here.]

### 2019-12-12 manuscript.Rmd
minor cleanups in bib file after merge, mention issues in contributions section

--- 2019-12-12 manuscript.Rmd
@@
 Starting with the Binder/Tidyverse base images eliminates a large part of the build time and in most cases results in a binder instance launching within a minute.
-`holepunch` as a side effect also creates a DESCRIPTION file which then turns any project into a research compendium [doi:10.1080/00031305.2017.1375986].
+`holepunch` as a side effect also creates a DESCRIPTION file which then turns any project into a research compendium \citep{marwick_packaging_2018}.
 The Dockerfile included with the project can also be used to launch a RStudio server locally independent of binder which is especially useful when more computational resources are required.
@@
 This articles was collaboratively written at \href{https://github.com/nuest/rockerverse-paper/}{https://github.com/nuest/rockerverse-paper/}.
-The \href{https://github.com/nuest/rockerverse-paper/graphs/contributors}{contributors page} and \href{https://github.com/nuest/rockerverse-paper/commits/master}{commit history} provide a detailed view on the respective contributions.
+The \href{https://github.com/nuest/rockerverse-paper/graphs/contributors}{contributors page}, \href{https://github.com/nuest/rockerverse-paper/commits/master}{commit history}, and \href{https://github.com/nuest/rockerverse-paper/issues/}{discussion issues} provide a detailed view on the respective contributions.
 

### 2019-12-13 manuscript.Rmd

@@
 
-The [Binder project](https://mybinder.readthedocs.io/en/latest/), maintained by the team behind Jupyter, makes it possible for users to create and share their computing environments with others.
-A deployed version of this service called binder hub allows anyone with access to a web browser and an internet connection to launch a temporary instance of these custom environments and execute any notebooks contained within.
+The fact that Docker images are portable and well defined make them useful when more than one person needs access to the same computing environment.
+This is even more useful when some of the users do not have the expertise to create such an environment themselves, and when these environments can be run in public or shared infrastructure.
+
+[`holepunch`](https://github.com/karthik/holepunch) is an R package that was designed to make sharing work environments accessible to novice R users based on _Binder_.
+The [Binder project](https://mybinder.readthedocs.io/en/latest/), maintained by the team behind Jupyter, makes it possible for users to create and share their computing environments with others \citep{jupyter_binder_2018}.
+A _BinderHub_ allows anyone with access to a web browser and an internet connection to launch a temporary instance of these custom environments and execute any workflows contained within.
 From a reproducibility standpoint, Binder makes it exceedingly easy to compile a paper, visualize data, and run small examples from papers or tutorials without the need for any local installation.
-
-To set up Binder for a project, a user typically starts at an instance of a Binderhub (examples include mybinder.org and binder.pangeo.io) and passes the location of a hosted Git repository.
-Using internal tools such as [repo2docker](https://repo2docker.readthedocs.io/en/latest/config_files.html), the binderhub builds a Dockerfile by parsing the contents of all code contained within.
+To set up Binder for a project, a user typically starts at an instance of a BinderHub and passes the location of a repository with a workspace, e.g., a hosted Git repository, or a data repository like Zenodo.
+Binder's core internal tool is [repo2docker](https://repo2docker.readthedocs.io/en/latest/config_files.html).
+It deterministically builds a Docker image by parsing the contents of a repository, e.g. project dependency configurations or simple configuration files.
+In the most powerful case, `repo2docker` builds a given Dockerfile.
 While this approach works well for most run of the mill Python projects, it is not so seamless for R projects.
-For any R projects that use the Tidyverse suite [citation in review.
-Will add by the time this gets submitted], the time and resources required to build all dependencies from source can often time out before completion, making it frustrating for the average R user.
+For any R projects that use the Tidyverse suite [citation in review, will add by the time this gets submitted], the time and resources required to build all dependencies from source can often time out before completion, making it frustrating for the average R user.
+`holepunch` removes some of these limitations by leveraging Rocker images that contain the Tidyverse along special Jupyter dependencies, and only installs additional packages from CRAN and Bioconductor that are not already part of these images.
+It short cicuits the configuration file parsing in `repo2docker` and starts with the Binder/Tidyverse base images, which eliminates a large part of the build time and in most cases results in a binder instance launching within a minute.
+`holepunch` as a side effect also creates a `DESCRIPTION` file which then turns any project into a research compendium \citep{marwick_packaging_2018}.
+The Dockerfile included with the project can also be used to launch a RStudio server locally independent of Binder which is especially useful when more or special computational resources are required.
 
@@
 
-One use for containers is to run on shared local hardware where teams manage their own high-performance servers.
+In **high-performance computing**, one use for containers is to run workflows on shared local hardware where teams manage their own high-performance servers.
 This can follow one of several design patterns: users may deploy containers to hardware as a work environment for a specific project, conatiners may provide per-user persistent environments, or a single container can act as a common multi-user environment for a server.
+In all cases, though, the containerized approach provides several advantages: First, users may use the same image and thus work environment on desktop and laptop computers, as well.
 The former models provide modularity, while the latter approach is most similar to a simple shared server.
-In all cases, though, the containerized approach provides several advantages: First, users may use the same image and thus work environment on desktop and laptop computers, as well.
 Second, software updates can be achieved by updating and redeploying the container, rather than tracking local installs on each server.
@@
 But learning and teaching these tools comes with a cost: the cost of deploying or having access to an environment with the softwares and drivers installed. 
-For people teaching R, it can become a barrier if they need to install local versions of the drivers, or to connect to remote instances which might or might not be made available by the IT. 
-Giving access to a sandbox for the most common database environments is the idea behind [`r-db`](https://github.com/ColinFay/r-db) [@ColinFay], a docker image that contains everything needed to connect, from R, to a database.
+For people teaching R, it can become a barrier if they need to install local versions of the drivers, or to connect to remote instances which might or might not be made available by IT services.
+Giving access to a sandbox for the most common database environments is the idea behind **[`r-db`](https://github.com/ColinFay/r-db)** [@ColinFay], a Docker image that contains everything needed to connect, from R, to a database.
 Notably, with `r-db`, the users don't have to install complex drivers or to configure their machine in a specific way. 
-On top of that, `r-db` comes with a comprehensive [online guide](http://colinfay.me/r-db/), explaining how to spin a docker instance of a specific database, and how to use `r-db` to connect to it. 
+On top of that, `r-db` comes with a comprehensive [online guide](http://colinfay.me/r-db/), explaining how to spin a Docker instance of a specific database, and how to use `r-db` to connect to it.
 Each package contained inside this image also has a series examples, so that the user can get started with the database right away.
+The `rocker/tidyverse` base image ensures that users can also readily use packages for analysis, display, and reporting.
 
-- [RCloud](https://github.com/att/rcloud/tree/master/docker) uses a `rocker/drd` base image for creation of collaborative data analysis and visualisation environments.
-- [ShinyProxy](https://www.shinyproxy.io/) [creates a container](https://github.com/openanalytics/shinyproxy/blob/master/src/main/java/eu/openanalytics/services/DockerService.java#L388) for each user
+**[RCloud](https://github.com/att/rcloud/tree/master/docker)** uses a `rocker/drd` base image for creation of collaborative data analysis and visualisation environments.
 

### 2019-12-13 manuscript.Rmd

@@
 - [`analogsea`](https://github.com/sckott/analogsea) (digital ocean R client)
-- [`plumber`](https://www.rplumber.io/docs/hosting.html#docker)
-- [production deployment of neural networks](https://github.com/tmobile/r-tensorflow-api) [@jnolis, @nolistic]
+- [`AzureContainers`](https://cran.r-project.org/package=AzureContainers): Umbrella package for working with containers in Microsoft's Azure cloud.
+Provides interfaces to three Azure services: Container Registry, Container Instances (for running individual containers), and Kubernetes Services (for orchestrated deployments)
+
+## Using R to power enterprise software in production environments [@jnolis, @nolistic]
+
+R has been historically viewed as a tool for analysis and scientific research--not for creating software that corporations can rely on to continously run in real time.
+However, thanks to advancements in R running as a web service, along with with the ability to deploy R in docker containers, modern enterprises are now capable of having real-time machine learning powered by R.
+
+The backbone of this work is the _[plumber](https://www.rplumber.io/docs/hosting.html#docker)_ package, created by Jeff Allen.
+The plumber package allows R to be used to host web services through RESTful APIs.
+When R is running on a server, a different server can ask R to make a computation through an HTTP request and get an immediate response.
+For instance, if a data scientist at a company creates a machine learning model to tell which customers are likely to make a repeat purchase, if they serve the model through a plumber API then someone else on the marketing team could write software that sends different emails depending on that real-time prediction.
+Plumber syntax is simple to read and understand: the API is created by commenting above function names in an R script.
+
+```
+#* Return the sum of two numbers
+#* @get /random
+function(){
+  runif()
+}
+```
+_Example of a plumber API to generated a random number by visiting `http://127.0.0.1/random`_
+
+Plumber becomes far more powerful with the addition of Docker.
+Docker can be used to create images that [include both R and plumber](https://www.rplumber.io/docs/hosting.html#docker).
+Then when the image is deployed, that creates the web endpoints that other services can hit and call the R code.
+This pattern of deploying code mimics those used by software engineering services created in Java, Python, and other modern languages.
+By using Docker to deploy plumber APIs, R can be used alongside those langages as a first class member of a software engineering technical stack.
+
+At T-Mobile, R is used to power real-time services that are called over a million times a day.
+T-Mobile created a set of neural network machine learning natural language processing models to help customer care agents manage text-based messages for customers.
+The models are convolutional neural networks that use the RStudio Keras package and are built on top of a Rocker Docker image.
+Since the models power tools for agents and customers, they need to have extremely high uptime and reliability--and they found that R is able to perform.
+The AI @ T-Mobile team open sourced [their docker images](https://github.com/tmobile/r-tensorflow-api  which are available on GitHub.
+
+## Deployment to the cloud [@hongooi]
+
+While Plumber provides the low-level infrastructure for turning a statistical model into a predictive service, for production purposes it is usually necessary to take into account considerations such as scalability, reliability and ease of management.
+A single container is limited in the volume of requests it can service.
+Furthermore, if the container goes down, or is occupied with processing a computationally intensive task, the service becomes unavailable until it is restarted.
+
+These issues can be mitigated by deploying not a single container, but a _cluster_ of containers, orchestrated as a single deployment.
+The deployment system of choice at the time of writing is [Kubernetes](https://kubernetes.io).
+This allows you to deploy a single image to a pool of servers, which can then serve requests in parallel.
+Client-side requests are made to the cluster ingress endpoint, which then redirects them to individual servers.
+The orchestrator can identify cluster nodes that have failed, and restart them automatically; it can also detect when there is no capacity to service more requests, and start additional servers and containers as needed.
+
+While these load-balancing and autoscaling properties of a Kubernetes cluster are invaluable, they are still limited by the hardware resources that it has access to.
+A cluster cannot scale out if it has run out of servers.
+This issue in turn can be addressed by deploying not to on-premise servers but to the cloud, where effectively unlimited computing resources are available (as long as one can pay for them).
+From a management perspective, cloud deployments have the additional advantage that hardware maintenance becomes the responsibility of the cloud provider, eliminating another pain point.
+Orchestrator services are available from all the major cloud service providers, including Amazon Web Services (AWS), Google Cloud Platform, and Microsoft Azure.
+
+An example of an R package for working with containers in the cloud is AzureContainers, which interacts with three Azure services: Container Instances (for running a single container), Container Registry (a private docker registry service), and Kubernetes Service (for Kubernetes on Azure).
+AzureContainers provides a lightweight yet powerful interface for working with Resource Manager, the framework for deploying and managing arbitrary resources in Azure.
+On the client side, it provides simple shells to the Docker, Kubectl and Helm tools that are commonly used for managing containers and Kubernetes clusters.
+
+Here is some simple code that pushes a Docker image to an Azure Container Registry, and then creates a deployment using Azure Kubernetes Service.
+This is a slightly modified version of the code from the "Deploying a prediction service with Plumber" AzureContainers vignette.
+The code makes use of a yaml file to define the deployment and predictive service, which is standard practice with Kubernetes; the yaml can be seen in the aforementioned vignette.
+
+```r
+library(AzureContainers)
+# create a resource group for our deployments
+deployresgrp <- AzureRMR::get_azure_login()$
+    get_subscription("subscription_id")$
+    create_resource_group("deployresgrp", location="australiaeast")
+
+### create a container registry
+deployreg_svc <- deployresgrp$create_acr("deployreg")
+
+# build our image (a random forest on the Boston housing data, available in the MASS package)
+call_docker("build -t bos-rf .")
+
+# upload the image to Azure
+deployreg <- deployreg_svc$get_docker_registry()
+deployreg$push("bos-rf")
+
+### create a Kubernetes cluster with 2 nodes
+deployclus_svc <- deployresgrp$create_aks("deployclus", agent_pools=aks_pools("pool1", 2))
+
+# grant the cluster pull access to the registry
+gr <- AzureGraph::get_graph_login()
+aks_app_id <- deployclus$properties$servicePrincipalProfile$clientID
+reg$add_role_assignment(gr$get_app(aks_app_id), "Acrpull")
+
+# get the cluster endpoint
+deployclus <- deployclus_svc$get_cluster()
+
+# create and start the service
+deployclus$create("bos-rf.yaml")
+```
+
+Once the service has been created, we can check on its status and obtain predictions:
+
+```r
+deployclus$get("deployment bos-rf")
+#> Kubernetes operation: get deployment bos-rf  --kubeconfig=".../kubeconfigxxxx"
+#> NAME      DESIRED   CURRENT   UP-TO-DATE   AVAILABLE   AGE
+#> bos-rf    1         1         1            1           5m
+
+deployclus$get("service bos-rf-svc")
+#> Kubernetes operation: get service bos-rf-svc  --kubeconfig=".../kubeconfigxxxx"
+#> NAME         TYPE           CLUSTER-IP   EXTERNAL-IP     PORT(S)          AGE
+#> bos-rf-svc   LoadBalancer   10.0.8.189   xxx.xxx.xxx.xxx 8000:32276/TCP   5m
+
+response <- httr::POST("http://xxx.xxx.xxx.xxx:8000/score",
+    body=list(df=MASS::Boston[1:10,]), encode="json")
+httr::content(response, simplifyVector=TRUE)
+#> [1] 25.9269 22.0636 34.1876 33.7737 34.8081 27.6394 21.8007 22.3577 16.7812 18.9785
+```
+
+Note that Plumber, by itself, provides no authentication or security features; anybody who knows the cluster's IP address can access the predictive service.
+At minimum, a basic authentication layer should be added to any Kubernetes deployment.
+Standard Kubernetes procedure involves creating an ingress controller, for example with the nginx or traefik reverse proxy software, and one can then add application-specific authentication layers on top of that (for example, to authenticate with a user's Azure Active Directory or LDAP credentials).
+More details can be found from most Kubernetes learning resources, and are outside the scope of this paper.
 

### 2019-12-13 manuscript.Rmd
Auto stash before merge of "master" and "origin/master"

@@
     email: nitesh.turaga@roswellpark.org
+  - name: Jacqueline Nolis
+    affiliation: Nolis, LLC
+    address:
+      - Seattle, WA, USA
+      #- \orcid{NA}
+    email: jacqueline@jnolis.com
   - name: Hong Ooi

### 2019-12-13 manuscript.Rmd
add sentence about CZI grant for Rocker Project, closes #50

@@
 Coinciding with the rise of R was the advent of [Docker](https://en.wikipedia.org/wiki/Docker_(software)) as a general tool for distribution and deployment of software.
-Combining both these topics, the Rocker Project ([https://www.rocker-project.org/](https://www.rocker-project.org/)) provides images with R (see the next Section for more details).
+Combining both these topics, the _Rocker Project_ ([https://www.rocker-project.org/](https://www.rocker-project.org/)) provides images with R (see the next Section for more details).
 The considerable uptake and continued evolution of the Rocker Project has lead to numerous projects extending or building upon Rocker images, ranging from reproducible research to production deployments.
@@
 
-# Containerization and Rocker [NN]
+# Containerization and Rocker
 
@@
 Also of note is a series of "versioned" containers which match the R release they contain with the _then-current_ set of packages via the MRAN Snapshot views of CRAN \citep{microsoft_cran_2019}.
-[Maybe at a later point add forward reference to other topics building on Rocker described here.]
+The Rocker Project's impact and importance was acknowledged by the Chan Zuckerberg Initiative's _Essential Open Source Software for Science_, who provide findung for the projects's sustainable maintenance, community growth, and targeting new hardware platforms including GPUs \citep{chan_zuckerberg_initiative_maintaining_2019}.
 

### 2019-12-13 manuscript.Rmd
enterprise software section - one sentence per line, add reference

@@
 At T-Mobile, R is used to power real-time services that are called over a million times a day.
-T-Mobile created a set of neural network machine learning natural language processing models to help customer care agents manage text-based messages for customers.
+T-Mobile created a set of neural network machine learning natural language processing models to help customer care agents manage text-based messages for customers \citep{t-mobile_enterprise_2018}.
 The models are convolutional neural networks that use the RStudio Keras package and are built on top of a Rocker Docker image.
 Since the models power tools for agents and customers, they need to have extremely high uptime and reliability--and they found that R is able to perform.
-The AI @ T-Mobile team open sourced [their docker images](https://github.com/tmobile/r-tensorflow-api  which are available on GitHub.
+The AI @ T-Mobile team open sourced their Docker images [on GitHub](https://github.com/tmobile/r-tensorflow-api).
 

### 2019-12-16 manuscript.Rmd
update author list and contributions section

@@
   - name: Daniel Nüst
+    # https://github.com/nuest
     affiliation: University of Münster
@@
     email: daniel.nuest@uni-muenster.de
-  - name: Carl Boettiger
-    #affiliation: -
-    #email: -
   - name: Robrecht Cannoodt
+    # https://github.com/rcannood
     affiliation: Ghent University
@@
     email: robrecht@cannoodt.dev
-  - name: Ellis Hughes
   - name: Dirk Eddelbuettel
+    # https://github.com/eddelbuettel
     affiliation: University of Illinois at Urbana-Champaign
@@
   - name: Colin Fay
+    # https://github.com/ColinFay
     affiliation: ThinkR
@@
     email: contact@colinfay.me
-  - name: Ben Marwick
-    #affiliation: -
-    #email: -
   - name: Karthik Ram
+    # https://github.com/karthik
     affiliation: Berkeley Institute for Data Science
@@
   - name: Noam Ross
+    # https://github.com/noamross
     affiliation: EcoHealth Alliance
@@
   - name: Nan Xiao
+    # https://github.com/nanxstats
     affiliation: Seven Bridges Genomics
@@
   - name: Lori Shepherd
+    # https://github.com/lshep
     affiliation: Roswell Park Comprehensive Cancer Center
@@
   - name: Nitesh Turaga
+    # https://github.com/nturaga
     affiliation: Roswell Park Comprehensive Cancer Center
@@
   - name: Jacqueline Nolis
+    # https://github.com/jnolis/
     affiliation: Nolis, LLC
@@
   - name: Hong Ooi
+    # https://github.com/Hong-Revo
     affiliation: Microsoft
@@
     email: hongooi@microsoft.com
+  - name: Ellis Hughes
+    # https://github.com/thebioengineer
+    affiliation: Fred Hutchinson Cancer Research Center 
+    address:
+      - Vaccine and Infectious Disease
+      - 1100 Fairview Ave. N., P.O. Box 19024
+      - Seattle, WA 98109-1024, USA
+      #- \orcid{}
+    email: ehhughes@fredhutch.org
+  - name: Ben Marwick
 abstract: >
@@
 
-## Interfaces for Docker in R [Ellis Hughes, Robrecht Cannoodt]
+## Interfaces for Docker in R
 
@@
 
-# Discussion/outlook/conclusion [@all, please add bullet points]
+# Conclusions
 
@@
 
-DN conceived of the presented idea and \href{https://github.com/nuest/rockerverse-paper/issues/3}{initialised the formation of the writing team} and wrote the section about images for R distributions.
+DN conceived of the presented idea and \href{https://github.com/nuest/rockerverse-paper/issues/3}{initialised the formation of the writing team}, wrote the section about images for R distributions, wrote the conclusions section, and revised all sections.
 CB ..
-RC and RC wrote the section on interfaces for Docker in R.
-DE ..
+RC and EH wrote the section on interfaces for Docker in R.
+DE wrote the introduction and section about Rocker.
 ME ..
-CF wrote the section about `r-online`, `dockerfiler`, `r-ci` and `r-db`.
+CF wrote the paragraphs about `r-online`, `dockerfiler`, `r-ci` and `r-db`.
 BW ..
@@
 LS \& NT wrote the section on Bioconductor.
-All authors contributed to the discussion and outlook section and approved the final version.
+All authors contributed minor edits, gave feedback on other sections, and approved the final version.
+
 This articles was collaboratively written at \href{https://github.com/nuest/rockerverse-paper/}{https://github.com/nuest/rockerverse-paper/}.

### 2019-12-16 manuscript.Rmd
add missing references

@@
 The R community keeps growing.
-This can be seen in the number of new packages on CRAN which keeps on rising, but also in the increased number of conferences, online courses, meetups, and unconferences, as well as in the uptake in education and industry increases [REFs].
-All this cements the role of R as the _lingua franca_ of statistics, data visualisation, and computational research.
-Coinciding with the rise of R was the advent of [Docker](https://en.wikipedia.org/wiki/Docker_(software)) as a general tool for distribution and deployment of software.
-Combining both these topics, the _Rocker Project_ ([https://www.rocker-project.org/](https://www.rocker-project.org/)) provides images with R (see the next Section for more details).
+This can be seen in the number of new packages on CRAN, which keeps on growing exponentially \citep{cran:2019}, but also in the numbers of conferences, open educational resources, meetups, unconferences, and companies taking up, as exemplified by the useR!
+conference series\footnote{\href{https://www.r-project.org/conferences/}{https://www.r-project.org/conferences/}}, the global growth of the R and R-Ladies user groups\footnote{\href{https://www.r-consortium.org/blog/2019/09/09/r-community-explorer-r-user-groups}{https://www.r-consortium.org/blog/2019/09/09/r-community-explorer-r-user-groups}, \href{https://www.r-consortium.org/blog/2019/08/12/r-community-explorer}{https://www.r-consortium.org/blog/2019/08/12/r-community-explorer}}, or the foundation and impact of the R&nbsp;Consortium\footnote{\href{https://www.r-consortium.org/news/announcements}{https://www.r-consortium.org/news/announcements}, \href{https://www.r-consortium.org/blog/2019/11/14/data-driven-tracking-and-discovery-of-r-consortium-activities}{https://www.r-consortium.org/blog/2019/11/14/data-driven-tracking-and-discovery-of-r-consortium-activities}}.
+All this cements the role of R as the _lingua&nbsp;franca_ of statistics, data visualisation, and computational research.
+Coinciding with the rise of R was the advent of [Docker](https://en.wikipedia.org/wiki/Docker_(software)) as a general tool for distribution and deployment of server applications.
+Combining both these topics, the _Rocker&nbsp;Project_ ([https://www.rocker-project.org/](https://www.rocker-project.org/)) provides images with R (see the next Section for more details).
 The considerable uptake and continued evolution of the Rocker Project has lead to numerous projects extending or building upon Rocker images, ranging from reproducible research to production deployments.
-This article presents this _Rockerverse_ of projects across all development stages: demonstrations, prototypes, and mature products.
+This article presents this _Rockerverse_ of projects across all development stages: early demonstrations, working prototypes, and mature products.
 We also introduce related activities connecting the R language and environment with other containerisation solutions.
@@
 But Docker improved on virtualization by accessing the host system---generally Linux---through a much thinner and smaller shim than a full operating system emulation or virtualization.
-This makes for more efficient use of system resources and allowed another order of magnitude in terms of scalability of deployment [REFERENCE!].   
+This makes for more efficient use of system resources \citep{felter_updated_2015} and allowed another order of magnitude in terms of scalability of deployment \citep[cf.][]{datadog_8_2018}.
 While Docker makes use of Linux kernel features, its importance was large enough so that some required aspects of running Docker have been added to other operating systems to support Docker there more efficiently too \citep{microsoft_linux_2019}.
@@
 **[pqR](http://www.pqr-project.org/)** is a "a pretty quick version of R".
-_pqR_ fixes some opinionated issues in the R language and is the basis for experimental features, e.g.
-automatic diffentiation\footnote{ \href{https://riotworkshop.github.io/abstracts/riot-2019-pqr.txt}{https://riotworkshop.github.io/abstracts/riot-2019-pqr.txt}}.
+_pqR_ attempts to improve R on some opinionated issues in the R language and is the basis for experimental features, e.g.
+automatic diffentiation\footnote{\href{https://riotworkshop.github.io/abstracts/riot-2019-pqr.txt}{https://riotworkshop.github.io/abstracts/riot-2019-pqr.txt}}.
 The source code development [on GitHub](https://github.com/radfordneal/pqR/) is a one man project and it does not provide any Docker images.
@@
 **FastR** is "A high-performance implementation of the R programming language, built on GraalVM" (https://github.com/oracle/fastr).
-It is developed by Oracle, connects R to the GraalVM ecosystem \citep{wikipedia_graalvm_2019}, and also claims superior performance but also targets full compatibility with base R \footnote{ \href{https://github.com/oracle/fastr}{https://github.com/oracle/fastr}}.
+It is developed by Oracle, connects R to the GraalVM ecosystem \citep{wikipedia_graalvm_2019}, and also claims superior performance but also targets full compatibility with base R \footnote{\href{https://github.com/oracle/fastr}{https://github.com/oracle/fastr}}.
 There are no offical Docker images provided, but `nuest/fastr` images and `Dockerfile` are independently maintained on [Docker Hub](https://cloud.docker.com/repository/docker/nuest/fastr/) and [GitHub](https://github.com/nuest/fastr-docker) respectively.
@@
 That might seem trivial, but all the code using the random number generator will not be exactly reproducible after this breaking change.
-Controlling the version of software is crucial for reproducible research [REF].
+Therefore, controlling the version of software is most crucial for reproducible research \citep[e.g.][]{boettiger_introduction_2015}.
 
@@
 
-Using **GPUs** (graphical processing units) as a specialised hardware from containerized common work environments is also possible [REF].
-GPUs are increasingly popular for compute-intensive machine learning tasks [REF].
+Using **GPUs** (graphical processing units) as a specialised hardware from containerized common work environments is also possible and useful \citep{haydel_enhancing_2015}.
+GPUs are increasingly popular for compute-intensive machine learning tasks, e.g. deep artificial neural networks \citep{schmidhuber_deep_2015}.
 Though in this case, containers are not completely portable between hardware environments.

### 2019-12-18 manuscript.Rmd
Updated the author info for jacqueline and heather nolis

@@
       - Seattle, WA, USA
+      - \orcid{0000-0001-9354-6501}
+    email: jacqueline@nolisllc.com
+  - name: Heather Nolis
+    # https://github.com/nolistic/
+    affiliation: T-Mobile
+    address:
+      - 12920 Se 38th St.
+      - Bellevue, WA, 98006
       #- \orcid{NA}
-    email: jacqueline@jnolis.com
+    email: heather.wensler1@t-mobile.com
   - name: Hong Ooi

### 2019-12-18 manuscript.Rmd
add section on Windows, closes #32

--- 2019-12-16 manuscript.Rmd
@@
 
-## Windows Images [@nuest]
+## Windows Images
 
-- [rocker-win](https://github.com/nuest/rocker-win)
-- is possible, only relevant in organisations with an existing Windows Server-based infrastructure, can meet policies then
-<!-- can also be relevant for standalone Win 10 Professional users who have Docker for Windows installed -- HO -->
+Docker containers on the Windows operating system were originally quite cumbersome to use.
+They required an extra tool, [Docker Toolbox](https://docs.docker.com/toolbox/), which by now only exists as a legacy solution for older Windows systems.
+Docker toolbox leverages `docker-machine` (see also Section&nbsp;\ref{interfaces} for an R package interfacing with `docker-machine`) to handle the process of creating a local virtual host which could host Docker Engine, while exposing the regular Docker CLI.
+The Docker CLI commands are forwarded to the virtual host transparently for the user.
+
+For current Windows Server (2016 and later) and Windows Desktop ([Docker Desktop](https://www.docker.com/products/docker-desktop) requires Windows 10) versions, Docker is supported natively\footnote{\href{https://www.docker.com/products/windows-containers}{https://www.docker.com/products/windows-containers}} and different base images are offered by Microsoft\footnote{\href{https://docs.microsoft.com/en-us/virtualization/windowscontainers/about/index}{https://docs.microsoft.com/en-us/virtualization/windowscontainers/about/index}}.
+The Docker CLI can be used in just the same way as on other operating systems but not every base image is supported on every Windows host\footnote{\href{https://docs.microsoft.com/en-us/virtualization/windowscontainers/deploy-containers/version-compatibility}{https://docs.microsoft.com/en-us/virtualization/windowscontainers/deploy-containers/version-compatibility}}.
+On Docker Desktop for Windows, the user can run both Linux-based and Windows-based containers, but only one of the Docker daemons can be used at a time\footnote{See "Switch between Windows and Linux containers" on \href{https://docs.docker.com/docker-for-windows/}{https://docs.docker.com/docker-for-windows/}.}.
+
+[`rocker-win`](https://github.com/nuest/rocker-win) is a proof of concept for running R in Windows-based containers.
+It provides selected R versions using three different base images using `microsoft/windowsservercore` and `mcr.microsoft.com/windows/servercore`.
+These base images match different Windows versions: Windows Server 2019, Windows Server 2016, and Windows Server, version 1803.
+The images are built using two CI services, [Travis CI](https://docs.travis-ci.com/user/reference/windows/) and [Appveyor](https://www.appveyor.com/docs/windows-images-software/), which provide the different Windows versions and therefore support different base images, and published automatically [on Docker Hub](https://hub.docker.com/r/nuest/rocker-win).
+The images are built from manually maintained seperate Dockerfiles, are tagged with both Windows-variant and R version, e.g.
+`nuest/rocker-win:ltsc2019-3.6.2` or `nuest/rocker-win:1803-latest`, and run the `R.exe` process by default.
+New R versions are only added on demand.
+The explorative project demonstrates the variety and pontentially confusing complexity of Windows on Docker, e.g., because of the different base images, but also provides a starting point for any Windows-constrained user to leverage containerisation, including examples for apps using \pkg{plumber\, \pkg{Shiny}, or R packages with system dependencies such as \pkg{sf}.
+The latter is quite comfortable actually, as R ships pre-compiled binaries for Windows.
+
+For Windows, just as originally for Linux, the cloud use cases drive the development and Docker and Microsoft collaborate closely\footnote{\href{https://www.docker.com/partners/microsoft}{https://www.docker.com/partners/microsoft}}.
+In most cases, individual developers or researchers worrying about reproducibility, will prefer the more widely and host independent Linux-based containers.
+Naturally, a Windows license is required for the host machine.
+But the developments in the `rocker-win` prototype show that containers can also be used for R workflows depending on Windows-only tools, for leveraging an existing Windows Server-based infrastructure where policies can otherwise not be met, or for streamlining interactions with system operating staff that runs Windows-based containers themselves.
 

### 2019-12-18 manuscript.Rmd
add missing label and link for Fedora

@@
 - https://github.com/jlisic/R-docker-centos
+- Fedora > easy to install: https://github.com/ismailsunni/dockeRs
 
@@
 ## Interfaces for Docker in R
+\label{interfaces}
 

### 2020-01-09 manuscript.Rmd
add idea to conclusion, closes #2
add Rockerverse hex sticker

@@
 The main contributions is a coherent picture of the current lay of the land of using containers in, with, and for R.
+
+```{r hexSticker, message=TRUE, warning=FALSE, cache=TRUE, include=FALSE}
+if(!file.exists("rockerverse.png")) {
+  library("magick")
+  library("here")
+
+  # https://commons.wikimedia.org/wiki/File:NASA-HS201427a-HubbleUltraDeepField2014-20140603.jpg
+  universe <- magick::image_read("https://upload.wikimedia.org/wikipedia/commons/6/69/NASA-HS201427a-HubbleUltraDeepField2014-20140603.jpg")
+  magick::image_write(universe, here("universe.png"))
+  # rockerLogoMask <-
+  magick::image_read("https://avatars2.githubusercontent.com/u/9100160?s=200&v=4") %>%
+    magick::image_scale("2100x2100") %>%
+    magick::image_channel("Matte") %>%
+    image_negate() %>%
+    magick::image_border("black", "150x0") %>%
+    magick::image_write(here("rocker_mask.png"))
+
+  # does not work: image_composite(universe, rockerLogoMask, operator = "CopyOpacity", compose_args = "-composite -alpha Off")
+  system2("convert", c(here("universe.png"), here("rocker_mask.png"),
+    "-alpha Off",
+    "-compose CopyOpacity",
+    "-composite",
+    here("whaleverse.png")))
+
+  library("hexSticker")
+  hexSticker::sticker(here("whaleverse.png"), package = "Rockerverse",
+          p_size = 20, s_x = 0.97, s_y = 0.95, s_width = 1,
+          h_fill = "#000000", h_color = "#ffffff",
+          url = "github.com/nuest/rockerverse-paper",
+          u_color = "#ffffff", u_size = 4,
+          filename = here("rockerverse.png"))
+
+  file.remove(c(here("universe.png"), here("whaleverse.png"), here("rocker_mask.png")))
+}
+```
+
+![Rockerverse hex sticker](`r here::here("rockerverse.png")`)
 
@@
 
-- next steps
-  - consolidation?
+- what are next steps?
 - support more complex set-ups with ready-to-use stacks, e.g. GPU, AI, ML
@@
   - will there be control packages for Singularity support?
-- ...
+- sustain the overview and structure in this article in form of a CRAN Task View
 
@@
 
+# Acknowledgements
+
+DN is supported by the project Opening Reproducible Research ([o2r](https://www.uni-muenster.de/forschungaz/project/12343)) funded by the German Research Foundation (DFG) under project number [PE&nbsp;1632/17-1](https://gepris.dfg.de/gepris/projekt/415851837).
+The funders had no role in data collection and analysis, decision to publish, or preparation of the manuscript.
+
 \bibliography{RJreferences}

### 2020-01-09 manuscript.Rmd
add rocker-win Reference

@@
 
-[`rocker-win`](https://github.com/nuest/rocker-win) is a proof of concept for running R in Windows-based containers.
+[`rocker-win`](https://github.com/nuest/rocker-win) is a proof of concept for running R in Windows-based containers \citep{nust_rocker-win_2019}.
 It provides selected R versions using three different base images using `microsoft/windowsservercore` and `mcr.microsoft.com/windows/servercore`.
@@
 
-## Non-Debian Linux images [NN]
+## Non-Debian Linux images
 
@@
 - https://github.com/jlisic/R-docker-centos
-- Fedora > easy to install: https://github.com/ismailsunni/dockeRs
 
@@
 ## Interfaces for Docker in R
-\label{interfaces}
 
@@
 # inserting a table in the R Journal currently fails
-# waiting for this PR to merge
-# https://github.com/rstudio/rticles/pull/261
+# waiting for new release of articles including this PR to be published: https://github.com/rstudio/rticles/pull/261
+# Until then, use remotes::install_github("rstudio/rticles")
 if (packageVersion("rticles") >= "0.12.3") {

### 2020-01-10 manuscript.Rmd
remove rize, add dockr and renv, see #35

@@
 - [`containerit`](https://github.com/o2r-project/containerit/)
+- [`dockr`](https://github.com/smaakage85/dockr)
 - [`dockertest`](https://github.com/traitecoevo/dockertest/)
+- [`renv` with Docker](https://rstudio.github.io/renv/articles/docker.html)
 
@@
 As an example, _liftr_ demonstrated its advantage for R Markdown-based computational workflow orchestration, by effortlessly containerizing 18 complex _Bioconductor_ workflows in the DockFlow project (https://dockflow.org) in 2017.
-
-- [`rize` for Shiny](https://github.com/cole-brokamp/rize)
 

### 2020-01-10 manuscript.Rmd
Remove section headline and add overview sentence

@@
 The main contributions is a coherent picture of the current lay of the land of using containers in, with, and for R.
+
+The article continues with a brief introduction of containerization basics and the Rocker Project, followed by descriptions of projects developing images with R installations.
+Then, we present applications, starting with the R packages specificly for interacting with Docker, second-level packages using containers mediately or only for specific features, up to complex use cases leveraging containers.
+We conclude with a reflection on the landscape of packages and applications and point out future directions of development.
 
@@
 
-# R Packages
-<!-- from packages closer to Docker to second level packages to more complex ones -->
+# Use cases and applications
+<!-- from packages closer to Docker to second level packages to more complex use cases -->
 
@@
 The reason behind this is that `dockerfiler`, on top of being scriptable from R, can leverage all the tools available in R to parse a `DESCRIPTION` file, to get system requirements, to list dependencies, versions, etc.
-
-# Use cases and applications
 

### 2020-01-10 manuscript.Rmd
write non-debian images section, closes #33
also mention R-hub in development section

@@
 
-- Alpine images
-- [Images used by R-Hub](https://github.com/r-hub/rhub-linux-builders) (overlap with CI?)
-- https://github.com/jlisic/R-docker-centos
+\label{rhub}
+The **R-hub** project provides _"a collection of services to help R package development", with the package builder as the most prominent one\citep{r-hub_docs_2019}.
+The builder allows R package developers to check their R package on different platforms and R versions using a web form or the package \pkg{rhub} \citep{csardi_rhub_2019}.
+The builder uses Docker containers to conduct these checks, taking advantages of their well-defined environments and sandboxing.
+The `Dockerfile`s and images are published on Docker Hub\footnote{\href{https://hub.docker.com/u/rhub}{https://hub.docker.com/u/rhub}} and GitHub\footnote{\href{https://github.com/r-hub/rhub-linux-builders}{https://github.com/r-hub/rhub-linux-builders}} respectively.
+The images comprise release and development builds of R running in the base images of the [Debian](https://www.debian.org/), [Ubuntu](https://ubuntu.com/), [Fedora](https://getfedora.org/), and [CentOS](https://centos.org/) linux distributions (Arch Linux is under development\footnote{\href{https://github.com/r-hub/rhub-linux-builders/pull/41}{https://github.com/r-hub/rhub-linux-builders/pull/41}}, `see `rhub::local_check_linux_images()` for a full list), and are intended for debugging (see Section&nbsp;\ref{development}) not for reproducible workflows.
+_The rationale_ of not re-using the Rocker image stack is the high number of operatings systems and configurations, which can be more uniformly maintained in a seperate suite of images.
+Also, a common structure across the respective image stacks of the operating systems, e.g., for `debian`, adding GCC in `debian-gcc` and then R's development version in `debian-gcc-devel`.
+The platform also has a specific user configuration differing from the setup covering typical applications of Rocker images, and the images include considerably more software, such as a fairly complete LaTeX installation, to provide an experience closer to CRAN.
+
+There are also independent projects installing R or RStudio on [**Alpine Linux**](https://www.alpinelinux.org/), but not beyond the proof of concept stage\footnote{\href{https://github.com/artemklevtsov/r-alpine}{https://github.com/artemklevtsov/r-alpine} or \href{https://www.github.com/velaco/alpine-r}{https://www.github.com/velaco/alpine-r} provide different use case images but are not recently maintained (development of the latter documented in \citet{ratesic_building_2018}); \href{https://github.com/cmplopes/alpine-r}{https://github.com/cmplopes/alpine-r} has only base R images, is relatively up-to-date but sparsely documented.}.
+Alpine images are often used for minimalistic and thereby small and secure containers, e.g., in hardware with limited storage, though this advantage quickly deteriorates if more features are needed, e.g.
+if installing RStudio, package building envrionments, or powerful data science tools.
+Furthermore the distribution uses the [`musl` C library](https://www.musl-libc.org/) instead of the `glibc` used in Debian/Ubuntu, for which more tooling and experiences exist\footnote{Cf.
+\href{https://github.com/rocker-org/rocker/issues/231}{https://github.com/rocker-org/rocker/issues/231}}.
+\href{https://github.com/CenterForStatistics-UGent/mountainr}{https://github.com/CenterForStatistics-UGent/mountainr} only has a single and quite short `Dockerfile`, showing the easy installation of the latest R from Alpine sources.
+These images are used ...
+
+[`altRnative`](https://github.com/ismailsunni/altRnative/) is an experimental R package for running the same code across multiple containerised versions of R, intended for comparison across operating systems and different implementations of R.
+It comes with a collection of `Dockerfile`s and corresponding images \footnote{\href{https://github.com/ismailsunni/dockeRs}{https://github.com/ismailsunni/dockeRs}}.
 
@@
 ## Development and debugging
+\label{development}
 
@@
 
-- **[R-Hub](https://github.com/r-hub/)**
-
-# Other containerisation platforms
-
-Docker is not the only containerisation software.
-An alternative stemming from the domain of high-perfomance computing is Singularity \citep{kurtzer_singularity_2017}.
-Singularity can run Docker images, and in the case of Rocker works out of the box if the main process is R, e.g., in `rocker/r-base`, but does not succeed in running images where there is an init script, e.g.
-in containers that by default run RStudio.
-In the latter case, a `Singularity` file, a set of instructions akin to a `Dockerfile` needs to be used.
+The R-hub Docker images (see Section&nbsp;\ref{rhub}) can also be used to debug problems locally using various combinations of Linux platforms, R versions, and compilers\footnote{See \href{https://r-hub.github.io/rhub/articles/local-debugging.html}{https://r-hub.github.io/rhub/articles/local-debugging.html} and \href{https://blog.r-hub.io/2019/04/25/r-devel-linux-x86-64-debian-clang/}{https://blog.r-hub.io/2019/04/25/r-devel-linux-x86-64-debian-clang/}.
+The images go beyond the configurations, or _flavours_, used by CRAN for checking packages\footnote{\href{https://cran.r-project.org/web/checks/check_flavors.html}{https://cran.r-project.org/web/checks/check_flavors.html}}, e.g., with CentOS-based images, but lack a container for checking on Windows or OS X.
+The images greatly support package developers to provide support on operating systems they are not familiar.
 

### 2020-01-10 manuscript.Rmd
explain images stacks and Dockerfile

@@
 This is a rather attractive, and novel, proposition which has lead to widespread adoption and use of Docker in a variety of domains, _e.g._, cloud computing infrastructure \citep{Bernstein2014}, data science \citep{boettiger_introduction_2015}, and edge computing \citep{alam_orchestration_2018} [more examples!].
-It provided to be a natural match for "cloud deployment" which runs (or at least appears to run) "seamlessly" without much explicit reference to the underlying machine, architecture or operating system: containers can be deployed with very little in terms of dependencies on the host system---only the container runtime is required.
+It provided to be a natural match for "cloud deployment" which runs (or at least appears to run) "seamlessly" without much explicit reference to the underlying machine, architecture or operating system: containers are portable and can be deployed with very little in terms of dependencies on the host system---only the container runtime is required.
+Images are normally build from plain text documents called `Dockerfile`.
+A `Dockerfile` has a specific set of instructions to create and document a well-defined environment, i.e., install specific software and expose specific ports.
 
@@
 The Rocker Project provides several lines of containers spanning to from building blocks with `R-release` or `R-devel`, via containers with RStudio~Server and Shiny~Server, to domain-specific containers such as `geospatial`.
+These containers form _image stacks_, building on top of each other for better maintainability and 
 Also of note is a series of "versioned" containers which match the R release they contain with the _then-current_ set of packages via the MRAN Snapshot views of CRAN \citep{microsoft_cran_2019}.

### 2020-01-10 manuscript.Rmd
move section about non-Docker tools up

@@
 
-# Container Images
+Docker is not the only containerisation software.
+An alternative stemming from the domain of high-perfomance computing is **Singularity** \citep{kurtzer_singularity_2017}.
+Singularity can run Docker images, and in the case of Rocker works out of the box if the main process is R, e.g., in `rocker/r-base`, but does not succeed in running images where there is an init script, e.g., in containers that by default run RStudio.
+In the latter case, a `Singularity` file, a recipe akin to a `Dockerfile`, needs to be used.
+To date, no comparable image stack to Rocker exists on [Singularity Hub](https://singularity-hub.org/).
+A further tool for running containers is [**podman**](https://github.com/containers/libpod), which also can build `Dockerfile`s and run Docker images.
+Proof of concepts for using podman to build and run Rocker containers exist\footnote{See \href{https://github.com/nuest/rodman}{https://github.com/nuest/rodman} and \href{https://github.com/rocker-org/rocker-versioned/issues/187}{https://github.com/rocker-org/rocker-versioned/issues/187}}.
+Yet the prevelance of Docker, especially in the broader user community beyond experts or niche systems, and the vast amount of blog posts and courses for Docker, currently caps specific development efforts for both Singularity and podman in the R community.
+This might quickly change when usability and spread increase, or security features such as rootless/unprivileged containers, which both these tools support out of the box, become more sought after.
+
+# Container images
 

### 2020-01-10 manuscript.Rmd
editorial and formatting changes

@@
 [http://packages.renjin.org/packages](http://packages.renjin.org/packages) for available packages.
-Packages are loaded on demand, i.e. at the first call to `library()`.
-Not all R packages, especially one linking to binary libraries, are available, e.g.
-`rgdal` \footnote{\href{http://packages.renjin.org/package/org.renjin.cran/rgdal/1.4-4/build/1}{http://packages.renjin.org/package/org.renjin.cran/rgdal/1.4-4/build/1}}.
+Packages are loaded on demand, i.e., at the first call to `library()`.
+Not all R packages, especially one linking to binary libraries, are available, e.g., `rgdal` \footnote{\href{http://packages.renjin.org/package/org.renjin.cran/rgdal/1.4-4/build/1}{http://packages.renjin.org/package/org.renjin.cran/rgdal/1.4-4/build/1}}.
 There are no offical Docker images for Renjin, but community-maintained images for selected releases only are available under `nuest/renjin` on Docker Hub and GitHub at [https://hub.docker.com/r/nuest/renjin](https://hub.docker.com/r/nuest/renjin) and [https://github.com/nuest/renjin-docker](https://github.com/nuest/renjin-docker) respectively.
@@
 **[pqR](http://www.pqr-project.org/)** is a "a pretty quick version of R".
-_pqR_ attempts to improve R on some opinionated issues in the R language and is the basis for experimental features, e.g.
-automatic diffentiation\footnote{\href{https://riotworkshop.github.io/abstracts/riot-2019-pqr.txt}{https://riotworkshop.github.io/abstracts/riot-2019-pqr.txt}}.
+_pqR_ attempts to improve R on some opinionated issues in the R language and is the basis for experimental features, e.g., automatic diffentiation\footnote{\href{https://riotworkshop.github.io/abstracts/riot-2019-pqr.txt}{https://riotworkshop.github.io/abstracts/riot-2019-pqr.txt}}.
 The source code development [on GitHub](https://github.com/radfordneal/pqR/) is a one man project and it does not provide any Docker images.
@@
 To simplify building and maintaining _Bioconductor_ images, we use a Ruby templating engine.
-A recent audit of the _Bioconductor_ Dockerfiles, following best practices from the Docker website, led to a reduction in the size and number of layers.
+A recent audit of the _Bioconductor_ `Dockerfile`s, following best practices from the Docker website, led to a reduction in the size and number of layers.
 The most important insights involve the ['union' file system used by Docker](https://docs.docker.com/storage/storagedriver/overlayfs-driver/#how-container-reads-and-writes-work-with-overlay-or-overlay2).
@@
 The Rocker images provides the _versioned stack_ for different the version of R since the project inception, which use the _then-current_ stable Debian base image \citep[cf.][]{RJ-2017-065}.
-Based on custom build phase hooks\footnote{\href{https://docs.docker.com/docker-hub/builds/advanced/}{https://docs.docker.com/docker-hub/builds/advanced/}}, i.e.
-small shell scripts executed at different phases in the automated build of Rocker images, there are also semantic version tags for the most recent freezed (i.e.
-using MRAN) versions with only the major and minor version\footnote{See \href{https://github.com/rocker-org/rocker-versioned/blob/master/VERSIONS.md}{https://github.com/rocker-org/rocker-versioned/blob/master/VERSIONS.md}, FIXME: \href{https://github.com/rocker-org/rocker-versioned/issues/42}{https://github.com/rocker-org/rocker-versioned/issues/42}.}.
+Based on custom build phase hooks\footnote{\href{https://docs.docker.com/docker-hub/builds/advanced/}{https://docs.docker.com/docker-hub/builds/advanced/}}, i.e., small shell scripts executed at different phases in the automated build of Rocker images, there are also semantic version tags for the most recent freezed (i.e., using MRAN) versions with only the major and minor version\footnote{See \href{https://github.com/rocker-org/rocker-versioned/blob/master/VERSIONS.md}{https://github.com/rocker-org/rocker-versioned/blob/master/VERSIONS.md}, FIXME: \href{https://github.com/rocker-org/rocker-versioned/issues/42}{https://github.com/rocker-org/rocker-versioned/issues/42}.}.
 So, at the time of writing this article, `rocker/r-ver:3` and `3.6` are aliases for `3.6.0`, because `3.6.1` is the latest release.
@@
 The Docker CLI commands are forwarded to the virtual host transparently for the user.
-
 For current Windows Server (2016 and later) and Windows Desktop ([Docker Desktop](https://www.docker.com/products/docker-desktop) requires Windows 10) versions, Docker is supported natively\footnote{\href{https://www.docker.com/products/windows-containers}{https://www.docker.com/products/windows-containers}} and different base images are offered by Microsoft\footnote{\href{https://docs.microsoft.com/en-us/virtualization/windowscontainers/about/index}{https://docs.microsoft.com/en-us/virtualization/windowscontainers/about/index}}.
@@
 
-[`rocker-win`](https://github.com/nuest/rocker-win) is a proof of concept for running R in Windows-based containers \citep{nust_rocker-win_2019}.
+**[`rocker-win`](https://github.com/nuest/rocker-win)** is a proof of concept for running R in Windows-based containers \citep{nust_rocker-win_2019}.
 It provides selected R versions using three different base images using `microsoft/windowsservercore` and `mcr.microsoft.com/windows/servercore`.
@@
 The images are built using two CI services, [Travis CI](https://docs.travis-ci.com/user/reference/windows/) and [Appveyor](https://www.appveyor.com/docs/windows-images-software/), which provide the different Windows versions and therefore support different base images, and published automatically [on Docker Hub](https://hub.docker.com/r/nuest/rocker-win).
-The images are built from manually maintained seperate Dockerfiles, are tagged with both Windows-variant and R version, e.g.
-`nuest/rocker-win:ltsc2019-3.6.2` or `nuest/rocker-win:1803-latest`, and run the `R.exe` process by default.
+The images are built from manually maintained seperate `Dockerfile`s, are tagged with both Windows-variant and R version, e.g., `nuest/rocker-win:ltsc2019-3.6.2` or `nuest/rocker-win:1803-latest`, and run the `R.exe` process by default.
 New R versions are only added on demand.
@@
 Each of these packages has particular advantages as they provide function wrappers for interacting with the Docker CLI at different stages of a container's life cycle.
-Examples of such interactions are installing the Docker software, creating Dockerfiles, building an image, and launching a container on a local machine or on the cloud.
+Examples of such interactions are installing the Docker software, creating `Dockerfile`s, building an image, and launching a container on a local machine or on the cloud.
 As such, the choice of which package is most useful depends on the use-case at hand.
@@
 
-**dockyard** ([https://github.com/thebioengineer/dockyard](https://github.com/thebioengineer/dockyard)) has the goal of lowering barrier to creating dockerfiles and docker images, and deploying docker containers.
-The way docker images are made is by reading and following a set of instructions found in dockerfiles/
 This requires stepping out of the R ecosystem and into an infrastructure mindset where the user needs to specify the environment they require.
-Because dockerfiles can be read as a series of steps, this lends itself well to the `tidyverse` style of programming where each step is piped into the next.
-dockyard has a 'dockerfile' object that either starts out as an empty dockerfile, or can use existing dockerfile to use as a template.
-This combined with a series of functions that are analogous to the steps commonly found in a dockerfile, it makes building docker images much more approachable to R users.
-There are also wrappers for commonly used steps for R programmers like adding libraries and copying files into the R docker containers.
-Once the dockerfile is created, dockyard has the functionality to build the image and deploy it.
+Because `Dockerfile`s can be read as a series of steps, this lends itself well to the `tidyverse` style of programming where each step is piped into the next.
+dockyard has a `Dockerfile` object that either starts out as an empty `Dockerfile`, or can use existing `Dockerfile` to use as a template.
+This combined with a series of functions that are analogous to the steps commonly found in a `Dockerfile`, it makes building Docker images much more approachable to R users.
+There are also wrappers for commonly used steps for R programmers like adding libraries and copying files into the R Docker containers.
+Once the `Dockerfile` is created, dockyard has the functionality to build the image and deploy it.
 
@@
 
-One of the key tool of Docker are dockerfiles, which can be thought of as "recipes" used to build a specific image. 
+One of the key tool of Docker are `Dockerfile`s, which can be thought of as "recipes" used to build a specific image. 
 But building these files can feel like a workflow break as it demands to open a separate file, and / or to write a small wrapper around R's `write()` function to add elements to the file. 
 At the same time, doing it by hand prevents from using programming language for what they are good for: iteration, and automation. 
-Iteration and automation for Dockerfiles is the very reason behind [`dockerfiler`](https://github.com/ColinFay/dockerfiler), an R package designed for building Dockerfiles straigth from R.
+Iteration and automation for `Dockerfile`s is the very reason behind [`dockerfiler`](https://github.com/ColinFay/dockerfiler), an R package designed for building `Dockerfile`s straight from R.
 
@@
 - [`googleComputeEngineR`](https://cloudyr.github.io/googleComputeEngineR/) (function `docker_run`)
-  - I don't think `docker_run` is main focus for containers for this package, its more how it uses Docker to launch custom environments that can be templated (e.g.
-rstudio, shiny etc.) and for parrallisation e.g.
-`library(future)`  using GCE VMs as endpoints.
+  - I don't think `docker_run` is main focus for containers for this package, its more how it uses Docker to launch custom environments that can be templated (e.g., rstudio, shiny etc.) and for parrallisation e.g., `library(future)`  using GCE VMs as endpoints.
 [@MarkEdmondson1234]
@@
 
-The controlled nature of containers, i.e.
-it is possible to define the software environment very well, even on remote machines, make them also useful for continuous integration (CI) and continuous delivery of applications.
+The controlled nature of containers, i.e., it is possible to define the software environment very well, even on remote machines, make them also useful for continuous integration (CI) and continuous delivery of applications.
 
@@
   - https://cloud.google.com/cloud-build/
-  - For me this is what makes Docker containers viable, as it builds the Dockerfiles on each GitHub commit.
+  - For me this is what makes Docker containers viable, as it builds the `Dockerfile`s on each GitHub commit.
 It couples with Google Container Registry to build private Docker images, for downstream applications.
@@
 Binder's core internal tool is [repo2docker](https://repo2docker.readthedocs.io/en/latest/config_files.html).
-It deterministically builds a Docker image by parsing the contents of a repository, e.g. project dependency configurations or simple configuration files.
-In the most powerful case, `repo2docker` builds a given Dockerfile.
+It deterministically builds a Docker image by parsing the contents of a repository, e.g., project dependency configurations or simple configuration files.
+In the most powerful case, `repo2docker` builds a given `Dockerfile`.
 While this approach works well for most run of the mill Python projects, it is not so seamless for R projects.
-For any R projects that use the Tidyverse suite [citation in review, will add by the time this gets submitted], the time and resources required to build all dependencies from source can often time out before completion, making it frustrating for the average R user.
+For any R projects that use the Tidyverse suite \citep{wickham_welcome_2019}, the time and resources required to build all dependencies from source can often time out before completion, making it frustrating for the average R user.
 `holepunch` removes some of these limitations by leveraging Rocker images that contain the Tidyverse along special Jupyter dependencies, and only installs additional packages from CRAN and Bioconductor that are not already part of these images.
@@
 `holepunch` as a side effect also creates a `DESCRIPTION` file which then turns any project into a research compendium \citep{marwick_packaging_2018}.
-The Dockerfile included with the project can also be used to launch a RStudio server locally independent of Binder which is especially useful when more or special computational resources are required.
-
-[`holepunch`](https://github.com/karthik/holepunch) is a R package that was designed to remove some of these limitations and make Binder more accessible to novice R users.
-`holepunch` achieves this by leveraging Rocker images that contains the Tidyverse along special Jupyter dependencies, and only installs additional packages from CRAN and Bioconductor that are not already part of these images.
-Starting with the Binder/Tidyverse base images eliminates a large part of the build time and in most cases results in a binder instance launching within a minute.
-`holepunch` as a side effect also creates a DESCRIPTION file which then turns any project into a research compendium \citep{marwick_packaging_2018}.
-The Dockerfile included with the project can also be used to launch a RStudio server locally independent of binder which is especially useful when more computational resources are required.
+The `Dockerfile` included with the project can also be used to launch a RStudio server locally, i.e., independent of Binder, which is especially useful when more or special computational resources can be provided there.
+The local image usage reduces the number of seperately managed environments and thereby reduces work and increases portability and reproducibility.
 
@@
 Using **GPUs** (graphical processing units) as a specialised hardware from containerized common work environments is also possible and useful \citep{haydel_enhancing_2015}.
-GPUs are increasingly popular for compute-intensive machine learning tasks, e.g. deep artificial neural networks \citep{schmidhuber_deep_2015}.
+GPUs are increasingly popular for compute-intensive machine learning tasks, e.g., deep artificial neural networks \citep{schmidhuber_deep_2015}.
 Though in this case, containers are not completely portable between hardware environments.
@@
 - `batchtools` \citep{Lang2017batchtools} can [schedule jobs with Docker Swarm](https://mllg.github.io/batchtools/reference/makeClusterFunctionsDocker.html)
-- scalable deployments, e.g. start with numerous Shiny talks mentioning Rocker at useR!2017
+- scalable deployments, e.g., start with numerous Shiny talks mentioning Rocker at useR!2017
 - [dynmethods](https://github.com/dynverse/dynmethods) [@rcannood]: In order to evaluate ±50 computational methods which all used different environments (R, Python, C++, ...), we wrapped each of them in a docker container and can execute these methods from R.
@@
 - what are next steps?
-- support more complex set-ups with ready-to-use stacks, e.g. GPU, AI, ML
+- support more complex set-ups with ready-to-use stacks, e.g., GPU, AI, ML
 - Missing pieces?
-- consolidation (e.g. via packages using `dockerfiler` and `stevedore`)
+- consolidation (e.g., via packages using `dockerfiler` and `stevedore`)
 - Common themes

### 2020-01-10 manuscript.Rmd
add list of not included projects to README

@@
 
-<!--
-**docker** ([https://bhaskarvk.github.io/docker](https://bhaskarvk.github.io/docker)
-docker has been removed from CRAN, seems unmaintained by authors. Maybe it should be removed from this list?
--->
-
+**dockyard** ([https://github.com/thebioengineer/dockyard](https://github.com/thebioengineer/dockyard)) has the goal of lowering barrier to creating `Dockerfile`s and Docker images, and deploying Docker containers.
+The way docker images are made is by reading and following a set of instructions found in `Dockerfile`s.
 This requires stepping out of the R ecosystem and into an infrastructure mindset where the user needs to specify the environment they require.

### 2020-01-10 manuscript.Rmd
fix footnote, no urldate for articles with DOI

@@
 The R-hub Docker images (see Section&nbsp;\ref{rhub}) can also be used to debug problems locally using various combinations of Linux platforms, R versions, and compilers\footnote{See \href{https://r-hub.github.io/rhub/articles/local-debugging.html}{https://r-hub.github.io/rhub/articles/local-debugging.html} and \href{https://blog.r-hub.io/2019/04/25/r-devel-linux-x86-64-debian-clang/}{https://blog.r-hub.io/2019/04/25/r-devel-linux-x86-64-debian-clang/}.
-The images go beyond the configurations, or _flavours_, used by CRAN for checking packages\footnote{\href{https://cran.r-project.org/web/checks/check_flavors.html}{https://cran.r-project.org/web/checks/check_flavors.html}}, e.g., with CentOS-based images, but lack a container for checking on Windows or OS X.
+The images go beyond the configurations, or _flavours_, used by CRAN for checking packages\footnote{\href{https://cran.r-project.org/web/checks/check_flavors.html}{https://cran.r-project.org/web/checks/check\_flavors.html}}, e.g., with CentOS-based images, but lack a container for checking on Windows or OS X.
 The images greatly support package developers to provide support on operating systems they are not familiar.

### 2020-01-10 manuscript.Rmd
add data science images section

@@
 It comes with a collection of `Dockerfile`s and corresponding images \footnote{\href{https://github.com/ismailsunni/dockeRs}{https://github.com/ismailsunni/dockeRs}}.
+
+## Data Science images
+
+The notion of _Data Science_ is widely discussed among all academic disciplines.
+Using computers and complex software to make sense of (large amounts of) collected or simulated data is a cross-cutting phenomenon leading to _X Data Science_ disciplines, seminars, and study majors, there are also Docker images with commonly used tools and software stacks pre-installed that are often used in combination for analysing data.
+A large number of bespoke computing environments are created this way.
+Here we only mention a few image stacks as examples that are not based on Rocker images but prominently include installations of R.
+
+The [**Jupyter Docker Stacks** project](https://github.com/jupyter/docker-stacks/) are a set of ready-to-run Docker images containing Jupyter applications and interactive computing tools \citep{project_jupyter_jupyter_2018}.
+The `jupyter/r-notebook` image includes R and "popular packages", and R is also included in the catchall `datascience-notebook` image\footnote{\href{https://jupyter-docker-stacks.readthedocs.io/en/latest/using/selecting.html}{https://jupyter-docker-stacks.readthedocs.io/en/latest/using/selecting.html}}.
+For example, these images allow users to quickly start a Jupyter Notebook server locally or build their own specialised images on top of stable toolsets.
+
+**Kaggle** provides the [`gcr.io/kaggle-images/rstats`](https://hub.docker.com/r/kaggle/rstats) image (previously `kaggle/rstats`) and [corresponding `Dockerfile`](https://github.com/Kaggle/docker-rstats) for usage in their Machine Learning competitions and easy access to the associated datasets.
+It includes machine learling libraries such as Tensorflow and Keras, and also configures the \pkg{reticulate} package.
+The image uses a base image with _all packages from CRAN_, `gcr.io/kaggle-images/rcran`, which requires a Google Cloud Build as Docker Hub would time out\footnote{Originally, a stacked collection of over 20 images with automated builds on Docker Hub was used, see \href{https://web.archive.org/web/20190606043353/http://blog.kaggle.com/2016/02/05/how-to-get-started-with-data-science-in-containers/}{https://web.archive.org/web/20190606043353/http://blog.kaggle.com/2016/02/05/how-to-get-started-with-data-science-in-containers/} and \href{https://hub.docker.com/r/kaggle/rcran/dockerfile}{https://hub.docker.com/r/kaggle/rcran/dockerfile}}.
+The final extracted image size is over 25GB, which makes it debatable if having everything available is actually convenient.
+
+```{bash kaggle_images, eval=FALSE, include=FALSE}
+docker pull gcr.io/kaggle-images/rcran
+docker pull kaggle/rcran # over 22 months old
+docker pull gcr.io/kaggle-images/rstats
+docker images | grep kaggle
+```
+
+As a further example, [**Radiant** project](https://radiant-rstats.github.io/docs/) provides several images, e.g., [`vnijs/rsm-msba-spark`](https://hub.docker.com/r/vnijs/rsm-msba-spark), for their browser-based business analytics interface based on \pkg{Shiny} (`Dockerfile` [on GitHub](https://github.com/radiant-rstats/docker)) and for use in education as part of an MSc course.
+As data science often applies a multitude of tools, this image favours inclusion over selection and features Python, Postgres, JupyterLab and Visual Studio Code besides R and RStudio, bringing the image size up to 9GB.
+
+```{bash radiant_images, eval=FALSE, include=FALSE}
+docker pull vnijs/rsm-msba-spark
+docker images | grep vnijs
+```
 

### 2020-01-10 manuscript.tex
update manuscript

--- 2019-10-18 manuscript.tex
@@
 % !TeX root = RJwrapper.tex
-\title{Containers and R: the Rockerverse and beyond}
-\author{by Daniel Nüst, Carl Boettiger, Robrecht Cannoodt, Dirk Eddelbuettel, Mark Edmondson, Colin Fay, Ben Marwick, Karthik Ram, Noam Ross, Nan Xiao, Lori Shepherd, Nitesh Turaga}
+\title{Containers and R---the Rockerverse and beyond}
+\author{by Daniel Nüst, Robrecht Cannoodt, Dirk Eddelbuettel, Mark Edmondson, Colin Fay, Karthik Ram, Noam Ross, Nan Xiao, Lori Shepherd, Nitesh Turaga, Jacqueline Nolis, Heather Nolis, Hong Ooi, Ellis Hughes, Ben Marwick}
 
@@
 
-The R community keeps growing: the number of new packages on CRAN keeps
-on rising, meetups, conferences, online courses and unconferences
-prosper, and the uptake in education and industry increases {[}REFs{]}.
-All this cements the role of R as the lingua franca of statistics, data
-visualisation, and computational research. Coinciding with the rise of R
-was the advent of Docker {[}CITATION?{]} as a general tool for
-development, distribution, and deployment of software. Combining both
-these topics, the \emph{Rocker
-Project}(\url{https://www.rocker-project.org/}) was introduced in 2014
-and provides a number of
-\href{https://en.wikipedia.org/wiki/Docker_(software)}{Docker} images
-for various use cases as described in \citet{RJ-2017-065}. The
-considerable uptake and continued evolution of the \emph{Rocker} suite
-of containers has lead to numerous projects extending or building upon
-\emph{Rocker} images ranging from reproducible research to production
-deployments. This article presents this \emph{Rockerverse} of projects.
-It also introduces related activities connecting the R language and
-environment with other containerisation solutions. The main
-contributions is a coherent picture of the current lay of the land of
-containers in, with, and for \texttt{R}. This diversity includes
-demonstrators, early prototypes, and mature projects.
+The R community keeps growing. This can be seen in the number of new
+packages on CRAN, which keeps on growing exponentially
+\citep{cran:2019}, but also in the numbers of conferences, open
+educational resources, meetups, unconferences, and companies taking up,
+as exemplified by the useR! conference
+series\footnote{\href{https://www.r-project.org/conferences/}{https://www.r-project.org/conferences/}},
+the global growth of the R and R-Ladies user
+groups\footnote{\href{https://www.r-consortium.org/blog/2019/09/09/r-community-explorer-r-user-groups}{https://www.r-consortium.org/blog/2019/09/09/r-community-explorer-r-user-groups}, \href{https://www.r-consortium.org/blog/2019/08/12/r-community-explorer}{https://www.r-consortium.org/blog/2019/08/12/r-community-explorer}},
+or the foundation and impact of the
+R~Consortium\footnote{\href{https://www.r-consortium.org/news/announcements}{https://www.r-consortium.org/news/announcements}, \href{https://www.r-consortium.org/blog/2019/11/14/data-driven-tracking-and-discovery-of-r-consortium-activities}{https://www.r-consortium.org/blog/2019/11/14/data-driven-tracking-and-discovery-of-r-consortium-activities}}.
+All this cements the role of R as the \emph{lingua~franca} of
+statistics, data visualisation, and computational research. Coinciding
+with the rise of R was the advent of
+\href{https://en.wikipedia.org/wiki/Docker_(software)}{Docker} as a
+general tool for distribution and deployment of server applications.
+Combining both these topics, the \emph{Rocker~Project}
+(\url{https://www.rocker-project.org/}) provides images with R (see the
+next Section for more details). The considerable uptake and continued
+evolution of the Rocker Project has lead to numerous projects extending
+or building upon Rocker images, ranging from reproducible research to
+production deployments. This article presents this \emph{Rockerverse} of
+projects across all development stages: early demonstrations, working
+prototypes, and mature products. We also introduce related activities
+connecting the R language and environment with other containerisation
+solutions. The main contributions is a coherent picture of the current
+lay of the land of using containers in, with, and for R.
+
+The article continues with a brief introduction of containerization
+basics and the Rocker Project, followed by descriptions of projects
+developing images with R installations. Then, we present applications,
+starting with the R packages specificly for interacting with Docker,
+second-level packages using containers mediately or only for specific
+features, up to complex use cases leveraging containers. We conclude
+with a reflection on the landscape of packages and applications and
+point out future directions of development.
+
+\begin{figure}
+\centering
+\includegraphics{/home/daniel/git/o2r/rockerverse-paper/rockerverse.png}
+\caption{Rockerverse hex sticker}
+\end{figure}
 
@@
 
-\emph{Do we need a generic intro here?} DE: Yes, and happy to take a
-stab.
+Docker, an application and service provide by the eponymous company, has
+in just a few short years risen to prominence for development, testing,
+deployment and distribution of computer software
+\citep[cf.][]{datadog_8_2018,munoz_history_2019}. While there are
+related approaches such as
+LXC\footnote{\href{https://en.wikipedia.org/wiki/LXC}{https://en.wikipedia.org/wiki/LXC}}
+or Singularity \citep{kurtzer_singularity_2017}, Docker has become
+synomymous with ``containerization''---the method of taking software
+artefacts and bundling them in such a way that use becomes standardized
+and portable across operationg systems. In doing so, Docker had
+recognised and validated the importance of one very important thread
+that had been emerging, namely virtualization. By allowing (one or
+possibly) multiple applications or services to run concurrently on one
+host machine without any fear of interference between them, an important
+scalability opportunity is being provided. But Docker improved on
+virtualization by accessing the host system---generally Linux---through
+a much thinner and smaller shim than a full operating system emulation
+or virtualization. This makes for more efficient use of system resources
+\citep{felter_updated_2015} and allowed another order of magnitude in
+terms of scalability of deployment \citep[cf.][]{datadog_8_2018}. While
+Docker makes use of Linux kernel features, its importance was large
+enough so that some required aspects of running Docker have been added
+to other operating systems to support Docker there more efficiently too
+\citep{microsoft_linux_2019}. The success even lead to standardisation
+and industry collaboration \citep{oci_open_2019}.
+
+The key accomplishment of Docker (as an ``application'') is to make a
+``bundled'' aggregation of software (the so-called ``image'') available
+to any system equipped to run Docker, without requiring much else from
+the host besides the actual Docker application installation. This is a
+rather attractive, and novel, proposition which has lead to widespread
+adoption and use of Docker in a variety of domains, \emph{e.g.}, cloud
+computing infrastructure \citep{Bernstein2014}, data science
+\citep{boettiger_introduction_2015}, and edge computing
+\citep{alam_orchestration_2018} {[}more examples!{]}. It provided to be
+a natural match for ``cloud deployment'' which runs (or at least appears
+to run) ``seamlessly'' without much explicit reference to the underlying
+machine, architecture or operating system: containers are portable and
+can be deployed with very little in terms of dependencies on the host
+system---only the container runtime is required. Images are normally
+build from plain text documents called \texttt{Dockerfile}. A
+\texttt{Dockerfile} has a specific set of instructions to create and
+document a well-defined environment, i.e., install specific software and
+expose specific ports.
+
+For statistical computing and analysis centered around R, the Rocker
+Project has provided a variety of Docker containers since its start in
+2014 \citep{RJ-2017-065}. The Rocker Project provides several lines of
+containers spanning to from building blocks with \texttt{R-release} or
+\texttt{R-devel}, via containers with RStudio\textasciitilde{}Server and
+Shiny\textasciitilde{}Server, to domain-specific containers such as
+\texttt{geospatial}. These containers form \emph{image stacks}, building
+on top of each other for better maintainability and Also of note is a
+series of ``versioned'' containers which match the R release they
+contain with the \emph{then-current} set of packages via the MRAN
+Snapshot views of CRAN \citep{microsoft_cran_2019}. The Rocker Project's
+impact and importance was acknowledged by the Chan Zuckerberg
+Initiative's \emph{Essential Open Source Software for Science}, who
+provide findung for the projects's sustainable maintenance, community
+growth, and targeting new hardware platforms including GPUs
+\citep{chan_zuckerberg_initiative_maintaining_2019}.
+
+Docker is not the only containerisation software. An alternative
+stemming from the domain of high-perfomance computing is
+\textbf{Singularity} \citep{kurtzer_singularity_2017}. Singularity can
+run Docker images, and in the case of Rocker works out of the box if the
+main process is R, e.g., in \texttt{rocker/r-base}, but does not succeed
+in running images where there is an init script, e.g., in containers
+that by default run RStudio. In the latter case, a \texttt{Singularity}
+file, a recipe akin to a \texttt{Dockerfile}, needs to be used. To date,
+no comparable image stack to Rocker exists on
+\href{https://singularity-hub.org/}{Singularity Hub}. A further tool for
+running containers is
+\href{https://github.com/containers/libpod}{\textbf{podman}}, which also
+can build \texttt{Dockerfile}s and run Docker images. Proof of concepts
+for using podman to build and run Rocker containers
+exist\footnote{See \href{https://github.com/nuest/rodman}{https://github.com/nuest/rodman} and \href{https://github.com/rocker-org/rocker-versioned/issues/187}{https://github.com/rocker-org/rocker-versioned/issues/187}}.
+Yet the prevelance of Docker, especially in the broader user community
+beyond experts or niche systems, and the vast amount of blog posts and
+courses for Docker, currently caps specific development efforts for both
+Singularity and podman in the R community. This might quickly change
+when usability and spread increase, or security features such as
+rootless/unprivileged containers, which both these tools support out of
+the box, become more sought after.
 
 \hypertarget{container-images}{%
-\section{Container Images}\label{container-images}}
+\section{Container images}\label{container-images}}
 
@@
 
-\emph{Microsoft R Open}
+\textbf{Microsoft R Open}
 (\href{https://github.com/nuest/mro-docker}{MRO}) is an R distribution
 formerly known as Revolution R Open (RRO) before Revolution Analytics
-was acquired by Microsoft. MRO is compatible with main R and it's
-packages.sss ``It includes additional capabilities for improved
+was acquired by Microsoft. MRO is compatible with main R and its
+packages. ``It includes additional capabilities for improved
 performance, reproducibility, and platform support.''
 \citep{microsoft_mro} Most notably these capabilities are the MRAN
-repository a.k.a. CRAN Time Machine, which is enabled by default, and
-the (optional) integration with
-\href{https://software.intel.com/en-us/mkl}{Intel® Math Kernel Library}
-(MKL) for multi-threading in linear algebra operations
+repository, which is enabled by default, and the (optional) integration
+with \href{https://software.intel.com/en-us/mkl}{Intel® Math Kernel
+Library} (MKL) for multi-threading in linear algebra operations
 \citep{microsoft_multithread}. MRO does not provide official Docker
@@
 
-\emph{Renjin} is an interpreter for the R language running on top of the
-\href{https://en.wikipedia.org/wiki/Java_virtual_machine}{Java Virtual
-Machine} (JVM) providing full two-way access between Java and R code
-\citep{wikipedia_renjin_2018}. It was developed to combine the benefits
-of R, such scripting and extension packages, with the JVM's advantages
-in the areas of security, cross-platform availability, and established
-position in enterprise settings. R extension packages need to be
-specially compiled and are distributed via the Java package manager
-\href{https://en.wikipedia.org/wiki/Apache_Maven}{Apache Maven},
+\textbf{Renjin} is an interpreter for the R language running on top of
+the \href{https://en.wikipedia.org/wiki/Java_virtual_machine}{Java
+Virtual Machine} (JVM) providing full two-way access between Java and R
+code \citep{wikipedia_renjin_2018}. It was developed to combine the
+benefits of R, such scripting and extension packages, with the JVM's
+advantages in the areas of security, cross-platform availability, and
+established position in enterprise settings. R extension packages need
+to be specially compiled and are distributed via the Java package
+manager \href{https://en.wikipedia.org/wiki/Apache_Maven}{Apache Maven},
 cf.~\url{http://packages.renjin.org/packages} for available packages.
-Packages are loaded on demand, i.e.~at the first call to
+Packages are loaded on demand, i.e., at the first call to
 \texttt{library()}. Not all R packages, especially one linking to binary
-libraries, are available, e.g.~\texttt{rgdal}
+libraries, are available, e.g., \texttt{rgdal}
 \footnote{\href{http://packages.renjin.org/package/org.renjin.cran/rgdal/1.4-4/build/1}{http://packages.renjin.org/package/org.renjin.cran/rgdal/1.4-4/build/1}}.
-
 There are no offical Docker images for Renjin, but community-maintained
@@
 
-\href{http://www.pqr-project.org/}{\emph{pqR}} is a ``a pretty quick
-version of R''. \emph{pqR} fixes some opinionated issues in the R
-language and is the basis for experimental features, e.g.~automatic
-diffentiation\footnote{ \href{https://riotworkshop.github.io/abstracts/riot-2019-pqr.txt}{https://riotworkshop.github.io/abstracts/riot-2019-pqr.txt}}.
+\textbf{\href{http://www.pqr-project.org/}{pqR}} is a ``a pretty quick
+version of R''. \emph{pqR} attempts to improve R on some opinionated
+issues in the R language and is the basis for experimental features,
+e.g., automatic
+diffentiation\footnote{\href{https://riotworkshop.github.io/abstracts/riot-2019-pqr.txt}{https://riotworkshop.github.io/abstracts/riot-2019-pqr.txt}}.
 The source code development
@@
 
-\emph{FastR} is ``A high-performance implementation of the R programming
-language, built on GraalVM'' (\url{https://github.com/oracle/fastr}). It
-is developed by Oracle, connects R to the GraalVM ecosystem
-\citep{wikipedia_graalvm_2019}, and also claims superior performance but
-also targets full compatibility with base R
-\footnote{ \href{https://github.com/oracle/fastr}{https://github.com/oracle/fastr}}.
-
-There are no offical Docker images provided, but
-\texttt{nuest/fastr-docker} provides image and \texttt{Dockerfile} on
-Docker Hub and GitHub at \href{}{} and
-\url{https://github.com/nuest/fastr-docker} respectively.
+\textbf{FastR} is ``A high-performance implementation of the R
+programming language, built on GraalVM''
+(\url{https://github.com/oracle/fastr}). It is developed by Oracle,
+connects R to the GraalVM ecosystem \citep{wikipedia_graalvm_2019}, and
+also claims superior performance but also targets full compatibility
+with base R
+\footnote{\href{https://github.com/oracle/fastr}{https://github.com/oracle/fastr}}.
+There are no offical Docker images provided, but \texttt{nuest/fastr}
+images and \texttt{Dockerfile} are independently maintained on
+\href{https://cloud.docker.com/repository/docker/nuest/fastr/}{Docker
+Hub} and \href{https://github.com/nuest/fastr-docker}{GitHub}
+respectively.
 
 While the images presented in this section are far from being as vetted,
-stable, and widely used as any of the \emph{Rocker} images, they
-demonstrate an important advantage of containerisation technology,
-namely the abiliy to transparently build portable stacks of open source
-software and make them easily accessible to users. All different
-distributions are published under GPL licenses. Since all of the
-different R distributions claim better performance as a core motivation,
-a comparision based on Docker images, potentially leveraging the
+stable, and widely used as any of the Rocker images, they demonstrate an
+important advantage of containerisation technology, namely the abiliy to
+transparently build portable stacks of open source software and make
+them easily accessible to users. All different distributions are
+published under GPL licenses. Since all of the different R distributions
+claim better performance as a core motivation, a comparision based on
+Docker images, potentially leveraging the
 \href{https://docs.docker.com/config/containers/resource_constraints/}{resource
@@
 
-\href{http://bioconductor.org/help/docker/}{\emph{Bioconductor}} is an
-open source, open development project for the analysis and comprehension
-of genomic data \citep{gentleman_bioconductor_2004}. The project
-consists of 1741 R software packages as of August 15th 2019, as well as
-packages containing annotation or experiment data. \_Bioconductor\_has a
-semiannual release cycle, each release is associated with a particular
-version of R. Docker images allow availability of current and past
-versions of \emph{Bioconductor} for convenience and reproducibility.
-
+\href{http://bioconductor.org/}{\emph{Bioconductor}} is an open source,
+open development project for the analysis and comprehension of genomic
+data \citep{gentleman_bioconductor_2004}. The project consists of 1741 R
+software packages as of August 15th 2019, as well as packages containing
+annotation or experiment data. \_Bioconductor\_has a semiannual release
+cycle, each release is associated with a particular version of R. Docker
+images allow availability of current and past versions of
+\emph{Bioconductor} for convenience and reproducibility.
 \emph{Bioconductor} `base' docker images are built on top of
@@
 \texttt{rocker/rstudio} and \texttt{rocker/r-ver} version tagging.
-\emph{Bioconductor} selects the desired version of R from \emph{Rocker},
-adds the BiocManager CRAN package for installing appropriate versions of
+\emph{Bioconductor} selects the desired version of R from Rocker, adds
+the BiocManager CRAN package for installing appropriate versions of
 \emph{Bioconductor} packages, and creates a \emph{Bioconductor} docker
@@
 maintained. All community maintained images build on top of the
-\emph{Bioconductor} base image and therefore indirectly the
-\emph{Rocker} images. To simplify building and maintaining
-\emph{Bioconductor} images, we use a Ruby templating engine.
-
-A recent audit of our Dockerfiles, following best practices from the
-Docker website, led to a reduction in the size and number of layers in
-images we produce. The most important insights involve the
+\emph{Bioconductor} base image and therefore indirectly the Rocker
+images. To simplify building and maintaining \emph{Bioconductor} images,
+we use a Ruby templating engine. A recent audit of the
+\emph{Bioconductor} \texttt{Dockerfile}s, following best practices from
+the Docker website, led to a reduction in the size and number of layers.
+The most important insights involve the
 \href{https://docs.docker.com/storage/storagedriver/overlayfs-driver/\#how-container-reads-and-writes-work-with-overlay-or-overlay2}{`union'
@@
 
-\hypertarget{images-for-historic-r-versions-nuestcolinfay-nuest}{%
-\subsection{\texorpdfstring{Images for (historic) R versions
-{[}@ColinFay,
-@nuest{]}}{Images for (historic) R versions , @nuest{[}@ColinFay, @nuest{]}}}\label{images-for-historic-r-versions-nuestcolinfay-nuest}}
+\hypertarget{images-for-historic-r-versions}{%
+\subsection{Images for (historic) R
+versions}\label{images-for-historic-r-versions}}
+
+As with any other software, each new version of R comes with its share
+of changes. Some are breaking changes, some are not, but the fact is
+that running a piece of code in a given version might give different
+results from another version can be an issue when it comes to
+reproducibility of workflows and stability of applications. For example,
+think about R random seed: starting with \texttt{R\ 3.6.0}, running
+\texttt{set.seed(2811);\ sample(1:1000,\ 2)} will not give the same
+result as if it was run inside an older version of R. That might seem
+trivial, but all the code using the random number generator will not be
+exactly reproducible after this breaking change. Therefore, controlling
+the version of software is most crucial for reproducible research
+\citep[e.g.][]{boettiger_introduction_2015}.
+
+Containers are perfectly suited to capture a specific configuration of a
+computing environment to prohibit such problems. The Rocker images
+provides the \emph{versioned stack} for different the version of R since
+the project inception, which use the \emph{then-current} stable Debian
+base image \citep[cf.][]{RJ-2017-065}. Based on custom build phase
+hooks\footnote{\href{https://docs.docker.com/docker-hub/builds/advanced/}{https://docs.docker.com/docker-hub/builds/advanced/}},
+i.e., small shell scripts executed at different phases in the automated
+build of Rocker images, there are also semantic version tags for the
+most recent freezed (i.e., using MRAN) versions with only the major and
+minor
+version\footnote{See \href{https://github.com/rocker-org/rocker-versioned/blob/master/VERSIONS.md}{https://github.com/rocker-org/rocker-versioned/blob/master/VERSIONS.md}, FIXME: \href{https://github.com/rocker-org/rocker-versioned/issues/42}{https://github.com/rocker-org/rocker-versioned/issues/42}.}.
+So, at the time of writing this article, \texttt{rocker/r-ver:3} and
+\texttt{3.6} are aliases for \texttt{3.6.0}, because \texttt{3.6.1} is
+the latest release. With the release of \texttt{3.6.2} and pinning of
+the MRAN version in \texttt{3.6.1}, the tags \texttt{3} and \texttt{3.6}
+will deliver the same image as \texttt{3.6.1}. These tags allow users to
+update the base image to retrieve bugfixes while reducing the risk of
+also introducing breaking changes.
+
+\href{https://github.com/ColinFay/ronline}{r-online} is an app for
+helping users to detect breaking changes between different R versions,
+and for historic exploration of R. With a standalone NodeJS app or
+\href{https://srv.colinfay.me/r-online}{online}, the user can compare a
+piece of code run in two separate versions of R. Internally, r-online
+opens one or two Docker instances with the given version of R based on
+Rocker images, executes a given piece of code, and returns the result to
+the user.
+
+A container also provides an isolated sandbox environment suitable for
+testing and evaluation, without interfering with the ``main'' working
+environment. This enables cross version testing and bugfixing. To take
+this even further, the Rocker contributors are discussing the provision
+of R versions reaching back futher than the project's own inception,
+reaching back as far as R \texttt{2.x} and \texttt{1.x}
+\footnote{\href{https://github.com/rocker-org/rocker-versioned/issues/138}{https://github.com/rocker-org/rocker-versioned/issues/138}}.
+The main challenges are finding a suitable base image with a matching OS
+version, or making the adjustments to compile R on more recent OS
+releases.
+
+\hypertarget{windows-images}{%
+\subsection{Windows Images}\label{windows-images}}
+
+Docker containers on the Windows operating system were originally quite
+cumbersome to use. They required an extra tool,
+\href{https://docs.docker.com/toolbox/}{Docker Toolbox}, which by now
+only exists as a legacy solution for older Windows systems. Docker
+toolbox leverages \texttt{docker-machine} (see also
+Section~\ref{interfaces} for an R package interfacing with
+\texttt{docker-machine}) to handle the process of creating a local
+virtual host which could host Docker Engine, while exposing the regular
+Docker CLI. The Docker CLI commands are forwarded to the virtual host
+transparently for the user. For current Windows Server (2016 and later)
+and Windows Desktop
+(\href{https://www.docker.com/products/docker-desktop}{Docker Desktop}
+requires Windows 10) versions, Docker is supported
+natively\footnote{\href{https://www.docker.com/products/windows-containers}{https://www.docker.com/products/windows-containers}}
+and different base images are offered by
+Microsoft\footnote{\href{https://docs.microsoft.com/en-us/virtualization/windowscontainers/about/index}{https://docs.microsoft.com/en-us/virtualization/windowscontainers/about/index}}.
+The Docker CLI can be used in just the same way as on other operating
+systems but not every base image is supported on every Windows
+host\footnote{\href{https://docs.microsoft.com/en-us/virtualization/windowscontainers/deploy-containers/version-compatibility}{https://docs.microsoft.com/en-us/virtualization/windowscontainers/deploy-containers/version-compatibility}}.
+On Docker Desktop for Windows, the user can run both Linux-based and
+Windows-based containers, but only one of the Docker daemons can be used
+at a
+time\footnote{See "Switch between Windows and Linux containers" on \href{https://docs.docker.com/docker-for-windows/}{https://docs.docker.com/docker-for-windows/}.}.
+
+\textbf{\href{https://github.com/nuest/rocker-win}{\texttt{rocker-win}}}
+is a proof of concept for running R in Windows-based containers
+\citep{nust_rocker-win_2019}. It provides selected R versions using
+three different base images using \texttt{microsoft/windowsservercore}
+and \texttt{mcr.microsoft.com/windows/servercore}. These base images
+match different Windows versions: Windows Server 2019, Windows Server
+2016, and Windows Server, version 1803. The images are built using two
+CI services,
+\href{https://docs.travis-ci.com/user/reference/windows/}{Travis CI} and
+\href{https://www.appveyor.com/docs/windows-images-software/}{Appveyor},
+which provide the different Windows versions and therefore support
+different base images, and published automatically
+\href{https://hub.docker.com/r/nuest/rocker-win}{on Docker Hub}. The
+images are built from manually maintained seperate \texttt{Dockerfile}s,
+are tagged with both Windows-variant and R version, e.g.,
+\texttt{nuest/rocker-win:ltsc2019-3.6.2} or
+\texttt{nuest/rocker-win:1803-latest}, and run the \texttt{R.exe}
+process by default. New R versions are only added on demand. The
+explorative project demonstrates the variety and pontentially confusing
+complexity of Windows on Docker, e.g., because of the different base
+images, but also provides a starting point for any Windows-constrained
+user to leverage containerisation, including examples for apps using
+\textbackslash{}pkg\{plumber, \pkg{Shiny}, or R packages with system
+dependencies such as \pkg{sf}. The latter is quite comfortable actually,
+as R ships pre-compiled binaries for Windows.
+
+For Windows, just as originally for Linux, the cloud use cases drive the
+development and Docker and Microsoft collaborate
+closely\footnote{\href{https://www.docker.com/partners/microsoft}{https://www.docker.com/partners/microsoft}}.
+In most cases, individual developers or researchers worrying about
+reproducibility, will prefer the more widely and host independent
+Linux-based containers. Naturally, a Windows license is required for the
+host machine. But the developments in the \texttt{rocker-win} prototype
+show that containers can also be used for R workflows depending on
+Windows-only tools, for leveraging an existing Windows Server-based
+infrastructure where policies can otherwise not be met, or for
+streamlining interactions with system operating staff that runs
+Windows-based containers themselves.
+
+\hypertarget{non-debian-linux-images}{%
+\subsection{Non-Debian Linux images}\label{non-debian-linux-images}}
+
+\label{rhub} The \textbf{R-hub} project provides \_``a collection of
+services to help R package development'', with the package builder as
+the most prominent one\citep{r-hub_docs_2019}. The builder allows R
+package developers to check their R package on different platforms and R
+versions using a web form or the package \pkg{rhub}
+\citep{csardi_rhub_2019}. The builder uses Docker containers to conduct
+these checks, taking advantages of their well-defined environments and
+sandboxing. The \texttt{Dockerfile}s and images are published on Docker
+Hub\footnote{\href{https://hub.docker.com/u/rhub}{https://hub.docker.com/u/rhub}}
+and
+GitHub\footnote{\href{https://github.com/r-hub/rhub-linux-builders}{https://github.com/r-hub/rhub-linux-builders}}
+respectively. The images comprise release and development builds of R
+running in the base images of the
+\href{https://www.debian.org/}{Debian},
+\href{https://ubuntu.com/}{Ubuntu},
+\href{https://getfedora.org/}{Fedora}, and
+\href{https://centos.org/}{CentOS} linux distributions (Arch Linux is
+under
+development\footnote{\href{https://github.com/r-hub/rhub-linux-builders/pull/41}{https://github.com/r-hub/rhub-linux-builders/pull/41}},
+\texttt{see}rhub::local\_check\_linux\_images()\texttt{for\ a\ full\ list),\ and\ are\ intended\ for\ debugging\ (see\ Section\&nbsp;\textbackslash{}ref\{development\})\ not\ for\ reproducible\ workflows.\ \_The\ rationale\_\ of\ not\ re-using\ the\ Rocker\ image\ stack\ is\ the\ high\ number\ of\ operatings\ systems\ and\ configurations,\ which\ can\ be\ more\ uniformly\ maintained\ in\ a\ seperate\ suite\ of\ images.\ Also,\ a\ common\ structure\ across\ the\ respective\ image\ stacks\ of\ the\ operating\ systems,\ e.g.,\ for}debian\texttt{,\ adding\ GCC\ in}debian-gcc\texttt{and\ then\ R\textquotesingle{}s\ development\ version\ in}debian-gcc-devel`.
+The platform also has a specific user configuration differing from the
+setup covering typical applications of Rocker images, and the images
+include considerably more software, such as a fairly complete LaTeX
+installation, to provide an experience closer to CRAN.
+
+There are also independent projects installing R or RStudio on
+\href{https://www.alpinelinux.org/}{\textbf{Alpine Linux}}, but not
+beyond the proof of concept
+stage\footnote{\href{https://github.com/artemklevtsov/r-alpine}{https://github.com/artemklevtsov/r-alpine} or \href{https://www.github.com/velaco/alpine-r}{https://www.github.com/velaco/alpine-r} provide different use case images but are not recently maintained (development of the latter documented in \citet{ratesic_building_2018}); \href{https://github.com/cmplopes/alpine-r}{https://github.com/cmplopes/alpine-r} has only base R images, is relatively up-to-date but sparsely documented.}.
+Alpine images are often used for minimalistic and thereby small and
+secure containers, e.g., in hardware with limited storage, though this
+advantage quickly deteriorates if more features are needed, e.g.~if
+installing RStudio, package building envrionments, or powerful data
+science tools. Furthermore the distribution uses the
+\href{https://www.musl-libc.org/}{\texttt{musl} C library} instead of
+the \texttt{glibc} used in Debian/Ubuntu, for which more tooling and
+experiences
+exist\footnote{Cf. \href{https://github.com/rocker-org/rocker/issues/231}{https://github.com/rocker-org/rocker/issues/231}}.
+\href{https://github.com/CenterForStatistics-UGent/mountainr}{https://github.com/CenterForStatistics-UGent/mountainr}
+only has a single and quite short \texttt{Dockerfile}, showing the easy
+installation of the latest R from Alpine sources. These images are used
+\ldots{}
+
+\href{https://github.com/ismailsunni/altRnative/}{\texttt{altRnative}}
+is an experimental R package for running the same code across multiple
+containerised versions of R, intended for comparison across operating
+systems and different implementations of R. It comes with a collection
+of \texttt{Dockerfile}s and corresponding images
+\footnote{\href{https://github.com/ismailsunni/dockeRs}{https://github.com/ismailsunni/dockeRs}}.
+
+\hypertarget{data-science-images}{%
+\subsection{Data Science images}\label{data-science-images}}
+
+The notion of \emph{Data Science} is widely discussed among all academic
+disciplines. Using computers and complex software to make sense of
+(large amounts of) collected or simulated data is a cross-cutting
+phenomenon leading to \emph{X Data Science} disciplines, seminars, and
+study majors, there are also Docker images with commonly used tools and
+software stacks pre-installed that are often used in combination for
+analysing data. A large number of bespoke computing environments are
+created this way. Here we only mention a few image stacks as examples
+that are not based on Rocker images but prominently include
+installations of R.
+
+The \href{https://github.com/jupyter/docker-stacks/}{\textbf{Jupyter
+Docker Stacks} project} are a set of ready-to-run Docker images
+containing Jupyter applications and interactive computing tools
+\citep{project_jupyter_jupyter_2018}. The \texttt{jupyter/r-notebook}
+image includes R and ``popular packages'', and R is also included in the
+catchall \texttt{datascience-notebook}
+image\footnote{\href{https://jupyter-docker-stacks.readthedocs.io/en/latest/using/selecting.html}{https://jupyter-docker-stacks.readthedocs.io/en/latest/using/selecting.html}}.
+For example, these images allow users to quickly start a Jupyter
+Notebook server locally or build their own specialised images on top of
+stable toolsets.
+
+\textbf{Kaggle} provides the
+\href{https://hub.docker.com/r/kaggle/rstats}{\texttt{gcr.io/kaggle-images/rstats}}
+image (previously \texttt{kaggle/rstats}) and
+\href{https://github.com/Kaggle/docker-rstats}{corresponding
+\texttt{Dockerfile}} for usage in their Machine Learning competitions
+and easy access to the associated datasets. It includes machine learling
+libraries such as Tensorflow and Keras, and also configures the
+\pkg{reticulate} package. The image uses a base image with \emph{all
+packages from CRAN}, \texttt{gcr.io/kaggle-images/rcran}, which requires
+a Google Cloud Build as Docker Hub would time
+out\footnote{Originally, a stacked collection of over 20 images with automated builds on Docker Hub was used, see \href{https://web.archive.org/web/20190606043353/http://blog.kaggle.com/2016/02/05/how-to-get-started-with-data-science-in-containers/}{https://web.archive.org/web/20190606043353/http://blog.kaggle.com/2016/02/05/how-to-get-started-with-data-science-in-containers/} and \href{https://hub.docker.com/r/kaggle/rcran/dockerfile}{https://hub.docker.com/r/kaggle/rcran/dockerfile}}.
+The final extracted image size is over 25GB, which makes it debatable if
+having everything available is actually convenient.
+
+As a further example,
+\href{https://radiant-rstats.github.io/docs/}{\textbf{Radiant} project}
+provides several images, e.g.,
+\href{https://hub.docker.com/r/vnijs/rsm-msba-spark}{\texttt{vnijs/rsm-msba-spark}},
+for their browser-based business analytics interface based on
+\pkg{Shiny} (\texttt{Dockerfile}
+\href{https://github.com/radiant-rstats/docker}{on GitHub}) and for use
+in education as part of an MSc course. As data science often applies a
+multitude of tools, this image favours inclusion over selection and
+features Python, Postgres, JupyterLab and Visual Studio Code besides R
+and RStudio, bringing the image size up to 9GB.
+
+\hypertarget{use-cases-and-applications}{%
+\section{Use cases and applications}\label{use-cases-and-applications}}
+
+\hypertarget{interfaces-for-docker-in-r}{%
+\subsection{Interfaces for Docker in
+R}\label{interfaces-for-docker-in-r}}
+
+Interfacing with the Docker daemon is typically done through the Docker
+Command Line Interface (CLI). However, moving back and forth between an
+R console and the command line can create friction in workflows and
+reduce reproducibility. A number of first-order R packages provide a
+interface to the Docker CLI, allowing to automate interaction with the
+Docker CLI from an R console.
+
+Each of these packages has particular advantages as they provide
+function wrappers for interacting with the Docker CLI at different
+stages of a container's life cycle. Examples of such interactions are
+installing the Docker software, creating \texttt{Dockerfile}s, building
+an image, and launching a container on a local machine or on the cloud.
+As such, the choice of which package is most useful depends on the
+use-case at hand.
+
+\begin{tabular}{l|l|l|l|l|l|l}
+\hline
+\rotatebox{-90}{Functionality} & \rotatebox{-90}{AzureContainers} & \rotatebox{-90}{babelwhale} & \rotatebox{-90}{dockermachine} & \rotatebox{-90}{dockyard} & \rotatebox{-90}{harbor} & \rotatebox{-90}{stevedore}\\
+\hline
+Generate a Dockerfile &  &  &  & \checkmark &  & \\
+\hline
+Build an image & \checkmark &  &  & \checkmark &  & \\
+\hline
+Execute a container locally or remotely & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark\\
+\hline
+Deploy or manage an instances in the cloud & \checkmark &  & \checkmark &  & \checkmark & \checkmark\\
+\hline
+Interact with an instance (e.g. file transfer) &  & \checkmark & \checkmark &  &  & \checkmark\\
+\hline
+Manage storage of images &  &  &  &  & \checkmark & \checkmark\\
+\hline
+Supports Docker and Singularity &  & \checkmark &  &  &  & \\
+\hline
+Direct access to Docker API instead of using the CLI &  &  &  &  &  & \checkmark\\
+\hline
+Installing Docker software &  &  & \checkmark &  &  & \\
+\hline
+\end{tabular}
+
+\textbf{stevedore} (\url{https://cran.r-project.org/package=stevedore})
+
+\textbf{dockyard} (\url{https://github.com/thebioengineer/dockyard}) has
+the goal of lowering barrier to creating \texttt{Dockerfile}s and Docker
+images, and deploying Docker containers. The way docker images are made
+is by reading and following a set of instructions found in
+\texttt{Dockerfile}s. This requires stepping out of the R ecosystem and
+into an infrastructure mindset where the user needs to specify the
+environment they require. Because \texttt{Dockerfile}s can be read as a
+series of steps, this lends itself well to the \texttt{tidyverse} style
+of programming where each step is piped into the next. dockyard has a
+\texttt{Dockerfile} object that either starts out as an empty
+\texttt{Dockerfile}, or can use existing \texttt{Dockerfile} to use as a
+template. This combined with a series of functions that are analogous to
+the steps commonly found in a \texttt{Dockerfile}, it makes building
+Docker images much more approachable to R users. There are also wrappers
+for commonly used steps for R programmers like adding libraries and
+copying files into the R Docker containers. Once the \texttt{Dockerfile}
+is created, dockyard has the functionality to build the image and deploy
+it.
+
+\textbf{dockermachine} (\url{https://github.com/cboettig/dockermachine})
+
+\textbf{AzureContainers}
+(\url{https://cran.r-project.org/package=AzureContainers}) is an
+interface to a number of container-related services in Microsoft's Azure
+cloud, namely Container Instances, Container Registry and Kubernetes
+Service. While it is mainly intended for working with Azure, as a
+convenience feature it includes lightweight, cross-platform shells to
+Docker, kubectl and helm. These can be used to create and manage
+arbitrary Docker images and containers, as well as Kubernetes clusters
+on any platform or cloud service.
+
+\textbf{babelwhale}
+(\url{https://cran.r-project.org/package=babelwhale}) allows executing
+and interacting with containers, which can use either Docker or
+Singularity as a backend. provides a unified interface to interact with
+Docker and Singularity containers. You can execute a command inside a
+container, mount a volume or copy a file.
+
+\textbf{harbor} (\url{https://github.com/wch/harbor}
+
+\hypertarget{capture-and-create-environments-nuest}{%
+\subsection{\texorpdfstring{Capture and create environments
+\citep{nuest}}{Capture and create environments {[}@nuest{]}}}\label{capture-and-create-environments-nuest}}
+
+Several second order R packages attempt to make the process of creating
+Docker images and using containers for specific tasks, such as running
+tests or rendering reproducible reports, easier.
 
@@
 \item
-  run code with Versions of R: \url{https://srv.colinfay.me/r-online}
-\item
-  old R versions:
-  \url{https://github.com/rocker-org/rocker-versioned/issues/138}
-\item
-  \href{https://github.com/rocker-org/rocker-versioned/\#version-tags}{semantic
-  version tags for Rocker}
-\item
-  debugging memory problems across multiple versions:
-  \url{https://github.com/wch/r-debug}
-\item
-  cross-R testing: Show how you can run the same script in multiple R
-  variants in containers (new contribution). Ideally the images should
-  be created using \texttt{containerit} and controlled with
-  \texttt{stevedore}/\texttt{docker}, i.e.~via an R script.
+  \href{https://github.com/ColinFay/dockerfiler/}{\texttt{dockerfiler}}
+\item
+  \href{https://github.com/o2r-project/containerit/}{\texttt{containerit}}
+\item
+  \href{https://github.com/smaakage85/dockr}{\texttt{dockr}}
+\item
+  \href{https://github.com/traitecoevo/dockertest/}{\texttt{dockertest}}
+\item
+  \href{https://rstudio.github.io/renv/articles/docker.html}{\texttt{renv}
+  with Docker}
 \end{itemize}
 
-\hypertarget{windows-images-nuest}{%
-\subsection{Windows Images {[}@nuest{]}}\label{windows-images-nuest}}
-
-\begin{itemize}
-\tightlist
-\item
-  \href{https://github.com/nuest/rocker-win}{rocker-win}
-\item
-  is possible, only relevant in organisations with an existing Windows
-  Server-based infrastructure, can meet policies then
-\end{itemize}
-
-\hypertarget{non-debian-linux-images-nn}{%
-\subsection{Non-Debian Linux images
-{[}NN{]}}\label{non-debian-linux-images-nn}}
-
-\begin{itemize}
-\tightlist
-\item
-  Alpine images
-\item
-  \href{https://github.com/r-hub/rhub-linux-builders}{Images used by
-  R-Hub} (overlap with CI?)
-\item
-  \url{https://github.com/jlisic/R-docker-centos}
-\end{itemize}
-
-\hypertarget{r-packages}{%
-\section{R Packages}\label{r-packages}}
-
-\hypertarget{packages-for-automation-and-packaging-nuest}{%
-\subsection{Packages for automation and packaging
-{[}@nuest{]}}\label{packages-for-automation-and-packaging-nuest}}
-
-\begin{itemize}
-\tightlist
-\item
-  \href{https://github.com/o2r-project/containerit/}{\texttt{containerit}}
-\item
-  \href{https://github.com/traitecoevo/dockertest/}{\texttt{dockertest}}
-\end{itemize}
-
-\hypertarget{liftr}{%
-\subsubsection{liftr}\label{liftr}}
-
-\href{https://nanx.me/liftr/}{\emph{liftr}} \citep{liftr2019} aims to
-solve the problem of persistent reproducible reporting in statistical
-computing. Currently, the R Markdown format and its backend compilation
-engine \emph{knitr} offer a \emph{de facto} standard for creating
-dynamic documents \citep{xie2018}. However, the reproducibility of such
-content authoring environments is often limited to individual machines
---- it is not easy to replicate the system environment (libraries, R
-versions, R packages) where the document was compiled. This issue
-becomes even more serious when it comes to collaborative document
-authoring and creating large-scale document building services.
+\textbf{\href{https://nanx.me/liftr/}{\emph{liftr}}} \citep{liftr2019}
+aims to solve the problem of persistent reproducible reporting in
+statistical computing. Currently, the R Markdown format and its backend
+compilation engine \emph{knitr} offer a \emph{de facto} standard for
+creating dynamic documents \citep{xie2018}. However, the reproducibility
+of such content authoring environments is often limited to individual
+machines --- it is not easy to replicate the system environment
+(libraries, R versions, R packages) where the document was compiled.
+This issue becomes even more serious when it comes to collaborative
+document authoring and creating large-scale document building services.
 \emph{liftr} solves this reproducibility problem by bringing Docker to
@@
 
+One of the key tool of Docker are \texttt{Dockerfile}s, which can be
+thought of as ``recipes'' used to build a specific image. But building
+these files can feel like a workflow break as it demands to open a
+separate file, and / or to write a small wrapper around R's
+\texttt{write()} function to add elements to the file. At the same time,
+doing it by hand prevents from using programming language for what they
+are good for: iteration, and automation. Iteration and automation for
+\texttt{Dockerfile}s is the very reason behind
+\href{https://github.com/ColinFay/dockerfiler}{\texttt{dockerfiler}}, an
+R package designed for building \texttt{Dockerfile}s straight from R.
+
+For example, the
+\textbf{\href{https://github.com/ThinkR-open/golem}{\texttt{golem}}}
+package makes an heavy use of \texttt{dockerfiler} when it comes to
+creating the \texttt{Dockerfile} for building production-grade Shiny
+applications and deploying them. The reason behind this is that
+\texttt{dockerfiler}, on top of being scriptable from R, can leverage
+all the tools available in R to parse a \texttt{DESCRIPTION} file, to
+get system requirements, to list dependencies, versions, etc.
+
+\hypertarget{deployment-to-cloud-services-div}{%
+\subsection{Deployment to cloud services
+{[}div{]}}\label{deployment-to-cloud-services-div}}
+
+The cloud is the natural environment of containers, and becomes the
+go-to mechanism to expose applications written in R to users.
+
 \begin{itemize}
@@
 \item
-  \href{https://github.com/cole-brokamp/rize}{\texttt{rize} for Shiny}
-\item
-  \href{https://github.com/ColinFay/dockerfiler}{\texttt{dockerfiler}}
-\end{itemize}
-
-\hypertarget{packages-for-control-and-provisioning-richfitz-rcannoodnuest-richfitz-rcannood}{%
-\subsection{\texorpdfstring{Packages for control and provisioning
-{[}@nuest, @richfitz,
-@rcannood{]}}{Packages for control and provisioning , @richfitz, @rcannood{[}@nuest, @richfitz, @rcannood{]}}}\label{packages-for-control-and-provisioning-richfitz-rcannoodnuest-richfitz-rcannood}}
+  \texttt{RSelenium}
+\item
+  \href{https://cloudyr.github.io/googleComputeEngineR/}{\texttt{googleComputeEngineR}}
+  (function \texttt{docker\_run})
 
@@
   \item
-  \href{https://github.com/richfitz/stevedore}{\texttt{stevedore}}
-\item
-  \href{https://cran.r-project.org/web/packages/babelwhale/index.html}{\texttt{babelwhale}}:
-  Running a Docker from R with Singularity or Docker as back-end. This
-  is really useful in HPC environments where a user might not have root
-  access but is able to install Singularity instead of Docker.
-  {[}@rcannood{]}
-\item
-  \href{https://bhaskarvk.github.io/docker/}{\texttt{docker}}
-\item
-  \texttt{RSelenium}
-\item
-  \href{https://cloudyr.github.io/googleComputeEngineR/}{\texttt{googleComputeEngineR}}
-  (function \texttt{docker\_run})
+    I don't think \texttt{docker\_run} is main focus for containers for
+    this package, its more how it uses Docker to launch custom
+    environments that can be templated (e.g., rstudio, shiny etc.) and
+    for parrallisation e.g., \texttt{library(future)} using GCE VMs as
+    endpoints. \citep{MarkEdmondson1234}
+  \end{itemize}
+\item
+  \href{https://github.com/sckott/analogsea}{\texttt{analogsea}}
+  (digital ocean R client)
+\item
+  \href{https://cran.r-project.org/package=AzureContainers}{\texttt{AzureContainers}}:
+  Umbrella package for working with containers in Microsoft's Azure
+  cloud. Provides interfaces to three Azure services: Container
+  Registry, Container Instances (for running individual containers), and
+  Kubernetes Services (for orchestrated deployments)
+\end{itemize}
+
+\hypertarget{using-r-to-power-enterprise-software-in-production-environments-nolisticjnolis-nolistic}{%
+\subsection{\texorpdfstring{Using R to power enterprise software in
+production environments
+\citep[\citet{nolistic}]{jnolis}}{Using R to power enterprise software in production environments , @nolistic{[}@jnolis, @nolistic{]}}}\label{using-r-to-power-enterprise-software-in-production-environments-nolisticjnolis-nolistic}}
+
+R has been historically viewed as a tool for analysis and scientific
+research--not for creating software that corporations can rely on to
+continously run in real time. However, thanks to advancements in R
+running as a web service, along with with the ability to deploy R in
+docker containers, modern enterprises are now capable of having
+real-time machine learning powered by R.
+
+The backbone of this work is the
+\emph{\href{https://www.rplumber.io/docs/hosting.html\#docker}{plumber}}
+package, created by Jeff Allen. The plumber package allows R to be used
+to host web services through RESTful APIs. When R is running on a
+server, a different server can ask R to make a computation through an
+HTTP request and get an immediate response. For instance, if a data
+scientist at a company creates a machine learning model to tell which
+customers are likely to make a repeat purchase, if they serve the model
+through a plumber API then someone else on the marketing team could
+write software that sends different emails depending on that real-time
+prediction. Plumber syntax is simple to read and understand: the API is
+created by commenting above function names in an R script.
+
+\begin{verbatim}
+#* Return the sum of two numbers
+#* @get /random
+function(){
+  runif()
+}
+\end{verbatim}
+
+\emph{Example of a plumber API to generated a random number by visiting
+\texttt{http://127.0.0.1/random}}
+
+Plumber becomes far more powerful with the addition of Docker. Docker
+can be used to create images that
+\href{https://www.rplumber.io/docs/hosting.html\#docker}{include both R
+and plumber}. Then when the image is deployed, that creates the web
+endpoints that other services can hit and call the R code. This pattern
+of deploying code mimics those used by software engineering services
+created in Java, Python, and other modern languages. By using Docker to
+deploy plumber APIs, R can be used alongside those langages as a first
+class member of a software engineering technical stack.
+
+At T-Mobile, R is used to power real-time services that are called over
+a million times a day. T-Mobile created a set of neural network machine
+learning natural language processing models to help customer care agents
+manage text-based messages for customers
+\citep{t-mobile_enterprise_2018}. The models are convolutional neural
+networks that use the RStudio Keras package and are built on top of a
+Rocker Docker image. Since the models power tools for agents and
+customers, they need to have extremely high uptime and reliability--and
+they found that R is able to perform. The AI @ T-Mobile team open
+sourced their Docker images
+\href{https://github.com/tmobile/r-tensorflow-api}{on GitHub}.
+
+\hypertarget{deployment-to-the-cloud-hongooi}{%
+\subsection{\texorpdfstring{Deployment to the cloud
+\citep{hongooi}}{Deployment to the cloud {[}@hongooi{]}}}\label{deployment-to-the-cloud-hongooi}}
+
+While Plumber provides the low-level infrastructure for turning a
+statistical model into a predictive service, for production purposes it
+is usually necessary to take into account considerations such as
+scalability, reliability and ease of management. A single container is
+limited in the volume of requests it can service. Furthermore, if the
+container goes down, or is occupied with processing a computationally
+intensive task, the service becomes unavailable until it is restarted.
+
+These issues can be mitigated by deploying not a single container, but a
+\emph{cluster} of containers, orchestrated as a single deployment. The
+deployment system of choice at the time of writing is
+\href{https://kubernetes.io}{Kubernetes}. This allows you to deploy a
+single image to a pool of servers, which can then serve requests in
+parallel. Client-side requests are made to the cluster ingress endpoint,
+which then redirects them to individual servers. The orchestrator can
+identify cluster nodes that have failed, and restart them automatically;
+it can also detect when there is no capacity to service more requests,
+and start additional servers and containers as needed.
+
+While these load-balancing and autoscaling properties of a Kubernetes
+cluster are invaluable, they are still limited by the hardware resources
+that it has access to. A cluster cannot scale out if it has run out of
+servers. This issue in turn can be addressed by deploying not to
+on-premise servers but to the cloud, where effectively unlimited
+computing resources are available (as long as one can pay for them).
+From a management perspective, cloud deployments have the additional
+advantage that hardware maintenance becomes the responsibility of the
+cloud provider, eliminating another pain point. Orchestrator services
+are available from all the major cloud service providers, including
+Amazon Web Services (AWS), Google Cloud Platform, and Microsoft Azure.
+
+An example of an R package for working with containers in the cloud is
+AzureContainers, which interacts with three Azure services: Container
+Instances (for running a single container), Container Registry (a
+private docker registry service), and Kubernetes Service (for Kubernetes
+on Azure). AzureContainers provides a lightweight yet powerful interface
+for working with Resource Manager, the framework for deploying and
+managing arbitrary resources in Azure. On the client side, it provides
+simple shells to the Docker, Kubectl and Helm tools that are commonly
+used for managing containers and Kubernetes clusters.
+
+Here is some simple code that pushes a Docker image to an Azure
+Container Registry, and then creates a deployment using Azure Kubernetes
+Service. This is a slightly modified version of the code from the
+``Deploying a prediction service with Plumber'' AzureContainers
+vignette. The code makes use of a yaml file to define the deployment and
+predictive service, which is standard practice with Kubernetes; the yaml
+can be seen in the aforementioned vignette.
+
+\begin{verbatim}
+library(AzureContainers)
+# create a resource group for our deployments
+deployresgrp <- AzureRMR::get_azure_login()$
+    get_subscription("subscription_id")$
+    create_resource_group("deployresgrp", location="australiaeast")
+
+### create a container registry
+deployreg_svc <- deployresgrp$create_acr("deployreg")
+
+# build our image (a random forest on the Boston housing data, available in the MASS package)
+call_docker("build -t bos-rf .")
+
+# upload the image to Azure
+deployreg <- deployreg_svc$get_docker_registry()
+deployreg$push("bos-rf")
+
+### create a Kubernetes cluster with 2 nodes
+deployclus_svc <- deployresgrp$create_aks("deployclus", agent_pools=aks_pools("pool1", 2))
+
+# grant the cluster pull access to the registry
+gr <- AzureGraph::get_graph_login()
+aks_app_id <- deployclus$properties$servicePrincipalProfile$clientID
+reg$add_role_assignment(gr$get_app(aks_app_id), "Acrpull")
+
+# get the cluster endpoint
+deployclus <- deployclus_svc$get_cluster()
+
+# create and start the service
+deployclus$create("bos-rf.yaml")
+\end{verbatim}
+
+Once the service has been created, we can check on its status and obtain
+predictions:
+
+\begin{verbatim}
+deployclus$get("deployment bos-rf")
+#> Kubernetes operation: get deployment bos-rf  --kubeconfig=".../kubeconfigxxxx"
+#> NAME      DESIRED   CURRENT   UP-TO-DATE   AVAILABLE   AGE
+#> bos-rf    1         1         1            1           5m
+
+deployclus$get("service bos-rf-svc")
+#> Kubernetes operation: get service bos-rf-svc  --kubeconfig=".../kubeconfigxxxx"
+#> NAME         TYPE           CLUSTER-IP   EXTERNAL-IP     PORT(S)          AGE
+#> bos-rf-svc   LoadBalancer   10.0.8.189   xxx.xxx.xxx.xxx 8000:32276/TCP   5m
+
+response <- httr::POST("http://xxx.xxx.xxx.xxx:8000/score",
+    body=list(df=MASS::Boston[1:10,]), encode="json")
+httr::content(response, simplifyVector=TRUE)
+#> [1] 25.9269 22.0636 34.1876 33.7737 34.8081 27.6394 21.8007 22.3577 16.7812 18.9785
+\end{verbatim}
+
+Note that Plumber, by itself, provides no authentication or security
+features; anybody who knows the cluster's IP address can access the
+predictive service. At minimum, a basic authentication layer should be
+added to any Kubernetes deployment. Standard Kubernetes procedure
+involves creating an ingress controller, for example with the nginx or
+traefik reverse proxy software, and one can then add
+application-specific authentication layers on top of that (for example,
+to authenticate with a user's Azure Active Directory or LDAP
+credentials). More details can be found from most Kubernetes learning
+resources, and are outside the scope of this paper.
+
+\hypertarget{continuous-integration-and-continuous-delivery-colinfaynoamross-colinfay}{%
+\subsection{\texorpdfstring{Continuous integration and continuous
+delivery
+\citep[\citet{ColinFay}]{noamross}}{Continuous integration and continuous delivery , @ColinFay{[}@noamross, @ColinFay{]}}}\label{continuous-integration-and-continuous-delivery-colinfaynoamross-colinfay}}
+
+The controlled nature of containers, i.e., it is possible to define the
+software environment very well, even on remote machines, make them also
+useful for continuous integration (CI) and continuous delivery of
+applications.
 
@@
 \item
-    I don't think \texttt{docker\_run} is main focus for containers for
-    this package, its more how it uses Docker to launch custom
-    environments that can be templated (e.g.~rstudio, shiny etc.) and
-    for parrallisation e.g.~\texttt{library(future)} using GCE VMs as
-    endpoints. {[}@MarkEdmondson1234{]}
-  \end{itemize}
-\item
-  \href{https://github.com/sckott/analogsea}{\texttt{analogsea}}
-\item
-  \href{https://github.com/wch/harbor/}{\texttt{harbor}}
-\item
-  \href{https://github.com/cboettig/dockermachine}{\texttt{dockermachine}}
-\end{itemize}
-
-\hypertarget{use-cases-and-applications}{%
-\section{Use cases and applications}\label{use-cases-and-applications}}
-
-\hypertarget{continuous-integration-and-continuous-delivery-colinfaynoamross-colinfay}{%
-\subsection{\texorpdfstring{Continuous integration and continuous
-delivery {[}@noamross,
-@ColinFay{]}}{Continuous integration and continuous delivery , @ColinFay{[}@noamross, @ColinFay{]}}}\label{continuous-integration-and-continuous-delivery-colinfaynoamross-colinfay}}
+  DevOps
 
@@
   \item
-  R-Hub
-\item
-  DevOps
+    \url{https://www.opencpu.org/posts/opencpu-with-docker/}
+  \end{itemize}
+\end{itemize}
+
+When doing continuous integration and continuous delivery, it's crucial
+to test in an environment that matches the production environment. Tools
+like \texttt{Gitlab-CI} are built on top of Docker images: the user
+specifies a base Docker image, and the whole tests are run inside this
+environment. But, as we just said, this environment has to be fixed, but
+it also have to contain the necessary toolkit. In order to achieve that,
+\href{https://github.com/ColinFay/r-ci}{\texttt{r-ci}} combines
+\texttt{rocker} versioning and a series of tools specifically designed
+for testing. That way, package builders can use this image as a base
+image for there testing environment, without having to install the
+necessary packages every time they need to run a new test.
 
@@
 \item
-    \url{https://www.opencpu.org/posts/opencpu-with-docker/}
-  \end{itemize}
-\item
-  \texttt{r-ci}: \url{https://github.com/ColinFay/r-ci}
-\item
   \href{https://github.com/dynverse/dynwrap_containers/blob/master/.travis.yml}{dynwrap}
-  {[}@rcannood{]}
+  \citep{rcannood}
 
@@
 \item
-  Google Cloud Build {[}@MarkEdmondson1234{]}
+  Google Cloud Build \citep{MarkEdmondson1234}
 
@@
     For me this is what makes Docker containers viable, as it builds the
-    Dockerfiles on each GitHub commit. It couples with Google Container
-    Registry to build private Docker images, for downstream
+    \texttt{Dockerfile}s on each GitHub commit. It couples with Google
+    Container Registry to build private Docker images, for downstream
     applications.
@@
 
-\hypertarget{common-or-public-work-environments-div}{%
-\subsection{Common or public work environments
-{[}div{]}}\label{common-or-public-work-environments-div}}
+\hypertarget{common-or-public-work-environments}{%
+\subsection{Common or public work
+environments}\label{common-or-public-work-environments}}
+
+The fact that Docker images are portable and well defined make them
+useful when more than one person needs access to the same computing
+environment. This is even more useful when some of the users do not have
+the expertise to create such an environment themselves, and when these
+environments can be run in public or shared infrastructure.
+
+\href{https://github.com/karthik/holepunch}{\texttt{holepunch}} is an R
+package that was designed to make sharing work environments accessible
+to novice R users based on \emph{Binder}. The
+\href{https://mybinder.readthedocs.io/en/latest/}{Binder project},
+maintained by the team behind Jupyter, makes it possible for users to
+create and share their computing environments with others
+\citep{jupyter_binder_2018}. A \emph{BinderHub} allows anyone with
+access to a web browser and an internet connection to launch a temporary
+instance of these custom environments and execute any workflows
+contained within. From a reproducibility standpoint, Binder makes it
+exceedingly easy to compile a paper, visualize data, and run small
+examples from papers or tutorials without the need for any local
+installation. To set up Binder for a project, a user typically starts at
+an instance of a BinderHub and passes the location of a repository with
+a workspace, e.g., a hosted Git repository, or a data repository like
+Zenodo. Binder's core internal tool is
+\href{https://repo2docker.readthedocs.io/en/latest/config_files.html}{repo2docker}.
+It deterministically builds a Docker image by parsing the contents of a
+repository, e.g., project dependency configurations or simple
+configuration files. In the most powerful case, \texttt{repo2docker}
+builds a given \texttt{Dockerfile}. While this approach works well for
+most run of the mill Python projects, it is not so seamless for R
+projects. For any R projects that use the Tidyverse suite
+\citep{wickham_welcome_2019}, the time and resources required to build
+all dependencies from source can often time out before completion,
+making it frustrating for the average R user. \texttt{holepunch} removes
+some of these limitations by leveraging Rocker images that contain the
+Tidyverse along special Jupyter dependencies, and only installs
+additional packages from CRAN and Bioconductor that are not already part
+of these images. It short cicuits the configuration file parsing in
+\texttt{repo2docker} and starts with the Binder/Tidyverse base images,
+which eliminates a large part of the build time and in most cases
+results in a binder instance launching within a minute.
+\texttt{holepunch} as a side effect also creates a \texttt{DESCRIPTION}
+file which then turns any project into a research compendium
+\citep{marwick_packaging_2018}. The \texttt{Dockerfile} included with
+the project can also be used to launch a RStudio server locally, i.e.,
+independent of Binder, which is especially useful when more or special
+computational resources can be provided there. The local image usage
+reduces the number of seperately managed environments and thereby
+reduces work and increases portability and reproducibility.
+
+In \textbf{high-performance computing}, one use for containers is to run
+workflows on shared local hardware where teams manage their own
+high-performance servers. This can follow one of several design
+patterns: users may deploy containers to hardware as a work environment
+for a specific project, conatiners may provide per-user persistent
+environments, or a single container can act as a common multi-user
+environment for a server. In all cases, though, the containerized
+approach provides several advantages: First, users may use the same
+image and thus work environment on desktop and laptop computers, as
+well. The former models provide modularity, while the latter approach is
+most similar to a simple shared server. Second, software updates can be
+achieved by updating and redeploying the container, rather than tracking
+local installs on each server. Third, the containerized environment can
+be quickly deployed to other hardware, cloud or local, if more resources
+are neccessary or in case of server destruction or failure. In any of
+these cases, users need a method to interact with the containers, be it
+and IDE, or command-like access and tools such as SSH, which is usually
+not part of standard container recipes and must be added. The Rocker
+project provides containers pre-installed with the RStudio IDE. In cases
+where users store nontrivial amounts of data for their projects, data
+needs to persist beyond the life of the container. This may be via in
+shared disks, attached network volumes, or in separate storage where it
+is uploaded between sessions. In the case of shared disks or
+network-attached volumes, care must be taken to persist user
+permissions, and of course backups are still neccessary. When working
+with multiple servers, an automation framework such as
+\href{https://www.ansible.com}{Ansible} may be useful for managing
+users, permisions, and disks along with containers.
+
+Using \textbf{GPUs} (graphical processing units) as a specialised
+hardware from containerized common work environments is also possible
+and useful \citep{haydel_enhancing_2015}. GPUs are increasingly popular
+for compute-intensive machine learning tasks, e.g., deep artificial
+neural networks \citep{schmidhuber_deep_2015}. Though in this case,
+containers are not completely portable between hardware environments.
+Containers running GPU software require drivers and libraries specific
+to GPU models and versions, and containers require a specialized runtime
+to connect to the underlying GPU hardware. For NVIDIA GPUs, the
+\href{https://github.com/NVIDIA/nvidia-docker}{NVIDIA Container Toolkit}
+includes a specialized runtime plugin for Docker and a set of base
+images with appropriate drivers and libraries. The Rocker project has
+(beta) images based on these that include GPU-enabled versions of
+machine-learning R packages: \url{https://github.com/rocker-org/ml}.
+
+\textbf{Teaching} is a further example where sharing a prepared
+computing environment can greatly improve the process, especially for
+courses that require access to a relatively complex setup of software
+tools, e.g., as in the case of database systems. R is a fantastic tool
+when it comes to interfacing with databases: almost every open source
+and proprietary database system has an R package that allows users to
+connect and interact with the it This flexibility is even broader now
+that we have tools like \texttt{DBI}, that allows to create a common API
+for interfacing these databases, or like \texttt{dbplyr}, which are
+designed to run \texttt{dplyr} code straight against the database. But
+learning and teaching these tools comes with a cost: the cost of
+deploying or having access to an environment with the softwares and
+drivers installed. For people teaching R, it can become a barrier if
+they need to install local versions of the drivers, or to connect to
+remote instances which might or might not be made available by IT
+services. Giving access to a sandbox for the most common database
+environments is the idea behind
+\textbf{\href{https://github.com/ColinFay/r-db}{\texttt{r-db}}}
+\citep{ColinFay}, a Docker image that contains everything needed to
+connect, from R, to a database. Notably, with \texttt{r-db}, the users
+don't have to install complex drivers or to configure their machine in a
+specific way. On top of that, \texttt{r-db} comes with a comprehensive
+\href{http://colinfay.me/r-db/}{online guide}, explaining how to spin a
+Docker instance of a specific database, and how to use \texttt{r-db} to
+connect to it. Each package contained inside this image also has a
+series examples, so that the user can get started with the database
+right away. The \texttt{rocker/tidyverse} base image ensures that users
+can also readily use packages for analysis, display, and reporting.
+
+\textbf{\href{https://github.com/att/rcloud/tree/master/docker}{RCloud}}
+uses a \texttt{rocker/drd} base image for creation of collaborative data
+analysis and visualisation environments.
+
+\hypertarget{processing-div}{%
+\subsection{Processing {[}div{]}}\label{processing-div}}
+
+The portability of containers becomes particularly useful when complex
+processing tasks shall be offloaded to a server.
 
@@
 \item
-  Binder
-
-  \begin{itemize}
-  \tightlist
-  \item
-    \href{https://github.com/karthik/holepunch}{\texttt{holepunch}}
-    streamlines making an R research compendium Binder-ready
-    {[}@karthik{]}
-  \item
-    \href{https://repo2docker.readthedocs.io/en/latest/config_files.html}{\texttt{repo2docker}}
-    (\texttt{install.R}, \texttt{DESCRPTION}, \texttt{runtime.txt})
-    based on the Rocker images {[}@nuest{]}
-  \end{itemize}
-\item
-  GPU {[}@noamross, @cboettig{]}
-\item
-  Gigantum stack
-\item
-  in education (ready to use teaching/learning environments)
-
-  \begin{itemize}
-  \tightlist
-  \item
-    \href{https://github.com/ColinFay/r-db}{\texttt{r-db}}
-    {[}@ColinFay{]}
-  \end{itemize}
-\item
-  \href{https://github.com/att/rcloud/tree/master/docker}{RCloud} ?
-\end{itemize}
-
-\hypertarget{deployment-processing-cloud-div}{%
-\subsection{Deployment, processing, cloud
-{[}div{]}}\label{deployment-processing-cloud-div}}
-
-\begin{itemize}
-\tightlist
-\item
-  Docker images for cloud services {[}@MarkEdmondson1234{]}
+  Docker images for cloud services \citep{MarkEdmondson1234}
 
@@
 \item
-  \href{https://www.rplumber.io/docs/hosting.html\#docker}{\texttt{plumber}}
-\item
   \texttt{batchtools} \citep{Lang2017batchtools} can
@@
 \item
-  scalable deployments, e.g.~start with numerous Shiny talks mentioning
+  scalable deployments, e.g., start with numerous Shiny talks mentioning
   Rocker at useR!2017
@@
   \href{https://github.com/dynverse/dynmethods}{dynmethods}
-  {[}@rcannood{]}: In order to evaluate ±50 computational methods which
-  all used different environments (R, Python, C++, \ldots), we wrapped
+  \citep{rcannood}: In order to evaluate ±50 computational methods which
+  all used different environments (R, Python, C++, \ldots{}), we wrapped
   each of them in a docker container and can execute these methods from
@@
   will only be pushed to docker hub if the integration test succeeds.
-\item
-  \href{https://www.shinyproxy.io/}{ShinyProxy}
-  \href{https://github.com/openanalytics/shinyproxy/blob/master/src/main/java/eu/openanalytics/services/DockerService.java\#L388}{creates
-  a container} for each user
 \end{itemize}
@@
 \hypertarget{research-compendia-benmarwick}{%
-\subsection{Research Compendia
-{[}@benmarwick{]}}\label{research-compendia-benmarwick}}
+\subsection{\texorpdfstring{Research Compendia
+\citep{benmarwick}}{Research Compendia {[}@benmarwick{]}}}\label{research-compendia-benmarwick}}
 
@@
 
-\hypertarget{other-containerisation-platforms-nn}{%
-\section{Other containerisation platforms
-{[}NN{]}}\label{other-containerisation-platforms-nn}}
+\hypertarget{development-and-debugging}{%
+\subsection{Development and debugging}\label{development-and-debugging}}
+
+\label{development}
+
+Containers can also serve as useful playgrounds to create environments
+ad-hoc or to provide very specific environments that are not often
+needed. Developers can readily start a local container based on a Rocker
+image to investigate a bug report, which might require them to install a
+specific combination of package versions or use a specific R version
+(see also versioned images above). Unlike recreating the reporter's
+environment on their regular machine, a container can simply be
+discarded after use, and there is no need to tediously roll back
+explorative changes to a previous state. Using the Rocker images with
+RStudio, the disposable environments lack no development comfort (see
+also the usage in research compendia).
+
+\textbf{\href{https://github.com/wch/r-debug}{\texttt{r-debug}}} is a
+purpose built Docker image for debugging R memory problems.
+\citet{eddelbuettel_debugging_2019} describes how a Docker container was
+used to debug an issue with a package only occuring with a particular
+version of Fortran, and using tools which are not readily available on
+all platforms (e.g., not on macOS).
+
+The R-hub Docker images (see Section~\ref{rhub}) can also be used to
+debug problems locally using various combinations of Linux platforms, R
+versions, and compilers\textbackslash{}footnote\{See
+\href{https://r-hub.github.io/rhub/articles/local-debugging.html}{https://r-hub.github.io/rhub/articles/local-debugging.html}
+and
+\href{https://blog.r-hub.io/2019/04/25/r-devel-linux-x86-64-debian-clang/}{https://blog.r-hub.io/2019/04/25/r-devel-linux-x86-64-debian-clang/}.
+The images go beyond the configurations, or \emph{flavours}, used by
+CRAN for checking
+packages\footnote{\href{https://cran.r-project.org/web/checks/check_flavors.html}{https://cran.r-project.org/web/checks/check\_flavors.html}},
+e.g., with CentOS-based images, but lack a container for checking on
+Windows or OS X. The images greatly support package developers to
+provide support on operating systems they are not familiar.
+
+\hypertarget{conclusions}{%
+\section{Conclusions}\label{conclusions}}
 
@@
 \item
-  R images for Singularity
-\item
-  Running Rocker images with
-
-  \begin{itemize}
-  \tightlist
-  \item
-    Singularity
-  \item
-    CoreOS rkt?
-  \end{itemize}
-\item
-  nix?
-\end{itemize}
-
-\hypertarget{discussionoutlookconclusion-please-add-bullet-pointsall-please-add-bullet-points}{%
-\section{\texorpdfstring{Discussion/outlook/conclusion {[}@all, please
-add bullet
-points{]}}{Discussion/outlook/conclusion , please add bullet points{[}@all, please add bullet points{]}}}\label{discussionoutlookconclusion-please-add-bullet-pointsall-please-add-bullet-points}}
-
-\begin{itemize}
-\tightlist
+  what are next steps?
+\item
+  support more complex set-ups with ready-to-use stacks, e.g., GPU, AI,
+  ML
 \item
@@
 \item
-  consolidation (e.g.~via packages using \texttt{dockerfiler} and
+  consolidation (e.g., via packages using \texttt{dockerfiler} and
   \texttt{stevedore})
@@
 \item
-  \ldots{}
+  running RStudio in a container is a common approach for
+  reproducibility - \emph{is there a need to run/control containers from
+  a UI integrated in RStudio?}
+\item
+  moving away from Docker focus towards containers?
+
+  \begin{itemize}
+  \tightlist
+  \item
+    is there a need for a wrapper for
+    \href{https://github.com/containers/libpod/blob/master/API.md}{podman's
+    Varlink-API} (similar to stevedore) and buildah to manage containers
+    and images respectively with alternative software stacks?
+  \item
+    will there be control packages for Singularity support?
+  \end{itemize}
+\item
+  sustain the overview and structure in this article in form of a CRAN
+  Task View
 \end{itemize}
@@
 DN conceived of the presented idea and
-\href{https://github.com/nuest/rockerverse-paper/issues/3}{initialised the formation of the writing team}.
-CB .. RC .. DE .. ME .. CF .. BW .. KR .. NR .. NX wrote the section on
-liftr. LS \& NT wrote the section on Bioconductor. All authors
-contributed to the discussion and outlook section and approved the final
-version. This articles was collaboratively written at
+\href{https://github.com/nuest/rockerverse-paper/issues/3}{initialised the formation of the writing team},
+wrote the section about images for R distributions, wrote the
+conclusions section, and revised all sections. CB .. RC and EH wrote the
+section on interfaces for Docker in R. DE wrote the introduction and
+section about Rocker. ME .. CF wrote the paragraphs about
+\texttt{r-online}, \texttt{dockerfiler}, \texttt{r-ci} and
+\texttt{r-db}. BW .. KR .. NR .. NX wrote the section on liftr. LS \& NT
+wrote the section on Bioconductor. All authors contributed minor edits,
+gave feedback on other sections, and approved the final version.
+
+This articles was collaboratively written at
 \href{https://github.com/nuest/rockerverse-paper/}{https://github.com/nuest/rockerverse-paper/}.
 The
-\href{https://github.com/nuest/rockerverse-paper/graphs/contributors}{contributors page}
+\href{https://github.com/nuest/rockerverse-paper/graphs/contributors}{contributors page},
+\href{https://github.com/nuest/rockerverse-paper/commits/master}{commit history},
 and
-\href{https://github.com/nuest/rockerverse-paper/commits/master}{commit history}
+\href{https://github.com/nuest/rockerverse-paper/issues/}{discussion issues}
 provide a detailed view on the respective contributions.
+
+\hypertarget{acknowledgements}{%
+\section{Acknowledgements}\label{acknowledgements}}
+
+DN is supported by the project Opening Reproducible Research
+(\href{https://www.uni-muenster.de/forschungaz/project/12343}{o2r})
+funded by the German Research Foundation (DFG) under project number
+\href{https://gepris.dfg.de/gepris/projekt/415851837}{PE~1632/17-1}. The
+funders had no role in data collection and analysis, decision to
+publish, or preparation of the manuscript.
 
@@
 Daniel Nüst\\
-Institute for Geoinformatics, University of Münster\\
-Heisenbergstr. 2, 48149 Münster, Germany\\
+University of Münster\\
+Institute for Geoinformatics\\ Heisenbergstr. 2\\ 48149 Münster, Germany\\ \orcid{0000-0002-0024-5046}\\
 }
 \href{mailto:daniel.nuest@uni-muenster.de}{\nolinkurl{daniel.nuest@uni-muenster.de}}
-
-\address{%
-Carl Boettiger\\
-\\
-\\
-}
 
@@
 Robrecht Cannoodt\\
-\\
-\\
-}
+Ghent University\\
+Data Mining and Modelling for Biomedicine group\\ VIB Center for Inflammation Research\\ Technologiepark 71\\ 9052 Ghent, Belgium\\ \orcid{0000-0003-3641-729X}\\
+}
+\href{mailto:robrecht@cannoodt.dev}{\nolinkurl{robrecht@cannoodt.dev}}
 
@@
 University of Illinois at Urbana-Champaign\\
-\\
+Department of Statistics\\ Illini Hall, 725 S Wright St\\ Champaign, IL 61820, USA\\ \orcid{0000-0001-6419-907X}\\
 }
@@
 ThinkR\\
-\\
+5O rue Arthur Rimbaud\\ 93300 Aubervilliers, France\\ \orcid{0000-0001-7343-1846}\\
 }
 \href{mailto:contact@colinfay.me}{\nolinkurl{contact@colinfay.me}}
+
+\address{%
+Karthik Ram\\
+Berkeley Institute for Data Science\\
+University of California\\ Berkeley, CA 94720, USA\\ \orcid{0000-0002-0233-1757}\\
+}
+\href{mailto:karthik.ram@berkeley.edu}{\nolinkurl{karthik.ram@berkeley.edu}}
+
+\address{%
+Noam Ross\\
+EcoHealth Alliance\\
+460 W 34th St., Ste. 1701\\ New York, NY 10001, USA\\ \orcid{0000-0002-0233-1757}\\
+}
+\href{mailto:ross@ecohealthalliance.org}{\nolinkurl{ross@ecohealthalliance.org}}
+
+\address{%
+Nan Xiao\\
+Seven Bridges Genomics\\
+529 Main St, Suite 6610\\ Charlestown, MA 02129, USA\\ \orcid{0000-0002-0250-5673}\\
+}
+\href{mailto:me@nanx.me}{\nolinkurl{me@nanx.me}}
+
+\address{%
+Lori Shepherd\\
+Roswell Park Comprehensive Cancer Center\\
+Elm \& Carlton Streets\\ Buffalo, New York, 14263, USA\\ \orcid{0000-0002-5910-4010}\\
+}
+\href{mailto:lori.shepherd@roswellpark.org}{\nolinkurl{lori.shepherd@roswellpark.org}}
+
+\address{%
+Nitesh Turaga\\
+Roswell Park Comprehensive Cancer Center\\
+Elm \& Carlton Streets\\ Buffalo, New York, 14263, USA\\ \orcid{0000-0002-0224-9817}\\
+}
+\href{mailto:nitesh.turaga@roswellpark.org}{\nolinkurl{nitesh.turaga@roswellpark.org}}
+
+\address{%
+Jacqueline Nolis\\
+Nolis, LLC\\
+Seattle, WA, USA\\ \orcid{0000-0001-9354-6501}\\
+}
+\href{mailto:jacqueline@nolisllc.com}{\nolinkurl{jacqueline@nolisllc.com}}
+
+\address{%
+Heather Nolis\\
+T-Mobile\\
+12920 Se 38th St.\\ Bellevue, WA, 98006\\
+}
+\href{mailto:heather.wensler1@t-mobile.com}{\nolinkurl{heather.wensler1@t-mobile.com}}
+
+\address{%
+Hong Ooi\\
+Microsoft\\
+Level 5, 4 Freshwater Place\\ Southbank, VIC 3006, Australia\\
+}
+\href{mailto:hongooi@microsoft.com}{\nolinkurl{hongooi@microsoft.com}}
+
+\address{%
+Ellis Hughes\\
+Fred Hutchinson Cancer Research Center\\
+Vaccine and Infectious Disease\\ 1100 Fairview Ave. N., P.O. Box 19024\\ Seattle, WA 98109-1024, USA\\
+}
+\href{mailto:ehhughes@fredhutch.org}{\nolinkurl{ehhughes@fredhutch.org}}
 
@@
 }
-
-\address{%
-Karthik Ram\\
-Berkeley Institute for Data Science, University of California, Berkeley,
-CA 94720 USA\\
-\\
-}
-\href{mailto:karthik.ram@berkeley.edu}{\nolinkurl{karthik.ram@berkeley.edu}}
-
-\address{%
-Noam Ross\\
-\\
-\\
-}
-
-\address{%
-Nan Xiao\\
-Seven Bridges Genomics\\
-529 Main St, Suite 6610 \newline Charlestown, MA 02129, USA
-\newline ORCiD: 0000-0002-0250-5673\\
-}
-\href{mailto:me@nanx.me}{\nolinkurl{me@nanx.me}}
-
-\address{%
-Lori Shepherd\\
-Roswell Park Comprehensive Cancer Center\\
-Elm \& Carlton Streets, Buffalo, New York, 14263 USA\\
-}
-\href{mailto:lori.shepherd@roswellpark.org}{\nolinkurl{lori.shepherd@roswellpark.org}}
-
-\address{%
-Nitesh Turaga\\
-Roswell Park Comprehensive Cancer Center\\
-Elm \& Carlton Streets, Buffalo, New York, 14263 USA\\
-}
-\href{mailto:nitesh.turaga@roswellpark.org}{\nolinkurl{nitesh.turaga@roswellpark.org}}

### 2020-01-10 manuscript.Rmd
add rstudio information on docker images and use cases from our customers

@@
     email: ehhughes@fredhutch.org
+  - name: Sean Lopp
+    # https://github.com/slopp
+    affiliation: RStudio, Inc
+    address:
+      - RStudio
+      - 250 Northern Ave
+      - Boston, MA 02210
+    email: sean@rstudio.com
   - name: Ben Marwick
@@
 
+RStudio also maintains a set of [Docker images](https://github.com/rstudio/r-docker) that provide an [opinionated R installation](https://github.com/rstudio/r-builds) built with the following considerations:
+
+- They use a minimal set of build and runtime dependencies. 
+- R is installed in a versioned directory
+- The R installation and profile is consistent across the different Linux distributions.
+
+Images are provided for multiple operating systems including SUSE, Ubuntu, CentOS, and Debian.
+These images are also responsible for creating pre-compiled binary R packages for Linux, which supplement the binaries built by CRAN.
+These pre-compiled packages dramatically decrease the installation time of R packages on Linux, which is especially important in decreasing the build time of Docker images.
+More informaiton is available [here](https://blog.rstudio.com/2019/11/07/package-manager-v1-1-no-interruptions/).
+
 There are also independent projects installing R or RStudio on [**Alpine Linux**](https://www.alpinelinux.org/), but not beyond the proof of concept stage\footnote{\href{https://github.com/artemklevtsov/r-alpine}{https://github.com/artemklevtsov/r-alpine} or \href{https://www.github.com/velaco/alpine-r}{https://www.github.com/velaco/alpine-r} provide different use case images but are not recently maintained (development of the latter documented in \citet{ratesic_building_2018}); \href{https://github.com/cmplopes/alpine-r}{https://github.com/cmplopes/alpine-r} has only base R images, is relatively up-to-date but sparsely documented.}.
@@
 ```
+
+A general description of what to consider when creating a Docker image used for data science projects is provided on [environments.rstudio.com](https://environments.rstudio.com/docker).
 
@@
 
+## Using Docker in a Data Science Platform
+
+In addition to embedding RStudio Open Source Server inside of Docker images, such as Rocker, many organizations use Docker through RStudio Team - RStudio's modular data science platform.
+These professional products allow teams of data scientists and their respective IT/DevOps groups to develop and deploy code in R and Python, using the benefits of containers, without requiring users to learn new tools or directly interact with containers.
+More information for scaling RStudio is [available here](https://solutions.rstudio.com/launcher/overview/).
+Notable patterns include:
+
+1) To deploy and manage RStudio Professional Products
+Many RStudio customers are managing professional products in Docker containers as an alternative to bare-metals servers or virtual machines.
+Best practices for [running RStudio with Docker containers](https://support.rstudio.com/hc/en-us/articles/360021594513-Running-RStudio-with-Docker-containers) as well as [Docker images](https://github.com/rstudio/rstudio-docker-products) for our commerical products are available.
+
+2) To run interactive RStudio sessions on a cluster
+In the RStudio Server Pro 1.2 release in 2019, we added new functionality called Launcher, which gives our users the ability to spawn RStudio sessions on Kubernetes.
+We provide [Docker images](https://support.rstudio.com/hc/en-us/articles/360019253393-Using-Docker-images-with-RStudio-Server-Pro-Launcher-and-Kubernetes) as a starting point for users to work with Kubernetes.
+Some of our customers are also using Launcher with Slurm clusters, and optionally, with Singularity containers.
+A key benefit of Launcher is the ability for R and Python users to interactively work in RStudio without learning about Docker at all - while still leveraging containers and Kubernetes.
+Launcher also enables users to submit batch jobs to Kubernetes clusters without writing specific deployment scripts.
+
+3) To deploy applications into production
+Most RStudio customers use [RStudio Connect](https://rstudio.com/products/connect/) to deploy dashboards, REST APIs, and other R and Python content into production.
+RStudio Connect uses an approach to isolate deployed R applications that is similar to Docker's filesystem and process namespacing.
+Many customers benefit from using [RStudio Package Manager](https://rstudio.com/products/package-manager/) to [manage their R environments and package versions with Docker](https://environments.rstudio.com/docker) to ensure that the images are built deterministically.
+
 ## Processing [div]

### 2020-01-13 manuscript.Rmd
add link for dockr

--- 2020-01-10 manuscript.Rmd
@@
 - [`containerit`](https://github.com/o2r-project/containerit/)
-- [`dockr`](https://github.com/smaakage85/dockr)
+- [`dockr`](https://github.com/smaakage85/dockr), see also http://smaakage85.netlify.com/2019/12/21/dockr-easy-containerization-for-r/
 - [`dockertest`](https://github.com/traitecoevo/dockertest/)

### 2020-01-13 manuscript.Rmd

@@
 - `RSelenium`
-- [`googleComputeEngineR`](https://cloudyr.github.io/googleComputeEngineR/) (function `docker_run`)
-  - I don't think `docker_run` is main focus for containers for this package, its more how it uses Docker to launch custom environments that can be templated (e.g., rstudio, shiny etc.) and for parrallisation e.g., `library(future)`  using GCE VMs as endpoints.
-[@MarkEdmondson1234]
+- [`googleComputeEngineR`](https://cloudyr.github.io/googleComputeEngineR/) (function `gce_vm_template()`)
+  - To enable quick deployments of key R services such as RStudio and Shiny onto cloud virtual machines (VMs), this package utilises Dockerfiles to move the labour of setting up those services from the user to a premade Docker image.
+For example, by specifying the template `template="rstudio"` in `gce_vm_template()/gce_vm()` an up to date RStudio Service image is launched.
+Specifying `template="rstudio-gpu"` will launch an RStudio Server image with a GPU attached, etc.
 - [`analogsea`](https://github.com/sckott/analogsea) (digital ocean R client)
+- [`plumber`](https://www.rplumber.io/docs/hosting.html#docker)
+- [production deployment of neural networks](https://github.com/tmobile/r-tensorflow-api) [@jnolis, @nolistic]
+- Kubernetes - A popular platform for managing Docker containers is [Kubernetes](https://kubernetes.io/), which is used for a wide variety of containerized applications.
+It may be your organisation has a Kubernetes cluster already for other applications.
+Docker containers are used within Kubernetes clusters to hold native code, for which Kubernetes creates a framework around network connections and scaling of resources up and down.
+An introduction on using [Kubernetes with R is published at this blog post](https://code.markedmondson.me/r-on-kubernetes-serverless-shiny-r-apis-and-scheduled-scripts/).
 - [`AzureContainers`](https://cran.r-project.org/package=AzureContainers): Umbrella package for working with containers in Microsoft's Azure cloud.
@@
   - For this project, we use travis-ci to build rocker-derived containers, test them, and only push them to docker hub (from travis-ci.org) if the integration tests succeed.
-- Google Cloud Build [@MarkEdmondson1234]
+- Google Cloud Build
   - https://cloud.google.com/cloud-build/
-  - For me this is what makes Docker containers viable, as it builds the `Dockerfile`s on each GitHub commit.
-It couples with Google Container Registry to build private Docker images, for downstream applications.
+  - Cloud Build is another continuous intergation service that helps move the workload of building the Docker images to an online service.
+Cloud Build can be set up to build the Dockerfiles on each GitHub commit or release.
+This means you do not need to build the Docker images locally, which can tie up resources since Docker images can be several GBs and take a long time to compile.
+Google Cloud Build works alongside Google Container Registry to allow you to build private and public Docker images, which allows you to build up your own dependency graphs for downstream applications.
 
@@
 
-- Docker images for cloud services [@MarkEdmondson1234]
-  - the most popular functionality for `googleComputeEngineR` is its use of Docker to enable parallel processing across VMS - [some demos here](https://cloudyr.github.io/googleComputeEngineR/articles/massive-parallel.html)
-- Google Cloud Run - CaaS (Containers as a Service) that lets you launch a Docker container without worrying about underlying infrastructure.
-An R implementation is shown here at [cloudRunR](https://github.com/MarkEdmondson1234/cloudRunR) which uses it to create a scalable R plumber API.
+- Docker images for cloud services
+  - A popular application of using Docker is to create the environments that can run in parallel for speeding up R code.
+For example, [`googleComputeEngineR`](https://CRAN.R-project.org/package=googleComputeEngineR)'s `gce_vm_cluster()` function can create clusters of 2 or more virtual machines, running multi-CPU architectures.
+Instead of running a local R script with the local CPU and RAM restrictions, the same code can be processed on all CPU threads of the cluster of machines in the cloud, all running in a Docker container with the same R environments.
+This is achieved through `googleComputeEngineR`'s integration with the R parralisation library [`library(future)` by Henrik Bengtsson]( https://CRAN.R-project.org/package=future).
+Local R computation can be thrown up to a multi-CPU and VM environment to achieve parrell computation in a few lines of R code in your local session.
+- [some demonstrations are available here](https://cloudyr.github.io/googleComputeEngineR/articles/massive-parallel.html).
+- [Google Cloud Run](https://cloud.run) - This is a CaaS (Containers as a Service) that lets you launch a Docker container without worrying about underlying infrastructure.
+This dispenses with the developer creating a cloud server to run the Docker image on, by abstracting away those servers to a more serverless configuration.
+Cloud Run lets you run your code on top of a managed or your own Kubernetes cluster running Knative, and can accept any Docker image.
+The service takes care of network ingress, scaling machines up and down to zero, authentication and authorisation, all features which are non-trivial for a developer to create on their own.
+This can be used to scale up R code to millions of instances if they need to, with little or no changes to existing code.
+An R implementation is shown here at [cloudRunR](https://github.com/MarkEdmondson1234/cloudRunR) which uses Cloud Run to create a scalable R plumber API.
 - `batchtools` \citep{Lang2017batchtools} can [schedule jobs with Docker Swarm](https://mllg.github.io/batchtools/reference/makeClusterFunctionsDocker.html)

### 2020-01-13 manuscript.Rmd
rewording of RStudio section

--- 2020-01-13 manuscript.Rmd
@@
 
+**RStudio** also maintains a set of [Docker images](https://github.com/rstudio/r-docker) for multiple operating systems including SUSE, Ubuntu, CentOS, and Debian.
+The image stacks include an [opinionated R installation](https://github.com/rstudio/r-builds) and ensure the R installation and profile is consistent across the different Linux distributions.
+The images also install R in a versioned directory and use a minimal set of build and runtime dependencies,
+Furthermore these images are responsible for creating pre-compiled binary R packages for Linux, which supplement the binaries built by CRAN.
+These pre-compiled packages dramatically decrease the installation time of R packages on Linux and subsequently can decrease the build time of Docker images \citep{lopp_package_2019}.
+
 There are also independent projects installing R or RStudio on [**Alpine Linux**](https://www.alpinelinux.org/), but not beyond the proof of concept stage\footnote{\href{https://github.com/artemklevtsov/r-alpine}{https://github.com/artemklevtsov/r-alpine} or \href{https://www.github.com/velaco/alpine-r}{https://www.github.com/velaco/alpine-r} provide different use case images but are not recently maintained (development of the latter documented in \citet{ratesic_building_2018}); \href{https://github.com/cmplopes/alpine-r}{https://github.com/cmplopes/alpine-r} has only base R images, is relatively up-to-date but sparsely documented.}.
@@
 ```
+
+A general description of what to consider when creating a Docker image used for data science projects in R is provided on [environments.rstudio.com](https://environments.rstudio.com/docker).
 
@@
 
+## Using Docker in a Data Science Platform
+
+In addition to embedding RStudio Open Source Server inside of Docker images, such as Rocker, RStudio offers the moduler data science platform [RStudio Team](https://rstudio.com/products/team/), a commercial product allowing teams of data scientists and their respective IT/DevOps groups to develop and deploy code in R and Python.
+RStudio team enables these groups to use the benefits of containers, without requiring users to learn new tools or directly interact with containers.
+Notable patterns include:
+
+1. _To deploy and manage RStudio Professional Products:_
+Many RStudio customers are managing professional products in Docker containers as an alternative to bare-metals servers or virtual machines.
+Best practices for [running RStudio with Docker containers](https://support.rstudio.com/hc/en-us/articles/360021594513-Running-RStudio-with-Docker-containers) as well as [Docker images](https://github.com/rstudio/rstudio-docker-products) for RStudio's commerical products are available.
+1. _To run interactive RStudio sessions on a cluster:_
+In the RStudio Server Pro 1.2 release in 2019, RStudio added new functionality called [Launcher](https://solutions.rstudio.com/launcher/overview/).
+It gives users the ability to spawn RStudio instances and background jobs in a scalable way on external clusters, e.g., [Kubernetes based on Docker images](https://support.rstudio.com/hc/en-us/articles/360019253393-Using-Docker-images-with-RStudio-Server-Pro-Launcher-and-Kubernetes) or [Slurm](https://slurm.schedmd.com/) clusters, and optionally, with Singularity containers.
+A key benefit of Launcher is the ability for R and Python users to interactively work in RStudio without learning about Docker at all - while still leveraging containers and Kubernetes.
+Users can submit batch jobs to Kubernetes clusters without writing specific deployment scripts.
+
+[RStudio Connect](https://rstudio.com/products/connect/) is a commercial publishing platform to deploy dashboards, REST APIs, and other R and Python content into production. 
+RStudio Connect uses an approach to isolate deployed R applications that is similar to Docker's filesystem and process namespacing and leverages the [RStudio Package Manager](https://rstudio.com/products/package-manager/) to manage R environments including system dependencies.
+RStudio Package manager may also be used to [manage their R environments and package versions with Docker](https://environments.rstudio.com/docker) and to ensure deterministic image builds.
+
 ## Processing [div]

### 2020-01-13 manuscript.Rmd
clarify containers vs. VMs in introduction

@@
 By allowing (one or possibly) multiple applications or services to run concurrently on one host machine without any fear of interference between them, an important scalability opportunity is being provided.
-But Docker improved on virtualization by accessing the host system---generally Linux---through a much thinner and smaller shim than a full operating system emulation or virtualization.
+But Docker improved this compartmentalization by accessing the host system---generally Linux---through a much thinner and smaller shim than a full operating system emulation or virtualization.
+This containerization is also called operating-system-level virtualization \ref{wikipedia_contributors_os-level_2020}.
+Typically a container runs one process, whereas virtualization may run whole operating systems at a larger footprint.
 This makes for more efficient use of system resources \citep{felter_updated_2015} and allowed another order of magnitude in terms of scalability of deployment \citep[cf.][]{datadog_8_2018}.
@@
 
-The key accomplishment of Docker (as an "application") is to make a "bundled" aggregation of software (the so-called "image") available to any system equipped to run Docker, without requiring much else from the host besides the actual Docker application installation.
-This is a rather attractive, and novel, proposition which has lead to widespread adoption and use of Docker in a variety of domains, _e.g._, cloud computing infrastructure \citep{Bernstein2014}, data science \citep{boettiger_introduction_2015}, and edge computing \citep{alam_orchestration_2018} [more examples!].
-It provided to be a natural match for "cloud deployment" which runs (or at least appears to run) "seamlessly" without much explicit reference to the underlying machine, architecture or operating system: containers are portable and can be deployed with very little in terms of dependencies on the host system---only the container runtime is required.
+The key accomplishment of Docker as an "application" is to make a "bundled" aggregation of software (the so-called "image") available to any system equipped to run Docker, without requiring much else from the host besides the actual Docker application installation.
+This is a rather attractive proposition and Docker's very easy to use user interface has lead to widespread adoption and use of Docker in a variety of domains, e.g., cloud computing infrastructure \citep{Bernstein2014}, data science \citep{boettiger_introduction_2015}, and edge computing \citep{alam_orchestration_2018} [more examples!].
+It provided to be a natural match for "cloud deployment" which runs, or at least appears to run, "seamlessly" without much explicit reference to the underlying machine, architecture or operating system: containers are portable and can be deployed with very little in terms of dependencies on the host system---only the container runtime is required.
 Images are normally build from plain text documents called `Dockerfile`.

### 2020-01-13 manuscript.Rmd
reformatting and fixing cross-references

@@
 **FastR** is "A high-performance implementation of the R programming language, built on GraalVM" (https://github.com/oracle/fastr).
-It is developed by Oracle, connects R to the GraalVM ecosystem \citep{wikipedia_graalvm_2019}, and also claims superior performance but also targets full compatibility with base R \footnote{\href{https://github.com/oracle/fastr}{https://github.com/oracle/fastr}}.
+It is developed by Oracle, connects R to the GraalVM ecosystem \citep{wikipedia_graalvm_2019}, and also claims superior performance but also targets full compatibility with base R\footnote{\href{https://github.com/oracle/fastr}{https://github.com/oracle/fastr}}.
 There are no offical Docker images provided, but `nuest/fastr` images and `Dockerfile` are independently maintained on [Docker Hub](https://cloud.docker.com/repository/docker/nuest/fastr/) and [GitHub](https://github.com/nuest/fastr-docker) respectively.
@@
 
-[_Bioconductor_](http://bioconductor.org/) is an open source, open development project for the analysis and comprehension of genomic data \citep{gentleman_bioconductor_2004}.
+_Bioconductor_ ([https://bioconductor.org/](https://bioconductor.org/)) is an open source, open development project for the analysis and comprehension of genomic data \citep{gentleman_bioconductor_2004}.
 The project consists of 1741 R software packages as of August 15th 2019, as well as packages containing annotation or experiment data.
@@
 They required an extra tool, [Docker Toolbox](https://docs.docker.com/toolbox/), which by now only exists as a legacy solution for older Windows systems.
-Docker toolbox leverages `docker-machine` (see also Section&nbsp;\ref{interfaces} for an R package interfacing with `docker-machine`) to handle the process of creating a local virtual host which could host Docker Engine, while exposing the regular Docker CLI.
+Docker toolbox leverages `docker-machine` (see also Section&nbsp;\nameref{interfaces} for an R package interfacing with `docker-machine`) to handle the process of creating a local virtual host which could host Docker Engine, while exposing the regular Docker CLI.
 The Docker CLI commands are forwarded to the virtual host transparently for the user.
@@
 
-**[`rocker-win`](https://github.com/nuest/rocker-win)** is a proof of concept for running R in Windows-based containers \citep{nust_rocker-win_2019}.
+**rocker-win** ([https://github.com/nuest/rocker-win](https://github.com/nuest/rocker-win)) is a proof of concept for running R in Windows-based containers \citep{nust_rocker-win_2019}.
 It provides selected R versions using three different base images using `microsoft/windowsservercore` and `mcr.microsoft.com/windows/servercore`.
@@
 \label{rhub}
-The **R-hub** project provides _"a collection of services to help R package development", with the package builder as the most prominent one\citep{r-hub_docs_2019}.
+The **R-hub** project provides _"a collection of services to help R package development"_, with the package builder as the most prominent one\citep{r-hub_docs_2019}.
 The builder allows R package developers to check their R package on different platforms and R versions using a web form or the package \pkg{rhub} \citep{csardi_rhub_2019}.
@@
 The `Dockerfile`s and images are published on Docker Hub\footnote{\href{https://hub.docker.com/u/rhub}{https://hub.docker.com/u/rhub}} and GitHub\footnote{\href{https://github.com/r-hub/rhub-linux-builders}{https://github.com/r-hub/rhub-linux-builders}} respectively.
-The images comprise release and development builds of R running in the base images of the [Debian](https://www.debian.org/), [Ubuntu](https://ubuntu.com/), [Fedora](https://getfedora.org/), and [CentOS](https://centos.org/) linux distributions (Arch Linux is under development\footnote{\href{https://github.com/r-hub/rhub-linux-builders/pull/41}{https://github.com/r-hub/rhub-linux-builders/pull/41}}, `see `rhub::local_check_linux_images()` for a full list), and are intended for debugging (see Section&nbsp;\ref{development}) not for reproducible workflows.
+The images comprise release and development builds of R running in the base images of the [Debian](https://www.debian.org/), [Ubuntu](https://ubuntu.com/), [Fedora](https://getfedora.org/), and [CentOS](https://centos.org/) linux distributions (Arch Linux is under development\footnote{\href{https://github.com/r-hub/rhub-linux-builders/pull/41}{https://github.com/r-hub/rhub-linux-builders/pull/41}}, see `rhub::local_check_linux_images()` for a full list), and are intended for debugging (see Section&nbsp;\nameref{development}) not for reproducible workflows.
 _The rationale_ of not re-using the Rocker image stack is the high number of operatings systems and configurations, which can be more uniformly maintained in a seperate suite of images.
@@
 
-**[_liftr_](https://nanx.me/liftr/)** \citep{liftr2019} aims to solve the problem of persistent reproducible reporting in statistical computing.
+**\pkg{liftr}** ([https://nanx.me/liftr/](https://nanx.me/liftr/)) \citep{liftr2019} aims to solve the problem of persistent reproducible reporting in statistical computing.
 Currently, the R Markdown format and its backend compilation engine _knitr_ offer a _de facto_ standard for creating dynamic documents \citep{xie2018}.
@@
 R has been historically viewed as a tool for analysis and scientific research--not for creating software that corporations can rely on to continously run in real time.
-However, thanks to advancements in R running as a web service, along with with the ability to deploy R in docker containers, modern enterprises are now capable of having real-time machine learning powered by R.
+However, thanks to advancements in R running as a web service, along with with the ability to deploy R in Docker containers, modern enterprises are now capable of having real-time machine learning powered by R.
 
-The backbone of this work is the _[plumber](https://www.rplumber.io/docs/hosting.html#docker)_ package, created by Jeff Allen.
+The backbone of this work is the \pkg{plumber} ([https://www.rplumber.io/](https://www.rplumber.io/)) package.
 The plumber package allows R to be used to host web services through RESTful APIs.
@@
 
-[`holepunch`](https://github.com/karthik/holepunch) is an R package that was designed to make sharing work environments accessible to novice R users based on _Binder_.
+**\pkg{holepunch}** ([https://github.com/karthik/holepunch](https://github.com/karthik/holepunch)) is an R package that was designed to make sharing work environments accessible to novice R users based on **Binder**.
 The [Binder project](https://mybinder.readthedocs.io/en/latest/), maintained by the team behind Jupyter, makes it possible for users to create and share their computing environments with others \citep{jupyter_binder_2018}.
@@
 To set up Binder for a project, a user typically starts at an instance of a BinderHub and passes the location of a repository with a workspace, e.g., a hosted Git repository, or a data repository like Zenodo.
-Binder's core internal tool is [repo2docker](https://repo2docker.readthedocs.io/en/latest/config_files.html).
+Binder's core internal tool is [`repo2docker`](https://repo2docker.readthedocs.io/en/latest/config_files.html).
 It deterministically builds a Docker image by parsing the contents of a repository, e.g., project dependency configurations or simple configuration files.
@@
 
-The R-hub Docker images (see Section&nbsp;\ref{rhub}) can also be used to debug problems locally using various combinations of Linux platforms, R versions, and compilers\footnote{See \href{https://r-hub.github.io/rhub/articles/local-debugging.html}{https://r-hub.github.io/rhub/articles/local-debugging.html} and \href{https://blog.r-hub.io/2019/04/25/r-devel-linux-x86-64-debian-clang/}{https://blog.r-hub.io/2019/04/25/r-devel-linux-x86-64-debian-clang/}.
+The R-hub Docker images (see Section&nbsp;\nameref{rhub}) can also be used to debug problems locally using various combinations of Linux platforms, R versions, and compilers\footnote{See \href{https://r-hub.github.io/rhub/articles/local-debugging.html}{https://r-hub.github.io/rhub/articles/local-debugging.html} and \href{https://blog.r-hub.io/2019/04/25/r-devel-linux-x86-64-debian-clang/}{https://blog.r-hub.io/2019/04/25/r-devel-linux-x86-64-debian-clang/}.
 The images go beyond the configurations, or _flavours_, used by CRAN for checking packages\footnote{\href{https://cran.r-project.org/web/checks/check_flavors.html}{https://cran.r-project.org/web/checks/check\_flavors.html}}, e.g., with CentOS-based images, but lack a container for checking on Windows or OS X.

### 2020-01-13 manuscript.Rmd
add section on dockermachine, use .Renviron for credentials, so ignore it

@@
 
-**dockermachine** ([https://github.com/cboettig/dockermachine](https://github.com/cboettig/dockermachine))
-<!-- From the github description:
-**dockermachine** ([https://github.com/cboettig/dockermachine](https://github.com/cboettig/dockermachine)) provides a convenient interface to docker-machine from the R command line.
-This will allow R users to easily create, control, and destroy cloud instances across a wide variety of providers.
--->
+**\pkg{dockermachine}** ([https://github.com/cboettig/dockermachine](https://github.com/cboettig/dockermachine)) is an R package to provide a convenient interface to [Docker Machine](https://docs.docker.com/machine/overview/) from R.
+Docker Machine's CLI tool `docker-machine` allows users to create and manage virtual host on local computers, local data centers, or at cloud providers.
+A local Docker installation can be configured to forward all commands issues on the local Docker CLI to a selected (remote) virtual host fully transparent.
+For local Docker Machine was especially crucial in early days of Docker, when no native Docker support was available for Mac or Windows computers, but remains relevant for provisioning on remote systems.
+The package has not received any updates for two years, but was functional with a current version of `docker-machine` (`0.16.2`) and potentially lowers the barriers for R users to run containers on various hosts, if using the Docker Machine CLI is perceived as a barrier.
 
-**AzureContainers** ([https://cran.r-project.org/package=AzureContainers](https://cran.r-project.org/package=AzureContainers)) is an interface to a number of container-related services in Microsoft's Azure cloud, namely Container Instances, Container Registry and Kubernetes Service.
-While it is mainly intended for working with Azure, as a convenience feature it includes lightweight, cross-platform shells to Docker, kubectl and helm.
-These can be used to create and manage arbitrary Docker images and containers, as well as Kubernetes clusters on any platform or cloud service.
-
-**babelwhale** ([https://cran.r-project.org/package=babelwhale](https://cran.r-project.org/package=babelwhale)) 
-allows executing and interacting with containers, which can use either Docker or Singularity as a backend.
-provides a unified interface to interact with Docker and Singularity containers. You can execute a command inside a container, mount a volume or copy a file.
-
-**harbor** ([https://github.com/wch/harbor](https://github.com/wch/harbor)
-<!-- from the github description
-is an R package for controlling docker containers on local and remote hosts.
--->
+```{r dockermachine, eval=FALSE, include=FALSE}
+library("dockermachine")
+dockermachine::machine_create(driver = "amazonec2")
+# machine started: https://console.aws.amazon.com/ec2/v2/home?region=us-east-1#Instances:sort=instanceId
+# get config on command line: $ docker-machine env machine
+# activate: $ $(docker-machine env test) does not work, because of "
+# run container: docker run -it -p 80:80 nginx
+# allow inbound HTTP traffic
+# http://ec2-3-85-85-61.compute-1.amazonaws.com/ shows "Welcome to nginx"!
+# Nice!
+# > dockermachine::machine_rm()
+# Instance shut down.
+```
 

### 2020-01-13 manuscript.Rmd
add first version of conclusions
closes #31 by mentioning rkt as potential future work

@@
 
-- what are next steps?
-- support more complex set-ups with ready-to-use stacks, e.g., GPU, AI, ML
-- Missing pieces?
-- consolidation (e.g., via packages using `dockerfiler` and `stevedore`)
-- Common themes
-  - reproducibility
-- will knowledge about containers continue to spread?
-- what is needed for even more containers with R?
-- the ability to move processing between services easily (e.g locally, one cloud providers VM, another cloud provider's Container-as-a-Service)
-- running RStudio in a container is a common approach for reproducibility - _is there a need to run/control containers from a UI integrated in RStudio?_
-- moving away from Docker focus towards containers?
-  - is there a need for a wrapper for [podman's Varlink-API](https://github.com/containers/libpod/blob/master/API.md) (similar to stevedore) and buildah to manage containers and images respectively with alternative software stacks?
-  - will there be control packages for Singularity support?
-- sustain the overview and structure in this article in form of a CRAN Task View
+This article is a snapshot of the R-corner in a universe of applications built many-faced piece of software, Docker.
+Docker became synonymous with applying the concept of containerisation to solve many challenges, such as reproducible environments in research, scalable deployments in heterogenous organisations, and the ability to move processing between machines easily (e.g., locally, one cloud providers VM, another cloud provider's Container-as-a-Service).
+These are probably the thre common themes described in this work.
+
+The R packages and use cases presented show the positive drive in the community to innovate, but also the challenge of keeping up with continuous evolving landscape.
+The use cases contributed by co-authors also have a degree of overlap.
+A common language and understanding of good practices is still taking shape.
+The easiness with which one can create a complex software systems, such as an independent Docker image stack, to serve one's specific needs currently leads to some fragmentation.
+It is unclear if Docker alone as a core technology is really a feasible level of collaboration and abstraction.
+Furthermore fragmentation may not be a bad sign but instead a reflection of a growing market, which is able to sustain multiple related efforts.
+With the maturing of core building blocks, such as the Rocker suite of images, more systems will be built successfully but will also be behind the curtains.
+
+However, at least _on the level of R packages_, some consolidation seems in order, e.g., to reduce the number of packages creating `Dockerfile`s from R code or controlling the Docker daemon with R code.
+It remains to be seen which approach to control Docker, via the Docker API as `stevedore` or via system calls as `dockyard`/`docker`/`dockr`, is more sustainable, or if the question will be answered by the endurance of maintainers and sufficient funding.
+Similarly, the capturing of environments and their serialization in form of a `Dockerfile` currently happens at different levels of abstraction and re-use of functionality seems reasonable, e.g.
+`liftr` could generate the environment with `containerit`, which in turn may use `dockerfiler` for low level objects representing a `Dockerfile`.
+In this consolidation, the Rocker Project could play the role of a unifying and coordinating entity in the future.
+Though for the moment, the sign of the times points to more experimentation and feature growth, e.g. images for GPU-based computing and artificial intelligence.
+Even with coding being more and more accepted as a required, and achievable skill, an easier access, for example by exposing containerisation benefits via simple user interfaces in the users' IDE, could be an important next step.
+Currently containerisation happens more in the background at the system level.
+
+New features, which make complex workflows accessible and reproducible, and variety in packages, even when they have overlapping features, are a signal and a support for a growing user base.
+This growth is possibly the the most important goal for the foreseeable future in the _Rockerverse_.
+The dominant position of Docker is a blessing and a curse for these goals.
+It could be wise to start experimenting with non-Docker containerisation tools now, e.g.
+R packages interfacing with other container engines such as [podman/buildah](https://github.com/containers/libpod) or [rkt](https://coreos.com/rkt/), or an R package for creating Singularity image recipe files.
+Such efforts help to avoid lock-in and to design sustainable workflows based on concepts of _containerisation_, not on their implementation in Docker.
+If adoption of containerisation and R continue to grow, the missing pieces for a success predominantly lie in (a) coordination and documentation of activities to reduce repeated work in favour of open collaboration, (b) the sharing of lessons learned from use cases to build common knowledge and language, and (c) a sustainable continuation and funding for all of development, community support, and education.
+A concrete effort by this work towards these pieces is to sustain the structure and captured status quo in a CRAN Task View _Containerization_.
 

### 2020-01-13 manuscript.Rmd
update images section

@@
 
-There are also independent projects installing R or RStudio on [**Alpine Linux**](https://www.alpinelinux.org/), but not beyond the proof of concept stage\footnote{\href{https://github.com/artemklevtsov/r-alpine}{https://github.com/artemklevtsov/r-alpine} or \href{https://www.github.com/velaco/alpine-r}{https://www.github.com/velaco/alpine-r} provide different use case images but are not recently maintained (development of the latter documented in \citet{ratesic_building_2018}); \href{https://github.com/cmplopes/alpine-r}{https://github.com/cmplopes/alpine-r} has only base R images, is relatively up-to-date but sparsely documented.}.
-Alpine images are often used for minimalistic and thereby small and secure containers, e.g., in hardware with limited storage, though this advantage quickly deteriorates if more features are needed, e.g.
-if installing RStudio, package building envrionments, or powerful data science tools.
+There are also independent projects installing R or RStudio on [**Alpine Linux**](https://www.alpinelinux.org/), but not beyond the proof of concept stage\footnote{\href{https://gitlab.com/artemklevtsov/r-alpine}{https://gitlab.com/artemklevtsov/r-alpine} or \href{https://www.github.com/velaco/alpine-r}{https://www.github.com/velaco/alpine-r} provide different use case images but are not recently maintained (development of the latter documented in \citet{ratesic_building_2018}); \href{https://github.com/cmplopes/alpine-r}{https://github.com/cmplopes/alpine-r} has only base R images, is relatively up-to-date but sparsely documented; \href{https://github.com/CenterForStatistics-UGent/mountainr}{https://github.com/CenterForStatistics-UGent/mountainr} is a ghost project with only a single short `Dockerfile`, yet it shows the easy installation of the latest R from Alpine sources}.
+In general, Alpine-based Docker images are often chosen for minimalistic and thereby small and secure containers, e.g., in hardware with limited storage.
+R-hub's proof of concept [`r-minimal`](https://github.com/r-hub/r-minimal) takes this to the extreme with compressed image size of 20MB and bespoke install scripts, e.g., to cleverly remove compilers after package installation.
+For R though, this advantage quickly deteriorates if more features are needed, e.g.
+if adding compilers, packages, or especially if installing powerful data science and communication tools, like RStudio or LaTeX.
 Furthermore the distribution uses the [`musl` C library](https://www.musl-libc.org/) instead of the `glibc` used in Debian/Ubuntu, for which more tooling and experiences exist\footnote{Cf.
 \href{https://github.com/rocker-org/rocker/issues/231}{https://github.com/rocker-org/rocker/issues/231}}.
-\href{https://github.com/CenterForStatistics-UGent/mountainr}{https://github.com/CenterForStatistics-UGent/mountainr} only has a single and quite short `Dockerfile`, showing the easy installation of the latest R from Alpine sources.
-These images are used ...
+In fact, using Debian's 'slim' variant and removing capabilities, one could probalby achieve similarly small image sizes on a more common stack.
 
-[`altRnative`](https://github.com/ismailsunni/altRnative/) is an experimental R package for running the same code across multiple containerised versions of R, intended for comparison across operating systems and different implementations of R.
-It comes with a collection of `Dockerfile`s and corresponding images \footnote{\href{https://github.com/ismailsunni/dockeRs}{https://github.com/ismailsunni/dockeRs}}.
+**\pkg{altRnative}** ([https://github.com/ismailsunni/altRnative/](https://github.com/ismailsunni/altRnative/)) is an experimental R package for running the same code across multiple containerised versions of R, intended for comparison across operating systems and different implementations of R.
+It comes with a collection of `Dockerfile`s and corresponding images in multiple combinations, currently including, e.g., MRO, FastR, Fedora, and TERR\footnote{See `Dockerfile`s at \href{https://github.com/ismailsunni/dockeRs}{https://github.com/ismailsunni/dockeRs}.}.
 
@@
 
-The notion of _Data Science_ is widely discussed among all academic disciplines.
-Using computers and complex software to make sense of (large amounts of) collected or simulated data is a cross-cutting phenomenon leading to _X Data Science_ disciplines, seminars, and study majors, there are also Docker images with commonly used tools and software stacks pre-installed that are often used in combination for analysing data.
-A large number of bespoke computing environments are created this way.
-Here we only mention a few image stacks as examples that are not based on Rocker images but prominently include installations of R.
+_Data Science_ is a widely discussed topic among all academic disciplines \citep[cf. e.g.,][]{donoho_50_2017}.
+Using computers and complex software to make sense of (large amounts of) collected or simulated data is a cross-cutting phenomenon leading to _X Data Science_ disciplines, seminars, and study majors.
+The practice to apply a combination tools and software stacks for analysing data and the cross-cutting skillsets connected with data science puts the management of computing environments and the challenges of reproducibility high up on the agenda.
+This awareness leads to a large number of bespoke Docker images but also some commonly used image stacks, ranging from base images with very common tools to 'catch all' images including _a lot_ of software.
+Here we only mention a few image stacks as examples, which are not based on Rocker images or stem from the R community, but prominently include installations of R.
+A general description of what to consider when creating a Docker image used for data science projects in R is provided on [environments.rstudio.com](https://environments.rstudio.com/docker).
 
@@
 ```
-
-A general description of what to consider when creating a Docker image used for data science projects in R is provided on [environments.rstudio.com](https://environments.rstudio.com/docker).
 

### 2020-01-13 manuscript.Rmd
update headlines (no more assignments to people) and improve package intro

@@
 ## Interfaces for Docker in R
+\label{interfaces}
 
-Interfacing with the Docker daemon is typically done through the Docker Command Line Interface (CLI).
+Interfacing with the Docker daemon is typically done through the [Docker Command Line Interface](https://docs.docker.com/engine/reference/commandline/cli/) (Docker CLI).
 However, moving back and forth between an R console and the command line can create friction in workflows and reduce reproducibility.
@@
 Each of these packages has particular advantages as they provide function wrappers for interacting with the Docker CLI at different stages of a container's life cycle.
-Examples of such interactions are installing the Docker software, creating `Dockerfile`s, building an image, and launching a container on a local machine or on the cloud.
-As such, the choice of which package is most useful depends on the use-case at hand.
-<!-- ^ in the examples listed above, the name of the packages which provide such functionality could be mentioned, this would highlight the variety in functionality already. -->
+Examples of such interactions are installing the Docker software, creating `Dockerfile`s (\pkg{dockerfiler}), \pkg{containerit}), building images and launching a containers (\pkg{stevedore}, \pkg{docker}) on a local machine or on the cloud.
+As such, the choice of which package is most useful depends on the use-case at hand, but also the users level of expertise.
 
@@
 
-**stevedore** ([https://cran.r-project.org/package=stevedore](https://cran.r-project.org/package=stevedore)) 
+**\pkg{stevedore}** ([https://cran.r-project.org/package=stevedore](https://cran.r-project.org/package=stevedore)) 
 <!-- from the CRAN description:
@@
 
-**dockyard** ([https://github.com/thebioengineer/dockyard](https://github.com/thebioengineer/dockyard)) has the goal of lowering barrier to creating `Dockerfile`s and Docker images, and deploying Docker containers.
+**\pkg{AzureContainers}** ([https://cran.r-project.org/package=AzureContainers](https://cran.r-project.org/package=AzureContainers)) is an interface to a number of container-related services in Microsoft's [Azure Cloud](https://azure.microsoft.com/), namely [Container Instances](https://azure.microsoft.com/en-us/services/container-instances/), [Container Registry](https://azure.microsoft.com/en-us/services/container-registry/), and [Kubernetes Service](https://azure.microsoft.com/en-us/services/kubernetes-service/).
+While it is mainly intended for working with Azure, as a convenience feature it includes lightweight, cross-platform shells to Docker and Kubernetes (tools `kubectl` and `helm`).
+These can be used to create and manage arbitrary Docker images and containers, as well as Kubernetes clusters on any platform or cloud service.
+
+**\pkg{babelwhale}** ([https://cran.r-project.org/package=babelwhale](https://cran.r-project.org/package=babelwhale)) 
+allows executing and interacting with containers, which can use either Docker or Singularity as a backend.
+The package provides a unified interface to interact with Docker and Singularity containers.
+Users can, for example, execute a command inside a container, mount a volume or copy a file.
+
+**\pkg{harbor}** ([https://github.com/wch/harbor](https://github.com/wch/harbor)
+<!-- from the github description
+is an R package for controlling docker containers on local and remote hosts.
+-->
+
+**\pkg{dockyard}** ([https://github.com/thebioengineer/dockyard](https://github.com/thebioengineer/dockyard)) has the goal of lowering barrier to creating `Dockerfile`s and Docker images, and deploying Docker containers.
 The way docker images are made is by reading and following a set of instructions found in `Dockerfile`s.
@@
 There are also wrappers for commonly used steps for R programmers like adding libraries and copying files into the R Docker containers.
-Once the `Dockerfile` is created, dockyard has the functionality to build the image and deploy it.
+Once the `Dockerfile` is created, dockyard has built-in functionality to build the image and run a container.
 
@@
 - [`renv` with Docker](https://rstudio.github.io/renv/articles/docker.html)
+- https://github.com/gigantum/base-images
 
@@
 
-## Deployment to cloud services [div]
+## Deployment to cloud services
 
@@
 
-## Using R to power enterprise software in production environments [@jnolis, @nolistic]
+## Using R to power enterprise software in production environments
 
@@
 
-## Deployment to the cloud [@hongooi]
+## Deployment to the cloud
 
@@
 
-## Using Docker in a Data Science Platform
+## Using Docker in a commercial Data Science Platform
 

### 2020-01-13 manuscript.Rmd
add a few more links and missing reference
update tex file

@@
 Again, all of these containers are being built on travis-ci, and will only be pushed to docker hub if the integration test succeeds.
+- \pkg{drake}, https://docs.ropensci.org/drake/index.html?q=docker#with-docker
 
@@
 
-...
+- https://github.com/benmarwick/rrtools
+- \pkg{renv}, https://rstudio.github.io/renv/articles/docker.html
 
@@
 
-DN conceived of the presented idea and \href{https://github.com/nuest/rockerverse-paper/issues/3}{initialised the formation of the writing team}, wrote the section about images for R distributions, wrote the conclusions section, and revised all sections.
+DN conceived of the presented idea and \href{https://github.com/nuest/rockerverse-paper/issues/3}{initialised the formation of the writing team}, wrote sections not mentioned below, and revised the full first draft.
 CB ..

### 2020-01-13 manuscript.tex
update tex file

@@
 \title{Containers and R---the Rockerverse and beyond}
-\author{by Daniel Nüst, Robrecht Cannoodt, Dirk Eddelbuettel, Mark Edmondson, Colin Fay, Karthik Ram, Noam Ross, Nan Xiao, Lori Shepherd, Nitesh Turaga, Jacqueline Nolis, Heather Nolis, Hong Ooi, Ellis Hughes, Ben Marwick}
+\author{by Daniel Nüst, Robrecht Cannoodt, Dirk Eddelbuettel, Mark Edmondson, Colin Fay, Karthik Ram, Noam Ross, Nan Xiao, Lori Shepherd, Nitesh Turaga, Jacqueline Nolis, Heather Nolis, Hong Ooi, Ellis Hughes, Sean Lopp, Ben Marwick}
 
@@
 host machine without any fear of interference between them, an important
-scalability opportunity is being provided. But Docker improved on
-virtualization by accessing the host system---generally Linux---through
-a much thinner and smaller shim than a full operating system emulation
-or virtualization. This makes for more efficient use of system resources
+scalability opportunity is being provided. But Docker improved this
+compartmentalization by accessing the host system---generally
+Linux---through a much thinner and smaller shim than a full operating
+system emulation or virtualization. This containerization is also called
+operating-system-level virtualization
+\ref{wikipedia_contributors_os-level_2020}. Typically a container runs
+one process, whereas virtualization may run whole operating systems at a
+larger footprint. This makes for more efficient use of system resources
 \citep{felter_updated_2015} and allowed another order of magnitude in
@@
 
-The key accomplishment of Docker (as an ``application'') is to make a
+The key accomplishment of Docker as an ``application'' is to make a
 ``bundled'' aggregation of software (the so-called ``image'') available
@@
 the host besides the actual Docker application installation. This is a
-rather attractive, and novel, proposition which has lead to widespread
-adoption and use of Docker in a variety of domains, \emph{e.g.}, cloud
-computing infrastructure \citep{Bernstein2014}, data science
-\citep{boettiger_introduction_2015}, and edge computing
+rather attractive proposition and Docker's very easy to use user
+interface has lead to widespread adoption and use of Docker in a variety
+of domains, e.g., cloud computing infrastructure \citep{Bernstein2014},
+data science \citep{boettiger_introduction_2015}, and edge computing
 \citep{alam_orchestration_2018} {[}more examples!{]}. It provided to be
-a natural match for ``cloud deployment'' which runs (or at least appears
-to run) ``seamlessly'' without much explicit reference to the underlying
+a natural match for ``cloud deployment'' which runs, or at least appears
+to run, ``seamlessly'' without much explicit reference to the underlying
 machine, architecture or operating system: containers are portable and
@@
 also claims superior performance but also targets full compatibility
-with base R
-\footnote{\href{https://github.com/oracle/fastr}{https://github.com/oracle/fastr}}.
+with base
+R\footnote{\href{https://github.com/oracle/fastr}{https://github.com/oracle/fastr}}.
 There are no offical Docker images provided, but \texttt{nuest/fastr}
@@
 
-\href{http://bioconductor.org/}{\emph{Bioconductor}} is an open source,
+\emph{Bioconductor} (\url{https://bioconductor.org/}) is an open source,
 open development project for the analysis and comprehension of genomic
@@
 toolbox leverages \texttt{docker-machine} (see also
-Section~\ref{interfaces} for an R package interfacing with
+Section~\nameref{interfaces} for an R package interfacing with
 \texttt{docker-machine}) to handle the process of creating a local
@@
 
-\textbf{\href{https://github.com/nuest/rocker-win}{\texttt{rocker-win}}}
-is a proof of concept for running R in Windows-based containers
+\textbf{rocker-win} (\url{https://github.com/nuest/rocker-win}) is a
+proof of concept for running R in Windows-based containers
 \citep{nust_rocker-win_2019}. It provides selected R versions using
@@
 
-\label{rhub} The \textbf{R-hub} project provides \_``a collection of
-services to help R package development'', with the package builder as
+\label{rhub} The \textbf{R-hub} project provides \emph{``a collection of
+services to help R package development''}, with the package builder as
 the most prominent one\citep{r-hub_docs_2019}. The builder allows R
@@
 development\footnote{\href{https://github.com/r-hub/rhub-linux-builders/pull/41}{https://github.com/r-hub/rhub-linux-builders/pull/41}},
-\texttt{see}rhub::local\_check\_linux\_images()\texttt{for\ a\ full\ list),\ and\ are\ intended\ for\ debugging\ (see\ Section\&nbsp;\textbackslash{}ref\{development\})\ not\ for\ reproducible\ workflows.\ \_The\ rationale\_\ of\ not\ re-using\ the\ Rocker\ image\ stack\ is\ the\ high\ number\ of\ operatings\ systems\ and\ configurations,\ which\ can\ be\ more\ uniformly\ maintained\ in\ a\ seperate\ suite\ of\ images.\ Also,\ a\ common\ structure\ across\ the\ respective\ image\ stacks\ of\ the\ operating\ systems,\ e.g.,\ for}debian\texttt{,\ adding\ GCC\ in}debian-gcc\texttt{and\ then\ R\textquotesingle{}s\ development\ version\ in}debian-gcc-devel`.
-The platform also has a specific user configuration differing from the
-setup covering typical applications of Rocker images, and the images
-include considerably more software, such as a fairly complete LaTeX
-installation, to provide an experience closer to CRAN.
+see \texttt{rhub::local\_check\_linux\_images()} for a full list), and
+are intended for debugging (see Section~\nameref{development}) not for
+reproducible workflows. \emph{The rationale} of not re-using the Rocker
+image stack is the high number of operatings systems and configurations,
+which can be more uniformly maintained in a seperate suite of images.
+Also, a common structure across the respective image stacks of the
+operating systems, e.g., for \texttt{debian}, adding GCC in
+\texttt{debian-gcc} and then R's development version in
+\texttt{debian-gcc-devel}. The platform also has a specific user
+configuration differing from the setup covering typical applications of
+Rocker images, and the images include considerably more software, such
+as a fairly complete LaTeX installation, to provide an experience closer
+to CRAN.
+
+\textbf{RStudio} also maintains a set of
+\href{https://github.com/rstudio/r-docker}{Docker images} for multiple
+operating systems including SUSE, Ubuntu, CentOS, and Debian. The image
+stacks include an \href{https://github.com/rstudio/r-builds}{opinionated
+R installation} and ensure the R installation and profile is consistent
+across the different Linux distributions. The images also install R in a
+versioned directory and use a minimal set of build and runtime
+dependencies, Furthermore these images are responsible for creating
+pre-compiled binary R packages for Linux, which supplement the binaries
+built by CRAN. These pre-compiled packages dramatically decrease the
+installation time of R packages on Linux and subsequently can decrease
+the build time of Docker images \citep{lopp_package_2019}.
 
@@
 beyond the proof of concept
-stage\footnote{\href{https://github.com/artemklevtsov/r-alpine}{https://github.com/artemklevtsov/r-alpine} or \href{https://www.github.com/velaco/alpine-r}{https://www.github.com/velaco/alpine-r} provide different use case images but are not recently maintained (development of the latter documented in \citet{ratesic_building_2018}); \href{https://github.com/cmplopes/alpine-r}{https://github.com/cmplopes/alpine-r} has only base R images, is relatively up-to-date but sparsely documented.}.
-Alpine images are often used for minimalistic and thereby small and
-secure containers, e.g., in hardware with limited storage, though this
-advantage quickly deteriorates if more features are needed, e.g.~if
-installing RStudio, package building envrionments, or powerful data
-science tools. Furthermore the distribution uses the
+stage\footnote{\href{https://gitlab.com/artemklevtsov/r-alpine}{https://gitlab.com/artemklevtsov/r-alpine} or \href{https://www.github.com/velaco/alpine-r}{https://www.github.com/velaco/alpine-r} provide different use case images but are not recently maintained (development of the latter documented in \citet{ratesic_building_2018}); \href{https://github.com/cmplopes/alpine-r}{https://github.com/cmplopes/alpine-r} has only base R images, is relatively up-to-date but sparsely documented; \href{https://github.com/CenterForStatistics-UGent/mountainr}{https://github.com/CenterForStatistics-UGent/mountainr} is a ghost project with only a single short `Dockerfile`, yet it shows the easy installation of the latest R from Alpine sources}.
+In general, Alpine-based Docker images are often chosen for minimalistic
+and thereby small and secure containers, e.g., in hardware with limited
+storage. R-hub's proof of concept
+\href{https://github.com/r-hub/r-minimal}{\texttt{r-minimal}} takes this
+to the extreme with compressed image size of 20MB and bespoke install
+scripts, e.g., to cleverly remove compilers after package installation.
+For R though, this advantage quickly deteriorates if more features are
+needed, e.g.~if adding compilers, packages, or especially if installing
+powerful data science and communication tools, like RStudio or LaTeX.
+Furthermore the distribution uses the
 \href{https://www.musl-libc.org/}{\texttt{musl} C library} instead of
@@
 exist\footnote{Cf. \href{https://github.com/rocker-org/rocker/issues/231}{https://github.com/rocker-org/rocker/issues/231}}.
-\href{https://github.com/CenterForStatistics-UGent/mountainr}{https://github.com/CenterForStatistics-UGent/mountainr}
-only has a single and quite short \texttt{Dockerfile}, showing the easy
-installation of the latest R from Alpine sources. These images are used
-\ldots{}
-
-\href{https://github.com/ismailsunni/altRnative/}{\texttt{altRnative}}
-is an experimental R package for running the same code across multiple
-containerised versions of R, intended for comparison across operating
-systems and different implementations of R. It comes with a collection
-of \texttt{Dockerfile}s and corresponding images
-\footnote{\href{https://github.com/ismailsunni/dockeRs}{https://github.com/ismailsunni/dockeRs}}.
+In fact, using Debian's `slim' variant and removing capabilities, one
+could probalby achieve similarly small image sizes on a more common
+stack.
+
+\textbf{\pkg{altRnative}}
+(\url{https://github.com/ismailsunni/altRnative/}) is an experimental R
+package for running the same code across multiple containerised versions
+of R, intended for comparison across operating systems and different
+implementations of R. It comes with a collection of \texttt{Dockerfile}s
+and corresponding images in multiple combinations, currently including,
+e.g., MRO, FastR, Fedora, and
+TERR\footnote{See `Dockerfile`s at \href{https://github.com/ismailsunni/dockeRs}{https://github.com/ismailsunni/dockeRs}.}.
 
@@
 
-The notion of \emph{Data Science} is widely discussed among all academic
-disciplines. Using computers and complex software to make sense of
-(large amounts of) collected or simulated data is a cross-cutting
-phenomenon leading to \emph{X Data Science} disciplines, seminars, and
-study majors, there are also Docker images with commonly used tools and
-software stacks pre-installed that are often used in combination for
-analysing data. A large number of bespoke computing environments are
-created this way. Here we only mention a few image stacks as examples
-that are not based on Rocker images but prominently include
-installations of R.
+\emph{Data Science} is a widely discussed topic among all academic
+disciplines \citep[cf. e.g.,][]{donoho_50_2017}. Using computers and
+complex software to make sense of (large amounts of) collected or
+simulated data is a cross-cutting phenomenon leading to \emph{X Data
+Science} disciplines, seminars, and study majors. The practice to apply
+a combination tools and software stacks for analysing data and the
+cross-cutting skillsets connected with data science puts the management
+of computing environments and the challenges of reproducibility high up
+on the agenda. This awareness leads to a large number of bespoke Docker
+images but also some commonly used image stacks, ranging from base
+images with very common tools to `catch all' images including \emph{a
+lot} of software. Here we only mention a few image stacks as examples,
+which are not based on Rocker images or stem from the R community, but
+prominently include installations of R. A general description of what to
+consider when creating a Docker image used for data science projects in
+R is provided on
+\href{https://environments.rstudio.com/docker}{environments.rstudio.com}.
 
@@
 
-Interfacing with the Docker daemon is typically done through the Docker
-Command Line Interface (CLI). However, moving back and forth between an
-R console and the command line can create friction in workflows and
-reduce reproducibility. A number of first-order R packages provide a
-interface to the Docker CLI, allowing to automate interaction with the
-Docker CLI from an R console.
+\label{interfaces}
+
+Interfacing with the Docker daemon is typically done through the
+\href{https://docs.docker.com/engine/reference/commandline/cli/}{Docker
+Command Line Interface} (Docker CLI). However, moving back and forth
+between an R console and the command line can create friction in
+workflows and reduce reproducibility. A number of first-order R packages
+provide a interface to the Docker CLI, allowing to automate interaction
+with the Docker CLI from an R console.
 
@@
 stages of a container's life cycle. Examples of such interactions are
-installing the Docker software, creating \texttt{Dockerfile}s, building
-an image, and launching a container on a local machine or on the cloud.
-As such, the choice of which package is most useful depends on the
-use-case at hand.
+installing the Docker software, creating \texttt{Dockerfile}s
+(\pkg{dockerfiler}), \pkg{containerit}), building images and launching a
+containers (\pkg{stevedore}, \pkg{docker}) on a local machine or on the
+cloud. As such, the choice of which package is most useful depends on
+the use-case at hand, but also the users level of expertise.
 
@@
 
-\textbf{stevedore} (\url{https://cran.r-project.org/package=stevedore})
-
-\textbf{dockyard} (\url{https://github.com/thebioengineer/dockyard}) has
-the goal of lowering barrier to creating \texttt{Dockerfile}s and Docker
-images, and deploying Docker containers. The way docker images are made
-is by reading and following a set of instructions found in
+\textbf{\pkg{stevedore}}
+(\url{https://cran.r-project.org/package=stevedore})
+
+\textbf{\pkg{AzureContainers}}
+(\url{https://cran.r-project.org/package=AzureContainers}) is an
+interface to a number of container-related services in Microsoft's
+\href{https://azure.microsoft.com/}{Azure Cloud}, namely
+\href{https://azure.microsoft.com/en-us/services/container-instances/}{Container
+Instances},
+\href{https://azure.microsoft.com/en-us/services/container-registry/}{Container
+Registry}, and
+\href{https://azure.microsoft.com/en-us/services/kubernetes-service/}{Kubernetes
+Service}. While it is mainly intended for working with Azure, as a
+convenience feature it includes lightweight, cross-platform shells to
+Docker and Kubernetes (tools \texttt{kubectl} and \texttt{helm}). These
+can be used to create and manage arbitrary Docker images and containers,
+as well as Kubernetes clusters on any platform or cloud service.
+
+\textbf{\pkg{babelwhale}}
+(\url{https://cran.r-project.org/package=babelwhale}) allows executing
+and interacting with containers, which can use either Docker or
+Singularity as a backend. The package provides a unified interface to
+interact with Docker and Singularity containers. Users can, for example,
+execute a command inside a container, mount a volume or copy a file.
+
+\textbf{\pkg{harbor}} (\url{https://github.com/wch/harbor}
+
+\textbf{\pkg{dockyard}}
+(\url{https://github.com/thebioengineer/dockyard}) has the goal of
+lowering barrier to creating \texttt{Dockerfile}s and Docker images, and
+deploying Docker containers. The way docker images are made is by
+reading and following a set of instructions found in
 \texttt{Dockerfile}s. This requires stepping out of the R ecosystem and
@@
 copying files into the R Docker containers. Once the \texttt{Dockerfile}
-is created, dockyard has the functionality to build the image and deploy
-it.
-
-\textbf{dockermachine} (\url{https://github.com/cboettig/dockermachine})
-
-\textbf{AzureContainers}
-(\url{https://cran.r-project.org/package=AzureContainers}) is an
-interface to a number of container-related services in Microsoft's Azure
-cloud, namely Container Instances, Container Registry and Kubernetes
-Service. While it is mainly intended for working with Azure, as a
-convenience feature it includes lightweight, cross-platform shells to
-Docker, kubectl and helm. These can be used to create and manage
-arbitrary Docker images and containers, as well as Kubernetes clusters
-on any platform or cloud service.
-
-\textbf{babelwhale}
-(\url{https://cran.r-project.org/package=babelwhale}) allows executing
-and interacting with containers, which can use either Docker or
-Singularity as a backend. provides a unified interface to interact with
-Docker and Singularity containers. You can execute a command inside a
-container, mount a volume or copy a file.
-
-\textbf{harbor} (\url{https://github.com/wch/harbor}
+is created, dockyard has built-in functionality to build the image and
+run a container.
+
+\textbf{\pkg{dockermachine}}
+(\url{https://github.com/cboettig/dockermachine}) is an R package to
+provide a convenient interface to
+\href{https://docs.docker.com/machine/overview/}{Docker Machine} from R.
+Docker Machine's CLI tool \texttt{docker-machine} allows users to create
+and manage virtual host on local computers, local data centers, or at
+cloud providers. A local Docker installation can be configured to
+forward all commands issues on the local Docker CLI to a selected
+(remote) virtual host fully transparent. For local Docker Machine was
+especially crucial in early days of Docker, when no native Docker
+support was available for Mac or Windows computers, but remains relevant
+for provisioning on remote systems. The package has not received any
+updates for two years, but was functional with a current version of
+\texttt{docker-machine} (\texttt{0.16.2}) and potentially lowers the
+barriers for R users to run containers on various hosts, if using the
+Docker Machine CLI is perceived as a barrier.
 
@@
 \item
-  \href{https://github.com/smaakage85/dockr}{\texttt{dockr}}
+  \href{https://github.com/smaakage85/dockr}{\texttt{dockr}}, see also
+  \url{http://smaakage85.netlify.com/2019/12/21/dockr-easy-containerization-for-r/}
 \item
@@
   with Docker}
+\item
+  \url{https://github.com/gigantum/base-images}
 \end{itemize}
 
-\textbf{\href{https://nanx.me/liftr/}{\emph{liftr}}} \citep{liftr2019}
+\textbf{\pkg{liftr}} (\url{https://nanx.me/liftr/}) \citep{liftr2019}
 aims to solve the problem of persistent reproducible reporting in
@@
 
-\hypertarget{deployment-to-cloud-services-div}{%
-\subsection{Deployment to cloud services
-{[}div{]}}\label{deployment-to-cloud-services-div}}
+\hypertarget{deployment-to-cloud-services}{%
+\subsection{Deployment to cloud
+services}\label{deployment-to-cloud-services}}
 
@@
   \href{https://cloudyr.github.io/googleComputeEngineR/}{\texttt{googleComputeEngineR}}
-  (function \texttt{docker\_run})
+  (function \texttt{gce\_vm\_template()})
 
@@
   \item
-    I don't think \texttt{docker\_run} is main focus for containers for
-    this package, its more how it uses Docker to launch custom
-    environments that can be templated (e.g., rstudio, shiny etc.) and
-    for parrallisation e.g., \texttt{library(future)} using GCE VMs as
-    endpoints. \citep{MarkEdmondson1234}
+    To enable quick deployments of key R services such as RStudio and
+    Shiny onto cloud virtual machines (VMs), this package utilises
+    Dockerfiles to move the labour of setting up those services from the
+    user to a premade Docker image. For example, by specifying the
+    template \texttt{template="rstudio"} in
+    \texttt{gce\_vm\_template()/gce\_vm()} an up to date RStudio Service
+    image is launched. Specifying \texttt{template="rstudio-gpu"} will
+    launch an RStudio Server image with a GPU attached, etc.\\
   \end{itemize}
@@
   (digital ocean R client)
+\item
+  \href{https://www.rplumber.io/docs/hosting.html\#docker}{\texttt{plumber}}
+\item
+  \href{https://github.com/tmobile/r-tensorflow-api}{production
+  deployment of neural networks} \citep[\citet{nolistic}]{jnolis}
+\item
+  Kubernetes - A popular platform for managing Docker containers is
+  \href{https://kubernetes.io/}{Kubernetes}, which is used for a wide
+  variety of containerized applications. It may be your organisation has
+  a Kubernetes cluster already for other applications. Docker containers
+  are used within Kubernetes clusters to hold native code, for which
+  Kubernetes creates a framework around network connections and scaling
+  of resources up and down. An introduction on using
+  \href{https://code.markedmondson.me/r-on-kubernetes-serverless-shiny-r-apis-and-scheduled-scripts/}{Kubernetes
+  with R is published at this blog post}.
 \item
@@
 
-\hypertarget{using-r-to-power-enterprise-software-in-production-environments-nolisticjnolis-nolistic}{%
-\subsection{\texorpdfstring{Using R to power enterprise software in
-production environments
-\citep[\citet{nolistic}]{jnolis}}{Using R to power enterprise software in production environments , @nolistic{[}@jnolis, @nolistic{]}}}\label{using-r-to-power-enterprise-software-in-production-environments-nolisticjnolis-nolistic}}
+\hypertarget{using-r-to-power-enterprise-software-in-production-environments}{%
+\subsection{Using R to power enterprise software in production
+environments}\label{using-r-to-power-enterprise-software-in-production-environments}}
 
@@
 running as a web service, along with with the ability to deploy R in
-docker containers, modern enterprises are now capable of having
+Docker containers, modern enterprises are now capable of having
 real-time machine learning powered by R.
 
-The backbone of this work is the
-\emph{\href{https://www.rplumber.io/docs/hosting.html\#docker}{plumber}}
-package, created by Jeff Allen. The plumber package allows R to be used
-to host web services through RESTful APIs. When R is running on a
-server, a different server can ask R to make a computation through an
-HTTP request and get an immediate response. For instance, if a data
+The backbone of this work is the \pkg{plumber}
+(\url{https://www.rplumber.io/}) package. The plumber package allows R
+to be used to host web services through RESTful APIs. When R is running
+on a server, a different server can ask R to make a computation through
+an HTTP request and get an immediate response. For instance, if a data
 scientist at a company creates a machine learning model to tell which
@@
 
-\hypertarget{deployment-to-the-cloud-hongooi}{%
-\subsection{\texorpdfstring{Deployment to the cloud
-\citep{hongooi}}{Deployment to the cloud {[}@hongooi{]}}}\label{deployment-to-the-cloud-hongooi}}
+\hypertarget{deployment-to-the-cloud}{%
+\subsection{Deployment to the cloud}\label{deployment-to-the-cloud}}
 
@@
 \item
-  Google Cloud Build \citep{MarkEdmondson1234}
+  Google Cloud Build
 
@@
   \item
-    For me this is what makes Docker containers viable, as it builds the
-    \texttt{Dockerfile}s on each GitHub commit. It couples with Google
-    Container Registry to build private Docker images, for downstream
-    applications.
+    Cloud Build is another continuous intergation service that helps
+    move the workload of building the Docker images to an online
+    service. Cloud Build can be set up to build the Dockerfiles on each
+    GitHub commit or release. This means you do not need to build the
+    Docker images locally, which can tie up resources since Docker
+    images can be several GBs and take a long time to compile. Google
+    Cloud Build works alongside Google Container Registry to allow you
+    to build private and public Docker images, which allows you to build
+    up your own dependency graphs for downstream applications.
   \end{itemize}
@@
 
-\href{https://github.com/karthik/holepunch}{\texttt{holepunch}} is an R
-package that was designed to make sharing work environments accessible
-to novice R users based on \emph{Binder}. The
+\textbf{\pkg{holepunch}} (\url{https://github.com/karthik/holepunch}) is
+an R package that was designed to make sharing work environments
+accessible to novice R users based on \textbf{Binder}. The
 \href{https://mybinder.readthedocs.io/en/latest/}{Binder project},
@@
 Zenodo. Binder's core internal tool is
-\href{https://repo2docker.readthedocs.io/en/latest/config_files.html}{repo2docker}.
+\href{https://repo2docker.readthedocs.io/en/latest/config_files.html}{\texttt{repo2docker}}.
 It deterministically builds a Docker image by parsing the contents of a
@@
 
+\hypertarget{using-docker-in-a-commercial-data-science-platform}{%
+\subsection{Using Docker in a commercial Data Science
+Platform}\label{using-docker-in-a-commercial-data-science-platform}}
+
+In addition to embedding RStudio Open Source Server inside of Docker
+images, such as Rocker, RStudio offers the moduler data science platform
+\href{https://rstudio.com/products/team/}{RStudio Team}, a commercial
+product allowing teams of data scientists and their respective IT/DevOps
+groups to develop and deploy code in R and Python. RStudio team enables
+these groups to use the benefits of containers, without requiring users
+to learn new tools or directly interact with containers. Notable
+patterns include:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  \emph{To deploy and manage RStudio Professional Products:} Many
+  RStudio customers are managing professional products in Docker
+  containers as an alternative to bare-metals servers or virtual
+  machines. Best practices for
+  \href{https://support.rstudio.com/hc/en-us/articles/360021594513-Running-RStudio-with-Docker-containers}{running
+  RStudio with Docker containers} as well as
+  \href{https://github.com/rstudio/rstudio-docker-products}{Docker
+  images} for RStudio's commerical products are available.
+\item
+  \emph{To run interactive RStudio sessions on a cluster:} In the
+  RStudio Server Pro 1.2 release in 2019, RStudio added new
+  functionality called
+  \href{https://solutions.rstudio.com/launcher/overview/}{Launcher}. It
+  gives users the ability to spawn RStudio instances and background jobs
+  in a scalable way on external clusters, e.g.,
+  \href{https://support.rstudio.com/hc/en-us/articles/360019253393-Using-Docker-images-with-RStudio-Server-Pro-Launcher-and-Kubernetes}{Kubernetes
+  based on Docker images} or \href{https://slurm.schedmd.com/}{Slurm}
+  clusters, and optionally, with Singularity containers. A key benefit
+  of Launcher is the ability for R and Python users to interactively
+  work in RStudio without learning about Docker at all - while still
+  leveraging containers and Kubernetes. Users can submit batch jobs to
+  Kubernetes clusters without writing specific deployment scripts.
+\end{enumerate}
+
+\href{https://rstudio.com/products/connect/}{RStudio Connect} is a
+commercial publishing platform to deploy dashboards, REST APIs, and
+other R and Python content into production. RStudio Connect uses an
+approach to isolate deployed R applications that is similar to Docker's
+filesystem and process namespacing and leverages the
+\href{https://rstudio.com/products/package-manager/}{RStudio Package
+Manager} to manage R environments including system dependencies. RStudio
+Package manager may also be used to
+\href{https://environments.rstudio.com/docker}{manage their R
+environments and package versions with Docker} and to ensure
+deterministic image builds.
+
 \hypertarget{processing-div}{%
@@
 \item
-  Docker images for cloud services \citep{MarkEdmondson1234}
+  Docker images for cloud services
 
@@
   \item
-    the most popular functionality for \texttt{googleComputeEngineR} is
-    its use of Docker to enable parallel processing across VMS -
+    A popular application of using Docker is to create the environments
+    that can run in parallel for speeding up R code. For example,
+    \href{https://CRAN.R-project.org/package=googleComputeEngineR}{\texttt{googleComputeEngineR}}'s
+    \texttt{gce\_vm\_cluster()} function can create clusters of 2 or
+    more virtual machines, running multi-CPU architectures. Instead of
+    running a local R script with the local CPU and RAM restrictions,
+    the same code can be processed on all CPU threads of the cluster of
+    machines in the cloud, all running in a Docker container with the
+    same R environments. This is achieved through
+    \texttt{googleComputeEngineR}'s integration with the R parralisation
+    library
+    \href{https://CRAN.R-project.org/package=future}{\texttt{library(future)}
+    by Henrik Bengtsson}. Local R computation can be thrown up to a
+    multi-CPU and VM environment to achieve parrell computation in a few
+    lines of R code in your local session. -
     \href{https://cloudyr.github.io/googleComputeEngineR/articles/massive-parallel.html}{some
-    demos here}
+    demonstrations are available here}.
   \end{itemize}
 \item
-  Google Cloud Run - CaaS (Containers as a Service) that lets you launch
-  a Docker container without worrying about underlying infrastructure.
-  An R implementation is shown here at
+  \href{https://cloud.run}{Google Cloud Run} - This is a CaaS
+  (Containers as a Service) that lets you launch a Docker container
+  without worrying about underlying infrastructure. This dispenses with
+  the developer creating a cloud server to run the Docker image on, by
+  abstracting away those servers to a more serverless configuration.
+  Cloud Run lets you run your code on top of a managed or your own
+  Kubernetes cluster running Knative, and can accept any Docker image.
+  The service takes care of network ingress, scaling machines up and
+  down to zero, authentication and authorisation, all features which are
+  non-trivial for a developer to create on their own. This can be used
+  to scale up R code to millions of instances if they need to, with
+  little or no changes to existing code. An R implementation is shown
+  here at
   \href{https://github.com/MarkEdmondson1234/cloudRunR}{cloudRunR} which
-  uses it to create a scalable R plumber API.
+  uses Cloud Run to create a scalable R plumber API.
 \item
@@
   will only be pushed to docker hub if the integration test succeeds.
+\item
+  \pkg{drake},
+  \url{https://docs.ropensci.org/drake/index.html?q=docker\#with-docker}
 \end{itemize}
@@
 
-\ldots{}
+\begin{itemize}
+\tightlist
+\item
+  \url{https://github.com/benmarwick/rrtools}
+\item
+  \pkg{renv}, \url{https://rstudio.github.io/renv/articles/docker.html}
+\end{itemize}
 
@@
 
-The R-hub Docker images (see Section~\ref{rhub}) can also be used to
+The R-hub Docker images (see Section~\nameref{rhub}) can also be used to
 debug problems locally using various combinations of Linux platforms, R
@@
 
-\begin{itemize}
-\tightlist
-\item
-  what are next steps?
-\item
-  support more complex set-ups with ready-to-use stacks, e.g., GPU, AI,
-  ML
-\item
-  Missing pieces?
-\item
-  consolidation (e.g., via packages using \texttt{dockerfiler} and
-  \texttt{stevedore})
-\item
-  Common themes
-
-  \begin{itemize}
-  \tightlist
-  \item
-    reproducibility
-  \end{itemize}
-\item
-  will knowledge about containers continue to spread?
-\item
-  what is needed for even more containers with R?
-\item
-  the ability to move processing between services easily (e.g locally,
-  one cloud providers VM, another cloud provider's
-  Container-as-a-Service)
-\item
-  running RStudio in a container is a common approach for
-  reproducibility - \emph{is there a need to run/control containers from
-  a UI integrated in RStudio?}
-\item
-  moving away from Docker focus towards containers?
-
-  \begin{itemize}
-  \tightlist
-  \item
-    is there a need for a wrapper for
-    \href{https://github.com/containers/libpod/blob/master/API.md}{podman's
-    Varlink-API} (similar to stevedore) and buildah to manage containers
-    and images respectively with alternative software stacks?
-  \item
-    will there be control packages for Singularity support?
-  \end{itemize}
-\item
-  sustain the overview and structure in this article in form of a CRAN
-  Task View
-\end{itemize}
+This article is a snapshot of the R-corner in a universe of applications
+built many-faced piece of software, Docker. Docker became synonymous
+with applying the concept of containerisation to solve many challenges,
+such as reproducible environments in research, scalable deployments in
+heterogenous organisations, and the ability to move processing between
+machines easily (e.g., locally, one cloud providers VM, another cloud
+provider's Container-as-a-Service). These are probably the thre common
+themes described in this work.
+
+The R packages and use cases presented show the positive drive in the
+community to innovate, but also the challenge of keeping up with
+continuous evolving landscape. The use cases contributed by co-authors
+also have a degree of overlap. A common language and understanding of
+good practices is still taking shape. The easiness with which one can
+create a complex software systems, such as an independent Docker image
+stack, to serve one's specific needs currently leads to some
+fragmentation. It is unclear if Docker alone as a core technology is
+really a feasible level of collaboration and abstraction. Furthermore
+fragmentation may not be a bad sign but instead a reflection of a
+growing market, which is able to sustain multiple related efforts. With
+the maturing of core building blocks, such as the Rocker suite of
+images, more systems will be built successfully but will also be behind
+the curtains.
+
+However, at least \emph{on the level of R packages}, some consolidation
+seems in order, e.g., to reduce the number of packages creating
+\texttt{Dockerfile}s from R code or controlling the Docker daemon with R
+code. It remains to be seen which approach to control Docker, via the
+Docker API as \texttt{stevedore} or via system calls as
+\texttt{dockyard}/\texttt{docker}/\texttt{dockr}, is more sustainable,
+or if the question will be answered by the endurance of maintainers and
+sufficient funding. Similarly, the capturing of environments and their
+serialization in form of a \texttt{Dockerfile} currently happens at
+different levels of abstraction and re-use of functionality seems
+reasonable, e.g.~\texttt{liftr} could generate the environment with
+\texttt{containerit}, which in turn may use \texttt{dockerfiler} for low
+level objects representing a \texttt{Dockerfile}. In this consolidation,
+the Rocker Project could play the role of a unifying and coordinating
+entity in the future. Though for the moment, the sign of the times
+points to more experimentation and feature growth, e.g.~images for
+GPU-based computing and artificial intelligence. Even with coding being
+more and more accepted as a required, and achievable skill, an easier
+access, for example by exposing containerisation benefits via simple
+user interfaces in the users' IDE, could be an important next step.
+Currently containerisation happens more in the background at the system
+level.
+
+New features, which make complex workflows accessible and reproducible,
+and variety in packages, even when they have overlapping features, are a
+signal and a support for a growing user base. This growth is possibly
+the the most important goal for the foreseeable future in the
+\emph{Rockerverse}. The dominant position of Docker is a blessing and a
+curse for these goals. It could be wise to start experimenting with
+non-Docker containerisation tools now, e.g.~R packages interfacing with
+other container engines such as
+\href{https://github.com/containers/libpod}{podman/buildah} or
+\href{https://coreos.com/rkt/}{rkt}, or an R package for creating
+Singularity image recipe files. Such efforts help to avoid lock-in and
+to design sustainable workflows based on concepts of
+\emph{containerisation}, not on their implementation in Docker. If
+adoption of containerisation and R continue to grow, the missing pieces
+for a success predominantly lie in (a) coordination and documentation of
+activities to reduce repeated work in favour of open collaboration, (b)
+the sharing of lessons learned from use cases to build common knowledge
+and language, and (c) a sustainable continuation and funding for all of
+development, community support, and education. A concrete effort by this
+work towards these pieces is to sustain the structure and captured
+status quo in a CRAN Task View \emph{Containerization}.
 
@@
 \href{https://github.com/nuest/rockerverse-paper/issues/3}{initialised the formation of the writing team},
-wrote the section about images for R distributions, wrote the
-conclusions section, and revised all sections. CB .. RC and EH wrote the
-section on interfaces for Docker in R. DE wrote the introduction and
-section about Rocker. ME .. CF wrote the paragraphs about
-\texttt{r-online}, \texttt{dockerfiler}, \texttt{r-ci} and
+wrote sections not mentioned below, and revised the full first draft. CB
+.. RC and EH wrote the section on interfaces for Docker in R. DE wrote
+the introduction and section about Rocker. ME .. CF wrote the paragraphs
+about \texttt{r-online}, \texttt{dockerfiler}, \texttt{r-ci} and
 \texttt{r-db}. BW .. KR .. NR .. NX wrote the section on liftr. LS \& NT
@@
 \address{%
+Sean Lopp\\
+RStudio, Inc\\
+RStudio\\ 250 Northern Ave\\ Boston, MA 02210\\
+}
+\href{mailto:sean@rstudio.com}{\nolinkurl{sean@rstudio.com}}
+
+\address{%
 Ben Marwick\\

### 2020-01-14 manuscript.Rmd
update abstract (match conclusions)

@@
   The Rocker project provides widely-used Docker images for R across different application scenarios.
-  This articles surveys describe downstream projects building upon Rocker.
-  We also look beyond Rocker to other projects connecting containerisation with R.
-  These use cases and the diversity of applications demonstrate the power of Rocker and containerisation for collaboration, effectivity, scalability, and transparency.
+  This articles surveys downstream projects building upon Rocker and presents the current state of R packages for managing Docker images and controlling containers.
+  We also look beyond Rocker to other projects connecting containerisation with R, namely alternative suites of images.
+  These use cases and the variety of applications demonstrate the power of Rocker specifically and containerisation in general.
+  We identified common themes across this diversity: reproducible environments, scalability and effectivity, and portability across clouds.
 # blank footnote? https://tex.stackexchange.com/questions/170511/footnotes-without-numbering
@@
 This article is a snapshot of the R-corner in a universe of applications built many-faced piece of software, Docker.
-Docker became synonymous with applying the concept of containerisation to solve many challenges, such as reproducible environments in research, scalable deployments in heterogenous organisations, and the ability to move processing between machines easily (e.g., locally, one cloud providers VM, another cloud provider's Container-as-a-Service).
-These are probably the thre common themes described in this work.
+`Dockerfile`s and Docker images are the go-to methods for collaboration between roles in an organisation, such as development and IT operations teams, and between parties in the communication of knowledge, such as research workflows or education.
+Docker became synonymous with applying the concept of containerisation to solve challenges of reproducible environments, e.g., in research and in development \& production, and of scalable deployments with the ability to move processing between machines easily (e.g., locally, one cloud providers VM, another cloud provider's Container-as-a-Service).
+Collaboration, reproducibility, and portability are the common themes behind R packages, use cases, and applications in this work.
 
-The R packages and use cases presented show the positive drive in the community to innovate, but also the challenge of keeping up with continuous evolving landscape.
+The R packages and use cases presented show the positive drive in the community to innovate, but also the challenge of keeping up with a continuously evolving landscape.
 The use cases contributed by co-authors also have a degree of overlap.
@@
 If adoption of containerisation and R continue to grow, the missing pieces for a success predominantly lie in (a) coordination and documentation of activities to reduce repeated work in favour of open collaboration, (b) the sharing of lessons learned from use cases to build common knowledge and language, and (c) a sustainable continuation and funding for all of development, community support, and education.
-A concrete effort by this work towards these pieces is to sustain the structure and captured status quo in a CRAN Task View _Containerization_.
+A concrete effort to work towards these pieces is to sustain the structure and captured status quo from this work in form of a CRAN Task View on containerization.
 
@@
 LS \& NT wrote the section on Bioconductor.
-All authors contributed minor edits, gave feedback on other sections, and approved the final version.
+All authors approved the final version.
 

### 2020-01-14 manuscript.Rmd
add section on outsider, see #54

@@
 
-## Research Compendia [@benmarwick]
+## Packaging programs for processing pipelines
+
+Docker is perfectly suited to package and execute any software and data, and is a good candidate to build complex processing pipelines independent of the used programming language.
+Because of the original use case (see&nbsp;\nameref{introduction}), Docker does not have a standard mechanism to chain a number of containers together, i.e., to define how environment variables, volume mounts, or ports can be used to pass input parameters and data into a container and how to get results out.
+Some packages, e.g.
+\pkg{containerit}, provide a Docker image that can be used similar as a CLI, but the usage is cumbersome\footnote{\href{https://o2r.info/containerit/articles/container.html}{https://o2r.info/containerit/articles/container.html}}.
+
+**\pkg{outsider}** ([https://antonellilab.github.io/outsider/](https://antonellilab.github.io/outsider/)) tackles the problem of integrating external programmes into an R workflow \citep{bennett_outsider_2019}.
+Installation and usage of external programmes can be cumbersome and even prohibited if platform dependent.
+Therefore \pkg{outsider} uses the platform-independent Docker images to encapsulate processes in _outsider modules_.
+Each outsider modules has a `Dockerfile` and an R package with functions for interacting with the encapsulated tool.
+A user only needs to use the given functions to install a module and can then integrate a tool into their own workflow seaminglessly using only R functions.
+The module takes care of transmitting arguments and sending and receiving files to and from a container.
+Outsider modules are hosted, e.g., on GitHub, and \pkg{outsider} contains discovery functions for them.
+<!-- illustrating outsider examples:
+https://github.com/DomBennett/om..astral/blob/master/examples/astral.R
+https://github.com/DomBennett/om..raxml#partitioned-dna-analysis
+-->
+
+## Research Compendia
 

### 2020-01-14 manuscript.Rmd
Add some links and fix errors caught by @slopp in #46

@@
 For statistical computing and analysis centered around R, the Rocker Project has provided a variety of Docker containers since its start in 2014 \citep{RJ-2017-065}.
-The Rocker Project provides several lines of containers spanning to from building blocks with `R-release` or `R-devel`, via containers with RStudio~Server and Shiny~Server, to domain-specific containers such as `geospatial`.
+The Rocker Project provides several lines of containers spanning to from building blocks with `R-release` or `R-devel`, via containers with [RStudio&nbsp;Server](https://rstudio.com/products/rstudio/) and [Shiny&nbsp;Server](https://rstudio.com/products/shiny/shiny-server/), to domain-specific containers such as [`rocker/geospatial`](https://github.com/rocker-org/geospatial) \citep{rocker_geospatial_2019}.
 These containers form _image stacks_, building on top of each other for better maintainability and 
@@
 
-**RStudio** also maintains a set of [Docker images](https://github.com/rstudio/r-docker) for multiple operating systems including SUSE, Ubuntu, CentOS, and Debian.
-The image stacks include an [opinionated R installation](https://github.com/rstudio/r-builds) and ensure the R installation and profile is consistent across the different Linux distributions.
-The images also install R in a versioned directory and use a minimal set of build and runtime dependencies,
+[**RStudio**](https://rstudio.com/) also maintains a set of [Docker images](https://github.com/rstudio/r-docker) for multiple operating systems including SUSE, Ubuntu, CentOS, and Debian.
+The image stack includes an [opinionated R installation](https://github.com/rstudio/r-builds) and ensures the R installation and profile is consistent across the different Linux distributions.
+The base image also installs R in a versioned directory and uses a minimal set of build and runtime dependencies.
 Furthermore these images are responsible for creating pre-compiled binary R packages for Linux, which supplement the binaries built by CRAN.
@@
 In the RStudio Server Pro 1.2 release in 2019, RStudio added new functionality called [Launcher](https://solutions.rstudio.com/launcher/overview/).
-It gives users the ability to spawn RStudio instances and background jobs in a scalable way on external clusters, e.g., [Kubernetes based on Docker images](https://support.rstudio.com/hc/en-us/articles/360019253393-Using-Docker-images-with-RStudio-Server-Pro-Launcher-and-Kubernetes) or [Slurm](https://slurm.schedmd.com/) clusters, and optionally, with Singularity containers.
+It gives users the ability to spawn R sessions and background jobs in a scalable way on external clusters, e.g., [Kubernetes based on Docker images](https://support.rstudio.com/hc/en-us/articles/360019253393-Using-Docker-images-with-RStudio-Server-Pro-Launcher-and-Kubernetes) or [Slurm](https://slurm.schedmd.com/) clusters, and optionally, with Singularity containers.
 A key benefit of Launcher is the ability for R and Python users to interactively work in RStudio without learning about Docker at all - while still leveraging containers and Kubernetes.

### 2020-01-14 manuscript.Rmd
Docker = lingua franca, also add labels, just in case, small fixes

@@
 # Introduction
+\label{intro}
 
@@
 All this cements the role of R as the _lingua&nbsp;franca_ of statistics, data visualisation, and computational research.
-Coinciding with the rise of R was the advent of [Docker](https://en.wikipedia.org/wiki/Docker_(software)) as a general tool for distribution and deployment of server applications.
+Coinciding with the rise of R was the advent of [Docker](https://en.wikipedia.org/wiki/Docker_(software)) as a general tool for distribution and deployment of server applications---in fact, Docker can be called the _lingua&nbsp;franca_ of describing computing environments and packaging software.
 Combining both these topics, the _Rocker&nbsp;Project_ ([https://www.rocker-project.org/](https://www.rocker-project.org/)) provides images with R (see the next Section for more details).
@@
 # Containerization and Rocker
+\label{containerisation}
+\label{rocker}
 
@@
 The key accomplishment of Docker as an "application" is to make a "bundled" aggregation of software (the so-called "image") available to any system equipped to run Docker, without requiring much else from the host besides the actual Docker application installation.
-This is a rather attractive proposition and Docker's very easy to use user interface has lead to widespread adoption and use of Docker in a variety of domains, e.g., cloud computing infrastructure \citep{Bernstein2014}, data science \citep{boettiger_introduction_2015}, and edge computing \citep{alam_orchestration_2018} [more examples!].
+This is a rather attractive proposition and Docker's very easy to use user interface has lead to widespread adoption and use of Docker in a variety of domains, e.g., cloud computing infrastructure \citep[e.g.,][]{Bernstein2014}, data science \citep[e.g.,][]{boettiger_introduction_2015}, and edge computing \citep[e.g.,][]{alam_orchestration_2018}.
 It provided to be a natural match for "cloud deployment" which runs, or at least appears to run, "seamlessly" without much explicit reference to the underlying machine, architecture or operating system: containers are portable and can be deployed with very little in terms of dependencies on the host system---only the container runtime is required.
@@
 The Rocker Project provides several lines of containers spanning to from building blocks with `R-release` or `R-devel`, via containers with [RStudio&nbsp;Server](https://rstudio.com/products/rstudio/) and [Shiny&nbsp;Server](https://rstudio.com/products/shiny/shiny-server/), to domain-specific containers such as [`rocker/geospatial`](https://github.com/rocker-org/geospatial) \citep{rocker_geospatial_2019}.
-These containers form _image stacks_, building on top of each other for better maintainability and 
+These containers form _image stacks_, building on top of each other for better maintainability (i.e. smaller `Dockerfile`s), composability, and to reduce build time.
 Also of note is a series of "versioned" containers which match the R release they contain with the _then-current_ set of packages via the MRAN Snapshot views of CRAN \citep{microsoft_cran_2019}.
@@
 An alternative stemming from the domain of high-perfomance computing is **Singularity** \citep{kurtzer_singularity_2017}.
-Singularity can run Docker images, and in the case of Rocker works out of the box if the main process is R, e.g., in `rocker/r-base`, but does not succeed in running images where there is an init script, e.g., in containers that by default run RStudio.
+Singularity can run Docker images, and in the case of Rocker works out of the box if the main process is R, e.g., in `rocker/r-base`, but does not succeed in running images where there is an init script, e.g., in containers that by default run RStudio&nbsp;Server.
 In the latter case, a `Singularity` file, a recipe akin to a `Dockerfile`, needs to be used.
@@
 # Container images
+\label{images}
 
 ## Images for alternative R distributions
+\label{alternatives}
 
@@
 ## Bioconductor
+\label{bioc}
 
@@
 ## Images for (historic) R versions
+\label{versions}
 
@@
 ## Windows Images
+\label{windows}
 
@@
 ## Non-Debian Linux images
+\label{nondebian}
 
@@
 ## Data Science images
+\label{datascience}
 
@@
 <!-- from packages closer to Docker to second level packages to more complex use cases -->
+\label{applications}
 
@@
 ## Capture and create environments [@nuest]
+\label{envs}
 
@@
 ## Using R to power enterprise software in production environments
+\label{enterprise}
 
@@
 ## Deployment to the cloud
+\label{deployment}
 
@@
 ## Continuous integration and continuous delivery [@noamross, @ColinFay]
+\label{cicd}
 
@@
 ## Common or public work environments
+\label{workenvs}
 
@@
 ## Using Docker in a commercial Data Science Platform
+\label{rstudio}
 
@@
 ## Processing [div]
+\label{processing}
 

### 2020-01-14 manuscript.Rmd
outsider text update

@@
 
-Docker is perfectly suited to package and execute any software and data, and is a good candidate to build complex processing pipelines independent of the used programming language.
-Because of the original use case (see&nbsp;\nameref{introduction}), Docker does not have a standard mechanism to chain a number of containers together, i.e., to define how environment variables, volume mounts, or ports can be used to pass input parameters and data into a container and how to get results out.
+Docker is perfectly suited for packaging and executing any software and data, and is a good candidate for building complex processing pipelines that are independent from the host programming language.
+Due to its original use case (see&nbsp;\nameref{introduction}), Docker has no standard mechanisms for chaining containers together; it lacks definitions and protocols for environment variables, volume mounts and/or ports that could enable transfer of input (parameters and data) and output (results) to and from containers.
 Some packages, e.g.
-\pkg{containerit}, provide a Docker image that can be used similar as a CLI, but the usage is cumbersome\footnote{\href{https://o2r.info/containerit/articles/container.html}{https://o2r.info/containerit/articles/container.html}}.
+\pkg{containerit}, do provide Docker images that can be used similarly to CLIs, but their usage is cumbersome\footnote{\href{https://o2r.info/containerit/articles/container.html}{https://o2r.info/containerit/articles/container.html}}.
 
-**\pkg{outsider}** ([https://antonellilab.github.io/outsider/](https://antonellilab.github.io/outsider/)) tackles the problem of integrating external programmes into an R workflow \citep{bennett_outsider_2019}.
-Installation and usage of external programmes can be cumbersome and even prohibited if platform dependent.
+**\pkg{outsider}** ([https://antonellilab.github.io/outsider/](https://antonellilab.github.io/outsider/)) tackles the problem of integrating external programs into an R workflow \citep{bennett_outsider_2020}.
+Installation and usage of external programs can be difficult, convoluted and even impossible if the platform is incompatible.
 Therefore \pkg{outsider} uses the platform-independent Docker images to encapsulate processes in _outsider modules_.
-Each outsider modules has a `Dockerfile` and an R package with functions for interacting with the encapsulated tool.
-A user only needs to use the given functions to install a module and can then integrate a tool into their own workflow seaminglessly using only R functions.
-The module takes care of transmitting arguments and sending and receiving files to and from a container.
-Outsider modules are hosted, e.g., on GitHub, and \pkg{outsider} contains discovery functions for them.
-<!-- illustrating outsider examples:
+Each outsider module has a `Dockerfile` and an R package with functions for interacting with the encapsulated tool.
+Using only R functions, an end-user can install a module with the \pkg{outsider} package and then call module code to integrate a tool into their own workflow seamlessly.
+The \pkg{outsider} package and the module manage the containers and handle the transmission of arguments and the transfer of files to and from a container.
+These functionalities also allow a user to launch module code on a remote machine via SSH, expanding the potential computational scale.
+Outsider modules can be hosted code-sharing services, e.g., on GitHub, and \pkg{outsider} contains discovery functions for them.
+
+<!-- @nuest, would you like me to expand on these in a particular way?
 https://github.com/DomBennett/om..astral/blob/master/examples/astral.R

### 2020-01-15 manuscript.Rmd
some reformatting, add some links, some reordering, some labels

--- 2020-01-14 manuscript.Rmd
@@
 
+**[Gigantum](http://gigantum.com/)** also its own image stack ([https://github.com/gigantum/base-images](https://github.com/gigantum/base-images)).
+...
+
 # Use cases and applications
@@
 
+**\pkg{harbor}** ([https://github.com/wch/harbor](https://github.com/wch/harbor) is not actively maintained anymore, but should be honorably mentioned as the first R package for managing Docker images and containers.
+It uses the \pkg{sys} package to run system commands against the Docker CLI, both locally and through an ssh connection, and has convenience functions, e.g., for listing and removing containers/images and for accessing logs.
+The output of container executions are converted to appropriate R types.
+The Docker CLI, while evolving quickly and not with great concern for introducing breaking changes, basic functionality is unchanged for a long time so `harbor::docker_run(image = "hello-world")` works like a charm.
+
+```{r harbor, eval=FALSE, include=FALSE}
+remotes::install_github("wch/harbor")
+harbor::docker_run(image = "hello-world")
+```
+
 **\pkg{stevedore}** ([https://cran.r-project.org/package=stevedore](https://cran.r-project.org/package=stevedore)) 
@@
 -->
+currently the most powerful Docker client in R.
 
@@
 Users can, for example, execute a command inside a container, mount a volume or copy a file.
-
-**\pkg{harbor}** ([https://github.com/wch/harbor](https://github.com/wch/harbor)
-<!-- from the github description
-is an R package for controlling docker containers on local and remote hosts.
--->
 
@@
 
-- [`dockerfiler`](https://github.com/ColinFay/dockerfiler/)
-- [`containerit`](https://github.com/o2r-project/containerit/)
-- [`dockr`](https://github.com/smaakage85/dockr), see also http://smaakage85.netlify.com/2019/12/21/dockr-easy-containerization-for-r/
-- [`dockertest`](https://github.com/traitecoevo/dockertest/)
+**\pkg{dockerfiler}** ([https://github.com/ColinFay/dockerfiler/](https://github.com/ColinFay/dockerfiler/)) ...
+
+**\pkg{containerit}** ([https://github.com/o2r-project/containerit/](https://github.com/o2r-project/containerit/)) ...
+
+**\pkg{dockr}** ([https://github.com/smaakage85/dockr](https://github.com/smaakage85/dockr)) ...
+see/cite http://smaakage85.netlify.com/2019/12/21/dockr-easy-containerization-for-r/
+<!-- from description on CRAN: Build a lightweight Docker r-base container image for your R package, in which the package itself is available.
+The Docker image seeks to mirror your R session as close as possible with respect to R specific dependencies.
+Both dependencies on CRAN R packages as well as local non-CRAN R packages will be included in the Docker container image.-->
+
 - [`renv` with Docker](https://rstudio.github.io/renv/articles/docker.html)
-- https://github.com/gigantum/base-images
 
@@
 
-## Deployment to cloud services
-
-The cloud is the natural environment of containers, and becomes the go-to mechanism to expose applications written in R to users.
-
-- `RSelenium`
-- [`googleComputeEngineR`](https://cloudyr.github.io/googleComputeEngineR/) (function `gce_vm_template()`)
-  - To enable quick deployments of key R services such as RStudio and Shiny onto cloud virtual machines (VMs), this package utilises Dockerfiles to move the labour of setting up those services from the user to a premade Docker image.
-For example, by specifying the template `template="rstudio"` in `gce_vm_template()/gce_vm()` an up to date RStudio Service image is launched.
-Specifying `template="rstudio-gpu"` will launch an RStudio Server image with a GPU attached, etc.
-- [`analogsea`](https://github.com/sckott/analogsea) (digital ocean R client)
-- [`plumber`](https://www.rplumber.io/docs/hosting.html#docker)
-- [production deployment of neural networks](https://github.com/tmobile/r-tensorflow-api) [@jnolis, @nolistic]
-- Kubernetes - A popular platform for managing Docker containers is [Kubernetes](https://kubernetes.io/), which is used for a wide variety of containerized applications.
-It may be your organisation has a Kubernetes cluster already for other applications.
-Docker containers are used within Kubernetes clusters to hold native code, for which Kubernetes creates a framework around network connections and scaling of resources up and down.
-An introduction on using [Kubernetes with R is published at this blog post](https://code.markedmondson.me/r-on-kubernetes-serverless-shiny-r-apis-and-scheduled-scripts/).
-- [`AzureContainers`](https://cran.r-project.org/package=AzureContainers): Umbrella package for working with containers in Microsoft's Azure cloud.
-Provides interfaces to three Azure services: Container Registry, Container Instances (for running individual containers), and Kubernetes Services (for orchestrated deployments)
-
 ## Using R to power enterprise software in production environments
@@
 \label{deployment}
+
+The cloud is the natural environment of containers, and is the natural mechanism to deploy R server applications.
+
+- `RSelenium`
+- [`googleComputeEngineR`](https://cloudyr.github.io/googleComputeEngineR/) (function `gce_vm_template()`)
+  - To enable quick deployments of key R services such as RStudio and Shiny onto cloud virtual machines (VMs), this package utilises Dockerfiles to move the labour of setting up those services from the user to a premade Docker image.
+For example, by specifying the template `template="rstudio"` in `gce_vm_template()/gce_vm()` an up to date RStudio Service image is launched.
+Specifying `template="rstudio-gpu"` will launch an RStudio Server image with a GPU attached, etc.
+- [`analogsea`](https://github.com/sckott/analogsea) (digital ocean R client)
+- [`plumber`](https://www.rplumber.io/docs/hosting.html#docker)
+- [production deployment of neural networks](https://github.com/tmobile/r-tensorflow-api) [@jnolis, @nolistic]
+- Kubernetes - A popular platform for managing Docker containers is [Kubernetes](https://kubernetes.io/), which is used for a wide variety of containerized applications.
+It may be your organisation has a Kubernetes cluster already for other applications.
+Docker containers are used within Kubernetes clusters to hold native code, for which Kubernetes creates a framework around network connections and scaling of resources up and down.
+An introduction on using [Kubernetes with R is published at this blog post](https://code.markedmondson.me/r-on-kubernetes-serverless-shiny-r-apis-and-scheduled-scripts/).
+- [`AzureContainers`](https://cran.r-project.org/package=AzureContainers): Umbrella package for working with containers in Microsoft's Azure cloud.
+Provides interfaces to three Azure services: Container Registry, Container Instances (for running individual containers), and Kubernetes Services (for orchestrated deployments)
 
@@
 Google Cloud Build works alongside Google Container Registry to allow you to build private and public Docker images, which allows you to build up your own dependency graphs for downstream applications.
+  - https://code.markedmondson.me/googleCloudRunner/articles/cloudbuild.html
 
@@
 An R implementation is shown here at [cloudRunR](https://github.com/MarkEdmondson1234/cloudRunR) which uses Cloud Run to create a scalable R plumber API.
+  - How does this relate to https://code.markedmondson.me/googleCloudRunner/index.html ?`
 - `batchtools` \citep{Lang2017batchtools} can [schedule jobs with Docker Swarm](https://mllg.github.io/batchtools/reference/makeClusterFunctionsDocker.html)
@@
 ## Packaging programs for processing pipelines
+\label{packaging}
 
@@
 ## Research Compendia
+\label{compendia}
 
+- researchcompendia.science
 - https://github.com/benmarwick/rrtools
 - \pkg{renv}, https://rstudio.github.io/renv/articles/docker.html
+- what to copy in and what not
 

### 2020-01-15 manuscript.Rmd
minor update of GPU paragraph

@@
 Using **GPUs** (graphical processing units) as a specialised hardware from containerized common work environments is also possible and useful \citep{haydel_enhancing_2015}.
-GPUs are increasingly popular for compute-intensive machine learning tasks, e.g., deep artificial neural networks \citep{schmidhuber_deep_2015}.
-Though in this case, containers are not completely portable between hardware environments.
+GPUs are increasingly popular for compute-intensive machine learning (ML) tasks, e.g., deep artificial neural networks \citep{schmidhuber_deep_2015}.
+Though in this case, containers are not completely portable between hardware environments, but the software stack for ML with GPUs is so complex to set up that a ready-to-use container is helpful.
 Containers running GPU software require drivers and libraries specific to GPU models and versions, and containers require a specialized runtime to connect to the underlying GPU hardware.
 For NVIDIA GPUs, the [NVIDIA Container Toolkit](https://github.com/NVIDIA/nvidia-docker) includes a specialized runtime plugin for Docker and a set of base images with appropriate drivers and libraries.
-The Rocker project has (beta) images based on these that include GPU-enabled versions of machine-learning R packages: https://github.com/rocker-org/ml.
+The Rocker project [has a repository](https://github.com/rocker-org/ml) with (beta) images based on these that include GPU-enabled versions of machine-learning R packages, e.g., `rocker/ml` and `rocker/tensorflow-gpu`.
 

### 2020-01-15 manuscript.Rmd
add section on development and debugging
closes #41

@@
 
-Containers can also serve as useful playgrounds to create environments ad-hoc or to provide very specific environments that are not often needed.
-Developers can readily start a local container based on a Rocker image to investigate a bug report, which might require them to install a specific combination of package versions or use a specific R version (see also versioned images above).
-Unlike recreating the reporter's environment on their regular machine, a container can simply be discarded after use, and there is no need to tediously roll back explorative changes to a previous state.
-Using the Rocker images with RStudio, the disposable environments lack no development comfort (see also the usage in research compendia).
+Containers can also serve as useful playgrounds to create environments ad-hoc or to provide very specific environments that are not needed or not easily available in day-to-day development.
+These environments may have specific versions of R, of R extension packages, and of system libraries used by R extension packages, and all of the above in a specific combination.
 
-**[`r-debug`](https://github.com/wch/r-debug)** is a purpose built Docker image for debugging R memory problems.
-\citet{eddelbuettel_debugging_2019} describes how a Docker container was used to debug an issue with a package only occuring with a particular version of Fortran, and using tools which are not readily available on all platforms (e.g., not on macOS).
+First, such containers can greatly facilitate **fixing bugs**, because developers can readily start and later discard a container to investigate a bug report without affecting their regular system, and potentially having to revert changes back.
+Using the Rocker images with RStudio, these disposable environments lack no development comfort (cf. Section&nbsp;\nameref{compendia}).
+\citet{eddelbuettel_debugging_2019} describes an example how a Docker container was used to debug an issue with a package only occuring with a particular version of Fortran, and using tools which are not readily available on all platforms (e.g., not on macOS).
+For the [R-spatial community](https://www.r-spatial.org/), using containers is particularly useful, because of the strong integration of system libraries in core packages for geospatial data modelling and analysis.
+Purpose built Docker are used to prepare for upcoming releases of system libraries, individual bug reports, and for the lowest supported versions of system libraries \footnote{\href{https://github.com/r-spatial/sf/tree/master/inst/docker}{https://github.com/r-spatial/sf/tree/master/inst/docker},
+\href{https://github.com/Nowosad/rspatial_proj6}{https://github.com/Nowosad/rspatial_proj6}, \href{https://github.com/r-spatial/sf/issues/1231#issuecomment-570076440}{https://github.com/r-spatial/sf/issues/1231\#issuecomment-570076440}}.
 
-The R-hub Docker images (see Section&nbsp;\nameref{rhub}) can also be used to debug problems locally using various combinations of Linux platforms, R versions, and compilers\footnote{See \href{https://r-hub.github.io/rhub/articles/local-debugging.html}{https://r-hub.github.io/rhub/articles/local-debugging.html} and \href{https://blog.r-hub.io/2019/04/25/r-devel-linux-x86-64-debian-clang/}{https://blog.r-hub.io/2019/04/25/r-devel-linux-x86-64-debian-clang/}.
+There are also special images for identifying problems beyond the mere R code, such as **debugging R memory problems**.
+The images significantly reduce the barrier to follow complex steps for fixing memory allocation bugs \citep[cf. Section~4.3 in][]{core_writing_1999}.
+These problems are hard to debug and critical, both because when they occur they lead to fatal crashing processes.
+[`rocker/r-devel-san`](https://github.com/rocker-org/r-devel-san) and [`rocker/r-devel-ubsan-clang`](https://github.com/rocker-org/r-devel-san-clang) are Docker images have a particularly configured version of R to trace such problems with gcc and clang compilers, respectively \citep[cf.~\pkg{sanitizers} for examples,][]{eddelbuettel_sanitizers_2014}.
+The image [`wch/r-debug`](https://github.com/wch/r-debug) is a purpose built Docker image with _multiple_ instrumented builds of R, each with a different diagnostic utility activated.
+
+Second, containers are useful for **testing** R code during development.
+To submit a package to CRAN, an R package must work with the development version of R, which must be compiled locally.
+That can be a challenge for some users.
+The R-hub service (see Section&nbsp;\nameref{rhub}) makes it easy to ensure that no errors occur, but to fix errors a local setup is still often warranted, e.g.
+using the image `rocker/r-devel`, and to test packages with native code, which can make the process more complex \citep[cf.][]{eckert_building_2018}.
+The R-hub Docker images can also be used to debug problems locally using various combinations of Linux platforms, R versions, and compilers\footnote{See \href{https://r-hub.github.io/rhub/articles/local-debugging.html}{https://r-hub.github.io/rhub/articles/local-debugging.html} and \href{https://blog.r-hub.io/2019/04/25/r-devel-linux-x86-64-debian-clang/}{https://blog.r-hub.io/2019/04/25/r-devel-linux-x86-64-debian-clang/}}.
 The images go beyond the configurations, or _flavours_, used by CRAN for checking packages\footnote{\href{https://cran.r-project.org/web/checks/check_flavors.html}{https://cran.r-project.org/web/checks/check\_flavors.html}}, e.g., with CentOS-based images, but lack a container for checking on Windows or OS X.
 The images greatly support package developers to provide support on operating systems they are not familiar.
+The package \pkg{dockertest} ([https://github.com/traitecoevo/dockertest/](https://github.com/traitecoevo/dockertest/)) is a proof of concept for automatically generating `Dockerfile`s and building images specifically to run tests\footnote{\pkg{dockertest} is not actively maintained, but mentioned still because of its interesting approach.}.
+These images are accompanied with a special launch script so the tested source code is not stored in the image but the currently checked in version from a local Git repository is cloned into the container at runtime.
+This approach clearly seperates test environment, test code, and current working copy of the code.
+
+Third, Docker images can be used **on CI platforms** to streamline the testing of packages.
+\citet{ye_docker_2019} describes how they speed up the process of testing by running tasks on [Travis&nbsp;CI](https://travis-ci.org/) within a container using `docker exec`, e.g., the package check or rendering of documentation.
+\citet{cardozo_faster_2018}, also on Travis&nbsp;CI saved time by re-using the testing image as the base iamge for an image intended for publication on Docker Hub.
+Especially for long running tests or complex system dependencies, these approaches seems to improve the development.
+Not due to a concern about time, but to control the environment used on a CI server, even this manuscript was rendered into a PDF and deployed to a GitHub-hosted website (see `.travis.yml` and `Dockerfile` in the manuscript repository).
+This gives on the one hand easy access after every update of the R Markdown source code, and on the other hand a second controlled environment making sure that the article renders successfully and correctly.
 

### 2020-01-15 manuscript.tex
update .tex file

@@
 The Rocker project provides widely-used Docker images for R across
-different application scenarios. This articles surveys describe
-downstream projects building upon Rocker. We also look beyond Rocker to
-other projects connecting containerisation with R. These use cases and
-the diversity of applications demonstrate the power of Rocker and
-containerisation for collaboration, effectivity, scalability, and
-transparency.
+different application scenarios. This articles surveys downstream
+projects building upon Rocker and presents the current state of R
+packages for managing Docker images and controlling containers. We also
+look beyond Rocker to other projects connecting containerisation with R,
+namely alternative suites of images. These use cases and the variety of
+applications demonstrate the power of Rocker specifically and
+containerisation in general. We identified common themes across this
+diversity: reproducible environments, scalability and effectivity, and
+portability across clouds.
 }
@@
 \section{Introduction}\label{introduction}}
+
+\label{intro}
 
@@
 \href{https://en.wikipedia.org/wiki/Docker_(software)}{Docker} as a
-general tool for distribution and deployment of server applications.
-Combining both these topics, the \emph{Rocker~Project}
+general tool for distribution and deployment of server applications---in
+fact, Docker can be called the \emph{lingua~franca} of describing
+computing environments and packaging software. Combining both these
+topics, the \emph{Rocker~Project}
 (\url{https://www.rocker-project.org/}) provides images with R (see the
@@
 Rocker}\label{containerization-and-rocker}}
+
+\label{containerisation} \label{rocker}
 
@@
 interface has lead to widespread adoption and use of Docker in a variety
-of domains, e.g., cloud computing infrastructure \citep{Bernstein2014},
-data science \citep{boettiger_introduction_2015}, and edge computing
-\citep{alam_orchestration_2018} {[}more examples!{]}. It provided to be
-a natural match for ``cloud deployment'' which runs, or at least appears
-to run, ``seamlessly'' without much explicit reference to the underlying
+of domains, e.g., cloud computing infrastructure
+\citep[e.g.,][]{Bernstein2014}, data science
+\citep[e.g.,][]{boettiger_introduction_2015}, and edge computing
+\citep[e.g.,][]{alam_orchestration_2018}. It provided to be a natural
+match for ``cloud deployment'' which runs, or at least appears to run,
+``seamlessly'' without much explicit reference to the underlying
 machine, architecture or operating system: containers are portable and
@@
 containers spanning to from building blocks with \texttt{R-release} or
-\texttt{R-devel}, via containers with RStudio\textasciitilde{}Server and
-Shiny\textasciitilde{}Server, to domain-specific containers such as
-\texttt{geospatial}. These containers form \emph{image stacks}, building
-on top of each other for better maintainability and Also of note is a
-series of ``versioned'' containers which match the R release they
-contain with the \emph{then-current} set of packages via the MRAN
-Snapshot views of CRAN \citep{microsoft_cran_2019}. The Rocker Project's
-impact and importance was acknowledged by the Chan Zuckerberg
-Initiative's \emph{Essential Open Source Software for Science}, who
-provide findung for the projects's sustainable maintenance, community
-growth, and targeting new hardware platforms including GPUs
-\citep{chan_zuckerberg_initiative_maintaining_2019}.
+\texttt{R-devel}, via containers with
+\href{https://rstudio.com/products/rstudio/}{RStudio~Server} and
+\href{https://rstudio.com/products/shiny/shiny-server/}{Shiny~Server},
+to domain-specific containers such as
+\href{https://github.com/rocker-org/geospatial}{\texttt{rocker/geospatial}}
+\citep{rocker_geospatial_2019}. These containers form \emph{image
+stacks}, building on top of each other for better maintainability
+(i.e.~smaller \texttt{Dockerfile}s), composability, and to reduce build
+time. Also of note is a series of ``versioned'' containers which match
+the R release they contain with the \emph{then-current} set of packages
+via the MRAN Snapshot views of CRAN \citep{microsoft_cran_2019}. The
+Rocker Project's impact and importance was acknowledged by the Chan
+Zuckerberg Initiative's \emph{Essential Open Source Software for
+Science}, who provide findung for the projects's sustainable
+maintenance, community growth, and targeting new hardware platforms
+including GPUs \citep{chan_zuckerberg_initiative_maintaining_2019}.
 
@@
 in running images where there is an init script, e.g., in containers
-that by default run RStudio. In the latter case, a \texttt{Singularity}
-file, a recipe akin to a \texttt{Dockerfile}, needs to be used. To date,
-no comparable image stack to Rocker exists on
+that by default run RStudio~Server. In the latter case, a
+\texttt{Singularity} file, a recipe akin to a \texttt{Dockerfile}, needs
+to be used. To date, no comparable image stack to Rocker exists on
 \href{https://singularity-hub.org/}{Singularity Hub}. A further tool for
@@
 
+\label{images}
+
 \hypertarget{images-for-alternative-r-distributions}{%
@@
 distributions}\label{images-for-alternative-r-distributions}}
+
+\label{alternatives}
 
@@
 
+\label{bioc}
+
 \emph{Bioconductor} (\url{https://bioconductor.org/}) is an open source,
@@
 
+\label{versions}
+
 As with any other software, each new version of R comes with its share
@@
 \subsection{Windows Images}\label{windows-images}}
+
+\label{windows}
 
@@
 \subsection{Non-Debian Linux images}\label{non-debian-linux-images}}
+
+\label{nondebian}
 
@@
 
-\textbf{RStudio} also maintains a set of
+\href{https://rstudio.com/}{\textbf{RStudio}} also maintains a set of
 \href{https://github.com/rstudio/r-docker}{Docker images} for multiple
 operating systems including SUSE, Ubuntu, CentOS, and Debian. The image
-stacks include an \href{https://github.com/rstudio/r-builds}{opinionated
-R installation} and ensure the R installation and profile is consistent
-across the different Linux distributions. The images also install R in a
-versioned directory and use a minimal set of build and runtime
-dependencies, Furthermore these images are responsible for creating
+stack includes an \href{https://github.com/rstudio/r-builds}{opinionated
+R installation} and ensures the R installation and profile is consistent
+across the different Linux distributions. The base image also installs R
+in a versioned directory and uses a minimal set of build and runtime
+dependencies. Furthermore these images are responsible for creating
 pre-compiled binary R packages for Linux, which supplement the binaries
@@
 
+\label{datascience}
+
 \emph{Data Science} is a widely discussed topic among all academic
@@
 
+\textbf{\href{http://gigantum.com/}{Gigantum}} also its own image stack
+(\url{https://github.com/gigantum/base-images}). \ldots{}
+
 \hypertarget{use-cases-and-applications}{%
 \section{Use cases and applications}\label{use-cases-and-applications}}
+
+\label{applications}
 
@@
 
+\textbf{\pkg{harbor}} (\url{https://github.com/wch/harbor} is not
+actively maintained anymore, but should be honorably mentioned as the
+first R package for managing Docker images and containers. It uses the
+\pkg{sys} package to run system commands against the Docker CLI, both
+locally and through an ssh connection, and has convenience functions,
+e.g., for listing and removing containers/images and for accessing logs.
+The output of container executions are converted to appropriate R types.
+The Docker CLI, while evolving quickly and not with great concern for
+introducing breaking changes, basic functionality is unchanged for a
+long time so \texttt{harbor::docker\_run(image\ =\ "hello-world")} works
+like a charm.
+
 \textbf{\pkg{stevedore}}
-(\url{https://cran.r-project.org/package=stevedore})
+(\url{https://cran.r-project.org/package=stevedore}) currently the most
+powerful Docker client in R.
 
@@
 execute a command inside a container, mount a volume or copy a file.
-
-\textbf{\pkg{harbor}} (\url{https://github.com/wch/harbor}
 
@@
 
+\label{envs}
+
 Several second order R packages attempt to make the process of creating
@@
 
+\textbf{\pkg{dockerfiler}}
+(\url{https://github.com/ColinFay/dockerfiler/}) \ldots{}
+
+\textbf{\pkg{containerit}}
+(\url{https://github.com/o2r-project/containerit/}) \ldots{}
+
+\textbf{\pkg{dockr}} (\url{https://github.com/smaakage85/dockr})
+\ldots{} see/cite
+\url{http://smaakage85.netlify.com/2019/12/21/dockr-easy-containerization-for-r/}
+
 \begin{itemize}
@@
 \item
-  \href{https://github.com/ColinFay/dockerfiler/}{\texttt{dockerfiler}}
-\item
-  \href{https://github.com/o2r-project/containerit/}{\texttt{containerit}}
-\item
-  \href{https://github.com/smaakage85/dockr}{\texttt{dockr}}, see also
-  \url{http://smaakage85.netlify.com/2019/12/21/dockr-easy-containerization-for-r/}
-\item
-  \href{https://github.com/traitecoevo/dockertest/}{\texttt{dockertest}}
-\item
   \href{https://rstudio.github.io/renv/articles/docker.html}{\texttt{renv}
   with Docker}
-\item
-  \url{https://github.com/gigantum/base-images}
 \end{itemize}
@@
 
-\hypertarget{deployment-to-cloud-services}{%
-\subsection{Deployment to cloud
-services}\label{deployment-to-cloud-services}}
-
-The cloud is the natural environment of containers, and becomes the
-go-to mechanism to expose applications written in R to users.
-
-\begin{itemize}
-\tightlist
-\item
-  \texttt{RSelenium}
-\item
-  \href{https://cloudyr.github.io/googleComputeEngineR/}{\texttt{googleComputeEngineR}}
-  (function \texttt{gce\_vm\_template()})
-
-  \begin{itemize}
-  \tightlist
-  \item
-    To enable quick deployments of key R services such as RStudio and
-    Shiny onto cloud virtual machines (VMs), this package utilises
-    Dockerfiles to move the labour of setting up those services from the
-    user to a premade Docker image. For example, by specifying the
-    template \texttt{template="rstudio"} in
-    \texttt{gce\_vm\_template()/gce\_vm()} an up to date RStudio Service
-    image is launched. Specifying \texttt{template="rstudio-gpu"} will
-    launch an RStudio Server image with a GPU attached, etc.\\
-  \end{itemize}
-\item
-  \href{https://github.com/sckott/analogsea}{\texttt{analogsea}}
-  (digital ocean R client)
-\item
-  \href{https://www.rplumber.io/docs/hosting.html\#docker}{\texttt{plumber}}
-\item
-  \href{https://github.com/tmobile/r-tensorflow-api}{production
-  deployment of neural networks} \citep[\citet{nolistic}]{jnolis}
-\item
-  Kubernetes - A popular platform for managing Docker containers is
-  \href{https://kubernetes.io/}{Kubernetes}, which is used for a wide
-  variety of containerized applications. It may be your organisation has
-  a Kubernetes cluster already for other applications. Docker containers
-  are used within Kubernetes clusters to hold native code, for which
-  Kubernetes creates a framework around network connections and scaling
-  of resources up and down. An introduction on using
-  \href{https://code.markedmondson.me/r-on-kubernetes-serverless-shiny-r-apis-and-scheduled-scripts/}{Kubernetes
-  with R is published at this blog post}.
-\item
-  \href{https://cran.r-project.org/package=AzureContainers}{\texttt{AzureContainers}}:
-  Umbrella package for working with containers in Microsoft's Azure
-  cloud. Provides interfaces to three Azure services: Container
-  Registry, Container Instances (for running individual containers), and
-  Kubernetes Services (for orchestrated deployments)
-\end{itemize}
-
 \hypertarget{using-r-to-power-enterprise-software-in-production-environments}{%
@@
 environments}\label{using-r-to-power-enterprise-software-in-production-environments}}
+
+\label{enterprise}
 
@@
 
+\label{deployment}
+
+The cloud is the natural environment of containers, and is the natural
+mechanism to deploy R server applications.
+
+\begin{itemize}
+\tightlist
+\item
+  \texttt{RSelenium}
+\item
+  \href{https://cloudyr.github.io/googleComputeEngineR/}{\texttt{googleComputeEngineR}}
+  (function \texttt{gce\_vm\_template()})
+
+  \begin{itemize}
+  \tightlist
+  \item
+    To enable quick deployments of key R services such as RStudio and
+    Shiny onto cloud virtual machines (VMs), this package utilises
+    Dockerfiles to move the labour of setting up those services from the
+    user to a premade Docker image. For example, by specifying the
+    template \texttt{template="rstudio"} in
+    \texttt{gce\_vm\_template()/gce\_vm()} an up to date RStudio Service
+    image is launched. Specifying \texttt{template="rstudio-gpu"} will
+    launch an RStudio Server image with a GPU attached, etc.\\
+  \end{itemize}
+\item
+  \href{https://github.com/sckott/analogsea}{\texttt{analogsea}}
+  (digital ocean R client)
+\item
+  \href{https://www.rplumber.io/docs/hosting.html\#docker}{\texttt{plumber}}
+\item
+  \href{https://github.com/tmobile/r-tensorflow-api}{production
+  deployment of neural networks} \citep[\citet{nolistic}]{jnolis}
+\item
+  Kubernetes - A popular platform for managing Docker containers is
+  \href{https://kubernetes.io/}{Kubernetes}, which is used for a wide
+  variety of containerized applications. It may be your organisation has
+  a Kubernetes cluster already for other applications. Docker containers
+  are used within Kubernetes clusters to hold native code, for which
+  Kubernetes creates a framework around network connections and scaling
+  of resources up and down. An introduction on using
+  \href{https://code.markedmondson.me/r-on-kubernetes-serverless-shiny-r-apis-and-scheduled-scripts/}{Kubernetes
+  with R is published at this blog post}.
+\item
+  \href{https://cran.r-project.org/package=AzureContainers}{\texttt{AzureContainers}}:
+  Umbrella package for working with containers in Microsoft's Azure
+  cloud. Provides interfaces to three Azure services: Container
+  Registry, Container Instances (for running individual containers), and
+  Kubernetes Services (for orchestrated deployments)
+\end{itemize}
+
 While Plumber provides the low-level infrastructure for turning a
@@
 
+\label{cicd}
+
 The controlled nature of containers, i.e., it is possible to define the
@@
     up your own dependency graphs for downstream applications.
+  \item
+    \url{https://code.markedmondson.me/googleCloudRunner/articles/cloudbuild.html}
   \end{itemize}
@@
 environments}\label{common-or-public-work-environments}}
+
+\label{workenvs}
 
@@
 and useful \citep{haydel_enhancing_2015}. GPUs are increasingly popular
-for compute-intensive machine learning tasks, e.g., deep artificial
+for compute-intensive machine learning (ML) tasks, e.g., deep artificial
 neural networks \citep{schmidhuber_deep_2015}. Though in this case,
-containers are not completely portable between hardware environments.
-Containers running GPU software require drivers and libraries specific
-to GPU models and versions, and containers require a specialized runtime
-to connect to the underlying GPU hardware. For NVIDIA GPUs, the
+containers are not completely portable between hardware environments,
+but the software stack for ML with GPUs is so complex to set up that a
+ready-to-use container is helpful. Containers running GPU software
+require drivers and libraries specific to GPU models and versions, and
+containers require a specialized runtime to connect to the underlying
+GPU hardware. For NVIDIA GPUs, the
 \href{https://github.com/NVIDIA/nvidia-docker}{NVIDIA Container Toolkit}
 includes a specialized runtime plugin for Docker and a set of base
-images with appropriate drivers and libraries. The Rocker project has
-(beta) images based on these that include GPU-enabled versions of
-machine-learning R packages: \url{https://github.com/rocker-org/ml}.
+images with appropriate drivers and libraries. The Rocker project
+\href{https://github.com/rocker-org/ml}{has a repository} with (beta)
+images based on these that include GPU-enabled versions of
+machine-learning R packages, e.g., \texttt{rocker/ml} and
+\texttt{rocker/tensorflow-gpu}.
 
@@
 
+\label{rstudio}
+
 In addition to embedding RStudio Open Source Server inside of Docker
@@
   \href{https://solutions.rstudio.com/launcher/overview/}{Launcher}. It
-  gives users the ability to spawn RStudio instances and background jobs
-  in a scalable way on external clusters, e.g.,
+  gives users the ability to spawn R sessions and background jobs in a
+  scalable way on external clusters, e.g.,
   \href{https://support.rstudio.com/hc/en-us/articles/360019253393-Using-Docker-images-with-RStudio-Server-Pro-Launcher-and-Kubernetes}{Kubernetes
@@
 \subsection{Processing {[}div{]}}\label{processing-div}}
+
+\label{processing}
 
@@
   uses Cloud Run to create a scalable R plumber API.
+
+  \begin{itemize}
+  \tightlist
+  \item
+    How does this relate to
+    \url{https://code.markedmondson.me/googleCloudRunner/index.html} ?`
+  \end{itemize}
 \item
@@
 
-\hypertarget{research-compendia-benmarwick}{%
-\subsection{\texorpdfstring{Research Compendia
-\citep{benmarwick}}{Research Compendia {[}@benmarwick{]}}}\label{research-compendia-benmarwick}}
+\hypertarget{packaging-programs-for-processing-pipelines}{%
+\subsection{Packaging programs for processing
+pipelines}\label{packaging-programs-for-processing-pipelines}}
+
+\label{packaging}
+
+Docker is perfectly suited to package and execute any software and data,
+and is a good candidate to build complex processing pipelines
+independent of the used programming language. Because of the original
+use case (see~\nameref{introduction}), Docker does not have a standard
+mechanism to chain a number of containers together, i.e., to define how
+environment variables, volume mounts, or ports can be used to pass input
+parameters and data into a container and how to get results out. Some
+packages, e.g.~\pkg{containerit}, provide a Docker image that can be
+used similar as a CLI, but the usage is
+cumbersome\footnote{\href{https://o2r.info/containerit/articles/container.html}{https://o2r.info/containerit/articles/container.html}}.
+
+\textbf{\pkg{outsider}} (\url{https://antonellilab.github.io/outsider/})
+tackles the problem of integrating external programmes into an R
+workflow \citep{bennett_outsider_2019}. Installation and usage of
+external programmes can be cumbersome and even prohibited if platform
+dependent. Therefore \pkg{outsider} uses the platform-independent Docker
+images to encapsulate processes in \emph{outsider modules}. Each
+outsider modules has a \texttt{Dockerfile} and an R package with
+functions for interacting with the encapsulated tool. A user only needs
+to use the given functions to install a module and can then integrate a
+tool into their own workflow seaminglessly using only R functions. The
+module takes care of transmitting arguments and sending and receiving
+files to and from a container. Outsider modules are hosted, e.g., on
+GitHub, and \pkg{outsider} contains discovery functions for them.
+
+\hypertarget{research-compendia}{%
+\subsection{Research Compendia}\label{research-compendia}}
+
+\label{compendia}
 
@@
 \item
+  researchcompendia.science
+\item
   \url{https://github.com/benmarwick/rrtools}
@@
   \pkg{renv}, \url{https://rstudio.github.io/renv/articles/docker.html}
+\item
+  what to copy in and what not
 \end{itemize}
@@
 Containers can also serve as useful playgrounds to create environments
-ad-hoc or to provide very specific environments that are not often
-needed. Developers can readily start a local container based on a Rocker
-image to investigate a bug report, which might require them to install a
-specific combination of package versions or use a specific R version
-(see also versioned images above). Unlike recreating the reporter's
-environment on their regular machine, a container can simply be
-discarded after use, and there is no need to tediously roll back
-explorative changes to a previous state. Using the Rocker images with
-RStudio, the disposable environments lack no development comfort (see
-also the usage in research compendia).
-
-\textbf{\href{https://github.com/wch/r-debug}{\texttt{r-debug}}} is a
-purpose built Docker image for debugging R memory problems.
-\citet{eddelbuettel_debugging_2019} describes how a Docker container was
-used to debug an issue with a package only occuring with a particular
-version of Fortran, and using tools which are not readily available on
-all platforms (e.g., not on macOS).
-
-The R-hub Docker images (see Section~\nameref{rhub}) can also be used to
-debug problems locally using various combinations of Linux platforms, R
-versions, and compilers\textbackslash{}footnote\{See
-\href{https://r-hub.github.io/rhub/articles/local-debugging.html}{https://r-hub.github.io/rhub/articles/local-debugging.html}
+ad-hoc or to provide very specific environments that are not needed or
+not easily available in day-to-day development. These environments may
+have specific versions of R, of R extension packages, and of system
+libraries used by R extension packages, and all of the above in a
+specific combination.
+
+First, such containers can greatly facilitate \textbf{fixing bugs},
+because developers can readily start and later discard a container to
+investigate a bug report without affecting their regular system, and
+potentially having to revert changes back. Using the Rocker images with
+RStudio, these disposable environments lack no development comfort
+(cf.~Section~\nameref{compendia}). \citet{eddelbuettel_debugging_2019}
+describes an example how a Docker container was used to debug an issue
+with a package only occuring with a particular version of Fortran, and
+using tools which are not readily available on all platforms (e.g., not
+on macOS). For the \href{https://www.r-spatial.org/}{R-spatial
+community}, using containers is particularly useful, because of the
+strong integration of system libraries in core packages for geospatial
+data modelling and analysis. Purpose built Docker are used to prepare
+for upcoming releases of system libraries, individual bug reports, and
+for the lowest supported versions of system libraries
+\textbackslash{}footnote\{\href{https://github.com/r-spatial/sf/tree/master/inst/docker}{https://github.com/r-spatial/sf/tree/master/inst/docker},
+\textbackslash{}href\{\url{https://github.com/Nowosad/rspatial_proj6}\}\{\url{https://github.com/Nowosad/rspatial_proj6}\},
+\href{https://github.com/r-spatial/sf/issues/1231#issuecomment-570076440}{https://github.com/r-spatial/sf/issues/1231\#issuecomment-570076440}\}.
+
+There are also special images for identifying problems beyond the mere R
+code, such as \textbf{debugging R memory problems}. The images
+significantly reduce the barrier to follow complex steps for fixing
+memory allocation bugs \citep[cf. Section~4.3 in][]{core_writing_1999}.
+These problems are hard to debug and critical, both because when they
+occur they lead to fatal crashing processes.
+\href{https://github.com/rocker-org/r-devel-san}{\texttt{rocker/r-devel-san}}
 and
-\href{https://blog.r-hub.io/2019/04/25/r-devel-linux-x86-64-debian-clang/}{https://blog.r-hub.io/2019/04/25/r-devel-linux-x86-64-debian-clang/}.
+\href{https://github.com/rocker-org/r-devel-san-clang}{\texttt{rocker/r-devel-ubsan-clang}}
+are Docker images have a particularly configured version of R to trace
+such problems with gcc and clang compilers, respectively
+\citep[cf.~\pkg{sanitizers} for examples,][]{eddelbuettel_sanitizers_2014}.
+The image \href{https://github.com/wch/r-debug}{\texttt{wch/r-debug}} is
+a purpose built Docker image with \emph{multiple} instrumented builds of
+R, each with a different diagnostic utility activated.
+
+Second, containers are useful for \textbf{testing} R code during
+development. To submit a package to CRAN, an R package must work with
+the development version of R, which must be compiled locally. That can
+be a challenge for some users. The R-hub service (see
+Section~\nameref{rhub}) makes it easy to ensure that no errors occur,
+but to fix errors a local setup is still often warranted, e.g.~using the
+image \texttt{rocker/r-devel}, and to test packages with native code,
+which can make the process more complex
+\citep[cf.][]{eckert_building_2018}. The R-hub Docker images can also be
+used to debug problems locally using various combinations of Linux
+platforms, R versions, and
+compilers\footnote{See \href{https://r-hub.github.io/rhub/articles/local-debugging.html}{https://r-hub.github.io/rhub/articles/local-debugging.html} and \href{https://blog.r-hub.io/2019/04/25/r-devel-linux-x86-64-debian-clang/}{https://blog.r-hub.io/2019/04/25/r-devel-linux-x86-64-debian-clang/}}.
 The images go beyond the configurations, or \emph{flavours}, used by
@@
 Windows or OS X. The images greatly support package developers to
-provide support on operating systems they are not familiar.
+provide support on operating systems they are not familiar. The package
+\pkg{dockertest} (\url{https://github.com/traitecoevo/dockertest/}) is a
+proof of concept for automatically generating \texttt{Dockerfile}s and
+building images specifically to run
+tests\footnote{\pkg{dockertest} is not actively maintained, but mentioned still because of its interesting approach.}.
+These images are accompanied with a special launch script so the tested
+source code is not stored in the image but the currently checked in
+version from a local Git repository is cloned into the container at
+runtime. This approach clearly seperates test environment, test code,
+and current working copy of the code.
+
+Third, Docker images can be used \textbf{on CI platforms} to streamline
+the testing of packages. \citet{ye_docker_2019} describes how they speed
+up the process of testing by running tasks on
+\href{https://travis-ci.org/}{Travis~CI} within a container using
+\texttt{docker\ exec}, e.g., the package check or rendering of
+documentation. \citet{cardozo_faster_2018}, also on Travis~CI saved time
+by re-using the testing image as the base iamge for an image intended
+for publication on Docker Hub. Especially for long running tests or
+complex system dependencies, these approaches seems to improve the
+development. Not due to a concern about time, but to control the
+environment used on a CI server, even this manuscript was rendered into
+a PDF and deployed to a GitHub-hosted website (see \texttt{.travis.yml}
+and \texttt{Dockerfile} in the manuscript repository). This gives on the
+one hand easy access after every update of the R Markdown source code,
+and on the other hand a second controlled environment making sure that
+the article renders successfully and correctly.
 
@@
 This article is a snapshot of the R-corner in a universe of applications
-built many-faced piece of software, Docker. Docker became synonymous
-with applying the concept of containerisation to solve many challenges,
-such as reproducible environments in research, scalable deployments in
-heterogenous organisations, and the ability to move processing between
+built many-faced piece of software, Docker. \texttt{Dockerfile}s and
+Docker images are the go-to methods for collaboration between roles in
+an organisation, such as development and IT operations teams, and
+between parties in the communication of knowledge, such as research
+workflows or education. Docker became synonymous with applying the
+concept of containerisation to solve challenges of reproducible
+environments, e.g., in research and in development \& production, and of
+scalable deployments with the ability to move processing between
 machines easily (e.g., locally, one cloud providers VM, another cloud
-provider's Container-as-a-Service). These are probably the thre common
-themes described in this work.
+provider's Container-as-a-Service). Collaboration, reproducibility, and
+portability are the common themes behind R packages, use cases, and
+applications in this work.
 
 The R packages and use cases presented show the positive drive in the
-community to innovate, but also the challenge of keeping up with
-continuous evolving landscape. The use cases contributed by co-authors
+community to innovate, but also the challenge of keeping up with a
+continuously evolving landscape. The use cases contributed by co-authors
 also have a degree of overlap. A common language and understanding of
@@
 and language, and (c) a sustainable continuation and funding for all of
-development, community support, and education. A concrete effort by this
-work towards these pieces is to sustain the structure and captured
-status quo in a CRAN Task View \emph{Containerization}.
+development, community support, and education. A concrete effort to work
+towards these pieces is to sustain the structure and captured status quo
+from this work in form of a CRAN Task View on containerization.
 
@@
 \texttt{r-db}. BW .. KR .. NR .. NX wrote the section on liftr. LS \& NT
-wrote the section on Bioconductor. All authors contributed minor edits,
-gave feedback on other sections, and approved the final version.
+wrote the section on Bioconductor. All authors approved the final
+version.
 

### 2020-01-15 manuscript.Rmd
small updates about windows images

@@
 The Docker CLI commands are forwarded to the virtual host transparently for the user.
+
 For current Windows Server (2016 and later) and Windows Desktop ([Docker Desktop](https://www.docker.com/products/docker-desktop) requires Windows 10) versions, Docker is supported natively\footnote{\href{https://www.docker.com/products/windows-containers}{https://www.docker.com/products/windows-containers}} and different base images are offered by Microsoft\footnote{\href{https://docs.microsoft.com/en-us/virtualization/windowscontainers/about/index}{https://docs.microsoft.com/en-us/virtualization/windowscontainers/about/index}}.
@@
 The images are built from manually maintained seperate `Dockerfile`s, are tagged with both Windows-variant and R version, e.g., `nuest/rocker-win:ltsc2019-3.6.2` or `nuest/rocker-win:1803-latest`, and run the `R.exe` process by default.
+All these images can be run on Docker for Desktop on Windows 10.
 New R versions are only added on demand.
-The explorative project demonstrates the variety and pontentially confusing complexity of Windows on Docker, e.g., because of the different base images, but also provides a starting point for any Windows-constrained user to leverage containerisation, including examples for apps using \pkg{plumber\, \pkg{Shiny}, or R packages with system dependencies such as \pkg{sf}.
-The latter is quite comfortable actually, as R ships pre-compiled binaries for Windows.
+The explorative rocker-win project demonstrates the variety and pontentially confusing complexity of Windows on Docker, e.g., because of the different base images, but also provides a starting point for any Windows-constrained user to leverage containerisation, including examples for apps using \pkg{plumber\, \pkg{Shiny}, or R packages with system dependencies such as \pkg{sf}.
+The latter is quite comfortable and fast actually, compared to installation from source on Linux, as R ships pre-compiled binaries for Windows.
+However, compared to the various Linux images, the footprint of Windows images is quite large.
 
@@
 In most cases, individual developers or researchers worrying about reproducibility, will prefer the more widely and host independent Linux-based containers.
-Naturally, a Windows license is required for the host machine.
+Naturally, a Windows license is required for the host machine, and the licensing impacts the potnetial to build images in clouds or redistribute exported images as archive files.
 But the developments in the `rocker-win` prototype show that containers can also be used for R workflows depending on Windows-only tools, for leveraging an existing Windows Server-based infrastructure where policies can otherwise not be met, or for streamlining interactions with system operating staff that runs Windows-based containers themselves.

### 2020-01-16 manuscript.Rmd
reformat renv link, mention sevenbridges in README
Add draft based on https://hackmd.io/@Sq-NZxXnQQGtVzoZDwyznA/Sk0ijR8IV

@@
 
-- [`renv` with Docker](https://rstudio.github.io/renv/articles/docker.html)
+**\pkg{renv}** ([https://rstudio.github.io/renv/](https://rstudio.github.io/renv/)) with Docker: https://rstudio.github.io/renv/articles/docker.html
 
@@
 - https://github.com/benmarwick/rrtools
+
+Containers provide a high degree of isolation that is often desirable when attempting to capture a specific computational environment so that others can reproduce and extend a research result.
+Many computationally intensive research projects depend on specific versions of original and third-party software packages in diverse languages, joined together to form a  pipeline through which data flows.
+New releases of even just a single piece of software in this pipeline can break the entire workflow, making it difficult for others to use, and making the original the results irreproducible.
+The chance of a substantial disruption like this is high in a multi-year research project where key pieces of third-party software may have several major updates over the duration of the project.
+Containers give the researcher an isolated environment to assemble these research pipelines with specific versions of software.
+Research workflows in containers are safe from contamination from other activities occuring on the researcher's computer, for example the installation of newest version of packages for teaching demonstrations.
+
+In many fields we see that a typical unit of research for a container is a research report or journal article, where the container holds the compendium, or self-contained set of data (or connections to data elsewhere) and code files needed to fully reproduce the article [@doi:10.1080/00031305.2017.1375986].
+Images of compendium containers can be hosted on services such as Docker Hub for convienient sharing among collaborators and others.
+Similarly, packages such as containerit and dockerfiler can be used to create a Dockerfile to be archived with a compendium on a data repository (e.g.
+Zenodo, Dataverse, Figshare, OSF).
+A typial compendium Dockerfile will pull a rocker image fixed to a specific version of R, and install R packages from the MRAN repository to ensure the package versions are tied to a specific date, rather than the most recent version.
+A future researchers can download then compendium from the repository and run the included Dockerfile build a new image that recreates the computational environment used to produce the original research results.
+
 - \pkg{renv}, https://rstudio.github.io/renv/articles/docker.html

### 2020-01-16 manuscript.Rmd
one sentence per line in compendia draft

@@
 
-- researchcompendia.science
-- https://github.com/benmarwick/rrtools
-
 Containers provide a high degree of isolation that is often desirable when attempting to capture a specific computational environment so that others can reproduce and extend a research result.
@@
 Images of compendium containers can be hosted on services such as Docker Hub for convienient sharing among collaborators and others. 
-Similarly, packages such as containerit and dockerfiler can be used to create a Dockerfile to be archived with a compendium on a data repository (e.g.
-Zenodo, Dataverse, Figshare, OSF).
+Similarly, packages such as containerit and dockerfiler can be used to create a Dockerfile to be archived with a compendium on a data repository (e.g. Zenodo, Dataverse, Figshare, OSF).
 A typial compendium Dockerfile will pull a rocker image fixed to a specific version of R, and install R packages from the MRAN repository to ensure the package versions are tied to a specific date, rather than the most recent version.
@@
 
+- Website researchcompendia.science
+- Gentlemen
+- R and friends (what do they recommend)
+- rrtools - https://github.com/benmarwick/rrtools
+- Some concrete examples, incl. Madjebe 
 - \pkg{renv}, https://rstudio.github.io/renv/articles/docker.html

### 2020-01-16 manuscript.Rmd
write research compendia section, closes #40

--- 2020-01-15 manuscript.Rmd
@@
 
-- [`renv` with Docker](https://rstudio.github.io/renv/articles/docker.html)
+**\pkg{renv}** ([https://rstudio.github.io/renv/](https://rstudio.github.io/renv/)) with Docker: https://rstudio.github.io/renv/articles/docker.html
+
+_version pinning_
 
@@
 
-- researchcompendia.science
-- https://github.com/benmarwick/rrtools
-- \pkg{renv}, https://rstudio.github.io/renv/articles/docker.html
-- what to copy in and what not
+Containers provide a high degree of isolation that is often desirable when attempting to capture a specific computational environment so that others can reproduce and extend a research result.
+Many computationally intensive research projects depend on specific versions of original and third-party software packages in diverse languages, joined together to form a pipeline through which data flows.
+New releases of even just a single piece of software in this pipeline can break the entire workflow, making it difficult to find the error and for others to reuse existing pipelines, and making the original the results irreproducible and not extandable.
+The chance of a substantial disruption like this is high in a multi-year research project where key pieces of third-party software may have several major updates over the duration of the project.
+The classical "paper" article is insufficient to adequately communicate the knowledge behind such research projects \citep[cf.][]{donoho_invitation_2010,marwick_how_2015}.
+\citet{gentleman_statistical_2007} coined the term _Research Compendium_ for a dynamic document together with supporting data and code.
+They use the R package system \cite{core_writing_1999} as for the functional prototype to structuring, validation, and distribution of research compendia.
+This concept has been taken up and extended\footnote{See full literature list at \href{https://research-compendium.science/}{https://research-compendium.science/}.}, not the least by applying containerisation and other methods for managing computing environments---see Section&nbsp;\nameref{envs}.
+Containers give the researcher an isolated environment to assemble these research pipelines with specific versions of software.
+Research workflows in containers are safe from contamination from other activities occuring on the researcher's computer, for example the installation of newest version of packages for teaching demonstrations.
+Given the users in this scenario, i.e., often academics with limited formal software development training, templates and assistance with containers around research compendia is essential.
+
+In many fields we see that a typical unit of research for a container is a research report or journal article, where the container holds the compendium, or self-contained set of data (or connections to data elsewhere) and code files needed to fully reproduce the article \citep{marwick_packaging_2018}.
+The package \pkg{rrtools} ([https://github.com/benmarwick/rrtools](https://github.com/benmarwick/rrtools)) provides a template and convienence functions to apply good practices for research compendia, including a starter `Dockerfile`.
+Images of compendium containers can be hosted on services such as Docker&nbsp;Hub for convienient sharing among collaborators and others. 
+Similarly, packages such as \pkg{containerit} and \pkg{dockerfiler} can be used to manage the `Dockerfile` to be archived with a compendium on a data repository (e.g.
+[Zenodo](https://zenodo.org/), [Dataverse](https://dataverse.org/), [Figshare](https://figshare.com/), [OSF](https://osf.io/)).
+A typial compendium's `Dockerfile` will pull a rocker image fixed to a specific version of R, and install R packages from the MRAN repository to ensure the package versions are tied to a specific date, rather than the most recent version.
+A future researchers can download then compendium from the repository and run the included `Dockerfile` build a new image that recreates the computational environment used to produce the original research results.
+If building the image fails, the human-readable instructions in a `Dockerfile` are a the starting point for rebuilding the environment.
+Further safeguarding practices are currently under development or not part of common practice yet, such as preservation of images \citep{emsley_framework_2018} and storing them alongside `Dockerfile`s \citep[cf.][]{nust_opening_2017}, or pinning of system libraries beyond the tagged base images, which may be seen as stable or dynamic depending on the applied time scale \citep[see discussion on `debian:testing` base image in][]{RJ-2017-065}.
 

### 2020-01-16 manuscript.Rmd
update conclusion, protected space for Docker Hub

--- 2020-01-16 manuscript.Rmd
@@
 Combining both these topics, the _Rocker&nbsp;Project_ ([https://www.rocker-project.org/](https://www.rocker-project.org/)) provides images with R (see the next Section for more details).
-The considerable uptake and continued evolution of the Rocker Project has lead to numerous projects extending or building upon Rocker images, ranging from reproducible research to production deployments.
+The considerable uptake and continued evolution of the Rocker Project has lead to numerous projects extending or building upon Rocker images, ranging from reproducible research\footnote{"Reproducible" in the sense of the _Claerbout/Donoho/Peng_ terminology \citep{barba_terminologies_2018}.} to production deployments.
 This article presents this _Rockerverse_ of projects across all development stages: early demonstrations, working prototypes, and mature products.
@@
 Most notably these capabilities are the MRAN repository, which is enabled by default, and the (optional) integration with [Intel® Math Kernel Library](https://software.intel.com/en-us/mkl) (MKL) for multi-threading in linear algebra operations \citep{microsoft_multithread}.
-MRO does not provide official Docker images, but a set of community-maintained `Dockerfile`s and Docker images `nuest/mro` are available on Docker Hub at [https://github.com/nuest/mro-docker](https://github.com/nuest/mro-docker) and on GitHub at [https://github.com/nuest/mro-docker](https://github.com/nuest/mro-docker) respectively.
+MRO does not provide official Docker images, but a set of community-maintained `Dockerfile`s and Docker images `nuest/mro` are available on Docker&nbsp;Hub at [https://github.com/nuest/mro-docker](https://github.com/nuest/mro-docker) and on GitHub at [https://github.com/nuest/mro-docker](https://github.com/nuest/mro-docker) respectively.
 The images are inspired by the Rocker images and can be used much in the same fashion, effectively a drop-in replacement allowing users to quickly evaluate if the benefits of MRO + Intel® MKL apply to their use case.
@@
 Not all R packages, especially one linking to binary libraries, are available, e.g., `rgdal` \footnote{\href{http://packages.renjin.org/package/org.renjin.cran/rgdal/1.4-4/build/1}{http://packages.renjin.org/package/org.renjin.cran/rgdal/1.4-4/build/1}}.
-There are no offical Docker images for Renjin, but community-maintained images for selected releases only are available under `nuest/renjin` on Docker Hub and GitHub at [https://hub.docker.com/r/nuest/renjin](https://hub.docker.com/r/nuest/renjin) and [https://github.com/nuest/renjin-docker](https://github.com/nuest/renjin-docker) respectively.
+There are no offical Docker images for Renjin, but community-maintained images for selected releases only are available under `nuest/renjin` on Docker&nbsp;Hub and GitHub at [https://hub.docker.com/r/nuest/renjin](https://hub.docker.com/r/nuest/renjin) and [https://github.com/nuest/renjin-docker](https://github.com/nuest/renjin-docker) respectively.
 These images expose the command line interface of Renjin in a similar fashion as Rocker images and allow an easy evaluation of Renjin's suitability, but are not intended for production use.
@@
 The source code development [on GitHub](https://github.com/radfordneal/pqR/) is a one man project and it does not provide any Docker images.
-But especially disruptive approaches may contribute to the development of the R ecosystem, so the `nuest/pqr` images on Docker Hub and GitHub at [https://hub.docker.com/r/nuest/pqr/](https://hub.docker.com/r/nuest/pqr/) and [https://github.com/nuest/pqr-docker](https://github.com/nuest/pqr-docker) respectively.
+But especially disruptive approaches may contribute to the development of the R ecosystem, so the `nuest/pqr` images on Docker&nbsp;Hub and GitHub at [https://hub.docker.com/r/nuest/pqr/](https://hub.docker.com/r/nuest/pqr/) and [https://github.com/nuest/pqr-docker](https://github.com/nuest/pqr-docker) respectively.
 
@@
 It is developed by Oracle, connects R to the GraalVM ecosystem \citep{wikipedia_graalvm_2019}, and also claims superior performance but also targets full compatibility with base R\footnote{\href{https://github.com/oracle/fastr}{https://github.com/oracle/fastr}}.
-There are no offical Docker images provided, but `nuest/fastr` images and `Dockerfile` are independently maintained on [Docker Hub](https://cloud.docker.com/repository/docker/nuest/fastr/) and [GitHub](https://github.com/nuest/fastr-docker) respectively.
+There are no offical Docker images provided, but `nuest/fastr` images and `Dockerfile` are independently maintained on [Docker&nbsp;Hub](https://cloud.docker.com/repository/docker/nuest/fastr/) and [GitHub](https://github.com/nuest/fastr-docker) respectively.
 
@@
 These base images match different Windows versions: Windows Server 2019, Windows Server 2016, and Windows Server, version 1803.
-The images are built using two CI services, [Travis CI](https://docs.travis-ci.com/user/reference/windows/) and [Appveyor](https://www.appveyor.com/docs/windows-images-software/), which provide the different Windows versions and therefore support different base images, and published automatically [on Docker Hub](https://hub.docker.com/r/nuest/rocker-win).
+The images are built using two CI services, [Travis CI](https://docs.travis-ci.com/user/reference/windows/) and [Appveyor](https://www.appveyor.com/docs/windows-images-software/), which provide the different Windows versions and therefore support different base images, and published automatically [on Docker&nbsp;Hub](https://hub.docker.com/r/nuest/rocker-win).
 The images are built from manually maintained seperate `Dockerfile`s, are tagged with both Windows-variant and R version, e.g., `nuest/rocker-win:ltsc2019-3.6.2` or `nuest/rocker-win:1803-latest`, and run the `R.exe` process by default.
@@
 The builder uses Docker containers to conduct these checks, taking advantages of their well-defined environments and sandboxing.
-The `Dockerfile`s and images are published on Docker Hub\footnote{\href{https://hub.docker.com/u/rhub}{https://hub.docker.com/u/rhub}} and GitHub\footnote{\href{https://github.com/r-hub/rhub-linux-builders}{https://github.com/r-hub/rhub-linux-builders}} respectively.
+The `Dockerfile`s and images are published on Docker&nbsp;Hub\footnote{\href{https://hub.docker.com/u/rhub}{https://hub.docker.com/u/rhub}} and GitHub\footnote{\href{https://github.com/r-hub/rhub-linux-builders}{https://github.com/r-hub/rhub-linux-builders}} respectively.
 The images comprise release and development builds of R running in the base images of the [Debian](https://www.debian.org/), [Ubuntu](https://ubuntu.com/), [Fedora](https://getfedora.org/), and [CentOS](https://centos.org/) linux distributions (Arch Linux is under development\footnote{\href{https://github.com/r-hub/rhub-linux-builders/pull/41}{https://github.com/r-hub/rhub-linux-builders/pull/41}}, see `rhub::local_check_linux_images()` for a full list), and are intended for debugging (see Section&nbsp;\nameref{development}) not for reproducible workflows.
@@
 It includes machine learling libraries such as Tensorflow and Keras, and also configures the \pkg{reticulate} package.
-The image uses a base image with _all packages from CRAN_, `gcr.io/kaggle-images/rcran`, which requires a Google Cloud Build as Docker Hub would time out\footnote{Originally, a stacked collection of over 20 images with automated builds on Docker Hub was used, see \href{https://web.archive.org/web/20190606043353/http://blog.kaggle.com/2016/02/05/how-to-get-started-with-data-science-in-containers/}{https://web.archive.org/web/20190606043353/http://blog.kaggle.com/2016/02/05/how-to-get-started-with-data-science-in-containers/} and \href{https://hub.docker.com/r/kaggle/rcran/dockerfile}{https://hub.docker.com/r/kaggle/rcran/dockerfile}}.
+The image uses a base image with _all packages from CRAN_, `gcr.io/kaggle-images/rcran`, which requires a Google Cloud Build as Docker&nbsp;Hub would time out\footnote{Originally, a stacked collection of over 20 images with automated builds on Docker~Hub was used, see \href{https://web.archive.org/web/20190606043353/http://blog.kaggle.com/2016/02/05/how-to-get-started-with-data-science-in-containers/}{https://web.archive.org/web/20190606043353/http://blog.kaggle.com/2016/02/05/how-to-get-started-with-data-science-in-containers/} and \href{https://hub.docker.com/r/kaggle/rcran/dockerfile}{https://hub.docker.com/r/kaggle/rcran/dockerfile}}.
 The final extracted image size is over 25GB, which makes it debatable if having everything available is actually convenient.
@@
 - [dynwrap](https://github.com/dynverse/dynwrap_containers/blob/master/.travis.yml) [@rcannood]
-  - For this project, we use travis-ci to build rocker-derived containers, test them, and only push them to docker hub (from travis-ci.org) if the integration tests succeed.
+  - For this project, we use travis-ci to build rocker-derived containers, test them, and only push them to Docker&nbsp;Hub (from travis-ci.org) if the integration tests succeed.
 - Google Cloud Build
@@
 - [dynmethods](https://github.com/dynverse/dynmethods) [@rcannood]: In order to evaluate ±50 computational methods which all used different environments (R, Python, C++, ...), we wrapped each of them in a docker container and can execute these methods from R.
-Again, all of these containers are being built on travis-ci, and will only be pushed to docker hub if the integration test succeeds.
+Again, all of these containers are being built on travis-ci, and will only be pushed to Docker&nbsp;Hub if the integration test succeeds.
 - \pkg{drake}, https://docs.ropensci.org/drake/index.html?q=docker#with-docker
@@
 - https://github.com/benmarwick/rrtools
-
-Containers provide a high degree of isolation that is often desirable when attempting to capture a specific computational environment so that others can reproduce and extend a research result.
-Many computationally intensive research projects depend on specific versions of original and third-party software packages in diverse languages, joined together to form a  pipeline through which data flows.
-New releases of even just a single piece of software in this pipeline can break the entire workflow, making it difficult for others to use, and making the original the results irreproducible.
-The chance of a substantial disruption like this is high in a multi-year research project where key pieces of third-party software may have several major updates over the duration of the project.
-Containers give the researcher an isolated environment to assemble these research pipelines with specific versions of software.
-Research workflows in containers are safe from contamination from other activities occuring on the researcher's computer, for example the installation of newest version of packages for teaching demonstrations.
-
-In many fields we see that a typical unit of research for a container is a research report or journal article, where the container holds the compendium, or self-contained set of data (or connections to data elsewhere) and code files needed to fully reproduce the article [@doi:10.1080/00031305.2017.1375986].
-Images of compendium containers can be hosted on services such as Docker Hub for convienient sharing among collaborators and others.
-Similarly, packages such as containerit and dockerfiler can be used to create a Dockerfile to be archived with a compendium on a data repository (e.g.
-Zenodo, Dataverse, Figshare, OSF).
-A typial compendium Dockerfile will pull a rocker image fixed to a specific version of R, and install R packages from the MRAN repository to ensure the package versions are tied to a specific date, rather than the most recent version.
-A future researchers can download then compendium from the repository and run the included Dockerfile build a new image that recreates the computational environment used to produce the original research results.
-
 - \pkg{renv}, https://rstudio.github.io/renv/articles/docker.html
@@
 \citet{ye_docker_2019} describes how they speed up the process of testing by running tasks on [Travis&nbsp;CI](https://travis-ci.org/) within a container using `docker exec`, e.g., the package check or rendering of documentation.
-\citet{cardozo_faster_2018}, also on Travis&nbsp;CI saved time by re-using the testing image as the base iamge for an image intended for publication on Docker Hub.
+\citet{cardozo_faster_2018}, also on Travis&nbsp;CI saved time by re-using the testing image as the base iamge for an image intended for publication on Docker&nbsp;Hub.
 Especially for long running tests or complex system dependencies, these approaches seems to improve the development.
@@
 The R packages and use cases presented show the positive drive in the community to innovate, but also the challenge of keeping up with a continuously evolving landscape.
-The use cases contributed by co-authors also have a degree of overlap.
-A common language and understanding of good practices is still taking shape.
-The easiness with which one can create a complex software systems, such as an independent Docker image stack, to serve one's specific needs currently leads to some fragmentation.
-It is unclear if Docker alone as a core technology is really a feasible level of collaboration and abstraction.
+The use cases contributed by co-authors also have a degree of overlap, which can be expected as a common language and understanding of good practices is still taking shape.
+Also, the ease with which one can create a complex software systems, such as an independent Docker image stack, to serve one's specific needs leads to some fragmentation.
 Furthermore fragmentation may not be a bad sign but instead a reflection of a growing market, which is able to sustain multiple related efforts.
 With the maturing of core building blocks, such as the Rocker suite of images, more systems will be built successfully but will also be behind the curtains.
-
-However, at least _on the level of R packages_, some consolidation seems in order, e.g., to reduce the number of packages creating `Dockerfile`s from R code or controlling the Docker daemon with R code.
+Docker alone, as a flexible core technology, is not a feasible level of collaboration and abstraction.
+Instead, the use cases and applications observed in this work provide a more useful division.
+
+Still, at least _on the level of R packages_ some consolidation seems in order, e.g., to reduce the number of packages creating `Dockerfile`s from R code or controlling the Docker daemon with R code.
 It remains to be seen which approach to control Docker, via the Docker API as `stevedore` or via system calls as `dockyard`/`docker`/`dockr`, is more sustainable, or if the question will be answered by the endurance of maintainers and sufficient funding.
@@
 New features, which make complex workflows accessible and reproducible, and variety in packages, even when they have overlapping features, are a signal and a support for a growing user base.
-This growth is possibly the the most important goal for the foreseeable future in the _Rockerverse_.
-The dominant position of Docker is a blessing and a curse for these goals.
+This growth is possibly the most important goal for the foreseeable future in the _Rockerverse_, and just like the Rocker images have matured over years of use and millions of runs the new ideas and prototypes will have to proof themselves.
+It should be noted that the dominant position of Docker is a blessing and a curse for these goals.
 It could be wise to start experimenting with non-Docker containerisation tools now, e.g.
@@
 If adoption of containerisation and R continue to grow, the missing pieces for a success predominantly lie in (a) coordination and documentation of activities to reduce repeated work in favour of open collaboration, (b) the sharing of lessons learned from use cases to build common knowledge and language, and (c) a sustainable continuation and funding for all of development, community support, and education.
-A concrete effort to work towards these pieces is to sustain the structure and captured status quo from this work in form of a CRAN Task View on containerization.
+A concrete effort to work towards these pieces is to sustain the structure and captured status quo from this work in form of a _CRAN Task View on containerization_.
 

### 2020-01-16 manuscript.Rmd
minor updates to links, fix footnote

@@
 
-**Microsoft R Open** ([MRO](https://github.com/nuest/mro-docker)) is an R distribution formerly known as Revolution R Open (RRO) before Revolution Analytics was acquired by Microsoft.
-MRO is compatible with main R and its packages.
-"It includes additional capabilities for improved performance, reproducibility, and platform support." \citep{microsoft_mro}
+**Microsoft R Open** (MRO) is an R distribution formerly known as Revolution R Open (RRO) before Revolution Analytics was acquired by Microsoft.
+MRO is compatible with main R and its packages and it "[..] includes additional capabilities for improved performance, reproducibility, and platform support." \citep{microsoft_mro}
 Most notably these capabilities are the MRAN repository, which is enabled by default, and the (optional) integration with [Intel® Math Kernel Library](https://software.intel.com/en-us/mkl) (MKL) for multi-threading in linear algebra operations \citep{microsoft_multithread}.
-MRO does not provide official Docker images, but a set of community-maintained `Dockerfile`s and Docker images `nuest/mro` are available on Docker&nbsp;Hub at [https://github.com/nuest/mro-docker](https://github.com/nuest/mro-docker) and on GitHub at [https://github.com/nuest/mro-docker](https://github.com/nuest/mro-docker) respectively.
+MRO does not provide official Docker images, but a set of community-maintained `Dockerfile`s and Docker images are provided by the poject `mro-docker` ([https://github.com/nuest/mro-docker](https://github.com/nuest/mro-docker)), e.g.
+the image [`nuest/mro`](https://hub.docker.com/repository/docker/nuest/mro/).
 The images are inspired by the Rocker images and can be used much in the same fashion, effectively a drop-in replacement allowing users to quickly evaluate if the benefits of MRO + Intel® MKL apply to their use case.
@@
 
-**[pqR](http://www.pqr-project.org/)** is a "a pretty quick version of R".
+**pqR** ([http://www.pqr-project.org/](http://www.pqr-project.org/)) is a "a pretty quick version of R".
 _pqR_ attempts to improve R on some opinionated issues in the R language and is the basis for experimental features, e.g., automatic diffentiation\footnote{\href{https://riotworkshop.github.io/abstracts/riot-2019-pqr.txt}{https://riotworkshop.github.io/abstracts/riot-2019-pqr.txt}}.
 The source code development [on GitHub](https://github.com/radfordneal/pqR/) is a one man project and it does not provide any Docker images.
-But especially disruptive approaches may contribute to the development of the R ecosystem, so the `nuest/pqr` images on Docker&nbsp;Hub and GitHub at [https://hub.docker.com/r/nuest/pqr/](https://hub.docker.com/r/nuest/pqr/) and [https://github.com/nuest/pqr-docker](https://github.com/nuest/pqr-docker) respectively.
-
-**FastR** is "A high-performance implementation of the R programming language, built on GraalVM" (https://github.com/oracle/fastr).
-It is developed by Oracle, connects R to the GraalVM ecosystem \citep{wikipedia_graalvm_2019}, and also claims superior performance but also targets full compatibility with base R\footnote{\href{https://github.com/oracle/fastr}{https://github.com/oracle/fastr}}.
-There are no offical Docker images provided, but `nuest/fastr` images and `Dockerfile` are independently maintained on [Docker&nbsp;Hub](https://cloud.docker.com/repository/docker/nuest/fastr/) and [GitHub](https://github.com/nuest/fastr-docker) respectively.
+But especially disruptive approaches may contribute to the development of the R ecosystem, so the `nuest/pqr` image was independently created, see Docker Hub at [https://hub.docker.com/r/nuest/pqr/](https://hub.docker.com/r/nuest/pqr/) and GitHub at [https://github.com/nuest/pqr-docker](https://github.com/nuest/pqr-docker).
+
+**FastR** ([https://github.com/oracle/fastr](https://github.com/oracle/fastr)) is "A high-performance implementation of the R programming language, built on GraalVM" \citep{oracle_labs_oraclefastr_2020}.
+It is developed by Oracle, connects R to the GraalVM ecosystem \citep{wikipedia_graalvm_2019}, and also claims superior performance but also targets full compatibility with base R \citep{oracle_labs_oraclefastr_2020}.
+There are no offical Docker images provided, but indepedently maintained experimental images and `Dockerfile`s are provided by [nuest/fastr-docker](https://github.com/nuest/fastr-docker), e.g.
+[`nuest/fastr` on Docker Hub](https://cloud.docker.com/repository/docker/nuest/fastr/).
 
@@
 **\pkg{renv}** ([https://rstudio.github.io/renv/](https://rstudio.github.io/renv/)) with Docker: https://rstudio.github.io/renv/articles/docker.html
+
+_version pinning_ with packages \pkg{remotes} and \pkg{versions} for R, and with system package managers for distros
 
@@
 
-## Using Docker in a commercial Data Science Platform
+## Using Docker in a commercial data science platform
 \label{rstudio}
@@
 
-## Processing [div]
+## Processing
 \label{processing}
@@
 For the [R-spatial community](https://www.r-spatial.org/), using containers is particularly useful, because of the strong integration of system libraries in core packages for geospatial data modelling and analysis.
-Purpose built Docker are used to prepare for upcoming releases of system libraries, individual bug reports, and for the lowest supported versions of system libraries \footnote{\href{https://github.com/r-spatial/sf/tree/master/inst/docker}{https://github.com/r-spatial/sf/tree/master/inst/docker},
-\href{https://github.com/Nowosad/rspatial_proj6}{https://github.com/Nowosad/rspatial_proj6}, \href{https://github.com/r-spatial/sf/issues/1231#issuecomment-570076440}{https://github.com/r-spatial/sf/issues/1231\#issuecomment-570076440}}.
+Purpose built Docker are used to prepare for upcoming releases of system libraries, individual bug reports, and for the lowest supported versions of system libraries\footnote{Cf.
+\href{https://github.com/r-spatial/sf/tree/master/inst/docker}{https://github.com/r-spatial/sf/tree/master/inst/docker}, \href{https://github.com/Nowosad/rspatial_proj6}{https://github.com/Nowosad/rspatial\_proj6}, and \href{https://github.com/r-spatial/sf/issues/1231}{https://github.com/r-spatial/sf/issues/1231}}.
 
@@
 
-New features, which make complex workflows accessible and reproducible, and variety in packages, even when they have overlapping features, are a signal and a support for a growing user base.
+New features, which make complex workflows accessible and reproducible, and the variety in packages connected with containerisation, even when they have overlapping features, are a signal and a support for a growing user base.
 This growth is possibly the most important goal for the foreseeable future in the _Rockerverse_, and just like the Rocker images have matured over years of use and millions of runs the new ideas and prototypes will have to proof themselves.

### 2020-01-16 manuscript.Rmd
once sentence per line

--- 2020-01-16 manuscript.Rmd
@@
 Combining both these topics, the _Rocker&nbsp;Project_ ([https://www.rocker-project.org/](https://www.rocker-project.org/)) provides images with R (see the next Section for more details).
-The considerable uptake and continued evolution of the Rocker Project has lead to numerous projects extending or building upon Rocker images, ranging from reproducible research\footnote{"Reproducible" in the sense of the _Claerbout/Donoho/Peng_ terminology \citep{barba_terminologies_2018}.} to production deployments.
+The considerable uptake and continued evolution of the Rocker Project has lead to numerous projects extending or building upon Rocker images, ranging from reproducible research to production deployments.
 This article presents this _Rockerverse_ of projects across all development stages: early demonstrations, working prototypes, and mature products.
@@
 Most notably these capabilities are the MRAN repository, which is enabled by default, and the (optional) integration with [Intel® Math Kernel Library](https://software.intel.com/en-us/mkl) (MKL) for multi-threading in linear algebra operations \citep{microsoft_multithread}.
-MRO does not provide official Docker images, but a set of community-maintained `Dockerfile`s and Docker images `nuest/mro` are available on Docker&nbsp;Hub at [https://github.com/nuest/mro-docker](https://github.com/nuest/mro-docker) and on GitHub at [https://github.com/nuest/mro-docker](https://github.com/nuest/mro-docker) respectively.
+MRO does not provide official Docker images, but a set of community-maintained `Dockerfile`s and Docker images `nuest/mro` are available on Docker Hub at [https://github.com/nuest/mro-docker](https://github.com/nuest/mro-docker) and on GitHub at [https://github.com/nuest/mro-docker](https://github.com/nuest/mro-docker) respectively.
 The images are inspired by the Rocker images and can be used much in the same fashion, effectively a drop-in replacement allowing users to quickly evaluate if the benefits of MRO + Intel® MKL apply to their use case.
@@
 Not all R packages, especially one linking to binary libraries, are available, e.g., `rgdal` \footnote{\href{http://packages.renjin.org/package/org.renjin.cran/rgdal/1.4-4/build/1}{http://packages.renjin.org/package/org.renjin.cran/rgdal/1.4-4/build/1}}.
-There are no offical Docker images for Renjin, but community-maintained images for selected releases only are available under `nuest/renjin` on Docker&nbsp;Hub and GitHub at [https://hub.docker.com/r/nuest/renjin](https://hub.docker.com/r/nuest/renjin) and [https://github.com/nuest/renjin-docker](https://github.com/nuest/renjin-docker) respectively.
+There are no offical Docker images for Renjin, but community-maintained images for selected releases only are available under `nuest/renjin` on Docker Hub and GitHub at [https://hub.docker.com/r/nuest/renjin](https://hub.docker.com/r/nuest/renjin) and [https://github.com/nuest/renjin-docker](https://github.com/nuest/renjin-docker) respectively.
 These images expose the command line interface of Renjin in a similar fashion as Rocker images and allow an easy evaluation of Renjin's suitability, but are not intended for production use.
@@
 The source code development [on GitHub](https://github.com/radfordneal/pqR/) is a one man project and it does not provide any Docker images.
-But especially disruptive approaches may contribute to the development of the R ecosystem, so the `nuest/pqr` images on Docker&nbsp;Hub and GitHub at [https://hub.docker.com/r/nuest/pqr/](https://hub.docker.com/r/nuest/pqr/) and [https://github.com/nuest/pqr-docker](https://github.com/nuest/pqr-docker) respectively.
+But especially disruptive approaches may contribute to the development of the R ecosystem, so the `nuest/pqr` images on Docker Hub and GitHub at [https://hub.docker.com/r/nuest/pqr/](https://hub.docker.com/r/nuest/pqr/) and [https://github.com/nuest/pqr-docker](https://github.com/nuest/pqr-docker) respectively.
 
@@
 It is developed by Oracle, connects R to the GraalVM ecosystem \citep{wikipedia_graalvm_2019}, and also claims superior performance but also targets full compatibility with base R\footnote{\href{https://github.com/oracle/fastr}{https://github.com/oracle/fastr}}.
-There are no offical Docker images provided, but `nuest/fastr` images and `Dockerfile` are independently maintained on [Docker&nbsp;Hub](https://cloud.docker.com/repository/docker/nuest/fastr/) and [GitHub](https://github.com/nuest/fastr-docker) respectively.
+There are no offical Docker images provided, but `nuest/fastr` images and `Dockerfile` are independently maintained on [Docker Hub](https://cloud.docker.com/repository/docker/nuest/fastr/) and [GitHub](https://github.com/nuest/fastr-docker) respectively.
 
@@
 These base images match different Windows versions: Windows Server 2019, Windows Server 2016, and Windows Server, version 1803.
-The images are built using two CI services, [Travis CI](https://docs.travis-ci.com/user/reference/windows/) and [Appveyor](https://www.appveyor.com/docs/windows-images-software/), which provide the different Windows versions and therefore support different base images, and published automatically [on Docker&nbsp;Hub](https://hub.docker.com/r/nuest/rocker-win).
+The images are built using two CI services, [Travis CI](https://docs.travis-ci.com/user/reference/windows/) and [Appveyor](https://www.appveyor.com/docs/windows-images-software/), which provide the different Windows versions and therefore support different base images, and published automatically [on Docker Hub](https://hub.docker.com/r/nuest/rocker-win).
 The images are built from manually maintained seperate `Dockerfile`s, are tagged with both Windows-variant and R version, e.g., `nuest/rocker-win:ltsc2019-3.6.2` or `nuest/rocker-win:1803-latest`, and run the `R.exe` process by default.
-All these images can be run on Docker for Desktop on Windows 10.
 New R versions are only added on demand.
-The explorative rocker-win project demonstrates the variety and pontentially confusing complexity of Windows on Docker, e.g., because of the different base images, but also provides a starting point for any Windows-constrained user to leverage containerisation, including examples for apps using \pkg{plumber\, \pkg{Shiny}, or R packages with system dependencies such as \pkg{sf}.
-The latter is quite comfortable and fast actually, compared to installation from source on Linux, as R ships pre-compiled binaries for Windows.
-However, compared to the various Linux images, the footprint of Windows images is quite large.
+The explorative project demonstrates the variety and pontentially confusing complexity of Windows on Docker, e.g., because of the different base images, but also provides a starting point for any Windows-constrained user to leverage containerisation, including examples for apps using \pkg{plumber\, \pkg{Shiny}, or R packages with system dependencies such as \pkg{sf}.
+The latter is quite comfortable actually, as R ships pre-compiled binaries for Windows.
 
@@
 In most cases, individual developers or researchers worrying about reproducibility, will prefer the more widely and host independent Linux-based containers.
-Naturally, a Windows license is required for the host machine, and the licensing impacts the potnetial to build images in clouds or redistribute exported images as archive files.
+Naturally, a Windows license is required for the host machine.
 But the developments in the `rocker-win` prototype show that containers can also be used for R workflows depending on Windows-only tools, for leveraging an existing Windows Server-based infrastructure where policies can otherwise not be met, or for streamlining interactions with system operating staff that runs Windows-based containers themselves.
@@
 The builder uses Docker containers to conduct these checks, taking advantages of their well-defined environments and sandboxing.
-The `Dockerfile`s and images are published on Docker&nbsp;Hub\footnote{\href{https://hub.docker.com/u/rhub}{https://hub.docker.com/u/rhub}} and GitHub\footnote{\href{https://github.com/r-hub/rhub-linux-builders}{https://github.com/r-hub/rhub-linux-builders}} respectively.
+The `Dockerfile`s and images are published on Docker Hub\footnote{\href{https://hub.docker.com/u/rhub}{https://hub.docker.com/u/rhub}} and GitHub\footnote{\href{https://github.com/r-hub/rhub-linux-builders}{https://github.com/r-hub/rhub-linux-builders}} respectively.
 The images comprise release and development builds of R running in the base images of the [Debian](https://www.debian.org/), [Ubuntu](https://ubuntu.com/), [Fedora](https://getfedora.org/), and [CentOS](https://centos.org/) linux distributions (Arch Linux is under development\footnote{\href{https://github.com/r-hub/rhub-linux-builders/pull/41}{https://github.com/r-hub/rhub-linux-builders/pull/41}}, see `rhub::local_check_linux_images()` for a full list), and are intended for debugging (see Section&nbsp;\nameref{development}) not for reproducible workflows.
@@
 It includes machine learling libraries such as Tensorflow and Keras, and also configures the \pkg{reticulate} package.
-The image uses a base image with _all packages from CRAN_, `gcr.io/kaggle-images/rcran`, which requires a Google Cloud Build as Docker&nbsp;Hub would time out\footnote{Originally, a stacked collection of over 20 images with automated builds on Docker~Hub was used, see \href{https://web.archive.org/web/20190606043353/http://blog.kaggle.com/2016/02/05/how-to-get-started-with-data-science-in-containers/}{https://web.archive.org/web/20190606043353/http://blog.kaggle.com/2016/02/05/how-to-get-started-with-data-science-in-containers/} and \href{https://hub.docker.com/r/kaggle/rcran/dockerfile}{https://hub.docker.com/r/kaggle/rcran/dockerfile}}.
+The image uses a base image with _all packages from CRAN_, `gcr.io/kaggle-images/rcran`, which requires a Google Cloud Build as Docker Hub would time out\footnote{Originally, a stacked collection of over 20 images with automated builds on Docker Hub was used, see \href{https://web.archive.org/web/20190606043353/http://blog.kaggle.com/2016/02/05/how-to-get-started-with-data-science-in-containers/}{https://web.archive.org/web/20190606043353/http://blog.kaggle.com/2016/02/05/how-to-get-started-with-data-science-in-containers/} and \href{https://hub.docker.com/r/kaggle/rcran/dockerfile}{https://hub.docker.com/r/kaggle/rcran/dockerfile}}.
 The final extracted image size is over 25GB, which makes it debatable if having everything available is actually convenient.
@@
 ```
-
-**[Gigantum](http://gigantum.com/)** also its own image stack ([https://github.com/gigantum/base-images](https://github.com/gigantum/base-images)).
-...
 
@@
 
-**\pkg{harbor}** ([https://github.com/wch/harbor](https://github.com/wch/harbor) is not actively maintained anymore, but should be honorably mentioned as the first R package for managing Docker images and containers.
-It uses the \pkg{sys} package to run system commands against the Docker CLI, both locally and through an ssh connection, and has convenience functions, e.g., for listing and removing containers/images and for accessing logs.
-The output of container executions are converted to appropriate R types.
-The Docker CLI, while evolving quickly and not with great concern for introducing breaking changes, basic functionality is unchanged for a long time so `harbor::docker_run(image = "hello-world")` works like a charm.
-
-```{r harbor, eval=FALSE, include=FALSE}
-remotes::install_github("wch/harbor")
-harbor::docker_run(image = "hello-world")
-```
-
 **\pkg{stevedore}** ([https://cran.r-project.org/package=stevedore](https://cran.r-project.org/package=stevedore)) 
@@
 -->
-currently the most powerful Docker client in R.
 
@@
 Users can, for example, execute a command inside a container, mount a volume or copy a file.
+
+**\pkg{harbor}** ([https://github.com/wch/harbor](https://github.com/wch/harbor)
+<!-- from the github description
+is an R package for controlling docker containers on local and remote hosts.
+-->
 
@@
 
-**\pkg{dockerfiler}** ([https://github.com/ColinFay/dockerfiler/](https://github.com/ColinFay/dockerfiler/)) ...
-
-**\pkg{containerit}** ([https://github.com/o2r-project/containerit/](https://github.com/o2r-project/containerit/)) ...
-
-**\pkg{dockr}** ([https://github.com/smaakage85/dockr](https://github.com/smaakage85/dockr)) ...
-see/cite http://smaakage85.netlify.com/2019/12/21/dockr-easy-containerization-for-r/
-<!-- from description on CRAN: Build a lightweight Docker r-base container image for your R package, in which the package itself is available.
-The Docker image seeks to mirror your R session as close as possible with respect to R specific dependencies.
-Both dependencies on CRAN R packages as well as local non-CRAN R packages will be included in the Docker container image.-->
-
-**\pkg{renv}** ([https://rstudio.github.io/renv/](https://rstudio.github.io/renv/)) with Docker: https://rstudio.github.io/renv/articles/docker.html
+- [`dockerfiler`](https://github.com/ColinFay/dockerfiler/)
+- [`containerit`](https://github.com/o2r-project/containerit/)
+- [`dockr`](https://github.com/smaakage85/dockr), see also http://smaakage85.netlify.com/2019/12/21/dockr-easy-containerization-for-r/
+- [`dockertest`](https://github.com/traitecoevo/dockertest/)
+- [`renv` with Docker](https://rstudio.github.io/renv/articles/docker.html)
+- https://github.com/gigantum/base-images
 
@@
 
+## Deployment to cloud services
+
+The cloud is the natural environment of containers, and becomes the go-to mechanism to expose applications written in R to users.
+
+- `RSelenium`
+- [`googleComputeEngineR`](https://cloudyr.github.io/googleComputeEngineR/) (function `gce_vm_template()`)
+  - To enable quick deployments of key R services such as RStudio and Shiny onto cloud virtual machines (VMs), this package utilises Dockerfiles to move the labour of setting up those services from the user to a premade Docker image.
+For example, by specifying the template `template="rstudio"` in `gce_vm_template()/gce_vm()` an up to date RStudio Service image is launched.
+Specifying `template="rstudio-gpu"` will launch an RStudio Server image with a GPU attached, etc.
+- [`analogsea`](https://github.com/sckott/analogsea) (digital ocean R client)
+- [`plumber`](https://www.rplumber.io/docs/hosting.html#docker)
+- [production deployment of neural networks](https://github.com/tmobile/r-tensorflow-api) [@jnolis, @nolistic]
+- Kubernetes - A popular platform for managing Docker containers is [Kubernetes](https://kubernetes.io/), which is used for a wide variety of containerized applications.
+It may be your organisation has a Kubernetes cluster already for other applications.
+Docker containers are used within Kubernetes clusters to hold native code, for which Kubernetes creates a framework around network connections and scaling of resources up and down.
+An introduction on using [Kubernetes with R is published at this blog post](https://code.markedmondson.me/r-on-kubernetes-serverless-shiny-r-apis-and-scheduled-scripts/).
+- [`AzureContainers`](https://cran.r-project.org/package=AzureContainers): Umbrella package for working with containers in Microsoft's Azure cloud.
+Provides interfaces to three Azure services: Container Registry, Container Instances (for running individual containers), and Kubernetes Services (for orchestrated deployments)
+
 ## Using R to power enterprise software in production environments
@@
 \label{deployment}
-
-The cloud is the natural environment of containers, and is the natural mechanism to deploy R server applications.
-
-- `RSelenium`
-- [`googleComputeEngineR`](https://cloudyr.github.io/googleComputeEngineR/) (function `gce_vm_template()`)
-  - To enable quick deployments of key R services such as RStudio and Shiny onto cloud virtual machines (VMs), this package utilises Dockerfiles to move the labour of setting up those services from the user to a premade Docker image.
-For example, by specifying the template `template="rstudio"` in `gce_vm_template()/gce_vm()` an up to date RStudio Service image is launched.
-Specifying `template="rstudio-gpu"` will launch an RStudio Server image with a GPU attached, etc.
-- [`analogsea`](https://github.com/sckott/analogsea) (digital ocean R client)
-- [`plumber`](https://www.rplumber.io/docs/hosting.html#docker)
-- [production deployment of neural networks](https://github.com/tmobile/r-tensorflow-api) [@jnolis, @nolistic]
-- Kubernetes - A popular platform for managing Docker containers is [Kubernetes](https://kubernetes.io/), which is used for a wide variety of containerized applications.
-It may be your organisation has a Kubernetes cluster already for other applications.
-Docker containers are used within Kubernetes clusters to hold native code, for which Kubernetes creates a framework around network connections and scaling of resources up and down.
-An introduction on using [Kubernetes with R is published at this blog post](https://code.markedmondson.me/r-on-kubernetes-serverless-shiny-r-apis-and-scheduled-scripts/).
-- [`AzureContainers`](https://cran.r-project.org/package=AzureContainers): Umbrella package for working with containers in Microsoft's Azure cloud.
-Provides interfaces to three Azure services: Container Registry, Container Instances (for running individual containers), and Kubernetes Services (for orchestrated deployments)
 
@@
 - [dynwrap](https://github.com/dynverse/dynwrap_containers/blob/master/.travis.yml) [@rcannood]
-  - For this project, we use travis-ci to build rocker-derived containers, test them, and only push them to Docker&nbsp;Hub (from travis-ci.org) if the integration tests succeed.
+  - For this project, we use travis-ci to build rocker-derived containers, test them, and only push them to docker hub (from travis-ci.org) if the integration tests succeed.
 - Google Cloud Build
@@
 Google Cloud Build works alongside Google Container Registry to allow you to build private and public Docker images, which allows you to build up your own dependency graphs for downstream applications.
-  - https://code.markedmondson.me/googleCloudRunner/articles/cloudbuild.html
 
@@
 Using **GPUs** (graphical processing units) as a specialised hardware from containerized common work environments is also possible and useful \citep{haydel_enhancing_2015}.
-GPUs are increasingly popular for compute-intensive machine learning (ML) tasks, e.g., deep artificial neural networks \citep{schmidhuber_deep_2015}.
-Though in this case, containers are not completely portable between hardware environments, but the software stack for ML with GPUs is so complex to set up that a ready-to-use container is helpful.
+GPUs are increasingly popular for compute-intensive machine learning tasks, e.g., deep artificial neural networks \citep{schmidhuber_deep_2015}.
+Though in this case, containers are not completely portable between hardware environments.
 Containers running GPU software require drivers and libraries specific to GPU models and versions, and containers require a specialized runtime to connect to the underlying GPU hardware.
 For NVIDIA GPUs, the [NVIDIA Container Toolkit](https://github.com/NVIDIA/nvidia-docker) includes a specialized runtime plugin for Docker and a set of base images with appropriate drivers and libraries.
-The Rocker project [has a repository](https://github.com/rocker-org/ml) with (beta) images based on these that include GPU-enabled versions of machine-learning R packages, e.g., `rocker/ml` and `rocker/tensorflow-gpu`.
+The Rocker project has (beta) images based on these that include GPU-enabled versions of machine-learning R packages: https://github.com/rocker-org/ml.
 
@@
 An R implementation is shown here at [cloudRunR](https://github.com/MarkEdmondson1234/cloudRunR) which uses Cloud Run to create a scalable R plumber API.
-  - How does this relate to https://code.markedmondson.me/googleCloudRunner/index.html ?`
 - `batchtools` \citep{Lang2017batchtools} can [schedule jobs with Docker Swarm](https://mllg.github.io/batchtools/reference/makeClusterFunctionsDocker.html)
@@
 - [dynmethods](https://github.com/dynverse/dynmethods) [@rcannood]: In order to evaluate ±50 computational methods which all used different environments (R, Python, C++, ...), we wrapped each of them in a docker container and can execute these methods from R.
-Again, all of these containers are being built on travis-ci, and will only be pushed to Docker&nbsp;Hub if the integration test succeeds.
+Again, all of these containers are being built on travis-ci, and will only be pushed to docker hub if the integration test succeeds.
 - \pkg{drake}, https://docs.ropensci.org/drake/index.html?q=docker#with-docker
@@
 ## Packaging programs for processing pipelines
-\label{packaging}
 
-Docker is perfectly suited to package and execute any software and data, and is a good candidate to build complex processing pipelines independent of the used programming language.
-Because of the original use case (see&nbsp;\nameref{introduction}), Docker does not have a standard mechanism to chain a number of containers together, i.e., to define how environment variables, volume mounts, or ports can be used to pass input parameters and data into a container and how to get results out.
+Docker is perfectly suited for packaging and executing any software and data, and is a good candidate for building complex processing pipelines that are independent from the host programming language.
+Due to its original use case (see&nbsp;\nameref{introduction}), Docker has no standard mechanisms for chaining containers together; it lacks definitions and protocols for environment variables, volume mounts and/or ports that could enable transfer of input (parameters and data) and output (results) to and from containers.
 Some packages, e.g.
-\pkg{containerit}, provide a Docker image that can be used similar as a CLI, but the usage is cumbersome\footnote{\href{https://o2r.info/containerit/articles/container.html}{https://o2r.info/containerit/articles/container.html}}.
+\pkg{containerit}, do provide Docker images that can be used similarly to CLIs, but their usage is cumbersome\footnote{\href{https://o2r.info/containerit/articles/container.html}{https://o2r.info/containerit/articles/container.html}}.
 
-**\pkg{outsider}** ([https://antonellilab.github.io/outsider/](https://antonellilab.github.io/outsider/)) tackles the problem of integrating external programmes into an R workflow \citep{bennett_outsider_2019}.
-Installation and usage of external programmes can be cumbersome and even prohibited if platform dependent.
+**\pkg{outsider}** ([https://antonellilab.github.io/outsider/](https://antonellilab.github.io/outsider/)) tackles the problem of integrating external programs into an R workflow \citep{bennett_outsider_2020}.
+Installation and usage of external programs can be difficult, convoluted and even impossible if the platform is incompatible. 
 Therefore \pkg{outsider} uses the platform-independent Docker images to encapsulate processes in _outsider modules_.
-Each outsider modules has a `Dockerfile` and an R package with functions for interacting with the encapsulated tool.
-A user only needs to use the given functions to install a module and can then integrate a tool into their own workflow seaminglessly using only R functions.
-The module takes care of transmitting arguments and sending and receiving files to and from a container.
-Outsider modules are hosted, e.g., on GitHub, and \pkg{outsider} contains discovery functions for them.
-<!-- illustrating outsider examples:
+Each outsider module has a `Dockerfile` and an R package with functions for interacting with the encapsulated tool.
+Using only R functions, an end-user can install a module with the \pkg{outsider} package and then call module code to integrate a tool into their own workflow seamlessly.
+The \pkg{outsider} package and module manage the containers and handle the transmission of arguments and the transfer of files to and from a container.
+These functionalities also allow a user to launch module code on a remote machine via SSH, expanding the potential computational scale.
+Outsider modules can be hosted code-sharing services, e.g., on GitHub, and \pkg{outsider} contains discovery functions for them.
+
+<!-- @nuest, would you like me to expand on these in a particular way?
 https://github.com/DomBennett/om..astral/blob/master/examples/astral.R
@@
 ## Research Compendia
-\label{compendia}
 
-- researchcompendia.science
 - https://github.com/benmarwick/rrtools
 - \pkg{renv}, https://rstudio.github.io/renv/articles/docker.html
-- what to copy in and what not
 
@@
 
-Containers can also serve as useful playgrounds to create environments ad-hoc or to provide very specific environments that are not needed or not easily available in day-to-day development.
-These environments may have specific versions of R, of R extension packages, and of system libraries used by R extension packages, and all of the above in a specific combination.
+Containers can also serve as useful playgrounds to create environments ad-hoc or to provide very specific environments that are not often needed.
+Developers can readily start a local container based on a Rocker image to investigate a bug report, which might require them to install a specific combination of package versions or use a specific R version (see also versioned images above).
+Unlike recreating the reporter's environment on their regular machine, a container can simply be discarded after use, and there is no need to tediously roll back explorative changes to a previous state.
+Using the Rocker images with RStudio, the disposable environments lack no development comfort (see also the usage in research compendia).
 
-First, such containers can greatly facilitate **fixing bugs**, because developers can readily start and later discard a container to investigate a bug report without affecting their regular system, and potentially having to revert changes back.
-Using the Rocker images with RStudio, these disposable environments lack no development comfort (cf. Section&nbsp;\nameref{compendia}).
-\citet{eddelbuettel_debugging_2019} describes an example how a Docker container was used to debug an issue with a package only occuring with a particular version of Fortran, and using tools which are not readily available on all platforms (e.g., not on macOS).
-For the [R-spatial community](https://www.r-spatial.org/), using containers is particularly useful, because of the strong integration of system libraries in core packages for geospatial data modelling and analysis.
-Purpose built Docker are used to prepare for upcoming releases of system libraries, individual bug reports, and for the lowest supported versions of system libraries \footnote{\href{https://github.com/r-spatial/sf/tree/master/inst/docker}{https://github.com/r-spatial/sf/tree/master/inst/docker},
-\href{https://github.com/Nowosad/rspatial_proj6}{https://github.com/Nowosad/rspatial_proj6}, \href{https://github.com/r-spatial/sf/issues/1231#issuecomment-570076440}{https://github.com/r-spatial/sf/issues/1231\#issuecomment-570076440}}.
+**[`r-debug`](https://github.com/wch/r-debug)** is a purpose built Docker image for debugging R memory problems.
+\citet{eddelbuettel_debugging_2019} describes how a Docker container was used to debug an issue with a package only occuring with a particular version of Fortran, and using tools which are not readily available on all platforms (e.g., not on macOS).
 
-There are also special images for identifying problems beyond the mere R code, such as **debugging R memory problems**.
-The images significantly reduce the barrier to follow complex steps for fixing memory allocation bugs \citep[cf. Section~4.3 in][]{core_writing_1999}.
-These problems are hard to debug and critical, both because when they occur they lead to fatal crashing processes.
-[`rocker/r-devel-san`](https://github.com/rocker-org/r-devel-san) and [`rocker/r-devel-ubsan-clang`](https://github.com/rocker-org/r-devel-san-clang) are Docker images have a particularly configured version of R to trace such problems with gcc and clang compilers, respectively \citep[cf.~\pkg{sanitizers} for examples,][]{eddelbuettel_sanitizers_2014}.
-The image [`wch/r-debug`](https://github.com/wch/r-debug) is a purpose built Docker image with _multiple_ instrumented builds of R, each with a different diagnostic utility activated.
-
-Second, containers are useful for **testing** R code during development.
-To submit a package to CRAN, an R package must work with the development version of R, which must be compiled locally.
-That can be a challenge for some users.
-The R-hub service (see Section&nbsp;\nameref{rhub}) makes it easy to ensure that no errors occur, but to fix errors a local setup is still often warranted, e.g.
-using the image `rocker/r-devel`, and to test packages with native code, which can make the process more complex \citep[cf.][]{eckert_building_2018}.
-The R-hub Docker images can also be used to debug problems locally using various combinations of Linux platforms, R versions, and compilers\footnote{See \href{https://r-hub.github.io/rhub/articles/local-debugging.html}{https://r-hub.github.io/rhub/articles/local-debugging.html} and \href{https://blog.r-hub.io/2019/04/25/r-devel-linux-x86-64-debian-clang/}{https://blog.r-hub.io/2019/04/25/r-devel-linux-x86-64-debian-clang/}}.
+The R-hub Docker images (see Section&nbsp;\nameref{rhub}) can also be used to debug problems locally using various combinations of Linux platforms, R versions, and compilers\footnote{See \href{https://r-hub.github.io/rhub/articles/local-debugging.html}{https://r-hub.github.io/rhub/articles/local-debugging.html} and \href{https://blog.r-hub.io/2019/04/25/r-devel-linux-x86-64-debian-clang/}{https://blog.r-hub.io/2019/04/25/r-devel-linux-x86-64-debian-clang/}.
 The images go beyond the configurations, or _flavours_, used by CRAN for checking packages\footnote{\href{https://cran.r-project.org/web/checks/check_flavors.html}{https://cran.r-project.org/web/checks/check\_flavors.html}}, e.g., with CentOS-based images, but lack a container for checking on Windows or OS X.
 The images greatly support package developers to provide support on operating systems they are not familiar.
-The package \pkg{dockertest} ([https://github.com/traitecoevo/dockertest/](https://github.com/traitecoevo/dockertest/)) is a proof of concept for automatically generating `Dockerfile`s and building images specifically to run tests\footnote{\pkg{dockertest} is not actively maintained, but mentioned still because of its interesting approach.}.
-These images are accompanied with a special launch script so the tested source code is not stored in the image but the currently checked in version from a local Git repository is cloned into the container at runtime.
-This approach clearly seperates test environment, test code, and current working copy of the code.
-
-Third, Docker images can be used **on CI platforms** to streamline the testing of packages.
-\citet{ye_docker_2019} describes how they speed up the process of testing by running tasks on [Travis&nbsp;CI](https://travis-ci.org/) within a container using `docker exec`, e.g., the package check or rendering of documentation.
-\citet{cardozo_faster_2018}, also on Travis&nbsp;CI saved time by re-using the testing image as the base iamge for an image intended for publication on Docker&nbsp;Hub.
-Especially for long running tests or complex system dependencies, these approaches seems to improve the development.
-Not due to a concern about time, but to control the environment used on a CI server, even this manuscript was rendered into a PDF and deployed to a GitHub-hosted website (see `.travis.yml` and `Dockerfile` in the manuscript repository).
-This gives on the one hand easy access after every update of the R Markdown source code, and on the other hand a second controlled environment making sure that the article renders successfully and correctly.
 
@@
 The R packages and use cases presented show the positive drive in the community to innovate, but also the challenge of keeping up with a continuously evolving landscape.
-The use cases contributed by co-authors also have a degree of overlap, which can be expected as a common language and understanding of good practices is still taking shape.
-Also, the ease with which one can create a complex software systems, such as an independent Docker image stack, to serve one's specific needs leads to some fragmentation.
+The use cases contributed by co-authors also have a degree of overlap.
+A common language and understanding of good practices is still taking shape.
+The easiness with which one can create a complex software systems, such as an independent Docker image stack, to serve one's specific needs currently leads to some fragmentation.
+It is unclear if Docker alone as a core technology is really a feasible level of collaboration and abstraction.
 Furthermore fragmentation may not be a bad sign but instead a reflection of a growing market, which is able to sustain multiple related efforts.
 With the maturing of core building blocks, such as the Rocker suite of images, more systems will be built successfully but will also be behind the curtains.
-Docker alone, as a flexible core technology, is not a feasible level of collaboration and abstraction.
-Instead, the use cases and applications observed in this work provide a more useful division.
 
-Still, at least _on the level of R packages_ some consolidation seems in order, e.g., to reduce the number of packages creating `Dockerfile`s from R code or controlling the Docker daemon with R code.
+However, at least _on the level of R packages_, some consolidation seems in order, e.g., to reduce the number of packages creating `Dockerfile`s from R code or controlling the Docker daemon with R code.
 It remains to be seen which approach to control Docker, via the Docker API as `stevedore` or via system calls as `dockyard`/`docker`/`dockr`, is more sustainable, or if the question will be answered by the endurance of maintainers and sufficient funding.
@@
 New features, which make complex workflows accessible and reproducible, and variety in packages, even when they have overlapping features, are a signal and a support for a growing user base.
-This growth is possibly the most important goal for the foreseeable future in the _Rockerverse_, and just like the Rocker images have matured over years of use and millions of runs the new ideas and prototypes will have to proof themselves.
-It should be noted that the dominant position of Docker is a blessing and a curse for these goals.
+This growth is possibly the the most important goal for the foreseeable future in the _Rockerverse_.
+The dominant position of Docker is a blessing and a curse for these goals.
 It could be wise to start experimenting with non-Docker containerisation tools now, e.g.
@@
 If adoption of containerisation and R continue to grow, the missing pieces for a success predominantly lie in (a) coordination and documentation of activities to reduce repeated work in favour of open collaboration, (b) the sharing of lessons learned from use cases to build common knowledge and language, and (c) a sustainable continuation and funding for all of development, community support, and education.
-A concrete effort to work towards these pieces is to sustain the structure and captured status quo from this work in form of a _CRAN Task View on containerization_.
+A concrete effort to work towards these pieces is to sustain the structure and captured status quo from this work in form of a CRAN Task View on containerization.
 

### 2020-01-16 manuscript.Rmd

--- 2020-01-16 manuscript.Rmd
@@
 
-Docker is perfectly suited to package and execute any software and data, and is a good candidate to build complex processing pipelines independent of the used programming language.
-Because of the original use case (see&nbsp;\nameref{introduction}), Docker does not have a standard mechanism to chain a number of containers together, i.e., to define how environment variables, volume mounts, or ports can be used to pass input parameters and data into a container and how to get results out.
+Docker is perfectly suited for packaging and executing any software and data, and is a good candidate for building complex processing pipelines that are independent from the host programming language.
+Due to its original use case (see&nbsp;\nameref{introduction}), Docker has no standard mechanisms for chaining containers together; it lacks definitions and protocols for environment variables, volume mounts and/or ports that could enable transfer of input (parameters and data) and output (results) to and from containers.
 Some packages, e.g.
-\pkg{containerit}, provide a Docker image that can be used similar as a CLI, but the usage is cumbersome\footnote{\href{https://o2r.info/containerit/articles/container.html}{https://o2r.info/containerit/articles/container.html}}.
-
-**\pkg{outsider}** ([https://antonellilab.github.io/outsider/](https://antonellilab.github.io/outsider/)) tackles the problem of integrating external programmes into an R workflow \citep{bennett_outsider_2019}.
-Installation and usage of external programmes can be cumbersome and even prohibited if platform dependent.
+\pkg{containerit}, do provide Docker images that can be used similarly to CLIs, but their usage is cumbersome\footnote{\href{https://o2r.info/containerit/articles/container.html}{https://o2r.info/containerit/articles/container.html}}.
+
+**\pkg{outsider}** ([https://antonellilab.github.io/outsider/](https://antonellilab.github.io/outsider/)) tackles the problem of integrating external programs into an R workflow \citep{bennett_outsider_2020}.
+Installation and usage of external programs can be difficult, convoluted and even impossible if the platform is incompatible. 
 Therefore \pkg{outsider} uses the platform-independent Docker images to encapsulate processes in _outsider modules_.
-Each outsider modules has a `Dockerfile` and an R package with functions for interacting with the encapsulated tool.
-A user only needs to use the given functions to install a module and can then integrate a tool into their own workflow seaminglessly using only R functions.
-The module takes care of transmitting arguments and sending and receiving files to and from a container.
-Outsider modules are hosted, e.g., on GitHub, and \pkg{outsider} contains discovery functions for them.
-<!-- illustrating outsider examples:
+Each outsider module has a `Dockerfile` and an R package with functions for interacting with the encapsulated tool.
+Using only R functions, an end-user can install a module with the \pkg{outsider} package and then call module code to integrate a tool into their own workflow seamlessly.
+The \pkg{outsider} package and module manage the containers and handle the transmission of arguments and the transfer of files to and from a container.
+These functionalities also allow a user to launch module code on a remote machine via SSH, expanding the potential computational scale.
+Outsider modules can be hosted code-sharing services, e.g., on GitHub, and \pkg{outsider} contains discovery functions for them.
+
+<!-- @nuest, would you like me to expand on these in a particular way?
 https://github.com/DomBennett/om..astral/blob/master/examples/astral.R

### 2020-01-16 manuscript.Rmd
add @DomBennett to author list

@@
   - name: Ben Marwick
+  - name: Dom Bennett
+    # https://github.com/DomBennett
+    affiliation: University of Gothenburg, Sweden
+    address:
+      - Department of Biological and Environmental Sciences
+      - Carl Skottsbergs gata 22B
+      - 413 19 Gothenburg
+      - \orcid{0000-0003-2722-1359}
+    email: dominic.john.bennett@gmail.com
 abstract: >
@@
 Outsider modules can be hosted code-sharing services, e.g., on GitHub, and \pkg{outsider} contains discovery functions for them.
-
-<!-- @nuest, would you like me to expand on these in a particular way?
+<!-- informative examples of how modules work, just as a reminder for @nuest
 https://github.com/DomBennett/om..astral/blob/master/examples/astral.R
@@
 LS \& NT wrote the section on Bioconductor.
+DB wrote the section on \pkg{outsider}.
 All authors approved the final version.

### 2020-01-16 manuscript.Rmd
add @benmarwick
update addresses

@@
       - Elm & Carlton Streets
-      - Buffalo, New York, 14263, USA
+      - Buffalo, NY, 14263, USA
       - \orcid{0000-0002-5910-4010}
@@
       - Elm & Carlton Streets
-      - Buffalo, New York, 14263, USA
+      - Buffalo, NY, 14263, USA
       - \orcid{0000-0002-0224-9817}
@@
       - 12920 Se 38th St.
-      - Bellevue, WA, 98006
+      - Bellevue, WA, 98006, USA
       #- \orcid{NA}
@@
     address:
-      - RStudio
       - 250 Northern Ave
-      - Boston, MA 02210
+      - Boston, MA 02210, USA
+      #- \orcid{}
     email: sean@rstudio.com
   - name: Ben Marwick
+    # https://github.com/benmarwick/
+    affiliation: University of Washington
+    address:
+      - Department of Anthropology
+      - Denny Hall 230, Spokane Ln
+      - Seattle, WA 98105, USA
+      - \orcid{0000-0001-7879-4531}
+    email: bmarwick@uw.edu
   - name: Dom Bennett
@@
       - Carl Skottsbergs gata 22B
-      - 413 19 Gothenburg
+      - 413 19 Gothenburg, Sweden
       - \orcid{0000-0003-2722-1359}
@@
 
-DN conceived of the presented idea and \href{https://github.com/nuest/rockerverse-paper/issues/3}{initialised the formation of the writing team}, wrote sections not mentioned below, and revised the full first draft.
+DN conceived of the presented idea and \href{https://github.com/nuest/rockerverse-paper/issues/3}{initialised the formation of the writing team}, wrote sections not mentioned below, and revised all sections.
+DE wrote the introduction and section about Rocker.
 CB ..
 RC and EH wrote the section on interfaces for Docker in R.
-DE wrote the introduction and section about Rocker.
 ME ..
 CF wrote the paragraphs about `r-online`, `dockerfiler`, `r-ci` and `r-db`.
-BW ..
+BW wrote the section on research compendia and made the project Binder-ready.
 KR ..
 NR ..
-NX wrote the section on liftr.
+NX wrote the section on \pkg{liftr}.
 LS \& NT wrote the section on Bioconductor.

### 2020-01-16 manuscript.Rmd
some housekeeping and reordering

@@
 abstract: >
-  The Rocker project provides widely-used Docker images for R across different application scenarios.
+  The Rocker&nbsp;Project provides widely-used Docker images for R across different application scenarios.
   This articles surveys downstream projects building upon Rocker and presents the current state of R packages for managing Docker images and controlling containers.
@@
 Combining both these topics, the _Rocker&nbsp;Project_ ([https://www.rocker-project.org/](https://www.rocker-project.org/)) provides images with R (see the next Section for more details).
-The considerable uptake and continued evolution of the Rocker Project has lead to numerous projects extending or building upon Rocker images, ranging from reproducible research\footnote{"Reproducible" in the sense of the _Claerbout/Donoho/Peng_ terminology \citep{barba_terminologies_2018}.} to production deployments.
+The considerable uptake and continued evolution of the Rocker&nbsp;Project has lead to numerous projects extending or building upon Rocker images, ranging from reproducible research\footnote{"Reproducible" in the sense of the _Claerbout/Donoho/Peng_ terminology \citep{barba_terminologies_2018}.} to production deployments.
 This article presents this _Rockerverse_ of projects across all development stages: early demonstrations, working prototypes, and mature products.
@@
 
-The article continues with a brief introduction of containerization basics and the Rocker Project, followed by descriptions of projects developing images with R installations.
+The article continues with a brief introduction of containerization basics and the Rocker&nbsp;Project, followed by descriptions of projects developing images with R installations.
 Then, we present applications, starting with the R packages specificly for interacting with Docker, second-level packages using containers mediately or only for specific features, up to complex use cases leveraging containers.
@@
 But Docker improved this compartmentalization by accessing the host system---generally Linux---through a much thinner and smaller shim than a full operating system emulation or virtualization.
-This containerization is also called operating-system-level virtualization \ref{wikipedia_contributors_os-level_2020}.
+This containerization is also called operating-system-level virtualization \citep{wikipedia_contributors_os-level_2020}.
 Typically a container runs one process, whereas virtualization may run whole operating systems at a larger footprint.
@@
 
-For statistical computing and analysis centered around R, the Rocker Project has provided a variety of Docker containers since its start in 2014 \citep{RJ-2017-065}.
-The Rocker Project provides several lines of containers spanning to from building blocks with `R-release` or `R-devel`, via containers with [RStudio&nbsp;Server](https://rstudio.com/products/rstudio/) and [Shiny&nbsp;Server](https://rstudio.com/products/shiny/shiny-server/), to domain-specific containers such as [`rocker/geospatial`](https://github.com/rocker-org/geospatial) \citep{rocker_geospatial_2019}.
+For statistical computing and analysis centered around R, the Rocker&nbsp;Project has provided a variety of Docker containers since its start in 2014 \citep{RJ-2017-065}.
+The Rocker&nbsp;Project provides several lines of containers spanning to from building blocks with `R-release` or `R-devel`, via containers with [RStudio&nbsp;Server](https://rstudio.com/products/rstudio/) and [Shiny&nbsp;Server](https://rstudio.com/products/shiny/shiny-server/), to domain-specific containers such as [`rocker/geospatial`](https://github.com/rocker-org/geospatial) \citep{rocker_geospatial_2019}.
 These containers form _image stacks_, building on top of each other for better maintainability (i.e. smaller `Dockerfile`s), composability, and to reduce build time.
 Also of note is a series of "versioned" containers which match the R release they contain with the _then-current_ set of packages via the MRAN Snapshot views of CRAN \citep{microsoft_cran_2019}.
-The Rocker Project's impact and importance was acknowledged by the Chan Zuckerberg Initiative's _Essential Open Source Software for Science_, who provide findung for the projects's sustainable maintenance, community growth, and targeting new hardware platforms including GPUs \citep{chan_zuckerberg_initiative_maintaining_2019}.
+The Rocker&nbsp;Project's impact and importance was acknowledged by the Chan Zuckerberg Initiative's _Essential Open Source Software for Science_, who provide findung for the projects's sustainable maintenance, community growth, and targeting new hardware platforms including GPUs \citep{chan_zuckerberg_initiative_maintaining_2019}.
 
@@
 The large number of extension packages provides access to an unrivaled variety of established and upcoming features.
-Nevertheless, special use cases and experimental projects exist to test approaches or provide features different to what _"base R"_ provides.
-These projects stem both form academia and industry.
-_Base R_, sometimes called `GNU-R`, is the R distribution maintained by the [R Core Team](https://www.r-project.org/contributors.html) and and provided via [CRAN](https://cran.r-project.org/).
+Nevertheless, special use cases and experimental projects in academia and industry exist to test new approaches or provide features different to what _GNU-R_, or Base R, provides.
+GNU-R is the R distribution maintained by the [R Core Team](https://www.r-project.org/contributors.html) and available via [CRAN](https://cran.r-project.org/).
+While the images presented in this section are far from being as vetted, stable, and widely used as any of the Rocker images, they demonstrate an important advantage of containerisation technology, namely the abiliy to transparently build portable stacks of open source software and make them easily accessible to others.
+All these distributions of R are published under GPL licenses.
+Since all of the R distributions claim better performance as a core motivation, a comparision based on Docker images, potentially leveraging the [resource restriction mechanisms](https://docs.docker.com/config/containers/resource_constraints/) of Docker to level the playing field, seems useful future work.
 
@@
 [`nuest/fastr` on Docker Hub](https://cloud.docker.com/repository/docker/nuest/fastr/).
-
-While the images presented in this section are far from being as vetted, stable, and widely used as any of the Rocker images, they demonstrate an important advantage of containerisation technology, namely the abiliy to transparently build portable stacks of open source software and make them easily accessible to users.
-All different distributions are published under GPL licenses.
-Since all of the different R distributions claim better performance as a core motivation, a comparision based on Docker images, potentially leveraging the [resource restriction mechanisms](https://docs.docker.com/config/containers/resource_constraints/) of Docker to level the playing field, seems useful future work.
 
@@
 In any of these cases, users need a method to interact with the containers, be it and IDE, or command-like access and tools such as SSH, which is usually not part of standard container recipes and must be added.
-The Rocker project provides containers pre-installed with the RStudio IDE.
+The Rocker&nbsp;Project provides containers pre-installed with the RStudio IDE.
 In cases where users store nontrivial amounts of data for their projects, data needs to persist beyond the life of the container.
@@
 For NVIDIA GPUs, the [NVIDIA Container Toolkit](https://github.com/NVIDIA/nvidia-docker) includes a specialized runtime plugin for Docker and a set of base images with appropriate drivers and libraries.
-The Rocker project [has a repository](https://github.com/rocker-org/ml) with (beta) images based on these that include GPU-enabled versions of machine-learning R packages, e.g., `rocker/ml` and `rocker/tensorflow-gpu`.
+The Rocker&nbsp;Project [has a repository](https://github.com/rocker-org/ml) with (beta) images based on these that include GPU-enabled versions of machine-learning R packages, e.g., `rocker/ml` and `rocker/tensorflow-gpu`.
 
@@
 `liftr` could generate the environment with `containerit`, which in turn may use `dockerfiler` for low level objects representing a `Dockerfile`.
-In this consolidation, the Rocker Project could play the role of a unifying and coordinating entity in the future.
+In this consolidation, the Rocker&nbsp;Project could play the role of a unifying and coordinating entity in the future.
 Though for the moment, the sign of the times points to more experimentation and feature growth, e.g. images for GPU-based computing and artificial intelligence.
@@
 
+```{r word_stats, eval=FALSE, include=FALSE}
+# devtools::install_github("benmarwick/wordcountaddin", type = "source", dependencies = TRUE)
+wordcountaddin::readability(here::here("manuscript.Rmd"))
+wordcountaddin::text_stats(here::here("manuscript.Rmd"))
+```
+
 \bibliography{RJreferences}

### 2020-01-16 manuscript.Rmd
small shortening of bioconductor section

@@
 The project consists of 1741 R software packages as of August 15th 2019, as well as packages containing annotation or experiment data.
-_Bioconductor_has a semiannual release cycle, each release is associated with a particular version of R.
+_Bioconductor_ has a semiannual release cycle, each release is associated with a particular version of R.
 Docker images allow availability of current and past versions of _Bioconductor_ for convenience and reproducibility.
-_Bioconductor_ 'base' docker images are built on top of `rocker/r-ver` and `rocker/rstudio`.
+_Bioconductor_'s 'base' docker images are built on top of `rocker/r-ver` and `rocker/rstudio` images.
 _Bioconductor_ installs packages based on the R version, and therefore uses `rocker/rstudio` and `rocker/r-ver` version tagging.
 _Bioconductor_ selects the desired version of R from Rocker, adds the BiocManager CRAN package for installing appropriate versions of _Bioconductor_ packages, and creates a _Bioconductor_ docker image with an informative tag (R\_version\_Bioc\_version).
-The images are summarized on the _Bioconductor_ web site (https://bioconductor.org/help/docker/), maintained on GitHub ([https://github.com/Bioconductor/bioc_docker](https://github.com/Bioconductor/bioc_docker)), and available to the community through [DockerHub](https://hub.docker.com/).
-Past and current combinations of R and _Bioconductor_ are therefore accessible via a specific docker tag.
+The images are summarized on the _Bioconductor_ web site (see [https://bioconductor.org/help/docker/](https://bioconductor.org/help/docker/)), created with `Dockerfile`s maintained on GitHub ([https://github.com/Bioconductor/bioconductor_docker](https://github.com/Bioconductor/bioconductor_docker)), and [distributed through DockerHub](https://hub.docker.com/u/bioconductor).
+Past and current combinations of R and _Bioconductor_ are therefore accessible via a specific image tag.
+A strength of this approach is that the responsibility for complex software configuration (including customized development) is shifted from the user to the experienced _Bioconductor_ core team.
+Images may be valuable in teaching, where participants pull pre-built images to avoid complicated configuration of their own computing environemnts.
 
@@
 `bioconductor_full` provides developers with a test environment like _Bioconductor_'s build system.
-
-Use of images suggests several interesting possibilities for the _Bioconductor_ project.
-Images may be valuable in teaching, where participants pull pre-built images to avoid complicated configuration of their own computing environemnts.
-An appeal of this over our current approach (providing Amazon Machine Instances for the duration of the course) is the utility of the image to participant after the course is over.
-`bioconductor_full` introduces a common system configuration, so it becomes increasingly sensible for _Bioconductor_ to distribute convenient _binary_ packages.
-Images also suggest approaches to more advanced computational models.
-For instance, we are exploring use of images for [Helm](https://helm.sh/)-orchestrated [Kubernetes](https://kubernetes.io/) clusters on the Google Cloud Platform.
-The user interacts with a manager image based on `bioconductor_full`, configured to perform map-reduce style computations via the BiocParallel package communicating with minimally-configured worker images.
-A strength of this approach is that the responsibility for complex software configuration (including customized development) is shifted from the user to the experienced _Bioconductor_ core team.
 

### 2020-01-16 manuscript.Rmd
add section about Gigantum image stack, add @davclark, extend conclusion
closes #29

Co-authored-by: Dav Clark <dav@gigantum.com>

@@
     email: dominic.john.bennett@gmail.com
+  - name: Dav Clark
+    # https://github.com/davclark
+    affiliation: Gigantum, Inc.
+    address:
+      - 1140 3rd Street NE
+      - Washington, D.C. 20002, USA
+      - \orcid{0000-0002-3982-4416}
+    email: dav@gigantum.com
 abstract: >
@@
 
-**[Gigantum](http://gigantum.com/)** also its own image stack ([https://github.com/gigantum/base-images](https://github.com/gigantum/base-images)).
-...
+**Gigantum** ([http://gigantum.com/](http://gigantum.com/) is a platform for open and decentralized data science with a focus on using automation and user-friendly tools for easy sharing of reproducible computational workflows.
+Gigantum builds on the locally running _Gigantum CLI_ and _Gigantum Desktop_ applications for development and execution of analyses, which can be stored and shared via _Gigantum Cloud_.
+The Gigantum apps are user friendly interfaces to a backend using Docker containers to package a Gigantum project ([https://github.com/gigantum/base-images](https://github.com/gigantum/base-images)).
+The available images include two with R based on Ubuntu Linux have the [`c2d4u` CRAN PPA](https://launchpad.net/~marutter/+archive/ubuntu/c2d4u3.5/) pre-configured for installation of binary R packages\footnote{\href{https://docs.gigantum.com/docs/using-r}{https://docs.gigantum.com/docs/using-r}}.
+<!-- mention lack of littler? -->
+The R images vary in the included authoring environment, i.e., Jupyter in `r-tidyverse` or both Jupyter \& RStudio in `rstudio-server`.
+The independent image stack can be traced back to the Gigantum environment and its features.
+The R images are based on Gigantum's `python3-minimal` image, originally to keep the existing front-end configuration, but also to provide a consistent Python-to-R interoperability.
+The `Dockerfile`s also use build args to specify bases, for example for different version of NVIDIA CUDA for GPU processing\footnote{See \href{https://github.com/gigantum/base-images/blob/master/_templates/python3-minimal-template/Dockerfile}{https://github.com/gigantum/base-images/blob/master/_templates/python3-minimal-template/Dockerfile} for the `Dockerfile` of `python3-minimal`.} so that appropriate GPU drivers can be enabled automatically when supported.
+Furthermore, Gigantum's focus lies on environment management via GUI and ensuring a smooth user interaction, e.g., with reliable and easy conflict detection and resulution.
+<!-- @DC: when a user interactively installs packages in Python, do you write that to the Dockerfile?
+UI-based environment management does not work for R packages, right?
+\href{https://docs.gigantum.com/docs/environment-management}{https://docs.gigantum.com/docs/environment-management} -->
+Because of these specifics, instructions from `rocker/r-ubuntu` could _not_ be readily re-used in this image stack (see Section&nbsp;\nameref{conclusion}).
+The image build and publication process is scripted with Python and JSON template configuration files, unlike Rocker which relies on more plain `Dockerfile`s.
+A minor reason in the inception of the images were also project constraints requiring a Rocker-incompatible licensing of the `Dockerfile`s, i.e., the MIT License.
 
@@
 The use cases contributed by co-authors also have a degree of overlap, which can be expected as a common language and understanding of good practices is still taking shape.
-Also, the ease with which one can create a complex software systems, such as an independent Docker image stack, to serve one's specific needs leads to some fragmentation.
-Furthermore fragmentation may not be a bad sign but instead a reflection of a growing market, which is able to sustain multiple related efforts.
+Also, the ease with which one can create a complex software systems with Docker, such as an independent Docker image stack, to serve one's specific needs leads to parallel developments.
+<!-- Tyler W./DC from Gigantum: ideas about ease-of-DIY together with difficulty of composition; shared starting points of Rocker -->
+This ease-of-DIY in combination with the difficulty to compose environments based on `Dockerfile`s alone is a further reason for **fragmentation**.
+Instructions can be outsourced into distributable scripts and then copied into the image during build, but that make `Dockerfile` hard to read and adds a layor of complexity.
+Despite the different image stacks presented here, the pervasiveness of Rocker can be traced back to its maintainers and the user community valueing collaboration and shared starting points over impulses to create individual solutions.
+Aside from that, fragmentation may not be a bad sign, but instead a reflection of a growing market, which is able to sustain multiple related efforts.
 With the maturing of core building blocks, such as the Rocker suite of images, more systems will be built successfully but will also be behind the curtains.
@@
 
-Still, at least _on the level of R packages_ some consolidation seems in order, e.g., to reduce the number of packages creating `Dockerfile`s from R code or controlling the Docker daemon with R code.
+Nonetheless, at least on the level of R packages some **consolidation** seems in order, e.g., to reduce the number of packages creating `Dockerfile`s from R code or controlling the Docker daemon with R code.
 It remains to be seen which approach to control Docker, via the Docker API as `stevedore` or via system calls as `dockyard`/`docker`/`dockr`, is more sustainable, or if the question will be answered by the endurance of maintainers and sufficient funding.
@@
 `liftr` could generate the environment with `containerit`, which in turn may use `dockerfiler` for low level objects representing a `Dockerfile`.
-In this consolidation, the Rocker&nbsp;Project could play the role of a unifying and coordinating entity in the future.
+In this consolidation, the Rocker&nbsp;Project could play the role of coordinating entity.
 Though for the moment, the sign of the times points to more experimentation and feature growth, e.g. images for GPU-based computing and artificial intelligence.
@@
 DB wrote the section on \pkg{outsider}.
+DC contributed content on Gigantum.
 All authors approved the final version.

### 2020-01-16 manuscript.Rmd
update author ordering and contributions

--- 2020-01-16 manuscript.Rmd
@@
     email: daniel.nuest@uni-muenster.de
+  - name: Dom Bennett
+    # https://github.com/DomBennett
+    affiliation: University of Gothenburg, Sweden
+    address:
+      - Department of Biological and Environmental Sciences
+      - Carl Skottsbergs gata 22B
+      - 413 19 Gothenburg, Sweden
+      - \orcid{0000-0003-2722-1359}
+    email: dominic.john.bennett@gmail.com
   - name: Robrecht Cannoodt
@@
     email: contact@colinfay.me
-  - name: Karthik Ram
-    # https://github.com/karthik
-    affiliation: Berkeley Institute for Data Science
-    address:
-      - University of California
-      - Berkeley, CA 94720, USA
-      - \orcid{0000-0002-0233-1757}
-    email: karthik.ram@berkeley.edu
-  - name: Noam Ross
-    # https://github.com/noamross
-    affiliation: EcoHealth Alliance
-    address:
-      - 460 W 34th St., Ste. 1701
-      - New York, NY 10001, USA
-      - \orcid{0000-0002-0233-1757}
-    email: ross@ecohealthalliance.org
-  - name: Nan Xiao
-    # https://github.com/nanxstats
-    affiliation: Seven Bridges Genomics
-    address:
-      - 529 Main St, Suite 6610
-      - Charlestown, MA 02129, USA
-      - \orcid{0000-0002-0250-5673}
-    email: me@nanx.me
-  - name: Lori Shepherd
-    # https://github.com/lshep
-    affiliation: Roswell Park Comprehensive Cancer Center
-    address:
-      - Elm & Carlton Streets
-      - Buffalo, NY, 14263, USA
-      - \orcid{0000-0002-5910-4010}
-    email: lori.shepherd@roswellpark.org
-  - name: Nitesh Turaga
-    # https://github.com/nturaga
-    affiliation: Roswell Park Comprehensive Cancer Center
-    address:
-      - Elm & Carlton Streets
-      - Buffalo, NY, 14263, USA
-      - \orcid{0000-0002-0224-9817}
-    email: nitesh.turaga@roswellpark.org
-  - name: Jacqueline Nolis
-    # https://github.com/jnolis/
-    affiliation: Nolis, LLC
-    address:
-      - Seattle, WA, USA
-      - \orcid{0000-0001-9354-6501}
-    email: jacqueline@nolisllc.com
-  - name: Heather Nolis
-    # https://github.com/nolistic/
-    affiliation: T-Mobile
-    address:
-      - 12920 Se 38th St.
-      - Bellevue, WA, 98006, USA
-      #- \orcid{NA}
-    email: heather.wensler1@t-mobile.com
-  - name: Hong Ooi
-    # https://github.com/Hong-Revo
-    affiliation: Microsoft
-    address:
-    - Level 5, 4 Freshwater Place
-    - Southbank, VIC 3006, Australia
-    email: hongooi@microsoft.com
   - name: Ellis Hughes
@@
     email: bmarwick@uw.edu
-  - name: Dom Bennett
-    # https://github.com/DomBennett
-    affiliation: University of Gothenburg, Sweden
-    address:
-      - Department of Biological and Environmental Sciences
-      - Carl Skottsbergs gata 22B
-      - 413 19 Gothenburg, Sweden
-      - \orcid{0000-0003-2722-1359}
-    email: dominic.john.bennett@gmail.com
+  - name: Heather Nolis
+    # https://github.com/nolistic/
+    affiliation: T-Mobile
+    address:
+      - 12920 Se 38th St.
+      - Bellevue, WA, 98006, USA
+      #- \orcid{NA}
+    email: heather.wensler1@t-mobile.com
+  - name: Jacqueline Nolis
+    # https://github.com/jnolis/
+    affiliation: Nolis, LLC
+    address:
+      - Seattle, WA, USA
+      - \orcid{0000-0001-9354-6501}
+    email: jacqueline@nolisllc.com
+  - name: Hong Ooi
+    # https://github.com/Hong-Revo
+    affiliation: Microsoft
+    address:
+    - Level 5, 4 Freshwater Place
+    - Southbank, VIC 3006, Australia
+    email: hongooi@microsoft.com
+  - name: Karthik Ram
+    # https://github.com/karthik
+    affiliation: Berkeley Institute for Data Science
+    address:
+      - University of California
+      - Berkeley, CA 94720, USA
+      - \orcid{0000-0002-0233-1757}
+    email: karthik.ram@berkeley.edu
+  - name: Noam Ross
+    # https://github.com/noamross
+    affiliation: EcoHealth Alliance
+    address:
+      - 460 W 34th St., Ste. 1701
+      - New York, NY 10001, USA
+      - \orcid{0000-0002-0233-1757}
+    email: ross@ecohealthalliance.org
+  - name: Lori Shepherd
+    # https://github.com/lshep
+    affiliation: Roswell Park Comprehensive Cancer Center
+    address:
+      - Elm & Carlton Streets
+      - Buffalo, NY, 14263, USA
+      - \orcid{0000-0002-5910-4010}
+    email: lori.shepherd@roswellpark.org
+  - name: Nitesh Turaga
+    # https://github.com/nturaga
+    affiliation: Roswell Park Comprehensive Cancer Center
+    address:
+      - Elm & Carlton Streets
+      - Buffalo, NY, 14263, USA
+      - \orcid{0000-0002-0224-9817}
+    email: nitesh.turaga@roswellpark.org
+  - name: Nan Xiao
+    # https://github.com/nanxstats
+    affiliation: Seven Bridges Genomics
+    address:
+      - 529 Main St, Suite 6610
+      - Charlestown, MA 02129, USA
+      - \orcid{0000-0002-0250-5673}
+    email: me@nanx.me
 abstract: >
@@
 
-DN conceived of the presented idea and \href{https://github.com/nuest/rockerverse-paper/issues/3}{initialised the formation of the writing team}, wrote sections not mentioned below, and revised all sections.
-DE wrote the introduction and section about Rocker.
-CB ..
-RC and EH wrote the section on interfaces for Docker in R.
-ME ..
-CF wrote the paragraphs about `r-online`, `dockerfiler`, `r-ci` and `r-db`.
-BW wrote the section on research compendia and made the project Binder-ready.
-KR ..
-NR ..
+_The ordering of authors following DN as first author is alphabetical._ DN conceived the article idea, \href{https://github.com/nuest/rockerverse-paper/issues/3}{initialised the formation of the writing team}, wrote sections not mentioned below, and revised all sections.
+DB wrote the section on \pkg{outsider}.
+DE wrote the introduction and the section about Containerization and Rocker.
+RC \& EH contributed to the section on interfaces for Docker in R.
+ME contributed to the section on processing and deployment to cloud services.
+CF wrote paragraphs about `r-online`, `dockerfiler`, `r-ci` and `r-db`.
+BM wrote the section on research compendia and made the project Binder-ready.
+HN \& JN co-wrote the section on enterprise production environments (\pkg{googleComputeEngineR}, Google Cloud Run, Kubernetes).
+KR wrote the section about \pkg{wholepunch}.
+NR wrote paragraphs about shared work environments and GPUs.
+LS \& NT wrote the section on Bioconductor.
 NX wrote the section on \pkg{liftr}.
-LS \& NT wrote the section on Bioconductor.
-DB wrote the section on \pkg{outsider}.
-All authors approved the final version.
-
-This articles was collaboratively written at \href{https://github.com/nuest/rockerverse-paper/}{https://github.com/nuest/rockerverse-paper/}.
+All authors approved the final version. This articles was collaboratively written at \href{https://github.com/nuest/rockerverse-paper/}{https://github.com/nuest/rockerverse-paper/}.
 The \href{https://github.com/nuest/rockerverse-paper/graphs/contributors}{contributors page}, \href{https://github.com/nuest/rockerverse-paper/commits/master}{commit history}, and \href{https://github.com/nuest/rockerverse-paper/issues/}{discussion issues} provide a detailed view on the respective contributions.

### 2020-01-16 manuscript.Rmd

--- 2020-01-16 manuscript.Rmd
@@
     email: robrecht@cannoodt.dev
+  - name: Dav Clark
+    # https://github.com/davclark
+    affiliation: Gigantum, Inc.
+    address:
+      - 1140 3rd Street NE
+      - Washington, D.C. 20002, USA
+      - \orcid{0000-0002-3982-4416}
+    email: dav@gigantum.com
   - name: Dirk Eddelbuettel
@@
     email: bmarwick@uw.edu
-  - name: Dom Bennett
-    # https://github.com/DomBennett
-    affiliation: University of Gothenburg, Sweden
-    address:
-      - Department of Biological and Environmental Sciences
-      - Carl Skottsbergs gata 22B
-      - 413 19 Gothenburg, Sweden
-      - \orcid{0000-0003-2722-1359}
-    email: dominic.john.bennett@gmail.com
-  - name: Dav Clark
-    # https://github.com/davclark
-    affiliation: Gigantum, Inc.
-    address:
-      - 1140 3rd Street NE
-      - Washington, D.C. 20002, USA
-      - \orcid{0000-0002-3982-4416}
-    email: dav@gigantum.com
+  - name: Heather Nolis
+    # https://github.com/nolistic/
+    affiliation: T-Mobile
+    address:
+      - 12920 Se 38th St.
+      - Bellevue, WA, 98006, USA
+      #- \orcid{NA}
+    email: heather.wensler1@t-mobile.com
+  - name: Jacqueline Nolis
+    # https://github.com/jnolis/
+    affiliation: Nolis, LLC
+    address:
+      - Seattle, WA, USA
+      - \orcid{0000-0001-9354-6501}
+    email: jacqueline@nolisllc.com
+  - name: Hong Ooi
+    # https://github.com/Hong-Revo
+    affiliation: Microsoft
+    address:
+    - Level 5, 4 Freshwater Place
+    - Southbank, VIC 3006, Australia
+    email: hongooi@microsoft.com
+  - name: Karthik Ram
+    # https://github.com/karthik
+    affiliation: Berkeley Institute for Data Science
+    address:
+      - University of California
+      - Berkeley, CA 94720, USA
+      - \orcid{0000-0002-0233-1757}
+    email: karthik.ram@berkeley.edu
+  - name: Noam Ross
+    # https://github.com/noamross
+    affiliation: EcoHealth Alliance
+    address:
+      - 460 W 34th St., Ste. 1701
+      - New York, NY 10001, USA
+      - \orcid{0000-0002-0233-1757}
+    email: ross@ecohealthalliance.org
+  - name: Lori Shepherd
+    # https://github.com/lshep
+    affiliation: Roswell Park Comprehensive Cancer Center
+    address:
+      - Elm & Carlton Streets
+      - Buffalo, NY, 14263, USA
+      - \orcid{0000-0002-5910-4010}
+    email: lori.shepherd@roswellpark.org
+  - name: Nitesh Turaga
+    # https://github.com/nturaga
+    affiliation: Roswell Park Comprehensive Cancer Center
+    address:
+      - Elm & Carlton Streets
+      - Buffalo, NY, 14263, USA
+      - \orcid{0000-0002-0224-9817}
+    email: nitesh.turaga@roswellpark.org
+  - name: Nan Xiao
+    # https://github.com/nanxstats
+    affiliation: Seven Bridges Genomics
+    address:
+      - 529 Main St, Suite 6610
+      - Charlestown, MA 02129, USA
+      - \orcid{0000-0002-0250-5673}
+    email: me@nanx.me
 abstract: >
@@
 
-DN conceived of the presented idea and \href{https://github.com/nuest/rockerverse-paper/issues/3}{initialised the formation of the writing team}, wrote sections not mentioned below, and revised all sections.
-DE wrote the introduction and section about Rocker.
-CB ..
-RC and EH wrote the section on interfaces for Docker in R.
-ME ..
-CF wrote the paragraphs about `r-online`, `dockerfiler`, `r-ci` and `r-db`.
-BW wrote the section on research compendia and made the project Binder-ready.
-KR ..
-NR ..
+_The ordering of authors following DN as first author is alphabetical._ DN conceived the article idea, \href{https://github.com/nuest/rockerverse-paper/issues/3}{initialised the formation of the writing team}, wrote sections not mentioned below, and revised all sections.
+DB wrote the section on \pkg{outsider}.
+DE wrote the introduction and the section about Containerization and Rocker.
+RC \& EH contributed to the section on interfaces for Docker in R.
+DC contributed content on Gigantum.
+ME contributed to the section on processing and deployment to cloud services.
+CF wrote paragraphs about `r-online`, `dockerfiler`, `r-ci` and `r-db`.
+BM wrote the section on research compendia and made the project Binder-ready.
+HN \& JN co-wrote the section on enterprise production environments (\pkg{googleComputeEngineR}, Google Cloud Run, Kubernetes).
+KR wrote the section about \pkg{wholepunch}.
+NR wrote paragraphs about shared work environments and GPUs.
+LS \& NT wrote the section on Bioconductor.
 NX wrote the section on \pkg{liftr}.
-LS \& NT wrote the section on Bioconductor.
-DB wrote the section on \pkg{outsider}.
-DC contributed content on Gigantum.
-All authors approved the final version.
-
-This articles was collaboratively written at \href{https://github.com/nuest/rockerverse-paper/}{https://github.com/nuest/rockerverse-paper/}.
+All authors approved the final version. This articles was collaboratively written at \href{https://github.com/nuest/rockerverse-paper/}{https://github.com/nuest/rockerverse-paper/}.
 The \href{https://github.com/nuest/rockerverse-paper/graphs/contributors}{contributors page}, \href{https://github.com/nuest/rockerverse-paper/commits/master}{commit history}, and \href{https://github.com/nuest/rockerverse-paper/issues/}{discussion issues} provide a detailed view on the respective contributions.

### 2020-01-16 manuscript.Rmd
formatting housekeeping

--- 2020-01-16 manuscript.Rmd
@@
 
-_The ordering of authors following DN as first author is alphabetical._ DN conceived the article idea, \href{https://github.com/nuest/rockerverse-paper/issues/3}{initialised the formation of the writing team}, wrote sections not mentioned below, and revised all sections.
+The ordering of authors following DN as first author is alphabetical.
+DN conceived the article idea, \href{https://github.com/nuest/rockerverse-paper/issues/3}{initialised the formation of the writing team}, wrote sections not mentioned below, and revised all sections.
 DB wrote the section on \pkg{outsider}.
@@
 ME contributed to the section on processing and deployment to cloud services.
-CF wrote paragraphs about `r-online`, `dockerfiler`, `r-ci` and `r-db`.
+CF wrote paragraphs about \pkg{r-online}, \pkg{dockerfiler}, \pkg{r-ci} and \pkg{r-db}.
 BM wrote the section on research compendia and made the project Binder-ready.

### 2020-01-16 manuscript.Rmd
update RCloud paragraph

@@
 
-**[RCloud](https://github.com/att/rcloud/tree/master/docker)** uses a `rocker/drd` base image for creation of collaborative data analysis and visualisation environments.
+**RCloud** ([https://rcloud.social](https://rcloud.social)) is a cloud-based platform for data analysis, visualisation and collaboration using R.
+It provides a `rocker/drd` base image for easy evaluation of the platform\footnote{\href{https://github.com/att/rcloud/tree/master/docker}{https://github.com/att/rcloud/tree/master/docker}}.
 

### 2020-01-16 manuscript.tex
update manuscript

@@
 \title{Containers and R---the Rockerverse and beyond}
-\author{by Daniel Nüst, Robrecht Cannoodt, Dirk Eddelbuettel, Mark Edmondson, Colin Fay, Karthik Ram, Noam Ross, Nan Xiao, Lori Shepherd, Nitesh Turaga, Jacqueline Nolis, Heather Nolis, Hong Ooi, Ellis Hughes, Sean Lopp, Ben Marwick}
+\author{by Daniel Nüst, Dom Bennett, Robrecht Cannoodt, Dirk Eddelbuettel, Mark Edmondson, Colin Fay, Ellis Hughes, Sean Lopp, Ben Marwick, Heather Nolis, Jacqueline Nolis, Hong Ooi, Karthik Ram, Noam Ross, Lori Shepherd, Nitesh Turaga, Nan Xiao}
 
@@
 \abstract{%
-The Rocker project provides widely-used Docker images for R across
+The Rocker~Project provides widely-used Docker images for R across
 different application scenarios. This articles surveys downstream
@@
 next Section for more details). The considerable uptake and continued
-evolution of the Rocker Project has lead to numerous projects extending
-or building upon Rocker images, ranging from reproducible research to
-production deployments. This article presents this \emph{Rockerverse} of
-projects across all development stages: early demonstrations, working
-prototypes, and mature products. We also introduce related activities
-connecting the R language and environment with other containerisation
-solutions. The main contributions is a coherent picture of the current
-lay of the land of using containers in, with, and for R.
+evolution of the Rocker~Project has lead to numerous projects extending
+or building upon Rocker images, ranging from reproducible
+research\textbackslash{}footnote\{``Reproducible'' in the sense of the
+\emph{Claerbout/Donoho/Peng} terminology
+\citep{barba_terminologies_2018}.\} to production deployments. This
+article presents this \emph{Rockerverse} of projects across all
+development stages: early demonstrations, working prototypes, and mature
+products. We also introduce related activities connecting the R language
+and environment with other containerisation solutions. The main
+contributions is a coherent picture of the current lay of the land of
+using containers in, with, and for R.
 
 The article continues with a brief introduction of containerization
-basics and the Rocker Project, followed by descriptions of projects
+basics and the Rocker~Project, followed by descriptions of projects
 developing images with R installations. Then, we present applications,
@@
 operating-system-level virtualization
-\ref{wikipedia_contributors_os-level_2020}. Typically a container runs
+\citep{wikipedia_contributors_os-level_2020}. Typically a container runs
 one process, whereas virtualization may run whole operating systems at a
@@
 
-For statistical computing and analysis centered around R, the Rocker
-Project has provided a variety of Docker containers since its start in
-2014 \citep{RJ-2017-065}. The Rocker Project provides several lines of
-containers spanning to from building blocks with \texttt{R-release} or
-\texttt{R-devel}, via containers with
+For statistical computing and analysis centered around R, the
+Rocker~Project has provided a variety of Docker containers since its
+start in 2014 \citep{RJ-2017-065}. The Rocker~Project provides several
+lines of containers spanning to from building blocks with
+\texttt{R-release} or \texttt{R-devel}, via containers with
 \href{https://rstudio.com/products/rstudio/}{RStudio~Server} and
@@
 via the MRAN Snapshot views of CRAN \citep{microsoft_cran_2019}. The
-Rocker Project's impact and importance was acknowledged by the Chan
+Rocker~Project's impact and importance was acknowledged by the Chan
 Zuckerberg Initiative's \emph{Essential Open Source Software for
@@
 variety of established and upcoming features. Nevertheless, special use
-cases and experimental projects exist to test approaches or provide
-features different to what \emph{``base R''} provides. These projects
-stem both form academia and industry. \emph{Base R}, sometimes called
-\texttt{GNU-R}, is the R distribution maintained by the
-\href{https://www.r-project.org/contributors.html}{R Core Team} and and
-provided via \href{https://cran.r-project.org/}{CRAN}.
-
-\textbf{Microsoft R Open}
-(\href{https://github.com/nuest/mro-docker}{MRO}) is an R distribution
-formerly known as Revolution R Open (RRO) before Revolution Analytics
-was acquired by Microsoft. MRO is compatible with main R and its
-packages. ``It includes additional capabilities for improved
-performance, reproducibility, and platform support.''
-\citep{microsoft_mro} Most notably these capabilities are the MRAN
-repository, which is enabled by default, and the (optional) integration
-with \href{https://software.intel.com/en-us/mkl}{Intel® Math Kernel
-Library} (MKL) for multi-threading in linear algebra operations
+cases and experimental projects in academia and industry exist to test
+new approaches or provide features different to what \emph{GNU-R}, or
+Base R, provides. GNU-R is the R distribution maintained by the
+\href{https://www.r-project.org/contributors.html}{R Core Team} and
+available via \href{https://cran.r-project.org/}{CRAN}. While the images
+presented in this section are far from being as vetted, stable, and
+widely used as any of the Rocker images, they demonstrate an important
+advantage of containerisation technology, namely the abiliy to
+transparently build portable stacks of open source software and make
+them easily accessible to others. All these distributions of R are
+published under GPL licenses. Since all of the R distributions claim
+better performance as a core motivation, a comparision based on Docker
+images, potentially leveraging the
+\href{https://docs.docker.com/config/containers/resource_constraints/}{resource
+restriction mechanisms} of Docker to level the playing field, seems
+useful future work.
+
+\textbf{Microsoft R Open} (MRO) is an R distribution formerly known as
+Revolution R Open (RRO) before Revolution Analytics was acquired by
+Microsoft. MRO is compatible with main R and its packages and it
+``{[}..{]} includes additional capabilities for improved performance,
+reproducibility, and platform support.'' \citep{microsoft_mro} Most
+notably these capabilities are the MRAN repository, which is enabled by
+default, and the (optional) integration with
+\href{https://software.intel.com/en-us/mkl}{Intel® Math Kernel Library}
+(MKL) for multi-threading in linear algebra operations
 \citep{microsoft_multithread}. MRO does not provide official Docker
 images, but a set of community-maintained \texttt{Dockerfile}s and
-Docker images \texttt{nuest/mro} are available on Docker Hub at
-\url{https://github.com/nuest/mro-docker} and on GitHub at
-\url{https://github.com/nuest/mro-docker} respectively. The images are
-inspired by the Rocker images and can be used much in the same fashion,
-effectively a drop-in replacement allowing users to quickly evaluate if
-the benefits of MRO + Intel® MKL apply to their use case. Version-tagged
-images are provided for the latest bugfix release of recent R versions.
-Extended license information about MKL is printed at every startup.
+Docker images are provided by the poject \texttt{mro-docker}
+(\url{https://github.com/nuest/mro-docker}), e.g.~the image
+\href{https://hub.docker.com/repository/docker/nuest/mro/}{\texttt{nuest/mro}}.
+The images are inspired by the Rocker images and can be used much in the
+same fashion, effectively a drop-in replacement allowing users to
+quickly evaluate if the benefits of MRO + Intel® MKL apply to their use
+case. Version-tagged images are provided for the latest bugfix release
+of recent R versions. Extended license information about MKL is printed
+at every startup.
 
@@
 images for selected releases only are available under
-\texttt{nuest/renjin} on Docker Hub and GitHub at
+\texttt{nuest/renjin} on Docker~Hub and GitHub at
 \url{https://hub.docker.com/r/nuest/renjin} and
@@
 
-\textbf{\href{http://www.pqr-project.org/}{pqR}} is a ``a pretty quick
+\textbf{pqR} (\url{http://www.pqr-project.org/}) is a ``a pretty quick
 version of R''. \emph{pqR} attempts to improve R on some opinionated
@@
 disruptive approaches may contribute to the development of the R
-ecosystem, so the \texttt{nuest/pqr} images on Docker Hub and GitHub at
-\url{https://hub.docker.com/r/nuest/pqr/} and
-\url{https://github.com/nuest/pqr-docker} respectively.
-
-\textbf{FastR} is ``A high-performance implementation of the R
-programming language, built on GraalVM''
-(\url{https://github.com/oracle/fastr}). It is developed by Oracle,
-connects R to the GraalVM ecosystem \citep{wikipedia_graalvm_2019}, and
-also claims superior performance but also targets full compatibility
-with base
-R\footnote{\href{https://github.com/oracle/fastr}{https://github.com/oracle/fastr}}.
-There are no offical Docker images provided, but \texttt{nuest/fastr}
-images and \texttt{Dockerfile} are independently maintained on
-\href{https://cloud.docker.com/repository/docker/nuest/fastr/}{Docker
-Hub} and \href{https://github.com/nuest/fastr-docker}{GitHub}
-respectively.
-
-While the images presented in this section are far from being as vetted,
-stable, and widely used as any of the Rocker images, they demonstrate an
-important advantage of containerisation technology, namely the abiliy to
-transparently build portable stacks of open source software and make
-them easily accessible to users. All different distributions are
-published under GPL licenses. Since all of the different R distributions
-claim better performance as a core motivation, a comparision based on
-Docker images, potentially leveraging the
-\href{https://docs.docker.com/config/containers/resource_constraints/}{resource
-restriction mechanisms} of Docker to level the playing field, seems
-useful future work.
+ecosystem, so the \texttt{nuest/pqr} image was independently created,
+see Docker Hub at \url{https://hub.docker.com/r/nuest/pqr/} and GitHub
+at \url{https://github.com/nuest/pqr-docker}.
+
+\textbf{FastR} (\url{https://github.com/oracle/fastr}) is ``A
+high-performance implementation of the R programming language, built on
+GraalVM'' \citep{oracle_labs_oraclefastr_2020}. It is developed by
+Oracle, connects R to the GraalVM ecosystem
+\citep{wikipedia_graalvm_2019}, and also claims superior performance but
+also targets full compatibility with base R
+\citep{oracle_labs_oraclefastr_2020}. There are no offical Docker images
+provided, but indepedently maintained experimental images and
+\texttt{Dockerfile}s are provided by
+\href{https://github.com/nuest/fastr-docker}{nuest/fastr-docker},
+e.g.~\href{https://cloud.docker.com/repository/docker/nuest/fastr/}{\texttt{nuest/fastr}
+on Docker Hub}.
 
@@
 software packages as of August 15th 2019, as well as packages containing
-annotation or experiment data. \_Bioconductor\_has a semiannual release
-cycle, each release is associated with a particular version of R. Docker
-images allow availability of current and past versions of
+annotation or experiment data. \emph{Bioconductor} has a semiannual
+release cycle, each release is associated with a particular version of
+R. Docker images allow availability of current and past versions of
 \emph{Bioconductor} for convenience and reproducibility.
-\emph{Bioconductor} `base' docker images are built on top of
-\texttt{rocker/r-ver} and \texttt{rocker/rstudio}. \emph{Bioconductor}
-installs packages based on the R version, and therefore uses
-\texttt{rocker/rstudio} and \texttt{rocker/r-ver} version tagging.
-\emph{Bioconductor} selects the desired version of R from Rocker, adds
-the BiocManager CRAN package for installing appropriate versions of
-\emph{Bioconductor} packages, and creates a \emph{Bioconductor} docker
-image with an informative tag (R\_version\_Bioc\_version). The images
-are summarized on the \emph{Bioconductor} web site
-(\url{https://bioconductor.org/help/docker/}), maintained on GitHub
-(\url{https://github.com/Bioconductor/bioc_docker}), and available to
-the community through \href{https://hub.docker.com/}{DockerHub}. Past
-and current combinations of R and \emph{Bioconductor} are therefore
-accessible via a specific docker tag.
+\emph{Bioconductor}'s `base' docker images are built on top of
+\texttt{rocker/r-ver} and \texttt{rocker/rstudio} images.
+\emph{Bioconductor} installs packages based on the R version, and
+therefore uses \texttt{rocker/rstudio} and \texttt{rocker/r-ver} version
+tagging. \emph{Bioconductor} selects the desired version of R from
+Rocker, adds the BiocManager CRAN package for installing appropriate
+versions of \emph{Bioconductor} packages, and creates a
+\emph{Bioconductor} docker image with an informative tag
+(R\_version\_Bioc\_version). The images are summarized on the
+\emph{Bioconductor} web site (see
+\url{https://bioconductor.org/help/docker/}), created with
+\texttt{Dockerfile}s maintained on GitHub
+(\url{https://github.com/Bioconductor/bioconductor_docker}), and
+\href{https://hub.docker.com/u/bioconductor}{distributed through
+DockerHub}. Past and current combinations of R and \emph{Bioconductor}
+are therefore accessible via a specific image tag. A strength of this
+approach is that the responsibility for complex software configuration
+(including customized development) is shifted from the user to the
+experienced \emph{Bioconductor} core team. Images may be valuable in
+teaching, where participants pull pre-built images to avoid complicated
+configuration of their own computing environemnts.
 
@@
 
-Use of images suggests several interesting possibilities for the
-\emph{Bioconductor} project. Images may be valuable in teaching, where
-participants pull pre-built images to avoid complicated configuration of
-their own computing environemnts. An appeal of this over our current
-approach (providing Amazon Machine Instances for the duration of the
-course) is the utility of the image to participant after the course is
-over. \texttt{bioconductor\_full} introduces a common system
-configuration, so it becomes increasingly sensible for
-\emph{Bioconductor} to distribute convenient \emph{binary} packages.
-Images also suggest approaches to more advanced computational models.
-For instance, we are exploring use of images for
-\href{https://helm.sh/}{Helm}-orchestrated
-\href{https://kubernetes.io/}{Kubernetes} clusters on the Google Cloud
-Platform. The user interacts with a manager image based on
-\texttt{bioconductor\_full}, configured to perform map-reduce style
-computations via the BiocParallel package communicating with
-minimally-configured worker images. A strength of this approach is that
-the responsibility for complex software configuration (including
-customized development) is shifted from the user to the experienced
-\emph{Bioconductor} core team.
-
 \hypertarget{images-for-historic-r-versions}{%
@@
 Docker CLI. The Docker CLI commands are forwarded to the virtual host
-transparently for the user. For current Windows Server (2016 and later)
-and Windows Desktop
+transparently for the user.
+
+For current Windows Server (2016 and later) and Windows Desktop
 (\href{https://www.docker.com/products/docker-desktop}{Docker Desktop}
@@
 different base images, and published automatically
-\href{https://hub.docker.com/r/nuest/rocker-win}{on Docker Hub}. The
+\href{https://hub.docker.com/r/nuest/rocker-win}{on Docker~Hub}. The
 images are built from manually maintained seperate \texttt{Dockerfile}s,
@@
 \texttt{nuest/rocker-win:1803-latest}, and run the \texttt{R.exe}
-process by default. New R versions are only added on demand. The
-explorative project demonstrates the variety and pontentially confusing
+process by default. All these images can be run on Docker for Desktop on
+Windows 10. New R versions are only added on demand. The explorative
+rocker-win project demonstrates the variety and pontentially confusing
 complexity of Windows on Docker, e.g., because of the different base
@@
 \textbackslash{}pkg\{plumber, \pkg{Shiny}, or R packages with system
-dependencies such as \pkg{sf}. The latter is quite comfortable actually,
-as R ships pre-compiled binaries for Windows.
+dependencies such as \pkg{sf}. The latter is quite comfortable and fast
+actually, compared to installation from source on Linux, as R ships
+pre-compiled binaries for Windows. However, compared to the various
+Linux images, the footprint of Windows images is quite large.
 
@@
 Linux-based containers. Naturally, a Windows license is required for the
-host machine. But the developments in the \texttt{rocker-win} prototype
-show that containers can also be used for R workflows depending on
-Windows-only tools, for leveraging an existing Windows Server-based
-infrastructure where policies can otherwise not be met, or for
-streamlining interactions with system operating staff that runs
-Windows-based containers themselves.
+host machine, and the licensing impacts the potnetial to build images in
+clouds or redistribute exported images as archive files. But the
+developments in the \texttt{rocker-win} prototype show that containers
+can also be used for R workflows depending on Windows-only tools, for
+leveraging an existing Windows Server-based infrastructure where
+policies can otherwise not be met, or for streamlining interactions with
+system operating staff that runs Windows-based containers themselves.
 
@@
 these checks, taking advantages of their well-defined environments and
-sandboxing. The \texttt{Dockerfile}s and images are published on Docker
-Hub\footnote{\href{https://hub.docker.com/u/rhub}{https://hub.docker.com/u/rhub}}
+sandboxing. The \texttt{Dockerfile}s and images are published on
+Docker~Hub\footnote{\href{https://hub.docker.com/u/rhub}{https://hub.docker.com/u/rhub}}
 and
@@
 packages from CRAN}, \texttt{gcr.io/kaggle-images/rcran}, which requires
-a Google Cloud Build as Docker Hub would time
-out\footnote{Originally, a stacked collection of over 20 images with automated builds on Docker Hub was used, see \href{https://web.archive.org/web/20190606043353/http://blog.kaggle.com/2016/02/05/how-to-get-started-with-data-science-in-containers/}{https://web.archive.org/web/20190606043353/http://blog.kaggle.com/2016/02/05/how-to-get-started-with-data-science-in-containers/} and \href{https://hub.docker.com/r/kaggle/rcran/dockerfile}{https://hub.docker.com/r/kaggle/rcran/dockerfile}}.
+a Google Cloud Build as Docker~Hub would time
+out\footnote{Originally, a stacked collection of over 20 images with automated builds on Docker~Hub was used, see \href{https://web.archive.org/web/20190606043353/http://blog.kaggle.com/2016/02/05/how-to-get-started-with-data-science-in-containers/}{https://web.archive.org/web/20190606043353/http://blog.kaggle.com/2016/02/05/how-to-get-started-with-data-science-in-containers/} and \href{https://hub.docker.com/r/kaggle/rcran/dockerfile}{https://hub.docker.com/r/kaggle/rcran/dockerfile}}.
 The final extracted image size is over 25GB, which makes it debatable if
@@
 
-\begin{itemize}
-\tightlist
-\item
-  \href{https://rstudio.github.io/renv/articles/docker.html}{\texttt{renv}
-  with Docker}
-\end{itemize}
+\textbf{\pkg{renv}} (\url{https://rstudio.github.io/renv/}) with Docker:
+\url{https://rstudio.github.io/renv/articles/docker.html}
+
+\emph{version pinning} with packages \pkg{remotes} and \pkg{versions}
+for R, and with system package managers for distros
 
@@
     For this project, we use travis-ci to build rocker-derived
-    containers, test them, and only push them to docker hub (from
+    containers, test them, and only push them to Docker~Hub (from
     travis-ci.org) if the integration tests succeed.
@@
 and IDE, or command-like access and tools such as SSH, which is usually
-not part of standard container recipes and must be added. The Rocker
-project provides containers pre-installed with the RStudio IDE. In cases
-where users store nontrivial amounts of data for their projects, data
-needs to persist beyond the life of the container. This may be via in
-shared disks, attached network volumes, or in separate storage where it
-is uploaded between sessions. In the case of shared disks or
-network-attached volumes, care must be taken to persist user
+not part of standard container recipes and must be added. The
+Rocker~Project provides containers pre-installed with the RStudio IDE.
+In cases where users store nontrivial amounts of data for their
+projects, data needs to persist beyond the life of the container. This
+may be via in shared disks, attached network volumes, or in separate
+storage where it is uploaded between sessions. In the case of shared
+disks or network-attached volumes, care must be taken to persist user
 permissions, and of course backups are still neccessary. When working
@@
 includes a specialized runtime plugin for Docker and a set of base
-images with appropriate drivers and libraries. The Rocker project
+images with appropriate drivers and libraries. The Rocker~Project
 \href{https://github.com/rocker-org/ml}{has a repository} with (beta)
@@
 
-\textbf{\href{https://github.com/att/rcloud/tree/master/docker}{RCloud}}
-uses a \texttt{rocker/drd} base image for creation of collaborative data
-analysis and visualisation environments.
+\textbf{RCloud} (\url{https://rcloud.social}) is a cloud-based platform
+for data analysis, visualisation and collaboration using R. It provides
+a \texttt{rocker/drd} base image for easy evaluation of the
+platform\footnote{\href{https://github.com/att/rcloud/tree/master/docker}{https://github.com/att/rcloud/tree/master/docker}}.
 
 \hypertarget{using-docker-in-a-commercial-data-science-platform}{%
-\subsection{Using Docker in a commercial Data Science
-Platform}\label{using-docker-in-a-commercial-data-science-platform}}
+\subsection{Using Docker in a commercial data science
+platform}\label{using-docker-in-a-commercial-data-science-platform}}
 
@@
 
-\hypertarget{processing-div}{%
-\subsection{Processing {[}div{]}}\label{processing-div}}
+\hypertarget{processing}{%
+\subsection{Processing}\label{processing}}
 
@@
   R. Again, all of these containers are being built on travis-ci, and
-  will only be pushed to docker hub if the integration test succeeds.
+  will only be pushed to Docker~Hub if the integration test succeeds.
 \item
@@
 
-Docker is perfectly suited to package and execute any software and data,
-and is a good candidate to build complex processing pipelines
-independent of the used programming language. Because of the original
-use case (see~\nameref{introduction}), Docker does not have a standard
-mechanism to chain a number of containers together, i.e., to define how
-environment variables, volume mounts, or ports can be used to pass input
-parameters and data into a container and how to get results out. Some
-packages, e.g.~\pkg{containerit}, provide a Docker image that can be
-used similar as a CLI, but the usage is
+Docker is perfectly suited for packaging and executing any software and
+data, and is a good candidate for building complex processing pipelines
+that are independent from the host programming language. Due to its
+original use case (see~\nameref{introduction}), Docker has no standard
+mechanisms for chaining containers together; it lacks definitions and
+protocols for environment variables, volume mounts and/or ports that
+could enable transfer of input (parameters and data) and output
+(results) to and from containers. Some packages, e.g.~\pkg{containerit},
+do provide Docker images that can be used similarly to CLIs, but their
+usage is
 cumbersome\footnote{\href{https://o2r.info/containerit/articles/container.html}{https://o2r.info/containerit/articles/container.html}}.
@@
 \textbf{\pkg{outsider}} (\url{https://antonellilab.github.io/outsider/})
-tackles the problem of integrating external programmes into an R
-workflow \citep{bennett_outsider_2019}. Installation and usage of
-external programmes can be cumbersome and even prohibited if platform
-dependent. Therefore \pkg{outsider} uses the platform-independent Docker
-images to encapsulate processes in \emph{outsider modules}. Each
-outsider modules has a \texttt{Dockerfile} and an R package with
-functions for interacting with the encapsulated tool. A user only needs
-to use the given functions to install a module and can then integrate a
-tool into their own workflow seaminglessly using only R functions. The
-module takes care of transmitting arguments and sending and receiving
-files to and from a container. Outsider modules are hosted, e.g., on
-GitHub, and \pkg{outsider} contains discovery functions for them.
+tackles the problem of integrating external programs into an R workflow
+\citep{bennett_outsider_2020}. Installation and usage of external
+programs can be difficult, convoluted and even impossible if the
+platform is incompatible. Therefore \pkg{outsider} uses the
+platform-independent Docker images to encapsulate processes in
+\emph{outsider modules}. Each outsider module has a \texttt{Dockerfile}
+and an R package with functions for interacting with the encapsulated
+tool. Using only R functions, an end-user can install a module with the
+\pkg{outsider} package and then call module code to integrate a tool
+into their own workflow seamlessly. The \pkg{outsider} package and
+module manage the containers and handle the transmission of arguments
+and the transfer of files to and from a container. These functionalities
+also allow a user to launch module code on a remote machine via SSH,
+expanding the potential computational scale. Outsider modules can be
+hosted code-sharing services, e.g., on GitHub, and \pkg{outsider}
+contains discovery functions for them.
 
@@
 for upcoming releases of system libraries, individual bug reports, and
-for the lowest supported versions of system libraries
-\textbackslash{}footnote\{\href{https://github.com/r-spatial/sf/tree/master/inst/docker}{https://github.com/r-spatial/sf/tree/master/inst/docker},
-\textbackslash{}href\{\url{https://github.com/Nowosad/rspatial_proj6}\}\{\url{https://github.com/Nowosad/rspatial_proj6}\},
-\href{https://github.com/r-spatial/sf/issues/1231#issuecomment-570076440}{https://github.com/r-spatial/sf/issues/1231\#issuecomment-570076440}\}.
+for the lowest supported versions of system
+libraries\footnote{Cf.
+\href{https://github.com/r-spatial/sf/tree/master/inst/docker}{https://github.com/r-spatial/sf/tree/master/inst/docker}, \href{https://github.com/Nowosad/rspatial_proj6}{https://github.com/Nowosad/rspatial\_proj6}, and \href{https://github.com/r-spatial/sf/issues/1231}{https://github.com/r-spatial/sf/issues/1231}}.
 
@@
 by re-using the testing image as the base iamge for an image intended
-for publication on Docker Hub. Especially for long running tests or
+for publication on Docker~Hub. Especially for long running tests or
 complex system dependencies, these approaches seems to improve the
@@
 continuously evolving landscape. The use cases contributed by co-authors
-also have a degree of overlap. A common language and understanding of
-good practices is still taking shape. The easiness with which one can
-create a complex software systems, such as an independent Docker image
-stack, to serve one's specific needs currently leads to some
-fragmentation. It is unclear if Docker alone as a core technology is
-really a feasible level of collaboration and abstraction. Furthermore
-fragmentation may not be a bad sign but instead a reflection of a
-growing market, which is able to sustain multiple related efforts. With
-the maturing of core building blocks, such as the Rocker suite of
-images, more systems will be built successfully but will also be behind
-the curtains.
-
-However, at least \emph{on the level of R packages}, some consolidation
+also have a degree of overlap, which can be expected as a common
+language and understanding of good practices is still taking shape.
+Also, the ease with which one can create a complex software systems,
+such as an independent Docker image stack, to serve one's specific needs
+leads to some fragmentation. Furthermore fragmentation may not be a bad
+sign but instead a reflection of a growing market, which is able to
+sustain multiple related efforts. With the maturing of core building
+blocks, such as the Rocker suite of images, more systems will be built
+successfully but will also be behind the curtains. Docker alone, as a
+flexible core technology, is not a feasible level of collaboration and
+abstraction. Instead, the use cases and applications observed in this
+work provide a more useful division.
+
+Still, at least \emph{on the level of R packages} some consolidation
 seems in order, e.g., to reduce the number of packages creating
@@
 level objects representing a \texttt{Dockerfile}. In this consolidation,
-the Rocker Project could play the role of a unifying and coordinating
+the Rocker~Project could play the role of a unifying and coordinating
 entity in the future. Though for the moment, the sign of the times
@@
 New features, which make complex workflows accessible and reproducible,
-and variety in packages, even when they have overlapping features, are a
-signal and a support for a growing user base. This growth is possibly
-the the most important goal for the foreseeable future in the
-\emph{Rockerverse}. The dominant position of Docker is a blessing and a
-curse for these goals. It could be wise to start experimenting with
-non-Docker containerisation tools now, e.g.~R packages interfacing with
-other container engines such as
+and the variety in packages connected with containerisation, even when
+they have overlapping features, are a signal and a support for a growing
+user base. This growth is possibly the most important goal for the
+foreseeable future in the \emph{Rockerverse}, and just like the Rocker
+images have matured over years of use and millions of runs the new ideas
+and prototypes will have to proof themselves. It should be noted that
+the dominant position of Docker is a blessing and a curse for these
+goals. It could be wise to start experimenting with non-Docker
+containerisation tools now, e.g.~R packages interfacing with other
+container engines such as
 \href{https://github.com/containers/libpod}{podman/buildah} or
@@
 towards these pieces is to sustain the structure and captured status quo
-from this work in form of a CRAN Task View on containerization.
+from this work in form of a \emph{CRAN Task View on containerization}.
 
@@
 
-DN conceived of the presented idea and
+The ordering of authors following DN as first author is alphabetical. DN
+conceived the article idea,
 \href{https://github.com/nuest/rockerverse-paper/issues/3}{initialised the formation of the writing team},
-wrote sections not mentioned below, and revised the full first draft. CB
-.. RC and EH wrote the section on interfaces for Docker in R. DE wrote
-the introduction and section about Rocker. ME .. CF wrote the paragraphs
-about \texttt{r-online}, \texttt{dockerfiler}, \texttt{r-ci} and
-\texttt{r-db}. BW .. KR .. NR .. NX wrote the section on liftr. LS \& NT
-wrote the section on Bioconductor. All authors approved the final
-version.
-
-This articles was collaboratively written at
+wrote sections not mentioned below, and revised all sections. DB wrote
+the section on \pkg{outsider}. DE wrote the introduction and the section
+about Containerization and Rocker. RC \& EH contributed to the section
+on interfaces for Docker in R. ME contributed to the section on
+processing and deployment to cloud services. CF wrote paragraphs about
+\pkg{r-online}, \pkg{dockerfiler}, \pkg{r-ci} and \pkg{r-db}. BM wrote
+the section on research compendia and made the project Binder-ready. HN
+\& JN co-wrote the section on enterprise production environments
+(\pkg{googleComputeEngineR}, Google Cloud Run, Kubernetes). KR wrote the
+section about \pkg{wholepunch}. NR wrote paragraphs about shared work
+environments and GPUs. LS \& NT wrote the section on Bioconductor. NX
+wrote the section on \pkg{liftr}. All authors approved the final
+version. This articles was collaboratively written at
 \href{https://github.com/nuest/rockerverse-paper/}{https://github.com/nuest/rockerverse-paper/}.
@@
 \address{%
+Dom Bennett\\
+University of Gothenburg, Sweden\\
+Department of Biological and Environmental Sciences\\ Carl Skottsbergs gata 22B\\ 413 19 Gothenburg, Sweden\\ \orcid{0000-0003-2722-1359}\\
+}
+\href{mailto:dominic.john.bennett@gmail.com}{\nolinkurl{dominic.john.bennett@gmail.com}}
+
+\address{%
 Robrecht Cannoodt\\
@@
 \address{%
+Ellis Hughes\\
+Fred Hutchinson Cancer Research Center\\
+Vaccine and Infectious Disease\\ 1100 Fairview Ave. N., P.O. Box 19024\\ Seattle, WA 98109-1024, USA\\
+}
+\href{mailto:ehhughes@fredhutch.org}{\nolinkurl{ehhughes@fredhutch.org}}
+
+\address{%
+Sean Lopp\\
+RStudio, Inc\\
+250 Northern Ave\\ Boston, MA 02210, USA\\
+}
+\href{mailto:sean@rstudio.com}{\nolinkurl{sean@rstudio.com}}
+
+\address{%
+Ben Marwick\\
+University of Washington\\
+Department of Anthropology\\ Denny Hall 230, Spokane Ln\\ Seattle, WA 98105, USA\\ \orcid{0000-0001-7879-4531}\\
+}
+\href{mailto:bmarwick@uw.edu}{\nolinkurl{bmarwick@uw.edu}}
+
+\address{%
+Heather Nolis\\
+T-Mobile\\
+12920 Se 38th St.\\ Bellevue, WA, 98006, USA\\
+}
+\href{mailto:heather.wensler1@t-mobile.com}{\nolinkurl{heather.wensler1@t-mobile.com}}
+
+\address{%
+Jacqueline Nolis\\
+Nolis, LLC\\
+Seattle, WA, USA\\ \orcid{0000-0001-9354-6501}\\
+}
+\href{mailto:jacqueline@nolisllc.com}{\nolinkurl{jacqueline@nolisllc.com}}
+
+\address{%
+Hong Ooi\\
+Microsoft\\
+Level 5, 4 Freshwater Place\\ Southbank, VIC 3006, Australia\\
+}
+\href{mailto:hongooi@microsoft.com}{\nolinkurl{hongooi@microsoft.com}}
+
+\address{%
 Karthik Ram\\
@@
 \address{%
+Lori Shepherd\\
+Roswell Park Comprehensive Cancer Center\\
+Elm \& Carlton Streets\\ Buffalo, NY, 14263, USA\\ \orcid{0000-0002-5910-4010}\\
+}
+\href{mailto:lori.shepherd@roswellpark.org}{\nolinkurl{lori.shepherd@roswellpark.org}}
+
+\address{%
+Nitesh Turaga\\
+Roswell Park Comprehensive Cancer Center\\
+Elm \& Carlton Streets\\ Buffalo, NY, 14263, USA\\ \orcid{0000-0002-0224-9817}\\
+}
+\href{mailto:nitesh.turaga@roswellpark.org}{\nolinkurl{nitesh.turaga@roswellpark.org}}
+
+\address{%
 Nan Xiao\\
@@
 \href{mailto:me@nanx.me}{\nolinkurl{me@nanx.me}}
-
-\address{%
-Lori Shepherd\\
-Roswell Park Comprehensive Cancer Center\\
-Elm \& Carlton Streets\\ Buffalo, New York, 14263, USA\\ \orcid{0000-0002-5910-4010}\\
-}
-\href{mailto:lori.shepherd@roswellpark.org}{\nolinkurl{lori.shepherd@roswellpark.org}}
-
-\address{%
-Nitesh Turaga\\
-Roswell Park Comprehensive Cancer Center\\
-Elm \& Carlton Streets\\ Buffalo, New York, 14263, USA\\ \orcid{0000-0002-0224-9817}\\
-}
-\href{mailto:nitesh.turaga@roswellpark.org}{\nolinkurl{nitesh.turaga@roswellpark.org}}
-
-\address{%
-Jacqueline Nolis\\
-Nolis, LLC\\
-Seattle, WA, USA\\ \orcid{0000-0001-9354-6501}\\
-}
-\href{mailto:jacqueline@nolisllc.com}{\nolinkurl{jacqueline@nolisllc.com}}
-
-\address{%
-Heather Nolis\\
-T-Mobile\\
-12920 Se 38th St.\\ Bellevue, WA, 98006\\
-}
-\href{mailto:heather.wensler1@t-mobile.com}{\nolinkurl{heather.wensler1@t-mobile.com}}
-
-\address{%
-Hong Ooi\\
-Microsoft\\
-Level 5, 4 Freshwater Place\\ Southbank, VIC 3006, Australia\\
-}
-\href{mailto:hongooi@microsoft.com}{\nolinkurl{hongooi@microsoft.com}}
-
-\address{%
-Ellis Hughes\\
-Fred Hutchinson Cancer Research Center\\
-Vaccine and Infectious Disease\\ 1100 Fairview Ave. N., P.O. Box 19024\\ Seattle, WA 98109-1024, USA\\
-}
-\href{mailto:ehhughes@fredhutch.org}{\nolinkurl{ehhughes@fredhutch.org}}
-
-\address{%
-Sean Lopp\\
-RStudio, Inc\\
-RStudio\\ 250 Northern Ave\\ Boston, MA 02210\\
-}
-\href{mailto:sean@rstudio.com}{\nolinkurl{sean@rstudio.com}}
-
-\address{%
-Ben Marwick\\
-\\
-\\
-}

### 2020-01-16 manuscript.Rmd
Some minor clarification and answering questions
I removed the GUI or CLI distinction - these are two ways of running the client, and seem extraneous to the core points here. But feel free to add back in (or even reject these changes).

--- 2020-01-16 manuscript.Rmd
@@
 **Gigantum** ([http://gigantum.com/](http://gigantum.com/) is a platform for open and decentralized data science with a focus on using automation and user-friendly tools for easy sharing of reproducible computational workflows.
-Gigantum builds on the locally running _Gigantum CLI_ and _Gigantum Desktop_ applications for development and execution of analyses, which can be stored and shared via _Gigantum Cloud_.
-The Gigantum apps are user friendly interfaces to a backend using Docker containers to package a Gigantum project ([https://github.com/gigantum/base-images](https://github.com/gigantum/base-images)).
+Gigantum builds on the _Gigantum Client_ (running either locally or on a remote server) for development and execution of data-focused _Projects_, which can be stored and shared via the _Gigantum Hub_ or via a zipfile export.
+The Client is a user friendly interface to a backend using Docker containers to package, build, and run Gigantum projects.
+It is configured to use a default set of Docker base images ([https://github.com/gigantum/base-images](https://github.com/gigantum/base-images)), and users are able to define and configure their own custom images.
 The available images include two with R based on Ubuntu Linux have the [`c2d4u` CRAN PPA](https://launchpad.net/~marutter/+archive/ubuntu/c2d4u3.5/) pre-configured for installation of binary R packages\footnote{\href{https://docs.gigantum.com/docs/using-r}{https://docs.gigantum.com/docs/using-r}}.
-<!-- mention lack of littler? -->
 The R images vary in the included authoring environment, i.e., Jupyter in `r-tidyverse` or both Jupyter \& RStudio in `rstudio-server`.
@@
 The `Dockerfile`s also use build args to specify bases, for example for different version of NVIDIA CUDA for GPU processing\footnote{See \href{https://github.com/gigantum/base-images/blob/master/_templates/python3-minimal-template/Dockerfile}{https://github.com/gigantum/base-images/blob/master/_templates/python3-minimal-template/Dockerfile} for the `Dockerfile` of `python3-minimal`.} so that appropriate GPU drivers can be enabled automatically when supported.
-Furthermore, Gigantum's focus lies on environment management via GUI and ensuring a smooth user interaction, e.g., with reliable and easy conflict detection and resulution.
-<!-- @DC: when a user interactively installs packages in Python, do you write that to the Dockerfile?
-UI-based environment management does not work for R packages, right?
-\href{https://docs.gigantum.com/docs/environment-management}{https://docs.gigantum.com/docs/environment-management} -->
+Furthermore, Gigantum's focus lies on environment management via GUI and ensuring a smooth user interaction, e.g., with reliable and easy conflict detection and resolution.
+For this reason, project repositories store authoritative package information in a separate file per package, allowing Git to directly detect conflicts and changes.
+A Dockerfile is generated from this description that inherits from the specified base image, and additional custom Docker commands may be appended.
 Because of these specifics, instructions from `rocker/r-ubuntu` could _not_ be readily re-used in this image stack (see Section&nbsp;\nameref{conclusion}).
-The image build and publication process is scripted with Python and JSON template configuration files, unlike Rocker which relies on more plain `Dockerfile`s.
+Both approaches enable Apt as an installation method, and this is exposed via the GUI-based environment management,\footnote{See \href{https://docs.gigantum.com/docs/environment-management}{https://docs.gigantum.com/docs/environment-management}} though Gigantum's default base images do not currently include `littler`.
+The image build and publication process is scripted with Python and JSON template configuration files, unlike Rocker which relies on plain `Dockerfile`s.
 A minor reason in the inception of the images were also project constraints requiring a Rocker-incompatible licensing of the `Dockerfile`s, i.e., the MIT License.

### 2020-01-16 manuscript.Rmd
extend jupyter docker stack description

--- 2020-01-16 manuscript.Rmd
@@
 
-The [**Jupyter Docker Stacks** project](https://github.com/jupyter/docker-stacks/) are a set of ready-to-run Docker images containing Jupyter applications and interactive computing tools \citep{project_jupyter_jupyter_2018}.
-The `jupyter/r-notebook` image includes R and "popular packages", and R is also included in the catchall `datascience-notebook` image\footnote{\href{https://jupyter-docker-stacks.readthedocs.io/en/latest/using/selecting.html}{https://jupyter-docker-stacks.readthedocs.io/en/latest/using/selecting.html}}.
+The [**Jupyter Docker Stacks**](https://github.com/jupyter/docker-stacks/) project are a set of ready-to-run Docker images containing Jupyter applications and interactive computing tools \citep{project_jupyter_jupyter_2018}.
+The `jupyter/r-notebook` image includes R and "popular packages", and naturally also the IRKernel ([https://irkernel.github.io/](https://irkernel.github.io/)), an R kernel for Jupyter so that Jupyter Notebooks can contain R code cells.
+R is also included in the catchall `jupyter/datascience-notebook` image\footnote{\href{https://jupyter-docker-stacks.readthedocs.io/en/latest/using/selecting.html}{https://jupyter-docker-stacks.readthedocs.io/en/latest/using/selecting.html}}.
 For example, these images allow users to quickly start a Jupyter Notebook server locally or build their own specialised images on top of stable toolsets.
+The R installation is installed using the Conda package manager, which can manage environments for various programming languages, pinning both the R version and the versions of R packages\footnote{See `jupyter/datascience-notebook` `Dockerfile` at \href{https://github.com/jupyter/docker-stacks/blob/master/datascience-notebook/Dockerfile#L47}{https://github.com/jupyter/docker-stacks/blob/master/datascience-notebook/Dockerfile#L47}.}.
 

### 2020-01-16 manuscript.Rmd
one sentence per line, url formatting

--- 2020-01-16 manuscript.Rmd
@@
 The R images are based on Gigantum's `python3-minimal` image, originally to keep the existing front-end configuration, but also to provide a consistent Python-to-R interoperability.
-The `Dockerfile`s also use build args to specify bases, for example for different version of NVIDIA CUDA for GPU processing\footnote{See \href{https://github.com/gigantum/base-images/blob/master/_templates/python3-minimal-template/Dockerfile}{https://github.com/gigantum/base-images/blob/master/_templates/python3-minimal-template/Dockerfile} for the `Dockerfile` of `python3-minimal`.} so that appropriate GPU drivers can be enabled automatically when supported.
+The `Dockerfile`s also use build args to specify bases, for example for different version of NVIDIA CUDA for GPU processing\footnote{See \href{https://github.com/gigantum/base-images/blob/master/_templates/python3-minimal-template/Dockerfile}{https://github.com/gigantum/base-images/blob/master/\_templates/python3-minimal-template/Dockerfile} for the `Dockerfile` of `python3-minimal`.} so that appropriate GPU drivers can be enabled automatically when supported.
 Furthermore, Gigantum's focus lies on environment management via GUI and ensuring a smooth user interaction, e.g., with reliable and easy conflict detection and resolution.

### 2020-01-16 manuscript.Rmd
minor change

@@
 For this reason, project repositories store authoritative package information in a separate file per package, allowing Git to directly detect conflicts and changes.
-A Dockerfile is generated from this description that inherits from the specified base image, and additional custom Docker commands may be appended.
+A Dockerfile is generated from this description that inherits from the specified base image, and additional custom Docker instructions may be appended by users, though Gigantum's default base images do not currently include the `littler` tool, which is used by Rocker to install packages within `Dockerfile`s.
 Because of these specifics, instructions from `rocker/r-ubuntu` could _not_ be readily re-used in this image stack (see Section&nbsp;\nameref{conclusion}).
-Both approaches enable Apt as an installation method, and this is exposed via the GUI-based environment management,\footnote{See \href{https://docs.gigantum.com/docs/environment-management}{https://docs.gigantum.com/docs/environment-management}} though Gigantum's default base images do not currently include `littler`.
+Both approaches enable `apt` as an installation method, and this is exposed via the GUI-based environment management,\footnote{See \href{https://docs.gigantum.com/docs/environment-management}{https://docs.gigantum.com/docs/environment-management}}.
 The image build and publication process is scripted with Python and JSON template configuration files, unlike Rocker which relies on plain `Dockerfile`s.

### 2020-01-17 manuscript.Rmd
Add a short section on the use of Rocker containers in Dodona

@@
 ---
-title: "Containers and R---the Rockerverse and beyond"
+title: "The Rockerverse: Images, Packages, and Applications for Containerization with R"
 author:
@@
     email: dav@gigantum.com
+  - name: Gergely Daroczi
+    # https://github.com/daroczig
+    address:
+      - \orcid{0000-0003-3149-8537}
+    email: daroczig@rapporter.net
   - name: Dirk Eddelbuettel
@@
     email: me@nanx.me
+  - name: Charlotte Van Petegem
+    # https://github.com/charvp
+    affiliation: Ghent University
+    address:
+      - Department WE02
+      - Krijgslaan 281, S9
+      - 9000 Gent, Belgium
+      - \orcid{0000-0003-0779-4897}
+    email: charlotte.vanpetegem@ugent.be
 abstract: >
@@
 For example, these images allow users to quickly start a Jupyter Notebook server locally or build their own specialised images on top of stable toolsets.
-The R installation is installed using the Conda package manager, which can manage environments for various programming languages, pinning both the R version and the versions of R packages\footnote{See `jupyter/datascience-notebook` `Dockerfile` at \href{https://github.com/jupyter/docker-stacks/blob/master/datascience-notebook/Dockerfile#L47}{https://github.com/jupyter/docker-stacks/blob/master/datascience-notebook/Dockerfile#L47}.}.
+R is installed using the Conda package manager, which can manage environments for various programming languages, pinning both the R version and the versions of R packages\footnote{See \texttt{jupyter/datascience-notebook}'s \texttt{Dockerfile} at \href{https://github.com/jupyter/docker-stacks/blob/master/datascience-notebook/Dockerfile\#L47}{https://github.com/jupyter/docker-stacks/blob/master/datascience-notebook/Dockerfile\#L47}.}.
 
@@
 The R images are based on Gigantum's `python3-minimal` image, originally to keep the existing front-end configuration, but also to provide a consistent Python-to-R interoperability.
-The `Dockerfile`s also use build args to specify bases, for example for different version of NVIDIA CUDA for GPU processing\footnote{See \href{https://github.com/gigantum/base-images/blob/master/_templates/python3-minimal-template/Dockerfile}{https://github.com/gigantum/base-images/blob/master/\_templates/python3-minimal-template/Dockerfile} for the `Dockerfile` of `python3-minimal`.} so that appropriate GPU drivers can be enabled automatically when supported.
+The `Dockerfile`s also use build args to specify bases, for example for different version of NVIDIA CUDA for GPU processing\footnote{See \href{https://github.com/gigantum/base-images/blob/master/_templates/python3-minimal-template/Dockerfile}{https://github.com/gigantum/base-images/blob/master/\_templates/python3-minimal-template/Dockerfile} for the \texttt{Dockerfile} of \texttt{python3-minimal}.} so that appropriate GPU drivers can be enabled automatically when supported.
 Furthermore, Gigantum's focus lies on environment management via GUI and ensuring a smooth user interaction, e.g., with reliable and easy conflict detection and resolution.
@@
 
-**\pkg{stevedore}** ([https://cran.r-project.org/package=stevedore](https://cran.r-project.org/package=stevedore)) 
-<!-- from the CRAN description:
-Work with containers over the Docker API.
-Rather than using system calls to interact with a docker client, using the API directly means that we can receive richer information from docker.
-The interface in the package is automatically generated using the 'OpenAPI' (a.k.a., 'swagger') specification, and all return values are checked in order to make them type stable.
--->
-currently the most powerful Docker client in R.
-
-**\pkg{AzureContainers}** ([https://cran.r-project.org/package=AzureContainers](https://cran.r-project.org/package=AzureContainers)) is an interface to a number of container-related services in Microsoft's [Azure Cloud](https://azure.microsoft.com/), namely [Container Instances](https://azure.microsoft.com/en-us/services/container-instances/), [Container Registry](https://azure.microsoft.com/en-us/services/container-registry/), and [Kubernetes Service](https://azure.microsoft.com/en-us/services/kubernetes-service/).
+**\pkg{stevedore}** ([https://cran.r-project.org/package=stevedore](https://cran.r-project.org/package=stevedore)) is currently the most powerful Docker client in R.
+It interfaces with the Docker daemon over the Docker HTTP API\footnote{} via a Unix socket on Linux or MacOS, over a named pipe on Windows, or over an HTTP/TCP connection.
+The package is the only one not using system calls to the `docker` CLI for managing images and containers and easily exposes connections to remote Docker daemons, which has to be configured on the Docker level otherwise.
+Using the API gives access to more information and is system independent and likely more reliable than parsing command line output.
+\pkg{stevedore}'s own interface is automatically generated based on the OpenAPI specification of the Docker daemon, but still similar to the Docker CLI.
+The interface is similar to R6 objects, in that a `stevedore_object` has a number of functions attached to it that can be called, and multiple specific versions of the Docker API can be supported thanks to the automatic generation\footnote{See \href{https://github.com/richfitz/stevedore/tree/master/inst/spec}{https://github.com/richfitz/stevedore/tree/master/inst/spec} and \href{https://github.com/richfitz/stevedore/blob/master/development.md}{https://github.com/richfitz/stevedore/blob/master/development.md}.}.
+
+**\pkg{AzureContainers}** ([https://cran.r-project.org/package=AzureContainers](https://cran.r-project.org/package=AzureContainers)) is an interface to a number of container-related services in Microsoft's [Azure Cloud](https://azure.microsoft.com/), see Section&nbsp;\nameref{azurecontainers}.
 While it is mainly intended for working with Azure, as a convenience feature it includes lightweight, cross-platform shells to Docker and Kubernetes (tools `kubectl` and `helm`).
@@
 However, thanks to advancements in R running as a web service, along with with the ability to deploy R in Docker containers, modern enterprises are now capable of having real-time machine learning powered by R.
-
-The backbone of this work is the \pkg{plumber} ([https://www.rplumber.io/](https://www.rplumber.io/)) package.
-The plumber package allows R to be used to host web services through RESTful APIs.
+One of the first implementations of having R respond to client reqests were TCP/IP and local socket servers, such as \CRANpkg{Rserve} and \CRANpkg{svSocket}, which allowed other local or remote processes to make binary requests to interact with R without the need to link to R at compile time and connected other programming languages with R.
+R could also empower web services since at least 2007, when [rApache](http://www.rapache.net) was presented at the useR! conference \citep{horner_rapache_2013}.
+<!-- website seems down, https://web.archive.org/web/20190627101203/http://rapache.net/manual.html -->
+rApache runs an R interpreter inside the Apache web server, which makes it possible and very efficient to build scalable R-based APIs and web services.
+
+Such rApache applications have been used at [rapporter.net](http://rapporter.net), transforming parametrized R Markdown-like documents into interactive web applications for reporting.
+A similar design was implemented at [CARD.com](https://www.card.com/), where the optimization of the acquisition flow and the real-time analysis of debit card transactions both relied on an efficient combination of Rserve and rApache running in Docker containers:
+the R session behind Rserve acted as a read-only, exteremely fast and scalabale, in-memory database for the many concurrent rApache processes responding to the live-scoring requests of various divisions of the company.
+Similar dockerized R scripts were responsible for the ETL processes and even the client-facing email, text message and push notification alerts sent in real-time based on card transactions.
+The related Docker images were made available at [https://github.com/cardcorp/card-rocker](https://github.com/cardcorp/card-rocker).
+The images extend `rocker/r-base`, e.g., with an SSH client and a workaround for being able to mount SSH keys from the host, Pandoc, the Amazon Web Services (AWS) SDK, and Java, which is required by the AWS SDK.
+The AWS SDK allowed to run R consumers reading from real-time data processing streams of [AWS Kinesis](https://aws.amazon.com/kinesis/).
+The applications were deployed on Amazon Elastic Container Service ([ECS](https://aws.amazon.com/ecs/)).
+The main learnings from using R in Docker was the imporance of not only pinning the R package versions via MRAN, but also moving away from Debian testing to an LTS distribution to minimize the risk of breaking the automated builds of the Docker images and production jobs.
+
+A more recent related work is the \pkg{plumber} ([https://www.rplumber.io/](https://www.rplumber.io/)) package, which makes it even easier for R to be used to host web services through RESTful APIs.
 When R is running on a server, a different server can ask R to make a computation through an HTTP request and get an immediate response.
@@
 The AI @ T-Mobile team open sourced their Docker images [on GitHub](https://github.com/tmobile/r-tensorflow-api).
+
+[RestRserve](http://restrserve.org) is an alternative R web API framework with the main focus on being high-performant and robust.
+It also ships automated Docker builds at https://hub.docker.com/r/rexyai/restrserve/dockerfile.
 
@@
 An introduction on using [Kubernetes with R is published at this blog post](https://code.markedmondson.me/r-on-kubernetes-serverless-shiny-r-apis-and-scheduled-scripts/).
-- [`AzureContainers`](https://cran.r-project.org/package=AzureContainers): Umbrella package for working with containers in Microsoft's Azure cloud.
-Provides interfaces to three Azure services: Container Registry, Container Instances (for running individual containers), and Kubernetes Services (for orchestrated deployments)
-
-While Plumber provides the low-level infrastructure for turning a statistical model into a predictive service, for production purposes it is usually necessary to take into account considerations such as scalability, reliability and ease of management.
-A single container is limited in the volume of requests it can service.
-Furthermore, if the container goes down, or is occupied with processing a computationally intensive task, the service becomes unavailable until it is restarted.
-
-These issues can be mitigated by deploying not a single container, but a _cluster_ of containers, orchestrated as a single deployment.
-The deployment system of choice at the time of writing is [Kubernetes](https://kubernetes.io).
-This allows you to deploy a single image to a pool of servers, which can then serve requests in parallel.
-Client-side requests are made to the cluster ingress endpoint, which then redirects them to individual servers.
-The orchestrator can identify cluster nodes that have failed, and restart them automatically; it can also detect when there is no capacity to service more requests, and start additional servers and containers as needed.
-
-While these load-balancing and autoscaling properties of a Kubernetes cluster are invaluable, they are still limited by the hardware resources that it has access to.
-A cluster cannot scale out if it has run out of servers.
-This issue in turn can be addressed by deploying not to on-premise servers but to the cloud, where effectively unlimited computing resources are available (as long as one can pay for them).
-From a management perspective, cloud deployments have the additional advantage that hardware maintenance becomes the responsibility of the cloud provider, eliminating another pain point.
-Orchestrator services are available from all the major cloud service providers, including Amazon Web Services (AWS), Google Cloud Platform, and Microsoft Azure.
-
-An example of an R package for working with containers in the cloud is AzureContainers, which interacts with three Azure services: Container Instances (for running a single container), Container Registry (a private docker registry service), and Kubernetes Service (for Kubernetes on Azure).
-AzureContainers provides a lightweight yet powerful interface for working with Resource Manager, the framework for deploying and managing arbitrary resources in Azure.
-On the client side, it provides simple shells to the Docker, Kubectl and Helm tools that are commonly used for managing containers and Kubernetes clusters.
-
-Here is some simple code that pushes a Docker image to an Azure Container Registry, and then creates a deployment using Azure Kubernetes Service.
-This is a slightly modified version of the code from the "Deploying a prediction service with Plumber" AzureContainers vignette.
-The code makes use of a yaml file to define the deployment and predictive service, which is standard practice with Kubernetes; the yaml can be seen in the aforementioned vignette.
-
-```r
-library(AzureContainers)
-# create a resource group for our deployments
-deployresgrp <- AzureRMR::get_azure_login()$
-    get_subscription("subscription_id")$
-    create_resource_group("deployresgrp", location="australiaeast")
-
-### create a container registry
-deployreg_svc <- deployresgrp$create_acr("deployreg")
-
-# build our image (a random forest on the Boston housing data, available in the MASS package)
-call_docker("build -t bos-rf .")
-
-# upload the image to Azure
-deployreg <- deployreg_svc$get_docker_registry()
-deployreg$push("bos-rf")
-
-### create a Kubernetes cluster with 2 nodes
-deployclus_svc <- deployresgrp$create_aks("deployclus", agent_pools=aks_pools("pool1", 2))
-
-# grant the cluster pull access to the registry
-gr <- AzureGraph::get_graph_login()
-aks_app_id <- deployclus$properties$servicePrincipalProfile$clientID
-reg$add_role_assignment(gr$get_app(aks_app_id), "Acrpull")
-
-# get the cluster endpoint
-deployclus <- deployclus_svc$get_cluster()
-
-# create and start the service
-deployclus$create("bos-rf.yaml")
-```
-
-Once the service has been created, we can check on its status and obtain predictions:
-
-```r
-deployclus$get("deployment bos-rf")
-#> Kubernetes operation: get deployment bos-rf  --kubeconfig=".../kubeconfigxxxx"
-#> NAME      DESIRED   CURRENT   UP-TO-DATE   AVAILABLE   AGE
-#> bos-rf    1         1         1            1           5m
-
-deployclus$get("service bos-rf-svc")
-#> Kubernetes operation: get service bos-rf-svc  --kubeconfig=".../kubeconfigxxxx"
-#> NAME         TYPE           CLUSTER-IP   EXTERNAL-IP     PORT(S)          AGE
-#> bos-rf-svc   LoadBalancer   10.0.8.189   xxx.xxx.xxx.xxx 8000:32276/TCP   5m
-
-response <- httr::POST("http://xxx.xxx.xxx.xxx:8000/score",
-    body=list(df=MASS::Boston[1:10,]), encode="json")
-httr::content(response, simplifyVector=TRUE)
-#> [1] 25.9269 22.0636 34.1876 33.7737 34.8081 27.6394 21.8007 22.3577 16.7812 18.9785
-```
-
-Note that Plumber, by itself, provides no authentication or security features; anybody who knows the cluster's IP address can access the predictive service.
-At minimum, a basic authentication layer should be added to any Kubernetes deployment.
-Standard Kubernetes procedure involves creating an ingress controller, for example with the nginx or traefik reverse proxy software, and one can then add application-specific authentication layers on top of that (for example, to authenticate with a user's Azure Active Directory or LDAP credentials).
-More details can be found from most Kubernetes learning resources, and are outside the scope of this paper.
+
+\label{azurecontainers}
+**\pkg{AzureContainers}** is an umbrealla package providing interfaces the three services of Microsoft's Azure Cloud, namely [Container Instances](https://azure.microsoft.com/en-us/services/container-instances/) for running individual containers, [Container Registry](https://azure.microsoft.com/en-us/services/container-registry/) for private image distribution, and [Kubernetes Service](https://azure.microsoft.com/en-us/services/kubernetes-service/) for orchestrated deployments.
+While a package like \pkg{plumber} provides the infrastructure for turning an R workflow a service, for production purposes it is usually necessary to take into account scalability, reliability and ease of management.
+A single container is limited in the volume of requests it can service, and if the container goes down or is occupied with processing a computationally intensive task, the service becomes unavailable.
+These issues can be mitigated by deploying not a single container, but a _cluster_ of containers, orchestrated as a single deployment, e.g., with [Kubernetes](https://kubernetes.io).
+A cluster features load-balancing, autoscaling of containers across numerous servers (in the cloud or on premise), and restarting failed ones.
+\pkg{AzureContainers} provides a lightweight yet powerful interface for working with Resource Manager, the framework for deploying and managing arbitrary resources in Azure.
+On the client side, it provides simple shells to the Docker, and to Kubernetes' `kubectl` and `helm` tools, that are needed for managing Kubernetes clusters.
+The package simplifies complex infrastructure management to a number of R funcction calls, given an Azure account with sufficient credit\footnote{See \emph{"Deploying a prediction service with Plumber"} vignette for details:  \href{https://cran.r-project.org/web/packages/AzureContainers/vignettes/vig01_plumber_deploy.html}{https://cran.r-project.org/web/packages/AzureContainers/vignettes/vig01\_plumber\_deploy.html}.}.
 
@@
 The `rocker/tidyverse` base image ensures that users can also readily use packages for analysis, display, and reporting.
+
+Another way this idea of a common environment and the partitioning that docker provides is used in teaching is in the **[Dodona](https://dodona.ugent.be)** project.
+Dodona is a web platform developed at Ghent University and is used to teach students basic programming skills.
+Dodona uses Docker containers to test submissions by students.
+This means that both the code testing the students' submissions and the submission itself are executed in a predictable environment, avoiding compatibility issues between the wide variety of configurations used by students.
+The containerization is also used to shield the Dodona servers from bad (or even malicious) code: memory, time and I/O limits are used to make sure students can't overload the system.
+The web application communicates with containers by sending configuration information as a JSON document over standard input. 
+Every Dodona docker image shares a `main.sh` file that passes through this information to the actual testing framework, while setting up some error handling. 
+The testing process in the docker containers sends back the test results by writing a JSON document to its standard output channel.
+In june 2019 R support was added to Dodona using an image derived from the Rocker `r-base` image that sets up the `runner` user and and `main.sh` file expected by Dodona\footnote{\href{https://github.com/dodona-edu/docker-images/blob/master/dodona-r.dockerfile}{https://github.com/dodona-edu/docker-images/blob/master/dodona-r.dockerfile}}.
+It also installs the packages required for the testing framework and the exercises so that this doesn't have to happen every time a student's submission is evaluated.
+The actual testing of R exercises is done using a custom framework loosely based on `testthat`.
+During the development of the testing framework it was found that the testthat framework did not provide enough information to its reporter system to send back all the fields required by Dodona to render its feedback.
+Right now, multiple statistics courses are developing exercises to automate the feedback for their lab classes.
 
@@
 Each outsider module has a `Dockerfile` and an R package with functions for interacting with the encapsulated tool.
-Using only R functions, an end-user can install a module with the \pkg{outsider} package and then call module code to integrate a tool into their own workflow seamlessly.
+Using only R functions, an end-user can install a module with the \pkg{outsider} package and then call module code to integrate a tool into their own R-based workflow seamlessly.
 The \pkg{outsider} package and module manage the containers and handle the transmission of arguments and the transfer of files to and from a container.
@@
 DB wrote the section on \pkg{outsider}.
+GD contributed the CARD.com use case.
 DE wrote the introduction and the section about Containerization and Rocker.

### 2020-01-17 manuscript.Rmd
add stevedore

--- 2020-01-16 manuscript.Rmd
@@
 
-**\pkg{stevedore}** ([https://cran.r-project.org/package=stevedore](https://cran.r-project.org/package=stevedore)) 
-<!-- from the CRAN description:
-Work with containers over the Docker API.
-Rather than using system calls to interact with a docker client, using the API directly means that we can receive richer information from docker.
-The interface in the package is automatically generated using the 'OpenAPI' (a.k.a., 'swagger') specification, and all return values are checked in order to make them type stable.
--->
-currently the most powerful Docker client in R.
+**\pkg{stevedore}** ([https://cran.r-project.org/package=stevedore](https://cran.r-project.org/package=stevedore)) is currently the most powerful Docker client in R.
+It interfaces with the Docker daemon over the Docker HTTP API\footnote{} via a Unix socket on Linux or MacOS, over a named pipe on Windows, or over an HTTP/TCP connection.
+The package is the only one not using system calls to the `docker` CLI for managing images and containers and easily exposes connections to remote Docker daemons, which has to be configured on the Docker level otherwise.
+Using the API gives access to more information and is system independent and likely more reliable than parsing command line output.
+\pkg{stevedore}'s own interface is automatically generated based on the OpenAPI specification of the Docker daemon, but still similar to the Docker CLI.
+The interface is similar to R6 objects, in that a `stevedore_object` has a number of functions attached to it that can be called, and multiple specific versions of the Docker API can be supported thanks to the automatic generation\footnote{See \href{https://github.com/richfitz/stevedore/tree/master/inst/spec}{https://github.com/richfitz/stevedore/tree/master/inst/spec} and \href{https://github.com/richfitz/stevedore/blob/master/development.md}{https://github.com/richfitz/stevedore/blob/master/development.md}.}.
 

### 2020-01-17 manuscript.Rmd
shorten AzureContainers

--- 2020-01-17 manuscript.Rmd
@@
 ---
-title: "The Rockerverse: Images, Packages, and Applications for Containerization with R"
+title: "Containers and R---the Rockerverse and beyond"
 author:
@@
     email: dav@gigantum.com
-  - name: Gergely Daroczi
-    # https://github.com/daroczig
-    address:
-      - \orcid{0000-0003-3149-8537}
-    email: daroczig@rapporter.net
   - name: Dirk Eddelbuettel
@@
     email: me@nanx.me
-  - name: Charlotte Van Petegem
-    # https://github.com/charvp
-    affiliation: Ghent University
-    address:
-      - Department WE02
-      - Krijgslaan 281, S9
-      - 9000 Gent, Belgium
-      - \orcid{0000-0003-0779-4897}
-    email: charlotte.vanpetegem@ugent.be
 abstract: >
@@
 For example, these images allow users to quickly start a Jupyter Notebook server locally or build their own specialised images on top of stable toolsets.
-R is installed using the Conda package manager, which can manage environments for various programming languages, pinning both the R version and the versions of R packages\footnote{See \texttt{jupyter/datascience-notebook}'s \texttt{Dockerfile} at \href{https://github.com/jupyter/docker-stacks/blob/master/datascience-notebook/Dockerfile\#L47}{https://github.com/jupyter/docker-stacks/blob/master/datascience-notebook/Dockerfile\#L47}.}.
+The R installation is installed using the Conda package manager, which can manage environments for various programming languages, pinning both the R version and the versions of R packages\footnote{See `jupyter/datascience-notebook` `Dockerfile` at \href{https://github.com/jupyter/docker-stacks/blob/master/datascience-notebook/Dockerfile#L47}{https://github.com/jupyter/docker-stacks/blob/master/datascience-notebook/Dockerfile#L47}.}.
 
@@
 The R images are based on Gigantum's `python3-minimal` image, originally to keep the existing front-end configuration, but also to provide a consistent Python-to-R interoperability.
-The `Dockerfile`s also use build args to specify bases, for example for different version of NVIDIA CUDA for GPU processing\footnote{See \href{https://github.com/gigantum/base-images/blob/master/_templates/python3-minimal-template/Dockerfile}{https://github.com/gigantum/base-images/blob/master/\_templates/python3-minimal-template/Dockerfile} for the \texttt{Dockerfile} of \texttt{python3-minimal}.} so that appropriate GPU drivers can be enabled automatically when supported.
+The `Dockerfile`s also use build args to specify bases, for example for different version of NVIDIA CUDA for GPU processing\footnote{See \href{https://github.com/gigantum/base-images/blob/master/_templates/python3-minimal-template/Dockerfile}{https://github.com/gigantum/base-images/blob/master/\_templates/python3-minimal-template/Dockerfile} for the `Dockerfile` of `python3-minimal`.} so that appropriate GPU drivers can be enabled automatically when supported.
 Furthermore, Gigantum's focus lies on environment management via GUI and ensuring a smooth user interaction, e.g., with reliable and easy conflict detection and resolution.
@@
 However, thanks to advancements in R running as a web service, along with with the ability to deploy R in Docker containers, modern enterprises are now capable of having real-time machine learning powered by R.
-One of the first implementations of having R respond to client reqests were TCP/IP and local socket servers, such as \CRANpkg{Rserve} and \CRANpkg{svSocket}, which allowed other local or remote processes to make binary requests to interact with R without the need to link to R at compile time and connected other programming languages with R.
-R could also empower web services since at least 2007, when [rApache](http://www.rapache.net) was presented at the useR! conference \citep{horner_rapache_2013}.
-<!-- website seems down, https://web.archive.org/web/20190627101203/http://rapache.net/manual.html -->
-rApache runs an R interpreter inside the Apache web server, which makes it possible and very efficient to build scalable R-based APIs and web services.
-
-Such rApache applications have been used at [rapporter.net](http://rapporter.net), transforming parametrized R Markdown-like documents into interactive web applications for reporting.
-A similar design was implemented at [CARD.com](https://www.card.com/), where the optimization of the acquisition flow and the real-time analysis of debit card transactions both relied on an efficient combination of Rserve and rApache running in Docker containers:
-the R session behind Rserve acted as a read-only, exteremely fast and scalabale, in-memory database for the many concurrent rApache processes responding to the live-scoring requests of various divisions of the company.
-Similar dockerized R scripts were responsible for the ETL processes and even the client-facing email, text message and push notification alerts sent in real-time based on card transactions.
-The related Docker images were made available at [https://github.com/cardcorp/card-rocker](https://github.com/cardcorp/card-rocker).
-The images extend `rocker/r-base`, e.g., with an SSH client and a workaround for being able to mount SSH keys from the host, Pandoc, the Amazon Web Services (AWS) SDK, and Java, which is required by the AWS SDK.
-The AWS SDK allowed to run R consumers reading from real-time data processing streams of [AWS Kinesis](https://aws.amazon.com/kinesis/).
-The applications were deployed on Amazon Elastic Container Service ([ECS](https://aws.amazon.com/ecs/)).
-The main learnings from using R in Docker was the imporance of not only pinning the R package versions via MRAN, but also moving away from Debian testing to an LTS distribution to minimize the risk of breaking the automated builds of the Docker images and production jobs.
-
-A more recent related work is the \pkg{plumber} ([https://www.rplumber.io/](https://www.rplumber.io/)) package, which makes it even easier for R to be used to host web services through RESTful APIs.
+
+The backbone of this work is the \pkg{plumber} ([https://www.rplumber.io/](https://www.rplumber.io/)) package.
+The plumber package allows R to be used to host web services through RESTful APIs.
 When R is running on a server, a different server can ask R to make a computation through an HTTP request and get an immediate response.
@@
 The AI @ T-Mobile team open sourced their Docker images [on GitHub](https://github.com/tmobile/r-tensorflow-api).
-
-[RestRserve](http://restrserve.org) is an alternative R web API framework with the main focus on being high-performant and robust.
-It also ships automated Docker builds at https://hub.docker.com/r/rexyai/restrserve/dockerfile.
 
@@
 On the client side, it provides simple shells to the Docker, and to Kubernetes' `kubectl` and `helm` tools, that are needed for managing Kubernetes clusters.
-The package simplifies complex infrastructure management to a number of R funcction calls, given an Azure account with sufficient credit\footnote{See \emph{"Deploying a prediction service with Plumber"} vignette for details:  \href{https://cran.r-project.org/web/packages/AzureContainers/vignettes/vig01_plumber_deploy.html}{https://cran.r-project.org/web/packages/AzureContainers/vignettes/vig01\_plumber\_deploy.html}.}.
+The package simplifies complex infrastructure management to a number of R funcction calls, given an Azure account with sufficient credit\footnots{See _"Deploying a prediction service with Plumber"_ vignette at \href{https://cran.r-project.org/web/packages/AzureContainers/vignettes/vig01_plumber_deploy.html}{https://cran.r-project.org/web/packages/AzureContainers/vignettes/vig01_plumber_deploy.html} for details.}.
 
@@
 The `rocker/tidyverse` base image ensures that users can also readily use packages for analysis, display, and reporting.
-
-Another way this idea of a common environment and the partitioning that docker provides is used in teaching is in the **[Dodona](https://dodona.ugent.be)** project.
-Dodona is a web platform developed at Ghent University and is used to teach students basic programming skills.
-Dodona uses Docker containers to test submissions by students.
-This means that both the code testing the students' submissions and the submission itself are executed in a predictable environment, avoiding compatibility issues between the wide variety of configurations used by students.
-The containerization is also used to shield the Dodona servers from bad (or even malicious) code: memory, time and I/O limits are used to make sure students can't overload the system.
-The web application communicates with containers by sending configuration information as a JSON document over standard input. 
-Every Dodona docker image shares a `main.sh` file that passes through this information to the actual testing framework, while setting up some error handling. 
-The testing process in the docker containers sends back the test results by writing a JSON document to its standard output channel.
-In june 2019 R support was added to Dodona using an image derived from the Rocker `r-base` image that sets up the `runner` user and and `main.sh` file expected by Dodona\footnote{\href{https://github.com/dodona-edu/docker-images/blob/master/dodona-r.dockerfile}{https://github.com/dodona-edu/docker-images/blob/master/dodona-r.dockerfile}}.
-It also installs the packages required for the testing framework and the exercises so that this doesn't have to happen every time a student's submission is evaluated.
-The actual testing of R exercises is done using a custom framework loosely based on `testthat`.
-During the development of the testing framework it was found that the testthat framework did not provide enough information to its reporter system to send back all the fields required by Dodona to render its feedback.
-Right now, multiple statistics courses are developing exercises to automate the feedback for their lab classes.
 
@@
 DB wrote the section on \pkg{outsider}.
-GD contributed the CARD.com use case.
 DE wrote the introduction and the section about Containerization and Rocker.

### 2020-01-18 manuscript.Rmd
mention Rserve, svSocket and RApache before plumber

@@
 
-The backbone of this work is the \pkg{plumber} ([https://www.rplumber.io/](https://www.rplumber.io/)) package.
-The plumber package allows R to be used to host web services through RESTful APIs.
+One of the first implementations of having R respond to client reqests were TCP/IP and local socket servers, such as \pkg{Rserve} ([https://cran.r-project.org/package=Rserve](https://cran.r-project.org/package=Rserve)) and \pkg{svSocket} ([https://cran.r-project.org/package=svSocket](https://cran.r-project.org/package=svSocket)), which allowed other local or remote processes, even written in other programming languages, such as C/C++ or Java, to make binary requests to interact with R without the need to link to R at compile time.
+
+R could also empower webserver since at least 2007, when [RApache](http://www.rapache.net) was presented at the useR!
+conference.
+Rapache runs an R interpreter inside the Apache web server and makes it possible and very efficient to build scalable R-based APIs and web services.
+
+A more recent related work is the \pkg{plumber} ([https://www.rplumber.io/](https://www.rplumber.io/)) package, which makes it even easier for R to be used to host web services through RESTful APIs.
 When R is running on a server, a different server can ask R to make a computation through an HTTP request and get an immediate response.

### 2020-01-18 manuscript.Rmd
add use-cases on Rserve/RApache + Docker

@@
 conference.
-Rapache runs an R interpreter inside the Apache web server and makes it possible and very efficient to build scalable R-based APIs and web services.
+Rapache runs an R interpreter inside the Apache web server, which makes it possible and very efficient to build scalable R-based APIs and web services.
+
+Such RApache applications have been used at [rapporter.net](http://rapporter.net), transforming parametrized R Markdown-like documents into interactive web applications for reporting.
+A similar design was implemented at CARD.com, where the optimization of the acquisition flow and the real-time analysis of debit card transactions both relied on an efficient combination of Rserve and RApache running in Docker containers: the R session behind Rserve acted as a read-only, exteremely fast and scalabale, in-memory database for the many concurrent RApache processes responding to the live-scoring requests of other components of the company.
+Similar dockerized R scripts were responsible for the ETL processes and even the client-facing email, text message and push notification alerts sent in real-time on the card transaction.
+The related Docker images were made available at [https://github.com/cardcorp/card-rocker](https://github.com/cardcorp/card-rocker), which are extending `rocker/r-base` eg with an SSH client and workaround for being able to mount SSH keys from the host, Pandoc, the AWS SDK and Java -- which latter was required at that time to be able to easily use the Amazon Web Services via their Java SDK, eg to run R consumers reading from a Kinesis stream, deployed as a dockerized ECS service.
 

### 2020-01-18 manuscript.Rmd
mention RestRserve - thanks @eddelbuettel
mention learnings from using rocker/r-base in prod

@@
 The related Docker images were made available at [https://github.com/cardcorp/card-rocker](https://github.com/cardcorp/card-rocker), which are extending `rocker/r-base` eg with an SSH client and workaround for being able to mount SSH keys from the host, Pandoc, the AWS SDK and Java -- which latter was required at that time to be able to easily use the Amazon Web Services via their Java SDK, eg to run R consumers reading from a Kinesis stream, deployed as a dockerized ECS service.
+The main learnings from using R in Docker was the imporance of not only pinning the R package versions via MRAN, but also moving away from Debian testing to an LTS distribution to minimize the risk of breaking the automated builds of the Docker images and/or production jobs.
 

### 2020-01-18 manuscript.Rmd
sneaking into the list of authors
fix formatting

--- 2020-01-18 manuscript.Rmd
@@
 
-One of the first implementations of having R respond to client reqests were TCP/IP and local socket servers, such as \pkg{Rserve} ([https://cran.r-project.org/package=Rserve](https://cran.r-project.org/package=Rserve)) and \pkg{svSocket} ([https://cran.r-project.org/package=svSocket](https://cran.r-project.org/package=svSocket)), which allowed other local or remote processes, even written in other programming languages, such as C/C++ or Java, to make binary requests to interact with R without the need to link to R at compile time.
-
-R could also empower webserver since at least 2007, when [RApache](http://www.rapache.net) was presented at the useR!
-conference.
-Rapache runs an R interpreter inside the Apache web server and makes it possible and very efficient to build scalable R-based APIs and web services.
-
-A more recent related work is the \pkg{plumber} ([https://www.rplumber.io/](https://www.rplumber.io/)) package, which makes it even easier for R to be used to host web services through RESTful APIs.
+The backbone of this work is the \pkg{plumber} ([https://www.rplumber.io/](https://www.rplumber.io/)) package.
+The plumber package allows R to be used to host web services through RESTful APIs.
 When R is running on a server, a different server can ask R to make a computation through an HTTP request and get an immediate response.
@@
 On the client side, it provides simple shells to the Docker, and to Kubernetes' `kubectl` and `helm` tools, that are needed for managing Kubernetes clusters.
-The package simplifies complex infrastructure management to a number of R funcction calls, given an Azure account with sufficient credit\footnots{See _"Deploying a prediction service with Plumber"_ vignette at \href{https://cran.r-project.org/web/packages/AzureContainers/vignettes/vig01_plumber_deploy.html}{https://cran.r-project.org/web/packages/AzureContainers/vignettes/vig01_plumber_deploy.html} for details.}.
+The package simplifies complex infrastructure management to a number of R funcction calls, given an Azure account with sufficient credit\footnote{See \emph{"Deploying a prediction service with Plumber"} vignette for details:  \href{https://cran.r-project.org/web/packages/AzureContainers/vignettes/vig01_plumber_deploy.html}{https://cran.r-project.org/web/packages/AzureContainers/vignettes/vig01\_plumber\_deploy.html}.}.
 

### 2020-01-18 manuscript.Rmd
update title
one sentence per line

--- 2020-01-18 manuscript.Rmd
@@
     email: me@nanx.me
+  - name: Gergely Daroczi
+    # https://github.com/daroczig
+    address:
+      - \orcid{0000-0003-3149-8537}
+    email: daroczig@rapporter.net
 abstract: >
@@
 
-R could also empower webserver since at least 2007, when [RApache](http://www.rapache.net) was presented at the useR!
-conference.
+R could also empower webserver since at least 2007, when [RApache](http://www.rapache.net) was presented at the useR! conference.
 Rapache runs an R interpreter inside the Apache web server, which makes it possible and very efficient to build scalable R-based APIs and web services.
@@
 Such RApache applications have been used at [rapporter.net](http://rapporter.net), transforming parametrized R Markdown-like documents into interactive web applications for reporting.
-A similar design was implemented at CARD.com, where the optimization of the acquisition flow and the real-time analysis of debit card transactions both relied on an efficient combination of Rserve and RApache running in Docker containers: the R session behind Rserve acted as a read-only, exteremely fast and scalabale, in-memory database for the many concurrent RApache processes responding to the live-scoring requests of other components of the company.
+A similar design was implemented at CARD.com, where the optimization of the acquisition flow and the real-time analysis of debit card transactions both relied on an efficient combination of Rserve and RApache running in Docker containers:
+the R session behind Rserve acted as a read-only, exteremely fast and scalabale, in-memory database for the many concurrent RApache processes responding to the live-scoring requests of other components of the company.
 Similar dockerized R scripts were responsible for the ETL processes and even the client-facing email, text message and push notification alerts sent in real-time on the card transaction.
-The related Docker images were made available at [https://github.com/cardcorp/card-rocker](https://github.com/cardcorp/card-rocker), which are extending `rocker/r-base` eg with an SSH client and workaround for being able to mount SSH keys from the host, Pandoc, the AWS SDK and Java -- which latter was required at that time to be able to easily use the Amazon Web Services via their Java SDK, eg to run R consumers reading from a Kinesis stream, deployed as a dockerized ECS service.
+The related Docker images were made available at [https://github.com/cardcorp/card-rocker](https://github.com/cardcorp/card-rocker), which are extending `rocker/r-base`, e.g., with an SSH client and workaround for being able to mount SSH keys from the host, Pandoc, the AWS SDK and Java -- which latter was required at that time to be able to easily use the Amazon Web Services via their Java SDK, eg to run R consumers reading from a Kinesis stream, deployed as a dockerized ECS service.
 The main learnings from using R in Docker was the imporance of not only pinning the R package versions via MRAN, but also moving away from Debian testing to an LTS distribution to minimize the risk of breaking the automated builds of the Docker images and/or production jobs.

### 2020-01-18 manuscript.Rmd
small update following up #62

--- 2020-01-18 manuscript.Rmd
@@
 However, thanks to advancements in R running as a web service, along with with the ability to deploy R in Docker containers, modern enterprises are now capable of having real-time machine learning powered by R.
+One of the first implementations of having R respond to client reqests were TCP/IP and local socket servers, such as \CRANpkg{Rserve} and \CRANpkg{svSocket}, which allowed other local or remote processes to make binary requests to interact with R without the need to link to R at compile time and connected other programming languages with R.
+R could also empower web services since at least 2007, when [rApache](http://www.rapache.net) was presented at the useR! conference \citep{horner_rapache_2013}.
+<!-- website seems down, https://web.archive.org/web/20190627101203/http://rapache.net/manual.html -->
+rApache runs an R interpreter inside the Apache web server, which makes it possible and very efficient to build scalable R-based APIs and web services.
 
-The backbone of this work is the \pkg{plumber} ([https://www.rplumber.io/](https://www.rplumber.io/)) package.
-The plumber package allows R to be used to host web services through RESTful APIs.
+Such rApache applications have been used at [rapporter.net](http://rapporter.net), transforming parametrized R Markdown-like documents into interactive web applications for reporting.
+A similar design was implemented at [CARD.com](https://www.card.com/), where the optimization of the acquisition flow and the real-time analysis of debit card transactions both relied on an efficient combination of Rserve and rApache running in Docker containers:
+the R session behind Rserve acted as a read-only, exteremely fast and scalabale, in-memory database for the many concurrent rApache processes responding to the live-scoring requests of various divisions of the company.
+Similar dockerized R scripts were responsible for the ETL processes and even the client-facing email, text message and push notification alerts sent in real-time based on card transactions.
+The related Docker images were made available at [https://github.com/cardcorp/card-rocker](https://github.com/cardcorp/card-rocker).
+The images extend `rocker/r-base`, e.g., with an SSH client and a workaround for being able to mount SSH keys from the host, Pandoc, the Amazon Web Services (AWS) SDK, and Java, which is required by the AWS SDK.
+The AWS SDK allowed to run R consumers reading from real-time data processing streams of [AWS Kinesis](https://aws.amazon.com/kinesis/).
+The applications were deployed on Amazon Elastic Container Service ([ECS](https://aws.amazon.com/ecs/)).
+The main learnings from using R in Docker was the imporance of not only pinning the R package versions via MRAN, but also moving away from Debian testing to an LTS distribution to minimize the risk of breaking the automated builds of the Docker images and production jobs.
+
+A more recent related work is the \pkg{plumber} ([https://www.rplumber.io/](https://www.rplumber.io/)) package, which makes it even easier for R to be used to host web services through RESTful APIs.
 When R is running on a server, a different server can ask R to make a computation through an HTTP request and get an immediate response.

### 2020-01-18 manuscript.Rmd
fix author order and add @daroczig to contribution section
some housekeeping

@@
     email: me@nanx.me
-  - name: Gergely Daroczi
-    # https://github.com/daroczig
-    address:
-      - \orcid{0000-0003-3149-8537}
-    email: daroczig@rapporter.net
 abstract: >
@@
 The Rocker&nbsp;Project provides several lines of containers spanning to from building blocks with `R-release` or `R-devel`, via containers with [RStudio&nbsp;Server](https://rstudio.com/products/rstudio/) and [Shiny&nbsp;Server](https://rstudio.com/products/shiny/shiny-server/), to domain-specific containers such as [`rocker/geospatial`](https://github.com/rocker-org/geospatial) \citep{rocker_geospatial_2019}.
-These containers form _image stacks_, building on top of each other for better maintainability (i.e. smaller `Dockerfile`s), composability, and to reduce build time.
+These containers form _image stacks_, building on top of each other for better maintainability (i.e., smaller `Dockerfile`s), composability, and to reduce build time.
 Also of note is a series of "versioned" containers which match the R release they contain with the _then-current_ set of packages via the MRAN Snapshot views of CRAN \citep{microsoft_cran_2019}.
@@
 Most notably these capabilities are the MRAN repository, which is enabled by default, and the (optional) integration with [Intel® Math Kernel Library](https://software.intel.com/en-us/mkl) (MKL) for multi-threading in linear algebra operations \citep{microsoft_multithread}.
-MRO does not provide official Docker images, but a set of community-maintained `Dockerfile`s and Docker images are provided by the poject `mro-docker` ([https://github.com/nuest/mro-docker](https://github.com/nuest/mro-docker)), e.g.
-the image [`nuest/mro`](https://hub.docker.com/repository/docker/nuest/mro/).
+MRO does not provide official Docker images, but a set of community-maintained `Dockerfile`s and Docker images are provided by the poject `mro-docker` ([https://github.com/nuest/mro-docker](https://github.com/nuest/mro-docker)), e.g., the image [`nuest/mro`](https://hub.docker.com/repository/docker/nuest/mro/).
 The images are inspired by the Rocker images and can be used much in the same fashion, effectively a drop-in replacement allowing users to quickly evaluate if the benefits of MRO + Intel® MKL apply to their use case.
@@
 It is developed by Oracle, connects R to the GraalVM ecosystem \citep{wikipedia_graalvm_2019}, and also claims superior performance but also targets full compatibility with base R \citep{oracle_labs_oraclefastr_2020}.
-There are no offical Docker images provided, but indepedently maintained experimental images and `Dockerfile`s are provided by [nuest/fastr-docker](https://github.com/nuest/fastr-docker), e.g.
-[`nuest/fastr` on Docker Hub](https://cloud.docker.com/repository/docker/nuest/fastr/).
+There are no offical Docker images provided, but indepedently maintained experimental images and `Dockerfile`s are provided by [nuest/fastr-docker](https://github.com/nuest/fastr-docker), e.g., [`nuest/fastr` on Docker Hub](https://cloud.docker.com/repository/docker/nuest/fastr/).
 
@@
 R-hub's proof of concept [`r-minimal`](https://github.com/r-hub/r-minimal) takes this to the extreme with compressed image size of 20MB and bespoke install scripts, e.g., to cleverly remove compilers after package installation.
-For R though, this advantage quickly deteriorates if more features are needed, e.g.
-if adding compilers, packages, or especially if installing powerful data science and communication tools, like RStudio or LaTeX.
+For R though, this advantage quickly deteriorates if more features are needed, e.g., if adding compilers, packages, or especially if installing powerful data science and communication tools, like RStudio or LaTeX.
 Furthermore the distribution uses the [`musl` C library](https://www.musl-libc.org/) instead of the `glibc` used in Debian/Ubuntu, for which more tooling and experiences exist\footnote{Cf.
@@
 
-_Data Science_ is a widely discussed topic among all academic disciplines \citep[cf. e.g.,][]{donoho_50_2017}.
+_Data Science_ is a widely discussed topic among all academic disciplines \citep[e.g.,][]{donoho_50_2017}.
 Using computers and complex software to make sense of (large amounts of) collected or simulated data is a cross-cutting phenomenon leading to _X Data Science_ disciplines, seminars, and study majors.
@@
   "cloud", "Deploy or manage an instances in the cloud",
-  "interact", "Interact with an instance (e.g. file transfer)",
+  "interact", "Interact with an instance (e.g., file transfer)",
   "images", "Manage storage of images",
@@
 Due to its original use case (see&nbsp;\nameref{introduction}), Docker has no standard mechanisms for chaining containers together; it lacks definitions and protocols for environment variables, volume mounts and/or ports that could enable transfer of input (parameters and data) and output (results) to and from containers.
-Some packages, e.g.
-\pkg{containerit}, do provide Docker images that can be used similarly to CLIs, but their usage is cumbersome\footnote{\href{https://o2r.info/containerit/articles/container.html}{https://o2r.info/containerit/articles/container.html}}.
+Some packages, e.g., \pkg{containerit}, do provide Docker images that can be used similarly to CLIs, but their usage is cumbersome\footnote{\href{https://o2r.info/containerit/articles/container.html}{https://o2r.info/containerit/articles/container.html}}.
 
@@
 That can be a challenge for some users.
-The R-hub service (see Section&nbsp;\nameref{rhub}) makes it easy to ensure that no errors occur, but to fix errors a local setup is still often warranted, e.g.
-using the image `rocker/r-devel`, and to test packages with native code, which can make the process more complex \citep[cf.][]{eckert_building_2018}.
+The R-hub service (see Section&nbsp;\nameref{rhub}) makes it easy to ensure that no errors occur, but to fix errors a local setup is still often warranted, e.g., using the image `rocker/r-devel`, and to test packages with native code, which can make the process more complex \citep[cf.][]{eckert_building_2018}.
 The R-hub Docker images can also be used to debug problems locally using various combinations of Linux platforms, R versions, and compilers\footnote{See \href{https://r-hub.github.io/rhub/articles/local-debugging.html}{https://r-hub.github.io/rhub/articles/local-debugging.html} and \href{https://blog.r-hub.io/2019/04/25/r-devel-linux-x86-64-debian-clang/}{https://blog.r-hub.io/2019/04/25/r-devel-linux-x86-64-debian-clang/}}.
@@
 It remains to be seen which approach to control Docker, via the Docker API as `stevedore` or via system calls as `dockyard`/`docker`/`dockr`, is more sustainable, or if the question will be answered by the endurance of maintainers and sufficient funding.
-Similarly, the capturing of environments and their serialization in form of a `Dockerfile` currently happens at different levels of abstraction and re-use of functionality seems reasonable, e.g.
-`liftr` could generate the environment with `containerit`, which in turn may use `dockerfiler` for low level objects representing a `Dockerfile`.
+Similarly, the capturing of environments and their serialization in form of a `Dockerfile` currently happens at different levels of abstraction and re-use of functionality seems reasonable, e.g., `liftr` could generate the environment with `containerit`, which in turn may use `dockerfiler` for low level objects representing a `Dockerfile`.
 In this consolidation, the Rocker&nbsp;Project could play the role of coordinating entity.
-Though for the moment, the sign of the times points to more experimentation and feature growth, e.g. images for GPU-based computing and artificial intelligence.
+Though for the moment, the sign of the times points to more experimentation and feature growth, e.g., images for GPU-based computing and artificial intelligence.
 Even with coding being more and more accepted as a required, and achievable skill, an easier access, for example by exposing containerisation benefits via simple user interfaces in the users' IDE, could be an important next step.
@@
 It should be noted that the dominant position of Docker is a blessing and a curse for these goals.
-It could be wise to start experimenting with non-Docker containerisation tools now, e.g.
-R packages interfacing with other container engines such as [podman/buildah](https://github.com/containers/libpod) or [rkt](https://coreos.com/rkt/), or an R package for creating Singularity image recipe files.
+It could be wise to start experimenting with non-Docker containerisation tools now, e.g., R packages interfacing with other container engines such as [podman/buildah](https://github.com/containers/libpod) or [rkt](https://coreos.com/rkt/), or an R package for creating Singularity image recipe files.
 Such efforts help to avoid lock-in and to design sustainable workflows based on concepts of _containerisation_, not on their implementation in Docker.
@@
 NX wrote the section on \pkg{liftr}.
-All authors approved the final version. This articles was collaboratively written at \href{https://github.com/nuest/rockerverse-paper/}{https://github.com/nuest/rockerverse-paper/}.
+All authors approved the final version.
+This articles was collaboratively written at \href{https://github.com/nuest/rockerverse-paper/}{https://github.com/nuest/rockerverse-paper/}.
 The \href{https://github.com/nuest/rockerverse-paper/graphs/contributors}{contributors page}, \href{https://github.com/nuest/rockerverse-paper/commits/master}{commit history}, and \href{https://github.com/nuest/rockerverse-paper/issues/}{discussion issues} provide a detailed view on the respective contributions.

### 2020-01-18 manuscript.Rmd
Added Whole Tale description and journal reproducibility audit use case

--- 2020-01-18 manuscript.Rmd
@@
     email: nitesh.turaga@roswellpark.org
+  - name: Craig Willis
+    # https://github.com/craig-willis
+    affiliation: University of Illinois at Urbana-Champaign
+    address:
+      - 501 E. Daniel St.
+      - Champaign, IL 61820, USA
+      - \orcid{0000-0002-6148-7196}
+    email: willis8@illinois.edu
   - name: Nan Xiao
@@
 A minor reason in the inception of the images were also project constraints requiring a Rocker-incompatible licensing of the `Dockerfile`s, i.e., the MIT License.
+
+**Whole Tale** (https://wholetale.org) is a National Science Foundation (NSF) funded project\footnote{NSF Award OAC-1541450} developing a scalable, open-source, multi-user platform for reproducible research \citep{brinckman2019, chard2019a}.
+A central goal of the platform is to enable researchers to easily create and publish executable research objects\footnote{In Whole Tale a `tale` is a research object that contains metadata, data (by copy or reference), code, narrative, documentation, provenance, and information about the computational environment to support computational reproducibility} associated with published research \citep{chard2019b}.
+For R users, Whole Tale attempts to combine the strengths of the Rocker Project's curated Docker images with the Jupyter Project's `repo2docker`\footnote{See \href{https://repo2docker.readthedocs.io/}{https://repo2docker.readthedocs.io/}} component to simplify the customization of R-based environments for researchers with limited experience with either Docker or Git.
+Using Whole Tale, researchers can create and publish Rocker-based reproducible research objects to a growing number of repositories including DataONE member nodes, Zenodo and soon Dataverse.
+Additionally, Whole Tale supports automatic data citation and is working on capabilities for image preservation and provenance capture to improve the transparency of publish computational research artifacts \citep{mecum2018, mcphillips2019}.
 
@@
 
+## Publication Reproducibility Audits
+
+A recommendation of the recent National Academies' report on _Reproducibility and Replicability in Science_ is that journals "consider ways to ensure computational reproducibility for publications that make claims based on computations" \citep{NASEM2019}.
+In fields such as Political Science and Economics, journals are increasingly adopting policies requiring authors to publish the code and data required to reproduce computational findings reported in published manuscripts, subject to independent verification \citep{Jacoby2017,Vilhuber2019,Alvarez2018,Christian2018,Eubank2016,King1995}.
+Problems with the computational environment, installation and availability of software dependencies are common.
+R is gaining popularity in these communities (in a sample of 105 replication packages published by the _American Journal of Political Science_ (AJPS) over 65% use R).
+Projects such as NSF Whole Tale use the Rocker Project community images with the goal of improving the reproducibility of published research artifacts and simplifying the publication and verification process for both authors and journals by reducing errors and time spent specifying the environment.
+
 ## Development and debugging
@@
 LS \& NT wrote the section on Bioconductor.
+CW wrote the sections on Whole Tale and contributed the publication reproducibility audit use case.
 NX wrote the section on \pkg{liftr}.
@@
 The funders had no role in data collection and analysis, decision to publish, or preparation of the manuscript.
+CW is supported by the Whole Tale projects (https://wholetale.org) funded by the US National Science Foundation (NSF) under award [OAC-1541450](https://www.nsf.gov/awardsearch/showAward?AWD_ID=1541450).
 

### 2020-01-20 manuscript.Rmd
add README.Rmd and move sticker there

--- 2020-01-18 manuscript.Rmd
@@
 We conclude with a reflection on the landscape of packages and applications and point out future directions of development.
-
-```{r hexSticker, message=TRUE, warning=FALSE, cache=TRUE, include=FALSE}
-if(!file.exists("rockerverse.png")) {
-  library("magick")
-  library("here")
-
-  # https://commons.wikimedia.org/wiki/File:NASA-HS201427a-HubbleUltraDeepField2014-20140603.jpg
-  universe <- magick::image_read("https://upload.wikimedia.org/wikipedia/commons/6/69/NASA-HS201427a-HubbleUltraDeepField2014-20140603.jpg")
-  magick::image_write(universe, here("universe.png"))
-  # rockerLogoMask <-
-  magick::image_read("https://avatars2.githubusercontent.com/u/9100160?s=200&v=4") %>%
-    magick::image_scale("2100x2100") %>%
-    magick::image_channel("Matte") %>%
-    image_negate() %>%
-    magick::image_border("black", "150x0") %>%
-    magick::image_write(here("rocker_mask.png"))
-
-  # does not work: image_composite(universe, rockerLogoMask, operator = "CopyOpacity", compose_args = "-composite -alpha Off")
-  system2("convert", c(here("universe.png"), here("rocker_mask.png"),
-    "-alpha Off",
-    "-compose CopyOpacity",
-    "-composite",
-    here("whaleverse.png")))
-
-  library("hexSticker")
-  hexSticker::sticker(here("whaleverse.png"), package = "Rockerverse",
-          p_size = 20, s_x = 0.97, s_y = 0.95, s_width = 1,
-          h_fill = "#000000", h_color = "#ffffff",
-          url = "github.com/nuest/rockerverse-paper",
-          u_color = "#ffffff", u_size = 4,
-          filename = here("rockerverse.png"))
-
-  file.remove(c(here("universe.png"), here("whaleverse.png"), here("rocker_mask.png")))
-}
-```
-
-![Rockerverse hex sticker](`r here::here("rockerverse.png")`)
 

### 2020-01-20 manuscript.Rmd
language improvements and clarifications
Co-authored-by: Dirk Eddelbuettel <edd@debian.org>

@@
   These use cases and the variety of applications demonstrate the power of Rocker specifically and containerisation in general.
-  We identified common themes across this diversity: reproducible environments, scalability and effectivity, and portability across clouds.
+  We identified common themes across this diversity: reproducible environments, scalability and efficiency, and portability across clouds.
 # blank footnote? https://tex.stackexchange.com/questions/170511/footnotes-without-numbering
@@
 The article continues with a brief introduction of containerization basics and the Rocker&nbsp;Project, followed by descriptions of projects developing images with R installations.
-Then, we present applications, starting with the R packages specificly for interacting with Docker, second-level packages using containers mediately or only for specific features, up to complex use cases leveraging containers.
+Then, we present applications, starting with the R packages specificly for interacting with Docker, second-level packages using containers indirectly or only for specific features, up to complex use cases leveraging containers.
 We conclude with a reflection on the landscape of packages and applications and point out future directions of development.
@@
 It provided to be a natural match for "cloud deployment" which runs, or at least appears to run, "seamlessly" without much explicit reference to the underlying machine, architecture or operating system: containers are portable and can be deployed with very little in terms of dependencies on the host system---only the container runtime is required.
-Images are normally build from plain text documents called `Dockerfile`.
+Images are normally built from plain text documents called `Dockerfile`.
 A `Dockerfile` has a specific set of instructions to create and document a well-defined environment, i.e., install specific software and expose specific ports.
@@
 
-## Images for alternative R distributions
+## Images for alternative R engines
 \label{alternatives}
@@
 The large number of extension packages provides access to an unrivaled variety of established and upcoming features.
-Nevertheless, special use cases and experimental projects in academia and industry exist to test new approaches or provide features different to what _GNU-R_, or Base R, provides.
-GNU-R is the R distribution maintained by the [R Core Team](https://www.r-project.org/contributors.html) and available via [CRAN](https://cran.r-project.org/).
+Nevertheless, special use cases and experimental projects in academia and industry exist to test new approaches, or provide features different to what the R implementation maintained by the [R Core Team](https://www.r-project.org/contributors.html) and available via [CRAN](https://cran.r-project.org/) provides.
 While the images presented in this section are far from being as vetted, stable, and widely used as any of the Rocker images, they demonstrate an important advantage of containerisation technology, namely the abiliy to transparently build portable stacks of open source software and make them easily accessible to others.
-All these distributions of R are published under GPL licenses.
-Since all of the R distributions claim better performance as a core motivation, a comparision based on Docker images, potentially leveraging the [resource restriction mechanisms](https://docs.docker.com/config/containers/resource_constraints/) of Docker to level the playing field, seems useful future work.
+All these alternative implementations of R are published under GPL licenses.
+Since all implementations of R claim better performance as a core motivation, a comparision based on Docker images, potentially leveraging the [resource restriction mechanisms](https://docs.docker.com/config/containers/resource_constraints/) of Docker to level the playing field, seems useful future work.
 
@@
 MRO is compatible with main R and its packages and it "[..] includes additional capabilities for improved performance, reproducibility, and platform support." \citep{microsoft_mro}
-Most notably these capabilities are the MRAN repository, which is enabled by default, and the (optional) integration with [Intel® Math Kernel Library](https://software.intel.com/en-us/mkl) (MKL) for multi-threading in linear algebra operations \citep{microsoft_multithread}.
+Most notably these capabilities are the MRAN repository, which is enabled by default, and some bundled packages such as \CRANpkg{checkpoint} and specific R packages maintained by Microsoft\footnote{See "Bundled Packages" on \href{https://mran.microsoft.com/rro/installed}{https://mran.microsoft.com/rro/installed}.}.
+Originally the optional integration with [Intel® Math Kernel Library](https://software.intel.com/en-us/mkl) (MKL) for multi-threading in linear algebra operations \citep{microsoft_multithread} was a distinguishing feature before it was made available publicly, and can be used in base R, too \citep{eddelbuettel_thinking_2018}.
+
 MRO does not provide official Docker images, but a set of community-maintained `Dockerfile`s and Docker images are provided by the poject `mro-docker` ([https://github.com/nuest/mro-docker](https://github.com/nuest/mro-docker)), e.g., the image [`nuest/mro`](https://hub.docker.com/repository/docker/nuest/mro/).
@@
 **Renjin** is an interpreter for the R language running on top of the [Java Virtual Machine](https://en.wikipedia.org/wiki/Java_virtual_machine) (JVM) providing full two-way access between Java and R code \citep{wikipedia_renjin_2018}.
-It was developed to combine the benefits of R, such scripting and extension packages, with the JVM's advantages in the areas of security, cross-platform availability, and established position in enterprise settings.
+It was developed to combine the benefits of R, such scripting and extension packages, with the JVM's advantages in the areas of security, cross-platform availability and integration into enterprise platforms.
 R extension packages need to be specially compiled and are distributed via the Java package manager [Apache Maven](https://en.wikipedia.org/wiki/Apache_Maven), cf.
@@
 The project consists of 1741 R software packages as of August 15th 2019, as well as packages containing annotation or experiment data.
-_Bioconductor_ has a semiannual release cycle, each release is associated with a particular version of R.
+_Bioconductor_ has a semi-annual release cycle, each release is associated with a particular version of R.
 Docker images allow availability of current and past versions of _Bioconductor_ for convenience and reproducibility.
@@
 All community maintained images build on top of the _Bioconductor_ base image and therefore indirectly the Rocker images.
-To simplify building and maintaining _Bioconductor_ images, we use a Ruby templating engine.
+To simplify building and maintaining _Bioconductor_ images a Ruby templating engine is used for the `Dockerfile`s.
 A recent audit of the _Bioconductor_ `Dockerfile`s, following best practices from the Docker website, led to a reduction in the size and number of layers.
@@
 New R versions are only added on demand.
-The explorative rocker-win project demonstrates the variety and pontentially confusing complexity of Windows on Docker, e.g., because of the different base images, but also provides a starting point for any Windows-constrained user to leverage containerisation, including examples for apps using \pkg{plumber\, \pkg{Shiny}, or R packages with system dependencies such as \pkg{sf}.
-The latter is quite comfortable and fast actually, compared to installation from source on Linux, as R ships pre-compiled binaries for Windows.
+The explorative rocker-win project demonstrates the variety of Windows on Docker, which is potentially confusing, e.g., because of the different base images.
+The proof of concept provides a starting point for any Windows-constrained user to leverage containerisation, including examples for apps using \CRANpkg{plumber}, \CRANpkg{Shiny}, or R packages with system dependencies such as \CRANpkg{sf}.
+The latter is quite comfortable and fast actually, compared to the default installation from source on Linux, because CRAN ships pre-compiled binaries for Windows.
 However, compared to the various Linux images, the footprint of Windows images is quite large.
@@
 In most cases, individual developers or researchers worrying about reproducibility, will prefer the more widely and host independent Linux-based containers.
-Naturally, a Windows license is required for the host machine, and the licensing impacts the potnetial to build images in clouds or redistribute exported images as archive files.
+Naturally, a Windows license is required for the host machine, and the licensing impacts the potential to build images in clouds or redistribute exported images as archive files.
 But the developments in the `rocker-win` prototype show that containers can also be used for R workflows depending on Windows-only tools, for leveraging an existing Windows Server-based infrastructure where policies can otherwise not be met, or for streamlining interactions with system operating staff that runs Windows-based containers themselves.
@@
 The **R-hub** project provides _"a collection of services to help R package development"_, with the package builder as the most prominent one\citep{r-hub_docs_2019}.
-The builder allows R package developers to check their R package on different platforms and R versions using a web form or the package \pkg{rhub} \citep{csardi_rhub_2019}.
+The builder allows R package developers to check their R package on different platforms and R versions using a web form or the package \CRANpkg{rhub} \citep{csardi_rhub_2019}.
 The builder uses Docker containers to conduct these checks, taking advantages of their well-defined environments and sandboxing.
@@
 **Kaggle** provides the [`gcr.io/kaggle-images/rstats`](https://hub.docker.com/r/kaggle/rstats) image (previously `kaggle/rstats`) and [corresponding `Dockerfile`](https://github.com/Kaggle/docker-rstats) for usage in their Machine Learning competitions and easy access to the associated datasets.
-It includes machine learling libraries such as Tensorflow and Keras, and also configures the \pkg{reticulate} package.
+It includes machine learling libraries such as Tensorflow and Keras, and also configures the \CRANpkg{reticulate} package.
 The image uses a base image with _all packages from CRAN_, `gcr.io/kaggle-images/rcran`, which requires a Google Cloud Build as Docker&nbsp;Hub would time out\footnote{Originally, a stacked collection of over 20 images with automated builds on Docker~Hub was used, see \href{https://web.archive.org/web/20190606043353/http://blog.kaggle.com/2016/02/05/how-to-get-started-with-data-science-in-containers/}{https://web.archive.org/web/20190606043353/http://blog.kaggle.com/2016/02/05/how-to-get-started-with-data-science-in-containers/} and \href{https://hub.docker.com/r/kaggle/rcran/dockerfile}{https://hub.docker.com/r/kaggle/rcran/dockerfile}}.
@@
 
-As a further example, [**Radiant** project](https://radiant-rstats.github.io/docs/) provides several images, e.g., [`vnijs/rsm-msba-spark`](https://hub.docker.com/r/vnijs/rsm-msba-spark), for their browser-based business analytics interface based on \pkg{Shiny} (`Dockerfile` [on GitHub](https://github.com/radiant-rstats/docker)) and for use in education as part of an MSc course.
+As a further example, [**Radiant** project](https://radiant-rstats.github.io/docs/) provides several images, e.g., [`vnijs/rsm-msba-spark`](https://hub.docker.com/r/vnijs/rsm-msba-spark), for their browser-based business analytics interface based on \CRANpkg{Shiny} (`Dockerfile` [on GitHub](https://github.com/radiant-rstats/docker)) and for use in education as part of an MSc course.
 As data science often applies a multitude of tools, this image favours inclusion over selection and features Python, Postgres, JupyterLab and Visual Studio Code besides R and RStudio, bringing the image size up to 9GB.
@@
 Each of these packages has particular advantages as they provide function wrappers for interacting with the Docker CLI at different stages of a container's life cycle.
-Examples of such interactions are installing the Docker software, creating `Dockerfile`s (\pkg{dockerfiler}), \pkg{containerit}), building images and launching a containers (\pkg{stevedore}, \pkg{docker}) on a local machine or on the cloud.
+Examples of such interactions are installing the Docker software, creating `Dockerfile`s (\CRANpkg{dockerfiler}, \pkg{containerit}), building images and launching a containers (\CRANpkg{stevedore}, \pkg{docker}) on a local machine or on the cloud.
 As such, the choice of which package is most useful depends on the use-case at hand, but also the users level of expertise.
@@
 **\pkg{harbor}** ([https://github.com/wch/harbor](https://github.com/wch/harbor) is not actively maintained anymore, but should be honorably mentioned as the first R package for managing Docker images and containers.
-It uses the \pkg{sys} package to run system commands against the Docker CLI, both locally and through an ssh connection, and has convenience functions, e.g., for listing and removing containers/images and for accessing logs.
+It uses the \CRANpkg{sys} package to run system commands against the Docker CLI, both locally and through an ssh connection, and has convenience functions, e.g., for listing and removing containers/images and for accessing logs.
 The output of container executions are converted to appropriate R types.
@@
 
-**\pkg{stevedore}** ([https://cran.r-project.org/package=stevedore](https://cran.r-project.org/package=stevedore)) is currently the most powerful Docker client in R.
-It interfaces with the Docker daemon over the Docker HTTP API\footnote{} via a Unix socket on Linux or MacOS, over a named pipe on Windows, or over an HTTP/TCP connection.
+**\CRANpkg{stevedore}** ([https://cran.r-project.org/package=stevedore](https://cran.r-project.org/package=stevedore)) is currently the most powerful Docker client in R.
+It interfaces with the Docker daemon over the Docker HTTP API\footnote{\href{https://docs.docker.com/engine/api/latest/}{https://docs.docker.com/engine/api/latest/}} via a Unix socket on Linux or MacOS, over a named pipe on Windows, or over an HTTP/TCP connection.
 The package is the only one not using system calls to the `docker` CLI for managing images and containers and easily exposes connections to remote Docker daemons, which has to be configured on the Docker level otherwise.
 Using the API gives access to more information and is system independent and likely more reliable than parsing command line output.
-\pkg{stevedore}'s own interface is automatically generated based on the OpenAPI specification of the Docker daemon, but still similar to the Docker CLI.
+\CRANpkg{stevedore}'s own interface is automatically generated based on the OpenAPI specification of the Docker daemon, but still similar to the Docker CLI.
 The interface is similar to R6 objects, in that a `stevedore_object` has a number of functions attached to it that can be called, and multiple specific versions of the Docker API can be supported thanks to the automatic generation\footnote{See \href{https://github.com/richfitz/stevedore/tree/master/inst/spec}{https://github.com/richfitz/stevedore/tree/master/inst/spec} and \href{https://github.com/richfitz/stevedore/blob/master/development.md}{https://github.com/richfitz/stevedore/blob/master/development.md}.}.
 
-**\pkg{AzureContainers}** ([https://cran.r-project.org/package=AzureContainers](https://cran.r-project.org/package=AzureContainers)) is an interface to a number of container-related services in Microsoft's [Azure Cloud](https://azure.microsoft.com/), see Section&nbsp;\nameref{azurecontainers}.
+**\CRANpkg{AzureContainers}** is an interface to a number of container-related services in Microsoft's [Azure Cloud](https://azure.microsoft.com/), see Section&nbsp;\nameref{azurecontainers}.
 While it is mainly intended for working with Azure, as a convenience feature it includes lightweight, cross-platform shells to Docker and Kubernetes (tools `kubectl` and `helm`).
@@
 
-**\pkg{babelwhale}** ([https://cran.r-project.org/package=babelwhale](https://cran.r-project.org/package=babelwhale)) 
-allows executing and interacting with containers, which can use either Docker or Singularity as a backend.
+**\CRANpkg{babelwhale}** allows executing and interacting with containers, which can use either Docker or Singularity as a backend \citep{cannoodt_babelwhale_2019}.
 The package provides a unified interface to interact with Docker and Singularity containers.
@@
 
-**\pkg{renv}** ([https://rstudio.github.io/renv/](https://rstudio.github.io/renv/)) with Docker: https://rstudio.github.io/renv/articles/docker.html
+**\CRANpkg{renv}** ([https://rstudio.github.io/renv/](https://rstudio.github.io/renv/)) with Docker: https://rstudio.github.io/renv/articles/docker.html
 
@@
 
-**\pkg{liftr}** ([https://nanx.me/liftr/](https://nanx.me/liftr/)) \citep{liftr2019} aims to solve the problem of persistent reproducible reporting in statistical computing.
+**\CRANpkg{liftr}** ([https://nanx.me/liftr/](https://nanx.me/liftr/)) \citep{liftr2019} aims to solve the problem of persistent reproducible reporting in statistical computing.
 Currently, the R Markdown format and its backend compilation engine _knitr_ offer a _de facto_ standard for creating dynamic documents \citep{xie2018}.
@@
 
-A more recent related work is the \pkg{plumber} ([https://www.rplumber.io/](https://www.rplumber.io/)) package, which makes it even easier for R to be used to host web services through RESTful APIs.
+A more recent related work is the \CRANpkg{plumber} ([https://www.rplumber.io/](https://www.rplumber.io/)) package, which makes it even easier for R to be used to host web services through RESTful APIs.
 When R is running on a server, a different server can ask R to make a computation through an HTTP request and get an immediate response.
@@
 \label{azurecontainers}
-**\pkg{AzureContainers}** is an umbrealla package providing interfaces the three services of Microsoft's Azure Cloud, namely [Container Instances](https://azure.microsoft.com/en-us/services/container-instances/) for running individual containers, [Container Registry](https://azure.microsoft.com/en-us/services/container-registry/) for private image distribution, and [Kubernetes Service](https://azure.microsoft.com/en-us/services/kubernetes-service/) for orchestrated deployments.
-While a package like \pkg{plumber} provides the infrastructure for turning an R workflow a service, for production purposes it is usually necessary to take into account scalability, reliability and ease of management.
+**\CRANpkg{AzureContainers}** is an umbrealla package providing interfaces the three services of Microsoft's Azure Cloud, namely [Container Instances](https://azure.microsoft.com/en-us/services/container-instances/) for running individual containers, [Container Registry](https://azure.microsoft.com/en-us/services/container-registry/) for private image distribution, and [Kubernetes Service](https://azure.microsoft.com/en-us/services/kubernetes-service/) for orchestrated deployments.
+While a package like \CRANpkg{plumber} provides the infrastructure for turning an R workflow a service, for production purposes it is usually necessary to take into account scalability, reliability and ease of management.
 A single container is limited in the volume of requests it can service, and if the container goes down or is occupied with processing a computationally intensive task, the service becomes unavailable.
@@
 A cluster features load-balancing, autoscaling of containers across numerous servers (in the cloud or on premise), and restarting failed ones.
-\pkg{AzureContainers} provides a lightweight yet powerful interface for working with Resource Manager, the framework for deploying and managing arbitrary resources in Azure.
+AzureContainers provides a lightweight yet powerful interface for working with Resource Manager, the framework for deploying and managing arbitrary resources in Azure.
 On the client side, it provides simple shells to the Docker, and to Kubernetes' `kubectl` and `helm` tools, that are needed for managing Kubernetes clusters.
@@
 - https://github.com/benmarwick/rrtools
-- \pkg{renv}, https://rstudio.github.io/renv/articles/docker.html
+- \CRANpkg{renv}, https://rstudio.github.io/renv/articles/docker.html
 - what to copy in and what not
@@
 These problems are hard to debug and critical, both because when they occur they lead to fatal crashing processes.
-[`rocker/r-devel-san`](https://github.com/rocker-org/r-devel-san) and [`rocker/r-devel-ubsan-clang`](https://github.com/rocker-org/r-devel-san-clang) are Docker images have a particularly configured version of R to trace such problems with gcc and clang compilers, respectively \citep[cf.~\pkg{sanitizers} for examples,][]{eddelbuettel_sanitizers_2014}.
+[`rocker/r-devel-san`](https://github.com/rocker-org/r-devel-san) and [`rocker/r-devel-ubsan-clang`](https://github.com/rocker-org/r-devel-san-clang) are Docker images have a particularly configured version of R to trace such problems with gcc and clang compilers, respectively \citep[cf.~\CRANpkg{sanitizers} for examples,][]{eddelbuettel_sanitizers_2014}.
 The image [`wch/r-debug`](https://github.com/wch/r-debug) is a purpose built Docker image with _multiple_ instrumented builds of R, each with a different diagnostic utility activated.
@@
 Docker became synonymous with applying the concept of containerisation to solve challenges of reproducible environments, e.g., in research and in development \& production, and of scalable deployments with the ability to move processing between machines easily (e.g., locally, one cloud providers VM, another cloud provider's Container-as-a-Service).
-Collaboration, reproducibility, and portability are the common themes behind R packages, use cases, and applications in this work.
+Reproducible environments, scalability and efficiency, and portability across clouds are the common themes behind R packages, use cases, and applications in this work.
 

### 2020-01-20 manuscript.Rmd
small clarification, see #62
Co-authored-by: Gergely Daroczi (@daroczig) <daroczig@rapporter.net>

@@
 A similar design was implemented at [CARD.com](https://www.card.com/), where the optimization of the acquisition flow and the real-time analysis of debit card transactions both relied on an efficient combination of Rserve and rApache running in Docker containers:
-the R session behind Rserve acted as a read-only, exteremely fast and scalabale, in-memory database for the many concurrent rApache processes responding to the live-scoring requests of various divisions of the company.
+the R session behind Rserve acted as a read-only, extremely fast and scalabale, in-memory database for the many concurrent rApache processes responding to the live-scoring requests of various divisions of the company.
 Similar dockerized R scripts were responsible for the ETL processes and even the client-facing email, text message and push notification alerts sent in real-time based on card transactions.
@@
 The applications were deployed on Amazon Elastic Container Service ([ECS](https://aws.amazon.com/ecs/)).
-The main learnings from using R in Docker was the imporance of not only pinning the R package versions via MRAN, but also moving away from Debian testing to an LTS distribution to minimize the risk of breaking the automated builds of the Docker images and production jobs.
+The main learnings from using R in Docker was the imporance of not only pinning the R package versions via MRAN, but also moving away from Debian testing to an LTS distribution.
+For the use case at hand, the priority was to have more control over upstream updates, and to minimize the risk of breaking the automated builds of the Docker images and production jobs.
 

### 2020-01-20 manuscript.Rmd
small text update
Co-authored-by: Dirk Eddelbuettel <edd@debian.org>

@@
 
-## Non-Debian Linux images
+## Non-Rocker Linux images
 \label{nondebian}
@@
 _Data Science_ is a widely discussed topic among all academic disciplines \citep[e.g.,][]{donoho_50_2017}.
-Using computers and complex software to make sense of (large amounts of) collected or simulated data is a cross-cutting phenomenon leading to _X Data Science_ disciplines, seminars, and study majors.
-The practice to apply a combination tools and software stacks for analysing data and the cross-cutting skillsets connected with data science puts the management of computing environments and the challenges of reproducibility high up on the agenda.
-This awareness leads to a large number of bespoke Docker images but also some commonly used image stacks, ranging from base images with very common tools to 'catch all' images including _a lot_ of software.
-Here we only mention a few image stacks as examples, which are not based on Rocker images or stem from the R community, but prominently include installations of R.
-A general description of what to consider when creating a Docker image used for data science projects in R is provided on [environments.rstudio.com](https://environments.rstudio.com/docker).
+The discussions shed a light on the tools and craftspersonship behind the analysis of data with computational methods.
+The practice of Data Science often involves a combination tools and software stacks and requires a cross-cutting skillset.
+This complexity and an inherent concern for openness and reproducibility in the Data Science community lead to Docker being used widely.
+<!--A general description of what to consider when creating a Docker image used for data science projects in R is provided on [environments.rstudio.com](https://environments.rstudio.com/docker).-->
+This section presents exemplary Docker images and image stacks featuring R intended for Data Science.
 
@@
 **Kaggle** provides the [`gcr.io/kaggle-images/rstats`](https://hub.docker.com/r/kaggle/rstats) image (previously `kaggle/rstats`) and [corresponding `Dockerfile`](https://github.com/Kaggle/docker-rstats) for usage in their Machine Learning competitions and easy access to the associated datasets.
-It includes machine learling libraries such as Tensorflow and Keras, and also configures the \CRANpkg{reticulate} package.
+It includes machine learning libraries such as Tensorflow and Keras (see also image `rocker/ml` in Section&nbsp;\nameref{rocker-gpu}), and also configures the \CRANpkg{reticulate} package.
 The image uses a base image with _all packages from CRAN_, `gcr.io/kaggle-images/rcran`, which requires a Google Cloud Build as Docker&nbsp;Hub would time out\footnote{Originally, a stacked collection of over 20 images with automated builds on Docker~Hub was used, see \href{https://web.archive.org/web/20190606043353/http://blog.kaggle.com/2016/02/05/how-to-get-started-with-data-science-in-containers/}{https://web.archive.org/web/20190606043353/http://blog.kaggle.com/2016/02/05/how-to-get-started-with-data-science-in-containers/} and \href{https://hub.docker.com/r/kaggle/rcran/dockerfile}{https://hub.docker.com/r/kaggle/rcran/dockerfile}}.
@@
 
+\label{rocker-gpu}
 Using **GPUs** (graphical processing units) as a specialised hardware from containerized common work environments is also possible and useful \citep{haydel_enhancing_2015}.

### 2020-01-20 manuscript.Rmd

@@
 
-- researchcompendia.science
-- https://github.com/benmarwick/rrtools
-- \CRANpkg{renv}, https://rstudio.github.io/renv/articles/docker.html
-- what to copy in and what not
+Containers provide a high degree of isolation that is often desirable when attempting to capture a specific computational environment so that others can reproduce and extend a research result.
+Many computationally intensive research projects depend on specific versions of original and third-party software packages in diverse languages, joined together to form a pipeline through which data flows.
+New releases of even just a single piece of software in this pipeline can break the entire workflow, making it difficult to find the error and for others to reuse existing pipelines, and making the original the results irreproducible and not extandable.
+The chance of a substantial disruption like this is high in a multi-year research project where key pieces of third-party software may have several major updates over the duration of the project.
+The classical "paper" article is insufficient to adequately communicate the knowledge behind such research projects \citep[cf.][]{donoho_invitation_2010,marwick_how_2015}.
+\citet{gentleman_statistical_2007} coined the term _Research Compendium_ for a dynamic document together with supporting data and code.
+They use the R package system \cite{core_writing_1999} as for the functional prototype to structuring, validation, and distribution of research compendia.
+This concept has been taken up and extended\footnote{See full literature list at \href{https://research-compendium.science/}{https://research-compendium.science/}.}, not the least by applying containerisation and other methods for managing computing environments---see Section&nbsp;\nameref{envs}.
+Containers give the researcher an isolated environment to assemble these research pipelines with specific versions of software.
+Research workflows in containers are safe from contamination from other activities occuring on the researcher's computer, for example the installation of newest version of packages for teaching demonstrations.
+Given the users in this scenario, i.e., often academics with limited formal software development training, templates and assistance with containers around research compendia is essential.
+
+In many fields we see that a typical unit of research for a container is a research report or journal article, where the container holds the compendium, or self-contained set of data (or connections to data elsewhere) and code files needed to fully reproduce the article \citep{marwick_packaging_2018}.
+The package \pkg{rrtools} ([https://github.com/benmarwick/rrtools](https://github.com/benmarwick/rrtools)) provides a template and convienence functions to apply good practices for research compendia, including a starter `Dockerfile`.
+Images of compendium containers can be hosted on services such as Docker&nbsp;Hub for convienient sharing among collaborators and others. 
+Similarly, packages such as \pkg{containerit} and \pkg{dockerfiler} can be used to manage the `Dockerfile` to be archived with a compendium on a data repository (e.g.
+[Zenodo](https://zenodo.org/), [Dataverse](https://dataverse.org/), [Figshare](https://figshare.com/), [OSF](https://osf.io/)).
+A typial compendium's `Dockerfile` will pull a rocker image fixed to a specific version of R, and install R packages from the MRAN repository to ensure the package versions are tied to a specific date, rather than the most recent version.
+A future researchers can download then compendium from the repository and run the included `Dockerfile` build a new image that recreates the computational environment used to produce the original research results.
+If building the image fails, the human-readable instructions in a `Dockerfile` are a the starting point for rebuilding the environment.
+Further safeguarding practices are currently under development or not part of common practice yet, such as preservation of images \citep{emsley_framework_2018} and storing them alongside `Dockerfile`s \citep[cf.][]{nust_opening_2017}, or pinning of system libraries beyond the tagged base images, which may be seen as stable or dynamic depending on the applied time scale \citep[see discussion on `debian:testing` base image in][]{RJ-2017-065}.
 

### 2020-01-20 manuscript.Rmd
one sentence per line

--- 2020-01-20 manuscript.Rmd
@@
   These use cases and the variety of applications demonstrate the power of Rocker specifically and containerisation in general.
-  We identified common themes across this diversity: reproducible environments, scalability and efficiency, and portability across clouds.
+  We identified common themes across this diversity: reproducible environments, scalability and effectivity, and portability across clouds.
 # blank footnote? https://tex.stackexchange.com/questions/170511/footnotes-without-numbering
@@
 The article continues with a brief introduction of containerization basics and the Rocker&nbsp;Project, followed by descriptions of projects developing images with R installations.
-Then, we present applications, starting with the R packages specificly for interacting with Docker, second-level packages using containers indirectly or only for specific features, up to complex use cases leveraging containers.
+Then, we present applications, starting with the R packages specificly for interacting with Docker, second-level packages using containers mediately or only for specific features, up to complex use cases leveraging containers.
 We conclude with a reflection on the landscape of packages and applications and point out future directions of development.
+
+```{r hexSticker, message=TRUE, warning=FALSE, cache=TRUE, include=FALSE}
+if(!file.exists("rockerverse.png")) {
+  library("magick")
+  library("here")
+
+  # https://commons.wikimedia.org/wiki/File:NASA-HS201427a-HubbleUltraDeepField2014-20140603.jpg
+  universe <- magick::image_read("https://upload.wikimedia.org/wikipedia/commons/6/69/NASA-HS201427a-HubbleUltraDeepField2014-20140603.jpg")
+  magick::image_write(universe, here("universe.png"))
+  # rockerLogoMask <-
+  magick::image_read("https://avatars2.githubusercontent.com/u/9100160?s=200&v=4") %>%
+    magick::image_scale("2100x2100") %>%
+    magick::image_channel("Matte") %>%
+    image_negate() %>%
+    magick::image_border("black", "150x0") %>%
+    magick::image_write(here("rocker_mask.png"))
+
+  # does not work: image_composite(universe, rockerLogoMask, operator = "CopyOpacity", compose_args = "-composite -alpha Off")
+  system2("convert", c(here("universe.png"), here("rocker_mask.png"),
+    "-alpha Off",
+    "-compose CopyOpacity",
+    "-composite",
+    here("whaleverse.png")))
+
+  library("hexSticker")
+  hexSticker::sticker(here("whaleverse.png"), package = "Rockerverse",
+          p_size = 20, s_x = 0.97, s_y = 0.95, s_width = 1,
+          h_fill = "#000000", h_color = "#ffffff",
+          url = "github.com/nuest/rockerverse-paper",
+          u_color = "#ffffff", u_size = 4,
+          filename = here("rockerverse.png"))
+
+  file.remove(c(here("universe.png"), here("whaleverse.png"), here("rocker_mask.png")))
+}
+```
+
+![Rockerverse hex sticker](`r here::here("rockerverse.png")`)
 
@@
 It provided to be a natural match for "cloud deployment" which runs, or at least appears to run, "seamlessly" without much explicit reference to the underlying machine, architecture or operating system: containers are portable and can be deployed with very little in terms of dependencies on the host system---only the container runtime is required.
-Images are normally built from plain text documents called `Dockerfile`.
+Images are normally build from plain text documents called `Dockerfile`.
 A `Dockerfile` has a specific set of instructions to create and document a well-defined environment, i.e., install specific software and expose specific ports.
@@
 The Rocker&nbsp;Project provides several lines of containers spanning to from building blocks with `R-release` or `R-devel`, via containers with [RStudio&nbsp;Server](https://rstudio.com/products/rstudio/) and [Shiny&nbsp;Server](https://rstudio.com/products/shiny/shiny-server/), to domain-specific containers such as [`rocker/geospatial`](https://github.com/rocker-org/geospatial) \citep{rocker_geospatial_2019}.
-These containers form _image stacks_, building on top of each other for better maintainability (i.e., smaller `Dockerfile`s), composability, and to reduce build time.
+These containers form _image stacks_, building on top of each other for better maintainability (i.e. smaller `Dockerfile`s), composability, and to reduce build time.
 Also of note is a series of "versioned" containers which match the R release they contain with the _then-current_ set of packages via the MRAN Snapshot views of CRAN \citep{microsoft_cran_2019}.
@@
 
-## Images for alternative R engines
+## Images for alternative R distributions
 \label{alternatives}
@@
 The large number of extension packages provides access to an unrivaled variety of established and upcoming features.
-Nevertheless, special use cases and experimental projects in academia and industry exist to test new approaches, or provide features different to what the R implementation maintained by the [R Core Team](https://www.r-project.org/contributors.html) and available via [CRAN](https://cran.r-project.org/) provides.
+Nevertheless, special use cases and experimental projects in academia and industry exist to test new approaches or provide features different to what _GNU-R_, or Base R, provides.
+GNU-R is the R distribution maintained by the [R Core Team](https://www.r-project.org/contributors.html) and available via [CRAN](https://cran.r-project.org/).
 While the images presented in this section are far from being as vetted, stable, and widely used as any of the Rocker images, they demonstrate an important advantage of containerisation technology, namely the abiliy to transparently build portable stacks of open source software and make them easily accessible to others.
-All these alternative implementations of R are published under GPL licenses.
-Since all implementations of R claim better performance as a core motivation, a comparision based on Docker images, potentially leveraging the [resource restriction mechanisms](https://docs.docker.com/config/containers/resource_constraints/) of Docker to level the playing field, seems useful future work.
+All these distributions of R are published under GPL licenses.
+Since all of the R distributions claim better performance as a core motivation, a comparision based on Docker images, potentially leveraging the [resource restriction mechanisms](https://docs.docker.com/config/containers/resource_constraints/) of Docker to level the playing field, seems useful future work.
 
@@
 MRO is compatible with main R and its packages and it "[..] includes additional capabilities for improved performance, reproducibility, and platform support." \citep{microsoft_mro}
-Most notably these capabilities are the MRAN repository, which is enabled by default, and some bundled packages such as \CRANpkg{checkpoint} and specific R packages maintained by Microsoft\footnote{See "Bundled Packages" on \href{https://mran.microsoft.com/rro/installed}{https://mran.microsoft.com/rro/installed}.}.
-Originally the optional integration with [Intel® Math Kernel Library](https://software.intel.com/en-us/mkl) (MKL) for multi-threading in linear algebra operations \citep{microsoft_multithread} was a distinguishing feature before it was made available publicly, and can be used in base R, too \citep{eddelbuettel_thinking_2018}.
-
-MRO does not provide official Docker images, but a set of community-maintained `Dockerfile`s and Docker images are provided by the poject `mro-docker` ([https://github.com/nuest/mro-docker](https://github.com/nuest/mro-docker)), e.g., the image [`nuest/mro`](https://hub.docker.com/repository/docker/nuest/mro/).
+Most notably these capabilities are the MRAN repository, which is enabled by default, and the (optional) integration with [Intel® Math Kernel Library](https://software.intel.com/en-us/mkl) (MKL) for multi-threading in linear algebra operations \citep{microsoft_multithread}.
+MRO does not provide official Docker images, but a set of community-maintained `Dockerfile`s and Docker images are provided by the poject `mro-docker` ([https://github.com/nuest/mro-docker](https://github.com/nuest/mro-docker)), e.g.
+the image [`nuest/mro`](https://hub.docker.com/repository/docker/nuest/mro/).
 The images are inspired by the Rocker images and can be used much in the same fashion, effectively a drop-in replacement allowing users to quickly evaluate if the benefits of MRO + Intel® MKL apply to their use case.
@@
 **Renjin** is an interpreter for the R language running on top of the [Java Virtual Machine](https://en.wikipedia.org/wiki/Java_virtual_machine) (JVM) providing full two-way access between Java and R code \citep{wikipedia_renjin_2018}.
-It was developed to combine the benefits of R, such scripting and extension packages, with the JVM's advantages in the areas of security, cross-platform availability and integration into enterprise platforms.
+It was developed to combine the benefits of R, such scripting and extension packages, with the JVM's advantages in the areas of security, cross-platform availability, and established position in enterprise settings.
 R extension packages need to be specially compiled and are distributed via the Java package manager [Apache Maven](https://en.wikipedia.org/wiki/Apache_Maven), cf.
@@
 It is developed by Oracle, connects R to the GraalVM ecosystem \citep{wikipedia_graalvm_2019}, and also claims superior performance but also targets full compatibility with base R \citep{oracle_labs_oraclefastr_2020}.
-There are no offical Docker images provided, but indepedently maintained experimental images and `Dockerfile`s are provided by [nuest/fastr-docker](https://github.com/nuest/fastr-docker), e.g., [`nuest/fastr` on Docker Hub](https://cloud.docker.com/repository/docker/nuest/fastr/).
+There are no offical Docker images provided, but indepedently maintained experimental images and `Dockerfile`s are provided by [nuest/fastr-docker](https://github.com/nuest/fastr-docker), e.g.
+[`nuest/fastr` on Docker Hub](https://cloud.docker.com/repository/docker/nuest/fastr/).
 
@@
 The project consists of 1741 R software packages as of August 15th 2019, as well as packages containing annotation or experiment data.
-_Bioconductor_ has a semi-annual release cycle, each release is associated with a particular version of R.
+_Bioconductor_ has a semiannual release cycle, each release is associated with a particular version of R.
 Docker images allow availability of current and past versions of _Bioconductor_ for convenience and reproducibility.
@@
 All community maintained images build on top of the _Bioconductor_ base image and therefore indirectly the Rocker images.
-To simplify building and maintaining _Bioconductor_ images a Ruby templating engine is used for the `Dockerfile`s.
+To simplify building and maintaining _Bioconductor_ images, we use a Ruby templating engine.
 A recent audit of the _Bioconductor_ `Dockerfile`s, following best practices from the Docker website, led to a reduction in the size and number of layers.
@@
 New R versions are only added on demand.
-The explorative rocker-win project demonstrates the variety of Windows on Docker, which is potentially confusing, e.g., because of the different base images.
-The proof of concept provides a starting point for any Windows-constrained user to leverage containerisation, including examples for apps using \CRANpkg{plumber}, \CRANpkg{Shiny}, or R packages with system dependencies such as \CRANpkg{sf}.
-The latter is quite comfortable and fast actually, compared to the default installation from source on Linux, because CRAN ships pre-compiled binaries for Windows.
+The explorative rocker-win project demonstrates the variety and pontentially confusing complexity of Windows on Docker, e.g., because of the different base images, but also provides a starting point for any Windows-constrained user to leverage containerisation, including examples for apps using \pkg{plumber\, \pkg{Shiny}, or R packages with system dependencies such as \pkg{sf}.
+The latter is quite comfortable and fast actually, compared to installation from source on Linux, as R ships pre-compiled binaries for Windows.
 However, compared to the various Linux images, the footprint of Windows images is quite large.
@@
 In most cases, individual developers or researchers worrying about reproducibility, will prefer the more widely and host independent Linux-based containers.
-Naturally, a Windows license is required for the host machine, and the licensing impacts the potential to build images in clouds or redistribute exported images as archive files.
+Naturally, a Windows license is required for the host machine, and the licensing impacts the potnetial to build images in clouds or redistribute exported images as archive files.
 But the developments in the `rocker-win` prototype show that containers can also be used for R workflows depending on Windows-only tools, for leveraging an existing Windows Server-based infrastructure where policies can otherwise not be met, or for streamlining interactions with system operating staff that runs Windows-based containers themselves.
@@
 The **R-hub** project provides _"a collection of services to help R package development"_, with the package builder as the most prominent one\citep{r-hub_docs_2019}.
-The builder allows R package developers to check their R package on different platforms and R versions using a web form or the package \CRANpkg{rhub} \citep{csardi_rhub_2019}.
+The builder allows R package developers to check their R package on different platforms and R versions using a web form or the package \pkg{rhub} \citep{csardi_rhub_2019}.
 The builder uses Docker containers to conduct these checks, taking advantages of their well-defined environments and sandboxing.
@@
 R-hub's proof of concept [`r-minimal`](https://github.com/r-hub/r-minimal) takes this to the extreme with compressed image size of 20MB and bespoke install scripts, e.g., to cleverly remove compilers after package installation.
-For R though, this advantage quickly deteriorates if more features are needed, e.g., if adding compilers, packages, or especially if installing powerful data science and communication tools, like RStudio or LaTeX.
+For R though, this advantage quickly deteriorates if more features are needed, e.g.
+if adding compilers, packages, or especially if installing powerful data science and communication tools, like RStudio or LaTeX.
 Furthermore the distribution uses the [`musl` C library](https://www.musl-libc.org/) instead of the `glibc` used in Debian/Ubuntu, for which more tooling and experiences exist\footnote{Cf.
@@
 
-_Data Science_ is a widely discussed topic among all academic disciplines \citep[e.g.,][]{donoho_50_2017}.
+_Data Science_ is a widely discussed topic among all academic disciplines \citep[cf. e.g.,][]{donoho_50_2017}.
 Using computers and complex software to make sense of (large amounts of) collected or simulated data is a cross-cutting phenomenon leading to _X Data Science_ disciplines, seminars, and study majors.
@@
 **Kaggle** provides the [`gcr.io/kaggle-images/rstats`](https://hub.docker.com/r/kaggle/rstats) image (previously `kaggle/rstats`) and [corresponding `Dockerfile`](https://github.com/Kaggle/docker-rstats) for usage in their Machine Learning competitions and easy access to the associated datasets.
-It includes machine learling libraries such as Tensorflow and Keras, and also configures the \CRANpkg{reticulate} package.
+It includes machine learling libraries such as Tensorflow and Keras, and also configures the \pkg{reticulate} package.
 The image uses a base image with _all packages from CRAN_, `gcr.io/kaggle-images/rcran`, which requires a Google Cloud Build as Docker&nbsp;Hub would time out\footnote{Originally, a stacked collection of over 20 images with automated builds on Docker~Hub was used, see \href{https://web.archive.org/web/20190606043353/http://blog.kaggle.com/2016/02/05/how-to-get-started-with-data-science-in-containers/}{https://web.archive.org/web/20190606043353/http://blog.kaggle.com/2016/02/05/how-to-get-started-with-data-science-in-containers/} and \href{https://hub.docker.com/r/kaggle/rcran/dockerfile}{https://hub.docker.com/r/kaggle/rcran/dockerfile}}.
@@
 
-As a further example, [**Radiant** project](https://radiant-rstats.github.io/docs/) provides several images, e.g., [`vnijs/rsm-msba-spark`](https://hub.docker.com/r/vnijs/rsm-msba-spark), for their browser-based business analytics interface based on \CRANpkg{Shiny} (`Dockerfile` [on GitHub](https://github.com/radiant-rstats/docker)) and for use in education as part of an MSc course.
+As a further example, [**Radiant** project](https://radiant-rstats.github.io/docs/) provides several images, e.g., [`vnijs/rsm-msba-spark`](https://hub.docker.com/r/vnijs/rsm-msba-spark), for their browser-based business analytics interface based on \pkg{Shiny} (`Dockerfile` [on GitHub](https://github.com/radiant-rstats/docker)) and for use in education as part of an MSc course.
 As data science often applies a multitude of tools, this image favours inclusion over selection and features Python, Postgres, JupyterLab and Visual Studio Code besides R and RStudio, bringing the image size up to 9GB.
@@
 
+**Whole Tale** (https://wholetale.org) is a National Science Foundation (NSF) funded project\footnote{NSF Award OAC-1541450} developing a scalable, open-source, multi-user platform for reproducible research \citep{brinckman2019, chard2019a}.
+A central goal of the platform is to enable researchers to easily create and publish executable research objects\footnote{In Whole Tale a `tale` is a research object that contains metadata, data (by copy or reference), code, narrative, documentation, provenance, and information about the computational environment to support computational reproducibility} associated with published research \citep{chard2019b}.
+For R users, Whole Tale attempts to combine the strengths of the Rocker Project's curated Docker images with the Jupyter Project's `repo2docker`\footnote{See \href{https://repo2docker.readthedocs.io/}{https://repo2docker.readthedocs.io/}} component to simplify the customization of R-based environments for researchers with limited experience with either Docker or Git.
+Using Whole Tale, researchers can create and publish Rocker-based reproducible research objects to a growing number of repositories including DataONE member nodes, Zenodo and soon Dataverse.
+Additionally, Whole Tale supports automatic data citation and is working on capabilities for image preservation and provenance capture to improve the transparency of publish computational research artifacts \citep{mecum2018, mcphillips2019}.
+
 # Use cases and applications
@@
 Each of these packages has particular advantages as they provide function wrappers for interacting with the Docker CLI at different stages of a container's life cycle.
-Examples of such interactions are installing the Docker software, creating `Dockerfile`s (\CRANpkg{dockerfiler}, \pkg{containerit}), building images and launching a containers (\CRANpkg{stevedore}, \pkg{docker}) on a local machine or on the cloud.
+Examples of such interactions are installing the Docker software, creating `Dockerfile`s (\pkg{dockerfiler}), \pkg{containerit}), building images and launching a containers (\pkg{stevedore}, \pkg{docker}) on a local machine or on the cloud.
 As such, the choice of which package is most useful depends on the use-case at hand, but also the users level of expertise.
@@
   "cloud", "Deploy or manage an instances in the cloud",
-  "interact", "Interact with an instance (e.g., file transfer)",
+  "interact", "Interact with an instance (e.g. file transfer)",
   "images", "Manage storage of images",
@@
 **\pkg{harbor}** ([https://github.com/wch/harbor](https://github.com/wch/harbor) is not actively maintained anymore, but should be honorably mentioned as the first R package for managing Docker images and containers.
-It uses the \CRANpkg{sys} package to run system commands against the Docker CLI, both locally and through an ssh connection, and has convenience functions, e.g., for listing and removing containers/images and for accessing logs.
+It uses the \pkg{sys} package to run system commands against the Docker CLI, both locally and through an ssh connection, and has convenience functions, e.g., for listing and removing containers/images and for accessing logs.
 The output of container executions are converted to appropriate R types.
@@
 
-**\CRANpkg{stevedore}** ([https://cran.r-project.org/package=stevedore](https://cran.r-project.org/package=stevedore)) is currently the most powerful Docker client in R.
-It interfaces with the Docker daemon over the Docker HTTP API\footnote{\href{https://docs.docker.com/engine/api/latest/}{https://docs.docker.com/engine/api/latest/}} via a Unix socket on Linux or MacOS, over a named pipe on Windows, or over an HTTP/TCP connection.
+**\pkg{stevedore}** ([https://cran.r-project.org/package=stevedore](https://cran.r-project.org/package=stevedore)) is currently the most powerful Docker client in R.
+It interfaces with the Docker daemon over the Docker HTTP API\footnote{} via a Unix socket on Linux or MacOS, over a named pipe on Windows, or over an HTTP/TCP connection.
 The package is the only one not using system calls to the `docker` CLI for managing images and containers and easily exposes connections to remote Docker daemons, which has to be configured on the Docker level otherwise.
 Using the API gives access to more information and is system independent and likely more reliable than parsing command line output.
-\CRANpkg{stevedore}'s own interface is automatically generated based on the OpenAPI specification of the Docker daemon, but still similar to the Docker CLI.
+\pkg{stevedore}'s own interface is automatically generated based on the OpenAPI specification of the Docker daemon, but still similar to the Docker CLI.
 The interface is similar to R6 objects, in that a `stevedore_object` has a number of functions attached to it that can be called, and multiple specific versions of the Docker API can be supported thanks to the automatic generation\footnote{See \href{https://github.com/richfitz/stevedore/tree/master/inst/spec}{https://github.com/richfitz/stevedore/tree/master/inst/spec} and \href{https://github.com/richfitz/stevedore/blob/master/development.md}{https://github.com/richfitz/stevedore/blob/master/development.md}.}.
 
-**\CRANpkg{AzureContainers}** is an interface to a number of container-related services in Microsoft's [Azure Cloud](https://azure.microsoft.com/), see Section&nbsp;\nameref{azurecontainers}.
+**\pkg{AzureContainers}** ([https://cran.r-project.org/package=AzureContainers](https://cran.r-project.org/package=AzureContainers)) is an interface to a number of container-related services in Microsoft's [Azure Cloud](https://azure.microsoft.com/), see Section&nbsp;\nameref{azurecontainers}.
 While it is mainly intended for working with Azure, as a convenience feature it includes lightweight, cross-platform shells to Docker and Kubernetes (tools `kubectl` and `helm`).
@@
 
-**\CRANpkg{babelwhale}** allows executing and interacting with containers, which can use either Docker or Singularity as a backend \citep{cannoodt_babelwhale_2019}.
+**\pkg{babelwhale}** ([https://cran.r-project.org/package=babelwhale](https://cran.r-project.org/package=babelwhale)) 
+allows executing and interacting with containers, which can use either Docker or Singularity as a backend.
 The package provides a unified interface to interact with Docker and Singularity containers.
@@
 
-**\CRANpkg{renv}** ([https://rstudio.github.io/renv/](https://rstudio.github.io/renv/)) with Docker: https://rstudio.github.io/renv/articles/docker.html
+**\pkg{renv}** ([https://rstudio.github.io/renv/](https://rstudio.github.io/renv/)) with Docker: https://rstudio.github.io/renv/articles/docker.html
 
@@
 
-**\CRANpkg{liftr}** ([https://nanx.me/liftr/](https://nanx.me/liftr/)) \citep{liftr2019} aims to solve the problem of persistent reproducible reporting in statistical computing.
+**\pkg{liftr}** ([https://nanx.me/liftr/](https://nanx.me/liftr/)) \citep{liftr2019} aims to solve the problem of persistent reproducible reporting in statistical computing.
 Currently, the R Markdown format and its backend compilation engine _knitr_ offer a _de facto_ standard for creating dynamic documents \citep{xie2018}.
@@
 A similar design was implemented at [CARD.com](https://www.card.com/), where the optimization of the acquisition flow and the real-time analysis of debit card transactions both relied on an efficient combination of Rserve and rApache running in Docker containers:
-the R session behind Rserve acted as a read-only, extremely fast and scalabale, in-memory database for the many concurrent rApache processes responding to the live-scoring requests of various divisions of the company.
+the R session behind Rserve acted as a read-only, exteremely fast and scalabale, in-memory database for the many concurrent rApache processes responding to the live-scoring requests of various divisions of the company.
 Similar dockerized R scripts were responsible for the ETL processes and even the client-facing email, text message and push notification alerts sent in real-time based on card transactions.
@@
 The applications were deployed on Amazon Elastic Container Service ([ECS](https://aws.amazon.com/ecs/)).
-The main learnings from using R in Docker was the imporance of not only pinning the R package versions via MRAN, but also moving away from Debian testing to an LTS distribution.
-For the use case at hand, the priority was to have more control over upstream updates, and to minimize the risk of breaking the automated builds of the Docker images and production jobs.
+The main learnings from using R in Docker was the imporance of not only pinning the R package versions via MRAN, but also moving away from Debian testing to an LTS distribution to minimize the risk of breaking the automated builds of the Docker images and production jobs.
 
-A more recent related work is the \CRANpkg{plumber} ([https://www.rplumber.io/](https://www.rplumber.io/)) package, which makes it even easier for R to be used to host web services through RESTful APIs.
+A more recent related work is the \pkg{plumber} ([https://www.rplumber.io/](https://www.rplumber.io/)) package, which makes it even easier for R to be used to host web services through RESTful APIs.
 When R is running on a server, a different server can ask R to make a computation through an HTTP request and get an immediate response.
@@
 \label{azurecontainers}
-**\CRANpkg{AzureContainers}** is an umbrealla package providing interfaces the three services of Microsoft's Azure Cloud, namely [Container Instances](https://azure.microsoft.com/en-us/services/container-instances/) for running individual containers, [Container Registry](https://azure.microsoft.com/en-us/services/container-registry/) for private image distribution, and [Kubernetes Service](https://azure.microsoft.com/en-us/services/kubernetes-service/) for orchestrated deployments.
-While a package like \CRANpkg{plumber} provides the infrastructure for turning an R workflow a service, for production purposes it is usually necessary to take into account scalability, reliability and ease of management.
+**\pkg{AzureContainers}** is an umbrealla package providing interfaces the three services of Microsoft's Azure Cloud, namely [Container Instances](https://azure.microsoft.com/en-us/services/container-instances/) for running individual containers, [Container Registry](https://azure.microsoft.com/en-us/services/container-registry/) for private image distribution, and [Kubernetes Service](https://azure.microsoft.com/en-us/services/kubernetes-service/) for orchestrated deployments.
+While a package like \pkg{plumber} provides the infrastructure for turning an R workflow a service, for production purposes it is usually necessary to take into account scalability, reliability and ease of management.
 A single container is limited in the volume of requests it can service, and if the container goes down or is occupied with processing a computationally intensive task, the service becomes unavailable.
@@
 A cluster features load-balancing, autoscaling of containers across numerous servers (in the cloud or on premise), and restarting failed ones.
-AzureContainers provides a lightweight yet powerful interface for working with Resource Manager, the framework for deploying and managing arbitrary resources in Azure.
+\pkg{AzureContainers} provides a lightweight yet powerful interface for working with Resource Manager, the framework for deploying and managing arbitrary resources in Azure.
 On the client side, it provides simple shells to the Docker, and to Kubernetes' `kubectl` and `helm` tools, that are needed for managing Kubernetes clusters.
@@
 Due to its original use case (see&nbsp;\nameref{introduction}), Docker has no standard mechanisms for chaining containers together; it lacks definitions and protocols for environment variables, volume mounts and/or ports that could enable transfer of input (parameters and data) and output (results) to and from containers.
-Some packages, e.g., \pkg{containerit}, do provide Docker images that can be used similarly to CLIs, but their usage is cumbersome\footnote{\href{https://o2r.info/containerit/articles/container.html}{https://o2r.info/containerit/articles/container.html}}.
+Some packages, e.g.
+\pkg{containerit}, do provide Docker images that can be used similarly to CLIs, but their usage is cumbersome\footnote{\href{https://o2r.info/containerit/articles/container.html}{https://o2r.info/containerit/articles/container.html}}.
 
@@
 - https://github.com/benmarwick/rrtools
-- \CRANpkg{renv}, https://rstudio.github.io/renv/articles/docker.html
+- \pkg{renv}, https://rstudio.github.io/renv/articles/docker.html
 - what to copy in and what not
+
+## Publication Reproducibility Audits
+
+A recommendation of the recent National Academies' report on _Reproducibility and Replicability in Science_ is that journals "consider ways to ensure computational reproducibility for publications that make claims based on computations" \citep{NASEM2019}.
+In fields such as Political Science and Economics, journals are increasingly adopting policies requiring authors to publish the code and data required to reproduce computational findings reported in published manuscripts, subject to independent verification \citep{Jacoby2017,Vilhuber2019,Alvarez2018,Christian2018,Eubank2016,King1995}.
+Problems with the computational environment, installation and availability of software dependencies are common.
+R is gaining popularity in these communities (in a sample of 105 replication packages published by the _American Journal of Political Science_ (AJPS) over 65% use R).
+Projects such as NSF Whole Tale use the Rocker Project community images with the goal of improving the reproducibility of published research artifacts and simplifying the publication and verification process for both authors and journals by reducing errors and time spent specifying the environment.
 
@@
 These problems are hard to debug and critical, both because when they occur they lead to fatal crashing processes.
-[`rocker/r-devel-san`](https://github.com/rocker-org/r-devel-san) and [`rocker/r-devel-ubsan-clang`](https://github.com/rocker-org/r-devel-san-clang) are Docker images have a particularly configured version of R to trace such problems with gcc and clang compilers, respectively \citep[cf.~\CRANpkg{sanitizers} for examples,][]{eddelbuettel_sanitizers_2014}.
+[`rocker/r-devel-san`](https://github.com/rocker-org/r-devel-san) and [`rocker/r-devel-ubsan-clang`](https://github.com/rocker-org/r-devel-san-clang) are Docker images have a particularly configured version of R to trace such problems with gcc and clang compilers, respectively \citep[cf.~\pkg{sanitizers} for examples,][]{eddelbuettel_sanitizers_2014}.
 The image [`wch/r-debug`](https://github.com/wch/r-debug) is a purpose built Docker image with _multiple_ instrumented builds of R, each with a different diagnostic utility activated.
@@
 That can be a challenge for some users.
-The R-hub service (see Section&nbsp;\nameref{rhub}) makes it easy to ensure that no errors occur, but to fix errors a local setup is still often warranted, e.g., using the image `rocker/r-devel`, and to test packages with native code, which can make the process more complex \citep[cf.][]{eckert_building_2018}.
+The R-hub service (see Section&nbsp;\nameref{rhub}) makes it easy to ensure that no errors occur, but to fix errors a local setup is still often warranted, e.g.
+using the image `rocker/r-devel`, and to test packages with native code, which can make the process more complex \citep[cf.][]{eckert_building_2018}.
 The R-hub Docker images can also be used to debug problems locally using various combinations of Linux platforms, R versions, and compilers\footnote{See \href{https://r-hub.github.io/rhub/articles/local-debugging.html}{https://r-hub.github.io/rhub/articles/local-debugging.html} and \href{https://blog.r-hub.io/2019/04/25/r-devel-linux-x86-64-debian-clang/}{https://blog.r-hub.io/2019/04/25/r-devel-linux-x86-64-debian-clang/}}.
@@
 Docker became synonymous with applying the concept of containerisation to solve challenges of reproducible environments, e.g., in research and in development \& production, and of scalable deployments with the ability to move processing between machines easily (e.g., locally, one cloud providers VM, another cloud provider's Container-as-a-Service).
-Reproducible environments, scalability and efficiency, and portability across clouds are the common themes behind R packages, use cases, and applications in this work.
+Collaboration, reproducibility, and portability are the common themes behind R packages, use cases, and applications in this work.
 
@@
 It remains to be seen which approach to control Docker, via the Docker API as `stevedore` or via system calls as `dockyard`/`docker`/`dockr`, is more sustainable, or if the question will be answered by the endurance of maintainers and sufficient funding.
-Similarly, the capturing of environments and their serialization in form of a `Dockerfile` currently happens at different levels of abstraction and re-use of functionality seems reasonable, e.g., `liftr` could generate the environment with `containerit`, which in turn may use `dockerfiler` for low level objects representing a `Dockerfile`.
+Similarly, the capturing of environments and their serialization in form of a `Dockerfile` currently happens at different levels of abstraction and re-use of functionality seems reasonable, e.g.
+`liftr` could generate the environment with `containerit`, which in turn may use `dockerfiler` for low level objects representing a `Dockerfile`.
 In this consolidation, the Rocker&nbsp;Project could play the role of coordinating entity.
-Though for the moment, the sign of the times points to more experimentation and feature growth, e.g., images for GPU-based computing and artificial intelligence.
+Though for the moment, the sign of the times points to more experimentation and feature growth, e.g. images for GPU-based computing and artificial intelligence.
 Even with coding being more and more accepted as a required, and achievable skill, an easier access, for example by exposing containerisation benefits via simple user interfaces in the users' IDE, could be an important next step.
@@
 It should be noted that the dominant position of Docker is a blessing and a curse for these goals.
-It could be wise to start experimenting with non-Docker containerisation tools now, e.g., R packages interfacing with other container engines such as [podman/buildah](https://github.com/containers/libpod) or [rkt](https://coreos.com/rkt/), or an R package for creating Singularity image recipe files.
+It could be wise to start experimenting with non-Docker containerisation tools now, e.g.
+R packages interfacing with other container engines such as [podman/buildah](https://github.com/containers/libpod) or [rkt](https://coreos.com/rkt/), or an R package for creating Singularity image recipe files.
 Such efforts help to avoid lock-in and to design sustainable workflows based on concepts of _containerisation_, not on their implementation in Docker.
@@
 LS \& NT wrote the section on Bioconductor.
+CW wrote the sections on Whole Tale and contributed the publication reproducibility audit use case.
 NX wrote the section on \pkg{liftr}.
-All authors approved the final version.
-This articles was collaboratively written at \href{https://github.com/nuest/rockerverse-paper/}{https://github.com/nuest/rockerverse-paper/}.
+All authors approved the final version. This articles was collaboratively written at \href{https://github.com/nuest/rockerverse-paper/}{https://github.com/nuest/rockerverse-paper/}.
 The \href{https://github.com/nuest/rockerverse-paper/graphs/contributors}{contributors page}, \href{https://github.com/nuest/rockerverse-paper/commits/master}{commit history}, and \href{https://github.com/nuest/rockerverse-paper/issues/}{discussion issues} provide a detailed view on the respective contributions.

### 2020-01-20 manuscript.Rmd
minor changes to Dodona use case

@@
     email: nitesh.turaga@roswellpark.org
-  - name: Craig Willis
-    # https://github.com/craig-willis
-    affiliation: University of Illinois at Urbana-Champaign
-    address:
-      - 501 E. Daniel St.
-      - Champaign, IL 61820, USA
-      - \orcid{0000-0002-6148-7196}
-    email: willis8@illinois.edu
   - name: Nan Xiao
@@
 
-**Whole Tale** (https://wholetale.org) is a National Science Foundation (NSF) funded project\footnote{NSF Award OAC-1541450} developing a scalable, open-source, multi-user platform for reproducible research \citep{brinckman2019, chard2019a}.
-A central goal of the platform is to enable researchers to easily create and publish executable research objects\footnote{In Whole Tale a `tale` is a research object that contains metadata, data (by copy or reference), code, narrative, documentation, provenance, and information about the computational environment to support computational reproducibility} associated with published research \citep{chard2019b}.
-For R users, Whole Tale attempts to combine the strengths of the Rocker Project's curated Docker images with the Jupyter Project's `repo2docker`\footnote{See \href{https://repo2docker.readthedocs.io/}{https://repo2docker.readthedocs.io/}} component to simplify the customization of R-based environments for researchers with limited experience with either Docker or Git.
-Using Whole Tale, researchers can create and publish Rocker-based reproducible research objects to a growing number of repositories including DataONE member nodes, Zenodo and soon Dataverse.
-Additionally, Whole Tale supports automatic data citation and is working on capabilities for image preservation and provenance capture to improve the transparency of publish computational research artifacts \citep{mecum2018, mcphillips2019}.
-
 # Use cases and applications
@@
 The `rocker/tidyverse` base image ensures that users can also readily use packages for analysis, display, and reporting.
+
+Another way this idea of a common environment and the partitioning that Docker provides is used in teaching is in the **[Dodona](https://dodona.ugent.be)** project.
+Dodona is a web platform developed at Ghent University and is used to teach students basic programming skills.
+Dodona uses Docker containers to test submissions by students.
+This means that both the code testing the students' submissions and the submission itself are executed in a predictable environment, avoiding compatibility issues between the wide variety of configurations used by students.
+The containerization is also used to shield the Dodona servers from bad (or even malicious) code: memory, time and I/O limits are used to make sure students can't overload the system.
+The web application communicates with containers by sending configuration information as a JSON document over standard input. 
+Every Dodona Docker image shares a `main.sh` file that passes through this information to the actual testing framework, while setting up some error handling.
+The testing process in the Docker containers sends back the test results by writing a JSON document to its standard output channel.
+In June 2019, R support was added to Dodona using an image derived from the `rocker/r-base` image that sets up the `runner` user and and `main.sh` file expected by Dodona\footnote{\href{https://github.com/dodona-edu/docker-images/blob/master/dodona-r.dockerfile}{https://github.com/dodona-edu/docker-images/blob/master/dodona-r.dockerfile}}.
+It also installs the packages required for the testing framework and the exercises so that this doesn't have to happen every time a student's submission is evaluated.
+The actual testing of R exercises is done using a custom framework loosely based on \CRANpkg{testthat}.
+During the development of the testing framework it was found that the \CRANpkg{testthat} framework did not provide enough information to its reporter system to send back all the fields required by Dodona to render its feedback.
+Right now, multiple statistics courses are developing exercises to automate the feedback for their lab classes.
 
@@
 
-## Publication Reproducibility Audits
-
-A recommendation of the recent National Academies' report on _Reproducibility and Replicability in Science_ is that journals "consider ways to ensure computational reproducibility for publications that make claims based on computations" \citep{NASEM2019}.
-In fields such as Political Science and Economics, journals are increasingly adopting policies requiring authors to publish the code and data required to reproduce computational findings reported in published manuscripts, subject to independent verification \citep{Jacoby2017,Vilhuber2019,Alvarez2018,Christian2018,Eubank2016,King1995}.
-Problems with the computational environment, installation and availability of software dependencies are common.
-R is gaining popularity in these communities (in a sample of 105 replication packages published by the _American Journal of Political Science_ (AJPS) over 65% use R).
-Projects such as NSF Whole Tale use the Rocker Project community images with the goal of improving the reproducibility of published research artifacts and simplifying the publication and verification process for both authors and journals by reducing errors and time spent specifying the environment.
-
 ## Development and debugging
@@
 LS \& NT wrote the section on Bioconductor.
-CW wrote the sections on Whole Tale and contributed the publication reproducibility audit use case.
 NX wrote the section on \pkg{liftr}.
@@
 The funders had no role in data collection and analysis, decision to publish, or preparation of the manuscript.
-CW is supported by the Whole Tale projects (https://wholetale.org) funded by the US National Science Foundation (NSF) under award [OAC-1541450](https://www.nsf.gov/awardsearch/showAward?AWD_ID=1541450).
 

### 2020-01-20 manuscript.Rmd

--- 2020-01-20 manuscript.Rmd
@@
 
+Another way this idea of a common environment and the partitioning that Docker provides is used in teaching is in the **[Dodona](https://dodona.ugent.be)** project.
+Dodona is a web platform developed at Ghent University and is used to teach students basic programming skills.
+Dodona uses Docker containers to test submissions by students.
+This means that both the code testing the students' submissions and the submission itself are executed in a predictable environment, avoiding compatibility issues between the wide variety of configurations used by students.
+The containerization is also used to shield the Dodona servers from bad (or even malicious) code: memory, time and I/O limits are used to make sure students can't overload the system.
+The web application communicates with containers by sending configuration information as a JSON document over standard input. 
+Every Dodona Docker image shares a `main.sh` file that passes through this information to the actual testing framework, while setting up some error handling.
+The testing process in the Docker containers sends back the test results by writing a JSON document to its standard output channel.
+In June 2019, R support was added to Dodona using an image derived from the `rocker/r-base` image that sets up the `runner` user and and `main.sh` file expected by Dodona\footnote{\href{https://github.com/dodona-edu/docker-images/blob/master/dodona-r.dockerfile}{https://github.com/dodona-edu/docker-images/blob/master/dodona-r.dockerfile}}.
+It also installs the packages required for the testing framework and the exercises so that this doesn't have to happen every time a student's submission is evaluated.
+The actual testing of R exercises is done using a custom framework loosely based on \CRANpkg{testthat}.
+During the development of the testing framework it was found that the \CRANpkg{testthat} framework did not provide enough information to its reporter system to send back all the fields required by Dodona to render its feedback.
+Right now, multiple statistics courses are developing exercises to automate the feedback for their lab classes.
+
 **RCloud** ([https://rcloud.social](https://rcloud.social)) is a cloud-based platform for data analysis, visualisation and collaboration using R.

### 2020-01-21 manuscript.Rmd
small rephrasing dodona
Update Bioconductor section issue #57
@nuest, this should be the update of the Bioconductor section to include
the new bioconductor_docker images

@@
 _Bioconductor_ ([https://bioconductor.org/](https://bioconductor.org/)) is an open source, open development project for the analysis and comprehension of genomic data \citep{gentleman_bioconductor_2004}.
-The project consists of 1741 R software packages as of August 15th 2019, as well as packages containing annotation or experiment data.
+The project consists of 1823 R software packages as of October 30th 2019, as well as packages containing annotation or experiment data.
 _Bioconductor_ has a semi-annual release cycle, each release is associated with a particular version of R.
@@
 _Bioconductor_'s 'base' docker images are built on top of `rocker/r-ver` and `rocker/rstudio` images.
-_Bioconductor_ installs packages based on the R version, and therefore uses `rocker/rstudio` and `rocker/r-ver` version tagging.
-_Bioconductor_ selects the desired version of R from Rocker, adds the BiocManager CRAN package for installing appropriate versions of _Bioconductor_ packages, and creates a _Bioconductor_ docker image with an informative tag (R\_version\_Bioc\_version).
+_Bioconductor_ installs packages based on the R version and the Bioconductor version, and therefore uses Bioconductor version tagging `devel` and `RELEASE_X_Y`.
+_Bioconductor_ selects the desired version of R from Rocker, adds the BiocManager CRAN package for installing appropriate versions of _Bioconductor_ packages.
+All community maintained images are supposed to be built on top of the _Bioconductor_ base image and therefore indirectly the Rocker images.
 The images are summarized on the _Bioconductor_ web site (see [https://bioconductor.org/help/docker/](https://bioconductor.org/help/docker/)), created with `Dockerfile`s maintained on GitHub ([https://github.com/Bioconductor/bioconductor_docker](https://github.com/Bioconductor/bioconductor_docker)), and [distributed through DockerHub](https://hub.docker.com/u/bioconductor).
-Past and current combinations of R and _Bioconductor_ are therefore accessible via a specific image tag.
+Past and current combinations of R and _Bioconductor_ are therefore accessible via a specific image tag `RELEASE_X_Y`.
 A strength of this approach is that the responsibility for complex software configuration (including customized development) is shifted from the user to the experienced _Bioconductor_ core team.
@@
 
-_Bioconductor_ has several images in addition to 'base', specific to various areas of research.
-The 'core' image installs the most commonly used _Bioconductor_ packages.
-_Bioconductor_ images for proteomics, metabolomics, and flow cytometry are community maintained.
-All community maintained images build on top of the _Bioconductor_ base image and therefore indirectly the Rocker images.
-To simplify building and maintaining _Bioconductor_ images a Ruby templating engine is used for the `Dockerfile`s.
-A recent audit of the _Bioconductor_ `Dockerfile`s, following best practices from the Docker website, led to a reduction in the size and number of layers.
+A recent audit of our deprecated _Bioconductor_ `Dockerfile`s, following best practices from the Docker website, led to a reduction in the size and number of layers.
 The most important insights involve the ['union' file system used by Docker](https://docs.docker.com/storage/storagedriver/overlayfs-driver/#how-container-reads-and-writes-work-with-overlay-or-overlay2).
@@
 
-A recent innovation is to produce a `bioconductor_full` image to emulate the _Bioconductor_ nightly Linux build machine.
-The image contains the _system dependencies_ needed to install and check almost all (1730 of 1741) _Bioconductor_ software packages.
+A recent innovation is to produce a `bioconductor_docker` image to emulate the _Bioconductor_ nightly Linux build machine as closely as possible.
+This image contains the build system environment variables.
+The image contains the _system dependencies_ needed to install and check almost all (1813 out of 1823) _Bioconductor_ software packages.
 Users no longer have to manage complciated system dependencies.
@@
 Many _R_ users pursue flexible work flows tailored to particular analysis needs, rather than standardized work flows.
-The `bioconductor_full` image is well-suited to this pattern.
-`bioconductor_full` provides developers with a test environment like _Bioconductor_'s build system.
+The `bioconductor_docker` image is well-suited to this pattern.
+`bioconductor_docker` provides developers with a test environment like _Bioconductor_'s build system.
 
@@
 The containerization is also used to shield the Dodona servers from bad (or even malicious) code: memory, time and I/O limits are used to make sure students can't overload the system.
-The web application communicates with containers by sending configuration information as a JSON document over standard input. 
+The web application managing the containers communicates with them by sending configuration information as a JSON document over standard input. 
 Every Dodona Docker image shares a `main.sh` file that passes through this information to the actual testing framework, while setting up some error handling.

### 2020-01-22 manuscript.Rmd
fix crossref

--- 2020-01-20 manuscript.Rmd
@@
 A Dockerfile is generated from this description that inherits from the specified base image, and additional custom Docker instructions may be appended by users, though Gigantum's default base images do not currently include the `littler` tool, which is used by Rocker to install packages within `Dockerfile`s.
-Because of these specifics, instructions from `rocker/r-ubuntu` could _not_ be readily re-used in this image stack (see Section&nbsp;\nameref{conclusion}).
+Because of these specifics, instructions from `rocker/r-ubuntu` could _not_ be readily re-used in this image stack (see Section&nbsp;\nameref{conclusions}).
 Both approaches enable `apt` as an installation method, and this is exposed via the GUI-based environment management,\footnote{See \href{https://docs.gigantum.com/docs/environment-management}{https://docs.gigantum.com/docs/environment-management}}.
@@
 # Conclusions
+\label{conclusions}
 

### 2020-01-22 manuscript.Rmd
update Bioconductor section, see #57

@@
 The project consists of 1823 R software packages as of October 30th 2019, as well as packages containing annotation or experiment data.
-_Bioconductor_ has a semi-annual release cycle, each release is associated with a particular version of R.
-Docker images allow availability of current and past versions of _Bioconductor_ for convenience and reproducibility.
-_Bioconductor_'s 'base' docker images are built on top of `rocker/r-ver` and `rocker/rstudio` images.
-_Bioconductor_ installs packages based on the R version and the Bioconductor version, and therefore uses Bioconductor version tagging `devel` and `RELEASE_X_Y`.
-_Bioconductor_ selects the desired version of R from Rocker, adds the BiocManager CRAN package for installing appropriate versions of _Bioconductor_ packages.
-All community maintained images are supposed to be built on top of the _Bioconductor_ base image and therefore indirectly the Rocker images.
-The images are summarized on the _Bioconductor_ web site (see [https://bioconductor.org/help/docker/](https://bioconductor.org/help/docker/)), created with `Dockerfile`s maintained on GitHub ([https://github.com/Bioconductor/bioconductor_docker](https://github.com/Bioconductor/bioconductor_docker)), and [distributed through DockerHub](https://hub.docker.com/u/bioconductor).
-Past and current combinations of R and _Bioconductor_ are therefore accessible via a specific image tag `RELEASE_X_Y`.
+_Bioconductor_ has a semi-annual release cycle, each release is associated with a particular version of R, and Docker images are provided for current and past versions of _Bioconductor_ for convenience and reproducibility.
+All images, included  are described on the _Bioconductor_ web site (see [https://bioconductor.org/help/docker/](https://bioconductor.org/help/docker/)), created with `Dockerfile`s maintained on GitHub, and distributed through Docker&nbsp;Hub\footnote{See \href{https://github.com/Bioconductor/bioconductor_docker}{https://github.com/Bioconductor/bioconductor\_docker} and \href{https://hub.docker.com/u/bioconductor}{https://hub.docker.com/u/bioconductor} respectively.}.
+_Bioconductor_'s 'base' Docker images are built on top of the `rocker/rstudio` image.
+_Bioconductor_ installs packages based on the R version in combination with the Bioconductor version, and therefore uses Bioconductor version tagging `devel` and `RELEASE_X_Y`, e.g., `RELEASE_3_10`.
+Past and current combinations of R and _Bioconductor_ will therefore be accessible via a specific image tags.
+The _Bioconductor_ `Dockerfile` selects the desired version of R from Rocker, adds required system dependencies, and uses the \CRANpkg{BiocManager} package for installing appropriate versions of _Bioconductor_ packages.
 A strength of this approach is that the responsibility for complex software configuration (including customized development) is shifted from the user to the experienced _Bioconductor_ core team.
-Images may be valuable in teaching, where participants pull pre-built images to avoid complicated configuration of their own computing environemnts.
-
-A recent audit of our deprecated _Bioconductor_ `Dockerfile`s, following best practices from the Docker website, led to a reduction in the size and number of layers.
-The most important insights involve the ['union' file system used by Docker](https://docs.docker.com/storage/storagedriver/overlayfs-driver/#how-container-reads-and-writes-work-with-overlay-or-overlay2).
-In this file system, once a layer (e.g., `RUN` statement) writes to a file path, the file path is never altered.
-A subsequent layer that might appear to remove or overwrite the path actually masks, rather than alters, the original.
-It is therefore important to clean up (e.g., cache removal) within each layer, and to avoid re-installing existing dependencies.
-
-A recent innovation is to produce a `bioconductor_docker` image to emulate the _Bioconductor_ nightly Linux build machine as closely as possible.
-This image contains the build system environment variables.
-The image contains the _system dependencies_ needed to install and check almost all (1813 out of 1823) _Bioconductor_ software packages.
-Users no longer have to manage complciated system dependencies.
-The image is configured so that `.libPaths()` has `/usr/local/lib/R/host-site-library` as the first location.
-Users mounting a location on the host file system to this location then persist installed packages across docker sessions or updates.
-Many _R_ users pursue flexible work flows tailored to particular analysis needs, rather than standardized work flows.
-The `bioconductor_docker` image is well-suited to this pattern.
-`bioconductor_docker` provides developers with a test environment like _Bioconductor_'s build system.
+A recent audit of the _Bioconductor_  image stack `Dockerfile` led to the deprecation of several community-maintained images, because the numerous specific images became too hard to understand, complex to maintain, and cumbersome to extent.
+As part of the simplification, a recent innovation is to produce a `bioconductor_docker:devel` image to emulate the _Bioconductor_ nightly Linux build machine as closely as possible.
+This image contains the build system environment variables and the _system dependencies_ needed to install and check almost all (1813 out of 1823) _Bioconductor_ software packages and saves users and package developers from managing these themselves.
+Furthermore the image is configured so that `.libPaths()` has `/usr/local/lib/R/host-site-library` as the first location.
+Users mounting a location on the host file system to this location can persist installed packages across Docker sessions or image updates.
+Many R users pursue flexible work flows tailored to particular analysis needs, rather than standardized work flows.
+The new `bioconductor_docker` image is well-suited for this pattern, while `bioconductor_docker:devel` provides developers with a test environment close to _Bioconductor_'s build system.
 

### 2020-01-22 manuscript.Rmd
Added repo2docker lessons learned

--- 2020-01-22 manuscript.Rmd
@@
 In most cases, individual developers or researchers worrying about reproducibility, will prefer the more widely and host independent Linux-based containers.
-Naturally, a Windows license is required for the host machine, and the licensing impacts the potential to build images in clouds or redistribute exported images as archive files.
+Naturally, a Windows license is required for the host machine, and the licensing impacts the potnetial to build images in clouds or redistribute exported images as archive files.
 But the developments in the `rocker-win` prototype show that containers can also be used for R workflows depending on Windows-only tools, for leveraging an existing Windows Server-based infrastructure where policies can otherwise not be met, or for streamlining interactions with system operating staff that runs Windows-based containers themselves.
 
-## Non-Rocker Linux images
+## Non-Debian Linux images
 \label{nondebian}
@@
 
-_Data Science_ is a widely discussed topic among all academic disciplines \citep[e.g.,][]{donoho_50_2017}.
-The discussions shed a light on the tools and craftspersonship behind the analysis of data with computational methods.
-The practice of Data Science often involves a combination tools and software stacks and requires a cross-cutting skillset.
-This complexity and an inherent concern for openness and reproducibility in the Data Science community lead to Docker being used widely.
-<!--A general description of what to consider when creating a Docker image used for data science projects in R is provided on [environments.rstudio.com](https://environments.rstudio.com/docker).-->
-This section presents exemplary Docker images and image stacks featuring R intended for Data Science.
+_Data Science_ is a widely discussed topic among all academic disciplines \citep[cf. e.g.,][]{donoho_50_2017}.
+Using computers and complex software to make sense of (large amounts of) collected or simulated data is a cross-cutting phenomenon leading to _X Data Science_ disciplines, seminars, and study majors.
+The practice to apply a combination tools and software stacks for analysing data and the cross-cutting skillsets connected with data science puts the management of computing environments and the challenges of reproducibility high up on the agenda.
+This awareness leads to a large number of bespoke Docker images but also some commonly used image stacks, ranging from base images with very common tools to 'catch all' images including _a lot_ of software.
+Here we only mention a few image stacks as examples, which are not based on Rocker images or stem from the R community, but prominently include installations of R.
+A general description of what to consider when creating a Docker image used for data science projects in R is provided on [environments.rstudio.com](https://environments.rstudio.com/docker).
 
@@
 **Kaggle** provides the [`gcr.io/kaggle-images/rstats`](https://hub.docker.com/r/kaggle/rstats) image (previously `kaggle/rstats`) and [corresponding `Dockerfile`](https://github.com/Kaggle/docker-rstats) for usage in their Machine Learning competitions and easy access to the associated datasets.
-It includes machine learning libraries such as Tensorflow and Keras (see also image `rocker/ml` in Section&nbsp;\nameref{rocker-gpu}), and also configures the \CRANpkg{reticulate} package.
+It includes machine learling libraries such as Tensorflow and Keras, and also configures the \pkg{reticulate} package.
 The image uses a base image with _all packages from CRAN_, `gcr.io/kaggle-images/rcran`, which requires a Google Cloud Build as Docker&nbsp;Hub would time out\footnote{Originally, a stacked collection of over 20 images with automated builds on Docker~Hub was used, see \href{https://web.archive.org/web/20190606043353/http://blog.kaggle.com/2016/02/05/how-to-get-started-with-data-science-in-containers/}{https://web.archive.org/web/20190606043353/http://blog.kaggle.com/2016/02/05/how-to-get-started-with-data-science-in-containers/} and \href{https://hub.docker.com/r/kaggle/rcran/dockerfile}{https://hub.docker.com/r/kaggle/rcran/dockerfile}}.
@@
 A Dockerfile is generated from this description that inherits from the specified base image, and additional custom Docker instructions may be appended by users, though Gigantum's default base images do not currently include the `littler` tool, which is used by Rocker to install packages within `Dockerfile`s.
-Because of these specifics, instructions from `rocker/r-ubuntu` could _not_ be readily re-used in this image stack (see Section&nbsp;\nameref{conclusions}).
+Because of these specifics, instructions from `rocker/r-ubuntu` could _not_ be readily re-used in this image stack (see Section&nbsp;\nameref{conclusion}).
 Both approaches enable `apt` as an installation method, and this is exposed via the GUI-based environment management,\footnote{See \href{https://docs.gigantum.com/docs/environment-management}{https://docs.gigantum.com/docs/environment-management}}.
@@
 
+**Whole Tale** (https://wholetale.org) is a National Science Foundation (NSF) funded project developing a scalable, open-source, multi-user platform for reproducible research \citep{brinckman2019, chard2019a}.
+A central goal of the platform is to enable researchers to easily create and publish executable research objects\footnote{In Whole Tale a `tale` is a research object that contains metadata, data (by copy or reference), code, narrative, documentation, provenance, and information about the computational environment to support computational reproducibility} associated with published research \citep{chard2019b}.
+For R users, Whole Tale attempts to combine the strengths of the Rocker Project's curated Docker images with the Jupyter Project's `repo2docker`\footnote{See \href{https://repo2docker.readthedocs.io/}{https://repo2docker.readthedocs.io/}} component to simplify the customization of R-based environments for researchers with limited experience with either Docker or Git.
+Using Whole Tale, researchers can create and publish Rocker-based reproducible research objects to a growing number of repositories including DataONE member nodes, Zenodo and soon Dataverse.
+Additionally, Whole Tale supports automatic data citation and is working on capabilities for image preservation and provenance capture to improve the transparency of published computational research artifacts \citep{mecum2018, mcphillips2019}.
+Multiple options have been discussed to allow users to change the base image used in `repo2docker` from the default Ubuntu LTS required to support the Rocker Project images. 
+Whole Tale implemented a custom `RockerBuildPack`\footnote{See \href{https://github.com/whole-tale/repo2docker\_wholetale}{https://github.com/whole-tale/repo2docker\_wholetale}} to support customization of the `rocker/geospatial` image through `repo2docker` composability \footnote{Composability refers to the ability to combine multiple package managers -- such as R, `pip`, and `conda`}.
+This works in part because Rocker images are based on a Debian distribution.
+
 # Use cases and applications
@@
 The applications were deployed on Amazon Elastic Container Service ([ECS](https://aws.amazon.com/ecs/)).
-The main learnings from using R in Docker was the imporance of not only pinning the R package versions via MRAN, but also moving away from Debian testing to an LTS distribution.
-For the use case at hand, the priority was to have more control over upstream updates, and to minimize the risk of breaking the automated builds of the Docker images and production jobs.
-
-A more recent related work is the \CRANpkg{plumber} ([https://www.rplumber.io/](https://www.rplumber.io/)) package, which makes it even easier for R to be used to host web services through RESTful APIs.
+The main learnings from using R in Docker was the imporance of not only pinning the R package versions via MRAN, but also moving away from Debian testing to an LTS distribution to minimize the risk of breaking the automated builds of the Docker images and production jobs.
+
+A more recent related work is the \pkg{plumber} ([https://www.rplumber.io/](https://www.rplumber.io/)) package, which makes it even easier for R to be used to host web services through RESTful APIs.
 When R is running on a server, a different server can ask R to make a computation through an HTTP request and get an immediate response.
@@
 
-\label{rocker-gpu}
 Using **GPUs** (graphical processing units) as a specialised hardware from containerized common work environments is also possible and useful \citep{haydel_enhancing_2015}.
@@
 The `rocker/tidyverse` base image ensures that users can also readily use packages for analysis, display, and reporting.
-
-Another way this idea of a common environment and the partitioning that Docker provides is used in teaching is in the **[Dodona](https://dodona.ugent.be)** project.
-Dodona is a web platform developed at Ghent University and is used to teach students basic programming skills.
-Dodona uses Docker containers to test submissions by students.
-This means that both the code testing the students' submissions and the submission itself are executed in a predictable environment, avoiding compatibility issues between the wide variety of configurations used by students.
-The containerization is also used to shield the Dodona servers from bad (or even malicious) code: memory, time and I/O limits are used to make sure students can't overload the system.
-The web application managing the containers communicates with them by sending configuration information as a JSON document over standard input. 
-Every Dodona Docker image shares a `main.sh` file that passes through this information to the actual testing framework, while setting up some error handling.
-The testing process in the Docker containers sends back the test results by writing a JSON document to its standard output channel.
-In June 2019, R support was added to Dodona using an image derived from the `rocker/r-base` image that sets up the `runner` user and and `main.sh` file expected by Dodona\footnote{\href{https://github.com/dodona-edu/docker-images/blob/master/dodona-r.dockerfile}{https://github.com/dodona-edu/docker-images/blob/master/dodona-r.dockerfile}}.
-It also installs the packages required for the testing framework and the exercises so that this doesn't have to happen every time a student's submission is evaluated.
-The actual testing of R exercises is done using a custom framework loosely based on \CRANpkg{testthat}.
-During the development of the testing framework it was found that the \CRANpkg{testthat} framework did not provide enough information to its reporter system to send back all the fields required by Dodona to render its feedback.
-Right now, multiple statistics courses are developing exercises to automate the feedback for their lab classes.
 
@@
 
-Containers provide a high degree of isolation that is often desirable when attempting to capture a specific computational environment so that others can reproduce and extend a research result.
-Many computationally intensive research projects depend on specific versions of original and third-party software packages in diverse languages, joined together to form a pipeline through which data flows.
-New releases of even just a single piece of software in this pipeline can break the entire workflow, making it difficult to find the error and for others to reuse existing pipelines, and making the original the results irreproducible and not extandable.
-The chance of a substantial disruption like this is high in a multi-year research project where key pieces of third-party software may have several major updates over the duration of the project.
-The classical "paper" article is insufficient to adequately communicate the knowledge behind such research projects \citep[cf.][]{donoho_invitation_2010,marwick_how_2015}.
-\citet{gentleman_statistical_2007} coined the term _Research Compendium_ for a dynamic document together with supporting data and code.
-They use the R package system \cite{core_writing_1999} as for the functional prototype to structuring, validation, and distribution of research compendia.
-This concept has been taken up and extended\footnote{See full literature list at \href{https://research-compendium.science/}{https://research-compendium.science/}.}, not the least by applying containerisation and other methods for managing computing environments---see Section&nbsp;\nameref{envs}.
-Containers give the researcher an isolated environment to assemble these research pipelines with specific versions of software.
-Research workflows in containers are safe from contamination from other activities occuring on the researcher's computer, for example the installation of newest version of packages for teaching demonstrations.
-Given the users in this scenario, i.e., often academics with limited formal software development training, templates and assistance with containers around research compendia is essential.
-
-In many fields we see that a typical unit of research for a container is a research report or journal article, where the container holds the compendium, or self-contained set of data (or connections to data elsewhere) and code files needed to fully reproduce the article \citep{marwick_packaging_2018}.
-The package \pkg{rrtools} ([https://github.com/benmarwick/rrtools](https://github.com/benmarwick/rrtools)) provides a template and convienence functions to apply good practices for research compendia, including a starter `Dockerfile`.
-Images of compendium containers can be hosted on services such as Docker&nbsp;Hub for convienient sharing among collaborators and others. 
-Similarly, packages such as \pkg{containerit} and \pkg{dockerfiler} can be used to manage the `Dockerfile` to be archived with a compendium on a data repository (e.g.
-[Zenodo](https://zenodo.org/), [Dataverse](https://dataverse.org/), [Figshare](https://figshare.com/), [OSF](https://osf.io/)).
-A typial compendium's `Dockerfile` will pull a rocker image fixed to a specific version of R, and install R packages from the MRAN repository to ensure the package versions are tied to a specific date, rather than the most recent version.
-A future researchers can download then compendium from the repository and run the included `Dockerfile` build a new image that recreates the computational environment used to produce the original research results.
-If building the image fails, the human-readable instructions in a `Dockerfile` are a the starting point for rebuilding the environment.
-Further safeguarding practices are currently under development or not part of common practice yet, such as preservation of images \citep{emsley_framework_2018} and storing them alongside `Dockerfile`s \citep[cf.][]{nust_opening_2017}, or pinning of system libraries beyond the tagged base images, which may be seen as stable or dynamic depending on the applied time scale \citep[see discussion on `debian:testing` base image in][]{RJ-2017-065}.
+- researchcompendia.science
+- https://github.com/benmarwick/rrtools
+- \pkg{renv}, https://rstudio.github.io/renv/articles/docker.html
+- what to copy in and what not
+
+## Publication Reproducibility Audits
+
+A recommendation of the recent National Academies' report on _Reproducibility and Replicability in Science_ is that journals "consider ways to ensure computational reproducibility for publications that make claims based on computations" \citep{NASEM2019}.
+In fields such as Political Science and Economics, journals are increasingly adopting policies requiring authors to publish the code and data required to reproduce computational findings reported in published manuscripts, subject to independent verification \citep{Jacoby2017,Vilhuber2019,Alvarez2018,Christian2018,Eubank2016,King1995}.
+Problems with the computational environment, installation and availability of software dependencies are common.
+R is gaining popularity in these communities (in a sample of 105 replication packages published by the _American Journal of Political Science_ (AJPS) over 65% use R).
+Projects such as NSF Whole Tale use the Rocker Project community images with the goal of improving the reproducibility of published research artifacts and simplifying the publication and verification process for both authors and journals by reducing errors and time spent specifying the environment.
 
@@
 # Conclusions
-\label{conclusions}
 

### 2020-01-22 README.Rmd

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Rockerverse

<!-- badges: start -->
[![Launch RStudio Binder](http://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/nuest/rockerverse-paper/master?urlpath=rstudio)
[![Build Status](https://travis-ci.org/nuest/rockerverse-paper.svg?branch=master)](https://travis-ci.org/nuest/rockerverse-paper)
<!-- badges: end -->

Draft for an article of everything related to containers and R, originally based on the blog post http://bit.ly/docker-r.

_Want to contribute?_ See https://github.com/nuest/rockerverse-paper/issues/3.

**Current draft: https://nuest.github.io/rockerverse-paper/master.pdf**

```{r hexSticker, message=TRUE, warning=FALSE, cache=TRUE, include=FALSE}
if(!file.exists("rockerverse.png")) {
  library("magick")
  library("here")

  # https://commons.wikimedia.org/wiki/File:NASA-HS201427a-HubbleUltraDeepField2014-20140603.jpg
  universe <- magick::image_read("https://upload.wikimedia.org/wikipedia/commons/6/69/NASA-HS201427a-HubbleUltraDeepField2014-20140603.jpg")
  magick::image_write(universe, here("universe.png"))
  # rockerLogoMask <-
  magick::image_read("https://avatars2.githubusercontent.com/u/9100160?s=200&v=4") %>%
    magick::image_scale("2100x2100") %>%
    magick::image_channel("Matte") %>%
    image_negate() %>%
    magick::image_border("black", "150x0") %>%
    magick::image_write(here("rocker_mask.png"))

  # does not work: image_composite(universe, rockerLogoMask, operator = "CopyOpacity", compose_args = "-composite -alpha Off")
  system2("convert", c(here("universe.png"), here("rocker_mask.png"),
    "-alpha Off",
    "-compose CopyOpacity",
    "-composite",
    here("whaleverse.png")))

  library("hexSticker")
  hexSticker::sticker(here("whaleverse.png"), package = "Rockerverse",
          p_size = 20, s_x = 0.97, s_y = 0.95, s_width = 1,
          h_fill = "#000000", h_color = "#ffffff",
          url = "github.com/nuest/rockerverse-paper",
          u_color = "#ffffff", u_size = 4,
          filename = here("rockerverse.png"))

  file.remove(c(here("universe.png"), here("whaleverse.png"), here("rocker_mask.png")))
}
```

![Rockerverse hex sticker](`r here::here("rockerverse.png")`)

## Render the manuscript with Docker

**Local build**

```bash
docker build --tag rockerverse-paper --file .binder/Dockerfile .
docker run -i -v $(pwd):/rockerverse --user $UID rockerverse-paper Rscript -e 'setwd("/rockerverse"); rmarkdown::render("manuscript.Rmd")'
```

**Image from Docker Hub**

```bash
docker run -i -v $(pwd):/rockerverse --user $UID nuest/rockerverse-paper Rscript -e 'setwd("/rockerverse"); rmarkdown::render("manuscript.Rmd")'
```

## Not included projects and packages

- `rize`, because discontinued, see #35
- `docker` ([https://bhaskarvk.github.io/docker](https://bhaskarvk.github.io/docker) has been removed from CRAN, seems unmaintained by authors
- https://github.com/svlentink/dockerfiles/blob/master/svlentink/r-base-alpine/Dockerfile because the Alpine image is quite a trivial one
- https://github.com/jlisic/R-docker-centos because CentOS images are also provided by R-hub
- https://github.com/cardcorp/card-rocker because last change 3 years ago, an all based on `rocker/r-base` - lesson learned would be interesting still, see [#15](https://github.com/nuest/rockerverse-paper/issues/15)
- `sevenbridges` includes several Dockerfiles and documentation about using images, and also seems to use Docker in the built workflows but via other tools, not directly, https://sbg.github.io/sevenbridges-r/, https://github.com/sbg/sevenbridges-r/search?p=4&q=docker&unscoped_q=docker
- https://github.com/rocker-jp because it seems to have stalled soon after inception

### 2020-01-22 manuscript.Rmd

--- 2020-01-22 manuscript.Rmd
@@
 A minor reason in the inception of the images were also project constraints requiring a Rocker-incompatible licensing of the `Dockerfile`s, i.e., the MIT License.
+
+**Whole Tale** (https://wholetale.org) is a National Science Foundation (NSF) funded project developing a scalable, open-source, multi-user platform for reproducible research \citep{brinckman2019, chard2019a}.
+A central goal of the platform is to enable researchers to easily create and publish executable research objects\footnote{In Whole Tale a `tale` is a research object that contains metadata, data (by copy or reference), code, narrative, documentation, provenance, and information about the computational environment to support computational reproducibility} associated with published research \citep{chard2019b}.
+For R users, Whole Tale attempts to combine the strengths of the Rocker Project's curated Docker images with the Jupyter Project's `repo2docker`\footnote{See \href{https://repo2docker.readthedocs.io/}{https://repo2docker.readthedocs.io/}} component to simplify the customization of R-based environments for researchers with limited experience with either Docker or Git.
+Using Whole Tale, researchers can create and publish Rocker-based reproducible research objects to a growing number of repositories including DataONE member nodes, Zenodo and soon Dataverse.
+Additionally, Whole Tale supports automatic data citation and is working on capabilities for image preservation and provenance capture to improve the transparency of published computational research artifacts \citep{mecum2018, mcphillips2019}.
+Multiple options have been discussed to allow users to change the base image used in `repo2docker` from the default Ubuntu LTS required to support the Rocker Project images. 
+Whole Tale implemented a custom `RockerBuildPack`\footnote{See \href{https://github.com/whole-tale/repo2docker\_wholetale}{https://github.com/whole-tale/repo2docker\_wholetale}} to support customization of the `rocker/geospatial` image through `repo2docker` composability \footnote{Composability refers to the ability to combine multiple package managers -- such as R, `pip`, and `conda`}.
+This works in part because Rocker images are based on a Debian distribution.
 

### 2020-01-22 manuscript.tex

@@
 % !TeX root = RJwrapper.tex
-\title{Containers and R---the Rockerverse and beyond}
-\author{by Daniel Nüst, Dom Bennett, Robrecht Cannoodt, Dirk Eddelbuettel, Mark Edmondson, Colin Fay, Ellis Hughes, Sean Lopp, Ben Marwick, Heather Nolis, Jacqueline Nolis, Hong Ooi, Karthik Ram, Noam Ross, Lori Shepherd, Nitesh Turaga, Nan Xiao}
+\title{The Rockerverse: Images, Packages, and Applications for Containerization
+with R}
+\author{by Daniel Nüst, Dom Bennett, Robrecht Cannoodt, Dav Clark, Gergely Daroczi, Dirk Eddelbuettel, Mark Edmondson, Colin Fay, Ellis Hughes, Sean Lopp, Ben Marwick, Heather Nolis, Jacqueline Nolis, Hong Ooi, Karthik Ram, Noam Ross, Lori Shepherd, Nitesh Turaga, Craig Willis, Nan Xiao}
 
@@
 \centering
-\includegraphics{/home/daniel/git/o2r/rockerverse-paper/rockerverse.png}
+\includegraphics{/rockerverse/rockerverse.png}
 \caption{Rockerverse hex sticker}
@@
 manager \href{https://en.wikipedia.org/wiki/Apache_Maven}{Apache Maven},
-cf.~\url{http://packages.renjin.org/packages} for available packages.
+cf. \url{http://packages.renjin.org/packages} for available packages.
 Packages are loaded on demand, i.e., at the first call to
@@
 \texttt{Dockerfile}s are provided by
-\href{https://github.com/nuest/fastr-docker}{nuest/fastr-docker},
-e.g.~\href{https://cloud.docker.com/repository/docker/nuest/fastr/}{\texttt{nuest/fastr}
+\href{https://github.com/nuest/fastr-docker}{nuest/fastr-docker}, e.g.
+\href{https://cloud.docker.com/repository/docker/nuest/fastr/}{\texttt{nuest/fastr}
 on Docker Hub}.
@@
 The \href{https://github.com/jupyter/docker-stacks/}{\textbf{Jupyter
-Docker Stacks} project} are a set of ready-to-run Docker images
+Docker Stacks}} project are a set of ready-to-run Docker images
 containing Jupyter applications and interactive computing tools
 \citep{project_jupyter_jupyter_2018}. The \texttt{jupyter/r-notebook}
-image includes R and ``popular packages'', and R is also included in the
-catchall \texttt{datascience-notebook}
+image includes R and ``popular packages'', and naturally also the
+IRKernel (\url{https://irkernel.github.io/}), an R kernel for Jupyter so
+that Jupyter Notebooks can contain R code cells. R is also included in
+the catchall \texttt{jupyter/datascience-notebook}
 image\footnote{\href{https://jupyter-docker-stacks.readthedocs.io/en/latest/using/selecting.html}{https://jupyter-docker-stacks.readthedocs.io/en/latest/using/selecting.html}}.
@@
 Notebook server locally or build their own specialised images on top of
-stable toolsets.
+stable toolsets. R is installed using the Conda package manager, which
+can manage environments for various programming languages, pinning both
+the R version and the versions of R
+packages\footnote{See \texttt{jupyter/datascience-notebook}'s \texttt{Dockerfile} at \href{https://github.com/jupyter/docker-stacks/blob/master/datascience-notebook/Dockerfile\#L47}{https://github.com/jupyter/docker-stacks/blob/master/datascience-notebook/Dockerfile\#L47}.}.
 
@@
 
-\textbf{\href{http://gigantum.com/}{Gigantum}} also its own image stack
-(\url{https://github.com/gigantum/base-images}). \ldots{}
+\textbf{Gigantum} (\url{http://gigantum.com/} is a platform for open and
+decentralized data science with a focus on using automation and
+user-friendly tools for easy sharing of reproducible computational
+workflows. Gigantum builds on the \emph{Gigantum Client} (running either
+locally or on a remote server) for development and execution of
+data-focused \emph{Projects}, which can be stored and shared via the
+\emph{Gigantum Hub} or via a zipfile export. The Client is a user
+friendly interface to a backend using Docker containers to package,
+build, and run Gigantum projects. It is configured to use a default set
+of Docker base images (\url{https://github.com/gigantum/base-images}),
+and users are able to define and configure their own custom images. The
+available images include two with R based on Ubuntu Linux have the
+\href{https://launchpad.net/~marutter/+archive/ubuntu/c2d4u3.5/}{\texttt{c2d4u}
+CRAN PPA} pre-configured for installation of binary R
+packages\footnote{\href{https://docs.gigantum.com/docs/using-r}{https://docs.gigantum.com/docs/using-r}}.
+The R images vary in the included authoring environment, i.e., Jupyter
+in \texttt{r-tidyverse} or both Jupyter \& RStudio in
+\texttt{rstudio-server}. The independent image stack can be traced back
+to the Gigantum environment and its features. The R images are based on
+Gigantum's \texttt{python3-minimal} image, originally to keep the
+existing front-end configuration, but also to provide a consistent
+Python-to-R interoperability. The \texttt{Dockerfile}s also use build
+args to specify bases, for example for different version of NVIDIA CUDA
+for GPU
+processing\footnote{See \href{https://github.com/gigantum/base-images/blob/master/_templates/python3-minimal-template/Dockerfile}{https://github.com/gigantum/base-images/blob/master/\_templates/python3-minimal-template/Dockerfile} for the \texttt{Dockerfile} of \texttt{python3-minimal}.}
+so that appropriate GPU drivers can be enabled automatically when
+supported. Furthermore, Gigantum's focus lies on environment management
+via GUI and ensuring a smooth user interaction, e.g., with reliable and
+easy conflict detection and resolution. For this reason, project
+repositories store authoritative package information in a separate file
+per package, allowing Git to directly detect conflicts and changes. A
+Dockerfile is generated from this description that inherits from the
+specified base image, and additional custom Docker instructions may be
+appended by users, though Gigantum's default base images do not
+currently include the \texttt{littler} tool, which is used by Rocker to
+install packages within \texttt{Dockerfile}s. Because of these
+specifics, instructions from \texttt{rocker/r-ubuntu} could \emph{not}
+be readily re-used in this image stack (see
+Section~\nameref{conclusion}). Both approaches enable \texttt{apt} as an
+installation method, and this is exposed via the GUI-based environment
+management,\footnote{See \href{https://docs.gigantum.com/docs/environment-management}{https://docs.gigantum.com/docs/environment-management}}.
+The image build and publication process is scripted with Python and JSON
+template configuration files, unlike Rocker which relies on plain
+\texttt{Dockerfile}s. A minor reason in the inception of the images were
+also project constraints requiring a Rocker-incompatible licensing of
+the \texttt{Dockerfile}s, i.e., the MIT License.
+
+\textbf{Whole Tale} (\url{https://wholetale.org}) is a National Science
+Foundation (NSF) funded project\footnote{NSF Award OAC-1541450}
+developing a scalable, open-source, multi-user platform for reproducible
+research \citep{brinckman2019, chard2019a}. A central goal of the
+platform is to enable researchers to easily create and publish
+executable research
+objects\footnote{In Whole Tale a `tale` is a research object that contains metadata, data (by copy or reference), code, narrative, documentation, provenance, and information about the computational environment to support computational reproducibility}
+associated with published research \citep{chard2019b}. For R users,
+Whole Tale attempts to combine the strengths of the Rocker Project's
+curated Docker images with the Jupyter Project's
+\texttt{repo2docker}\footnote{See \href{https://repo2docker.readthedocs.io/}{https://repo2docker.readthedocs.io/}}
+component to simplify the customization of R-based environments for
+researchers with limited experience with either Docker or Git. Using
+Whole Tale, researchers can create and publish Rocker-based reproducible
+research objects to a growing number of repositories including DataONE
+member nodes, Zenodo and soon Dataverse. Additionally, Whole Tale
+supports automatic data citation and is working on capabilities for
+image preservation and provenance capture to improve the transparency of
+publish computational research artifacts
+\citep{mecum2018, mcphillips2019}.
 
@@
 \textbf{\pkg{stevedore}}
-(\url{https://cran.r-project.org/package=stevedore}) currently the most
-powerful Docker client in R.
+(\url{https://cran.r-project.org/package=stevedore}) is currently the
+most powerful Docker client in R. It interfaces with the Docker daemon
+over the Docker HTTP API\footnote{} via a Unix socket on Linux or MacOS,
+over a named pipe on Windows, or over an HTTP/TCP connection. The
+package is the only one not using system calls to the \texttt{docker}
+CLI for managing images and containers and easily exposes connections to
+remote Docker daemons, which has to be configured on the Docker level
+otherwise. Using the API gives access to more information and is system
+independent and likely more reliable than parsing command line output.
+\pkg{stevedore}'s own interface is automatically generated based on the
+OpenAPI specification of the Docker daemon, but still similar to the
+Docker CLI. The interface is similar to R6 objects, in that a
+\texttt{stevedore\_object} has a number of functions attached to it that
+can be called, and multiple specific versions of the Docker API can be
+supported thanks to the automatic
+generation\footnote{See \href{https://github.com/richfitz/stevedore/tree/master/inst/spec}{https://github.com/richfitz/stevedore/tree/master/inst/spec} and \href{https://github.com/richfitz/stevedore/blob/master/development.md}{https://github.com/richfitz/stevedore/blob/master/development.md}.}.
 
@@
 interface to a number of container-related services in Microsoft's
-\href{https://azure.microsoft.com/}{Azure Cloud}, namely
-\href{https://azure.microsoft.com/en-us/services/container-instances/}{Container
-Instances},
-\href{https://azure.microsoft.com/en-us/services/container-registry/}{Container
-Registry}, and
-\href{https://azure.microsoft.com/en-us/services/kubernetes-service/}{Kubernetes
-Service}. While it is mainly intended for working with Azure, as a
-convenience feature it includes lightweight, cross-platform shells to
-Docker and Kubernetes (tools \texttt{kubectl} and \texttt{helm}). These
-can be used to create and manage arbitrary Docker images and containers,
-as well as Kubernetes clusters on any platform or cloud service.
+\href{https://azure.microsoft.com/}{Azure Cloud}, see
+Section~\nameref{azurecontainers}. While it is mainly intended for
+working with Azure, as a convenience feature it includes lightweight,
+cross-platform shells to Docker and Kubernetes (tools \texttt{kubectl}
+and \texttt{helm}). These can be used to create and manage arbitrary
+Docker images and containers, as well as Kubernetes clusters on any
+platform or cloud service.
 
@@
 Docker containers, modern enterprises are now capable of having
-real-time machine learning powered by R.
-
-The backbone of this work is the \pkg{plumber}
-(\url{https://www.rplumber.io/}) package. The plumber package allows R
-to be used to host web services through RESTful APIs. When R is running
-on a server, a different server can ask R to make a computation through
-an HTTP request and get an immediate response. For instance, if a data
-scientist at a company creates a machine learning model to tell which
-customers are likely to make a repeat purchase, if they serve the model
-through a plumber API then someone else on the marketing team could
-write software that sends different emails depending on that real-time
-prediction. Plumber syntax is simple to read and understand: the API is
-created by commenting above function names in an R script.
+real-time machine learning powered by R. One of the first
+implementations of having R respond to client reqests were TCP/IP and
+local socket servers, such as \CRANpkg{Rserve} and \CRANpkg{svSocket},
+which allowed other local or remote processes to make binary requests to
+interact with R without the need to link to R at compile time and
+connected other programming languages with R. R could also empower web
+services since at least 2007, when
+\href{http://www.rapache.net}{rApache} was presented at the useR!
+conference \citep{horner_rapache_2013}. rApache runs an R interpreter
+inside the Apache web server, which makes it possible and very efficient
+to build scalable R-based APIs and web services.
+
+Such rApache applications have been used at
+\href{http://rapporter.net}{rapporter.net}, transforming parametrized R
+Markdown-like documents into interactive web applications for reporting.
+A similar design was implemented at
+\href{https://www.card.com/}{CARD.com}, where the optimization of the
+acquisition flow and the real-time analysis of debit card transactions
+both relied on an efficient combination of Rserve and rApache running in
+Docker containers: the R session behind Rserve acted as a read-only,
+exteremely fast and scalabale, in-memory database for the many
+concurrent rApache processes responding to the live-scoring requests of
+various divisions of the company. Similar dockerized R scripts were
+responsible for the ETL processes and even the client-facing email, text
+message and push notification alerts sent in real-time based on card
+transactions. The related Docker images were made available at
+\url{https://github.com/cardcorp/card-rocker}. The images extend
+\texttt{rocker/r-base}, e.g., with an SSH client and a workaround for
+being able to mount SSH keys from the host, Pandoc, the Amazon Web
+Services (AWS) SDK, and Java, which is required by the AWS SDK. The AWS
+SDK allowed to run R consumers reading from real-time data processing
+streams of \href{https://aws.amazon.com/kinesis/}{AWS Kinesis}. The
+applications were deployed on Amazon Elastic Container Service
+(\href{https://aws.amazon.com/ecs/}{ECS}). The main learnings from using
+R in Docker was the imporance of not only pinning the R package versions
+via MRAN, but also moving away from Debian testing to an LTS
+distribution to minimize the risk of breaking the automated builds of
+the Docker images and production jobs.
+
+A more recent related work is the \pkg{plumber}
+(\url{https://www.rplumber.io/}) package, which makes it even easier for
+R to be used to host web services through RESTful APIs. When R is
+running on a server, a different server can ask R to make a computation
+through an HTTP request and get an immediate response. For instance, if
+a data scientist at a company creates a machine learning model to tell
+which customers are likely to make a repeat purchase, if they serve the
+model through a plumber API then someone else on the marketing team
+could write software that sends different emails depending on that
+real-time prediction. Plumber syntax is simple to read and understand:
+the API is created by commenting above function names in an R script.
 
@@
 
+\href{http://restrserve.org}{RestRserve} is an alternative R web API
+framework with the main focus on being high-performant and robust. It
+also ships automated Docker builds at
+\url{https://hub.docker.com/r/rexyai/restrserve/dockerfile}.
+
 \hypertarget{deployment-to-the-cloud}{%
@@
   with R is published at this blog post}.
-\item
-  \href{https://cran.r-project.org/package=AzureContainers}{\texttt{AzureContainers}}:
-  Umbrella package for working with containers in Microsoft's Azure
-  cloud. Provides interfaces to three Azure services: Container
-  Registry, Container Instances (for running individual containers), and
-  Kubernetes Services (for orchestrated deployments)
 \end{itemize}
 
-While Plumber provides the low-level infrastructure for turning a
-statistical model into a predictive service, for production purposes it
-is usually necessary to take into account considerations such as
-scalability, reliability and ease of management. A single container is
-limited in the volume of requests it can service. Furthermore, if the
-container goes down, or is occupied with processing a computationally
-intensive task, the service becomes unavailable until it is restarted.
-
-These issues can be mitigated by deploying not a single container, but a
-\emph{cluster} of containers, orchestrated as a single deployment. The
-deployment system of choice at the time of writing is
-\href{https://kubernetes.io}{Kubernetes}. This allows you to deploy a
-single image to a pool of servers, which can then serve requests in
-parallel. Client-side requests are made to the cluster ingress endpoint,
-which then redirects them to individual servers. The orchestrator can
-identify cluster nodes that have failed, and restart them automatically;
-it can also detect when there is no capacity to service more requests,
-and start additional servers and containers as needed.
-
-While these load-balancing and autoscaling properties of a Kubernetes
-cluster are invaluable, they are still limited by the hardware resources
-that it has access to. A cluster cannot scale out if it has run out of
-servers. This issue in turn can be addressed by deploying not to
-on-premise servers but to the cloud, where effectively unlimited
-computing resources are available (as long as one can pay for them).
-From a management perspective, cloud deployments have the additional
-advantage that hardware maintenance becomes the responsibility of the
-cloud provider, eliminating another pain point. Orchestrator services
-are available from all the major cloud service providers, including
-Amazon Web Services (AWS), Google Cloud Platform, and Microsoft Azure.
-
-An example of an R package for working with containers in the cloud is
-AzureContainers, which interacts with three Azure services: Container
-Instances (for running a single container), Container Registry (a
-private docker registry service), and Kubernetes Service (for Kubernetes
-on Azure). AzureContainers provides a lightweight yet powerful interface
-for working with Resource Manager, the framework for deploying and
-managing arbitrary resources in Azure. On the client side, it provides
-simple shells to the Docker, Kubectl and Helm tools that are commonly
-used for managing containers and Kubernetes clusters.
-
-Here is some simple code that pushes a Docker image to an Azure
-Container Registry, and then creates a deployment using Azure Kubernetes
-Service. This is a slightly modified version of the code from the
-``Deploying a prediction service with Plumber'' AzureContainers
-vignette. The code makes use of a yaml file to define the deployment and
-predictive service, which is standard practice with Kubernetes; the yaml
-can be seen in the aforementioned vignette.
-
-\begin{verbatim}
-library(AzureContainers)
-# create a resource group for our deployments
-deployresgrp <- AzureRMR::get_azure_login()$
-    get_subscription("subscription_id")$
-    create_resource_group("deployresgrp", location="australiaeast")
-
-### create a container registry
-deployreg_svc <- deployresgrp$create_acr("deployreg")
-
-# build our image (a random forest on the Boston housing data, available in the MASS package)
-call_docker("build -t bos-rf .")
-
-# upload the image to Azure
-deployreg <- deployreg_svc$get_docker_registry()
-deployreg$push("bos-rf")
-
-### create a Kubernetes cluster with 2 nodes
-deployclus_svc <- deployresgrp$create_aks("deployclus", agent_pools=aks_pools("pool1", 2))
-
-# grant the cluster pull access to the registry
-gr <- AzureGraph::get_graph_login()
-aks_app_id <- deployclus$properties$servicePrincipalProfile$clientID
-reg$add_role_assignment(gr$get_app(aks_app_id), "Acrpull")
-
-# get the cluster endpoint
-deployclus <- deployclus_svc$get_cluster()
-
-# create and start the service
-deployclus$create("bos-rf.yaml")
-\end{verbatim}
-
-Once the service has been created, we can check on its status and obtain
-predictions:
-
-\begin{verbatim}
-deployclus$get("deployment bos-rf")
-#> Kubernetes operation: get deployment bos-rf  --kubeconfig=".../kubeconfigxxxx"
-#> NAME      DESIRED   CURRENT   UP-TO-DATE   AVAILABLE   AGE
-#> bos-rf    1         1         1            1           5m
-
-deployclus$get("service bos-rf-svc")
-#> Kubernetes operation: get service bos-rf-svc  --kubeconfig=".../kubeconfigxxxx"
-#> NAME         TYPE           CLUSTER-IP   EXTERNAL-IP     PORT(S)          AGE
-#> bos-rf-svc   LoadBalancer   10.0.8.189   xxx.xxx.xxx.xxx 8000:32276/TCP   5m
-
-response <- httr::POST("http://xxx.xxx.xxx.xxx:8000/score",
-    body=list(df=MASS::Boston[1:10,]), encode="json")
-httr::content(response, simplifyVector=TRUE)
-#> [1] 25.9269 22.0636 34.1876 33.7737 34.8081 27.6394 21.8007 22.3577 16.7812 18.9785
-\end{verbatim}
-
-Note that Plumber, by itself, provides no authentication or security
-features; anybody who knows the cluster's IP address can access the
-predictive service. At minimum, a basic authentication layer should be
-added to any Kubernetes deployment. Standard Kubernetes procedure
-involves creating an ingress controller, for example with the nginx or
-traefik reverse proxy software, and one can then add
-application-specific authentication layers on top of that (for example,
-to authenticate with a user's Azure Active Directory or LDAP
-credentials). More details can be found from most Kubernetes learning
-resources, and are outside the scope of this paper.
+\label{azurecontainers} \textbf{\pkg{AzureContainers}} is an umbrealla
+package providing interfaces the three services of Microsoft's Azure
+Cloud, namely
+\href{https://azure.microsoft.com/en-us/services/container-instances/}{Container
+Instances} for running individual containers,
+\href{https://azure.microsoft.com/en-us/services/container-registry/}{Container
+Registry} for private image distribution, and
+\href{https://azure.microsoft.com/en-us/services/kubernetes-service/}{Kubernetes
+Service} for orchestrated deployments. While a package like
+\pkg{plumber} provides the infrastructure for turning an R workflow a
+service, for production purposes it is usually necessary to take into
+account scalability, reliability and ease of management. A single
+container is limited in the volume of requests it can service, and if
+the container goes down or is occupied with processing a computationally
+intensive task, the service becomes unavailable. These issues can be
+mitigated by deploying not a single container, but a \emph{cluster} of
+containers, orchestrated as a single deployment, e.g., with
+\href{https://kubernetes.io}{Kubernetes}. A cluster features
+load-balancing, autoscaling of containers across numerous servers (in
+the cloud or on premise), and restarting failed ones.
+\pkg{AzureContainers} provides a lightweight yet powerful interface for
+working with Resource Manager, the framework for deploying and managing
+arbitrary resources in Azure. On the client side, it provides simple
+shells to the Docker, and to Kubernetes' \texttt{kubectl} and
+\texttt{helm} tools, that are needed for managing Kubernetes clusters.
+The package simplifies complex infrastructure management to a number of
+R funcction calls, given an Azure account with sufficient
+credit\footnote{See \emph{"Deploying a prediction service with Plumber"} vignette for details:  \href{https://cran.r-project.org/web/packages/AzureContainers/vignettes/vig01_plumber_deploy.html}{https://cran.r-project.org/web/packages/AzureContainers/vignettes/vig01\_plumber\_deploy.html}.}.
 
@@
 could enable transfer of input (parameters and data) and output
-(results) to and from containers. Some packages, e.g.~\pkg{containerit},
+(results) to and from containers. Some packages, e.g. \pkg{containerit},
 do provide Docker images that can be used similarly to CLIs, but their
@@
 \pkg{outsider} package and then call module code to integrate a tool
-into their own workflow seamlessly. The \pkg{outsider} package and
-module manage the containers and handle the transmission of arguments
-and the transfer of files to and from a container. These functionalities
-also allow a user to launch module code on a remote machine via SSH,
-expanding the potential computational scale. Outsider modules can be
-hosted code-sharing services, e.g., on GitHub, and \pkg{outsider}
-contains discovery functions for them.
+into their own R-based workflow seamlessly. The \pkg{outsider} package
+and module manage the containers and handle the transmission of
+arguments and the transfer of files to and from a container. These
+functionalities also allow a user to launch module code on a remote
+machine via SSH, expanding the potential computational scale. Outsider
+modules can be hosted code-sharing services, e.g., on GitHub, and
+\pkg{outsider} contains discovery functions for them.
 
@@
 \end{itemize}
+
+\hypertarget{publication-reproducibility-audits}{%
+\subsection{Publication Reproducibility
+Audits}\label{publication-reproducibility-audits}}
+
+A recommendation of the recent National Academies' report on
+\emph{Reproducibility and Replicability in Science} is that journals
+``consider ways to ensure computational reproducibility for publications
+that make claims based on computations'' \citep{NASEM2019}. In fields
+such as Political Science and Economics, journals are increasingly
+adopting policies requiring authors to publish the code and data
+required to reproduce computational findings reported in published
+manuscripts, subject to independent verification
+\citep{Jacoby2017,Vilhuber2019,Alvarez2018,Christian2018,Eubank2016,King1995}.
+Problems with the computational environment, installation and
+availability of software dependencies are common. R is gaining
+popularity in these communities (in a sample of 105 replication packages
+published by the \emph{American Journal of Political Science} (AJPS)
+over 65\% use R). Projects such as NSF Whole Tale use the Rocker Project
+community images with the goal of improving the reproducibility of
+published research artifacts and simplifying the publication and
+verification process for both authors and journals by reducing errors
+and time spent specifying the environment.
 
@@
 language and understanding of good practices is still taking shape.
-Also, the ease with which one can create a complex software systems,
-such as an independent Docker image stack, to serve one's specific needs
-leads to some fragmentation. Furthermore fragmentation may not be a bad
-sign but instead a reflection of a growing market, which is able to
-sustain multiple related efforts. With the maturing of core building
-blocks, such as the Rocker suite of images, more systems will be built
-successfully but will also be behind the curtains. Docker alone, as a
-flexible core technology, is not a feasible level of collaboration and
-abstraction. Instead, the use cases and applications observed in this
-work provide a more useful division.
-
-Still, at least \emph{on the level of R packages} some consolidation
-seems in order, e.g., to reduce the number of packages creating
-\texttt{Dockerfile}s from R code or controlling the Docker daemon with R
-code. It remains to be seen which approach to control Docker, via the
-Docker API as \texttt{stevedore} or via system calls as
-\texttt{dockyard}/\texttt{docker}/\texttt{dockr}, is more sustainable,
-or if the question will be answered by the endurance of maintainers and
-sufficient funding. Similarly, the capturing of environments and their
-serialization in form of a \texttt{Dockerfile} currently happens at
-different levels of abstraction and re-use of functionality seems
-reasonable, e.g.~\texttt{liftr} could generate the environment with
-\texttt{containerit}, which in turn may use \texttt{dockerfiler} for low
-level objects representing a \texttt{Dockerfile}. In this consolidation,
-the Rocker~Project could play the role of a unifying and coordinating
-entity in the future. Though for the moment, the sign of the times
-points to more experimentation and feature growth, e.g.~images for
-GPU-based computing and artificial intelligence. Even with coding being
-more and more accepted as a required, and achievable skill, an easier
-access, for example by exposing containerisation benefits via simple
-user interfaces in the users' IDE, could be an important next step.
-Currently containerisation happens more in the background at the system
-level.
+Also, the ease with which one can create a complex software systems with
+Docker, such as an independent Docker image stack, to serve one's
+specific needs leads to parallel developments. This ease-of-DIY in
+combination with the difficulty to compose environments based on
+\texttt{Dockerfile}s alone is a further reason for
+\textbf{fragmentation}. Instructions can be outsourced into
+distributable scripts and then copied into the image during build, but
+that make \texttt{Dockerfile} hard to read and adds a layor of
+complexity. Despite the different image stacks presented here, the
+pervasiveness of Rocker can be traced back to its maintainers and the
+user community valueing collaboration and shared starting points over
+impulses to create individual solutions. Aside from that, fragmentation
+may not be a bad sign, but instead a reflection of a growing market,
+which is able to sustain multiple related efforts. With the maturing of
+core building blocks, such as the Rocker suite of images, more systems
+will be built successfully but will also be behind the curtains. Docker
+alone, as a flexible core technology, is not a feasible level of
+collaboration and abstraction. Instead, the use cases and applications
+observed in this work provide a more useful division.
+
+Nonetheless, at least on the level of R packages some
+\textbf{consolidation} seems in order, e.g., to reduce the number of
+packages creating \texttt{Dockerfile}s from R code or controlling the
+Docker daemon with R code. It remains to be seen which approach to
+control Docker, via the Docker API as \texttt{stevedore} or via system
+calls as \texttt{dockyard}/\texttt{docker}/\texttt{dockr}, is more
+sustainable, or if the question will be answered by the endurance of
+maintainers and sufficient funding. Similarly, the capturing of
+environments and their serialization in form of a \texttt{Dockerfile}
+currently happens at different levels of abstraction and re-use of
+functionality seems reasonable, e.g. \texttt{liftr} could generate the
+environment with \texttt{containerit}, which in turn may use
+\texttt{dockerfiler} for low level objects representing a
+\texttt{Dockerfile}. In this consolidation, the Rocker~Project could
+play the role of coordinating entity. Though for the moment, the sign of
+the times points to more experimentation and feature growth, e.g.~images
+for GPU-based computing and artificial intelligence. Even with coding
+being more and more accepted as a required, and achievable skill, an
+easier access, for example by exposing containerisation benefits via
+simple user interfaces in the users' IDE, could be an important next
+step. Currently containerisation happens more in the background at the
+system level.
 
@@
 wrote sections not mentioned below, and revised all sections. DB wrote
-the section on \pkg{outsider}. DE wrote the introduction and the section
-about Containerization and Rocker. RC \& EH contributed to the section
-on interfaces for Docker in R. ME contributed to the section on
+the section on \pkg{outsider}. GD contributed the CARD.com use case. DE
+wrote the introduction and the section about Containerization and
+Rocker. RC \& EH contributed to the section on interfaces for Docker in
+R. DC contributed content on Gigantum. ME contributed to the section on
 processing and deployment to cloud services. CF wrote paragraphs about
@@
 section about \pkg{wholepunch}. NR wrote paragraphs about shared work
-environments and GPUs. LS \& NT wrote the section on Bioconductor. NX
-wrote the section on \pkg{liftr}. All authors approved the final
-version. This articles was collaboratively written at
+environments and GPUs. LS \& NT wrote the section on Bioconductor. CW
+wrote the sections on Whole Tale and contributed the publication
+reproducibility audit use case. NX wrote the section on \pkg{liftr}. All
+authors approved the final version. This articles was collaboratively
+written at
 \href{https://github.com/nuest/rockerverse-paper/}{https://github.com/nuest/rockerverse-paper/}.
@@
 funders had no role in data collection and analysis, decision to
-publish, or preparation of the manuscript.
+publish, or preparation of the manuscript. CW is supported by the Whole
+Tale projects (\url{https://wholetale.org}) funded by the US National
+Science Foundation (NSF) under award
+\href{https://www.nsf.gov/awardsearch/showAward?AWD_ID=1541450}{OAC-1541450}.
 
@@
 \address{%
+Dav Clark\\
+Gigantum, Inc.\\
+1140 3rd Street NE\\ Washington, D.C. 20002, USA\\ \orcid{0000-0002-3982-4416}\\
+}
+\href{mailto:dav@gigantum.com}{\nolinkurl{dav@gigantum.com}}
+
+\address{%
+Gergely Daroczi\\
+\\
+\orcid{0000-0003-3149-8537}\\
+}
+\href{mailto:daroczig@rapporter.net}{\nolinkurl{daroczig@rapporter.net}}
+
+\address{%
 Dirk Eddelbuettel\\
@@
 \address{%
+Craig Willis\\
+University of Illinois at Urbana-Champaign\\
+501 E. Daniel St.\\ Champaign, IL 61820, USA\\ \orcid{0000-0002-6148-7196}\\
+}
+\href{mailto:willis8@illinois.edu}{\nolinkurl{willis8@illinois.edu}}
+
+\address{%
 Nan Xiao\\

### 2020-01-22 manuscript.Rmd
expand geospatial use case
also minor fix in bibfile

@@
 
-Containers can also serve as useful playgrounds to create environments ad-hoc or to provide very specific environments that are not needed or not easily available in day-to-day development.
+Containers can also serve as useful playgrounds to create environments ad-hoc or to provide very specific environments that are not needed or not easily available in day-to-day development for the purposes of developing R packages.
 These environments may have specific versions of R, of R extension packages, and of system libraries used by R extension packages, and all of the above in a specific combination.
@@
 \citet{eddelbuettel_debugging_2019} describes an example how a Docker container was used to debug an issue with a package only occuring with a particular version of Fortran, and using tools which are not readily available on all platforms (e.g., not on macOS).
-For the [R-spatial community](https://www.r-spatial.org/), using containers is particularly useful, because of the strong integration of system libraries in core packages for geospatial data modelling and analysis.
+
+Second, the strong integration of **system libraries in core packages** in the [R-spatial community](https://www.r-spatial.org/) makes containers essential for stable and proactive development of common classes for geospatial data modelling and analysis.
+For example, GDAL \citep{gdal_2019} is a crucial library in the geospatial domain.
+GDAL is a system dependency allowing R packages such as \CRANpkg{sf}, which provides the core data model for geospatial vector data, or \CRANpkg{rgdal}, to accomodate users to be able to read and write hundreds of different spatial raster and vector formats.
+\CRANpkg{sf} and \CRANpkg{rgdal} have hundreds of indirect reverse imports and dependencies and therefore the maintainers spend a lot of efforts not to break these.
+```{r geospatial_deps, eval=FALSE, include=FALSE}
+length(tools::dependsOnPkgs(c("rgdal", "sf"), dependencies = c("most"), recursive = TRUE))
+```
 Purpose built Docker are used to prepare for upcoming releases of system libraries, individual bug reports, and for the lowest supported versions of system libraries\footnote{Cf.
@@
 
-There are also special images for identifying problems beyond the mere R code, such as **debugging R memory problems**.
+Third, there are special purpose images for identifying problems beyond the mere R code, such as **debugging R memory problems**.
 The images significantly reduce the barrier to follow complex steps for fixing memory allocation bugs \citep[cf. Section~4.3 in][]{core_writing_1999}.
@@
 
-Second, containers are useful for **testing** R code during development.
+Fourth, containers are useful for **testing** R code during development.
 To submit a package to CRAN, an R package must work with the development version of R, which must be compiled locally.
@@
 
-Third, Docker images can be used **on CI platforms** to streamline the testing of packages.
+Fifth, Docker images can be used **on CI platforms** to streamline the testing of packages.
 \citet{ye_docker_2019} describes how they speed up the process of testing by running tasks on [Travis&nbsp;CI](https://travis-ci.org/) within a container using `docker exec`, e.g., the package check or rendering of documentation.

### 2020-01-22 manuscript.Rmd
little polish of the Research Compendia section text

--- 2020-01-22 manuscript.Rmd
@@
 Many computationally intensive research projects depend on specific versions of original and third-party software packages in diverse languages, joined together to form a pipeline through which data flows.
-New releases of even just a single piece of software in this pipeline can break the entire workflow, making it difficult to find the error and for others to reuse existing pipelines, and making the original the results irreproducible and not extandable.
+New releases of even just a single piece of software in this pipeline can break the entire workflow, making it difficult to find the error and difficult for others to reuse existing pipelines.
+These breakages can make the original the results irreproducible and not extandable.
 The chance of a substantial disruption like this is high in a multi-year research project where key pieces of third-party software may have several major updates over the duration of the project.
@@
 \citet{gentleman_statistical_2007} coined the term _Research Compendium_ for a dynamic document together with supporting data and code.
-They use the R package system \cite{core_writing_1999} as for the functional prototype to structuring, validation, and distribution of research compendia.
+They use the R package system \cite{core_writing_1999} as the functional prototype for structuring, validation, and distribution of research compendia.
 This concept has been taken up and extended\footnote{See full literature list at \href{https://research-compendium.science/}{https://research-compendium.science/}.}, not the least by applying containerisation and other methods for managing computing environments---see Section&nbsp;\nameref{envs}.
@@
 A typial compendium's `Dockerfile` will pull a rocker image fixed to a specific version of R, and install R packages from the MRAN repository to ensure the package versions are tied to a specific date, rather than the most recent version.
-A future researchers can download then compendium from the repository and run the included `Dockerfile` build a new image that recreates the computational environment used to produce the original research results.
-If building the image fails, the human-readable instructions in a `Dockerfile` are a the starting point for rebuilding the environment.
+Future researchers can download the compendium from the repository and run the included `Dockerfile` to build a new image that recreates the computational environment used to produce the original research results.
+If building the image fails, the human-readable instructions in a `Dockerfile` are the starting point for rebuilding the environment.
 Further safeguarding practices are currently under development or not part of common practice yet, such as preservation of images \citep{emsley_framework_2018} and storing them alongside `Dockerfile`s \citep[cf.][]{nust_opening_2017}, or pinning of system libraries beyond the tagged base images, which may be seen as stable or dynamic depending on the applied time scale \citep[see discussion on `debian:testing` base image in][]{RJ-2017-065}.
@@
 R is gaining popularity in these communities (in a sample of 105 replication packages published by the _American Journal of Political Science_ (AJPS) over 65% use R).
-Projects such as NSF Whole Tale use the Rocker Project community images with the goal of improving the reproducibility of published research artifacts and simplifying the publication and verification process for both authors and journals by reducing errors and time spent specifying the environment.
+The NSF-funded Whole Tale project uses the Rocker Project community images with the goal of improving the reproducibility of published research artifacts and simplifying the publication and verification process for both authors and journals by reducing errors and time spent specifying the environment.
 
@@
 HN \& JN co-wrote the section on enterprise production environments (\pkg{googleComputeEngineR}, Google Cloud Run, Kubernetes).
-KR wrote the section about \pkg{wholepunch}.
+KR wrote the section about \pkg{holepunch}.
 NR wrote paragraphs about shared work environments and GPUs.

### 2020-01-23 manuscript.Rmd
combine rc and audit section, see #64 #43

@@
 
-## Research Compendia
+## Packaging research reproducibly
 \label{compendia}
@@
 The classical "paper" article is insufficient to adequately communicate the knowledge behind such research projects \citep[cf.][]{donoho_invitation_2010,marwick_how_2015}.
-\citet{gentleman_statistical_2007} coined the term _Research Compendium_ for a dynamic document together with supporting data and code.
-They use the R package system \cite{core_writing_1999} as the functional prototype for structuring, validation, and distribution of research compendia.
+\citet{gentleman_statistical_2007} coined the term **Research Compendium** for a dynamic document together with supporting data and code.
+They use the R package system \cite{core_writing_1999} as for the functional prototype to structuring, validation, and distribution of research compendia.
 This concept has been taken up and extended\footnote{See full literature list at \href{https://research-compendium.science/}{https://research-compendium.science/}.}, not the least by applying containerisation and other methods for managing computing environments---see Section&nbsp;\nameref{envs}.
@@
 
-## Publication Reproducibility Audits
-
-A recommendation of the recent National Academies' report on _Reproducibility and Replicability in Science_ is that journals "consider ways to ensure computational reproducibility for publications that make claims based on computations" \citep{NASEM2019}.
+A recommendation of the recent National Academies' report on _Reproducibility and Replicability in Science_ is that journals _"consider ways to ensure computational reproducibility for publications that make claims based on computations"_ \citep{NASEM2019}.
 In fields such as Political Science and Economics, journals are increasingly adopting policies requiring authors to publish the code and data required to reproduce computational findings reported in published manuscripts, subject to independent verification \citep{Jacoby2017,Vilhuber2019,Alvarez2018,Christian2018,Eubank2016,King1995}.
 Problems with the computational environment, installation and availability of software dependencies are common.
-R is gaining popularity in these communities (in a sample of 105 replication packages published by the _American Journal of Political Science_ (AJPS) over 65% use R).
+R is gaining popularity in these communities, such as creating a research compendium.
+In a sample of 105 replication packages published by the _American Journal of Political Science_ (AJPS) over 65% use R.
 The NSF-funded Whole Tale project uses the Rocker Project community images with the goal of improving the reproducibility of published research artifacts and simplifying the publication and verification process for both authors and journals by reducing errors and time spent specifying the environment.

### 2020-01-23 manuscript.Rmd
reorder whole tale content

@@
 
-**Whole Tale** (https://wholetale.org) is a National Science Foundation (NSF) funded project developing a scalable, open-source, multi-user platform for reproducible research \citep{brinckman2019, chard2019a}.
-A central goal of the platform is to enable researchers to easily create and publish executable research objects\footnote{In Whole Tale a `tale` is a research object that contains metadata, data (by copy or reference), code, narrative, documentation, provenance, and information about the computational environment to support computational reproducibility} associated with published research \citep{chard2019b}.
-For R users, Whole Tale attempts to combine the strengths of the Rocker Project's curated Docker images with the Jupyter Project's `repo2docker`\footnote{See \href{https://repo2docker.readthedocs.io/}{https://repo2docker.readthedocs.io/}} component to simplify the customization of R-based environments for researchers with limited experience with either Docker or Git.
-Using Whole Tale, researchers can create and publish Rocker-based reproducible research objects to a growing number of repositories including DataONE member nodes, Zenodo and soon Dataverse.
-Additionally, Whole Tale supports automatic data citation and is working on capabilities for image preservation and provenance capture to improve the transparency of published computational research artifacts \citep{mecum2018, mcphillips2019}.
+**Whole&nbsp;Tale** (https://wholetale.org) is a National Science Foundation (NSF) funded project developing a scalable, open-source, multi-user platform for reproducible research \citep{brinckman2019, chard2019a}.
+A central goal of the platform is to enable researchers to easily create and publish executable research objects\footnote{In Whole~Tale a _tale_ is a research object that contains metadata, data (by copy or reference), code, narrative, documentation, provenance, and information about the computational environment to support computational reproducibility} associated with published research \citep{chard2019b}.
+Using Whole&nbsp;Tale, researchers can create and publish Rocker-based reproducible research objects to a growing number of repositories including DataONE member nodes, Zenodo and soon Dataverse.
+Additionally, Whole&nbsp;Tale supports automatic data citation and is working on capabilities for image preservation and provenance capture to improve the transparency of published computational research artifacts \citep{mecum2018, mcphillips2019}.
+For R users, Whole&nbsp;Tale attempts to combine the strengths of the Rocker Project's curated Docker images with the Jupyter Project's `repo2docker`\footnote{See \href{https://repo2docker.readthedocs.io/}{https://repo2docker.readthedocs.io/}} component to simplify the customization of R-based environments for researchers with limited experience with either Docker or Git.
 Multiple options have been discussed to allow users to change the base image used in `repo2docker` from the default Ubuntu LTS required to support the Rocker Project images. 
-Whole Tale implemented a custom `RockerBuildPack`\footnote{See \href{https://github.com/whole-tale/repo2docker\_wholetale}{https://github.com/whole-tale/repo2docker\_wholetale}} to support customization of the `rocker/geospatial` image through `repo2docker` composability \footnote{Composability refers to the ability to combine multiple package managers -- such as R, `pip`, and `conda`}.
-This works in part because Rocker images are based on a Debian distribution.
+Whole&nbsp;Tale implemented a custom `RockerBuildPack`\footnote{See \href{https://github.com/whole-tale/repo2docker\_wholetale}{https://github.com/whole-tale/repo2docker\_wholetale}} to support customization of the `rocker/geospatial` image through `repo2docker` composability\footnote{Composability refers to the ability to combine multiple package managers -- such as R, `pip`, and `conda`}.
+This works in part because Rocker images are based on a Debian distribution, so the instructions created by `repo2docker` for Ubuntu work because of compatible shell and package manager.
 

### 2020-01-23 manuscript.Rmd
add link to useR talk for card.com use case

@@
 Such rApache applications have been used at [rapporter.net](http://rapporter.net), transforming parametrized R Markdown-like documents into interactive web applications for reporting.
-A similar design was implemented at [CARD.com](https://www.card.com/), where the optimization of the acquisition flow and the real-time analysis of debit card transactions both relied on an efficient combination of Rserve and rApache running in Docker containers:
+A similar design was implemented at [CARD.com](https://www.card.com/), where the optimization of the acquisition flow and the real-time analysis of debit card transactions both relied on an efficient combination of \CRANpkg{Rserve} and rApache running in Docker containers:
 the R session behind Rserve acted as a read-only, extremely fast and scalabale, in-memory database for the many concurrent rApache processes responding to the live-scoring requests of various divisions of the company.
@@
 The images extend `rocker/r-base`, e.g., with an SSH client and a workaround for being able to mount SSH keys from the host, Pandoc, the Amazon Web Services (AWS) SDK, and Java, which is required by the AWS SDK.
-The AWS SDK allowed to run R consumers reading from real-time data processing streams of [AWS Kinesis](https://aws.amazon.com/kinesis/).
+The AWS SDK allowed to run R consumers reading from real-time data processing streams of [AWS Kinesis](https://aws.amazon.com/kinesis/)\footnote{See useR!2017 talk _Stream processing with R in AWS_ at \href{https://static.sched.com/hosted_files/user2017/2f/AWR%20Kinesis%20at%20useR%202017.pdf}{https://static.sched.com/hosted_files/user2017/2f/AWR Kinesis at use 02017.pdf}.}.
 The applications were deployed on Amazon Elastic Container Service ([ECS](https://aws.amazon.com/ecs/)).

### 2020-01-23 manuscript.Rmd
some small changes, formatting fixes

@@
 The output of container executions are converted to appropriate R types.
-The Docker CLI, while evolving quickly and not with great concern for introducing breaking changes, basic functionality is unchanged for a long time so `harbor::docker_run(image = "hello-world")` works like a charm.
-
+The Docker CLI's basic functionality, while evolving quickly and with small concern for avoiding breaking changes, is unchanged for a long time so a core function such as `harbor::docker_run(image = "hello-world")` still works depsite the stopped development.
 ```{r harbor, eval=FALSE, include=FALSE}
@@
 For any R projects that use the Tidyverse suite \citep{wickham_welcome_2019}, the time and resources required to build all dependencies from source can often time out before completion, making it frustrating for the average R user.
-`holepunch` removes some of these limitations by leveraging Rocker images that contain the Tidyverse along special Jupyter dependencies, and only installs additional packages from CRAN and Bioconductor that are not already part of these images.
+\pkg{holepunch} removes some of these limitations by leveraging Rocker images that contain the Tidyverse along special Jupyter dependencies, and only installs additional packages from CRAN and Bioconductor that are not already part of these images.
 It short cicuits the configuration file parsing in `repo2docker` and starts with the Binder/Tidyverse base images, which eliminates a large part of the build time and in most cases results in a binder instance launching within a minute.
-`holepunch` as a side effect also creates a `DESCRIPTION` file which then turns any project into a research compendium \citep{marwick_packaging_2018}.
+\pkg{holepunch} as a side effect also creates a `DESCRIPTION` file which then turns any project into a research compendium \citep{marwick_packaging_2018}.
 The `Dockerfile` included with the project can also be used to launch a RStudio server locally, i.e., independent of Binder, which is especially useful when more or special computational resources can be provided there.
@@
 In any of these cases, users need a method to interact with the containers, be it and IDE, or command-like access and tools such as SSH, which is usually not part of standard container recipes and must be added.
-The Rocker&nbsp;Project provides containers pre-installed with the RStudio IDE.
+The Rocker&nbsp;Project provides containers pre-installed with the RStudio&nbsp;IDE.
 In cases where users store nontrivial amounts of data for their projects, data needs to persist beyond the life of the container.
@@
 R is a fantastic tool when it comes to interfacing with databases: almost every open source and proprietary database system has an R package that allows users to connect and interact with the it 
-This flexibility is even broader now that we have tools like `DBI`, that allows to create a common API for interfacing these databases, or like `dbplyr`, which are designed to run `dplyr` code straight against the database.
+This flexibility is even broader now that we have tools like \CRANpkg{DBI}, that allows to create a common API for interfacing these databases, or like \CRANpkg{dbplyr}, which are designed to run \CRANpkg{dplyr} code straight against the database.
 But learning and teaching these tools comes with a cost: the cost of deploying or having access to an environment with the softwares and drivers installed.
 For people teaching R, it can become a barrier if they need to install local versions of the drivers, or to connect to remote instances which might or might not be made available by IT services.
-Giving access to a sandbox for the most common database environments is the idea behind **[`r-db`](https://github.com/ColinFay/r-db)** [@ColinFay], a Docker image that contains everything needed to connect, from R, to a database.
+Giving access to a sandbox for the most common database environments is the idea behind [`r-db`](https://github.com/ColinFay/r-db) [@ColinFay], a Docker image that contains everything needed to connect, from R, to a database.
 Notably, with `r-db`, the users don't have to install complex drivers or to configure their machine in a specific way.
@@
 
-Another way this idea of a common environment and the partitioning that Docker provides is used in teaching is in the **[Dodona](https://dodona.ugent.be)** project.
+Another way this idea of a common environment and partitioning containers in teaching is in the **[Dodona](https://dodona.ugent.be)** project.
 Dodona is a web platform developed at Ghent University and is used to teach students basic programming skills.
@@
 In a sample of 105 replication packages published by the _American Journal of Political Science_ (AJPS) over 65% use R.
-The NSF-funded Whole Tale project uses the Rocker Project community images with the goal of improving the reproducibility of published research artifacts and simplifying the publication and verification process for both authors and journals by reducing errors and time spent specifying the environment.
+The NSF-funded Whole Tale project uses the Rocker Project community images with the goal of improving the reproducibility of published research artifacts and simplifying the publication and verification process for both authors and reviewers by reducing errors and time spent specifying the environment.
 
@@
 Nonetheless, at least on the level of R packages some **consolidation** seems in order, e.g., to reduce the number of packages creating `Dockerfile`s from R code or controlling the Docker daemon with R code.
-It remains to be seen which approach to control Docker, via the Docker API as `stevedore` or via system calls as `dockyard`/`docker`/`dockr`, is more sustainable, or if the question will be answered by the endurance of maintainers and sufficient funding.
-Similarly, the capturing of environments and their serialization in form of a `Dockerfile` currently happens at different levels of abstraction and re-use of functionality seems reasonable, e.g., `liftr` could generate the environment with `containerit`, which in turn may use `dockerfiler` for low level objects representing a `Dockerfile`.
+It remains to be seen which approach to control Docker, via the Docker API as \pkg{stevedore} or via system calls as \pkg{dockyard}/\pkg{docker}/\pkg{dockr}, is more sustainable, or if the question will be answered by the endurance of maintainers and sufficient funding.
+Similarly, the capturing of environments and their serialization in form of a `Dockerfile` currently happens at different levels of abstraction and re-use of functionality seems reasonable, e.g., \pkg{liftr} could generate the environment with \pkg{containerit}, which in turn may use \pkg{dockerfiler} for low level R objects representing a `Dockerfile` and its instructions.
 In this consolidation, the Rocker&nbsp;Project could play the role of coordinating entity.

### 2020-01-23 manuscript.Rmd
join ci and deployment sections, split up RStudio product mentions, shorten AzureContainers

@@
 
-## Deployment to the cloud
+## Continuous integration, continuous delivery, and deployment
+\label{cicd}
 \label{deployment}
@@
 The cloud is the natural environment of containers, and is the natural mechanism to deploy R server applications.
+More and more continuous integration (CI) and continuous delivery (CD) services also use containers, opening up new options for use.
+The controlled nature of containers, i.e., the possibility to define the software environment very well, even on remote machines, is crucial here, for example to match test or build environments with production environments.
 
@@
 \label{azurecontainers}
-**\CRANpkg{AzureContainers}** is an umbrealla package providing interfaces the three services of Microsoft's Azure Cloud, namely [Container Instances](https://azure.microsoft.com/en-us/services/container-instances/) for running individual containers, [Container Registry](https://azure.microsoft.com/en-us/services/container-registry/) for private image distribution, and [Kubernetes Service](https://azure.microsoft.com/en-us/services/kubernetes-service/) for orchestrated deployments.
+**\CRANpkg{AzureContainers}** is an umbrella package providing interfaces three commercial services of Microsoft's Azure Cloud, namely [Container Instances](https://azure.microsoft.com/en-us/services/container-instances/) for running individual containers, [Container Registry](https://azure.microsoft.com/en-us/services/container-registry/) for private image distribution, and [Kubernetes Service](https://azure.microsoft.com/en-us/services/kubernetes-service/) for orchestrated deployments.
 While a package like \CRANpkg{plumber} provides the infrastructure for turning an R workflow a service, for production purposes it is usually necessary to take into account scalability, reliability and ease of management.
-A single container is limited in the volume of requests it can service, and if the container goes down or is occupied with processing a computationally intensive task, the service becomes unavailable.
-These issues can be mitigated by deploying not a single container, but a _cluster_ of containers, orchestrated as a single deployment, e.g., with [Kubernetes](https://kubernetes.io).
+A _cluster_ of containers, orchestrated as a single deployment, e.g., with [Kubernetes](https://kubernetes.io), can mitigate limitations on request volumes or a container occupied with computationally intensive task.
 A cluster features load-balancing, autoscaling of containers across numerous servers (in the cloud or on premise), and restarting failed ones.
-AzureContainers provides a lightweight yet powerful interface for working with Resource Manager, the framework for deploying and managing arbitrary resources in Azure.
-On the client side, it provides simple shells to the Docker, and to Kubernetes' `kubectl` and `helm` tools, that are needed for managing Kubernetes clusters.
-The package simplifies complex infrastructure management to a number of R funcction calls, given an Azure account with sufficient credit\footnote{See \emph{"Deploying a prediction service with Plumber"} vignette for details:  \href{https://cran.r-project.org/web/packages/AzureContainers/vignettes/vig01_plumber_deploy.html}{https://cran.r-project.org/web/packages/AzureContainers/vignettes/vig01\_plumber\_deploy.html}.}.
-
-## Continuous integration and continuous delivery [@noamross, @ColinFay]
-\label{cicd}
+AzureContainers provides an R-based interface to these features for Azure.
+On the client side, it provides simple wrappers to Docker and to Kubernetes' management tools `kubectl` and `helm`.
+The package simplifies complex infrastructure management to a number of R function calls, given an Azure account with sufficient credit\footnote{See \emph{"Deploying a prediction service with Plumber"} vignette for details:  \href{https://cran.r-project.org/web/packages/AzureContainers/vignettes/vig01_plumber_deploy.html}{https://cran.r-project.org/web/packages/AzureContainers/vignettes/vig01\_plumber\_deploy.html}.}.
+
+The prevelance of Docker in industry naturally leads to usage of containers with R in such settings as well, as customers already manage platforms in Docker containers.
+These products often entail a high amount of open source software in combination with proprietary layers addign the relevant commercializable features.
+One such example is RStudio's data science platform [RStudio Team](https://rstudio.com/products/team/).
+It allows teams of data scientists and their respective IT/DevOps groups to develop and deploy code in R and Python around the RStudio Open Source Server inside of Docker images, without requiring users to learn new tools or directly interact with containers.
+The best practices for [running RStudio with Docker containers](https://support.rstudio.com/hc/en-us/articles/360021594513-Running-RStudio-with-Docker-containers) as well as [Docker images](https://github.com/rstudio/rstudio-docker-products) for RStudio's commerical products are openly available.
 
@@
 
-## Using Docker in a commercial data science platform
+<!-- ## Commercial data science platforms
 \label{rstudio}
-
-In addition to embedding RStudio Open Source Server inside of Docker images, such as Rocker, RStudio offers the moduler data science platform [RStudio Team](https://rstudio.com/products/team/), a commercial product allowing teams of data scientists and their respective IT/DevOps groups to develop and deploy code in R and Python.
-RStudio team enables these groups to use the benefits of containers, without requiring users to learn new tools or directly interact with containers.
-Notable patterns include:
-
-1. _To deploy and manage RStudio Professional Products:_
-Many RStudio customers are managing professional products in Docker containers as an alternative to bare-metals servers or virtual machines.
-Best practices for [running RStudio with Docker containers](https://support.rstudio.com/hc/en-us/articles/360021594513-Running-RStudio-with-Docker-containers) as well as [Docker images](https://github.com/rstudio/rstudio-docker-products) for RStudio's commerical products are available.
-1. _To run interactive RStudio sessions on a cluster:_
-In the RStudio Server Pro 1.2 release in 2019, RStudio added new functionality called [Launcher](https://solutions.rstudio.com/launcher/overview/).
-It gives users the ability to spawn R sessions and background jobs in a scalable way on external clusters, e.g., [Kubernetes based on Docker images](https://support.rstudio.com/hc/en-us/articles/360019253393-Using-Docker-images-with-RStudio-Server-Pro-Launcher-and-Kubernetes) or [Slurm](https://slurm.schedmd.com/) clusters, and optionally, with Singularity containers.
-A key benefit of Launcher is the ability for R and Python users to interactively work in RStudio without learning about Docker at all - while still leveraging containers and Kubernetes.
-Users can submit batch jobs to Kubernetes clusters without writing specific deployment scripts.
-
 [RStudio Connect](https://rstudio.com/products/connect/) is a commercial publishing platform to deploy dashboards, REST APIs, and other R and Python content into production.
 RStudio Connect uses an approach to isolate deployed R applications that is similar to Docker's filesystem and process namespacing and leverages the [RStudio Package Manager](https://rstudio.com/products/package-manager/) to manage R environments including system dependencies.
-RStudio Package manager may also be used to [manage their R environments and package versions with Docker](https://environments.rstudio.com/docker) and to ensure deterministic image builds.
+-->
 

### 2020-01-23 manuscript.Rmd
shorten liftr section and move dockerfile stuff

@@
 
-## Capture and create environments [@nuest]
+## Capture and create environments
 \label{envs}
@@
 Several second order R packages attempt to make the process of creating Docker images and using containers for specific tasks, such as running tests or rendering reproducible reports, easier.
+While authoring and managing an environment with Docker by hand is possible and feasible for experts\footnote{See, e.g., this tutorial by RStudio on how to manage environments and package versions and to ensure deterministic image builds with Docker: \href{https://environments.rstudio.com/docker}{https://environments.rstudio.com/docker}.}, the following examples show the power of automation when environments become too cumbersome or acquiring the skills is not possible in due course.
 
 **\pkg{dockerfiler}** ([https://github.com/ColinFay/dockerfiler/](https://github.com/ColinFay/dockerfiler/)) ...
+At the same time, doing it by hand prevents from using programming language for what they are good for: iteration, and automation.
+Iteration and automation for `Dockerfile`s is the very reason behind [`dockerfiler`](https://github.com/ColinFay/dockerfiler), an R package designed for building `Dockerfile`s straight from R.
+For example, the **[`golem`](https://github.com/ThinkR-open/golem)** package makes an heavy use of `dockerfiler` when it comes to creating the `Dockerfile` for building production-grade Shiny applications and deploying them.
+The reason behind this is that `dockerfiler`, on top of being scriptable from R, can leverage all the tools available in R to parse a `DESCRIPTION` file, to get system requirements, to list dependencies, versions, etc.
 
@@
 
-**\CRANpkg{liftr}** ([https://nanx.me/liftr/](https://nanx.me/liftr/)) \citep{liftr2019} aims to solve the problem of persistent reproducible reporting in statistical computing.
+**\CRANpkg{liftr}** aims to solve the problem of persistent reproducible reporting in statistical computing\citep[\href{https://nanx.me/liftr/}{https://nanx.me/liftr/}, ][]{liftr2019}.
 Currently, the R Markdown format and its backend compilation engine _knitr_ offer a _de facto_ standard for creating dynamic documents \citep{xie2018}.
-However, the reproducibility of such content authoring environments is often limited to individual machines --- it is not easy to replicate the system environment (libraries, R versions, R packages) where the document was compiled.
-This issue becomes even more serious when it comes to collaborative document authoring and creating large-scale document building services.
-_liftr_ solves this reproducibility problem by bringing Docker to the game.
-In essence, _liftr_ helps R Markdown users create and manage Docker containers for rendering the documents, thus make the computations utterly reproducible across machines and systems.
-
-On implementation, with no side effects, _liftr_ extended and introduced new metadata fields to R Markdown, allowing users to declare the dependencies for rendering the document.
-_liftr_ parses such fields and generates a `Dockerfile` for creating Docker containers.
-_liftr_ then helps render the document inside the created Docker container.
-This workflow is summarized in Figure \ref{figure:liftr}.
-
-\begin{figure}[htbp]
-  \centering
-  \includegraphics[width=\textwidth]{liftr-workflow}
-  \caption{The liftr workflow for rendering containerized R Markdown documents.}
-  \label{figure:liftr}
-\end{figure}
-
-_knitr_ and R Markdown are used as the template engine to generate the `Dockerfile`.
-Features such as caching container layers for saving image build time, automatic housekeeping for fault-tolerant builds, and Docker status check are supported by _liftr_.
-Four RStudio addins are also offered by _liftr_ to allow push-button compilation of documents and provide better IDE integrations.
-
-Three basic principles are followed to design the _liftr_ package since its inception.
-
-1. Continuous reproducibility.
-Continuous integration, continuous delivery, and continuous deployment are well-accepted practices in software engineering.
-Similarly, it is believed by the authors of _liftr_ that ensuring computational reproducibility means a continuous process instead of creating static data/code archives or a one-time deal.
-Specifically, the software packages used in data analysis should be upgraded regularly in a manageable way.
-Therefore, _liftr_ supports specifying particular versions of package dependencies, while users are encouraged to always use the latest version of packages (without a version number) by default.
-
-2. Document first.
-Many data analysis workflows could be wrapped as either R packages or dynamic documents.
-In _liftr_, the endpoint of dynamic report creation is the focus of containerization, because this offers more possibilities for organizing both computations and documentation.
-Users are encouraged to start thinking from the visible research output from the first day.
-
-3. Minimal footprint.
-R Markdown and Docker are already complex software systems.
-Making them work together seamlessly can be complicated.
-Therefore, API designs such as function arguments are simplified while being kept as expressive and flexible as possible.
-
-In summary, _liftr_ tries to redefine the meaning of computational reproducibility by offering system-level reproducibility for data analysis.
-It provided a practical way for achieving it --- a new perspective on how reproducible research could be done in reality.
-Further, sharing system environments for data analysis also becomes extremely easy, since users only need to share the R Markdown document (with a few extra metadata fields), and compile them with _liftr_.
-As an example, _liftr_ demonstrated its advantage for R Markdown-based computational workflow orchestration, by effortlessly containerizing 18 complex _Bioconductor_ workflows in the DockFlow project (https://dockflow.org) in 2017.
-
-One of the key tool of Docker are `Dockerfile`s, which can be thought of as "recipes" used to build a specific image.
-But building these files can feel like a workflow break as it demands to open a separate file, and / or to write a small wrapper around R's `write()` function to add elements to the file.
-At the same time, doing it by hand prevents from using programming language for what they are good for: iteration, and automation.
-Iteration and automation for `Dockerfile`s is the very reason behind [`dockerfiler`](https://github.com/ColinFay/dockerfiler), an R package designed for building `Dockerfile`s straight from R.
-
-For example, the **[`golem`](https://github.com/ThinkR-open/golem)** package makes an heavy use of `dockerfiler` when it comes to creating the `Dockerfile` for building production-grade Shiny applications and deploying them.
-The reason behind this is that `dockerfiler`, on top of being scriptable from R, can leverage all the tools available in R to parse a `DESCRIPTION` file, to get system requirements, to list dependencies, versions, etc.
+The reproducibility authoring environments is often limited to individual machines and becomes an issue for collaborative documents and large-scale platforms for processing documents.
+\CRANpkg{liftr} helps R Markdown users to create and manage Docker containers for rendering the documents.
+It introduced new fields to the R Markdown document header, allowing users to declare the dependencies, including versions, for rendering the document.
+The package then generates a `Dockerfile` from this metadata and provides a utitility function to render the document inside a Docker container, e.g., `render_docker("foo.Rmd")`.
+An RStudio addins even allows compilation of documents with a single push of a button.
+\CRAKpkg{liftr} makes the document the main and sole workflow control file and the only file that needs to be shared between collaborators for consistent workflow environments, e.g.
+demonstrated in the DockFlow project (https://dockflow.org).
 

### 2020-01-23 manuscript.Rmd
update environments section

@@
 While authoring and managing an environment with Docker by hand is possible and feasible for experts\footnote{See, e.g., this tutorial by RStudio on how to manage environments and package versions and to ensure deterministic image builds with Docker: \href{https://environments.rstudio.com/docker}{https://environments.rstudio.com/docker}.}, the following examples show the power of automation when environments become too cumbersome or acquiring the skills is not possible in due course.
+Especially _version pinning_, with packages \pkg{remotes} and \pkg{versions} for R or by using MRAN, and with system package managers for different operating systems, can greatly increase the reproducibility of built images and are common approaches.
 
-**\pkg{dockerfiler}** ([https://github.com/ColinFay/dockerfiler/](https://github.com/ColinFay/dockerfiler/)) ...
-At the same time, doing it by hand prevents from using programming language for what they are good for: iteration, and automation.
-Iteration and automation for `Dockerfile`s is the very reason behind [`dockerfiler`](https://github.com/ColinFay/dockerfiler), an R package designed for building `Dockerfile`s straight from R.
-For example, the **[`golem`](https://github.com/ThinkR-open/golem)** package makes an heavy use of `dockerfiler` when it comes to creating the `Dockerfile` for building production-grade Shiny applications and deploying them.
-The reason behind this is that `dockerfiler`, on top of being scriptable from R, can leverage all the tools available in R to parse a `DESCRIPTION` file, to get system requirements, to list dependencies, versions, etc.
+**\pkg{dockerfiler}** ([https://github.com/ColinFay/dockerfiler/](https://github.com/ColinFay/dockerfiler/)) is an R package designed for building `Dockerfile`s straight from R.
+A scripted creation of a `Dockerfile` enables iteration and automation, for example for packaging applications for deployment (see \nameref{deployment}).
+Being scriptable from R developers can leverage the tools available in R to parse a `DESCRIPTION` file, to get system requirements, to list dependencies, versions, etc.
+**\pkg{containerit}** ([https://github.com/o2r-project/containerit/](https://github.com/o2r-project/containerit/)) attempts to take this one step further and includes these tools to automatically create a `Dockerfile` that can execute a given workflow \citep{nust_containerit_2019}.
+\pkg{containerit} accepts R `sessionInfo` objects as input and provides helper functions to derive these from workflows, e.g., an R script or R Markdown document, by analysing the session state at the end of the workflow.
+It relies on the \pkg{sysreqs} ([https://github.com/r-hub/sysreqs/](https://github.com/r-hub/sysreqs/)) package and it's mapping of package system dependencies to platform specific installation package names\footnote{See \href{https://sysreqs.r-hub.io/}{https://sysreqs.r-hub.io/}.}.
+\pkg{containerit} uses \CRANpkg{stevedore} to streamline the user interaction and improve the created `Dockerfile`s, e.g., by running a container for the desired base image to extract the already available R packages.
+**\pkg{dockr}** ([https://github.com/smaakage85/dockr](https://github.com/smaakage85/dockr)) is a very similar package attempting to mirror a given R session, including local non-CRAN packages\citep{kjeldgaard_dockr_2019}.
+Users can manually add statements for non-R dependencies to the `Dockerfile`.
+**\CRANpkg{liftr}** aims to solve the problem of persistent reproducible reporting in statistical computing based on the R Markdown format \citep{xie2018} for dynamic documents \citep[\href{https://nanx.me/liftr/}{https://nanx.me/liftr/}, ][]{liftr2019}.
+The irreproducibility of authoring environments can become an issue for collaborative documents and large-scale platforms for processing documents.
+\CRAKpkg{liftr} makes the document the main and sole workflow control file and the only file that needs to be shared between collaborators for consistent environments, e.g.
+demonstrated in the DockFlow project (https://dockflow.org).
+It introduces new fields to the R Markdown document header, allowing users to manually declare the dependencies, including versions, for rendering the document.
+The package then generates a `Dockerfile` from this metadata and provides a utitility function to render the document inside a Docker container, i.e., `render_docker("foo.Rmd")`.
+An RStudio addins even allows compilation of documents with a single push of a button.
 
-**\pkg{containerit}** ([https://github.com/o2r-project/containerit/](https://github.com/o2r-project/containerit/)) ...
-
-**\pkg{dockr}** ([https://github.com/smaakage85/dockr](https://github.com/smaakage85/dockr)) ...
-see/cite http://smaakage85.netlify.com/2019/12/21/dockr-easy-containerization-for-r/
-<!-- from description on CRAN: Build a lightweight Docker r-base container image for your R package, in which the package itself is available.
-The Docker image seeks to mirror your R session as close as possible with respect to R specific dependencies.
-Both dependencies on CRAN R packages as well as local non-CRAN R packages will be included in the Docker container image.-->
-
-**\CRANpkg{renv}** ([https://rstudio.github.io/renv/](https://rstudio.github.io/renv/)) with Docker: https://rstudio.github.io/renv/articles/docker.html
-
-_version pinning_ with packages \pkg{remotes} and \pkg{versions} for R, and with system package managers for distros
-
-**\CRANpkg{liftr}** aims to solve the problem of persistent reproducible reporting in statistical computing\citep[\href{https://nanx.me/liftr/}{https://nanx.me/liftr/}, ][]{liftr2019}.
-Currently, the R Markdown format and its backend compilation engine _knitr_ offer a _de facto_ standard for creating dynamic documents \citep{xie2018}.
-The reproducibility authoring environments is often limited to individual machines and becomes an issue for collaborative documents and large-scale platforms for processing documents.
-\CRANpkg{liftr} helps R Markdown users to create and manage Docker containers for rendering the documents.
-It introduced new fields to the R Markdown document header, allowing users to declare the dependencies, including versions, for rendering the document.
-The package then generates a `Dockerfile` from this metadata and provides a utitility function to render the document inside a Docker container, e.g., `render_docker("foo.Rmd")`.
-An RStudio addins even allows compilation of documents with a single push of a button.
-\CRAKpkg{liftr} makes the document the main and sole workflow control file and the only file that needs to be shared between collaborators for consistent workflow environments, e.g.
-demonstrated in the DockFlow project (https://dockflow.org).
+The package **\CRANpkg{renv}** ([https://rstudio.github.io/renv/](https://rstudio.github.io/renv/)) helps users to manage the state of the R library in a reproducible way, further providing isolation and portability \citep{renv2019}.
+The package does not cover system dependencies, though, the \CRANpkg{renv}-based environment can be transferred into a container either by restoring the environment based on the main configuration file `renv.lock` or by storing the \CRANpkg{renv}-cache on the host and not in the container \citep{ushey_using_2019}.
 

### 2020-01-23 manuscript.Rmd
update enterprise section
restructure, shorten a bit

@@
 
-R has been historically viewed as a tool for analysis and scientific research--not for creating software that corporations can rely on to continously run in real time.
+R has been historically viewed as a tool for analysis and scientific research, but not for creating software that corporations can rely on for production services.
 However, thanks to advancements in R running as a web service, along with with the ability to deploy R in Docker containers, modern enterprises are now capable of having real-time machine learning powered by R.
-One of the first implementations of having R respond to client reqests were TCP/IP and local socket servers, such as \CRANpkg{Rserve} and \CRANpkg{svSocket}, which allowed other local or remote processes to make binary requests to interact with R without the need to link to R at compile time and connected other programming languages with R.
-R could also empower web services since at least 2007, when [rApache](http://www.rapache.net) was presented at the useR! conference \citep{horner_rapache_2013}.
-<!-- website seems down, https://web.archive.org/web/20190627101203/http://rapache.net/manual.html -->
-rApache runs an R interpreter inside the Apache web server, which makes it possible and very efficient to build scalable R-based APIs and web services.
+A number of packages and projects enabled R to respond to client reqests over TCP/IP and local socket servers, such as \CRANpkg{Rserve}, \CRANpkg{svSocket}, [rApache](http://www.rapache.net) and more recently \CRANpkg{plumber} ([https://www.rplumber.io/](https://www.rplumber.io/)) and \pkg{RestRserve} ([http://restrserve.org](http://restrserve.org)), which even processes incoming requests in parallel with forked processes using \CRANpkg{Rserve}.
+The latter two also provide documentation for deployment with Docker or ready to use automated builds of images\footnote{See \href{https://www.rplumber.io/docs/hosting.html#docker}{https://www.rplumber.io/docs/hosting.html#docker}, \href{https://hub.docker.com/r/trestletech/plumber/}{https://hub.docker.com/r/trestletech/plumber/} and \href{https://hub.docker.com/r/rexyai/restrserve/}{https://hub.docker.com/r/rexyai/restrserve/}.}.
+These software allow other (remote) processes and programming languages to interact with R and to expose R-based function in a service architecture with HTTP APIs.
+APIs based on these package can be deployed with scalability and high availability using containers.
+This pattern of deploying code matches those used by software engineering services created in more established languages in the enterprise domain, such as Java or Python, and R can be used alongside those langages as a first class member of a software engineering technical stack.
 
-Such rApache applications have been used at [rapporter.net](http://rapporter.net), transforming parametrized R Markdown-like documents into interactive web applications for reporting.
-A similar design was implemented at [CARD.com](https://www.card.com/), where the optimization of the acquisition flow and the real-time analysis of debit card transactions both relied on an efficient combination of \CRANpkg{Rserve} and rApache running in Docker containers:
-the R session behind Rserve acted as a read-only, extremely fast and scalabale, in-memory database for the many concurrent rApache processes responding to the live-scoring requests of various divisions of the company.
-Similar dockerized R scripts were responsible for the ETL processes and even the client-facing email, text message and push notification alerts sent in real-time based on card transactions.
+CARD.com implemented a web application for the optimization of the acquisition flow and the real-time analysis of debit card transactions.
+The software used \CRANpkg{Rserve} and rApache and was deployed in Docker containers.
+The R session behind \CRANpkg{Rserve} acted as a read-only in-memory database, which was extremely fast and scalabale, for the many concurrent rApache processes responding to the live-scoring requests of various divisions of the company.
+Similarly dockerized R scripts were responsible for the ETL processes and even the client-facing email, text message and push notification alerts sent in real-time based on card transactions.
 The related Docker images were made available at [https://github.com/cardcorp/card-rocker](https://github.com/cardcorp/card-rocker).
-The images extend `rocker/r-base`, e.g., with an SSH client and a workaround for being able to mount SSH keys from the host, Pandoc, the Amazon Web Services (AWS) SDK, and Java, which is required by the AWS SDK.
+The images extend `rocker/r-base` and additionally entailed an SSH client and a workaround for being able to mount SSH keys from the host, Pandoc, the Amazon Web Services (AWS) SDK, and Java, which is required by the AWS SDK.
 The AWS SDK allowed to run R consumers reading from real-time data processing streams of [AWS Kinesis](https://aws.amazon.com/kinesis/)\footnote{See useR!2017 talk _Stream processing with R in AWS_ at \href{https://static.sched.com/hosted_files/user2017/2f/AWR%20Kinesis%20at%20useR%202017.pdf}{https://static.sched.com/hosted_files/user2017/2f/AWR Kinesis at use 02017.pdf}.}.
 The applications were deployed on Amazon Elastic Container Service ([ECS](https://aws.amazon.com/ecs/)).
-The main learnings from using R in Docker was the imporance of not only pinning the R package versions via MRAN, but also moving away from Debian testing to an LTS distribution.
-For the use case at hand, the priority was to have more control over upstream updates, and to minimize the risk of breaking the automated builds of the Docker images and production jobs.
+The main learnings from using R in Docker was the imporance of not only pinning the R package versions via MRAN, but also moving away from Debian testing to a distribution with long-term support.
+For the use case at hand, this switch served the priority to have more control over upstream updates, and to minimize the risk of breaking the automated builds of the Docker images and production jobs.
 
-A more recent related work is the \CRANpkg{plumber} ([https://www.rplumber.io/](https://www.rplumber.io/)) package, which makes it even easier for R to be used to host web services through RESTful APIs.
-When R is running on a server, a different server can ask R to make a computation through an HTTP request and get an immediate response.
-For instance, if a data scientist at a company creates a machine learning model to tell which customers are likely to make a repeat purchase, if they serve the model through a plumber API then someone else on the marketing team could write software that sends different emails depending on that real-time prediction.
-Plumber syntax is simple to read and understand: the API is created by commenting above function names in an R script.
-
-```
-#* Return the sum of two numbers
-#* @get /random
-function(){
-  runif()
-}
-```
-_Example of a plumber API to generated a random number by visiting `http://127.0.0.1/random`_
-
-Plumber becomes far more powerful with the addition of Docker.
-Docker can be used to create images that [include both R and plumber](https://www.rplumber.io/docs/hosting.html#docker).
-Then when the image is deployed, that creates the web endpoints that other services can hit and call the R code.
-This pattern of deploying code mimics those used by software engineering services created in Java, Python, and other modern languages.
-By using Docker to deploy plumber APIs, R can be used alongside those langages as a first class member of a software engineering technical stack.
-
-At T-Mobile, R is used to power real-time services that are called over a million times a day.
-T-Mobile created a set of neural network machine learning natural language processing models to help customer care agents manage text-based messages for customers \citep{t-mobile_enterprise_2018}.
-The models are convolutional neural networks that use the RStudio Keras package and are built on top of a Rocker Docker image.
-Since the models power tools for agents and customers, they need to have extremely high uptime and reliability--and they found that R is able to perform.
-The AI @ T-Mobile team open sourced their Docker images [on GitHub](https://github.com/tmobile/r-tensorflow-api).
-
-[RestRserve](http://restrserve.org) is an alternative R web API framework with the main focus on being high-performant and robust.
-It also ships automated Docker builds at https://hub.docker.com/r/rexyai/restrserve/dockerfile.
+The AI @ T-Mobile team created a set of neural network machine learning natural language processing models to help customer care agents manage text-based messages for customers \citep{t-mobile_enterprise_2018}.
+For example, one model quickly identifies if a message is from a customer or not \citep[see \CRANpkg{Shiny}-based \href{https://secure.message.t-mobile.com/v1/shiny/is-customer/app/}{demo}, ][]{nolis_small_2019}, others tell which customers are likely to make a repeat purchase.
+If a data scientist creates a machine learning model and exposes it through a \CRANpkg{plumber} API, then someone else on the marketing team could write software that sends different emails depending on that real-time prediction.
+The models are convolutional neural networks that use the \CRANpkg{keras} package and run in a Rocker Docker image.
+The corresponding `Dockerfile`s are published [on GitHub](https://github.com/tmobile/r-tensorflow-api).
+Since the models power tools for agents and customers, they need to have extremely high uptime and reliability.
+The AI @ T-Mobile team found that the models performed well and today these models power real-time services that are called over a million times a day.
 

### 2020-01-23 manuscript.Rmd

--- 2020-01-23 manuscript.Rmd
@@
     # https://github.com/DomBennett
-    affiliation: University of Gothenburg, Sweden
+    affiliation: Gothenburg Global Biodiversity Centre, Sweden
     address:
-      - Department of Biological and Environmental Sciences
       - Carl Skottsbergs gata 22B
-      - 413 19 Gothenburg, Sweden
+      - 413 19 Göteborg, Sweden
       - \orcid{0000-0003-2722-1359}
@@
 
-**\pkg{outsider}** ([https://antonellilab.github.io/outsider/](https://antonellilab.github.io/outsider/)) tackles the problem of integrating external programs into an R workflow \citep{bennett_outsider_2020}.
+**\pkg{outsider}** ([https://docs.ropensci.org/outsider/](https://docs.ropensci.org/outsider/)) tackles the problem of integrating external programs into an R workflow \citep{bennett_outsider_2020}.
 Installation and usage of external programs can be difficult, convoluted and even impossible if the platform is incompatible.

### 2020-01-23 README.Rmd
add ShinyProxy, update list in README

@@
 - https://github.com/jlisic/R-docker-centos because CentOS images are also provided by R-hub
-- https://github.com/cardcorp/card-rocker because last change 3 years ago, an all based on `rocker/r-base` - lesson learned would be interesting still, see [#15](https://github.com/nuest/rockerverse-paper/issues/15)
 - `sevenbridges` includes several Dockerfiles and documentation about using images, and also seems to use Docker in the built workflows but via other tools, not directly, https://sbg.github.io/sevenbridges-r/, https://github.com/sbg/sevenbridges-r/search?p=4&q=docker&unscoped_q=docker
 - https://github.com/rocker-jp because it seems to have stalled soon after inception
+- https://github.com/thomasp85/fiery because no examples/docs for Docker (though probably trivial anyway)
+- [RStudio Connect](https://rstudio.com/products/connect/) because while it does use a similar mechanism to seperate processes, it does _not_ use Docker and the comparison would require a lot of background explanation, and not all details are available because the software is proprietary

### 2020-01-23 manuscript.Rmd
add ShinyProxy, update list in README

--- 2020-01-23 manuscript.Rmd
@@
 
-The cloud is the natural environment of containers, and is the natural mechanism to deploy R server applications.
+The cloud is the natural environment of containers, and subsequently containers are the go-to mechanism to deploy R server applications.
 More and more continuous integration (CI) and continuous delivery (CD) services also use containers, opening up new options for use.
@@
 
+As a first example, \CRANpkg{Shiny} is the most popular package for creating interactive online dashboards with R and it enables users with very diverse backgrounds to create stable and user friendly web applications.
+The ShinyProxy ([https://www.shinyproxy.io/](https://www.shinyproxy.io/)) is an open source tool to deploy Shiny apps in an enterprise context.
+They feature single sign-on, but also in scientific use cases \citep[e.g., ][]{savini_epiexplorer_2019,glouzon_structurexplor_2017}.
+ShinyProxy uses Docker containers to isolate user sessions and to achieve scalability for multi-user scenarious with multiple apps.
+ShinyProxy itself is written in Java to accomodate corporate requirements and may itself run in a container for stability and availability.
+The tool is build on ContainerProxy ([https://www.containerproxy.io/](https://www.containerproxy.io/)), which provides similar features for executign long-running R jobs or interactive R sessions.
+The started containers can run on a regular Docker host but also in clusters.
 - `RSelenium`

### 2020-01-23 manuscript.Rmd
minor edits and fixes. update conclusion

@@
 For R users, Whole&nbsp;Tale attempts to combine the strengths of the Rocker Project's curated Docker images with the Jupyter Project's `repo2docker`\footnote{See \href{https://repo2docker.readthedocs.io/}{https://repo2docker.readthedocs.io/}} component to simplify the customization of R-based environments for researchers with limited experience with either Docker or Git.
-Multiple options have been discussed to allow users to change the base image used in `repo2docker` from the default Ubuntu LTS required to support the Rocker Project images. 
+Multiple options have been discussed to allow users to change the base image used in `repo2docker` from the default Ubuntu LTS (long-term support) required to support the Rocker Project images. 
 Whole&nbsp;Tale implemented a custom `RockerBuildPack`\footnote{See \href{https://github.com/whole-tale/repo2docker\_wholetale}{https://github.com/whole-tale/repo2docker\_wholetale}} to support customization of the `rocker/geospatial` image through `repo2docker` composability\footnote{Composability refers to the ability to combine multiple package managers -- such as R, `pip`, and `conda`}.
@@
 **\pkg{harbor}** ([https://github.com/wch/harbor](https://github.com/wch/harbor) is not actively maintained anymore, but should be honorably mentioned as the first R package for managing Docker images and containers.
-It uses the \CRANpkg{sys} package to run system commands against the Docker CLI, both locally and through an ssh connection, and has convenience functions, e.g., for listing and removing containers/images and for accessing logs.
+It uses the \CRANpkg{sys} package to run system commands against the Docker CLI, both locally and through an SSH connection, and has convenience functions, e.g., for listing and removing containers/images and for accessing logs.
 The output of container executions are converted to appropriate R types.
@@
 The irreproducibility of authoring environments can become an issue for collaborative documents and large-scale platforms for processing documents.
-\CRAKpkg{liftr} makes the document the main and sole workflow control file and the only file that needs to be shared between collaborators for consistent environments, e.g.
+\CRANpkg{liftr} makes the document the main and sole workflow control file and the only file that needs to be shared between collaborators for consistent environments, e.g.
 demonstrated in the DockFlow project (https://dockflow.org).
@@
 The package then generates a `Dockerfile` from this metadata and provides a utitility function to render the document inside a Docker container, i.e., `render_docker("foo.Rmd")`.
-An RStudio addins even allows compilation of documents with a single push of a button.
+An RStudio addin even allows compilation of documents with a single push of a button.
 
@@
 The classical "paper" article is insufficient to adequately communicate the knowledge behind such research projects \citep[cf.][]{donoho_invitation_2010,marwick_how_2015}.
+
 \citet{gentleman_statistical_2007} coined the term **Research Compendium** for a dynamic document together with supporting data and code.
@@
 Given the users in this scenario, i.e., often academics with limited formal software development training, templates and assistance with containers around research compendia is essential.
-
 In many fields we see that a typical unit of research for a container is a research report or journal article, where the container holds the compendium, or self-contained set of data (or connections to data elsewhere) and code files needed to fully reproduce the article \citep{marwick_packaging_2018}.
@@
 
-This article is a snapshot of the R-corner in a universe of applications built many-faced piece of software, Docker.
+This article is a snapshot of the R-corner in a universe of applications built with a many-faced piece of software, Docker.
 `Dockerfile`s and Docker images are the go-to methods for collaboration between roles in an organisation, such as development and IT operations teams, and between parties in the communication of knowledge, such as research workflows or education.
 Docker became synonymous with applying the concept of containerisation to solve challenges of reproducible environments, e.g., in research and in development \& production, and of scalable deployments with the ability to move processing between machines easily (e.g., locally, one cloud providers VM, another cloud provider's Container-as-a-Service).
-Reproducible environments, scalability and efficiency, and portability across clouds are the common themes behind R packages, use cases, and applications in this work.
+Reproducible environments, scalability and efficiency, and portability across platforms/infrastructures are the common themes behind R packages, use cases, and applications in this work.
 
-The R packages and use cases presented show the positive drive in the community to innovate, but also the challenge of keeping up with a continuously evolving landscape.
+The R packages and use cases presented show the growing number of users, developers, and real-world applications in the community and the resulting innovations.
+But the applications also point to the challenge of keeping up with a continuously evolving landscape.
 The use cases contributed by co-authors also have a degree of overlap, which can be expected as a common language and understanding of good practices is still taking shape.
@@
 New features, which make complex workflows accessible and reproducible, and the variety in packages connected with containerisation, even when they have overlapping features, are a signal and a support for a growing user base.
-This growth is possibly the most important goal for the foreseeable future in the _Rockerverse_, and just like the Rocker images have matured over years of use and millions of runs the new ideas and prototypes will have to proof themselves.
+This growth is possibly the most important goal for the foreseeable future in the _Rockerverse_, and just like the Rocker images have matured over years of use and millions of runs, the new ideas and prototypes will have to proof themselves.
 It should be noted that the dominant position of Docker is a blessing and a curse for these goals.
-It could be wise to start experimenting with non-Docker containerisation tools now, e.g., R packages interfacing with other container engines such as [podman/buildah](https://github.com/containers/libpod) or [rkt](https://coreos.com/rkt/), or an R package for creating Singularity image recipe files.
-Such efforts help to avoid lock-in and to design sustainable workflows based on concepts of _containerisation_, not on their implementation in Docker.
+It could be wise to start experimenting with non-Docker containerisation tools now, e.g., R packages interfacing with other container engines such as [podman/buildah](https://github.com/containers/libpod) or [rkt](https://coreos.com/rkt/), or an R package for creating `Singularity` files.
+Such efforts can help to avoid lock-in and to design sustainable workflows based on concepts of _containerisation_, not on their implementation in Docker.
 If adoption of containerisation and R continue to grow, the missing pieces for a success predominantly lie in (a) coordination and documentation of activities to reduce repeated work in favour of open collaboration, (b) the sharing of lessons learned from use cases to build common knowledge and language, and (c) a sustainable continuation and funding for all of development, community support, and education.

### 2020-01-23 manuscript.Rmd
update geospatial use case, move r-ci to testing section

@@
 
-## Development and debugging
+## Development, debugging, and testing
 \label{development}
@@
 
-First, such containers can greatly facilitate **fixing bugs**, because developers can readily start and later discard a container to investigate a bug report without affecting their regular system, and potentially having to revert changes back.
+First, such containers can greatly facilitate **fixing bugs and quick evaluation**, because developers and users can readily start and later discard a container to investigate a bug report or try out a piece of software \citep[cf.][]{ooms_opencpu_2017}.
+The container does not affect their regular system.
 Using the Rocker images with RStudio, these disposable environments lack no development comfort (cf. Section&nbsp;\nameref{compendia}).
+\citet{ooms_opencpu_2017} describes how `docker exec` can be used to get a root shell in a container for cusotmization during software evaluation.
 \citet{eddelbuettel_debugging_2019} describes an example how a Docker container was used to debug an issue with a package only occuring with a particular version of Fortran, and using tools which are not readily available on all platforms (e.g., not on macOS).
@@
 For example, GDAL \citep{gdal_2019} is a crucial library in the geospatial domain.
-GDAL is a system dependency allowing R packages such as \CRANpkg{sf}, which provides the core data model for geospatial vector data, or \CRANpkg{rgdal}, to accomodate users to be able to read and write hundreds of different spatial raster and vector formats.
+GDAL is a system dependen^cy allowing R packages such as \CRANpkg{sf}, which provides the core data model for geospatial vector data, or \CRANpkg{rgdal}, to accomodate users to be able to read and write hundreds of different spatial raster and vector formats.
 \CRANpkg{sf} and \CRANpkg{rgdal} have hundreds of indirect reverse imports and dependencies and therefore the maintainers spend a lot of efforts not to break these.
+Purpose built Docker are used to prepare for upcoming releases of system libraries, individual bug reports, and for the lowest supported versions of system libraries\footnote{Cf.
+\href{https://github.com/r-spatial/sf/tree/master/inst/docker}{https://github.com/r-spatial/sf/tree/master/inst/docker}, \href{https://github.com/Nowosad/rspatial_proj6}{https://github.com/Nowosad/rspatial\_proj6}, and \href{https://github.com/r-spatial/sf/issues/1231}{https://github.com/r-spatial/sf/issues/1231}}.
 ```{r geospatial_deps, eval=FALSE, include=FALSE}
@@
 ```
-Purpose built Docker are used to prepare for upcoming releases of system libraries, individual bug reports, and for the lowest supported versions of system libraries\footnote{Cf.
-\href{https://github.com/r-spatial/sf/tree/master/inst/docker}{https://github.com/r-spatial/sf/tree/master/inst/docker}, \href{https://github.com/Nowosad/rspatial_proj6}{https://github.com/Nowosad/rspatial\_proj6}, and \href{https://github.com/r-spatial/sf/issues/1231}{https://github.com/r-spatial/sf/issues/1231}}.
 
@@
 \citet{ye_docker_2019} describes how they speed up the process of testing by running tasks on [Travis&nbsp;CI](https://travis-ci.org/) within a container using `docker exec`, e.g., the package check or rendering of documentation.
-\citet{cardozo_faster_2018}, also on Travis&nbsp;CI saved time by re-using the testing image as the base iamge for an image intended for publication on Docker&nbsp;Hub.
-Especially for long running tests or complex system dependencies, these approaches seems to improve the development.
-Not due to a concern about time, but to control the environment used on a CI server, even this manuscript was rendered into a PDF and deployed to a GitHub-hosted website (see `.travis.yml` and `Dockerfile` in the manuscript repository).
+\citet{cardozo_faster_2018} saved time, also on Travis&nbsp;CI, by re-using the testing image as the base for an image intended for publication on Docker&nbsp;Hub.
+[`r-ci`](https://github.com/ColinFay/r-ci) is in turn used with [GitLab&nbsp;CI](https://docs.gitlab.com/ee/ci/), which itself is built on top of Docker images: the user specifies a base Docker image, and the whole tests are run inside this environment.
+The `r-ci` image stack combines `rocker` versioning and a series of tools specifically designed for testing in a fixed environment with a customized list of preinstalled packages.
+Especially for long running tests or complex system dependencies, these approaches to seperate installation of build dependencies with code testing streamline the development process.
+Not due to a concern about time, but to control the environment used on a CI server, even this manuscript is rendered into a PDF and deployed to a GitHub-hosted website with every change (see `.travis.yml` and `Dockerfile` in the manuscript repository).
 This gives on the one hand easy access after every update of the R Markdown source code, and on the other hand a second controlled environment making sure that the article renders successfully and correctly.

### 2020-01-23 manuscript.Rmd
WIP safe for processing and CI/CD/deployment sections

@@
 The started containers can run on a regular Docker host but also in clusters.
+
+Second, containers can be used to create **platform installation packages** in a DevOps setting.
+The [OpenCPU](https://www.opencpu.org/) system provides an HTTP API for data analysis based on R.
+\citet{ooms_opencpu_2017} describes how various platform-specific installation files for OpenCPU are created using Docker&nbsp;Hub: the automated builds install the software stack from source on different operating systems; afterwards a script file downloads the images and extracts the OpenCPU binaries.
+
+- [dynwrap](https://github.com/dynverse/dynwrap_containers/blob/master/.travis.yml) [@rcannood]
+  - For this project, we use travis-ci to build rocker-derived containers, test them, and only push them to Docker&nbsp;Hub (from travis-ci.org) if the integration tests succeed.
+
+For example, the **[`golem`](https://github.com/ThinkR-open/golem)** package makes an heavy use of `dockerfiler` when it comes to creating the `Dockerfile` for building production-grade Shiny applications and deploying them.
+
 - `RSelenium`
@@
 The best practices for [running RStudio with Docker containers](https://support.rstudio.com/hc/en-us/articles/360021594513-Running-RStudio-with-Docker-containers) as well as [Docker images](https://github.com/rstudio/rstudio-docker-products) for RStudio's commerical products are openly available.
-
-The controlled nature of containers, i.e., it is possible to define the software environment very well, even on remote machines, make them also useful for continuous integration (CI) and continuous delivery of applications.
-
-- DevOps
-  - https://www.opencpu.org/posts/opencpu-with-docker/
-
-When doing continuous integration and continuous delivery, it's crucial to test in an environment that matches the production environment.
-Tools like `Gitlab-CI` are built on top of Docker images: the user specifies a base Docker image, and the whole tests are run inside this environment.
-But, as we just said, this environment has to be fixed, but it also have to contain the necessary toolkit.
-In order to achieve that, [`r-ci`](https://github.com/ColinFay/r-ci) combines `rocker` versioning and a series of tools specifically designed for testing.
-That way, package builders can use this image as a base image for there testing environment, without having to install the necessary packages every time they need to run a new test.
-
-- [dynwrap](https://github.com/dynverse/dynwrap_containers/blob/master/.travis.yml) [@rcannood]
-  - For this project, we use travis-ci to build rocker-derived containers, test them, and only push them to Docker&nbsp;Hub (from travis-ci.org) if the integration tests succeed.
-- Google Cloud Build
-  - https://cloud.google.com/cloud-build/
-  - Cloud Build is another continuous intergation service that helps move the workload of building the Docker images to an online service.
-Cloud Build can be set up to build the Dockerfiles on each GitHub commit or release.
-This means you do not need to build the Docker images locally, which can tie up resources since Docker images can be several GBs and take a long time to compile.
-Google Cloud Build works alongside Google Container Registry to allow you to build private and public Docker images, which allows you to build up your own dependency graphs for downstream applications.
-  - https://code.markedmondson.me/googleCloudRunner/articles/cloudbuild.html
 
@@
 
-The portability of containers becomes particularly useful when complex processing tasks shall be offloaded to a server.
+The portability of containerized environments becomes particularly useful for improving expensive processing of data or shipping complex processing pipelines.
 
-- Docker images for cloud services
-  - A popular application of using Docker is to create the environments that can run in parallel for speeding up R code.
+First, it is possible to **offload complex processing to a server**.
+...
+
+Second, when processes can be **run in parallel** for speeding up R code.
+
 For example, [`googleComputeEngineR`](https://CRAN.R-project.org/package=googleComputeEngineR)'s `gce_vm_cluster()` function can create clusters of 2 or more virtual machines, running multi-CPU architectures.
@@
 An R implementation is shown here at [cloudRunR](https://github.com/MarkEdmondson1234/cloudRunR) which uses Cloud Run to create a scalable R plumber API.
-  - How does this relate to https://code.markedmondson.me/googleCloudRunner/index.html ?`
-- `batchtools` \citep{Lang2017batchtools} can [schedule jobs with Docker Swarm](https://mllg.github.io/batchtools/reference/makeClusterFunctionsDocker.html)
+https://code.markedmondson.me/googleCloudRunner/index.html
+
+[Google Cloud Build](https://cloud.google.com/cloud-build/) and the Google Container Registry are a continuous intergation service respectively image registry that offload building of images to the cloud, while serving the needs of commercial environments  that helps move the workload of building the Docker images to an online service, while  Cloud Build can be set up to build the Dockerfiles on each GitHub commit or release.
+This means you do not need to build the Docker images locally, which can tie up resources since Docker images can be several GBs and take a long time to compile.
+Google Cloud Build works alongside  to allow you to build private and public Docker images, which allows you to build up your own dependency graphs for downstream applications.
+  - https://code.markedmondson.me/googleCloudRunner/articles/cloudbuild.html
+
+- \CRANpkg{batchtools} \citep{Lang2017batchtools} can [schedule jobs with Docker Swarm](https://mllg.github.io/batchtools/reference/makeClusterFunctionsDocker.html)
 - scalable deployments, e.g., start with numerous Shiny talks mentioning Rocker at useR!2017
@@
 Again, all of these containers are being built on travis-ci, and will only be pushed to Docker&nbsp;Hub if the integration test succeeds.
-- \pkg{drake}, https://docs.ropensci.org/drake/index.html?q=docker#with-docker
+- \CRANpkg{drake}, https://docs.ropensci.org/drake/index.html?q=docker#with-docker
 
-## Packaging programs for processing pipelines
-\label{packaging}
+In the RStudio Server Pro 1.2 release in 2019, RStudio added new functionality called [Launcher](https://solutions.rstudio.com/launcher/overview/).
+It gives users the ability to spawn R sessions and background jobs in a scalable way on external clusters, e.g., [Kubernetes based on Docker images](https://support.rstudio.com/hc/en-us/articles/360019253393-Using-Docker-images-with-RStudio-Server-Pro-Launcher-and-Kubernetes) or [Slurm](https://slurm.schedmd.com/) clusters, and optionally, with Singularity containers.
+A key benefit of Launcher is the ability for R and Python users to interactively work in RStudio without learning about Docker at all - while still leveraging containers and Kubernetes.
+Users can submit batch jobs to Kubernetes clusters without writing specific deployment scripts.
 
-Docker is perfectly suited for packaging and executing any software and data, and is a good candidate for building complex processing pipelines that are independent from the host programming language.
+\label{pipelines}
+Third, containers are perfectly suited for **packaging and executing software pipelines** and required data.
+Containers allow building complex processing pipelines that are independent from the host programming language.
 Due to its original use case (see&nbsp;\nameref{introduction}), Docker has no standard mechanisms for chaining containers together; it lacks definitions and protocols for environment variables, volume mounts and/or ports that could enable transfer of input (parameters and data) and output (results) to and from containers.
 Some packages, e.g., \pkg{containerit}, do provide Docker images that can be used similarly to CLIs, but their usage is cumbersome\footnote{\href{https://o2r.info/containerit/articles/container.html}{https://o2r.info/containerit/articles/container.html}}.
-
 **\pkg{outsider}** ([https://antonellilab.github.io/outsider/](https://antonellilab.github.io/outsider/)) tackles the problem of integrating external programs into an R workflow \citep{bennett_outsider_2020}.

### 2020-01-23 manuscript.Rmd
shorten dockyard, rephrase docker-machine

@@
 
-**\pkg{dockyard}** ([https://github.com/thebioengineer/dockyard](https://github.com/thebioengineer/dockyard)) has the goal of lowering barrier to creating `Dockerfile`s and Docker images, and deploying Docker containers.
-The way docker images are made is by reading and following a set of instructions found in `Dockerfile`s.
-This requires stepping out of the R ecosystem and into an infrastructure mindset where the user needs to specify the environment they require.
-Because `Dockerfile`s can be read as a series of steps, this lends itself well to the `tidyverse` style of programming where each step is piped into the next.
-dockyard has a `Dockerfile` object that either starts out as an empty `Dockerfile`, or can use existing `Dockerfile` to use as a template.
-This combined with a series of functions that are analogous to the steps commonly found in a `Dockerfile`, it makes building Docker images much more approachable to R users.
-There are also wrappers for commonly used steps for R programmers like adding libraries and copying files into the R Docker containers.
-Once the `Dockerfile` is created, dockyard has built-in functionality to build the image and run a container.
+**\pkg{dockyard}** ([https://github.com/thebioengineer/dockyard](https://github.com/thebioengineer/dockyard)) has the goal of lowering barrier to creating `Dockerfile`s, building Docker images, and deploying Docker containers.
+The package follows the increasingly used piping paradigm of the `tidyverse` style of programming for chaining R functions representing the instructions in a `Dockerfile`.
+An existing `Dockerfile` can be used as a template.
+\pkg{dockyard} also includes wrappers for common steps, such as installing an R package or copying files, and build-in functions for building an image running a container, to make using Docker even more approachable to R users with a single API.
 
-**\pkg{dockermachine}** ([https://github.com/cboettig/dockermachine](https://github.com/cboettig/dockermachine)) is an R package to provide a convenient interface to [Docker Machine](https://docs.docker.com/machine/overview/) from R.
-Docker Machine's CLI tool `docker-machine` allows users to create and manage virtual host on local computers, local data centers, or at cloud providers.
-A local Docker installation can be configured to forward all commands issues on the local Docker CLI to a selected (remote) virtual host fully transparent.
-For local Docker Machine was especially crucial in early days of Docker, when no native Docker support was available for Mac or Windows computers, but remains relevant for provisioning on remote systems.
-The package has not received any updates for two years, but was functional with a current version of `docker-machine` (`0.16.2`) and potentially lowers the barriers for R users to run containers on various hosts, if using the Docker Machine CLI is perceived as a barrier.
-
+**\pkg{dockermachine}** ([https://github.com/cboettig/dockermachine](https://github.com/cboettig/dockermachine)) is an R package to provide a convenient interface to [Docker&nbsp;Machine](https://docs.docker.com/machine/overview/) from R.
+The CLI tool `docker-machine` allows users to create and manage virtual host on local computers, local data centers, or at cloud providers.
+A local Docker installation can be configured to transparntly forward all commands issued on the local Docker CLI to a selected (remote) virtual host.
+Docker&nbsp;Machine was especially crucial for local use in early days of Docker, when no native support was available for Mac or Windows computers, but remains relevant for provisioning on remote systems.
+The package has not received any updates for two years, but is functional with a current version of `docker-machine` (`0.16.2`).
+It potentially lowers the barriers for R users to run containers on various hosts, if using the Docker&nbsp;Machine CLI directly is perceived as a barrier, or enables scripted workflows with remote processing.
 ```{r dockermachine, eval=FALSE, include=FALSE}

### 2020-01-23 manuscript.Rmd
update teaching paragraph

@@
 
-**Teaching** is a further example where sharing a prepared computing environment can greatly improve the process, especially for courses that require access to a relatively complex setup of software tools, e.g., as in the case of database systems.
-R is a fantastic tool when it comes to interfacing with databases: almost every open source and proprietary database system has an R package that allows users to connect and interact with the it 
-This flexibility is even broader now that we have tools like \CRANpkg{DBI}, that allows to create a common API for interfacing these databases, or like \CRANpkg{dbplyr}, which are designed to run \CRANpkg{dplyr} code straight against the database.
-But learning and teaching these tools comes with a cost: the cost of deploying or having access to an environment with the softwares and drivers installed.
-For people teaching R, it can become a barrier if they need to install local versions of the drivers, or to connect to remote instances which might or might not be made available by IT services.
-Giving access to a sandbox for the most common database environments is the idea behind [`r-db`](https://github.com/ColinFay/r-db) [@ColinFay], a Docker image that contains everything needed to connect, from R, to a database.
+Teaching is a further example where shared computing environments and sandboxing can greatly improve the process.
+First, **prepared environments for teaching** are especially helpful for courses that require access to a relatively complex setup of software tools, e.g., database systems.
+R is very useful tool for interfacing with databases, because almost every open source and proprietary database system has an R package that allows users to connect and interact with the it.
+This flexibility is even broadened by \CRANpkg{DBI}, which allows to create a common API for interfacing these databases, or the \CRANpkg{dbplyr} package, which runs \CRANpkg{dplyr} code straight against the database as queries.
+But learning and teaching these tools comes with the cost of deploying or having access to an environment with the software and drivers installed.
+For people teaching R, it can become a barrier if they need to install local versions of database drivers, or to connect to remote instances which might or might not be made available by IT services.
+Giving access to a sandbox for the most common database environments is the idea behind [`r-db`](https://github.com/ColinFay/r-db), a Docker image that contains everything needed to connect to a database from R.
 Notably, with `r-db`, the users don't have to install complex drivers or to configure their machine in a specific way.
-On top of that, `r-db` comes with a comprehensive [online guide](http://colinfay.me/r-db/), explaining how to spin a Docker instance of a specific database, and how to use `r-db` to connect to it.
-Each package contained inside this image also has a series examples, so that the user can get started with the database right away.
 The `rocker/tidyverse` base image ensures that users can also readily use packages for analysis, display, and reporting.
-
-Another way this idea of a common environment and partitioning containers in teaching is in the **[Dodona](https://dodona.ugent.be)** project.
-Dodona is a web platform developed at Ghent University and is used to teach students basic programming skills.
-Dodona uses Docker containers to test submissions by students.
+Second, the idea of a common environment and partitioning allow using **containers in teaching for secure execution and automated testing** of submissions by students.
+[Dodona](https://dodona.ugent.be) is a web platform developed at Ghent University and is used to teach students basic programming skills, and it uses Docker containers to test submissions by students.
 This means that both the code testing the students' submissions and the submission itself are executed in a predictable environment, avoiding compatibility issues between the wide variety of configurations used by students.
-The containerization is also used to shield the Dodona servers from bad (or even malicious) code: memory, time and I/O limits are used to make sure students can't overload the system.
+The containerization is also used to shield the Dodona servers from bad or even malicious code: memory, time and I/O&nbsp;limits are used to make sure students can't overload the system.
 The web application managing the containers communicates with them by sending configuration information as a JSON document over standard input. 
@@
 It provides a `rocker/drd` base image for easy evaluation of the platform\footnote{\href{https://github.com/att/rcloud/tree/master/docker}{https://github.com/att/rcloud/tree/master/docker}}.
-
-<!-- ## Commercial data science platforms
-\label{rstudio}
-[RStudio Connect](https://rstudio.com/products/connect/) is a commercial publishing platform to deploy dashboards, REST APIs, and other R and Python content into production.
-RStudio Connect uses an approach to isolate deployed R applications that is similar to Docker's filesystem and process namespacing and leverages the [RStudio Package Manager](https://rstudio.com/products/package-manager/) to manage R environments including system dependencies.
--->
 

### 2020-01-23 manuscript.Rmd
update author contributions, fix tex error, and update .tex file

@@
 A number of packages and projects enabled R to respond to client reqests over TCP/IP and local socket servers, such as \CRANpkg{Rserve}, \CRANpkg{svSocket}, [rApache](http://www.rapache.net) and more recently \CRANpkg{plumber} ([https://www.rplumber.io/](https://www.rplumber.io/)) and \pkg{RestRserve} ([http://restrserve.org](http://restrserve.org)), which even processes incoming requests in parallel with forked processes using \CRANpkg{Rserve}.
-The latter two also provide documentation for deployment with Docker or ready to use automated builds of images\footnote{See \href{https://www.rplumber.io/docs/hosting.html#docker}{https://www.rplumber.io/docs/hosting.html#docker}, \href{https://hub.docker.com/r/trestletech/plumber/}{https://hub.docker.com/r/trestletech/plumber/} and \href{https://hub.docker.com/r/rexyai/restrserve/}{https://hub.docker.com/r/rexyai/restrserve/}.}.
+The latter two also provide documentation for deployment with Docker or ready to use automated builds of images\footnote{See \href{https://www.rplumber.io/docs/hosting.html\#docker}{https://www.rplumber.io/docs/hosting.html\#docker}, \href{https://hub.docker.com/r/trestletech/plumber/}{https://hub.docker.com/r/trestletech/plumber/} and \href{https://hub.docker.com/r/rexyai/restrserve/}{https://hub.docker.com/r/rexyai/restrserve/}.}.
 These software allow other (remote) processes and programming languages to interact with R and to expose R-based function in a service architecture with HTTP APIs.
@@
 CF wrote paragraphs about \pkg{r-online}, \pkg{dockerfiler}, \pkg{r-ci} and \pkg{r-db}.
+SL contributed content on RStudio's usage of Docker.
 BM wrote the section on research compendia and made the project Binder-ready.
@@
 CW wrote the sections on Whole Tale and contributed the publication reproducibility audit use case.
-NX wrote the section on \pkg{liftr}.
+NX contributed content on \pkg{liftr}.
 All authors approved the final version.

### 2020-01-23 manuscript.tex
update author contributions, fix tex error, and update .tex file

@@
 with R}
-\author{by Daniel Nüst, Dom Bennett, Robrecht Cannoodt, Dav Clark, Gergely Daroczi, Dirk Eddelbuettel, Mark Edmondson, Colin Fay, Ellis Hughes, Sean Lopp, Ben Marwick, Heather Nolis, Jacqueline Nolis, Hong Ooi, Karthik Ram, Noam Ross, Lori Shepherd, Nitesh Turaga, Craig Willis, Nan Xiao}
+\author{by Daniel Nüst, Dom Bennett, Robrecht Cannoodt, Dav Clark, Gergely Daroczi, Dirk Eddelbuettel, Mark Edmondson, Colin Fay, Ellis Hughes, Sean Lopp, Ben Marwick, Heather Nolis, Jacqueline Nolis, Hong Ooi, Karthik Ram, Noam Ross, Lori Shepherd, Nitesh Turaga, Craig Willis, Nan Xiao, Charlotte Van Petegem}
 
@@
 containerisation in general. We identified common themes across this
-diversity: reproducible environments, scalability and effectivity, and
+diversity: reproducible environments, scalability and efficiency, and
 portability across clouds.
@@
 starting with the R packages specificly for interacting with Docker,
-second-level packages using containers mediately or only for specific
+second-level packages using containers indirectly or only for specific
 features, up to complex use cases leveraging containers. We conclude
@@
 point out future directions of development.
-
-\begin{figure}
-\centering
-\includegraphics{/rockerverse/rockerverse.png}
-\caption{Rockerverse hex sticker}
-\end{figure}
 
@@
 system---only the container runtime is required. Images are normally
-build from plain text documents called \texttt{Dockerfile}. A
+built from plain text documents called \texttt{Dockerfile}. A
 \texttt{Dockerfile} has a specific set of instructions to create and
@@
 \citep{rocker_geospatial_2019}. These containers form \emph{image
-stacks}, building on top of each other for better maintainability
-(i.e.~smaller \texttt{Dockerfile}s), composability, and to reduce build
-time. Also of note is a series of ``versioned'' containers which match
-the R release they contain with the \emph{then-current} set of packages
-via the MRAN Snapshot views of CRAN \citep{microsoft_cran_2019}. The
+stacks}, building on top of each other for better maintainability (i.e.,
+smaller \texttt{Dockerfile}s), composability, and to reduce build time.
+Also of note is a series of ``versioned'' containers which match the R
+release they contain with the \emph{then-current} set of packages via
+the MRAN Snapshot views of CRAN \citep{microsoft_cran_2019}. The
 Rocker~Project's impact and importance was acknowledged by the Chan
@@
 
-\hypertarget{images-for-alternative-r-distributions}{%
+\hypertarget{images-for-alternative-r-engines}{%
 \subsection{Images for alternative R
-distributions}\label{images-for-alternative-r-distributions}}
+engines}\label{images-for-alternative-r-engines}}
 
@@
 cases and experimental projects in academia and industry exist to test
-new approaches or provide features different to what \emph{GNU-R}, or
-Base R, provides. GNU-R is the R distribution maintained by the
+new approaches, or provide features different to what the R
+implementation maintained by the
 \href{https://www.r-project.org/contributors.html}{R Core Team} and
-available via \href{https://cran.r-project.org/}{CRAN}. While the images
-presented in this section are far from being as vetted, stable, and
-widely used as any of the Rocker images, they demonstrate an important
-advantage of containerisation technology, namely the abiliy to
+available via \href{https://cran.r-project.org/}{CRAN} provides. While
+the images presented in this section are far from being as vetted,
+stable, and widely used as any of the Rocker images, they demonstrate an
+important advantage of containerisation technology, namely the abiliy to
 transparently build portable stacks of open source software and make
-them easily accessible to others. All these distributions of R are
-published under GPL licenses. Since all of the R distributions claim
-better performance as a core motivation, a comparision based on Docker
-images, potentially leveraging the
+them easily accessible to others. All these alternative implementations
+of R are published under GPL licenses. Since all implementations of R
+claim better performance as a core motivation, a comparision based on
+Docker images, potentially leveraging the
 \href{https://docs.docker.com/config/containers/resource_constraints/}{resource
@@
 notably these capabilities are the MRAN repository, which is enabled by
-default, and the (optional) integration with
+default, and some bundled packages such as \CRANpkg{checkpoint} and
+specific R packages maintained by
+Microsoft\footnote{See "Bundled Packages" on \href{https://mran.microsoft.com/rro/installed}{https://mran.microsoft.com/rro/installed}.}.
+Originally the optional integration with
 \href{https://software.intel.com/en-us/mkl}{Intel® Math Kernel Library}
 (MKL) for multi-threading in linear algebra operations
-\citep{microsoft_multithread}. MRO does not provide official Docker
-images, but a set of community-maintained \texttt{Dockerfile}s and
-Docker images are provided by the poject \texttt{mro-docker}
-(\url{https://github.com/nuest/mro-docker}), e.g.~the image
+\citep{microsoft_multithread} was a distinguishing feature before it was
+made available publicly, and can be used in base R, too
+\citep{eddelbuettel_thinking_2018}.
+
+MRO does not provide official Docker images, but a set of
+community-maintained \texttt{Dockerfile}s and Docker images are provided
+by the poject \texttt{mro-docker}
+(\url{https://github.com/nuest/mro-docker}), e.g., the image
 \href{https://hub.docker.com/repository/docker/nuest/mro/}{\texttt{nuest/mro}}.
@@
 benefits of R, such scripting and extension packages, with the JVM's
-advantages in the areas of security, cross-platform availability, and
-established position in enterprise settings. R extension packages need
-to be specially compiled and are distributed via the Java package
-manager \href{https://en.wikipedia.org/wiki/Apache_Maven}{Apache Maven},
-cf. \url{http://packages.renjin.org/packages} for available packages.
+advantages in the areas of security, cross-platform availability and
+integration into enterprise platforms. R extension packages need to be
+specially compiled and are distributed via the Java package manager
+\href{https://en.wikipedia.org/wiki/Apache_Maven}{Apache Maven},
+cf.~\url{http://packages.renjin.org/packages} for available packages.
 Packages are loaded on demand, i.e., at the first call to
@@
 \texttt{Dockerfile}s are provided by
-\href{https://github.com/nuest/fastr-docker}{nuest/fastr-docker}, e.g.
+\href{https://github.com/nuest/fastr-docker}{nuest/fastr-docker}, e.g.,
 \href{https://cloud.docker.com/repository/docker/nuest/fastr/}{\texttt{nuest/fastr}
@@
 open development project for the analysis and comprehension of genomic
-data \citep{gentleman_bioconductor_2004}. The project consists of 1741 R
-software packages as of August 15th 2019, as well as packages containing
-annotation or experiment data. \emph{Bioconductor} has a semiannual
-release cycle, each release is associated with a particular version of
-R. Docker images allow availability of current and past versions of
-\emph{Bioconductor} for convenience and reproducibility.
-\emph{Bioconductor}'s `base' docker images are built on top of
-\texttt{rocker/r-ver} and \texttt{rocker/rstudio} images.
-\emph{Bioconductor} installs packages based on the R version, and
-therefore uses \texttt{rocker/rstudio} and \texttt{rocker/r-ver} version
-tagging. \emph{Bioconductor} selects the desired version of R from
-Rocker, adds the BiocManager CRAN package for installing appropriate
-versions of \emph{Bioconductor} packages, and creates a
-\emph{Bioconductor} docker image with an informative tag
-(R\_version\_Bioc\_version). The images are summarized on the
-\emph{Bioconductor} web site (see
+data \citep{gentleman_bioconductor_2004}. The project consists of 1823 R
+software packages as of October 30th 2019, as well as packages
+containing annotation or experiment data. \emph{Bioconductor} has a
+semi-annual release cycle, each release is associated with a particular
+version of R, and Docker images are provided for current and past
+versions of \emph{Bioconductor} for convenience and reproducibility. All
+images, included are described on the \emph{Bioconductor} web site (see
 \url{https://bioconductor.org/help/docker/}), created with
-\texttt{Dockerfile}s maintained on GitHub
-(\url{https://github.com/Bioconductor/bioconductor_docker}), and
-\href{https://hub.docker.com/u/bioconductor}{distributed through
-DockerHub}. Past and current combinations of R and \emph{Bioconductor}
-are therefore accessible via a specific image tag. A strength of this
+\texttt{Dockerfile}s maintained on GitHub, and distributed through
+Docker~Hub\footnote{See \href{https://github.com/Bioconductor/bioconductor_docker}{https://github.com/Bioconductor/bioconductor\_docker} and \href{https://hub.docker.com/u/bioconductor}{https://hub.docker.com/u/bioconductor} respectively.}.
+\emph{Bioconductor}'s `base' Docker images are built on top of the
+\texttt{rocker/rstudio} image. \emph{Bioconductor} installs packages
+based on the R version in combination with the Bioconductor version, and
+therefore uses Bioconductor version tagging \texttt{devel} and
+\texttt{RELEASE\_X\_Y}, e.g., \texttt{RELEASE\_3\_10}. Past and current
+combinations of R and \emph{Bioconductor} will therefore be accessible
+via a specific image tags. The \emph{Bioconductor} \texttt{Dockerfile}
+selects the desired version of R from Rocker, adds required system
+dependencies, and uses the \CRANpkg{BiocManager} package for installing
+appropriate versions of \emph{Bioconductor} packages. A strength of this
 approach is that the responsibility for complex software configuration
 (including customized development) is shifted from the user to the
-experienced \emph{Bioconductor} core team. Images may be valuable in
-teaching, where participants pull pre-built images to avoid complicated
-configuration of their own computing environemnts.
-
-\emph{Bioconductor} has several images in addition to `base', specific
-to various areas of research. The `core' image installs the most
-commonly used \emph{Bioconductor} packages. \emph{Bioconductor} images
-for proteomics, metabolomics, and flow cytometry are community
-maintained. All community maintained images build on top of the
-\emph{Bioconductor} base image and therefore indirectly the Rocker
-images. To simplify building and maintaining \emph{Bioconductor} images,
-we use a Ruby templating engine. A recent audit of the
-\emph{Bioconductor} \texttt{Dockerfile}s, following best practices from
-the Docker website, led to a reduction in the size and number of layers.
-The most important insights involve the
-\href{https://docs.docker.com/storage/storagedriver/overlayfs-driver/\#how-container-reads-and-writes-work-with-overlay-or-overlay2}{`union'
-file system used by Docker}. In this file system, once a layer (e.g.,
-\texttt{RUN} statement) writes to a file path, the file path is never
-altered. A subsequent layer that might appear to remove or overwrite the
-path actually masks, rather than alters, the original. It is therefore
-important to clean up (e.g., cache removal) within each layer, and to
-avoid re-installing existing dependencies.
-
-A recent innovation is to produce a \texttt{bioconductor\_full} image to
-emulate the \emph{Bioconductor} nightly Linux build machine. The image
-contains the \emph{system dependencies} needed to install and check
-almost all (1730 of 1741) \emph{Bioconductor} software packages. Users
-no longer have to manage complciated system dependencies. The image is
-configured so that \texttt{.libPaths()} has
+experienced \emph{Bioconductor} core team. A recent audit of the
+\emph{Bioconductor} image stack \texttt{Dockerfile} led to the
+deprecation of several community-maintained images, because the numerous
+specific images became too hard to understand, complex to maintain, and
+cumbersome to extent. As part of the simplification, a recent innovation
+is to produce a \texttt{bioconductor\_docker:devel} image to emulate the
+\emph{Bioconductor} nightly Linux build machine as closely as possible.
+This image contains the build system environment variables and the
+\emph{system dependencies} needed to install and check almost all (1813
+out of 1823) \emph{Bioconductor} software packages and saves users and
+package developers from managing these themselves. Furthermore the image
+is configured so that \texttt{.libPaths()} has
 \texttt{/usr/local/lib/R/host-site-library} as the first location. Users
-mounting a location on the host file system to this location then
-persist installed packages across docker sessions or updates. Many
-\emph{R} users pursue flexible work flows tailored to particular
-analysis needs, rather than standardized work flows. The
-\texttt{bioconductor\_full} image is well-suited to this pattern.
-\texttt{bioconductor\_full} provides developers with a test environment
-like \emph{Bioconductor}'s build system.
+mounting a location on the host file system to this location can persist
+installed packages across Docker sessions or image updates. Many R users
+pursue flexible work flows tailored to particular analysis needs, rather
+than standardized work flows. The new \texttt{bioconductor\_docker}
+image is well-suited for this pattern, while
+\texttt{bioconductor\_docker:devel} provides developers with a test
+environment close to \emph{Bioconductor}'s build system.
 
@@
 Windows 10. New R versions are only added on demand. The explorative
-rocker-win project demonstrates the variety and pontentially confusing
-complexity of Windows on Docker, e.g., because of the different base
-images, but also provides a starting point for any Windows-constrained
-user to leverage containerisation, including examples for apps using
-\textbackslash{}pkg\{plumber, \pkg{Shiny}, or R packages with system
-dependencies such as \pkg{sf}. The latter is quite comfortable and fast
-actually, compared to installation from source on Linux, as R ships
-pre-compiled binaries for Windows. However, compared to the various
-Linux images, the footprint of Windows images is quite large.
+rocker-win project demonstrates the variety of Windows on Docker, which
+is potentially confusing, e.g., because of the different base images.
+The proof of concept provides a starting point for any
+Windows-constrained user to leverage containerisation, including
+examples for apps using \CRANpkg{plumber}, \CRANpkg{Shiny}, or R
+packages with system dependencies such as \CRANpkg{sf}. The latter is
+quite comfortable and fast actually, compared to the default
+installation from source on Linux, because CRAN ships pre-compiled
+binaries for Windows. However, compared to the various Linux images, the
+footprint of Windows images is quite large.
 
@@
 Linux-based containers. Naturally, a Windows license is required for the
-host machine, and the licensing impacts the potnetial to build images in
+host machine, and the licensing impacts the potential to build images in
 clouds or redistribute exported images as archive files. But the
@@
 
-\hypertarget{non-debian-linux-images}{%
-\subsection{Non-Debian Linux images}\label{non-debian-linux-images}}
+\hypertarget{non-rocker-linux-images}{%
+\subsection{Non-Rocker Linux images}\label{non-rocker-linux-images}}
 
@@
 package developers to check their R package on different platforms and R
-versions using a web form or the package \pkg{rhub}
+versions using a web form or the package \CRANpkg{rhub}
 \citep{csardi_rhub_2019}. The builder uses Docker containers to conduct
@@
 For R though, this advantage quickly deteriorates if more features are
-needed, e.g.~if adding compilers, packages, or especially if installing
+needed, e.g., if adding compilers, packages, or especially if installing
 powerful data science and communication tools, like RStudio or LaTeX.
@@
 \emph{Data Science} is a widely discussed topic among all academic
-disciplines \citep[cf. e.g.,][]{donoho_50_2017}. Using computers and
-complex software to make sense of (large amounts of) collected or
-simulated data is a cross-cutting phenomenon leading to \emph{X Data
-Science} disciplines, seminars, and study majors. The practice to apply
-a combination tools and software stacks for analysing data and the
-cross-cutting skillsets connected with data science puts the management
-of computing environments and the challenges of reproducibility high up
-on the agenda. This awareness leads to a large number of bespoke Docker
-images but also some commonly used image stacks, ranging from base
-images with very common tools to `catch all' images including \emph{a
-lot} of software. Here we only mention a few image stacks as examples,
-which are not based on Rocker images or stem from the R community, but
-prominently include installations of R. A general description of what to
-consider when creating a Docker image used for data science projects in
-R is provided on
-\href{https://environments.rstudio.com/docker}{environments.rstudio.com}.
+disciplines \citep[e.g.,][]{donoho_50_2017}. The discussions shed a
+light on the tools and craftspersonship behind the analysis of data with
+computational methods. The practice of Data Science often involves a
+combination tools and software stacks and requires a cross-cutting
+skillset. This complexity and an inherent concern for openness and
+reproducibility in the Data Science community lead to Docker being used
+widely. This section presents exemplary Docker images and image stacks
+featuring R intended for Data Science.
 
@@
 \texttt{Dockerfile}} for usage in their Machine Learning competitions
-and easy access to the associated datasets. It includes machine learling
-libraries such as Tensorflow and Keras, and also configures the
-\pkg{reticulate} package. The image uses a base image with \emph{all
-packages from CRAN}, \texttt{gcr.io/kaggle-images/rcran}, which requires
-a Google Cloud Build as Docker~Hub would time
+and easy access to the associated datasets. It includes machine learning
+libraries such as Tensorflow and Keras (see also image
+\texttt{rocker/ml} in Section~\nameref{rocker-gpu}), and also configures
+the \CRANpkg{reticulate} package. The image uses a base image with
+\emph{all packages from CRAN}, \texttt{gcr.io/kaggle-images/rcran},
+which requires a Google Cloud Build as Docker~Hub would time
 out\footnote{Originally, a stacked collection of over 20 images with automated builds on Docker~Hub was used, see \href{https://web.archive.org/web/20190606043353/http://blog.kaggle.com/2016/02/05/how-to-get-started-with-data-science-in-containers/}{https://web.archive.org/web/20190606043353/http://blog.kaggle.com/2016/02/05/how-to-get-started-with-data-science-in-containers/} and \href{https://hub.docker.com/r/kaggle/rcran/dockerfile}{https://hub.docker.com/r/kaggle/rcran/dockerfile}}.
@@
 for their browser-based business analytics interface based on
-\pkg{Shiny} (\texttt{Dockerfile}
+\CRANpkg{Shiny} (\texttt{Dockerfile}
 \href{https://github.com/radiant-rstats/docker}{on GitHub}) and for use
@@
 be readily re-used in this image stack (see
-Section~\nameref{conclusion}). Both approaches enable \texttt{apt} as an
-installation method, and this is exposed via the GUI-based environment
+Section~\nameref{conclusions}). Both approaches enable \texttt{apt} as
+an installation method, and this is exposed via the GUI-based
+environment
 management,\footnote{See \href{https://docs.gigantum.com/docs/environment-management}{https://docs.gigantum.com/docs/environment-management}}.
@@
 
-\textbf{Whole Tale} (\url{https://wholetale.org}) is a National Science
-Foundation (NSF) funded project\footnote{NSF Award OAC-1541450}
-developing a scalable, open-source, multi-user platform for reproducible
-research \citep{brinckman2019, chard2019a}. A central goal of the
-platform is to enable researchers to easily create and publish
-executable research
-objects\footnote{In Whole Tale a `tale` is a research object that contains metadata, data (by copy or reference), code, narrative, documentation, provenance, and information about the computational environment to support computational reproducibility}
-associated with published research \citep{chard2019b}. For R users,
-Whole Tale attempts to combine the strengths of the Rocker Project's
-curated Docker images with the Jupyter Project's
+\textbf{Whole~Tale} (\url{https://wholetale.org}) is a National Science
+Foundation (NSF) funded project developing a scalable, open-source,
+multi-user platform for reproducible research
+\citep{brinckman2019, chard2019a}. A central goal of the platform is to
+enable researchers to easily create and publish executable research
+objects\textbackslash{}footnote\{In Whole\textasciitilde{}Tale a
+\emph{tale} is a research object that contains metadata, data (by copy
+or reference), code, narrative, documentation, provenance, and
+information about the computational environment to support computational
+reproducibility\} associated with published research \citep{chard2019b}.
+Using Whole~Tale, researchers can create and publish Rocker-based
+reproducible research objects to a growing number of repositories
+including DataONE member nodes, Zenodo and soon Dataverse. Additionally,
+Whole~Tale supports automatic data citation and is working on
+capabilities for image preservation and provenance capture to improve
+the transparency of published computational research artifacts
+\citep{mecum2018, mcphillips2019}. For R users, Whole~Tale attempts to
+combine the strengths of the Rocker Project's curated Docker images with
+the Jupyter Project's
 \texttt{repo2docker}\footnote{See \href{https://repo2docker.readthedocs.io/}{https://repo2docker.readthedocs.io/}}
 component to simplify the customization of R-based environments for
-researchers with limited experience with either Docker or Git. Using
-Whole Tale, researchers can create and publish Rocker-based reproducible
-research objects to a growing number of repositories including DataONE
-member nodes, Zenodo and soon Dataverse. Additionally, Whole Tale
-supports automatic data citation and is working on capabilities for
-image preservation and provenance capture to improve the transparency of
-publish computational research artifacts
-\citep{mecum2018, mcphillips2019}.
+researchers with limited experience with either Docker or Git. Multiple
+options have been discussed to allow users to change the base image used
+in \texttt{repo2docker} from the default Ubuntu LTS (long-term support)
+required to support the Rocker Project images. Whole~Tale implemented a
+custom
+\texttt{RockerBuildPack}\footnote{See \href{https://github.com/whole-tale/repo2docker\_wholetale}{https://github.com/whole-tale/repo2docker\_wholetale}}
+to support customization of the \texttt{rocker/geospatial} image through
+\texttt{repo2docker}
+composability\footnote{Composability refers to the ability to combine multiple package managers -- such as R, `pip`, and `conda`}.
+This works in part because Rocker images are based on a Debian
+distribution, so the instructions created by \texttt{repo2docker} for
+Ubuntu work because of compatible shell and package manager.
 
@@
 installing the Docker software, creating \texttt{Dockerfile}s
-(\pkg{dockerfiler}), \pkg{containerit}), building images and launching a
-containers (\pkg{stevedore}, \pkg{docker}) on a local machine or on the
-cloud. As such, the choice of which package is most useful depends on
-the use-case at hand, but also the users level of expertise.
+(\CRANpkg{dockerfiler}, \pkg{containerit}), building images and
+launching a containers (\CRANpkg{stevedore}, \pkg{docker}) on a local
+machine or on the cloud. As such, the choice of which package is most
+useful depends on the use-case at hand, but also the users level of
+expertise.
 
@@
 \hline
-Interact with an instance (e.g. file transfer) &  & \checkmark & \checkmark &  &  & \checkmark\\
+Interact with an instance (e.g., file transfer) &  & \checkmark & \checkmark &  &  & \checkmark\\
 \hline
@@
 first R package for managing Docker images and containers. It uses the
-\pkg{sys} package to run system commands against the Docker CLI, both
-locally and through an ssh connection, and has convenience functions,
-e.g., for listing and removing containers/images and for accessing logs.
-The output of container executions are converted to appropriate R types.
-The Docker CLI, while evolving quickly and not with great concern for
-introducing breaking changes, basic functionality is unchanged for a
-long time so \texttt{harbor::docker\_run(image\ =\ "hello-world")} works
-like a charm.
-
-\textbf{\pkg{stevedore}}
+\CRANpkg{sys} package to run system commands against the Docker CLI,
+both locally and through an SSH connection, and has convenience
+functions, e.g., for listing and removing containers/images and for
+accessing logs. The output of container executions are converted to
+appropriate R types. The Docker CLI's basic functionality, while
+evolving quickly and with small concern for avoiding breaking changes,
+is unchanged for a long time so a core function such as
+\texttt{harbor::docker\_run(image\ =\ "hello-world")} still works
+depsite the stopped development.
+
+\textbf{\CRANpkg{stevedore}}
 (\url{https://cran.r-project.org/package=stevedore}) is currently the
 most powerful Docker client in R. It interfaces with the Docker daemon
-over the Docker HTTP API\footnote{} via a Unix socket on Linux or MacOS,
-over a named pipe on Windows, or over an HTTP/TCP connection. The
-package is the only one not using system calls to the \texttt{docker}
-CLI for managing images and containers and easily exposes connections to
-remote Docker daemons, which has to be configured on the Docker level
-otherwise. Using the API gives access to more information and is system
-independent and likely more reliable than parsing command line output.
-\pkg{stevedore}'s own interface is automatically generated based on the
-OpenAPI specification of the Docker daemon, but still similar to the
-Docker CLI. The interface is similar to R6 objects, in that a
+over the Docker HTTP
+API\footnote{\href{https://docs.docker.com/engine/api/latest/}{https://docs.docker.com/engine/api/latest/}}
+via a Unix socket on Linux or MacOS, over a named pipe on Windows, or
+over an HTTP/TCP connection. The package is the only one not using
+system calls to the \texttt{docker} CLI for managing images and
+containers and easily exposes connections to remote Docker daemons,
+which has to be configured on the Docker level otherwise. Using the API
+gives access to more information and is system independent and likely
+more reliable than parsing command line output. \CRANpkg{stevedore}'s
+own interface is automatically generated based on the OpenAPI
+specification of the Docker daemon, but still similar to the Docker CLI.
+The interface is similar to R6 objects, in that a
 \texttt{stevedore\_object} has a number of functions attached to it that
@@
 
-\textbf{\pkg{AzureContainers}}
-(\url{https://cran.r-project.org/package=AzureContainers}) is an
-interface to a number of container-related services in Microsoft's
+\textbf{\CRANpkg{AzureContainers}} is an interface to a number of
+container-related services in Microsoft's
 \href{https://azure.microsoft.com/}{Azure Cloud}, see
@@
 
-\textbf{\pkg{babelwhale}}
-(\url{https://cran.r-project.org/package=babelwhale}) allows executing
-and interacting with containers, which can use either Docker or
-Singularity as a backend. The package provides a unified interface to
-interact with Docker and Singularity containers. Users can, for example,
-execute a command inside a container, mount a volume or copy a file.
+\textbf{\CRANpkg{babelwhale}} allows executing and interacting with
+containers, which can use either Docker or Singularity as a backend
+\citep{cannoodt_babelwhale_2019}. The package provides a unified
+interface to interact with Docker and Singularity containers. Users can,
+for example, execute a command inside a container, mount a volume or
+copy a file.
 
@@
 (\url{https://github.com/thebioengineer/dockyard}) has the goal of
-lowering barrier to creating \texttt{Dockerfile}s and Docker images, and
-deploying Docker containers. The way docker images are made is by
-reading and following a set of instructions found in
-\texttt{Dockerfile}s. This requires stepping out of the R ecosystem and
-into an infrastructure mindset where the user needs to specify the
-environment they require. Because \texttt{Dockerfile}s can be read as a
-series of steps, this lends itself well to the \texttt{tidyverse} style
-of programming where each step is piped into the next. dockyard has a
-\texttt{Dockerfile} object that either starts out as an empty
-\texttt{Dockerfile}, or can use existing \texttt{Dockerfile} to use as a
-template. This combined with a series of functions that are analogous to
-the steps commonly found in a \texttt{Dockerfile}, it makes building
-Docker images much more approachable to R users. There are also wrappers
-for commonly used steps for R programmers like adding libraries and
-copying files into the R Docker containers. Once the \texttt{Dockerfile}
-is created, dockyard has built-in functionality to build the image and
-run a container.
+lowering barrier to creating \texttt{Dockerfile}s, building Docker
+images, and deploying Docker containers. The package follows the
+increasingly used piping paradigm of the \texttt{tidyverse} style of
+programming for chaining R functions representing the instructions in a
+\texttt{Dockerfile}. An existing \texttt{Dockerfile} can be used as a
+template. \pkg{dockyard} also includes wrappers for common steps, such
+as installing an R package or copying files, and build-in functions for
+building an image running a container, to make using Docker even more
+approachable to R users with a single API.
 
@@
 provide a convenient interface to
-\href{https://docs.docker.com/machine/overview/}{Docker Machine} from R.
-Docker Machine's CLI tool \texttt{docker-machine} allows users to create
-and manage virtual host on local computers, local data centers, or at
-cloud providers. A local Docker installation can be configured to
-forward all commands issues on the local Docker CLI to a selected
-(remote) virtual host fully transparent. For local Docker Machine was
-especially crucial in early days of Docker, when no native Docker
-support was available for Mac or Windows computers, but remains relevant
-for provisioning on remote systems. The package has not received any
-updates for two years, but was functional with a current version of
-\texttt{docker-machine} (\texttt{0.16.2}) and potentially lowers the
-barriers for R users to run containers on various hosts, if using the
-Docker Machine CLI is perceived as a barrier.
-
-\hypertarget{capture-and-create-environments-nuest}{%
-\subsection{\texorpdfstring{Capture and create environments
-\citep{nuest}}{Capture and create environments {[}@nuest{]}}}\label{capture-and-create-environments-nuest}}
+\href{https://docs.docker.com/machine/overview/}{Docker~Machine} from R.
+The CLI tool \texttt{docker-machine} allows users to create and manage
+virtual host on local computers, local data centers, or at cloud
+providers. A local Docker installation can be configured to transparntly
+forward all commands issued on the local Docker CLI to a selected
+(remote) virtual host. Docker~Machine was especially crucial for local
+use in early days of Docker, when no native support was available for
+Mac or Windows computers, but remains relevant for provisioning on
+remote systems. The package has not received any updates for two years,
+but is functional with a current version of \texttt{docker-machine}
+(\texttt{0.16.2}). It potentially lowers the barriers for R users to run
+containers on various hosts, if using the Docker~Machine CLI directly is
+perceived as a barrier, or enables scripted workflows with remote
+processing.
+
+\hypertarget{capture-and-create-environments}{%
+\subsection{Capture and create
+environments}\label{capture-and-create-environments}}
 
@@
 Docker images and using containers for specific tasks, such as running
-tests or rendering reproducible reports, easier.
+tests or rendering reproducible reports, easier. While authoring and
+managing an environment with Docker by hand is possible and feasible for
+experts\footnote{See, e.g., this tutorial by RStudio on how to manage environments and package versions and to ensure deterministic image builds with Docker: \href{https://environments.rstudio.com/docker}{https://environments.rstudio.com/docker}.},
+the following examples show the power of automation when environments
+become too cumbersome or acquiring the skills is not possible in due
+course. Especially \emph{version pinning}, with packages \pkg{remotes}
+and \pkg{versions} for R or by using MRAN, and with system package
+managers for different operating systems, can greatly increase the
+reproducibility of built images and are common approaches.
 
 \textbf{\pkg{dockerfiler}}
-(\url{https://github.com/ColinFay/dockerfiler/}) \ldots{}
-
+(\url{https://github.com/ColinFay/dockerfiler/}) is an R package
+designed for building \texttt{Dockerfile}s straight from R. A scripted
+creation of a \texttt{Dockerfile} enables iteration and automation, for
+example for packaging applications for deployment (see
+\nameref{deployment}). Being scriptable from R developers can leverage
+the tools available in R to parse a \texttt{DESCRIPTION} file, to get
+system requirements, to list dependencies, versions, etc.
 \textbf{\pkg{containerit}}
-(\url{https://github.com/o2r-project/containerit/}) \ldots{}
-
-\textbf{\pkg{dockr}} (\url{https://github.com/smaakage85/dockr})
-\ldots{} see/cite
-\url{http://smaakage85.netlify.com/2019/12/21/dockr-easy-containerization-for-r/}
-
-\textbf{\pkg{renv}} (\url{https://rstudio.github.io/renv/}) with Docker:
-\url{https://rstudio.github.io/renv/articles/docker.html}
-
-\emph{version pinning} with packages \pkg{remotes} and \pkg{versions}
-for R, and with system package managers for distros
-
-\textbf{\pkg{liftr}} (\url{https://nanx.me/liftr/}) \citep{liftr2019}
-aims to solve the problem of persistent reproducible reporting in
-statistical computing. Currently, the R Markdown format and its backend
-compilation engine \emph{knitr} offer a \emph{de facto} standard for
-creating dynamic documents \citep{xie2018}. However, the reproducibility
-of such content authoring environments is often limited to individual
-machines --- it is not easy to replicate the system environment
-(libraries, R versions, R packages) where the document was compiled.
-This issue becomes even more serious when it comes to collaborative
-document authoring and creating large-scale document building services.
-\emph{liftr} solves this reproducibility problem by bringing Docker to
-the game. In essence, \emph{liftr} helps R Markdown users create and
-manage Docker containers for rendering the documents, thus make the
-computations utterly reproducible across machines and systems.
-
-On implementation, with no side effects, \emph{liftr} extended and
-introduced new metadata fields to R Markdown, allowing users to declare
-the dependencies for rendering the document. \emph{liftr} parses such
-fields and generates a \texttt{Dockerfile} for creating Docker
-containers. \emph{liftr} then helps render the document inside the
-created Docker container. This workflow is summarized in Figure
-\ref{figure:liftr}.
-
-\begin{figure}[htbp]
-  \centering
-  \includegraphics[width=\textwidth]{liftr-workflow}
-  \caption{The liftr workflow for rendering containerized R Markdown documents.}
-  \label{figure:liftr}
-\end{figure}
-
-\emph{knitr} and R Markdown are used as the template engine to generate
-the \texttt{Dockerfile}. Features such as caching container layers for
-saving image build time, automatic housekeeping for fault-tolerant
-builds, and Docker status check are supported by \emph{liftr}. Four
-RStudio addins are also offered by \emph{liftr} to allow push-button
-compilation of documents and provide better IDE integrations.
-
-Three basic principles are followed to design the \emph{liftr} package
-since its inception.
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
+(\url{https://github.com/o2r-project/containerit/}) attempts to take
+this one step further and includes these tools to automatically create a
+\texttt{Dockerfile} that can execute a given workflow
+\citep{nust_containerit_2019}. \pkg{containerit} accepts R
+\texttt{sessionInfo} objects as input and provides helper functions to
+derive these from workflows, e.g., an R script or R Markdown document,
+by analysing the session state at the end of the workflow. It relies on
+the \pkg{sysreqs} (\url{https://github.com/r-hub/sysreqs/}) package and
+it's mapping of package system dependencies to platform specific
+installation package
+names\footnote{See \href{https://sysreqs.r-hub.io/}{https://sysreqs.r-hub.io/}.}.
+\pkg{containerit} uses \CRANpkg{stevedore} to streamline the user
+interaction and improve the created \texttt{Dockerfile}s, e.g., by
+running a container for the desired base image to extract the already
+available R packages. \textbf{\pkg{dockr}}
+(\url{https://github.com/smaakage85/dockr}) is a very similar package
+attempting to mirror a given R session, including local non-CRAN
+packages\citep{kjeldgaard_dockr_2019}. Users can manually add statements
+for non-R dependencies to the \texttt{Dockerfile}.
+\textbf{\CRANpkg{liftr}} aims to solve the problem of persistent
+reproducible reporting in statistical computing based on the R Markdown
+format \citep{xie2018} for dynamic documents
+\citep[\href{https://nanx.me/liftr/}{https://nanx.me/liftr/}, ][]{liftr2019}.
+The irreproducibility of authoring environments can become an issue for
+collaborative documents and large-scale platforms for processing
+documents. \CRANpkg{liftr} makes the document the main and sole workflow
+control file and the only file that needs to be shared between
+collaborators for consistent environments, e.g.~demonstrated in the
+DockFlow project (\url{https://dockflow.org}). It introduces new fields
+to the R Markdown document header, allowing users to manually declare
+the dependencies, including versions, for rendering the document. The
+package then generates a \texttt{Dockerfile} from this metadata and
+provides a utitility function to render the document inside a Docker
+container, i.e., \texttt{render\_docker("foo.Rmd")}. An RStudio addin
+even allows compilation of documents with a single push of a button.
+
+The package \textbf{\CRANpkg{renv}}
+(\url{https://rstudio.github.io/renv/}) helps users to manage the state
+of the R library in a reproducible way, further providing isolation and
+portability \citep{renv2019}. The package does not cover system
+dependencies, though, the \CRANpkg{renv}-based environment can be
+transferred into a container either by restoring the environment based
+on the main configuration file \texttt{renv.lock} or by storing the
+\CRANpkg{renv}-cache on the host and not in the container
+\citep{ushey_using_2019}.
+
+\hypertarget{using-r-to-power-enterprise-software-in-production-environments}{%
+\subsection{Using R to power enterprise software in production
+environments}\label{using-r-to-power-enterprise-software-in-production-environments}}
+
+\label{enterprise}
+
+R has been historically viewed as a tool for analysis and scientific
+research, but not for creating software that corporations can rely on
+for production services. However, thanks to advancements in R running as
+a web service, along with with the ability to deploy R in Docker
+containers, modern enterprises are now capable of having real-time
+machine learning powered by R. A number of packages and projects enabled
+R to respond to client reqests over TCP/IP and local socket servers,
+such as \CRANpkg{Rserve}, \CRANpkg{svSocket},
+\href{http://www.rapache.net}{rApache} and more recently
+\CRANpkg{plumber} (\url{https://www.rplumber.io/}) and \pkg{RestRserve}
+(\url{http://restrserve.org}), which even processes incoming requests in
+parallel with forked processes using \CRANpkg{Rserve}. The latter two
+also provide documentation for deployment with Docker or ready to use
+automated builds of
+images\footnote{See \href{https://www.rplumber.io/docs/hosting.html\#docker}{https://www.rplumber.io/docs/hosting.html\#docker}, \href{https://hub.docker.com/r/trestletech/plumber/}{https://hub.docker.com/r/trestletech/plumber/} and \href{https://hub.docker.com/r/rexyai/restrserve/}{https://hub.docker.com/r/rexyai/restrserve/}.}.
+These software allow other (remote) processes and programming languages
+to interact with R and to expose R-based function in a service
+architecture with HTTP APIs. APIs based on these package can be deployed
+with scalability and high availability using containers. This pattern of
+deploying code matches those used by software engineering services
+created in more established languages in the enterprise domain, such as
+Java or Python, and R can be used alongside those langages as a first
+class member of a software engineering technical stack.
+
+CARD.com implemented a web application for the optimization of the
+acquisition flow and the real-time analysis of debit card transactions.
+The software used \CRANpkg{Rserve} and rApache and was deployed in
+Docker containers. The R session behind \CRANpkg{Rserve} acted as a
+read-only in-memory database, which was extremely fast and scalabale,
+for the many concurrent rApache processes responding to the live-scoring
+requests of various divisions of the company. Similarly dockerized R
+scripts were responsible for the ETL processes and even the
+client-facing email, text message and push notification alerts sent in
+real-time based on card transactions. The related Docker images were
+made available at \url{https://github.com/cardcorp/card-rocker}. The
+images extend \texttt{rocker/r-base} and additionally entailed an SSH
+client and a workaround for being able to mount SSH keys from the host,
+Pandoc, the Amazon Web Services (AWS) SDK, and Java, which is required
+by the AWS SDK. The AWS SDK allowed to run R consumers reading from
+real-time data processing streams of
+\href{https://aws.amazon.com/kinesis/}{AWS
+Kinesis}\textbackslash{}footnote\{See useR!2017 talk \emph{Stream
+processing with R in AWS} at
+\textbackslash{}href\{\url{https://static.sched.com/hosted_files/user2017/2f/AWR\%20Kinesis\%20at\%20useR\%202017.pdf}\}\{\url{https://static.sched.com/hosted_files/user2017/2f/AWR}
+Kinesis at use 02017.pdf\}.\}. The applications were deployed on Amazon
+Elastic Container Service (\href{https://aws.amazon.com/ecs/}{ECS}). The
+main learnings from using R in Docker was the imporance of not only
+pinning the R package versions via MRAN, but also moving away from
+Debian testing to a distribution with long-term support. For the use
+case at hand, this switch served the priority to have more control over
+upstream updates, and to minimize the risk of breaking the automated
+builds of the Docker images and production jobs.
+
+The AI @ T-Mobile team created a set of neural network machine learning
+natural language processing models to help customer care agents manage
+text-based messages for customers \citep{t-mobile_enterprise_2018}. For
+example, one model quickly identifies if a message is from a customer or
+not
+\citep[see \CRANpkg{Shiny}-based \href{https://secure.message.t-mobile.com/v1/shiny/is-customer/app/}{demo}, ][]{nolis_small_2019},
+others tell which customers are likely to make a repeat purchase. If a
+data scientist creates a machine learning model and exposes it through a
+\CRANpkg{plumber} API, then someone else on the marketing team could
+write software that sends different emails depending on that real-time
+prediction. The models are convolutional neural networks that use the
+\CRANpkg{keras} package and run in a Rocker Docker image. The
+corresponding \texttt{Dockerfile}s are published
+\href{https://github.com/tmobile/r-tensorflow-api}{on GitHub}. Since the
+models power tools for agents and customers, they need to have extremely
+high uptime and reliability. The AI @ T-Mobile team found that the
+models performed well and today these models power real-time services
+that are called over a million times a day.
+
+\hypertarget{continuous-integration-continuous-delivery-and-deployment}{%
+\subsection{Continuous integration, continuous delivery, and
+deployment}\label{continuous-integration-continuous-delivery-and-deployment}}
+
+\label{cicd} \label{deployment}
+
+The cloud is the natural environment of containers, and subsequently
+containers are the go-to mechanism to deploy R server applications. More
+and more continuous integration (CI) and continuous delivery (CD)
+services also use containers, opening up new options for use. The
+controlled nature of containers, i.e., the possibility to define the
+software environment very well, even on remote machines, is crucial
+here, for example to match test or build environments with production
+environments.
+
+As a first example, \CRANpkg{Shiny} is the most popular package for
+creating interactive online dashboards with R and it enables users with
+very diverse backgrounds to create stable and user friendly web
+applications. The ShinyProxy (\url{https://www.shinyproxy.io/}) is an
+open source tool to deploy Shiny apps in an enterprise context. They
+feature single sign-on, but also in scientific use cases
+\citep[e.g., ][]{savini_epiexplorer_2019,glouzon_structurexplor_2017}.
+ShinyProxy uses Docker containers to isolate user sessions and to
+achieve scalability for multi-user scenarious with multiple apps.
+ShinyProxy itself is written in Java to accomodate corporate
+requirements and may itself run in a container for stability and
+availability. The tool is build on ContainerProxy
+(\url{https://www.containerproxy.io/}), which provides similar features
+for executign long-running R jobs or interactive R sessions. The started
+containers can run on a regular Docker host but also in clusters.
+
+Second, containers can be used to create \textbf{platform installation
+packages} in a DevOps setting. The
+\href{https://www.opencpu.org/}{OpenCPU} system provides an HTTP API for
+data analysis based on R. \citet{ooms_opencpu_2017} describes how
+various platform-specific installation files for OpenCPU are created
+using Docker~Hub: the automated builds install the software stack from
+source on different operating systems; afterwards a script file
+downloads the images and extracts the OpenCPU binaries.
+
+\begin{itemize}
+\tightlist
 \item
-  Continuous reproducibility. Continuous integration, continuous
-  delivery, and continuous deployment are well-accepted practices in
-  software engineering. Similarly, it is believed by the authors of
-  \emph{liftr} that ensuring computational reproducibility means a
-  continuous process instead of creating static data/code archives or a
-  one-time deal. Specifically, the software packages used in data
-  analysis should be upgraded regularly in a manageable way. Therefore,
-  \emph{liftr} supports specifying particular versions of package
-  dependencies, while users are encouraged to always use the latest
-  version of packages (without a version number) by default.
+  \href{https://github.com/dynverse/dynwrap_containers/blob/master/.travis.yml}{dynwrap}
+  \citep{rcannood}
+
+  \begin{itemize}
+  \tightlist
   \item
-  Document first. Many data analysis workflows could be wrapped as
-  either R packages or dynamic documents. In \emph{liftr}, the endpoint
-  of dynamic report creation is the focus of containerization, because
-  this offers more possibilities for organizing both computations and
-  documentation. Users are encouraged to start thinking from the visible
-  research output from the first day.
-\item
-  Minimal footprint. R Markdown and Docker are already complex software
-  systems. Making them work together seamlessly can be complicated.
-  Therefore, API designs such as function arguments are simplified while
-  being kept as expressive and flexible as possible.
-\end{enumerate}
-
-In summary, \emph{liftr} tries to redefine the meaning of computational
-reproducibility by offering system-level reproducibility for data
-analysis. It provided a practical way for achieving it --- a new
-perspective on how reproducible research could be done in reality.
-Further, sharing system environments for data analysis also becomes
-extremely easy, since users only need to share the R Markdown document
-(with a few extra metadata fields), and compile them with \emph{liftr}.
-As an example, \emph{liftr} demonstrated its advantage for R
-Markdown-based computational workflow orchestration, by effortlessly
-containerizing 18 complex \emph{Bioconductor} workflows in the DockFlow
-project (\url{https://dockflow.org}) in 2017.
-
-One of the key tool of Docker are \texttt{Dockerfile}s, which can be
-thought of as ``recipes'' used to build a specific image. But building
-these files can feel like a workflow break as it demands to open a
-separate file, and / or to write a small wrapper around R's
-\texttt{write()} function to add elements to the file. At the same time,
-doing it by hand prevents from using programming language for what they
-are good for: iteration, and automation. Iteration and automation for
-\texttt{Dockerfile}s is the very reason behind
-\href{https://github.com/ColinFay/dockerfiler}{\texttt{dockerfiler}}, an
-R package designed for building \texttt{Dockerfile}s straight from R.
+    For this project, we use travis-ci to build rocker-derived
+    containers, test them, and only push them to Docker~Hub (from
+    travis-ci.org) if the integration tests succeed.
+  \end{itemize}
+\end{itemize}
 
@@
 creating the \texttt{Dockerfile} for building production-grade Shiny
-applications and deploying them. The reason behind this is that
-\texttt{dockerfiler}, on top of being scriptable from R, can leverage
-all the tools available in R to parse a \texttt{DESCRIPTION} file, to
-get system requirements, to list dependencies, versions, etc.
-
-\hypertarget{using-r-to-power-enterprise-software-in-production-environments}{%
-\subsection{Using R to power enterprise software in production
-environments}\label{using-r-to-power-enterprise-software-in-production-environments}}
-
-\label{enterprise}
-
-R has been historically viewed as a tool for analysis and scientific
-research--not for creating software that corporations can rely on to
-continously run in real time. However, thanks to advancements in R
-running as a web service, along with with the ability to deploy R in
-Docker containers, modern enterprises are now capable of having
-real-time machine learning powered by R. One of the first
-implementations of having R respond to client reqests were TCP/IP and
-local socket servers, such as \CRANpkg{Rserve} and \CRANpkg{svSocket},
-which allowed other local or remote processes to make binary requests to
-interact with R without the need to link to R at compile time and
-connected other programming languages with R. R could also empower web
-services since at least 2007, when
-\href{http://www.rapache.net}{rApache} was presented at the useR!
-conference \citep{horner_rapache_2013}. rApache runs an R interpreter
-inside the Apache web server, which makes it possible and very efficient
-to build scalable R-based APIs and web services.
-
-Such rApache applications have been used at
-\href{http://rapporter.net}{rapporter.net}, transforming parametrized R
-Markdown-like documents into interactive web applications for reporting.
-A similar design was implemented at
-\href{https://www.card.com/}{CARD.com}, where the optimization of the
-acquisition flow and the real-time analysis of debit card transactions
-both relied on an efficient combination of Rserve and rApache running in
-Docker containers: the R session behind Rserve acted as a read-only,
-exteremely fast and scalabale, in-memory database for the many
-concurrent rApache processes responding to the live-scoring requests of
-various divisions of the company. Similar dockerized R scripts were
-responsible for the ETL processes and even the client-facing email, text
-message and push notification alerts sent in real-time based on card
-transactions. The related Docker images were made available at
-\url{https://github.com/cardcorp/card-rocker}. The images extend
-\texttt{rocker/r-base}, e.g., with an SSH client and a workaround for
-being able to mount SSH keys from the host, Pandoc, the Amazon Web
-Services (AWS) SDK, and Java, which is required by the AWS SDK. The AWS
-SDK allowed to run R consumers reading from real-time data processing
-streams of \href{https://aws.amazon.com/kinesis/}{AWS Kinesis}. The
-applications were deployed on Amazon Elastic Container Service
-(\href{https://aws.amazon.com/ecs/}{ECS}). The main learnings from using
-R in Docker was the imporance of not only pinning the R package versions
-via MRAN, but also moving away from Debian testing to an LTS
-distribution to minimize the risk of breaking the automated builds of
-the Docker images and production jobs.
-
-A more recent related work is the \pkg{plumber}
-(\url{https://www.rplumber.io/}) package, which makes it even easier for
-R to be used to host web services through RESTful APIs. When R is
-running on a server, a different server can ask R to make a computation
-through an HTTP request and get an immediate response. For instance, if
-a data scientist at a company creates a machine learning model to tell
-which customers are likely to make a repeat purchase, if they serve the
-model through a plumber API then someone else on the marketing team
-could write software that sends different emails depending on that
-real-time prediction. Plumber syntax is simple to read and understand:
-the API is created by commenting above function names in an R script.
-
-\begin{verbatim}
-#* Return the sum of two numbers
-#* @get /random
-function(){
-  runif()
-}
-\end{verbatim}
-
-\emph{Example of a plumber API to generated a random number by visiting
-\texttt{http://127.0.0.1/random}}
-
-Plumber becomes far more powerful with the addition of Docker. Docker
-can be used to create images that
-\href{https://www.rplumber.io/docs/hosting.html\#docker}{include both R
-and plumber}. Then when the image is deployed, that creates the web
-endpoints that other services can hit and call the R code. This pattern
-of deploying code mimics those used by software engineering services
-created in Java, Python, and other modern languages. By using Docker to
-deploy plumber APIs, R can be used alongside those langages as a first
-class member of a software engineering technical stack.
-
-At T-Mobile, R is used to power real-time services that are called over
-a million times a day. T-Mobile created a set of neural network machine
-learning natural language processing models to help customer care agents
-manage text-based messages for customers
-\citep{t-mobile_enterprise_2018}. The models are convolutional neural
-networks that use the RStudio Keras package and are built on top of a
-Rocker Docker image. Since the models power tools for agents and
-customers, they need to have extremely high uptime and reliability--and
-they found that R is able to perform. The AI @ T-Mobile team open
-sourced their Docker images
-\href{https://github.com/tmobile/r-tensorflow-api}{on GitHub}.
-
-\href{http://restrserve.org}{RestRserve} is an alternative R web API
-framework with the main focus on being high-performant and robust. It
-also ships automated Docker builds at
-\url{https://hub.docker.com/r/rexyai/restrserve/dockerfile}.
-
-\hypertarget{deployment-to-the-cloud}{%
-\subsection{Deployment to the cloud}\label{deployment-to-the-cloud}}
-
-\label{deployment}
-
-The cloud is the natural environment of containers, and is the natural
-mechanism to deploy R server applications.
+applications and deploying them.
 
@@
 
-\label{azurecontainers} \textbf{\pkg{AzureContainers}} is an umbrealla
-package providing interfaces the three services of Microsoft's Azure
-Cloud, namely
+\label{azurecontainers} \textbf{\CRANpkg{AzureContainers}} is an
+umbrella package providing interfaces three commercial services of
+Microsoft's Azure Cloud, namely
 \href{https://azure.microsoft.com/en-us/services/container-instances/}{Container
@@
 Service} for orchestrated deployments. While a package like
-\pkg{plumber} provides the infrastructure for turning an R workflow a
-service, for production purposes it is usually necessary to take into
-account scalability, reliability and ease of management. A single
-container is limited in the volume of requests it can service, and if
-the container goes down or is occupied with processing a computationally
-intensive task, the service becomes unavailable. These issues can be
-mitigated by deploying not a single container, but a \emph{cluster} of
-containers, orchestrated as a single deployment, e.g., with
-\href{https://kubernetes.io}{Kubernetes}. A cluster features
-load-balancing, autoscaling of containers across numerous servers (in
-the cloud or on premise), and restarting failed ones.
-\pkg{AzureContainers} provides a lightweight yet powerful interface for
-working with Resource Manager, the framework for deploying and managing
-arbitrary resources in Azure. On the client side, it provides simple
-shells to the Docker, and to Kubernetes' \texttt{kubectl} and
-\texttt{helm} tools, that are needed for managing Kubernetes clusters.
-The package simplifies complex infrastructure management to a number of
-R funcction calls, given an Azure account with sufficient
+\CRANpkg{plumber} provides the infrastructure for turning an R workflow
+a service, for production purposes it is usually necessary to take into
+account scalability, reliability and ease of management. A
+\emph{cluster} of containers, orchestrated as a single deployment, e.g.,
+with \href{https://kubernetes.io}{Kubernetes}, can mitigate limitations
+on request volumes or a container occupied with computationally
+intensive task. A cluster features load-balancing, autoscaling of
+containers across numerous servers (in the cloud or on premise), and
+restarting failed ones. AzureContainers provides an R-based interface to
+these features for Azure. On the client side, it provides simple
+wrappers to Docker and to Kubernetes' management tools \texttt{kubectl}
+and \texttt{helm}. The package simplifies complex infrastructure
+management to a number of R function calls, given an Azure account with
+sufficient
 credit\footnote{See \emph{"Deploying a prediction service with Plumber"} vignette for details:  \href{https://cran.r-project.org/web/packages/AzureContainers/vignettes/vig01_plumber_deploy.html}{https://cran.r-project.org/web/packages/AzureContainers/vignettes/vig01\_plumber\_deploy.html}.}.
 
-\hypertarget{continuous-integration-and-continuous-delivery-colinfaynoamross-colinfay}{%
-\subsection{\texorpdfstring{Continuous integration and continuous
-delivery
-\citep[\citet{ColinFay}]{noamross}}{Continuous integration and continuous delivery , @ColinFay{[}@noamross, @ColinFay{]}}}\label{continuous-integration-and-continuous-delivery-colinfaynoamross-colinfay}}
-
-\label{cicd}
-
-The controlled nature of containers, i.e., it is possible to define the
-software environment very well, even on remote machines, make them also
-useful for continuous integration (CI) and continuous delivery of
-applications.
-
-\begin{itemize}
-\tightlist
-\item
-  DevOps
-
-  \begin{itemize}
-  \tightlist
-  \item
-    \url{https://www.opencpu.org/posts/opencpu-with-docker/}
-  \end{itemize}
-\end{itemize}
-
-When doing continuous integration and continuous delivery, it's crucial
-to test in an environment that matches the production environment. Tools
-like \texttt{Gitlab-CI} are built on top of Docker images: the user
-specifies a base Docker image, and the whole tests are run inside this
-environment. But, as we just said, this environment has to be fixed, but
-it also have to contain the necessary toolkit. In order to achieve that,
-\href{https://github.com/ColinFay/r-ci}{\texttt{r-ci}} combines
-\texttt{rocker} versioning and a series of tools specifically designed
-for testing. That way, package builders can use this image as a base
-image for there testing environment, without having to install the
-necessary packages every time they need to run a new test.
-
-\begin{itemize}
-\tightlist
-\item
-  \href{https://github.com/dynverse/dynwrap_containers/blob/master/.travis.yml}{dynwrap}
-  \citep{rcannood}
-
-  \begin{itemize}
-  \tightlist
-  \item
-    For this project, we use travis-ci to build rocker-derived
-    containers, test them, and only push them to Docker~Hub (from
-    travis-ci.org) if the integration tests succeed.
-  \end{itemize}
-\item
-  Google Cloud Build
-
-  \begin{itemize}
-  \tightlist
-  \item
-    \url{https://cloud.google.com/cloud-build/}
-  \item
-    Cloud Build is another continuous intergation service that helps
-    move the workload of building the Docker images to an online
-    service. Cloud Build can be set up to build the Dockerfiles on each
-    GitHub commit or release. This means you do not need to build the
-    Docker images locally, which can tie up resources since Docker
-    images can be several GBs and take a long time to compile. Google
-    Cloud Build works alongside Google Container Registry to allow you
-    to build private and public Docker images, which allows you to build
-    up your own dependency graphs for downstream applications.
-  \item
-    \url{https://code.markedmondson.me/googleCloudRunner/articles/cloudbuild.html}
-  \end{itemize}
-\end{itemize}
+The prevelance of Docker in industry naturally leads to usage of
+containers with R in such settings as well, as customers already manage
+platforms in Docker containers. These products often entail a high
+amount of open source software in combination with proprietary layers
+addign the relevant commercializable features. One such example is
+RStudio's data science platform
+\href{https://rstudio.com/products/team/}{RStudio Team}. It allows teams
+of data scientists and their respective IT/DevOps groups to develop and
+deploy code in R and Python around the RStudio Open Source Server inside
+of Docker images, without requiring users to learn new tools or directly
+interact with containers. The best practices for
+\href{https://support.rstudio.com/hc/en-us/articles/360021594513-Running-RStudio-with-Docker-containers}{running
+RStudio with Docker containers} as well as
+\href{https://github.com/rstudio/rstudio-docker-products}{Docker images}
+for RStudio's commerical products are openly available.
 
@@
 all dependencies from source can often time out before completion,
-making it frustrating for the average R user. \texttt{holepunch} removes
+making it frustrating for the average R user. \pkg{holepunch} removes
 some of these limitations by leveraging Rocker images that contain the
@@
 which eliminates a large part of the build time and in most cases
-results in a binder instance launching within a minute.
-\texttt{holepunch} as a side effect also creates a \texttt{DESCRIPTION}
-file which then turns any project into a research compendium
+results in a binder instance launching within a minute. \pkg{holepunch}
+as a side effect also creates a \texttt{DESCRIPTION} file which then
+turns any project into a research compendium
 \citep{marwick_packaging_2018}. The \texttt{Dockerfile} included with
@@
 not part of standard container recipes and must be added. The
-Rocker~Project provides containers pre-installed with the RStudio IDE.
+Rocker~Project provides containers pre-installed with the RStudio~IDE.
 In cases where users store nontrivial amounts of data for their
@@
 
-Using \textbf{GPUs} (graphical processing units) as a specialised
-hardware from containerized common work environments is also possible
-and useful \citep{haydel_enhancing_2015}. GPUs are increasingly popular
-for compute-intensive machine learning (ML) tasks, e.g., deep artificial
-neural networks \citep{schmidhuber_deep_2015}. Though in this case,
-containers are not completely portable between hardware environments,
-but the software stack for ML with GPUs is so complex to set up that a
-ready-to-use container is helpful. Containers running GPU software
-require drivers and libraries specific to GPU models and versions, and
-containers require a specialized runtime to connect to the underlying
-GPU hardware. For NVIDIA GPUs, the
+\label{rocker-gpu} Using \textbf{GPUs} (graphical processing units) as a
+specialised hardware from containerized common work environments is also
+possible and useful \citep{haydel_enhancing_2015}. GPUs are increasingly
+popular for compute-intensive machine learning (ML) tasks, e.g., deep
+artificial neural networks \citep{schmidhuber_deep_2015}. Though in this
+case, containers are not completely portable between hardware
+environments, but the software stack for ML with GPUs is so complex to
+set up that a ready-to-use container is helpful. Containers running GPU
+software require drivers and libraries specific to GPU models and
+versions, and containers require a specialized runtime to connect to the
+underlying GPU hardware. For NVIDIA GPUs, the
 \href{https://github.com/NVIDIA/nvidia-docker}{NVIDIA Container Toolkit}
@@
 
-\textbf{Teaching} is a further example where sharing a prepared
-computing environment can greatly improve the process, especially for
-courses that require access to a relatively complex setup of software
-tools, e.g., as in the case of database systems. R is a fantastic tool
-when it comes to interfacing with databases: almost every open source
-and proprietary database system has an R package that allows users to
-connect and interact with the it This flexibility is even broader now
-that we have tools like \texttt{DBI}, that allows to create a common API
-for interfacing these databases, or like \texttt{dbplyr}, which are
-designed to run \texttt{dplyr} code straight against the database. But
-learning and teaching these tools comes with a cost: the cost of
-deploying or having access to an environment with the softwares and
+Teaching is a further example where shared computing environments and
+sandboxing can greatly improve the process. First, \textbf{prepared
+environments for teaching} are especially helpful for courses that
+require access to a relatively complex setup of software tools, e.g.,
+database systems. R is very useful tool for interfacing with databases,
+because almost every open source and proprietary database system has an
+R package that allows users to connect and interact with the it. This
+flexibility is even broadened by \CRANpkg{DBI}, which allows to create a
+common API for interfacing these databases, or the \CRANpkg{dbplyr}
+package, which runs \CRANpkg{dplyr} code straight against the database
+as queries. But learning and teaching these tools comes with the cost of
+deploying or having access to an environment with the software and
 drivers installed. For people teaching R, it can become a barrier if
-they need to install local versions of the drivers, or to connect to
-remote instances which might or might not be made available by IT
+they need to install local versions of database drivers, or to connect
+to remote instances which might or might not be made available by IT
 services. Giving access to a sandbox for the most common database
 environments is the idea behind
-\textbf{\href{https://github.com/ColinFay/r-db}{\texttt{r-db}}}
-\citep{ColinFay}, a Docker image that contains everything needed to
-connect, from R, to a database. Notably, with \texttt{r-db}, the users
-don't have to install complex drivers or to configure their machine in a
-specific way. On top of that, \texttt{r-db} comes with a comprehensive
-\href{http://colinfay.me/r-db/}{online guide}, explaining how to spin a
-Docker instance of a specific database, and how to use \texttt{r-db} to
-connect to it. Each package contained inside this image also has a
-series examples, so that the user can get started with the database
-right away. The \texttt{rocker/tidyverse} base image ensures that users
-can also readily use packages for analysis, display, and reporting.
+\href{https://github.com/ColinFay/r-db}{\texttt{r-db}}, a Docker image
+that contains everything needed to connect to a database from R.
+Notably, with \texttt{r-db}, the users don't have to install complex
+drivers or to configure their machine in a specific way. The
+\texttt{rocker/tidyverse} base image ensures that users can also readily
+use packages for analysis, display, and reporting. Second, the idea of a
+common environment and partitioning allow using \textbf{containers in
+teaching for secure execution and automated testing} of submissions by
+students. \href{https://dodona.ugent.be}{Dodona} is a web platform
+developed at Ghent University and is used to teach students basic
+programming skills, and it uses Docker containers to test submissions by
+students. This means that both the code testing the students'
+submissions and the submission itself are executed in a predictable
+environment, avoiding compatibility issues between the wide variety of
+configurations used by students. The containerization is also used to
+shield the Dodona servers from bad or even malicious code: memory, time
+and I/O~limits are used to make sure students can't overload the system.
+The web application managing the containers communicates with them by
+sending configuration information as a JSON document over standard
+input. Every Dodona Docker image shares a \texttt{main.sh} file that
+passes through this information to the actual testing framework, while
+setting up some error handling. The testing process in the Docker
+containers sends back the test results by writing a JSON document to its
+standard output channel. In June 2019, R support was added to Dodona
+using an image derived from the \texttt{rocker/r-base} image that sets
+up the \texttt{runner} user and and \texttt{main.sh} file expected by
+Dodona\footnote{\href{https://github.com/dodona-edu/docker-images/blob/master/dodona-r.dockerfile}{https://github.com/dodona-edu/docker-images/blob/master/dodona-r.dockerfile}}.
+It also installs the packages required for the testing framework and the
+exercises so that this doesn't have to happen every time a student's
+submission is evaluated. The actual testing of R exercises is done using
+a custom framework loosely based on \CRANpkg{testthat}. During the
+development of the testing framework it was found that the
+\CRANpkg{testthat} framework did not provide enough information to its
+reporter system to send back all the fields required by Dodona to render
+its feedback. Right now, multiple statistics courses are developing
+exercises to automate the feedback for their lab classes.
 
@@
 
-\hypertarget{using-docker-in-a-commercial-data-science-platform}{%
-\subsection{Using Docker in a commercial data science
-platform}\label{using-docker-in-a-commercial-data-science-platform}}
-
-\label{rstudio}
-
-In addition to embedding RStudio Open Source Server inside of Docker
-images, such as Rocker, RStudio offers the moduler data science platform
-\href{https://rstudio.com/products/team/}{RStudio Team}, a commercial
-product allowing teams of data scientists and their respective IT/DevOps
-groups to develop and deploy code in R and Python. RStudio team enables
-these groups to use the benefits of containers, without requiring users
-to learn new tools or directly interact with containers. Notable
-patterns include:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  \emph{To deploy and manage RStudio Professional Products:} Many
-  RStudio customers are managing professional products in Docker
-  containers as an alternative to bare-metals servers or virtual
-  machines. Best practices for
-  \href{https://support.rstudio.com/hc/en-us/articles/360021594513-Running-RStudio-with-Docker-containers}{running
-  RStudio with Docker containers} as well as
-  \href{https://github.com/rstudio/rstudio-docker-products}{Docker
-  images} for RStudio's commerical products are available.
-\item
-  \emph{To run interactive RStudio sessions on a cluster:} In the
-  RStudio Server Pro 1.2 release in 2019, RStudio added new
-  functionality called
-  \href{https://solutions.rstudio.com/launcher/overview/}{Launcher}. It
-  gives users the ability to spawn R sessions and background jobs in a
-  scalable way on external clusters, e.g.,
-  \href{https://support.rstudio.com/hc/en-us/articles/360019253393-Using-Docker-images-with-RStudio-Server-Pro-Launcher-and-Kubernetes}{Kubernetes
-  based on Docker images} or \href{https://slurm.schedmd.com/}{Slurm}
-  clusters, and optionally, with Singularity containers. A key benefit
-  of Launcher is the ability for R and Python users to interactively
-  work in RStudio without learning about Docker at all - while still
-  leveraging containers and Kubernetes. Users can submit batch jobs to
-  Kubernetes clusters without writing specific deployment scripts.
-\end{enumerate}
-
-\href{https://rstudio.com/products/connect/}{RStudio Connect} is a
-commercial publishing platform to deploy dashboards, REST APIs, and
-other R and Python content into production. RStudio Connect uses an
-approach to isolate deployed R applications that is similar to Docker's
-filesystem and process namespacing and leverages the
-\href{https://rstudio.com/products/package-manager/}{RStudio Package
-Manager} to manage R environments including system dependencies. RStudio
-Package manager may also be used to
-\href{https://environments.rstudio.com/docker}{manage their R
-environments and package versions with Docker} and to ensure
-deterministic image builds.
-
 \hypertarget{processing}{%
@@
 
-The portability of containers becomes particularly useful when complex
-processing tasks shall be offloaded to a server.
-
-\begin{itemize}
-\tightlist
-\item
-  Docker images for cloud services
-
-  \begin{itemize}
-  \tightlist
-  \item
-    A popular application of using Docker is to create the environments
-    that can run in parallel for speeding up R code. For example,
+The portability of containerized environments becomes particularly
+useful for improving expensive processing of data or shipping complex
+processing pipelines.
+
+First, it is possible to \textbf{offload complex processing to a
+server}. \ldots{}
+
+Second, when processes can be \textbf{run in parallel} for speeding up R
+code.
+
+For example,
 \href{https://CRAN.R-project.org/package=googleComputeEngineR}{\texttt{googleComputeEngineR}}'s
-    \texttt{gce\_vm\_cluster()} function can create clusters of 2 or
-    more virtual machines, running multi-CPU architectures. Instead of
-    running a local R script with the local CPU and RAM restrictions,
-    the same code can be processed on all CPU threads of the cluster of
-    machines in the cloud, all running in a Docker container with the
-    same R environments. This is achieved through
-    \texttt{googleComputeEngineR}'s integration with the R parralisation
-    library
+\texttt{gce\_vm\_cluster()} function can create clusters of 2 or more
+virtual machines, running multi-CPU architectures. Instead of running a
+local R script with the local CPU and RAM restrictions, the same code
+can be processed on all CPU threads of the cluster of machines in the
+cloud, all running in a Docker container with the same R environments.
+This is achieved through \texttt{googleComputeEngineR}'s integration
+with the R parralisation library
 \href{https://CRAN.R-project.org/package=future}{\texttt{library(future)}
@@
 \href{https://cloudyr.github.io/googleComputeEngineR/articles/massive-parallel.html}{some
-    demonstrations are available here}.
-  \end{itemize}
-\item
-  \href{https://cloud.run}{Google Cloud Run} - This is a CaaS
-  (Containers as a Service) that lets you launch a Docker container
-  without worrying about underlying infrastructure. This dispenses with
-  the developer creating a cloud server to run the Docker image on, by
-  abstracting away those servers to a more serverless configuration.
-  Cloud Run lets you run your code on top of a managed or your own
-  Kubernetes cluster running Knative, and can accept any Docker image.
-  The service takes care of network ingress, scaling machines up and
-  down to zero, authentication and authorisation, all features which are
-  non-trivial for a developer to create on their own. This can be used
-  to scale up R code to millions of instances if they need to, with
-  little or no changes to existing code. An R implementation is shown
-  here at
+demonstrations are available here}. - \href{https://cloud.run}{Google
+Cloud Run} - This is a CaaS (Containers as a Service) that lets you
+launch a Docker container without worrying about underlying
+infrastructure. This dispenses with the developer creating a cloud
+server to run the Docker image on, by abstracting away those servers to
+a more serverless configuration. Cloud Run lets you run your code on top
+of a managed or your own Kubernetes cluster running Knative, and can
+accept any Docker image. The service takes care of network ingress,
+scaling machines up and down to zero, authentication and authorisation,
+all features which are non-trivial for a developer to create on their
+own. This can be used to scale up R code to millions of instances if
+they need to, with little or no changes to existing code. An R
+implementation is shown here at
 \href{https://github.com/MarkEdmondson1234/cloudRunR}{cloudRunR} which
 uses Cloud Run to create a scalable R plumber API.
+\url{https://code.markedmondson.me/googleCloudRunner/index.html}
+
+\href{https://cloud.google.com/cloud-build/}{Google Cloud Build} and the
+Google Container Registry are a continuous intergation service
+respectively image registry that offload building of images to the
+cloud, while serving the needs of commercial environments that helps
+move the workload of building the Docker images to an online service,
+while Cloud Build can be set up to build the Dockerfiles on each GitHub
+commit or release. This means you do not need to build the Docker images
+locally, which can tie up resources since Docker images can be several
+GBs and take a long time to compile. Google Cloud Build works alongside
+to allow you to build private and public Docker images, which allows you
+to build up your own dependency graphs for downstream applications. -
+\url{https://code.markedmondson.me/googleCloudRunner/articles/cloudbuild.html}
 
@@
 \item
-    How does this relate to
-    \url{https://code.markedmondson.me/googleCloudRunner/index.html} ?`
-  \end{itemize}
-\item
-  \texttt{batchtools} \citep{Lang2017batchtools} can
+  \CRANpkg{batchtools} \citep{Lang2017batchtools} can
   \href{https://mllg.github.io/batchtools/reference/makeClusterFunctionsDocker.html}{schedule
@@
 \item
-  \pkg{drake},
+  \CRANpkg{drake},
   \url{https://docs.ropensci.org/drake/index.html?q=docker\#with-docker}
@@
 
-\hypertarget{packaging-programs-for-processing-pipelines}{%
-\subsection{Packaging programs for processing
-pipelines}\label{packaging-programs-for-processing-pipelines}}
-
-\label{packaging}
-
-Docker is perfectly suited for packaging and executing any software and
-data, and is a good candidate for building complex processing pipelines
-that are independent from the host programming language. Due to its
-original use case (see~\nameref{introduction}), Docker has no standard
-mechanisms for chaining containers together; it lacks definitions and
-protocols for environment variables, volume mounts and/or ports that
-could enable transfer of input (parameters and data) and output
-(results) to and from containers. Some packages, e.g. \pkg{containerit},
-do provide Docker images that can be used similarly to CLIs, but their
-usage is
+In the RStudio Server Pro 1.2 release in 2019, RStudio added new
+functionality called
+\href{https://solutions.rstudio.com/launcher/overview/}{Launcher}. It
+gives users the ability to spawn R sessions and background jobs in a
+scalable way on external clusters, e.g.,
+\href{https://support.rstudio.com/hc/en-us/articles/360019253393-Using-Docker-images-with-RStudio-Server-Pro-Launcher-and-Kubernetes}{Kubernetes
+based on Docker images} or \href{https://slurm.schedmd.com/}{Slurm}
+clusters, and optionally, with Singularity containers. A key benefit of
+Launcher is the ability for R and Python users to interactively work in
+RStudio without learning about Docker at all - while still leveraging
+containers and Kubernetes. Users can submit batch jobs to Kubernetes
+clusters without writing specific deployment scripts.
+
+\label{pipelines} Third, containers are perfectly suited for
+\textbf{packaging and executing software pipelines} and required data.
+Containers allow building complex processing pipelines that are
+independent from the host programming language. Due to its original use
+case (see~\nameref{introduction}), Docker has no standard mechanisms for
+chaining containers together; it lacks definitions and protocols for
+environment variables, volume mounts and/or ports that could enable
+transfer of input (parameters and data) and output (results) to and from
+containers. Some packages, e.g., \pkg{containerit}, do provide Docker
+images that can be used similarly to CLIs, but their usage is
 cumbersome\footnote{\href{https://o2r.info/containerit/articles/container.html}{https://o2r.info/containerit/articles/container.html}}.
-
 \textbf{\pkg{outsider}} (\url{https://antonellilab.github.io/outsider/})
@@
 
-\hypertarget{research-compendia}{%
-\subsection{Research Compendia}\label{research-compendia}}
+\hypertarget{packaging-research-reproducibly}{%
+\subsection{Packaging research
+reproducibly}\label{packaging-research-reproducibly}}
 
@@
 
-\begin{itemize}
-\tightlist
-\item
-  researchcompendia.science
-\item
-  \url{https://github.com/benmarwick/rrtools}
-\item
-  \pkg{renv}, \url{https://rstudio.github.io/renv/articles/docker.html}
-\item
-  what to copy in and what not
-\end{itemize}
-
-\hypertarget{publication-reproducibility-audits}{%
-\subsection{Publication Reproducibility
-Audits}\label{publication-reproducibility-audits}}
+Containers provide a high degree of isolation that is often desirable
+when attempting to capture a specific computational environment so that
+others can reproduce and extend a research result. Many computationally
+intensive research projects depend on specific versions of original and
+third-party software packages in diverse languages, joined together to
+form a pipeline through which data flows. New releases of even just a
+single piece of software in this pipeline can break the entire workflow,
+making it difficult to find the error and difficult for others to reuse
+existing pipelines. These breakages can make the original the results
+irreproducible and not extandable. The chance of a substantial
+disruption like this is high in a multi-year research project where key
+pieces of third-party software may have several major updates over the
+duration of the project. The classical ``paper'' article is insufficient
+to adequately communicate the knowledge behind such research projects
+\citep[cf.][]{donoho_invitation_2010,marwick_how_2015}.
+
+\citet{gentleman_statistical_2007} coined the term \textbf{Research
+Compendium} for a dynamic document together with supporting data and
+code. They use the R package system \cite{core_writing_1999} as for the
+functional prototype to structuring, validation, and distribution of
+research compendia. This concept has been taken up and
+extended\footnote{See full literature list at \href{https://research-compendium.science/}{https://research-compendium.science/}.},
+not the least by applying containerisation and other methods for
+managing computing environments---see Section~\nameref{envs}. Containers
+give the researcher an isolated environment to assemble these research
+pipelines with specific versions of software. Research workflows in
+containers are safe from contamination from other activities occuring on
+the researcher's computer, for example the installation of newest
+version of packages for teaching demonstrations. Given the users in this
+scenario, i.e., often academics with limited formal software development
+training, templates and assistance with containers around research
+compendia is essential. In many fields we see that a typical unit of
+research for a container is a research report or journal article, where
+the container holds the compendium, or self-contained set of data (or
+connections to data elsewhere) and code files needed to fully reproduce
+the article \citep{marwick_packaging_2018}. The package \pkg{rrtools}
+(\url{https://github.com/benmarwick/rrtools}) provides a template and
+convienence functions to apply good practices for research compendia,
+including a starter \texttt{Dockerfile}. Images of compendium containers
+can be hosted on services such as Docker~Hub for convienient sharing
+among collaborators and others. Similarly, packages such as
+\pkg{containerit} and \pkg{dockerfiler} can be used to manage the
+\texttt{Dockerfile} to be archived with a compendium on a data
+repository (e.g.~\href{https://zenodo.org/}{Zenodo},
+\href{https://dataverse.org/}{Dataverse},
+\href{https://figshare.com/}{Figshare}, \href{https://osf.io/}{OSF}). A
+typial compendium's \texttt{Dockerfile} will pull a rocker image fixed
+to a specific version of R, and install R packages from the MRAN
+repository to ensure the package versions are tied to a specific date,
+rather than the most recent version. Future researchers can download the
+compendium from the repository and run the included \texttt{Dockerfile}
+to build a new image that recreates the computational environment used
+to produce the original research results. If building the image fails,
+the human-readable instructions in a \texttt{Dockerfile} are the
+starting point for rebuilding the environment. Further safeguarding
+practices are currently under development or not part of common practice
+yet, such as preservation of images \citep{emsley_framework_2018} and
+storing them alongside \texttt{Dockerfile}s
+\citep[cf.][]{nust_opening_2017}, or pinning of system libraries beyond
+the tagged base images, which may be seen as stable or dynamic depending
+on the applied time scale
+\citep[see discussion on `debian:testing` base image in][]{RJ-2017-065}.
 
@@
 \emph{Reproducibility and Replicability in Science} is that journals
-``consider ways to ensure computational reproducibility for publications
-that make claims based on computations'' \citep{NASEM2019}. In fields
-such as Political Science and Economics, journals are increasingly
-adopting policies requiring authors to publish the code and data
-required to reproduce computational findings reported in published
-manuscripts, subject to independent verification
+\emph{``consider ways to ensure computational reproducibility for
+publications that make claims based on computations''}
+\citep{NASEM2019}. In fields such as Political Science and Economics,
+journals are increasingly adopting policies requiring authors to publish
+the code and data required to reproduce computational findings reported
+in published manuscripts, subject to independent verification
 \citep{Jacoby2017,Vilhuber2019,Alvarez2018,Christian2018,Eubank2016,King1995}.
@@
 availability of software dependencies are common. R is gaining
-popularity in these communities (in a sample of 105 replication packages
-published by the \emph{American Journal of Political Science} (AJPS)
-over 65\% use R). Projects such as NSF Whole Tale use the Rocker Project
-community images with the goal of improving the reproducibility of
-published research artifacts and simplifying the publication and
-verification process for both authors and journals by reducing errors
-and time spent specifying the environment.
-
-\hypertarget{development-and-debugging}{%
-\subsection{Development and debugging}\label{development-and-debugging}}
+popularity in these communities, such as creating a research compendium.
+In a sample of 105 replication packages published by the \emph{American
+Journal of Political Science} (AJPS) over 65\% use R. The NSF-funded
+Whole Tale project uses the Rocker Project community images with the
+goal of improving the reproducibility of published research artifacts
+and simplifying the publication and verification process for both
+authors and reviewers by reducing errors and time spent specifying the
+environment.
+
+\hypertarget{development-debugging-and-testing}{%
+\subsection{Development, debugging, and
+testing}\label{development-debugging-and-testing}}
 
@@
 ad-hoc or to provide very specific environments that are not needed or
-not easily available in day-to-day development. These environments may
-have specific versions of R, of R extension packages, and of system
-libraries used by R extension packages, and all of the above in a
-specific combination.
-
-First, such containers can greatly facilitate \textbf{fixing bugs},
-because developers can readily start and later discard a container to
-investigate a bug report without affecting their regular system, and
-potentially having to revert changes back. Using the Rocker images with
-RStudio, these disposable environments lack no development comfort
-(cf.~Section~\nameref{compendia}). \citet{eddelbuettel_debugging_2019}
-describes an example how a Docker container was used to debug an issue
-with a package only occuring with a particular version of Fortran, and
-using tools which are not readily available on all platforms (e.g., not
-on macOS). For the \href{https://www.r-spatial.org/}{R-spatial
-community}, using containers is particularly useful, because of the
-strong integration of system libraries in core packages for geospatial
-data modelling and analysis. Purpose built Docker are used to prepare
-for upcoming releases of system libraries, individual bug reports, and
-for the lowest supported versions of system
+not easily available in day-to-day development for the purposes of
+developing R packages. These environments may have specific versions of
+R, of R extension packages, and of system libraries used by R extension
+packages, and all of the above in a specific combination.
+
+First, such containers can greatly facilitate \textbf{fixing bugs and
+quick evaluation}, because developers and users can readily start and
+later discard a container to investigate a bug report or try out a piece
+of software \citep[cf.][]{ooms_opencpu_2017}. The container does not
+affect their regular system. Using the Rocker images with RStudio, these
+disposable environments lack no development comfort
+(cf.~Section~\nameref{compendia}). \citet{ooms_opencpu_2017} describes
+how \texttt{docker\ exec} can be used to get a root shell in a container
+for cusotmization during software evaluation.
+\citet{eddelbuettel_debugging_2019} describes an example how a Docker
+container was used to debug an issue with a package only occuring with a
+particular version of Fortran, and using tools which are not readily
+available on all platforms (e.g., not on macOS).
+
+Second, the strong integration of \textbf{system libraries in core
+packages} in the \href{https://www.r-spatial.org/}{R-spatial community}
+makes containers essential for stable and proactive development of
+common classes for geospatial data modelling and analysis. For example,
+GDAL \citep{gdal_2019} is a crucial library in the geospatial domain.
+GDAL is a system dependen\^{}cy allowing R packages such as
+\CRANpkg{sf}, which provides the core data model for geospatial vector
+data, or \CRANpkg{rgdal}, to accomodate users to be able to read and
+write hundreds of different spatial raster and vector formats.
+\CRANpkg{sf} and \CRANpkg{rgdal} have hundreds of indirect reverse
+imports and dependencies and therefore the maintainers spend a lot of
+efforts not to break these. Purpose built Docker are used to prepare for
+upcoming releases of system libraries, individual bug reports, and for
+the lowest supported versions of system
 libraries\footnote{Cf.
@@
 
-There are also special images for identifying problems beyond the mere R
-code, such as \textbf{debugging R memory problems}. The images
-significantly reduce the barrier to follow complex steps for fixing
-memory allocation bugs \citep[cf. Section~4.3 in][]{core_writing_1999}.
-These problems are hard to debug and critical, both because when they
-occur they lead to fatal crashing processes.
+Third, there are special purpose images for identifying problems beyond
+the mere R code, such as \textbf{debugging R memory problems}. The
+images significantly reduce the barrier to follow complex steps for
+fixing memory allocation bugs
+\citep[cf. Section~4.3 in][]{core_writing_1999}. These problems are hard
+to debug and critical, both because when they occur they lead to fatal
+crashing processes.
 \href{https://github.com/rocker-org/r-devel-san}{\texttt{rocker/r-devel-san}}
@@
 such problems with gcc and clang compilers, respectively
-\citep[cf.~\pkg{sanitizers} for examples,][]{eddelbuettel_sanitizers_2014}.
+\citep[cf.~\CRANpkg{sanitizers} for examples,][]{eddelbuettel_sanitizers_2014}.
 The image \href{https://github.com/wch/r-debug}{\texttt{wch/r-debug}} is
@@
 
-Second, containers are useful for \textbf{testing} R code during
+Fourth, containers are useful for \textbf{testing} R code during
 development. To submit a package to CRAN, an R package must work with
@@
 Section~\nameref{rhub}) makes it easy to ensure that no errors occur,
-but to fix errors a local setup is still often warranted, e.g.~using the
-image \texttt{rocker/r-devel}, and to test packages with native code,
-which can make the process more complex
+but to fix errors a local setup is still often warranted, e.g., using
+the image \texttt{rocker/r-devel}, and to test packages with native
+code, which can make the process more complex
 \citep[cf.][]{eckert_building_2018}. The R-hub Docker images can also be
@@
 
-Third, Docker images can be used \textbf{on CI platforms} to streamline
+Fifth, Docker images can be used \textbf{on CI platforms} to streamline
 the testing of packages. \citet{ye_docker_2019} describes how they speed
@@
 \texttt{docker\ exec}, e.g., the package check or rendering of
-documentation. \citet{cardozo_faster_2018}, also on Travis~CI saved time
-by re-using the testing image as the base iamge for an image intended
-for publication on Docker~Hub. Especially for long running tests or
-complex system dependencies, these approaches seems to improve the
-development. Not due to a concern about time, but to control the
-environment used on a CI server, even this manuscript was rendered into
-a PDF and deployed to a GitHub-hosted website (see \texttt{.travis.yml}
-and \texttt{Dockerfile} in the manuscript repository). This gives on the
-one hand easy access after every update of the R Markdown source code,
-and on the other hand a second controlled environment making sure that
-the article renders successfully and correctly.
+documentation. \citet{cardozo_faster_2018} saved time, also on
+Travis~CI, by re-using the testing image as the base for an image
+intended for publication on Docker~Hub.
+\href{https://github.com/ColinFay/r-ci}{\texttt{r-ci}} is in turn used
+with \href{https://docs.gitlab.com/ee/ci/}{GitLab~CI}, which itself is
+built on top of Docker images: the user specifies a base Docker image,
+and the whole tests are run inside this environment. The \texttt{r-ci}
+image stack combines \texttt{rocker} versioning and a series of tools
+specifically designed for testing in a fixed environment with a
+customized list of preinstalled packages. Especially for long running
+tests or complex system dependencies, these approaches to seperate
+installation of build dependencies with code testing streamline the
+development process. Not due to a concern about time, but to control the
+environment used on a CI server, even this manuscript is rendered into a
+PDF and deployed to a GitHub-hosted website with every change (see
+\texttt{.travis.yml} and \texttt{Dockerfile} in the manuscript
+repository). This gives on the one hand easy access after every update
+of the R Markdown source code, and on the other hand a second controlled
+environment making sure that the article renders successfully and
+correctly.
 
@@
 
+\label{conclusions}
+
 This article is a snapshot of the R-corner in a universe of applications
-built many-faced piece of software, Docker. \texttt{Dockerfile}s and
-Docker images are the go-to methods for collaboration between roles in
-an organisation, such as development and IT operations teams, and
+built with a many-faced piece of software, Docker. \texttt{Dockerfile}s
+and Docker images are the go-to methods for collaboration between roles
+in an organisation, such as development and IT operations teams, and
 between parties in the communication of knowledge, such as research
@@
 machines easily (e.g., locally, one cloud providers VM, another cloud
-provider's Container-as-a-Service). Collaboration, reproducibility, and
-portability are the common themes behind R packages, use cases, and
-applications in this work.
-
-The R packages and use cases presented show the positive drive in the
-community to innovate, but also the challenge of keeping up with a
-continuously evolving landscape. The use cases contributed by co-authors
-also have a degree of overlap, which can be expected as a common
-language and understanding of good practices is still taking shape.
-Also, the ease with which one can create a complex software systems with
-Docker, such as an independent Docker image stack, to serve one's
-specific needs leads to parallel developments. This ease-of-DIY in
-combination with the difficulty to compose environments based on
-\texttt{Dockerfile}s alone is a further reason for
+provider's Container-as-a-Service). Reproducible environments,
+scalability and efficiency, and portability across
+platforms/infrastructures are the common themes behind R packages, use
+cases, and applications in this work.
+
+The R packages and use cases presented show the growing number of users,
+developers, and real-world applications in the community and the
+resulting innovations. But the applications also point to the challenge
+of keeping up with a continuously evolving landscape. The use cases
+contributed by co-authors also have a degree of overlap, which can be
+expected as a common language and understanding of good practices is
+still taking shape. Also, the ease with which one can create a complex
+software systems with Docker, such as an independent Docker image stack,
+to serve one's specific needs leads to parallel developments. This
+ease-of-DIY in combination with the difficulty to compose environments
+based on \texttt{Dockerfile}s alone is a further reason for
 \textbf{fragmentation}. Instructions can be outsourced into
@@
 Docker daemon with R code. It remains to be seen which approach to
-control Docker, via the Docker API as \texttt{stevedore} or via system
-calls as \texttt{dockyard}/\texttt{docker}/\texttt{dockr}, is more
-sustainable, or if the question will be answered by the endurance of
-maintainers and sufficient funding. Similarly, the capturing of
-environments and their serialization in form of a \texttt{Dockerfile}
-currently happens at different levels of abstraction and re-use of
-functionality seems reasonable, e.g. \texttt{liftr} could generate the
-environment with \texttt{containerit}, which in turn may use
-\texttt{dockerfiler} for low level objects representing a
-\texttt{Dockerfile}. In this consolidation, the Rocker~Project could
-play the role of coordinating entity. Though for the moment, the sign of
-the times points to more experimentation and feature growth, e.g.~images
-for GPU-based computing and artificial intelligence. Even with coding
-being more and more accepted as a required, and achievable skill, an
-easier access, for example by exposing containerisation benefits via
-simple user interfaces in the users' IDE, could be an important next
-step. Currently containerisation happens more in the background at the
-system level.
+control Docker, via the Docker API as \pkg{stevedore} or via system
+calls as \pkg{dockyard}/\pkg{docker}/\pkg{dockr}, is more sustainable,
+or if the question will be answered by the endurance of maintainers and
+sufficient funding. Similarly, the capturing of environments and their
+serialization in form of a \texttt{Dockerfile} currently happens at
+different levels of abstraction and re-use of functionality seems
+reasonable, e.g., \pkg{liftr} could generate the environment with
+\pkg{containerit}, which in turn may use \pkg{dockerfiler} for low level
+R objects representing a \texttt{Dockerfile} and its instructions. In
+this consolidation, the Rocker~Project could play the role of
+coordinating entity. Though for the moment, the sign of the times points
+to more experimentation and feature growth, e.g., images for GPU-based
+computing and artificial intelligence. Even with coding being more and
+more accepted as a required, and achievable skill, an easier access, for
+example by exposing containerisation benefits via simple user interfaces
+in the users' IDE, could be an important next step. Currently
+containerisation happens more in the background at the system level.
 
@@
 foreseeable future in the \emph{Rockerverse}, and just like the Rocker
-images have matured over years of use and millions of runs the new ideas
-and prototypes will have to proof themselves. It should be noted that
-the dominant position of Docker is a blessing and a curse for these
+images have matured over years of use and millions of runs, the new
+ideas and prototypes will have to proof themselves. It should be noted
+that the dominant position of Docker is a blessing and a curse for these
 goals. It could be wise to start experimenting with non-Docker
-containerisation tools now, e.g.~R packages interfacing with other
+containerisation tools now, e.g., R packages interfacing with other
 container engines such as
@@
 \href{https://coreos.com/rkt/}{rkt}, or an R package for creating
-Singularity image recipe files. Such efforts help to avoid lock-in and
+\texttt{Singularity} files. Such efforts can help to avoid lock-in and
 to design sustainable workflows based on concepts of
@@
 processing and deployment to cloud services. CF wrote paragraphs about
-\pkg{r-online}, \pkg{dockerfiler}, \pkg{r-ci} and \pkg{r-db}. BM wrote
-the section on research compendia and made the project Binder-ready. HN
-\& JN co-wrote the section on enterprise production environments
+\pkg{r-online}, \pkg{dockerfiler}, \pkg{r-ci} and \pkg{r-db}. SL
+contributed content on RStudio's usage of Docker. BM wrote the section
+on research compendia and made the project Binder-ready. HN \& JN
+co-wrote the section on enterprise production environments
 (\pkg{googleComputeEngineR}, Google Cloud Run, Kubernetes). KR wrote the
-section about \pkg{wholepunch}. NR wrote paragraphs about shared work
+section about \pkg{holepunch}. NR wrote paragraphs about shared work
 environments and GPUs. LS \& NT wrote the section on Bioconductor. CW
 wrote the sections on Whole Tale and contributed the publication
-reproducibility audit use case. NX wrote the section on \pkg{liftr}. All
-authors approved the final version. This articles was collaboratively
-written at
+reproducibility audit use case. NX contributed content on \pkg{liftr}.
+All authors approved the final version. This articles was
+collaboratively written at
 \href{https://github.com/nuest/rockerverse-paper/}{https://github.com/nuest/rockerverse-paper/}.
@@
 \href{mailto:me@nanx.me}{\nolinkurl{me@nanx.me}}
+
+\address{%
+Charlotte Van Petegem\\
+Ghent University\\
+Department WE02\\ Krijgslaan 281, S9\\ 9000 Gent, Belgium\\ \orcid{0000-0003-0779-4897}\\
+}
+\href{mailto:charlotte.vanpetegem@ugent.be}{\nolinkurl{charlotte.vanpetegem@ugent.be}}

### 2020-01-23 manuscript.Rmd

@@
 Some packages, e.g., \pkg{containerit}, do provide Docker images that can be used similarly to CLIs, but their usage is cumbersome\footnote{\href{https://o2r.info/containerit/articles/container.html}{https://o2r.info/containerit/articles/container.html}}.
-**\pkg{outsider}** ([https://antonellilab.github.io/outsider/](https://antonellilab.github.io/outsider/)) tackles the problem of integrating external programs into an R workflow \citep{bennett_outsider_2020}.
+**\pkg{outsider}** ([https://docs.ropensci.org/outsider/](https://docs.ropensci.org/outsider/)) tackles the problem of integrating external programs into an R workflow \citep{bennett_outsider_2020}.
 Installation and usage of external programs can be difficult, convoluted and even impossible if the platform is incompatible.

### 2020-01-23 manuscript.tex

@@
 cumbersome\footnote{\href{https://o2r.info/containerit/articles/container.html}{https://o2r.info/containerit/articles/container.html}}.
-\textbf{\pkg{outsider}} (\url{https://antonellilab.github.io/outsider/})
+\textbf{\pkg{outsider}} (\url{https://docs.ropensci.org/outsider/})
 tackles the problem of integrating external programs into an R workflow
@@
 Dom Bennett\\
-University of Gothenburg, Sweden\\
-Department of Biological and Environmental Sciences\\ Carl Skottsbergs gata 22B\\ 413 19 Gothenburg, Sweden\\ \orcid{0000-0003-2722-1359}\\
+Gothenburg Global Biodiversity Centre, Sweden\\
+Carl Skottsbergs gata 22B\\ 413 19 Göteborg, Sweden\\ \orcid{0000-0003-2722-1359}\\
 }

### 2020-01-26 manuscript.Rmd
added some 'dockr' remarks

@@
     email: charlotte.vanpetegem@ugent.be
+  - name: Lars Kjeldgaard
+    # https://github.com/smaakage85
+    affiliation: Danish Tax Authorities
+    address:
+      - Oestbanegade 123
+      - 2100, Koebenhavn Oe
+    email: lars_kjeldgaard@hotmail.com
 abstract: >
@@
 \pkg{containerit} uses \CRANpkg{stevedore} to streamline the user interaction and improve the created `Dockerfile`s, e.g., by running a container for the desired base image to extract the already available R packages.
-**\pkg{dockr}** ([https://github.com/smaakage85/dockr](https://github.com/smaakage85/dockr)) is a very similar package attempting to mirror a given R session, including local non-CRAN packages\citep{kjeldgaard_dockr_2019}.
+**\pkg{dockr}** ([https://github.com/smaakage85/dockr](https://github.com/smaakage85/dockr)) is a similar package.
+It is often desirable to organize R code in the R package structure.
+\pkg{dockr} facilitates the creation of a Docker container image for your R
+package, in which the package itself is available.
+The container image mirrors your current R session and includes all of the R dependencies needed to run your R package - also local non-CRAN packages\citep{kjeldgaard_dockr_2019}.
 Users can manually add statements for non-R dependencies to the `Dockerfile`.

### 2020-01-27 manuscript.Rmd
DE second author

--- 2020-01-23 manuscript.Rmd
@@
     email: daniel.nuest@uni-muenster.de
+  - name: Dirk Eddelbuettel
+    # https://github.com/eddelbuettel
+    affiliation: University of Illinois at Urbana-Champaign
+    address:
+      - Department of Statistics
+      - Illini Hall, 725 S Wright St
+      - Champaign, IL 61820, USA
+      - \orcid{0000-0001-6419-907X}
+    email: dirkd@eddelbuettel.com
   - name: Dom Bennett
@@
     email: daroczig@rapporter.net
-  - name: Dirk Eddelbuettel
-    # https://github.com/eddelbuettel
-    affiliation: University of Illinois at Urbana-Champaign
-    address:
-      - Department of Statistics
-      - Illini Hall, 725 S Wright St
-      - Champaign, IL 61820, USA
-      - \orcid{0000-0001-6419-907X}
-    email: dirkd@eddelbuettel.com
   - name: Mark Edmondson
@@
 
-The ordering of authors following DN as first author is alphabetical.
+The ordering of authors following DN and DE is alphabetical.
 DN conceived the article idea, \href{https://github.com/nuest/rockerverse-paper/issues/3}{initialised the formation of the writing team}, wrote sections not mentioned below, and revised all sections.
@@
 GD contributed the CARD.com use case.
-DE wrote the introduction and the section about Containerization and Rocker.
+DE wrote the introduction and the section about Containerization and Rocker and reviewed all sections.
 RC \& EH contributed to the section on interfaces for Docker in R.

### 2020-01-27 manuscript.Rmd
finish processing section, closes #39

@@
 
-- `RSelenium`
 - [`googleComputeEngineR`](https://cloudyr.github.io/googleComputeEngineR/) (function `gce_vm_template()`)
@@
 The portability of containerized environments becomes particularly useful for improving expensive processing of data or shipping complex processing pipelines.
-
-First, it is possible to **offload complex processing to a server**.
-...
-
-Second, when processes can be **run in parallel** for speeding up R code.
-
-For example, [`googleComputeEngineR`](https://CRAN.R-project.org/package=googleComputeEngineR)'s `gce_vm_cluster()` function can create clusters of 2 or more virtual machines, running multi-CPU architectures.
-Instead of running a local R script with the local CPU and RAM restrictions, the same code can be processed on all CPU threads of the cluster of machines in the cloud, all running in a Docker container with the same R environments.
-This is achieved through `googleComputeEngineR`'s integration with the R parralisation library [`library(future)` by Henrik Bengtsson]( https://CRAN.R-project.org/package=future).
-Local R computation can be thrown up to a multi-CPU and VM environment to achieve parrell computation in a few lines of R code in your local session.
-  - [some demonstrations are available here](https://cloudyr.github.io/googleComputeEngineR/articles/massive-parallel.html).
-- [Google Cloud Run](https://cloud.run) - This is a CaaS (Containers as a Service) that lets you launch a Docker container without worrying about underlying infrastructure.
-This dispenses with the developer creating a cloud server to run the Docker image on, by abstracting away those servers to a more serverless configuration.
-Cloud Run lets you run your code on top of a managed or your own Kubernetes cluster running Knative, and can accept any Docker image.
-The service takes care of network ingress, scaling machines up and down to zero, authentication and authorisation, all features which are non-trivial for a developer to create on their own.
-This can be used to scale up R code to millions of instances if they need to, with little or no changes to existing code.
-An R implementation is shown here at [cloudRunR](https://github.com/MarkEdmondson1234/cloudRunR) which uses Cloud Run to create a scalable R plumber API.
-https://code.markedmondson.me/googleCloudRunner/index.html
-
-[Google Cloud Build](https://cloud.google.com/cloud-build/) and the Google Container Registry are a continuous intergation service respectively image registry that offload building of images to the cloud, while serving the needs of commercial environments  that helps move the workload of building the Docker images to an online service, while  Cloud Build can be set up to build the Dockerfiles on each GitHub commit or release.
-This means you do not need to build the Docker images locally, which can tie up resources since Docker images can be several GBs and take a long time to compile.
-Google Cloud Build works alongside  to allow you to build private and public Docker images, which allows you to build up your own dependency graphs for downstream applications.
-  - https://code.markedmondson.me/googleCloudRunner/articles/cloudbuild.html
-
-- \CRANpkg{batchtools} \citep{Lang2017batchtools} can [schedule jobs with Docker Swarm](https://mllg.github.io/batchtools/reference/makeClusterFunctionsDocker.html)
-- scalable deployments, e.g., start with numerous Shiny talks mentioning Rocker at useR!2017
-- [dynmethods](https://github.com/dynverse/dynmethods) [@rcannood]: In order to evaluate ±50 computational methods which all used different environments (R, Python, C++, ...), we wrapped each of them in a docker container and can execute these methods from R.
-Again, all of these containers are being built on travis-ci, and will only be pushed to Docker&nbsp;Hub if the integration test succeeds.
-- \CRANpkg{drake}, https://docs.ropensci.org/drake/index.html?q=docker#with-docker
-
-In the RStudio Server Pro 1.2 release in 2019, RStudio added new functionality called [Launcher](https://solutions.rstudio.com/launcher/overview/).
-It gives users the ability to spawn R sessions and background jobs in a scalable way on external clusters, e.g., [Kubernetes based on Docker images](https://support.rstudio.com/hc/en-us/articles/360019253393-Using-Docker-images-with-RStudio-Server-Pro-Launcher-and-Kubernetes) or [Slurm](https://slurm.schedmd.com/) clusters, and optionally, with Singularity containers.
-A key benefit of Launcher is the ability for R and Python users to interactively work in RStudio without learning about Docker at all - while still leveraging containers and Kubernetes.
-Users can submit batch jobs to Kubernetes clusters without writing specific deployment scripts.
+First, it is possible to **offload complex processing to a server** or clouds, and also parallel executing of processes for speeding up or serving of many users.
+**\CRANpkg{batchtools}** provides a parallel implementation of Map for various schedulers \citep{Lang2017batchtools}.
+The package can [schedule jobs with Docker Swarm](https://mllg.github.io/batchtools/reference/makeClusterFunctionsDocker.html)
+**\CRANpkg{googleComputeEngineR}** has the function `gce_vm_cluster()` to create clusters of 2 or more virtual machines, running multi-CPU architectures.
+Instead of running a local R script with the local CPU and RAM restrictions, the same code can be processed on all CPU threads of the cluster of machines in the cloud, all running a Docker container with the same R environments.
+\CRANpkg{googleComputeEngineR} integrates with the R parralisation package \CRANpkg{future} to enable this with only a few lines or R code\footnote{\href{https://cloudyr.github.io/googleComputeEngineR/articles/massive-parallel.html}{https://cloudyr.github.io/googleComputeEngineR/articles/massive-parallel.html}}.
+[Google Cloud Run](https://cloud.run) is a CaaS (Containers as a Service) platform.
+Users can launch containers using any Docker image without worrying about underlying infrastructure in a so called serverless configuration.
+The service takes care of network ingress, scaling machines up and down, authentication and authorisation---all features which are non-trivial for a developer to create on their own.
+This can be used to scale up R code to millions of instances if need be with little or no changes to existing code, as demonstrated by the proof of concept `cloudRunR` ([https://github.com/MarkEdmondson1234/cloudRunR](https://github.com/MarkEdmondson1234/cloudRunR)), which uses Cloud Run to create a scalable R-based API using \CRANpkg{plumber}.
+[Google Cloud Build](https://cloud.google.com/cloud-build/) and the Google Container Registry are a continuous integration service respectively image registry that offload building of images to the cloud, while serving the needs of commercial environments such as private Docker images or image stacks.
+As Googl Cloud Build itself can run any container, the package \pkg{googleCloudRunner} demonstrates how R can be used as the control language for one time or batch processing jobs and scheduling of jobs\footnote{\href{https://code.markedmondson.me/googleCloudRunner/articles/cloudbuild.html}{https://code.markedmondson.me/googleCloudRunner/articles/cloudbuild.html}}.
+\label{drake} **\CRANpkg{drake}** is a workflow manager for data science projects \citep{landau_drake_2019}.
+It features implicit parallel computing and automated detection of the parts of the work that actually needs to be reexecuted.
+\CRANpkg{drake} has been demonstrated to run inside containers for high reproducibility\footnote{See for example \href{https://github.com/joelnitta/pleurosoriopsis}{https://github.com/joelnitta/pleurosoriopsis} or \href{https://gitlab.com/ecohealthalliance/drake-gitlab-docker-example}{https://gitlab.com/ecohealthalliance/drake-gitlab-docker-example}, the latter even running in a CI, see~\nameref{cicd}.}, but also how to use \CRANpkg{future} package's function `makeClusterPSOCK()` to send parts of the workflow to a Docker image for execution\footnote{\href{https://docs.ropensci.org/drake/index.html?q=docker\#with-docker}{https://docs.ropensci.org/drake/index.html?q=docker\#with-docker}} \citep[see package's function documentation;~][]{future_2020}.
+In the latter case, the container control code must be written by the user and the \CRANpkg{future} package ensures the host and worker can connect for communicating over socket connections.
+**RStudio Server Pro** includes a functionality called [Launcher](https://solutions.rstudio.com/launcher/overview/) (since version&nbsp;1.2, released in 2019).
+It gives users the ability to spawn R sessions and background/batch jobs in a scalable way on external clusters, e.g., [Kubernetes based on Docker images](https://support.rstudio.com/hc/en-us/articles/360019253393-Using-Docker-images-with-RStudio-Server-Pro-Launcher-and-Kubernetes) or [Slurm](https://slurm.schedmd.com/) clusters, and optionally, with Singularity containers.
+A benefit of the proprietary Launcher software is the ability for R and Python users to leverage containerisation's advantages in RStudio without writing specific deployment scripts or learning about Docker or managing clusters at all.
 
 \label{pipelines}
-Third, containers are perfectly suited for **packaging and executing software pipelines** and required data.
+Second, containers are perfectly suited for **packaging and executing software pipelines** and required data.
 Containers allow building complex processing pipelines that are independent from the host programming language.
@@
 A typial compendium's `Dockerfile` will pull a rocker image fixed to a specific version of R, and install R packages from the MRAN repository to ensure the package versions are tied to a specific date, rather than the most recent version.
+A more extreme case is the _dynverse_ project ([https://dynverse.org/](https://dynverse.org/)), which packages over 50 computational methods with different environments (R, Python, C++, etc.) in Docker images, which can be executed from R.
 Future researchers can download the compendium from the repository and run the included `Dockerfile` to build a new image that recreates the computational environment used to produce the original research results.
@@
 This approach clearly seperates test environment, test code, and current working copy of the code.
+\label{rselenium} Another use case where a container helps to standardise tests across operating systems is detailed the vignettes of the package \CRANpkg{RSelenium} \citep{rselenium_2019}.
+The package recommends Docker for running the [Selenium](https://selenium.dev/) Server application needed to execute test suites on browser-based user interfaces and webpages, but requires users to manually manage the Docker containers.
 
@@
 Currently containerisation happens more in the background at the system level.
+Furthermore, a maturing of the Rockerverse packages for managing containers may lead to their adoption where currently manual coding is required, e.g.
+in the case of \CRANpkg{RSelentium} or \CRANpkg{drake} (see Sections&nbsp;\nameref{rselenium} and \nameref{drake} respectively).
 

### 2020-01-27 manuscript.Rmd
move Whole Tale description into Binder use case, closes #38

@@
 
-**Whole&nbsp;Tale** (https://wholetale.org) is a National Science Foundation (NSF) funded project developing a scalable, open-source, multi-user platform for reproducible research \citep{brinckman2019, chard2019a}.
-A central goal of the platform is to enable researchers to easily create and publish executable research objects\footnote{In Whole~Tale a _tale_ is a research object that contains metadata, data (by copy or reference), code, narrative, documentation, provenance, and information about the computational environment to support computational reproducibility} associated with published research \citep{chard2019b}.
-Using Whole&nbsp;Tale, researchers can create and publish Rocker-based reproducible research objects to a growing number of repositories including DataONE member nodes, Zenodo and soon Dataverse.
-Additionally, Whole&nbsp;Tale supports automatic data citation and is working on capabilities for image preservation and provenance capture to improve the transparency of published computational research artifacts \citep{mecum2018, mcphillips2019}.
-For R users, Whole&nbsp;Tale attempts to combine the strengths of the Rocker Project's curated Docker images with the Jupyter Project's `repo2docker`\footnote{See \href{https://repo2docker.readthedocs.io/}{https://repo2docker.readthedocs.io/}} component to simplify the customization of R-based environments for researchers with limited experience with either Docker or Git.
-Multiple options have been discussed to allow users to change the base image used in `repo2docker` from the default Ubuntu LTS (long-term support) required to support the Rocker Project images. 
-Whole&nbsp;Tale implemented a custom `RockerBuildPack`\footnote{See \href{https://github.com/whole-tale/repo2docker\_wholetale}{https://github.com/whole-tale/repo2docker\_wholetale}} to support customization of the `rocker/geospatial` image through `repo2docker` composability\footnote{Composability refers to the ability to combine multiple package managers -- such as R, `pip`, and `conda`}.
-This works in part because Rocker images are based on a Debian distribution, so the instructions created by `repo2docker` for Ubuntu work because of compatible shell and package manager.
-
 # Use cases and applications
@@
 
-**\pkg{holepunch}** ([https://github.com/karthik/holepunch](https://github.com/karthik/holepunch)) is an R package that was designed to make sharing work environments accessible to novice R users based on **Binder**.
 The [Binder project](https://mybinder.readthedocs.io/en/latest/), maintained by the team behind Jupyter, makes it possible for users to create and share their computing environments with others \citep{jupyter_binder_2018}.
@@
 To set up Binder for a project, a user typically starts at an instance of a BinderHub and passes the location of a repository with a workspace, e.g., a hosted Git repository, or a data repository like Zenodo.
-Binder's core internal tool is [`repo2docker`](https://repo2docker.readthedocs.io/en/latest/config_files.html).
-It deterministically builds a Docker image by parsing the contents of a repository, e.g., project dependency configurations or simple configuration files.
+Binder's core internal tool is `repo2docker`.
+It deterministically builds a Docker image by parsing the contents of a repository, e.g., project dependency configurations or simple configuration files\footnote{See supported file types at \href{https://repo2docker.readthedocs.io/en/latest/config_files.html}{https://repo2docker.readthedocs.io/en/latest/config_files.html}.}.
 In the most powerful case, `repo2docker` builds a given `Dockerfile`.
 While this approach works well for most run of the mill Python projects, it is not so seamless for R projects.
+This is partly because `repo2docker` does not support arbitrary base images due to the complex auto-generation of the `Dockerfile` instructions.
+Two approaches make using Binder easier. for R users.
+First, **\pkg{holepunch}** ([https://github.com/karthik/holepunch](https://github.com/karthik/holepunch)) is an R package that was designed to make sharing work environments accessible to novice R users based on **Binder**.
 For any R projects that use the Tidyverse suite \citep{wickham_welcome_2019}, the time and resources required to build all dependencies from source can often time out before completion, making it frustrating for the average R user.
@@
 The local image usage reduces the number of seperately managed environments and thereby reduces work and increases portability and reproducibility.
+Second, the **Whole&nbsp;Tale** project (https://wholetale.org) combines the strengths of the Rocker Project's curated Docker images with `repo2docker`.
+Whole&nbsp;Tale is a National Science Foundation (NSF) funded project developing a scalable, open-source, multi-user platform for reproducible research \citep{brinckman2019, chard2019a}.
+A central goal of the platform is to enable researchers to easily create and publish executable research objects\footnote{In Whole~Tale a _tale_ is a research object that contains metadata, data (by copy or reference), code, narrative, documentation, provenance, and information about the computational environment to support computational reproducibility} associated with published research \citep{chard2019b}.
+Using Whole&nbsp;Tale, researchers can create and publish Rocker-based reproducible research objects to a growing number of repositories including DataONE member nodes, Zenodo and soon Dataverse.
+Additionally, Whole&nbsp;Tale supports automatic data citation and is working on capabilities for image preservation and provenance capture to improve the transparency of published computational research artifacts \citep{mecum2018, mcphillips2019}.
+For R users, Whole&nbsp;Tale extends the Jupyter Project's `repo2docker` component to simplify the customization of R-based environments for researchers with limited experience with either Docker or Git.
+Multiple options have been discussed to allow users to change the base image used in `repo2docker` from the default Ubuntu LTS (long-term support) required to support the Rocker Project images. 
+Whole&nbsp;Tale implemented a custom `RockerBuildPack`\footnote{See \href{https://github.com/whole-tale/repo2docker\_wholetale}{https://github.com/whole-tale/repo2docker\_wholetale}} to support customization of the `rocker/geospatial` image through `repo2docker` composability\footnote{Composability refers to the ability to combine multiple package managers -- such as R, `pip`, and `conda`}.
+This works in part because Rocker images are based on a Debian distribution, so the instructions created by `repo2docker` for Ubuntu work because of compatible shell and package manager.
 

### 2020-01-27 manuscript.Rmd
wrap up deployment/CD section, closes #36

@@
 \CRANpkg{stevedore}'s own interface is automatically generated based on the OpenAPI specification of the Docker daemon, but still similar to the Docker CLI.
-The interface is similar to R6 objects, in that a `stevedore_object` has a number of functions attached to it that can be called, and multiple specific versions of the Docker API can be supported thanks to the automatic generation\footnote{See \href{https://github.com/richfitz/stevedore/tree/master/inst/spec}{https://github.com/richfitz/stevedore/tree/master/inst/spec} and \href{https://github.com/richfitz/stevedore/blob/master/development.md}{https://github.com/richfitz/stevedore/blob/master/development.md}.}.
+The interface is similar to R6 objects, in that a `stevedore_object` has a number of functions attached to it that can be called, and multiple specific versions of the Docker API can be supported thanks to the automatic generation\footnote{See \href{https://github.com/richfitz/stevedore/blob/master/development.md}{https://github.com/richfitz/stevedore/blob/master/development.md}.}.
 
-**\CRANpkg{AzureContainers}** is an interface to a number of container-related services in Microsoft's [Azure Cloud](https://azure.microsoft.com/), see Section&nbsp;\nameref{azurecontainers}.
+**\CRANpkg{AzureContainers}** is an interface to a number of container-related services in Microsoft's [Azure Cloud](https://azure.microsoft.com/) \citep{AzureContainers_2019}.
 While it is mainly intended for working with Azure, as a convenience feature it includes lightweight, cross-platform shells to Docker and Kubernetes (tools `kubectl` and `helm`).
@@
 
-## Continuous integration, continuous delivery, and deployment
-\label{cicd}
+## Deployment and continuous delivery
 \label{deployment}
@@
 More and more continuous integration (CI) and continuous delivery (CD) services also use containers, opening up new options for use.
-The controlled nature of containers, i.e., the possibility to define the software environment very well, even on remote machines, is crucial here, for example to match test or build environments with production environments.
+The controlled nature of containers, i.e., the possibility to abstract internal software environment from a minimal dependency outside of the container is crucial, for example to match test or build environments with production environments or transfer runnable entities to as-a-service infrastructures.
 
-As a first example, \CRANpkg{Shiny} is the most popular package for creating interactive online dashboards with R and it enables users with very diverse backgrounds to create stable and user friendly web applications.
-The ShinyProxy ([https://www.shinyproxy.io/](https://www.shinyproxy.io/)) is an open source tool to deploy Shiny apps in an enterprise context.
+First, different packages use containers for the **deployment of R and \CRANpkg{Shiny} apps**.
+\CRANpkg{Shiny} is a popular package for creating interactive online dashboards with R and it enables users with very diverse backgrounds to create stable and user friendly web applications.
+For example, _ShinyProxy_ ([https://www.shinyproxy.io/](https://www.shinyproxy.io/)) is an open source tool to deploy Shiny apps in an enterprise context.
 They feature single sign-on, but also in scientific use cases \citep[e.g., ][]{savini_epiexplorer_2019,glouzon_structurexplor_2017}.
@@
 The started containers can run on a regular Docker host but also in clusters.
+Another example is the package \CRANpkg{golem} ([https://github.com/ThinkR-open/golem](https://github.com/ThinkR-open/golem)), which makes an heavy use of `dockerfiler` when it comes to creating the `Dockerfile` for building production-grade Shiny applications and deploying them.
+\CRANpkg{googleComputeEngineR} ([https://cloudyr.github.io/googleComputeEngineR/](https://cloudyr.github.io/googleComputeEngineR/)) enables quick deployments of key R services, such as RStudio and Shiny, onto cloud virtual machines (VMs) with Google Cloud Compute Engine \citep{googleComputeEngineR_2019}.
+The package utilises `Dockerfile`s to move the labour of setting up those services from the user to a premade Docker image, which is configured and run in the cloud VM.
+For example, by specifying the template `template="rstudio"` in functions `gce_vm_template()/gce_vm()` an up to date RStudio Service image is launched for development work, while specifying `template="rstudio-gpu"` will launch an RStudio Server image with a GPU attached, etc.
 
@@
 
-- [dynwrap](https://github.com/dynverse/dynwrap_containers/blob/master/.travis.yml) [@rcannood]
-  - For this project, we use travis-ci to build rocker-derived containers, test them, and only push them to Docker&nbsp;Hub (from travis-ci.org) if the integration tests succeed.
-
-For example, the **[`golem`](https://github.com/ThinkR-open/golem)** package makes an heavy use of `dockerfiler` when it comes to creating the `Dockerfile` for building production-grade Shiny applications and deploying them.
-
-- [`googleComputeEngineR`](https://cloudyr.github.io/googleComputeEngineR/) (function `gce_vm_template()`)
-  - To enable quick deployments of key R services such as RStudio and Shiny onto cloud virtual machines (VMs), this package utilises Dockerfiles to move the labour of setting up those services from the user to a premade Docker image.
-For example, by specifying the template `template="rstudio"` in `gce_vm_template()/gce_vm()` an up to date RStudio Service image is launched.
-Specifying `template="rstudio-gpu"` will launch an RStudio Server image with a GPU attached, etc.
-- [`analogsea`](https://github.com/sckott/analogsea) (digital ocean R client)
-- [`plumber`](https://www.rplumber.io/docs/hosting.html#docker)
-- [production deployment of neural networks](https://github.com/tmobile/r-tensorflow-api) [@jnolis, @nolistic]
-- Kubernetes - A popular platform for managing Docker containers is [Kubernetes](https://kubernetes.io/), which is used for a wide variety of containerized applications.
-It may be your organisation has a Kubernetes cluster already for other applications.
+Third, containers can greatly facilitate the **deployment to existing infrastructures**.
+_Kubernetes_ ([https://kubernetes.io/](https://kubernetes.io/)) is a container-orchestration system for managing container-based application deployment and scaling.
+A _cluster_ of containers, orchestrated as a single deployment, e.g., with Kubernetes, can mitigate limitations on request volumes or a container occupied with computationally intensive task.
+A cluster features load-balancing, autoscaling of containers across numerous servers (in the cloud or on premise), and restarting failed ones.
+It may be your organisation has a Kubernetes cluster already for other applications, or you may use one of the different provides for managed Kubernetes clusters.
 Docker containers are used within Kubernetes clusters to hold native code, for which Kubernetes creates a framework around network connections and scaling of resources up and down.
-An introduction on using [Kubernetes with R is published at this blog post](https://code.markedmondson.me/r-on-kubernetes-serverless-shiny-r-apis-and-scheduled-scripts/).
-
-\label{azurecontainers}
-**\CRANpkg{AzureContainers}** is an umbrella package providing interfaces three commercial services of Microsoft's Azure Cloud, namely [Container Instances](https://azure.microsoft.com/en-us/services/container-instances/) for running individual containers, [Container Registry](https://azure.microsoft.com/en-us/services/container-registry/) for private image distribution, and [Kubernetes Service](https://azure.microsoft.com/en-us/services/kubernetes-service/) for orchestrated deployments.
+R applications, big parallel tasks, or scheduled batch jobs can be deployed in a scalable way using Kubernetes, and the deployment can even be triggered by changes to code repositories (i.e., CD), see blog post "R on Kubernetes" \citep{edmondson_r_2018}.
+The package \pkg{googleKubernetesR} ([https://github.com/RhysJackson/googleKubernetesR](https://github.com/RhysJackson/googleKubernetesR)) is a proof of concept for wrapping the Google Kubernetes Engine API, Google's hosted Kubernetes solution, in an easy to use R package.
+The package \CRANpkg{analogsea} provides a solution to programmatically create and destroy cloud VMs on the [Digital Ocean](https://www.digitalocean.com/) platform \citep{analogsea_2019}.
+It also includes R wrapper functions to install Docker in such a VM, manage images, and control containers straight from R functions.
+These functions are translated to Docker CLI commands and transferred transparently to the respective remove machine using SSH.
+\CRANpkg{AzureContainers} is an umbrella package providing interfaces three commercial services of Microsoft's Azure Cloud, namely [Container Instances](https://azure.microsoft.com/en-us/services/container-instances/) for running individual containers, [Container Registry](https://azure.microsoft.com/en-us/services/container-registry/) for private image distribution, and [Kubernetes Service](https://azure.microsoft.com/en-us/services/kubernetes-service/) for orchestrated deployments.
 While a package like \CRANpkg{plumber} provides the infrastructure for turning an R workflow a service, for production purposes it is usually necessary to take into account scalability, reliability and ease of management.
-A _cluster_ of containers, orchestrated as a single deployment, e.g., with [Kubernetes](https://kubernetes.io), can mitigate limitations on request volumes or a container occupied with computationally intensive task.
-A cluster features load-balancing, autoscaling of containers across numerous servers (in the cloud or on premise), and restarting failed ones.
-AzureContainers provides an R-based interface to these features for Azure.
-On the client side, it provides simple wrappers to Docker and to Kubernetes' management tools `kubectl` and `helm`.
-The package simplifies complex infrastructure management to a number of R function calls, given an Azure account with sufficient credit\footnote{See \emph{"Deploying a prediction service with Plumber"} vignette for details:  \href{https://cran.r-project.org/web/packages/AzureContainers/vignettes/vig01_plumber_deploy.html}{https://cran.r-project.org/web/packages/AzureContainers/vignettes/vig01\_plumber\_deploy.html}.}.
+AzureContainers provides an R-based interface to these features and thereby simplifies complex infrastructure management to a number of R function calls, given an Azure account with sufficient credit\footnote{See \emph{"Deploying a prediction service with Plumber"} vignette for details:  \href{https://cran.r-project.org/web/packages/AzureContainers/vignettes/vig01_plumber_deploy.html}{https://cran.r-project.org/web/packages/AzureContainers/vignettes/vig01\_plumber\_deploy.html}.}.
 
 The prevelance of Docker in industry naturally leads to usage of containers with R in such settings as well, as customers already manage platforms in Docker containers.
-These products often entail a high amount of open source software in combination with proprietary layers addign the relevant commercializable features.
+These products often entail a high amount of open source software in combination with proprietary layers adding the relevant commercializable features.
 One such example is RStudio's data science platform [RStudio Team](https://rstudio.com/products/team/).
 It allows teams of data scientists and their respective IT/DevOps groups to develop and deploy code in R and Python around the RStudio Open Source Server inside of Docker images, without requiring users to learn new tools or directly interact with containers.
-The best practices for [running RStudio with Docker containers](https://support.rstudio.com/hc/en-us/articles/360021594513-Running-RStudio-with-Docker-containers) as well as [Docker images](https://github.com/rstudio/rstudio-docker-products) for RStudio's commerical products are openly available.
+The best practices for [running RStudio with Docker containers](https://support.rstudio.com/hc/en-us/articles/360021594513-Running-RStudio-with-Docker-containers) as well as [Docker images](https://github.com/rstudio/rstudio-docker-products) for RStudio's commerical products are publicly available.
 
@@
 Binder's core internal tool is `repo2docker`.
-It deterministically builds a Docker image by parsing the contents of a repository, e.g., project dependency configurations or simple configuration files\footnote{See supported file types at \href{https://repo2docker.readthedocs.io/en/latest/config_files.html}{https://repo2docker.readthedocs.io/en/latest/config_files.html}.}.
+It deterministically builds a Docker image by parsing the contents of a repository, e.g., project dependency configurations or simple configuration files\footnote{See supported file types at \href{https://repo2docker.readthedocs.io/en/latest/config\_files.html}{https://repo2docker.readthedocs.io/en/latest/config\_files.html}.}.
 In the most powerful case, `repo2docker` builds a given `Dockerfile`.
@@
 A more extreme case is the _dynverse_ project ([https://dynverse.org/](https://dynverse.org/)), which packages over 50 computational methods with different environments (R, Python, C++, etc.) in Docker images, which can be executed from R.
+_dynverse_ uses a CI (cf.&nbsp;\nameref{cicd}) to build Rocker-derived containers, test them, and push them to Docker&nbsp;Hub if the tests succeed.
+
 Future researchers can download the compendium from the repository and run the included `Dockerfile` to build a new image that recreates the computational environment used to produce the original research results.
@@
 Further safeguarding practices are currently under development or not part of common practice yet, such as preservation of images \citep{emsley_framework_2018} and storing them alongside `Dockerfile`s \citep[cf.][]{nust_opening_2017}, or pinning of system libraries beyond the tagged base images, which may be seen as stable or dynamic depending on the applied time scale \citep[see discussion on `debian:testing` base image in][]{RJ-2017-065}.
-
 A recommendation of the recent National Academies' report on _Reproducibility and Replicability in Science_ is that journals _"consider ways to ensure computational reproducibility for publications that make claims based on computations"_ \citep{NASEM2019}.
@@
 in the case of \CRANpkg{RSelentium} or \CRANpkg{drake} (see Sections&nbsp;\nameref{rselenium} and \nameref{drake} respectively).
+In specific cases, e.g., for \CRANpkg{analogsea}, the interaction with the Docker daemon may remain to specific to reuse first order packages to control Docker.
 
@@
 DE wrote the introduction and the section about Containerization and Rocker and reviewed all sections.
-RC \& EH contributed to the section on interfaces for Docker in R.
+RC \& EH contributed to the section on interfaces for Docker in R (_dynverse_ and `dynwrap`).
 DC contributed content on Gigantum.
@@
 BM wrote the section on research compendia and made the project Binder-ready.
-HN \& JN co-wrote the section on enterprise production environments (\pkg{googleComputeEngineR}, Google Cloud Run, Kubernetes).
+HN \& JN co-wrote the section on the T-Mobile use case.
 KR wrote the section about \pkg{holepunch}.

### 2020-01-27 manuscript.Rmd
minor fixes

@@
 The images extend `rocker/r-base` and additionally entailed an SSH client and a workaround for being able to mount SSH keys from the host, Pandoc, the Amazon Web Services (AWS) SDK, and Java, which is required by the AWS SDK.
-The AWS SDK allowed to run R consumers reading from real-time data processing streams of [AWS Kinesis](https://aws.amazon.com/kinesis/)\footnote{See useR!2017 talk _Stream processing with R in AWS_ at \href{https://static.sched.com/hosted_files/user2017/2f/AWR%20Kinesis%20at%20useR%202017.pdf}{https://static.sched.com/hosted_files/user2017/2f/AWR Kinesis at use 02017.pdf}.}.
+The AWS SDK allowed to run R consumers reading from real-time data processing streams of [AWS Kinesis](https://aws.amazon.com/kinesis/) \footnote{See useR!2017 talk \href{https://static.sched.com/hosted\_files/user2017/2f/AWR Kinesis at useR 2017.pdf}{"Stream processing with R in AWS"}.}.
 The applications were deployed on Amazon Elastic Container Service ([ECS](https://aws.amazon.com/ecs/)).
@@
 
-The [Binder project](https://mybinder.readthedocs.io/en/latest/), maintained by the team behind Jupyter, makes it possible for users to create and share their computing environments with others \citep{jupyter_binder_2018}.
+The [**Binder**](https://mybinder.readthedocs.io/en/latest/) project, maintained by the team behind Jupyter, makes it possible for users to **create and share computing environments** with others \citep{jupyter_binder_2018}.
 A _BinderHub_ allows anyone with access to a web browser and an internet connection to launch a temporary instance of these custom environments and execute any workflows contained within.
@@
 Two approaches make using Binder easier. for R users.
-First, **\pkg{holepunch}** ([https://github.com/karthik/holepunch](https://github.com/karthik/holepunch)) is an R package that was designed to make sharing work environments accessible to novice R users based on **Binder**.
+First, **\pkg{holepunch}** ([https://github.com/karthik/holepunch](https://github.com/karthik/holepunch)) is an R package that was designed to make sharing work environments accessible to novice R users based on Binder.
 For any R projects that use the Tidyverse suite \citep{wickham_welcome_2019}, the time and resources required to build all dependencies from source can often time out before completion, making it frustrating for the average R user.
@@
 The service takes care of network ingress, scaling machines up and down, authentication and authorisation---all features which are non-trivial for a developer to create on their own.
-This can be used to scale up R code to millions of instances if need be with little or no changes to existing code, as demonstrated by the proof of concept `cloudRunR` ([https://github.com/MarkEdmondson1234/cloudRunR](https://github.com/MarkEdmondson1234/cloudRunR)), which uses Cloud Run to create a scalable R-based API using \CRANpkg{plumber}.
+This can be used to scale up R code to millions of instances if need be with little or no changes to existing code, as demonstrated by the proof of concept `cloudRunR`\footnote{\href{https://github.com/MarkEdmondson1234/cloudRunR}{https://github.com/MarkEdmondson1234/cloudRunR}}, which uses Cloud Run to create a scalable R-based API using \CRANpkg{plumber}.
 [Google Cloud Build](https://cloud.google.com/cloud-build/) and the Google Container Registry are a continuous integration service respectively image registry that offload building of images to the cloud, while serving the needs of commercial environments such as private Docker images or image stacks.
@@
 For example, GDAL \citep{gdal_2019} is a crucial library in the geospatial domain.
-GDAL is a system dependen^cy allowing R packages such as \CRANpkg{sf}, which provides the core data model for geospatial vector data, or \CRANpkg{rgdal}, to accomodate users to be able to read and write hundreds of different spatial raster and vector formats.
+GDAL is a system dependency allowing R packages such as \CRANpkg{sf}, which provides the core data model for geospatial vector data, or \CRANpkg{rgdal}, to accomodate users to be able to read and write hundreds of different spatial raster and vector formats.
 \CRANpkg{sf} and \CRANpkg{rgdal} have hundreds of indirect reverse imports and dependencies and therefore the maintainers spend a lot of efforts not to break these.

### 2020-01-27 manuscript.tex
update tex file

@@
 with R}
-\author{by Daniel Nüst, Dom Bennett, Robrecht Cannoodt, Dav Clark, Gergely Daroczi, Dirk Eddelbuettel, Mark Edmondson, Colin Fay, Ellis Hughes, Sean Lopp, Ben Marwick, Heather Nolis, Jacqueline Nolis, Hong Ooi, Karthik Ram, Noam Ross, Lori Shepherd, Nitesh Turaga, Craig Willis, Nan Xiao, Charlotte Van Petegem}
+\author{by Daniel Nüst, Dirk Eddelbuettel, Dom Bennett, Robrecht Cannoodt, Dav Clark, Gergely Daroczi, Mark Edmondson, Colin Fay, Ellis Hughes, Sean Lopp, Ben Marwick, Heather Nolis, Jacqueline Nolis, Hong Ooi, Karthik Ram, Noam Ross, Lori Shepherd, Nitesh Turaga, Craig Willis, Nan Xiao, Charlotte Van Petegem}
 
@@
 
-\textbf{Whole~Tale} (\url{https://wholetale.org}) is a National Science
-Foundation (NSF) funded project developing a scalable, open-source,
-multi-user platform for reproducible research
-\citep{brinckman2019, chard2019a}. A central goal of the platform is to
-enable researchers to easily create and publish executable research
-objects\textbackslash{}footnote\{In Whole\textasciitilde{}Tale a
-\emph{tale} is a research object that contains metadata, data (by copy
-or reference), code, narrative, documentation, provenance, and
-information about the computational environment to support computational
-reproducibility\} associated with published research \citep{chard2019b}.
-Using Whole~Tale, researchers can create and publish Rocker-based
-reproducible research objects to a growing number of repositories
-including DataONE member nodes, Zenodo and soon Dataverse. Additionally,
-Whole~Tale supports automatic data citation and is working on
-capabilities for image preservation and provenance capture to improve
-the transparency of published computational research artifacts
-\citep{mecum2018, mcphillips2019}. For R users, Whole~Tale attempts to
-combine the strengths of the Rocker Project's curated Docker images with
-the Jupyter Project's
-\texttt{repo2docker}\footnote{See \href{https://repo2docker.readthedocs.io/}{https://repo2docker.readthedocs.io/}}
-component to simplify the customization of R-based environments for
-researchers with limited experience with either Docker or Git. Multiple
-options have been discussed to allow users to change the base image used
-in \texttt{repo2docker} from the default Ubuntu LTS (long-term support)
-required to support the Rocker Project images. Whole~Tale implemented a
-custom
-\texttt{RockerBuildPack}\footnote{See \href{https://github.com/whole-tale/repo2docker\_wholetale}{https://github.com/whole-tale/repo2docker\_wholetale}}
-to support customization of the \texttt{rocker/geospatial} image through
-\texttt{repo2docker}
-composability\footnote{Composability refers to the ability to combine multiple package managers -- such as R, `pip`, and `conda`}.
-This works in part because Rocker images are based on a Debian
-distribution, so the instructions created by \texttt{repo2docker} for
-Ubuntu work because of compatible shell and package manager.
-
 \hypertarget{use-cases-and-applications}{%
@@
 supported thanks to the automatic
-generation\footnote{See \href{https://github.com/richfitz/stevedore/tree/master/inst/spec}{https://github.com/richfitz/stevedore/tree/master/inst/spec} and \href{https://github.com/richfitz/stevedore/blob/master/development.md}{https://github.com/richfitz/stevedore/blob/master/development.md}.}.
+generation\footnote{See \href{https://github.com/richfitz/stevedore/blob/master/development.md}{https://github.com/richfitz/stevedore/blob/master/development.md}.}.
 
@@
 container-related services in Microsoft's
-\href{https://azure.microsoft.com/}{Azure Cloud}, see
-Section~\nameref{azurecontainers}. While it is mainly intended for
-working with Azure, as a convenience feature it includes lightweight,
+\href{https://azure.microsoft.com/}{Azure Cloud}
+\citep{AzureContainers_2019}. While it is mainly intended for working
+with Azure, as a convenience feature it includes lightweight,
 cross-platform shells to Docker and Kubernetes (tools \texttt{kubectl}
@@
 real-time data processing streams of
-\href{https://aws.amazon.com/kinesis/}{AWS
-Kinesis}\textbackslash{}footnote\{See useR!2017 talk \emph{Stream
-processing with R in AWS} at
-\textbackslash{}href\{\url{https://static.sched.com/hosted_files/user2017/2f/AWR\%20Kinesis\%20at\%20useR\%202017.pdf}\}\{\url{https://static.sched.com/hosted_files/user2017/2f/AWR}
-Kinesis at use 02017.pdf\}.\}. The applications were deployed on Amazon
-Elastic Container Service (\href{https://aws.amazon.com/ecs/}{ECS}). The
-main learnings from using R in Docker was the imporance of not only
-pinning the R package versions via MRAN, but also moving away from
-Debian testing to a distribution with long-term support. For the use
-case at hand, this switch served the priority to have more control over
-upstream updates, and to minimize the risk of breaking the automated
-builds of the Docker images and production jobs.
+\href{https://aws.amazon.com/kinesis/}{AWS Kinesis}
+\footnote{See useR!2017 talk \href{https://static.sched.com/hosted\_files/user2017/2f/AWR Kinesis at useR 2017.pdf}{"Stream processing with R in AWS"}.}.
+The applications were deployed on Amazon Elastic Container Service
+(\href{https://aws.amazon.com/ecs/}{ECS}). The main learnings from using
+R in Docker was the imporance of not only pinning the R package versions
+via MRAN, but also moving away from Debian testing to a distribution
+with long-term support. For the use case at hand, this switch served the
+priority to have more control over upstream updates, and to minimize the
+risk of breaking the automated builds of the Docker images and
+production jobs.
 
@@
 
-\hypertarget{continuous-integration-continuous-delivery-and-deployment}{%
-\subsection{Continuous integration, continuous delivery, and
-deployment}\label{continuous-integration-continuous-delivery-and-deployment}}
-
-\label{cicd} \label{deployment}
+\hypertarget{deployment-and-continuous-delivery}{%
+\subsection{Deployment and continuous
+delivery}\label{deployment-and-continuous-delivery}}
+
+\label{deployment}
 
@@
 services also use containers, opening up new options for use. The
-controlled nature of containers, i.e., the possibility to define the
-software environment very well, even on remote machines, is crucial
-here, for example to match test or build environments with production
-environments.
-
-As a first example, \CRANpkg{Shiny} is the most popular package for
+controlled nature of containers, i.e., the possibility to abstract
+internal software environment from a minimal dependency outside of the
+container is crucial, for example to match test or build environments
+with production environments or transfer runnable entities to
+as-a-service infrastructures.
+
+First, different packages use containers for the \textbf{deployment of R
+and \CRANpkg{Shiny} apps}. \CRANpkg{Shiny} is a popular package for
 creating interactive online dashboards with R and it enables users with
 very diverse backgrounds to create stable and user friendly web
-applications. The ShinyProxy (\url{https://www.shinyproxy.io/}) is an
-open source tool to deploy Shiny apps in an enterprise context. They
-feature single sign-on, but also in scientific use cases
+applications. For example, \emph{ShinyProxy}
+(\url{https://www.shinyproxy.io/}) is an open source tool to deploy
+Shiny apps in an enterprise context. They feature single sign-on, but
+also in scientific use cases
 \citep[e.g., ][]{savini_epiexplorer_2019,glouzon_structurexplor_2017}.
@@
 containers can run on a regular Docker host but also in clusters.
+Another example is the package \CRANpkg{golem}
+(\url{https://github.com/ThinkR-open/golem}), which makes an heavy use
+of \texttt{dockerfiler} when it comes to creating the
+\texttt{Dockerfile} for building production-grade Shiny applications and
+deploying them. \CRANpkg{googleComputeEngineR}
+(\url{https://cloudyr.github.io/googleComputeEngineR/}) enables quick
+deployments of key R services, such as RStudio and Shiny, onto cloud
+virtual machines (VMs) with Google Cloud Compute Engine
+\citep{googleComputeEngineR_2019}. The package utilises
+\texttt{Dockerfile}s to move the labour of setting up those services
+from the user to a premade Docker image, which is configured and run in
+the cloud VM. For example, by specifying the template
+\texttt{template="rstudio"} in functions
+\texttt{gce\_vm\_template()/gce\_vm()} an up to date RStudio Service
+image is launched for development work, while specifying
+\texttt{template="rstudio-gpu"} will launch an RStudio Server image with
+a GPU attached, etc.
 
@@
 
-\begin{itemize}
-\tightlist
-\item
-  \href{https://github.com/dynverse/dynwrap_containers/blob/master/.travis.yml}{dynwrap}
-  \citep{rcannood}
-
-  \begin{itemize}
-  \tightlist
-  \item
-    For this project, we use travis-ci to build rocker-derived
-    containers, test them, and only push them to Docker~Hub (from
-    travis-ci.org) if the integration tests succeed.
-  \end{itemize}
-\end{itemize}
-
-For example, the
-\textbf{\href{https://github.com/ThinkR-open/golem}{\texttt{golem}}}
-package makes an heavy use of \texttt{dockerfiler} when it comes to
-creating the \texttt{Dockerfile} for building production-grade Shiny
-applications and deploying them.
-
-\begin{itemize}
-\tightlist
-\item
-  \texttt{RSelenium}
-\item
-  \href{https://cloudyr.github.io/googleComputeEngineR/}{\texttt{googleComputeEngineR}}
-  (function \texttt{gce\_vm\_template()})
-
-  \begin{itemize}
-  \tightlist
-  \item
-    To enable quick deployments of key R services such as RStudio and
-    Shiny onto cloud virtual machines (VMs), this package utilises
-    Dockerfiles to move the labour of setting up those services from the
-    user to a premade Docker image. For example, by specifying the
-    template \texttt{template="rstudio"} in
-    \texttt{gce\_vm\_template()/gce\_vm()} an up to date RStudio Service
-    image is launched. Specifying \texttt{template="rstudio-gpu"} will
-    launch an RStudio Server image with a GPU attached, etc.\\
-  \end{itemize}
-\item
-  \href{https://github.com/sckott/analogsea}{\texttt{analogsea}}
-  (digital ocean R client)
-\item
-  \href{https://www.rplumber.io/docs/hosting.html\#docker}{\texttt{plumber}}
-\item
-  \href{https://github.com/tmobile/r-tensorflow-api}{production
-  deployment of neural networks} \citep[\citet{nolistic}]{jnolis}
-\item
-  Kubernetes - A popular platform for managing Docker containers is
-  \href{https://kubernetes.io/}{Kubernetes}, which is used for a wide
-  variety of containerized applications. It may be your organisation has
-  a Kubernetes cluster already for other applications. Docker containers
-  are used within Kubernetes clusters to hold native code, for which
-  Kubernetes creates a framework around network connections and scaling
-  of resources up and down. An introduction on using
-  \href{https://code.markedmondson.me/r-on-kubernetes-serverless-shiny-r-apis-and-scheduled-scripts/}{Kubernetes
-  with R is published at this blog post}.
-\end{itemize}
-
-\label{azurecontainers} \textbf{\CRANpkg{AzureContainers}} is an
-umbrella package providing interfaces three commercial services of
-Microsoft's Azure Cloud, namely
+Third, containers can greatly facilitate the \textbf{deployment to
+existing infrastructures}. \emph{Kubernetes}
+(\url{https://kubernetes.io/}) is a container-orchestration system for
+managing container-based application deployment and scaling. A
+\emph{cluster} of containers, orchestrated as a single deployment, e.g.,
+with Kubernetes, can mitigate limitations on request volumes or a
+container occupied with computationally intensive task. A cluster
+features load-balancing, autoscaling of containers across numerous
+servers (in the cloud or on premise), and restarting failed ones. It may
+be your organisation has a Kubernetes cluster already for other
+applications, or you may use one of the different provides for managed
+Kubernetes clusters. Docker containers are used within Kubernetes
+clusters to hold native code, for which Kubernetes creates a framework
+around network connections and scaling of resources up and down. R
+applications, big parallel tasks, or scheduled batch jobs can be
+deployed in a scalable way using Kubernetes, and the deployment can even
+be triggered by changes to code repositories (i.e., CD), see blog post
+``R on Kubernetes'' \citep{edmondson_r_2018}. The package
+\pkg{googleKubernetesR}
+(\url{https://github.com/RhysJackson/googleKubernetesR}) is a proof of
+concept for wrapping the Google Kubernetes Engine API, Google's hosted
+Kubernetes solution, in an easy to use R package. The package
+\CRANpkg{analogsea} provides a solution to programmatically create and
+destroy cloud VMs on the \href{https://www.digitalocean.com/}{Digital
+Ocean} platform \citep{analogsea_2019}. It also includes R wrapper
+functions to install Docker in such a VM, manage images, and control
+containers straight from R functions. These functions are translated to
+Docker CLI commands and transferred transparently to the respective
+remove machine using SSH. \CRANpkg{AzureContainers} is an umbrella
+package providing interfaces three commercial services of Microsoft's
+Azure Cloud, namely
 \href{https://azure.microsoft.com/en-us/services/container-instances/}{Container
@@
 a service, for production purposes it is usually necessary to take into
-account scalability, reliability and ease of management. A
-\emph{cluster} of containers, orchestrated as a single deployment, e.g.,
-with \href{https://kubernetes.io}{Kubernetes}, can mitigate limitations
-on request volumes or a container occupied with computationally
-intensive task. A cluster features load-balancing, autoscaling of
-containers across numerous servers (in the cloud or on premise), and
-restarting failed ones. AzureContainers provides an R-based interface to
-these features for Azure. On the client side, it provides simple
-wrappers to Docker and to Kubernetes' management tools \texttt{kubectl}
-and \texttt{helm}. The package simplifies complex infrastructure
-management to a number of R function calls, given an Azure account with
-sufficient
+account scalability, reliability and ease of management. AzureContainers
+provides an R-based interface to these features and thereby simplifies
+complex infrastructure management to a number of R function calls, given
+an Azure account with sufficient
 credit\footnote{See \emph{"Deploying a prediction service with Plumber"} vignette for details:  \href{https://cran.r-project.org/web/packages/AzureContainers/vignettes/vig01_plumber_deploy.html}{https://cran.r-project.org/web/packages/AzureContainers/vignettes/vig01\_plumber\_deploy.html}.}.
@@
 amount of open source software in combination with proprietary layers
-addign the relevant commercializable features. One such example is
+adding the relevant commercializable features. One such example is
 RStudio's data science platform
@@
 \href{https://github.com/rstudio/rstudio-docker-products}{Docker images}
-for RStudio's commerical products are openly available.
+for RStudio's commerical products are publicly available.
 
@@
 
-\textbf{\pkg{holepunch}} (\url{https://github.com/karthik/holepunch}) is
-an R package that was designed to make sharing work environments
-accessible to novice R users based on \textbf{Binder}. The
-\href{https://mybinder.readthedocs.io/en/latest/}{Binder project},
-maintained by the team behind Jupyter, makes it possible for users to
-create and share their computing environments with others
+The \href{https://mybinder.readthedocs.io/en/latest/}{\textbf{Binder}}
+project, maintained by the team behind Jupyter, makes it possible for
+users to \textbf{create and share computing environments} with others
 \citep{jupyter_binder_2018}. A \emph{BinderHub} allows anyone with
@@
 a workspace, e.g., a hosted Git repository, or a data repository like
-Zenodo. Binder's core internal tool is
-\href{https://repo2docker.readthedocs.io/en/latest/config_files.html}{\texttt{repo2docker}}.
-It deterministically builds a Docker image by parsing the contents of a
+Zenodo. Binder's core internal tool is \texttt{repo2docker}. It
+deterministically builds a Docker image by parsing the contents of a
 repository, e.g., project dependency configurations or simple
-configuration files. In the most powerful case, \texttt{repo2docker}
-builds a given \texttt{Dockerfile}. While this approach works well for
-most run of the mill Python projects, it is not so seamless for R
-projects. For any R projects that use the Tidyverse suite
+configuration
+files\footnote{See supported file types at \href{https://repo2docker.readthedocs.io/en/latest/config\_files.html}{https://repo2docker.readthedocs.io/en/latest/config\_files.html}.}.
+In the most powerful case, \texttt{repo2docker} builds a given
+\texttt{Dockerfile}. While this approach works well for most run of the
+mill Python projects, it is not so seamless for R projects. This is
+partly because \texttt{repo2docker} does not support arbitrary base
+images due to the complex auto-generation of the \texttt{Dockerfile}
+instructions. Two approaches make using Binder easier. for R users.
+First, \textbf{\pkg{holepunch}}
+(\url{https://github.com/karthik/holepunch}) is an R package that was
+designed to make sharing work environments accessible to novice R users
+based on Binder. For any R projects that use the Tidyverse suite
 \citep{wickham_welcome_2019}, the time and resources required to build
@@
 reduces the number of seperately managed environments and thereby
-reduces work and increases portability and reproducibility.
+reduces work and increases portability and reproducibility. Second, the
+\textbf{Whole~Tale} project (\url{https://wholetale.org}) combines the
+strengths of the Rocker Project's curated Docker images with
+\texttt{repo2docker}. Whole~Tale is a National Science Foundation (NSF)
+funded project developing a scalable, open-source, multi-user platform
+for reproducible research \citep{brinckman2019, chard2019a}. A central
+goal of the platform is to enable researchers to easily create and
+publish executable research objects\textbackslash{}footnote\{In
+Whole\textasciitilde{}Tale a \emph{tale} is a research object that
+contains metadata, data (by copy or reference), code, narrative,
+documentation, provenance, and information about the computational
+environment to support computational reproducibility\} associated with
+published research \citep{chard2019b}. Using Whole~Tale, researchers can
+create and publish Rocker-based reproducible research objects to a
+growing number of repositories including DataONE member nodes, Zenodo
+and soon Dataverse. Additionally, Whole~Tale supports automatic data
+citation and is working on capabilities for image preservation and
+provenance capture to improve the transparency of published
+computational research artifacts \citep{mecum2018, mcphillips2019}. For
+R users, Whole~Tale extends the Jupyter Project's \texttt{repo2docker}
+component to simplify the customization of R-based environments for
+researchers with limited experience with either Docker or Git. Multiple
+options have been discussed to allow users to change the base image used
+in \texttt{repo2docker} from the default Ubuntu LTS (long-term support)
+required to support the Rocker Project images. Whole~Tale implemented a
+custom
+\texttt{RockerBuildPack}\footnote{See \href{https://github.com/whole-tale/repo2docker\_wholetale}{https://github.com/whole-tale/repo2docker\_wholetale}}
+to support customization of the \texttt{rocker/geospatial} image through
+\texttt{repo2docker}
+composability\footnote{Composability refers to the ability to combine multiple package managers -- such as R, `pip`, and `conda`}.
+This works in part because Rocker images are based on a Debian
+distribution, so the instructions created by \texttt{repo2docker} for
+Ubuntu work because of compatible shell and package manager.
 
@@
 useful for improving expensive processing of data or shipping complex
-processing pipelines.
-
-First, it is possible to \textbf{offload complex processing to a
-server}. \ldots{}
-
-Second, when processes can be \textbf{run in parallel} for speeding up R
-code.
-
-For example,
-\href{https://CRAN.R-project.org/package=googleComputeEngineR}{\texttt{googleComputeEngineR}}'s
-\texttt{gce\_vm\_cluster()} function can create clusters of 2 or more
+processing pipelines. First, it is possible to \textbf{offload complex
+processing to a server} or clouds, and also parallel executing of
+processes for speeding up or serving of many users.
+\textbf{\CRANpkg{batchtools}} provides a parallel implementation of Map
+for various schedulers \citep{Lang2017batchtools}. The package can
+\href{https://mllg.github.io/batchtools/reference/makeClusterFunctionsDocker.html}{schedule
+jobs with Docker Swarm} \textbf{\CRANpkg{googleComputeEngineR}} has the
+function \texttt{gce\_vm\_cluster()} to create clusters of 2 or more
 virtual machines, running multi-CPU architectures. Instead of running a
@@
 can be processed on all CPU threads of the cluster of machines in the
-cloud, all running in a Docker container with the same R environments.
-This is achieved through \texttt{googleComputeEngineR}'s integration
-with the R parralisation library
-\href{https://CRAN.R-project.org/package=future}{\texttt{library(future)}
-by Henrik Bengtsson}. Local R computation can be thrown up to a
-multi-CPU and VM environment to achieve parrell computation in a few
-lines of R code in your local session. -
-\href{https://cloudyr.github.io/googleComputeEngineR/articles/massive-parallel.html}{some
-demonstrations are available here}. - \href{https://cloud.run}{Google
-Cloud Run} - This is a CaaS (Containers as a Service) that lets you
-launch a Docker container without worrying about underlying
-infrastructure. This dispenses with the developer creating a cloud
-server to run the Docker image on, by abstracting away those servers to
-a more serverless configuration. Cloud Run lets you run your code on top
-of a managed or your own Kubernetes cluster running Knative, and can
-accept any Docker image. The service takes care of network ingress,
-scaling machines up and down to zero, authentication and authorisation,
-all features which are non-trivial for a developer to create on their
-own. This can be used to scale up R code to millions of instances if
-they need to, with little or no changes to existing code. An R
-implementation is shown here at
-\href{https://github.com/MarkEdmondson1234/cloudRunR}{cloudRunR} which
-uses Cloud Run to create a scalable R plumber API.
-\url{https://code.markedmondson.me/googleCloudRunner/index.html}
-
-\href{https://cloud.google.com/cloud-build/}{Google Cloud Build} and the
-Google Container Registry are a continuous intergation service
-respectively image registry that offload building of images to the
-cloud, while serving the needs of commercial environments that helps
-move the workload of building the Docker images to an online service,
-while Cloud Build can be set up to build the Dockerfiles on each GitHub
-commit or release. This means you do not need to build the Docker images
-locally, which can tie up resources since Docker images can be several
-GBs and take a long time to compile. Google Cloud Build works alongside
-to allow you to build private and public Docker images, which allows you
-to build up your own dependency graphs for downstream applications. -
-\url{https://code.markedmondson.me/googleCloudRunner/articles/cloudbuild.html}
-
-\begin{itemize}
-\tightlist
-\item
-  \CRANpkg{batchtools} \citep{Lang2017batchtools} can
-  \href{https://mllg.github.io/batchtools/reference/makeClusterFunctionsDocker.html}{schedule
-  jobs with Docker Swarm}
-\item
-  scalable deployments, e.g., start with numerous Shiny talks mentioning
-  Rocker at useR!2017
-\item
-  \href{https://github.com/dynverse/dynmethods}{dynmethods}
-  \citep{rcannood}: In order to evaluate ±50 computational methods which
-  all used different environments (R, Python, C++, \ldots{}), we wrapped
-  each of them in a docker container and can execute these methods from
-  R. Again, all of these containers are being built on travis-ci, and
-  will only be pushed to Docker~Hub if the integration test succeeds.
-\item
-  \CRANpkg{drake},
-  \url{https://docs.ropensci.org/drake/index.html?q=docker\#with-docker}
-\end{itemize}
-
-In the RStudio Server Pro 1.2 release in 2019, RStudio added new
-functionality called
-\href{https://solutions.rstudio.com/launcher/overview/}{Launcher}. It
-gives users the ability to spawn R sessions and background jobs in a
-scalable way on external clusters, e.g.,
+cloud, all running a Docker container with the same R environments.
+\CRANpkg{googleComputeEngineR} integrates with the R parralisation
+package \CRANpkg{future} to enable this with only a few lines or R
+code\footnote{\href{https://cloudyr.github.io/googleComputeEngineR/articles/massive-parallel.html}{https://cloudyr.github.io/googleComputeEngineR/articles/massive-parallel.html}}.
+\href{https://cloud.run}{Google Cloud Run} is a CaaS (Containers as a
+Service) platform. Users can launch containers using any Docker image
+without worrying about underlying infrastructure in a so called
+serverless configuration. The service takes care of network ingress,
+scaling machines up and down, authentication and authorisation---all
+features which are non-trivial for a developer to create on their own.
+This can be used to scale up R code to millions of instances if need be
+with little or no changes to existing code, as demonstrated by the proof
+of concept
+\texttt{cloudRunR}\footnote{\href{https://github.com/MarkEdmondson1234/cloudRunR}{https://github.com/MarkEdmondson1234/cloudRunR}},
+which uses Cloud Run to create a scalable R-based API using
+\CRANpkg{plumber}. \href{https://cloud.google.com/cloud-build/}{Google
+Cloud Build} and the Google Container Registry are a continuous
+integration service respectively image registry that offload building of
+images to the cloud, while serving the needs of commercial environments
+such as private Docker images or image stacks. As Googl Cloud Build
+itself can run any container, the package \pkg{googleCloudRunner}
+demonstrates how R can be used as the control language for one time or
+batch processing jobs and scheduling of
+jobs\footnote{\href{https://code.markedmondson.me/googleCloudRunner/articles/cloudbuild.html}{https://code.markedmondson.me/googleCloudRunner/articles/cloudbuild.html}}.
+\label{drake} \textbf{\CRANpkg{drake}} is a workflow manager for data
+science projects \citep{landau_drake_2019}. It features implicit
+parallel computing and automated detection of the parts of the work that
+actually needs to be reexecuted. \CRANpkg{drake} has been demonstrated
+to run inside containers for high
+reproducibility\footnote{See for example \href{https://github.com/joelnitta/pleurosoriopsis}{https://github.com/joelnitta/pleurosoriopsis} or \href{https://gitlab.com/ecohealthalliance/drake-gitlab-docker-example}{https://gitlab.com/ecohealthalliance/drake-gitlab-docker-example}, the latter even running in a CI, see~\nameref{cicd}.},
+but also how to use \CRANpkg{future} package's function
+\texttt{makeClusterPSOCK()} to send parts of the workflow to a Docker
+image for
+execution\footnote{\href{https://docs.ropensci.org/drake/index.html?q=docker\#with-docker}{https://docs.ropensci.org/drake/index.html?q=docker\#with-docker}}
+\citep[see package's function documentation;~][]{future_2020}. In the
+latter case, the container control code must be written by the user and
+the \CRANpkg{future} package ensures the host and worker can connect for
+communicating over socket connections. \textbf{RStudio Server Pro}
+includes a functionality called
+\href{https://solutions.rstudio.com/launcher/overview/}{Launcher} (since
+version~1.2, released in 2019). It gives users the ability to spawn R
+sessions and background/batch jobs in a scalable way on external
+clusters, e.g.,
 \href{https://support.rstudio.com/hc/en-us/articles/360019253393-Using-Docker-images-with-RStudio-Server-Pro-Launcher-and-Kubernetes}{Kubernetes
 based on Docker images} or \href{https://slurm.schedmd.com/}{Slurm}
-clusters, and optionally, with Singularity containers. A key benefit of
-Launcher is the ability for R and Python users to interactively work in
-RStudio without learning about Docker at all - while still leveraging
-containers and Kubernetes. Users can submit batch jobs to Kubernetes
-clusters without writing specific deployment scripts.
-
-\label{pipelines} Third, containers are perfectly suited for
+clusters, and optionally, with Singularity containers. A benefit of the
+proprietary Launcher software is the ability for R and Python users to
+leverage containerisation's advantages in RStudio without writing
+specific deployment scripts or learning about Docker or managing
+clusters at all.
+
+\label{pipelines} Second, containers are perfectly suited for
 \textbf{packaging and executing software pipelines} and required data.
@@
 repository to ensure the package versions are tied to a specific date,
-rather than the most recent version. Future researchers can download the
-compendium from the repository and run the included \texttt{Dockerfile}
-to build a new image that recreates the computational environment used
-to produce the original research results. If building the image fails,
-the human-readable instructions in a \texttt{Dockerfile} are the
-starting point for rebuilding the environment. Further safeguarding
-practices are currently under development or not part of common practice
-yet, such as preservation of images \citep{emsley_framework_2018} and
-storing them alongside \texttt{Dockerfile}s
-\citep[cf.][]{nust_opening_2017}, or pinning of system libraries beyond
-the tagged base images, which may be seen as stable or dynamic depending
-on the applied time scale
+rather than the most recent version. A more extreme case is the
+\emph{dynverse} project (\url{https://dynverse.org/}), which packages
+over 50 computational methods with different environments (R, Python,
+C++, etc.) in Docker images, which can be executed from R.
+\emph{dynverse} uses a CI (cf.~\nameref{cicd}) to build Rocker-derived
+containers, test them, and push them to Docker~Hub if the tests succeed.
+
+Future researchers can download the compendium from the repository and
+run the included \texttt{Dockerfile} to build a new image that recreates
+the computational environment used to produce the original research
+results. If building the image fails, the human-readable instructions in
+a \texttt{Dockerfile} are the starting point for rebuilding the
+environment. Further safeguarding practices are currently under
+development or not part of common practice yet, such as preservation of
+images \citep{emsley_framework_2018} and storing them alongside
+\texttt{Dockerfile}s \citep[cf.][]{nust_opening_2017}, or pinning of
+system libraries beyond the tagged base images, which may be seen as
+stable or dynamic depending on the applied time scale
 \citep[see discussion on `debian:testing` base image in][]{RJ-2017-065}.
-
 A recommendation of the recent National Academies' report on
@@
 GDAL \citep{gdal_2019} is a crucial library in the geospatial domain.
-GDAL is a system dependen\^{}cy allowing R packages such as
-\CRANpkg{sf}, which provides the core data model for geospatial vector
-data, or \CRANpkg{rgdal}, to accomodate users to be able to read and
-write hundreds of different spatial raster and vector formats.
-\CRANpkg{sf} and \CRANpkg{rgdal} have hundreds of indirect reverse
-imports and dependencies and therefore the maintainers spend a lot of
-efforts not to break these. Purpose built Docker are used to prepare for
-upcoming releases of system libraries, individual bug reports, and for
-the lowest supported versions of system
+GDAL is a system dependency allowing R packages such as \CRANpkg{sf},
+which provides the core data model for geospatial vector data, or
+\CRANpkg{rgdal}, to accomodate users to be able to read and write
+hundreds of different spatial raster and vector formats. \CRANpkg{sf}
+and \CRANpkg{rgdal} have hundreds of indirect reverse imports and
+dependencies and therefore the maintainers spend a lot of efforts not to
+break these. Purpose built Docker are used to prepare for upcoming
+releases of system libraries, individual bug reports, and for the lowest
+supported versions of system
 libraries\footnote{Cf.
@@
 runtime. This approach clearly seperates test environment, test code,
-and current working copy of the code.
+and current working copy of the code. \label{rselenium} Another use case
+where a container helps to standardise tests across operating systems is
+detailed the vignettes of the package \CRANpkg{RSelenium}
+\citep{rselenium_2019}. The package recommends Docker for running the
+\href{https://selenium.dev/}{Selenium} Server application needed to
+execute test suites on browser-based user interfaces and webpages, but
+requires users to manually manage the Docker containers.
 
@@
 containerisation happens more in the background at the system level.
+Furthermore, a maturing of the Rockerverse packages for managing
+containers may lead to their adoption where currently manual coding is
+required, e.g.~in the case of \CRANpkg{RSelentium} or \CRANpkg{drake}
+(see Sections~\nameref{rselenium} and \nameref{drake} respectively). In
+specific cases, e.g., for \CRANpkg{analogsea}, the interaction with the
+Docker daemon may remain to specific to reuse first order packages to
+control Docker.
 
@@
 
-The ordering of authors following DN as first author is alphabetical. DN
+The ordering of authors following DN and DE is alphabetical. DN
 conceived the article idea,
@@
 the section on \pkg{outsider}. GD contributed the CARD.com use case. DE
-wrote the introduction and the section about Containerization and
-Rocker. RC \& EH contributed to the section on interfaces for Docker in
-R. DC contributed content on Gigantum. ME contributed to the section on
+wrote the introduction and the section about Containerization and Rocker
+and reviewed all sections. RC \& EH contributed to the section on
+interfaces for Docker in R (\emph{dynverse} and \texttt{dynwrap}). DC
+contributed content on Gigantum. ME contributed to the section on
 processing and deployment to cloud services. CF wrote paragraphs about
@@
 on research compendia and made the project Binder-ready. HN \& JN
-co-wrote the section on enterprise production environments
-(\pkg{googleComputeEngineR}, Google Cloud Run, Kubernetes). KR wrote the
-section about \pkg{holepunch}. NR wrote paragraphs about shared work
+co-wrote the section on the T-Mobile use case. KR wrote the section
+about \pkg{holepunch}. NR wrote paragraphs about shared work
 environments and GPUs. LS \& NT wrote the section on Bioconductor. CW
@@
 \address{%
+Dirk Eddelbuettel\\
+University of Illinois at Urbana-Champaign\\
+Department of Statistics\\ Illini Hall, 725 S Wright St\\ Champaign, IL 61820, USA\\ \orcid{0000-0001-6419-907X}\\
+}
+\href{mailto:dirkd@eddelbuettel.com}{\nolinkurl{dirkd@eddelbuettel.com}}
+
+\address{%
 Dom Bennett\\
@@
 \address{%
-Dirk Eddelbuettel\\
-University of Illinois at Urbana-Champaign\\
-Department of Statistics\\ Illini Hall, 725 S Wright St\\ Champaign, IL 61820, USA\\ \orcid{0000-0001-6419-907X}\\
-}
-\href{mailto:dirkd@eddelbuettel.com}{\nolinkurl{dirkd@eddelbuettel.com}}
-
-\address{%
 Mark Edmondson\\

### 2020-01-27 manuscript.Rmd
fix output of table to latex

@@
 # Until then, use remotes::install_github("rstudio/rticles")
-if (packageVersion("rticles") >= "0.12.3") {
 requireNamespace("kableExtra")
-  knitr::kable(tab, escape = FALSE) %>%
+knitr::kable(tab, format = "latex", escape = FALSE) %>%
   kableExtra::row_spec(0, angle = -90)
-} else {
-  cat("\n\n[Functionality overview table omitted due to a small bug in rticles]\n\n")
-}
 ```

### 2020-01-27 manuscript.Rmd
fix cross refs

@@
 It features implicit parallel computing and automated detection of the parts of the work that actually needs to be reexecuted.
-\CRANpkg{drake} has been demonstrated to run inside containers for high reproducibility\footnote{See for example \href{https://github.com/joelnitta/pleurosoriopsis}{https://github.com/joelnitta/pleurosoriopsis} or \href{https://gitlab.com/ecohealthalliance/drake-gitlab-docker-example}{https://gitlab.com/ecohealthalliance/drake-gitlab-docker-example}, the latter even running in a CI, see~\nameref{cicd}.}, but also how to use \CRANpkg{future} package's function `makeClusterPSOCK()` to send parts of the workflow to a Docker image for execution\footnote{\href{https://docs.ropensci.org/drake/index.html?q=docker\#with-docker}{https://docs.ropensci.org/drake/index.html?q=docker\#with-docker}} \citep[see package's function documentation;~][]{future_2020}.
+\CRANpkg{drake} has been demonstrated to run inside containers for high reproducibility\footnote{See for example \href{https://github.com/joelnitta/pleurosoriopsis}{https://github.com/joelnitta/pleurosoriopsis} or \href{https://gitlab.com/ecohealthalliance/drake-gitlab-docker-example}{https://gitlab.com/ecohealthalliance/drake-gitlab-docker-example}, the latter even running in a continuous integration platform (cf.~\nameref{development}.}, but also how to use \CRANpkg{future} package's function `makeClusterPSOCK()` to send parts of the workflow to a Docker image for execution\footnote{\href{https://docs.ropensci.org/drake/index.html?q=docker\#with-docker}{https://docs.ropensci.org/drake/index.html?q=docker\#with-docker}} \citep[see package's function documentation;~][]{future_2020}.
 In the latter case, the container control code must be written by the user and the \CRANpkg{future} package ensures the host and worker can connect for communicating over socket connections.
@@
 A more extreme case is the _dynverse_ project ([https://dynverse.org/](https://dynverse.org/)), which packages over 50 computational methods with different environments (R, Python, C++, etc.) in Docker images, which can be executed from R.
-_dynverse_ uses a CI (cf.&nbsp;\nameref{cicd}) to build Rocker-derived containers, test them, and push them to Docker&nbsp;Hub if the tests succeed.
+_dynverse_ uses a continuous integreation (CI) platform to build Rocker-derived images, test them, and publish them on Docker&nbsp;Hub if the tests succeed.
 

### 2020-01-27 manuscript.Rmd
add preprint template

@@
 
+```{r preprint_pdf, eval=FALSE, include=FALSE}
+# run this chunk, get the PDF, then revert changes in touched files
+library("rmarkdown")
+library("rticles")
+unlink('manuscript_cache', recursive = TRUE)
+
+# function is a copy of rticles::rjournal_article() but...
+# - using RJWrapper_preprint_template.tex (which includes header.tex commands)
+# - directly referencing unexported functions from rticles
+# - direclty setting the keep_tex option
+rjournal_style_preprint <- function (..., citation_package = "natbib") {
+    rmarkdown::pandoc_available("2.2", TRUE)
+    base <- rticles:::pdf_document_format("rjournal_article", highlight = NULL, 
+        citation_package = citation_package, ...)
+    base$pandoc$to <- "latex"
+    base$pandoc$ext <- ".tex"
+    base$pandoc$keep_tex <- TRUE # added
+    base$post_processor <- function(metadata, utf8_input, output_file, 
+        clean, verbose) {
+      filename <- basename(output_file)
+      if (filename != (filename2 <- gsub("_", "-", filename))) {
+          file.rename(filename, filename2)
+          filename <- filename2
+      }
+      m <- list(filename = xfun::sans_ext(filename))
+      h <- rticles:::get_list_element(metadata, c("output", "rticles::rjournal_article", 
+          "includes", "in_header"))
+      h <- c(h, if (length(preamble <- unlist(metadata[c("preamble", 
+          "header-includes")]))) {
+          f <- tempfile(fileext = ".tex")
+          on.exit(unlink(f), add = TRUE)
+          xfun::write_utf8(preamble, f)
+          f
+      })
+      #t <- find_resource("rjournal_article", "RJwrapper.tex")
+      t <- here::here("RJwrapper_preprint_template.tex")
+      rticles:::template_pandoc(m, t, "RJwrapper.tex", h, verbose)
+      tinytex::latexmk("RJwrapper.tex", base$pandoc$latex_engine, 
+          clean = clean)
+    }
+    base$knitr$opts_chunk$comment <- "#>"
+    hilight_source <- rticles:::knitr_fun("hilight_source")
+    hook_chunk = function(x, options) {
+        if (rticles:::output_asis(x, options)) 
+            return(x)
+        paste0("```{=latex}\n\\begin{Schunk}\n", x, "\\end{Schunk}\n```")
+    }
+    hook_input <- function(x, options) paste(c("\\begin{Sinput}", 
+        hilight_source(x, "sweave", options), "\\end{Sinput}", 
+        ""), collapse = "\n")
+    hook_output <- function(x, options) paste("\\begin{Soutput}\n", 
+        x, "\\end{Soutput}\n", sep = "")
+    base$knitr$knit_hooks <- rticles:::merge_list(base$knitr$knit_hooks, 
+        list(chunk = hook_chunk, source = hook_input, output = hook_output, 
+            message = hook_output, warning = hook_output, plot = knitr::hook_plot_tex))
+    base
+}
+
+rmarkdown::render("manuscript.Rmd", output_format = rjournal_style_preprint())
+```
+
 \bibliography{RJreferences}

### 2020-01-27 README.Rmd
add some missing aspects based on @jennybc 's docker-why repo, closes #69

--- 2020-01-22 README.Rmd
@@
 - https://github.com/jlisic/R-docker-centos because CentOS images are also provided by R-hub
-- https://github.com/cardcorp/card-rocker because last change 3 years ago, an all based on `rocker/r-base` - lesson learned would be interesting still, see [#15](https://github.com/nuest/rockerverse-paper/issues/15)
 - `sevenbridges` includes several Dockerfiles and documentation about using images, and also seems to use Docker in the built workflows but via other tools, not directly, https://sbg.github.io/sevenbridges-r/, https://github.com/sbg/sevenbridges-r/search?p=4&q=docker&unscoped_q=docker
 - https://github.com/rocker-jp because it seems to have stalled soon after inception
+- https://github.com/thomasp85/fiery because no examples/docs for Docker (though probably trivial anyway)
+- [RStudio Connect](https://rstudio.com/products/connect/) because while it does use a similar mechanism to seperate processes, it does _not_ use Docker and the comparison would require a lot of background explanation, and not all details are available because the software is proprietary
+- https://github.com/ecohealthalliance/reservoir is an invididual specific data science image container for a group - add if it fits
+- https://itsalocke.com/blog/building-an-r-training-environment/ (one single container for many users in teaching, uses rocker/tidyverse)

### 2020-01-27 manuscript.Rmd
add some missing aspects based on @jennybc 's docker-why repo, closes #69

@@
 Several second order R packages attempt to make the process of creating Docker images and using containers for specific tasks, such as running tests or rendering reproducible reports, easier.
-While authoring and managing an environment with Docker by hand is possible and feasible for experts\footnote{See, e.g., this tutorial by RStudio on how to manage environments and package versions and to ensure deterministic image builds with Docker: \href{https://environments.rstudio.com/docker}{https://environments.rstudio.com/docker}.}, the following examples show the power of automation when environments become too cumbersome or acquiring the skills is not possible in due course.
+While authoring and managing an environment with Docker by hand is possible and feasible for experts\footnote{See, e.g., this tutorial by RStudio on how to manage environments and package versions and to ensure deterministic image builds with Docker: \href{https://environments.rstudio.com/docker}{https://environments.rstudio.com/docker}.}, the following examples show the power of automation when environments become too cumbersome.
 Especially _version pinning_, with packages \pkg{remotes} and \pkg{versions} for R or by using MRAN, and with system package managers for different operating systems, can greatly increase the reproducibility of built images and are common approaches.
@@
 
-The package **\CRANpkg{renv}** ([https://rstudio.github.io/renv/](https://rstudio.github.io/renv/)) helps users to manage the state of the R library in a reproducible way, further providing isolation and portability \citep{renv2019}.
-The package does not cover system dependencies, though, the \CRANpkg{renv}-based environment can be transferred into a container either by restoring the environment based on the main configuration file `renv.lock` or by storing the \CRANpkg{renv}-cache on the host and not in the container \citep{ushey_using_2019}.
+System dependencies are the domain of Docker, but for a full description of the computing environment one must also manage the R package.
+This can be achiedved with MRAN, or with the package **\CRANpkg{renv}** ([https://rstudio.github.io/renv/](https://rstudio.github.io/renv/)), which helps users to manage the state of the R library in a reproducible way, further providing isolation and portability \citep{renv2019}.
+While \CRANpkg{renv} does not cover system dependencies, the \CRANpkg{renv}-based environment can be transferred into a container either by restoring the environment based on the main configuration file `renv.lock` or by storing the \CRANpkg{renv}-cache on the host and not in the container \citep{ushey_using_2019}.
+With both the system dependencies and R packages consciously managed in a Docker image, users can start using containers as the _only_ environment for their workflows, which allows them to work indepenently of physical computers\footnote{Allowing them to be digital "nomads", cf.
+J.~Bryan's \href{https://github.com/jennybc/docker-why}{https://github.com/jennybc/docker-why}.} and to assert a specific degree of confidence in the stability of a developed software \citep[cf.
+\texttt{README.Rmd} in][]{marwick_research_2017}.
 
@@
 AzureContainers provides an R-based interface to these features and thereby simplifies complex infrastructure management to a number of R function calls, given an Azure account with sufficient credit\footnote{See \emph{"Deploying a prediction service with Plumber"} vignette for details:  \href{https://cran.r-project.org/web/packages/AzureContainers/vignettes/vig01_plumber_deploy.html}{https://cran.r-project.org/web/packages/AzureContainers/vignettes/vig01\_plumber\_deploy.html}.}.
+[Heroku](https://www.heroku.com/) is a further cloud platform as a service provider, who supports container-based applications.
+`heroku-docker-r` ([https://github.com/virtualstaticvoid/heroku-docker-r](https://github.com/virtualstaticvoid/heroku-docker-r)) is an independent project providing a template for deploying R applications based on Heroku's image stack, including multiple examples for interfacing R with other programming languages.
+Yet the approach requires a manual management of the computing environment.
 
@@
 Teaching is a further example where shared computing environments and sandboxing can greatly improve the process.
-First, **prepared environments for teaching** are especially helpful for courses that require access to a relatively complex setup of software tools, e.g., database systems.
+First, **prepared environments for teaching** are especially helpful for (a) introductory courses where the first step of installation and configuration can be hurdles for students \citep{cetinkaya-rundel_infrastructure_2018}, and (b) courses that require access to a relatively complex setup of software tools, e.g., database systems.
+\citet{cetinkaya-rundel_infrastructure_2018} describe how a Docker-based deployment of RStudio (i) avoided problems with troubleshooting individual students computers and greatly increase engagement through very quickly showing tangible outcomes, e.g., a visualisation, and (ii) reduced demand on teaching and IT staff.
+Each student get's acces to a personal RStudio instance running in a container after authentication with the university login, which gives the benefits of sandboxing and possibility of limiting resources.
+\citet{cetinkaya-rundel_infrastructure_2018} found that for the courses at hand, actual usage of the UI is intermittent so a single cloud based VM with 4&nbsp;cores and 28GB&nbsp;RAM sufficed for over 100 containers.
+An example for mitigating _complex setups_ is teaching databases.
 R is very useful tool for interfacing with databases, because almost every open source and proprietary database system has an R package that allows users to connect and interact with the it.
@@
 New releases of even just a single piece of software in this pipeline can break the entire workflow, making it difficult to find the error and difficult for others to reuse existing pipelines.
-These breakages can make the original the results irreproducible and not extandable.
+These breakages can make the original the results irreproducible and not ¸
 The chance of a substantial disruption like this is high in a multi-year research project where key pieces of third-party software may have several major updates over the duration of the project.
@@
 This concept has been taken up and extended\footnote{See full literature list at \href{https://research-compendium.science/}{https://research-compendium.science/}.}, not the least by applying containerisation and other methods for managing computing environments---see Section&nbsp;\nameref{envs}.
-Containers give the researcher an isolated environment to assemble these research pipelines with specific versions of software.
+Containers give the researcher an isolated environment to assemble these research pipelines with specific versions of software to minimize problems with breaking changes and make workflows easier to share \citep[cf.][]{boettiger_introduction_2015,marwick_packaging_2018}.
 Research workflows in containers are safe from contamination from other activities occuring on the researcher's computer, for example the installation of newest version of packages for teaching demonstrations.
@@
 A more extreme case is the _dynverse_ project ([https://dynverse.org/](https://dynverse.org/)), which packages over 50 computational methods with different environments (R, Python, C++, etc.) in Docker images, which can be executed from R.
-_dynverse_ uses a continuous integreation (CI) platform to build Rocker-derived images, test them, and publish them on Docker&nbsp;Hub if the tests succeed.
+_dynverse_ uses a CI platform (see&nbsp;\nameref{ci}) to build Rocker-derived images, test them, and publish them on Docker&nbsp;Hub if the tests succeed.
 
@@
 If building the image fails, the human-readable instructions in a `Dockerfile` are the starting point for rebuilding the environment.
+When combinded with CI (see&nbsp;\nameref{ci}), a research compendium set-up can enable _continuous analysis_ with easier verification of reproducibility and audits trails \citep{beaulieu-jones_reproducibility_2017}.
+
 Further safeguarding practices are currently under development or not part of common practice yet, such as preservation of images \citep{emsley_framework_2018} and storing them alongside `Dockerfile`s \citep[cf.][]{nust_opening_2017}, or pinning of system libraries beyond the tagged base images, which may be seen as stable or dynamic depending on the applied time scale \citep[see discussion on `debian:testing` base image in][]{RJ-2017-065}.
@@
 
-Fifth, Docker images can be used **on CI platforms** to streamline the testing of packages.
+\label{ci}
+Fifth, Docker images can be used **on continuous integration (CI) platforms** to streamline the testing of packages.
 \citet{ye_docker_2019} describes how they speed up the process of testing by running tasks on [Travis&nbsp;CI](https://travis-ci.org/) within a container using `docker exec`, e.g., the package check or rendering of documentation.
@@
 Though for the moment, the sign of the times points to more experimentation and feature growth, e.g., images for GPU-based computing and artificial intelligence.
-Even with coding being more and more accepted as a required, and achievable skill, an easier access, for example by exposing containerisation benefits via simple user interfaces in the users' IDE, could be an important next step.
-Currently containerisation happens more in the background at the system level.
+Even with coding being more and more accepted as a required, and achievable skill, an easier access, for example by exposing containerisation benefits via simple user interfaces in the users' IDE, could be an important next step since currently containerisation happens more in the background for UI-based development (e.g., a `rocker/rstudio` image in the cloud).
 Furthermore, a maturing of the Rockerverse packages for managing containers may lead to their adoption where currently manual coding is required, e.g.

### 2020-01-27 manuscript.Rmd
remove image sections, closes #65

@@
 ---
-title: "The Rockerverse: Images, Packages, and Applications for Containerization with R"
+title: "The Rockerverse: Packages and Applications for Containerization with R"
 author:
@@
   This articles surveys downstream projects building upon Rocker and presents the current state of R packages for managing Docker images and controlling containers.
-  We also look beyond Rocker to other projects connecting containerisation with R, namely alternative suites of images.
   These use cases and the variety of applications demonstrate the power of Rocker specifically and containerisation in general.
@@
 
-The article continues with a brief introduction of containerization basics and the Rocker&nbsp;Project, followed by descriptions of projects developing images with R installations.
-Then, we present applications, starting with the R packages specificly for interacting with Docker, second-level packages using containers indirectly or only for specific features, up to complex use cases leveraging containers.
+The article continues with a brief introduction of containerization basics and the Rocker&nbsp;Project, followed by use cases and applications, starting with the R packages specificly for interacting with Docker, second-level packages using containers indirectly or only for specific features, up to complex use cases leveraging containers.
 We conclude with a reflection on the landscape of packages and applications and point out future directions of development.
@@
 
-For statistical computing and analysis centered around R, the Rocker&nbsp;Project has provided a variety of Docker containers since its start in 2014 \citep{RJ-2017-065}.
+For statistical computing and analysis centered around R, the **Rocker&nbsp;Project** has provided a variety of Docker containers since its start in 2014 \citep{RJ-2017-065}.
 The Rocker&nbsp;Project provides several lines of containers spanning to from building blocks with `R-release` or `R-devel`, via containers with [RStudio&nbsp;Server](https://rstudio.com/products/rstudio/) and [Shiny&nbsp;Server](https://rstudio.com/products/shiny/shiny-server/), to domain-specific containers such as [`rocker/geospatial`](https://github.com/rocker-org/geospatial) \citep{rocker_geospatial_2019}.
@@
 
-# Container images
-\label{images}
+# Use cases and applications
+<!-- from packages closer to Docker to second level packages to more complex use cases -->
+\label{applications}
 
-## Images for alternative R engines
-\label{alternatives}
+## Image stacks for communities of practice
 
-As outlined above, R is a widely-used language with a large community.
-The large number of extension packages provides access to an unrivaled variety of established and upcoming features.
-Nevertheless, special use cases and experimental projects in academia and industry exist to test new approaches, or provide features different to what the R implementation maintained by the [R Core Team](https://www.r-project.org/contributors.html) and available via [CRAN](https://cran.r-project.org/) provides.
-While the images presented in this section are far from being as vetted, stable, and widely used as any of the Rocker images, they demonstrate an important advantage of containerisation technology, namely the abiliy to transparently build portable stacks of open source software and make them easily accessible to others.
-All these alternative implementations of R are published under GPL licenses.
-Since all implementations of R claim better performance as a core motivation, a comparision based on Docker images, potentially leveraging the [resource restriction mechanisms](https://docs.docker.com/config/containers/resource_constraints/) of Docker to level the playing field, seems useful future work.
-
-**Microsoft R Open** (MRO) is an R distribution formerly known as Revolution R Open (RRO) before Revolution Analytics was acquired by Microsoft.
-MRO is compatible with main R and its packages and it "[..] includes additional capabilities for improved performance, reproducibility, and platform support." \citep{microsoft_mro}
-Most notably these capabilities are the MRAN repository, which is enabled by default, and some bundled packages such as \CRANpkg{checkpoint} and specific R packages maintained by Microsoft\footnote{See "Bundled Packages" on \href{https://mran.microsoft.com/rro/installed}{https://mran.microsoft.com/rro/installed}.}.
-Originally the optional integration with [Intel® Math Kernel Library](https://software.intel.com/en-us/mkl) (MKL) for multi-threading in linear algebra operations \citep{microsoft_multithread} was a distinguishing feature before it was made available publicly, and can be used in base R, too \citep{eddelbuettel_thinking_2018}.
-
-MRO does not provide official Docker images, but a set of community-maintained `Dockerfile`s and Docker images are provided by the poject `mro-docker` ([https://github.com/nuest/mro-docker](https://github.com/nuest/mro-docker)), e.g., the image [`nuest/mro`](https://hub.docker.com/repository/docker/nuest/mro/).
-The images are inspired by the Rocker images and can be used much in the same fashion, effectively a drop-in replacement allowing users to quickly evaluate if the benefits of MRO + Intel® MKL apply to their use case.
-Version-tagged images are provided for the latest bugfix release of recent R versions.
-Extended license information about MKL is printed at every startup.
-
-**Renjin** is an interpreter for the R language running on top of the [Java Virtual Machine](https://en.wikipedia.org/wiki/Java_virtual_machine) (JVM) providing full two-way access between Java and R code \citep{wikipedia_renjin_2018}.
-It was developed to combine the benefits of R, such scripting and extension packages, with the JVM's advantages in the areas of security, cross-platform availability and integration into enterprise platforms.
-R extension packages need to be specially compiled and are distributed via the Java package manager [Apache Maven](https://en.wikipedia.org/wiki/Apache_Maven), cf.
-[http://packages.renjin.org/packages](http://packages.renjin.org/packages) for available packages.
-Packages are loaded on demand, i.e., at the first call to `library()`.
-Not all R packages, especially one linking to binary libraries, are available, e.g., `rgdal` \footnote{\href{http://packages.renjin.org/package/org.renjin.cran/rgdal/1.4-4/build/1}{http://packages.renjin.org/package/org.renjin.cran/rgdal/1.4-4/build/1}}.
-There are no offical Docker images for Renjin, but community-maintained images for selected releases only are available under `nuest/renjin` on Docker&nbsp;Hub and GitHub at [https://hub.docker.com/r/nuest/renjin](https://hub.docker.com/r/nuest/renjin) and [https://github.com/nuest/renjin-docker](https://github.com/nuest/renjin-docker) respectively.
-These images expose the command line interface of Renjin in a similar fashion as Rocker images and allow an easy evaluation of Renjin's suitability, but are not intended for production use.
-
-**pqR** ([http://www.pqr-project.org/](http://www.pqr-project.org/)) is a "a pretty quick version of R".
-_pqR_ attempts to improve R on some opinionated issues in the R language and is the basis for experimental features, e.g., automatic diffentiation\footnote{\href{https://riotworkshop.github.io/abstracts/riot-2019-pqr.txt}{https://riotworkshop.github.io/abstracts/riot-2019-pqr.txt}}.
-The source code development [on GitHub](https://github.com/radfordneal/pqR/) is a one man project and it does not provide any Docker images.
-But especially disruptive approaches may contribute to the development of the R ecosystem, so the `nuest/pqr` image was independently created, see Docker Hub at [https://hub.docker.com/r/nuest/pqr/](https://hub.docker.com/r/nuest/pqr/) and GitHub at [https://github.com/nuest/pqr-docker](https://github.com/nuest/pqr-docker).
-
-**FastR** ([https://github.com/oracle/fastr](https://github.com/oracle/fastr)) is "A high-performance implementation of the R programming language, built on GraalVM" \citep{oracle_labs_oraclefastr_2020}.
-It is developed by Oracle, connects R to the GraalVM ecosystem \citep{wikipedia_graalvm_2019}, and also claims superior performance but also targets full compatibility with base R \citep{oracle_labs_oraclefastr_2020}.
-There are no offical Docker images provided, but indepedently maintained experimental images and `Dockerfile`s are provided by [nuest/fastr-docker](https://github.com/nuest/fastr-docker), e.g., [`nuest/fastr` on Docker Hub](https://cloud.docker.com/repository/docker/nuest/fastr/).
-
-## Bioconductor
-\label{bioc}
-
-_Bioconductor_ ([https://bioconductor.org/](https://bioconductor.org/)) is an open source, open development project for the analysis and comprehension of genomic data \citep{gentleman_bioconductor_2004}.
+**Bioconductor** ([https://bioconductor.org/](https://bioconductor.org/)) is an open source, open development project for the analysis and comprehension of genomic data \citep{gentleman_bioconductor_2004}.
 The project consists of 1823 R software packages as of October 30th 2019, as well as packages containing annotation or experiment data.
@@
 
-## Images for (historic) R versions
-\label{versions}
-
-As with any other software, each new version of R comes with its share of changes.
-Some are breaking changes, some are not, but the fact is that running a piece of code in a given version might give different results from another version can be an issue when it comes to reproducibility of workflows and stability of applications.
-For example, think about R random seed: starting with `R 3.6.0`, running `set.seed(2811); sample(1:1000, 2)` will not give the same result as if it was run inside an older version of R.
-That might seem trivial, but all the code using the random number generator will not be exactly reproducible after this breaking change.
-Therefore, controlling the version of software is most crucial for reproducible research \citep[e.g.][]{boettiger_introduction_2015}.
-
-Containers are perfectly suited to capture a specific configuration of a computing environment to prohibit such problems.
-The Rocker images provides the _versioned stack_ for different the version of R since the project inception, which use the _then-current_ stable Debian base image \citep[cf.][]{RJ-2017-065}.
-Based on custom build phase hooks\footnote{\href{https://docs.docker.com/docker-hub/builds/advanced/}{https://docs.docker.com/docker-hub/builds/advanced/}}, i.e., small shell scripts executed at different phases in the automated build of Rocker images, there are also semantic version tags for the most recent freezed (i.e., using MRAN) versions with only the major and minor version\footnote{See \href{https://github.com/rocker-org/rocker-versioned/blob/master/VERSIONS.md}{https://github.com/rocker-org/rocker-versioned/blob/master/VERSIONS.md}, FIXME: \href{https://github.com/rocker-org/rocker-versioned/issues/42}{https://github.com/rocker-org/rocker-versioned/issues/42}.}.
-So, at the time of writing this article, `rocker/r-ver:3` and `3.6` are aliases for `3.6.0`, because `3.6.1` is the latest release.
-With the release of `3.6.2` and pinning of the MRAN version in `3.6.1`, the tags `3` and `3.6` will deliver the same image as `3.6.1`.
-These tags allow users to update the base image to retrieve bugfixes while reducing the risk of also introducing breaking changes.
-
-[r-online](https://github.com/ColinFay/ronline) is an app for helping users to detect breaking changes between different R versions, and for historic exploration of R.
-With a standalone NodeJS app or [online](https://srv.colinfay.me/r-online), the user can compare a piece of code run in two separate versions of R.
-Internally, r-online opens one or two Docker instances with the given version of R based on Rocker images, executes a given piece of code, and returns the result to the user.
-
-A container also provides an isolated sandbox environment suitable for testing and evaluation, without interfering with the "main" working environment.
-This enables cross version testing and bugfixing.
-To take this even further, the Rocker contributors are discussing the provision of R versions reaching back futher than the project's own inception, reaching back as far as R `2.x` and `1.x` \footnote{\href{https://github.com/rocker-org/rocker-versioned/issues/138}{https://github.com/rocker-org/rocker-versioned/issues/138}}.
-The main challenges are finding a suitable base image with a matching OS version, or making the adjustments to compile R on more recent OS releases.
-
-## Windows Images
-\label{windows}
-
-Docker containers on the Windows operating system were originally quite cumbersome to use.
-They required an extra tool, [Docker Toolbox](https://docs.docker.com/toolbox/), which by now only exists as a legacy solution for older Windows systems.
-Docker toolbox leverages `docker-machine` (see also Section&nbsp;\nameref{interfaces} for an R package interfacing with `docker-machine`) to handle the process of creating a local virtual host which could host Docker Engine, while exposing the regular Docker CLI.
-The Docker CLI commands are forwarded to the virtual host transparently for the user.
-
-For current Windows Server (2016 and later) and Windows Desktop ([Docker Desktop](https://www.docker.com/products/docker-desktop) requires Windows 10) versions, Docker is supported natively\footnote{\href{https://www.docker.com/products/windows-containers}{https://www.docker.com/products/windows-containers}} and different base images are offered by Microsoft\footnote{\href{https://docs.microsoft.com/en-us/virtualization/windowscontainers/about/index}{https://docs.microsoft.com/en-us/virtualization/windowscontainers/about/index}}.
-The Docker CLI can be used in just the same way as on other operating systems but not every base image is supported on every Windows host\footnote{\href{https://docs.microsoft.com/en-us/virtualization/windowscontainers/deploy-containers/version-compatibility}{https://docs.microsoft.com/en-us/virtualization/windowscontainers/deploy-containers/version-compatibility}}.
-On Docker Desktop for Windows, the user can run both Linux-based and Windows-based containers, but only one of the Docker daemons can be used at a time\footnote{See "Switch between Windows and Linux containers" on \href{https://docs.docker.com/docker-for-windows/}{https://docs.docker.com/docker-for-windows/}.}.
-
-**rocker-win** ([https://github.com/nuest/rocker-win](https://github.com/nuest/rocker-win)) is a proof of concept for running R in Windows-based containers \citep{nust_rocker-win_2019}.
-It provides selected R versions using three different base images using `microsoft/windowsservercore` and `mcr.microsoft.com/windows/servercore`.
-These base images match different Windows versions: Windows Server 2019, Windows Server 2016, and Windows Server, version 1803.
-The images are built using two CI services, [Travis CI](https://docs.travis-ci.com/user/reference/windows/) and [Appveyor](https://www.appveyor.com/docs/windows-images-software/), which provide the different Windows versions and therefore support different base images, and published automatically [on Docker&nbsp;Hub](https://hub.docker.com/r/nuest/rocker-win).
-The images are built from manually maintained seperate `Dockerfile`s, are tagged with both Windows-variant and R version, e.g., `nuest/rocker-win:ltsc2019-3.6.2` or `nuest/rocker-win:1803-latest`, and run the `R.exe` process by default.
-All these images can be run on Docker for Desktop on Windows 10.
-New R versions are only added on demand.
-The explorative rocker-win project demonstrates the variety of Windows on Docker, which is potentially confusing, e.g., because of the different base images.
-The proof of concept provides a starting point for any Windows-constrained user to leverage containerisation, including examples for apps using \CRANpkg{plumber}, \CRANpkg{Shiny}, or R packages with system dependencies such as \CRANpkg{sf}.
-The latter is quite comfortable and fast actually, compared to the default installation from source on Linux, because CRAN ships pre-compiled binaries for Windows.
-However, compared to the various Linux images, the footprint of Windows images is quite large.
-
-For Windows, just as originally for Linux, the cloud use cases drive the development and Docker and Microsoft collaborate closely\footnote{\href{https://www.docker.com/partners/microsoft}{https://www.docker.com/partners/microsoft}}.
-In most cases, individual developers or researchers worrying about reproducibility, will prefer the more widely and host independent Linux-based containers.
-Naturally, a Windows license is required for the host machine, and the licensing impacts the potential to build images in clouds or redistribute exported images as archive files.
-But the developments in the `rocker-win` prototype show that containers can also be used for R workflows depending on Windows-only tools, for leveraging an existing Windows Server-based infrastructure where policies can otherwise not be met, or for streamlining interactions with system operating staff that runs Windows-based containers themselves.
-
-## Non-Rocker Linux images
-\label{nondebian}
-
-\label{rhub}
-The **R-hub** project provides _"a collection of services to help R package development"_, with the package builder as the most prominent one\citep{r-hub_docs_2019}.
-The builder allows R package developers to check their R package on different platforms and R versions using a web form or the package \CRANpkg{rhub} \citep{csardi_rhub_2019}.
-The builder uses Docker containers to conduct these checks, taking advantages of their well-defined environments and sandboxing.
-The `Dockerfile`s and images are published on Docker&nbsp;Hub\footnote{\href{https://hub.docker.com/u/rhub}{https://hub.docker.com/u/rhub}} and GitHub\footnote{\href{https://github.com/r-hub/rhub-linux-builders}{https://github.com/r-hub/rhub-linux-builders}} respectively.
-The images comprise release and development builds of R running in the base images of the [Debian](https://www.debian.org/), [Ubuntu](https://ubuntu.com/), [Fedora](https://getfedora.org/), and [CentOS](https://centos.org/) linux distributions (Arch Linux is under development\footnote{\href{https://github.com/r-hub/rhub-linux-builders/pull/41}{https://github.com/r-hub/rhub-linux-builders/pull/41}}, see `rhub::local_check_linux_images()` for a full list), and are intended for debugging (see Section&nbsp;\nameref{development}) not for reproducible workflows.
-_The rationale_ of not re-using the Rocker image stack is the high number of operatings systems and configurations, which can be more uniformly maintained in a seperate suite of images.
-Also, a common structure across the respective image stacks of the operating systems, e.g., for `debian`, adding GCC in `debian-gcc` and then R's development version in `debian-gcc-devel`.
-The platform also has a specific user configuration differing from the setup covering typical applications of Rocker images, and the images include considerably more software, such as a fairly complete LaTeX installation, to provide an experience closer to CRAN.
-
-[**RStudio**](https://rstudio.com/) also maintains a set of [Docker images](https://github.com/rstudio/r-docker) for multiple operating systems including SUSE, Ubuntu, CentOS, and Debian.
-The image stack includes an [opinionated R installation](https://github.com/rstudio/r-builds) and ensures the R installation and profile is consistent across the different Linux distributions.
-The base image also installs R in a versioned directory and uses a minimal set of build and runtime dependencies.
-Furthermore these images are responsible for creating pre-compiled binary R packages for Linux, which supplement the binaries built by CRAN.
-These pre-compiled packages dramatically decrease the installation time of R packages on Linux and subsequently can decrease the build time of Docker images \citep{lopp_package_2019}.
-
-There are also independent projects installing R or RStudio on [**Alpine Linux**](https://www.alpinelinux.org/), but not beyond the proof of concept stage\footnote{\href{https://gitlab.com/artemklevtsov/r-alpine}{https://gitlab.com/artemklevtsov/r-alpine} or \href{https://www.github.com/velaco/alpine-r}{https://www.github.com/velaco/alpine-r} provide different use case images but are not recently maintained (development of the latter documented in \citet{ratesic_building_2018}); \href{https://github.com/cmplopes/alpine-r}{https://github.com/cmplopes/alpine-r} has only base R images, is relatively up-to-date but sparsely documented; \href{https://github.com/CenterForStatistics-UGent/mountainr}{https://github.com/CenterForStatistics-UGent/mountainr} is a ghost project with only a single short `Dockerfile`, yet it shows the easy installation of the latest R from Alpine sources}.
-In general, Alpine-based Docker images are often chosen for minimalistic and thereby small and secure containers, e.g., in hardware with limited storage.
-R-hub's proof of concept [`r-minimal`](https://github.com/r-hub/r-minimal) takes this to the extreme with compressed image size of 20MB and bespoke install scripts, e.g., to cleverly remove compilers after package installation.
-For R though, this advantage quickly deteriorates if more features are needed, e.g., if adding compilers, packages, or especially if installing powerful data science and communication tools, like RStudio or LaTeX.
-Furthermore the distribution uses the [`musl` C library](https://www.musl-libc.org/) instead of the `glibc` used in Debian/Ubuntu, for which more tooling and experiences exist\footnote{Cf.
-\href{https://github.com/rocker-org/rocker/issues/231}{https://github.com/rocker-org/rocker/issues/231}}.
-In fact, using Debian's 'slim' variant and removing capabilities, one could probalby achieve similarly small image sizes on a more common stack.
-
-**\pkg{altRnative}** ([https://github.com/ismailsunni/altRnative/](https://github.com/ismailsunni/altRnative/)) is an experimental R package for running the same code across multiple containerised versions of R, intended for comparison across operating systems and different implementations of R.
-It comes with a collection of `Dockerfile`s and corresponding images in multiple combinations, currently including, e.g., MRO, FastR, Fedora, and TERR\footnote{See `Dockerfile`s at \href{https://github.com/ismailsunni/dockeRs}{https://github.com/ismailsunni/dockeRs}.}.
-
-## Data Science images
 \label{datascience}
-
-_Data Science_ is a widely discussed topic among all academic disciplines \citep[e.g.,][]{donoho_50_2017}.
+**Data Science** is a widely discussed topic among all academic disciplines \citep[e.g.,][]{donoho_50_2017}.
 The discussions shed a light on the tools and craftspersonship behind the analysis of data with computational methods.
@@
 <!--A general description of what to consider when creating a Docker image used for data science projects in R is provided on [environments.rstudio.com](https://environments.rstudio.com/docker).-->
-This section presents exemplary Docker images and image stacks featuring R intended for Data Science.
-
-The [**Jupyter Docker Stacks**](https://github.com/jupyter/docker-stacks/) project are a set of ready-to-run Docker images containing Jupyter applications and interactive computing tools \citep{project_jupyter_jupyter_2018}.
+The remaineder of this section presents exemplary Docker images and image stacks featuring R intended for Data Science.
+The [_Jupyter Docker Stacks_](https://github.com/jupyter/docker-stacks/) project are a set of ready-to-run Docker images containing Jupyter applications and interactive computing tools \citep{project_jupyter_jupyter_2018}.
 The `jupyter/r-notebook` image includes R and "popular packages", and naturally also the IRKernel ([https://irkernel.github.io/](https://irkernel.github.io/)), an R kernel for Jupyter so that Jupyter Notebooks can contain R code cells.
@@
 R is installed using the Conda package manager, which can manage environments for various programming languages, pinning both the R version and the versions of R packages\footnote{See \texttt{jupyter/datascience-notebook}'s \texttt{Dockerfile} at \href{https://github.com/jupyter/docker-stacks/blob/master/datascience-notebook/Dockerfile\#L47}{https://github.com/jupyter/docker-stacks/blob/master/datascience-notebook/Dockerfile\#L47}.}.
-
-**Kaggle** provides the [`gcr.io/kaggle-images/rstats`](https://hub.docker.com/r/kaggle/rstats) image (previously `kaggle/rstats`) and [corresponding `Dockerfile`](https://github.com/Kaggle/docker-rstats) for usage in their Machine Learning competitions and easy access to the associated datasets.
+_Kaggle_ provides the [`gcr.io/kaggle-images/rstats`](https://hub.docker.com/r/kaggle/rstats) image (previously `kaggle/rstats`) and [corresponding `Dockerfile`](https://github.com/Kaggle/docker-rstats) for usage in their Machine Learning competitions and easy access to the associated datasets.
 It includes machine learning libraries such as Tensorflow and Keras (see also image `rocker/ml` in Section&nbsp;\nameref{rocker-gpu}), and also configures the \CRANpkg{reticulate} package.
@@
 The final extracted image size is over 25GB, which makes it debatable if having everything available is actually convenient.
-
-```{bash kaggle_images, eval=FALSE, include=FALSE}
-docker pull gcr.io/kaggle-images/rcran
-docker pull kaggle/rcran # over 22 months old
-docker pull gcr.io/kaggle-images/rstats
-docker images | grep kaggle
-```
-
-As a further example, [**Radiant** project](https://radiant-rstats.github.io/docs/) provides several images, e.g., [`vnijs/rsm-msba-spark`](https://hub.docker.com/r/vnijs/rsm-msba-spark), for their browser-based business analytics interface based on \CRANpkg{Shiny} (`Dockerfile` [on GitHub](https://github.com/radiant-rstats/docker)) and for use in education as part of an MSc course.
+As a further example, [_Radiant_](https://radiant-rstats.github.io/docs/) project provides several images, e.g., [`vnijs/rsm-msba-spark`](https://hub.docker.com/r/vnijs/rsm-msba-spark), for their browser-based business analytics interface based on \CRANpkg{Shiny} (`Dockerfile` [on GitHub](https://github.com/radiant-rstats/docker)) and for use in education as part of an MSc course.
 As data science often applies a multitude of tools, this image favours inclusion over selection and features Python, Postgres, JupyterLab and Visual Studio Code besides R and RStudio, bringing the image size up to 9GB.
-
-```{bash radiant_images, eval=FALSE, include=FALSE}
-docker pull vnijs/rsm-msba-spark
-docker images | grep vnijs
-```
-
-**Gigantum** ([http://gigantum.com/](http://gigantum.com/) is a platform for open and decentralized data science with a focus on using automation and user-friendly tools for easy sharing of reproducible computational workflows.
+_Gigantum_ ([http://gigantum.com/](http://gigantum.com/)) is a platform for open and decentralized data science with a focus on using automation and user-friendly tools for easy sharing of reproducible computational workflows.
 Gigantum builds on the _Gigantum Client_ (running either locally or on a remote server) for development and execution of data-focused _Projects_, which can be stored and shared via the _Gigantum Hub_ or via a zipfile export.
@@
 
-# Use cases and applications
-<!-- from packages closer to Docker to second level packages to more complex use cases -->
-\label{applications}
+```{bash kaggle_images, eval=FALSE, include=FALSE}
+docker pull gcr.io/kaggle-images/rcran
+docker pull kaggle/rcran # over 22 months old
+docker pull gcr.io/kaggle-images/rstats
+docker images | grep kaggle
+```
+
+```{bash radiant_images, eval=FALSE, include=FALSE}
+docker pull vnijs/rsm-msba-spark
+docker images | grep vnijs
+```
 
@@
 
-System dependencies are the domain of Docker, but for a full description of the computing environment one must also manage the R package.
-This can be achiedved with MRAN, or with the package **\CRANpkg{renv}** ([https://rstudio.github.io/renv/](https://rstudio.github.io/renv/)), which helps users to manage the state of the R library in a reproducible way, further providing isolation and portability \citep{renv2019}.
+System dependencies are the domain of Docker, but for a full description of the computing environment one must also manage the R version and the the R packages.
+R&nbsp;versions are available via the versioned Rocker image stack \citep{RJ-2017-065}.
+[r-online](https://github.com/ColinFay/ronline) leverages these images and provides an app for helping users to detect breaking changes between different R versions, and for historic exploration of R.
+With a standalone NodeJS app or [online](https://srv.colinfay.me/r-online), the user can compare a piece of code run in two separate versions of R.
+Internally, r-online opens one or two Docker instances with the given version of R based on Rocker images, executes a given piece of code, and returns the result to the user.
+R&nbsp;package management can be achieved with MRAN, or with packages such as \CRANpkg{checkpoint} and **\CRANpkg{renv}**, which can naturally be applied within images and containers.
+For example, \CRANpkg{renv} ([https://rstudio.github.io/renv/](https://rstudio.github.io/renv/)) helps users to manage the state of the R library in a reproducible way, further providing isolation and portability \citep{renv2019}.
 While \CRANpkg{renv} does not cover system dependencies, the \CRANpkg{renv}-based environment can be transferred into a container either by restoring the environment based on the main configuration file `renv.lock` or by storing the \CRANpkg{renv}-cache on the host and not in the container \citep{ushey_using_2019}.
@@
 That can be a challenge for some users.
-The R-hub service (see Section&nbsp;\nameref{rhub}) makes it easy to ensure that no errors occur, but to fix errors a local setup is still often warranted, e.g., using the image `rocker/r-devel`, and to test packages with native code, which can make the process more complex \citep[cf.][]{eckert_building_2018}.
+The **R-hub** project provides _"a collection of services to help R package development"_, with the package builder as the most prominent one\citep{r-hub_docs_2019}.
+R-hub makes it easy to ensure that no errors occur, but to fix errors a local setup is still often warranted, e.g., using the image `rocker/r-devel`, and to test packages with native code, which can make the process more complex \citep[cf.][]{eckert_building_2018}.
 The R-hub Docker images can also be used to debug problems locally using various combinations of Linux platforms, R versions, and compilers\footnote{See \href{https://r-hub.github.io/rhub/articles/local-debugging.html}{https://r-hub.github.io/rhub/articles/local-debugging.html} and \href{https://blog.r-hub.io/2019/04/25/r-devel-linux-x86-64-debian-clang/}{https://blog.r-hub.io/2019/04/25/r-devel-linux-x86-64-debian-clang/}}.

### 2020-01-27 manuscript.Rmd
update comments

@@
 
-# inserting a table in the R Journal currently fails
-# waiting for new release of articles including this PR to be published: https://github.com/rstudio/rticles/pull/261
-# Until then, use remotes::install_github("rstudio/rticles")
+# insert table
 requireNamespace("kableExtra")

### 2020-01-27 manuscript.tex
update tex file

@@
 % !TeX root = RJwrapper.tex
-\title{The Rockerverse: Images, Packages, and Applications for Containerization
-with R}
+\title{The Rockerverse: Packages and Applications for Containerization with R}
 \author{by Daniel Nüst, Dirk Eddelbuettel, Dom Bennett, Robrecht Cannoodt, Dav Clark, Gergely Daroczi, Mark Edmondson, Colin Fay, Ellis Hughes, Sean Lopp, Ben Marwick, Heather Nolis, Jacqueline Nolis, Hong Ooi, Karthik Ram, Noam Ross, Lori Shepherd, Nitesh Turaga, Craig Willis, Nan Xiao, Charlotte Van Petegem}
@@
 projects building upon Rocker and presents the current state of R
-packages for managing Docker images and controlling containers. We also
-look beyond Rocker to other projects connecting containerisation with R,
-namely alternative suites of images. These use cases and the variety of
-applications demonstrate the power of Rocker specifically and
-containerisation in general. We identified common themes across this
-diversity: reproducible environments, scalability and efficiency, and
-portability across clouds.
+packages for managing Docker images and controlling containers. These
+use cases and the variety of applications demonstrate the power of
+Rocker specifically and containerisation in general. We identified
+common themes across this diversity: reproducible environments,
+scalability and efficiency, and portability across clouds.
 }
@@
 The article continues with a brief introduction of containerization
-basics and the Rocker~Project, followed by descriptions of projects
-developing images with R installations. Then, we present applications,
+basics and the Rocker~Project, followed by use cases and applications,
 starting with the R packages specificly for interacting with Docker,
@@
 For statistical computing and analysis centered around R, the
-Rocker~Project has provided a variety of Docker containers since its
-start in 2014 \citep{RJ-2017-065}. The Rocker~Project provides several
-lines of containers spanning to from building blocks with
+\textbf{Rocker~Project} has provided a variety of Docker containers
+since its start in 2014 \citep{RJ-2017-065}. The Rocker~Project provides
+several lines of containers spanning to from building blocks with
 \texttt{R-release} or \texttt{R-devel}, via containers with
@@
 
-\hypertarget{container-images}{%
-\section{Container images}\label{container-images}}
-
-\label{images}
-
-\hypertarget{images-for-alternative-r-engines}{%
-\subsection{Images for alternative R
-engines}\label{images-for-alternative-r-engines}}
-
-\label{alternatives}
-
-As outlined above, R is a widely-used language with a large community.
-The large number of extension packages provides access to an unrivaled
-variety of established and upcoming features. Nevertheless, special use
-cases and experimental projects in academia and industry exist to test
-new approaches, or provide features different to what the R
-implementation maintained by the
-\href{https://www.r-project.org/contributors.html}{R Core Team} and
-available via \href{https://cran.r-project.org/}{CRAN} provides. While
-the images presented in this section are far from being as vetted,
-stable, and widely used as any of the Rocker images, they demonstrate an
-important advantage of containerisation technology, namely the abiliy to
-transparently build portable stacks of open source software and make
-them easily accessible to others. All these alternative implementations
-of R are published under GPL licenses. Since all implementations of R
-claim better performance as a core motivation, a comparision based on
-Docker images, potentially leveraging the
-\href{https://docs.docker.com/config/containers/resource_constraints/}{resource
-restriction mechanisms} of Docker to level the playing field, seems
-useful future work.
-
-\textbf{Microsoft R Open} (MRO) is an R distribution formerly known as
-Revolution R Open (RRO) before Revolution Analytics was acquired by
-Microsoft. MRO is compatible with main R and its packages and it
-``{[}..{]} includes additional capabilities for improved performance,
-reproducibility, and platform support.'' \citep{microsoft_mro} Most
-notably these capabilities are the MRAN repository, which is enabled by
-default, and some bundled packages such as \CRANpkg{checkpoint} and
-specific R packages maintained by
-Microsoft\footnote{See "Bundled Packages" on \href{https://mran.microsoft.com/rro/installed}{https://mran.microsoft.com/rro/installed}.}.
-Originally the optional integration with
-\href{https://software.intel.com/en-us/mkl}{Intel® Math Kernel Library}
-(MKL) for multi-threading in linear algebra operations
-\citep{microsoft_multithread} was a distinguishing feature before it was
-made available publicly, and can be used in base R, too
-\citep{eddelbuettel_thinking_2018}.
-
-MRO does not provide official Docker images, but a set of
-community-maintained \texttt{Dockerfile}s and Docker images are provided
-by the poject \texttt{mro-docker}
-(\url{https://github.com/nuest/mro-docker}), e.g., the image
-\href{https://hub.docker.com/repository/docker/nuest/mro/}{\texttt{nuest/mro}}.
-The images are inspired by the Rocker images and can be used much in the
-same fashion, effectively a drop-in replacement allowing users to
-quickly evaluate if the benefits of MRO + Intel® MKL apply to their use
-case. Version-tagged images are provided for the latest bugfix release
-of recent R versions. Extended license information about MKL is printed
-at every startup.
-
-\textbf{Renjin} is an interpreter for the R language running on top of
-the \href{https://en.wikipedia.org/wiki/Java_virtual_machine}{Java
-Virtual Machine} (JVM) providing full two-way access between Java and R
-code \citep{wikipedia_renjin_2018}. It was developed to combine the
-benefits of R, such scripting and extension packages, with the JVM's
-advantages in the areas of security, cross-platform availability and
-integration into enterprise platforms. R extension packages need to be
-specially compiled and are distributed via the Java package manager
-\href{https://en.wikipedia.org/wiki/Apache_Maven}{Apache Maven},
-cf.~\url{http://packages.renjin.org/packages} for available packages.
-Packages are loaded on demand, i.e., at the first call to
-\texttt{library()}. Not all R packages, especially one linking to binary
-libraries, are available, e.g., \texttt{rgdal}
-\footnote{\href{http://packages.renjin.org/package/org.renjin.cran/rgdal/1.4-4/build/1}{http://packages.renjin.org/package/org.renjin.cran/rgdal/1.4-4/build/1}}.
-There are no offical Docker images for Renjin, but community-maintained
-images for selected releases only are available under
-\texttt{nuest/renjin} on Docker~Hub and GitHub at
-\url{https://hub.docker.com/r/nuest/renjin} and
-\url{https://github.com/nuest/renjin-docker} respectively. These images
-expose the command line interface of Renjin in a similar fashion as
-Rocker images and allow an easy evaluation of Renjin's suitability, but
-are not intended for production use.
-
-\textbf{pqR} (\url{http://www.pqr-project.org/}) is a ``a pretty quick
-version of R''. \emph{pqR} attempts to improve R on some opinionated
-issues in the R language and is the basis for experimental features,
-e.g., automatic
-diffentiation\footnote{\href{https://riotworkshop.github.io/abstracts/riot-2019-pqr.txt}{https://riotworkshop.github.io/abstracts/riot-2019-pqr.txt}}.
-The source code development
-\href{https://github.com/radfordneal/pqR/}{on GitHub} is a one man
-project and it does not provide any Docker images. But especially
-disruptive approaches may contribute to the development of the R
-ecosystem, so the \texttt{nuest/pqr} image was independently created,
-see Docker Hub at \url{https://hub.docker.com/r/nuest/pqr/} and GitHub
-at \url{https://github.com/nuest/pqr-docker}.
-
-\textbf{FastR} (\url{https://github.com/oracle/fastr}) is ``A
-high-performance implementation of the R programming language, built on
-GraalVM'' \citep{oracle_labs_oraclefastr_2020}. It is developed by
-Oracle, connects R to the GraalVM ecosystem
-\citep{wikipedia_graalvm_2019}, and also claims superior performance but
-also targets full compatibility with base R
-\citep{oracle_labs_oraclefastr_2020}. There are no offical Docker images
-provided, but indepedently maintained experimental images and
-\texttt{Dockerfile}s are provided by
-\href{https://github.com/nuest/fastr-docker}{nuest/fastr-docker}, e.g.,
-\href{https://cloud.docker.com/repository/docker/nuest/fastr/}{\texttt{nuest/fastr}
-on Docker Hub}.
-
-\hypertarget{bioconductor}{%
-\subsection{Bioconductor}\label{bioconductor}}
-
-\label{bioc}
-
-\emph{Bioconductor} (\url{https://bioconductor.org/}) is an open source,
-open development project for the analysis and comprehension of genomic
-data \citep{gentleman_bioconductor_2004}. The project consists of 1823 R
-software packages as of October 30th 2019, as well as packages
+\hypertarget{use-cases-and-applications}{%
+\section{Use cases and applications}\label{use-cases-and-applications}}
+
+\label{applications}
+
+\hypertarget{image-stacks-for-communities-of-practice}{%
+\subsection{Image stacks for communities of
+practice}\label{image-stacks-for-communities-of-practice}}
+
+\textbf{Bioconductor} (\url{https://bioconductor.org/}) is an open
+source, open development project for the analysis and comprehension of
+genomic data \citep{gentleman_bioconductor_2004}. The project consists
+of 1823 R software packages as of October 30th 2019, as well as packages
 containing annotation or experiment data. \emph{Bioconductor} has a
@@
 
-\hypertarget{images-for-historic-r-versions}{%
-\subsection{Images for (historic) R
-versions}\label{images-for-historic-r-versions}}
-
-\label{versions}
-
-As with any other software, each new version of R comes with its share
-of changes. Some are breaking changes, some are not, but the fact is
-that running a piece of code in a given version might give different
-results from another version can be an issue when it comes to
-reproducibility of workflows and stability of applications. For example,
-think about R random seed: starting with \texttt{R\ 3.6.0}, running
-\texttt{set.seed(2811);\ sample(1:1000,\ 2)} will not give the same
-result as if it was run inside an older version of R. That might seem
-trivial, but all the code using the random number generator will not be
-exactly reproducible after this breaking change. Therefore, controlling
-the version of software is most crucial for reproducible research
-\citep[e.g.][]{boettiger_introduction_2015}.
-
-Containers are perfectly suited to capture a specific configuration of a
-computing environment to prohibit such problems. The Rocker images
-provides the \emph{versioned stack} for different the version of R since
-the project inception, which use the \emph{then-current} stable Debian
-base image \citep[cf.][]{RJ-2017-065}. Based on custom build phase
-hooks\footnote{\href{https://docs.docker.com/docker-hub/builds/advanced/}{https://docs.docker.com/docker-hub/builds/advanced/}},
-i.e., small shell scripts executed at different phases in the automated
-build of Rocker images, there are also semantic version tags for the
-most recent freezed (i.e., using MRAN) versions with only the major and
-minor
-version\footnote{See \href{https://github.com/rocker-org/rocker-versioned/blob/master/VERSIONS.md}{https://github.com/rocker-org/rocker-versioned/blob/master/VERSIONS.md}, FIXME: \href{https://github.com/rocker-org/rocker-versioned/issues/42}{https://github.com/rocker-org/rocker-versioned/issues/42}.}.
-So, at the time of writing this article, \texttt{rocker/r-ver:3} and
-\texttt{3.6} are aliases for \texttt{3.6.0}, because \texttt{3.6.1} is
-the latest release. With the release of \texttt{3.6.2} and pinning of
-the MRAN version in \texttt{3.6.1}, the tags \texttt{3} and \texttt{3.6}
-will deliver the same image as \texttt{3.6.1}. These tags allow users to
-update the base image to retrieve bugfixes while reducing the risk of
-also introducing breaking changes.
-
-\href{https://github.com/ColinFay/ronline}{r-online} is an app for
-helping users to detect breaking changes between different R versions,
-and for historic exploration of R. With a standalone NodeJS app or
-\href{https://srv.colinfay.me/r-online}{online}, the user can compare a
-piece of code run in two separate versions of R. Internally, r-online
-opens one or two Docker instances with the given version of R based on
-Rocker images, executes a given piece of code, and returns the result to
-the user.
-
-A container also provides an isolated sandbox environment suitable for
-testing and evaluation, without interfering with the ``main'' working
-environment. This enables cross version testing and bugfixing. To take
-this even further, the Rocker contributors are discussing the provision
-of R versions reaching back futher than the project's own inception,
-reaching back as far as R \texttt{2.x} and \texttt{1.x}
-\footnote{\href{https://github.com/rocker-org/rocker-versioned/issues/138}{https://github.com/rocker-org/rocker-versioned/issues/138}}.
-The main challenges are finding a suitable base image with a matching OS
-version, or making the adjustments to compile R on more recent OS
-releases.
-
-\hypertarget{windows-images}{%
-\subsection{Windows Images}\label{windows-images}}
-
-\label{windows}
-
-Docker containers on the Windows operating system were originally quite
-cumbersome to use. They required an extra tool,
-\href{https://docs.docker.com/toolbox/}{Docker Toolbox}, which by now
-only exists as a legacy solution for older Windows systems. Docker
-toolbox leverages \texttt{docker-machine} (see also
-Section~\nameref{interfaces} for an R package interfacing with
-\texttt{docker-machine}) to handle the process of creating a local
-virtual host which could host Docker Engine, while exposing the regular
-Docker CLI. The Docker CLI commands are forwarded to the virtual host
-transparently for the user.
-
-For current Windows Server (2016 and later) and Windows Desktop
-(\href{https://www.docker.com/products/docker-desktop}{Docker Desktop}
-requires Windows 10) versions, Docker is supported
-natively\footnote{\href{https://www.docker.com/products/windows-containers}{https://www.docker.com/products/windows-containers}}
-and different base images are offered by
-Microsoft\footnote{\href{https://docs.microsoft.com/en-us/virtualization/windowscontainers/about/index}{https://docs.microsoft.com/en-us/virtualization/windowscontainers/about/index}}.
-The Docker CLI can be used in just the same way as on other operating
-systems but not every base image is supported on every Windows
-host\footnote{\href{https://docs.microsoft.com/en-us/virtualization/windowscontainers/deploy-containers/version-compatibility}{https://docs.microsoft.com/en-us/virtualization/windowscontainers/deploy-containers/version-compatibility}}.
-On Docker Desktop for Windows, the user can run both Linux-based and
-Windows-based containers, but only one of the Docker daemons can be used
-at a
-time\footnote{See "Switch between Windows and Linux containers" on \href{https://docs.docker.com/docker-for-windows/}{https://docs.docker.com/docker-for-windows/}.}.
-
-\textbf{rocker-win} (\url{https://github.com/nuest/rocker-win}) is a
-proof of concept for running R in Windows-based containers
-\citep{nust_rocker-win_2019}. It provides selected R versions using
-three different base images using \texttt{microsoft/windowsservercore}
-and \texttt{mcr.microsoft.com/windows/servercore}. These base images
-match different Windows versions: Windows Server 2019, Windows Server
-2016, and Windows Server, version 1803. The images are built using two
-CI services,
-\href{https://docs.travis-ci.com/user/reference/windows/}{Travis CI} and
-\href{https://www.appveyor.com/docs/windows-images-software/}{Appveyor},
-which provide the different Windows versions and therefore support
-different base images, and published automatically
-\href{https://hub.docker.com/r/nuest/rocker-win}{on Docker~Hub}. The
-images are built from manually maintained seperate \texttt{Dockerfile}s,
-are tagged with both Windows-variant and R version, e.g.,
-\texttt{nuest/rocker-win:ltsc2019-3.6.2} or
-\texttt{nuest/rocker-win:1803-latest}, and run the \texttt{R.exe}
-process by default. All these images can be run on Docker for Desktop on
-Windows 10. New R versions are only added on demand. The explorative
-rocker-win project demonstrates the variety of Windows on Docker, which
-is potentially confusing, e.g., because of the different base images.
-The proof of concept provides a starting point for any
-Windows-constrained user to leverage containerisation, including
-examples for apps using \CRANpkg{plumber}, \CRANpkg{Shiny}, or R
-packages with system dependencies such as \CRANpkg{sf}. The latter is
-quite comfortable and fast actually, compared to the default
-installation from source on Linux, because CRAN ships pre-compiled
-binaries for Windows. However, compared to the various Linux images, the
-footprint of Windows images is quite large.
-
-For Windows, just as originally for Linux, the cloud use cases drive the
-development and Docker and Microsoft collaborate
-closely\footnote{\href{https://www.docker.com/partners/microsoft}{https://www.docker.com/partners/microsoft}}.
-In most cases, individual developers or researchers worrying about
-reproducibility, will prefer the more widely and host independent
-Linux-based containers. Naturally, a Windows license is required for the
-host machine, and the licensing impacts the potential to build images in
-clouds or redistribute exported images as archive files. But the
-developments in the \texttt{rocker-win} prototype show that containers
-can also be used for R workflows depending on Windows-only tools, for
-leveraging an existing Windows Server-based infrastructure where
-policies can otherwise not be met, or for streamlining interactions with
-system operating staff that runs Windows-based containers themselves.
-
-\hypertarget{non-rocker-linux-images}{%
-\subsection{Non-Rocker Linux images}\label{non-rocker-linux-images}}
-
-\label{nondebian}
-
-\label{rhub} The \textbf{R-hub} project provides \emph{``a collection of
-services to help R package development''}, with the package builder as
-the most prominent one\citep{r-hub_docs_2019}. The builder allows R
-package developers to check their R package on different platforms and R
-versions using a web form or the package \CRANpkg{rhub}
-\citep{csardi_rhub_2019}. The builder uses Docker containers to conduct
-these checks, taking advantages of their well-defined environments and
-sandboxing. The \texttt{Dockerfile}s and images are published on
-Docker~Hub\footnote{\href{https://hub.docker.com/u/rhub}{https://hub.docker.com/u/rhub}}
-and
-GitHub\footnote{\href{https://github.com/r-hub/rhub-linux-builders}{https://github.com/r-hub/rhub-linux-builders}}
-respectively. The images comprise release and development builds of R
-running in the base images of the
-\href{https://www.debian.org/}{Debian},
-\href{https://ubuntu.com/}{Ubuntu},
-\href{https://getfedora.org/}{Fedora}, and
-\href{https://centos.org/}{CentOS} linux distributions (Arch Linux is
-under
-development\footnote{\href{https://github.com/r-hub/rhub-linux-builders/pull/41}{https://github.com/r-hub/rhub-linux-builders/pull/41}},
-see \texttt{rhub::local\_check\_linux\_images()} for a full list), and
-are intended for debugging (see Section~\nameref{development}) not for
-reproducible workflows. \emph{The rationale} of not re-using the Rocker
-image stack is the high number of operatings systems and configurations,
-which can be more uniformly maintained in a seperate suite of images.
-Also, a common structure across the respective image stacks of the
-operating systems, e.g., for \texttt{debian}, adding GCC in
-\texttt{debian-gcc} and then R's development version in
-\texttt{debian-gcc-devel}. The platform also has a specific user
-configuration differing from the setup covering typical applications of
-Rocker images, and the images include considerably more software, such
-as a fairly complete LaTeX installation, to provide an experience closer
-to CRAN.
-
-\href{https://rstudio.com/}{\textbf{RStudio}} also maintains a set of
-\href{https://github.com/rstudio/r-docker}{Docker images} for multiple
-operating systems including SUSE, Ubuntu, CentOS, and Debian. The image
-stack includes an \href{https://github.com/rstudio/r-builds}{opinionated
-R installation} and ensures the R installation and profile is consistent
-across the different Linux distributions. The base image also installs R
-in a versioned directory and uses a minimal set of build and runtime
-dependencies. Furthermore these images are responsible for creating
-pre-compiled binary R packages for Linux, which supplement the binaries
-built by CRAN. These pre-compiled packages dramatically decrease the
-installation time of R packages on Linux and subsequently can decrease
-the build time of Docker images \citep{lopp_package_2019}.
-
-There are also independent projects installing R or RStudio on
-\href{https://www.alpinelinux.org/}{\textbf{Alpine Linux}}, but not
-beyond the proof of concept
-stage\footnote{\href{https://gitlab.com/artemklevtsov/r-alpine}{https://gitlab.com/artemklevtsov/r-alpine} or \href{https://www.github.com/velaco/alpine-r}{https://www.github.com/velaco/alpine-r} provide different use case images but are not recently maintained (development of the latter documented in \citet{ratesic_building_2018}); \href{https://github.com/cmplopes/alpine-r}{https://github.com/cmplopes/alpine-r} has only base R images, is relatively up-to-date but sparsely documented; \href{https://github.com/CenterForStatistics-UGent/mountainr}{https://github.com/CenterForStatistics-UGent/mountainr} is a ghost project with only a single short `Dockerfile`, yet it shows the easy installation of the latest R from Alpine sources}.
-In general, Alpine-based Docker images are often chosen for minimalistic
-and thereby small and secure containers, e.g., in hardware with limited
-storage. R-hub's proof of concept
-\href{https://github.com/r-hub/r-minimal}{\texttt{r-minimal}} takes this
-to the extreme with compressed image size of 20MB and bespoke install
-scripts, e.g., to cleverly remove compilers after package installation.
-For R though, this advantage quickly deteriorates if more features are
-needed, e.g., if adding compilers, packages, or especially if installing
-powerful data science and communication tools, like RStudio or LaTeX.
-Furthermore the distribution uses the
-\href{https://www.musl-libc.org/}{\texttt{musl} C library} instead of
-the \texttt{glibc} used in Debian/Ubuntu, for which more tooling and
-experiences
-exist\footnote{Cf. \href{https://github.com/rocker-org/rocker/issues/231}{https://github.com/rocker-org/rocker/issues/231}}.
-In fact, using Debian's `slim' variant and removing capabilities, one
-could probalby achieve similarly small image sizes on a more common
-stack.
-
-\textbf{\pkg{altRnative}}
-(\url{https://github.com/ismailsunni/altRnative/}) is an experimental R
-package for running the same code across multiple containerised versions
-of R, intended for comparison across operating systems and different
-implementations of R. It comes with a collection of \texttt{Dockerfile}s
-and corresponding images in multiple combinations, currently including,
-e.g., MRO, FastR, Fedora, and
-TERR\footnote{See `Dockerfile`s at \href{https://github.com/ismailsunni/dockeRs}{https://github.com/ismailsunni/dockeRs}.}.
-
-\hypertarget{data-science-images}{%
-\subsection{Data Science images}\label{data-science-images}}
-
-\label{datascience}
-
-\emph{Data Science} is a widely discussed topic among all academic
-disciplines \citep[e.g.,][]{donoho_50_2017}. The discussions shed a
-light on the tools and craftspersonship behind the analysis of data with
-computational methods. The practice of Data Science often involves a
-combination tools and software stacks and requires a cross-cutting
-skillset. This complexity and an inherent concern for openness and
-reproducibility in the Data Science community lead to Docker being used
-widely. This section presents exemplary Docker images and image stacks
-featuring R intended for Data Science.
-
-The \href{https://github.com/jupyter/docker-stacks/}{\textbf{Jupyter
-Docker Stacks}} project are a set of ready-to-run Docker images
-containing Jupyter applications and interactive computing tools
+\label{datascience} \textbf{Data Science} is a widely discussed topic
+among all academic disciplines \citep[e.g.,][]{donoho_50_2017}. The
+discussions shed a light on the tools and craftspersonship behind the
+analysis of data with computational methods. The practice of Data
+Science often involves a combination tools and software stacks and
+requires a cross-cutting skillset. This complexity and an inherent
+concern for openness and reproducibility in the Data Science community
+lead to Docker being used widely. The remaineder of this section
+presents exemplary Docker images and image stacks featuring R intended
+for Data Science. The
+\href{https://github.com/jupyter/docker-stacks/}{\emph{Jupyter Docker
+Stacks}} project are a set of ready-to-run Docker images containing
+Jupyter applications and interactive computing tools
 \citep{project_jupyter_jupyter_2018}. The \texttt{jupyter/r-notebook}
@@
 packages\footnote{See \texttt{jupyter/datascience-notebook}'s \texttt{Dockerfile} at \href{https://github.com/jupyter/docker-stacks/blob/master/datascience-notebook/Dockerfile\#L47}{https://github.com/jupyter/docker-stacks/blob/master/datascience-notebook/Dockerfile\#L47}.}.
-
-\textbf{Kaggle} provides the
+\emph{Kaggle} provides the
 \href{https://hub.docker.com/r/kaggle/rstats}{\texttt{gcr.io/kaggle-images/rstats}}
@@
 The final extracted image size is over 25GB, which makes it debatable if
-having everything available is actually convenient.
-
-As a further example,
-\href{https://radiant-rstats.github.io/docs/}{\textbf{Radiant} project}
-provides several images, e.g.,
+having everything available is actually convenient. As a further
+example, \href{https://radiant-rstats.github.io/docs/}{\emph{Radiant}}
+project provides several images, e.g.,
 \href{https://hub.docker.com/r/vnijs/rsm-msba-spark}{\texttt{vnijs/rsm-msba-spark}},
@@
 features Python, Postgres, JupyterLab and Visual Studio Code besides R
-and RStudio, bringing the image size up to 9GB.
-
-\textbf{Gigantum} (\url{http://gigantum.com/} is a platform for open and
-decentralized data science with a focus on using automation and
-user-friendly tools for easy sharing of reproducible computational
-workflows. Gigantum builds on the \emph{Gigantum Client} (running either
-locally or on a remote server) for development and execution of
-data-focused \emph{Projects}, which can be stored and shared via the
-\emph{Gigantum Hub} or via a zipfile export. The Client is a user
-friendly interface to a backend using Docker containers to package,
-build, and run Gigantum projects. It is configured to use a default set
-of Docker base images (\url{https://github.com/gigantum/base-images}),
-and users are able to define and configure their own custom images. The
-available images include two with R based on Ubuntu Linux have the
+and RStudio, bringing the image size up to 9GB. \emph{Gigantum}
+(\url{http://gigantum.com/}) is a platform for open and decentralized
+data science with a focus on using automation and user-friendly tools
+for easy sharing of reproducible computational workflows. Gigantum
+builds on the \emph{Gigantum Client} (running either locally or on a
+remote server) for development and execution of data-focused
+\emph{Projects}, which can be stored and shared via the \emph{Gigantum
+Hub} or via a zipfile export. The Client is a user friendly interface to
+a backend using Docker containers to package, build, and run Gigantum
+projects. It is configured to use a default set of Docker base images
+(\url{https://github.com/gigantum/base-images}), and users are able to
+define and configure their own custom images. The available images
+include two with R based on Ubuntu Linux have the
 \href{https://launchpad.net/~marutter/+archive/ubuntu/c2d4u3.5/}{\texttt{c2d4u}
@@
 
-\hypertarget{use-cases-and-applications}{%
-\section{Use cases and applications}\label{use-cases-and-applications}}
-
-\label{applications}
-
 \hypertarget{interfaces-for-docker-in-r}{%
@@
 the following examples show the power of automation when environments
-become too cumbersome or acquiring the skills is not possible in due
-course. Especially \emph{version pinning}, with packages \pkg{remotes}
-and \pkg{versions} for R or by using MRAN, and with system package
-managers for different operating systems, can greatly increase the
-reproducibility of built images and are common approaches.
+become too cumbersome. Especially \emph{version pinning}, with packages
+\pkg{remotes} and \pkg{versions} for R or by using MRAN, and with system
+package managers for different operating systems, can greatly increase
+the reproducibility of built images and are common approaches.
 
@@
 
-The package \textbf{\CRANpkg{renv}}
-(\url{https://rstudio.github.io/renv/}) helps users to manage the state
-of the R library in a reproducible way, further providing isolation and
-portability \citep{renv2019}. The package does not cover system
-dependencies, though, the \CRANpkg{renv}-based environment can be
-transferred into a container either by restoring the environment based
-on the main configuration file \texttt{renv.lock} or by storing the
-\CRANpkg{renv}-cache on the host and not in the container
-\citep{ushey_using_2019}.
+System dependencies are the domain of Docker, but for a full description
+of the computing environment one must also manage the R version and the
+the R packages. R~versions are available via the versioned Rocker image
+stack \citep{RJ-2017-065}.
+\href{https://github.com/ColinFay/ronline}{r-online} leverages these
+images and provides an app for helping users to detect breaking changes
+between different R versions, and for historic exploration of R. With a
+standalone NodeJS app or
+\href{https://srv.colinfay.me/r-online}{online}, the user can compare a
+piece of code run in two separate versions of R. Internally, r-online
+opens one or two Docker instances with the given version of R based on
+Rocker images, executes a given piece of code, and returns the result to
+the user. R~package management can be achieved with MRAN, or with
+packages such as \CRANpkg{checkpoint} and \textbf{\CRANpkg{renv}}, which
+can naturally be applied within images and containers. For example,
+\CRANpkg{renv} (\url{https://rstudio.github.io/renv/}) helps users to
+manage the state of the R library in a reproducible way, further
+providing isolation and portability \citep{renv2019}. While
+\CRANpkg{renv} does not cover system dependencies, the
+\CRANpkg{renv}-based environment can be transferred into a container
+either by restoring the environment based on the main configuration file
+\texttt{renv.lock} or by storing the \CRANpkg{renv}-cache on the host
+and not in the container \citep{ushey_using_2019}. With both the system
+dependencies and R packages consciously managed in a Docker image, users
+can start using containers as the \emph{only} environment for their
+workflows, which allows them to work indepenently of physical
+computers\footnote{Allowing them to be digital "nomads", cf. J.~Bryan's \href{https://github.com/jennybc/docker-why}{https://github.com/jennybc/docker-why}.}
+and to assert a specific degree of confidence in the stability of a
+developed software
+\citep[cf. \texttt{README.Rmd} in][]{marwick_research_2017}.
 
@@
 credit\footnote{See \emph{"Deploying a prediction service with Plumber"} vignette for details:  \href{https://cran.r-project.org/web/packages/AzureContainers/vignettes/vig01_plumber_deploy.html}{https://cran.r-project.org/web/packages/AzureContainers/vignettes/vig01\_plumber\_deploy.html}.}.
+\href{https://www.heroku.com/}{Heroku} is a further cloud platform as a
+service provider, who supports container-based applications.
+\texttt{heroku-docker-r}
+(\url{https://github.com/virtualstaticvoid/heroku-docker-r}) is an
+independent project providing a template for deploying R applications
+based on Heroku's image stack, including multiple examples for
+interfacing R with other programming languages. Yet the approach
+requires a manual management of the computing environment.
 
@@
 sandboxing can greatly improve the process. First, \textbf{prepared
-environments for teaching} are especially helpful for courses that
-require access to a relatively complex setup of software tools, e.g.,
-database systems. R is very useful tool for interfacing with databases,
-because almost every open source and proprietary database system has an
-R package that allows users to connect and interact with the it. This
-flexibility is even broadened by \CRANpkg{DBI}, which allows to create a
-common API for interfacing these databases, or the \CRANpkg{dbplyr}
-package, which runs \CRANpkg{dplyr} code straight against the database
-as queries. But learning and teaching these tools comes with the cost of
-deploying or having access to an environment with the software and
-drivers installed. For people teaching R, it can become a barrier if
-they need to install local versions of database drivers, or to connect
-to remote instances which might or might not be made available by IT
-services. Giving access to a sandbox for the most common database
-environments is the idea behind
-\href{https://github.com/ColinFay/r-db}{\texttt{r-db}}, a Docker image
-that contains everything needed to connect to a database from R.
+environments for teaching} are especially helpful for (a) introductory
+courses where the first step of installation and configuration can be
+hurdles for students \citep{cetinkaya-rundel_infrastructure_2018}, and
+(b) courses that require access to a relatively complex setup of
+software tools, e.g., database systems.
+\citet{cetinkaya-rundel_infrastructure_2018} describe how a Docker-based
+deployment of RStudio (i) avoided problems with troubleshooting
+individual students computers and greatly increase engagement through
+very quickly showing tangible outcomes, e.g., a visualisation, and (ii)
+reduced demand on teaching and IT staff. Each student get's acces to a
+personal RStudio instance running in a container after authentication
+with the university login, which gives the benefits of sandboxing and
+possibility of limiting resources.
+\citet{cetinkaya-rundel_infrastructure_2018} found that for the courses
+at hand, actual usage of the UI is intermittent so a single cloud based
+VM with 4~cores and 28GB~RAM sufficed for over 100 containers. An
+example for mitigating \emph{complex setups} is teaching databases. R is
+very useful tool for interfacing with databases, because almost every
+open source and proprietary database system has an R package that allows
+users to connect and interact with the it. This flexibility is even
+broadened by \CRANpkg{DBI}, which allows to create a common API for
+interfacing these databases, or the \CRANpkg{dbplyr} package, which runs
+\CRANpkg{dplyr} code straight against the database as queries. But
+learning and teaching these tools comes with the cost of deploying or
+having access to an environment with the software and drivers installed.
+For people teaching R, it can become a barrier if they need to install
+local versions of database drivers, or to connect to remote instances
+which might or might not be made available by IT services. Giving access
+to a sandbox for the most common database environments is the idea
+behind \href{https://github.com/ColinFay/r-db}{\texttt{r-db}}, a Docker
+image that contains everything needed to connect to a database from R.
 Notably, with \texttt{r-db}, the users don't have to install complex
@@
 to run inside containers for high
-reproducibility\footnote{See for example \href{https://github.com/joelnitta/pleurosoriopsis}{https://github.com/joelnitta/pleurosoriopsis} or \href{https://gitlab.com/ecohealthalliance/drake-gitlab-docker-example}{https://gitlab.com/ecohealthalliance/drake-gitlab-docker-example}, the latter even running in a CI, see~\nameref{cicd}.},
+reproducibility\footnote{See for example \href{https://github.com/joelnitta/pleurosoriopsis}{https://github.com/joelnitta/pleurosoriopsis} or \href{https://gitlab.com/ecohealthalliance/drake-gitlab-docker-example}{https://gitlab.com/ecohealthalliance/drake-gitlab-docker-example}, the latter even running in a continuous integration platform (cf.~\nameref{development}.},
 but also how to use \CRANpkg{future} package's function
@@
 existing pipelines. These breakages can make the original the results
-irreproducible and not extandable. The chance of a substantial
-disruption like this is high in a multi-year research project where key
-pieces of third-party software may have several major updates over the
-duration of the project. The classical ``paper'' article is insufficient
-to adequately communicate the knowledge behind such research projects
+irreproducible and not ¸ The chance of a substantial disruption like
+this is high in a multi-year research project where key pieces of
+third-party software may have several major updates over the duration of
+the project. The classical ``paper'' article is insufficient to
+adequately communicate the knowledge behind such research projects
 \citep[cf.][]{donoho_invitation_2010,marwick_how_2015}.
@@
 give the researcher an isolated environment to assemble these research
-pipelines with specific versions of software. Research workflows in
-containers are safe from contamination from other activities occuring on
-the researcher's computer, for example the installation of newest
-version of packages for teaching demonstrations. Given the users in this
-scenario, i.e., often academics with limited formal software development
-training, templates and assistance with containers around research
-compendia is essential. In many fields we see that a typical unit of
-research for a container is a research report or journal article, where
-the container holds the compendium, or self-contained set of data (or
-connections to data elsewhere) and code files needed to fully reproduce
-the article \citep{marwick_packaging_2018}. The package \pkg{rrtools}
+pipelines with specific versions of software to minimize problems with
+breaking changes and make workflows easier to share
+\citep[cf.][]{boettiger_introduction_2015,marwick_packaging_2018}.
+Research workflows in containers are safe from contamination from other
+activities occuring on the researcher's computer, for example the
+installation of newest version of packages for teaching demonstrations.
+Given the users in this scenario, i.e., often academics with limited
+formal software development training, templates and assistance with
+containers around research compendia is essential. In many fields we see
+that a typical unit of research for a container is a research report or
+journal article, where the container holds the compendium, or
+self-contained set of data (or connections to data elsewhere) and code
+files needed to fully reproduce the article
+\citep{marwick_packaging_2018}. The package \pkg{rrtools}
 (\url{https://github.com/benmarwick/rrtools}) provides a template and
@@
 C++, etc.) in Docker images, which can be executed from R.
-\emph{dynverse} uses a CI (cf.~\nameref{cicd}) to build Rocker-derived
-containers, test them, and push them to Docker~Hub if the tests succeed.
+\emph{dynverse} uses a CI platform (see~\nameref{ci}) to build
+Rocker-derived images, test them, and publish them on Docker~Hub if the
+tests succeed.
 
@@
 a \texttt{Dockerfile} are the starting point for rebuilding the
-environment. Further safeguarding practices are currently under
-development or not part of common practice yet, such as preservation of
-images \citep{emsley_framework_2018} and storing them alongside
+environment. When combinded with CI (see~\nameref{ci}), a research
+compendium set-up can enable \emph{continuous analysis} with easier
+verification of reproducibility and audits trails
+\citep{beaulieu-jones_reproducibility_2017}.
+
+Further safeguarding practices are currently under development or not
+part of common practice yet, such as preservation of images
+\citep{emsley_framework_2018} and storing them alongside
 \texttt{Dockerfile}s \citep[cf.][]{nust_opening_2017}, or pinning of
@@
 the development version of R, which must be compiled locally. That can
-be a challenge for some users. The R-hub service (see
-Section~\nameref{rhub}) makes it easy to ensure that no errors occur,
-but to fix errors a local setup is still often warranted, e.g., using
-the image \texttt{rocker/r-devel}, and to test packages with native
-code, which can make the process more complex
-\citep[cf.][]{eckert_building_2018}. The R-hub Docker images can also be
-used to debug problems locally using various combinations of Linux
-platforms, R versions, and
+be a challenge for some users. The \textbf{R-hub} project provides
+\emph{``a collection of services to help R package development''}, with
+the package builder as the most prominent one\citep{r-hub_docs_2019}.
+R-hub makes it easy to ensure that no errors occur, but to fix errors a
+local setup is still often warranted, e.g., using the image
+\texttt{rocker/r-devel}, and to test packages with native code, which
+can make the process more complex \citep[cf.][]{eckert_building_2018}.
+The R-hub Docker images can also be used to debug problems locally using
+various combinations of Linux platforms, R versions, and
 compilers\footnote{See \href{https://r-hub.github.io/rhub/articles/local-debugging.html}{https://r-hub.github.io/rhub/articles/local-debugging.html} and \href{https://blog.r-hub.io/2019/04/25/r-devel-linux-x86-64-debian-clang/}{https://blog.r-hub.io/2019/04/25/r-devel-linux-x86-64-debian-clang/}}.
@@
 
-Fifth, Docker images can be used \textbf{on CI platforms} to streamline
-the testing of packages. \citet{ye_docker_2019} describes how they speed
-up the process of testing by running tasks on
-\href{https://travis-ci.org/}{Travis~CI} within a container using
-\texttt{docker\ exec}, e.g., the package check or rendering of
-documentation. \citet{cardozo_faster_2018} saved time, also on
-Travis~CI, by re-using the testing image as the base for an image
-intended for publication on Docker~Hub.
+\label{ci} Fifth, Docker images can be used \textbf{on continuous
+integration (CI) platforms} to streamline the testing of packages.
+\citet{ye_docker_2019} describes how they speed up the process of
+testing by running tasks on \href{https://travis-ci.org/}{Travis~CI}
+within a container using \texttt{docker\ exec}, e.g., the package check
+or rendering of documentation. \citet{cardozo_faster_2018} saved time,
+also on Travis~CI, by re-using the testing image as the base for an
+image intended for publication on Docker~Hub.
 \href{https://github.com/ColinFay/r-ci}{\texttt{r-ci}} is in turn used
@@
 example by exposing containerisation benefits via simple user interfaces
-in the users' IDE, could be an important next step. Currently
-containerisation happens more in the background at the system level.
-Furthermore, a maturing of the Rockerverse packages for managing
-containers may lead to their adoption where currently manual coding is
-required, e.g.~in the case of \CRANpkg{RSelentium} or \CRANpkg{drake}
-(see Sections~\nameref{rselenium} and \nameref{drake} respectively). In
+in the users' IDE, could be an important next step since currently
+containerisation happens more in the background for UI-based development
+(e.g., a \texttt{rocker/rstudio} image in the cloud). Furthermore, a
+maturing of the Rockerverse packages for managing containers may lead to
+their adoption where currently manual coding is required, e.g.~in the
+case of \CRANpkg{RSelentium} or \CRANpkg{drake} (see
+Sections~\nameref{rselenium} and \nameref{drake} respectively). In
 specific cases, e.g., for \CRANpkg{analogsea}, the interaction with the

### 2020-01-28 manuscript.Rmd
reorder use cases, make R packages own chapter
also remove some unneeded fields in bib file
more explicit package imports, set locale to fix encoding warning in tex

@@
 ```{r setup, include=FALSE}
-library(tidyverse)
+Sys.setlocale('LC_ALL','C')
 ```
@@
 
-# Use cases and applications
-<!-- from packages closer to Docker to second level packages to more complex use cases -->
-\label{applications}
-
-## Image stacks for communities of practice
-
-**Bioconductor** ([https://bioconductor.org/](https://bioconductor.org/)) is an open source, open development project for the analysis and comprehension of genomic data \citep{gentleman_bioconductor_2004}.
-The project consists of 1823 R software packages as of October 30th 2019, as well as packages containing annotation or experiment data.
-_Bioconductor_ has a semi-annual release cycle, each release is associated with a particular version of R, and Docker images are provided for current and past versions of _Bioconductor_ for convenience and reproducibility.
-All images, included  are described on the _Bioconductor_ web site (see [https://bioconductor.org/help/docker/](https://bioconductor.org/help/docker/)), created with `Dockerfile`s maintained on GitHub, and distributed through Docker&nbsp;Hub\footnote{See \href{https://github.com/Bioconductor/bioconductor_docker}{https://github.com/Bioconductor/bioconductor\_docker} and \href{https://hub.docker.com/u/bioconductor}{https://hub.docker.com/u/bioconductor} respectively.}.
-_Bioconductor_'s 'base' Docker images are built on top of the `rocker/rstudio` image.
-_Bioconductor_ installs packages based on the R version in combination with the Bioconductor version, and therefore uses Bioconductor version tagging `devel` and `RELEASE_X_Y`, e.g., `RELEASE_3_10`.
-Past and current combinations of R and _Bioconductor_ will therefore be accessible via a specific image tags.
-The _Bioconductor_ `Dockerfile` selects the desired version of R from Rocker, adds required system dependencies, and uses the \CRANpkg{BiocManager} package for installing appropriate versions of _Bioconductor_ packages.
-A strength of this approach is that the responsibility for complex software configuration (including customized development) is shifted from the user to the experienced _Bioconductor_ core team.
-A recent audit of the _Bioconductor_  image stack `Dockerfile` led to the deprecation of several community-maintained images, because the numerous specific images became too hard to understand, complex to maintain, and cumbersome to extent.
-As part of the simplification, a recent innovation is to produce a `bioconductor_docker:devel` image to emulate the _Bioconductor_ nightly Linux build machine as closely as possible.
-This image contains the build system environment variables and the _system dependencies_ needed to install and check almost all (1813 out of 1823) _Bioconductor_ software packages and saves users and package developers from managing these themselves.
-Furthermore the image is configured so that `.libPaths()` has `/usr/local/lib/R/host-site-library` as the first location.
-Users mounting a location on the host file system to this location can persist installed packages across Docker sessions or image updates.
-Many R users pursue flexible work flows tailored to particular analysis needs, rather than standardized work flows.
-The new `bioconductor_docker` image is well-suited for this pattern, while `bioconductor_docker:devel` provides developers with a test environment close to _Bioconductor_'s build system.
-
-\label{datascience}
-**Data Science** is a widely discussed topic among all academic disciplines \citep[e.g.,][]{donoho_50_2017}.
-The discussions shed a light on the tools and craftspersonship behind the analysis of data with computational methods.
-The practice of Data Science often involves a combination tools and software stacks and requires a cross-cutting skillset.
-This complexity and an inherent concern for openness and reproducibility in the Data Science community lead to Docker being used widely.
-<!--A general description of what to consider when creating a Docker image used for data science projects in R is provided on [environments.rstudio.com](https://environments.rstudio.com/docker).-->
-The remaineder of this section presents exemplary Docker images and image stacks featuring R intended for Data Science.
-The [_Jupyter Docker Stacks_](https://github.com/jupyter/docker-stacks/) project are a set of ready-to-run Docker images containing Jupyter applications and interactive computing tools \citep{project_jupyter_jupyter_2018}.
-The `jupyter/r-notebook` image includes R and "popular packages", and naturally also the IRKernel ([https://irkernel.github.io/](https://irkernel.github.io/)), an R kernel for Jupyter so that Jupyter Notebooks can contain R code cells.
-R is also included in the catchall `jupyter/datascience-notebook` image\footnote{\href{https://jupyter-docker-stacks.readthedocs.io/en/latest/using/selecting.html}{https://jupyter-docker-stacks.readthedocs.io/en/latest/using/selecting.html}}.
-For example, these images allow users to quickly start a Jupyter Notebook server locally or build their own specialised images on top of stable toolsets.
-R is installed using the Conda package manager, which can manage environments for various programming languages, pinning both the R version and the versions of R packages\footnote{See \texttt{jupyter/datascience-notebook}'s \texttt{Dockerfile} at \href{https://github.com/jupyter/docker-stacks/blob/master/datascience-notebook/Dockerfile\#L47}{https://github.com/jupyter/docker-stacks/blob/master/datascience-notebook/Dockerfile\#L47}.}.
-_Kaggle_ provides the [`gcr.io/kaggle-images/rstats`](https://hub.docker.com/r/kaggle/rstats) image (previously `kaggle/rstats`) and [corresponding `Dockerfile`](https://github.com/Kaggle/docker-rstats) for usage in their Machine Learning competitions and easy access to the associated datasets.
-It includes machine learning libraries such as Tensorflow and Keras (see also image `rocker/ml` in Section&nbsp;\nameref{rocker-gpu}), and also configures the \CRANpkg{reticulate} package.
-The image uses a base image with _all packages from CRAN_, `gcr.io/kaggle-images/rcran`, which requires a Google Cloud Build as Docker&nbsp;Hub would time out\footnote{Originally, a stacked collection of over 20 images with automated builds on Docker~Hub was used, see \href{https://web.archive.org/web/20190606043353/http://blog.kaggle.com/2016/02/05/how-to-get-started-with-data-science-in-containers/}{https://web.archive.org/web/20190606043353/http://blog.kaggle.com/2016/02/05/how-to-get-started-with-data-science-in-containers/} and \href{https://hub.docker.com/r/kaggle/rcran/dockerfile}{https://hub.docker.com/r/kaggle/rcran/dockerfile}}.
-The final extracted image size is over 25GB, which makes it debatable if having everything available is actually convenient.
-As a further example, [_Radiant_](https://radiant-rstats.github.io/docs/) project provides several images, e.g., [`vnijs/rsm-msba-spark`](https://hub.docker.com/r/vnijs/rsm-msba-spark), for their browser-based business analytics interface based on \CRANpkg{Shiny} (`Dockerfile` [on GitHub](https://github.com/radiant-rstats/docker)) and for use in education as part of an MSc course.
-As data science often applies a multitude of tools, this image favours inclusion over selection and features Python, Postgres, JupyterLab and Visual Studio Code besides R and RStudio, bringing the image size up to 9GB.
-_Gigantum_ ([http://gigantum.com/](http://gigantum.com/)) is a platform for open and decentralized data science with a focus on using automation and user-friendly tools for easy sharing of reproducible computational workflows.
-Gigantum builds on the _Gigantum Client_ (running either locally or on a remote server) for development and execution of data-focused _Projects_, which can be stored and shared via the _Gigantum Hub_ or via a zipfile export.
-The Client is a user friendly interface to a backend using Docker containers to package, build, and run Gigantum projects.
-It is configured to use a default set of Docker base images ([https://github.com/gigantum/base-images](https://github.com/gigantum/base-images)), and users are able to define and configure their own custom images.
-The available images include two with R based on Ubuntu Linux have the [`c2d4u` CRAN PPA](https://launchpad.net/~marutter/+archive/ubuntu/c2d4u3.5/) pre-configured for installation of binary R packages\footnote{\href{https://docs.gigantum.com/docs/using-r}{https://docs.gigantum.com/docs/using-r}}.
-The R images vary in the included authoring environment, i.e., Jupyter in `r-tidyverse` or both Jupyter \& RStudio in `rstudio-server`.
-The independent image stack can be traced back to the Gigantum environment and its features.
-The R images are based on Gigantum's `python3-minimal` image, originally to keep the existing front-end configuration, but also to provide a consistent Python-to-R interoperability.
-The `Dockerfile`s also use build args to specify bases, for example for different version of NVIDIA CUDA for GPU processing\footnote{See \href{https://github.com/gigantum/base-images/blob/master/_templates/python3-minimal-template/Dockerfile}{https://github.com/gigantum/base-images/blob/master/\_templates/python3-minimal-template/Dockerfile} for the \texttt{Dockerfile} of \texttt{python3-minimal}.} so that appropriate GPU drivers can be enabled automatically when supported.
-Furthermore, Gigantum's focus lies on environment management via GUI and ensuring a smooth user interaction, e.g., with reliable and easy conflict detection and resolution.
-For this reason, project repositories store authoritative package information in a separate file per package, allowing Git to directly detect conflicts and changes.
-A Dockerfile is generated from this description that inherits from the specified base image, and additional custom Docker instructions may be appended by users, though Gigantum's default base images do not currently include the `littler` tool, which is used by Rocker to install packages within `Dockerfile`s.
-Because of these specifics, instructions from `rocker/r-ubuntu` could _not_ be readily re-used in this image stack (see Section&nbsp;\nameref{conclusions}).
-Both approaches enable `apt` as an installation method, and this is exposed via the GUI-based environment management,\footnote{See \href{https://docs.gigantum.com/docs/environment-management}{https://docs.gigantum.com/docs/environment-management}}.
-The image build and publication process is scripted with Python and JSON template configuration files, unlike Rocker which relies on plain `Dockerfile`s.
-A minor reason in the inception of the images were also project constraints requiring a Rocker-incompatible licensing of the `Dockerfile`s, i.e., the MIT License.
-
-```{bash kaggle_images, eval=FALSE, include=FALSE}
-docker pull gcr.io/kaggle-images/rcran
-docker pull kaggle/rcran # over 22 months old
-docker pull gcr.io/kaggle-images/rstats
-docker images | grep kaggle
-```
-
-```{bash radiant_images, eval=FALSE, include=FALSE}
-docker pull vnijs/rsm-msba-spark
-docker images | grep vnijs
-```
-
-## Interfaces for Docker in R
+# Interfaces for Docker in R
 \label{interfaces}
@@
 ```{r clientfunctionalities, echo=FALSE, results='asis', warning=FALSE, message=FALSE}
+library("tibble")
+library("tidyr")
+library("dplyr")
+
 functionalities <- tribble(
@@
 
+# Use cases and applications
+\label{applications}
+
+## Image stacks for communities of practice
+
+**Bioconductor** ([https://bioconductor.org/](https://bioconductor.org/)) is an open source, open development project for the analysis and comprehension of genomic data \citep{gentleman_bioconductor_2004}.
+The project consists of 1823 R software packages as of October 30th 2019, as well as packages containing annotation or experiment data.
+_Bioconductor_ has a semi-annual release cycle, each release is associated with a particular version of R, and Docker images are provided for current and past versions of _Bioconductor_ for convenience and reproducibility.
+All images, included  are described on the _Bioconductor_ web site (see [https://bioconductor.org/help/docker/](https://bioconductor.org/help/docker/)), created with `Dockerfile`s maintained on GitHub, and distributed through Docker&nbsp;Hub\footnote{See \href{https://github.com/Bioconductor/bioconductor_docker}{https://github.com/Bioconductor/bioconductor\_docker} and \href{https://hub.docker.com/u/bioconductor}{https://hub.docker.com/u/bioconductor} respectively.}.
+_Bioconductor_'s 'base' Docker images are built on top of the `rocker/rstudio` image.
+_Bioconductor_ installs packages based on the R version in combination with the Bioconductor version, and therefore uses Bioconductor version tagging `devel` and `RELEASE_X_Y`, e.g., `RELEASE_3_10`.
+Past and current combinations of R and _Bioconductor_ will therefore be accessible via a specific image tags.
+The _Bioconductor_ `Dockerfile` selects the desired version of R from Rocker, adds required system dependencies, and uses the \CRANpkg{BiocManager} package for installing appropriate versions of _Bioconductor_ packages.
+A strength of this approach is that the responsibility for complex software configuration (including customized development) is shifted from the user to the experienced _Bioconductor_ core team.
+A recent audit of the _Bioconductor_  image stack `Dockerfile` led to the deprecation of several community-maintained images, because the numerous specific images became too hard to understand, complex to maintain, and cumbersome to extent.
+As part of the simplification, a recent innovation is to produce a `bioconductor_docker:devel` image to emulate the _Bioconductor_ nightly Linux build machine as closely as possible.
+This image contains the build system environment variables and the _system dependencies_ needed to install and check almost all (1813 out of 1823) _Bioconductor_ software packages and saves users and package developers from managing these themselves.
+Furthermore the image is configured so that `.libPaths()` has `/usr/local/lib/R/host-site-library` as the first location.
+Users mounting a location on the host file system to this location can persist installed packages across Docker sessions or image updates.
+Many R users pursue flexible work flows tailored to particular analysis needs, rather than standardized work flows.
+The new `bioconductor_docker` image is well-suited for this pattern, while `bioconductor_docker:devel` provides developers with a test environment close to _Bioconductor_'s build system.
+
+\label{datascience}
+**Data Science** is a widely discussed topic among all academic disciplines \citep[e.g.,][]{donoho_50_2017}.
+The discussions shed a light on the tools and craftspersonship behind the analysis of data with computational methods.
+The practice of Data Science often involves a combination tools and software stacks and requires a cross-cutting skillset.
+This complexity and an inherent concern for openness and reproducibility in the Data Science community lead to Docker being used widely.
+<!--A general description of what to consider when creating a Docker image used for data science projects in R is provided on [environments.rstudio.com](https://environments.rstudio.com/docker).-->
+The remaineder of this section presents exemplary Docker images and image stacks featuring R intended for Data Science.
+The [_Jupyter Docker Stacks_](https://github.com/jupyter/docker-stacks/) project are a set of ready-to-run Docker images containing Jupyter applications and interactive computing tools \citep{project_jupyter_jupyter_2018}.
+The `jupyter/r-notebook` image includes R and "popular packages", and naturally also the IRKernel ([https://irkernel.github.io/](https://irkernel.github.io/)), an R kernel for Jupyter so that Jupyter Notebooks can contain R code cells.
+R is also included in the catchall `jupyter/datascience-notebook` image\footnote{\href{https://jupyter-docker-stacks.readthedocs.io/en/latest/using/selecting.html}{https://jupyter-docker-stacks.readthedocs.io/en/latest/using/selecting.html}}.
+For example, these images allow users to quickly start a Jupyter Notebook server locally or build their own specialised images on top of stable toolsets.
+R is installed using the Conda package manager, which can manage environments for various programming languages, pinning both the R version and the versions of R packages\footnote{See \texttt{jupyter/datascience-notebook}'s \texttt{Dockerfile} at \href{https://github.com/jupyter/docker-stacks/blob/master/datascience-notebook/Dockerfile\#L47}{https://github.com/jupyter/docker-stacks/blob/master/datascience-notebook/Dockerfile\#L47}.}.
+_Kaggle_ provides the [`gcr.io/kaggle-images/rstats`](https://hub.docker.com/r/kaggle/rstats) image (previously `kaggle/rstats`) and [corresponding `Dockerfile`](https://github.com/Kaggle/docker-rstats) for usage in their Machine Learning competitions and easy access to the associated datasets.
+It includes machine learning libraries such as Tensorflow and Keras (see also image `rocker/ml` in Section&nbsp;\nameref{rocker-gpu}), and also configures the \CRANpkg{reticulate} package.
+The image uses a base image with _all packages from CRAN_, `gcr.io/kaggle-images/rcran`, which requires a Google Cloud Build as Docker&nbsp;Hub would time out\footnote{Originally, a stacked collection of over 20 images with automated builds on Docker~Hub was used, see \href{https://web.archive.org/web/20190606043353/http://blog.kaggle.com/2016/02/05/how-to-get-started-with-data-science-in-containers/}{https://web.archive.org/web/20190606043353/http://blog.kaggle.com/2016/02/05/how-to-get-started-with-data-science-in-containers/} and \href{https://hub.docker.com/r/kaggle/rcran/dockerfile}{https://hub.docker.com/r/kaggle/rcran/dockerfile}}.
+The final extracted image size is over 25GB, which makes it debatable if having everything available is actually convenient.
+As a further example, [_Radiant_](https://radiant-rstats.github.io/docs/) project provides several images, e.g., [`vnijs/rsm-msba-spark`](https://hub.docker.com/r/vnijs/rsm-msba-spark), for their browser-based business analytics interface based on \CRANpkg{Shiny} (`Dockerfile` [on GitHub](https://github.com/radiant-rstats/docker)) and for use in education as part of an MSc course.
+As data science often applies a multitude of tools, this image favours inclusion over selection and features Python, Postgres, JupyterLab and Visual Studio Code besides R and RStudio, bringing the image size up to 9GB.
+_Gigantum_ ([http://gigantum.com/](http://gigantum.com/)) is a platform for open and decentralized data science with a focus on using automation and user-friendly tools for easy sharing of reproducible computational workflows.
+Gigantum builds on the _Gigantum Client_ (running either locally or on a remote server) for development and execution of data-focused _Projects_, which can be stored and shared via the _Gigantum Hub_ or via a zipfile export.
+The Client is a user friendly interface to a backend using Docker containers to package, build, and run Gigantum projects.
+It is configured to use a default set of Docker base images ([https://github.com/gigantum/base-images](https://github.com/gigantum/base-images)), and users are able to define and configure their own custom images.
+The available images include two with R based on Ubuntu Linux have the [`c2d4u` CRAN PPA](https://launchpad.net/~marutter/+archive/ubuntu/c2d4u3.5/) pre-configured for installation of binary R packages\footnote{\href{https://docs.gigantum.com/docs/using-r}{https://docs.gigantum.com/docs/using-r}}.
+The R images vary in the included authoring environment, i.e., Jupyter in `r-tidyverse` or both Jupyter \& RStudio in `rstudio-server`.
+The independent image stack can be traced back to the Gigantum environment and its features.
+The R images are based on Gigantum's `python3-minimal` image, originally to keep the existing front-end configuration, but also to provide a consistent Python-to-R interoperability.
+The `Dockerfile`s also use build args to specify bases, for example for different version of NVIDIA CUDA for GPU processing\footnote{See \href{https://github.com/gigantum/base-images/blob/master/_templates/python3-minimal-template/Dockerfile}{https://github.com/gigantum/base-images/blob/master/\_templates/python3-minimal-template/Dockerfile} for the \texttt{Dockerfile} of \texttt{python3-minimal}.} so that appropriate GPU drivers can be enabled automatically when supported.
+Furthermore, Gigantum's focus lies on environment management via GUI and ensuring a smooth user interaction, e.g., with reliable and easy conflict detection and resolution.
+For this reason, project repositories store authoritative package information in a separate file per package, allowing Git to directly detect conflicts and changes.
+A Dockerfile is generated from this description that inherits from the specified base image, and additional custom Docker instructions may be appended by users, though Gigantum's default base images do not currently include the `littler` tool, which is used by Rocker to install packages within `Dockerfile`s.
+Because of these specifics, instructions from `rocker/r-ubuntu` could _not_ be readily re-used in this image stack (see Section&nbsp;\nameref{conclusions}).
+Both approaches enable `apt` as an installation method, and this is exposed via the GUI-based environment management,\footnote{See \href{https://docs.gigantum.com/docs/environment-management}{https://docs.gigantum.com/docs/environment-management}}.
+The image build and publication process is scripted with Python and JSON template configuration files, unlike Rocker which relies on plain `Dockerfile`s.
+A minor reason in the inception of the images were also project constraints requiring a Rocker-incompatible licensing of the `Dockerfile`s, i.e., the MIT License.
+
+```{bash kaggle_images, eval=FALSE, include=FALSE}
+docker pull gcr.io/kaggle-images/rcran
+docker pull kaggle/rcran # over 22 months old
+docker pull gcr.io/kaggle-images/rstats
+docker images | grep kaggle
+```
+
+```{bash radiant_images, eval=FALSE, include=FALSE}
+docker pull vnijs/rsm-msba-spark
+docker images | grep vnijs
+```
+
 ## Capture and create environments
@@
 
+Community maintained images provide a solid basis for extension as part of individual requirements.
 Several second order R packages attempt to make the process of creating Docker images and using containers for specific tasks, such as running tests or rendering reproducible reports, easier.
@@
 
-## Using R to power enterprise software in production environments
-\label{enterprise}
+## Development, debugging, and testing
+\label{development}
 
-R has been historically viewed as a tool for analysis and scientific research, but not for creating software that corporations can rely on for production services.
-However, thanks to advancements in R running as a web service, along with with the ability to deploy R in Docker containers, modern enterprises are now capable of having real-time machine learning powered by R.
-A number of packages and projects enabled R to respond to client reqests over TCP/IP and local socket servers, such as \CRANpkg{Rserve}, \CRANpkg{svSocket}, [rApache](http://www.rapache.net) and more recently \CRANpkg{plumber} ([https://www.rplumber.io/](https://www.rplumber.io/)) and \pkg{RestRserve} ([http://restrserve.org](http://restrserve.org)), which even processes incoming requests in parallel with forked processes using \CRANpkg{Rserve}.
-The latter two also provide documentation for deployment with Docker or ready to use automated builds of images\footnote{See \href{https://www.rplumber.io/docs/hosting.html\#docker}{https://www.rplumber.io/docs/hosting.html\#docker}, \href{https://hub.docker.com/r/trestletech/plumber/}{https://hub.docker.com/r/trestletech/plumber/} and \href{https://hub.docker.com/r/rexyai/restrserve/}{https://hub.docker.com/r/rexyai/restrserve/}.}.
-These software allow other (remote) processes and programming languages to interact with R and to expose R-based function in a service architecture with HTTP APIs.
-APIs based on these package can be deployed with scalability and high availability using containers.
-This pattern of deploying code matches those used by software engineering services created in more established languages in the enterprise domain, such as Java or Python, and R can be used alongside those langages as a first class member of a software engineering technical stack.
+Containers can also serve as useful playgrounds to create environments ad-hoc or to provide very specific environments that are not needed or not easily available in day-to-day development for the purposes of developing R packages.
+These environments may have specific versions of R, of R extension packages, and of system libraries used by R extension packages, and all of the above in a specific combination.
 
-CARD.com implemented a web application for the optimization of the acquisition flow and the real-time analysis of debit card transactions.
-The software used \CRANpkg{Rserve} and rApache and was deployed in Docker containers.
-The R session behind \CRANpkg{Rserve} acted as a read-only in-memory database, which was extremely fast and scalabale, for the many concurrent rApache processes responding to the live-scoring requests of various divisions of the company.
-Similarly dockerized R scripts were responsible for the ETL processes and even the client-facing email, text message and push notification alerts sent in real-time based on card transactions.
-The related Docker images were made available at [https://github.com/cardcorp/card-rocker](https://github.com/cardcorp/card-rocker).
-The images extend `rocker/r-base` and additionally entailed an SSH client and a workaround for being able to mount SSH keys from the host, Pandoc, the Amazon Web Services (AWS) SDK, and Java, which is required by the AWS SDK.
-The AWS SDK allowed to run R consumers reading from real-time data processing streams of [AWS Kinesis](https://aws.amazon.com/kinesis/) \footnote{See useR!2017 talk \href{https://static.sched.com/hosted\_files/user2017/2f/AWR Kinesis at useR 2017.pdf}{"Stream processing with R in AWS"}.}.
-The applications were deployed on Amazon Elastic Container Service ([ECS](https://aws.amazon.com/ecs/)).
-The main learnings from using R in Docker was the imporance of not only pinning the R package versions via MRAN, but also moving away from Debian testing to a distribution with long-term support.
-For the use case at hand, this switch served the priority to have more control over upstream updates, and to minimize the risk of breaking the automated builds of the Docker images and production jobs.
+First, such containers can greatly facilitate **fixing bugs and quick evaluation**, because developers and users can readily start and later discard a container to investigate a bug report or try out a piece of software \citep[cf.][]{ooms_opencpu_2017}.
+The container does not affect their regular system.
+Using the Rocker images with RStudio, these disposable environments lack no development comfort (cf. Section&nbsp;\nameref{compendia}).
+\citet{ooms_opencpu_2017} describes how `docker exec` can be used to get a root shell in a container for cusotmization during software evaluation.
+\citet{eddelbuettel_debugging_2019} describes an example how a Docker container was used to debug an issue with a package only occuring with a particular version of Fortran, and using tools which are not readily available on all platforms (e.g., not on macOS).
 
-The AI @ T-Mobile team created a set of neural network machine learning natural language processing models to help customer care agents manage text-based messages for customers \citep{t-mobile_enterprise_2018}.
-For example, one model quickly identifies if a message is from a customer or not \citep[see \CRANpkg{Shiny}-based \href{https://secure.message.t-mobile.com/v1/shiny/is-customer/app/}{demo}, ][]{nolis_small_2019}, others tell which customers are likely to make a repeat purchase.
-If a data scientist creates a machine learning model and exposes it through a \CRANpkg{plumber} API, then someone else on the marketing team could write software that sends different emails depending on that real-time prediction.
-The models are convolutional neural networks that use the \CRANpkg{keras} package and run in a Rocker Docker image.
-The corresponding `Dockerfile`s are published [on GitHub](https://github.com/tmobile/r-tensorflow-api).
-Since the models power tools for agents and customers, they need to have extremely high uptime and reliability.
-The AI @ T-Mobile team found that the models performed well and today these models power real-time services that are called over a million times a day.
+Second, the strong integration of **system libraries in core packages** in the [R-spatial community](https://www.r-spatial.org/) makes containers essential for stable and proactive development of common classes for geospatial data modelling and analysis.
+For example, GDAL \citep{gdal_2019} is a crucial library in the geospatial domain.
+GDAL is a system dependency allowing R packages such as \CRANpkg{sf}, which provides the core data model for geospatial vector data, or \CRANpkg{rgdal}, to accomodate users to be able to read and write hundreds of different spatial raster and vector formats.
+\CRANpkg{sf} and \CRANpkg{rgdal} have hundreds of indirect reverse imports and dependencies and therefore the maintainers spend a lot of efforts not to break these.
+Purpose built Docker are used to prepare for upcoming releases of system libraries, individual bug reports, and for the lowest supported versions of system libraries\footnote{Cf.
+\href{https://github.com/r-spatial/sf/tree/master/inst/docker}{https://github.com/r-spatial/sf/tree/master/inst/docker}, \href{https://github.com/Nowosad/rspatial_proj6}{https://github.com/Nowosad/rspatial\_proj6}, and \href{https://github.com/r-spatial/sf/issues/1231}{https://github.com/r-spatial/sf/issues/1231}}.
+```{r geospatial_deps, eval=FALSE, include=FALSE}
+length(tools::dependsOnPkgs(c("rgdal", "sf"), dependencies = c("most"), recursive = TRUE))
+```
+
+Third, there are special purpose images for identifying problems beyond the mere R code, such as **debugging R memory problems**.
+The images significantly reduce the barrier to follow complex steps for fixing memory allocation bugs \citep[cf. Section~4.3 in][]{core_writing_1999}.
+These problems are hard to debug and critical, both because when they occur they lead to fatal crashing processes.
+[`rocker/r-devel-san`](https://github.com/rocker-org/r-devel-san) and [`rocker/r-devel-ubsan-clang`](https://github.com/rocker-org/r-devel-san-clang) are Docker images have a particularly configured version of R to trace such problems with gcc and clang compilers, respectively \citep[cf.~\CRANpkg{sanitizers} for examples,][]{eddelbuettel_sanitizers_2014}.
+The image [`wch/r-debug`](https://github.com/wch/r-debug) is a purpose built Docker image with _multiple_ instrumented builds of R, each with a different diagnostic utility activated.
+
+Fourth, containers are useful for **testing** R code during development.
+To submit a package to CRAN, an R package must work with the development version of R, which must be compiled locally.
+That can be a challenge for some users.
+The **R-hub** project provides _"a collection of services to help R package development"_, with the package builder as the most prominent one\citep{r-hub_docs_2019}.
+R-hub makes it easy to ensure that no errors occur, but to fix errors a local setup is still often warranted, e.g., using the image `rocker/r-devel`, and to test packages with native code, which can make the process more complex \citep[cf.][]{eckert_building_2018}.
+The R-hub Docker images can also be used to debug problems locally using various combinations of Linux platforms, R versions, and compilers\footnote{See \href{https://r-hub.github.io/rhub/articles/local-debugging.html}{https://r-hub.github.io/rhub/articles/local-debugging.html} and \href{https://blog.r-hub.io/2019/04/25/r-devel-linux-x86-64-debian-clang/}{https://blog.r-hub.io/2019/04/25/r-devel-linux-x86-64-debian-clang/}}.
+The images go beyond the configurations, or _flavours_, used by CRAN for checking packages\footnote{\href{https://cran.r-project.org/web/checks/check_flavors.html}{https://cran.r-project.org/web/checks/check\_flavors.html}}, e.g., with CentOS-based images, but lack a container for checking on Windows or OS X.
+The images greatly support package developers to provide support on operating systems they are not familiar.
+The package \pkg{dockertest} ([https://github.com/traitecoevo/dockertest/](https://github.com/traitecoevo/dockertest/)) is a proof of concept for automatically generating `Dockerfile`s and building images specifically to run tests\footnote{\pkg{dockertest} is not actively maintained, but mentioned still because of its interesting approach.}.
+These images are accompanied with a special launch script so the tested source code is not stored in the image but the currently checked in version from a local Git repository is cloned into the container at runtime.
+This approach clearly seperates test environment, test code, and current working copy of the code.
+\label{rselenium} Another use case where a container helps to standardise tests across operating systems is detailed the vignettes of the package \CRANpkg{RSelenium} \citep{rselenium_2019}.
+The package recommends Docker for running the [Selenium](https://selenium.dev/) Server application needed to execute test suites on browser-based user interfaces and webpages, but requires users to manually manage the Docker containers.
+
+\label{ci}
+Fifth, Docker images can be used **on continuous integration (CI) platforms** to streamline the testing of packages.
+\citet{ye_docker_2019} describes how they speed up the process of testing by running tasks on [Travis&nbsp;CI](https://travis-ci.org/) within a container using `docker exec`, e.g., the package check or rendering of documentation.
+\citet{cardozo_faster_2018} saved time, also on Travis&nbsp;CI, by re-using the testing image as the base for an image intended for publication on Docker&nbsp;Hub.
+[`r-ci`](https://github.com/ColinFay/r-ci) is in turn used with [GitLab&nbsp;CI](https://docs.gitlab.com/ee/ci/), which itself is built on top of Docker images: the user specifies a base Docker image, and the whole tests are run inside this environment.
+The `r-ci` image stack combines `rocker` versioning and a series of tools specifically designed for testing in a fixed environment with a customized list of preinstalled packages.
+Especially for long running tests or complex system dependencies, these approaches to seperate installation of build dependencies with code testing streamline the development process.
+Not due to a concern about time, but to control the environment used on a CI server, even this manuscript is rendered into a PDF and deployed to a GitHub-hosted website with every change (see `.travis.yml` and `Dockerfile` in the manuscript repository).
+This gives on the one hand easy access after every update of the R Markdown source code, and on the other hand a second controlled environment making sure that the article renders successfully and correctly.
+
+## Processing
+
+The portability of containerized environments becomes particularly useful for improving expensive processing of data or shipping complex processing pipelines.
+First, it is possible to **offload complex processing to a server** or clouds, and also parallel executing of processes for speeding up or serving of many users.
+**\CRANpkg{batchtools}** provides a parallel implementation of Map for various schedulers \citep{Lang2017batchtools}.
+The package can [schedule jobs with Docker Swarm](https://mllg.github.io/batchtools/reference/makeClusterFunctionsDocker.html)
+**\CRANpkg{googleComputeEngineR}** has the function `gce_vm_cluster()` to create clusters of 2 or more virtual machines, running multi-CPU architectures.
+Instead of running a local R script with the local CPU and RAM restrictions, the same code can be processed on all CPU threads of the cluster of machines in the cloud, all running a Docker container with the same R environments.
+\CRANpkg{googleComputeEngineR} integrates with the R parralisation package \CRANpkg{future} to enable this with only a few lines or R code\footnote{\href{https://cloudyr.github.io/googleComputeEngineR/articles/massive-parallel.html}{https://cloudyr.github.io/googleComputeEngineR/articles/massive-parallel.html}}.
+[Google Cloud Run](https://cloud.run) is a CaaS (Containers as a Service) platform.
+Users can launch containers using any Docker image without worrying about underlying infrastructure in a so called serverless configuration.
+The service takes care of network ingress, scaling machines up and down, authentication and authorisation---all features which are non-trivial for a developer to create on their own.
+This can be used to scale up R code to millions of instances if need be with little or no changes to existing code, as demonstrated by the proof of concept `cloudRunR`\footnote{\href{https://github.com/MarkEdmondson1234/cloudRunR}{https://github.com/MarkEdmondson1234/cloudRunR}}, which uses Cloud Run to create a scalable R-based API using \CRANpkg{plumber}.
+[Google Cloud Build](https://cloud.google.com/cloud-build/) and the Google Container Registry are a continuous integration service respectively image registry that offload building of images to the cloud, while serving the needs of commercial environments such as private Docker images or image stacks.
+As Googl Cloud Build itself can run any container, the package \pkg{googleCloudRunner} demonstrates how R can be used as the control language for one time or batch processing jobs and scheduling of jobs\footnote{\href{https://code.markedmondson.me/googleCloudRunner/articles/cloudbuild.html}{https://code.markedmondson.me/googleCloudRunner/articles/cloudbuild.html}}.
+\label{drake} **\CRANpkg{drake}** is a workflow manager for data science projects \citep{landau_drake_2019}.
+It features implicit parallel computing and automated detection of the parts of the work that actually needs to be reexecuted.
+\CRANpkg{drake} has been demonstrated to run inside containers for high reproducibility\footnote{See for example \href{https://github.com/joelnitta/pleurosoriopsis}{https://github.com/joelnitta/pleurosoriopsis} or \href{https://gitlab.com/ecohealthalliance/drake-gitlab-docker-example}{https://gitlab.com/ecohealthalliance/drake-gitlab-docker-example}, the latter even running in a continuous integration platform (cf.~\nameref{development}.}, but also how to use \CRANpkg{future} package's function `makeClusterPSOCK()` to send parts of the workflow to a Docker image for execution\footnote{\href{https://docs.ropensci.org/drake/index.html?q=docker\#with-docker}{https://docs.ropensci.org/drake/index.html?q=docker\#with-docker}} \citep[see package's function documentation;~][]{future_2020}.
+In the latter case, the container control code must be written by the user and the \CRANpkg{future} package ensures the host and worker can connect for communicating over socket connections.
+**RStudio Server Pro** includes a functionality called [Launcher](https://solutions.rstudio.com/launcher/overview/) (since version&nbsp;1.2, released in 2019).
+It gives users the ability to spawn R sessions and background/batch jobs in a scalable way on external clusters, e.g., [Kubernetes based on Docker images](https://support.rstudio.com/hc/en-us/articles/360019253393-Using-Docker-images-with-RStudio-Server-Pro-Launcher-and-Kubernetes) or [Slurm](https://slurm.schedmd.com/) clusters, and optionally, with Singularity containers.
+A benefit of the proprietary Launcher software is the ability for R and Python users to leverage containerisation's advantages in RStudio without writing specific deployment scripts or learning about Docker or managing clusters at all.
+
+\label{pipelines}
+Second, containers are perfectly suited for **packaging and executing software pipelines** and required data.
+Containers allow building complex processing pipelines that are independent from the host programming language.
+Due to its original use case (see&nbsp;\nameref{introduction}), Docker has no standard mechanisms for chaining containers together; it lacks definitions and protocols for environment variables, volume mounts and/or ports that could enable transfer of input (parameters and data) and output (results) to and from containers.
+Some packages, e.g., \pkg{containerit}, do provide Docker images that can be used similarly to CLIs, but their usage is cumbersome\footnote{\href{https://o2r.info/containerit/articles/container.html}{https://o2r.info/containerit/articles/container.html}}.
+**\pkg{outsider}** ([https://docs.ropensci.org/outsider/](https://docs.ropensci.org/outsider/)) tackles the problem of integrating external programs into an R workflow \citep{bennett_outsider_2020}.
+Installation and usage of external programs can be difficult, convoluted and even impossible if the platform is incompatible.
+Therefore \pkg{outsider} uses the platform-independent Docker images to encapsulate processes in _outsider modules_.
+Each outsider module has a `Dockerfile` and an R package with functions for interacting with the encapsulated tool.
+Using only R functions, an end-user can install a module with the \pkg{outsider} package and then call module code to integrate a tool into their own R-based workflow seamlessly.
+The \pkg{outsider} package and module manage the containers and handle the transmission of arguments and the transfer of files to and from a container.
+These functionalities also allow a user to launch module code on a remote machine via SSH, expanding the potential computational scale.
+Outsider modules can be hosted code-sharing services, e.g., on GitHub, and \pkg{outsider} contains discovery functions for them.
+<!-- informative examples of how modules work, just as a reminder for @nuest
+https://github.com/DomBennett/om..astral/blob/master/examples/astral.R
+https://github.com/DomBennett/om..raxml#partitioned-dna-analysis
+-->
 
@@
 The best practices for [running RStudio with Docker containers](https://support.rstudio.com/hc/en-us/articles/360021594513-Running-RStudio-with-Docker-containers) as well as [Docker images](https://github.com/rstudio/rstudio-docker-products) for RStudio's commerical products are publicly available.
+## Using R to power enterprise software in production environments
+\label{enterprise}
+
+R has been historically viewed as a tool for analysis and scientific research, but not for creating software that corporations can rely on for production services.
+However, thanks to advancements in R running as a web service, along with with the ability to deploy R in Docker containers, modern enterprises are now capable of having real-time machine learning powered by R.
+A number of packages and projects enabled R to respond to client reqests over TCP/IP and local socket servers, such as \CRANpkg{Rserve}, \CRANpkg{svSocket}, [rApache](http://www.rapache.net) and more recently \CRANpkg{plumber} ([https://www.rplumber.io/](https://www.rplumber.io/)) and \pkg{RestRserve} ([http://restrserve.org](http://restrserve.org)), which even processes incoming requests in parallel with forked processes using \CRANpkg{Rserve}.
+The latter two also provide documentation for deployment with Docker or ready to use automated builds of images\footnote{See \href{https://www.rplumber.io/docs/hosting.html\#docker}{https://www.rplumber.io/docs/hosting.html\#docker}, \href{https://hub.docker.com/r/trestletech/plumber/}{https://hub.docker.com/r/trestletech/plumber/} and \href{https://hub.docker.com/r/rexyai/restrserve/}{https://hub.docker.com/r/rexyai/restrserve/}.}.
+These software allow other (remote) processes and programming languages to interact with R and to expose R-based function in a service architecture with HTTP APIs.
+APIs based on these package can be deployed with scalability and high availability using containers.
+This pattern of deploying code matches those used by software engineering services created in more established languages in the enterprise domain, such as Java or Python, and R can be used alongside those langages as a first class member of a software engineering technical stack.
+
+CARD.com implemented a web application for the optimization of the acquisition flow and the real-time analysis of debit card transactions.
+The software used \CRANpkg{Rserve} and rApache and was deployed in Docker containers.
+The R session behind \CRANpkg{Rserve} acted as a read-only in-memory database, which was extremely fast and scalabale, for the many concurrent rApache processes responding to the live-scoring requests of various divisions of the company.
+Similarly dockerized R scripts were responsible for the ETL processes and even the client-facing email, text message and push notification alerts sent in real-time based on card transactions.
+The related Docker images were made available at [https://github.com/cardcorp/card-rocker](https://github.com/cardcorp/card-rocker).
+The images extend `rocker/r-base` and additionally entailed an SSH client and a workaround for being able to mount SSH keys from the host, Pandoc, the Amazon Web Services (AWS) SDK, and Java, which is required by the AWS SDK.
+The AWS SDK allowed to run R consumers reading from real-time data processing streams of [AWS Kinesis](https://aws.amazon.com/kinesis/) \footnote{See useR!2017 talk \href{https://static.sched.com/hosted\_files/user2017/2f/AWR Kinesis at useR 2017.pdf}{"Stream processing with R in AWS"}.}.
+The applications were deployed on Amazon Elastic Container Service ([ECS](https://aws.amazon.com/ecs/)).
+The main learnings from using R in Docker was the imporance of not only pinning the R package versions via MRAN, but also moving away from Debian testing to a distribution with long-term support.
+For the use case at hand, this switch served the priority to have more control over upstream updates, and to minimize the risk of breaking the automated builds of the Docker images and production jobs.
+
+The AI @ T-Mobile team created a set of neural network machine learning natural language processing models to help customer care agents manage text-based messages for customers \citep{t-mobile_enterprise_2018}.
+For example, one model quickly identifies if a message is from a customer or not \citep[see \CRANpkg{Shiny}-based \href{https://secure.message.t-mobile.com/v1/shiny/is-customer/app/}{demo}, ][]{nolis_small_2019}, others tell which customers are likely to make a repeat purchase.
+If a data scientist creates a machine learning model and exposes it through a \CRANpkg{plumber} API, then someone else on the marketing team could write software that sends different emails depending on that real-time prediction.
+The models are convolutional neural networks that use the \CRANpkg{keras} package and run in a Rocker Docker image.
+The corresponding `Dockerfile`s are published [on GitHub](https://github.com/tmobile/r-tensorflow-api).
+Since the models power tools for agents and customers, they need to have extremely high uptime and reliability.
+The AI @ T-Mobile team found that the models performed well and today these models power real-time services that are called over a million times a day.
 
@@
 
-## Processing
-\label{processing}
-
-The portability of containerized environments becomes particularly useful for improving expensive processing of data or shipping complex processing pipelines.
-First, it is possible to **offload complex processing to a server** or clouds, and also parallel executing of processes for speeding up or serving of many users.
-**\CRANpkg{batchtools}** provides a parallel implementation of Map for various schedulers \citep{Lang2017batchtools}.
-The package can [schedule jobs with Docker Swarm](https://mllg.github.io/batchtools/reference/makeClusterFunctionsDocker.html)
-**\CRANpkg{googleComputeEngineR}** has the function `gce_vm_cluster()` to create clusters of 2 or more virtual machines, running multi-CPU architectures.
-Instead of running a local R script with the local CPU and RAM restrictions, the same code can be processed on all CPU threads of the cluster of machines in the cloud, all running a Docker container with the same R environments.
-\CRANpkg{googleComputeEngineR} integrates with the R parralisation package \CRANpkg{future} to enable this with only a few lines or R code\footnote{\href{https://cloudyr.github.io/googleComputeEngineR/articles/massive-parallel.html}{https://cloudyr.github.io/googleComputeEngineR/articles/massive-parallel.html}}.
-[Google Cloud Run](https://cloud.run) is a CaaS (Containers as a Service) platform.
-Users can launch containers using any Docker image without worrying about underlying infrastructure in a so called serverless configuration.
-The service takes care of network ingress, scaling machines up and down, authentication and authorisation---all features which are non-trivial for a developer to create on their own.
-This can be used to scale up R code to millions of instances if need be with little or no changes to existing code, as demonstrated by the proof of concept `cloudRunR`\footnote{\href{https://github.com/MarkEdmondson1234/cloudRunR}{https://github.com/MarkEdmondson1234/cloudRunR}}, which uses Cloud Run to create a scalable R-based API using \CRANpkg{plumber}.
-[Google Cloud Build](https://cloud.google.com/cloud-build/) and the Google Container Registry are a continuous integration service respectively image registry that offload building of images to the cloud, while serving the needs of commercial environments such as private Docker images or image stacks.
-As Googl Cloud Build itself can run any container, the package \pkg{googleCloudRunner} demonstrates how R can be used as the control language for one time or batch processing jobs and scheduling of jobs\footnote{\href{https://code.markedmondson.me/googleCloudRunner/articles/cloudbuild.html}{https://code.markedmondson.me/googleCloudRunner/articles/cloudbuild.html}}.
-\label{drake} **\CRANpkg{drake}** is a workflow manager for data science projects \citep{landau_drake_2019}.
-It features implicit parallel computing and automated detection of the parts of the work that actually needs to be reexecuted.
-\CRANpkg{drake} has been demonstrated to run inside containers for high reproducibility\footnote{See for example \href{https://github.com/joelnitta/pleurosoriopsis}{https://github.com/joelnitta/pleurosoriopsis} or \href{https://gitlab.com/ecohealthalliance/drake-gitlab-docker-example}{https://gitlab.com/ecohealthalliance/drake-gitlab-docker-example}, the latter even running in a continuous integration platform (cf.~\nameref{development}.}, but also how to use \CRANpkg{future} package's function `makeClusterPSOCK()` to send parts of the workflow to a Docker image for execution\footnote{\href{https://docs.ropensci.org/drake/index.html?q=docker\#with-docker}{https://docs.ropensci.org/drake/index.html?q=docker\#with-docker}} \citep[see package's function documentation;~][]{future_2020}.
-In the latter case, the container control code must be written by the user and the \CRANpkg{future} package ensures the host and worker can connect for communicating over socket connections.
-**RStudio Server Pro** includes a functionality called [Launcher](https://solutions.rstudio.com/launcher/overview/) (since version&nbsp;1.2, released in 2019).
-It gives users the ability to spawn R sessions and background/batch jobs in a scalable way on external clusters, e.g., [Kubernetes based on Docker images](https://support.rstudio.com/hc/en-us/articles/360019253393-Using-Docker-images-with-RStudio-Server-Pro-Launcher-and-Kubernetes) or [Slurm](https://slurm.schedmd.com/) clusters, and optionally, with Singularity containers.
-A benefit of the proprietary Launcher software is the ability for R and Python users to leverage containerisation's advantages in RStudio without writing specific deployment scripts or learning about Docker or managing clusters at all.
-
-\label{pipelines}
-Second, containers are perfectly suited for **packaging and executing software pipelines** and required data.
-Containers allow building complex processing pipelines that are independent from the host programming language.
-Due to its original use case (see&nbsp;\nameref{introduction}), Docker has no standard mechanisms for chaining containers together; it lacks definitions and protocols for environment variables, volume mounts and/or ports that could enable transfer of input (parameters and data) and output (results) to and from containers.
-Some packages, e.g., \pkg{containerit}, do provide Docker images that can be used similarly to CLIs, but their usage is cumbersome\footnote{\href{https://o2r.info/containerit/articles/container.html}{https://o2r.info/containerit/articles/container.html}}.
-**\pkg{outsider}** ([https://docs.ropensci.org/outsider/](https://docs.ropensci.org/outsider/)) tackles the problem of integrating external programs into an R workflow \citep{bennett_outsider_2020}.
-Installation and usage of external programs can be difficult, convoluted and even impossible if the platform is incompatible.
-Therefore \pkg{outsider} uses the platform-independent Docker images to encapsulate processes in _outsider modules_.
-Each outsider module has a `Dockerfile` and an R package with functions for interacting with the encapsulated tool.
-Using only R functions, an end-user can install a module with the \pkg{outsider} package and then call module code to integrate a tool into their own R-based workflow seamlessly.
-The \pkg{outsider} package and module manage the containers and handle the transmission of arguments and the transfer of files to and from a container.
-These functionalities also allow a user to launch module code on a remote machine via SSH, expanding the potential computational scale.
-Outsider modules can be hosted code-sharing services, e.g., on GitHub, and \pkg{outsider} contains discovery functions for them.
-<!-- informative examples of how modules work, just as a reminder for @nuest
-https://github.com/DomBennett/om..astral/blob/master/examples/astral.R
-https://github.com/DomBennett/om..raxml#partitioned-dna-analysis
--->
-
 ## Packaging research reproducibly
@@
 
-## Development, debugging, and testing
-\label{development}
-
-Containers can also serve as useful playgrounds to create environments ad-hoc or to provide very specific environments that are not needed or not easily available in day-to-day development for the purposes of developing R packages.
-These environments may have specific versions of R, of R extension packages, and of system libraries used by R extension packages, and all of the above in a specific combination.
-
-First, such containers can greatly facilitate **fixing bugs and quick evaluation**, because developers and users can readily start and later discard a container to investigate a bug report or try out a piece of software \citep[cf.][]{ooms_opencpu_2017}.
-The container does not affect their regular system.
-Using the Rocker images with RStudio, these disposable environments lack no development comfort (cf. Section&nbsp;\nameref{compendia}).
-\citet{ooms_opencpu_2017} describes how `docker exec` can be used to get a root shell in a container for cusotmization during software evaluation.
-\citet{eddelbuettel_debugging_2019} describes an example how a Docker container was used to debug an issue with a package only occuring with a particular version of Fortran, and using tools which are not readily available on all platforms (e.g., not on macOS).
-
-Second, the strong integration of **system libraries in core packages** in the [R-spatial community](https://www.r-spatial.org/) makes containers essential for stable and proactive development of common classes for geospatial data modelling and analysis.
-For example, GDAL \citep{gdal_2019} is a crucial library in the geospatial domain.
-GDAL is a system dependency allowing R packages such as \CRANpkg{sf}, which provides the core data model for geospatial vector data, or \CRANpkg{rgdal}, to accomodate users to be able to read and write hundreds of different spatial raster and vector formats.
-\CRANpkg{sf} and \CRANpkg{rgdal} have hundreds of indirect reverse imports and dependencies and therefore the maintainers spend a lot of efforts not to break these.
-Purpose built Docker are used to prepare for upcoming releases of system libraries, individual bug reports, and for the lowest supported versions of system libraries\footnote{Cf.
-\href{https://github.com/r-spatial/sf/tree/master/inst/docker}{https://github.com/r-spatial/sf/tree/master/inst/docker}, \href{https://github.com/Nowosad/rspatial_proj6}{https://github.com/Nowosad/rspatial\_proj6}, and \href{https://github.com/r-spatial/sf/issues/1231}{https://github.com/r-spatial/sf/issues/1231}}.
-```{r geospatial_deps, eval=FALSE, include=FALSE}
-length(tools::dependsOnPkgs(c("rgdal", "sf"), dependencies = c("most"), recursive = TRUE))
-```
-
-Third, there are special purpose images for identifying problems beyond the mere R code, such as **debugging R memory problems**.
-The images significantly reduce the barrier to follow complex steps for fixing memory allocation bugs \citep[cf. Section~4.3 in][]{core_writing_1999}.
-These problems are hard to debug and critical, both because when they occur they lead to fatal crashing processes.
-[`rocker/r-devel-san`](https://github.com/rocker-org/r-devel-san) and [`rocker/r-devel-ubsan-clang`](https://github.com/rocker-org/r-devel-san-clang) are Docker images have a particularly configured version of R to trace such problems with gcc and clang compilers, respectively \citep[cf.~\CRANpkg{sanitizers} for examples,][]{eddelbuettel_sanitizers_2014}.
-The image [`wch/r-debug`](https://github.com/wch/r-debug) is a purpose built Docker image with _multiple_ instrumented builds of R, each with a different diagnostic utility activated.
-
-Fourth, containers are useful for **testing** R code during development.
-To submit a package to CRAN, an R package must work with the development version of R, which must be compiled locally.
-That can be a challenge for some users.
-The **R-hub** project provides _"a collection of services to help R package development"_, with the package builder as the most prominent one\citep{r-hub_docs_2019}.
-R-hub makes it easy to ensure that no errors occur, but to fix errors a local setup is still often warranted, e.g., using the image `rocker/r-devel`, and to test packages with native code, which can make the process more complex \citep[cf.][]{eckert_building_2018}.
-The R-hub Docker images can also be used to debug problems locally using various combinations of Linux platforms, R versions, and compilers\footnote{See \href{https://r-hub.github.io/rhub/articles/local-debugging.html}{https://r-hub.github.io/rhub/articles/local-debugging.html} and \href{https://blog.r-hub.io/2019/04/25/r-devel-linux-x86-64-debian-clang/}{https://blog.r-hub.io/2019/04/25/r-devel-linux-x86-64-debian-clang/}}.
-The images go beyond the configurations, or _flavours_, used by CRAN for checking packages\footnote{\href{https://cran.r-project.org/web/checks/check_flavors.html}{https://cran.r-project.org/web/checks/check\_flavors.html}}, e.g., with CentOS-based images, but lack a container for checking on Windows or OS X.
-The images greatly support package developers to provide support on operating systems they are not familiar.
-The package \pkg{dockertest} ([https://github.com/traitecoevo/dockertest/](https://github.com/traitecoevo/dockertest/)) is a proof of concept for automatically generating `Dockerfile`s and building images specifically to run tests\footnote{\pkg{dockertest} is not actively maintained, but mentioned still because of its interesting approach.}.
-These images are accompanied with a special launch script so the tested source code is not stored in the image but the currently checked in version from a local Git repository is cloned into the container at runtime.
-This approach clearly seperates test environment, test code, and current working copy of the code.
-\label{rselenium} Another use case where a container helps to standardise tests across operating systems is detailed the vignettes of the package \CRANpkg{RSelenium} \citep{rselenium_2019}.
-The package recommends Docker for running the [Selenium](https://selenium.dev/) Server application needed to execute test suites on browser-based user interfaces and webpages, but requires users to manually manage the Docker containers.
-
-\label{ci}
-Fifth, Docker images can be used **on continuous integration (CI) platforms** to streamline the testing of packages.
-\citet{ye_docker_2019} describes how they speed up the process of testing by running tasks on [Travis&nbsp;CI](https://travis-ci.org/) within a container using `docker exec`, e.g., the package check or rendering of documentation.
-\citet{cardozo_faster_2018} saved time, also on Travis&nbsp;CI, by re-using the testing image as the base for an image intended for publication on Docker&nbsp;Hub.
-[`r-ci`](https://github.com/ColinFay/r-ci) is in turn used with [GitLab&nbsp;CI](https://docs.gitlab.com/ee/ci/), which itself is built on top of Docker images: the user specifies a base Docker image, and the whole tests are run inside this environment.
-The `r-ci` image stack combines `rocker` versioning and a series of tools specifically designed for testing in a fixed environment with a customized list of preinstalled packages.
-Especially for long running tests or complex system dependencies, these approaches to seperate installation of build dependencies with code testing streamline the development process.
-Not due to a concern about time, but to control the environment used on a CI server, even this manuscript is rendered into a PDF and deployed to a GitHub-hosted website with every change (see `.travis.yml` and `Dockerfile` in the manuscript repository).
-This gives on the one hand easy access after every update of the R Markdown source code, and on the other hand a second controlled environment making sure that the article renders successfully and correctly.
-
 # Conclusions
-\label{conclusions}
 

### 2020-01-28 manuscript.Rmd
ignore preprint file
document steps for submission to arXiv

@@
 rmarkdown::render("manuscript.Rmd", output_format = rjournal_style_preprint())
+file.rename("RJwrapper.pdf", "rockerverse-preprint.pdf") # file for own archival
+
+# for submission to arXiv, copy the following files into a new folder and run pdfLaTeX directly on RJwrapper.text (works with RStudio): figures/*, manuscript.tex, RJournal.sty, RJreferences.bib, RJwrapper.tex
+# then tip the following files and submit to arXiv:
+# - /figures
+# - manuscript.tex
+# - RJournal.sty
+# - RJwrapper.bbl
+# - RJwrapper.tex
 ```

### 2020-01-29 manuscript.Rmd
fix footnote

--- 2020-01-28 manuscript.Rmd
@@
 abstract: >
-  The Rocker&nbsp;Project provides widely-used Docker images for R across different application scenarios.
+  The Rocker Project provides widely-used Docker images for R across different application scenarios.
   This articles surveys downstream projects building upon Rocker and presents the current state of R packages for managing Docker images and controlling containers.
@@
 Combining both these topics, the _Rocker&nbsp;Project_ ([https://www.rocker-project.org/](https://www.rocker-project.org/)) provides images with R (see the next Section for more details).
-The considerable uptake and continued evolution of the Rocker&nbsp;Project has lead to numerous projects extending or building upon Rocker images, ranging from reproducible research\footnote{"Reproducible" in the sense of the _Claerbout/Donoho/Peng_ terminology \citep{barba_terminologies_2018}.} to production deployments.
+The considerable uptake and continued evolution of the Rocker&nbsp;Project has lead to numerous projects extending or building upon Rocker images, ranging from reproducible research\footnote{"Reproducible" in the sense of the \emph{Claerbout/Donoho/Peng} terminology \citep{barba_terminologies_2018}.} to production deployments.
 This article presents this _Rockerverse_ of projects across all development stages: early demonstrations, working prototypes, and mature products.
@@
 Whole&nbsp;Tale is a National Science Foundation (NSF) funded project developing a scalable, open-source, multi-user platform for reproducible research \citep{brinckman2019, chard2019a}.
-A central goal of the platform is to enable researchers to easily create and publish executable research objects\footnote{In Whole~Tale a _tale_ is a research object that contains metadata, data (by copy or reference), code, narrative, documentation, provenance, and information about the computational environment to support computational reproducibility} associated with published research \citep{chard2019b}.
+A central goal of the platform is to enable researchers to easily create and publish executable research objects\footnote{In Whole~Tale a \emph{tale} is a research object that contains metadata, data (by copy or reference), code, narrative, documentation, provenance, and information about the computational environment to support computational reproducibility.} associated with published research \citep{chard2019b}.
 Using Whole&nbsp;Tale, researchers can create and publish Rocker-based reproducible research objects to a growing number of repositories including DataONE member nodes, Zenodo and soon Dataverse.
@@
 rmarkdown::render("manuscript.Rmd", output_format = rjournal_style_preprint())
+file.rename("RJwrapper.pdf", "rockerverse-preprint.pdf") # file for own archival
+
+# for submission to arXiv, copy the following files into a new folder and run pdfLaTeX directly on RJwrapper.tex (works with RStudio): figures/*, manuscript.tex, RJournal.sty, RJreferences.bib, RJwrapper.tex
+# then zip the following files and submit to arXiv:
+# - /figures
+# - manuscript.tex
+# - RJournal.sty
+# - RJwrapper.bbl
+# - RJwrapper.tex
 ```

### 2020-01-31 README.Rmd
use more suitable font for sticker

@@
 ```{r hexSticker, message=TRUE, warning=FALSE, cache=TRUE, include=FALSE}
-if(!file.exists("rockerverse.png")) {
 library("magick")
@@
 # https://commons.wikimedia.org/wiki/File:NASA-HS201427a-HubbleUltraDeepField2014-20140603.jpg
+if(!file.exists("universe.png")) {
   universe <- magick::image_read("https://upload.wikimedia.org/wikipedia/commons/6/69/NASA-HS201427a-HubbleUltraDeepField2014-20140603.jpg")
   magick::image_write(universe, here("universe.png"))
-  # rockerLogoMask <-
+}
+
 magick::image_read("https://avatars2.githubusercontent.com/u/9100160?s=200&v=4") %>%
@@
 
+library("sysfonts")
+if(!file.exists(here::here("FinalFrontierOldStyle-8Pg.ttf"))) {
+  stop("Download font Final Frontier Old Style, e.g. from https://www.fontspace.com/allen-r-walden/final-frontier-old-style and put the file in the repo root.")
+}
+sysfonts::font_add("finalfrontier", here::here("FinalFrontierOldStyle-8Pg.ttf"))
+
 library("hexSticker")
-  hexSticker::sticker(here("whaleverse.png"), package = "Rockerverse",
-          p_size = 20, s_x = 0.97, s_y = 0.95, s_width = 1,
+hexSticker::sticker(here::here("whaleverse.png"), package = "ROCKERVERSE",
+        p_family = "finalfrontier",
+        p_size = 29, p_y = 1.3,
+        s_x = 0.97, s_y = 0.95, s_width = 1,
         h_fill = "#000000", h_color = "#ffffff",
@@
         u_color = "#ffffff", u_size = 4,
-          filename = here("rockerverse.png"))
+        filename = here::here("rockerverse.png"))
 
-  file.remove(c(here("universe.png"), here("whaleverse.png"), here("rocker_mask.png")))
-}
+#file.remove(c(here("universe.png"), here("whaleverse.png"), here("rocker_mask.png")))
 ```
@@
 ![Rockerverse hex sticker](`r here::here("rockerverse.png")`)
+
+Hex sticker uses the font [Final Frontier Old Style](https://www.fontspace.com/allen-r-walden/final-frontier-old-style) and the [Hubble Ultra-Deep Field](https://en.wikipedia.org/wiki/Hubble_Ultra-Deep_Field) combined with the Docker logo as background.
 

### 2020-01-31 README.Rmd
fix sticker path
high-res sticker option

@@
 
-Draft for an article of everything related to containers and R, originally based on the blog post http://bit.ly/docker-r.
+Article about everything related to containers and R, originally based on the blog post http://bit.ly/docker-r and now **available as a preprint on arXiv.org: https://arxiv.org/abs/2001.10641**.
 
@@
 library("hexSticker")
+factor <- 1 # for print: 4
 hexSticker::sticker(here::here("whaleverse.png"), package = "ROCKERVERSE",
         p_family = "finalfrontier",
-        p_size = 29, p_y = 1.3,
+        p_size = 29 * factor, p_y = 1.3,
         s_x = 0.97, s_y = 0.95, s_width = 1,
@@
         url = "github.com/nuest/rockerverse-paper",
-        u_color = "#ffffff", u_size = 4,
-        filename = here::here("rockerverse.png"))
+        u_color = "#ffffff", u_size = 4 * factor,
+        filename = here::here(ifelse(
+          factor == 1,
+          "rockerverse.png",
+          paste0("rockerverse_", 300 * factor, "dpi.png"))),
+        dpi = 300 * factor)
 
@@
 
-![Rockerverse hex sticker](`r here::here("rockerverse.png")`)
+![Rockerverse hex sticker](rockerverse.png)
 
-Hex sticker uses the font [Final Frontier Old Style](https://www.fontspace.com/allen-r-walden/final-frontier-old-style) and the [Hubble Ultra-Deep Field](https://en.wikipedia.org/wiki/Hubble_Ultra-Deep_Field) combined with the Docker logo as background.
+The hex sticker uses the font [Final Frontier Old Style](https://www.fontspace.com/allen-r-walden/final-frontier-old-style) and the [Hubble Ultra-Deep Field](https://en.wikipedia.org/wiki/Hubble_Ultra-Deep_Field) combined with the Docker logo as background.
 

### 2020-02-04 manuscript.Rmd
fix Dirk's email, add snipped for email list
proofreading changes

@@
 ---
-title: "The Rockerverse: Packages and Applications for Containerization with R"
+title: "The Rockerverse: Packages and Applications for Containerisation with R"
 author:
@@
       - \orcid{0000-0001-6419-907X}
-    email: dirkd@eddelbuettel.com
+    email: dirk@eddelbuettel.com
   - name: Dom Bennett
@@
 abstract: >
-  The Rocker Project provides widely-used Docker images for R across different application scenarios.
-  This articles surveys downstream projects building upon Rocker and presents the current state of R packages for managing Docker images and controlling containers.
+  The Rocker Project provides widely used Docker images for R across different application scenarios.
+  This articles surveys downstream projects that build upon Rocker images and presents the current state of R packages for managing Docker images and controlling containers.
   These use cases and the variety of applications demonstrate the power of Rocker specifically and containerisation in general.
-  We identified common themes across this diversity: reproducible environments, scalability and efficiency, and portability across clouds.
-# blank footnote? https://tex.stackexchange.com/questions/170511/footnotes-without-numbering
+  Across the diverse ways to use containers, we identified common themes: reproducible environments, scalability and efficiency, and portability across clouds.
 output:
@@
 
-The R community keeps growing.
-This can be seen in the number of new packages on CRAN, which keeps on growing exponentially \citep{cran:2019}, but also in the numbers of conferences, open educational resources, meetups, unconferences, and companies taking up, as exemplified by the useR!
+The R community continues to grow.
+This can be seen in the number of new packages on CRAN, which is still on growing exponentially \citep{cran:2019}, but also in the numbers of conferences, open educational resources, meetups, unconferences, and companies that are adopting R, as exemplified by the useR!
 conference series\footnote{\href{https://www.r-project.org/conferences/}{https://www.r-project.org/conferences/}}, the global growth of the R and R-Ladies user groups\footnote{\href{https://www.r-consortium.org/blog/2019/09/09/r-community-explorer-r-user-groups}{https://www.r-consortium.org/blog/2019/09/09/r-community-explorer-r-user-groups}, \href{https://www.r-consortium.org/blog/2019/08/12/r-community-explorer}{https://www.r-consortium.org/blog/2019/08/12/r-community-explorer}}, or the foundation and impact of the R&nbsp;Consortium\footnote{\href{https://www.r-consortium.org/news/announcements}{https://www.r-consortium.org/news/announcements}, \href{https://www.r-consortium.org/blog/2019/11/14/data-driven-tracking-and-discovery-of-r-consortium-activities}{https://www.r-consortium.org/blog/2019/11/14/data-driven-tracking-and-discovery-of-r-consortium-activities}}.
-All this cements the role of R as the _lingua&nbsp;franca_ of statistics, data visualisation, and computational research.
-Coinciding with the rise of R was the advent of [Docker](https://en.wikipedia.org/wiki/Docker_(software)) as a general tool for distribution and deployment of server applications---in fact, Docker can be called the _lingua&nbsp;franca_ of describing computing environments and packaging software.
-Combining both these topics, the _Rocker&nbsp;Project_ ([https://www.rocker-project.org/](https://www.rocker-project.org/)) provides images with R (see the next Section for more details).
-The considerable uptake and continued evolution of the Rocker&nbsp;Project has lead to numerous projects extending or building upon Rocker images, ranging from reproducible research\footnote{"Reproducible" in the sense of the \emph{Claerbout/Donoho/Peng} terminology \citep{barba_terminologies_2018}.} to production deployments.
-This article presents this _Rockerverse_ of projects across all development stages: early demonstrations, working prototypes, and mature products.
-We also introduce related activities connecting the R language and environment with other containerisation solutions.
-The main contributions is a coherent picture of the current lay of the land of using containers in, with, and for R.
+These trends cement the role of R as the _lingua&nbsp;franca_ of statistics, data visualisation, and computational research.
+Also though, coinciding with the rise of R has been the advent of [Docker](https://en.wikipedia.org/wiki/Docker_(software)) as a general tool for distributing and deploying of server applications---in fact, Docker can be called the _lingua&nbsp;franca_ of describing computing environments and packaging software.
+Combining both these topics, the _Rocker&nbsp;Project_ ([https://www.rocker-project.org/](https://www.rocker-project.org/)) provides images with R (see the next section for more details).
+The considerable uptake and continued evolution of the Rocker&nbsp;Project has led to numerous projects that extend or build upon Rocker images, ranging from reproducible\footnote{"Reproducible" in the sense of the \emph{Claerbout/Donoho/Peng} terminology \citep{barba_terminologies_2018}.} research to production deployments.
+As such, this article presents this _Rockerverse_ of projects across all development stages: early demonstrations, working prototypes, and mature products.
+We also introduce related activities that connect the R language and environment with other containerisation solutions.
+Our main contributions is a coherent picture of the current status of using containers in, with, and for R.
 
-The article continues with a brief introduction of containerization basics and the Rocker&nbsp;Project, followed by use cases and applications, starting with the R packages specificly for interacting with Docker, second-level packages using containers indirectly or only for specific features, up to complex use cases leveraging containers.
-We conclude with a reflection on the landscape of packages and applications and point out future directions of development.
+The article continues with a brief introduction of containerisation basics and the Rocker&nbsp;Project, followed by use cases and applications, starting with the R packages specificly for interacting with Docker, next the second-level packages that use containers indirectly or only for specific features, and finally some complex use cases that leverage containers.
+We conclude with by reflecting on the landscape of packages and applications and point out future directions of development.
 
-# Containerization and Rocker
+# Containerisation and Rocker
 \label{containerisation}
@@
 
-Docker, an application and service provide by the eponymous company, has in just a few short years risen to prominence for development, testing, deployment and distribution of computer software \citep[cf.][]{datadog_8_2018,munoz_history_2019}.
-While there are related approaches such as LXC\footnote{\href{https://en.wikipedia.org/wiki/LXC}{https://en.wikipedia.org/wiki/LXC}} or Singularity \citep{kurtzer_singularity_2017}, Docker has become synomymous with "containerization"---the method of taking software artefacts and bundling them in such a way that use becomes standardized and portable across operationg systems.
-In doing so, Docker had recognised and validated the importance of one very important thread that had been emerging, namely virtualization.
-By allowing (one or possibly) multiple applications or services to run concurrently on one host machine without any fear of interference between them, an important scalability opportunity is being provided.
-But Docker improved this compartmentalization by accessing the host system---generally Linux---through a much thinner and smaller shim than a full operating system emulation or virtualization.
-This containerization is also called operating-system-level virtualization \citep{wikipedia_contributors_os-level_2020}.
-Typically a container runs one process, whereas virtualization may run whole operating systems at a larger footprint.
+Docker, an application and service provided by the eponymous company, has, in just a few short years, risen to prominence for developing, testing, deploying and distributing computer software \citep[cf.][]{datadog_8_2018,munoz_history_2019}.
+While related approaches exist, such as LXC\footnote{\href{https://en.wikipedia.org/wiki/LXC}{https://en.wikipedia.org/wiki/LXC}} or Singularity \citep{kurtzer_singularity_2017}, Docker has become synomymous with "containerisation"---the method of taking software artefacts and bundling them in such a way that use becomes standardized and portable across operationg systems.
+In doing so, Docker had recognised and validated the importance of one very important thread that had been emerging, namely virtualisation.
+By allowing (one or possibly) multiple applications or services to run concurrently on one host machine without any fear of interference between them, Docker provides an important scalability opportunity.
+Beyond this though, Docker has improved this compartmentalisation by accessing the host system---generally Linux---through a much thinner and smaller shim than a full operating system emulation or virtualisation.
+This containerisation, also called operating-system-level virtualisation \citep{wikipedia_contributors_os-level_2020}, 
+makes efficient use of operating system resources \citep{felter_updated_2015} and allows another order of magnitude in terms of scalability of deployment \citep[cf.][]{datadog_8_2018} because, while a container typically runs one process, virtualisation may run whole operating systems at a larger footprint.
 This makes for more efficient use of system resources \citep{felter_updated_2015} and allowed another order of magnitude in terms of scalability of deployment \citep[cf.][]{datadog_8_2018}.
-While Docker makes use of Linux kernel features, its importance was large enough so that some required aspects of running Docker have been added to other operating systems to support Docker there more efficiently too \citep{microsoft_linux_2019}.
-The success even lead to standardisation and industry collaboration \citep{oci_open_2019}.
+While Docker makes use of Linux kernel features, it has become important enough that some required aspects of running Docker have been added to other operating systems so that those systems can more efficiently support Docker \citep{microsoft_linux_2019}.
+The success of Docker has even led to standardisation and industry collaboration \citep{oci_open_2019}.
 
-The key accomplishment of Docker as an "application" is to make a "bundled" aggregation of software (the so-called "image") available to any system equipped to run Docker, without requiring much else from the host besides the actual Docker application installation.
-This is a rather attractive proposition and Docker's very easy to use user interface has lead to widespread adoption and use of Docker in a variety of domains, e.g., cloud computing infrastructure \citep[e.g.,][]{Bernstein2014}, data science \citep[e.g.,][]{boettiger_introduction_2015}, and edge computing \citep[e.g.,][]{alam_orchestration_2018}.
-It provided to be a natural match for "cloud deployment" which runs, or at least appears to run, "seamlessly" without much explicit reference to the underlying machine, architecture or operating system: containers are portable and can be deployed with very little in terms of dependencies on the host system---only the container runtime is required.
-Images are normally built from plain text documents called `Dockerfile`.
-A `Dockerfile` has a specific set of instructions to create and document a well-defined environment, i.e., install specific software and expose specific ports.
+The key accomplishment of Docker as an "application" is to make a "bundled" aggregation of software, the so-called "image", available to any system equipped to run Docker, without requiring much else from the host besides the actual Docker application installation.
+This is a rather attractive proposition, and Docker's very easy to operate user interface has led to widespread adoption and use of Docker in a variety of domains, e.g., cloud computing infrastructure \citep[e.g.,][]{Bernstein2014}, data science \citep[e.g.,][]{boettiger_introduction_2015}, and edge computing \citep[e.g.,][]{alam_orchestration_2018}.
+It has also proven to be a natural match for "cloud deployment" which runs, or at least appears to run, "seamlessly" without much explicit reference to the underlying machine, architecture or operating system:
+Containers are portable and can be deployed with very little dependencies on the host system---only the container runtime is required.
+These Docker images are normally built from plain text documents called `Dockerfile`s; a `Dockerfile` has a specific set of instructions to create and document a well-defined environment, i.e., install specific software and expose specific ports.
 
-For statistical computing and analysis centered around R, the **Rocker&nbsp;Project** has provided a variety of Docker containers since its start in 2014 \citep{RJ-2017-065}.
+For statistical computing and analysis centered around R, the **Rocker&nbsp;Project** has provided a variety of Docker containers since it began in 2014 \citep{RJ-2017-065}.
 The Rocker&nbsp;Project provides several lines of containers spanning to from building blocks with `R-release` or `R-devel`, via containers with [RStudio&nbsp;Server](https://rstudio.com/products/rstudio/) and [Shiny&nbsp;Server](https://rstudio.com/products/shiny/shiny-server/), to domain-specific containers such as [`rocker/geospatial`](https://github.com/rocker-org/geospatial) \citep{rocker_geospatial_2019}.
-These containers form _image stacks_, building on top of each other for better maintainability (i.e., smaller `Dockerfile`s), composability, and to reduce build time.
+These containers form _image stacks_, building on top of each other for easier maintainability (i.e., smaller `Dockerfile`s), better composability, and to reduce build time.
 Also of note is a series of "versioned" containers which match the R release they contain with the _then-current_ set of packages via the MRAN Snapshot views of CRAN \citep{microsoft_cran_2019}.
-The Rocker&nbsp;Project's impact and importance was acknowledged by the Chan Zuckerberg Initiative's _Essential Open Source Software for Science_, who provide findung for the projects's sustainable maintenance, community growth, and targeting new hardware platforms including GPUs \citep{chan_zuckerberg_initiative_maintaining_2019}.
+The Rocker&nbsp;Project's impact and importance was acknowledged by the Chan Zuckerberg Initiative's _Essential Open Source Software for Science_, which provides funding for the projects's sustainable maintenance, community growth, and targeting new hardware platforms including GPUs \citep{chan_zuckerberg_initiative_maintaining_2019}.
 
@@
 An alternative stemming from the domain of high-perfomance computing is **Singularity** \citep{kurtzer_singularity_2017}.
-Singularity can run Docker images, and in the case of Rocker works out of the box if the main process is R, e.g., in `rocker/r-base`, but does not succeed in running images where there is an init script, e.g., in containers that by default run RStudio&nbsp;Server.
-In the latter case, a `Singularity` file, a recipe akin to a `Dockerfile`, needs to be used.
+Singularity can run Docker images.
+Rocker images work out of the box if the main process is R, e.g., in `rocker/r-base`, but Singularity does not succeed in running images where there is an init script, e.g., in containers that by default run RStudio&nbsp;Server.
+In the latter case, a `Singularity` file, a recipe akin to a `Dockerfile`, needs to be used to make necessary adjustments.
 To date, no comparable image stack to Rocker exists on [Singularity Hub](https://singularity-hub.org/).
-A further tool for running containers is [**podman**](https://github.com/containers/libpod), which also can build `Dockerfile`s and run Docker images.
-Proof of concepts for using podman to build and run Rocker containers exist\footnote{See \href{https://github.com/nuest/rodman}{https://github.com/nuest/rodman} and \href{https://github.com/rocker-org/rocker-versioned/issues/187}{https://github.com/rocker-org/rocker-versioned/issues/187}}.
-Yet the prevelance of Docker, especially in the broader user community beyond experts or niche systems, and the vast amount of blog posts and courses for Docker, currently caps specific development efforts for both Singularity and podman in the R community.
-This might quickly change when usability and spread increase, or security features such as rootless/unprivileged containers, which both these tools support out of the box, become more sought after.
+A further tool for running containers is [**podman**](https://github.com/containers/libpod), which can also build `Dockerfile`s and run Docker images.
+Proof of concepts exists for using podman to build and run Rocker containers\footnote{See \href{https://github.com/nuest/rodman}{https://github.com/nuest/rodman} and \href{https://github.com/rocker-org/rocker-versioned/issues/187}{https://github.com/rocker-org/rocker-versioned/issues/187}}, but the prevelance of Docker, especially in the broader user community beyond experts or niche systems and the vast amount of blog posts and courses for Docker currently cap specific development efforts for both Singularity and podman in the R community.
+This might quickly change if the usability and spread of Singularity or podman increase, or if security features such as rootless/unprivileged containers, which both these tools support out of the box, become more sought after.
 
@@
 However, moving back and forth between an R console and the command line can create friction in workflows and reduce reproducibility.
-A number of first-order R packages provide a interface to the Docker CLI, allowing to automate interaction with the Docker CLI from an R console.
+A number of first-order R packages provide an interface to the Docker CLI, allowing for the interaction with the Docker CLI from an R console, e.g., to be automated.
 
 Each of these packages has particular advantages as they provide function wrappers for interacting with the Docker CLI at different stages of a container's life cycle.
-Examples of such interactions are installing the Docker software, creating `Dockerfile`s (\CRANpkg{dockerfiler}, \pkg{containerit}), building images and launching a containers (\CRANpkg{stevedore}, \pkg{docker}) on a local machine or on the cloud.
-As such, the choice of which package is most useful depends on the use-case at hand, but also the users level of expertise.
+Examples of such interactions are installing the Docker software, creating `Dockerfile`s (\CRANpkg{dockerfiler}, \pkg{containerit}), building images, and launching a containers (\CRANpkg{stevedore}, \pkg{docker}) on a local machine or on the cloud.
+As such, the choice of which package is most useful depends on the use case at hand as well as on the user's level of expertise.
 
@@
   ~package, ~id,  
+  "harbor", c("execute", "cloud", "images"),
   "stevedore", c("api", "execute", "cloud", "images", "interact"), # build?
-  "dockyard", c("dockerfile", "build", "execute"), # cloud?
-  "dockermachine", c("install", "execute", "cloud", "interact"),
   "AzureContainers", c("build", "execute", "cloud"),
   "babelwhale", c("execute", "interact", "othercontainers"),
-  "harbor", c("execute", "cloud", "images")
+  "dockyard", c("dockerfile", "build", "execute"), # cloud?
+  "dockermachine", c("install", "execute", "cloud", "interact")
 )
@@
 
-**\pkg{harbor}** ([https://github.com/wch/harbor](https://github.com/wch/harbor) is not actively maintained anymore, but should be honorably mentioned as the first R package for managing Docker images and containers.
-It uses the \CRANpkg{sys} package to run system commands against the Docker CLI, both locally and through an SSH connection, and has convenience functions, e.g., for listing and removing containers/images and for accessing logs.
+**\pkg{harbor}** ([https://github.com/wch/harbor](https://github.com/wch/harbor) is no longer actively maintained, but it should be honorably mentioned as the first R package for managing Docker images and containers.
+It uses the \CRANpkg{sys} package to run system commands against the Docker CLI, both locally and through an SSH connection, and it has convenience functions, e.g., for listing and removing containers/images and for accessing logs.
 The output of container executions are converted to appropriate R types.
-The Docker CLI's basic functionality, while evolving quickly and with small concern for avoiding breaking changes, is unchanged for a long time so a core function such as `harbor::docker_run(image = "hello-world")` still works depsite the stopped development.
+The Docker CLI's basic functionality, although it evolves quickly and with little concern for avoiding breaking changes, has remained unchanged in core functions, meaning that a core function such as `harbor::docker_run(image = "hello-world")` still works depsite its stopped development.
 ```{r harbor, eval=FALSE, include=FALSE}
@@
 It interfaces with the Docker daemon over the Docker HTTP API\footnote{\href{https://docs.docker.com/engine/api/latest/}{https://docs.docker.com/engine/api/latest/}} via a Unix socket on Linux or MacOS, over a named pipe on Windows, or over an HTTP/TCP connection.
-The package is the only one not using system calls to the `docker` CLI for managing images and containers and easily exposes connections to remote Docker daemons, which has to be configured on the Docker level otherwise.
-Using the API gives access to more information and is system independent and likely more reliable than parsing command line output.
-\CRANpkg{stevedore}'s own interface is automatically generated based on the OpenAPI specification of the Docker daemon, but still similar to the Docker CLI.
+The package is the only one not using system calls to the `docker` CLI tool for managing images and containers.
+The package thereby enables connections to remote Docker instances without direct configuration of the local Docker daemon.
+Furthermore using the API gives access to information in a structured way, is system independent, and is likely more reliable than parsing command line output.
+\CRANpkg{stevedore}'s own interface is automatically generated based on the OpenAPI specification of the Docker daemon, but it is still similar to the Docker CLI.
 The interface is similar to R6 objects, in that a `stevedore_object` has a number of functions attached to it that can be called, and multiple specific versions of the Docker API can be supported thanks to the automatic generation\footnote{See \href{https://github.com/richfitz/stevedore/blob/master/development.md}{https://github.com/richfitz/stevedore/blob/master/development.md}.}.
@@
 
-**\CRANpkg{babelwhale}** allows executing and interacting with containers, which can use either Docker or Singularity as a backend \citep{cannoodt_babelwhale_2019}.
-The package provides a unified interface to interact with Docker and Singularity containers.
-Users can, for example, execute a command inside a container, mount a volume or copy a file.
+**\CRANpkg{babelwhale}** provides a unified interface to interact with Docker and Singularity containers \citep{cannoodt_babelwhale_2019}.
+Users can, for example, execute a command inside a container, mount a volume, or copy a file with the same R commands for both container runtimes.
 
-**\pkg{dockyard}** ([https://github.com/thebioengineer/dockyard](https://github.com/thebioengineer/dockyard)) has the goal of lowering barrier to creating `Dockerfile`s, building Docker images, and deploying Docker containers.
+**\pkg{dockyard}** ([https://github.com/thebioengineer/dockyard](https://github.com/thebioengineer/dockyard)) has the goal of lowering the barrier to creating `Dockerfile`s, building Docker images, and deploying Docker containers.
 The package follows the increasingly used piping paradigm of the `tidyverse` style of programming for chaining R functions representing the instructions in a `Dockerfile`.
 An existing `Dockerfile` can be used as a template.
-\pkg{dockyard} also includes wrappers for common steps, such as installing an R package or copying files, and build-in functions for building an image running a container, to make using Docker even more approachable to R users with a single API.
+\pkg{dockyard} also includes wrappers for common steps, such as installing an R package or copying files, and provides built-in functions for building an image and running a container, which make Docker more approachable within a single R-based user interface.
 
 **\pkg{dockermachine}** ([https://github.com/cboettig/dockermachine](https://github.com/cboettig/dockermachine)) is an R package to provide a convenient interface to [Docker&nbsp;Machine](https://docs.docker.com/machine/overview/) from R.
-The CLI tool `docker-machine` allows users to create and manage virtual host on local computers, local data centers, or at cloud providers.
+The CLI tool `docker-machine` allows users to create and manage a virtual host on local computers, local data centers, or at cloud providers.
 A local Docker installation can be configured to transparntly forward all commands issued on the local Docker CLI to a selected (remote) virtual host.
-Docker&nbsp;Machine was especially crucial for local use in early days of Docker, when no native support was available for Mac or Windows computers, but remains relevant for provisioning on remote systems.
-The package has not received any updates for two years, but is functional with a current version of `docker-machine` (`0.16.2`).
-It potentially lowers the barriers for R users to run containers on various hosts, if using the Docker&nbsp;Machine CLI directly is perceived as a barrier, or enables scripted workflows with remote processing.
+Docker&nbsp;Machine was especially crucial for local use in the early days of Docker, when no native support was available for Mac or Windows computers, but it remains relevant for provisioning on remote systems.
+The package has not received any updates for two years, but it is functional with a current version of `docker-machine` (`0.16.2`).
+It potentially lowers the barriers for R users to run containers on various hosts if they perceive that using the Docker&nbsp;Machine CLI directly as a barrier and it enables scripted workflows with remote processing.
 ```{r dockermachine, eval=FALSE, include=FALSE}
@@
 **Bioconductor** ([https://bioconductor.org/](https://bioconductor.org/)) is an open source, open development project for the analysis and comprehension of genomic data \citep{gentleman_bioconductor_2004}.
-The project consists of 1823 R software packages as of October 30th 2019, as well as packages containing annotation or experiment data.
-_Bioconductor_ has a semi-annual release cycle, each release is associated with a particular version of R, and Docker images are provided for current and past versions of _Bioconductor_ for convenience and reproducibility.
-All images, included  are described on the _Bioconductor_ web site (see [https://bioconductor.org/help/docker/](https://bioconductor.org/help/docker/)), created with `Dockerfile`s maintained on GitHub, and distributed through Docker&nbsp;Hub\footnote{See \href{https://github.com/Bioconductor/bioconductor_docker}{https://github.com/Bioconductor/bioconductor\_docker} and \href{https://hub.docker.com/u/bioconductor}{https://hub.docker.com/u/bioconductor} respectively.}.
-_Bioconductor_'s 'base' Docker images are built on top of the `rocker/rstudio` image.
-_Bioconductor_ installs packages based on the R version in combination with the Bioconductor version, and therefore uses Bioconductor version tagging `devel` and `RELEASE_X_Y`, e.g., `RELEASE_3_10`.
+As of October 30th 2019, the project consists of 1823 R software packages, as well as packages containing annotation or experiment data.
+_Bioconductor_ has a semi-annual release cycle, where each release is associated with a particular version of R, and Docker images are provided for current and past versions of _Bioconductor_ for convenience and reproducibility.
+All images, which are described on the _Bioconductor_ web site (see [https://bioconductor.org/help/docker/](https://bioconductor.org/help/docker/)), are created with `Dockerfile`s maintained on GitHub and distributed through Docker&nbsp;Hub\footnote{See \href{https://github.com/Bioconductor/bioconductor_docker}{https://github.com/Bioconductor/bioconductor\_docker} and \href{https://hub.docker.com/u/bioconductor}{https://hub.docker.com/u/bioconductor} respectively.}.
+_Bioconductor_'s "base" Docker images are built on top of the `rocker/rstudio` image.
+_Bioconductor_ installs packages based on the R version in combination with the Bioconductor version and, therefore, uses Bioconductor version tagging `devel` and `RELEASE_X_Y`, e.g., `RELEASE_3_10`.
 Past and current combinations of R and _Bioconductor_ will therefore be accessible via a specific image tags.
+
 The _Bioconductor_ `Dockerfile` selects the desired version of R from Rocker, adds required system dependencies, and uses the \CRANpkg{BiocManager} package for installing appropriate versions of _Bioconductor_ packages.
 A strength of this approach is that the responsibility for complex software configuration (including customized development) is shifted from the user to the experienced _Bioconductor_ core team.
-A recent audit of the _Bioconductor_  image stack `Dockerfile` led to the deprecation of several community-maintained images, because the numerous specific images became too hard to understand, complex to maintain, and cumbersome to extent.
+However, a recent audit of the _Bioconductor_  image stack `Dockerfile` led to the deprecation of several community-maintained images, because the numerous specific images became too hard to understand, complex to maintain, and cumbersome to customise.
 As part of the simplification, a recent innovation is to produce a `bioconductor_docker:devel` image to emulate the _Bioconductor_ nightly Linux build machine as closely as possible.
-This image contains the build system environment variables and the _system dependencies_ needed to install and check almost all (1813 out of 1823) _Bioconductor_ software packages and saves users and package developers from managing these themselves.
-Furthermore the image is configured so that `.libPaths()` has `/usr/local/lib/R/host-site-library` as the first location.
-Users mounting a location on the host file system to this location can persist installed packages across Docker sessions or image updates.
-Many R users pursue flexible work flows tailored to particular analysis needs, rather than standardized work flows.
-The new `bioconductor_docker` image is well-suited for this pattern, while `bioconductor_docker:devel` provides developers with a test environment close to _Bioconductor_'s build system.
+This image contains the build system environment variables and the _system dependencies_ needed to install and check almost all (1813 out of 1823) _Bioconductor_ software packages, and it saves users and package developers from managing these themselves.
+Furthermore, the image is configured so that `.libPaths()` has `/usr/local/lib/R/host-site-library` as the first location.
+Users mounting a location on the host file system to this location can persistently manage installed packages across Docker containers or image updates.
+Many R users pursue flexible work flows tailored to particular analysis needs rather than standardized work flows.
+The new `bioconductor_docker` image is well suited for this preference, while `bioconductor_docker:devel` provides developers with a test environment close to _Bioconductor_'s build system.
 
 \label{datascience}
-**Data Science** is a widely discussed topic among all academic disciplines \citep[e.g.,][]{donoho_50_2017}.
-The discussions shed a light on the tools and craftspersonship behind the analysis of data with computational methods.
-The practice of Data Science often involves a combination tools and software stacks and requires a cross-cutting skillset.
-This complexity and an inherent concern for openness and reproducibility in the Data Science community lead to Docker being used widely.
+**Data science** is a widely discussed topic in all academic disciplines \citep[e.g.,][]{donoho_50_2017}.
+These discussions have shed a light on the tools and craftspersonship behind the analysis of data with computational methods.
+The practice of data science often involves combining tools and software stacks and requires a cross-cutting skillset.
+This complexity and an inherent concern for openness and reproducibility in the data science community led to Docker being used widely.
 <!--A general description of what to consider when creating a Docker image used for data science projects in R is provided on [environments.rstudio.com](https://environments.rstudio.com/docker).-->
-The remaineder of this section presents exemplary Docker images and image stacks featuring R intended for Data Science.
-The [_Jupyter Docker Stacks_](https://github.com/jupyter/docker-stacks/) project are a set of ready-to-run Docker images containing Jupyter applications and interactive computing tools \citep{project_jupyter_jupyter_2018}.
-The `jupyter/r-notebook` image includes R and "popular packages", and naturally also the IRKernel ([https://irkernel.github.io/](https://irkernel.github.io/)), an R kernel for Jupyter so that Jupyter Notebooks can contain R code cells.
+The remaineder of this section presents exemple Docker images and image stacks featuring R intended for data science.
+
+- The [_Jupyter Docker Stacks_](https://github.com/jupyter/docker-stacks/) project is a set of ready-to-run Docker images containing Jupyter applications and interactive computing tools \citep{project_jupyter_jupyter_2018}.
+The `jupyter/r-notebook` image includes R and "popular packages", and naturally also the IRKernel ([https://irkernel.github.io/](https://irkernel.github.io/)), an R kernel for Jupyter, so that Jupyter Notebooks can contain R code cells.
 R is also included in the catchall `jupyter/datascience-notebook` image\footnote{\href{https://jupyter-docker-stacks.readthedocs.io/en/latest/using/selecting.html}{https://jupyter-docker-stacks.readthedocs.io/en/latest/using/selecting.html}}.
 For example, these images allow users to quickly start a Jupyter Notebook server locally or build their own specialised images on top of stable toolsets.
-R is installed using the Conda package manager, which can manage environments for various programming languages, pinning both the R version and the versions of R packages\footnote{See \texttt{jupyter/datascience-notebook}'s \texttt{Dockerfile} at \href{https://github.com/jupyter/docker-stacks/blob/master/datascience-notebook/Dockerfile\#L47}{https://github.com/jupyter/docker-stacks/blob/master/datascience-notebook/Dockerfile\#L47}.}.
-_Kaggle_ provides the [`gcr.io/kaggle-images/rstats`](https://hub.docker.com/r/kaggle/rstats) image (previously `kaggle/rstats`) and [corresponding `Dockerfile`](https://github.com/Kaggle/docker-rstats) for usage in their Machine Learning competitions and easy access to the associated datasets.
-It includes machine learning libraries such as Tensorflow and Keras (see also image `rocker/ml` in Section&nbsp;\nameref{rocker-gpu}), and also configures the \CRANpkg{reticulate} package.
-The image uses a base image with _all packages from CRAN_, `gcr.io/kaggle-images/rcran`, which requires a Google Cloud Build as Docker&nbsp;Hub would time out\footnote{Originally, a stacked collection of over 20 images with automated builds on Docker~Hub was used, see \href{https://web.archive.org/web/20190606043353/http://blog.kaggle.com/2016/02/05/how-to-get-started-with-data-science-in-containers/}{https://web.archive.org/web/20190606043353/http://blog.kaggle.com/2016/02/05/how-to-get-started-with-data-science-in-containers/} and \href{https://hub.docker.com/r/kaggle/rcran/dockerfile}{https://hub.docker.com/r/kaggle/rcran/dockerfile}}.
-The final extracted image size is over 25GB, which makes it debatable if having everything available is actually convenient.
-As a further example, [_Radiant_](https://radiant-rstats.github.io/docs/) project provides several images, e.g., [`vnijs/rsm-msba-spark`](https://hub.docker.com/r/vnijs/rsm-msba-spark), for their browser-based business analytics interface based on \CRANpkg{Shiny} (`Dockerfile` [on GitHub](https://github.com/radiant-rstats/docker)) and for use in education as part of an MSc course.
-As data science often applies a multitude of tools, this image favours inclusion over selection and features Python, Postgres, JupyterLab and Visual Studio Code besides R and RStudio, bringing the image size up to 9GB.
-_Gigantum_ ([http://gigantum.com/](http://gigantum.com/)) is a platform for open and decentralized data science with a focus on using automation and user-friendly tools for easy sharing of reproducible computational workflows.
+R is installed using the Conda package manager\footnote{\href{https://conda.io/}{https://conda.io/}}, which can manage environments for various programming languages, pinning both the R version and the versions of R packages\footnote{See \texttt{jupyter/datascience-notebook}'s \texttt{Dockerfile} at \href{https://github.com/jupyter/docker-stacks/blob/master/datascience-notebook/Dockerfile\#L47}{https://github.com/jupyter/docker-stacks/blob/master/datascience-notebook/Dockerfile\#L47}.}.
+- _Kaggle_ provides the [`gcr.io/kaggle-images/rstats`](https://hub.docker.com/r/kaggle/rstats) image (previously `kaggle/rstats`) and [corresponding `Dockerfile`](https://github.com/Kaggle/docker-rstats) for usage in their Machine Learning competitions and easy access to the associated datasets.
+It includes machine learning libraries such as Tensorflow and Keras (see also image `rocker/ml` in Section&nbsp;\nameref{rocker-gpu}), and it also configures the \CRANpkg{reticulate} package.
+The image uses a base image with _all packages from CRAN_, `gcr.io/kaggle-images/rcran`, which requires a Google Cloud Build because Docker&nbsp;Hub would time out\footnote{Originally, a stacked collection of over 20~images with automated builds on Docker~Hub was used, see \href{https://web.archive.org/web/20190606043353/http://blog.kaggle.com/2016/02/05/how-to-get-started-with-data-science-in-containers/}{https://web.archive.org/web/20190606043353/http://blog.kaggle.com/2016/02/05/how-to-get-started-with-data-science-in-containers/} and \href{https://hub.docker.com/r/kaggle/rcran/dockerfile}{https://hub.docker.com/r/kaggle/rcran/dockerfile}}.
+The final extracted image size is over 25&nbsp;GB, which calls into question whether having everything available is actually convenient.
+<!--As a further example, -->
+- The [_Radiant_](https://radiant-rstats.github.io/docs/) project provides several images, e.g., [`vnijs/rsm-msba-spark`](https://hub.docker.com/r/vnijs/rsm-msba-spark), for their browser-based business analytics interface based on \CRANpkg{Shiny} (`Dockerfile` [on GitHub](https://github.com/radiant-rstats/docker)) and for use in education as part of an MSc course.
+As data science often applies a multitude of tools, this image favours inclusion over selection and features Python, Postgres, JupyterLab and Visual Studio Code besides R and RStudio, bringing the image size up to 9&nbsp;GB.
+- _Gigantum_ ([http://gigantum.com/](http://gigantum.com/)) is a platform for open and decentralized data science with a focus on using automation and user-friendly tools for easy sharing of reproducible computational workflows.
 Gigantum builds on the _Gigantum Client_ (running either locally or on a remote server) for development and execution of data-focused _Projects_, which can be stored and shared via the _Gigantum Hub_ or via a zipfile export.
-The Client is a user friendly interface to a backend using Docker containers to package, build, and run Gigantum projects.
+The Client is a user-friendly interface to a backend using Docker containers to package, build, and run Gigantum projects.
 It is configured to use a default set of Docker base images ([https://github.com/gigantum/base-images](https://github.com/gigantum/base-images)), and users are able to define and configure their own custom images.
-The available images include two with R based on Ubuntu Linux have the [`c2d4u` CRAN PPA](https://launchpad.net/~marutter/+archive/ubuntu/c2d4u3.5/) pre-configured for installation of binary R packages\footnote{\href{https://docs.gigantum.com/docs/using-r}{https://docs.gigantum.com/docs/using-r}}.
+The available images include two with R based on Ubuntu Linux and these have the [`c2d4u` CRAN PPA](https://launchpad.net/~marutter/+archive/ubuntu/c2d4u3.5/) pre-configured for installation of binary R packages\footnote{\href{https://docs.gigantum.com/docs/using-r}{https://docs.gigantum.com/docs/using-r}}.
 The R images vary in the included authoring environment, i.e., Jupyter in `r-tidyverse` or both Jupyter \& RStudio in `rstudio-server`.
@@
 The R images are based on Gigantum's `python3-minimal` image, originally to keep the existing front-end configuration, but also to provide a consistent Python-to-R interoperability.
-The `Dockerfile`s also use build args to specify bases, for example for different version of NVIDIA CUDA for GPU processing\footnote{See \href{https://github.com/gigantum/base-images/blob/master/_templates/python3-minimal-template/Dockerfile}{https://github.com/gigantum/base-images/blob/master/\_templates/python3-minimal-template/Dockerfile} for the \texttt{Dockerfile} of \texttt{python3-minimal}.} so that appropriate GPU drivers can be enabled automatically when supported.
+The `Dockerfile`s also use build args to specify bases, for example for different versions of NVIDIA CUDA for GPU processing\footnote{See \href{https://github.com/gigantum/base-images/blob/master/_templates/python3-minimal-template/Dockerfile}{https://github.com/gigantum/base-images/blob/master/\_templates/python3-minimal-template/Dockerfile} for the \texttt{Dockerfile} of \texttt{python3-minimal}.}, so that appropriate GPU drivers can be enabled automatically when supported.
 Furthermore, Gigantum's focus lies on environment management via GUI and ensuring a smooth user interaction, e.g., with reliable and easy conflict detection and resolution.
@@
 The image build and publication process is scripted with Python and JSON template configuration files, unlike Rocker which relies on plain `Dockerfile`s.
-A minor reason in the inception of the images were also project constraints requiring a Rocker-incompatible licensing of the `Dockerfile`s, i.e., the MIT License.
+A further reason in the creation of an independent image stack were project constraints requiring a Rocker-incompatible licensing of the `Dockerfile`s, i.e., the MIT License.
 
@@
 
-Community maintained images provide a solid basis for extension as part of individual requirements.
-Several second order R packages attempt to make the process of creating Docker images and using containers for specific tasks, such as running tests or rendering reproducible reports, easier.
-While authoring and managing an environment with Docker by hand is possible and feasible for experts\footnote{See, e.g., this tutorial by RStudio on how to manage environments and package versions and to ensure deterministic image builds with Docker: \href{https://environments.rstudio.com/docker}{https://environments.rstudio.com/docker}.}, the following examples show the power of automation when environments become too cumbersome.
-Especially _version pinning_, with packages \pkg{remotes} and \pkg{versions} for R or by using MRAN, and with system package managers for different operating systems, can greatly increase the reproducibility of built images and are common approaches.
+Community-maintained images provide a solid basis so users can meet their own individual requirements.
+Several second-order R packages attempt to streamline the process of creating Docker images and using containers for specific tasks, such as running tests or rendering reproducible reports.
+While authoring and managing an environment with Docker by hand is possible and feasible for experts\footnote{See, e.g., this tutorial by RStudio on how to manage environments and package versions and to ensure deterministic image builds with Docker: \href{https://environments.rstudio.com/docker}{https://environments.rstudio.com/docker}.}, the following examples show that when environments become too cumbersome to create manually, automation is a powerful tool.
+In particular, the practice of _version pinning_, with system package managers for different operating systems and with packages \pkg{remotes} and \pkg{versions} or by using MRAN for R, can greatly increase the reproducibility of built images and are common approaches.
 
@@
 A scripted creation of a `Dockerfile` enables iteration and automation, for example for packaging applications for deployment (see \nameref{deployment}).
-Being scriptable from R developers can leverage the tools available in R to parse a `DESCRIPTION` file, to get system requirements, to list dependencies, versions, etc.
+Developers can retrieve system requirements and package dependencies to write a `Dockerfile`, for example, by leveraging the tools available in R to parse a `DESCRIPTION` file.
+
 **\pkg{containerit}** ([https://github.com/o2r-project/containerit/](https://github.com/o2r-project/containerit/)) attempts to take this one step further and includes these tools to automatically create a `Dockerfile` that can execute a given workflow \citep{nust_containerit_2019}.
@@
 \pkg{containerit} uses \CRANpkg{stevedore} to streamline the user interaction and improve the created `Dockerfile`s, e.g., by running a container for the desired base image to extract the already available R packages.
+
 **\pkg{dockr}** ([https://github.com/smaakage85/dockr](https://github.com/smaakage85/dockr)) is a very similar package attempting to mirror a given R session, including local non-CRAN packages\citep{kjeldgaard_dockr_2019}.
 Users can manually add statements for non-R dependencies to the `Dockerfile`.
+
 **\CRANpkg{liftr}** aims to solve the problem of persistent reproducible reporting in statistical computing based on the R Markdown format \citep{xie2018} for dynamic documents \citep[\href{https://nanx.me/liftr/}{https://nanx.me/liftr/}, ][]{liftr2019}.
@@
 The package then generates a `Dockerfile` from this metadata and provides a utitility function to render the document inside a Docker container, i.e., `render_docker("foo.Rmd")`.
-An RStudio addin even allows compilation of documents with a single push of a button.
+An RStudio addin even allows compilation of documents with the single push of a button.
 
 System dependencies are the domain of Docker, but for a full description of the computing environment one must also manage the R version and the the R packages.
-R&nbsp;versions are available via the versioned Rocker image stack \citep{RJ-2017-065}.
-[r-online](https://github.com/ColinFay/ronline) leverages these images and provides an app for helping users to detect breaking changes between different R versions, and for historic exploration of R.
+R versions are available via the versioned Rocker image stack \citep{RJ-2017-065}.
+[r-online](https://github.com/ColinFay/ronline) leverages these images and provides an app for helping users to detect breaking changes between different R versions and for historic exploration of R.
 With a standalone NodeJS app or [online](https://srv.colinfay.me/r-online), the user can compare a piece of code run in two separate versions of R.
 Internally, r-online opens one or two Docker instances with the given version of R based on Rocker images, executes a given piece of code, and returns the result to the user.
-R&nbsp;package management can be achieved with MRAN, or with packages such as \CRANpkg{checkpoint} and **\CRANpkg{renv}**, which can naturally be applied within images and containers.
+Regarding R package management, this can be achieved with MRAN, or with packages such as \CRANpkg{checkpoint} and **\CRANpkg{renv}**, which can naturally be applied within images and containers.
 For example, \CRANpkg{renv} ([https://rstudio.github.io/renv/](https://rstudio.github.io/renv/)) helps users to manage the state of the R library in a reproducible way, further providing isolation and portability \citep{renv2019}.
@@
 
-Containers can also serve as useful playgrounds to create environments ad-hoc or to provide very specific environments that are not needed or not easily available in day-to-day development for the purposes of developing R packages.
+Containers can also serve as useful playgrounds to create environments ad hoc or to provide very specific environments that are not needed or not easily available in day-to-day development for the purposes of developing R packages.
 These environments may have specific versions of R, of R extension packages, and of system libraries used by R extension packages, and all of the above in a specific combination.
@@
 Using the Rocker images with RStudio, these disposable environments lack no development comfort (cf. Section&nbsp;\nameref{compendia}).
-\citet{ooms_opencpu_2017} describes how `docker exec` can be used to get a root shell in a container for cusotmization during software evaluation.
-\citet{eddelbuettel_debugging_2019} describes an example how a Docker container was used to debug an issue with a package only occuring with a particular version of Fortran, and using tools which are not readily available on all platforms (e.g., not on macOS).
+\citet{ooms_opencpu_2017} describes how `docker exec` can be used to get a root shell in a container for cusotmisation during software evaluation.
+\citet{eddelbuettel_debugging_2019} describes an example of how a Docker container was used to debug an issue with a package only occuring with a particular version of Fortran, and using tools which are not readily available on all platforms (e.g., not on macOS).
 
@@
 GDAL is a system dependency allowing R packages such as \CRANpkg{sf}, which provides the core data model for geospatial vector data, or \CRANpkg{rgdal}, to accomodate users to be able to read and write hundreds of different spatial raster and vector formats.
-\CRANpkg{sf} and \CRANpkg{rgdal} have hundreds of indirect reverse imports and dependencies and therefore the maintainers spend a lot of efforts not to break these.
-Purpose built Docker are used to prepare for upcoming releases of system libraries, individual bug reports, and for the lowest supported versions of system libraries\footnote{Cf.
+\CRANpkg{sf} and \CRANpkg{rgdal} have hundreds of indirect reverse imports and dependencies and, therefore, the maintainers spend a lot of effort trying not to break them.
+Purpose-built Docker are used to prepare for upcoming releases of system libraries, individual bug reports, and for the lowest supported versions of system libraries\footnote{Cf.
 \href{https://github.com/r-spatial/sf/tree/master/inst/docker}{https://github.com/r-spatial/sf/tree/master/inst/docker}, \href{https://github.com/Nowosad/rspatial_proj6}{https://github.com/Nowosad/rspatial\_proj6}, and \href{https://github.com/r-spatial/sf/issues/1231}{https://github.com/r-spatial/sf/issues/1231}}.
@@
 
-Third, there are special purpose images for identifying problems beyond the mere R code, such as **debugging R memory problems**.
-The images significantly reduce the barrier to follow complex steps for fixing memory allocation bugs \citep[cf. Section~4.3 in][]{core_writing_1999}.
-These problems are hard to debug and critical, both because when they occur they lead to fatal crashing processes.
-[`rocker/r-devel-san`](https://github.com/rocker-org/r-devel-san) and [`rocker/r-devel-ubsan-clang`](https://github.com/rocker-org/r-devel-san-clang) are Docker images have a particularly configured version of R to trace such problems with gcc and clang compilers, respectively \citep[cf.~\CRANpkg{sanitizers} for examples,][]{eddelbuettel_sanitizers_2014}.
-The image [`wch/r-debug`](https://github.com/wch/r-debug) is a purpose built Docker image with _multiple_ instrumented builds of R, each with a different diagnostic utility activated.
+Third, special-purpose images exist for identifying problems beyond the mere R code, such as **debugging R memory problems**.
+These images significantly reduce the barriers to following complex steps for fixing memory allocation bugs \citep[cf. Section~4.3 in][]{core_writing_1999}.
+These problems are hard to debug and critical, because when they do occur they lead to fatal crashing processes.
+[`rocker/r-devel-san`](https://github.com/rocker-org/r-devel-san) and [`rocker/r-devel-ubsan-clang`](https://github.com/rocker-org/r-devel-san-clang) are Docker images that have a particularly configured version of R to trace such problems with gcc and clang compilers, respectively \citep[cf.~\CRANpkg{sanitizers} for examples,][]{eddelbuettel_sanitizers_2014}.
+[`wch/r-debug`](https://github.com/wch/r-debug) is a purpose-built Docker image with _multiple_ instrumented builds of R, each with a different diagnostic utility activated.
 
 Fourth, containers are useful for **testing** R code during development.
-To submit a package to CRAN, an R package must work with the development version of R, which must be compiled locally.
-That can be a challenge for some users.
-The **R-hub** project provides _"a collection of services to help R package development"_, with the package builder as the most prominent one\citep{r-hub_docs_2019}.
-R-hub makes it easy to ensure that no errors occur, but to fix errors a local setup is still often warranted, e.g., using the image `rocker/r-devel`, and to test packages with native code, which can make the process more complex \citep[cf.][]{eckert_building_2018}.
+To submit a package to CRAN, an R package must work with the development version of R, which must be compiled locally; this can be a challenge for some users.
+The **R-hub** project provides _"a collection of services to help R package development"_, with the package builder as the most prominent one \citep{r-hub_docs_2019}.
+R-hub makes it easy to ensure that no errors occur, but fixing errors still often warrants a local setup, e.g., using the image `rocker/r-devel`, as is testing packages with native code, which can make the process more complex \citep[cf.][]{eckert_building_2018}.
 The R-hub Docker images can also be used to debug problems locally using various combinations of Linux platforms, R versions, and compilers\footnote{See \href{https://r-hub.github.io/rhub/articles/local-debugging.html}{https://r-hub.github.io/rhub/articles/local-debugging.html} and \href{https://blog.r-hub.io/2019/04/25/r-devel-linux-x86-64-debian-clang/}{https://blog.r-hub.io/2019/04/25/r-devel-linux-x86-64-debian-clang/}}.
-The images go beyond the configurations, or _flavours_, used by CRAN for checking packages\footnote{\href{https://cran.r-project.org/web/checks/check_flavors.html}{https://cran.r-project.org/web/checks/check\_flavors.html}}, e.g., with CentOS-based images, but lack a container for checking on Windows or OS X.
-The images greatly support package developers to provide support on operating systems they are not familiar.
+The images go beyond the configurations, or _flavours_, used by CRAN for checking packages\footnote{\href{https://cran.r-project.org/web/checks/check_flavors.html}{https://cran.r-project.org/web/checks/check\_flavors.html}}, e.g., with CentOS-based images, but they lack a container for checking on Windows or OS X.
+The images greatly support package developers to provide support on operating systems with which they are not familiar.
 The package \pkg{dockertest} ([https://github.com/traitecoevo/dockertest/](https://github.com/traitecoevo/dockertest/)) is a proof of concept for automatically generating `Dockerfile`s and building images specifically to run tests\footnote{\pkg{dockertest} is not actively maintained, but mentioned still because of its interesting approach.}.
-These images are accompanied with a special launch script so the tested source code is not stored in the image but the currently checked in version from a local Git repository is cloned into the container at runtime.
+These images are accompanied with a special launch script so the tested source code is not stored in the image; instead, the currently checked in version from a local Git repository is cloned into the container at runtime.
 This approach clearly seperates test environment, test code, and current working copy of the code.
-\label{rselenium} Another use case where a container helps to standardise tests across operating systems is detailed the vignettes of the package \CRANpkg{RSelenium} \citep{rselenium_2019}.
-The package recommends Docker for running the [Selenium](https://selenium.dev/) Server application needed to execute test suites on browser-based user interfaces and webpages, but requires users to manually manage the Docker containers.
+\label{rselenium} Another use case where a container can help to standardise tests across operating systems is detailed the vignettes of the package \CRANpkg{RSelenium} \citep{rselenium_2019}.
+The package recommends Docker for running the [Selenium](https://selenium.dev/) Server application needed to execute test suites on browser-based user interfaces and webpages, but it requires users to manually manage the containers.
 
@@
 \citet{ye_docker_2019} describes how they speed up the process of testing by running tasks on [Travis&nbsp;CI](https://travis-ci.org/) within a container using `docker exec`, e.g., the package check or rendering of documentation.
-\citet{cardozo_faster_2018} saved time, also on Travis&nbsp;CI, by re-using the testing image as the base for an image intended for publication on Docker&nbsp;Hub.
-[`r-ci`](https://github.com/ColinFay/r-ci) is in turn used with [GitLab&nbsp;CI](https://docs.gitlab.com/ee/ci/), which itself is built on top of Docker images: the user specifies a base Docker image, and the whole tests are run inside this environment.
+\citet{cardozo_faster_2018} also saved time with Travis&nbsp;CI by re-using the testing image as the basis for an image intended for publication on Docker&nbsp;Hub.
+[`r-ci`](https://github.com/ColinFay/r-ci) is, in turn, used with [GitLab&nbsp;CI](https://docs.gitlab.com/ee/ci/), which itself is built on top of Docker images: the user specifies a base Docker image and control code, and the whole set of tests is run inside a container.
 The `r-ci` image stack combines `rocker` versioning and a series of tools specifically designed for testing in a fixed environment with a customized list of preinstalled packages.
-Especially for long running tests or complex system dependencies, these approaches to seperate installation of build dependencies with code testing streamline the development process.
-Not due to a concern about time, but to control the environment used on a CI server, even this manuscript is rendered into a PDF and deployed to a GitHub-hosted website with every change (see `.travis.yml` and `Dockerfile` in the manuscript repository).
-This gives on the one hand easy access after every update of the R Markdown source code, and on the other hand a second controlled environment making sure that the article renders successfully and correctly.
+Especially for long-running tests or complex system dependencies, these approaches to seperate installation of build dependencies with code testing streamline the development process.
+Furthermore, with each change, even this manuscript is rendered into a PDF and deployed to a GitHub-hosted website (see `.travis.yml` and `Dockerfile` in the manuscript repository), not because of concern about time, but to control the environment used on a CI server.
+This gives, on the one hand, easy access after every update of the R Markdown source code and, on the other hand a second controlled environment to make sure that the article renders successfully and correctly.
 
@@
 
-The portability of containerized environments becomes particularly useful for improving expensive processing of data or shipping complex processing pipelines.
-First, it is possible to **offload complex processing to a server** or clouds, and also parallel executing of processes for speeding up or serving of many users.
-**\CRANpkg{batchtools}** provides a parallel implementation of Map for various schedulers \citep{Lang2017batchtools}.
-The package can [schedule jobs with Docker Swarm](https://mllg.github.io/batchtools/reference/makeClusterFunctionsDocker.html)
-**\CRANpkg{googleComputeEngineR}** has the function `gce_vm_cluster()` to create clusters of 2 or more virtual machines, running multi-CPU architectures.
+The portability of containerised environments becomes particularly useful for improving expensive processing of data or shipping complex processing pipelines.
+First, it is possible to **offload complex processing to a server** or clouds and also to execute proccesses in parallel to speed up or to serve many users.
+**\CRANpkg{batchtools}** provides a parallel implementation of the Map function for various schedulers \citep{Lang2017batchtools}.
+For example, the package can [schedule jobs with Docker Swarm](https://mllg.github.io/batchtools/reference/makeClusterFunctionsDocker.html).
+**\CRANpkg{googleComputeEngineR}** has the function `gce_vm_cluster()` to create clusters of 2 or more virtual machines, running multi-CPU architectures \citep{googleComputeEngineR_2019}.
 Instead of running a local R script with the local CPU and RAM restrictions, the same code can be processed on all CPU threads of the cluster of machines in the cloud, all running a Docker container with the same R environments.
-\CRANpkg{googleComputeEngineR} integrates with the R parralisation package \CRANpkg{future} to enable this with only a few lines or R code\footnote{\href{https://cloudyr.github.io/googleComputeEngineR/articles/massive-parallel.html}{https://cloudyr.github.io/googleComputeEngineR/articles/massive-parallel.html}}.
+\CRANpkg{googleComputeEngineR} integrates with the R parallelisation package \CRANpkg{future} to enable this with only a few lines or R code\footnote{\href{https://cloudyr.github.io/googleComputeEngineR/articles/massive-parallel.html}{https://cloudyr.github.io/googleComputeEngineR/articles/massive-parallel.html}}.
 [Google Cloud Run](https://cloud.run) is a CaaS (Containers as a Service) platform.
-Users can launch containers using any Docker image without worrying about underlying infrastructure in a so called serverless configuration.
-The service takes care of network ingress, scaling machines up and down, authentication and authorisation---all features which are non-trivial for a developer to create on their own.
+Users can launch containers using any Docker image without worrying about underlying infrastructure in a so-called serverless configuration.
+The service takes care of network ingress, scaling machines up and down, authentication, and authorisation---all features which are non-trivial for a developer to build and maintain on their own.
 This can be used to scale up R code to millions of instances if need be with little or no changes to existing code, as demonstrated by the proof of concept `cloudRunR`\footnote{\href{https://github.com/MarkEdmondson1234/cloudRunR}{https://github.com/MarkEdmondson1234/cloudRunR}}, which uses Cloud Run to create a scalable R-based API using \CRANpkg{plumber}.
-[Google Cloud Build](https://cloud.google.com/cloud-build/) and the Google Container Registry are a continuous integration service respectively image registry that offload building of images to the cloud, while serving the needs of commercial environments such as private Docker images or image stacks.
-As Googl Cloud Build itself can run any container, the package \pkg{googleCloudRunner} demonstrates how R can be used as the control language for one time or batch processing jobs and scheduling of jobs\footnote{\href{https://code.markedmondson.me/googleCloudRunner/articles/cloudbuild.html}{https://code.markedmondson.me/googleCloudRunner/articles/cloudbuild.html}}.
+[Google Cloud Build](https://cloud.google.com/cloud-build/) and the Google Container Registry are a continuous integration service and an image registry, respectively, that offload building of images to the cloud, while serving the needs of commercial environments such as private Docker images or image stacks.
+As Google Cloud Build itself can run any container, the package \pkg{googleCloudRunner} demonstrates how R can be used as the control language for one-time or batch processing jobs and scheduling of jobs\footnote{\href{https://code.markedmondson.me/googleCloudRunner/articles/cloudbuild.html}{https://code.markedmondson.me/googleCloudRunner/articles/cloudbuild.html}}.
 \label{drake} **\CRANpkg{drake}** is a workflow manager for data science projects \citep{landau_drake_2019}.
 It features implicit parallel computing and automated detection of the parts of the work that actually needs to be reexecuted.
-\CRANpkg{drake} has been demonstrated to run inside containers for high reproducibility\footnote{See for example \href{https://github.com/joelnitta/pleurosoriopsis}{https://github.com/joelnitta/pleurosoriopsis} or \href{https://gitlab.com/ecohealthalliance/drake-gitlab-docker-example}{https://gitlab.com/ecohealthalliance/drake-gitlab-docker-example}, the latter even running in a continuous integration platform (cf.~\nameref{development}.}, but also how to use \CRANpkg{future} package's function `makeClusterPSOCK()` to send parts of the workflow to a Docker image for execution\footnote{\href{https://docs.ropensci.org/drake/index.html?q=docker\#with-docker}{https://docs.ropensci.org/drake/index.html?q=docker\#with-docker}} \citep[see package's function documentation;~][]{future_2020}.
-In the latter case, the container control code must be written by the user and the \CRANpkg{future} package ensures the host and worker can connect for communicating over socket connections.
+\CRANpkg{drake} has been demonstrated to run inside containers for high reproducibility\footnote{See for example \href{https://github.com/joelnitta/pleurosoriopsis}{https://github.com/joelnitta/pleurosoriopsis} or \href{https://gitlab.com/ecohealthalliance/drake-gitlab-docker-example}{https://gitlab.com/ecohealthalliance/drake-gitlab-docker-example}, the latter even running in a continuous integration platform (cf.~\nameref{development}.}.
+Futhermore, \CRANpkg{drake} workflwos have been shown to use \CRANpkg{future} package's function `makeClusterPSOCK()` for sending parts of the workflow to a Docker image for execution\footnote{\href{https://docs.ropensci.org/drake/index.html?q=docker\#with-docker}{https://docs.ropensci.org/drake/index.html?q=docker\#with-docker}} \citep[see package's function documentation;~][]{future_2020}.
+In the latter case, the container control code must be written by the user, and the \CRANpkg{future} package ensures that the host and worker can connect for communicating over socket connections.
 **RStudio Server Pro** includes a functionality called [Launcher](https://solutions.rstudio.com/launcher/overview/) (since version&nbsp;1.2, released in 2019).
@@
 Second, containers are perfectly suited for **packaging and executing software pipelines** and required data.
-Containers allow building complex processing pipelines that are independent from the host programming language.
-Due to its original use case (see&nbsp;\nameref{introduction}), Docker has no standard mechanisms for chaining containers together; it lacks definitions and protocols for environment variables, volume mounts and/or ports that could enable transfer of input (parameters and data) and output (results) to and from containers.
-Some packages, e.g., \pkg{containerit}, do provide Docker images that can be used similarly to CLIs, but their usage is cumbersome\footnote{\href{https://o2r.info/containerit/articles/container.html}{https://o2r.info/containerit/articles/container.html}}.
-**\pkg{outsider}** ([https://docs.ropensci.org/outsider/](https://docs.ropensci.org/outsider/)) tackles the problem of integrating external programs into an R workflow \citep{bennett_outsider_2020}.
+Containers allow for building complex processing pipelines that are independent of the host programming language.
+Due to its original use case (see&nbsp;\nameref{introduction}), Docker has no standard mechanisms for chaining containers together; it lacks definitions and protocols for how to use environment variables, volume mounts, and/or ports that could enable transfer of input (parameters and data) and output (results) to and from containers.
+Some packages, e.g., \pkg{containerit}, provide Docker images that can be used very similar to a CLI, but this usage is cumbersome\footnote{\href{https://o2r.info/containerit/articles/container.html}{https://o2r.info/containerit/articles/container.html}}.
+**\pkg{outsider}** ([https://docs.ropensci.org/outsider/](https://docs.ropensci.org/outsider/)) tackles the problem of integrating external programs into an R workflow without thee need for users to directly interact with containers \citep{bennett_outsider_2020}.
 Installation and usage of external programs can be difficult, convoluted and even impossible if the platform is incompatible.
-Therefore \pkg{outsider} uses the platform-independent Docker images to encapsulate processes in _outsider modules_.
+Therefore, \pkg{outsider} uses the platform-independent Docker images to encapsulate processes in _outsider modules_.
 Each outsider module has a `Dockerfile` and an R package with functions for interacting with the encapsulated tool.
-Using only R functions, an end-user can install a module with the \pkg{outsider} package and then call module code to integrate a tool into their own R-based workflow seamlessly.
+Using only R functions, an end user can install a module with the \pkg{outsider} package and then call module code to seamlessly integrate a tool into their own R-based workflow.
 The \pkg{outsider} package and module manage the containers and handle the transmission of arguments and the transfer of files to and from a container.
@@
 
-The cloud is the natural environment of containers, and subsequently containers are the go-to mechanism to deploy R server applications.
+The cloud is the natural environment of containers, and, subsequently, containers are the go-to mechanism to deploy R server applications.
 More and more continuous integration (CI) and continuous delivery (CD) services also use containers, opening up new options for use.
@@
 First, different packages use containers for the **deployment of R and \CRANpkg{Shiny} apps**.
-\CRANpkg{Shiny} is a popular package for creating interactive online dashboards with R and it enables users with very diverse backgrounds to create stable and user friendly web applications.
-For example, _ShinyProxy_ ([https://www.shinyproxy.io/](https://www.shinyproxy.io/)) is an open source tool to deploy Shiny apps in an enterprise context.
-They feature single sign-on, but also in scientific use cases \citep[e.g., ][]{savini_epiexplorer_2019,glouzon_structurexplor_2017}.
+\CRANpkg{Shiny} is a popular package for creating interactive online dashboards with R, and it enables users with very diverse backgrounds to create stable and user-friendly web applications.
+_ShinyProxy_ ([https://www.shinyproxy.io/](https://www.shinyproxy.io/)) is an open source tool to deploy Shiny apps in an enterprise context, where it features single sign-on, but it can also be used in scientific use cases \citep[e.g., ][]{savini_epiexplorer_2019,glouzon_structurexplor_2017}.
 ShinyProxy uses Docker containers to isolate user sessions and to achieve scalability for multi-user scenarious with multiple apps.
 ShinyProxy itself is written in Java to accomodate corporate requirements and may itself run in a container for stability and availability.
-The tool is build on ContainerProxy ([https://www.containerproxy.io/](https://www.containerproxy.io/)), which provides similar features for executign long-running R jobs or interactive R sessions.
+The tool is built on ContainerProxy ([https://www.containerproxy.io/](https://www.containerproxy.io/)), which provides similar features for executing long-running R jobs or interactive R sessions.
 The started containers can run on a regular Docker host but also in clusters.
-Another example is the package \CRANpkg{golem} ([https://github.com/ThinkR-open/golem](https://github.com/ThinkR-open/golem)), which makes an heavy use of `dockerfiler` when it comes to creating the `Dockerfile` for building production-grade Shiny applications and deploying them.
+Another example is the package \CRANpkg{golem} ([https://github.com/ThinkR-open/golem](https://github.com/ThinkR-open/golem)), which makes an heavy use of `dockerfiler` when it comes to creating the `Dockerfile` for building and deploying production-grade Shiny applications.
 \CRANpkg{googleComputeEngineR} ([https://cloudyr.github.io/googleComputeEngineR/](https://cloudyr.github.io/googleComputeEngineR/)) enables quick deployments of key R services, such as RStudio and Shiny, onto cloud virtual machines (VMs) with Google Cloud Compute Engine \citep{googleComputeEngineR_2019}.
 The package utilises `Dockerfile`s to move the labour of setting up those services from the user to a premade Docker image, which is configured and run in the cloud VM.
-For example, by specifying the template `template="rstudio"` in functions `gce_vm_template()/gce_vm()` an up to date RStudio Service image is launched for development work, while specifying `template="rstudio-gpu"` will launch an RStudio Server image with a GPU attached, etc.
+For example, by specifying the template `template="rstudio"` in functions `gce_vm_template()/gce_vm()` an up-to-date RStudio Server image is launched for development work, whereas specifying `template="rstudio-gpu"` will launch an RStudio Server image with a GPU attached, etc.
 
@@
 The [OpenCPU](https://www.opencpu.org/) system provides an HTTP API for data analysis based on R.
-\citet{ooms_opencpu_2017} describes how various platform-specific installation files for OpenCPU are created using Docker&nbsp;Hub: the automated builds install the software stack from source on different operating systems; afterwards a script file downloads the images and extracts the OpenCPU binaries.
+\citet{ooms_opencpu_2017} describes how various platform-specific installation files for OpenCPU are created using Docker&nbsp;Hub.
+The automated builds install the software stack from the source code on different operating systems; afterwards a script file downloads the images and extracts the OpenCPU binaries.
 
@@
 A cluster features load-balancing, autoscaling of containers across numerous servers (in the cloud or on premise), and restarting failed ones.
-It may be your organisation has a Kubernetes cluster already for other applications, or you may use one of the different provides for managed Kubernetes clusters.
+Many organisations already use a Kubernetes cluster for other applications, or a managed cluster can be acquired from service providers.
 Docker containers are used within Kubernetes clusters to hold native code, for which Kubernetes creates a framework around network connections and scaling of resources up and down.
-R applications, big parallel tasks, or scheduled batch jobs can be deployed in a scalable way using Kubernetes, and the deployment can even be triggered by changes to code repositories (i.e., CD), see blog post "R on Kubernetes" \citep{edmondson_r_2018}.
-The package \pkg{googleKubernetesR} ([https://github.com/RhysJackson/googleKubernetesR](https://github.com/RhysJackson/googleKubernetesR)) is a proof of concept for wrapping the Google Kubernetes Engine API, Google's hosted Kubernetes solution, in an easy to use R package.
-The package \CRANpkg{analogsea} provides a solution to programmatically create and destroy cloud VMs on the [Digital Ocean](https://www.digitalocean.com/) platform \citep{analogsea_2019}.
+Kubernetes can thereby host R applications, big parallel tasks, or scheduled batch jobs in a scalable way, and the deployment can even be triggered by changes to code repositories \citep[i.e., CD, see][]{edmondson_r_2018}.
+The package \pkg{googleKubernetesR} ([https://github.com/RhysJackson/googleKubernetesR](https://github.com/RhysJackson/googleKubernetesR)) is a proof of concept for wrapping the Google Kubernetes Engine API, Google's hosted Kubernetes solution, in an easy-to-use R package.
+The package \CRANpkg{analogsea} provides a way to programmatically create and destroy cloud VMs on the [Digital Ocean](https://www.digitalocean.com/) platform \citep{analogsea_2019}.
 It also includes R wrapper functions to install Docker in such a VM, manage images, and control containers straight from R functions.
 These functions are translated to Docker CLI commands and transferred transparently to the respective remove machine using SSH.
-\CRANpkg{AzureContainers} is an umbrella package providing interfaces three commercial services of Microsoft's Azure Cloud, namely [Container Instances](https://azure.microsoft.com/en-us/services/container-instances/) for running individual containers, [Container Registry](https://azure.microsoft.com/en-us/services/container-registry/) for private image distribution, and [Kubernetes Service](https://azure.microsoft.com/en-us/services/kubernetes-service/) for orchestrated deployments.
-While a package like \CRANpkg{plumber} provides the infrastructure for turning an R workflow a service, for production purposes it is usually necessary to take into account scalability, reliability and ease of management.
-AzureContainers provides an R-based interface to these features and thereby simplifies complex infrastructure management to a number of R function calls, given an Azure account with sufficient credit\footnote{See \emph{"Deploying a prediction service with Plumber"} vignette for details:  \href{https://cran.r-project.org/web/packages/AzureContainers/vignettes/vig01_plumber_deploy.html}{https://cran.r-project.org/web/packages/AzureContainers/vignettes/vig01\_plumber\_deploy.html}.}.
-[Heroku](https://www.heroku.com/) is a further cloud platform as a service provider, who supports container-based applications.
+\CRANpkg{AzureContainers} is an umbrella package that provides interfaces for three commercial services of Microsoft's Azure Cloud, namely [Container Instances](https://azure.microsoft.com/en-us/services/container-instances/) for running individual containers, [Container Registry](https://azure.microsoft.com/en-us/services/container-registry/) for private image distribution, and [Kubernetes Service](https://azure.microsoft.com/en-us/services/kubernetes-service/) for orchestrated deployments.
+While a package like \CRANpkg{plumber} provides the infrastructure for turning an R workflow into a web service, for production purposes it is usually necessary to take into account scalability, reliability and ease of management.
+AzureContainers provides an R-based interface to these features and, thereby, simplifies complex infrastructure management to a number of R function calls, given an Azure account with sufficient credit\footnote{See \emph{"Deploying a prediction service with Plumber"} vignette for details:  \href{https://cran.r-project.org/web/packages/AzureContainers/vignettes/vig01_plumber_deploy.html}{https://cran.r-project.org/web/packages/AzureContainers/vignettes/vig01\_plumber\_deploy.html}.}.
+[Heroku](https://www.heroku.com/) is another cloud platform as a service provider, and it supports container-based applications.
 `heroku-docker-r` ([https://github.com/virtualstaticvoid/heroku-docker-r](https://github.com/virtualstaticvoid/heroku-docker-r)) is an independent project providing a template for deploying R applications based on Heroku's image stack, including multiple examples for interfacing R with other programming languages.
@@
 
-The prevelance of Docker in industry naturally leads to usage of containers with R in such settings as well, as customers already manage platforms in Docker containers.
-These products often entail a high amount of open source software in combination with proprietary layers adding the relevant commercializable features.
+The prevelance of Docker in industry naturally leads to the use of containers with R in such settings as well, as companies already manage platforms in Docker containers.
+These products often entail a large amount of open source software in combination with proprietary layers adding the relevant commercializable features.
 One such example is RStudio's data science platform [RStudio Team](https://rstudio.com/products/team/).
@@
 The best practices for [running RStudio with Docker containers](https://support.rstudio.com/hc/en-us/articles/360021594513-Running-RStudio-with-Docker-containers) as well as [Docker images](https://github.com/rstudio/rstudio-docker-products) for RStudio's commerical products are publicly available.
+
 ## Using R to power enterprise software in production environments
@@
 However, thanks to advancements in R running as a web service, along with with the ability to deploy R in Docker containers, modern enterprises are now capable of having real-time machine learning powered by R.
-A number of packages and projects enabled R to respond to client reqests over TCP/IP and local socket servers, such as \CRANpkg{Rserve}, \CRANpkg{svSocket}, [rApache](http://www.rapache.net) and more recently \CRANpkg{plumber} ([https://www.rplumber.io/](https://www.rplumber.io/)) and \pkg{RestRserve} ([http://restrserve.org](http://restrserve.org)), which even processes incoming requests in parallel with forked processes using \CRANpkg{Rserve}.
-The latter two also provide documentation for deployment with Docker or ready to use automated builds of images\footnote{See \href{https://www.rplumber.io/docs/hosting.html\#docker}{https://www.rplumber.io/docs/hosting.html\#docker}, \href{https://hub.docker.com/r/trestletech/plumber/}{https://hub.docker.com/r/trestletech/plumber/} and \href{https://hub.docker.com/r/rexyai/restrserve/}{https://hub.docker.com/r/rexyai/restrserve/}.}.
+A number of packages and projects have enabled R to respond to client reqests over TCP/IP and local socket servers, such as \CRANpkg{Rserve}, \CRANpkg{svSocket}, [rApache](http://www.rapache.net) and more recently \CRANpkg{plumber} ([https://www.rplumber.io/](https://www.rplumber.io/)) and \pkg{RestRserve} ([http://restrserve.org](http://restrserve.org)), which even processes incoming requests in parallel with forked processes using \CRANpkg{Rserve}.
+The latter two also provide documentation for deployment with Docker or ready-to-use images with automated builds\footnote{See \href{https://www.rplumber.io/docs/hosting.html\#docker}{https://www.rplumber.io/docs/hosting.html\#docker}, \href{https://hub.docker.com/r/trestletech/plumber/}{https://hub.docker.com/r/trestletech/plumber/} and \href{https://hub.docker.com/r/rexyai/restrserve/}{https://hub.docker.com/r/rexyai/restrserve/}.}.
 These software allow other (remote) processes and programming languages to interact with R and to expose R-based function in a service architecture with HTTP APIs.
-APIs based on these package can be deployed with scalability and high availability using containers.
-This pattern of deploying code matches those used by software engineering services created in more established languages in the enterprise domain, such as Java or Python, and R can be used alongside those langages as a first class member of a software engineering technical stack.
+APIs based on these packages can be deployed with scalability and high availability using containers.
+This pattern of deploying code matches those used by software engineering services created in more established languages in the enterprise domain, such as Java or Python, and R can be used alongside those langages as a first-class member of a software engineering technical stack.
 
-CARD.com implemented a web application for the optimization of the acquisition flow and the real-time analysis of debit card transactions.
+CARD.com implemented a web application for the optimisation of the acquisition flow and the real-time analysis of debit card transactions.
 The software used \CRANpkg{Rserve} and rApache and was deployed in Docker containers.
@@
 The related Docker images were made available at [https://github.com/cardcorp/card-rocker](https://github.com/cardcorp/card-rocker).
-The images extend `rocker/r-base` and additionally entailed an SSH client and a workaround for being able to mount SSH keys from the host, Pandoc, the Amazon Web Services (AWS) SDK, and Java, which is required by the AWS SDK.
-The AWS SDK allowed to run R consumers reading from real-time data processing streams of [AWS Kinesis](https://aws.amazon.com/kinesis/) \footnote{See useR!2017 talk \href{https://static.sched.com/hosted\_files/user2017/2f/AWR Kinesis at useR 2017.pdf}{"Stream processing with R in AWS"}.}.
+The images extended `rocker/r-base` and additionally entailed an SSH client and a workaround for being able to mount SSH keys from the host, Pandoc, the Amazon Web Services (AWS) SDK, and Java, which is required by the AWS SDK.
+The AWS SDK allowed for running R consumers reading from real-time data processing streams of [AWS Kinesis](https://aws.amazon.com/kinesis/) \footnote{See useR!2017 talk \href{https://static.sched.com/hosted\_files/user2017/2f/AWR Kinesis at useR 2017.pdf}{"Stream processing with R in AWS"}.}.
 The applications were deployed on Amazon Elastic Container Service ([ECS](https://aws.amazon.com/ecs/)).
-The main learnings from using R in Docker was the imporance of not only pinning the R package versions via MRAN, but also moving away from Debian testing to a distribution with long-term support.
-For the use case at hand, this switch served the priority to have more control over upstream updates, and to minimize the risk of breaking the automated builds of the Docker images and production jobs.
+The main take-aways from using R in Docker were not only that pinning the R package versions via MRAN is important, but also that moving away from Debian testing to a distribution with long-term support can be necessary.
+For the use case at hand, this switch allowed for more control over upstream updates and for minimising the risk of breaking the automated builds of the Docker images and production jobs.
 
 The AI @ T-Mobile team created a set of neural network machine learning natural language processing models to help customer care agents manage text-based messages for customers \citep{t-mobile_enterprise_2018}.
-For example, one model quickly identifies if a message is from a customer or not \citep[see \CRANpkg{Shiny}-based \href{https://secure.message.t-mobile.com/v1/shiny/is-customer/app/}{demo}, ][]{nolis_small_2019}, others tell which customers are likely to make a repeat purchase.
+For example, one model quickly identifies whether a message is from a customer \citep[see \CRANpkg{Shiny}-based \href{https://secure.message.t-mobile.com/v1/shiny/is-customer/app/}{demo} further described by ][]{nolis_small_2019}, and others tell which customers are likely to make a repeat purchase.
 If a data scientist creates a machine learning model and exposes it through a \CRANpkg{plumber} API, then someone else on the marketing team could write software that sends different emails depending on that real-time prediction.
@@
 Since the models power tools for agents and customers, they need to have extremely high uptime and reliability.
-The AI @ T-Mobile team found that the models performed well and today these models power real-time services that are called over a million times a day.
+The AI @ T-Mobile team found that the models performed well, and today these models power real-time services that are called over a million times a day.
 
@@
 The fact that Docker images are portable and well defined make them useful when more than one person needs access to the same computing environment.
-This is even more useful when some of the users do not have the expertise to create such an environment themselves, and when these environments can be run in public or shared infrastructure.
+This is even more useful when some of the users do not have the expertise to create such an environment themselves, and when these environments can be run in public or using shared infrastructure.
+For example, _RCloud_ ([https://rcloud.social](https://rcloud.social)) is a cloud-based platform for data analysis, visualisation and collaboration using R.
+It provides a `rocker/drd` base image for easy evaluation of the platform\footnote{\href{https://github.com/att/rcloud/tree/master/docker}{https://github.com/att/rcloud/tree/master/docker}}.
 
@@
 Binder's core internal tool is `repo2docker`.
-It deterministically builds a Docker image by parsing the contents of a repository, e.g., project dependency configurations or simple configuration files\footnote{See supported file types at \href{https://repo2docker.readthedocs.io/en/latest/config\_files.html}{https://repo2docker.readthedocs.io/en/latest/config\_files.html}.}.
+It deterministically builds a Docker image by parsing the contents of a repository, e.g., project dependency configurations or simple configuration files\footnote{See supported file types at \href{https://repo2docker.readthedocs.io/en/latest/config\_files.html}{https://repo2docker.readthedocs.io/en/latest/config\_files.html}.
+For R, the }.
 In the most powerful case, `repo2docker` builds a given `Dockerfile`.
-While this approach works well for most run of the mill Python projects, it is not so seamless for R projects.
+While this approach works well for most run-of-the-mill Python projects, it is not so seamless for R projects.
 This is partly because `repo2docker` does not support arbitrary base images due to the complex auto-generation of the `Dockerfile` instructions.
-Two approaches make using Binder easier. for R users.
+Two approaches make using Binder easier for R users.
 First, **\pkg{holepunch}** ([https://github.com/karthik/holepunch](https://github.com/karthik/holepunch)) is an R package that was designed to make sharing work environments accessible to novice R users based on Binder.
@@
 \pkg{holepunch} removes some of these limitations by leveraging Rocker images that contain the Tidyverse along special Jupyter dependencies, and only installs additional packages from CRAN and Bioconductor that are not already part of these images.
-It short cicuits the configuration file parsing in `repo2docker` and starts with the Binder/Tidyverse base images, which eliminates a large part of the build time and in most cases results in a binder instance launching within a minute.
-\pkg{holepunch} as a side effect also creates a `DESCRIPTION` file which then turns any project into a research compendium \citep{marwick_packaging_2018}.
-The `Dockerfile` included with the project can also be used to launch a RStudio server locally, i.e., independent of Binder, which is especially useful when more or special computational resources can be provided there.
-The local image usage reduces the number of seperately managed environments and thereby reduces work and increases portability and reproducibility.
+It short cicuits the configuration file parsing in `repo2docker` and starts with the Binder/Tidyverse base images, which eliminates a large part of the build time and, in most cases, results in a Binder instance launching within a minute.
+\pkg{holepunch} also creates a `DESCRIPTION` file for essential metadata and dependency specification, and thereby turns turns any project into a research compendium (see&nbsp;\nameref{compendia}).
+The `Dockerfile` included with the project can also be used to launch an RStudio Server instance locally, i.e., independent of Binder, which is especially useful when more or special computational resources can be provided there.
+The local image usage reduces the number of seperately managed environments and, thereby, reduces work and increases portability and reproducibility.
 Second, the **Whole&nbsp;Tale** project (https://wholetale.org) combines the strengths of the Rocker Project's curated Docker images with `repo2docker`.
-Whole&nbsp;Tale is a National Science Foundation (NSF) funded project developing a scalable, open-source, multi-user platform for reproducible research \citep{brinckman2019, chard2019a}.
+Whole&nbsp;Tale is a National Science Foundation (NSF) funded project developing a scalable, open source, multi-user platform for reproducible research \citep{brinckman2019, chard2019a}.
 A central goal of the platform is to enable researchers to easily create and publish executable research objects\footnote{In Whole~Tale a \emph{tale} is a research object that contains metadata, data (by copy or reference), code, narrative, documentation, provenance, and information about the computational environment to support computational reproducibility.} associated with published research \citep{chard2019b}.
 Using Whole&nbsp;Tale, researchers can create and publish Rocker-based reproducible research objects to a growing number of repositories including DataONE member nodes, Zenodo and soon Dataverse.
-Additionally, Whole&nbsp;Tale supports automatic data citation and is working on capabilities for image preservation and provenance capture to improve the transparency of published computational research artifacts \citep{mecum2018, mcphillips2019}.
-For R users, Whole&nbsp;Tale extends the Jupyter Project's `repo2docker` component to simplify the customization of R-based environments for researchers with limited experience with either Docker or Git.
+Additionally, Whole&nbsp;Tale supports automatic data citation and is working on capabilities for image preservation and provenance capture to improve the transparency of published computational research artefacts \citep{mecum2018, mcphillips2019}.
+For R users, Whole&nbsp;Tale extends the Jupyter Project's `repo2docker` component to simplify the customisation of R-based environments for researchers with limited experience with either Docker or Git.
 Multiple options have been discussed to allow users to change the base image used in `repo2docker` from the default Ubuntu LTS (long-term support) required to support the Rocker Project images. 
-Whole&nbsp;Tale implemented a custom `RockerBuildPack`\footnote{See \href{https://github.com/whole-tale/repo2docker\_wholetale}{https://github.com/whole-tale/repo2docker\_wholetale}} to support customization of the `rocker/geospatial` image through `repo2docker` composability\footnote{Composability refers to the ability to combine multiple package managers -- such as R, `pip`, and `conda`}.
+Whole&nbsp;Tale implemented a custom `RockerBuildPack`\footnote{See \href{https://github.com/whole-tale/repo2docker\_wholetale}{https://github.com/whole-tale/repo2docker\_wholetale}} to support customisation of the `rocker/geospatial` image through `repo2docker` composability\footnote{Composability refers to the ability to combine multiple package managers -- such as R, `pip`, and `conda`}.
 This works in part because Rocker images are based on a Debian distribution, so the instructions created by `repo2docker` for Ubuntu work because of compatible shell and package manager.
@@
 In **high-performance computing**, one use for containers is to run workflows on shared local hardware where teams manage their own high-performance servers.
-This can follow one of several design patterns: users may deploy containers to hardware as a work environment for a specific project, conatiners may provide per-user persistent environments, or a single container can act as a common multi-user environment for a server.
-In all cases, though, the containerized approach provides several advantages: First, users may use the same image and thus work environment on desktop and laptop computers, as well.
-The former models provide modularity, while the latter approach is most similar to a simple shared server.
-Second, software updates can be achieved by updating and redeploying the container, rather than tracking local installs on each server.
-Third, the containerized environment can be quickly deployed to other hardware, cloud or local, if more resources are neccessary or in case of server destruction or failure.
-In any of these cases, users need a method to interact with the containers, be it and IDE, or command-like access and tools such as SSH, which is usually not part of standard container recipes and must be added.
+This can follow one of several design patterns:
+Users may deploy containers to hardware as a work environment for a specific project, containers may provide per-user persistent environments, or a single container can act as a common multi-user environment for a server.
+In all cases, though, the containerised approach provides several advantages:
+First, users may use the same image and thus work environment on desktop and laptop computers.
+The first to patterns provide modularity, while the last approach is most similar to a simple shared server.
+Second, software updates can be achieved by updating and redeploying the container rather than by tracking local installs on each server.
+Third, the containerised environment can be quickly deployed to other hardware, cloud or local, if more resources are neccessary or in case of server destruction or failure.
+In any of these cases, users need a method to interact with the containers, be it an IDE exposed over an HTTP port or command-line access via tools such as SSH.
+A suitable method must be added to the container recipes.
 The Rocker&nbsp;Project provides containers pre-installed with the RStudio&nbsp;IDE.
-In cases where users store nontrivial amounts of data for their projects, data needs to persist beyond the life of the container.
-This may be via in shared disks, attached network volumes, or in separate storage where it is uploaded between sessions.
-In the case of shared disks or network-attached volumes, care must be taken to persist user permissions, and of course backups are still neccessary.
+In cases where users store nontrivial amounts of data for their projects, the data needs to persist beyond the life of the container.
+This may be in shared disks, attached network volumes, or in separate storage where it is uploaded between sessions.
+In the case of shared disks or network-attached volumes, care must be taken to match user permissions, and of course backups are still neccessary.
 When working with multiple servers, an automation framework such as [Ansible](https://www.ansible.com) may be useful for managing users, permisions, and disks along with containers.
@@
 \label{rocker-gpu}
-Using **GPUs** (graphical processing units) as a specialised hardware from containerized common work environments is also possible and useful \citep{haydel_enhancing_2015}.
+Using **GPUs** (graphical processing units) as a specialised hardware from containerised common work environments is also possible and useful \citep{haydel_enhancing_2015}.
 GPUs are increasingly popular for compute-intensive machine learning (ML) tasks, e.g., deep artificial neural networks \citep{schmidhuber_deep_2015}.
-Though in this case, containers are not completely portable between hardware environments, but the software stack for ML with GPUs is so complex to set up that a ready-to-use container is helpful.
+Although in this case containers are not completely portable between hardware environments, but the software stack for ML with GPUs is so complex to set up that a ready-to-use container is helpful.
 Containers running GPU software require drivers and libraries specific to GPU models and versions, and containers require a specialized runtime to connect to the underlying GPU hardware.
@@
 
-Teaching is a further example where shared computing environments and sandboxing can greatly improve the process.
-First, **prepared environments for teaching** are especially helpful for (a) introductory courses where the first step of installation and configuration can be hurdles for students \citep{cetinkaya-rundel_infrastructure_2018}, and (b) courses that require access to a relatively complex setup of software tools, e.g., database systems.
-\citet{cetinkaya-rundel_infrastructure_2018} describe how a Docker-based deployment of RStudio (i) avoided problems with troubleshooting individual students computers and greatly increase engagement through very quickly showing tangible outcomes, e.g., a visualisation, and (ii) reduced demand on teaching and IT staff.
-Each student get's acces to a personal RStudio instance running in a container after authentication with the university login, which gives the benefits of sandboxing and possibility of limiting resources.
-\citet{cetinkaya-rundel_infrastructure_2018} found that for the courses at hand, actual usage of the UI is intermittent so a single cloud based VM with 4&nbsp;cores and 28GB&nbsp;RAM sufficed for over 100 containers.
+**Teaching** is a further example of where shared computing environments and sandboxing can greatly improve the process.
+First, _prepared environments for teaching_ are especially helpful for (a) introductory courses, where students often struggle with the first step of installation and configuration \citep{cetinkaya-rundel_infrastructure_2018}, and (b) courses that require access to a relatively complex setup of software tools, e.g., database systems.
+\citet{cetinkaya-rundel_infrastructure_2018} describe how a Docker-based deployment of RStudio (i) avoided problems with troubleshooting individual students' computers and greatly increased engagement through very quickly showing tangible outcomes, e.g., a visualisation, and (ii) reduced demand on teaching and IT staff.
+Each student received access to a personal RStudio instance running in a container after authentication with the university login, which gives the benefits of sandboxing and the possibility of limiting resources.
+\citet{cetinkaya-rundel_infrastructure_2018} found that for the courses at hand, actual usage of the UI is intermittent so a single cloud-based VM with four cores and 28&nbsp;GB RAM sufficed for over 100 containers.
 An example for mitigating _complex setups_ is teaching databases.
 R is very useful tool for interfacing with databases, because almost every open source and proprietary database system has an R package that allows users to connect and interact with the it.
-This flexibility is even broadened by \CRANpkg{DBI}, which allows to create a common API for interfacing these databases, or the \CRANpkg{dbplyr} package, which runs \CRANpkg{dplyr} code straight against the database as queries.
+This flexibility is even broadened by \CRANpkg{DBI}, which allows for creating a common API for interfacing these databases, or the \CRANpkg{dbplyr} package, which runs \CRANpkg{dplyr} code straight against the database as queries.
 But learning and teaching these tools comes with the cost of deploying or having access to an environment with the software and drivers installed.
-For people teaching R, it can become a barrier if they need to install local versions of database drivers, or to connect to remote instances which might or might not be made available by IT services.
-Giving access to a sandbox for the most common database environments is the idea behind [`r-db`](https://github.com/ColinFay/r-db), a Docker image that contains everything needed to connect to a database from R.
-Notably, with `r-db`, the users don't have to install complex drivers or to configure their machine in a specific way.
+For people teaching R, it can become a barrier if they need to install local versions of database drivers or connect to remote instances which might or might not be made available by IT services.
+Giving access to a sandbox for the most common environments for teaching databases is the idea behind [`r-db`](https://github.com/ColinFay/r-db), a Docker image that contains everything needed to connect to a database from R.
+Notably, with `r-db`, users do not have to install complex drivers or configure their machine in a specific way.
 The `rocker/tidyverse` base image ensures that users can also readily use packages for analysis, display, and reporting.
-Second, the idea of a common environment and partitioning allow using **containers in teaching for secure execution and automated testing** of submissions by students.
-[Dodona](https://dodona.ugent.be) is a web platform developed at Ghent University and is used to teach students basic programming skills, and it uses Docker containers to test submissions by students.
+
+Second, the idea of a common environment and partitioning allows for using _containers in teaching for secure execution and automated testing_ of submissions by students.
+[Dodona](https://dodona.ugent.be) is a web platform developed at Ghent University that is used to teach students basic programming skills, and it uses Docker containers to test submissions by students.
 This means that both the code testing the students' submissions and the submission itself are executed in a predictable environment, avoiding compatibility issues between the wide variety of configurations used by students.
-The containerization is also used to shield the Dodona servers from bad or even malicious code: memory, time and I/O&nbsp;limits are used to make sure students can't overload the system.
+The containerisation is also used to shield the Dodona servers from bad or even malicious code: memory, time and I/O&nbsp;limits are used to make sure students cannot overload the system.
 The web application managing the containers communicates with them by sending configuration information as a JSON document over standard input. 
@@
 In June 2019, R support was added to Dodona using an image derived from the `rocker/r-base` image that sets up the `runner` user and and `main.sh` file expected by Dodona\footnote{\href{https://github.com/dodona-edu/docker-images/blob/master/dodona-r.dockerfile}{https://github.com/dodona-edu/docker-images/blob/master/dodona-r.dockerfile}}.
-It also installs the packages required for the testing framework and the exercises so that this doesn't have to happen every time a student's submission is evaluated.
+It also installs the packages required for the testing framework and the exercises so that this does not have to happen every time a student's submission is evaluated.
 The actual testing of R exercises is done using a custom framework loosely based on \CRANpkg{testthat}.
-During the development of the testing framework it was found that the \CRANpkg{testthat} framework did not provide enough information to its reporter system to send back all the fields required by Dodona to render its feedback.
+During the development of the testing framework, it was found that the \CRANpkg{testthat} framework did not provide enough information to its reporter system to send back all the fields required by Dodona to render its feedback.
 Right now, multiple statistics courses are developing exercises to automate the feedback for their lab classes.
-
-**RCloud** ([https://rcloud.social](https://rcloud.social)) is a cloud-based platform for data analysis, visualisation and collaboration using R.
-It provides a `rocker/drd` base image for easy evaluation of the platform\footnote{\href{https://github.com/att/rcloud/tree/master/docker}{https://github.com/att/rcloud/tree/master/docker}}.
 
@@
 New releases of even just a single piece of software in this pipeline can break the entire workflow, making it difficult to find the error and difficult for others to reuse existing pipelines.
-These breakages can make the original the results irreproducible and not ¸
-The chance of a substantial disruption like this is high in a multi-year research project where key pieces of third-party software may have several major updates over the duration of the project.
+These breakages can make the original the results irreproducible and, and the chance of a substantial disruption like this is high in a multi-year research project where key pieces of third-party software may have several major updates over the duration of the project.
 The classical "paper" article is insufficient to adequately communicate the knowledge behind such research projects \citep[cf.][]{donoho_invitation_2010,marwick_how_2015}.
@@
 \citet{gentleman_statistical_2007} coined the term **Research Compendium** for a dynamic document together with supporting data and code.
-They use the R package system \cite{core_writing_1999} as for the functional prototype to structuring, validation, and distribution of research compendia.
-This concept has been taken up and extended\footnote{See full literature list at \href{https://research-compendium.science/}{https://research-compendium.science/}.}, not the least by applying containerisation and other methods for managing computing environments---see Section&nbsp;\nameref{envs}.
+They used the R package system \cite{core_writing_1999} for the functional prototype all the way to structuring, validation, and distribution of research compendia.
+This concept has been taken up and extended\footnote{See full literature list at \href{https://research-compendium.science/}{https://research-compendium.science/}.}, not in the least by applying containerisation and other methods for managing computing environments---see Section&nbsp;\nameref{envs}.
 Containers give the researcher an isolated environment to assemble these research pipelines with specific versions of software to minimize problems with breaking changes and make workflows easier to share \citep[cf.][]{boettiger_introduction_2015,marwick_packaging_2018}.
-Research workflows in containers are safe from contamination from other activities occuring on the researcher's computer, for example the installation of newest version of packages for teaching demonstrations.
+Research workflows in containers are safe from contamination from other activities that occur on the researcher's computer, for example the installation of newest version of packages for teaching demonstrations or specific versions for evaluation of other's works.
 Given the users in this scenario, i.e., often academics with limited formal software development training, templates and assistance with containers around research compendia is essential.
@@
 A more extreme case is the _dynverse_ project ([https://dynverse.org/](https://dynverse.org/)), which packages over 50 computational methods with different environments (R, Python, C++, etc.) in Docker images, which can be executed from R.
-_dynverse_ uses a CI platform (see&nbsp;\nameref{ci}) to build Rocker-derived images, test them, and publish them on Docker&nbsp;Hub if the tests succeed.
+_dynverse_ uses a CI platform (see&nbsp;\nameref{ci}) to build Rocker-derived images, test them, and, if the tests succeed, publish them on Docker&nbsp;Hub.
 
@@
 
-Further safeguarding practices are currently under development or not part of common practice yet, such as preservation of images \citep{emsley_framework_2018} and storing them alongside `Dockerfile`s \citep[cf.][]{nust_opening_2017}, or pinning of system libraries beyond the tagged base images, which may be seen as stable or dynamic depending on the applied time scale \citep[see discussion on `debian:testing` base image in][]{RJ-2017-065}.
+Further safeguarding practices are currently under development or not part of common practice yet, such as preserving images \citep{emsley_framework_2018}, storing both images and `Dockerfile`s \citep[cf.][]{nust_opening_2017}, or pinning system libraries beyond the tagged base images, which may be seen as stable or dynamic depending on the applied time scale \citep[see discussion on \texttt{debian:testing} base image in][]{RJ-2017-065}.
 A recommendation of the recent National Academies' report on _Reproducibility and Replicability in Science_ is that journals _"consider ways to ensure computational reproducibility for publications that make claims based on computations"_ \citep{NASEM2019}.
-In fields such as Political Science and Economics, journals are increasingly adopting policies requiring authors to publish the code and data required to reproduce computational findings reported in published manuscripts, subject to independent verification \citep{Jacoby2017,Vilhuber2019,Alvarez2018,Christian2018,Eubank2016,King1995}.
+In fields such as political science and economics, journals are increasingly adopting policies that require authors to publish the code and data required to reproduce computational findings reported in published manuscripts, subject to independent verification \citep{Jacoby2017,Vilhuber2019,Alvarez2018,Christian2018,Eubank2016,King1995}.
 Problems with the computational environment, installation and availability of software dependencies are common.
-R is gaining popularity in these communities, such as creating a research compendium.
-In a sample of 105 replication packages published by the _American Journal of Political Science_ (AJPS) over 65% use R.
-The NSF-funded Whole Tale project uses the Rocker Project community images with the goal of improving the reproducibility of published research artifacts and simplifying the publication and verification process for both authors and reviewers by reducing errors and time spent specifying the environment.
+R is gaining popularity in these communities, such as for creating a research compendium.
+In a sample of 105 replication packages published by the _American Journal of Political Science_ (AJPS), over 65% use R.
+The NSF-funded Whole Tale project, which was mentioned above, uses the Rocker Project community images with the goal of improving the reproducibility of published research artefacts and simplifying the publication and verification process for both authors and reviewers by reducing errors and time spent specifying the environment.
 
@@
 
-This article is a snapshot of the R-corner in a universe of applications built with a many-faced piece of software, Docker.
+This article is a snapshot of the R corner in a universe of applications built with a many-faced piece of software, Docker.
 `Dockerfile`s and Docker images are the go-to methods for collaboration between roles in an organisation, such as development and IT operations teams, and between parties in the communication of knowledge, such as research workflows or education.
-Docker became synonymous with applying the concept of containerisation to solve challenges of reproducible environments, e.g., in research and in development \& production, and of scalable deployments with the ability to move processing between machines easily (e.g., locally, one cloud providers VM, another cloud provider's Container-as-a-Service).
+Docker has become synonymous with applying the concept of containerisation to solve challenges of reproducible environments, e.g., in research and in development \& production, and of scalable deployments as it can easily move processing between machines (e.g., locally, a cloud provider's VM, another cloud provider's Container-as-a-Service).
 Reproducible environments, scalability and efficiency, and portability across platforms/infrastructures are the common themes behind R packages, use cases, and applications in this work.
 
-The R packages and use cases presented show the growing number of users, developers, and real-world applications in the community and the resulting innovations.
+The R packages and use cases presented above show the growing number of users, developers, and real-world applications in the community and the resulting innovations.
 But the applications also point to the challenge of keeping up with a continuously evolving landscape.
@@
 <!-- Tyler W./DC from Gigantum: ideas about ease-of-DIY together with difficulty of composition; shared starting points of Rocker -->
-This ease-of-DIY in combination with the difficulty to compose environments based on `Dockerfile`s alone is a further reason for **fragmentation**.
-Instructions can be outsourced into distributable scripts and then copied into the image during build, but that make `Dockerfile` hard to read and adds a layor of complexity.
+This ease-of-DIY in combination with the difficulty of reusing parts from or composing multiple `Dockerfile`s is a further reason for **fragmentation**.
+Instructions can be outsourced into distributable scripts and then copied into the image during build, but that makes `Dockerfile`s harder to read.
+Scripts added to a `Dockerfile` also add a layer of complexity and increase the risk of incomplete recipes.
 Despite the different image stacks presented here, the pervasiveness of Rocker can be traced back to its maintainers and the user community valueing collaboration and shared starting points over impulses to create individual solutions.
-Aside from that, fragmentation may not be a bad sign, but instead a reflection of a growing market, which is able to sustain multiple related efforts.
-With the maturing of core building blocks, such as the Rocker suite of images, more systems will be built successfully but will also be behind the curtains.
+Aside from that, fragmentation may not be a bad sign but may instead be a reflection of a growing market that is able to sustain multiple related efforts.
+With the maturing of core building blocks, such as the Rocker suite of images, more systems will be built successfully, but they may also be built behind the curtains.
 Docker alone, as a flexible core technology, is not a feasible level of collaboration and abstraction.
@@
 Nonetheless, at least on the level of R packages some **consolidation** seems in order, e.g., to reduce the number of packages creating `Dockerfile`s from R code or controlling the Docker daemon with R code.
-It remains to be seen which approach to control Docker, via the Docker API as \pkg{stevedore} or via system calls as \pkg{dockyard}/\pkg{docker}/\pkg{dockr}, is more sustainable, or if the question will be answered by the endurance of maintainers and sufficient funding.
-Similarly, the capturing of environments and their serialization in form of a `Dockerfile` currently happens at different levels of abstraction and re-use of functionality seems reasonable, e.g., \pkg{liftr} could generate the environment with \pkg{containerit}, which in turn may use \pkg{dockerfiler} for low level R objects representing a `Dockerfile` and its instructions.
-In this consolidation, the Rocker&nbsp;Project could play the role of coordinating entity.
-Though for the moment, the sign of the times points to more experimentation and feature growth, e.g., images for GPU-based computing and artificial intelligence.
-Even with coding being more and more accepted as a required, and achievable skill, an easier access, for example by exposing containerisation benefits via simple user interfaces in the users' IDE, could be an important next step since currently containerisation happens more in the background for UI-based development (e.g., a `rocker/rstudio` image in the cloud).
-Furthermore, a maturing of the Rockerverse packages for managing containers may lead to their adoption where currently manual coding is required, e.g.
+It remains to be seen which approach to control Docker, via the Docker API as \pkg{stevedore} or via system calls as \pkg{dockyard}/\pkg{docker}/\pkg{dockr}, is more sustainable, or whether the question will be answered by the endurance of maintainers and sufficient funding.
+Similarly, the capturing of environments and their serialisation in form of a `Dockerfile` currently happens at different levels of abstraction, and re-use of functionality seems reasonable, e.g., \pkg{liftr} could generate the environment with \pkg{containerit}, which in turn may use \pkg{dockerfiler} for low-level R objects representing a `Dockerfile` and its instructions.
+In this consolidation, the Rocker&nbsp;Project could play the role of a coordinating entity.
+Nonetheless, for the moment, it seems that Rocker Project will be used more experimentation and feature growth, e.g., images for GPU-based computing and artificial intelligence.
+Even with coding being more and more accepted as a required and achievable skill, an easier access, for example by exposing containerisation benefits via simple user interfaces in the users' IDE, could be an important next step, since currently containerisation happens more in the background for UI-based development (e.g., a `rocker/rstudio` image in the cloud).
+Furthermore, a maturing of the Rockerverse packages for managing containers may lead to them being adopted in situations where manual coding is currently required, e.g.
 in the case of \CRANpkg{RSelentium} or \CRANpkg{drake} (see Sections&nbsp;\nameref{rselenium} and \nameref{drake} respectively).
-In specific cases, e.g., for \CRANpkg{analogsea}, the interaction with the Docker daemon may remain to specific to reuse first order packages to control Docker.
+In specific cases, e.g., for \CRANpkg{analogsea}, the interaction with the Docker daemon may remain too specific to re-use first-order packages to control Docker.
 
-New features, which make complex workflows accessible and reproducible, and the variety in packages connected with containerisation, even when they have overlapping features, are a signal and a support for a growing user base.
-This growth is possibly the most important goal for the foreseeable future in the _Rockerverse_, and just like the Rocker images have matured over years of use and millions of runs, the new ideas and prototypes will have to proof themselves.
-It should be noted that the dominant position of Docker is a blessing and a curse for these goals.
-It could be wise to start experimenting with non-Docker containerisation tools now, e.g., R packages interfacing with other container engines such as [podman/buildah](https://github.com/containers/libpod) or [rkt](https://coreos.com/rkt/), or an R package for creating `Singularity` files.
-Such efforts can help to avoid lock-in and to design sustainable workflows based on concepts of _containerisation_, not on their implementation in Docker.
-If adoption of containerisation and R continue to grow, the missing pieces for a success predominantly lie in (a) coordination and documentation of activities to reduce repeated work in favour of open collaboration, (b) the sharing of lessons learned from use cases to build common knowledge and language, and (c) a sustainable continuation and funding for all of development, community support, and education.
-A concrete effort to work towards these pieces is to sustain the structure and captured status quo from this work in form of a _CRAN Task View on containerization_.
+New features which make complex workflows accessible and reproducible and the variety in packages connected with containerisation, even when they have overlapping features, are a signal and a support for a growing user base.
+This growth is possibly the most important goal for the foreseeable future in the _Rockerverse_, and, just like the Rocker images have matured over years of use and millions of runs, the new ideas and prototypes will have to prove themselves.
+It should be noted that the dominant position is that Docker is a blessing and a curse for these goals.
+It might be wise to start experimenting with non-Docker containerisation tools now, e.g., R packages interfacing with other container engines such as [podman/buildah](https://github.com/containers/libpod) or [rkt](https://coreos.com/rkt/), or an R package for creating `Singularity` files.
+Such efforts might help to avoid lock-in and to design sustainable workflows based on concepts of _containerisation_, not on their implementation in Docker.
+If adoption of containerisation and R continue to grow, the missing pieces for a success predominantly lie in (a) coordination and documentation of activities to reduce repeated work in favour of open collaboration, (b) the sharing of lessons learned from use cases to build common knowledge and language, and (c) a sustainable continuation and funding for development, community support, and education.
+A first concrete effort to work towards these missing pieces should be sustain the structure and captured status quo from this work in form of a _CRAN Task View on containerisation_.
 
@@
 GD contributed the CARD.com use case.
-DE wrote the introduction and the section about Containerization and Rocker and reviewed all sections.
+DE wrote the introduction and the section about containerisation and Rocker Project, and reviewed all sections.
 RC \& EH contributed to the section on interfaces for Docker in R (_dynverse_ and `dynwrap`).
@@
 SL contributed content on RStudio's usage of Docker.
-BM wrote the section on research compendia and made the project Binder-ready.
+BM wrote the section on research compendia and made the project Binder ready.
 HN \& JN co-wrote the section on the T-Mobile use case.
@@
 CW is supported by the Whole Tale projects (https://wholetale.org) funded by the US National Science Foundation (NSF) under award [OAC-1541450](https://www.nsf.gov/awardsearch/showAward?AWD_ID=1541450).
+We would like to thank Celeste R. Brennecka from the Science Writing Support Service of the University of Münster for her editorial support.
 
@@
 
+```{r collab_emails, eval=FALSE, include=FALSE}
+front_matter <- rmarkdown::yaml_front_matter("manuscript.Rmd")
+emails <- sapply(X = front_matter$author, FUN = function(a) { a$email })
+cat(emails, sep = "\n")
+```
+
 \bibliography{RJreferences}

### 2020-02-05 manuscript.Rmd
add conclusion summary to abstract, give examples for use cases

@@
   This articles surveys downstream projects that build upon Rocker images and presents the current state of R packages for managing Docker images and controlling containers.
-  These use cases and the variety of applications demonstrate the power of Rocker specifically and containerisation in general.
+  These use cases cover diverse topics such as package development, reproducible research, collaborative work, cloud-based data processing, and production deployment of services.
+  The variety of applications demonstrates the power of Rocker specifically and containerisation in general.
   Across the diverse ways to use containers, we identified common themes: reproducible environments, scalability and efficiency, and portability across clouds.
+  We conclude that the current growth and diversification of use cases is likely to continue its positive impact, but see the need for consolidating the Rockerverse ecosystem of packages, developing common practices for applications, and exploring alternative containerisation software.
 output:

### 2020-02-18 manuscript.Rmd
Update dockr paragraph, authors in alphabetical order

--- 2020-01-29 manuscript.Rmd
@@
 ---
-title: "The Rockerverse: Packages and Applications for Containerization with R"
+title: "The Rockerverse: Images, Packages, and Applications for Containerization with R"
 author:
@@
     email: daniel.nuest@uni-muenster.de
-  - name: Dirk Eddelbuettel
-    # https://github.com/eddelbuettel
-    affiliation: University of Illinois at Urbana-Champaign
-    address:
-      - Department of Statistics
-      - Illini Hall, 725 S Wright St
-      - Champaign, IL 61820, USA
-      - \orcid{0000-0001-6419-907X}
-    email: dirkd@eddelbuettel.com
   - name: Dom Bennett
@@
     email: daroczig@rapporter.net
+  - name: Dirk Eddelbuettel
+    # https://github.com/eddelbuettel
+    affiliation: University of Illinois at Urbana-Champaign
+    address:
+      - Department of Statistics
+      - Illini Hall, 725 S Wright St
+      - Champaign, IL 61820, USA
+      - \orcid{0000-0001-6419-907X}
+    email: dirkd@eddelbuettel.com
   - name: Mark Edmondson
@@
     email: karthik.ram@berkeley.edu
+  - name: Lars Kjeldgaard
+    # https://github.com/smaakage85
+    affiliation: Danish Tax Authorities
+    address:
+      - Oestbanegade 123
+      - 2100, Koebenhavn Oe
+    email: lars_kjeldgaard@hotmail.com
   - name: Noam Ross
@@
 abstract: >
-  The Rocker Project provides widely-used Docker images for R across different application scenarios.
+  The Rocker&nbsp;Project provides widely-used Docker images for R across different application scenarios.
   This articles surveys downstream projects building upon Rocker and presents the current state of R packages for managing Docker images and controlling containers.
+  We also look beyond Rocker to other projects connecting containerisation with R, namely alternative suites of images.
   These use cases and the variety of applications demonstrate the power of Rocker specifically and containerisation in general.
@@
 ```{r setup, include=FALSE}
-Sys.setlocale('LC_ALL','C')
+library(tidyverse)
 ```
@@
 Combining both these topics, the _Rocker&nbsp;Project_ ([https://www.rocker-project.org/](https://www.rocker-project.org/)) provides images with R (see the next Section for more details).
-The considerable uptake and continued evolution of the Rocker&nbsp;Project has lead to numerous projects extending or building upon Rocker images, ranging from reproducible research\footnote{"Reproducible" in the sense of the \emph{Claerbout/Donoho/Peng} terminology \citep{barba_terminologies_2018}.} to production deployments.
+The considerable uptake and continued evolution of the Rocker&nbsp;Project has lead to numerous projects extending or building upon Rocker images, ranging from reproducible research\footnote{"Reproducible" in the sense of the _Claerbout/Donoho/Peng_ terminology \citep{barba_terminologies_2018}.} to production deployments.
 This article presents this _Rockerverse_ of projects across all development stages: early demonstrations, working prototypes, and mature products.
@@
 
-The article continues with a brief introduction of containerization basics and the Rocker&nbsp;Project, followed by use cases and applications, starting with the R packages specificly for interacting with Docker, second-level packages using containers indirectly or only for specific features, up to complex use cases leveraging containers.
+The article continues with a brief introduction of containerization basics and the Rocker&nbsp;Project, followed by descriptions of projects developing images with R installations.
+Then, we present applications, starting with the R packages specificly for interacting with Docker, second-level packages using containers indirectly or only for specific features, up to complex use cases leveraging containers.
 We conclude with a reflection on the landscape of packages and applications and point out future directions of development.
@@
 
-For statistical computing and analysis centered around R, the **Rocker&nbsp;Project** has provided a variety of Docker containers since its start in 2014 \citep{RJ-2017-065}.
+For statistical computing and analysis centered around R, the Rocker&nbsp;Project has provided a variety of Docker containers since its start in 2014 \citep{RJ-2017-065}.
 The Rocker&nbsp;Project provides several lines of containers spanning to from building blocks with `R-release` or `R-devel`, via containers with [RStudio&nbsp;Server](https://rstudio.com/products/rstudio/) and [Shiny&nbsp;Server](https://rstudio.com/products/shiny/shiny-server/), to domain-specific containers such as [`rocker/geospatial`](https://github.com/rocker-org/geospatial) \citep{rocker_geospatial_2019}.
@@
 
-# Interfaces for Docker in R
+# Container images
+\label{images}
+
+## Images for alternative R engines
+\label{alternatives}
+
+As outlined above, R is a widely-used language with a large community.
+The large number of extension packages provides access to an unrivaled variety of established and upcoming features.
+Nevertheless, special use cases and experimental projects in academia and industry exist to test new approaches, or provide features different to what the R implementation maintained by the [R Core Team](https://www.r-project.org/contributors.html) and available via [CRAN](https://cran.r-project.org/) provides.
+While the images presented in this section are far from being as vetted, stable, and widely used as any of the Rocker images, they demonstrate an important advantage of containerisation technology, namely the abiliy to transparently build portable stacks of open source software and make them easily accessible to others.
+All these alternative implementations of R are published under GPL licenses.
+Since all implementations of R claim better performance as a core motivation, a comparision based on Docker images, potentially leveraging the [resource restriction mechanisms](https://docs.docker.com/config/containers/resource_constraints/) of Docker to level the playing field, seems useful future work.
+
+**Microsoft R Open** (MRO) is an R distribution formerly known as Revolution R Open (RRO) before Revolution Analytics was acquired by Microsoft.
+MRO is compatible with main R and its packages and it "[..] includes additional capabilities for improved performance, reproducibility, and platform support." \citep{microsoft_mro}
+Most notably these capabilities are the MRAN repository, which is enabled by default, and some bundled packages such as \CRANpkg{checkpoint} and specific R packages maintained by Microsoft\footnote{See "Bundled Packages" on \href{https://mran.microsoft.com/rro/installed}{https://mran.microsoft.com/rro/installed}.}.
+Originally the optional integration with [Intel® Math Kernel Library](https://software.intel.com/en-us/mkl) (MKL) for multi-threading in linear algebra operations \citep{microsoft_multithread} was a distinguishing feature before it was made available publicly, and can be used in base R, too \citep{eddelbuettel_thinking_2018}.
+
+MRO does not provide official Docker images, but a set of community-maintained `Dockerfile`s and Docker images are provided by the poject `mro-docker` ([https://github.com/nuest/mro-docker](https://github.com/nuest/mro-docker)), e.g., the image [`nuest/mro`](https://hub.docker.com/repository/docker/nuest/mro/).
+The images are inspired by the Rocker images and can be used much in the same fashion, effectively a drop-in replacement allowing users to quickly evaluate if the benefits of MRO + Intel® MKL apply to their use case.
+Version-tagged images are provided for the latest bugfix release of recent R versions.
+Extended license information about MKL is printed at every startup.
+
+**Renjin** is an interpreter for the R language running on top of the [Java Virtual Machine](https://en.wikipedia.org/wiki/Java_virtual_machine) (JVM) providing full two-way access between Java and R code \citep{wikipedia_renjin_2018}.
+It was developed to combine the benefits of R, such scripting and extension packages, with the JVM's advantages in the areas of security, cross-platform availability and integration into enterprise platforms.
+R extension packages need to be specially compiled and are distributed via the Java package manager [Apache Maven](https://en.wikipedia.org/wiki/Apache_Maven), cf.
+[http://packages.renjin.org/packages](http://packages.renjin.org/packages) for available packages.
+Packages are loaded on demand, i.e., at the first call to `library()`.
+Not all R packages, especially one linking to binary libraries, are available, e.g., `rgdal` \footnote{\href{http://packages.renjin.org/package/org.renjin.cran/rgdal/1.4-4/build/1}{http://packages.renjin.org/package/org.renjin.cran/rgdal/1.4-4/build/1}}.
+There are no offical Docker images for Renjin, but community-maintained images for selected releases only are available under `nuest/renjin` on Docker&nbsp;Hub and GitHub at [https://hub.docker.com/r/nuest/renjin](https://hub.docker.com/r/nuest/renjin) and [https://github.com/nuest/renjin-docker](https://github.com/nuest/renjin-docker) respectively.
+These images expose the command line interface of Renjin in a similar fashion as Rocker images and allow an easy evaluation of Renjin's suitability, but are not intended for production use.
+
+**pqR** ([http://www.pqr-project.org/](http://www.pqr-project.org/)) is a "a pretty quick version of R".
+_pqR_ attempts to improve R on some opinionated issues in the R language and is the basis for experimental features, e.g., automatic diffentiation\footnote{\href{https://riotworkshop.github.io/abstracts/riot-2019-pqr.txt}{https://riotworkshop.github.io/abstracts/riot-2019-pqr.txt}}.
+The source code development [on GitHub](https://github.com/radfordneal/pqR/) is a one man project and it does not provide any Docker images.
+But especially disruptive approaches may contribute to the development of the R ecosystem, so the `nuest/pqr` image was independently created, see Docker Hub at [https://hub.docker.com/r/nuest/pqr/](https://hub.docker.com/r/nuest/pqr/) and GitHub at [https://github.com/nuest/pqr-docker](https://github.com/nuest/pqr-docker).
+
+**FastR** ([https://github.com/oracle/fastr](https://github.com/oracle/fastr)) is "A high-performance implementation of the R programming language, built on GraalVM" \citep{oracle_labs_oraclefastr_2020}.
+It is developed by Oracle, connects R to the GraalVM ecosystem \citep{wikipedia_graalvm_2019}, and also claims superior performance but also targets full compatibility with base R \citep{oracle_labs_oraclefastr_2020}.
+There are no offical Docker images provided, but indepedently maintained experimental images and `Dockerfile`s are provided by [nuest/fastr-docker](https://github.com/nuest/fastr-docker), e.g., [`nuest/fastr` on Docker Hub](https://cloud.docker.com/repository/docker/nuest/fastr/).
+
+## Bioconductor
+\label{bioc}
+
+_Bioconductor_ ([https://bioconductor.org/](https://bioconductor.org/)) is an open source, open development project for the analysis and comprehension of genomic data \citep{gentleman_bioconductor_2004}.
+The project consists of 1823 R software packages as of October 30th 2019, as well as packages containing annotation or experiment data.
+_Bioconductor_ has a semi-annual release cycle, each release is associated with a particular version of R, and Docker images are provided for current and past versions of _Bioconductor_ for convenience and reproducibility.
+All images, included  are described on the _Bioconductor_ web site (see [https://bioconductor.org/help/docker/](https://bioconductor.org/help/docker/)), created with `Dockerfile`s maintained on GitHub, and distributed through Docker&nbsp;Hub\footnote{See \href{https://github.com/Bioconductor/bioconductor_docker}{https://github.com/Bioconductor/bioconductor\_docker} and \href{https://hub.docker.com/u/bioconductor}{https://hub.docker.com/u/bioconductor} respectively.}.
+_Bioconductor_'s 'base' Docker images are built on top of the `rocker/rstudio` image.
+_Bioconductor_ installs packages based on the R version in combination with the Bioconductor version, and therefore uses Bioconductor version tagging `devel` and `RELEASE_X_Y`, e.g., `RELEASE_3_10`.
+Past and current combinations of R and _Bioconductor_ will therefore be accessible via a specific image tags.
+The _Bioconductor_ `Dockerfile` selects the desired version of R from Rocker, adds required system dependencies, and uses the \CRANpkg{BiocManager} package for installing appropriate versions of _Bioconductor_ packages.
+A strength of this approach is that the responsibility for complex software configuration (including customized development) is shifted from the user to the experienced _Bioconductor_ core team.
+A recent audit of the _Bioconductor_  image stack `Dockerfile` led to the deprecation of several community-maintained images, because the numerous specific images became too hard to understand, complex to maintain, and cumbersome to extent.
+As part of the simplification, a recent innovation is to produce a `bioconductor_docker:devel` image to emulate the _Bioconductor_ nightly Linux build machine as closely as possible.
+This image contains the build system environment variables and the _system dependencies_ needed to install and check almost all (1813 out of 1823) _Bioconductor_ software packages and saves users and package developers from managing these themselves.
+Furthermore the image is configured so that `.libPaths()` has `/usr/local/lib/R/host-site-library` as the first location.
+Users mounting a location on the host file system to this location can persist installed packages across Docker sessions or image updates.
+Many R users pursue flexible work flows tailored to particular analysis needs, rather than standardized work flows.
+The new `bioconductor_docker` image is well-suited for this pattern, while `bioconductor_docker:devel` provides developers with a test environment close to _Bioconductor_'s build system.
+
+## Images for (historic) R versions
+\label{versions}
+
+As with any other software, each new version of R comes with its share of changes.
+Some are breaking changes, some are not, but the fact is that running a piece of code in a given version might give different results from another version can be an issue when it comes to reproducibility of workflows and stability of applications.
+For example, think about R random seed: starting with `R 3.6.0`, running `set.seed(2811); sample(1:1000, 2)` will not give the same result as if it was run inside an older version of R.
+That might seem trivial, but all the code using the random number generator will not be exactly reproducible after this breaking change.
+Therefore, controlling the version of software is most crucial for reproducible research \citep[e.g.][]{boettiger_introduction_2015}.
+
+Containers are perfectly suited to capture a specific configuration of a computing environment to prohibit such problems.
+The Rocker images provides the _versioned stack_ for different the version of R since the project inception, which use the _then-current_ stable Debian base image \citep[cf.][]{RJ-2017-065}.
+Based on custom build phase hooks\footnote{\href{https://docs.docker.com/docker-hub/builds/advanced/}{https://docs.docker.com/docker-hub/builds/advanced/}}, i.e., small shell scripts executed at different phases in the automated build of Rocker images, there are also semantic version tags for the most recent freezed (i.e., using MRAN) versions with only the major and minor version\footnote{See \href{https://github.com/rocker-org/rocker-versioned/blob/master/VERSIONS.md}{https://github.com/rocker-org/rocker-versioned/blob/master/VERSIONS.md}, FIXME: \href{https://github.com/rocker-org/rocker-versioned/issues/42}{https://github.com/rocker-org/rocker-versioned/issues/42}.}.
+So, at the time of writing this article, `rocker/r-ver:3` and `3.6` are aliases for `3.6.0`, because `3.6.1` is the latest release.
+With the release of `3.6.2` and pinning of the MRAN version in `3.6.1`, the tags `3` and `3.6` will deliver the same image as `3.6.1`.
+These tags allow users to update the base image to retrieve bugfixes while reducing the risk of also introducing breaking changes.
+
+[r-online](https://github.com/ColinFay/ronline) is an app for helping users to detect breaking changes between different R versions, and for historic exploration of R.
+With a standalone NodeJS app or [online](https://srv.colinfay.me/r-online), the user can compare a piece of code run in two separate versions of R.
+Internally, r-online opens one or two Docker instances with the given version of R based on Rocker images, executes a given piece of code, and returns the result to the user.
+
+A container also provides an isolated sandbox environment suitable for testing and evaluation, without interfering with the "main" working environment.
+This enables cross version testing and bugfixing.
+To take this even further, the Rocker contributors are discussing the provision of R versions reaching back futher than the project's own inception, reaching back as far as R `2.x` and `1.x` \footnote{\href{https://github.com/rocker-org/rocker-versioned/issues/138}{https://github.com/rocker-org/rocker-versioned/issues/138}}.
+The main challenges are finding a suitable base image with a matching OS version, or making the adjustments to compile R on more recent OS releases.
+
+## Windows Images
+\label{windows}
+
+Docker containers on the Windows operating system were originally quite cumbersome to use.
+They required an extra tool, [Docker Toolbox](https://docs.docker.com/toolbox/), which by now only exists as a legacy solution for older Windows systems.
+Docker toolbox leverages `docker-machine` (see also Section&nbsp;\nameref{interfaces} for an R package interfacing with `docker-machine`) to handle the process of creating a local virtual host which could host Docker Engine, while exposing the regular Docker CLI.
+The Docker CLI commands are forwarded to the virtual host transparently for the user.
+
+For current Windows Server (2016 and later) and Windows Desktop ([Docker Desktop](https://www.docker.com/products/docker-desktop) requires Windows 10) versions, Docker is supported natively\footnote{\href{https://www.docker.com/products/windows-containers}{https://www.docker.com/products/windows-containers}} and different base images are offered by Microsoft\footnote{\href{https://docs.microsoft.com/en-us/virtualization/windowscontainers/about/index}{https://docs.microsoft.com/en-us/virtualization/windowscontainers/about/index}}.
+The Docker CLI can be used in just the same way as on other operating systems but not every base image is supported on every Windows host\footnote{\href{https://docs.microsoft.com/en-us/virtualization/windowscontainers/deploy-containers/version-compatibility}{https://docs.microsoft.com/en-us/virtualization/windowscontainers/deploy-containers/version-compatibility}}.
+On Docker Desktop for Windows, the user can run both Linux-based and Windows-based containers, but only one of the Docker daemons can be used at a time\footnote{See "Switch between Windows and Linux containers" on \href{https://docs.docker.com/docker-for-windows/}{https://docs.docker.com/docker-for-windows/}.}.
+
+**rocker-win** ([https://github.com/nuest/rocker-win](https://github.com/nuest/rocker-win)) is a proof of concept for running R in Windows-based containers \citep{nust_rocker-win_2019}.
+It provides selected R versions using three different base images using `microsoft/windowsservercore` and `mcr.microsoft.com/windows/servercore`.
+These base images match different Windows versions: Windows Server 2019, Windows Server 2016, and Windows Server, version 1803.
+The images are built using two CI services, [Travis CI](https://docs.travis-ci.com/user/reference/windows/) and [Appveyor](https://www.appveyor.com/docs/windows-images-software/), which provide the different Windows versions and therefore support different base images, and published automatically [on Docker&nbsp;Hub](https://hub.docker.com/r/nuest/rocker-win).
+The images are built from manually maintained seperate `Dockerfile`s, are tagged with both Windows-variant and R version, e.g., `nuest/rocker-win:ltsc2019-3.6.2` or `nuest/rocker-win:1803-latest`, and run the `R.exe` process by default.
+All these images can be run on Docker for Desktop on Windows 10.
+New R versions are only added on demand.
+The explorative rocker-win project demonstrates the variety of Windows on Docker, which is potentially confusing, e.g., because of the different base images.
+The proof of concept provides a starting point for any Windows-constrained user to leverage containerisation, including examples for apps using \CRANpkg{plumber}, \CRANpkg{Shiny}, or R packages with system dependencies such as \CRANpkg{sf}.
+The latter is quite comfortable and fast actually, compared to the default installation from source on Linux, because CRAN ships pre-compiled binaries for Windows.
+However, compared to the various Linux images, the footprint of Windows images is quite large.
+
+For Windows, just as originally for Linux, the cloud use cases drive the development and Docker and Microsoft collaborate closely\footnote{\href{https://www.docker.com/partners/microsoft}{https://www.docker.com/partners/microsoft}}.
+In most cases, individual developers or researchers worrying about reproducibility, will prefer the more widely and host independent Linux-based containers.
+Naturally, a Windows license is required for the host machine, and the licensing impacts the potential to build images in clouds or redistribute exported images as archive files.
+But the developments in the `rocker-win` prototype show that containers can also be used for R workflows depending on Windows-only tools, for leveraging an existing Windows Server-based infrastructure where policies can otherwise not be met, or for streamlining interactions with system operating staff that runs Windows-based containers themselves.
+
+## Non-Rocker Linux images
+\label{nondebian}
+
+\label{rhub}
+The **R-hub** project provides _"a collection of services to help R package development"_, with the package builder as the most prominent one\citep{r-hub_docs_2019}.
+The builder allows R package developers to check their R package on different platforms and R versions using a web form or the package \CRANpkg{rhub} \citep{csardi_rhub_2019}.
+The builder uses Docker containers to conduct these checks, taking advantages of their well-defined environments and sandboxing.
+The `Dockerfile`s and images are published on Docker&nbsp;Hub\footnote{\href{https://hub.docker.com/u/rhub}{https://hub.docker.com/u/rhub}} and GitHub\footnote{\href{https://github.com/r-hub/rhub-linux-builders}{https://github.com/r-hub/rhub-linux-builders}} respectively.
+The images comprise release and development builds of R running in the base images of the [Debian](https://www.debian.org/), [Ubuntu](https://ubuntu.com/), [Fedora](https://getfedora.org/), and [CentOS](https://centos.org/) linux distributions (Arch Linux is under development\footnote{\href{https://github.com/r-hub/rhub-linux-builders/pull/41}{https://github.com/r-hub/rhub-linux-builders/pull/41}}, see `rhub::local_check_linux_images()` for a full list), and are intended for debugging (see Section&nbsp;\nameref{development}) not for reproducible workflows.
+_The rationale_ of not re-using the Rocker image stack is the high number of operatings systems and configurations, which can be more uniformly maintained in a seperate suite of images.
+Also, a common structure across the respective image stacks of the operating systems, e.g., for `debian`, adding GCC in `debian-gcc` and then R's development version in `debian-gcc-devel`.
+The platform also has a specific user configuration differing from the setup covering typical applications of Rocker images, and the images include considerably more software, such as a fairly complete LaTeX installation, to provide an experience closer to CRAN.
+
+[**RStudio**](https://rstudio.com/) also maintains a set of [Docker images](https://github.com/rstudio/r-docker) for multiple operating systems including SUSE, Ubuntu, CentOS, and Debian.
+The image stack includes an [opinionated R installation](https://github.com/rstudio/r-builds) and ensures the R installation and profile is consistent across the different Linux distributions.
+The base image also installs R in a versioned directory and uses a minimal set of build and runtime dependencies.
+Furthermore these images are responsible for creating pre-compiled binary R packages for Linux, which supplement the binaries built by CRAN.
+These pre-compiled packages dramatically decrease the installation time of R packages on Linux and subsequently can decrease the build time of Docker images \citep{lopp_package_2019}.
+
+There are also independent projects installing R or RStudio on [**Alpine Linux**](https://www.alpinelinux.org/), but not beyond the proof of concept stage\footnote{\href{https://gitlab.com/artemklevtsov/r-alpine}{https://gitlab.com/artemklevtsov/r-alpine} or \href{https://www.github.com/velaco/alpine-r}{https://www.github.com/velaco/alpine-r} provide different use case images but are not recently maintained (development of the latter documented in \citet{ratesic_building_2018}); \href{https://github.com/cmplopes/alpine-r}{https://github.com/cmplopes/alpine-r} has only base R images, is relatively up-to-date but sparsely documented; \href{https://github.com/CenterForStatistics-UGent/mountainr}{https://github.com/CenterForStatistics-UGent/mountainr} is a ghost project with only a single short `Dockerfile`, yet it shows the easy installation of the latest R from Alpine sources}.
+In general, Alpine-based Docker images are often chosen for minimalistic and thereby small and secure containers, e.g., in hardware with limited storage.
+R-hub's proof of concept [`r-minimal`](https://github.com/r-hub/r-minimal) takes this to the extreme with compressed image size of 20MB and bespoke install scripts, e.g., to cleverly remove compilers after package installation.
+For R though, this advantage quickly deteriorates if more features are needed, e.g., if adding compilers, packages, or especially if installing powerful data science and communication tools, like RStudio or LaTeX.
+Furthermore the distribution uses the [`musl` C library](https://www.musl-libc.org/) instead of the `glibc` used in Debian/Ubuntu, for which more tooling and experiences exist\footnote{Cf.
+\href{https://github.com/rocker-org/rocker/issues/231}{https://github.com/rocker-org/rocker/issues/231}}.
+In fact, using Debian's 'slim' variant and removing capabilities, one could probalby achieve similarly small image sizes on a more common stack.
+
+**\pkg{altRnative}** ([https://github.com/ismailsunni/altRnative/](https://github.com/ismailsunni/altRnative/)) is an experimental R package for running the same code across multiple containerised versions of R, intended for comparison across operating systems and different implementations of R.
+It comes with a collection of `Dockerfile`s and corresponding images in multiple combinations, currently including, e.g., MRO, FastR, Fedora, and TERR\footnote{See `Dockerfile`s at \href{https://github.com/ismailsunni/dockeRs}{https://github.com/ismailsunni/dockeRs}.}.
+
+## Data Science images
+\label{datascience}
+
+_Data Science_ is a widely discussed topic among all academic disciplines \citep[e.g.,][]{donoho_50_2017}.
+The discussions shed a light on the tools and craftspersonship behind the analysis of data with computational methods.
+The practice of Data Science often involves a combination tools and software stacks and requires a cross-cutting skillset.
+This complexity and an inherent concern for openness and reproducibility in the Data Science community lead to Docker being used widely.
+<!--A general description of what to consider when creating a Docker image used for data science projects in R is provided on [environments.rstudio.com](https://environments.rstudio.com/docker).-->
+This section presents exemplary Docker images and image stacks featuring R intended for Data Science.
+
+The [**Jupyter Docker Stacks**](https://github.com/jupyter/docker-stacks/) project are a set of ready-to-run Docker images containing Jupyter applications and interactive computing tools \citep{project_jupyter_jupyter_2018}.
+The `jupyter/r-notebook` image includes R and "popular packages", and naturally also the IRKernel ([https://irkernel.github.io/](https://irkernel.github.io/)), an R kernel for Jupyter so that Jupyter Notebooks can contain R code cells.
+R is also included in the catchall `jupyter/datascience-notebook` image\footnote{\href{https://jupyter-docker-stacks.readthedocs.io/en/latest/using/selecting.html}{https://jupyter-docker-stacks.readthedocs.io/en/latest/using/selecting.html}}.
+For example, these images allow users to quickly start a Jupyter Notebook server locally or build their own specialised images on top of stable toolsets.
+R is installed using the Conda package manager, which can manage environments for various programming languages, pinning both the R version and the versions of R packages\footnote{See \texttt{jupyter/datascience-notebook}'s \texttt{Dockerfile} at \href{https://github.com/jupyter/docker-stacks/blob/master/datascience-notebook/Dockerfile\#L47}{https://github.com/jupyter/docker-stacks/blob/master/datascience-notebook/Dockerfile\#L47}.}.
+
+**Kaggle** provides the [`gcr.io/kaggle-images/rstats`](https://hub.docker.com/r/kaggle/rstats) image (previously `kaggle/rstats`) and [corresponding `Dockerfile`](https://github.com/Kaggle/docker-rstats) for usage in their Machine Learning competitions and easy access to the associated datasets.
+It includes machine learning libraries such as Tensorflow and Keras (see also image `rocker/ml` in Section&nbsp;\nameref{rocker-gpu}), and also configures the \CRANpkg{reticulate} package.
+The image uses a base image with _all packages from CRAN_, `gcr.io/kaggle-images/rcran`, which requires a Google Cloud Build as Docker&nbsp;Hub would time out\footnote{Originally, a stacked collection of over 20 images with automated builds on Docker~Hub was used, see \href{https://web.archive.org/web/20190606043353/http://blog.kaggle.com/2016/02/05/how-to-get-started-with-data-science-in-containers/}{https://web.archive.org/web/20190606043353/http://blog.kaggle.com/2016/02/05/how-to-get-started-with-data-science-in-containers/} and \href{https://hub.docker.com/r/kaggle/rcran/dockerfile}{https://hub.docker.com/r/kaggle/rcran/dockerfile}}.
+The final extracted image size is over 25GB, which makes it debatable if having everything available is actually convenient.
+
+```{bash kaggle_images, eval=FALSE, include=FALSE}
+docker pull gcr.io/kaggle-images/rcran
+docker pull kaggle/rcran # over 22 months old
+docker pull gcr.io/kaggle-images/rstats
+docker images | grep kaggle
+```
+
+As a further example, [**Radiant** project](https://radiant-rstats.github.io/docs/) provides several images, e.g., [`vnijs/rsm-msba-spark`](https://hub.docker.com/r/vnijs/rsm-msba-spark), for their browser-based business analytics interface based on \CRANpkg{Shiny} (`Dockerfile` [on GitHub](https://github.com/radiant-rstats/docker)) and for use in education as part of an MSc course.
+As data science often applies a multitude of tools, this image favours inclusion over selection and features Python, Postgres, JupyterLab and Visual Studio Code besides R and RStudio, bringing the image size up to 9GB.
+
+```{bash radiant_images, eval=FALSE, include=FALSE}
+docker pull vnijs/rsm-msba-spark
+docker images | grep vnijs
+```
+
+**Gigantum** ([http://gigantum.com/](http://gigantum.com/) is a platform for open and decentralized data science with a focus on using automation and user-friendly tools for easy sharing of reproducible computational workflows.
+Gigantum builds on the _Gigantum Client_ (running either locally or on a remote server) for development and execution of data-focused _Projects_, which can be stored and shared via the _Gigantum Hub_ or via a zipfile export.
+The Client is a user friendly interface to a backend using Docker containers to package, build, and run Gigantum projects.
+It is configured to use a default set of Docker base images ([https://github.com/gigantum/base-images](https://github.com/gigantum/base-images)), and users are able to define and configure their own custom images.
+The available images include two with R based on Ubuntu Linux have the [`c2d4u` CRAN PPA](https://launchpad.net/~marutter/+archive/ubuntu/c2d4u3.5/) pre-configured for installation of binary R packages\footnote{\href{https://docs.gigantum.com/docs/using-r}{https://docs.gigantum.com/docs/using-r}}.
+The R images vary in the included authoring environment, i.e., Jupyter in `r-tidyverse` or both Jupyter \& RStudio in `rstudio-server`.
+The independent image stack can be traced back to the Gigantum environment and its features.
+The R images are based on Gigantum's `python3-minimal` image, originally to keep the existing front-end configuration, but also to provide a consistent Python-to-R interoperability.
+The `Dockerfile`s also use build args to specify bases, for example for different version of NVIDIA CUDA for GPU processing\footnote{See \href{https://github.com/gigantum/base-images/blob/master/_templates/python3-minimal-template/Dockerfile}{https://github.com/gigantum/base-images/blob/master/\_templates/python3-minimal-template/Dockerfile} for the \texttt{Dockerfile} of \texttt{python3-minimal}.} so that appropriate GPU drivers can be enabled automatically when supported.
+Furthermore, Gigantum's focus lies on environment management via GUI and ensuring a smooth user interaction, e.g., with reliable and easy conflict detection and resolution.
+For this reason, project repositories store authoritative package information in a separate file per package, allowing Git to directly detect conflicts and changes.
+A Dockerfile is generated from this description that inherits from the specified base image, and additional custom Docker instructions may be appended by users, though Gigantum's default base images do not currently include the `littler` tool, which is used by Rocker to install packages within `Dockerfile`s.
+Because of these specifics, instructions from `rocker/r-ubuntu` could _not_ be readily re-used in this image stack (see Section&nbsp;\nameref{conclusions}).
+Both approaches enable `apt` as an installation method, and this is exposed via the GUI-based environment management,\footnote{See \href{https://docs.gigantum.com/docs/environment-management}{https://docs.gigantum.com/docs/environment-management}}.
+The image build and publication process is scripted with Python and JSON template configuration files, unlike Rocker which relies on plain `Dockerfile`s.
+A minor reason in the inception of the images were also project constraints requiring a Rocker-incompatible licensing of the `Dockerfile`s, i.e., the MIT License.
+
+**Whole&nbsp;Tale** (https://wholetale.org) is a National Science Foundation (NSF) funded project developing a scalable, open-source, multi-user platform for reproducible research \citep{brinckman2019, chard2019a}.
+A central goal of the platform is to enable researchers to easily create and publish executable research objects\footnote{In Whole~Tale a _tale_ is a research object that contains metadata, data (by copy or reference), code, narrative, documentation, provenance, and information about the computational environment to support computational reproducibility} associated with published research \citep{chard2019b}.
+Using Whole&nbsp;Tale, researchers can create and publish Rocker-based reproducible research objects to a growing number of repositories including DataONE member nodes, Zenodo and soon Dataverse.
+Additionally, Whole&nbsp;Tale supports automatic data citation and is working on capabilities for image preservation and provenance capture to improve the transparency of published computational research artifacts \citep{mecum2018, mcphillips2019}.
+For R users, Whole&nbsp;Tale attempts to combine the strengths of the Rocker Project's curated Docker images with the Jupyter Project's `repo2docker`\footnote{See \href{https://repo2docker.readthedocs.io/}{https://repo2docker.readthedocs.io/}} component to simplify the customization of R-based environments for researchers with limited experience with either Docker or Git.
+Multiple options have been discussed to allow users to change the base image used in `repo2docker` from the default Ubuntu LTS (long-term support) required to support the Rocker Project images. 
+Whole&nbsp;Tale implemented a custom `RockerBuildPack`\footnote{See \href{https://github.com/whole-tale/repo2docker\_wholetale}{https://github.com/whole-tale/repo2docker\_wholetale}} to support customization of the `rocker/geospatial` image through `repo2docker` composability\footnote{Composability refers to the ability to combine multiple package managers -- such as R, `pip`, and `conda`}.
+This works in part because Rocker images are based on a Debian distribution, so the instructions created by `repo2docker` for Ubuntu work because of compatible shell and package manager.
+
+# Use cases and applications
+<!-- from packages closer to Docker to second level packages to more complex use cases -->
+\label{applications}
+
+## Interfaces for Docker in R
 \label{interfaces}
@@
 ```{r clientfunctionalities, echo=FALSE, results='asis', warning=FALSE, message=FALSE}
-library("tibble")
-library("tidyr")
-library("dplyr")
-
 functionalities <- tribble(
@@
 
-# insert table
+# inserting a table in the R Journal currently fails
+# waiting for new release of articles including this PR to be published: https://github.com/rstudio/rticles/pull/261
+# Until then, use remotes::install_github("rstudio/rticles")
+if (packageVersion("rticles") >= "0.12.3") {
   requireNamespace("kableExtra")
-knitr::kable(tab, format = "latex", escape = FALSE) %>%
+  knitr::kable(tab, escape = FALSE) %>%
     kableExtra::row_spec(0, angle = -90)
+} else {
+  cat("\n\n[Functionality overview table omitted due to a small bug in rticles]\n\n")
+}
 ```
@@
 \CRANpkg{stevedore}'s own interface is automatically generated based on the OpenAPI specification of the Docker daemon, but still similar to the Docker CLI.
-The interface is similar to R6 objects, in that a `stevedore_object` has a number of functions attached to it that can be called, and multiple specific versions of the Docker API can be supported thanks to the automatic generation\footnote{See \href{https://github.com/richfitz/stevedore/blob/master/development.md}{https://github.com/richfitz/stevedore/blob/master/development.md}.}.
+The interface is similar to R6 objects, in that a `stevedore_object` has a number of functions attached to it that can be called, and multiple specific versions of the Docker API can be supported thanks to the automatic generation\footnote{See \href{https://github.com/richfitz/stevedore/tree/master/inst/spec}{https://github.com/richfitz/stevedore/tree/master/inst/spec} and \href{https://github.com/richfitz/stevedore/blob/master/development.md}{https://github.com/richfitz/stevedore/blob/master/development.md}.}.
 
-**\CRANpkg{AzureContainers}** is an interface to a number of container-related services in Microsoft's [Azure Cloud](https://azure.microsoft.com/) \citep{AzureContainers_2019}.
+**\CRANpkg{AzureContainers}** is an interface to a number of container-related services in Microsoft's [Azure Cloud](https://azure.microsoft.com/), see Section&nbsp;\nameref{azurecontainers}.
 While it is mainly intended for working with Azure, as a convenience feature it includes lightweight, cross-platform shells to Docker and Kubernetes (tools `kubectl` and `helm`).
@@
 
-# Use cases and applications
-\label{applications}
-
-## Image stacks for communities of practice
-
-**Bioconductor** ([https://bioconductor.org/](https://bioconductor.org/)) is an open source, open development project for the analysis and comprehension of genomic data \citep{gentleman_bioconductor_2004}.
-The project consists of 1823 R software packages as of October 30th 2019, as well as packages containing annotation or experiment data.
-_Bioconductor_ has a semi-annual release cycle, each release is associated with a particular version of R, and Docker images are provided for current and past versions of _Bioconductor_ for convenience and reproducibility.
-All images, included  are described on the _Bioconductor_ web site (see [https://bioconductor.org/help/docker/](https://bioconductor.org/help/docker/)), created with `Dockerfile`s maintained on GitHub, and distributed through Docker&nbsp;Hub\footnote{See \href{https://github.com/Bioconductor/bioconductor_docker}{https://github.com/Bioconductor/bioconductor\_docker} and \href{https://hub.docker.com/u/bioconductor}{https://hub.docker.com/u/bioconductor} respectively.}.
-_Bioconductor_'s 'base' Docker images are built on top of the `rocker/rstudio` image.
-_Bioconductor_ installs packages based on the R version in combination with the Bioconductor version, and therefore uses Bioconductor version tagging `devel` and `RELEASE_X_Y`, e.g., `RELEASE_3_10`.
-Past and current combinations of R and _Bioconductor_ will therefore be accessible via a specific image tags.
-The _Bioconductor_ `Dockerfile` selects the desired version of R from Rocker, adds required system dependencies, and uses the \CRANpkg{BiocManager} package for installing appropriate versions of _Bioconductor_ packages.
-A strength of this approach is that the responsibility for complex software configuration (including customized development) is shifted from the user to the experienced _Bioconductor_ core team.
-A recent audit of the _Bioconductor_  image stack `Dockerfile` led to the deprecation of several community-maintained images, because the numerous specific images became too hard to understand, complex to maintain, and cumbersome to extent.
-As part of the simplification, a recent innovation is to produce a `bioconductor_docker:devel` image to emulate the _Bioconductor_ nightly Linux build machine as closely as possible.
-This image contains the build system environment variables and the _system dependencies_ needed to install and check almost all (1813 out of 1823) _Bioconductor_ software packages and saves users and package developers from managing these themselves.
-Furthermore the image is configured so that `.libPaths()` has `/usr/local/lib/R/host-site-library` as the first location.
-Users mounting a location on the host file system to this location can persist installed packages across Docker sessions or image updates.
-Many R users pursue flexible work flows tailored to particular analysis needs, rather than standardized work flows.
-The new `bioconductor_docker` image is well-suited for this pattern, while `bioconductor_docker:devel` provides developers with a test environment close to _Bioconductor_'s build system.
-
-\label{datascience}
-**Data Science** is a widely discussed topic among all academic disciplines \citep[e.g.,][]{donoho_50_2017}.
-The discussions shed a light on the tools and craftspersonship behind the analysis of data with computational methods.
-The practice of Data Science often involves a combination tools and software stacks and requires a cross-cutting skillset.
-This complexity and an inherent concern for openness and reproducibility in the Data Science community lead to Docker being used widely.
-<!--A general description of what to consider when creating a Docker image used for data science projects in R is provided on [environments.rstudio.com](https://environments.rstudio.com/docker).-->
-The remaineder of this section presents exemplary Docker images and image stacks featuring R intended for Data Science.
-The [_Jupyter Docker Stacks_](https://github.com/jupyter/docker-stacks/) project are a set of ready-to-run Docker images containing Jupyter applications and interactive computing tools \citep{project_jupyter_jupyter_2018}.
-The `jupyter/r-notebook` image includes R and "popular packages", and naturally also the IRKernel ([https://irkernel.github.io/](https://irkernel.github.io/)), an R kernel for Jupyter so that Jupyter Notebooks can contain R code cells.
-R is also included in the catchall `jupyter/datascience-notebook` image\footnote{\href{https://jupyter-docker-stacks.readthedocs.io/en/latest/using/selecting.html}{https://jupyter-docker-stacks.readthedocs.io/en/latest/using/selecting.html}}.
-For example, these images allow users to quickly start a Jupyter Notebook server locally or build their own specialised images on top of stable toolsets.
-R is installed using the Conda package manager, which can manage environments for various programming languages, pinning both the R version and the versions of R packages\footnote{See \texttt{jupyter/datascience-notebook}'s \texttt{Dockerfile} at \href{https://github.com/jupyter/docker-stacks/blob/master/datascience-notebook/Dockerfile\#L47}{https://github.com/jupyter/docker-stacks/blob/master/datascience-notebook/Dockerfile\#L47}.}.
-_Kaggle_ provides the [`gcr.io/kaggle-images/rstats`](https://hub.docker.com/r/kaggle/rstats) image (previously `kaggle/rstats`) and [corresponding `Dockerfile`](https://github.com/Kaggle/docker-rstats) for usage in their Machine Learning competitions and easy access to the associated datasets.
-It includes machine learning libraries such as Tensorflow and Keras (see also image `rocker/ml` in Section&nbsp;\nameref{rocker-gpu}), and also configures the \CRANpkg{reticulate} package.
-The image uses a base image with _all packages from CRAN_, `gcr.io/kaggle-images/rcran`, which requires a Google Cloud Build as Docker&nbsp;Hub would time out\footnote{Originally, a stacked collection of over 20 images with automated builds on Docker~Hub was used, see \href{https://web.archive.org/web/20190606043353/http://blog.kaggle.com/2016/02/05/how-to-get-started-with-data-science-in-containers/}{https://web.archive.org/web/20190606043353/http://blog.kaggle.com/2016/02/05/how-to-get-started-with-data-science-in-containers/} and \href{https://hub.docker.com/r/kaggle/rcran/dockerfile}{https://hub.docker.com/r/kaggle/rcran/dockerfile}}.
-The final extracted image size is over 25GB, which makes it debatable if having everything available is actually convenient.
-As a further example, [_Radiant_](https://radiant-rstats.github.io/docs/) project provides several images, e.g., [`vnijs/rsm-msba-spark`](https://hub.docker.com/r/vnijs/rsm-msba-spark), for their browser-based business analytics interface based on \CRANpkg{Shiny} (`Dockerfile` [on GitHub](https://github.com/radiant-rstats/docker)) and for use in education as part of an MSc course.
-As data science often applies a multitude of tools, this image favours inclusion over selection and features Python, Postgres, JupyterLab and Visual Studio Code besides R and RStudio, bringing the image size up to 9GB.
-_Gigantum_ ([http://gigantum.com/](http://gigantum.com/)) is a platform for open and decentralized data science with a focus on using automation and user-friendly tools for easy sharing of reproducible computational workflows.
-Gigantum builds on the _Gigantum Client_ (running either locally or on a remote server) for development and execution of data-focused _Projects_, which can be stored and shared via the _Gigantum Hub_ or via a zipfile export.
-The Client is a user friendly interface to a backend using Docker containers to package, build, and run Gigantum projects.
-It is configured to use a default set of Docker base images ([https://github.com/gigantum/base-images](https://github.com/gigantum/base-images)), and users are able to define and configure their own custom images.
-The available images include two with R based on Ubuntu Linux have the [`c2d4u` CRAN PPA](https://launchpad.net/~marutter/+archive/ubuntu/c2d4u3.5/) pre-configured for installation of binary R packages\footnote{\href{https://docs.gigantum.com/docs/using-r}{https://docs.gigantum.com/docs/using-r}}.
-The R images vary in the included authoring environment, i.e., Jupyter in `r-tidyverse` or both Jupyter \& RStudio in `rstudio-server`.
-The independent image stack can be traced back to the Gigantum environment and its features.
-The R images are based on Gigantum's `python3-minimal` image, originally to keep the existing front-end configuration, but also to provide a consistent Python-to-R interoperability.
-The `Dockerfile`s also use build args to specify bases, for example for different version of NVIDIA CUDA for GPU processing\footnote{See \href{https://github.com/gigantum/base-images/blob/master/_templates/python3-minimal-template/Dockerfile}{https://github.com/gigantum/base-images/blob/master/\_templates/python3-minimal-template/Dockerfile} for the \texttt{Dockerfile} of \texttt{python3-minimal}.} so that appropriate GPU drivers can be enabled automatically when supported.
-Furthermore, Gigantum's focus lies on environment management via GUI and ensuring a smooth user interaction, e.g., with reliable and easy conflict detection and resolution.
-For this reason, project repositories store authoritative package information in a separate file per package, allowing Git to directly detect conflicts and changes.
-A Dockerfile is generated from this description that inherits from the specified base image, and additional custom Docker instructions may be appended by users, though Gigantum's default base images do not currently include the `littler` tool, which is used by Rocker to install packages within `Dockerfile`s.
-Because of these specifics, instructions from `rocker/r-ubuntu` could _not_ be readily re-used in this image stack (see Section&nbsp;\nameref{conclusions}).
-Both approaches enable `apt` as an installation method, and this is exposed via the GUI-based environment management,\footnote{See \href{https://docs.gigantum.com/docs/environment-management}{https://docs.gigantum.com/docs/environment-management}}.
-The image build and publication process is scripted with Python and JSON template configuration files, unlike Rocker which relies on plain `Dockerfile`s.
-A minor reason in the inception of the images were also project constraints requiring a Rocker-incompatible licensing of the `Dockerfile`s, i.e., the MIT License.
-
-```{bash kaggle_images, eval=FALSE, include=FALSE}
-docker pull gcr.io/kaggle-images/rcran
-docker pull kaggle/rcran # over 22 months old
-docker pull gcr.io/kaggle-images/rstats
-docker images | grep kaggle
-```
-
-```{bash radiant_images, eval=FALSE, include=FALSE}
-docker pull vnijs/rsm-msba-spark
-docker images | grep vnijs
-```
-
 ## Capture and create environments
@@
 
-Community maintained images provide a solid basis for extension as part of individual requirements.
 Several second order R packages attempt to make the process of creating Docker images and using containers for specific tasks, such as running tests or rendering reproducible reports, easier.
-While authoring and managing an environment with Docker by hand is possible and feasible for experts\footnote{See, e.g., this tutorial by RStudio on how to manage environments and package versions and to ensure deterministic image builds with Docker: \href{https://environments.rstudio.com/docker}{https://environments.rstudio.com/docker}.}, the following examples show the power of automation when environments become too cumbersome.
+While authoring and managing an environment with Docker by hand is possible and feasible for experts\footnote{See, e.g., this tutorial by RStudio on how to manage environments and package versions and to ensure deterministic image builds with Docker: \href{https://environments.rstudio.com/docker}{https://environments.rstudio.com/docker}.}, the following examples show the power of automation when environments become too cumbersome or acquiring the skills is not possible in due course.
 Especially _version pinning_, with packages \pkg{remotes} and \pkg{versions} for R or by using MRAN, and with system package managers for different operating systems, can greatly increase the reproducibility of built images and are common approaches.
@@
 \pkg{containerit} uses \CRANpkg{stevedore} to streamline the user interaction and improve the created `Dockerfile`s, e.g., by running a container for the desired base image to extract the already available R packages.
-**\pkg{dockr}** ([https://github.com/smaakage85/dockr](https://github.com/smaakage85/dockr)) is a very similar package attempting to mirror a given R session, including local non-CRAN packages\citep{kjeldgaard_dockr_2019}.
+**\pkg{dockr}** ([https://github.com/smaakage85/dockr](https://github.com/smaakage85/dockr)) is a similar package focusing on the generation of Docker images for R packages, in which the package itself and all of the R dependencies, including local non-CRAN packages, are available \citep{kjeldgaard_dockr_2019}.
+\pkg{dockr} facilitates the organisation of code in the R package structure and the resulting Docker image mirrors the package versions of the current R session.
 Users can manually add statements for non-R dependencies to the `Dockerfile`.
@@
 
-System dependencies are the domain of Docker, but for a full description of the computing environment one must also manage the R version and the the R packages.
-R&nbsp;versions are available via the versioned Rocker image stack \citep{RJ-2017-065}.
-[r-online](https://github.com/ColinFay/ronline) leverages these images and provides an app for helping users to detect breaking changes between different R versions, and for historic exploration of R.
-With a standalone NodeJS app or [online](https://srv.colinfay.me/r-online), the user can compare a piece of code run in two separate versions of R.
-Internally, r-online opens one or two Docker instances with the given version of R based on Rocker images, executes a given piece of code, and returns the result to the user.
-R&nbsp;package management can be achieved with MRAN, or with packages such as \CRANpkg{checkpoint} and **\CRANpkg{renv}**, which can naturally be applied within images and containers.
-For example, \CRANpkg{renv} ([https://rstudio.github.io/renv/](https://rstudio.github.io/renv/)) helps users to manage the state of the R library in a reproducible way, further providing isolation and portability \citep{renv2019}.
-While \CRANpkg{renv} does not cover system dependencies, the \CRANpkg{renv}-based environment can be transferred into a container either by restoring the environment based on the main configuration file `renv.lock` or by storing the \CRANpkg{renv}-cache on the host and not in the container \citep{ushey_using_2019}.
-With both the system dependencies and R packages consciously managed in a Docker image, users can start using containers as the _only_ environment for their workflows, which allows them to work indepenently of physical computers\footnote{Allowing them to be digital "nomads", cf.
-J.~Bryan's \href{https://github.com/jennybc/docker-why}{https://github.com/jennybc/docker-why}.} and to assert a specific degree of confidence in the stability of a developed software \citep[cf.
-\texttt{README.Rmd} in][]{marwick_research_2017}.
+The package **\CRANpkg{renv}** ([https://rstudio.github.io/renv/](https://rstudio.github.io/renv/)) helps users to manage the state of the R library in a reproducible way, further providing isolation and portability \citep{renv2019}.
+The package does not cover system dependencies, though, the \CRANpkg{renv}-based environment can be transferred into a container either by restoring the environment based on the main configuration file `renv.lock` or by storing the \CRANpkg{renv}-cache on the host and not in the container \citep{ushey_using_2019}.
 
-## Development, debugging, and testing
-\label{development}
-
-Containers can also serve as useful playgrounds to create environments ad-hoc or to provide very specific environments that are not needed or not easily available in day-to-day development for the purposes of developing R packages.
-These environments may have specific versions of R, of R extension packages, and of system libraries used by R extension packages, and all of the above in a specific combination.
-
-First, such containers can greatly facilitate **fixing bugs and quick evaluation**, because developers and users can readily start and later discard a container to investigate a bug report or try out a piece of software \citep[cf.][]{ooms_opencpu_2017}.
-The container does not affect their regular system.
-Using the Rocker images with RStudio, these disposable environments lack no development comfort (cf. Section&nbsp;\nameref{compendia}).
-\citet{ooms_opencpu_2017} describes how `docker exec` can be used to get a root shell in a container for cusotmization during software evaluation.
-\citet{eddelbuettel_debugging_2019} describes an example how a Docker container was used to debug an issue with a package only occuring with a particular version of Fortran, and using tools which are not readily available on all platforms (e.g., not on macOS).
-
-Second, the strong integration of **system libraries in core packages** in the [R-spatial community](https://www.r-spatial.org/) makes containers essential for stable and proactive development of common classes for geospatial data modelling and analysis.
-For example, GDAL \citep{gdal_2019} is a crucial library in the geospatial domain.
-GDAL is a system dependency allowing R packages such as \CRANpkg{sf}, which provides the core data model for geospatial vector data, or \CRANpkg{rgdal}, to accomodate users to be able to read and write hundreds of different spatial raster and vector formats.
-\CRANpkg{sf} and \CRANpkg{rgdal} have hundreds of indirect reverse imports and dependencies and therefore the maintainers spend a lot of efforts not to break these.
-Purpose built Docker are used to prepare for upcoming releases of system libraries, individual bug reports, and for the lowest supported versions of system libraries\footnote{Cf.
-\href{https://github.com/r-spatial/sf/tree/master/inst/docker}{https://github.com/r-spatial/sf/tree/master/inst/docker}, \href{https://github.com/Nowosad/rspatial_proj6}{https://github.com/Nowosad/rspatial\_proj6}, and \href{https://github.com/r-spatial/sf/issues/1231}{https://github.com/r-spatial/sf/issues/1231}}.
-```{r geospatial_deps, eval=FALSE, include=FALSE}
-length(tools::dependsOnPkgs(c("rgdal", "sf"), dependencies = c("most"), recursive = TRUE))
-```
-
-Third, there are special purpose images for identifying problems beyond the mere R code, such as **debugging R memory problems**.
-The images significantly reduce the barrier to follow complex steps for fixing memory allocation bugs \citep[cf. Section~4.3 in][]{core_writing_1999}.
-These problems are hard to debug and critical, both because when they occur they lead to fatal crashing processes.
-[`rocker/r-devel-san`](https://github.com/rocker-org/r-devel-san) and [`rocker/r-devel-ubsan-clang`](https://github.com/rocker-org/r-devel-san-clang) are Docker images have a particularly configured version of R to trace such problems with gcc and clang compilers, respectively \citep[cf.~\CRANpkg{sanitizers} for examples,][]{eddelbuettel_sanitizers_2014}.
-The image [`wch/r-debug`](https://github.com/wch/r-debug) is a purpose built Docker image with _multiple_ instrumented builds of R, each with a different diagnostic utility activated.
-
-Fourth, containers are useful for **testing** R code during development.
-To submit a package to CRAN, an R package must work with the development version of R, which must be compiled locally.
-That can be a challenge for some users.
-The **R-hub** project provides _"a collection of services to help R package development"_, with the package builder as the most prominent one\citep{r-hub_docs_2019}.
-R-hub makes it easy to ensure that no errors occur, but to fix errors a local setup is still often warranted, e.g., using the image `rocker/r-devel`, and to test packages with native code, which can make the process more complex \citep[cf.][]{eckert_building_2018}.
-The R-hub Docker images can also be used to debug problems locally using various combinations of Linux platforms, R versions, and compilers\footnote{See \href{https://r-hub.github.io/rhub/articles/local-debugging.html}{https://r-hub.github.io/rhub/articles/local-debugging.html} and \href{https://blog.r-hub.io/2019/04/25/r-devel-linux-x86-64-debian-clang/}{https://blog.r-hub.io/2019/04/25/r-devel-linux-x86-64-debian-clang/}}.
-The images go beyond the configurations, or _flavours_, used by CRAN for checking packages\footnote{\href{https://cran.r-project.org/web/checks/check_flavors.html}{https://cran.r-project.org/web/checks/check\_flavors.html}}, e.g., with CentOS-based images, but lack a container for checking on Windows or OS X.
-The images greatly support package developers to provide support on operating systems they are not familiar.
-The package \pkg{dockertest} ([https://github.com/traitecoevo/dockertest/](https://github.com/traitecoevo/dockertest/)) is a proof of concept for automatically generating `Dockerfile`s and building images specifically to run tests\footnote{\pkg{dockertest} is not actively maintained, but mentioned still because of its interesting approach.}.
-These images are accompanied with a special launch script so the tested source code is not stored in the image but the currently checked in version from a local Git repository is cloned into the container at runtime.
-This approach clearly seperates test environment, test code, and current working copy of the code.
-\label{rselenium} Another use case where a container helps to standardise tests across operating systems is detailed the vignettes of the package \CRANpkg{RSelenium} \citep{rselenium_2019}.
-The package recommends Docker for running the [Selenium](https://selenium.dev/) Server application needed to execute test suites on browser-based user interfaces and webpages, but requires users to manually manage the Docker containers.
-
-\label{ci}
-Fifth, Docker images can be used **on continuous integration (CI) platforms** to streamline the testing of packages.
-\citet{ye_docker_2019} describes how they speed up the process of testing by running tasks on [Travis&nbsp;CI](https://travis-ci.org/) within a container using `docker exec`, e.g., the package check or rendering of documentation.
-\citet{cardozo_faster_2018} saved time, also on Travis&nbsp;CI, by re-using the testing image as the base for an image intended for publication on Docker&nbsp;Hub.
-[`r-ci`](https://github.com/ColinFay/r-ci) is in turn used with [GitLab&nbsp;CI](https://docs.gitlab.com/ee/ci/), which itself is built on top of Docker images: the user specifies a base Docker image, and the whole tests are run inside this environment.
-The `r-ci` image stack combines `rocker` versioning and a series of tools specifically designed for testing in a fixed environment with a customized list of preinstalled packages.
-Especially for long running tests or complex system dependencies, these approaches to seperate installation of build dependencies with code testing streamline the development process.
-Not due to a concern about time, but to control the environment used on a CI server, even this manuscript is rendered into a PDF and deployed to a GitHub-hosted website with every change (see `.travis.yml` and `Dockerfile` in the manuscript repository).
-This gives on the one hand easy access after every update of the R Markdown source code, and on the other hand a second controlled environment making sure that the article renders successfully and correctly.
-
-## Processing
-
-The portability of containerized environments becomes particularly useful for improving expensive processing of data or shipping complex processing pipelines.
-First, it is possible to **offload complex processing to a server** or clouds, and also parallel executing of processes for speeding up or serving of many users.
-**\CRANpkg{batchtools}** provides a parallel implementation of Map for various schedulers \citep{Lang2017batchtools}.
-The package can [schedule jobs with Docker Swarm](https://mllg.github.io/batchtools/reference/makeClusterFunctionsDocker.html)
-**\CRANpkg{googleComputeEngineR}** has the function `gce_vm_cluster()` to create clusters of 2 or more virtual machines, running multi-CPU architectures.
-Instead of running a local R script with the local CPU and RAM restrictions, the same code can be processed on all CPU threads of the cluster of machines in the cloud, all running a Docker container with the same R environments.
-\CRANpkg{googleComputeEngineR} integrates with the R parralisation package \CRANpkg{future} to enable this with only a few lines or R code\footnote{\href{https://cloudyr.github.io/googleComputeEngineR/articles/massive-parallel.html}{https://cloudyr.github.io/googleComputeEngineR/articles/massive-parallel.html}}.
-[Google Cloud Run](https://cloud.run) is a CaaS (Containers as a Service) platform.
-Users can launch containers using any Docker image without worrying about underlying infrastructure in a so called serverless configuration.
-The service takes care of network ingress, scaling machines up and down, authentication and authorisation---all features which are non-trivial for a developer to create on their own.
-This can be used to scale up R code to millions of instances if need be with little or no changes to existing code, as demonstrated by the proof of concept `cloudRunR`\footnote{\href{https://github.com/MarkEdmondson1234/cloudRunR}{https://github.com/MarkEdmondson1234/cloudRunR}}, which uses Cloud Run to create a scalable R-based API using \CRANpkg{plumber}.
-[Google Cloud Build](https://cloud.google.com/cloud-build/) and the Google Container Registry are a continuous integration service respectively image registry that offload building of images to the cloud, while serving the needs of commercial environments such as private Docker images or image stacks.
-As Googl Cloud Build itself can run any container, the package \pkg{googleCloudRunner} demonstrates how R can be used as the control language for one time or batch processing jobs and scheduling of jobs\footnote{\href{https://code.markedmondson.me/googleCloudRunner/articles/cloudbuild.html}{https://code.markedmondson.me/googleCloudRunner/articles/cloudbuild.html}}.
-\label{drake} **\CRANpkg{drake}** is a workflow manager for data science projects \citep{landau_drake_2019}.
-It features implicit parallel computing and automated detection of the parts of the work that actually needs to be reexecuted.
-\CRANpkg{drake} has been demonstrated to run inside containers for high reproducibility\footnote{See for example \href{https://github.com/joelnitta/pleurosoriopsis}{https://github.com/joelnitta/pleurosoriopsis} or \href{https://gitlab.com/ecohealthalliance/drake-gitlab-docker-example}{https://gitlab.com/ecohealthalliance/drake-gitlab-docker-example}, the latter even running in a continuous integration platform (cf.~\nameref{development}.}, but also how to use \CRANpkg{future} package's function `makeClusterPSOCK()` to send parts of the workflow to a Docker image for execution\footnote{\href{https://docs.ropensci.org/drake/index.html?q=docker\#with-docker}{https://docs.ropensci.org/drake/index.html?q=docker\#with-docker}} \citep[see package's function documentation;~][]{future_2020}.
-In the latter case, the container control code must be written by the user and the \CRANpkg{future} package ensures the host and worker can connect for communicating over socket connections.
-**RStudio Server Pro** includes a functionality called [Launcher](https://solutions.rstudio.com/launcher/overview/) (since version&nbsp;1.2, released in 2019).
-It gives users the ability to spawn R sessions and background/batch jobs in a scalable way on external clusters, e.g., [Kubernetes based on Docker images](https://support.rstudio.com/hc/en-us/articles/360019253393-Using-Docker-images-with-RStudio-Server-Pro-Launcher-and-Kubernetes) or [Slurm](https://slurm.schedmd.com/) clusters, and optionally, with Singularity containers.
-A benefit of the proprietary Launcher software is the ability for R and Python users to leverage containerisation's advantages in RStudio without writing specific deployment scripts or learning about Docker or managing clusters at all.
-
-\label{pipelines}
-Second, containers are perfectly suited for **packaging and executing software pipelines** and required data.
-Containers allow building complex processing pipelines that are independent from the host programming language.
-Due to its original use case (see&nbsp;\nameref{introduction}), Docker has no standard mechanisms for chaining containers together; it lacks definitions and protocols for environment variables, volume mounts and/or ports that could enable transfer of input (parameters and data) and output (results) to and from containers.
-Some packages, e.g., \pkg{containerit}, do provide Docker images that can be used similarly to CLIs, but their usage is cumbersome\footnote{\href{https://o2r.info/containerit/articles/container.html}{https://o2r.info/containerit/articles/container.html}}.
-**\pkg{outsider}** ([https://docs.ropensci.org/outsider/](https://docs.ropensci.org/outsider/)) tackles the problem of integrating external programs into an R workflow \citep{bennett_outsider_2020}.
-Installation and usage of external programs can be difficult, convoluted and even impossible if the platform is incompatible.
-Therefore \pkg{outsider} uses the platform-independent Docker images to encapsulate processes in _outsider modules_.
-Each outsider module has a `Dockerfile` and an R package with functions for interacting with the encapsulated tool.
-Using only R functions, an end-user can install a module with the \pkg{outsider} package and then call module code to integrate a tool into their own R-based workflow seamlessly.
-The \pkg{outsider} package and module manage the containers and handle the transmission of arguments and the transfer of files to and from a container.
-These functionalities also allow a user to launch module code on a remote machine via SSH, expanding the potential computational scale.
-Outsider modules can be hosted code-sharing services, e.g., on GitHub, and \pkg{outsider} contains discovery functions for them.
-<!-- informative examples of how modules work, just as a reminder for @nuest
-https://github.com/DomBennett/om..astral/blob/master/examples/astral.R
-https://github.com/DomBennett/om..raxml#partitioned-dna-analysis
--->
-
-## Deployment and continuous delivery
-\label{deployment}
-
-The cloud is the natural environment of containers, and subsequently containers are the go-to mechanism to deploy R server applications.
-More and more continuous integration (CI) and continuous delivery (CD) services also use containers, opening up new options for use.
-The controlled nature of containers, i.e., the possibility to abstract internal software environment from a minimal dependency outside of the container is crucial, for example to match test or build environments with production environments or transfer runnable entities to as-a-service infrastructures.
-
-First, different packages use containers for the **deployment of R and \CRANpkg{Shiny} apps**.
-\CRANpkg{Shiny} is a popular package for creating interactive online dashboards with R and it enables users with very diverse backgrounds to create stable and user friendly web applications.
-For example, _ShinyProxy_ ([https://www.shinyproxy.io/](https://www.shinyproxy.io/)) is an open source tool to deploy Shiny apps in an enterprise context.
-They feature single sign-on, but also in scientific use cases \citep[e.g., ][]{savini_epiexplorer_2019,glouzon_structurexplor_2017}.
-ShinyProxy uses Docker containers to isolate user sessions and to achieve scalability for multi-user scenarious with multiple apps.
-ShinyProxy itself is written in Java to accomodate corporate requirements and may itself run in a container for stability and availability.
-The tool is build on ContainerProxy ([https://www.containerproxy.io/](https://www.containerproxy.io/)), which provides similar features for executign long-running R jobs or interactive R sessions.
-The started containers can run on a regular Docker host but also in clusters.
-Another example is the package \CRANpkg{golem} ([https://github.com/ThinkR-open/golem](https://github.com/ThinkR-open/golem)), which makes an heavy use of `dockerfiler` when it comes to creating the `Dockerfile` for building production-grade Shiny applications and deploying them.
-\CRANpkg{googleComputeEngineR} ([https://cloudyr.github.io/googleComputeEngineR/](https://cloudyr.github.io/googleComputeEngineR/)) enables quick deployments of key R services, such as RStudio and Shiny, onto cloud virtual machines (VMs) with Google Cloud Compute Engine \citep{googleComputeEngineR_2019}.
-The package utilises `Dockerfile`s to move the labour of setting up those services from the user to a premade Docker image, which is configured and run in the cloud VM.
-For example, by specifying the template `template="rstudio"` in functions `gce_vm_template()/gce_vm()` an up to date RStudio Service image is launched for development work, while specifying `template="rstudio-gpu"` will launch an RStudio Server image with a GPU attached, etc.
-
-Second, containers can be used to create **platform installation packages** in a DevOps setting.
-The [OpenCPU](https://www.opencpu.org/) system provides an HTTP API for data analysis based on R.
-\citet{ooms_opencpu_2017} describes how various platform-specific installation files for OpenCPU are created using Docker&nbsp;Hub: the automated builds install the software stack from source on different operating systems; afterwards a script file downloads the images and extracts the OpenCPU binaries.
-
-Third, containers can greatly facilitate the **deployment to existing infrastructures**.
-_Kubernetes_ ([https://kubernetes.io/](https://kubernetes.io/)) is a container-orchestration system for managing container-based application deployment and scaling.
-A _cluster_ of containers, orchestrated as a single deployment, e.g., with Kubernetes, can mitigate limitations on request volumes or a container occupied with computationally intensive task.
-A cluster features load-balancing, autoscaling of containers across numerous servers (in the cloud or on premise), and restarting failed ones.
-It may be your organisation has a Kubernetes cluster already for other applications, or you may use one of the different provides for managed Kubernetes clusters.
-Docker containers are used within Kubernetes clusters to hold native code, for which Kubernetes creates a framework around network connections and scaling of resources up and down.
-R applications, big parallel tasks, or scheduled batch jobs can be deployed in a scalable way using Kubernetes, and the deployment can even be triggered by changes to code repositories (i.e., CD), see blog post "R on Kubernetes" \citep{edmondson_r_2018}.
-The package \pkg{googleKubernetesR} ([https://github.com/RhysJackson/googleKubernetesR](https://github.com/RhysJackson/googleKubernetesR)) is a proof of concept for wrapping the Google Kubernetes Engine API, Google's hosted Kubernetes solution, in an easy to use R package.
-The package \CRANpkg{analogsea} provides a solution to programmatically create and destroy cloud VMs on the [Digital Ocean](https://www.digitalocean.com/) platform \citep{analogsea_2019}.
-It also includes R wrapper functions to install Docker in such a VM, manage images, and control containers straight from R functions.
-These functions are translated to Docker CLI commands and transferred transparently to the respective remove machine using SSH.
-\CRANpkg{AzureContainers} is an umbrella package providing interfaces three commercial services of Microsoft's Azure Cloud, namely [Container Instances](https://azure.microsoft.com/en-us/services/container-instances/) for running individual containers, [Container Registry](https://azure.microsoft.com/en-us/services/container-registry/) for private image distribution, and [Kubernetes Service](https://azure.microsoft.com/en-us/services/kubernetes-service/) for orchestrated deployments.
-While a package like \CRANpkg{plumber} provides the infrastructure for turning an R workflow a service, for production purposes it is usually necessary to take into account scalability, reliability and ease of management.
-AzureContainers provides an R-based interface to these features and thereby simplifies complex infrastructure management to a number of R function calls, given an Azure account with sufficient credit\footnote{See \emph{"Deploying a prediction service with Plumber"} vignette for details:  \href{https://cran.r-project.org/web/packages/AzureContainers/vignettes/vig01_plumber_deploy.html}{https://cran.r-project.org/web/packages/AzureContainers/vignettes/vig01\_plumber\_deploy.html}.}.
-[Heroku](https://www.heroku.com/) is a further cloud platform as a service provider, who supports container-based applications.
-`heroku-docker-r` ([https://github.com/virtualstaticvoid/heroku-docker-r](https://github.com/virtualstaticvoid/heroku-docker-r)) is an independent project providing a template for deploying R applications based on Heroku's image stack, including multiple examples for interfacing R with other programming languages.
-Yet the approach requires a manual management of the computing environment.
-
-The prevelance of Docker in industry naturally leads to usage of containers with R in such settings as well, as customers already manage platforms in Docker containers.
-These products often entail a high amount of open source software in combination with proprietary layers adding the relevant commercializable features.
-One such example is RStudio's data science platform [RStudio Team](https://rstudio.com/products/team/).
-It allows teams of data scientists and their respective IT/DevOps groups to develop and deploy code in R and Python around the RStudio Open Source Server inside of Docker images, without requiring users to learn new tools or directly interact with containers.
-The best practices for [running RStudio with Docker containers](https://support.rstudio.com/hc/en-us/articles/360021594513-Running-RStudio-with-Docker-containers) as well as [Docker images](https://github.com/rstudio/rstudio-docker-products) for RStudio's commerical products are publicly available.
 ## Using R to power enterprise software in production environments
@@
 The images extend `rocker/r-base` and additionally entailed an SSH client and a workaround for being able to mount SSH keys from the host, Pandoc, the Amazon Web Services (AWS) SDK, and Java, which is required by the AWS SDK.
-The AWS SDK allowed to run R consumers reading from real-time data processing streams of [AWS Kinesis](https://aws.amazon.com/kinesis/) \footnote{See useR!2017 talk \href{https://static.sched.com/hosted\_files/user2017/2f/AWR Kinesis at useR 2017.pdf}{"Stream processing with R in AWS"}.}.
+The AWS SDK allowed to run R consumers reading from real-time data processing streams of [AWS Kinesis](https://aws.amazon.com/kinesis/)\footnote{See useR!2017 talk _Stream processing with R in AWS_ at \href{https://static.sched.com/hosted_files/user2017/2f/AWR%20Kinesis%20at%20useR%202017.pdf}{https://static.sched.com/hosted_files/user2017/2f/AWR Kinesis at use 02017.pdf}.}.
 The applications were deployed on Amazon Elastic Container Service ([ECS](https://aws.amazon.com/ecs/)).
@@
 
+## Continuous integration, continuous delivery, and deployment
+\label{cicd}
+\label{deployment}
+
+The cloud is the natural environment of containers, and subsequently containers are the go-to mechanism to deploy R server applications.
+More and more continuous integration (CI) and continuous delivery (CD) services also use containers, opening up new options for use.
+The controlled nature of containers, i.e., the possibility to define the software environment very well, even on remote machines, is crucial here, for example to match test or build environments with production environments.
+
+As a first example, \CRANpkg{Shiny} is the most popular package for creating interactive online dashboards with R and it enables users with very diverse backgrounds to create stable and user friendly web applications.
+The ShinyProxy ([https://www.shinyproxy.io/](https://www.shinyproxy.io/)) is an open source tool to deploy Shiny apps in an enterprise context.
+They feature single sign-on, but also in scientific use cases \citep[e.g., ][]{savini_epiexplorer_2019,glouzon_structurexplor_2017}.
+ShinyProxy uses Docker containers to isolate user sessions and to achieve scalability for multi-user scenarious with multiple apps.
+ShinyProxy itself is written in Java to accomodate corporate requirements and may itself run in a container for stability and availability.
+The tool is build on ContainerProxy ([https://www.containerproxy.io/](https://www.containerproxy.io/)), which provides similar features for executign long-running R jobs or interactive R sessions.
+The started containers can run on a regular Docker host but also in clusters.
+
+Second, containers can be used to create **platform installation packages** in a DevOps setting.
+The [OpenCPU](https://www.opencpu.org/) system provides an HTTP API for data analysis based on R.
+\citet{ooms_opencpu_2017} describes how various platform-specific installation files for OpenCPU are created using Docker&nbsp;Hub: the automated builds install the software stack from source on different operating systems; afterwards a script file downloads the images and extracts the OpenCPU binaries.
+
+- [dynwrap](https://github.com/dynverse/dynwrap_containers/blob/master/.travis.yml) [@rcannood]
+  - For this project, we use travis-ci to build rocker-derived containers, test them, and only push them to Docker&nbsp;Hub (from travis-ci.org) if the integration tests succeed.
+
+For example, the **[`golem`](https://github.com/ThinkR-open/golem)** package makes an heavy use of `dockerfiler` when it comes to creating the `Dockerfile` for building production-grade Shiny applications and deploying them.
+
+- `RSelenium`
+- [`googleComputeEngineR`](https://cloudyr.github.io/googleComputeEngineR/) (function `gce_vm_template()`)
+  - To enable quick deployments of key R services such as RStudio and Shiny onto cloud virtual machines (VMs), this package utilises Dockerfiles to move the labour of setting up those services from the user to a premade Docker image.
+For example, by specifying the template `template="rstudio"` in `gce_vm_template()/gce_vm()` an up to date RStudio Service image is launched.
+Specifying `template="rstudio-gpu"` will launch an RStudio Server image with a GPU attached, etc.
+- [`analogsea`](https://github.com/sckott/analogsea) (digital ocean R client)
+- [`plumber`](https://www.rplumber.io/docs/hosting.html#docker)
+- [production deployment of neural networks](https://github.com/tmobile/r-tensorflow-api) [@jnolis, @nolistic]
+- Kubernetes - A popular platform for managing Docker containers is [Kubernetes](https://kubernetes.io/), which is used for a wide variety of containerized applications.
+It may be your organisation has a Kubernetes cluster already for other applications.
+Docker containers are used within Kubernetes clusters to hold native code, for which Kubernetes creates a framework around network connections and scaling of resources up and down.
+An introduction on using [Kubernetes with R is published at this blog post](https://code.markedmondson.me/r-on-kubernetes-serverless-shiny-r-apis-and-scheduled-scripts/).
+
+\label{azurecontainers}
+**\CRANpkg{AzureContainers}** is an umbrella package providing interfaces three commercial services of Microsoft's Azure Cloud, namely [Container Instances](https://azure.microsoft.com/en-us/services/container-instances/) for running individual containers, [Container Registry](https://azure.microsoft.com/en-us/services/container-registry/) for private image distribution, and [Kubernetes Service](https://azure.microsoft.com/en-us/services/kubernetes-service/) for orchestrated deployments.
+While a package like \CRANpkg{plumber} provides the infrastructure for turning an R workflow a service, for production purposes it is usually necessary to take into account scalability, reliability and ease of management.
+A _cluster_ of containers, orchestrated as a single deployment, e.g., with [Kubernetes](https://kubernetes.io), can mitigate limitations on request volumes or a container occupied with computationally intensive task.
+A cluster features load-balancing, autoscaling of containers across numerous servers (in the cloud or on premise), and restarting failed ones.
+AzureContainers provides an R-based interface to these features for Azure.
+On the client side, it provides simple wrappers to Docker and to Kubernetes' management tools `kubectl` and `helm`.
+The package simplifies complex infrastructure management to a number of R function calls, given an Azure account with sufficient credit\footnote{See \emph{"Deploying a prediction service with Plumber"} vignette for details:  \href{https://cran.r-project.org/web/packages/AzureContainers/vignettes/vig01_plumber_deploy.html}{https://cran.r-project.org/web/packages/AzureContainers/vignettes/vig01\_plumber\_deploy.html}.}.
+
+The prevelance of Docker in industry naturally leads to usage of containers with R in such settings as well, as customers already manage platforms in Docker containers.
+These products often entail a high amount of open source software in combination with proprietary layers addign the relevant commercializable features.
+One such example is RStudio's data science platform [RStudio Team](https://rstudio.com/products/team/).
+It allows teams of data scientists and their respective IT/DevOps groups to develop and deploy code in R and Python around the RStudio Open Source Server inside of Docker images, without requiring users to learn new tools or directly interact with containers.
+The best practices for [running RStudio with Docker containers](https://support.rstudio.com/hc/en-us/articles/360021594513-Running-RStudio-with-Docker-containers) as well as [Docker images](https://github.com/rstudio/rstudio-docker-products) for RStudio's commerical products are openly available.
+
 ## Common or public work environments
@@
 
-The [**Binder**](https://mybinder.readthedocs.io/en/latest/) project, maintained by the team behind Jupyter, makes it possible for users to **create and share computing environments** with others \citep{jupyter_binder_2018}.
+**\pkg{holepunch}** ([https://github.com/karthik/holepunch](https://github.com/karthik/holepunch)) is an R package that was designed to make sharing work environments accessible to novice R users based on **Binder**.
+The [Binder project](https://mybinder.readthedocs.io/en/latest/), maintained by the team behind Jupyter, makes it possible for users to create and share their computing environments with others \citep{jupyter_binder_2018}.
 A _BinderHub_ allows anyone with access to a web browser and an internet connection to launch a temporary instance of these custom environments and execute any workflows contained within.
@@
 To set up Binder for a project, a user typically starts at an instance of a BinderHub and passes the location of a repository with a workspace, e.g., a hosted Git repository, or a data repository like Zenodo.
-Binder's core internal tool is `repo2docker`.
-It deterministically builds a Docker image by parsing the contents of a repository, e.g., project dependency configurations or simple configuration files\footnote{See supported file types at \href{https://repo2docker.readthedocs.io/en/latest/config\_files.html}{https://repo2docker.readthedocs.io/en/latest/config\_files.html}.}.
+Binder's core internal tool is [`repo2docker`](https://repo2docker.readthedocs.io/en/latest/config_files.html).
+It deterministically builds a Docker image by parsing the contents of a repository, e.g., project dependency configurations or simple configuration files.
 In the most powerful case, `repo2docker` builds a given `Dockerfile`.
 While this approach works well for most run of the mill Python projects, it is not so seamless for R projects.
-This is partly because `repo2docker` does not support arbitrary base images due to the complex auto-generation of the `Dockerfile` instructions.
-Two approaches make using Binder easier. for R users.
-First, **\pkg{holepunch}** ([https://github.com/karthik/holepunch](https://github.com/karthik/holepunch)) is an R package that was designed to make sharing work environments accessible to novice R users based on Binder.
 For any R projects that use the Tidyverse suite \citep{wickham_welcome_2019}, the time and resources required to build all dependencies from source can often time out before completion, making it frustrating for the average R user.
@@
 The local image usage reduces the number of seperately managed environments and thereby reduces work and increases portability and reproducibility.
-Second, the **Whole&nbsp;Tale** project (https://wholetale.org) combines the strengths of the Rocker Project's curated Docker images with `repo2docker`.
-Whole&nbsp;Tale is a National Science Foundation (NSF) funded project developing a scalable, open-source, multi-user platform for reproducible research \citep{brinckman2019, chard2019a}.
-A central goal of the platform is to enable researchers to easily create and publish executable research objects\footnote{In Whole~Tale a \emph{tale} is a research object that contains metadata, data (by copy or reference), code, narrative, documentation, provenance, and information about the computational environment to support computational reproducibility.} associated with published research \citep{chard2019b}.
-Using Whole&nbsp;Tale, researchers can create and publish Rocker-based reproducible research objects to a growing number of repositories including DataONE member nodes, Zenodo and soon Dataverse.
-Additionally, Whole&nbsp;Tale supports automatic data citation and is working on capabilities for image preservation and provenance capture to improve the transparency of published computational research artifacts \citep{mecum2018, mcphillips2019}.
-For R users, Whole&nbsp;Tale extends the Jupyter Project's `repo2docker` component to simplify the customization of R-based environments for researchers with limited experience with either Docker or Git.
-Multiple options have been discussed to allow users to change the base image used in `repo2docker` from the default Ubuntu LTS (long-term support) required to support the Rocker Project images. 
-Whole&nbsp;Tale implemented a custom `RockerBuildPack`\footnote{See \href{https://github.com/whole-tale/repo2docker\_wholetale}{https://github.com/whole-tale/repo2docker\_wholetale}} to support customization of the `rocker/geospatial` image through `repo2docker` composability\footnote{Composability refers to the ability to combine multiple package managers -- such as R, `pip`, and `conda`}.
-This works in part because Rocker images are based on a Debian distribution, so the instructions created by `repo2docker` for Ubuntu work because of compatible shell and package manager.
 
@@
 Teaching is a further example where shared computing environments and sandboxing can greatly improve the process.
-First, **prepared environments for teaching** are especially helpful for (a) introductory courses where the first step of installation and configuration can be hurdles for students \citep{cetinkaya-rundel_infrastructure_2018}, and (b) courses that require access to a relatively complex setup of software tools, e.g., database systems.
-\citet{cetinkaya-rundel_infrastructure_2018} describe how a Docker-based deployment of RStudio (i) avoided problems with troubleshooting individual students computers and greatly increase engagement through very quickly showing tangible outcomes, e.g., a visualisation, and (ii) reduced demand on teaching and IT staff.
-Each student get's acces to a personal RStudio instance running in a container after authentication with the university login, which gives the benefits of sandboxing and possibility of limiting resources.
-\citet{cetinkaya-rundel_infrastructure_2018} found that for the courses at hand, actual usage of the UI is intermittent so a single cloud based VM with 4&nbsp;cores and 28GB&nbsp;RAM sufficed for over 100 containers.
-An example for mitigating _complex setups_ is teaching databases.
+First, **prepared environments for teaching** are especially helpful for courses that require access to a relatively complex setup of software tools, e.g., database systems.
 R is very useful tool for interfacing with databases, because almost every open source and proprietary database system has an R package that allows users to connect and interact with the it.
@@
 
+## Processing
+\label{processing}
+
+The portability of containerized environments becomes particularly useful for improving expensive processing of data or shipping complex processing pipelines.
+
+First, it is possible to **offload complex processing to a server**.
+...
+
+Second, when processes can be **run in parallel** for speeding up R code.
+
+For example, [`googleComputeEngineR`](https://CRAN.R-project.org/package=googleComputeEngineR)'s `gce_vm_cluster()` function can create clusters of 2 or more virtual machines, running multi-CPU architectures.
+Instead of running a local R script with the local CPU and RAM restrictions, the same code can be processed on all CPU threads of the cluster of machines in the cloud, all running in a Docker container with the same R environments.
+This is achieved through `googleComputeEngineR`'s integration with the R parralisation library [`library(future)` by Henrik Bengtsson]( https://CRAN.R-project.org/package=future).
+Local R computation can be thrown up to a multi-CPU and VM environment to achieve parrell computation in a few lines of R code in your local session.
+  - [some demonstrations are available here](https://cloudyr.github.io/googleComputeEngineR/articles/massive-parallel.html).
+- [Google Cloud Run](https://cloud.run) - This is a CaaS (Containers as a Service) that lets you launch a Docker container without worrying about underlying infrastructure.
+This dispenses with the developer creating a cloud server to run the Docker image on, by abstracting away those servers to a more serverless configuration.
+Cloud Run lets you run your code on top of a managed or your own Kubernetes cluster running Knative, and can accept any Docker image.
+The service takes care of network ingress, scaling machines up and down to zero, authentication and authorisation, all features which are non-trivial for a developer to create on their own.
+This can be used to scale up R code to millions of instances if they need to, with little or no changes to existing code.
+An R implementation is shown here at [cloudRunR](https://github.com/MarkEdmondson1234/cloudRunR) which uses Cloud Run to create a scalable R plumber API.
+https://code.markedmondson.me/googleCloudRunner/index.html
+
+[Google Cloud Build](https://cloud.google.com/cloud-build/) and the Google Container Registry are a continuous intergation service respectively image registry that offload building of images to the cloud, while serving the needs of commercial environments  that helps move the workload of building the Docker images to an online service, while  Cloud Build can be set up to build the Dockerfiles on each GitHub commit or release.
+This means you do not need to build the Docker images locally, which can tie up resources since Docker images can be several GBs and take a long time to compile.
+Google Cloud Build works alongside  to allow you to build private and public Docker images, which allows you to build up your own dependency graphs for downstream applications.
+  - https://code.markedmondson.me/googleCloudRunner/articles/cloudbuild.html
+
+- \CRANpkg{batchtools} \citep{Lang2017batchtools} can [schedule jobs with Docker Swarm](https://mllg.github.io/batchtools/reference/makeClusterFunctionsDocker.html)
+- scalable deployments, e.g., start with numerous Shiny talks mentioning Rocker at useR!2017
+- [dynmethods](https://github.com/dynverse/dynmethods) [@rcannood]: In order to evaluate ±50 computational methods which all used different environments (R, Python, C++, ...), we wrapped each of them in a docker container and can execute these methods from R.
+Again, all of these containers are being built on travis-ci, and will only be pushed to Docker&nbsp;Hub if the integration test succeeds.
+- \CRANpkg{drake}, https://docs.ropensci.org/drake/index.html?q=docker#with-docker
+
+In the RStudio Server Pro 1.2 release in 2019, RStudio added new functionality called [Launcher](https://solutions.rstudio.com/launcher/overview/).
+It gives users the ability to spawn R sessions and background jobs in a scalable way on external clusters, e.g., [Kubernetes based on Docker images](https://support.rstudio.com/hc/en-us/articles/360019253393-Using-Docker-images-with-RStudio-Server-Pro-Launcher-and-Kubernetes) or [Slurm](https://slurm.schedmd.com/) clusters, and optionally, with Singularity containers.
+A key benefit of Launcher is the ability for R and Python users to interactively work in RStudio without learning about Docker at all - while still leveraging containers and Kubernetes.
+Users can submit batch jobs to Kubernetes clusters without writing specific deployment scripts.
+
+\label{pipelines}
+Third, containers are perfectly suited for **packaging and executing software pipelines** and required data.
+Containers allow building complex processing pipelines that are independent from the host programming language.
+Due to its original use case (see&nbsp;\nameref{introduction}), Docker has no standard mechanisms for chaining containers together; it lacks definitions and protocols for environment variables, volume mounts and/or ports that could enable transfer of input (parameters and data) and output (results) to and from containers.
+Some packages, e.g., \pkg{containerit}, do provide Docker images that can be used similarly to CLIs, but their usage is cumbersome\footnote{\href{https://o2r.info/containerit/articles/container.html}{https://o2r.info/containerit/articles/container.html}}.
+**\pkg{outsider}** ([https://docs.ropensci.org/outsider/](https://docs.ropensci.org/outsider/)) tackles the problem of integrating external programs into an R workflow \citep{bennett_outsider_2020}.
+Installation and usage of external programs can be difficult, convoluted and even impossible if the platform is incompatible.
+Therefore \pkg{outsider} uses the platform-independent Docker images to encapsulate processes in _outsider modules_.
+Each outsider module has a `Dockerfile` and an R package with functions for interacting with the encapsulated tool.
+Using only R functions, an end-user can install a module with the \pkg{outsider} package and then call module code to integrate a tool into their own R-based workflow seamlessly.
+The \pkg{outsider} package and module manage the containers and handle the transmission of arguments and the transfer of files to and from a container.
+These functionalities also allow a user to launch module code on a remote machine via SSH, expanding the potential computational scale.
+Outsider modules can be hosted code-sharing services, e.g., on GitHub, and \pkg{outsider} contains discovery functions for them.
+<!-- informative examples of how modules work, just as a reminder for @nuest
+https://github.com/DomBennett/om..astral/blob/master/examples/astral.R
+https://github.com/DomBennett/om..raxml#partitioned-dna-analysis
+-->
+
 ## Packaging research reproducibly
@@
 New releases of even just a single piece of software in this pipeline can break the entire workflow, making it difficult to find the error and difficult for others to reuse existing pipelines.
-These breakages can make the original the results irreproducible and not ¸
+These breakages can make the original the results irreproducible and not extandable.
 The chance of a substantial disruption like this is high in a multi-year research project where key pieces of third-party software may have several major updates over the duration of the project.
@@
 This concept has been taken up and extended\footnote{See full literature list at \href{https://research-compendium.science/}{https://research-compendium.science/}.}, not the least by applying containerisation and other methods for managing computing environments---see Section&nbsp;\nameref{envs}.
-Containers give the researcher an isolated environment to assemble these research pipelines with specific versions of software to minimize problems with breaking changes and make workflows easier to share \citep[cf.][]{boettiger_introduction_2015,marwick_packaging_2018}.
+Containers give the researcher an isolated environment to assemble these research pipelines with specific versions of software.
 Research workflows in containers are safe from contamination from other activities occuring on the researcher's computer, for example the installation of newest version of packages for teaching demonstrations.
@@
 A typial compendium's `Dockerfile` will pull a rocker image fixed to a specific version of R, and install R packages from the MRAN repository to ensure the package versions are tied to a specific date, rather than the most recent version.
-A more extreme case is the _dynverse_ project ([https://dynverse.org/](https://dynverse.org/)), which packages over 50 computational methods with different environments (R, Python, C++, etc.) in Docker images, which can be executed from R.
-_dynverse_ uses a CI platform (see&nbsp;\nameref{ci}) to build Rocker-derived images, test them, and publish them on Docker&nbsp;Hub if the tests succeed.
-
 Future researchers can download the compendium from the repository and run the included `Dockerfile` to build a new image that recreates the computational environment used to produce the original research results.
 If building the image fails, the human-readable instructions in a `Dockerfile` are the starting point for rebuilding the environment.
-When combinded with CI (see&nbsp;\nameref{ci}), a research compendium set-up can enable _continuous analysis_ with easier verification of reproducibility and audits trails \citep{beaulieu-jones_reproducibility_2017}.
+Further safeguarding practices are currently under development or not part of common practice yet, such as preservation of images \citep{emsley_framework_2018} and storing them alongside `Dockerfile`s \citep[cf.][]{nust_opening_2017}, or pinning of system libraries beyond the tagged base images, which may be seen as stable or dynamic depending on the applied time scale \citep[see discussion on `debian:testing` base image in][]{RJ-2017-065}.
 
-Further safeguarding practices are currently under development or not part of common practice yet, such as preservation of images \citep{emsley_framework_2018} and storing them alongside `Dockerfile`s \citep[cf.][]{nust_opening_2017}, or pinning of system libraries beyond the tagged base images, which may be seen as stable or dynamic depending on the applied time scale \citep[see discussion on `debian:testing` base image in][]{RJ-2017-065}.
 A recommendation of the recent National Academies' report on _Reproducibility and Replicability in Science_ is that journals _"consider ways to ensure computational reproducibility for publications that make claims based on computations"_ \citep{NASEM2019}.
@@
 
+## Development, debugging, and testing
+\label{development}
+
+Containers can also serve as useful playgrounds to create environments ad-hoc or to provide very specific environments that are not needed or not easily available in day-to-day development for the purposes of developing R packages.
+These environments may have specific versions of R, of R extension packages, and of system libraries used by R extension packages, and all of the above in a specific combination.
+
+First, such containers can greatly facilitate **fixing bugs and quick evaluation**, because developers and users can readily start and later discard a container to investigate a bug report or try out a piece of software \citep[cf.][]{ooms_opencpu_2017}.
+The container does not affect their regular system.
+Using the Rocker images with RStudio, these disposable environments lack no development comfort (cf. Section&nbsp;\nameref{compendia}).
+\citet{ooms_opencpu_2017} describes how `docker exec` can be used to get a root shell in a container for cusotmization during software evaluation.
+\citet{eddelbuettel_debugging_2019} describes an example how a Docker container was used to debug an issue with a package only occuring with a particular version of Fortran, and using tools which are not readily available on all platforms (e.g., not on macOS).
+
+Second, the strong integration of **system libraries in core packages** in the [R-spatial community](https://www.r-spatial.org/) makes containers essential for stable and proactive development of common classes for geospatial data modelling and analysis.
+For example, GDAL \citep{gdal_2019} is a crucial library in the geospatial domain.
+GDAL is a system dependen^cy allowing R packages such as \CRANpkg{sf}, which provides the core data model for geospatial vector data, or \CRANpkg{rgdal}, to accomodate users to be able to read and write hundreds of different spatial raster and vector formats.
+\CRANpkg{sf} and \CRANpkg{rgdal} have hundreds of indirect reverse imports and dependencies and therefore the maintainers spend a lot of efforts not to break these.
+Purpose built Docker are used to prepare for upcoming releases of system libraries, individual bug reports, and for the lowest supported versions of system libraries\footnote{Cf.
+\href{https://github.com/r-spatial/sf/tree/master/inst/docker}{https://github.com/r-spatial/sf/tree/master/inst/docker}, \href{https://github.com/Nowosad/rspatial_proj6}{https://github.com/Nowosad/rspatial\_proj6}, and \href{https://github.com/r-spatial/sf/issues/1231}{https://github.com/r-spatial/sf/issues/1231}}.
+```{r geospatial_deps, eval=FALSE, include=FALSE}
+length(tools::dependsOnPkgs(c("rgdal", "sf"), dependencies = c("most"), recursive = TRUE))
+```
+
+Third, there are special purpose images for identifying problems beyond the mere R code, such as **debugging R memory problems**.
+The images significantly reduce the barrier to follow complex steps for fixing memory allocation bugs \citep[cf. Section~4.3 in][]{core_writing_1999}.
+These problems are hard to debug and critical, both because when they occur they lead to fatal crashing processes.
+[`rocker/r-devel-san`](https://github.com/rocker-org/r-devel-san) and [`rocker/r-devel-ubsan-clang`](https://github.com/rocker-org/r-devel-san-clang) are Docker images have a particularly configured version of R to trace such problems with gcc and clang compilers, respectively \citep[cf.~\CRANpkg{sanitizers} for examples,][]{eddelbuettel_sanitizers_2014}.
+The image [`wch/r-debug`](https://github.com/wch/r-debug) is a purpose built Docker image with _multiple_ instrumented builds of R, each with a different diagnostic utility activated.
+
+Fourth, containers are useful for **testing** R code during development.
+To submit a package to CRAN, an R package must work with the development version of R, which must be compiled locally.
+That can be a challenge for some users.
+The R-hub service (see Section&nbsp;\nameref{rhub}) makes it easy to ensure that no errors occur, but to fix errors a local setup is still often warranted, e.g., using the image `rocker/r-devel`, and to test packages with native code, which can make the process more complex \citep[cf.][]{eckert_building_2018}.
+The R-hub Docker images can also be used to debug problems locally using various combinations of Linux platforms, R versions, and compilers\footnote{See \href{https://r-hub.github.io/rhub/articles/local-debugging.html}{https://r-hub.github.io/rhub/articles/local-debugging.html} and \href{https://blog.r-hub.io/2019/04/25/r-devel-linux-x86-64-debian-clang/}{https://blog.r-hub.io/2019/04/25/r-devel-linux-x86-64-debian-clang/}}.
+The images go beyond the configurations, or _flavours_, used by CRAN for checking packages\footnote{\href{https://cran.r-project.org/web/checks/check_flavors.html}{https://cran.r-project.org/web/checks/check\_flavors.html}}, e.g., with CentOS-based images, but lack a container for checking on Windows or OS X.
+The images greatly support package developers to provide support on operating systems they are not familiar.
+The package \pkg{dockertest} ([https://github.com/traitecoevo/dockertest/](https://github.com/traitecoevo/dockertest/)) is a proof of concept for automatically generating `Dockerfile`s and building images specifically to run tests\footnote{\pkg{dockertest} is not actively maintained, but mentioned still because of its interesting approach.}.
+These images are accompanied with a special launch script so the tested source code is not stored in the image but the currently checked in version from a local Git repository is cloned into the container at runtime.
+This approach clearly seperates test environment, test code, and current working copy of the code.
+
+Fifth, Docker images can be used **on CI platforms** to streamline the testing of packages.
+\citet{ye_docker_2019} describes how they speed up the process of testing by running tasks on [Travis&nbsp;CI](https://travis-ci.org/) within a container using `docker exec`, e.g., the package check or rendering of documentation.
+\citet{cardozo_faster_2018} saved time, also on Travis&nbsp;CI, by re-using the testing image as the base for an image intended for publication on Docker&nbsp;Hub.
+[`r-ci`](https://github.com/ColinFay/r-ci) is in turn used with [GitLab&nbsp;CI](https://docs.gitlab.com/ee/ci/), which itself is built on top of Docker images: the user specifies a base Docker image, and the whole tests are run inside this environment.
+The `r-ci` image stack combines `rocker` versioning and a series of tools specifically designed for testing in a fixed environment with a customized list of preinstalled packages.
+Especially for long running tests or complex system dependencies, these approaches to seperate installation of build dependencies with code testing streamline the development process.
+Not due to a concern about time, but to control the environment used on a CI server, even this manuscript is rendered into a PDF and deployed to a GitHub-hosted website with every change (see `.travis.yml` and `Dockerfile` in the manuscript repository).
+This gives on the one hand easy access after every update of the R Markdown source code, and on the other hand a second controlled environment making sure that the article renders successfully and correctly.
+
 # Conclusions
+\label{conclusions}
 
@@
 Though for the moment, the sign of the times points to more experimentation and feature growth, e.g., images for GPU-based computing and artificial intelligence.
-Even with coding being more and more accepted as a required, and achievable skill, an easier access, for example by exposing containerisation benefits via simple user interfaces in the users' IDE, could be an important next step since currently containerisation happens more in the background for UI-based development (e.g., a `rocker/rstudio` image in the cloud).
-Furthermore, a maturing of the Rockerverse packages for managing containers may lead to their adoption where currently manual coding is required, e.g.
-in the case of \CRANpkg{RSelentium} or \CRANpkg{drake} (see Sections&nbsp;\nameref{rselenium} and \nameref{drake} respectively).
-In specific cases, e.g., for \CRANpkg{analogsea}, the interaction with the Docker daemon may remain to specific to reuse first order packages to control Docker.
+Even with coding being more and more accepted as a required, and achievable skill, an easier access, for example by exposing containerisation benefits via simple user interfaces in the users' IDE, could be an important next step.
+Currently containerisation happens more in the background at the system level.
 
@@
 
-The ordering of authors following DN and DE is alphabetical.
+The ordering of authors following DN as first author is alphabetical.
 DN conceived the article idea, \href{https://github.com/nuest/rockerverse-paper/issues/3}{initialised the formation of the writing team}, wrote sections not mentioned below, and revised all sections.
@@
 GD contributed the CARD.com use case.
-DE wrote the introduction and the section about Containerization and Rocker and reviewed all sections.
-RC \& EH contributed to the section on interfaces for Docker in R (_dynverse_ and `dynwrap`).
+DE wrote the introduction and the section about Containerization and Rocker.
+RC \& EH contributed to the section on interfaces for Docker in R.
 DC contributed content on Gigantum.
@@
 BM wrote the section on research compendia and made the project Binder-ready.
-HN \& JN co-wrote the section on the T-Mobile use case.
+HN \& JN co-wrote the section on enterprise production environments (\pkg{googleComputeEngineR}, Google Cloud Run, Kubernetes).
 KR wrote the section about \pkg{holepunch}.
@@
 
-```{r preprint_pdf, eval=FALSE, include=FALSE}
-# run this chunk, get the PDF, then revert changes in touched files
-library("rmarkdown")
-library("rticles")
-unlink('manuscript_cache', recursive = TRUE)
-
-# function is a copy of rticles::rjournal_article() but...
-# - using RJWrapper_preprint_template.tex (which includes header.tex commands)
-# - directly referencing unexported functions from rticles
-# - direclty setting the keep_tex option
-rjournal_style_preprint <- function (..., citation_package = "natbib") {
-    rmarkdown::pandoc_available("2.2", TRUE)
-    base <- rticles:::pdf_document_format("rjournal_article", highlight = NULL, 
-        citation_package = citation_package, ...)
-    base$pandoc$to <- "latex"
-    base$pandoc$ext <- ".tex"
-    base$pandoc$keep_tex <- TRUE # added
-    base$post_processor <- function(metadata, utf8_input, output_file, 
-        clean, verbose) {
-      filename <- basename(output_file)
-      if (filename != (filename2 <- gsub("_", "-", filename))) {
-          file.rename(filename, filename2)
-          filename <- filename2
-      }
-      m <- list(filename = xfun::sans_ext(filename))
-      h <- rticles:::get_list_element(metadata, c("output", "rticles::rjournal_article", 
-          "includes", "in_header"))
-      h <- c(h, if (length(preamble <- unlist(metadata[c("preamble", 
-          "header-includes")]))) {
-          f <- tempfile(fileext = ".tex")
-          on.exit(unlink(f), add = TRUE)
-          xfun::write_utf8(preamble, f)
-          f
-      })
-      #t <- find_resource("rjournal_article", "RJwrapper.tex")
-      t <- here::here("RJwrapper_preprint_template.tex")
-      rticles:::template_pandoc(m, t, "RJwrapper.tex", h, verbose)
-      tinytex::latexmk("RJwrapper.tex", base$pandoc$latex_engine, 
-          clean = clean)
-    }
-    base$knitr$opts_chunk$comment <- "#>"
-    hilight_source <- rticles:::knitr_fun("hilight_source")
-    hook_chunk = function(x, options) {
-        if (rticles:::output_asis(x, options)) 
-            return(x)
-        paste0("```{=latex}\n\\begin{Schunk}\n", x, "\\end{Schunk}\n```")
-    }
-    hook_input <- function(x, options) paste(c("\\begin{Sinput}", 
-        hilight_source(x, "sweave", options), "\\end{Sinput}", 
-        ""), collapse = "\n")
-    hook_output <- function(x, options) paste("\\begin{Soutput}\n", 
-        x, "\\end{Soutput}\n", sep = "")
-    base$knitr$knit_hooks <- rticles:::merge_list(base$knitr$knit_hooks, 
-        list(chunk = hook_chunk, source = hook_input, output = hook_output, 
-            message = hook_output, warning = hook_output, plot = knitr::hook_plot_tex))
-    base
-}
-
-rmarkdown::render("manuscript.Rmd", output_format = rjournal_style_preprint())
-file.rename("RJwrapper.pdf", "rockerverse-preprint.pdf") # file for own archival
-
-# for submission to arXiv, copy the following files into a new folder and run pdfLaTeX directly on RJwrapper.tex (works with RStudio): figures/*, manuscript.tex, RJournal.sty, RJreferences.bib, RJwrapper.tex
-# then zip the following files and submit to arXiv:
-# - /figures
-# - manuscript.tex
-# - RJournal.sty
-# - RJwrapper.bbl
-# - RJwrapper.tex
-```
-
 \bibliography{RJreferences}

### 2020-02-22 manuscript.Rmd
fix author order, add contribution statement
added CyVerse content
Added a paragraph in the # Common or public work environments lines 726-743, and added Jason Williams and Tyson Swetnam as co-authors

--- 2020-02-05 manuscript.Rmd
@@
     email: lori.shepherd@roswellpark.org
+  - name: Tyson Lee Swetnam
+    # https://github.com/tyson-swetnam
+    affiliation: University of Arizona
+    address:
+      - 1657 E Helen St.
+      - Tucson, AZ, 85721, USA
+      - \orcid{0000-0002-6639-7181}
+    email: tswetnam@arizona.edu
+  - name: Jason Williams
+    # https://github.com/jasonjwilliamsny
+    affiliation: Cold Spring Harbor Laboratory
+    address:
+      - 1 Bungtown Rd.
+      - Cold Spring Harbor, NY, 11724, USA
+      - \orcid{0000-0003-3049-2010}
+    email: williams@cshl.edu       
   - name: Nitesh Turaga
@@
 
+[CyVerse](https://cyverse.org) is an open source, NSF-funded cyberinfrastructure
+platform for the life sciences \citep{merchant2016iplant}. CyVerse supports a
+browser-based 'data science workbench' called the [Discovery Environment (DE)](https://cyverse-de.github.io/), which uses a combination of [HTCondor](https://research.cs.wisc.edu/htcondor/)
+and Kubernetes for orchestrating container based analysis, it integrates with external HPC (High Performance Computing), NSF-XSEDE through [TAPIS] (https://www.tacc.utexas.edu/tapis) API's.
+In addition, CyVerse provides a multi-petabyte Data Store based on [iRODS](https://irods.org/, this allows users to share and manage hundreds of GB's of data.
+The DE uses an [OAuth](https://oauth.net/2/) system for single sign on (SSO) for its 70+ thousand registered user.
+Rocker integration in the DE is designed to provide life science researchers with scalable,
+compute-intensive R analysis capabilities needed for large and complex datasets
+(e.g. genomics/multi-omics, GWAS, phenotypic data, spatial data etc.). The DE runs Docker
+containers on demand, with users able to integrate containers from DockerHub or
+other registries \citep{devisetty2016bringing}. These capabilities give users
+flexibility similar to Binder, but allows containers to run on larger computational resources (RAM, CPU, Disk), and for longer periods of time (days to weeks).. The
+Rocker Project's RStudio and Shiny containers are integrated into the DE by
+building new layers from official Rocker images (including a reverse proxy using
+`nginx` to handle communication with CyVerse authentication systems).
+All running instances are protected through this authentication system and users can invite other registered users to securely access the same instance.
+\citep{rstudio_proxy_2020}).
+Users can reference the same base image and add packages to further customize their R environment, the Rocker base image is cached on all nodes in the Kubernetes environment to speed up launching.
+CyVerse maintains its own [GitHub Organization](https://github.com/cyverse-vice/) with Dockerfiles and configuration scripts auto-built on
+DockerHub. For additional reproducibility, CyVerse operates a private
+Docker registry where a user's published (publicly available) containers are
+held. CyVerse user accounts are free (with some limitations for non-US users)
+and its ReadTheDocs documentation repository [CyVerse Learning Center](https://learning.cyverse.org/) allows community members to document their containers
+to support training and education.
+
 ## Packaging research reproducibly

### 2020-02-25 manuscript.Rmd

--- 2020-02-05 manuscript.Rmd
@@
     email: ehhughes@fredhutch.org
+  - name: Lars Kjeldgaard
+    # https://github.com/smaakage85
+    affiliation: Danish Tax Authorities
+    address:
+      - Oestbanegade 123
+      - 2100, Koebenhavn Oe
+    email: lars_kjeldgaard@hotmail.com
   - name: Sean Lopp
@@
 
-**\pkg{dockr}** ([https://github.com/smaakage85/dockr](https://github.com/smaakage85/dockr)) is a very similar package attempting to mirror a given R session, including local non-CRAN packages\citep{kjeldgaard_dockr_2019}.
+**\pkg{dockr}** ([https://github.com/smaakage85/dockr](https://github.com/smaakage85/dockr)) is a similar package focusing on the generation of Docker images for R packages, in which the package itself and all of the R dependencies, including local non-CRAN packages, are available \citep{kjeldgaard_dockr_2019}.
+\pkg{dockr} facilitates the organisation of code in the R package structure and the resulting Docker image mirrors the package versions of the current R session.
 Users can manually add statements for non-R dependencies to the `Dockerfile`.
@@
 NX contributed content on \pkg{liftr}.
+LK contributed content on \pkg{dockr}.
 All authors approved the final version.

### 2020-02-25 manuscript.Rmd
Fix author order
Reformat CyVerse, minor rewording

--- 2020-02-22 manuscript.Rmd
@@
     email: tswetnam@arizona.edu      
+  - name: Nitesh Turaga
+    # https://github.com/nturaga
+    affiliation: Roswell Park Comprehensive Cancer Center
+    address:
+      - Elm & Carlton Streets
+      - Buffalo, NY, 14263, USA
+      - \orcid{0000-0002-0224-9817}
+    email: nitesh.turaga@roswellpark.org
   - name: Jason Williams
@@
     email: williams@cshl.edu 
-  - name: Nitesh Turaga
-    # https://github.com/nturaga
-    affiliation: Roswell Park Comprehensive Cancer Center
-    address:
-      - Elm & Carlton Streets
-      - Buffalo, NY, 14263, USA
-      - \orcid{0000-0002-0224-9817}
-    email: nitesh.turaga@roswellpark.org
   - name: Craig Willis
@@
 
-[CyVerse](https://cyverse.org) is an open source, NSF-funded cyberinfrastructure
-platform for the life sciences \citep{merchant2016iplant}. CyVerse supports a
-browser-based 'data science workbench' called the [Discovery Environment (DE)](https://cyverse-de.github.io/), which uses a combination of [HTCondor](https://research.cs.wisc.edu/htcondor/)
-and Kubernetes for orchestrating container based analysis, it integrates with external HPC (High Performance Computing), NSF-XSEDE through [TAPIS] (https://www.tacc.utexas.edu/tapis) API's.
+[CyVerse](https://cyverse.org) is an open source, NSF-funded cyberinfrastructure platform for the life sciences \citep{merchant2016iplant}.
+CyVerse supports a browser-based 'data science workbench' called the [Discovery Environment (DE)](https://cyverse-de.github.io/), which uses a combination of [HTCondor](https://research.cs.wisc.edu/htcondor/) and Kubernetes for orchestrating container based analysis, it integrates with external HPC (High Performance Computing), NSF-XSEDE through [TAPIS] (https://www.tacc.utexas.edu/tapis) API's.
 In addition, CyVerse provides a multi-petabyte Data Store based on [iRODS](https://irods.org/, this allows users to share and manage hundreds of GB's of data.
-The DE uses an [OAuth](https://oauth.net/2/) system for single sign on (SSO) for its 70+ thousand registered user.
-Rocker integration in the DE is designed to provide life science researchers with scalable,
-compute-intensive R analysis capabilities needed for large and complex datasets
-(e.g. genomics/multi-omics, GWAS, phenotypic data, spatial data etc.). The DE runs Docker
-containers on demand, with users able to integrate containers from DockerHub or
-other registries \citep{devisetty2016bringing}. These capabilities give users
-flexibility similar to Binder, but allows containers to run on larger computational resources (RAM, CPU, Disk), and for longer periods of time (days to weeks).. The
-Rocker Project's RStudio and Shiny containers are integrated into the DE by
-building new layers from official Rocker images (including a reverse proxy using
-`nginx` to handle communication with CyVerse authentication systems).
-All running instances are protected through this authentication system and users can invite other registered users to securely access the same instance.
-\citep{rstudio_proxy_2020}).
-Users can reference the same base image and add packages to further customize their R environment, the Rocker base image is cached on all nodes in the Kubernetes environment to speed up launching.
-CyVerse maintains its own [GitHub Organization](https://github.com/cyverse-vice/) with Dockerfiles and configuration scripts auto-built on
-DockerHub. For additional reproducibility, CyVerse operates a private
-Docker registry where a user's published (publicly available) containers are
-held. CyVerse user accounts are free (with some limitations for non-US users)
-and its ReadTheDocs documentation repository [CyVerse Learning Center](https://learning.cyverse.org/) allows community members to document their containers
-to support training and education.
+The DE uses an [OAuth](https://oauth.net/2/) system for single sign on (SSO) for its 70+ thousand registered users.
+Rocker integration in the DE is designed to provide life science researchers with scalable, compute-intensive R analysis capabilities needed for large and complex datasets (e.g., genomics/multi-omics, GWAS, phenotypic data, spatial data etc.).
+The DE runs Docker containers on demand, with users able to integrate containers from DockerHub or other registries \citep{devisetty2016bringing}.
+These capabilities give users flexibility similar to Binder, but allows containers to run on larger computational resources (RAM, CPU, Disk), and for longer periods of time (days to weeks).
+The Rocker Project's RStudio and Shiny containers are integrated into the DE by deriving new images from Rocker images.
+These images include a reverse proxy using `nginx` to handle communication with CyVerse authentication systems.
+All running instances are protected through this authentication system and users can invite other registered users to securely access the same instance \citep{rstudio_proxy_2020}.
+Users can reference the same base image and add packages to further customize their R environment.
+The Rocker base image is cached on all nodes in the Kubernetes environment to speed up launching.
+CyVerse maintains its own [GitHub Organization](https://github.com/cyverse-vice/) with Dockerfiles and configuration scripts auto-built on DockerHub.
+For additional reproducibility, CyVerse operates a private Docker image registry where a user's published, i.e., publicly available, containers are held.
+CyVerse user accounts are free (with some limitations for non-US users) and the documentation repository [CyVerse Learning Center](https://learning.cyverse.org/) allows community members to document their containers to support training and education.
 

### 2020-02-26 manuscript.Rmd
fix rendering

@@
 
-[CyVerse](https://cyverse.org) is an open source, NSF-funded cyberinfrastructure platform for the life sciences \citep{merchant2016iplant}.
-CyVerse supports a browser-based 'data science workbench' called the [Discovery Environment (DE)](https://cyverse-de.github.io/), which uses a combination of [HTCondor](https://research.cs.wisc.edu/htcondor/) and Kubernetes for orchestrating container based analysis, it integrates with external HPC (High Performance Computing), NSF-XSEDE through [TAPIS] (https://www.tacc.utexas.edu/tapis) API's.
-In addition, CyVerse provides a multi-petabyte Data Store based on [iRODS](https://irods.org/, this allows users to share and manage hundreds of GB's of data.
+[CyVerse](https://cyverse.org) is an open source, NSF-funded cyberinfrastructure platform for the life sciences \citep{merchant_iplant_2016}.
+CyVerse supports a browser-based 'data science workbench' called the [Discovery Environment (DE)](https://cyverse-de.github.io/), which uses a combination of [HTCondor](https://research.cs.wisc.edu/htcondor/) and Kubernetes for orchestrating container based analysis, it integrates with external HPC (High Performance Computing), NSF-XSEDE through [TAPIS](https://www.tacc.utexas.edu/tapis) API's.
+In addition, CyVerse provides a multi-petabyte Data Store based on [iRODS](https://irods.org/), this allows users to share and manage hundreds of GB's of data.
 The DE uses an [OAuth](https://oauth.net/2/) system for single sign on (SSO) for its 70+ thousand registered users.
 Rocker integration in the DE is designed to provide life science researchers with scalable, compute-intensive R analysis capabilities needed for large and complex datasets (e.g., genomics/multi-omics, GWAS, phenotypic data, spatial data etc.).
-The DE runs Docker containers on demand, with users able to integrate containers from DockerHub or other registries \citep{devisetty2016bringing}.
+The DE runs Docker containers on demand, with users able to integrate containers from DockerHub or other registries \citep{devisetty_bringing_2016}.
 These capabilities give users flexibility similar to Binder, but allows containers to run on larger computational resources (RAM, CPU, Disk), and for longer periods of time (days to weeks).

### 2020-03-02 manuscript.Rmd

--- 2020-02-25 manuscript.Rmd
@@
 
+\label{binder}
 The [**Binder**](https://mybinder.readthedocs.io/en/latest/) project, maintained by the team behind Jupyter, makes it possible for users to **create and share computing environments** with others \citep{jupyter_binder_2018}.
@@
 In the case of shared disks or network-attached volumes, care must be taken to match user permissions, and of course backups are still neccessary.
-When working with multiple servers, an automation framework such as [Ansible](https://www.ansible.com) may be useful for managing users, permisions, and disks along with containers.
+
+[CyVerse](https://cyverse.org) is an open source, NSF-funded cyberinfrastructure platform for the life sciences providing easy access to computing and storage resources \citep{merchant_iplant_2016}.
+CyVerse's a browser-based 'data science workbench' is called [Discovery Environment](https://cyverse-de.github.io/) (DE).
+DE uses a combination of [HTCondor](https://research.cs.wisc.edu/htcondor/) and Kubernetes for orchestrating container-based analysis and integrates with external HPC, i.e., [NSF-XSEDE](https://www.xsede.org/), through [TAPIS](https://www.tacc.utexas.edu/tapis) API's.
+In addition, CyVerse provides a multi-petabyte Data Store based on [iRODS](https://irods.org/) with shared access by users.
+The DE runs Docker containers on demand, with users able to integrate bespoke containers from DockerHub or other registries \citep{devisetty_bringing_2016}.
+Rocker integration in the DE is designed to provide life science researchers with scalable, compute-intensive R analysis capabilities needed for large and complex datasets (e.g., genomics/multi-omics, GWAS, phenotypic data, spatial data etc.).
+These capabilities give users flexibility similar to Binder, but allows containers to run on larger computational resources (RAM, CPU, Disk), and for longer periods of time (days to weeks).
+The Rocker Project's RStudio and Shiny containers are integrated into the DE by deriving new images from Rocker images\footnote{See \href{https://github.com/cyverse-vice/}{https://github.com/cyverse-vice/} for \texttt{Dockerfile}s and configuration scripts; images are auto-built on DockerHub at \href{https://hub.docker.com/u/cyversevice}{https://hub.docker.com/u/cyversevice}.}.
+These images include a reverse proxy using `nginx` to handle communication with CyVerse's authentication systems, which allow users to invite other registered users to securely access the same instance \citep{rstudio_proxy_2020}.
+The images further include specific tools, such as the CLI utility `icommands` to access iRODS.
+Users can reference the same base image and add packages to further customize their R environment, or they can use CyVerse's image registry where a user's published, i.e., publicly available, containers are held for additional reproducibility.
+CyVerse user accounts are free (with some limitations for non-US users) and the [CyVerse Learning Center](https://learning.cyverse.org/) allows community members to document their containers to support training and education.
 
@@
 Right now, multiple statistics courses are developing exercises to automate the feedback for their lab classes.
-
-[CyVerse](https://cyverse.org) is an open source, NSF-funded cyberinfrastructure platform for the life sciences \citep{merchant2016iplant}.
-CyVerse supports a browser-based 'data science workbench' called the [Discovery Environment (DE)](https://cyverse-de.github.io/), which uses a combination of [HTCondor](https://research.cs.wisc.edu/htcondor/) and Kubernetes for orchestrating container based analysis, it integrates with external HPC (High Performance Computing), NSF-XSEDE through [TAPIS] (https://www.tacc.utexas.edu/tapis) API's.
-In addition, CyVerse provides a multi-petabyte Data Store based on [iRODS](https://irods.org/, this allows users to share and manage hundreds of GB's of data.
-The DE uses an [OAuth](https://oauth.net/2/) system for single sign on (SSO) for its 70+ thousand registered users.
-Rocker integration in the DE is designed to provide life science researchers with scalable, compute-intensive R analysis capabilities needed for large and complex datasets (e.g., genomics/multi-omics, GWAS, phenotypic data, spatial data etc.).
-The DE runs Docker containers on demand, with users able to integrate containers from DockerHub or other registries \citep{devisetty2016bringing}.
-These capabilities give users flexibility similar to Binder, but allows containers to run on larger computational resources (RAM, CPU, Disk), and for longer periods of time (days to weeks).
-The Rocker Project's RStudio and Shiny containers are integrated into the DE by deriving new images from Rocker images.
-These images include a reverse proxy using `nginx` to handle communication with CyVerse authentication systems.
-All running instances are protected through this authentication system and users can invite other registered users to securely access the same instance \citep{rstudio_proxy_2020}.
-Users can reference the same base image and add packages to further customize their R environment.
-The Rocker base image is cached on all nodes in the Kubernetes environment to speed up launching.
-CyVerse maintains its own [GitHub Organization](https://github.com/cyverse-vice/) with Dockerfiles and configuration scripts auto-built on DockerHub.
-For additional reproducibility, CyVerse operates a private Docker image registry where a user's published, i.e., publicly available, containers are held.
-CyVerse user accounts are free (with some limitations for non-US users) and the documentation repository [CyVerse Learning Center](https://learning.cyverse.org/) allows community members to document their containers to support training and education.
 
@@
 in the case of \CRANpkg{RSelentium} or \CRANpkg{drake} (see Sections&nbsp;\nameref{rselenium} and \nameref{drake} respectively).
-In specific cases, e.g., for \CRANpkg{analogsea}, the interaction with the Docker daemon may remain too specific to re-use first-order packages to control Docker.
+In some cases, e.g., for \CRANpkg{analogsea}, the interaction with the Docker daemon may remain too specific to re-use first-order packages to control Docker.
 

### 2020-03-02 manuscript.Rmd
Small edits, PS to authors

@@
     email: lori.shepherd@roswellpark.org
+  - name: Péter Sólymos
+    # https://github.com/psolymos
+    affiliation: Analythium Solutions
+    address:
+      - \#258 150 Chippewa Road 
+      - Sherwood Park, AB, T8A 6A2, Canada
+      - \orcid{0000-0001-7337-1740}
+    email: peter@analythium.io
   - name: Tyson Lee Swetnam
@@
 Also though, coinciding with the rise of R has been the advent of [Docker](https://en.wikipedia.org/wiki/Docker_(software)) as a general tool for distributing and deploying of server applications---in fact, Docker can be called the _lingua&nbsp;franca_ of describing computing environments and packaging software.
-Combining both these topics, the _Rocker&nbsp;Project_ ([https://www.rocker-project.org/](https://www.rocker-project.org/)) provides images with R (see the next section for more details).
+Combining both these topics, the _Rocker&nbsp;Project_ ([https://www.rocker-project.org/](https://www.rocker-project.org/)) provides Docker images with R (see the next section for more details).
 The considerable uptake and continued evolution of the Rocker&nbsp;Project has led to numerous projects that extend or build upon Rocker images, ranging from reproducible\footnote{"Reproducible" in the sense of the \emph{Claerbout/Donoho/Peng} terminology \citep{barba_terminologies_2018}.} research to production deployments.
@@
 
-The article continues with a brief introduction of containerisation basics and the Rocker&nbsp;Project, followed by use cases and applications, starting with the R packages specificly for interacting with Docker, next the second-level packages that use containers indirectly or only for specific features, and finally some complex use cases that leverage containers.
+The article continues with a brief introduction of containerisation basics and the Rocker&nbsp;Project, followed by use cases and applications, starting with the R packages specifically for interacting with Docker, next the second-level packages that use containers indirectly or only for specific features, and finally some complex use cases that leverage containers.
 We conclude with by reflecting on the landscape of packages and applications and point out future directions of development.
@@
 Docker, an application and service provided by the eponymous company, has, in just a few short years, risen to prominence for developing, testing, deploying and distributing computer software \citep[cf.][]{datadog_8_2018,munoz_history_2019}.
-While related approaches exist, such as LXC\footnote{\href{https://en.wikipedia.org/wiki/LXC}{https://en.wikipedia.org/wiki/LXC}} or Singularity \citep{kurtzer_singularity_2017}, Docker has become synomymous with "containerisation"---the method of taking software artefacts and bundling them in such a way that use becomes standardized and portable across operationg systems.
+While related approaches exist, such as LXC\footnote{\href{https://en.wikipedia.org/wiki/LXC}{https://en.wikipedia.org/wiki/LXC}} or Singularity \citep{kurtzer_singularity_2017}, Docker has become synomymous with "containerisation"---the method of taking software artefacts and bundling them in such a way that use becomes standardized and portable across operating systems.
 In doing so, Docker had recognised and validated the importance of one very important thread that had been emerging, namely virtualisation.
@@
 For statistical computing and analysis centered around R, the **Rocker&nbsp;Project** has provided a variety of Docker containers since it began in 2014 \citep{RJ-2017-065}.
-The Rocker&nbsp;Project provides several lines of containers spanning to from building blocks with `R-release` or `R-devel`, via containers with [RStudio&nbsp;Server](https://rstudio.com/products/rstudio/) and [Shiny&nbsp;Server](https://rstudio.com/products/shiny/shiny-server/), to domain-specific containers such as [`rocker/geospatial`](https://github.com/rocker-org/geospatial) \citep{rocker_geospatial_2019}.
+The Rocker&nbsp;Project provides several lines of containers spanning from building blocks with `R-release` or `R-devel`, via containers with [RStudio&nbsp;Server](https://rstudio.com/products/rstudio/) and [Shiny&nbsp;Server](https://rstudio.com/products/shiny/shiny-server/), to domain-specific containers such as [`rocker/geospatial`](https://github.com/rocker-org/geospatial) \citep{rocker_geospatial_2019}.
 These containers form _image stacks_, building on top of each other for easier maintainability (i.e., smaller `Dockerfile`s), better composability, and to reduce build time.
@@
 Each of these packages has particular advantages as they provide function wrappers for interacting with the Docker CLI at different stages of a container's life cycle.
-Examples of such interactions are installing the Docker software, creating `Dockerfile`s (\CRANpkg{dockerfiler}, \pkg{containerit}), building images, and launching a containers (\CRANpkg{stevedore}, \pkg{docker}) on a local machine or on the cloud.
+Examples of such interactions are installing the Docker software, creating `Dockerfile`s (\CRANpkg{dockerfiler}, \pkg{containerit}), building images, and launching containers (\CRANpkg{stevedore}, \pkg{docker}) on a local machine or in the cloud.
 As such, the choice of which package is most useful depends on the use case at hand as well as on the user's level of expertise.
@@
   "execute", "Execute a container locally or remotely",
-  "cloud", "Deploy or manage an instances in the cloud",
+  "cloud", "Deploy or manage instances in the cloud",
   "interact", "Interact with an instance (e.g., file transfer)",
@@
 
-**\pkg{harbor}** ([https://github.com/wch/harbor](https://github.com/wch/harbor) is no longer actively maintained, but it should be honorably mentioned as the first R package for managing Docker images and containers.
+**\pkg{harbor}** ([https://github.com/wch/harbor](https://github.com/wch/harbor)) is no longer actively maintained, but it should be honorably mentioned as the first R package for managing Docker images and containers.
 It uses the \CRANpkg{sys} package to run system commands against the Docker CLI, both locally and through an SSH connection, and it has convenience functions, e.g., for listing and removing containers/images and for accessing logs.
-The output of container executions are converted to appropriate R types.
+The outputs of container executions are converted to appropriate R types.
 The Docker CLI's basic functionality, although it evolves quickly and with little concern for avoiding breaking changes, has remained unchanged in core functions, meaning that a core function such as `harbor::docker_run(image = "hello-world")` still works depsite its stopped development.
@@
 The practice of data science often involves combining tools and software stacks and requires a cross-cutting skillset.
-This complexity and an inherent concern for openness and reproducibility in the data science community led to Docker being used widely.
+This complexity and an inherent concern for openness and reproducibility in the data science community has led to Docker being used widely.
 <!--A general description of what to consider when creating a Docker image used for data science projects in R is provided on [environments.rstudio.com](https://environments.rstudio.com/docker).-->
@@
 [r-online](https://github.com/ColinFay/ronline) leverages these images and provides an app for helping users to detect breaking changes between different R versions and for historic exploration of R.
-With a standalone NodeJS app or [online](https://srv.colinfay.me/r-online), the user can compare a piece of code run in two separate versions of R.
+With a standalone NodeJS app or [r-online](https://srv.colinfay.me/r-online), the user can compare a piece of code run in two separate versions of R.
 Internally, r-online opens one or two Docker instances with the given version of R based on Rocker images, executes a given piece of code, and returns the result to the user.
@@
 \CRANpkg{sf} and \CRANpkg{rgdal} have hundreds of indirect reverse imports and dependencies and, therefore, the maintainers spend a lot of effort trying not to break them.
-Purpose-built Docker are used to prepare for upcoming releases of system libraries, individual bug reports, and for the lowest supported versions of system libraries\footnote{Cf.
+Purpose-built Docker images are used to prepare for upcoming releases of system libraries, individual bug reports, and for the lowest supported versions of system libraries\footnote{Cf.
 \href{https://github.com/r-spatial/sf/tree/master/inst/docker}{https://github.com/r-spatial/sf/tree/master/inst/docker}, \href{https://github.com/Nowosad/rspatial_proj6}{https://github.com/Nowosad/rspatial\_proj6}, and \href{https://github.com/r-spatial/sf/issues/1231}{https://github.com/r-spatial/sf/issues/1231}}.
@@
 The package \pkg{dockertest} ([https://github.com/traitecoevo/dockertest/](https://github.com/traitecoevo/dockertest/)) is a proof of concept for automatically generating `Dockerfile`s and building images specifically to run tests\footnote{\pkg{dockertest} is not actively maintained, but mentioned still because of its interesting approach.}.
-These images are accompanied with a special launch script so the tested source code is not stored in the image; instead, the currently checked in version from a local Git repository is cloned into the container at runtime.
+These images are accompanied by a special launch script so the tested source code is not stored in the image; instead, the currently checked in version from a local Git repository is cloned into the container at runtime.
 This approach clearly seperates test environment, test code, and current working copy of the code.
@@
 First, it is possible to **offload complex processing to a server** or clouds and also to execute proccesses in parallel to speed up or to serve many users.
-**\CRANpkg{batchtools}** provides a parallel implementation of the Map function for various schedulers \citep{Lang2017batchtools}.
+**\CRANpkg{batchtools}** provides a parallel implementation of the `Map` function for various schedulers \citep{Lang2017batchtools}.
 For example, the package can [schedule jobs with Docker Swarm](https://mllg.github.io/batchtools/reference/makeClusterFunctionsDocker.html).
@@
 
-The cloud is the natural environment of containers, and, subsequently, containers are the go-to mechanism to deploy R server applications.
+The cloud is the natural environment for containers, and, therefore, containers are the go-to mechanism for deploying R server applications.
 More and more continuous integration (CI) and continuous delivery (CD) services also use containers, opening up new options for use.
@@
 R has been historically viewed as a tool for analysis and scientific research, but not for creating software that corporations can rely on for production services.
-However, thanks to advancements in R running as a web service, along with with the ability to deploy R in Docker containers, modern enterprises are now capable of having real-time machine learning powered by R.
+However, thanks to advancements in R running as a web service, along with the ability to deploy R in Docker containers, modern enterprises are now capable of having real-time machine learning powered by R.
 A number of packages and projects have enabled R to respond to client reqests over TCP/IP and local socket servers, such as \CRANpkg{Rserve}, \CRANpkg{svSocket}, [rApache](http://www.rapache.net) and more recently \CRANpkg{plumber} ([https://www.rplumber.io/](https://www.rplumber.io/)) and \pkg{RestRserve} ([http://restrserve.org](http://restrserve.org)), which even processes incoming requests in parallel with forked processes using \CRANpkg{Rserve}.
@@
 If a data scientist creates a machine learning model and exposes it through a \CRANpkg{plumber} API, then someone else on the marketing team could write software that sends different emails depending on that real-time prediction.
-The models are convolutional neural networks that use the \CRANpkg{keras} package and run in a Rocker Docker image.
+The models are convolutional neural networks that use the \CRANpkg{keras} package and run in a Rocker Docker container.
 The corresponding `Dockerfile`s are published [on GitHub](https://github.com/tmobile/r-tensorflow-api).
@@
 This is partly because `repo2docker` does not support arbitrary base images due to the complex auto-generation of the `Dockerfile` instructions.
+
 Two approaches make using Binder easier for R users.
@@
 The local image usage reduces the number of seperately managed environments and, thereby, reduces work and increases portability and reproducibility.
+
 Second, the **Whole&nbsp;Tale** project (https://wholetale.org) combines the strengths of the Rocker Project's curated Docker images with `repo2docker`.
@@
 An example for mitigating _complex setups_ is teaching databases.
-R is very useful tool for interfacing with databases, because almost every open source and proprietary database system has an R package that allows users to connect and interact with the it.
+R is very useful tool for interfacing with databases, because almost every open source and proprietary database system has an R package that allows users to connect and interact with it.
 This flexibility is even broadened by \CRANpkg{DBI}, which allows for creating a common API for interfacing these databases, or the \CRANpkg{dbplyr} package, which runs \CRANpkg{dplyr} code straight against the database as queries.
@@
 \citet{gentleman_statistical_2007} coined the term **Research Compendium** for a dynamic document together with supporting data and code.
-They used the R package system \cite{core_writing_1999} for the functional prototype all the way to structuring, validation, and distribution of research compendia.
+They used the R package system \citep{core_writing_1999} for the functional prototype all the way to structuring, validating, and distributing research compendia.
 This concept has been taken up and extended\footnote{See full literature list at \href{https://research-compendium.science/}{https://research-compendium.science/}.}, not in the least by applying containerisation and other methods for managing computing environments---see Section&nbsp;\nameref{envs}.
@@
 The R packages and use cases presented above show the growing number of users, developers, and real-world applications in the community and the resulting innovations.
-But the applications also point to the challenge of keeping up with a continuously evolving landscape.
+But the applications also point to the challenges of keeping up with a continuously evolving landscape.
 The use cases contributed by co-authors also have a degree of overlap, which can be expected as a common language and understanding of good practices is still taking shape.
@@
 Scripts added to a `Dockerfile` also add a layer of complexity and increase the risk of incomplete recipes.
-Despite the different image stacks presented here, the pervasiveness of Rocker can be traced back to its maintainers and the user community valueing collaboration and shared starting points over impulses to create individual solutions.
+Despite the different image stacks presented here, the pervasiveness of Rocker can be traced back to its maintainers and the user community valuing collaboration and shared starting points over impulses to create individual solutions.
 Aside from that, fragmentation may not be a bad sign but may instead be a reflection of a growing market that is able to sustain multiple related efforts.
@@
 It remains to be seen which approach to control Docker, via the Docker API as \pkg{stevedore} or via system calls as \pkg{dockyard}/\pkg{docker}/\pkg{dockr}, is more sustainable, or whether the question will be answered by the endurance of maintainers and sufficient funding.
-Similarly, the capturing of environments and their serialisation in form of a `Dockerfile` currently happens at different levels of abstraction, and re-use of functionality seems reasonable, e.g., \pkg{liftr} could generate the environment with \pkg{containerit}, which in turn may use \pkg{dockerfiler} for low-level R objects representing a `Dockerfile` and its instructions.
+Similarly, capturing environments and their serialisation in form of a `Dockerfile` currently is happening at different levels of abstraction, and re-use of functionality seems reasonable, e.g., \pkg{liftr} could generate the environment with \pkg{containerit}, which in turn may use \pkg{dockerfiler} for low-level R objects representing a `Dockerfile` and its instructions.
 In this consolidation, the Rocker&nbsp;Project could play the role of a coordinating entity.
@@
 Even with coding being more and more accepted as a required and achievable skill, an easier access, for example by exposing containerisation benefits via simple user interfaces in the users' IDE, could be an important next step, since currently containerisation happens more in the background for UI-based development (e.g., a `rocker/rstudio` image in the cloud).
-Furthermore, a maturing of the Rockerverse packages for managing containers may lead to them being adopted in situations where manual coding is currently required, e.g.
+Furthermore, the maturing of the Rockerverse packages for managing containers may lead to them being adopted in situations where manual coding is currently required, e.g.
 in the case of \CRANpkg{RSelentium} or \CRANpkg{drake} (see Sections&nbsp;\nameref{rselenium} and \nameref{drake} respectively).
@@
 LK contributed content on \pkg{dockr}.
+PS wrote the paragraph about OpenFaaS R templates and GitLab CI/CD pipelines with Shinyproxy.
 All authors approved the final version.

### 2020-03-02 manuscript.Rmd
GitLab/Shinyproxy learnings added

@@
 The started containers can run on a regular Docker host but also in clusters.
+Continuous integration and deployment (CI/CD) for Shiny applications using Shinyproxy on virtual machines
+can be achieved e.g. via [GitLab pipelines](https://docs.gitlab.com/ee/ci/pipelines.html). 
+A pipeline can include building and checking R packages and Shiny apps. After the code has passed
+checks, Docker images are built and pushed to GitLab's container registry
+(public or private, depending on project settings). The pipeline is finished off by sending an HTTP
+request to the virtual machine where and endpoint (webhook) is waiting to execute the
+deployment script (updating configurations, pulling new Docker images, restarting services).
+
 Another example is the package \CRANpkg{golem} ([https://github.com/ThinkR-open/golem](https://github.com/ThinkR-open/golem)), which makes an heavy use of `dockerfiler` when it comes to creating the `Dockerfile` for building and deploying production-grade Shiny applications.

### 2020-03-02 manuscript.Rmd
OpenFaaS R template paragraph added

@@
 
+Parallel efforts to enable R for different cloud providers leads to some level of code fragmentation.
+This fact and to avoid vendor lock in motivated the [OpenFaaS project](https://www.openfaas.com/), that facilitates the deployment of functions and microservices to Kubernetes or Docker Swarm.
+OpenFaaS is language agnostic, it provides auto-scaling, metrics, and an API gateway.
+Reduced boilerplate code is achieved via templates.
+R templates\footnote{OpenFaaS R templates: \href{https://github.com/analythium/openfaas-rstats-templates}{https://github.com/analythium/openfaas-rstats-templates}}
+are provided based on Rocker's and R-hub's [r-minimal](https://github.com/r-hub/r-minimal) Docker images.
+The templates use multi-stage Docker builds\footnote{\herf{https://docs.docker.com/develop/develop-images/multistage-build/}{https://docs.docker.com/develop/develop-images/multistage-build/}} to combine R base images with the OpenFaaS 'watchdog', a tiny Golang webserver that marshals an HTTP request accepted on the API Gateway and invokes the application.
+The R session use \CRANpkg{plumber} or similar packages for the API endpoint with packages and data preloaded, thus minimizing response times.
+
 The prevelance of Docker in industry naturally leads to the use of containers with R in such settings as well, as companies already manage platforms in Docker containers.
@@
 LK contributed content on \pkg{dockr}.
-PS wrote the paragraph about OpenFaaS R templates and GitLab CI/CD pipelines with Shinyproxy.
+PS wrote the paragraphs about GitLab CI/CD pipelines with Shinyproxy and OpenFaaS R templates.
 All authors approved the final version.

### 2020-03-02 manuscript.Rmd
\http issue fixed

@@
 are provided based on Rocker's and R-hub's [r-minimal](https://github.com/r-hub/r-minimal) Docker images.
-The templates use multi-stage Docker builds\footnote{\herf{https://docs.docker.com/develop/develop-images/multistage-build/}{https://docs.docker.com/develop/develop-images/multistage-build/}} to combine R base images with the OpenFaaS 'watchdog', a tiny Golang webserver that marshals an HTTP request accepted on the API Gateway and invokes the application.
+The templates use [multi-stage Docker builds](https://docs.docker.com/develop/develop-images/multistage-build/) to combine R base images with the OpenFaaS 'watchdog', a tiny Golang webserver that marshals an HTTP request accepted on the API Gateway and invokes the application.
 The R session use \CRANpkg{plumber} or similar packages for the API endpoint with packages and data preloaded, thus minimizing response times.

### 2020-03-02 manuscript.Rmd
Finsihing touches

--- 2020-03-02 manuscript.Rmd
@@
 
+Parallel efforts to enable R for different cloud providers leads to some level of code fragmentation.
+This fact and to avoid lock-in motivated the [OpenFaaS project](https://www.openfaas.com/), that facilitates the deployment of functions and microservices to Kubernetes or Docker Swarm.
+OpenFaaS is language agnostic, it provides auto-scaling, metrics, and an API gateway.
+Reduced boilerplate code is achieved via templates.
+R templates\footnote{OpenFaaS R templates: \href{https://github.com/analythium/openfaas-rstats-templates}{https://github.com/analythium/openfaas-rstats-templates}}
+are provided based on Rocker's Debian and R-hub's [r-minimal](https://github.com/r-hub/r-minimal) Alpine images.
+The templates use [multi-stage Docker builds](https://docs.docker.com/develop/develop-images/multistage-build/) to combine R base images with the OpenFaaS 'watchdog', a tiny Golang webserver that marshals an HTTP request and invokes the application.
+The R session uses \CRANpkg{plumber} or similar packages for the API endpoint with packages and data preloaded, thus minimizing response times.
+
 The prevelance of Docker in industry naturally leads to the use of containers with R in such settings as well, as companies already manage platforms in Docker containers.

### 2020-03-02 manuscript.Rmd
one sentence per line, slight shortening

@@
 The started containers can run on a regular Docker host but also in clusters.
-Continuous integration and deployment (CI/CD) for Shiny applications using Shinyproxy on virtual machines
-can be achieved e.g. via [GitLab pipelines](https://docs.gitlab.com/ee/ci/pipelines.html). 
-A pipeline can include building and checking R packages and Shiny apps. After the code has passed
-checks, Docker images are built and pushed to GitLab's container registry
-(public or private, depending on project settings). The pipeline is finished off by sending an HTTP
-request to the virtual machine where and endpoint (webhook) is waiting to execute the
-deployment script (updating configurations, pulling new Docker images, restarting services).
+Continuous integration and deployment (CI/CD) for Shiny applications using Shinyproxy can be achieved, e.g., via [GitLab pipelines](https://docs.gitlab.com/ee/ci/pipelines.html).
+A pipeline can include building and checking R packages and Shiny apps.
+After the code has passed the checks, Docker images are built and pushed to GitLab's container registry.
+The pipeline finishes with triggering a webhook on the server, where the
+deployment script is executed.
+The script updates configurations, pulls the new Docker images, and restarts the services.
 
@@
 
-Parallel efforts to enable R for different cloud providers leads to some level of code fragmentation.
-This fact and to avoid lock-in motivated the [OpenFaaS project](https://www.openfaas.com/), that facilitates the deployment of functions and microservices to Kubernetes or Docker Swarm.
-OpenFaaS is language agnostic, it provides auto-scaling, metrics, and an API gateway.
+Independent integrations of R for different cloud providers lead to repeated efforts and code fragmentation.
+To mitigate these problems and to avoid vendor lock-in motivated the [OpenFaaS project](https://www.openfaas.com/).
+OpenFaas facilitates the deployment of functions and microservices to Kubernetes or Docker Swarm.
+It is language agnostic and provides auto-scaling, metrics, and an API gateway.
 Reduced boilerplate code is achieved via templates.
-R templates\footnote{OpenFaaS R templates: \href{https://github.com/analythium/openfaas-rstats-templates}{https://github.com/analythium/openfaas-rstats-templates}}
+Template for R\footnote{See OpenFaaS R templates at \href{https://github.com/analythium/openfaas-rstats-templates}{https://github.com/analythium/openfaas-rstats-templates}.}
 are provided based on Rocker's Debian and R-hub's [r-minimal](https://github.com/r-hub/r-minimal) Alpine images.
-The templates use [multi-stage Docker builds](https://docs.docker.com/develop/develop-images/multistage-build/) to combine R base images with the OpenFaaS 'watchdog', a tiny Golang webserver that marshals an HTTP request and invokes the application.
+The templates use [multi-stage Docker builds](https://docs.docker.com/develop/develop-images/multistage-build/) to combine R base images with the OpenFaaS 'watchdog', a tiny Golang webserver.
+The watchdog marshals an HTTP request and invokes the actual application.
 The R session uses \CRANpkg{plumber} or similar packages for the API endpoint with packages and data preloaded, thus minimizing response times.

### 2020-03-02 manuscript.Rmd
closes #71 (together with previously merged PR)
update CRAN references, closes #63

@@
 **\pkg{harbor}** ([https://github.com/wch/harbor](https://github.com/wch/harbor)) is no longer actively maintained, but it should be honorably mentioned as the first R package for managing Docker images and containers.
-It uses the \CRANpkg{sys} package to run system commands against the Docker CLI, both locally and through an SSH connection, and it has convenience functions, e.g., for listing and removing containers/images and for accessing logs.
+It uses the \CRANpkg{sys} package \citep{cran_sys} to run system commands against the Docker CLI, both locally and through an SSH connection, and it has convenience functions, e.g., for listing and removing containers/images and for accessing logs.
 The outputs of container executions are converted to appropriate R types.
@@
 
-**\CRANpkg{stevedore}** ([https://cran.r-project.org/package=stevedore](https://cran.r-project.org/package=stevedore)) is currently the most powerful Docker client in R.
+**\CRANpkg{stevedore}** is currently the most powerful Docker client in R \citep{cran_stevedore}.
 It interfaces with the Docker daemon over the Docker HTTP API\footnote{\href{https://docs.docker.com/engine/api/latest/}{https://docs.docker.com/engine/api/latest/}} via a Unix socket on Linux or MacOS, over a named pipe on Windows, or over an HTTP/TCP connection.
@@
 
-The _Bioconductor_ `Dockerfile` selects the desired version of R from Rocker, adds required system dependencies, and uses the \CRANpkg{BiocManager} package for installing appropriate versions of _Bioconductor_ packages.
+The _Bioconductor_ `Dockerfile` selects the desired version of R from Rocker, adds required system dependencies, and uses the \CRANpkg{BiocManager} package for installing appropriate versions of _Bioconductor_ packages \citep{cran_biocmanager}.
 A strength of this approach is that the responsibility for complex software configuration (including customized development) is shifted from the user to the experienced _Bioconductor_ core team.
@@
 - _Kaggle_ provides the [`gcr.io/kaggle-images/rstats`](https://hub.docker.com/r/kaggle/rstats) image (previously `kaggle/rstats`) and [corresponding `Dockerfile`](https://github.com/Kaggle/docker-rstats) for usage in their Machine Learning competitions and easy access to the associated datasets.
-It includes machine learning libraries such as Tensorflow and Keras (see also image `rocker/ml` in Section&nbsp;\nameref{rocker-gpu}), and it also configures the \CRANpkg{reticulate} package.
+It includes machine learning libraries such as Tensorflow and Keras (see also image `rocker/ml` in Section&nbsp;\nameref{rocker-gpu}), and it also configures the \CRANpkg{reticulate} package \citep{cran_reticulate}.
 The image uses a base image with _all packages from CRAN_, `gcr.io/kaggle-images/rcran`, which requires a Google Cloud Build because Docker&nbsp;Hub would time out\footnote{Originally, a stacked collection of over 20~images with automated builds on Docker~Hub was used, see \href{https://web.archive.org/web/20190606043353/http://blog.kaggle.com/2016/02/05/how-to-get-started-with-data-science-in-containers/}{https://web.archive.org/web/20190606043353/http://blog.kaggle.com/2016/02/05/how-to-get-started-with-data-science-in-containers/} and \href{https://hub.docker.com/r/kaggle/rcran/dockerfile}{https://hub.docker.com/r/kaggle/rcran/dockerfile}}.
@@
 <!--As a further example, -->
-- The [_Radiant_](https://radiant-rstats.github.io/docs/) project provides several images, e.g., [`vnijs/rsm-msba-spark`](https://hub.docker.com/r/vnijs/rsm-msba-spark), for their browser-based business analytics interface based on \CRANpkg{Shiny} (`Dockerfile` [on GitHub](https://github.com/radiant-rstats/docker)) and for use in education as part of an MSc course.
+- The [_Radiant_](https://radiant-rstats.github.io/docs/) project provides several images, e.g., [`vnijs/rsm-msba-spark`](https://hub.docker.com/r/vnijs/rsm-msba-spark), for their browser-based business analytics interface based on \CRANpkg{Shiny} \citep{cran_shiny}, and for use in education as part of an MSc course\footnote{`Dockerfile` available on GitHub: \href{https://github.com/radiant-rstats/docker}{https://github.com/radiant-rstats/docker}.}.
 As data science often applies a multitude of tools, this image favours inclusion over selection and features Python, Postgres, JupyterLab and Visual Studio Code besides R and RStudio, bringing the image size up to 9&nbsp;GB.
@@
 
-**\pkg{dockerfiler}** ([https://github.com/ColinFay/dockerfiler/](https://github.com/ColinFay/dockerfiler/)) is an R package designed for building `Dockerfile`s straight from R.
+**\CRANpkg{dockerfiler}** is an R package designed for building `Dockerfile`s straight from R \citep{cran_dockerfiler}.
 A scripted creation of a `Dockerfile` enables iteration and automation, for example for packaging applications for deployment (see \nameref{deployment}).
@@
 
-**\pkg{dockr}** ([https://github.com/smaakage85/dockr](https://github.com/smaakage85/dockr)) is a similar package focusing on the generation of Docker images for R packages, in which the package itself and all of the R dependencies, including local non-CRAN packages, are available \citep{kjeldgaard_dockr_2019}.
+**\CRANpkg{dockr}** is a similar package focusing on the generation of Docker images for R packages, in which the package itself and all of the R dependencies, including local non-CRAN packages, are available \citep{cran_dockr,kjeldgaard_dockr_2019}.
 \pkg{dockr} facilitates the organisation of code in the R package structure and the resulting Docker image mirrors the package versions of the current R session.
@@
 Internally, r-online opens one or two Docker instances with the given version of R based on Rocker images, executes a given piece of code, and returns the result to the user.
-Regarding R package management, this can be achieved with MRAN, or with packages such as \CRANpkg{checkpoint} and **\CRANpkg{renv}**, which can naturally be applied within images and containers.
-For example, \CRANpkg{renv} ([https://rstudio.github.io/renv/](https://rstudio.github.io/renv/)) helps users to manage the state of the R library in a reproducible way, further providing isolation and portability \citep{renv2019}.
+Regarding R package management, this can be achieved with MRAN, or with packages such as \CRANpkg{checkpoint} \citep{cran_checkpoint} and \CRANpkg{renv} \citep{renv2019}, which can naturally be applied within images and containers.
+For example, \CRANpkg{renv} helps users to manage the state of the R library in a reproducible way, further providing isolation and portability.
 While \CRANpkg{renv} does not cover system dependencies, the \CRANpkg{renv}-based environment can be transferred into a container either by restoring the environment based on the main configuration file `renv.lock` or by storing the \CRANpkg{renv}-cache on the host and not in the container \citep{ushey_using_2019}.
@@
 For example, GDAL \citep{gdal_2019} is a crucial library in the geospatial domain.
-GDAL is a system dependency allowing R packages such as \CRANpkg{sf}, which provides the core data model for geospatial vector data, or \CRANpkg{rgdal}, to accomodate users to be able to read and write hundreds of different spatial raster and vector formats.
+GDAL is a system dependency allowing R packages such as \CRANpkg{sf}, which provides the core data model for geospatial vector data, or \CRANpkg{rgdal}, to accomodate users to be able to read and write hundreds of different spatial raster and vector formats \citep{pebesma_simple_2018,cran_rgdal}.
 \CRANpkg{sf} and \CRANpkg{rgdal} have hundreds of indirect reverse imports and dependencies and, therefore, the maintainers spend a lot of effort trying not to break them.
@@
 Especially for long-running tests or complex system dependencies, these approaches to seperate installation of build dependencies with code testing streamline the development process.
+Containers can also simplify the integration of R software into larger, multi-language CI pipelines.
 Furthermore, with each change, even this manuscript is rendered into a PDF and deployed to a GitHub-hosted website (see `.travis.yml` and `Dockerfile` in the manuscript repository), not because of concern about time, but to control the environment used on a CI server.
@@
 Instead of running a local R script with the local CPU and RAM restrictions, the same code can be processed on all CPU threads of the cluster of machines in the cloud, all running a Docker container with the same R environments.
-\CRANpkg{googleComputeEngineR} integrates with the R parallelisation package \CRANpkg{future} to enable this with only a few lines or R code\footnote{\href{https://cloudyr.github.io/googleComputeEngineR/articles/massive-parallel.html}{https://cloudyr.github.io/googleComputeEngineR/articles/massive-parallel.html}}.
+\CRANpkg{googleComputeEngineR} integrates with the R parallelisation package \CRANpkg{future} \citep{cran_future} to enable this with only a few lines or R code\footnote{\href{https://cloudyr.github.io/googleComputeEngineR/articles/massive-parallel.html}{https://cloudyr.github.io/googleComputeEngineR/articles/massive-parallel.html}}.
 [Google Cloud Run](https://cloud.run) is a CaaS (Containers as a Service) platform.
@@
 The service takes care of network ingress, scaling machines up and down, authentication, and authorisation---all features which are non-trivial for a developer to build and maintain on their own.
-This can be used to scale up R code to millions of instances if need be with little or no changes to existing code, as demonstrated by the proof of concept `cloudRunR`\footnote{\href{https://github.com/MarkEdmondson1234/cloudRunR}{https://github.com/MarkEdmondson1234/cloudRunR}}, which uses Cloud Run to create a scalable R-based API using \CRANpkg{plumber}.
+This can be used to scale up R code to millions of instances if need be with little or no changes to existing code, as demonstrated by the proof of concept `cloudRunR`\footnote{\href{https://github.com/MarkEdmondson1234/cloudRunR}{https://github.com/MarkEdmondson1234/cloudRunR}}, which uses Cloud Run to create a scalable R-based API using \CRANpkg{plumber} \citep{cran_plumber}.
 [Google Cloud Build](https://cloud.google.com/cloud-build/) and the Google Container Registry are a continuous integration service and an image registry, respectively, that offload building of images to the cloud, while serving the needs of commercial environments such as private Docker images or image stacks.
@@
 First, different packages use containers for the **deployment of R and \CRANpkg{Shiny} apps**.
-\CRANpkg{Shiny} is a popular package for creating interactive online dashboards with R, and it enables users with very diverse backgrounds to create stable and user-friendly web applications.
+\CRANpkg{Shiny} is a popular package for creating interactive online dashboards with R, and it enables users with very diverse backgrounds to create stable and user-friendly web applications \citep{cran_shiny}.
 _ShinyProxy_ ([https://www.shinyproxy.io/](https://www.shinyproxy.io/)) is an open source tool to deploy Shiny apps in an enterprise context, where it features single sign-on, but it can also be used in scientific use cases \citep[e.g., ][]{savini_epiexplorer_2019,glouzon_structurexplor_2017}.
@@
 
-Another example is the package \CRANpkg{golem} ([https://github.com/ThinkR-open/golem](https://github.com/ThinkR-open/golem)), which makes an heavy use of `dockerfiler` when it comes to creating the `Dockerfile` for building and deploying production-grade Shiny applications.
-\CRANpkg{googleComputeEngineR} ([https://cloudyr.github.io/googleComputeEngineR/](https://cloudyr.github.io/googleComputeEngineR/)) enables quick deployments of key R services, such as RStudio and Shiny, onto cloud virtual machines (VMs) with Google Cloud Compute Engine \citep{googleComputeEngineR_2019}.
+Another example is the package \CRANpkg{golem}, which makes an heavy use of `dockerfiler` when it comes to creating the `Dockerfile` for building and deploying production-grade Shiny applications \citep{cran_golem}.
+\CRANpkg{googleComputeEngineR} enables quick deployments of key R services, such as RStudio and Shiny, onto cloud virtual machines (VMs) with Google Cloud Compute Engine \citep{googleComputeEngineR_2019}.
 The package utilises `Dockerfile`s to move the labour of setting up those services from the user to a premade Docker image, which is configured and run in the cloud VM.
@@
 However, thanks to advancements in R running as a web service, along with the ability to deploy R in Docker containers, modern enterprises are now capable of having real-time machine learning powered by R.
-A number of packages and projects have enabled R to respond to client reqests over TCP/IP and local socket servers, such as \CRANpkg{Rserve}, \CRANpkg{svSocket}, [rApache](http://www.rapache.net) and more recently \CRANpkg{plumber} ([https://www.rplumber.io/](https://www.rplumber.io/)) and \pkg{RestRserve} ([http://restrserve.org](http://restrserve.org)), which even processes incoming requests in parallel with forked processes using \CRANpkg{Rserve}.
+A number of packages and projects have enabled R to respond to client reqests over TCP/IP and local socket servers, such as \CRANpkg{Rserve} \citep{cran_rserve}, \CRANpkg{svSocket} \citep{grosjean_sciviews_2019}, [rApache](http://www.rapache.net) and more recently \CRANpkg{plumber} ([https://www.rplumber.io/](https://www.rplumber.io/)) and \pkg{RestRserve} ([http://restrserve.org](http://restrserve.org)), which even processes incoming requests in parallel with forked processes using \CRANpkg{Rserve}.
 The latter two also provide documentation for deployment with Docker or ready-to-use images with automated builds\footnote{See \href{https://www.rplumber.io/docs/hosting.html\#docker}{https://www.rplumber.io/docs/hosting.html\#docker}, \href{https://hub.docker.com/r/trestletech/plumber/}{https://hub.docker.com/r/trestletech/plumber/} and \href{https://hub.docker.com/r/rexyai/restrserve/}{https://hub.docker.com/r/rexyai/restrserve/}.}.
@@
 If a data scientist creates a machine learning model and exposes it through a \CRANpkg{plumber} API, then someone else on the marketing team could write software that sends different emails depending on that real-time prediction.
-The models are convolutional neural networks that use the \CRANpkg{keras} package and run in a Rocker Docker container.
+The models are convolutional neural networks that use the \CRANpkg{keras} package \citep{cran_keras} and run in a Rocker Docker container.
 The corresponding `Dockerfile`s are published [on GitHub](https://github.com/tmobile/r-tensorflow-api).
@@
 R is very useful tool for interfacing with databases, because almost every open source and proprietary database system has an R package that allows users to connect and interact with it.
-This flexibility is even broadened by \CRANpkg{DBI}, which allows for creating a common API for interfacing these databases, or the \CRANpkg{dbplyr} package, which runs \CRANpkg{dplyr} code straight against the database as queries.
+This flexibility is even broadened by \CRANpkg{DBI} \citep{cran_dbi}, which allows for creating a common API for interfacing these databases, or the \CRANpkg{dbplyr} package \citep{cran_dbplyr}, which runs \CRANpkg{dplyr} \citep{cran_dplyr} code straight against the database as queries.
 But learning and teaching these tools comes with the cost of deploying or having access to an environment with the software and drivers installed.
@@
 It also installs the packages required for the testing framework and the exercises so that this does not have to happen every time a student's submission is evaluated.
-The actual testing of R exercises is done using a custom framework loosely based on \CRANpkg{testthat}.
+The actual testing of R exercises is done using a custom framework loosely based on \CRANpkg{testthat} \citep{wickham_testthat_2011}.
 During the development of the testing framework, it was found that the \CRANpkg{testthat} framework did not provide enough information to its reporter system to send back all the fields required by Dodona to render its feedback.
@@
 Furthermore, the maturing of the Rockerverse packages for managing containers may lead to them being adopted in situations where manual coding is currently required, e.g.
-in the case of \CRANpkg{RSelentium} or \CRANpkg{drake} (see Sections&nbsp;\nameref{rselenium} and \nameref{drake} respectively).
+in the case of \CRANpkg{RSelenium} or \CRANpkg{drake} (see Sections&nbsp;\nameref{rselenium} and \nameref{drake} respectively).
 In some cases, e.g., for \CRANpkg{analogsea}, the interaction with the Docker daemon may remain too specific to re-use first-order packages to control Docker.

### 2020-03-02 manuscript.Rmd
updat preprint footer

@@
 
-# for submission to arXiv, copy the following files into a new folder and run pdfLaTeX directly on RJwrapper.tex (works with RStudio): figures/*, manuscript.tex, RJournal.sty, RJreferences.bib, RJwrapper.tex
-# then zip the following files and submit to arXiv:
+# For submission to arXiv:
+# 1. Copy the following files into a new folder:
+# - figures/*
+# - manuscript.tex
+# - RJournal.sty
+# - RJreferences.bib
+# - RJwrapper.tex
+# 2. Run pdfLaTeX directly on RJwrapper.tex (works with RStudio)
+# 3. Zip the following files and submit to arXiv:
 # - /figures

### 2020-03-02 manuscript.Rmd
add little snippet for getting a list of GitHub handles
add author contributions for Jason Williams, Tyson Swetnam - fix incorrect link in CyVerse section

@@
       - Seattle, WA 98109-1024, USA
-      #- \orcid{}
     email: ehhughes@fredhutch.org
@@
       - Boston, MA 02210, USA
-      #- \orcid{}
     email: sean@rstudio.com
   - name: Ben Marwick
-    # https://github.com/benmarwick/
+    # https://github.com/benmarwick
     affiliation: University of Washington
@@
   - name: Heather Nolis
-    # https://github.com/nolistic/
+    # https://github.com/nolistic
     affiliation: T-Mobile
@@
       - Bellevue, WA, 98006, USA
-      #- \orcid{NA}
     email: heather.wensler1@t-mobile.com
   - name: Jacqueline Nolis
-    # https://github.com/jnolis/
+    # https://github.com/jnolis
     affiliation: Nolis, LLC
@@
 [CyVerse](https://cyverse.org) is an open source, NSF-funded cyberinfrastructure platform for the life sciences providing easy access to computing and storage resources \citep{merchant_iplant_2016}.
-CyVerse's a browser-based 'data science workbench' is called [Discovery Environment](https://cyverse-de.github.io/) (DE).
+CyVerse's a browser-based 'data science workbench' is called [Discovery Environment](https://cyverse.org/discovery-environment) (DE).
 DE uses a combination of [HTCondor](https://research.cs.wisc.edu/htcondor/) and Kubernetes for orchestrating container-based analysis and integrates with external HPC, i.e., [NSF-XSEDE](https://www.xsede.org/), through [TAPIS](https://www.tacc.utexas.edu/tapis) API's.
@@
 PS wrote the paragraphs about GitLab CI/CD pipelines with Shinyproxy and OpenFaaS R templates.
+CV wrote the section on the usage of Docker containers in Dodona.
+TS and JW wrote the section on CyVerse. 
 All authors approved the final version.
@@
 
+```{bash collab_handles, eval=FALSE, include=FALSE}
+cat manuscript.Rmd | grep '    # https://github.com/' | sed 's|    # https://github.com/|@|'
+```
+
 \bibliography{RJreferences}

### 2020-03-02 manuscript.Rmd

@@
 [CyVerse](https://cyverse.org) is an open source, NSF-funded cyberinfrastructure platform for the life sciences providing easy access to computing and storage resources \citep{merchant_iplant_2016}.
-CyVerse's a browser-based 'data science workbench' is called [Discovery Environment](https://cyverse.org/discovery-environment) (DE).
+CyVerse's a browser-based 'data science workbench' is called [Discovery Environment](https://cyverse-de.github.io/) (DE).
 DE uses a combination of [HTCondor](https://research.cs.wisc.edu/htcondor/) and Kubernetes for orchestrating container-based analysis and integrates with external HPC, i.e., [NSF-XSEDE](https://www.xsede.org/), through [TAPIS](https://www.tacc.utexas.edu/tapis) API's.
@@
 CV wrote the section on the usage of Docker containers in Dodona.
-TS and JW wrote the section on CyVerse. 
 All authors approved the final version.
@@
 We would like to thank Celeste R. Brennecka from the Science Writing Support Service of the University of Münster for her editorial support.
+LS and NT are supported by US NIH / NHGRI awards U41HG00405 and U24HG010263.
 

### 2020-03-03 manuscript.Rmd
Added funding for Karthik Ram
Fix order in acknowledgements
fix author ordering

--- 2020-03-02 manuscript.Rmd
@@
     email: nitesh.turaga@roswellpark.org
+  - name: Charlotte Van Petegem
+    # https://github.com/charvp
+    affiliation: Ghent University
+    address:
+      - Department WE02
+      - Krijgslaan 281, S9
+      - 9000 Gent, Belgium
+      - \orcid{0000-0003-0779-4897}
+    email: charlotte.vanpetegem@ugent.be
   - name: Jason Williams
@@
 CF wrote paragraphs about \pkg{r-online}, \pkg{dockerfiler}, \pkg{r-ci} and \pkg{r-db}.
+LK contributed content on \pkg{dockr}.
 SL contributed content on RStudio's usage of Docker.
@@
 LS \& NT wrote the section on Bioconductor.
+PS wrote the paragraphs about GitLab CI/CD pipelines with Shinyproxy and OpenFaaS templates.
+TS \& JW wrote the section on CyVerse.
+CvP wrote the section on the usage of Docker containers in Dodona.
 CW wrote the sections on Whole Tale and contributed the publication reproducibility audit use case.
 NX contributed content on \pkg{liftr}.
-LK contributed content on \pkg{dockr}.
-PS wrote the paragraphs about GitLab CI/CD pipelines with Shinyproxy and OpenFaaS R templates.
-CV wrote the section on the usage of Docker containers in Dodona.
-TS and JW wrote the section on CyVerse. 
 All authors approved the final version.
@@
 The funders had no role in data collection and analysis, decision to publish, or preparation of the manuscript.
+KR was supported in part  by a grant from The Leona M. and Harry B. Helmsley Charitable Trust, award number 2016PG-BRI004.
+LS and NT are supported by US NIH / NHGRI awards U41HG00405 and U24HG010263.
 CW is supported by the Whole Tale projects (https://wholetale.org) funded by the US National Science Foundation (NSF) under award [OAC-1541450](https://www.nsf.gov/awardsearch/showAward?AWD_ID=1541450).

### 2020-03-03 manuscript.Rmd
revise relation of Rocker and repo2docker in WT
Co-authored-by: Dirk Eddelbuettel <edd@debian.org>

@@
 Additionally, Whole&nbsp;Tale supports automatic data citation and is working on capabilities for image preservation and provenance capture to improve the transparency of published computational research artefacts \citep{mecum2018, mcphillips2019}.
-For R users, Whole&nbsp;Tale extends the Jupyter Project's `repo2docker` component to simplify the customisation of R-based environments for researchers with limited experience with either Docker or Git.
-Multiple options have been discussed to allow users to change the base image used in `repo2docker` from the default Ubuntu LTS (long-term support) required to support the Rocker Project images.
-Whole&nbsp;Tale implemented a custom `RockerBuildPack`\footnote{See \href{https://github.com/whole-tale/repo2docker\_wholetale}{https://github.com/whole-tale/repo2docker\_wholetale}} to support customisation of the `rocker/geospatial` image through `repo2docker` composability\footnote{Composability refers to the ability to combine multiple package managers -- such as R, `pip`, and `conda`}.
-This works in part because Rocker images are based on a Debian distribution, so the instructions created by `repo2docker` for Ubuntu work because of compatible shell and package manager.
+For R users, Whole&nbsp;Tale extends the Jupyter Project's `repo2docker` tool to simplify the customisation of R-based environments for researchers with limited experience with either Docker or Git.
+Multiple options have been discussed to allow users to change the Ubuntu LTS (long-term support, currently _Bionic Beaver_) base image, `buildpack-deps:bionic`, used in `repo2docker`.
+Whole&nbsp;Tale implemented a custom `RockerBuildPack`\footnote{See \href{https://github.com/whole-tale/repo2docker\_wholetale}{https://github.com/whole-tale/repo2docker\_wholetale}.}.
+The build pack combines a `rocker/geospatial` image with `repo2docker`'s composability\footnote{Composability refers to the ability to combine multiple package managers and their configuration files, such as R, `pip`, and `conda`; see Section~\nameref{binder} for details.}.
+This works because both Rocker images and the `repo2docker` base image use distributions with APT \citep{wikipedia_contributors_apt_2020} so that the instructions created by the latter work because of the compatible shell and package manager.
 

### 2020-03-03 manuscript.Rmd
Text improvements and corrections, spellcheck
Co-authored-by: Dirk Eddelbuettel <edd@debian.org>

@@
   The Rocker Project provides widely used Docker images for R across different application scenarios.
-  This articles surveys downstream projects that build upon Rocker images and presents the current state of R packages for managing Docker images and controlling containers.
+  This articles surveys downstream projects that build upon the Rocker Project images and presents the current state of R packages for managing Docker images and controlling containers.
   These use cases cover diverse topics such as package development, reproducible research, collaborative work, cloud-based data processing, and production deployment of services.
-  The variety of applications demonstrates the power of Rocker specifically and containerisation in general.
+  The variety of applications demonstrates the power of the Rocker Project specifically and containerisation in general.
   Across the diverse ways to use containers, we identified common themes: reproducible environments, scalability and efficiency, and portability across clouds.
@@
 These trends cement the role of R as the _lingua&nbsp;franca_ of statistics, data visualisation, and computational research.
-Also though, coinciding with the rise of R has been the advent of [Docker](https://en.wikipedia.org/wiki/Docker_(software)) as a general tool for distributing and deploying of server applications---in fact, Docker can be called the _lingua&nbsp;franca_ of describing computing environments and packaging software.
+During the last few years, and coinciding with the rise of R, has been the advent of [Docker](https://en.wikipedia.org/wiki/Docker_(software)) as a general tool for distributing and deploying of server applications---in fact, Docker can be called the _lingua&nbsp;franca_ of describing computing environments and packaging software.
 Combining both these topics, the _Rocker&nbsp;Project_ ([https://www.rocker-project.org/](https://www.rocker-project.org/)) provides Docker images with R (see the next section for more details).
 The considerable uptake and continued evolution of the Rocker&nbsp;Project has led to numerous projects that extend or build upon Rocker images, ranging from reproducible\footnote{"Reproducible" in the sense of the \emph{Claerbout/Donoho/Peng} terminology \citep{barba_terminologies_2018}.} research to production deployments.
-As such, this article presents this _Rockerverse_ of projects across all development stages: early demonstrations, working prototypes, and mature products.
+As such, this article presents what we may call the _Rockerverse_ of projects across all development stages: early demonstrations, working prototypes, and mature products.
 We also introduce related activities that connect the R language and environment with other containerisation solutions.
@@
 The article continues with a brief introduction of containerisation basics and the Rocker&nbsp;Project, followed by use cases and applications, starting with the R packages specifically for interacting with Docker, next the second-level packages that use containers indirectly or only for specific features, and finally some complex use cases that leverage containers.
-We conclude with by reflecting on the landscape of packages and applications and point out future directions of development.
+We conclude by reflecting on the landscape of packages and applications and point out future directions of development.
 
@@
 This containerisation, also called operating-system-level virtualisation \citep{wikipedia_contributors_os-level_2020},
-makes efficient use of operating system resources \citep{felter_updated_2015} and allows another order of magnitude in terms of scalability of deployment \citep[cf.][]{datadog_8_2018} because, while a container typically runs one process, virtualisation may run whole operating systems at a larger footprint.
-This makes for more efficient use of system resources \citep{felter_updated_2015} and allowed another order of magnitude in terms of scalability of deployment \citep[cf.][]{datadog_8_2018}.
+makes efficient use of operating system resources \citep{felter_updated_2015} and allows another order of magnitude in terms of scalability of deployment \citep[cf.][]{datadog_8_2018}, because virtualisation may emulate a whole operating system, a container typically runs only one process.
+The single process together with sharing the host's kernel results in a reduced footprint and faster start times.
 While Docker makes use of Linux kernel features, it has become important enough that some required aspects of running Docker have been added to other operating systems so that those systems can more efficiently support Docker \citep{microsoft_linux_2019}.
-The success of Docker has even led to standardisation and industry collaboration \citep{oci_open_2019}.
+The success of Docker has even paved the way for industry collaboration and standardisation \citep{oci_open_2019}.
 
@@
 In the latter case, a `Singularity` file, a recipe akin to a `Dockerfile`, needs to be used to make necessary adjustments.
-To date, no comparable image stack to Rocker exists on [Singularity Hub](https://singularity-hub.org/).
+To date, no comparable image stack to the Rocker Project's images exists on [Singularity Hub](https://singularity-hub.org/).
 A further tool for running containers is [**podman**](https://github.com/containers/libpod), which can also build `Dockerfile`s and run Docker images.
@@
 
-Interfacing with the Docker daemon is typically done through the [Docker Command Line Interface](https://docs.docker.com/engine/reference/commandline/cli/) (Docker CLI).
-However, moving back and forth between an R console and the command line can create friction in workflows and reduce reproducibility.
-A number of first-order R packages provide an interface to the Docker CLI, allowing for the interaction with the Docker CLI from an R console, e.g., to be automated.
-
-Each of these packages has particular advantages as they provide function wrappers for interacting with the Docker CLI at different stages of a container's life cycle.
-Examples of such interactions are installing the Docker software, creating `Dockerfile`s (\CRANpkg{dockerfiler}, \pkg{containerit}), building images, and launching containers (\CRANpkg{stevedore}, \pkg{docker}) on a local machine or in the cloud.
+Users interact with the Docker daemon typically through the [Docker Command Line Interface](https://docs.docker.com/engine/reference/commandline/cli/) (Docker CLI).
+However, moving back and forth between an R console and the command line can create friction in workflows and reduce reproducibility because of manual steps.
+A number of first-order R packages provide an interface to the Docker CLI, allowing for the interaction with the Docker CLI from an R console.
+Table&nbsp;\ref{tab:clients} gives an overview of packages with client functionality, each of which provides functions for interacting with the Docker daemon.
+The packages focus on different aspects and support different stages of a container's life cycle.
 As such, the choice of which package is most useful depends on the use case at hand as well as on the user's level of expertise.
@@
   functionalities %>% crossing(package = package_functionalities$package),
-  package_functionalities %>% unnest(id) %>% mutate(check = "\\checkmark"),#\u2714 # \\checkmark
+  package_functionalities %>% unnest(id) %>% mutate(check = "\\checkmark"),#\u2714 # \\checkmark # ✓
   by = c("id", "package")
@@
 
-# insert table
-requireNamespace("kableExtra")
-knitr::kable(tab, format = "latex", escape = FALSE) %>%
+# insert table, position caption below as described in https://github.com/haozhu233/kableExtra/issues/221#issuecomment-400473530
+library("kableExtra")
+library("xtable")
+#knitr::kable(tab,
+#             format = "latex",
+#             caption = "R packages with Docker client functionality.",
+#             label = "clients",
+#             escape = FALSE) %>%
+#  kableExtra::row_spec(0, angle = -90)
+xtable2kable <- function(x) {
+  label(x) <- "tab:clients"
+  out <- capture.output(print(x, 
+                              table.placement = NULL,
+                              booktabs = TRUE,
+                              caption.placement = "bottom",
+                              include.rownames = FALSE,
+                              hline.after = c(-1:nrow(x)),
+                              # next analoguosu to kable's "escape = FALSE", see https://stackoverflow.com/a/42355065/261210
+                              sanitize.text.function = function (x) x
+                              ))[-(1:2)]
+  out <- paste(out, collapse = "\n")
+  structure(out, format = "latex", class = "knitr_kable")
+}
+
+xtable(tab, caption = "R packages with Docker client functionality.") %>%
+  xtable2kable() %>%
   kableExtra::row_spec(0, angle = -90)
@@
 <!--A general description of what to consider when creating a Docker image used for data science projects in R is provided on [environments.rstudio.com](https://environments.rstudio.com/docker).-->
-The remaineder of this section presents exemple Docker images and image stacks featuring R intended for data science.
+The remainder of this section presents example Docker images and image stacks featuring R intended for data science.
 
@@
 Because of these specifics, instructions from `rocker/r-ubuntu` could _not_ be readily re-used in this image stack (see Section&nbsp;\nameref{conclusions}).
-Both approaches enable `apt` as an installation method, and this is exposed via the GUI-based environment management,\footnote{See \href{https://docs.gigantum.com/docs/environment-management}{https://docs.gigantum.com/docs/environment-management}}.
-The image build and publication process is scripted with Python and JSON template configuration files, unlike Rocker which relies on plain `Dockerfile`s.
+Both approaches enable the `apt` package manager \citep{wikipedia_contributors_apt_2020} as an installation method, and this is exposed via the GUI-based environment management,\footnote{See \href{https://docs.gigantum.com/docs/environment-management}{https://docs.gigantum.com/docs/environment-management}}.
+The image build and publication process is scripted with Python and JSON template configuration files, unlike Rocker images which rely on plain `Dockerfile`s.
 A further reason in the creation of an independent image stack were project constraints requiring a Rocker-incompatible licensing of the `Dockerfile`s, i.e., the MIT License.
@@
 Using the Rocker images with RStudio, these disposable environments lack no development comfort (cf. Section&nbsp;\nameref{compendia}).
-\citet{ooms_opencpu_2017} describes how `docker exec` can be used to get a root shell in a container for cusotmisation during software evaluation.
-\citet{eddelbuettel_debugging_2019} describes an example of how a Docker container was used to debug an issue with a package only occuring with a particular version of Fortran, and using tools which are not readily available on all platforms (e.g., not on macOS).
+\citet{ooms_opencpu_2017} describes how `docker exec` can be used to get a root shell in a container for customisation during software evaluation.
+\citet{eddelbuettel_debugging_2019} describes an example of how a Docker container was used to debug an issue with a package only occurring with a particular version of Fortran, and using tools which are not readily available on all platforms (e.g., not on macOS).
 
@@
 For example, GDAL \citep{gdal_2019} is a crucial library in the geospatial domain.
-GDAL is a system dependency allowing R packages such as \CRANpkg{sf}, which provides the core data model for geospatial vector data, or \CRANpkg{rgdal}, to accomodate users to be able to read and write hundreds of different spatial raster and vector formats \citep{pebesma_simple_2018,cran_rgdal}.
+GDAL is a system dependency allowing R packages such as \CRANpkg{sf}, which provides the core data model for geospatial vector data, or \CRANpkg{rgdal}, to accommodate users to be able to read and write hundreds of different spatial raster and vector formats \citep{pebesma_simple_2018,cran_rgdal}.
 \CRANpkg{sf} and \CRANpkg{rgdal} have hundreds of indirect reverse imports and dependencies and, therefore, the maintainers spend a lot of effort trying not to break them.
@@
 These images are accompanied by a special launch script so the tested source code is not stored in the image; instead, the currently checked in version from a local Git repository is cloned into the container at runtime.
-This approach clearly seperates test environment, test code, and current working copy of the code.
+This approach clearly separates test environment, test code, and current working copy of the code.
 \label{rselenium} Another use case where a container can help to standardise tests across operating systems is detailed the vignettes of the package \CRANpkg{RSelenium} \citep{rselenium_2019}.
@@
 The `r-ci` image stack combines `rocker` versioning and a series of tools specifically designed for testing in a fixed environment with a customized list of preinstalled packages.
-Especially for long-running tests or complex system dependencies, these approaches to seperate installation of build dependencies with code testing streamline the development process.
+Especially for long-running tests or complex system dependencies, these approaches to separate installation of build dependencies with code testing streamline the development process.
 Containers can also simplify the integration of R software into larger, multi-language CI pipelines.
@@
 The portability of containerised environments becomes particularly useful for improving expensive processing of data or shipping complex processing pipelines.
-First, it is possible to **offload complex processing to a server** or clouds and also to execute proccesses in parallel to speed up or to serve many users.
+First, it is possible to **offload complex processing to a server** or clouds and also to execute processes in parallel to speed up or to serve many users.
 **\CRANpkg{batchtools}** provides a parallel implementation of the `Map` function for various schedulers \citep{Lang2017batchtools}.
@@
 \label{drake} **\CRANpkg{drake}** is a workflow manager for data science projects \citep{landau_drake_2019}.
-It features implicit parallel computing and automated detection of the parts of the work that actually needs to be reexecuted.
+It features implicit parallel computing and automated detection of the parts of the work that actually needs to be re-executed.
 \CRANpkg{drake} has been demonstrated to run inside containers for high reproducibility\footnote{See for example \href{https://github.com/joelnitta/pleurosoriopsis}{https://github.com/joelnitta/pleurosoriopsis} or \href{https://gitlab.com/ecohealthalliance/drake-gitlab-docker-example}{https://gitlab.com/ecohealthalliance/drake-gitlab-docker-example}, the latter even running in a continuous integration platform (cf.~\nameref{development}.}.
-Futhermore, \CRANpkg{drake} workflwos have been shown to use \CRANpkg{future} package's function `makeClusterPSOCK()` for sending parts of the workflow to a Docker image for execution\footnote{\href{https://docs.ropensci.org/drake/index.html?q=docker\#with-docker}{https://docs.ropensci.org/drake/index.html?q=docker\#with-docker}} \citep[see package's function documentation;~][]{future_2020}.
+Furthermore, \CRANpkg{drake} workflows have been shown to use \CRANpkg{future} package's function `makeClusterPSOCK()` for sending parts of the workflow to a Docker image for execution\footnote{\href{https://docs.ropensci.org/drake/index.html?q=docker\#with-docker}{https://docs.ropensci.org/drake/index.html?q=docker\#with-docker}} \citep[see package's function documentation;~][]{future_2020}.
 In the latter case, the container control code must be written by the user, and the \CRANpkg{future} package ensures that the host and worker can connect for communicating over socket connections.
@@
 _ShinyProxy_ ([https://www.shinyproxy.io/](https://www.shinyproxy.io/)) is an open source tool to deploy Shiny apps in an enterprise context, where it features single sign-on, but it can also be used in scientific use cases \citep[e.g., ][]{savini_epiexplorer_2019,glouzon_structurexplor_2017}.
-ShinyProxy uses Docker containers to isolate user sessions and to achieve scalability for multi-user scenarious with multiple apps.
-ShinyProxy itself is written in Java to accomodate corporate requirements and may itself run in a container for stability and availability.
+ShinyProxy uses Docker containers to isolate user sessions and to achieve scalability for multi-user scenarios with multiple apps.
+ShinyProxy itself is written in Java to accommodate corporate requirements and may itself run in a container for stability and availability.
 The tool is built on ContainerProxy ([https://www.containerproxy.io/](https://www.containerproxy.io/)), which provides similar features for executing long-running R jobs or interactive R sessions.
@@
 are provided based on Rocker's Debian and R-hub's [r-minimal](https://github.com/r-hub/r-minimal) Alpine images.
-The templates use [multi-stage Docker builds](https://docs.docker.com/develop/develop-images/multistage-build/) to combine R base images with the OpenFaaS 'watchdog', a tiny Golang webserver.
+The templates use [multi-stage Docker builds](https://docs.docker.com/develop/develop-images/multistage-build/) to combine R base images with the OpenFaaS 'watchdog', a tiny Golang web server.
 The watchdog marshals an HTTP request and invokes the actual application.
@@
 
-The prevelance of Docker in industry naturally leads to the use of containers with R in such settings as well, as companies already manage platforms in Docker containers.
-These products often entail a large amount of open source software in combination with proprietary layers adding the relevant commercializable features.
+The prevalence of Docker in industry naturally leads to the use of containers with R in such settings as well, as companies already manage platforms in Docker containers.
+These products often entail a large amount of open source software in combination with proprietary layers adding the relevant commercialisation features.
 One such example is RStudio's data science platform [RStudio Team](https://rstudio.com/products/team/).
 It allows teams of data scientists and their respective IT/DevOps groups to develop and deploy code in R and Python around the RStudio Open Source Server inside of Docker images, without requiring users to learn new tools or directly interact with containers.
-The best practices for [running RStudio with Docker containers](https://support.rstudio.com/hc/en-us/articles/360021594513-Running-RStudio-with-Docker-containers) as well as [Docker images](https://github.com/rstudio/rstudio-docker-products) for RStudio's commerical products are publicly available.
+The best practices for [running RStudio with Docker containers](https://support.rstudio.com/hc/en-us/articles/360021594513-Running-RStudio-with-Docker-containers) as well as [Docker images](https://github.com/rstudio/rstudio-docker-products) for RStudio's commercial products are publicly available.
 
@@
 However, thanks to advancements in R running as a web service, along with the ability to deploy R in Docker containers, modern enterprises are now capable of having real-time machine learning powered by R.
-A number of packages and projects have enabled R to respond to client reqests over TCP/IP and local socket servers, such as \CRANpkg{Rserve} \citep{cran_rserve}, \CRANpkg{svSocket} \citep{grosjean_sciviews_2019}, [rApache](http://www.rapache.net) and more recently \CRANpkg{plumber} ([https://www.rplumber.io/](https://www.rplumber.io/)) and \pkg{RestRserve} ([http://restrserve.org](http://restrserve.org)), which even processes incoming requests in parallel with forked processes using \CRANpkg{Rserve}.
+A number of packages and projects have enabled R to respond to client requests over TCP/IP and local socket servers, such as \CRANpkg{Rserve} \citep{cran_rserve}, \CRANpkg{svSocket} \citep{grosjean_sciviews_2019}, [rApache](http://www.rapache.net) and more recently \CRANpkg{plumber} ([https://www.rplumber.io/](https://www.rplumber.io/)) and \pkg{RestRserve} ([http://restrserve.org](http://restrserve.org)), which even processes incoming requests in parallel with forked processes using \CRANpkg{Rserve}.
 The latter two also provide documentation for deployment with Docker or ready-to-use images with automated builds\footnote{See \href{https://www.rplumber.io/docs/hosting.html\#docker}{https://www.rplumber.io/docs/hosting.html\#docker}, \href{https://hub.docker.com/r/trestletech/plumber/}{https://hub.docker.com/r/trestletech/plumber/} and \href{https://hub.docker.com/r/rexyai/restrserve/}{https://hub.docker.com/r/rexyai/restrserve/}.}.
@@
 APIs based on these packages can be deployed with scalability and high availability using containers.
-This pattern of deploying code matches those used by software engineering services created in more established languages in the enterprise domain, such as Java or Python, and R can be used alongside those langages as a first-class member of a software engineering technical stack.
+This pattern of deploying code matches those used by software engineering services created in more established languages in the enterprise domain, such as Java or Python, and R can be used alongside those languages as a first-class member of a software engineering technical stack.
 
@@
 The R session behind \CRANpkg{Rserve} acted as a read-only in-memory database, which was extremely fast and scalabale, for the many concurrent rApache processes responding to the live-scoring requests of various divisions of the company.
-Similarly dockerized R scripts were responsible for the ETL processes and even the client-facing email, text message and push notification alerts sent in real-time based on card transactions.
+Similarly deodorised R scripts were responsible for the ETL processes and even the client-facing email, text message and push notification alerts sent in real-time based on card transactions.
 The related Docker images were made available at [https://github.com/cardcorp/card-rocker](https://github.com/cardcorp/card-rocker).
@@
 The applications were deployed on Amazon Elastic Container Service ([ECS](https://aws.amazon.com/ecs/)).
-The main take-aways from using R in Docker were not only that pinning the R package versions via MRAN is important, but also that moving away from Debian testing to a distribution with long-term support can be necessary.
+The main takeaways from using R in Docker were not only that pinning the R package versions via MRAN is important, but also that moving away from Debian testing to a distribution with long-term support can be necessary.
 For the use case at hand, this switch allowed for more control over upstream updates and for minimising the risk of breaking the automated builds of the Docker images and production jobs.
@@
 If a data scientist creates a machine learning model and exposes it through a \CRANpkg{plumber} API, then someone else on the marketing team could write software that sends different emails depending on that real-time prediction.
-The models are convolutional neural networks that use the \CRANpkg{keras} package \citep{cran_keras} and run in a Rocker Docker container.
+The models are convolutional neural networks that use the \CRANpkg{keras} package \citep{cran_keras} and run in a Rocker container.
 The corresponding `Dockerfile`s are published [on GitHub](https://github.com/tmobile/r-tensorflow-api).
@@
 \pkg{holepunch} removes some of these limitations by leveraging Rocker images that contain the Tidyverse along special Jupyter dependencies, and only installs additional packages from CRAN and Bioconductor that are not already part of these images.
-It short cicuits the configuration file parsing in `repo2docker` and starts with the Binder/Tidyverse base images, which eliminates a large part of the build time and, in most cases, results in a Binder instance launching within a minute.
+It short circuits the configuration file parsing in `repo2docker` and starts with the Binder/Tidyverse base images, which eliminates a large part of the build time and, in most cases, results in a Binder instance launching within a minute.
 \pkg{holepunch} also creates a `DESCRIPTION` file for essential metadata and dependency specification, and thereby turns turns any project into a research compendium (see&nbsp;\nameref{compendia}).
 The `Dockerfile` included with the project can also be used to launch an RStudio Server instance locally, i.e., independent of Binder, which is especially useful when more or special computational resources can be provided there.
-The local image usage reduces the number of seperately managed environments and, thereby, reduces work and increases portability and reproducibility.
+The local image usage reduces the number of separately managed environments and, thereby, reduces work and increases portability and reproducibility.
 
@@
 Second, software updates can be achieved by updating and redeploying the container rather than by tracking local installs on each server.
-Third, the containerised environment can be quickly deployed to other hardware, cloud or local, if more resources are neccessary or in case of server destruction or failure.
+Third, the containerised environment can be quickly deployed to other hardware, cloud or local, if more resources are necessary or in case of server destruction or failure.
 In any of these cases, users need a method to interact with the containers, be it an IDE exposed over an HTTP port or command-line access via tools such as SSH.
@@
 This may be in shared disks, attached network volumes, or in separate storage where it is uploaded between sessions.
-In the case of shared disks or network-attached volumes, care must be taken to match user permissions, and of course backups are still neccessary.
+In the case of shared disks or network-attached volumes, care must be taken to match user permissions, and of course backups are still necessary.
 
@@
 The DE runs Docker containers on demand, with users able to integrate bespoke containers from DockerHub or other registries \citep{devisetty_bringing_2016}.
-Rocker integration in the DE is designed to provide life science researchers with scalable, compute-intensive R analysis capabilities needed for large and complex datasets (e.g., genomics/multi-omics, GWAS, phenotypic data, spatial data etc.).
+Rocker images integration in the DE is designed to provide life science researchers with scalable, compute-intensive R analysis capabilities needed for large and complex datasets (e.g., genomics/multi-omics, GWAS, phenotypic data, spatial data etc.).
 These capabilities give users flexibility similar to Binder, but allows containers to run on larger computational resources (RAM, CPU, Disk), and for longer periods of time (days to weeks).
@@
 Containers give the researcher an isolated environment to assemble these research pipelines with specific versions of software to minimize problems with breaking changes and make workflows easier to share \citep[cf.][]{boettiger_introduction_2015,marwick_packaging_2018}.
-Research workflows in containers are safe from contamination from other activities that occur on the researcher's computer, for example the installation of newest version of packages for teaching demonstrations or specific versions for evaluation of other's works.
+Research workflows in containers are safe from contamination from other activities that occur on the researcher's computer, for example the installation of newest version of packages for teaching demonstrations or specific versions for evaluation of others' works.
 Given the users in this scenario, i.e., often academics with limited formal software development training, templates and assistance with containers around research compendia is essential.
 In many fields we see that a typical unit of research for a container is a research report or journal article, where the container holds the compendium, or self-contained set of data (or connections to data elsewhere) and code files needed to fully reproduce the article \citep{marwick_packaging_2018}.
-The package \pkg{rrtools} ([https://github.com/benmarwick/rrtools](https://github.com/benmarwick/rrtools)) provides a template and convienence functions to apply good practices for research compendia, including a starter `Dockerfile`.
-Images of compendium containers can be hosted on services such as Docker&nbsp;Hub for convienient sharing among collaborators and others.
+The package \pkg{rrtools} ([https://github.com/benmarwick/rrtools](https://github.com/benmarwick/rrtools)) provides a template and convenience functions to apply good practices for research compendia, including a starter `Dockerfile`.
+Images of compendium containers can be hosted on services such as Docker&nbsp;Hub for convenient sharing among collaborators and others.
 Similarly, packages such as \pkg{containerit} and \pkg{dockerfiler} can be used to manage the `Dockerfile` to be archived with a compendium on a data repository (e.g.
 [Zenodo](https://zenodo.org/), [Dataverse](https://dataverse.org/), [Figshare](https://figshare.com/), [OSF](https://osf.io/)).
-A typial compendium's `Dockerfile` will pull a rocker image fixed to a specific version of R, and install R packages from the MRAN repository to ensure the package versions are tied to a specific date, rather than the most recent version.
+A typical compendium's `Dockerfile` will pull a rocker image fixed to a specific version of R, and install R packages from the MRAN repository to ensure the package versions are tied to a specific date, rather than the most recent version.
 A more extreme case is the _dynverse_ project ([https://dynverse.org/](https://dynverse.org/)), which packages over 50 computational methods with different environments (R, Python, C++, etc.) in Docker images, which can be executed from R.
@@
 If building the image fails, the human-readable instructions in a `Dockerfile` are the starting point for rebuilding the environment.
-When combinded with CI (see&nbsp;\nameref{ci}), a research compendium set-up can enable _continuous analysis_ with easier verification of reproducibility and audits trails \citep{beaulieu-jones_reproducibility_2017}.
+When combined with CI (see&nbsp;\nameref{ci}), a research compendium set-up can enable _continuous analysis_ with easier verification of reproducibility and audits trails \citep{beaulieu-jones_reproducibility_2017}.
 
@@
 Scripts added to a `Dockerfile` also add a layer of complexity and increase the risk of incomplete recipes.
-Despite the different image stacks presented here, the pervasiveness of Rocker can be traced back to its maintainers and the user community valuing collaboration and shared starting points over impulses to create individual solutions.
+Despite the different image stacks presented here, the pervasiveness of Rocker images can be traced back to its maintainers and the user community valuing collaboration and shared starting points over impulses to create individual solutions.
 Aside from that, fragmentation may not be a bad sign but may instead be a reflection of a growing market that is able to sustain multiple related efforts.
-With the maturing of core building blocks, such as the Rocker suite of images, more systems will be built successfully, but they may also be built behind the curtains.
+With the maturing of core building blocks, such as the Rocker Project's suite of images, more systems will be built successfully, but they may also be built behind the curtains.
 Docker alone, as a flexible core technology, is not a feasible level of collaboration and abstraction.
@@
 Similarly, capturing environments and their serialisation in form of a `Dockerfile` currently is happening at different levels of abstraction, and re-use of functionality seems reasonable, e.g., \pkg{liftr} could generate the environment with \pkg{containerit}, which in turn may use \pkg{dockerfiler} for low-level R objects representing a `Dockerfile` and its instructions.
-In this consolidation, the Rocker&nbsp;Project could play the role of a coordinating entity.
-Nonetheless, for the moment, it seems that Rocker Project will be used more experimentation and feature growth, e.g., images for GPU-based computing and artificial intelligence.
+In this consolidation of R packages, the Rocker&nbsp;Project could play the role of a coordinating entity.
+Nonetheless, for the moment, it seems that the Rocker Project will focus on maintaining and extending its image stacks, e.g., images for GPU-based computing and artificial intelligence.
 Even with coding being more and more accepted as a required and achievable skill, an easier access, for example by exposing containerisation benefits via simple user interfaces in the users' IDE, could be an important next step, since currently containerisation happens more in the background for UI-based development (e.g., a `rocker/rstudio` image in the cloud).
@@
 DN conceived the article idea, \href{https://github.com/nuest/rockerverse-paper/issues/3}{initialised the formation of the writing team}, wrote sections not mentioned below, and revised all sections.
+DE wrote the introduction and the section about containerisation and the Rocker Project, and reviewed all sections.
 DB wrote the section on \pkg{outsider}.
 GD contributed the CARD.com use case.
-DE wrote the introduction and the section about containerisation and Rocker Project, and reviewed all sections.
 RC \& EH contributed to the section on interfaces for Docker in R (_dynverse_ and `dynwrap`).

### 2020-03-03 manuscript.Rmd
remove rkt, which just got stopped
https://github.com/rkt/rkt/issues/4024

@@
 It should be noted that the dominant position is that Docker is a blessing and a curse for these goals.
-It might be wise to start experimenting with non-Docker containerisation tools now, e.g., R packages interfacing with other container engines such as [podman/buildah](https://github.com/containers/libpod) or [rkt](https://coreos.com/rkt/), or an R package for creating `Singularity` files.
+It might be wise to start experimenting with non-Docker containerisation tools now, e.g., R packages interfacing with other container engines, such as [podman/buildah](https://github.com/containers/libpod), or an R package for creating `Singularity` files.
 Such efforts might help to avoid lock-in and to design sustainable workflows based on concepts of _containerisation_, not on their implementation in Docker.

### 2020-03-03 README.Rmd
rename main file to something meaningful
following submission guidelines

@@
 docker build --tag rockerverse-paper --file .binder/Dockerfile .
-docker run -i -v $(pwd):/rockerverse --user $UID rockerverse-paper Rscript -e 'setwd("/rockerverse"); rmarkdown::render("manuscript.Rmd")'
+docker run -i -v $(pwd):/rockerverse --user $UID rockerverse-paper Rscript -e 'setwd("/rockerverse"); rmarkdown::render("rockerverse.Rmd")'
 ```
@@
 ```bash
-docker run -i -v $(pwd):/rockerverse --user $UID nuest/rockerverse-paper Rscript -e 'setwd("/rockerverse"); rmarkdown::render("manuscript.Rmd")'
+docker run -i -v $(pwd):/rockerverse --user $UID nuest/rockerverse-paper Rscript -e 'setwd("/rockerverse"); rmarkdown::render("rockerverse.Rmd")'
 ```

### 2020-03-03 rockerverse.Rmd
rename main file to something meaningful
following submission guidelines

--- 2020-03-03 manuscript.Rmd
@@
 # devtools::install_github("benmarwick/wordcountaddin", type = "source", dependencies = TRUE)
-wordcountaddin::readability(here::here("manuscript.Rmd"))
-wordcountaddin::text_stats(here::here("manuscript.Rmd"))
+wordcountaddin::readability(here::here("rockerverse.Rmd"))
+wordcountaddin::text_stats(here::here("rockerverse.Rmd"))
 ```
@@
 
-rmarkdown::render("manuscript.Rmd", output_format = rjournal_style_preprint())
+rmarkdown::render("rockerverse.Rmd", output_format = rjournal_style_preprint())
 file.rename("RJwrapper.pdf", "rockerverse-preprint.pdf") # file for own archival
@@
 # - figures/*
-# - manuscript.tex
+# - rockerverse.tex
 # - RJournal.sty
@@
 # - /figures
-# - manuscript.tex
+# - rockerverse.tex
 # - RJournal.sty
@@
 ```{r collab_emails, eval=FALSE, include=FALSE}
-front_matter <- rmarkdown::yaml_front_matter("manuscript.Rmd")
+front_matter <- rmarkdown::yaml_front_matter("rockerverse.Rmd")
 emails <- sapply(X = front_matter$author, FUN = function(a) { a$email })
@@
 ```{bash collab_handles, eval=FALSE, include=FALSE}
-cat manuscript.Rmd | grep '    # https://github.com/' | sed 's|    # https://github.com/|@|'
+cat rockerverse.Rmd | grep '    # https://github.com/' | sed 's|    # https://github.com/|@|'
 ```

### 2020-03-03 rockerverse.tex
rename main file to something meaningful
following submission guidelines

% !TeX root = RJwrapper.tex
\title{The Rockerverse: Packages and Applications for Containerisation with R}
\author{by Daniel Nüst, Dirk Eddelbuettel, Dom Bennett, Robrecht Cannoodt, Dav Clark, Gergely Daroczi, Mark Edmondson, Colin Fay, Ellis Hughes, Lars Kjeldgaard, Sean Lopp, Ben Marwick, Heather Nolis, Jacqueline Nolis, Hong Ooi, Karthik Ram, Noam Ross, Lori Shepherd, Péter Sólymos, Tyson Lee Swetnam, Nitesh Turaga, Charlotte Van Petegem, Jason Williams, Craig Willis, Nan Xiao}

\maketitle

\abstract{%
The Rocker Project provides widely used Docker images for R across
different application scenarios. This articles surveys downstream
projects that build upon the Rocker Project images and presents the
current state of R packages for managing Docker images and controlling
containers. These use cases cover diverse topics such as package
development, reproducible research, collaborative work, cloud-based data
processing, and production deployment of services. The variety of
applications demonstrates the power of the Rocker Project specifically
and containerisation in general. Across the diverse ways to use
containers, we identified common themes: reproducible environments,
scalability and efficiency, and portability across clouds. We conclude
that the current growth and diversification of use cases is likely to
continue its positive impact, but see the need for consolidating the
Rockerverse ecosystem of packages, developing common practices for
applications, and exploring alternative containerisation software.
}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

\label{intro}

The R community continues to grow. This can be seen in the number of new
packages on CRAN, which is still on growing exponentially
\citep{cran:2019}, but also in the numbers of conferences, open
educational resources, meetups, unconferences, and companies that are
adopting R, as exemplified by the useR! conference
series\footnote{\href{https://www.r-project.org/conferences/}{https://www.r-project.org/conferences/}},
the global growth of the R and R-Ladies user
groups\footnote{\href{https://www.r-consortium.org/blog/2019/09/09/r-community-explorer-r-user-groups}{https://www.r-consortium.org/blog/2019/09/09/r-community-explorer-r-user-groups}, \href{https://www.r-consortium.org/blog/2019/08/12/r-community-explorer}{https://www.r-consortium.org/blog/2019/08/12/r-community-explorer}},
or the foundation and impact of the
R~Consortium\footnote{\href{https://www.r-consortium.org/news/announcements}{https://www.r-consortium.org/news/announcements}, \href{https://www.r-consortium.org/blog/2019/11/14/data-driven-tracking-and-discovery-of-r-consortium-activities}{https://www.r-consortium.org/blog/2019/11/14/data-driven-tracking-and-discovery-of-r-consortium-activities}}.
These trends cement the role of R as the \emph{lingua~franca} of
statistics, data visualisation, and computational research. During the
last few years, and coinciding with the rise of R, has been the advent
of \href{https://en.wikipedia.org/wiki/Docker_(software)}{Docker} as a
general tool for distributing and deploying of server applications---in
fact, Docker can be called the \emph{lingua~franca} of describing
computing environments and packaging software. Combining both these
topics, the \emph{Rocker~Project}
(\url{https://www.rocker-project.org/}) provides Docker images with R
(see the next section for more details). The considerable uptake and
continued evolution of the Rocker~Project has led to numerous projects
that extend or build upon Rocker images, ranging from
reproducible\footnote{"Reproducible" in the sense of the \emph{Claerbout/Donoho/Peng} terminology \citep{barba_terminologies_2018}.}
research to production deployments. As such, this article presents what
we may call the \emph{Rockerverse} of projects across all development
stages: early demonstrations, working prototypes, and mature products.
We also introduce related activities that connect the R language and
environment with other containerisation solutions. Our main
contributions is a coherent picture of the current status of using
containers in, with, and for R.

The article continues with a brief introduction of containerisation
basics and the Rocker~Project, followed by use cases and applications,
starting with the R packages specifically for interacting with Docker,
next the second-level packages that use containers indirectly or only
for specific features, and finally some complex use cases that leverage
containers. We conclude by reflecting on the landscape of packages and
applications and point out future directions of development.

\hypertarget{containerisation-and-rocker}{%
\section{Containerisation and
Rocker}\label{containerisation-and-rocker}}

\label{containerisation} \label{rocker}

Docker, an application and service provided by the eponymous company,
has, in just a few short years, risen to prominence for developing,
testing, deploying and distributing computer software
\citep[cf.][]{datadog_8_2018,munoz_history_2019}. While related
approaches exist, such as
LXC\footnote{\href{https://en.wikipedia.org/wiki/LXC}{https://en.wikipedia.org/wiki/LXC}}
or Singularity \citep{kurtzer_singularity_2017}, Docker has become
synomymous with ``containerisation''---the method of taking software
artefacts and bundling them in such a way that use becomes standardized
and portable across operating systems. In doing so, Docker had
recognised and validated the importance of one very important thread
that had been emerging, namely virtualisation. By allowing (one or
possibly) multiple applications or services to run concurrently on one
host machine without any fear of interference between them, Docker
provides an important scalability opportunity. Beyond this though,
Docker has improved this compartmentalisation by accessing the host
system---generally Linux---through a much thinner and smaller shim than
a full operating system emulation or virtualisation. This
containerisation, also called operating-system-level virtualisation
\citep{wikipedia_contributors_os-level_2020}, makes efficient use of
operating system resources \citep{felter_updated_2015} and allows
another order of magnitude in terms of scalability of deployment
\citep[cf.][]{datadog_8_2018}, because virtualisation may emulate a
whole operating system, a container typically runs only one process. The
single process together with sharing the host's kernel results in a
reduced footprint and faster start times. While Docker makes use of
Linux kernel features, it has become important enough that some required
aspects of running Docker have been added to other operating systems so
that those systems can more efficiently support Docker
\citep{microsoft_linux_2019}. The success of Docker has even paved the
way for industry collaboration and standardisation
\citep{oci_open_2019}.

The key accomplishment of Docker as an ``application'' is to make a
``bundled'' aggregation of software, the so-called ``image'', available
to any system equipped to run Docker, without requiring much else from
the host besides the actual Docker application installation. This is a
rather attractive proposition, and Docker's very easy to operate user
interface has led to widespread adoption and use of Docker in a variety
of domains, e.g., cloud computing infrastructure
\citep[e.g.,][]{Bernstein2014}, data science
\citep[e.g.,][]{boettiger_introduction_2015}, and edge computing
\citep[e.g.,][]{alam_orchestration_2018}. It has also proven to be a
natural match for ``cloud deployment'' which runs, or at least appears
to run, ``seamlessly'' without much explicit reference to the underlying
machine, architecture or operating system: Containers are portable and
can be deployed with very little dependencies on the host system---only
the container runtime is required. These Docker images are normally
built from plain text documents called \texttt{Dockerfile}s; a
\texttt{Dockerfile} has a specific set of instructions to create and
document a well-defined environment, i.e., install specific software and
expose specific ports.

For statistical computing and analysis centered around R, the
\textbf{Rocker~Project} has provided a variety of Docker containers
since it began in 2014 \citep{RJ-2017-065}. The Rocker~Project provides
several lines of containers spanning from building blocks with
\texttt{R-release} or \texttt{R-devel}, via containers with
\href{https://rstudio.com/products/rstudio/}{RStudio~Server} and
\href{https://rstudio.com/products/shiny/shiny-server/}{Shiny~Server},
to domain-specific containers such as
\href{https://github.com/rocker-org/geospatial}{\texttt{rocker/geospatial}}
\citep{rocker_geospatial_2019}. These containers form \emph{image
stacks}, building on top of each other for easier maintainability (i.e.,
smaller \texttt{Dockerfile}s), better composability, and to reduce build
time. Also of note is a series of ``versioned'' containers which match
the R release they contain with the \emph{then-current} set of packages
via the MRAN Snapshot views of CRAN \citep{microsoft_cran_2019}. The
Rocker~Project's impact and importance was acknowledged by the Chan
Zuckerberg Initiative's \emph{Essential Open Source Software for
Science}, which provides funding for the projects's sustainable
maintenance, community growth, and targeting new hardware platforms
including GPUs \citep{chan_zuckerberg_initiative_maintaining_2019}.

Docker is not the only containerisation software. An alternative
stemming from the domain of high-perfomance computing is
\textbf{Singularity} \citep{kurtzer_singularity_2017}. Singularity can
run Docker images. Rocker images work out of the box if the main process
is R, e.g., in \texttt{rocker/r-base}, but Singularity does not succeed
in running images where there is an init script, e.g., in containers
that by default run RStudio~Server. In the latter case, a
\texttt{Singularity} file, a recipe akin to a \texttt{Dockerfile}, needs
to be used to make necessary adjustments. To date, no comparable image
stack to the Rocker Project's images exists on
\href{https://singularity-hub.org/}{Singularity Hub}. A further tool for
running containers is
\href{https://github.com/containers/libpod}{\textbf{podman}}, which can
also build \texttt{Dockerfile}s and run Docker images. Proof of concepts
exists for using podman to build and run Rocker
containers\footnote{See \href{https://github.com/nuest/rodman}{https://github.com/nuest/rodman} and \href{https://github.com/rocker-org/rocker-versioned/issues/187}{https://github.com/rocker-org/rocker-versioned/issues/187}},
but the prevelance of Docker, especially in the broader user community
beyond experts or niche systems and the vast amount of blog posts and
courses for Docker currently cap specific development efforts for both
Singularity and podman in the R community. This might quickly change if
the usability and spread of Singularity or podman increase, or if
security features such as rootless/unprivileged containers, which both
these tools support out of the box, become more sought after.

\hypertarget{interfaces-for-docker-in-r}{%
\section{Interfaces for Docker in R}\label{interfaces-for-docker-in-r}}

\label{interfaces}

Users interact with the Docker daemon typically through the
\href{https://docs.docker.com/engine/reference/commandline/cli/}{Docker
Command Line Interface} (Docker CLI). However, moving back and forth
between an R console and the command line can create friction in
workflows and reduce reproducibility because of manual steps. A number
of first-order R packages provide an interface to the Docker CLI,
allowing for the interaction with the Docker CLI from an R console.
Table~\ref{tab:clients} gives an overview of packages with client
functionality, each of which provides functions for interacting with the
Docker daemon. The packages focus on different aspects and support
different stages of a container's life cycle. As such, the choice of
which package is most useful depends on the use case at hand as well as
on the user's level of expertise.

\begin{table}
\centering
\begin{tabular}{lllllll}
  \toprule
\rotatebox{-90}{Functionality} & \rotatebox{-90}{AzureContainers} & \rotatebox{-90}{babelwhale} & \rotatebox{-90}{dockermachine} & \rotatebox{-90}{dockyard} & \rotatebox{-90}{harbor} & \rotatebox{-90}{stevedore }\\ 
  \midrule
Generate a Dockerfile &  &  &  & \checkmark &  &  \\ 
   \midrule
Build an image & \checkmark &  &  & \checkmark &  &  \\ 
   \midrule
Execute a container locally or remotely & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\ 
   \midrule
Deploy or manage instances in the cloud & \checkmark &  & \checkmark &  & \checkmark & \checkmark \\ 
   \midrule
Interact with an instance (e.g., file transfer) &  & \checkmark & \checkmark &  &  & \checkmark \\ 
   \midrule
Manage storage of images &  &  &  &  & \checkmark & \checkmark \\ 
   \midrule
Supports Docker and Singularity &  & \checkmark &  &  &  &  \\ 
   \midrule
Direct access to Docker API instead of using the CLI &  &  &  &  &  & \checkmark \\ 
   \midrule
Installing Docker software &  &  & \checkmark &  &  &  \\ 
   \bottomrule
\end{tabular}
\caption{R packages with Docker client functionality.} 
\label{tab:clients}
\end{table}

\textbf{\pkg{harbor}} (\url{https://github.com/wch/harbor}) is no longer
actively maintained, but it should be honorably mentioned as the first R
package for managing Docker images and containers. It uses the
\CRANpkg{sys} package \citep{cran_sys} to run system commands against
the Docker CLI, both locally and through an SSH connection, and it has
convenience functions, e.g., for listing and removing containers/images
and for accessing logs. The outputs of container executions are
converted to appropriate R types. The Docker CLI's basic functionality,
although it evolves quickly and with little concern for avoiding
breaking changes, has remained unchanged in core functions, meaning that
a core function such as
\texttt{harbor::docker\_run(image\ =\ "hello-world")} still works
depsite its stopped development.

\textbf{\CRANpkg{stevedore}} is currently the most powerful Docker
client in R \citep{cran_stevedore}. It interfaces with the Docker daemon
over the Docker HTTP
API\footnote{\href{https://docs.docker.com/engine/api/latest/}{https://docs.docker.com/engine/api/latest/}}
via a Unix socket on Linux or MacOS, over a named pipe on Windows, or
over an HTTP/TCP connection. The package is the only one not using
system calls to the \texttt{docker} CLI tool for managing images and
containers. The package thereby enables connections to remote Docker
instances without direct configuration of the local Docker daemon.
Furthermore using the API gives access to information in a structured
way, is system independent, and is likely more reliable than parsing
command line output. \CRANpkg{stevedore}'s own interface is
automatically generated based on the OpenAPI specification of the Docker
daemon, but it is still similar to the Docker CLI. The interface is
similar to R6 objects, in that a \texttt{stevedore\_object} has a number
of functions attached to it that can be called, and multiple specific
versions of the Docker API can be supported thanks to the automatic
generation\footnote{See \href{https://github.com/richfitz/stevedore/blob/master/development.md}{https://github.com/richfitz/stevedore/blob/master/development.md}.}.

\textbf{\CRANpkg{AzureContainers}} is an interface to a number of
container-related services in Microsoft's
\href{https://azure.microsoft.com/}{Azure Cloud}
\citep{AzureContainers_2019}. While it is mainly intended for working
with Azure, as a convenience feature it includes lightweight,
cross-platform shells to Docker and Kubernetes (tools \texttt{kubectl}
and \texttt{helm}). These can be used to create and manage arbitrary
Docker images and containers, as well as Kubernetes clusters on any
platform or cloud service.

\textbf{\CRANpkg{babelwhale}} provides a unified interface to interact
with Docker and Singularity containers \citep{cannoodt_babelwhale_2019}.
Users can, for example, execute a command inside a container, mount a
volume, or copy a file with the same R commands for both container
runtimes.

\textbf{\pkg{dockyard}}
(\url{https://github.com/thebioengineer/dockyard}) has the goal of
lowering the barrier to creating \texttt{Dockerfile}s, building Docker
images, and deploying Docker containers. The package follows the
increasingly used piping paradigm of the \texttt{tidyverse} style of
programming for chaining R functions representing the instructions in a
\texttt{Dockerfile}. An existing \texttt{Dockerfile} can be used as a
template. \pkg{dockyard} also includes wrappers for common steps, such
as installing an R package or copying files, and provides built-in
functions for building an image and running a container, which make
Docker more approachable within a single R-based user interface.

\textbf{\pkg{dockermachine}}
(\url{https://github.com/cboettig/dockermachine}) is an R package to
provide a convenient interface to
\href{https://docs.docker.com/machine/overview/}{Docker~Machine} from R.
The CLI tool \texttt{docker-machine} allows users to create and manage a
virtual host on local computers, local data centers, or at cloud
providers. A local Docker installation can be configured to transparntly
forward all commands issued on the local Docker CLI to a selected
(remote) virtual host. Docker~Machine was especially crucial for local
use in the early days of Docker, when no native support was available
for Mac or Windows computers, but it remains relevant for provisioning
on remote systems. The package has not received any updates for two
years, but it is functional with a current version of
\texttt{docker-machine} (\texttt{0.16.2}). It potentially lowers the
barriers for R users to run containers on various hosts if they perceive
that using the Docker~Machine CLI directly as a barrier and it enables
scripted workflows with remote processing.

\hypertarget{use-cases-and-applications}{%
\section{Use cases and applications}\label{use-cases-and-applications}}

\label{applications}

\hypertarget{image-stacks-for-communities-of-practice}{%
\subsection{Image stacks for communities of
practice}\label{image-stacks-for-communities-of-practice}}

\textbf{Bioconductor} (\url{https://bioconductor.org/}) is an open
source, open development project for the analysis and comprehension of
genomic data \citep{gentleman_bioconductor_2004}. As of October 30th
2019, the project consists of 1823 R software packages, as well as
packages containing annotation or experiment data. \emph{Bioconductor}
has a semi-annual release cycle, where each release is associated with a
particular version of R, and Docker images are provided for current and
past versions of \emph{Bioconductor} for convenience and
reproducibility. All images, which are described on the
\emph{Bioconductor} web site (see
\url{https://bioconductor.org/help/docker/}), are created with
\texttt{Dockerfile}s maintained on GitHub and distributed through
Docker~Hub\footnote{See \href{https://github.com/Bioconductor/bioconductor_docker}{https://github.com/Bioconductor/bioconductor\_docker} and \href{https://hub.docker.com/u/bioconductor}{https://hub.docker.com/u/bioconductor} respectively.}.
\emph{Bioconductor}'s ``base'' Docker images are built on top of the
\texttt{rocker/rstudio} image. \emph{Bioconductor} installs packages
based on the R version in combination with the Bioconductor version and,
therefore, uses Bioconductor version tagging \texttt{devel} and
\texttt{RELEASE\_X\_Y}, e.g., \texttt{RELEASE\_3\_10}. Past and current
combinations of R and \emph{Bioconductor} will therefore be accessible
via a specific image tags.

The \emph{Bioconductor} \texttt{Dockerfile} selects the desired version
of R from Rocker, adds required system dependencies, and uses the
\CRANpkg{BiocManager} package for installing appropriate versions of
\emph{Bioconductor} packages \citep{cran_biocmanager}. A strength of
this approach is that the responsibility for complex software
configuration (including customized development) is shifted from the
user to the experienced \emph{Bioconductor} core team. However, a recent
audit of the \emph{Bioconductor} image stack \texttt{Dockerfile} led to
the deprecation of several community-maintained images, because the
numerous specific images became too hard to understand, complex to
maintain, and cumbersome to customise. As part of the simplification, a
recent innovation is to produce a \texttt{bioconductor\_docker:devel}
image to emulate the \emph{Bioconductor} nightly Linux build machine as
closely as possible. This image contains the build system environment
variables and the \emph{system dependencies} needed to install and check
almost all (1813 out of 1823) \emph{Bioconductor} software packages, and
it saves users and package developers from managing these themselves.
Furthermore, the image is configured so that \texttt{.libPaths()} has
\texttt{/usr/local/lib/R/host-site-library} as the first location. Users
mounting a location on the host file system to this location can
persistently manage installed packages across Docker containers or image
updates. Many R users pursue flexible work flows tailored to particular
analysis needs rather than standardized work flows. The new
\texttt{bioconductor\_docker} image is well suited for this preference,
while \texttt{bioconductor\_docker:devel} provides developers with a
test environment close to \emph{Bioconductor}'s build system.

\label{datascience} \textbf{Data science} is a widely discussed topic in
all academic disciplines \citep[e.g.,][]{donoho_50_2017}. These
discussions have shed a light on the tools and craftspersonship behind
the analysis of data with computational methods. The practice of data
science often involves combining tools and software stacks and requires
a cross-cutting skillset. This complexity and an inherent concern for
openness and reproducibility in the data science community has led to
Docker being used widely. The remainder of this section presents example
Docker images and image stacks featuring R intended for data science.

\begin{itemize}
\tightlist
\item
  The \href{https://github.com/jupyter/docker-stacks/}{\emph{Jupyter
  Docker Stacks}} project is a set of ready-to-run Docker images
  containing Jupyter applications and interactive computing tools
  \citep{project_jupyter_jupyter_2018}. The \texttt{jupyter/r-notebook}
  image includes R and ``popular packages'', and naturally also the
  IRKernel (\url{https://irkernel.github.io/}), an R kernel for Jupyter,
  so that Jupyter Notebooks can contain R code cells. R is also included
  in the catchall \texttt{jupyter/datascience-notebook}
  image\footnote{\href{https://jupyter-docker-stacks.readthedocs.io/en/latest/using/selecting.html}{https://jupyter-docker-stacks.readthedocs.io/en/latest/using/selecting.html}}.
  For example, these images allow users to quickly start a Jupyter
  Notebook server locally or build their own specialised images on top
  of stable toolsets. R is installed using the Conda package
  manager\footnote{\href{https://conda.io/}{https://conda.io/}}, which
  can manage environments for various programming languages, pinning
  both the R version and the versions of R
  packages\footnote{See \texttt{jupyter/datascience-notebook}'s \texttt{Dockerfile} at \href{https://github.com/jupyter/docker-stacks/blob/master/datascience-notebook/Dockerfile\#L47}{https://github.com/jupyter/docker-stacks/blob/master/datascience-notebook/Dockerfile\#L47}.}.
\item
  \emph{Kaggle} provides the
  \href{https://hub.docker.com/r/kaggle/rstats}{\texttt{gcr.io/kaggle-images/rstats}}
  image (previously \texttt{kaggle/rstats}) and
  \href{https://github.com/Kaggle/docker-rstats}{corresponding
  \texttt{Dockerfile}} for usage in their Machine Learning competitions
  and easy access to the associated datasets. It includes machine
  learning libraries such as Tensorflow and Keras (see also image
  \texttt{rocker/ml} in Section~\nameref{rocker-gpu}), and it also
  configures the \CRANpkg{reticulate} package \citep{cran_reticulate}.
  The image uses a base image with \emph{all packages from CRAN},
  \texttt{gcr.io/kaggle-images/rcran}, which requires a Google Cloud
  Build because Docker~Hub would time
  out\footnote{Originally, a stacked collection of over 20~images with automated builds on Docker~Hub was used, see \href{https://web.archive.org/web/20190606043353/http://blog.kaggle.com/2016/02/05/how-to-get-started-with-data-science-in-containers/}{https://web.archive.org/web/20190606043353/http://blog.kaggle.com/2016/02/05/how-to-get-started-with-data-science-in-containers/} and \href{https://hub.docker.com/r/kaggle/rcran/dockerfile}{https://hub.docker.com/r/kaggle/rcran/dockerfile}}.
  The final extracted image size is over 25~GB, which calls into
  question whether having everything available is actually convenient. 
\item
  The \href{https://radiant-rstats.github.io/docs/}{\emph{Radiant}}
  project provides several images, e.g.,
  \href{https://hub.docker.com/r/vnijs/rsm-msba-spark}{\texttt{vnijs/rsm-msba-spark}},
  for their browser-based business analytics interface based on
  \CRANpkg{Shiny} \citep{cran_shiny}, and for use in education as part
  of an MSc
  course\footnote{`Dockerfile` available on GitHub: \href{https://github.com/radiant-rstats/docker}{https://github.com/radiant-rstats/docker}.}.
  As data science often applies a multitude of tools, this image favours
  inclusion over selection and features Python, Postgres, JupyterLab and
  Visual Studio Code besides R and RStudio, bringing the image size up
  to 9~GB.
\item
  \emph{Gigantum} (\url{http://gigantum.com/}) is a platform for open
  and decentralized data science with a focus on using automation and
  user-friendly tools for easy sharing of reproducible computational
  workflows. Gigantum builds on the \emph{Gigantum Client} (running
  either locally or on a remote server) for development and execution of
  data-focused \emph{Projects}, which can be stored and shared via the
  \emph{Gigantum Hub} or via a zipfile export. The Client is a
  user-friendly interface to a backend using Docker containers to
  package, build, and run Gigantum projects. It is configured to use a
  default set of Docker base images
  (\url{https://github.com/gigantum/base-images}), and users are able to
  define and configure their own custom images. The available images
  include two with R based on Ubuntu Linux and these have the
  \href{https://launchpad.net/~marutter/+archive/ubuntu/c2d4u3.5/}{\texttt{c2d4u}
  CRAN PPA} pre-configured for installation of binary R
  packages\footnote{\href{https://docs.gigantum.com/docs/using-r}{https://docs.gigantum.com/docs/using-r}}.
  The R images vary in the included authoring environment, i.e., Jupyter
  in \texttt{r-tidyverse} or both Jupyter \& RStudio in
  \texttt{rstudio-server}. The independent image stack can be traced
  back to the Gigantum environment and its features. The R images are
  based on Gigantum's \texttt{python3-minimal} image, originally to keep
  the existing front-end configuration, but also to provide a consistent
  Python-to-R interoperability. The \texttt{Dockerfile}s also use build
  args to specify bases, for example for different versions of NVIDIA
  CUDA for GPU
  processing\footnote{See \href{https://github.com/gigantum/base-images/blob/master/_templates/python3-minimal-template/Dockerfile}{https://github.com/gigantum/base-images/blob/master/\_templates/python3-minimal-template/Dockerfile} for the \texttt{Dockerfile} of \texttt{python3-minimal}.},
  so that appropriate GPU drivers can be enabled automatically when
  supported. Furthermore, Gigantum's focus lies on environment
  management via GUI and ensuring a smooth user interaction, e.g., with
  reliable and easy conflict detection and resolution. For this reason,
  project repositories store authoritative package information in a
  separate file per package, allowing Git to directly detect conflicts
  and changes. A Dockerfile is generated from this description that
  inherits from the specified base image, and additional custom Docker
  instructions may be appended by users, though Gigantum's default base
  images do not currently include the \texttt{littler} tool, which is
  used by Rocker to install packages within \texttt{Dockerfile}s.
  Because of these specifics, instructions from \texttt{rocker/r-ubuntu}
  could \emph{not} be readily re-used in this image stack (see
  Section~\nameref{conclusions}). Both approaches enable the
  \texttt{apt} package manager \citep{wikipedia_contributors_apt_2020}
  as an installation method, and this is exposed via the GUI-based
  environment
  management,\footnote{See \href{https://docs.gigantum.com/docs/environment-management}{https://docs.gigantum.com/docs/environment-management}}.
  The image build and publication process is scripted with Python and
  JSON template configuration files, unlike Rocker images which rely on
  plain \texttt{Dockerfile}s. A further reason in the creation of an
  independent image stack were project constraints requiring a
  Rocker-incompatible licensing of the \texttt{Dockerfile}s, i.e., the
  MIT License.
\end{itemize}

\hypertarget{capture-and-create-environments}{%
\subsection{Capture and create
environments}\label{capture-and-create-environments}}

\label{envs}

Community-maintained images provide a solid basis so users can meet
their own individual requirements. Several second-order R packages
attempt to streamline the process of creating Docker images and using
containers for specific tasks, such as running tests or rendering
reproducible reports. While authoring and managing an environment with
Docker by hand is possible and feasible for
experts\footnote{See, e.g., this tutorial by RStudio on how to manage environments and package versions and to ensure deterministic image builds with Docker: \href{https://environments.rstudio.com/docker}{https://environments.rstudio.com/docker}.},
the following examples show that when environments become too cumbersome
to create manually, automation is a powerful tool. In particular, the
practice of \emph{version pinning}, with system package managers for
different operating systems and with packages \pkg{remotes} and
\pkg{versions} or by using MRAN for R, can greatly increase the
reproducibility of built images and are common approaches.

\textbf{\CRANpkg{dockerfiler}} is an R package designed for building
\texttt{Dockerfile}s straight from R \citep{cran_dockerfiler}. A
scripted creation of a \texttt{Dockerfile} enables iteration and
automation, for example for packaging applications for deployment (see
\nameref{deployment}). Developers can retrieve system requirements and
package dependencies to write a \texttt{Dockerfile}, for example, by
leveraging the tools available in R to parse a \texttt{DESCRIPTION}
file.

\textbf{\pkg{containerit}}
(\url{https://github.com/o2r-project/containerit/}) attempts to take
this one step further and includes these tools to automatically create a
\texttt{Dockerfile} that can execute a given workflow
\citep{nust_containerit_2019}. \pkg{containerit} accepts R
\texttt{sessionInfo} objects as input and provides helper functions to
derive these from workflows, e.g., an R script or R Markdown document,
by analysing the session state at the end of the workflow. It relies on
the \pkg{sysreqs} (\url{https://github.com/r-hub/sysreqs/}) package and
it's mapping of package system dependencies to platform specific
installation package
names\footnote{See \href{https://sysreqs.r-hub.io/}{https://sysreqs.r-hub.io/}.}.
\pkg{containerit} uses \CRANpkg{stevedore} to streamline the user
interaction and improve the created \texttt{Dockerfile}s, e.g., by
running a container for the desired base image to extract the already
available R packages.

\textbf{\CRANpkg{dockr}} is a similar package focusing on the generation
of Docker images for R packages, in which the package itself and all of
the R dependencies, including local non-CRAN packages, are available
\citep{cran_dockr,kjeldgaard_dockr_2019}. \pkg{dockr} facilitates the
organisation of code in the R package structure and the resulting Docker
image mirrors the package versions of the current R session. Users can
manually add statements for non-R dependencies to the
\texttt{Dockerfile}.

\textbf{\CRANpkg{liftr}} aims to solve the problem of persistent
reproducible reporting in statistical computing based on the R Markdown
format \citep{xie2018} for dynamic documents
\citep[\href{https://nanx.me/liftr/}{https://nanx.me/liftr/}, ][]{liftr2019}.
The irreproducibility of authoring environments can become an issue for
collaborative documents and large-scale platforms for processing
documents. \CRANpkg{liftr} makes the document the main and sole workflow
control file and the only file that needs to be shared between
collaborators for consistent environments, e.g.~demonstrated in the
DockFlow project (\url{https://dockflow.org}). It introduces new fields
to the R Markdown document header, allowing users to manually declare
the dependencies, including versions, for rendering the document. The
package then generates a \texttt{Dockerfile} from this metadata and
provides a utitility function to render the document inside a Docker
container, i.e., \texttt{render\_docker("foo.Rmd")}. An RStudio addin
even allows compilation of documents with the single push of a button.

System dependencies are the domain of Docker, but for a full description
of the computing environment one must also manage the R version and the
the R packages. R versions are available via the versioned Rocker image
stack \citep{RJ-2017-065}.
\href{https://github.com/ColinFay/ronline}{r-online} leverages these
images and provides an app for helping users to detect breaking changes
between different R versions and for historic exploration of R. With a
standalone NodeJS app or
\href{https://srv.colinfay.me/r-online}{r-online}, the user can compare
a piece of code run in two separate versions of R. Internally, r-online
opens one or two Docker instances with the given version of R based on
Rocker images, executes a given piece of code, and returns the result to
the user. Regarding R package management, this can be achieved with
MRAN, or with packages such as \CRANpkg{checkpoint}
\citep{cran_checkpoint} and \CRANpkg{renv} \citep{renv2019}, which can
naturally be applied within images and containers. For example,
\CRANpkg{renv} helps users to manage the state of the R library in a
reproducible way, further providing isolation and portability. While
\CRANpkg{renv} does not cover system dependencies, the
\CRANpkg{renv}-based environment can be transferred into a container
either by restoring the environment based on the main configuration file
\texttt{renv.lock} or by storing the \CRANpkg{renv}-cache on the host
and not in the container \citep{ushey_using_2019}. With both the system
dependencies and R packages consciously managed in a Docker image, users
can start using containers as the \emph{only} environment for their
workflows, which allows them to work indepenently of physical
computers\footnote{Allowing them to be digital "nomads", cf. J.~Bryan's \href{https://github.com/jennybc/docker-why}{https://github.com/jennybc/docker-why}.}
and to assert a specific degree of confidence in the stability of a
developed software
\citep[cf. \texttt{README.Rmd} in][]{marwick_research_2017}.

\hypertarget{development-debugging-and-testing}{%
\subsection{Development, debugging, and
testing}\label{development-debugging-and-testing}}

\label{development}

Containers can also serve as useful playgrounds to create environments
ad hoc or to provide very specific environments that are not needed or
not easily available in day-to-day development for the purposes of
developing R packages. These environments may have specific versions of
R, of R extension packages, and of system libraries used by R extension
packages, and all of the above in a specific combination.

First, such containers can greatly facilitate \textbf{fixing bugs and
quick evaluation}, because developers and users can readily start and
later discard a container to investigate a bug report or try out a piece
of software \citep[cf.][]{ooms_opencpu_2017}. The container does not
affect their regular system. Using the Rocker images with RStudio, these
disposable environments lack no development comfort
(cf.~Section~\nameref{compendia}). \citet{ooms_opencpu_2017} describes
how \texttt{docker\ exec} can be used to get a root shell in a container
for customisation during software evaluation.
\citet{eddelbuettel_debugging_2019} describes an example of how a Docker
container was used to debug an issue with a package only occurring with
a particular version of Fortran, and using tools which are not readily
available on all platforms (e.g., not on macOS).

Second, the strong integration of \textbf{system libraries in core
packages} in the \href{https://www.r-spatial.org/}{R-spatial community}
makes containers essential for stable and proactive development of
common classes for geospatial data modelling and analysis. For example,
GDAL \citep{gdal_2019} is a crucial library in the geospatial domain.
GDAL is a system dependency allowing R packages such as \CRANpkg{sf},
which provides the core data model for geospatial vector data, or
\CRANpkg{rgdal}, to accommodate users to be able to read and write
hundreds of different spatial raster and vector formats
\citep{pebesma_simple_2018,cran_rgdal}. \CRANpkg{sf} and \CRANpkg{rgdal}
have hundreds of indirect reverse imports and dependencies and,
therefore, the maintainers spend a lot of effort trying not to break
them. Purpose-built Docker images are used to prepare for upcoming
releases of system libraries, individual bug reports, and for the lowest
supported versions of system
libraries\footnote{Cf. \href{https://github.com/r-spatial/sf/tree/master/inst/docker}{https://github.com/r-spatial/sf/tree/master/inst/docker}, \href{https://github.com/Nowosad/rspatial_proj6}{https://github.com/Nowosad/rspatial\_proj6}, and \href{https://github.com/r-spatial/sf/issues/1231}{https://github.com/r-spatial/sf/issues/1231}}.

Third, special-purpose images exist for identifying problems beyond the
mere R code, such as \textbf{debugging R memory problems}. These images
significantly reduce the barriers to following complex steps for fixing
memory allocation bugs \citep[cf. Section~4.3 in][]{core_writing_1999}.
These problems are hard to debug and critical, because when they do
occur they lead to fatal crashing processes.
\href{https://github.com/rocker-org/r-devel-san}{\texttt{rocker/r-devel-san}}
and
\href{https://github.com/rocker-org/r-devel-san-clang}{\texttt{rocker/r-devel-ubsan-clang}}
are Docker images that have a particularly configured version of R to
trace such problems with gcc and clang compilers, respectively
\citep[cf.~\CRANpkg{sanitizers} for examples,][]{eddelbuettel_sanitizers_2014}.
\href{https://github.com/wch/r-debug}{\texttt{wch/r-debug}} is a
purpose-built Docker image with \emph{multiple} instrumented builds of
R, each with a different diagnostic utility activated.

Fourth, containers are useful for \textbf{testing} R code during
development. To submit a package to CRAN, an R package must work with
the development version of R, which must be compiled locally; this can
be a challenge for some users. The \textbf{R-hub} project provides
\emph{``a collection of services to help R package development''}, with
the package builder as the most prominent one \citep{r-hub_docs_2019}.
R-hub makes it easy to ensure that no errors occur, but fixing errors
still often warrants a local setup, e.g., using the image
\texttt{rocker/r-devel}, as is testing packages with native code, which
can make the process more complex \citep[cf.][]{eckert_building_2018}.
The R-hub Docker images can also be used to debug problems locally using
various combinations of Linux platforms, R versions, and
compilers\footnote{See \href{https://r-hub.github.io/rhub/articles/local-debugging.html}{https://r-hub.github.io/rhub/articles/local-debugging.html} and \href{https://blog.r-hub.io/2019/04/25/r-devel-linux-x86-64-debian-clang/}{https://blog.r-hub.io/2019/04/25/r-devel-linux-x86-64-debian-clang/}}.
The images go beyond the configurations, or \emph{flavours}, used by
CRAN for checking
packages\footnote{\href{https://cran.r-project.org/web/checks/check_flavors.html}{https://cran.r-project.org/web/checks/check\_flavors.html}},
e.g., with CentOS-based images, but they lack a container for checking
on Windows or OS X. The images greatly support package developers to
provide support on operating systems with which they are not familiar.
The package \pkg{dockertest}
(\url{https://github.com/traitecoevo/dockertest/}) is a proof of concept
for automatically generating \texttt{Dockerfile}s and building images
specifically to run
tests\footnote{\pkg{dockertest} is not actively maintained, but mentioned still because of its interesting approach.}.
These images are accompanied by a special launch script so the tested
source code is not stored in the image; instead, the currently checked
in version from a local Git repository is cloned into the container at
runtime. This approach clearly separates test environment, test code,
and current working copy of the code. \label{rselenium} Another use case
where a container can help to standardise tests across operating systems
is detailed the vignettes of the package \CRANpkg{RSelenium}
\citep{rselenium_2019}. The package recommends Docker for running the
\href{https://selenium.dev/}{Selenium} Server application needed to
execute test suites on browser-based user interfaces and webpages, but
it requires users to manually manage the containers.

\label{ci} Fifth, Docker images can be used \textbf{on continuous
integration (CI) platforms} to streamline the testing of packages.
\citet{ye_docker_2019} describes how they speed up the process of
testing by running tasks on \href{https://travis-ci.org/}{Travis~CI}
within a container using \texttt{docker\ exec}, e.g., the package check
or rendering of documentation. \citet{cardozo_faster_2018} also saved
time with Travis~CI by re-using the testing image as the basis for an
image intended for publication on Docker~Hub.
\href{https://github.com/ColinFay/r-ci}{\texttt{r-ci}} is, in turn, used
with \href{https://docs.gitlab.com/ee/ci/}{GitLab~CI}, which itself is
built on top of Docker images: the user specifies a base Docker image
and control code, and the whole set of tests is run inside a container.
The \texttt{r-ci} image stack combines \texttt{rocker} versioning and a
series of tools specifically designed for testing in a fixed environment
with a customized list of preinstalled packages. Especially for
long-running tests or complex system dependencies, these approaches to
separate installation of build dependencies with code testing streamline
the development process. Containers can also simplify the integration of
R software into larger, multi-language CI pipelines. Furthermore, with
each change, even this manuscript is rendered into a PDF and deployed to
a GitHub-hosted website (see \texttt{.travis.yml} and
\texttt{Dockerfile} in the manuscript repository), not because of
concern about time, but to control the environment used on a CI server.
This gives, on the one hand, easy access after every update of the R
Markdown source code and, on the other hand a second controlled
environment to make sure that the article renders successfully and
correctly.

\hypertarget{processing}{%
\subsection{Processing}\label{processing}}

The portability of containerised environments becomes particularly
useful for improving expensive processing of data or shipping complex
processing pipelines. First, it is possible to \textbf{offload complex
processing to a server} or clouds and also to execute processes in
parallel to speed up or to serve many users.
\textbf{\CRANpkg{batchtools}} provides a parallel implementation of the
\texttt{Map} function for various schedulers \citep{Lang2017batchtools}.
For example, the package can
\href{https://mllg.github.io/batchtools/reference/makeClusterFunctionsDocker.html}{schedule
jobs with Docker Swarm}. \textbf{\CRANpkg{googleComputeEngineR}} has the
function \texttt{gce\_vm\_cluster()} to create clusters of 2 or more
virtual machines, running multi-CPU architectures
\citep{googleComputeEngineR_2019}. Instead of running a local R script
with the local CPU and RAM restrictions, the same code can be processed
on all CPU threads of the cluster of machines in the cloud, all running
a Docker container with the same R environments.
\CRANpkg{googleComputeEngineR} integrates with the R parallelisation
package \CRANpkg{future} \citep{cran_future} to enable this with only a
few lines or R
code\footnote{\href{https://cloudyr.github.io/googleComputeEngineR/articles/massive-parallel.html}{https://cloudyr.github.io/googleComputeEngineR/articles/massive-parallel.html}}.
\href{https://cloud.run}{Google Cloud Run} is a CaaS (Containers as a
Service) platform. Users can launch containers using any Docker image
without worrying about underlying infrastructure in a so-called
serverless configuration. The service takes care of network ingress,
scaling machines up and down, authentication, and authorisation---all
features which are non-trivial for a developer to build and maintain on
their own. This can be used to scale up R code to millions of instances
if need be with little or no changes to existing code, as demonstrated
by the proof of concept
\texttt{cloudRunR}\footnote{\href{https://github.com/MarkEdmondson1234/cloudRunR}{https://github.com/MarkEdmondson1234/cloudRunR}},
which uses Cloud Run to create a scalable R-based API using
\CRANpkg{plumber} \citep{cran_plumber}.
\href{https://cloud.google.com/cloud-build/}{Google Cloud Build} and the
Google Container Registry are a continuous integration service and an
image registry, respectively, that offload building of images to the
cloud, while serving the needs of commercial environments such as
private Docker images or image stacks. As Google Cloud Build itself can
run any container, the package \pkg{googleCloudRunner} demonstrates how
R can be used as the control language for one-time or batch processing
jobs and scheduling of
jobs\footnote{\href{https://code.markedmondson.me/googleCloudRunner/articles/cloudbuild.html}{https://code.markedmondson.me/googleCloudRunner/articles/cloudbuild.html}}.
\label{drake} \textbf{\CRANpkg{drake}} is a workflow manager for data
science projects \citep{landau_drake_2019}. It features implicit
parallel computing and automated detection of the parts of the work that
actually needs to be re-executed. \CRANpkg{drake} has been demonstrated
to run inside containers for high
reproducibility\footnote{See for example \href{https://github.com/joelnitta/pleurosoriopsis}{https://github.com/joelnitta/pleurosoriopsis} or \href{https://gitlab.com/ecohealthalliance/drake-gitlab-docker-example}{https://gitlab.com/ecohealthalliance/drake-gitlab-docker-example}, the latter even running in a continuous integration platform (cf.~\nameref{development}.}.
Furthermore, \CRANpkg{drake} workflows have been shown to use
\CRANpkg{future} package's function \texttt{makeClusterPSOCK()} for
sending parts of the workflow to a Docker image for
execution\footnote{\href{https://docs.ropensci.org/drake/index.html?q=docker\#with-docker}{https://docs.ropensci.org/drake/index.html?q=docker\#with-docker}}
\citep[see package's function documentation;~][]{future_2020}. In the
latter case, the container control code must be written by the user, and
the \CRANpkg{future} package ensures that the host and worker can
connect for communicating over socket connections. \textbf{RStudio
Server Pro} includes a functionality called
\href{https://solutions.rstudio.com/launcher/overview/}{Launcher} (since
version~1.2, released in 2019). It gives users the ability to spawn R
sessions and background/batch jobs in a scalable way on external
clusters, e.g.,
\href{https://support.rstudio.com/hc/en-us/articles/360019253393-Using-Docker-images-with-RStudio-Server-Pro-Launcher-and-Kubernetes}{Kubernetes
based on Docker images} or \href{https://slurm.schedmd.com/}{Slurm}
clusters, and optionally, with Singularity containers. A benefit of the
proprietary Launcher software is the ability for R and Python users to
leverage containerisation's advantages in RStudio without writing
specific deployment scripts or learning about Docker or managing
clusters at all.

\label{pipelines} Second, containers are perfectly suited for
\textbf{packaging and executing software pipelines} and required data.
Containers allow for building complex processing pipelines that are
independent of the host programming language. Due to its original use
case (see~\nameref{introduction}), Docker has no standard mechanisms for
chaining containers together; it lacks definitions and protocols for how
to use environment variables, volume mounts, and/or ports that could
enable transfer of input (parameters and data) and output (results) to
and from containers. Some packages, e.g., \pkg{containerit}, provide
Docker images that can be used very similar to a CLI, but this usage is
cumbersome\footnote{\href{https://o2r.info/containerit/articles/container.html}{https://o2r.info/containerit/articles/container.html}}.
\textbf{\pkg{outsider}} (\url{https://docs.ropensci.org/outsider/})
tackles the problem of integrating external programs into an R workflow
without thee need for users to directly interact with containers
\citep{bennett_outsider_2020}. Installation and usage of external
programs can be difficult, convoluted and even impossible if the
platform is incompatible. Therefore, \pkg{outsider} uses the
platform-independent Docker images to encapsulate processes in
\emph{outsider modules}. Each outsider module has a \texttt{Dockerfile}
and an R package with functions for interacting with the encapsulated
tool. Using only R functions, an end user can install a module with the
\pkg{outsider} package and then call module code to seamlessly integrate
a tool into their own R-based workflow. The \pkg{outsider} package and
module manage the containers and handle the transmission of arguments
and the transfer of files to and from a container. These functionalities
also allow a user to launch module code on a remote machine via SSH,
expanding the potential computational scale. Outsider modules can be
hosted code-sharing services, e.g., on GitHub, and \pkg{outsider}
contains discovery functions for them.

\hypertarget{deployment-and-continuous-delivery}{%
\subsection{Deployment and continuous
delivery}\label{deployment-and-continuous-delivery}}

\label{deployment}

The cloud is the natural environment for containers, and, therefore,
containers are the go-to mechanism for deploying R server applications.
More and more continuous integration (CI) and continuous delivery (CD)
services also use containers, opening up new options for use. The
controlled nature of containers, i.e., the possibility to abstract
internal software environment from a minimal dependency outside of the
container is crucial, for example to match test or build environments
with production environments or transfer runnable entities to
as-a-service infrastructures.

First, different packages use containers for the \textbf{deployment of R
and \CRANpkg{Shiny} apps}. \CRANpkg{Shiny} is a popular package for
creating interactive online dashboards with R, and it enables users with
very diverse backgrounds to create stable and user-friendly web
applications \citep{cran_shiny}. \emph{ShinyProxy}
(\url{https://www.shinyproxy.io/}) is an open source tool to deploy
Shiny apps in an enterprise context, where it features single sign-on,
but it can also be used in scientific use cases
\citep[e.g., ][]{savini_epiexplorer_2019,glouzon_structurexplor_2017}.
ShinyProxy uses Docker containers to isolate user sessions and to
achieve scalability for multi-user scenarios with multiple apps.
ShinyProxy itself is written in Java to accommodate corporate
requirements and may itself run in a container for stability and
availability. The tool is built on ContainerProxy
(\url{https://www.containerproxy.io/}), which provides similar features
for executing long-running R jobs or interactive R sessions. The started
containers can run on a regular Docker host but also in clusters.
Continuous integration and deployment (CI/CD) for Shiny applications
using Shinyproxy can be achieved, e.g., via
\href{https://docs.gitlab.com/ee/ci/pipelines.html}{GitLab pipelines}. A
pipeline can include building and checking R packages and Shiny apps.
After the code has passed the checks, Docker images are built and pushed
to GitLab's container registry. The pipeline finishes with triggering a
webhook on the server, where the deployment script is executed. The
script updates configurations, pulls the new Docker images, and restarts
the services.

Another example is the package \CRANpkg{golem}, which makes an heavy use
of \texttt{dockerfiler} when it comes to creating the
\texttt{Dockerfile} for building and deploying production-grade Shiny
applications \citep{cran_golem}. \CRANpkg{googleComputeEngineR} enables
quick deployments of key R services, such as RStudio and Shiny, onto
cloud virtual machines (VMs) with Google Cloud Compute Engine
\citep{googleComputeEngineR_2019}. The package utilises
\texttt{Dockerfile}s to move the labour of setting up those services
from the user to a premade Docker image, which is configured and run in
the cloud VM. For example, by specifying the template
\texttt{template="rstudio"} in functions
\texttt{gce\_vm\_template()/gce\_vm()} an up-to-date RStudio Server
image is launched for development work, whereas specifying
\texttt{template="rstudio-gpu"} will launch an RStudio Server image with
a GPU attached, etc.

Second, containers can be used to create \textbf{platform installation
packages} in a DevOps setting. The
\href{https://www.opencpu.org/}{OpenCPU} system provides an HTTP API for
data analysis based on R. \citet{ooms_opencpu_2017} describes how
various platform-specific installation files for OpenCPU are created
using Docker~Hub. The automated builds install the software stack from
the source code on different operating systems; afterwards a script file
downloads the images and extracts the OpenCPU binaries.

Third, containers can greatly facilitate the \textbf{deployment to
existing infrastructures}. \emph{Kubernetes}
(\url{https://kubernetes.io/}) is a container-orchestration system for
managing container-based application deployment and scaling. A
\emph{cluster} of containers, orchestrated as a single deployment, e.g.,
with Kubernetes, can mitigate limitations on request volumes or a
container occupied with computationally intensive task. A cluster
features load-balancing, autoscaling of containers across numerous
servers (in the cloud or on premise), and restarting failed ones. Many
organisations already use a Kubernetes cluster for other applications,
or a managed cluster can be acquired from service providers. Docker
containers are used within Kubernetes clusters to hold native code, for
which Kubernetes creates a framework around network connections and
scaling of resources up and down. Kubernetes can thereby host R
applications, big parallel tasks, or scheduled batch jobs in a scalable
way, and the deployment can even be triggered by changes to code
repositories \citep[i.e., CD, see][]{edmondson_r_2018}. The package
\pkg{googleKubernetesR}
(\url{https://github.com/RhysJackson/googleKubernetesR}) is a proof of
concept for wrapping the Google Kubernetes Engine API, Google's hosted
Kubernetes solution, in an easy-to-use R package. The package
\CRANpkg{analogsea} provides a way to programmatically create and
destroy cloud VMs on the \href{https://www.digitalocean.com/}{Digital
Ocean} platform \citep{analogsea_2019}. It also includes R wrapper
functions to install Docker in such a VM, manage images, and control
containers straight from R functions. These functions are translated to
Docker CLI commands and transferred transparently to the respective
remove machine using SSH. \CRANpkg{AzureContainers} is an umbrella
package that provides interfaces for three commercial services of
Microsoft's Azure Cloud, namely
\href{https://azure.microsoft.com/en-us/services/container-instances/}{Container
Instances} for running individual containers,
\href{https://azure.microsoft.com/en-us/services/container-registry/}{Container
Registry} for private image distribution, and
\href{https://azure.microsoft.com/en-us/services/kubernetes-service/}{Kubernetes
Service} for orchestrated deployments. While a package like
\CRANpkg{plumber} provides the infrastructure for turning an R workflow
into a web service, for production purposes it is usually necessary to
take into account scalability, reliability and ease of management.
AzureContainers provides an R-based interface to these features and,
thereby, simplifies complex infrastructure management to a number of R
function calls, given an Azure account with sufficient
credit\footnote{See \emph{"Deploying a prediction service with Plumber"} vignette for details:  \href{https://cran.r-project.org/web/packages/AzureContainers/vignettes/vig01_plumber_deploy.html}{https://cran.r-project.org/web/packages/AzureContainers/vignettes/vig01\_plumber\_deploy.html}.}.
\href{https://www.heroku.com/}{Heroku} is another cloud platform as a
service provider, and it supports container-based applications.
\texttt{heroku-docker-r}
(\url{https://github.com/virtualstaticvoid/heroku-docker-r}) is an
independent project providing a template for deploying R applications
based on Heroku's image stack, including multiple examples for
interfacing R with other programming languages. Yet the approach
requires a manual management of the computing environment.

Independent integrations of R for different cloud providers lead to
repeated efforts and code fragmentation. To mitigate these problems and
to avoid vendor lock-in motivated the
\href{https://www.openfaas.com/}{OpenFaaS project}. OpenFaas facilitates
the deployment of functions and microservices to Kubernetes or Docker
Swarm. It is language agnostic and provides auto-scaling, metrics, and
an API gateway. Reduced boilerplate code is achieved via templates.
Template for
R\footnote{See OpenFaaS R templates at \href{https://github.com/analythium/openfaas-rstats-templates}{https://github.com/analythium/openfaas-rstats-templates}.}
are provided based on Rocker's Debian and R-hub's
\href{https://github.com/r-hub/r-minimal}{r-minimal} Alpine images. The
templates use
\href{https://docs.docker.com/develop/develop-images/multistage-build/}{multi-stage
Docker builds} to combine R base images with the OpenFaaS `watchdog', a
tiny Golang web server. The watchdog marshals an HTTP request and
invokes the actual application. The R session uses \CRANpkg{plumber} or
similar packages for the API endpoint with packages and data preloaded,
thus minimizing response times.

The prevalence of Docker in industry naturally leads to the use of
containers with R in such settings as well, as companies already manage
platforms in Docker containers. These products often entail a large
amount of open source software in combination with proprietary layers
adding the relevant commercialisation features. One such example is
RStudio's data science platform
\href{https://rstudio.com/products/team/}{RStudio Team}. It allows teams
of data scientists and their respective IT/DevOps groups to develop and
deploy code in R and Python around the RStudio Open Source Server inside
of Docker images, without requiring users to learn new tools or directly
interact with containers. The best practices for
\href{https://support.rstudio.com/hc/en-us/articles/360021594513-Running-RStudio-with-Docker-containers}{running
RStudio with Docker containers} as well as
\href{https://github.com/rstudio/rstudio-docker-products}{Docker images}
for RStudio's commercial products are publicly available.

\hypertarget{using-r-to-power-enterprise-software-in-production-environments}{%
\subsection{Using R to power enterprise software in production
environments}\label{using-r-to-power-enterprise-software-in-production-environments}}

\label{enterprise}

R has been historically viewed as a tool for analysis and scientific
research, but not for creating software that corporations can rely on
for production services. However, thanks to advancements in R running as
a web service, along with the ability to deploy R in Docker containers,
modern enterprises are now capable of having real-time machine learning
powered by R. A number of packages and projects have enabled R to
respond to client requests over TCP/IP and local socket servers, such as
\CRANpkg{Rserve} \citep{cran_rserve}, \CRANpkg{svSocket}
\citep{grosjean_sciviews_2019}, \href{http://www.rapache.net}{rApache}
and more recently \CRANpkg{plumber} (\url{https://www.rplumber.io/}) and
\pkg{RestRserve} (\url{http://restrserve.org}), which even processes
incoming requests in parallel with forked processes using
\CRANpkg{Rserve}. The latter two also provide documentation for
deployment with Docker or ready-to-use images with automated
builds\footnote{See \href{https://www.rplumber.io/docs/hosting.html\#docker}{https://www.rplumber.io/docs/hosting.html\#docker}, \href{https://hub.docker.com/r/trestletech/plumber/}{https://hub.docker.com/r/trestletech/plumber/} and \href{https://hub.docker.com/r/rexyai/restrserve/}{https://hub.docker.com/r/rexyai/restrserve/}.}.
These software allow other (remote) processes and programming languages
to interact with R and to expose R-based function in a service
architecture with HTTP APIs. APIs based on these packages can be
deployed with scalability and high availability using containers. This
pattern of deploying code matches those used by software engineering
services created in more established languages in the enterprise domain,
such as Java or Python, and R can be used alongside those languages as a
first-class member of a software engineering technical stack.

CARD.com implemented a web application for the optimisation of the
acquisition flow and the real-time analysis of debit card transactions.
The software used \CRANpkg{Rserve} and rApache and was deployed in
Docker containers. The R session behind \CRANpkg{Rserve} acted as a
read-only in-memory database, which was extremely fast and scalabale,
for the many concurrent rApache processes responding to the live-scoring
requests of various divisions of the company. Similarly deodorised R
scripts were responsible for the ETL processes and even the
client-facing email, text message and push notification alerts sent in
real-time based on card transactions. The related Docker images were
made available at \url{https://github.com/cardcorp/card-rocker}. The
images extended \texttt{rocker/r-base} and additionally entailed an SSH
client and a workaround for being able to mount SSH keys from the host,
Pandoc, the Amazon Web Services (AWS) SDK, and Java, which is required
by the AWS SDK. The AWS SDK allowed for running R consumers reading from
real-time data processing streams of
\href{https://aws.amazon.com/kinesis/}{AWS Kinesis}
\footnote{See useR!2017 talk \href{https://static.sched.com/hosted\_files/user2017/2f/AWR Kinesis at useR 2017.pdf}{"Stream processing with R in AWS"}.}.
The applications were deployed on Amazon Elastic Container Service
(\href{https://aws.amazon.com/ecs/}{ECS}). The main takeaways from using
R in Docker were not only that pinning the R package versions via MRAN
is important, but also that moving away from Debian testing to a
distribution with long-term support can be necessary. For the use case
at hand, this switch allowed for more control over upstream updates and
for minimising the risk of breaking the automated builds of the Docker
images and production jobs.

The AI @ T-Mobile team created a set of neural network machine learning
natural language processing models to help customer care agents manage
text-based messages for customers \citep{t-mobile_enterprise_2018}. For
example, one model quickly identifies whether a message is from a
customer
\citep[see \CRANpkg{Shiny}-based \href{https://secure.message.t-mobile.com/v1/shiny/is-customer/app/}{demo} further described by ][]{nolis_small_2019},
and others tell which customers are likely to make a repeat purchase. If
a data scientist creates a machine learning model and exposes it through
a \CRANpkg{plumber} API, then someone else on the marketing team could
write software that sends different emails depending on that real-time
prediction. The models are convolutional neural networks that use the
\CRANpkg{keras} package \citep{cran_keras} and run in a Rocker
container. The corresponding \texttt{Dockerfile}s are published
\href{https://github.com/tmobile/r-tensorflow-api}{on GitHub}. Since the
models power tools for agents and customers, they need to have extremely
high uptime and reliability. The AI @ T-Mobile team found that the
models performed well, and today these models power real-time services
that are called over a million times a day.

\hypertarget{common-or-public-work-environments}{%
\subsection{Common or public work
environments}\label{common-or-public-work-environments}}

\label{workenvs}

The fact that Docker images are portable and well defined make them
useful when more than one person needs access to the same computing
environment. This is even more useful when some of the users do not have
the expertise to create such an environment themselves, and when these
environments can be run in public or using shared infrastructure. For
example, \emph{RCloud} (\url{https://rcloud.social}) is a cloud-based
platform for data analysis, visualisation and collaboration using R. It
provides a \texttt{rocker/drd} base image for easy evaluation of the
platform\footnote{\href{https://github.com/att/rcloud/tree/master/docker}{https://github.com/att/rcloud/tree/master/docker}}.

\label{binder} The
\href{https://mybinder.readthedocs.io/en/latest/}{\textbf{Binder}}
project, maintained by the team behind Jupyter, makes it possible for
users to \textbf{create and share computing environments} with others
\citep{jupyter_binder_2018}. A \emph{BinderHub} allows anyone with
access to a web browser and an internet connection to launch a temporary
instance of these custom environments and execute any workflows
contained within. From a reproducibility standpoint, Binder makes it
exceedingly easy to compile a paper, visualize data, and run small
examples from papers or tutorials without the need for any local
installation. To set up Binder for a project, a user typically starts at
an instance of a BinderHub and passes the location of a repository with
a workspace, e.g., a hosted Git repository, or a data repository like
Zenodo. Binder's core internal tool is \texttt{repo2docker}. It
deterministically builds a Docker image by parsing the contents of a
repository, e.g., project dependency configurations or simple
configuration
files\footnote{See supported file types at \href{https://repo2docker.readthedocs.io/en/latest/config\_files.html}{https://repo2docker.readthedocs.io/en/latest/config\_files.html}. For R, the }.
In the most powerful case, \texttt{repo2docker} builds a given
\texttt{Dockerfile}. While this approach works well for most
run-of-the-mill Python projects, it is not so seamless for R projects.
This is partly because \texttt{repo2docker} does not support arbitrary
base images due to the complex auto-generation of the
\texttt{Dockerfile} instructions.

Two approaches make using Binder easier for R users. First,
\textbf{\pkg{holepunch}} (\url{https://github.com/karthik/holepunch}) is
an R package that was designed to make sharing work environments
accessible to novice R users based on Binder. For any R projects that
use the Tidyverse suite \citep{wickham_welcome_2019}, the time and
resources required to build all dependencies from source can often time
out before completion, making it frustrating for the average R user.
\pkg{holepunch} removes some of these limitations by leveraging Rocker
images that contain the Tidyverse along special Jupyter dependencies,
and only installs additional packages from CRAN and Bioconductor that
are not already part of these images. It short circuits the
configuration file parsing in \texttt{repo2docker} and starts with the
Binder/Tidyverse base images, which eliminates a large part of the build
time and, in most cases, results in a Binder instance launching within a
minute. \pkg{holepunch} also creates a \texttt{DESCRIPTION} file for
essential metadata and dependency specification, and thereby turns turns
any project into a research compendium (see~\nameref{compendia}). The
\texttt{Dockerfile} included with the project can also be used to launch
an RStudio Server instance locally, i.e., independent of Binder, which
is especially useful when more or special computational resources can be
provided there. The local image usage reduces the number of separately
managed environments and, thereby, reduces work and increases
portability and reproducibility.

Second, the \textbf{Whole~Tale} project (\url{https://wholetale.org})
combines the strengths of the Rocker Project's curated Docker images
with \texttt{repo2docker}. Whole~Tale is a National Science Foundation
(NSF) funded project developing a scalable, open source, multi-user
platform for reproducible research \citep{brinckman2019, chard2019a}. A
central goal of the platform is to enable researchers to easily create
and publish executable research
objects\footnote{In Whole~Tale a \emph{tale} is a research object that contains metadata, data (by copy or reference), code, narrative, documentation, provenance, and information about the computational environment to support computational reproducibility.}
associated with published research \citep{chard2019b}. Using Whole~Tale,
researchers can create and publish Rocker-based reproducible research
objects to a growing number of repositories including DataONE member
nodes, Zenodo and soon Dataverse. Additionally, Whole~Tale supports
automatic data citation and is working on capabilities for image
preservation and provenance capture to improve the transparency of
published computational research artefacts
\citep{mecum2018, mcphillips2019}. For R users, Whole~Tale extends the
Jupyter Project's \texttt{repo2docker} tool to simplify the
customisation of R-based environments for researchers with limited
experience with either Docker or Git. Multiple options have been
discussed to allow users to change the Ubuntu LTS (long-term support,
currently \emph{Bionic Beaver}) base image,
\texttt{buildpack-deps:bionic}, used in \texttt{repo2docker}. Whole~Tale
implemented a custom
\texttt{RockerBuildPack}\footnote{See \href{https://github.com/whole-tale/repo2docker\_wholetale}{https://github.com/whole-tale/repo2docker\_wholetale}.}.
The build pack combines a \texttt{rocker/geospatial} image with
\texttt{repo2docker}'s
composability\footnote{Composability refers to the ability to combine multiple package managers and their configuration files, such as R, `pip`, and `conda`; see Section~\nameref{binder} for details.}.
This works because both Rocker images and the \texttt{repo2docker} base
image use distributions with APT \citep{wikipedia_contributors_apt_2020}
so that the instructions created by the latter work because of the
compatible shell and package manager.

In \textbf{high-performance computing}, one use for containers is to run
workflows on shared local hardware where teams manage their own
high-performance servers. This can follow one of several design
patterns: Users may deploy containers to hardware as a work environment
for a specific project, containers may provide per-user persistent
environments, or a single container can act as a common multi-user
environment for a server. In all cases, though, the containerised
approach provides several advantages: First, users may use the same
image and thus work environment on desktop and laptop computers. The
first to patterns provide modularity, while the last approach is most
similar to a simple shared server. Second, software updates can be
achieved by updating and redeploying the container rather than by
tracking local installs on each server. Third, the containerised
environment can be quickly deployed to other hardware, cloud or local,
if more resources are necessary or in case of server destruction or
failure. In any of these cases, users need a method to interact with the
containers, be it an IDE exposed over an HTTP port or command-line
access via tools such as SSH. A suitable method must be added to the
container recipes. The Rocker~Project provides containers pre-installed
with the RStudio~IDE. In cases where users store nontrivial amounts of
data for their projects, the data needs to persist beyond the life of
the container. This may be in shared disks, attached network volumes, or
in separate storage where it is uploaded between sessions. In the case
of shared disks or network-attached volumes, care must be taken to match
user permissions, and of course backups are still necessary.

\href{https://cyverse.org}{CyVerse} is an open source, NSF-funded
cyberinfrastructure platform for the life sciences providing easy access
to computing and storage resources \citep{merchant_iplant_2016}.
CyVerse's a browser-based `data science workbench' is called
\href{https://cyverse.org/discovery-environment}{Discovery Environment}
(DE). DE uses a combination of
\href{https://research.cs.wisc.edu/htcondor/}{HTCondor} and Kubernetes
for orchestrating container-based analysis and integrates with external
HPC, i.e., \href{https://www.xsede.org/}{NSF-XSEDE}, through
\href{https://www.tacc.utexas.edu/tapis}{TAPIS} API's. In addition,
CyVerse provides a multi-petabyte Data Store based on
\href{https://irods.org/}{iRODS} with shared access by users. The DE
runs Docker containers on demand, with users able to integrate bespoke
containers from DockerHub or other registries
\citep{devisetty_bringing_2016}. Rocker images integration in the DE is
designed to provide life science researchers with scalable,
compute-intensive R analysis capabilities needed for large and complex
datasets (e.g., genomics/multi-omics, GWAS, phenotypic data, spatial
data etc.). These capabilities give users flexibility similar to Binder,
but allows containers to run on larger computational resources (RAM,
CPU, Disk), and for longer periods of time (days to weeks). The Rocker
Project's RStudio and Shiny containers are integrated into the DE by
deriving new images from Rocker
images\footnote{See \href{https://github.com/cyverse-vice/}{https://github.com/cyverse-vice/} for \texttt{Dockerfile}s and configuration scripts; images are auto-built on DockerHub at \href{https://hub.docker.com/u/cyversevice}{https://hub.docker.com/u/cyversevice}.}.
These images include a reverse proxy using \texttt{nginx} to handle
communication with CyVerse's authentication systems, which allow users
to invite other registered users to securely access the same instance
\citep{rstudio_proxy_2020}. The images further include specific tools,
such as the CLI utility \texttt{icommands} to access iRODS. Users can
reference the same base image and add packages to further customize
their R environment, or they can use CyVerse's image registry where a
user's published, i.e., publicly available, containers are held for
additional reproducibility. CyVerse user accounts are free (with some
limitations for non-US users) and the
\href{https://learning.cyverse.org/}{CyVerse Learning Center} allows
community members to document their containers to support training and
education.

\label{rocker-gpu} Using \textbf{GPUs} (graphical processing units) as a
specialised hardware from containerised common work environments is also
possible and useful \citep{haydel_enhancing_2015}. GPUs are increasingly
popular for compute-intensive machine learning (ML) tasks, e.g., deep
artificial neural networks \citep{schmidhuber_deep_2015}. Although in
this case containers are not completely portable between hardware
environments, but the software stack for ML with GPUs is so complex to
set up that a ready-to-use container is helpful. Containers running GPU
software require drivers and libraries specific to GPU models and
versions, and containers require a specialized runtime to connect to the
underlying GPU hardware. For NVIDIA GPUs, the
\href{https://github.com/NVIDIA/nvidia-docker}{NVIDIA Container Toolkit}
includes a specialized runtime plugin for Docker and a set of base
images with appropriate drivers and libraries. The Rocker~Project
\href{https://github.com/rocker-org/ml}{has a repository} with (beta)
images based on these that include GPU-enabled versions of
machine-learning R packages, e.g., \texttt{rocker/ml} and
\texttt{rocker/tensorflow-gpu}.

\textbf{Teaching} is a further example of where shared computing
environments and sandboxing can greatly improve the process. First,
\emph{prepared environments for teaching} are especially helpful for (a)
introductory courses, where students often struggle with the first step
of installation and configuration
\citep{cetinkaya-rundel_infrastructure_2018}, and (b) courses that
require access to a relatively complex setup of software tools, e.g.,
database systems. \citet{cetinkaya-rundel_infrastructure_2018} describe
how a Docker-based deployment of RStudio (i) avoided problems with
troubleshooting individual students' computers and greatly increased
engagement through very quickly showing tangible outcomes, e.g., a
visualisation, and (ii) reduced demand on teaching and IT staff. Each
student received access to a personal RStudio instance running in a
container after authentication with the university login, which gives
the benefits of sandboxing and the possibility of limiting resources.
\citet{cetinkaya-rundel_infrastructure_2018} found that for the courses
at hand, actual usage of the UI is intermittent so a single cloud-based
VM with four cores and 28~GB RAM sufficed for over 100 containers. An
example for mitigating \emph{complex setups} is teaching databases. R is
very useful tool for interfacing with databases, because almost every
open source and proprietary database system has an R package that allows
users to connect and interact with it. This flexibility is even
broadened by \CRANpkg{DBI} \citep{cran_dbi}, which allows for creating a
common API for interfacing these databases, or the \CRANpkg{dbplyr}
package \citep{cran_dbplyr}, which runs \CRANpkg{dplyr}
\citep{cran_dplyr} code straight against the database as queries. But
learning and teaching these tools comes with the cost of deploying or
having access to an environment with the software and drivers installed.
For people teaching R, it can become a barrier if they need to install
local versions of database drivers or connect to remote instances which
might or might not be made available by IT services. Giving access to a
sandbox for the most common environments for teaching databases is the
idea behind \href{https://github.com/ColinFay/r-db}{\texttt{r-db}}, a
Docker image that contains everything needed to connect to a database
from R. Notably, with \texttt{r-db}, users do not have to install
complex drivers or configure their machine in a specific way. The
\texttt{rocker/tidyverse} base image ensures that users can also readily
use packages for analysis, display, and reporting.

Second, the idea of a common environment and partitioning allows for
using \emph{containers in teaching for secure execution and automated
testing} of submissions by students.
\href{https://dodona.ugent.be}{Dodona} is a web platform developed at
Ghent University that is used to teach students basic programming
skills, and it uses Docker containers to test submissions by students.
This means that both the code testing the students' submissions and the
submission itself are executed in a predictable environment, avoiding
compatibility issues between the wide variety of configurations used by
students. The containerisation is also used to shield the Dodona servers
from bad or even malicious code: memory, time and I/O~limits are used to
make sure students cannot overload the system. The web application
managing the containers communicates with them by sending configuration
information as a JSON document over standard input. Every Dodona Docker
image shares a \texttt{main.sh} file that passes through this
information to the actual testing framework, while setting up some error
handling. The testing process in the Docker containers sends back the
test results by writing a JSON document to its standard output channel.
In June 2019, R support was added to Dodona using an image derived from
the \texttt{rocker/r-base} image that sets up the \texttt{runner} user
and and \texttt{main.sh} file expected by
Dodona\footnote{\href{https://github.com/dodona-edu/docker-images/blob/master/dodona-r.dockerfile}{https://github.com/dodona-edu/docker-images/blob/master/dodona-r.dockerfile}}.
It also installs the packages required for the testing framework and the
exercises so that this does not have to happen every time a student's
submission is evaluated. The actual testing of R exercises is done using
a custom framework loosely based on \CRANpkg{testthat}
\citep{wickham_testthat_2011}. During the development of the testing
framework, it was found that the \CRANpkg{testthat} framework did not
provide enough information to its reporter system to send back all the
fields required by Dodona to render its feedback. Right now, multiple
statistics courses are developing exercises to automate the feedback for
their lab classes.

\hypertarget{packaging-research-reproducibly}{%
\subsection{Packaging research
reproducibly}\label{packaging-research-reproducibly}}

\label{compendia}

Containers provide a high degree of isolation that is often desirable
when attempting to capture a specific computational environment so that
others can reproduce and extend a research result. Many computationally
intensive research projects depend on specific versions of original and
third-party software packages in diverse languages, joined together to
form a pipeline through which data flows. New releases of even just a
single piece of software in this pipeline can break the entire workflow,
making it difficult to find the error and difficult for others to reuse
existing pipelines. These breakages can make the original the results
irreproducible and, and the chance of a substantial disruption like this
is high in a multi-year research project where key pieces of third-party
software may have several major updates over the duration of the
project. The classical ``paper'' article is insufficient to adequately
communicate the knowledge behind such research projects
\citep[cf.][]{donoho_invitation_2010,marwick_how_2015}.

\citet{gentleman_statistical_2007} coined the term \textbf{Research
Compendium} for a dynamic document together with supporting data and
code. They used the R package system \citep{core_writing_1999} for the
functional prototype all the way to structuring, validating, and
distributing research compendia. This concept has been taken up and
extended\footnote{See full literature list at \href{https://research-compendium.science/}{https://research-compendium.science/}.},
not in the least by applying containerisation and other methods for
managing computing environments---see Section~\nameref{envs}. Containers
give the researcher an isolated environment to assemble these research
pipelines with specific versions of software to minimize problems with
breaking changes and make workflows easier to share
\citep[cf.][]{boettiger_introduction_2015,marwick_packaging_2018}.
Research workflows in containers are safe from contamination from other
activities that occur on the researcher's computer, for example the
installation of newest version of packages for teaching demonstrations
or specific versions for evaluation of others' works. Given the users in
this scenario, i.e., often academics with limited formal software
development training, templates and assistance with containers around
research compendia is essential. In many fields we see that a typical
unit of research for a container is a research report or journal
article, where the container holds the compendium, or self-contained set
of data (or connections to data elsewhere) and code files needed to
fully reproduce the article \citep{marwick_packaging_2018}. The package
\pkg{rrtools} (\url{https://github.com/benmarwick/rrtools}) provides a
template and convenience functions to apply good practices for research
compendia, including a starter \texttt{Dockerfile}. Images of compendium
containers can be hosted on services such as Docker~Hub for convenient
sharing among collaborators and others. Similarly, packages such as
\pkg{containerit} and \pkg{dockerfiler} can be used to manage the
\texttt{Dockerfile} to be archived with a compendium on a data
repository (e.g.~\href{https://zenodo.org/}{Zenodo},
\href{https://dataverse.org/}{Dataverse},
\href{https://figshare.com/}{Figshare}, \href{https://osf.io/}{OSF}). A
typical compendium's \texttt{Dockerfile} will pull a rocker image fixed
to a specific version of R, and install R packages from the MRAN
repository to ensure the package versions are tied to a specific date,
rather than the most recent version. A more extreme case is the
\emph{dynverse} project (\url{https://dynverse.org/}), which packages
over 50 computational methods with different environments (R, Python,
C++, etc.) in Docker images, which can be executed from R.
\emph{dynverse} uses a CI platform (see~\nameref{ci}) to build
Rocker-derived images, test them, and, if the tests succeed, publish
them on Docker~Hub.

Future researchers can download the compendium from the repository and
run the included \texttt{Dockerfile} to build a new image that recreates
the computational environment used to produce the original research
results. If building the image fails, the human-readable instructions in
a \texttt{Dockerfile} are the starting point for rebuilding the
environment. When combined with CI (see~\nameref{ci}), a research
compendium set-up can enable \emph{continuous analysis} with easier
verification of reproducibility and audits trails
\citep{beaulieu-jones_reproducibility_2017}.

Further safeguarding practices are currently under development or not
part of common practice yet, such as preserving images
\citep{emsley_framework_2018}, storing both images and
\texttt{Dockerfile}s \citep[cf.][]{nust_opening_2017}, or pinning system
libraries beyond the tagged base images, which may be seen as stable or
dynamic depending on the applied time scale
\citep[see discussion on \texttt{debian:testing} base image in][]{RJ-2017-065}.
A recommendation of the recent National Academies' report on
\emph{Reproducibility and Replicability in Science} is that journals
\emph{``consider ways to ensure computational reproducibility for
publications that make claims based on computations''}
\citep{NASEM2019}. In fields such as political science and economics,
journals are increasingly adopting policies that require authors to
publish the code and data required to reproduce computational findings
reported in published manuscripts, subject to independent verification
\citep{Jacoby2017,Vilhuber2019,Alvarez2018,Christian2018,Eubank2016,King1995}.
Problems with the computational environment, installation and
availability of software dependencies are common. R is gaining
popularity in these communities, such as for creating a research
compendium. In a sample of 105 replication packages published by the
\emph{American Journal of Political Science} (AJPS), over 65\% use R.
The NSF-funded Whole Tale project, which was mentioned above, uses the
Rocker Project community images with the goal of improving the
reproducibility of published research artefacts and simplifying the
publication and verification process for both authors and reviewers by
reducing errors and time spent specifying the environment.

\hypertarget{conclusions}{%
\section{Conclusions}\label{conclusions}}

This article is a snapshot of the R corner in a universe of applications
built with a many-faced piece of software, Docker. \texttt{Dockerfile}s
and Docker images are the go-to methods for collaboration between roles
in an organisation, such as development and IT operations teams, and
between parties in the communication of knowledge, such as research
workflows or education. Docker has become synonymous with applying the
concept of containerisation to solve challenges of reproducible
environments, e.g., in research and in development \& production, and of
scalable deployments as it can easily move processing between machines
(e.g., locally, a cloud provider's VM, another cloud provider's
Container-as-a-Service). Reproducible environments, scalability and
efficiency, and portability across platforms/infrastructures are the
common themes behind R packages, use cases, and applications in this
work.

The R packages and use cases presented above show the growing number of
users, developers, and real-world applications in the community and the
resulting innovations. But the applications also point to the challenges
of keeping up with a continuously evolving landscape. The use cases
contributed by co-authors also have a degree of overlap, which can be
expected as a common language and understanding of good practices is
still taking shape. Also, the ease with which one can create a complex
software systems with Docker, such as an independent Docker image stack,
to serve one's specific needs leads to parallel developments. This
ease-of-DIY in combination with the difficulty of reusing parts from or
composing multiple \texttt{Dockerfile}s is a further reason for
\textbf{fragmentation}. Instructions can be outsourced into
distributable scripts and then copied into the image during build, but
that makes \texttt{Dockerfile}s harder to read. Scripts added to a
\texttt{Dockerfile} also add a layer of complexity and increase the risk
of incomplete recipes. Despite the different image stacks presented
here, the pervasiveness of Rocker images can be traced back to its
maintainers and the user community valuing collaboration and shared
starting points over impulses to create individual solutions. Aside from
that, fragmentation may not be a bad sign but may instead be a
reflection of a growing market that is able to sustain multiple related
efforts. With the maturing of core building blocks, such as the Rocker
Project's suite of images, more systems will be built successfully, but
they may also be built behind the curtains. Docker alone, as a flexible
core technology, is not a feasible level of collaboration and
abstraction. Instead, the use cases and applications observed in this
work provide a more useful division.

Nonetheless, at least on the level of R packages some
\textbf{consolidation} seems in order, e.g., to reduce the number of
packages creating \texttt{Dockerfile}s from R code or controlling the
Docker daemon with R code. It remains to be seen which approach to
control Docker, via the Docker API as \pkg{stevedore} or via system
calls as \pkg{dockyard}/\pkg{docker}/\pkg{dockr}, is more sustainable,
or whether the question will be answered by the endurance of maintainers
and sufficient funding. Similarly, capturing environments and their
serialisation in form of a \texttt{Dockerfile} currently is happening at
different levels of abstraction, and re-use of functionality seems
reasonable, e.g., \pkg{liftr} could generate the environment with
\pkg{containerit}, which in turn may use \pkg{dockerfiler} for low-level
R objects representing a \texttt{Dockerfile} and its instructions. In
this consolidation of R packages, the Rocker~Project could play the role
of a coordinating entity. Nonetheless, for the moment, it seems that the
Rocker Project will focus on maintaining and extending its image stacks,
e.g., images for GPU-based computing and artificial intelligence. Even
with coding being more and more accepted as a required and achievable
skill, an easier access, for example by exposing containerisation
benefits via simple user interfaces in the users' IDE, could be an
important next step, since currently containerisation happens more in
the background for UI-based development (e.g., a \texttt{rocker/rstudio}
image in the cloud). Furthermore, the maturing of the Rockerverse
packages for managing containers may lead to them being adopted in
situations where manual coding is currently required, e.g.~in the case
of \CRANpkg{RSelenium} or \CRANpkg{drake} (see
Sections~\nameref{rselenium} and \nameref{drake} respectively). In some
cases, e.g., for \CRANpkg{analogsea}, the interaction with the Docker
daemon may remain too specific to re-use first-order packages to control
Docker.

New features which make complex workflows accessible and reproducible
and the variety in packages connected with containerisation, even when
they have overlapping features, are a signal and a support for a growing
user base. This growth is possibly the most important goal for the
foreseeable future in the \emph{Rockerverse}, and, just like the Rocker
images have matured over years of use and millions of runs, the new
ideas and prototypes will have to prove themselves. It should be noted
that the dominant position is that Docker is a blessing and a curse for
these goals. It might be wise to start experimenting with non-Docker
containerisation tools now, e.g., R packages interfacing with other
container engines, such as
\href{https://github.com/containers/libpod}{podman/buildah}, or an R
package for creating \texttt{Singularity} files. Such efforts might help
to avoid lock-in and to design sustainable workflows based on concepts
of \emph{containerisation}, not on their implementation in Docker. If
adoption of containerisation and R continue to grow, the missing pieces
for a success predominantly lie in (a) coordination and documentation of
activities to reduce repeated work in favour of open collaboration, (b)
the sharing of lessons learned from use cases to build common knowledge
and language, and (c) a sustainable continuation and funding for
development, community support, and education. A first concrete effort
to work towards these missing pieces should be sustain the structure and
captured status quo from this work in form of a \emph{CRAN Task View on
containerisation}.

\hypertarget{author-contributions}{%
\section{Author contributions}\label{author-contributions}}

The ordering of authors following DN and DE is alphabetical. DN
conceived the article idea,
\href{https://github.com/nuest/rockerverse-paper/issues/3}{initialised the formation of the writing team},
wrote sections not mentioned below, and revised all sections. DE wrote
the introduction and the section about containerisation and the Rocker
Project, and reviewed all sections. DB wrote the section on
\pkg{outsider}. GD contributed the CARD.com use case. RC \& EH
contributed to the section on interfaces for Docker in R
(\emph{dynverse} and \texttt{dynwrap}). DC contributed content on
Gigantum. ME contributed to the section on processing and deployment to
cloud services. CF wrote paragraphs about \pkg{r-online},
\pkg{dockerfiler}, \pkg{r-ci} and \pkg{r-db}. LK contributed content on
\pkg{dockr}. SL contributed content on RStudio's usage of Docker. BM
wrote the section on research compendia and made the project Binder
ready. HN \& JN co-wrote the section on the T-Mobile use case. KR wrote
the section about \pkg{holepunch}. NR wrote paragraphs about shared work
environments and GPUs. LS \& NT wrote the section on Bioconductor. PS
wrote the paragraphs about GitLab CI/CD pipelines with Shinyproxy and
OpenFaaS templates. TS \& JW wrote the section on CyVerse. CvP wrote the
section on the usage of Docker containers in Dodona. CW wrote the
sections on Whole Tale and contributed the publication reproducibility
audit use case. NX contributed content on \pkg{liftr}. All authors
approved the final version. This articles was collaboratively written at
\href{https://github.com/nuest/rockerverse-paper/}{https://github.com/nuest/rockerverse-paper/}.
The
\href{https://github.com/nuest/rockerverse-paper/graphs/contributors}{contributors page},
\href{https://github.com/nuest/rockerverse-paper/commits/master}{commit history},
and
\href{https://github.com/nuest/rockerverse-paper/issues/}{discussion issues}
provide a detailed view on the respective contributions.

\hypertarget{acknowledgements}{%
\section{Acknowledgements}\label{acknowledgements}}

DN is supported by the project Opening Reproducible Research
(\href{https://www.uni-muenster.de/forschungaz/project/12343}{o2r})
funded by the German Research Foundation (DFG) under project number
\href{https://gepris.dfg.de/gepris/projekt/415851837}{PE~1632/17-1}. The
funders had no role in data collection and analysis, decision to
publish, or preparation of the manuscript. KR was supported in part by a
grant from The Leona M. and Harry B. Helmsley Charitable Trust, award
number 2016PG-BRI004. LS and NT are supported by US NIH / NHGRI awards
U41HG00405 and U24HG010263. CW is supported by the Whole Tale projects
(\url{https://wholetale.org}) funded by the US National Science
Foundation (NSF) under award
\href{https://www.nsf.gov/awardsearch/showAward?AWD_ID=1541450}{OAC-1541450}.
We would like to thank Celeste R. Brennecka from the Science Writing
Support Service of the University of Münster for her editorial support.

\bibliography{RJreferences}

\address{%
Daniel Nüst\\
University of Münster\\
Institute for Geoinformatics\\ Heisenbergstr. 2\\ 48149 Münster, Germany\\ \orcid{0000-0002-0024-5046}\\
}
\href{mailto:daniel.nuest@uni-muenster.de}{\nolinkurl{daniel.nuest@uni-muenster.de}}

\address{%
Dirk Eddelbuettel\\
University of Illinois at Urbana-Champaign\\
Department of Statistics\\ Illini Hall, 725 S Wright St\\ Champaign, IL 61820, USA\\ \orcid{0000-0001-6419-907X}\\
}
\href{mailto:dirk@eddelbuettel.com}{\nolinkurl{dirk@eddelbuettel.com}}

\address{%
Dom Bennett\\
Gothenburg Global Biodiversity Centre, Sweden\\
Carl Skottsbergs gata 22B\\ 413 19 Göteborg, Sweden\\ \orcid{0000-0003-2722-1359}\\
}
\href{mailto:dominic.john.bennett@gmail.com}{\nolinkurl{dominic.john.bennett@gmail.com}}

\address{%
Robrecht Cannoodt\\
Ghent University\\
Data Mining and Modelling for Biomedicine group\\ VIB Center for Inflammation Research\\ Technologiepark 71\\ 9052 Ghent, Belgium\\ \orcid{0000-0003-3641-729X}\\
}
\href{mailto:robrecht@cannoodt.dev}{\nolinkurl{robrecht@cannoodt.dev}}

\address{%
Dav Clark\\
Gigantum, Inc.\\
1140 3rd Street NE\\ Washington, D.C. 20002, USA\\ \orcid{0000-0002-3982-4416}\\
}
\href{mailto:dav@gigantum.com}{\nolinkurl{dav@gigantum.com}}

\address{%
Gergely Daroczi\\
\\
\orcid{0000-0003-3149-8537}\\
}
\href{mailto:daroczig@rapporter.net}{\nolinkurl{daroczig@rapporter.net}}

\address{%
Mark Edmondson\\
IIH Nordic A/S, Google Developer Expert for GCP\\
\\
}
\href{mailto:mark@markedmondson.me}{\nolinkurl{mark@markedmondson.me}}

\address{%
Colin Fay\\
ThinkR\\
5O rue Arthur Rimbaud\\ 93300 Aubervilliers, France\\ \orcid{0000-0001-7343-1846}\\
}
\href{mailto:contact@colinfay.me}{\nolinkurl{contact@colinfay.me}}

\address{%
Ellis Hughes\\
Fred Hutchinson Cancer Research Center\\
Vaccine and Infectious Disease\\ 1100 Fairview Ave. N., P.O. Box 19024\\ Seattle, WA 98109-1024, USA\\
}
\href{mailto:ehhughes@fredhutch.org}{\nolinkurl{ehhughes@fredhutch.org}}

\address{%
Lars Kjeldgaard\\
Danish Tax Authorities\\
Oestbanegade 123\\ 2100, Koebenhavn Oe\\
}
\href{mailto:lars_kjeldgaard@hotmail.com}{\nolinkurl{lars\_kjeldgaard@hotmail.com}}

\address{%
Sean Lopp\\
RStudio, Inc\\
250 Northern Ave\\ Boston, MA 02210, USA\\
}
\href{mailto:sean@rstudio.com}{\nolinkurl{sean@rstudio.com}}

\address{%
Ben Marwick\\
University of Washington\\
Department of Anthropology\\ Denny Hall 230, Spokane Ln\\ Seattle, WA 98105, USA\\ \orcid{0000-0001-7879-4531}\\
}
\href{mailto:bmarwick@uw.edu}{\nolinkurl{bmarwick@uw.edu}}

\address{%
Heather Nolis\\
T-Mobile\\
12920 Se 38th St.\\ Bellevue, WA, 98006, USA\\
}
\href{mailto:heather.wensler1@t-mobile.com}{\nolinkurl{heather.wensler1@t-mobile.com}}

\address{%
Jacqueline Nolis\\
Nolis, LLC\\
Seattle, WA, USA\\ \orcid{0000-0001-9354-6501}\\
}
\href{mailto:jacqueline@nolisllc.com}{\nolinkurl{jacqueline@nolisllc.com}}

\address{%
Hong Ooi\\
Microsoft\\
Level 5, 4 Freshwater Place\\ Southbank, VIC 3006, Australia\\
}
\href{mailto:hongooi@microsoft.com}{\nolinkurl{hongooi@microsoft.com}}

\address{%
Karthik Ram\\
Berkeley Institute for Data Science\\
University of California\\ Berkeley, CA 94720, USA\\ \orcid{0000-0002-0233-1757}\\
}
\href{mailto:karthik.ram@berkeley.edu}{\nolinkurl{karthik.ram@berkeley.edu}}

\address{%
Noam Ross\\
EcoHealth Alliance\\
460 W 34th St., Ste. 1701\\ New York, NY 10001, USA\\ \orcid{0000-0002-0233-1757}\\
}
\href{mailto:ross@ecohealthalliance.org}{\nolinkurl{ross@ecohealthalliance.org}}

\address{%
Lori Shepherd\\
Roswell Park Comprehensive Cancer Center\\
Elm \& Carlton Streets\\ Buffalo, NY, 14263, USA\\ \orcid{0000-0002-5910-4010}\\
}
\href{mailto:lori.shepherd@roswellpark.org}{\nolinkurl{lori.shepherd@roswellpark.org}}

\address{%
Péter Sólymos\\
Analythium Solutions\\
\#258 150 Chippewa Road\\ Sherwood Park, AB, T8A 6A2, Canada\\ \orcid{0000-0001-7337-1740}\\
}
\href{mailto:peter@analythium.io}{\nolinkurl{peter@analythium.io}}

\address{%
Tyson Lee Swetnam\\
University of Arizona\\
1657 E Helen St.\\ Tucson, AZ, 85721, USA\\ \orcid{0000-0002-6639-7181}\\
}
\href{mailto:tswetnam@arizona.edu}{\nolinkurl{tswetnam@arizona.edu}}

\address{%
Nitesh Turaga\\
Roswell Park Comprehensive Cancer Center\\
Elm \& Carlton Streets\\ Buffalo, NY, 14263, USA\\ \orcid{0000-0002-0224-9817}\\
}
\href{mailto:nitesh.turaga@roswellpark.org}{\nolinkurl{nitesh.turaga@roswellpark.org}}

\address{%
Charlotte Van Petegem\\
Ghent University\\
Department WE02\\ Krijgslaan 281, S9\\ 9000 Gent, Belgium\\ \orcid{0000-0003-0779-4897}\\
}
\href{mailto:charlotte.vanpetegem@ugent.be}{\nolinkurl{charlotte.vanpetegem@ugent.be}}

\address{%
Jason Williams\\
Cold Spring Harbor Laboratory\\
1 Bungtown Rd.\\ Cold Spring Harbor, NY, 11724, USA\\ \orcid{0000-0003-3049-2010}\\
}
\href{mailto:williams@cshl.edu}{\nolinkurl{williams@cshl.edu}}

\address{%
Craig Willis\\
University of Illinois at Urbana-Champaign\\
501 E. Daniel St.\\ Champaign, IL 61820, USA\\ \orcid{0000-0002-6148-7196}\\
}
\href{mailto:willis8@illinois.edu}{\nolinkurl{willis8@illinois.edu}}

\address{%
Nan Xiao\\
Seven Bridges Genomics\\
529 Main St, Suite 6610\\ Charlestown, MA 02129, USA\\ \orcid{0000-0002-0250-5673}\\
}
\href{mailto:me@nanx.me}{\nolinkurl{me@nanx.me}}

### 2020-03-03 README.Rmd
move helper functions to README, add rockerverse.R
preparing for submission, see #26

also add missing used packages to DESCRIPTION

@@
 - https://itsalocke.com/blog/building-an-r-training-environment/ (one single container for many users in teaching, uses rocker/tidyverse)
+
+## Helpers
+
+The [README.Rmd](README.Rmd) file's source includes some useful little code snippets, e.g. count words or get a list of all contributor's GitHub handles.
+
+```{r word_stats, eval=FALSE, include=FALSE}
+# devtools::install_github("benmarwick/wordcountaddin", type = "source", dependencies = TRUE)
+wordcountaddin::readability(here::here("rockerverse.Rmd"))
+wordcountaddin::text_stats(here::here("rockerverse.Rmd"))
+```
+
+```{r preprint_pdf, eval=FALSE, include=FALSE}
+# run this chunk, get the PDF, then revert changes in touched files
+library("rmarkdown")
+library("rticles")
+unlink('manuscript_cache', recursive = TRUE)
+
+# function is a copy of rticles::rjournal_article() but...
+# - using RJWrapper_preprint_template.tex (which includes header.tex commands)
+# - directly referencing unexported functions from rticles
+# - direclty setting the keep_tex option
+rjournal_style_preprint <- function (..., citation_package = "natbib") {
+    rmarkdown::pandoc_available("2.2", TRUE)
+    base <- rticles:::pdf_document_format("rjournal_article", highlight = NULL,
+        citation_package = citation_package, ...)
+    base$pandoc$to <- "latex"
+    base$pandoc$ext <- ".tex"
+    base$pandoc$keep_tex <- TRUE # added
+    base$post_processor <- function(metadata, utf8_input, output_file,
+        clean, verbose) {
+      filename <- basename(output_file)
+      if (filename != (filename2 <- gsub("_", "-", filename))) {
+          file.rename(filename, filename2)
+          filename <- filename2
+      }
+      m <- list(filename = xfun::sans_ext(filename))
+      h <- rticles:::get_list_element(metadata, c("output", "rticles::rjournal_article",
+          "includes", "in_header"))
+      h <- c(h, if (length(preamble <- unlist(metadata[c("preamble",
+          "header-includes")]))) {
+          f <- tempfile(fileext = ".tex")
+          on.exit(unlink(f), add = TRUE)
+          xfun::write_utf8(preamble, f)
+          f
+      })
+      #t <- find_resource("rjournal_article", "RJwrapper.tex")
+      t <- here::here("RJwrapper_preprint_template.tex")
+      rticles:::template_pandoc(m, t, "RJwrapper.tex", h, verbose)
+      tinytex::latexmk("RJwrapper.tex", base$pandoc$latex_engine,
+          clean = clean)
+    }
+    base$knitr$opts_chunk$comment <- "#>"
+    hilight_source <- rticles:::knitr_fun("hilight_source")
+    hook_chunk = function(x, options) {
+        if (rticles:::output_asis(x, options))
+            return(x)
+        paste0("```{=latex}\n\\begin{Schunk}\n", x, "\\end{Schunk}\n```")
+    }
+    hook_input <- function(x, options) paste(c("\\begin{Sinput}",
+        hilight_source(x, "sweave", options), "\\end{Sinput}",
+        ""), collapse = "\n")
+    hook_output <- function(x, options) paste("\\begin{Soutput}\n",
+        x, "\\end{Soutput}\n", sep = "")
+    base$knitr$knit_hooks <- rticles:::merge_list(base$knitr$knit_hooks,
+        list(chunk = hook_chunk, source = hook_input, output = hook_output,
+            message = hook_output, warning = hook_output, plot = knitr::hook_plot_tex))
+    base
+}
+
+rmarkdown::render("rockerverse.Rmd", output_format = rjournal_style_preprint())
+file.rename("RJwrapper.pdf", "rockerverse-preprint.pdf") # file for own archival
+
+# For submission to arXiv:
+# 1. Copy the following files into a new folder:
+# - figures/*
+# - rockerverse.tex
+# - RJournal.sty
+# - RJreferences.bib
+# - RJwrapper.tex
+# 2. Run pdfLaTeX directly on RJwrapper.tex (works with RStudio)
+# 3. Zip the following files and submit to arXiv:
+# - /figures
+# - rockerverse.tex
+# - RJournal.sty
+# - RJwrapper.bbl
+# - RJwrapper.tex
+```
+
+```{r collab_emails, eval=FALSE, include=FALSE}
+front_matter <- rmarkdown::yaml_front_matter("rockerverse.Rmd")
+emails <- sapply(X = front_matter$author, FUN = function(a) { a$email })
+cat(emails, sep = "\n")
+```
+
+```{bash collab_handles, eval=FALSE, include=FALSE}
+cat rockerverse.Rmd | grep '    # https://github.com/' | sed 's|    # https://github.com/|@|'
+```

### 2020-03-03 rockerverse.Rmd
move helper functions to README, add rockerverse.R
preparing for submission, see #26

also add missing used packages to DESCRIPTION

--- 2020-03-03 manuscript.Rmd
@@
 
-```{r clientfunctionalities, echo=FALSE, results='asis', warning=FALSE, message=FALSE}
+```{r clients_table, echo=FALSE, results='asis', warning=FALSE, message=FALSE}
 library("tibble")
@@
 The Docker CLI's basic functionality, although it evolves quickly and with little concern for avoiding breaking changes, has remained unchanged in core functions, meaning that a core function such as `harbor::docker_run(image = "hello-world")` still works depsite its stopped development.
-```{r harbor, eval=FALSE, include=FALSE}
+
+<!-- TRY OUT harbor
 remotes::install_github("wch/harbor")
 harbor::docker_run(image = "hello-world")
-```
+-->
 
@@
 It potentially lowers the barriers for R users to run containers on various hosts if they perceive that using the Docker&nbsp;Machine CLI directly as a barrier and it enables scripted workflows with remote processing.
-```{r dockermachine, eval=FALSE, include=FALSE}
+
+<!-- TRY OUT dockermachine
 library("dockermachine")
@@
 # Instance shut down.
-```
+-->
 
@@
 This complexity and an inherent concern for openness and reproducibility in the data science community has led to Docker being used widely.
-<!--A general description of what to consider when creating a Docker image used for data science projects in R is provided on [environments.rstudio.com](https://environments.rstudio.com/docker).-->
 The remainder of this section presents example Docker images and image stacks featuring R intended for data science.
@@
 The final extracted image size is over 25&nbsp;GB, which calls into question whether having everything available is actually convenient.
-<!--As a further example, -->
 - The [_Radiant_](https://radiant-rstats.github.io/docs/) project provides several images, e.g., [`vnijs/rsm-msba-spark`](https://hub.docker.com/r/vnijs/rsm-msba-spark), for their browser-based business analytics interface based on \CRANpkg{Shiny} \citep{cran_shiny}, and for use in education as part of an MSc course\footnote{`Dockerfile` available on GitHub: \href{https://github.com/radiant-rstats/docker}{https://github.com/radiant-rstats/docker}.}.
@@
 
-```{bash kaggle_images, eval=FALSE, include=FALSE}
+<!-- CHECK IMAGES SIZES
 docker pull gcr.io/kaggle-images/rcran
@@
 docker images | grep kaggle
-```
 
-```{bash radiant_images, eval=FALSE, include=FALSE}
 docker pull vnijs/rsm-msba-spark
 docker images | grep vnijs
-```
+-->
 
@@
 \href{https://github.com/r-spatial/sf/tree/master/inst/docker}{https://github.com/r-spatial/sf/tree/master/inst/docker}, \href{https://github.com/Nowosad/rspatial_proj6}{https://github.com/Nowosad/rspatial\_proj6}, and \href{https://github.com/r-spatial/sf/issues/1231}{https://github.com/r-spatial/sf/issues/1231}}.
-```{r geospatial_deps, eval=FALSE, include=FALSE}
+
+<!-- GET THE NUMBER OF REVERSE IMPORTS AND DEPS FOR rgdal AND sf
 length(tools::dependsOnPkgs(c("rgdal", "sf"), dependencies = c("most"), recursive = TRUE))
-```
+-->
 
@@
 Outsider modules can be hosted code-sharing services, e.g., on GitHub, and \pkg{outsider} contains discovery functions for them.
-<!-- informative examples of how modules work, just as a reminder for @nuest
-https://github.com/DomBennett/om..astral/blob/master/examples/astral.R
-https://github.com/DomBennett/om..raxml#partitioned-dna-analysis
--->
 
@@
 Also, the ease with which one can create a complex software systems with Docker, such as an independent Docker image stack, to serve one's specific needs leads to parallel developments.
-<!-- Tyler W./DC from Gigantum: ideas about ease-of-DIY together with difficulty of composition; shared starting points of Rocker -->
 This ease-of-DIY in combination with the difficulty of reusing parts from or composing multiple `Dockerfile`s is a further reason for **fragmentation**.
@@
 
-```{r word_stats, eval=FALSE, include=FALSE}
-# devtools::install_github("benmarwick/wordcountaddin", type = "source", dependencies = TRUE)
-wordcountaddin::readability(here::here("manuscript.Rmd"))
-wordcountaddin::text_stats(here::here("manuscript.Rmd"))
+\bibliography{RJreferences}
+
+```{r purl_code, eval=TRUE, include=FALSE}
+# generate the reproducible R script every time the R Markdown document is compiled
+library("callr")
+callr::r_safe(function() { knitr::purl("rockerverse.Rmd") })
 ```
-
-```{r preprint_pdf, eval=FALSE, include=FALSE}
-# run this chunk, get the PDF, then revert changes in touched files
-library("rmarkdown")
-library("rticles")
-unlink('manuscript_cache', recursive = TRUE)
-
-# function is a copy of rticles::rjournal_article() but...
-# - using RJWrapper_preprint_template.tex (which includes header.tex commands)
-# - directly referencing unexported functions from rticles
-# - direclty setting the keep_tex option
-rjournal_style_preprint <- function (..., citation_package = "natbib") {
-    rmarkdown::pandoc_available("2.2", TRUE)
-    base <- rticles:::pdf_document_format("rjournal_article", highlight = NULL,
-        citation_package = citation_package, ...)
-    base$pandoc$to <- "latex"
-    base$pandoc$ext <- ".tex"
-    base$pandoc$keep_tex <- TRUE # added
-    base$post_processor <- function(metadata, utf8_input, output_file,
-        clean, verbose) {
-      filename <- basename(output_file)
-      if (filename != (filename2 <- gsub("_", "-", filename))) {
-          file.rename(filename, filename2)
-          filename <- filename2
-      }
-      m <- list(filename = xfun::sans_ext(filename))
-      h <- rticles:::get_list_element(metadata, c("output", "rticles::rjournal_article",
-          "includes", "in_header"))
-      h <- c(h, if (length(preamble <- unlist(metadata[c("preamble",
-          "header-includes")]))) {
-          f <- tempfile(fileext = ".tex")
-          on.exit(unlink(f), add = TRUE)
-          xfun::write_utf8(preamble, f)
-          f
-      })
-      #t <- find_resource("rjournal_article", "RJwrapper.tex")
-      t <- here::here("RJwrapper_preprint_template.tex")
-      rticles:::template_pandoc(m, t, "RJwrapper.tex", h, verbose)
-      tinytex::latexmk("RJwrapper.tex", base$pandoc$latex_engine,
-          clean = clean)
-    }
-    base$knitr$opts_chunk$comment <- "#>"
-    hilight_source <- rticles:::knitr_fun("hilight_source")
-    hook_chunk = function(x, options) {
-        if (rticles:::output_asis(x, options))
-            return(x)
-        paste0("```{=latex}\n\\begin{Schunk}\n", x, "\\end{Schunk}\n```")
-    }
-    hook_input <- function(x, options) paste(c("\\begin{Sinput}",
-        hilight_source(x, "sweave", options), "\\end{Sinput}",
-        ""), collapse = "\n")
-    hook_output <- function(x, options) paste("\\begin{Soutput}\n",
-        x, "\\end{Soutput}\n", sep = "")
-    base$knitr$knit_hooks <- rticles:::merge_list(base$knitr$knit_hooks,
-        list(chunk = hook_chunk, source = hook_input, output = hook_output,
-            message = hook_output, warning = hook_output, plot = knitr::hook_plot_tex))
-    base
-}
-
-rmarkdown::render("manuscript.Rmd", output_format = rjournal_style_preprint())
-file.rename("RJwrapper.pdf", "rockerverse-preprint.pdf") # file for own archival
-
-# For submission to arXiv:
-# 1. Copy the following files into a new folder:
-# - figures/*
-# - manuscript.tex
-# - RJournal.sty
-# - RJreferences.bib
-# - RJwrapper.tex
-# 2. Run pdfLaTeX directly on RJwrapper.tex (works with RStudio)
-# 3. Zip the following files and submit to arXiv:
-# - /figures
-# - manuscript.tex
-# - RJournal.sty
-# - RJwrapper.bbl
-# - RJwrapper.tex
-```
-
-```{r collab_emails, eval=FALSE, include=FALSE}
-front_matter <- rmarkdown::yaml_front_matter("manuscript.Rmd")
-emails <- sapply(X = front_matter$author, FUN = function(a) { a$email })
-cat(emails, sep = "\n")
-```
-
-```{bash collab_handles, eval=FALSE, include=FALSE}
-cat manuscript.Rmd | grep '    # https://github.com/' | sed 's|    # https://github.com/|@|'
-```
-
-\bibliography{RJreferences}

### 2020-03-03 rockerverse.R

@@ -1,0 +1,78 @@
+## ----setup, include=FALSE-----------------------------------------------------
+Sys.setlocale('LC_ALL','C')
+
+
+## ----clients_table, echo=FALSE, results='asis', warning=FALSE, message=FALSE----
+library("tibble")
+library("tidyr")
+library("dplyr")
+
+functionalities <- tribble(
+  ~id, ~Functionality,
+  "dockerfile", "Generate a Dockerfile",
+  "build", "Build an image",
+  "execute", "Execute a container locally or remotely",
+  "cloud", "Deploy or manage instances in the cloud",
+  "interact", "Interact with an instance (e.g., file transfer)",
+  "images", "Manage storage of images",
+  "othercontainers", "Supports Docker and Singularity",
+  "api", "Direct access to Docker API instead of using the CLI",
+  "install", "Installing Docker software"
+)
+
+# some functionalities might be missing. please add them if you find that any are missing!
+package_functionalities <- tribble(
+  ~package, ~id,
+  "harbor", c("execute", "cloud", "images"),
+  "stevedore", c("api", "execute", "cloud", "images", "interact"), # build?
+  "AzureContainers", c("build", "execute", "cloud"),
+  "babelwhale", c("execute", "interact", "othercontainers"),
+  "dockyard", c("dockerfile", "build", "execute"), # cloud?
+  "dockermachine", c("install", "execute", "cloud", "interact")
+)
+
+tab <-
+  full_join(
+  functionalities %>% crossing(package = package_functionalities$package),
+  package_functionalities %>% unnest(id) %>% mutate(check = "\\checkmark"),#\u2714 # \\checkmark # ✓
+  by = c("id", "package")
+) %>%
+  mutate(check = ifelse(is.na(check), "", check)) %>%
+  spread(package, check) %>%
+  slice(match(functionalities$id, id)) %>%
+  select(-id)
+
+# insert table, position caption below as described in https://github.com/haozhu233/kableExtra/issues/221#issuecomment-400473530
+library("kableExtra")
+library("xtable")
+#knitr::kable(tab,
+#             format = "latex",
+#             caption = "R packages with Docker client functionality.",
+#             label = "clients",
+#             escape = FALSE) %>%
+#  kableExtra::row_spec(0, angle = -90)
+xtable2kable <- function(x) {
+  label(x) <- "tab:clients"
+  out <- capture.output(print(x, 
+                              table.placement = NULL,
+                              booktabs = TRUE,
+                              caption.placement = "bottom",
[…]

### 2020-03-03 rockerverse.Rmd
use \code{..} where appropriate, use DOI URLs
preparing for submission, #26

@@
 The outputs of container executions are converted to appropriate R types.
-The Docker CLI's basic functionality, although it evolves quickly and with little concern for avoiding breaking changes, has remained unchanged in core functions, meaning that a core function such as `harbor::docker_run(image = "hello-world")` still works depsite its stopped development.
+The Docker CLI's basic functionality, although it evolves quickly and with little concern for avoiding breaking changes, has remained unchanged in core functions, meaning that a core function such as \code{harbor::docker\_run(image = "hello-world")} still works depsite its stopped development.
 
@@
 \CRANpkg{stevedore}'s own interface is automatically generated based on the OpenAPI specification of the Docker daemon, but it is still similar to the Docker CLI.
-The interface is similar to R6 objects, in that a `stevedore_object` has a number of functions attached to it that can be called, and multiple specific versions of the Docker API can be supported thanks to the automatic generation\footnote{See \href{https://github.com/richfitz/stevedore/blob/master/development.md}{https://github.com/richfitz/stevedore/blob/master/development.md}.}.
+The interface is similar to R6 objects, in that an object of class \code{"stevedore\_object"} has a number of functions attached to it that can be called, and multiple specific versions of the Docker API can be supported thanks to the automatic generation\footnote{See \href{https://github.com/richfitz/stevedore/blob/master/development.md}{https://github.com/richfitz/stevedore/blob/master/development.md}.}.
 
@@
 **\pkg{dockyard}** ([https://github.com/thebioengineer/dockyard](https://github.com/thebioengineer/dockyard)) has the goal of lowering the barrier to creating `Dockerfile`s, building Docker images, and deploying Docker containers.
-The package follows the increasingly used piping paradigm of the `tidyverse` style of programming for chaining R functions representing the instructions in a `Dockerfile`.
+The package follows the increasingly used piping paradigm of the Tidyverse-style \citep{wickham_welcome_2019} of programming for chaining R functions representing the instructions in a `Dockerfile`.
 An existing `Dockerfile` can be used as a template.
@@
 This image contains the build system environment variables and the _system dependencies_ needed to install and check almost all (1813 out of 1823) _Bioconductor_ software packages, and it saves users and package developers from managing these themselves.
-Furthermore, the image is configured so that `.libPaths()` has `/usr/local/lib/R/host-site-library` as the first location.
+Furthermore, the image is configured so that \code{.libPaths()} has \samp{/usr/local/lib/R/host-site-library} as the first location.
 Users mounting a location on the host file system to this location can persistently manage installed packages across Docker containers or image updates.
@@
 **\pkg{containerit}** ([https://github.com/o2r-project/containerit/](https://github.com/o2r-project/containerit/)) attempts to take this one step further and includes these tools to automatically create a `Dockerfile` that can execute a given workflow \citep{nust_containerit_2019}.
-\pkg{containerit} accepts R `sessionInfo` objects as input and provides helper functions to derive these from workflows, e.g., an R script or R Markdown document, by analysing the session state at the end of the workflow.
+\pkg{containerit} accepts an R object of classes \code{"sessionInfo"} or \code{"session\_info"} as input and provides helper functions to derive these from workflows, e.g., an R script or R Markdown document, by analysing the session state at the end of the workflow.
 It relies on the \pkg{sysreqs} ([https://github.com/r-hub/sysreqs/](https://github.com/r-hub/sysreqs/)) package and it's mapping of package system dependencies to platform specific installation package names\footnote{See \href{https://sysreqs.r-hub.io/}{https://sysreqs.r-hub.io/}.}.
@@
 It introduces new fields to the R Markdown document header, allowing users to manually declare the dependencies, including versions, for rendering the document.
-The package then generates a `Dockerfile` from this metadata and provides a utitility function to render the document inside a Docker container, i.e., `render_docker("foo.Rmd")`.
+The package then generates a `Dockerfile` from this metadata and provides a utitility function to render the document inside a Docker container, i.e., \code{render\_docker("foo.Rmd")}.
 An RStudio addin even allows compilation of documents with the single push of a button.
@@
 For example, the package can [schedule jobs with Docker Swarm](https://mllg.github.io/batchtools/reference/makeClusterFunctionsDocker.html).
-**\CRANpkg{googleComputeEngineR}** has the function `gce_vm_cluster()` to create clusters of 2 or more virtual machines, running multi-CPU architectures \citep{googleComputeEngineR_2019}.
+**\CRANpkg{googleComputeEngineR}** has the function \code{gce\_vm\_cluster()} to create clusters of 2 or more virtual machines, running multi-CPU architectures \citep{googleComputeEngineR_2019}.
 Instead of running a local R script with the local CPU and RAM restrictions, the same code can be processed on all CPU threads of the cluster of machines in the cloud, all running a Docker container with the same R environments.
@@
 \CRANpkg{drake} has been demonstrated to run inside containers for high reproducibility\footnote{See for example \href{https://github.com/joelnitta/pleurosoriopsis}{https://github.com/joelnitta/pleurosoriopsis} or \href{https://gitlab.com/ecohealthalliance/drake-gitlab-docker-example}{https://gitlab.com/ecohealthalliance/drake-gitlab-docker-example}, the latter even running in a continuous integration platform (cf.~\nameref{development}.}.
-Furthermore, \CRANpkg{drake} workflows have been shown to use \CRANpkg{future} package's function `makeClusterPSOCK()` for sending parts of the workflow to a Docker image for execution\footnote{\href{https://docs.ropensci.org/drake/index.html?q=docker\#with-docker}{https://docs.ropensci.org/drake/index.html?q=docker\#with-docker}} \citep[see package's function documentation;~][]{future_2020}.
+Furthermore, \CRANpkg{drake} workflows have been shown to use \CRANpkg{future} package's function \code{makeClusterPSOCK()} for sending parts of the workflow to a Docker image for execution\footnote{\href{https://docs.ropensci.org/drake/index.html?q=docker\#with-docker}{https://docs.ropensci.org/drake/index.html?q=docker\#with-docker}} \citep[see package's function documentation;~][]{future_2020}.
 In the latter case, the container control code must be written by the user, and the \CRANpkg{future} package ensures that the host and worker can connect for communicating over socket connections.
@@
 
-Another example is the package \CRANpkg{golem}, which makes an heavy use of `dockerfiler` when it comes to creating the `Dockerfile` for building and deploying production-grade Shiny applications \citep{cran_golem}.
+Another example is the package \CRANpkg{golem}, which makes an heavy use of \pkg{dockerfiler} when it comes to creating the `Dockerfile` for building and deploying production-grade Shiny applications \citep{cran_golem}.
 \CRANpkg{googleComputeEngineR} enables quick deployments of key R services, such as RStudio and Shiny, onto cloud virtual machines (VMs) with Google Cloud Compute Engine \citep{googleComputeEngineR_2019}.
 The package utilises `Dockerfile`s to move the labour of setting up those services from the user to a premade Docker image, which is configured and run in the cloud VM.
-For example, by specifying the template `template="rstudio"` in functions `gce_vm_template()/gce_vm()` an up-to-date RStudio Server image is launched for development work, whereas specifying `template="rstudio-gpu"` will launch an RStudio Server image with a GPU attached, etc.
+For example, by specifying the template `template="rstudio"` in functions \code{gce\_vm\_template()} and \code{gce\_vm()} an up-to-date RStudio Server image is launched for development work, whereas specifying `template="rstudio-gpu"` will launch an RStudio Server image with a GPU attached, etc.
 

### 2020-03-03 rockerverse.tex
move helper functions to README, add rockerverse.R
preparing for submission, see #26

also add missing used packages to DESCRIPTION
use \code{..} where appropriate, use DOI URLs
preparing for submission, #26

@@
 a core function such as
-\texttt{harbor::docker\_run(image\ =\ "hello-world")} still works
-depsite its stopped development.
+\code{harbor::docker\_run(image = "hello-world")} still works depsite
+its stopped development.
 
@@
 daemon, but it is still similar to the Docker CLI. The interface is
-similar to R6 objects, in that a \texttt{stevedore\_object} has a number
-of functions attached to it that can be called, and multiple specific
-versions of the Docker API can be supported thanks to the automatic
+similar to R6 objects, in that an object of class
+\code{"stevedore\_object"} has a number of functions attached to it that
+can be called, and multiple specific versions of the Docker API can be
+supported thanks to the automatic
 generation\footnote{See \href{https://github.com/richfitz/stevedore/blob/master/development.md}{https://github.com/richfitz/stevedore/blob/master/development.md}.}.
@@
 images, and deploying Docker containers. The package follows the
-increasingly used piping paradigm of the \texttt{tidyverse} style of
-programming for chaining R functions representing the instructions in a
-\texttt{Dockerfile}. An existing \texttt{Dockerfile} can be used as a
-template. \pkg{dockyard} also includes wrappers for common steps, such
-as installing an R package or copying files, and provides built-in
-functions for building an image and running a container, which make
-Docker more approachable within a single R-based user interface.
+increasingly used piping paradigm of the Tidyverse-style
+\citep{wickham_welcome_2019} of programming for chaining R functions
+representing the instructions in a \texttt{Dockerfile}. An existing
+\texttt{Dockerfile} can be used as a template. \pkg{dockyard} also
+includes wrappers for common steps, such as installing an R package or
+copying files, and provides built-in functions for building an image and
+running a container, which make Docker more approachable within a single
+R-based user interface.
 
@@
 it saves users and package developers from managing these themselves.
-Furthermore, the image is configured so that \texttt{.libPaths()} has
-\texttt{/usr/local/lib/R/host-site-library} as the first location. Users
+Furthermore, the image is configured so that \code{.libPaths()} has
+\samp{/usr/local/lib/R/host-site-library} as the first location. Users
 mounting a location on the host file system to this location can
@@
 \texttt{Dockerfile} that can execute a given workflow
-\citep{nust_containerit_2019}. \pkg{containerit} accepts R
-\texttt{sessionInfo} objects as input and provides helper functions to
-derive these from workflows, e.g., an R script or R Markdown document,
-by analysing the session state at the end of the workflow. It relies on
-the \pkg{sysreqs} (\url{https://github.com/r-hub/sysreqs/}) package and
-it's mapping of package system dependencies to platform specific
-installation package
+\citep{nust_containerit_2019}. \pkg{containerit} accepts an R object of
+classes \code{"sessionInfo"} or \code{"session\_info"} as input and
+provides helper functions to derive these from workflows, e.g., an R
+script or R Markdown document, by analysing the session state at the end
+of the workflow. It relies on the \pkg{sysreqs}
+(\url{https://github.com/r-hub/sysreqs/}) package and it's mapping of
+package system dependencies to platform specific installation package
 names\footnote{See \href{https://sysreqs.r-hub.io/}{https://sysreqs.r-hub.io/}.}.
@@
 provides a utitility function to render the document inside a Docker
-container, i.e., \texttt{render\_docker("foo.Rmd")}. An RStudio addin
-even allows compilation of documents with the single push of a button.
+container, i.e., \code{render\_docker("foo.Rmd")}. An RStudio addin even
+allows compilation of documents with the single push of a button.
 
@@
 jobs with Docker Swarm}. \textbf{\CRANpkg{googleComputeEngineR}} has the
-function \texttt{gce\_vm\_cluster()} to create clusters of 2 or more
+function \code{gce\_vm\_cluster()} to create clusters of 2 or more
 virtual machines, running multi-CPU architectures
@@
 Furthermore, \CRANpkg{drake} workflows have been shown to use
-\CRANpkg{future} package's function \texttt{makeClusterPSOCK()} for
+\CRANpkg{future} package's function \code{makeClusterPSOCK()} for
 sending parts of the workflow to a Docker image for
@@
 Another example is the package \CRANpkg{golem}, which makes an heavy use
-of \texttt{dockerfiler} when it comes to creating the
-\texttt{Dockerfile} for building and deploying production-grade Shiny
-applications \citep{cran_golem}. \CRANpkg{googleComputeEngineR} enables
-quick deployments of key R services, such as RStudio and Shiny, onto
-cloud virtual machines (VMs) with Google Cloud Compute Engine
+of \pkg{dockerfiler} when it comes to creating the \texttt{Dockerfile}
+for building and deploying production-grade Shiny applications
+\citep{cran_golem}. \CRANpkg{googleComputeEngineR} enables quick
+deployments of key R services, such as RStudio and Shiny, onto cloud
+virtual machines (VMs) with Google Cloud Compute Engine
 \citep{googleComputeEngineR_2019}. The package utilises
@@
 the cloud VM. For example, by specifying the template
-\texttt{template="rstudio"} in functions
-\texttt{gce\_vm\_template()/gce\_vm()} an up-to-date RStudio Server
-image is launched for development work, whereas specifying
-\texttt{template="rstudio-gpu"} will launch an RStudio Server image with
-a GPU attached, etc.
+\texttt{template="rstudio"} in functions \code{gce\_vm\_template()} and
+\code{gce\_vm()} an up-to-date RStudio Server image is launched for
+development work, whereas specifying \texttt{template="rstudio-gpu"}
+will launch an RStudio Server image with a GPU attached, etc.
 

### 2020-03-03 rockerverse.Rmd
only first mention uses \CRANpkg
preparing for submission, see #26

@@
 Furthermore using the API gives access to information in a structured way, is system independent, and is likely more reliable than parsing command line output.
-\CRANpkg{stevedore}'s own interface is automatically generated based on the OpenAPI specification of the Docker daemon, but it is still similar to the Docker CLI.
+\pkg{stevedore}'s own interface is automatically generated based on the OpenAPI specification of the Docker daemon, but it is still similar to the Docker CLI.
 The interface is similar to R6 objects, in that an object of class \code{"stevedore\_object"} has a number of functions attached to it that can be called, and multiple specific versions of the Docker API can be supported thanks to the automatic generation\footnote{See \href{https://github.com/richfitz/stevedore/blob/master/development.md}{https://github.com/richfitz/stevedore/blob/master/development.md}.}.
@@
 It relies on the \pkg{sysreqs} ([https://github.com/r-hub/sysreqs/](https://github.com/r-hub/sysreqs/)) package and it's mapping of package system dependencies to platform specific installation package names\footnote{See \href{https://sysreqs.r-hub.io/}{https://sysreqs.r-hub.io/}.}.
-\pkg{containerit} uses \CRANpkg{stevedore} to streamline the user interaction and improve the created `Dockerfile`s, e.g., by running a container for the desired base image to extract the already available R packages.
+\pkg{containerit} uses \pkg{stevedore} to streamline the user interaction and improve the created `Dockerfile`s, e.g., by running a container for the desired base image to extract the already available R packages.
 
@@
 The irreproducibility of authoring environments can become an issue for collaborative documents and large-scale platforms for processing documents.
-\CRANpkg{liftr} makes the document the main and sole workflow control file and the only file that needs to be shared between collaborators for consistent environments, e.g.
+\pkg{liftr} makes the document the main and sole workflow control file and the only file that needs to be shared between collaborators for consistent environments, e.g.
 demonstrated in the DockFlow project (https://dockflow.org).
@@
 Regarding R package management, this can be achieved with MRAN, or with packages such as \CRANpkg{checkpoint} \citep{cran_checkpoint} and \CRANpkg{renv} \citep{renv2019}, which can naturally be applied within images and containers.
-For example, \CRANpkg{renv} helps users to manage the state of the R library in a reproducible way, further providing isolation and portability.
-While \CRANpkg{renv} does not cover system dependencies, the \CRANpkg{renv}-based environment can be transferred into a container either by restoring the environment based on the main configuration file `renv.lock` or by storing the \CRANpkg{renv}-cache on the host and not in the container \citep{ushey_using_2019}.
+For example, \pkg{renv} helps users to manage the state of the R library in a reproducible way, further providing isolation and portability.
+While \pkg{renv} does not cover system dependencies, the \pkg{renv}-based environment can be transferred into a container either by restoring the environment based on the main configuration file `renv.lock` or by storing the \pkg{renv}-cache on the host and not in the container \citep{ushey_using_2019}.
 With both the system dependencies and R packages consciously managed in a Docker image, users can start using containers as the _only_ environment for their workflows, which allows them to work indepenently of physical computers\footnote{Allowing them to be digital "nomads", cf.
@@
 GDAL is a system dependency allowing R packages such as \CRANpkg{sf}, which provides the core data model for geospatial vector data, or \CRANpkg{rgdal}, to accommodate users to be able to read and write hundreds of different spatial raster and vector formats \citep{pebesma_simple_2018,cran_rgdal}.
-\CRANpkg{sf} and \CRANpkg{rgdal} have hundreds of indirect reverse imports and dependencies and, therefore, the maintainers spend a lot of effort trying not to break them.
+\pkg{sf} and \pkg{rgdal} have hundreds of indirect reverse imports and dependencies and, therefore, the maintainers spend a lot of effort trying not to break them.
 Purpose-built Docker images are used to prepare for upcoming releases of system libraries, individual bug reports, and for the lowest supported versions of system libraries\footnote{Cf.
@@
 Instead of running a local R script with the local CPU and RAM restrictions, the same code can be processed on all CPU threads of the cluster of machines in the cloud, all running a Docker container with the same R environments.
-\CRANpkg{googleComputeEngineR} integrates with the R parallelisation package \CRANpkg{future} \citep{cran_future} to enable this with only a few lines or R code\footnote{\href{https://cloudyr.github.io/googleComputeEngineR/articles/massive-parallel.html}{https://cloudyr.github.io/googleComputeEngineR/articles/massive-parallel.html}}.
+\pkg{googleComputeEngineR} integrates with the R parallelisation package \CRANpkg{future} \citep{cran_future} to enable this with only a few lines or R code\footnote{\href{https://cloudyr.github.io/googleComputeEngineR/articles/massive-parallel.html}{https://cloudyr.github.io/googleComputeEngineR/articles/massive-parallel.html}}.
 [Google Cloud Run](https://cloud.run) is a CaaS (Containers as a Service) platform.
@@
 It features implicit parallel computing and automated detection of the parts of the work that actually needs to be re-executed.
-\CRANpkg{drake} has been demonstrated to run inside containers for high reproducibility\footnote{See for example \href{https://github.com/joelnitta/pleurosoriopsis}{https://github.com/joelnitta/pleurosoriopsis} or \href{https://gitlab.com/ecohealthalliance/drake-gitlab-docker-example}{https://gitlab.com/ecohealthalliance/drake-gitlab-docker-example}, the latter even running in a continuous integration platform (cf.~\nameref{development}.}.
-Furthermore, \CRANpkg{drake} workflows have been shown to use \CRANpkg{future} package's function \code{makeClusterPSOCK()} for sending parts of the workflow to a Docker image for execution\footnote{\href{https://docs.ropensci.org/drake/index.html?q=docker\#with-docker}{https://docs.ropensci.org/drake/index.html?q=docker\#with-docker}} \citep[see package's function documentation;~][]{future_2020}.
-In the latter case, the container control code must be written by the user, and the \CRANpkg{future} package ensures that the host and worker can connect for communicating over socket connections.
+\pkg{drake} has been demonstrated to run inside containers for high reproducibility\footnote{See for example \href{https://github.com/joelnitta/pleurosoriopsis}{https://github.com/joelnitta/pleurosoriopsis} or \href{https://gitlab.com/ecohealthalliance/drake-gitlab-docker-example}{https://gitlab.com/ecohealthalliance/drake-gitlab-docker-example}, the latter even running in a continuous integration platform (cf.~\nameref{development}.}.
+Furthermore, \pkg{drake} workflows have been shown to use \pkg{future} package's function \code{makeClusterPSOCK()} for sending parts of the workflow to a Docker image for execution\footnote{\href{https://docs.ropensci.org/drake/index.html?q=docker\#with-docker}{https://docs.ropensci.org/drake/index.html?q=docker\#with-docker}} \citep[see package's function documentation;~][]{future_2020}.
+In the latter case, the container control code must be written by the user, and the \pkg{future} package ensures that the host and worker can connect for communicating over socket connections.
 **RStudio Server Pro** includes a functionality called [Launcher](https://solutions.rstudio.com/launcher/overview/) (since version&nbsp;1.2, released in 2019).
@@
 
-First, different packages use containers for the **deployment of R and \CRANpkg{Shiny} apps**.
-\CRANpkg{Shiny} is a popular package for creating interactive online dashboards with R, and it enables users with very diverse backgrounds to create stable and user-friendly web applications \citep{cran_shiny}.
+First, different packages use containers for the **deployment of R and \pkg{Shiny} apps**.
+\pkg{Shiny} is a popular package for creating interactive online dashboards with R, and it enables users with very diverse backgrounds to create stable and user-friendly web applications \citep{cran_shiny}.
 _ShinyProxy_ ([https://www.shinyproxy.io/](https://www.shinyproxy.io/)) is an open source tool to deploy Shiny apps in an enterprise context, where it features single sign-on, but it can also be used in scientific use cases \citep[e.g., ][]{savini_epiexplorer_2019,glouzon_structurexplor_2017}.
@@
 These functions are translated to Docker CLI commands and transferred transparently to the respective remove machine using SSH.
-\CRANpkg{AzureContainers} is an umbrella package that provides interfaces for three commercial services of Microsoft's Azure Cloud, namely [Container Instances](https://azure.microsoft.com/en-us/services/container-instances/) for running individual containers, [Container Registry](https://azure.microsoft.com/en-us/services/container-registry/) for private image distribution, and [Kubernetes Service](https://azure.microsoft.com/en-us/services/kubernetes-service/) for orchestrated deployments.
-While a package like \CRANpkg{plumber} provides the infrastructure for turning an R workflow into a web service, for production purposes it is usually necessary to take into account scalability, reliability and ease of management.
+\pkg{AzureContainers} is an umbrella package that provides interfaces for three commercial services of Microsoft's Azure Cloud, namely [Container Instances](https://azure.microsoft.com/en-us/services/container-instances/) for running individual containers, [Container Registry](https://azure.microsoft.com/en-us/services/container-registry/) for private image distribution, and [Kubernetes Service](https://azure.microsoft.com/en-us/services/kubernetes-service/) for orchestrated deployments.
+While a package like \pkg{plumber} provides the infrastructure for turning an R workflow into a web service, for production purposes it is usually necessary to take into account scalability, reliability and ease of management.
 AzureContainers provides an R-based interface to these features and, thereby, simplifies complex infrastructure management to a number of R function calls, given an Azure account with sufficient credit\footnote{See \emph{"Deploying a prediction service with Plumber"} vignette for details:  \href{https://cran.r-project.org/web/packages/AzureContainers/vignettes/vig01_plumber_deploy.html}{https://cran.r-project.org/web/packages/AzureContainers/vignettes/vig01\_plumber\_deploy.html}.}.
@@
 The watchdog marshals an HTTP request and invokes the actual application.
-The R session uses \CRANpkg{plumber} or similar packages for the API endpoint with packages and data preloaded, thus minimizing response times.
+The R session uses \pkg{plumber} or similar packages for the API endpoint with packages and data preloaded, thus minimizing response times.
 
@@
 However, thanks to advancements in R running as a web service, along with the ability to deploy R in Docker containers, modern enterprises are now capable of having real-time machine learning powered by R.
-A number of packages and projects have enabled R to respond to client requests over TCP/IP and local socket servers, such as \CRANpkg{Rserve} \citep{cran_rserve}, \CRANpkg{svSocket} \citep{grosjean_sciviews_2019}, [rApache](http://www.rapache.net) and more recently \CRANpkg{plumber} ([https://www.rplumber.io/](https://www.rplumber.io/)) and \pkg{RestRserve} ([http://restrserve.org](http://restrserve.org)), which even processes incoming requests in parallel with forked processes using \CRANpkg{Rserve}.
+A number of packages and projects have enabled R to respond to client requests over TCP/IP and local socket servers, such as \CRANpkg{Rserve} \citep{cran_rserve}, \CRANpkg{svSocket} \citep{grosjean_sciviews_2019}, [rApache](http://www.rapache.net) and more recently \pkg{plumber} ([https://www.rplumber.io/](https://www.rplumber.io/)) and \pkg{RestRserve} ([http://restrserve.org](http://restrserve.org)), which even processes incoming requests in parallel with forked processes using \CRANpkg{Rserve}.
 The latter two also provide documentation for deployment with Docker or ready-to-use images with automated builds\footnote{See \href{https://www.rplumber.io/docs/hosting.html\#docker}{https://www.rplumber.io/docs/hosting.html\#docker}, \href{https://hub.docker.com/r/trestletech/plumber/}{https://hub.docker.com/r/trestletech/plumber/} and \href{https://hub.docker.com/r/rexyai/restrserve/}{https://hub.docker.com/r/rexyai/restrserve/}.}.
@@
 CARD.com implemented a web application for the optimisation of the acquisition flow and the real-time analysis of debit card transactions.
-The software used \CRANpkg{Rserve} and rApache and was deployed in Docker containers.
-The R session behind \CRANpkg{Rserve} acted as a read-only in-memory database, which was extremely fast and scalabale, for the many concurrent rApache processes responding to the live-scoring requests of various divisions of the company.
+The software used \pkg{Rserve} and rApache and was deployed in Docker containers.
+The R session behind \pkg{Rserve} acted as a read-only in-memory database, which was extremely fast and scalabale, for the many concurrent rApache processes responding to the live-scoring requests of various divisions of the company.
 Similarly deodorised R scripts were responsible for the ETL processes and even the client-facing email, text message and push notification alerts sent in real-time based on card transactions.
@@
 The AI @ T-Mobile team created a set of neural network machine learning natural language processing models to help customer care agents manage text-based messages for customers \citep{t-mobile_enterprise_2018}.
-For example, one model quickly identifies whether a message is from a customer \citep[see \CRANpkg{Shiny}-based \href{https://secure.message.t-mobile.com/v1/shiny/is-customer/app/}{demo} further described by ][]{nolis_small_2019}, and others tell which customers are likely to make a repeat purchase.
-If a data scientist creates a machine learning model and exposes it through a \CRANpkg{plumber} API, then someone else on the marketing team could write software that sends different emails depending on that real-time prediction.
+For example, one model quickly identifies whether a message is from a customer \citep[see \pkg{Shiny}-based \href{https://secure.message.t-mobile.com/v1/shiny/is-customer/app/}{demo} further described by ][]{nolis_small_2019}, and others tell which customers are likely to make a repeat purchase.
+If a data scientist creates a machine learning model and exposes it through a \pkg{plumber} API, then someone else on the marketing team could write software that sends different emails depending on that real-time prediction.
 The models are convolutional neural networks that use the \CRANpkg{keras} package \citep{cran_keras} and run in a Rocker container.
@@
 The actual testing of R exercises is done using a custom framework loosely based on \CRANpkg{testthat} \citep{wickham_testthat_2011}.
-During the development of the testing framework, it was found that the \CRANpkg{testthat} framework did not provide enough information to its reporter system to send back all the fields required by Dodona to render its feedback.
+During the development of the testing framework, it was found that the \pkg{testthat} framework did not provide enough information to its reporter system to send back all the fields required by Dodona to render its feedback.
 Right now, multiple statistics courses are developing exercises to automate the feedback for their lab classes.
@@
 Furthermore, the maturing of the Rockerverse packages for managing containers may lead to them being adopted in situations where manual coding is currently required, e.g.
-in the case of \CRANpkg{RSelenium} or \CRANpkg{drake} (see Sections&nbsp;\nameref{rselenium} and \nameref{drake} respectively).
-In some cases, e.g., for \CRANpkg{analogsea}, the interaction with the Docker daemon may remain too specific to re-use first-order packages to control Docker.
+in the case of \pkg{RSelenium} or \pkg{drake} (see Sections&nbsp;\nameref{rselenium} and \nameref{drake} respectively).
+In some cases, e.g., for \pkg{analogsea}, the interaction with the Docker daemon may remain too specific to re-use first-order packages to control Docker.
 

### 2020-03-03 rockerverse.tex
only first mention uses \CRANpkg
preparing for submission, see #26

@@
 way, is system independent, and is likely more reliable than parsing
-command line output. \CRANpkg{stevedore}'s own interface is
-automatically generated based on the OpenAPI specification of the Docker
-daemon, but it is still similar to the Docker CLI. The interface is
-similar to R6 objects, in that an object of class
-\code{"stevedore\_object"} has a number of functions attached to it that
-can be called, and multiple specific versions of the Docker API can be
-supported thanks to the automatic
+command line output. \pkg{stevedore}'s own interface is automatically
+generated based on the OpenAPI specification of the Docker daemon, but
+it is still similar to the Docker CLI. The interface is similar to R6
+objects, in that an object of class \code{"stevedore\_object"} has a
+number of functions attached to it that can be called, and multiple
+specific versions of the Docker API can be supported thanks to the
+automatic
 generation\footnote{See \href{https://github.com/richfitz/stevedore/blob/master/development.md}{https://github.com/richfitz/stevedore/blob/master/development.md}.}.
@@
 names\footnote{See \href{https://sysreqs.r-hub.io/}{https://sysreqs.r-hub.io/}.}.
-\pkg{containerit} uses \CRANpkg{stevedore} to streamline the user
+\pkg{containerit} uses \pkg{stevedore} to streamline the user
 interaction and improve the created \texttt{Dockerfile}s, e.g., by
@@
 collaborative documents and large-scale platforms for processing
-documents. \CRANpkg{liftr} makes the document the main and sole workflow
+documents. \pkg{liftr} makes the document the main and sole workflow
 control file and the only file that needs to be shared between
@@
 naturally be applied within images and containers. For example,
-\CRANpkg{renv} helps users to manage the state of the R library in a
+\pkg{renv} helps users to manage the state of the R library in a
 reproducible way, further providing isolation and portability. While
-\CRANpkg{renv} does not cover system dependencies, the
-\CRANpkg{renv}-based environment can be transferred into a container
-either by restoring the environment based on the main configuration file
-\texttt{renv.lock} or by storing the \CRANpkg{renv}-cache on the host
-and not in the container \citep{ushey_using_2019}. With both the system
-dependencies and R packages consciously managed in a Docker image, users
-can start using containers as the \emph{only} environment for their
-workflows, which allows them to work indepenently of physical
+\pkg{renv} does not cover system dependencies, the \pkg{renv}-based
+environment can be transferred into a container either by restoring the
+environment based on the main configuration file \texttt{renv.lock} or
+by storing the \pkg{renv}-cache on the host and not in the container
+\citep{ushey_using_2019}. With both the system dependencies and R
+packages consciously managed in a Docker image, users can start using
+containers as the \emph{only} environment for their workflows, which
+allows them to work indepenently of physical
 computers\footnote{Allowing them to be digital "nomads", cf. J.~Bryan's \href{https://github.com/jennybc/docker-why}{https://github.com/jennybc/docker-why}.}
@@
 hundreds of different spatial raster and vector formats
-\citep{pebesma_simple_2018,cran_rgdal}. \CRANpkg{sf} and \CRANpkg{rgdal}
-have hundreds of indirect reverse imports and dependencies and,
-therefore, the maintainers spend a lot of effort trying not to break
-them. Purpose-built Docker images are used to prepare for upcoming
-releases of system libraries, individual bug reports, and for the lowest
-supported versions of system
+\citep{pebesma_simple_2018,cran_rgdal}. \pkg{sf} and \pkg{rgdal} have
+hundreds of indirect reverse imports and dependencies and, therefore,
+the maintainers spend a lot of effort trying not to break them.
+Purpose-built Docker images are used to prepare for upcoming releases of
+system libraries, individual bug reports, and for the lowest supported
+versions of system
 libraries\footnote{Cf.
@@
 a Docker container with the same R environments.
-\CRANpkg{googleComputeEngineR} integrates with the R parallelisation
-package \CRANpkg{future} \citep{cran_future} to enable this with only a
-few lines or R
+\pkg{googleComputeEngineR} integrates with the R parallelisation package
+\CRANpkg{future} \citep{cran_future} to enable this with only a few
+lines or R
 code\footnote{\href{https://cloudyr.github.io/googleComputeEngineR/articles/massive-parallel.html}{https://cloudyr.github.io/googleComputeEngineR/articles/massive-parallel.html}}.
@@
 parallel computing and automated detection of the parts of the work that
-actually needs to be re-executed. \CRANpkg{drake} has been demonstrated
-to run inside containers for high
+actually needs to be re-executed. \pkg{drake} has been demonstrated to
+run inside containers for high
 reproducibility\footnote{See for example \href{https://github.com/joelnitta/pleurosoriopsis}{https://github.com/joelnitta/pleurosoriopsis} or \href{https://gitlab.com/ecohealthalliance/drake-gitlab-docker-example}{https://gitlab.com/ecohealthalliance/drake-gitlab-docker-example}, the latter even running in a continuous integration platform (cf.~\nameref{development}.}.
-Furthermore, \CRANpkg{drake} workflows have been shown to use
-\CRANpkg{future} package's function \code{makeClusterPSOCK()} for
-sending parts of the workflow to a Docker image for
+Furthermore, \pkg{drake} workflows have been shown to use \pkg{future}
+package's function \code{makeClusterPSOCK()} for sending parts of the
+workflow to a Docker image for
 execution\footnote{\href{https://docs.ropensci.org/drake/index.html?q=docker\#with-docker}{https://docs.ropensci.org/drake/index.html?q=docker\#with-docker}}
@@
 latter case, the container control code must be written by the user, and
-the \CRANpkg{future} package ensures that the host and worker can
-connect for communicating over socket connections. \textbf{RStudio
-Server Pro} includes a functionality called
+the \pkg{future} package ensures that the host and worker can connect
+for communicating over socket connections. \textbf{RStudio Server Pro}
+includes a functionality called
 \href{https://solutions.rstudio.com/launcher/overview/}{Launcher} (since
@@
 First, different packages use containers for the \textbf{deployment of R
-and \CRANpkg{Shiny} apps}. \CRANpkg{Shiny} is a popular package for
-creating interactive online dashboards with R, and it enables users with
-very diverse backgrounds to create stable and user-friendly web
-applications \citep{cran_shiny}. \emph{ShinyProxy}
-(\url{https://www.shinyproxy.io/}) is an open source tool to deploy
-Shiny apps in an enterprise context, where it features single sign-on,
-but it can also be used in scientific use cases
+and \pkg{Shiny} apps}. \pkg{Shiny} is a popular package for creating
+interactive online dashboards with R, and it enables users with very
+diverse backgrounds to create stable and user-friendly web applications
+\citep{cran_shiny}. \emph{ShinyProxy} (\url{https://www.shinyproxy.io/})
+is an open source tool to deploy Shiny apps in an enterprise context,
+where it features single sign-on, but it can also be used in scientific
+use cases
 \citep[e.g., ][]{savini_epiexplorer_2019,glouzon_structurexplor_2017}.
@@
 Docker CLI commands and transferred transparently to the respective
-remove machine using SSH. \CRANpkg{AzureContainers} is an umbrella
-package that provides interfaces for three commercial services of
-Microsoft's Azure Cloud, namely
+remove machine using SSH. \pkg{AzureContainers} is an umbrella package
+that provides interfaces for three commercial services of Microsoft's
+Azure Cloud, namely
 \href{https://azure.microsoft.com/en-us/services/container-instances/}{Container
@@
 Service} for orchestrated deployments. While a package like
-\CRANpkg{plumber} provides the infrastructure for turning an R workflow
-into a web service, for production purposes it is usually necessary to
-take into account scalability, reliability and ease of management.
+\pkg{plumber} provides the infrastructure for turning an R workflow into
+a web service, for production purposes it is usually necessary to take
+into account scalability, reliability and ease of management.
 AzureContainers provides an R-based interface to these features and,
@@
 tiny Golang web server. The watchdog marshals an HTTP request and
-invokes the actual application. The R session uses \CRANpkg{plumber} or
+invokes the actual application. The R session uses \pkg{plumber} or
 similar packages for the API endpoint with packages and data preloaded,
@@
 \citep{grosjean_sciviews_2019}, \href{http://www.rapache.net}{rApache}
-and more recently \CRANpkg{plumber} (\url{https://www.rplumber.io/}) and
+and more recently \pkg{plumber} (\url{https://www.rplumber.io/}) and
 \pkg{RestRserve} (\url{http://restrserve.org}), which even processes
@@
 acquisition flow and the real-time analysis of debit card transactions.
-The software used \CRANpkg{Rserve} and rApache and was deployed in
-Docker containers. The R session behind \CRANpkg{Rserve} acted as a
-read-only in-memory database, which was extremely fast and scalabale,
-for the many concurrent rApache processes responding to the live-scoring
-requests of various divisions of the company. Similarly deodorised R
-scripts were responsible for the ETL processes and even the
-client-facing email, text message and push notification alerts sent in
-real-time based on card transactions. The related Docker images were
-made available at \url{https://github.com/cardcorp/card-rocker}. The
-images extended \texttt{rocker/r-base} and additionally entailed an SSH
-client and a workaround for being able to mount SSH keys from the host,
-Pandoc, the Amazon Web Services (AWS) SDK, and Java, which is required
-by the AWS SDK. The AWS SDK allowed for running R consumers reading from
-real-time data processing streams of
-\href{https://aws.amazon.com/kinesis/}{AWS Kinesis}
+The software used \pkg{Rserve} and rApache and was deployed in Docker
+containers. The R session behind \pkg{Rserve} acted as a read-only
+in-memory database, which was extremely fast and scalabale, for the many
+concurrent rApache processes responding to the live-scoring requests of
+various divisions of the company. Similarly deodorised R scripts were
+responsible for the ETL processes and even the client-facing email, text
+message and push notification alerts sent in real-time based on card
+transactions. The related Docker images were made available at
+\url{https://github.com/cardcorp/card-rocker}. The images extended
+\texttt{rocker/r-base} and additionally entailed an SSH client and a
+workaround for being able to mount SSH keys from the host, Pandoc, the
+Amazon Web Services (AWS) SDK, and Java, which is required by the AWS
+SDK. The AWS SDK allowed for running R consumers reading from real-time
+data processing streams of \href{https://aws.amazon.com/kinesis/}{AWS
+Kinesis}
 \footnote{See useR!2017 talk \href{https://static.sched.com/hosted\_files/user2017/2f/AWR Kinesis at useR 2017.pdf}{"Stream processing with R in AWS"}.}.
@@
 customer
-\citep[see \CRANpkg{Shiny}-based \href{https://secure.message.t-mobile.com/v1/shiny/is-customer/app/}{demo} further described by ][]{nolis_small_2019},
+\citep[see \pkg{Shiny}-based \href{https://secure.message.t-mobile.com/v1/shiny/is-customer/app/}{demo} further described by ][]{nolis_small_2019},
 and others tell which customers are likely to make a repeat purchase. If
 a data scientist creates a machine learning model and exposes it through
-a \CRANpkg{plumber} API, then someone else on the marketing team could
-write software that sends different emails depending on that real-time
+a \pkg{plumber} API, then someone else on the marketing team could write
+software that sends different emails depending on that real-time
 prediction. The models are convolutional neural networks that use the
@@
 \citep{wickham_testthat_2011}. During the development of the testing
-framework, it was found that the \CRANpkg{testthat} framework did not
+framework, it was found that the \pkg{testthat} framework did not
 provide enough information to its reporter system to send back all the
@@
 situations where manual coding is currently required, e.g.~in the case
-of \CRANpkg{RSelenium} or \CRANpkg{drake} (see
-Sections~\nameref{rselenium} and \nameref{drake} respectively). In some
-cases, e.g., for \CRANpkg{analogsea}, the interaction with the Docker
-daemon may remain too specific to re-use first-order packages to control
-Docker.
+of \pkg{RSelenium} or \pkg{drake} (see Sections~\nameref{rselenium} and
+\nameref{drake} respectively). In some cases, e.g., for \pkg{analogsea},
+the interaction with the Docker daemon may remain too specific to re-use
+first-order packages to control Docker.
 

### 2020-03-03 rockerverse.Rmd
corrected typos in the CyVerse section
corrected some grammar

@@
 [CyVerse](https://cyverse.org) is an open source, NSF-funded cyberinfrastructure platform for the life sciences providing easy access to computing and storage resources \citep{merchant_iplant_2016}.
-CyVerse's a browser-based 'data science workbench' is called [Discovery Environment](https://cyverse.org/discovery-environment) (DE).
-DE uses a combination of [HTCondor](https://research.cs.wisc.edu/htcondor/) and Kubernetes for orchestrating container-based analysis and integrates with external HPC, i.e., [NSF-XSEDE](https://www.xsede.org/), through [TAPIS](https://www.tacc.utexas.edu/tapis) API's.
-In addition, CyVerse provides a multi-petabyte Data Store based on [iRODS](https://irods.org/) with shared access by users.
+CyVerse has a browser-based 'data science workbench' called the [Discovery Environment](https://cyverse.org/discovery-environment) (DE).
+The DE uses a combination of [HTCondor](https://research.cs.wisc.edu/htcondor/) and Kubernetes for orchestrating container-based analysis and integrates with external HPC, i.e., [NSF-XSEDE](https://www.xsede.org/), through [TAPIS](https://www.tacc.utexas.edu/tapis)(TACC-API's).
+CyVerse hosts a multi-petabyte Data Store based on [iRODS](https://irods.org/) with shared access by its users.
 The DE runs Docker containers on demand, with users able to integrate bespoke containers from DockerHub or other registries \citep{devisetty_bringing_2016}.
-Rocker images integration in the DE is designed to provide life science researchers with scalable, compute-intensive R analysis capabilities needed for large and complex datasets (e.g., genomics/multi-omics, GWAS, phenotypic data, spatial data etc.).
-These capabilities give users flexibility similar to Binder, but allows containers to run on larger computational resources (RAM, CPU, Disk), and for longer periods of time (days to weeks).
-The Rocker Project's RStudio and Shiny containers are integrated into the DE by deriving new images from Rocker images\footnote{See \href{https://github.com/cyverse-vice/}{https://github.com/cyverse-vice/} for \texttt{Dockerfile}s and configuration scripts; images are auto-built on DockerHub at \href{https://hub.docker.com/u/cyversevice}{https://hub.docker.com/u/cyversevice}.}.
-These images include a reverse proxy using `nginx` to handle communication with CyVerse's authentication systems, which allow users to invite other registered users to securely access the same instance \citep{rstudio_proxy_2020}.
-The images further include specific tools, such as the CLI utility `icommands` to access iRODS.
-Users can reference the same base image and add packages to further customize their R environment, or they can use CyVerse's image registry where a user's published, i.e., publicly available, containers are held for additional reproducibility.
-CyVerse user accounts are free (with some limitations for non-US users) and the [CyVerse Learning Center](https://learning.cyverse.org/) allows community members to document their containers to support training and education.
+Rocker image integration in the DE is designed to provide researchers with scalable, compute-intensive, R analysis capabilities for large and complex datasets (e.g., genomics/multi-omics, GWAS, phenotypic data, geospatial data, etc.).
+These capabilities give users flexibility similar to Binder, but allow containers to be run on larger computational resources (RAM, CPU, Disk, GPU), and for longer periods of time (days to weeks).
+The Rocker Project's RStudio and Shiny are integrated into the DE by deriving new images from Rocker images\footnote{See \href{https://github.com/cyverse-vice/}{https://github.com/cyverse-vice/} for \texttt{Dockerfile}s and configuration scripts; images are auto-built on DockerHub at \href{https://hub.docker.com/u/cyversevice}{https://hub.docker.com/u/cyversevice}.}.
+These new images include a reverse proxy using `nginx` to handle communication with CyVerse's authentication system \citep{rstudio_proxy_2020}; CyVerse also allows owners to invite other registered users to securely access the same instance.
+The CyVerse Rocker images further include tools for connecting to its Data Store, such as the CLI utility `icommands` for iRODS.
+CyVerse accounts are free (with some limitations for non-US users), and the [CyVerse Learning Center](https://learning.cyverse.org/) provides community members with information about the platform, incuding training and education opportunities.
 

### 2020-03-03 rockerverse.Rmd
Minor changes to contributions and acknowledgements

@@
 CvP wrote the section on the usage of Docker containers in Dodona.
-CW wrote the sections on Whole Tale and contributed the publication reproducibility audit use case.
+CW wrote the sections on Whole Tale and contributed content about publication reproducibility audits.
 NX contributed content on \pkg{liftr}.
@@
 LS and NT are supported by US NIH / NHGRI awards U41HG00405 and U24HG010263.
-CW is supported by the Whole Tale projects (https://wholetale.org) funded by the US National Science Foundation (NSF) under award [OAC-1541450](https://www.nsf.gov/awardsearch/showAward?AWD_ID=1541450).
+CW is supported by the Whole Tale project (https://wholetale.org) funded by the US National Science Foundation (NSF) under award [OAC-1541450](https://www.nsf.gov/awardsearch/showAward?AWD_ID=1541450).
 We would like to thank Celeste R. Brennecka from the Science Writing Support Service of the University of Münster for her editorial support.

### 2020-03-04 rockerverse.Rmd
addition of the 'PrarieLearn' experience with rocker-pl

--- 2020-03-03 rockerverse.Rmd
@@
 These trends cement the role of R as the _lingua&nbsp;franca_ of statistics, data visualisation, and computational research.
-During the last few years, and coinciding with the rise of R, has been the advent of [Docker](https://en.wikipedia.org/wiki/Docker_(software)) as a general tool for distributing and deploying of server applications---in fact, Docker can be called the _lingua&nbsp;franca_ of describing computing environments and packaging software.
+The last few years, coinciding with the rise of R, have also seen the rise of [Docker](https://en.wikipedia.org/wiki/Docker_(software)) as a general tool for distributing and deploying of server applications---in fact, Docker can be called the _lingua&nbsp;franca_ of describing computing environments and packaging software.
 Combining both these topics, the _Rocker&nbsp;Project_ ([https://www.rocker-project.org/](https://www.rocker-project.org/)) provides Docker images with R (see the next section for more details).
@@
 
+Another example of a Docker-based teaching and testing platform is [PrairieLearn](https://github.com/PrairieLearn) which is being developed at the University of Illinois at Urbana-Champaign \citep{prairielearn:2018} where it has already been in extensive use across several faculties (along with initial use on some other campuses).
+It uses Docker containers a key components, both internally for its operations (programmed mainly in Python as well as in Javascript), as well as for two reference containers providing, respectively, Python and R autograders.
+A key design decision made by PrairieLearn permits _external_ grading container to be supplied and accessed via a well-defined interface of invoking, essentially, a single script `run.sh` expecting JSON based configuration, relying on a well-defined file layout containing support files, exam questions, supplementary data as well as student submissions, and returning per-question evaluations as JSON result files which are evaluated, aggregated and recorded in SQL.
+The [Data Science Programming Methods](https://stat430.com) course \citep{stat430:2019} uses this via the custom [`rocker-pl`](https://github.com/stat430dspm/rocker-pl) container \citep{rocker-pl:2019}.\footnote{The reference R container was unavailable at the time, and also relies on a heavier CentOS-based build so that a lighter and alternative was established.}
+The `rocker-pl` container extends `rocker/r-base` with both the [`plr`](https://github.com/stat430dspm/plr) \citep{pkg:plr:2019} custom R package for integration into PrarieLearn testing and question evaluation, along with the actual R packages used in instruction and testing for the course in question.
+As automated grading of submitted student answers is close to the well-understood problem of unit testing, the \CRANpkg{tinytest} package \citep{CRAN:tinytest} is used as it offers both its core features for testing as well as clean extensibility which package \CRANpkg{ttdo} \citep{CRAN:ttdo} utilizes to render differences between incorrect answers and reference answers by invoking the \CRANpkg{diffobj} package \citep{CRAN:diffobj} for context-sensitive colorized displays of these differences.
+The setup, described in more detail by \citet{paper:r_autograder}, is an excellent illustration of both the versatility and flexibility offered by Docker-based approaches in teaching and testing.
+
 ## Packaging research reproducibly

### 2020-03-04 rockerverse.tex
addition of the 'PrarieLearn' experience with rocker-pl

@@
 These trends cement the role of R as the \emph{lingua~franca} of
-statistics, data visualisation, and computational research. During the
-last few years, and coinciding with the rise of R, has been the advent
-of \href{https://en.wikipedia.org/wiki/Docker_(software)}{Docker} as a
+statistics, data visualisation, and computational research. The last few
+years, coinciding with the rise of R, have also seen the rise of
+\href{https://en.wikipedia.org/wiki/Docker_(software)}{Docker} as a
 general tool for distributing and deploying of server applications---in
@@
 cyberinfrastructure platform for the life sciences providing easy access
-to computing and storage resources \citep{merchant_iplant_2016}.
-CyVerse's a browser-based `data science workbench' is called
+to computing and storage resources \citep{merchant_iplant_2016}. CyVerse
+has a browser-based `data science workbench' called the
 \href{https://cyverse.org/discovery-environment}{Discovery Environment}
-(DE). DE uses a combination of
+(DE). The DE uses a combination of
 \href{https://research.cs.wisc.edu/htcondor/}{HTCondor} and Kubernetes
@@
 HPC, i.e., \href{https://www.xsede.org/}{NSF-XSEDE}, through
-\href{https://www.tacc.utexas.edu/tapis}{TAPIS} API's. In addition,
-CyVerse provides a multi-petabyte Data Store based on
-\href{https://irods.org/}{iRODS} with shared access by users. The DE
+\href{https://www.tacc.utexas.edu/tapis}{TAPIS}(TACC-API's). CyVerse
+hosts a multi-petabyte Data Store based on
+\href{https://irods.org/}{iRODS} with shared access by its users. The DE
 runs Docker containers on demand, with users able to integrate bespoke
 containers from DockerHub or other registries
-\citep{devisetty_bringing_2016}. Rocker images integration in the DE is
-designed to provide life science researchers with scalable,
-compute-intensive R analysis capabilities needed for large and complex
-datasets (e.g., genomics/multi-omics, GWAS, phenotypic data, spatial
-data etc.). These capabilities give users flexibility similar to Binder,
-but allows containers to run on larger computational resources (RAM,
-CPU, Disk), and for longer periods of time (days to weeks). The Rocker
-Project's RStudio and Shiny containers are integrated into the DE by
-deriving new images from Rocker
+\citep{devisetty_bringing_2016}. Rocker image integration in the DE is
+designed to provide researchers with scalable, compute-intensive, R
+analysis capabilities for large and complex datasets (e.g.,
+genomics/multi-omics, GWAS, phenotypic data, geospatial data, etc.).
+These capabilities give users flexibility similar to Binder, but allow
+containers to be run on larger computational resources (RAM, CPU, Disk,
+GPU), and for longer periods of time (days to weeks). The Rocker
+Project's RStudio and Shiny are integrated into the DE by deriving new
+images from Rocker
 images\footnote{See \href{https://github.com/cyverse-vice/}{https://github.com/cyverse-vice/} for \texttt{Dockerfile}s and configuration scripts; images are auto-built on DockerHub at \href{https://hub.docker.com/u/cyversevice}{https://hub.docker.com/u/cyversevice}.}.
-These images include a reverse proxy using \texttt{nginx} to handle
-communication with CyVerse's authentication systems, which allow users
-to invite other registered users to securely access the same instance
-\citep{rstudio_proxy_2020}. The images further include specific tools,
-such as the CLI utility \texttt{icommands} to access iRODS. Users can
-reference the same base image and add packages to further customize
-their R environment, or they can use CyVerse's image registry where a
-user's published, i.e., publicly available, containers are held for
-additional reproducibility. CyVerse user accounts are free (with some
-limitations for non-US users) and the
-\href{https://learning.cyverse.org/}{CyVerse Learning Center} allows
-community members to document their containers to support training and
-education.
+These new images include a reverse proxy using \texttt{nginx} to handle
+communication with CyVerse's authentication system
+\citep{rstudio_proxy_2020}; CyVerse also allows owners to invite other
+registered users to securely access the same instance. The CyVerse
+Rocker images further include tools for connecting to its Data Store,
+such as the CLI utility \texttt{icommands} for iRODS. CyVerse accounts
+are free (with some limitations for non-US users), and the
+\href{https://learning.cyverse.org/}{CyVerse Learning Center} provides
+community members with information about the platform, incuding training
+and education opportunities.
 
@@
 their lab classes.
+
+Another example of a Docker-based teaching and testing platform is
+\href{https://github.com/PrairieLearn}{PrairieLearn} which is being
+developed at the University of Illinois at Urbana-Champaign
+\citep{prairielearn:2018} where it has already been in extensive use
+across several faculties (along with initial use on some other
+campuses). It uses Docker containers a key components, both internally
+for its operations (programmed mainly in Python as well as in
+Javascript), as well as for two reference containers providing,
+respectively, Python and R autograders. A key design decision made by
+PrairieLearn permits \emph{external} grading container to be supplied
+and accessed via a well-defined interface of invoking, essentially, a
+single script \texttt{run.sh} expecting JSON based configuration,
+relying on a well-defined file layout containing support files, exam
+questions, supplementary data as well as student submissions, and
+returning per-question evaluations as JSON result files which are
+evaluated, aggregated and recorded in SQL. The
+\href{https://stat430.com}{Data Science Programming Methods} course
+\citep{stat430:2019} uses this via the custom
+\href{https://github.com/stat430dspm/rocker-pl}{\texttt{rocker-pl}}
+container
+\citep{rocker-pl:2019}.\footnote{The reference R container was unavailable at the time, and also relies on a heavier CentOS-based build so that a lighter and alternative was established.}
+The \texttt{rocker-pl} container extends \texttt{rocker/r-base} with
+both the \href{https://github.com/stat430dspm/plr}{\texttt{plr}}
+\citep{pkg:plr:2019} custom R package for integration into PrarieLearn
+testing and question evaluation, along with the actual R packages used
+in instruction and testing for the course in question. As automated
+grading of submitted student answers is close to the well-understood
+problem of unit testing, the \CRANpkg{tinytest} package
+\citep{CRAN:tinytest} is used as it offers both its core features for
+testing as well as clean extensibility which package \CRANpkg{ttdo}
+\citep{CRAN:ttdo} utilizes to render differences between incorrect
+answers and reference answers by invoking the \CRANpkg{diffobj} package
+\citep{CRAN:diffobj} for context-sensitive colorized displays of these
+differences. The setup, described in more detail by
+\citet{paper:r_autograder}, is an excellent illustration of both the
+versatility and flexibility offered by Docker-based approaches in
+teaching and testing.
 

### 2020-03-04 rockerverse.Rmd
add teaching use case

--- 2020-03-03 rockerverse.Rmd
@@
 
-**Teaching** is a further example of where shared computing environments and sandboxing can greatly improve the process.
-First, _prepared environments for teaching_ are especially helpful for (a) introductory courses, where students often struggle with the first step of installation and configuration \citep{cetinkaya-rundel_infrastructure_2018}, and (b) courses that require access to a relatively complex setup of software tools, e.g., database systems.
+## Teaching
+\label{workenvs}
+
+Two use cases demonstrate the practical usefulness and advantages of containerisation in the context of teaching.
+On the one hand a special case of shared computing environments (see Section&nbsp;\ref{workenvs}), and on the other hand leveraging sandboxing and controlled environments for autograding.
+
+**Prepared environments for teaching** are especially helpful for (a) introductory courses, where students often struggle with the first step of installation and configuration \citep{cetinkaya-rundel_infrastructure_2018}, and (b) courses that require access to a relatively complex setup of software tools, e.g., database systems.
 \citet{cetinkaya-rundel_infrastructure_2018} describe how a Docker-based deployment of RStudio (i) avoided problems with troubleshooting individual students' computers and greatly increased engagement through very quickly showing tangible outcomes, e.g., a visualisation, and (ii) reduced demand on teaching and IT staff.
@@
 
-Second, the idea of a common environment and partitioning allows for using _containers in teaching for secure execution and automated testing_ of submissions by students.
-[Dodona](https://dodona.ugent.be) is a web platform developed at Ghent University that is used to teach students basic programming skills, and it uses Docker containers to test submissions by students.
+The idea of a common environment and partitioning allows for using **containers in teaching for secure execution and automated testing** of submissions by students.
+First, [Dodona](https://dodona.ugent.be) is a web platform developed at Ghent University that is used to teach students basic programming skills, and it uses Docker containers to test submissions by students.
 This means that both the code testing the students' submissions and the submission itself are executed in a predictable environment, avoiding compatibility issues between the wide variety of configurations used by students.
@@
 Right now, multiple statistics courses are developing exercises to automate the feedback for their lab classes.
+
+Second, [PrairieLearn](https://github.com/PrairieLearn) is another example of a Docker-based teaching and testing platform.
+PrairieLearn is being developed at the University of Illinois at Urbana-Champaign \citep{prairielearn:2018} and has been in extensive use across several faculties along with initial use on some other campuses.
+¸It uses Docker containers a key components, both internally for its operations (programmed mainly in Python as well as in Javascript), as well as for two reference containers providing, respectively, Python and R autograders.
+A key design decision made by PrairieLearn permits _external_ grading containers to be supplied and accessed via a well-defined interface of invoking, essentially, a single script, `run.sh`.
+This script relies on a well-defined file layout containing JSON-based configuration, support files, exam questions, supplementary data, and student submissions.
+It returns per-question evaluations as JSON result files, which PrarieLearn evaluates, aggregates and records in a database.
+The [Data Science Programming Methods](https://stat430.com) course \citep{stat430:2019} uses this via the custom [`rocker-pl`](https://github.com/stat430dspm/rocker-pl) container \citep{rocker-pl:2019}.\footnote{The reference R container was unavailable at the time, and also relies on a heavier CentOS-based build so that a lighter alternative was established.}
+The `rocker-pl` image extends `rocker/r-base` with the \pkg{plr} R package \citep{pkg:plr:2019} for integration into PrarieLearn testing and question evaluation, along with the actual R packages used in instruction and testing for the course in question.
+As automated grading of submitted student answers is close to the well-understood problem of unit testing, the \CRANpkg{tinytest} package \citep{CRAN:tinytest} is used for both its core features for testing as well as clean extensibility.
+The package \CRANpkg{ttdo} \citep{CRAN:ttdo} utilizes the extensibility of \pkg{tinytest} to display context-sensitive colorized differences between incorrect answers and reference answers using the \CRANpkg{diffobj} package \citep{CRAN:diffobj}.
+The setup, described in more detail by \citet{paper:r_autograder}, is an excellent illustration of both the versatility and flexibility offered by Docker-based approaches in teaching and testing.
 

### 2020-03-04 rockerverse.tex
add teaching use case

@@
 
-\textbf{Teaching} is a further example of where shared computing
-environments and sandboxing can greatly improve the process. First,
-\emph{prepared environments for teaching} are especially helpful for (a)
-introductory courses, where students often struggle with the first step
-of installation and configuration
+\hypertarget{teaching}{%
+\subsection{Teaching}\label{teaching}}
+
+\label{workenvs}
+
+Two use cases demonstrate the practical usefulness and advantages of
+containerisation in the context of teaching. On the one hand a special
+case of shared computing environments (see Section~\ref{workenvs}), and
+on the other hand leveraging sandboxing and controlled environments for
+autograding.
+
+\textbf{Prepared environments for teaching} are especially helpful for
+(a) introductory courses, where students often struggle with the first
+step of installation and configuration
 \citep{cetinkaya-rundel_infrastructure_2018}, and (b) courses that
@@
 
-Second, the idea of a common environment and partitioning allows for
-using \emph{containers in teaching for secure execution and automated
-testing} of submissions by students.
+The idea of a common environment and partitioning allows for using
+\textbf{containers in teaching for secure execution and automated
+testing} of submissions by students. First,
 \href{https://dodona.ugent.be}{Dodona} is a web platform developed at
@@
 
-Another example of a Docker-based teaching and testing platform is
-\href{https://github.com/PrairieLearn}{PrairieLearn} which is being
-developed at the University of Illinois at Urbana-Champaign
-\citep{prairielearn:2018} where it has already been in extensive use
-across several faculties (along with initial use on some other
-campuses). It uses Docker containers a key components, both internally
-for its operations (programmed mainly in Python as well as in
-Javascript), as well as for two reference containers providing,
-respectively, Python and R autograders. A key design decision made by
-PrairieLearn permits \emph{external} grading container to be supplied
-and accessed via a well-defined interface of invoking, essentially, a
-single script \texttt{run.sh} expecting JSON based configuration,
-relying on a well-defined file layout containing support files, exam
-questions, supplementary data as well as student submissions, and
-returning per-question evaluations as JSON result files which are
-evaluated, aggregated and recorded in SQL. The
+Second, \href{https://github.com/PrairieLearn}{PrairieLearn} is another
+example of a Docker-based teaching and testing platform. PrairieLearn is
+being developed at the University of Illinois at Urbana-Champaign
+\citep{prairielearn:2018} and has been in extensive use across several
+faculties along with initial use on some other campuses. ¸It uses Docker
+containers a key components, both internally for its operations
+(programmed mainly in Python as well as in Javascript), as well as for
+two reference containers providing, respectively, Python and R
+autograders. A key design decision made by PrairieLearn permits
+\emph{external} grading containers to be supplied and accessed via a
+well-defined interface of invoking, essentially, a single script,
+\texttt{run.sh}. This script relies on a well-defined file layout
+containing JSON-based configuration, support files, exam questions,
+supplementary data, and student submissions. It returns per-question
+evaluations as JSON result files, which PrarieLearn evaluates,
+aggregates and records in a database. The
 \href{https://stat430.com}{Data Science Programming Methods} course
@@
 container
-\citep{rocker-pl:2019}.\footnote{The reference R container was unavailable at the time, and also relies on a heavier CentOS-based build so that a lighter and alternative was established.}
-The \texttt{rocker-pl} container extends \texttt{rocker/r-base} with
-both the \href{https://github.com/stat430dspm/plr}{\texttt{plr}}
-\citep{pkg:plr:2019} custom R package for integration into PrarieLearn
-testing and question evaluation, along with the actual R packages used
-in instruction and testing for the course in question. As automated
-grading of submitted student answers is close to the well-understood
-problem of unit testing, the \CRANpkg{tinytest} package
-\citep{CRAN:tinytest} is used as it offers both its core features for
-testing as well as clean extensibility which package \CRANpkg{ttdo}
-\citep{CRAN:ttdo} utilizes to render differences between incorrect
-answers and reference answers by invoking the \CRANpkg{diffobj} package
-\citep{CRAN:diffobj} for context-sensitive colorized displays of these
-differences. The setup, described in more detail by
+\citep{rocker-pl:2019}.\footnote{The reference R container was unavailable at the time, and also relies on a heavier CentOS-based build so that a lighter alternative was established.}
+The \texttt{rocker-pl} image extends \texttt{rocker/r-base} with the
+\pkg{plr} R package \citep{pkg:plr:2019} for integration into
+PrarieLearn testing and question evaluation, along with the actual R
+packages used in instruction and testing for the course in question. As
+automated grading of submitted student answers is close to the
+well-understood problem of unit testing, the \CRANpkg{tinytest} package
+\citep{CRAN:tinytest} is used for both its core features for testing as
+well as clean extensibility. The package \CRANpkg{ttdo}
+\citep{CRAN:ttdo} utilizes the extensibility of \pkg{tinytest} to
+display context-sensitive colorized differences between incorrect
+answers and reference answers using the \CRANpkg{diffobj} package
+\citep{CRAN:diffobj}. The setup, described in more detail by
 \citet{paper:r_autograder}, is an excellent illustration of both the
@@
 section on the usage of Docker containers in Dodona. CW wrote the
-sections on Whole Tale and contributed the publication reproducibility
-audit use case. NX contributed content on \pkg{liftr}. All authors
-approved the final version. This articles was collaboratively written at
+sections on Whole Tale and contributed content about publication
+reproducibility audits. NX contributed content on \pkg{liftr}. All
+authors approved the final version. This articles was collaboratively
+written at
 \href{https://github.com/nuest/rockerverse-paper/}{https://github.com/nuest/rockerverse-paper/}.
@@
 number 2016PG-BRI004. LS and NT are supported by US NIH / NHGRI awards
-U41HG00405 and U24HG010263. CW is supported by the Whole Tale projects
+U41HG00405 and U24HG010263. CW is supported by the Whole Tale project
 (\url{https://wholetale.org}) funded by the US National Science

### 2020-03-04 rockerverse.Rmd
fix some widows and orphans
https://en.wikipedia.org/wiki/Widows_and_orphans

@@
 Docker is not the only containerisation software.
-An alternative stemming from the domain of high-perfomance computing is **Singularity** \citep{kurtzer_singularity_2017}.
-Singularity can run Docker images.
+**Singularity** stems from the domain of high-perfomance computing \citep{kurtzer_singularity_2017} and can also run Docker images.
 Rocker images work out of the box if the main process is R, e.g., in `rocker/r-base`, but Singularity does not succeed in running images where there is an init script, e.g., in containers that by default run RStudio&nbsp;Server.
@@
 
-The _Bioconductor_ `Dockerfile` selects the desired version of R from Rocker, adds required system dependencies, and uses the \CRANpkg{BiocManager} package for installing appropriate versions of _Bioconductor_ packages \citep{cran_biocmanager}.
-A strength of this approach is that the responsibility for complex software configuration (including customized development) is shifted from the user to the experienced _Bioconductor_ core team.
+The _Bioconductor_ `Dockerfile` selects the desired R version from Rocker images, adds required system dependencies, and uses the \CRANpkg{BiocManager} package for installing appropriate versions of _Bioconductor_ packages \citep{cran_biocmanager}.
+A strength of this approach is that the responsibility for complex software configuration and customization is shifted from the user to the experienced _Bioconductor_ core team.
 However, a recent audit of the _Bioconductor_  image stack `Dockerfile` led to the deprecation of several community-maintained images, because the numerous specific images became too hard to understand, complex to maintain, and cumbersome to customise.
-As part of the simplification, a recent innovation is to produce a `bioconductor_docker:devel` image to emulate the _Bioconductor_ nightly Linux build machine as closely as possible.
-This image contains the build system environment variables and the _system dependencies_ needed to install and check almost all (1813 out of 1823) _Bioconductor_ software packages, and it saves users and package developers from managing these themselves.
+As part of the simplification, a recent innovation is the `bioconductor_docker:devel` image, which emulates the _Bioconductor_ environment for nightly builds as closely as possible.
+This image contains the environment variables and the system dependencies needed to install and check almost all _Bioconductor_ software packages (1813 out of 1823).
+It saves users and package developers from creating this environment themselves.
 Furthermore, the image is configured so that \code{.libPaths()} has \samp{/usr/local/lib/R/host-site-library} as the first location.
@@
 
-**\CRANpkg{liftr}** aims to solve the problem of persistent reproducible reporting in statistical computing based on the R Markdown format \citep{xie2018} for dynamic documents \citep[\href{https://nanx.me/liftr/}{https://nanx.me/liftr/}, ][]{liftr2019}.
+**\CRANpkg{liftr}** \citep{liftr2019} aims to solve the problem of persistent reproducible reporting in statistical computing based on the R Markdown format \citep{xie2018}.
 The irreproducibility of authoring environments can become an issue for collaborative documents and large-scale platforms for processing documents.
-\pkg{liftr} makes the document the main and sole workflow control file and the only file that needs to be shared between collaborators for consistent environments, e.g.
+\pkg{liftr} makes the dynamic R Markdown document the main and sole workflow control file and the only file that needs to be shared between collaborators for consistent environments, e.g.
 demonstrated in the DockFlow project (https://dockflow.org).
-It introduces new fields to the R Markdown document header, allowing users to manually declare the dependencies, including versions, for rendering the document.
+It introduces new fields to the document header, allowing users to manually declare the versioned dependencies required for rendering the document.
 The package then generates a `Dockerfile` from this metadata and provides a utitility function to render the document inside a Docker container, i.e., \code{render\_docker("foo.Rmd")}.
@@
 
-Containers can also serve as useful playgrounds to create environments ad hoc or to provide very specific environments that are not needed or not easily available in day-to-day development for the purposes of developing R packages.
+Containers can also serve as playgrounds and provide specific or ad hoc environments for the purposes of developing R packages.
 These environments may have specific versions of R, of R extension packages, and of system libraries used by R extension packages, and all of the above in a specific combination.
 
-First, such containers can greatly facilitate **fixing bugs and quick evaluation**, because developers and users can readily start and later discard a container to investigate a bug report or try out a piece of software \citep[cf.][]{ooms_opencpu_2017}.
-The container does not affect their regular system.
+First, such containers can greatly facilitate **fixing bugs and code evaluation**, because developers and users can readily start a container to investigate a bug report or try out a piece of software \citep[cf.][]{ooms_opencpu_2017}.
+The container can later be discarded and does not affect their regular system.
 Using the Rocker images with RStudio, these disposable environments lack no development comfort (cf. Section&nbsp;\nameref{compendia}).
-\citet{ooms_opencpu_2017} describes how `docker exec` can be used to get a root shell in a container for customisation during software evaluation.
+\citet{ooms_opencpu_2017} describes how `docker exec` can be used to get a root shell in a container for customisation during software evaluation without writing a `Dockerfile`.
 \citet{eddelbuettel_debugging_2019} describes an example of how a Docker container was used to debug an issue with a package only occurring with a particular version of Fortran, and using tools which are not readily available on all platforms (e.g., not on macOS).
@@
 
-The prevalence of Docker in industry naturally leads to the use of containers with R in such settings as well, as companies already manage platforms in Docker containers.
+The prevalence of Docker in industry naturally leads to the use of R in containers, as companies already manage platforms in Docker containers.
 These products often entail a large amount of open source software in combination with proprietary layers adding the relevant commercialisation features.
@@
 
-The AI @ T-Mobile team created a set of neural network machine learning natural language processing models to help customer care agents manage text-based messages for customers \citep{t-mobile_enterprise_2018}.
-For example, one model quickly identifies whether a message is from a customer \citep[see \pkg{Shiny}-based \href{https://secure.message.t-mobile.com/v1/shiny/is-customer/app/}{demo} further described by ][]{nolis_small_2019}, and others tell which customers are likely to make a repeat purchase.
-If a data scientist creates a machine learning model and exposes it through a \pkg{plumber} API, then someone else on the marketing team could write software that sends different emails depending on that real-time prediction.
+The AI @ T-Mobile team created a set of machine learning models for natural language processing to help customer care agents manage text-based messages from customers \citep{t-mobile_enterprise_2018}.
+For example, one model identifies whether a message is from a customer \citep[see \pkg{Shiny}-based \href{https://secure.message.t-mobile.com/v1/shiny/is-customer/app/}{demo} further described by ][]{nolis_small_2019}, and others tell which customers are likely to make a repeat purchase.
+If a data scientist creates a such a model and exposes it through a \pkg{plumber} API, then someone else on the marketing team can write software that sends different emails depending on that real-time prediction.
 The models are convolutional neural networks that use the \CRANpkg{keras} package \citep{cran_keras} and run in a Rocker container.
@@
 This article is a snapshot of the R corner in a universe of applications built with a many-faced piece of software, Docker.
-`Dockerfile`s and Docker images are the go-to methods for collaboration between roles in an organisation, such as development and IT operations teams, and between parties in the communication of knowledge, such as research workflows or education.
-Docker has become synonymous with applying the concept of containerisation to solve challenges of reproducible environments, e.g., in research and in development \& production, and of scalable deployments as it can easily move processing between machines (e.g., locally, a cloud provider's VM, another cloud provider's Container-as-a-Service).
-Reproducible environments, scalability and efficiency, and portability across platforms/infrastructures are the common themes behind R packages, use cases, and applications in this work.
+`Dockerfile`s and Docker images are the go-to methods for collaboration between roles in an organisation, such as developers and IT operators, and between participants in the communication of knowledge, such as researchers or students.
+Docker has become synonymous with applying the concept of containerisation to solve challenges of reproducible environments, e.g., in research and in development \& production, and of scalable deployments because it can easily move processing between machines, e.g., locally, a cloud provider's VM, another cloud provider's Container-as-a-Service.
+Reproducible environments, scalability \& efficiency, and portability across infrastructures are the common themes behind R packages, use cases, and applications in this work.
 
-The R packages and use cases presented above show the growing number of users, developers, and real-world applications in the community and the resulting innovations.
+The projects presented above show the growing number of users, developers, and real-world applications in the community and the resulting innovations.
 But the applications also point to the challenges of keeping up with a continuously evolving landscape.
-The use cases contributed by co-authors also have a degree of overlap, which can be expected as a common language and understanding of good practices is still taking shape.
-Also, the ease with which one can create a complex software systems with Docker, such as an independent Docker image stack, to serve one's specific needs leads to parallel developments.
+Some use cases have considerable overlap, which can be expected as a common language and understanding of good practices is still taking shape.
+Also, the ease with which one can create a complex software systems with Docker to serve one's specific needs, such as an independent Docker image stack, leads to parallel developments.
 This ease-of-DIY in combination with the difficulty of reusing parts from or composing multiple `Dockerfile`s is a further reason for **fragmentation**.
@@
 Aside from that, fragmentation may not be a bad sign but may instead be a reflection of a growing market that is able to sustain multiple related efforts.
-With the maturing of core building blocks, such as the Rocker Project's suite of images, more systems will be built successfully, but they may also be built behind the curtains.
+With the maturing of core building blocks, such as the Rocker suite of images, more working systems will be built, but they may simply work behind the curtains.
 Docker alone, as a flexible core technology, is not a feasible level of collaboration and abstraction.
@@
 This articles was collaboratively written at \href{https://github.com/nuest/rockerverse-paper/}{https://github.com/nuest/rockerverse-paper/}.
-The \href{https://github.com/nuest/rockerverse-paper/graphs/contributors}{contributors page}, \href{https://github.com/nuest/rockerverse-paper/commits/master}{commit history}, and \href{https://github.com/nuest/rockerverse-paper/issues/}{discussion issues} provide a detailed view on the respective contributions.
+The \href{https://github.com/nuest/rockerverse-paper/graphs/contributors}{contributors page} and \href{https://github.com/nuest/rockerverse-paper/issues/}{discussion issues} provide details on the respective contributions.
 

### 2020-03-04 rockerverse.tex
fix some widows and orphans
https://en.wikipedia.org/wiki/Widows_and_orphans

@@
 
-Docker is not the only containerisation software. An alternative
-stemming from the domain of high-perfomance computing is
-\textbf{Singularity} \citep{kurtzer_singularity_2017}. Singularity can
-run Docker images. Rocker images work out of the box if the main process
-is R, e.g., in \texttt{rocker/r-base}, but Singularity does not succeed
-in running images where there is an init script, e.g., in containers
-that by default run RStudio~Server. In the latter case, a
-\texttt{Singularity} file, a recipe akin to a \texttt{Dockerfile}, needs
-to be used to make necessary adjustments. To date, no comparable image
-stack to the Rocker Project's images exists on
+Docker is not the only containerisation software. \textbf{Singularity}
+stems from the domain of high-perfomance computing
+\citep{kurtzer_singularity_2017} and can also run Docker images. Rocker
+images work out of the box if the main process is R, e.g., in
+\texttt{rocker/r-base}, but Singularity does not succeed in running
+images where there is an init script, e.g., in containers that by
+default run RStudio~Server. In the latter case, a \texttt{Singularity}
+file, a recipe akin to a \texttt{Dockerfile}, needs to be used to make
+necessary adjustments. To date, no comparable image stack to the Rocker
+Project's images exists on
 \href{https://singularity-hub.org/}{Singularity Hub}. A further tool for
@@
 
-The \emph{Bioconductor} \texttt{Dockerfile} selects the desired version
-of R from Rocker, adds required system dependencies, and uses the
-\CRANpkg{BiocManager} package for installing appropriate versions of
+The \emph{Bioconductor} \texttt{Dockerfile} selects the desired R
+version from Rocker images, adds required system dependencies, and uses
+the \CRANpkg{BiocManager} package for installing appropriate versions of
 \emph{Bioconductor} packages \citep{cran_biocmanager}. A strength of
 this approach is that the responsibility for complex software
-configuration (including customized development) is shifted from the
-user to the experienced \emph{Bioconductor} core team. However, a recent
-audit of the \emph{Bioconductor} image stack \texttt{Dockerfile} led to
-the deprecation of several community-maintained images, because the
-numerous specific images became too hard to understand, complex to
-maintain, and cumbersome to customise. As part of the simplification, a
-recent innovation is to produce a \texttt{bioconductor\_docker:devel}
-image to emulate the \emph{Bioconductor} nightly Linux build machine as
-closely as possible. This image contains the build system environment
-variables and the \emph{system dependencies} needed to install and check
-almost all (1813 out of 1823) \emph{Bioconductor} software packages, and
-it saves users and package developers from managing these themselves.
+configuration and customization is shifted from the user to the
+experienced \emph{Bioconductor} core team. However, a recent audit of
+the \emph{Bioconductor} image stack \texttt{Dockerfile} led to the
+deprecation of several community-maintained images, because the numerous
+specific images became too hard to understand, complex to maintain, and
+cumbersome to customise. As part of the simplification, a recent
+innovation is the \texttt{bioconductor\_docker:devel} image, which
+emulates the \emph{Bioconductor} environment for nightly builds as
+closely as possible. This image contains the environment variables and
+the system dependencies needed to install and check almost all
+\emph{Bioconductor} software packages (1813 out of 1823). It saves users
+and package developers from creating this environment themselves.
 Furthermore, the image is configured so that \code{.libPaths()} has
@@
 
-\textbf{\CRANpkg{liftr}} aims to solve the problem of persistent
-reproducible reporting in statistical computing based on the R Markdown
-format \citep{xie2018} for dynamic documents
-\citep[\href{https://nanx.me/liftr/}{https://nanx.me/liftr/}, ][]{liftr2019}.
-The irreproducibility of authoring environments can become an issue for
-collaborative documents and large-scale platforms for processing
-documents. \pkg{liftr} makes the document the main and sole workflow
-control file and the only file that needs to be shared between
-collaborators for consistent environments, e.g.~demonstrated in the
-DockFlow project (\url{https://dockflow.org}). It introduces new fields
-to the R Markdown document header, allowing users to manually declare
-the dependencies, including versions, for rendering the document. The
-package then generates a \texttt{Dockerfile} from this metadata and
-provides a utitility function to render the document inside a Docker
-container, i.e., \code{render\_docker("foo.Rmd")}. An RStudio addin even
-allows compilation of documents with the single push of a button.
+\textbf{\CRANpkg{liftr}} \citep{liftr2019} aims to solve the problem of
+persistent reproducible reporting in statistical computing based on the
+R Markdown format \citep{xie2018}. The irreproducibility of authoring
+environments can become an issue for collaborative documents and
+large-scale platforms for processing documents. \pkg{liftr} makes the
+dynamic R Markdown document the main and sole workflow control file and
+the only file that needs to be shared between collaborators for
+consistent environments, e.g.~demonstrated in the DockFlow project
+(\url{https://dockflow.org}). It introduces new fields to the document
+header, allowing users to manually declare the versioned dependencies
+required for rendering the document. The package then generates a
+\texttt{Dockerfile} from this metadata and provides a utitility function
+to render the document inside a Docker container, i.e.,
+\code{render\_docker("foo.Rmd")}. An RStudio addin even allows
+compilation of documents with the single push of a button.
 
@@
 
-Containers can also serve as useful playgrounds to create environments
-ad hoc or to provide very specific environments that are not needed or
-not easily available in day-to-day development for the purposes of
-developing R packages. These environments may have specific versions of
-R, of R extension packages, and of system libraries used by R extension
-packages, and all of the above in a specific combination.
+Containers can also serve as playgrounds and provide specific or ad hoc
+environments for the purposes of developing R packages. These
+environments may have specific versions of R, of R extension packages,
+and of system libraries used by R extension packages, and all of the
+above in a specific combination.
 
 First, such containers can greatly facilitate \textbf{fixing bugs and
-quick evaluation}, because developers and users can readily start and
-later discard a container to investigate a bug report or try out a piece
-of software \citep[cf.][]{ooms_opencpu_2017}. The container does not
-affect their regular system. Using the Rocker images with RStudio, these
-disposable environments lack no development comfort
+code evaluation}, because developers and users can readily start a
+container to investigate a bug report or try out a piece of software
+\citep[cf.][]{ooms_opencpu_2017}. The container can later be discarded
+and does not affect their regular system. Using the Rocker images with
+RStudio, these disposable environments lack no development comfort
 (cf.~Section~\nameref{compendia}). \citet{ooms_opencpu_2017} describes
 how \texttt{docker\ exec} can be used to get a root shell in a container
-for customisation during software evaluation.
-\citet{eddelbuettel_debugging_2019} describes an example of how a Docker
-container was used to debug an issue with a package only occurring with
-a particular version of Fortran, and using tools which are not readily
-available on all platforms (e.g., not on macOS).
+for customisation during software evaluation without writing a
+\texttt{Dockerfile}. \citet{eddelbuettel_debugging_2019} describes an
+example of how a Docker container was used to debug an issue with a
+package only occurring with a particular version of Fortran, and using
+tools which are not readily available on all platforms (e.g., not on
+macOS).
 
@@
 
-The prevalence of Docker in industry naturally leads to the use of
-containers with R in such settings as well, as companies already manage
-platforms in Docker containers. These products often entail a large
-amount of open source software in combination with proprietary layers
-adding the relevant commercialisation features. One such example is
-RStudio's data science platform
-\href{https://rstudio.com/products/team/}{RStudio Team}. It allows teams
-of data scientists and their respective IT/DevOps groups to develop and
-deploy code in R and Python around the RStudio Open Source Server inside
-of Docker images, without requiring users to learn new tools or directly
-interact with containers. The best practices for
+The prevalence of Docker in industry naturally leads to the use of R in
+containers, as companies already manage platforms in Docker containers.
+These products often entail a large amount of open source software in
+combination with proprietary layers adding the relevant
+commercialisation features. One such example is RStudio's data science
+platform \href{https://rstudio.com/products/team/}{RStudio Team}. It
+allows teams of data scientists and their respective IT/DevOps groups to
+develop and deploy code in R and Python around the RStudio Open Source
+Server inside of Docker images, without requiring users to learn new
+tools or directly interact with containers. The best practices for
 \href{https://support.rstudio.com/hc/en-us/articles/360021594513-Running-RStudio-with-Docker-containers}{running
@@
 
-The AI @ T-Mobile team created a set of neural network machine learning
-natural language processing models to help customer care agents manage
-text-based messages for customers \citep{t-mobile_enterprise_2018}. For
-example, one model quickly identifies whether a message is from a
-customer
+The AI @ T-Mobile team created a set of machine learning models for
+natural language processing to help customer care agents manage
+text-based messages from customers \citep{t-mobile_enterprise_2018}. For
+example, one model identifies whether a message is from a customer
 \citep[see \pkg{Shiny}-based \href{https://secure.message.t-mobile.com/v1/shiny/is-customer/app/}{demo} further described by ][]{nolis_small_2019},
 and others tell which customers are likely to make a repeat purchase. If
-a data scientist creates a machine learning model and exposes it through
-a \pkg{plumber} API, then someone else on the marketing team could write
+a data scientist creates a such a model and exposes it through a
+\pkg{plumber} API, then someone else on the marketing team can write
 software that sends different emails depending on that real-time
@@
 and Docker images are the go-to methods for collaboration between roles
-in an organisation, such as development and IT operations teams, and
-between parties in the communication of knowledge, such as research
-workflows or education. Docker has become synonymous with applying the
-concept of containerisation to solve challenges of reproducible
-environments, e.g., in research and in development \& production, and of
-scalable deployments as it can easily move processing between machines
-(e.g., locally, a cloud provider's VM, another cloud provider's
-Container-as-a-Service). Reproducible environments, scalability and
-efficiency, and portability across platforms/infrastructures are the
-common themes behind R packages, use cases, and applications in this
-work.
-
-The R packages and use cases presented above show the growing number of
-users, developers, and real-world applications in the community and the
+in an organisation, such as developers and IT operators, and between
+participants in the communication of knowledge, such as researchers or
+students. Docker has become synonymous with applying the concept of
+containerisation to solve challenges of reproducible environments, e.g.,
+in research and in development \& production, and of scalable
+deployments because it can easily move processing between machines,
+e.g., locally, a cloud provider's VM, another cloud provider's
+Container-as-a-Service. Reproducible environments, scalability \&
+efficiency, and portability across infrastructures are the common themes
+behind R packages, use cases, and applications in this work.
+
+The projects presented above show the growing number of users,
+developers, and real-world applications in the community and the
 resulting innovations. But the applications also point to the challenges
-of keeping up with a continuously evolving landscape. The use cases
-contributed by co-authors also have a degree of overlap, which can be
-expected as a common language and understanding of good practices is
-still taking shape. Also, the ease with which one can create a complex
-software systems with Docker, such as an independent Docker image stack,
-to serve one's specific needs leads to parallel developments. This
-ease-of-DIY in combination with the difficulty of reusing parts from or
-composing multiple \texttt{Dockerfile}s is a further reason for
-\textbf{fragmentation}. Instructions can be outsourced into
-distributable scripts and then copied into the image during build, but
-that makes \texttt{Dockerfile}s harder to read. Scripts added to a
-\texttt{Dockerfile} also add a layer of complexity and increase the risk
-of incomplete recipes. Despite the different image stacks presented
-here, the pervasiveness of Rocker images can be traced back to its
-maintainers and the user community valuing collaboration and shared
-starting points over impulses to create individual solutions. Aside from
-that, fragmentation may not be a bad sign but may instead be a
-reflection of a growing market that is able to sustain multiple related
-efforts. With the maturing of core building blocks, such as the Rocker
-Project's suite of images, more systems will be built successfully, but
-they may also be built behind the curtains. Docker alone, as a flexible
-core technology, is not a feasible level of collaboration and
+of keeping up with a continuously evolving landscape. Some use cases
+have considerable overlap, which can be expected as a common language
+and understanding of good practices is still taking shape. Also, the
+ease with which one can create a complex software systems with Docker to
+serve one's specific needs, such as an independent Docker image stack,
+leads to parallel developments. This ease-of-DIY in combination with the
+difficulty of reusing parts from or composing multiple
+\texttt{Dockerfile}s is a further reason for \textbf{fragmentation}.
+Instructions can be outsourced into distributable scripts and then
+copied into the image during build, but that makes \texttt{Dockerfile}s
+harder to read. Scripts added to a \texttt{Dockerfile} also add a layer
+of complexity and increase the risk of incomplete recipes. Despite the
+different image stacks presented here, the pervasiveness of Rocker
+images can be traced back to its maintainers and the user community
+valuing collaboration and shared starting points over impulses to create
+individual solutions. Aside from that, fragmentation may not be a bad
+sign but may instead be a reflection of a growing market that is able to
+sustain multiple related efforts. With the maturing of core building
+blocks, such as the Rocker suite of images, more working systems will be
+built, but they may simply work behind the curtains. Docker alone, as a
+flexible core technology, is not a feasible level of collaboration and
 abstraction. Instead, the use cases and applications observed in this
@@
 The
-\href{https://github.com/nuest/rockerverse-paper/graphs/contributors}{contributors page},
-\href{https://github.com/nuest/rockerverse-paper/commits/master}{commit history},
+\href{https://github.com/nuest/rockerverse-paper/graphs/contributors}{contributors page}
 and
 \href{https://github.com/nuest/rockerverse-paper/issues/}{discussion issues}
-provide a detailed view on the respective contributions.
+provide details on the respective contributions.
 

### 2020-03-08 rockerverse.Rmd
fix multiply defined label
small edit thanks to Alton

@@
 ## Teaching
-\label{workenvs}
 
@@
 PrairieLearn is being developed at the University of Illinois at Urbana-Champaign \citep{prairielearn:2018} and has been in extensive use across several faculties along with initial use on some other campuses.
-¸It uses Docker containers a key components, both internally for its operations (programmed mainly in Python as well as in Javascript), as well as for two reference containers providing, respectively, Python and R autograders.
+It uses Docker containers as key components, both internally for its operations (programmed mainly in Python as well as in Javascript), as well as for two reference containers providing, respectively, Python and R autograders.
 A key design decision made by PrairieLearn permits _external_ grading containers to be supplied and accessed via a well-defined interface of invoking, essentially, a single script, `run.sh`.
-This script relies on a well-defined file layout containing JSON-based configuration, support files, exam questions, supplementary data, and student submissions.
+This script relies on a well-defined file layout containing JSON-based configurations, support files, exam questions, supplementary data, and student submissions.
 It returns per-question evaluations as JSON result files, which PrarieLearn evaluates, aggregates and records in a database.
@@
 The package \CRANpkg{ttdo} \citep{CRAN:ttdo} utilizes the extensibility of \pkg{tinytest} to display context-sensitive colorized differences between incorrect answers and reference answers using the \CRANpkg{diffobj} package \citep{CRAN:diffobj}.
+Additionally, \CRANpkg{ttdo} addresses the issue of insufficient information collection that Dodona faced by allowing for the collection of arbitrary, test specific attributes for additional logging and feedback.
 The setup, described in more detail by \citet{paper:r_autograder}, is an excellent illustration of both the versatility and flexibility offered by Docker-based approaches in teaching and testing.

### 2020-03-09 rockerverse.Rmd
add ó for @daroczig
Update rockerverse.Rmd

@@
     email: dav@gigantum.com
-  - name: Gergely Daroczi
+  - name: Gergely Daróczi
     # https://github.com/daroczig
@@
   - name: Mark Edmondson
-    affiliation: IIH Nordic A/S, Google Developer Expert for GCP
+    # https://github.com/MarkEdmondson1234
+    affiliation: IIH Nordic A/S, Google Developer Expert for Google Cloud Platform
     email: mark@markedmondson.me
+    address:
+      - Artillerivej 86
+      - 2300 København S, Denmark
+      - \orcid{0000-0002-8434-3881}
   - name: Colin Fay
@@
       - New York, NY 10001, USA
-      - \orcid{0000-0002-0233-1757}
+      - \orcid{0000-0002-2136-0000}
     email: ross@ecohealthalliance.org
@@
   "dockyard", c("dockerfile", "build", "execute"), # cloud?
-  "dockermachine", c("install", "execute", "cloud", "interact")
+  "dockermachine", c("install", "execute", "cloud", "interact"),
+  "googleCloudRunner", c("build", "execute", "cloud")
 )
@@
 The CLI tool `docker-machine` allows users to create and manage a virtual host on local computers, local data centers, or at cloud providers.
-A local Docker installation can be configured to transparntly forward all commands issued on the local Docker CLI to a selected (remote) virtual host.
+A local Docker installation can be configured to transparently forward all commands issued on the local Docker CLI to a selected (remote) virtual host.
 Docker&nbsp;Machine was especially crucial for local use in the early days of Docker, when no native support was available for Mac or Windows computers, but it remains relevant for provisioning on remote systems.

### 2020-03-09 rockerverse.Rmd
Add NR support to awknowledgements
use citation for dynverse

--- 2020-03-08 rockerverse.Rmd
@@
   - name: Mark Edmondson
+    # https://github.com/MarkEdmondson1234
     affiliation: IIH Nordic A/S, Google Developer Expert for GCP
@@
 A typical compendium's `Dockerfile` will pull a rocker image fixed to a specific version of R, and install R packages from the MRAN repository to ensure the package versions are tied to a specific date, rather than the most recent version.
-A more extreme case is the _dynverse_ project ([https://dynverse.org/](https://dynverse.org/)), which packages over 50 computational methods with different environments (R, Python, C++, etc.) in Docker images, which can be executed from R.
+A more extreme case is the _dynverse_ project \citep{saelens_comparisonsinglecelltrajectory_2019}, which packages over 50 computational methods with different environments (R, Python, C++, etc.) in Docker images, which can be executed from R.
 _dynverse_ uses a CI platform (see&nbsp;\nameref{ci}) to build Rocker-derived images, test them, and, if the tests succeed, publish them on Docker&nbsp;Hub.
@@
 CW is supported by the Whole Tale project (https://wholetale.org) funded by the US National Science Foundation (NSF) under award [OAC-1541450](https://www.nsf.gov/awardsearch/showAward?AWD_ID=1541450).
+NR is supported in part by the Chan-Zuckerberg Initiative Essential Open Source Software for Science program.
 We would like to thank Celeste R. Brennecka from the Science Writing Support Service of the University of Münster for her editorial support.

### 2020-03-09 rockerverse.Rmd
fix typos and grammatical errors

@@
   The Rocker Project provides widely used Docker images for R across different application scenarios.
-  This articles surveys downstream projects that build upon the Rocker Project images and presents the current state of R packages for managing Docker images and controlling containers.
+  This article surveys downstream projects that build upon the Rocker Project images and presents the current state of R packages for managing Docker images and controlling containers.
   These use cases cover diverse topics such as package development, reproducible research, collaborative work, cloud-based data processing, and production deployment of services.
@@
 We also introduce related activities that connect the R language and environment with other containerisation solutions.
-Our main contributions is a coherent picture of the current status of using containers in, with, and for R.
+Our main contribution is a coherent picture of the current status of using containers in, with, and for R.
 
@@
 Docker, an application and service provided by the eponymous company, has, in just a few short years, risen to prominence for developing, testing, deploying and distributing computer software \citep[cf.][]{datadog_8_2018,munoz_history_2019}.
-While related approaches exist, such as LXC\footnote{\href{https://en.wikipedia.org/wiki/LXC}{https://en.wikipedia.org/wiki/LXC}} or Singularity \citep{kurtzer_singularity_2017}, Docker has become synomymous with "containerisation"---the method of taking software artefacts and bundling them in such a way that use becomes standardized and portable across operating systems.
+While related approaches exist, such as LXC\footnote{\href{https://en.wikipedia.org/wiki/LXC}{https://en.wikipedia.org/wiki/LXC}} or Singularity \citep{kurtzer_singularity_2017}, Docker has become synonymous with "containerisation"---the method of taking software artefacts and bundling them in such a way that use becomes standardized and portable across operating systems.
 In doing so, Docker had recognised and validated the importance of one very important thread that had been emerging, namely virtualisation.
@@
 
-For statistical computing and analysis centered around R, the **Rocker&nbsp;Project** has provided a variety of Docker containers since it began in 2014 \citep{RJ-2017-065}.
+For statistical computing and analysis centred around R, the **Rocker&nbsp;Project** has provided a variety of Docker containers since it began in 2014 \citep{RJ-2017-065}.
 The Rocker&nbsp;Project provides several lines of containers spanning from building blocks with `R-release` or `R-devel`, via containers with [RStudio&nbsp;Server](https://rstudio.com/products/rstudio/) and [Shiny&nbsp;Server](https://rstudio.com/products/shiny/shiny-server/), to domain-specific containers such as [`rocker/geospatial`](https://github.com/rocker-org/geospatial) \citep{rocker_geospatial_2019}.
@@
 Also of note is a series of "versioned" containers which match the R release they contain with the _then-current_ set of packages via the MRAN Snapshot views of CRAN \citep{microsoft_cran_2019}.
-The Rocker&nbsp;Project's impact and importance was acknowledged by the Chan Zuckerberg Initiative's _Essential Open Source Software for Science_, which provides funding for the projects's sustainable maintenance, community growth, and targeting new hardware platforms including GPUs \citep{chan_zuckerberg_initiative_maintaining_2019}.
+The Rocker&nbsp;Project's impact and importance was acknowledged by the Chan Zuckerberg Initiative's _Essential Open Source Software for Science_, which provides funding for the project's sustainable maintenance, community growth, and targeting new hardware platforms including GPUs \citep{chan_zuckerberg_initiative_maintaining_2019}.
 
 Docker is not the only containerisation software.
-**Singularity** stems from the domain of high-perfomance computing \citep{kurtzer_singularity_2017} and can also run Docker images.
+**Singularity** stems from the domain of high-performance computing \citep{kurtzer_singularity_2017} and can also run Docker images.
 Rocker images work out of the box if the main process is R, e.g., in `rocker/r-base`, but Singularity does not succeed in running images where there is an init script, e.g., in containers that by default run RStudio&nbsp;Server.
@@
 A further tool for running containers is [**podman**](https://github.com/containers/libpod), which can also build `Dockerfile`s and run Docker images.
-Proof of concepts exists for using podman to build and run Rocker containers\footnote{See \href{https://github.com/nuest/rodman}{https://github.com/nuest/rodman} and \href{https://github.com/rocker-org/rocker-versioned/issues/187}{https://github.com/rocker-org/rocker-versioned/issues/187}}, but the prevelance of Docker, especially in the broader user community beyond experts or niche systems and the vast amount of blog posts and courses for Docker currently cap specific development efforts for both Singularity and podman in the R community.
+Proof of concepts exists for using podman to build and run Rocker containers\footnote{See \href{https://github.com/nuest/rodman}{https://github.com/nuest/rodman} and \href{https://github.com/rocker-org/rocker-versioned/issues/187}{https://github.com/rocker-org/rocker-versioned/issues/187}}, but the prevalence of Docker, especially in the broader user community beyond experts or niche systems and the vast amount of blog posts and courses for Docker currently cap specific development efforts for both Singularity and podman in the R community.
 This might quickly change if the usability and spread of Singularity or podman increase, or if security features such as rootless/unprivileged containers, which both these tools support out of the box, become more sought after.
@@
 
-**\pkg{harbor}** ([https://github.com/wch/harbor](https://github.com/wch/harbor)) is no longer actively maintained, but it should be honorably mentioned as the first R package for managing Docker images and containers.
+**\pkg{harbor}** ([https://github.com/wch/harbor](https://github.com/wch/harbor)) is no longer actively maintained, but it should be honourably mentioned as the first R package for managing Docker images and containers.
 It uses the \CRANpkg{sys} package \citep{cran_sys} to run system commands against the Docker CLI, both locally and through an SSH connection, and it has convenience functions, e.g., for listing and removing containers/images and for accessing logs.
 The outputs of container executions are converted to appropriate R types.
-The Docker CLI's basic functionality, although it evolves quickly and with little concern for avoiding breaking changes, has remained unchanged in core functions, meaning that a core function such as \code{harbor::docker\_run(image = "hello-world")} still works depsite its stopped development.
+The Docker CLI's basic functionality, although it evolves quickly and with little concern for avoiding breaking changes, has remained unchanged in core functions, meaning that a core function such as \code{harbor::docker\_run(image = "hello-world")} still works despite its stopped development.
 
@@
 **\pkg{dockermachine}** ([https://github.com/cboettig/dockermachine](https://github.com/cboettig/dockermachine)) is an R package to provide a convenient interface to [Docker&nbsp;Machine](https://docs.docker.com/machine/overview/) from R.
-The CLI tool `docker-machine` allows users to create and manage a virtual host on local computers, local data centers, or at cloud providers.
-A local Docker installation can be configured to transparntly forward all commands issued on the local Docker CLI to a selected (remote) virtual host.
+The CLI tool `docker-machine` allows users to create and manage a virtual host on local computers, local data centres, or at cloud providers.
+A local Docker installation can be configured to transparently forward all commands issued on the local Docker CLI to a selected (remote) virtual host.
 Docker&nbsp;Machine was especially crucial for local use in the early days of Docker, when no native support was available for Mac or Windows computers, but it remains relevant for provisioning on remote systems.
@@
 
-**Bioconductor** ([https://bioconductor.org/](https://bioconductor.org/)) is an open source, open development project for the analysis and comprehension of genomic data \citep{gentleman_bioconductor_2004}.
+**Bioconductor** ([https://bioconductor.org/](https://bioconductor.org/)) is an open-source, open development project for the analysis and comprehension of genomic data \citep{gentleman_bioconductor_2004}.
 As of October 30th 2019, the project consists of 1823 R software packages, as well as packages containing annotation or experiment data.
@@
 _Bioconductor_ installs packages based on the R version in combination with the Bioconductor version and, therefore, uses Bioconductor version tagging `devel` and `RELEASE_X_Y`, e.g., `RELEASE_3_10`.
-Past and current combinations of R and _Bioconductor_ will therefore be accessible via a specific image tags.
+Past and current combinations of R and _Bioconductor_ will therefore be accessible via specific image tags.
 
@@
 Users mounting a location on the host file system to this location can persistently manage installed packages across Docker containers or image updates.
-Many R users pursue flexible work flows tailored to particular analysis needs rather than standardized work flows.
+Many R users pursue flexible workflows tailored to particular analysis needs rather than standardized workflows.
 The new `bioconductor_docker` image is well suited for this preference, while `bioconductor_docker:devel` provides developers with a test environment close to _Bioconductor_'s build system.
@@
 **Data science** is a widely discussed topic in all academic disciplines \citep[e.g.,][]{donoho_50_2017}.
-These discussions have shed a light on the tools and craftspersonship behind the analysis of data with computational methods.
+These discussions have shed light on the tools and craftspersonship behind the analysis of data with computational methods.
 The practice of data science often involves combining tools and software stacks and requires a cross-cutting skillset.
@@
 The independent image stack can be traced back to the Gigantum environment and its features.
-The R images are based on Gigantum's `python3-minimal` image, originally to keep the existing front-end configuration, but also to provide a consistent Python-to-R interoperability.
+The R images are based on Gigantum's `python3-minimal` image, originally to keep the existing front-end configuration, but also to provide consistent Python-to-R interoperability.
 The `Dockerfile`s also use build args to specify bases, for example for different versions of NVIDIA CUDA for GPU processing\footnote{See \href{https://github.com/gigantum/base-images/blob/master/_templates/python3-minimal-template/Dockerfile}{https://github.com/gigantum/base-images/blob/master/\_templates/python3-minimal-template/Dockerfile} for the \texttt{Dockerfile} of \texttt{python3-minimal}.}, so that appropriate GPU drivers can be enabled automatically when supported.
@@
 Because of these specifics, instructions from `rocker/r-ubuntu` could _not_ be readily re-used in this image stack (see Section&nbsp;\nameref{conclusions}).
-Both approaches enable the `apt` package manager \citep{wikipedia_contributors_apt_2020} as an installation method, and this is exposed via the GUI-based environment management,\footnote{See \href{https://docs.gigantum.com/docs/environment-management}{https://docs.gigantum.com/docs/environment-management}}.
+Both approaches enable the `apt` package manager \citep{wikipedia_contributors_apt_2020} as an installation method, and this is exposed via the GUI-based environment management\footnote{See \href{https://docs.gigantum.com/docs/environment-management}{https://docs.gigantum.com/docs/environment-management}}.
 The image build and publication process is scripted with Python and JSON template configuration files, unlike Rocker images which rely on plain `Dockerfile`s.
@@
 \pkg{containerit} accepts an R object of classes \code{"sessionInfo"} or \code{"session\_info"} as input and provides helper functions to derive these from workflows, e.g., an R script or R Markdown document, by analysing the session state at the end of the workflow.
-It relies on the \pkg{sysreqs} ([https://github.com/r-hub/sysreqs/](https://github.com/r-hub/sysreqs/)) package and it's mapping of package system dependencies to platform specific installation package names\footnote{See \href{https://sysreqs.r-hub.io/}{https://sysreqs.r-hub.io/}.}.
+It relies on the \pkg{sysreqs} ([https://github.com/r-hub/sysreqs/](https://github.com/r-hub/sysreqs/)) package and it's mapping of package system dependencies to platform-specific installation package names\footnote{See \href{https://sysreqs.r-hub.io/}{https://sysreqs.r-hub.io/}.}.
 \pkg{containerit} uses \pkg{stevedore} to streamline the user interaction and improve the created `Dockerfile`s, e.g., by running a container for the desired base image to extract the already available R packages.
@@
 It introduces new fields to the document header, allowing users to manually declare the versioned dependencies required for rendering the document.
-The package then generates a `Dockerfile` from this metadata and provides a utitility function to render the document inside a Docker container, i.e., \code{render\_docker("foo.Rmd")}.
+The package then generates a `Dockerfile` from this metadata and provides a utility function to render the document inside a Docker container, i.e., \code{render\_docker("foo.Rmd")}.
 An RStudio addin even allows compilation of documents with the single push of a button.
 
-System dependencies are the domain of Docker, but for a full description of the computing environment one must also manage the R version and the the R packages.
+System dependencies are the domain of Docker, but for a full description of the computing environment, one must also manage the R version and the R packages.
 R versions are available via the versioned Rocker image stack \citep{RJ-2017-065}.
@@
 While \pkg{renv} does not cover system dependencies, the \pkg{renv}-based environment can be transferred into a container either by restoring the environment based on the main configuration file `renv.lock` or by storing the \pkg{renv}-cache on the host and not in the container \citep{ushey_using_2019}.
-With both the system dependencies and R packages consciously managed in a Docker image, users can start using containers as the _only_ environment for their workflows, which allows them to work indepenently of physical computers\footnote{Allowing them to be digital "nomads", cf.
+With both the system dependencies and R packages consciously managed in a Docker image, users can start using containers as the _only_ environment for their workflows, which allows them to work independently of physical computers\footnote{Allowing them to be digital "nomads", cf.
 J.~Bryan's \href{https://github.com/jennybc/docker-why}{https://github.com/jennybc/docker-why}.} and to assert a specific degree of confidence in the stability of a developed software \citep[cf.
@@
 These images significantly reduce the barriers to following complex steps for fixing memory allocation bugs \citep[cf. Section~4.3 in][]{core_writing_1999}.
-These problems are hard to debug and critical, because when they do occur they lead to fatal crashing processes.
+These problems are hard to debug and critical, because when they do occur they lead to fatal crashes.
 [`rocker/r-devel-san`](https://github.com/rocker-org/r-devel-san) and [`rocker/r-devel-ubsan-clang`](https://github.com/rocker-org/r-devel-san-clang) are Docker images that have a particularly configured version of R to trace such problems with gcc and clang compilers, respectively \citep[cf.~\CRANpkg{sanitizers} for examples,][]{eddelbuettel_sanitizers_2014}.
@@
 These images are accompanied by a special launch script so the tested source code is not stored in the image; instead, the currently checked in version from a local Git repository is cloned into the container at runtime.
-This approach clearly separates test environment, test code, and current working copy of the code.
+This approach separates the test environment, test code, and current working copy of the code.
 \label{rselenium} Another use case where a container can help to standardise tests across operating systems is detailed the vignettes of the package \CRANpkg{RSelenium} \citep{rselenium_2019}.
@@
 Furthermore, with each change, even this manuscript is rendered into a PDF and deployed to a GitHub-hosted website (see `.travis.yml` and `Dockerfile` in the manuscript repository), not because of concern about time, but to control the environment used on a CI server.
-This gives, on the one hand, easy access after every update of the R Markdown source code and, on the other hand a second controlled environment to make sure that the article renders successfully and correctly.
+This gives, on the one hand, easy access after every update of the R Markdown source code and, on the other hand, a second controlled environment to make sure that the article renders successfully and correctly.
 
@@
 Instead of running a local R script with the local CPU and RAM restrictions, the same code can be processed on all CPU threads of the cluster of machines in the cloud, all running a Docker container with the same R environments.
-\pkg{googleComputeEngineR} integrates with the R parallelisation package \CRANpkg{future} \citep{cran_future} to enable this with only a few lines or R code\footnote{\href{https://cloudyr.github.io/googleComputeEngineR/articles/massive-parallel.html}{https://cloudyr.github.io/googleComputeEngineR/articles/massive-parallel.html}}.
+\pkg{googleComputeEngineR} integrates with the R parallelisation package \CRANpkg{future} \citep{cran_future} to enable this with only a few lines of R code\footnote{\href{https://cloudyr.github.io/googleComputeEngineR/articles/massive-parallel.html}{https://cloudyr.github.io/googleComputeEngineR/articles/massive-parallel.html}}.
 [Google Cloud Run](https://cloud.run) is a CaaS (Containers as a Service) platform.
@@
 Containers allow for building complex processing pipelines that are independent of the host programming language.
-Due to its original use case (see&nbsp;\nameref{introduction}), Docker has no standard mechanisms for chaining containers together; it lacks definitions and protocols for how to use environment variables, volume mounts, and/or ports that could enable transfer of input (parameters and data) and output (results) to and from containers.
+Due to its original use case (see&nbsp;\nameref{introduction}), Docker has no standard mechanisms for chaining containers together; it lacks definitions and protocols for how to use environment variables, volume mounts, and/or ports that could enable the transfer of input (parameters and data) and output (results) to and from containers.
 Some packages, e.g., \pkg{containerit}, provide Docker images that can be used very similar to a CLI, but this usage is cumbersome\footnote{\href{https://o2r.info/containerit/articles/container.html}{https://o2r.info/containerit/articles/container.html}}.
-**\pkg{outsider}** ([https://docs.ropensci.org/outsider/](https://docs.ropensci.org/outsider/)) tackles the problem of integrating external programs into an R workflow without thee need for users to directly interact with containers \citep{bennett_outsider_2020}.
+**\pkg{outsider}** ([https://docs.ropensci.org/outsider/](https://docs.ropensci.org/outsider/)) tackles the problem of integrating external programs into an R workflow without the need for users to directly interact with containers \citep{bennett_outsider_2020}.
 Installation and usage of external programs can be difficult, convoluted and even impossible if the platform is incompatible.
@@
 Each outsider module has a `Dockerfile` and an R package with functions for interacting with the encapsulated tool.
-Using only R functions, an end user can install a module with the \pkg{outsider} package and then call module code to seamlessly integrate a tool into their own R-based workflow.
+Using only R functions, an end-user can install a module with the \pkg{outsider} package and then call module code to seamlessly integrate a tool into their own R-based workflow.
 The \pkg{outsider} package and module manage the containers and handle the transmission of arguments and the transfer of files to and from a container.
@@
 \pkg{Shiny} is a popular package for creating interactive online dashboards with R, and it enables users with very diverse backgrounds to create stable and user-friendly web applications \citep{cran_shiny}.
-_ShinyProxy_ ([https://www.shinyproxy.io/](https://www.shinyproxy.io/)) is an open source tool to deploy Shiny apps in an enterprise context, where it features single sign-on, but it can also be used in scientific use cases \citep[e.g., ][]{savini_epiexplorer_2019,glouzon_structurexplor_2017}.
+_ShinyProxy_ ([https://www.shinyproxy.io/](https://www.shinyproxy.io/)) is an open-source tool to deploy Shiny apps in an enterprise context, where it features single sign-on, but it can also be used in scientific use cases \citep[e.g., ][]{savini_epiexplorer_2019,glouzon_structurexplor_2017}.
 ShinyProxy uses Docker containers to isolate user sessions and to achieve scalability for multi-user scenarios with multiple apps.
@@
 
-Another example is the package \CRANpkg{golem}, which makes an heavy use of \pkg{dockerfiler} when it comes to creating the `Dockerfile` for building and deploying production-grade Shiny applications \citep{cran_golem}.
+Another example is the package \CRANpkg{golem}, which makes heavy use of \pkg{dockerfiler} when it comes to creating the `Dockerfile` for building and deploying production-grade Shiny applications \citep{cran_golem}.
 \CRANpkg{googleComputeEngineR} enables quick deployments of key R services, such as RStudio and Shiny, onto cloud virtual machines (VMs) with Google Cloud Compute Engine \citep{googleComputeEngineR_2019}.
@@
 _Kubernetes_ ([https://kubernetes.io/](https://kubernetes.io/)) is a container-orchestration system for managing container-based application deployment and scaling.
-A _cluster_ of containers, orchestrated as a single deployment, e.g., with Kubernetes, can mitigate limitations on request volumes or a container occupied with computationally intensive task.
+A _cluster_ of containers, orchestrated as a single deployment, e.g., with Kubernetes, can mitigate limitations on request volumes or a container occupied with a computationally intensive task.
 A cluster features load-balancing, autoscaling of containers across numerous servers (in the cloud or on premise), and restarting failed ones.
@@
 It also includes R wrapper functions to install Docker in such a VM, manage images, and control containers straight from R functions.
-These functions are translated to Docker CLI commands and transferred transparently to the respective remove machine using SSH.
+These functions are translated to Docker CLI commands and transferred transparently to the respective remote machine using SSH.
 \pkg{AzureContainers} is an umbrella package that provides interfaces for three commercial services of Microsoft's Azure Cloud, namely [Container Instances](https://azure.microsoft.com/en-us/services/container-instances/) for running individual containers, [Container Registry](https://azure.microsoft.com/en-us/services/container-registry/) for private image distribution, and [Kubernetes Service](https://azure.microsoft.com/en-us/services/kubernetes-service/) for orchestrated deployments.
@@
 `heroku-docker-r` ([https://github.com/virtualstaticvoid/heroku-docker-r](https://github.com/virtualstaticvoid/heroku-docker-r)) is an independent project providing a template for deploying R applications based on Heroku's image stack, including multiple examples for interfacing R with other programming languages.
-Yet the approach requires a manual management of the computing environment.
+Yet the approach requires manual management of the computing environment.
 
@@
 OpenFaas facilitates the deployment of functions and microservices to Kubernetes or Docker Swarm.
-It is language agnostic and provides auto-scaling, metrics, and an API gateway.
+It is language-agnostic and provides auto-scaling, metrics, and an API gateway.
 Reduced boilerplate code is achieved via templates.
-Template for R\footnote{See OpenFaaS R templates at \href{https://github.com/analythium/openfaas-rstats-templates}{https://github.com/analythium/openfaas-rstats-templates}.}
+Templates for R\footnote{See OpenFaaS R templates at \href{https://github.com/analythium/openfaas-rstats-templates}{https://github.com/analythium/openfaas-rstats-templates}.}
 are provided based on Rocker's Debian and R-hub's [r-minimal](https://github.com/r-hub/r-minimal) Alpine images.
@@
 The prevalence of Docker in industry naturally leads to the use of R in containers, as companies already manage platforms in Docker containers.
-These products often entail a large amount of open source software in combination with proprietary layers adding the relevant commercialisation features.
+These products often entail a large amount of open-source software in combination with proprietary layers adding the relevant commercialisation features.
 One such example is RStudio's data science platform [RStudio Team](https://rstudio.com/products/team/).
-It allows teams of data scientists and their respective IT/DevOps groups to develop and deploy code in R and Python around the RStudio Open Source Server inside of Docker images, without requiring users to learn new tools or directly interact with containers.
+It allows teams of data scientists and their respective IT/DevOps groups to develop and deploy code in R and Python around the RStudio Open-Source Server inside of Docker images, without requiring users to learn new tools or directly interact with containers.
 The best practices for [running RStudio with Docker containers](https://support.rstudio.com/hc/en-us/articles/360021594513-Running-RStudio-with-Docker-containers) as well as [Docker images](https://github.com/rstudio/rstudio-docker-products) for RStudio's commercial products are publicly available.
@@
 The software used \pkg{Rserve} and rApache and was deployed in Docker containers.
-The R session behind \pkg{Rserve} acted as a read-only in-memory database, which was extremely fast and scalabale, for the many concurrent rApache processes responding to the live-scoring requests of various divisions of the company.
+The R session behind \pkg{Rserve} acted as a read-only in-memory database, which was extremely fast and scalable, for the many concurrent rApache processes responding to the live-scoring requests of various divisions of the company.
 Similarly deodorised R scripts were responsible for the ETL processes and even the client-facing email, text message and push notification alerts sent in real-time based on card transactions.
@@
 For any R projects that use the Tidyverse suite \citep{wickham_welcome_2019}, the time and resources required to build all dependencies from source can often time out before completion, making it frustrating for the average R user.
-\pkg{holepunch} removes some of these limitations by leveraging Rocker images that contain the Tidyverse along special Jupyter dependencies, and only installs additional packages from CRAN and Bioconductor that are not already part of these images.
+\pkg{holepunch} removes some of these limitations by leveraging Rocker images that contain the Tidyverse along with special Jupyter dependencies, and only installs additional packages from CRAN and Bioconductor that are not already part of these images.
 It short circuits the configuration file parsing in `repo2docker` and starts with the Binder/Tidyverse base images, which eliminates a large part of the build time and, in most cases, results in a Binder instance launching within a minute.
-\pkg{holepunch} also creates a `DESCRIPTION` file for essential metadata and dependency specification, and thereby turns turns any project into a research compendium (see&nbsp;\nameref{compendia}).
+\pkg{holepunch} also creates a `DESCRIPTION` file for essential metadata and dependency specification, and thereby turns any project into a research compendium (see&nbsp;\nameref{compendia}).
 The `Dockerfile` included with the project can also be used to launch an RStudio Server instance locally, i.e., independent of Binder, which is especially useful when more or special computational resources can be provided there.
@@
 Second, the **Whole&nbsp;Tale** project (https://wholetale.org) combines the strengths of the Rocker Project's curated Docker images with `repo2docker`.
-Whole&nbsp;Tale is a National Science Foundation (NSF) funded project developing a scalable, open source, multi-user platform for reproducible research \citep{brinckman2019, chard2019a}.
+Whole&nbsp;Tale is a National Science Foundation (NSF) funded project developing a scalable, open-source, multi-user platform for reproducible research \citep{brinckman2019, chard2019a}.
 A central goal of the platform is to enable researchers to easily create and publish executable research objects\footnote{In Whole~Tale a \emph{tale} is a research object that contains metadata, data (by copy or reference), code, narrative, documentation, provenance, and information about the computational environment to support computational reproducibility.} associated with published research \citep{chard2019b}.
@@
 
-[CyVerse](https://cyverse.org) is an open source, NSF-funded cyberinfrastructure platform for the life sciences providing easy access to computing and storage resources \citep{merchant_iplant_2016}.
+[CyVerse](https://cyverse.org) is an open-source, NSF-funded cyberinfrastructure platform for the life sciences providing easy access to computing and storage resources \citep{merchant_iplant_2016}.
 CyVerse has a browser-based 'data science workbench' called the [Discovery Environment](https://cyverse.org/discovery-environment) (DE).
@@
 The CyVerse Rocker images further include tools for connecting to its Data Store, such as the CLI utility `icommands` for iRODS.
-CyVerse accounts are free (with some limitations for non-US users), and the [CyVerse Learning Center](https://learning.cyverse.org/) provides community members with information about the platform, incuding training and education opportunities.
+CyVerse accounts are free (with some limitations for non-US users), and the [CyVerse Learning Center](https://learning.cyverse.org/) provides community members with information about the platform, including training and education opportunities.
 
 \label{rocker-gpu}
-Using **GPUs** (graphical processing units) as a specialised hardware from containerised common work environments is also possible and useful \citep{haydel_enhancing_2015}.
+Using **GPUs** (graphical processing units) as specialised hardware from containerised common work environments is also possible and useful \citep{haydel_enhancing_2015}.
 GPUs are increasingly popular for compute-intensive machine learning (ML) tasks, e.g., deep artificial neural networks \citep{schmidhuber_deep_2015}.
@@
 Two use cases demonstrate the practical usefulness and advantages of containerisation in the context of teaching.
-On the one hand a special case of shared computing environments (see Section&nbsp;\ref{workenvs}), and on the other hand leveraging sandboxing and controlled environments for autograding.
+On the one hand a special case of shared computing environments (see Section&nbsp;\ref{workenvs}), and on the other hand leveraging sandboxing and controlled environments for auto-grading.
 
@@
 An example for mitigating _complex setups_ is teaching databases.
-R is very useful tool for interfacing with databases, because almost every open source and proprietary database system has an R package that allows users to connect and interact with it.
+R is very useful tool for interfacing with databases, because almost every open-source and proprietary database system has an R package that allows users to connect and interact with it.
 This flexibility is even broadened by \CRANpkg{DBI} \citep{cran_dbi}, which allows for creating a common API for interfacing these databases, or the \CRANpkg{dbplyr} package \citep{cran_dbplyr}, which runs \CRANpkg{dplyr} \citep{cran_dplyr} code straight against the database as queries.
@@
 The testing process in the Docker containers sends back the test results by writing a JSON document to its standard output channel.
-In June 2019, R support was added to Dodona using an image derived from the `rocker/r-base` image that sets up the `runner` user and and `main.sh` file expected by Dodona\footnote{\href{https://github.com/dodona-edu/docker-images/blob/master/dodona-r.dockerfile}{https://github.com/dodona-edu/docker-images/blob/master/dodona-r.dockerfile}}.
+In June 2019, R support was added to Dodona using an image derived from the `rocker/r-base` image that sets up the `runner` user and `main.sh` file expected by Dodona\footnote{\href{https://github.com/dodona-edu/docker-images/blob/master/dodona-r.dockerfile}{https://github.com/dodona-edu/docker-images/blob/master/dodona-r.dockerfile}}.
 It also installs the packages required for the testing framework and the exercises so that this does not have to happen every time a student's submission is evaluated.
@@
 PrairieLearn is being developed at the University of Illinois at Urbana-Champaign \citep{prairielearn:2018} and has been in extensive use across several faculties along with initial use on some other campuses.
-It uses Docker containers as key components, both internally for its operations (programmed mainly in Python as well as in Javascript), as well as for two reference containers providing, respectively, Python and R autograders.
+It uses Docker containers as key components, both internally for its operations (programmed mainly in Python as well as in Javascript), as well as for two reference containers providing, respectively, Python and R auto-graders.
 A key design decision made by PrairieLearn permits _external_ grading containers to be supplied and accessed via a well-defined interface of invoking, essentially, a single script, `run.sh`.
@@
 As automated grading of submitted student answers is close to the well-understood problem of unit testing, the \CRANpkg{tinytest} package \citep{CRAN:tinytest} is used for both its core features for testing as well as clean extensibility.
-The package \CRANpkg{ttdo} \citep{CRAN:ttdo} utilizes the extensibility of \pkg{tinytest} to display context-sensitive colorized differences between incorrect answers and reference answers using the \CRANpkg{diffobj} package \citep{CRAN:diffobj}.
+The package \CRANpkg{ttdo} \citep{CRAN:ttdo} utilizes the extensibility of \pkg{tinytest} to display context-sensitive colourized differences between incorrect answers and reference answers using the \CRANpkg{diffobj} package \citep{CRAN:diffobj}.
 Additionally, \CRANpkg{ttdo} addresses the issue of insufficient information collection that Dodona faced by allowing for the collection of arbitrary, test specific attributes for additional logging and feedback.
@@
 Containers give the researcher an isolated environment to assemble these research pipelines with specific versions of software to minimize problems with breaking changes and make workflows easier to share \citep[cf.][]{boettiger_introduction_2015,marwick_packaging_2018}.
-Research workflows in containers are safe from contamination from other activities that occur on the researcher's computer, for example the installation of newest version of packages for teaching demonstrations or specific versions for evaluation of others' works.
+Research workflows in containers are safe from contamination from other activities that occur on the researcher's computer, for example the installation of the newest version of packages for teaching demonstrations or specific versions for evaluation of others' works.
 Given the users in this scenario, i.e., often academics with limited formal software development training, templates and assistance with containers around research compendia is essential.
-In many fields we see that a typical unit of research for a container is a research report or journal article, where the container holds the compendium, or self-contained set of data (or connections to data elsewhere) and code files needed to fully reproduce the article \citep{marwick_packaging_2018}.
+In many fields, we see that a typical unit of research for a container is a research report or journal article, where the container holds the compendium, or self-contained set of data (or connections to data elsewhere) and code files needed to fully reproduce the article \citep{marwick_packaging_2018}.
 The package \pkg{rrtools} ([https://github.com/benmarwick/rrtools](https://github.com/benmarwick/rrtools)) provides a template and convenience functions to apply good practices for research compendia, including a starter `Dockerfile`.
@@
 Some use cases have considerable overlap, which can be expected as a common language and understanding of good practices is still taking shape.
-Also, the ease with which one can create a complex software systems with Docker to serve one's specific needs, such as an independent Docker image stack, leads to parallel developments.
+Also, the ease with which one can create complex software systems with Docker to serve one's specific needs, such as an independent Docker image stack, leads to parallel developments.
 This ease-of-DIY in combination with the difficulty of reusing parts from or composing multiple `Dockerfile`s is a further reason for **fragmentation**.
@@
 
-New features which make complex workflows accessible and reproducible and the variety in packages connected with containerisation, even when they have overlapping features, are a signal and a support for a growing user base.
+New features which make complex workflows accessible and reproducible and the variety in packages connected with containerisation, even when they have overlapping features, are a signal and support for a growing user base.
 This growth is possibly the most important goal for the foreseeable future in the _Rockerverse_, and, just like the Rocker images have matured over years of use and millions of runs, the new ideas and prototypes will have to prove themselves.
@@
 If adoption of containerisation and R continue to grow, the missing pieces for a success predominantly lie in (a) coordination and documentation of activities to reduce repeated work in favour of open collaboration, (b) the sharing of lessons learned from use cases to build common knowledge and language, and (c) a sustainable continuation and funding for development, community support, and education.
-A first concrete effort to work towards these missing pieces should be sustain the structure and captured status quo from this work in form of a _CRAN Task View on containerisation_.
+A first concrete effort to work towards these missing pieces should be sustaining the structure and captured status quo from this work in the form of a _CRAN Task View on containerisation_.
 
@@
 DB wrote the section on \pkg{outsider}.
-GD contributed the CARD.com use case.
+GD contributed to the CARD.com use case.
 RC \& EH contributed to the section on interfaces for Docker in R (_dynverse_ and `dynwrap`).

### 2020-03-09 rockerverse.Rmd
undo 'to'

@@
     # https://github.com/MarkEdmondson1234
-    affiliation: IIH Nordic A/S, Google Developer Expert for GCP
+    affiliation: IIH Nordic A/S, Google Developer Expert for Google Cloud Platform
     email: mark@markedmondson.me
+    address:
+      - Artillerivej 86
+      - 2300 København S, Denmark
+      - \orcid{0000-0002-8434-3881}
   - name: Colin Fay
@@
 DB wrote the section on \pkg{outsider}.
-GD contributed to the CARD.com use case.
+GD contributed the CARD.com use case.
 RC \& EH contributed to the section on interfaces for Docker in R (_dynverse_ and `dynwrap`).

### 2020-03-09 rockerverse.Rmd
googleCloudRunner on CRAN

@@
 [Google Cloud Build](https://cloud.google.com/cloud-build/) and the Google Container Registry are a continuous integration service and an image registry, respectively, that offload building of images to the cloud, while serving the needs of commercial environments such as private Docker images or image stacks.
-As Google Cloud Build itself can run any container, the package \pkg{googleCloudRunner} demonstrates how R can be used as the control language for one-time or batch processing jobs and scheduling of jobs\footnote{\href{https://code.markedmondson.me/googleCloudRunner/articles/cloudbuild.html}{https://code.markedmondson.me/googleCloudRunner/articles/cloudbuild.html}}.
+As Google Cloud Build itself can run any container, the package \CRANpkg{googleCloudRunner} \citep{cran:googleCloudRunner} demonstrates how R can be used as the control language for one-time or batch processing jobs and scheduling of jobs\footnote{\href{https://code.markedmondson.me/googleCloudRunner/articles/cloudbuild.html}{https://code.markedmondson.me/googleCloudRunner/articles/cloudbuild.html}}.
 \label{drake} **\CRANpkg{drake}** is a workflow manager for data science projects \citep{landau_drake_2019}.

### 2020-03-09 rockerverse.tex
fix multiply defined label
googleCloudRunner on CRAN

@@
 \title{The Rockerverse: Packages and Applications for Containerisation with R}
-\author{by Daniel Nüst, Dirk Eddelbuettel, Dom Bennett, Robrecht Cannoodt, Dav Clark, Gergely Daroczi, Mark Edmondson, Colin Fay, Ellis Hughes, Lars Kjeldgaard, Sean Lopp, Ben Marwick, Heather Nolis, Jacqueline Nolis, Hong Ooi, Karthik Ram, Noam Ross, Lori Shepherd, Péter Sólymos, Tyson Lee Swetnam, Nitesh Turaga, Charlotte Van Petegem, Jason Williams, Craig Willis, Nan Xiao}
+\author{by Daniel Nüst, Dirk Eddelbuettel, Dom Bennett, Robrecht Cannoodt, Dav Clark, Gergely Daróczi, Mark Edmondson, Colin Fay, Ellis Hughes, Lars Kjeldgaard, Sean Lopp, Ben Marwick, Heather Nolis, Jacqueline Nolis, Hong Ooi, Karthik Ram, Noam Ross, Lori Shepherd, Péter Sólymos, Tyson Lee Swetnam, Nitesh Turaga, Charlotte Van Petegem, Jason Williams, Craig Willis, Nan Xiao}
 
@@
 The Rocker Project provides widely used Docker images for R across
-different application scenarios. This articles surveys downstream
+different application scenarios. This article surveys downstream
 projects that build upon the Rocker Project images and presents the
@@
 We also introduce related activities that connect the R language and
-environment with other containerisation solutions. Our main
-contributions is a coherent picture of the current status of using
-containers in, with, and for R.
+environment with other containerisation solutions. Our main contribution
+is a coherent picture of the current status of using containers in,
+with, and for R.
 
@@
 or Singularity \citep{kurtzer_singularity_2017}, Docker has become
-synomymous with ``containerisation''---the method of taking software
+synonymous with ``containerisation''---the method of taking software
 artefacts and bundling them in such a way that use becomes standardized
@@
 
-For statistical computing and analysis centered around R, the
+For statistical computing and analysis centred around R, the
 \textbf{Rocker~Project} has provided a variety of Docker containers
@@
 Zuckerberg Initiative's \emph{Essential Open Source Software for
-Science}, which provides funding for the projects's sustainable
+Science}, which provides funding for the project's sustainable
 maintenance, community growth, and targeting new hardware platforms
@@
 Docker is not the only containerisation software. \textbf{Singularity}
-stems from the domain of high-perfomance computing
+stems from the domain of high-performance computing
 \citep{kurtzer_singularity_2017} and can also run Docker images. Rocker
@@
 containers\footnote{See \href{https://github.com/nuest/rodman}{https://github.com/nuest/rodman} and \href{https://github.com/rocker-org/rocker-versioned/issues/187}{https://github.com/rocker-org/rocker-versioned/issues/187}},
-but the prevelance of Docker, especially in the broader user community
+but the prevalence of Docker, especially in the broader user community
 beyond experts or niche systems and the vast amount of blog posts and
@@
 \centering
-\begin{tabular}{lllllll}
+\begin{tabular}{llllllll}
   \toprule
-\rotatebox{-90}{Functionality} & \rotatebox{-90}{AzureContainers} & \rotatebox{-90}{babelwhale} & \rotatebox{-90}{dockermachine} & \rotatebox{-90}{dockyard} & \rotatebox{-90}{harbor} & \rotatebox{-90}{stevedore }\\
+\rotatebox{-90}{Functionality} & \rotatebox{-90}{AzureContainers} & \rotatebox{-90}{babelwhale} & \rotatebox{-90}{dockermachine} & \rotatebox{-90}{dockyard} & \rotatebox{-90}{googleCloudRunner} & \rotatebox{-90}{harbor} & \rotatebox{-90}{stevedore }\\
   \midrule
-Generate a Dockerfile &  &  &  & \checkmark &  &  \\ 
+Generate a Dockerfile &  &  &  & \checkmark &  &  &  \\ 
    \midrule
-Build an image & \checkmark &  &  & \checkmark &  &  \\ 
+Build an image & \checkmark &  &  & \checkmark & \checkmark &  &  \\ 
    \midrule
-Execute a container locally or remotely & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\ 
+Execute a container locally or remotely & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\ 
    \midrule
-Deploy or manage instances in the cloud & \checkmark &  & \checkmark &  & \checkmark & \checkmark \\ 
+Deploy or manage instances in the cloud & \checkmark &  & \checkmark &  & \checkmark & \checkmark & \checkmark \\ 
    \midrule
-Interact with an instance (e.g., file transfer) &  & \checkmark & \checkmark &  &  & \checkmark \\ 
+Interact with an instance (e.g., file transfer) &  & \checkmark & \checkmark &  &  &  & \checkmark \\ 
    \midrule
-Manage storage of images &  &  &  &  & \checkmark & \checkmark \\ 
+Manage storage of images &  &  &  &  &  & \checkmark & \checkmark \\ 
    \midrule
-Supports Docker and Singularity &  & \checkmark &  &  &  &  \\ 
+Supports Docker and Singularity &  & \checkmark &  &  &  &  &  \\ 
    \midrule
-Direct access to Docker API instead of using the CLI &  &  &  &  &  & \checkmark \\ 
+Direct access to Docker API instead of using the CLI &  &  &  &  &  &  & \checkmark \\ 
    \midrule
-Installing Docker software &  &  & \checkmark &  &  &  \\ 
+Installing Docker software &  &  & \checkmark &  &  &  &  \\ 
    \bottomrule
@@
 \textbf{\pkg{harbor}} (\url{https://github.com/wch/harbor}) is no longer
-actively maintained, but it should be honorably mentioned as the first R
-package for managing Docker images and containers. It uses the
+actively maintained, but it should be honourably mentioned as the first
+R package for managing Docker images and containers. It uses the
 \CRANpkg{sys} package \citep{cran_sys} to run system commands against
@@
 a core function such as
-\code{harbor::docker\_run(image = "hello-world")} still works depsite
+\code{harbor::docker\_run(image = "hello-world")} still works despite
 its stopped development.
@@
 The CLI tool \texttt{docker-machine} allows users to create and manage a
-virtual host on local computers, local data centers, or at cloud
-providers. A local Docker installation can be configured to transparntly
-forward all commands issued on the local Docker CLI to a selected
-(remote) virtual host. Docker~Machine was especially crucial for local
-use in the early days of Docker, when no native support was available
-for Mac or Windows computers, but it remains relevant for provisioning
-on remote systems. The package has not received any updates for two
-years, but it is functional with a current version of
+virtual host on local computers, local data centres, or at cloud
+providers. A local Docker installation can be configured to
+transparently forward all commands issued on the local Docker CLI to a
+selected (remote) virtual host. Docker~Machine was especially crucial
+for local use in the early days of Docker, when no native support was
+available for Mac or Windows computers, but it remains relevant for
+provisioning on remote systems. The package has not received any updates
+for two years, but it is functional with a current version of
 \texttt{docker-machine} (\texttt{0.16.2}). It potentially lowers the
@@
 
-\textbf{Bioconductor} (\url{https://bioconductor.org/}) is an open
-source, open development project for the analysis and comprehension of
-genomic data \citep{gentleman_bioconductor_2004}. As of October 30th
+\textbf{Bioconductor} (\url{https://bioconductor.org/}) is an
+open-source, open development project for the analysis and comprehension
+of genomic data \citep{gentleman_bioconductor_2004}. As of October 30th
 2019, the project consists of 1823 R software packages, as well as
@@
 combinations of R and \emph{Bioconductor} will therefore be accessible
-via a specific image tags.
+via specific image tags.
 
@@
 persistently manage installed packages across Docker containers or image
-updates. Many R users pursue flexible work flows tailored to particular
-analysis needs rather than standardized work flows. The new
+updates. Many R users pursue flexible workflows tailored to particular
+analysis needs rather than standardized workflows. The new
 \texttt{bioconductor\_docker} image is well suited for this preference,
@@
 all academic disciplines \citep[e.g.,][]{donoho_50_2017}. These
-discussions have shed a light on the tools and craftspersonship behind
-the analysis of data with computational methods. The practice of data
+discussions have shed light on the tools and craftspersonship behind the
+analysis of data with computational methods. The practice of data
 science often involves combining tools and software stacks and requires
@@
   based on Gigantum's \texttt{python3-minimal} image, originally to keep
-  the existing front-end configuration, but also to provide a consistent
+  the existing front-end configuration, but also to provide consistent
   Python-to-R interoperability. The \texttt{Dockerfile}s also use build
@@
   environment
-  management,\footnote{See \href{https://docs.gigantum.com/docs/environment-management}{https://docs.gigantum.com/docs/environment-management}}.
+  management\footnote{See \href{https://docs.gigantum.com/docs/environment-management}{https://docs.gigantum.com/docs/environment-management}}.
   The image build and publication process is scripted with Python and
@@
 (\url{https://github.com/r-hub/sysreqs/}) package and it's mapping of
-package system dependencies to platform specific installation package
+package system dependencies to platform-specific installation package
 names\footnote{See \href{https://sysreqs.r-hub.io/}{https://sysreqs.r-hub.io/}.}.
@@
 required for rendering the document. The package then generates a
-\texttt{Dockerfile} from this metadata and provides a utitility function
+\texttt{Dockerfile} from this metadata and provides a utility function
 to render the document inside a Docker container, i.e.,
@@
 System dependencies are the domain of Docker, but for a full description
-of the computing environment one must also manage the R version and the
-the R packages. R versions are available via the versioned Rocker image
+of the computing environment, one must also manage the R version and the
+R packages. R versions are available via the versioned Rocker image
 stack \citep{RJ-2017-065}.
@@
 containers as the \emph{only} environment for their workflows, which
-allows them to work indepenently of physical
+allows them to work independently of physical
 computers\footnote{Allowing them to be digital "nomads", cf. J.~Bryan's \href{https://github.com/jennybc/docker-why}{https://github.com/jennybc/docker-why}.}
@@
 These problems are hard to debug and critical, because when they do
-occur they lead to fatal crashing processes.
+occur they lead to fatal crashes.
 \href{https://github.com/rocker-org/r-devel-san}{\texttt{rocker/r-devel-san}}
@@
 in version from a local Git repository is cloned into the container at
-runtime. This approach clearly separates test environment, test code,
-and current working copy of the code. \label{rselenium} Another use case
+runtime. This approach separates the test environment, test code, and
+current working copy of the code. \label{rselenium} Another use case
 where a container can help to standardise tests across operating systems
@@
 This gives, on the one hand, easy access after every update of the R
-Markdown source code and, on the other hand a second controlled
+Markdown source code and, on the other hand, a second controlled
 environment to make sure that the article renders successfully and
@@
 \CRANpkg{future} \citep{cran_future} to enable this with only a few
-lines or R
+lines of R
 code\footnote{\href{https://cloudyr.github.io/googleComputeEngineR/articles/massive-parallel.html}{https://cloudyr.github.io/googleComputeEngineR/articles/massive-parallel.html}}.
@@
 private Docker images or image stacks. As Google Cloud Build itself can
-run any container, the package \pkg{googleCloudRunner} demonstrates how
-R can be used as the control language for one-time or batch processing
-jobs and scheduling of
+run any container, the package \CRANpkg{googleCloudRunner}
+\citep{cran:googleCloudRunner} demonstrates how R can be used as the
+control language for one-time or batch processing jobs and scheduling of
 jobs\footnote{\href{https://code.markedmondson.me/googleCloudRunner/articles/cloudbuild.html}{https://code.markedmondson.me/googleCloudRunner/articles/cloudbuild.html}}.
@@
 to use environment variables, volume mounts, and/or ports that could
-enable transfer of input (parameters and data) and output (results) to
-and from containers. Some packages, e.g., \pkg{containerit}, provide
+enable the transfer of input (parameters and data) and output (results)
+to and from containers. Some packages, e.g., \pkg{containerit}, provide
 Docker images that can be used very similar to a CLI, but this usage is
@@
 tackles the problem of integrating external programs into an R workflow
-without thee need for users to directly interact with containers
+without the need for users to directly interact with containers
 \citep{bennett_outsider_2020}. Installation and usage of external
@@
 and an R package with functions for interacting with the encapsulated
-tool. Using only R functions, an end user can install a module with the
+tool. Using only R functions, an end-user can install a module with the
 \pkg{outsider} package and then call module code to seamlessly integrate
@@
 \citep{cran_shiny}. \emph{ShinyProxy} (\url{https://www.shinyproxy.io/})
-is an open source tool to deploy Shiny apps in an enterprise context,
+is an open-source tool to deploy Shiny apps in an enterprise context,
 where it features single sign-on, but it can also be used in scientific
@@
 
-Another example is the package \CRANpkg{golem}, which makes an heavy use
-of \pkg{dockerfiler} when it comes to creating the \texttt{Dockerfile}
-for building and deploying production-grade Shiny applications
+Another example is the package \CRANpkg{golem}, which makes heavy use of
+\pkg{dockerfiler} when it comes to creating the \texttt{Dockerfile} for
+building and deploying production-grade Shiny applications
 \citep{cran_golem}. \CRANpkg{googleComputeEngineR} enables quick
@@
 with Kubernetes, can mitigate limitations on request volumes or a
-container occupied with computationally intensive task. A cluster
+container occupied with a computationally intensive task. A cluster
 features load-balancing, autoscaling of containers across numerous
@@
 Docker CLI commands and transferred transparently to the respective
-remove machine using SSH. \pkg{AzureContainers} is an umbrella package
+remote machine using SSH. \pkg{AzureContainers} is an umbrella package
 that provides interfaces for three commercial services of Microsoft's
@@
 interfacing R with other programming languages. Yet the approach
-requires a manual management of the computing environment.
+requires manual management of the computing environment.
 
@@
 the deployment of functions and microservices to Kubernetes or Docker
-Swarm. It is language agnostic and provides auto-scaling, metrics, and
+Swarm. It is language-agnostic and provides auto-scaling, metrics, and
 an API gateway. Reduced boilerplate code is achieved via templates.
-Template for
+Templates for
 R\footnote{See OpenFaaS R templates at \href{https://github.com/analythium/openfaas-rstats-templates}{https://github.com/analythium/openfaas-rstats-templates}.}
@@
 containers, as companies already manage platforms in Docker containers.
-These products often entail a large amount of open source software in
+These products often entail a large amount of open-source software in
 combination with proprietary layers adding the relevant
@@
 allows teams of data scientists and their respective IT/DevOps groups to
-develop and deploy code in R and Python around the RStudio Open Source
+develop and deploy code in R and Python around the RStudio Open-Source
 Server inside of Docker images, without requiring users to learn new
@@
 containers. The R session behind \pkg{Rserve} acted as a read-only
-in-memory database, which was extremely fast and scalabale, for the many
+in-memory database, which was extremely fast and scalable, for the many
 concurrent rApache processes responding to the live-scoring requests of
@@
 \pkg{holepunch} removes some of these limitations by leveraging Rocker
-images that contain the Tidyverse along special Jupyter dependencies,
-and only installs additional packages from CRAN and Bioconductor that
-are not already part of these images. It short circuits the
-configuration file parsing in \texttt{repo2docker} and starts with the
-Binder/Tidyverse base images, which eliminates a large part of the build
-time and, in most cases, results in a Binder instance launching within a
-minute. \pkg{holepunch} also creates a \texttt{DESCRIPTION} file for
-essential metadata and dependency specification, and thereby turns turns
-any project into a research compendium (see~\nameref{compendia}). The
-\texttt{Dockerfile} included with the project can also be used to launch
-an RStudio Server instance locally, i.e., independent of Binder, which
-is especially useful when more or special computational resources can be
-provided there. The local image usage reduces the number of separately
-managed environments and, thereby, reduces work and increases
-portability and reproducibility.
+images that contain the Tidyverse along with special Jupyter
+dependencies, and only installs additional packages from CRAN and
+Bioconductor that are not already part of these images. It short
+circuits the configuration file parsing in \texttt{repo2docker} and
+starts with the Binder/Tidyverse base images, which eliminates a large
+part of the build time and, in most cases, results in a Binder instance
+launching within a minute. \pkg{holepunch} also creates a
+\texttt{DESCRIPTION} file for essential metadata and dependency
+specification, and thereby turns any project into a research compendium
+(see~\nameref{compendia}). The \texttt{Dockerfile} included with the
+project can also be used to launch an RStudio Server instance locally,
+i.e., independent of Binder, which is especially useful when more or
+special computational resources can be provided there. The local image
+usage reduces the number of separately managed environments and,
+thereby, reduces work and increases portability and reproducibility.
 
@@
 with \texttt{repo2docker}. Whole~Tale is a National Science Foundation
-(NSF) funded project developing a scalable, open source, multi-user
+(NSF) funded project developing a scalable, open-source, multi-user
 platform for reproducible research \citep{brinckman2019, chard2019a}. A
@@
 
-\href{https://cyverse.org}{CyVerse} is an open source, NSF-funded
+\href{https://cyverse.org}{CyVerse} is an open-source, NSF-funded
 cyberinfrastructure platform for the life sciences providing easy access
@@
 \href{https://learning.cyverse.org/}{CyVerse Learning Center} provides
-community members with information about the platform, incuding training
-and education opportunities.
-
-\label{rocker-gpu} Using \textbf{GPUs} (graphical processing units) as a
+community members with information about the platform, including
+training and education opportunities.
+
+\label{rocker-gpu} Using \textbf{GPUs} (graphical processing units) as
 specialised hardware from containerised common work environments is also
@@
 
-\label{workenvs}
-
 Two use cases demonstrate the practical usefulness and advantages of
@@
 on the other hand leveraging sandboxing and controlled environments for
-autograding.
+auto-grading.
 
@@
 very useful tool for interfacing with databases, because almost every
-open source and proprietary database system has an R package that allows
+open-source and proprietary database system has an R package that allows
 users to connect and interact with it. This flexibility is even
@@
 the \texttt{rocker/r-base} image that sets up the \texttt{runner} user
-and and \texttt{main.sh} file expected by
+and \texttt{main.sh} file expected by
 Dodona\footnote{\href{https://github.com/dodona-edu/docker-images/blob/master/dodona-r.dockerfile}{https://github.com/dodona-edu/docker-images/blob/master/dodona-r.dockerfile}}.
@@
 \citep{prairielearn:2018} and has been in extensive use across several
-faculties along with initial use on some other campuses. ¸It uses Docker
-containers a key components, both internally for its operations
+faculties along with initial use on some other campuses. It uses Docker
+containers as key components, both internally for its operations
 (programmed mainly in Python as well as in Javascript), as well as for
 two reference containers providing, respectively, Python and R
-autograders. A key design decision made by PrairieLearn permits
+auto-graders. A key design decision made by PrairieLearn permits
 \emph{external} grading containers to be supplied and accessed via a
@@
 \texttt{run.sh}. This script relies on a well-defined file layout
-containing JSON-based configuration, support files, exam questions,
+containing JSON-based configurations, support files, exam questions,
 supplementary data, and student submissions. It returns per-question
@@
 \citep{CRAN:ttdo} utilizes the extensibility of \pkg{tinytest} to
-display context-sensitive colorized differences between incorrect
+display context-sensitive colourized differences between incorrect
 answers and reference answers using the \CRANpkg{diffobj} package
-\citep{CRAN:diffobj}. The setup, described in more detail by
+\citep{CRAN:diffobj}. Additionally, \CRANpkg{ttdo} addresses the issue
+of insufficient information collection that Dodona faced by allowing for
+the collection of arbitrary, test specific attributes for additional
+logging and feedback. The setup, described in more detail by
 \citet{paper:r_autograder}, is an excellent illustration of both the
@@
 activities that occur on the researcher's computer, for example the
-installation of newest version of packages for teaching demonstrations
-or specific versions for evaluation of others' works. Given the users in
-this scenario, i.e., often academics with limited formal software
-development training, templates and assistance with containers around
-research compendia is essential. In many fields we see that a typical
-unit of research for a container is a research report or journal
-article, where the container holds the compendium, or self-contained set
-of data (or connections to data elsewhere) and code files needed to
-fully reproduce the article \citep{marwick_packaging_2018}. The package
-\pkg{rrtools} (\url{https://github.com/benmarwick/rrtools}) provides a
-template and convenience functions to apply good practices for research
-compendia, including a starter \texttt{Dockerfile}. Images of compendium
-containers can be hosted on services such as Docker~Hub for convenient
-sharing among collaborators and others. Similarly, packages such as
+installation of the newest version of packages for teaching
+demonstrations or specific versions for evaluation of others' works.
+Given the users in this scenario, i.e., often academics with limited
+formal software development training, templates and assistance with
+containers around research compendia is essential. In many fields, we
+see that a typical unit of research for a container is a research report
+or journal article, where the container holds the compendium, or
+self-contained set of data (or connections to data elsewhere) and code
+files needed to fully reproduce the article
+\citep{marwick_packaging_2018}. The package \pkg{rrtools}
+(\url{https://github.com/benmarwick/rrtools}) provides a template and
+convenience functions to apply good practices for research compendia,
+including a starter \texttt{Dockerfile}. Images of compendium containers
+can be hosted on services such as Docker~Hub for convenient sharing
+among collaborators and others. Similarly, packages such as
 \pkg{containerit} and \pkg{dockerfiler} can be used to manage the
@@
 rather than the most recent version. A more extreme case is the
-\emph{dynverse} project (\url{https://dynverse.org/}), which packages
-over 50 computational methods with different environments (R, Python,
-C++, etc.) in Docker images, which can be executed from R.
-\emph{dynverse} uses a CI platform (see~\nameref{ci}) to build
-Rocker-derived images, test them, and, if the tests succeed, publish
-them on Docker~Hub.
+\emph{dynverse} project
+\citep{saelens_comparisonsinglecelltrajectory_2019}, which packages over
+50 computational methods with different environments (R, Python, C++,
+etc.) in Docker images, which can be executed from R. \emph{dynverse}
+uses a CI platform (see~\nameref{ci}) to build Rocker-derived images,
+test them, and, if the tests succeed, publish them on Docker~Hub.
 
@@
 and understanding of good practices is still taking shape. Also, the
-ease with which one can create a complex software systems with Docker to
+ease with which one can create complex software systems with Docker to
 serve one's specific needs, such as an independent Docker image stack,
@@
 and the variety in packages connected with containerisation, even when
-they have overlapping features, are a signal and a support for a growing
+they have overlapping features, are a signal and support for a growing
 user base. This growth is possibly the most important goal for the
@@
 development, community support, and education. A first concrete effort
-to work towards these missing pieces should be sustain the structure and
-captured status quo from this work in form of a \emph{CRAN Task View on
-containerisation}.
+to work towards these missing pieces should be sustaining the structure
+and captured status quo from this work in the form of a \emph{CRAN Task
+View on containerisation}.
 
@@
 \href{https://www.nsf.gov/awardsearch/showAward?AWD_ID=1541450}{OAC-1541450}.
-We would like to thank Celeste R. Brennecka from the Science Writing
-Support Service of the University of Münster for her editorial support.
+NR is supported in part by the Chan-Zuckerberg Initiative Essential Open
+Source Software for Science program. We would like to thank Celeste R.
+Brennecka from the Science Writing Support Service of the University of
+Münster for her editorial support.
 
@@
 \address{%
-Gergely Daroczi\\
+Gergely Daróczi\\
 \\
@@
 Mark Edmondson\\
-IIH Nordic A/S, Google Developer Expert for GCP\\
-\\
+IIH Nordic A/S, Google Developer Expert for Google Cloud Platform\\
+Artillerivej 86\\ 2300 København S, Denmark\\ \orcid{0000-0002-8434-3881}\\
 }
@@
 EcoHealth Alliance\\
-460 W 34th St., Ste. 1701\\ New York, NY 10001, USA\\ \orcid{0000-0002-0233-1757}\\
+460 W 34th St., Ste. 1701\\ New York, NY 10001, USA\\ \orcid{0000-0002-2136-0000}\\
 }

### 2020-03-11 rockerverse.Rmd
Update rockerverse.Rmd

@@
 These can be used to create and manage arbitrary Docker images and containers, as well as Kubernetes clusters on any platform or cloud service.
+
+**\pkg{googleCloudRunner}** is an interface with [Google Cloud Platform](https://cloud.google.com/) container related services, with tools to make it easier for R users to interact with them for common use cases.
+It includes deployment functions for creating R APIs using the Docker based [Cloud Run](https://cloud.run) service; create long running batch jobs calling any Docker image including Rocker via [Cloud Build](https://cloud.google.com/cloud-build/) and schedule services using [Cloud Scheduler](https://cloud.google.com/scheduler/).
 

### 2020-03-12 rockerverse.Rmd
Full citation on first mention of googleCloudRunner
also one sentence per line and consistent use of -based and -related

--- 2020-03-09 rockerverse.Rmd
@@
 These can be used to create and manage arbitrary Docker images and containers, as well as Kubernetes clusters on any platform or cloud service.
+
+**\CRANpkg{googleCloudRunner}** is an interface with [Google Cloud Platform](https://cloud.google.com/) container-related services, with tools to make it easier for R users to interact with them for common use cases \citep{cran:googleCloudRunner}.
+It includes deployment functions for creating R APIs using the Docker-based [Cloud Run](https://cloud.run) service.
+Users can create long running batch jobs calling any Docker image including Rocker via [Cloud Build](https://cloud.google.com/cloud-build/) and schedule services using [Cloud Scheduler](https://cloud.google.com/scheduler/).
 

### 2020-03-12 rockerverse.tex
update tex file

@@
 platform or cloud service.
+
+\textbf{\CRANpkg{googleCloudRunner}} is an interface with
+\href{https://cloud.google.com/}{Google Cloud Platform}
+container-related services, with tools to make it easier for R users to
+interact with them for common use cases \citep{cran:googleCloudRunner}.
+It includes deployment functions for creating R APIs using the
+Docker-based \href{https://cloud.run}{Cloud Run} service. Users can
+create long running batch jobs calling any Docker image including Rocker
+via \href{https://cloud.google.com/cloud-build/}{Cloud Build} and
+schedule services using \href{https://cloud.google.com/scheduler/}{Cloud
+Scheduler}.
 
@@
 private Docker images or image stacks. As Google Cloud Build itself can
-run any container, the package \CRANpkg{googleCloudRunner}
-\citep{cran:googleCloudRunner} demonstrates how R can be used as the
-control language for one-time or batch processing jobs and scheduling of
+run any container, the package \pkg{googleCloudRunner} demonstrates how
+R can be used as the control language for one-time or batch processing
+jobs and scheduling of
 jobs\footnote{\href{https://code.markedmondson.me/googleCloudRunner/articles/cloudbuild.html}{https://code.markedmondson.me/googleCloudRunner/articles/cloudbuild.html}}.

### 2020-03-12 rockerverse.Rmd
fix contribution by @thebioengineer
See https://github.com/nuest/rockerverse-paper/issues/26#issuecomment-597935834

@@
 GD contributed the CARD.com use case.
-RC \& EH contributed to the section on interfaces for Docker in R (_dynverse_ and `dynwrap`).
+RC contributed to the section on interfaces for Docker in R (_dynverse_ and `dynwrap`).
 DC contributed content on Gigantum.
@@
 CF wrote paragraphs about \pkg{r-online}, \pkg{dockerfiler}, \pkg{r-ci} and \pkg{r-db}.
+EH contributed content on \pkg{dockyard}.
 LK contributed content on \pkg{dockr}.
 SL contributed content on RStudio's usage of Docker.
-BM wrote the section on research compendia and made the project Binder ready.
+BM wrote the section on research compendia and made the project Binder-ready.
 HN \& JN co-wrote the section on the T-Mobile use case.

### 2020-03-12 rockerverse.tex
fix contribution by @thebioengineer
See https://github.com/nuest/rockerverse-paper/issues/26#issuecomment-597935834

@@
 Project, and reviewed all sections. DB wrote the section on
-\pkg{outsider}. GD contributed the CARD.com use case. RC \& EH
-contributed to the section on interfaces for Docker in R
-(\emph{dynverse} and \texttt{dynwrap}). DC contributed content on
-Gigantum. ME contributed to the section on processing and deployment to
-cloud services. CF wrote paragraphs about \pkg{r-online},
-\pkg{dockerfiler}, \pkg{r-ci} and \pkg{r-db}. LK contributed content on
-\pkg{dockr}. SL contributed content on RStudio's usage of Docker. BM
-wrote the section on research compendia and made the project Binder
-ready. HN \& JN co-wrote the section on the T-Mobile use case. KR wrote
-the section about \pkg{holepunch}. NR wrote paragraphs about shared work
-environments and GPUs. LS \& NT wrote the section on Bioconductor. PS
-wrote the paragraphs about GitLab CI/CD pipelines with Shinyproxy and
-OpenFaaS templates. TS \& JW wrote the section on CyVerse. CvP wrote the
-section on the usage of Docker containers in Dodona. CW wrote the
-sections on Whole Tale and contributed content about publication
-reproducibility audits. NX contributed content on \pkg{liftr}. All
-authors approved the final version. This articles was collaboratively
-written at
+\pkg{outsider}. GD contributed the CARD.com use case. RC contributed to
+the section on interfaces for Docker in R (\emph{dynverse} and
+\texttt{dynwrap}). DC contributed content on Gigantum. ME contributed to
+the section on processing and deployment to cloud services. CF wrote
+paragraphs about \pkg{r-online}, \pkg{dockerfiler}, \pkg{r-ci} and
+\pkg{r-db}. EH contributed content on \pkg{dockyard}. LK contributed
+content on \pkg{dockr}. SL contributed content on RStudio's usage of
+Docker. BM wrote the section on research compendia and made the project
+Binder-ready. HN \& JN co-wrote the section on the T-Mobile use case. KR
+wrote the section about \pkg{holepunch}. NR wrote paragraphs about
+shared work environments and GPUs. LS \& NT wrote the section on
+Bioconductor. PS wrote the paragraphs about GitLab CI/CD pipelines with
+Shinyproxy and OpenFaaS templates. TS \& JW wrote the section on
+CyVerse. CvP wrote the section on the usage of Docker containers in
+Dodona. CW wrote the sections on Whole Tale and contributed content
+about publication reproducibility audits. NX contributed content on
+\pkg{liftr}. All authors approved the final version. This articles was
+collaboratively written at
 \href{https://github.com/nuest/rockerverse-paper/}{https://github.com/nuest/rockerverse-paper/}.

### 2020-06-09 README.Rmd

@@
 
-Article about everything related to containers and R, originally based on the blog post http://bit.ly/docker-r and now **available as a preprint on arXiv.org: https://arxiv.org/abs/2001.10641**.
-
-_Want to contribute?_ See https://github.com/nuest/rockerverse-paper/issues/3.
+Article about everything related to containers and R, originally based on [this blog post](http://bit.ly/docker-r) and now **available as a preprint on arXiv.org: https://arxiv.org/abs/2001.10641**.
 
@@
 # - RJournal.sty
-# - RJreferences.bib
+# - rockerverse.bib
 # - RJwrapper.tex
@@
 ```
+
+```{r submission_zipfile, eval=FALSE, include=FALSE}
+# Please upload a zip file of the directory containing a motivating letter (in PDF format) if required, the sources of your article; the *.tex, *.bib and *.R files should have the same file name before the extension, which should be your last or family name in lower-case letters, or the last or family names of the authors joined using hyphens.
+Your title should not include dashes or hyphens.
+Make sure to also include RJwrapper.tex and RJwrapper.pdf.
+zip::zip("rockerverse.zip", c("rockerverse.tex",
+                              "rockerverse.bib",
+                              "rockerverse.R",
+                              "figures/orcid.pdf",
+                              "figures/Rlogo.pdf",
+                              "RJwrapper.tex",
+                              "RJwrapper.pdf"))
+```

### 2020-06-09 rockerverse.Rmd
fix Celeste's affiliation
Added pointers to 1-Click app and webhooks

@@
 The started containers can run on a regular Docker host but also in clusters.
-Continuous integration and deployment (CI/CD) for Shiny applications using Shinyproxy can be achieved, e.g., via [GitLab pipelines](https://docs.gitlab.com/ee/ci/pipelines.html).
+Continuous integration and deployment (CI/CD) for Shiny applications using Shinyproxy can be achieved, e.g., via GitLab pipelines or with a combination of GitHub and Docker Hub.
 A pipeline can include building and checking R packages and Shiny apps.
-After the code has passed the checks, Docker images are built and pushed to GitLab's container registry.
-The pipeline finishes with triggering a webhook on the server, where the
+After the code has passed the checks, Docker images are built and pushed to the container registry.
+The pipeline finishes with triggering a [webhook](https://github.com/analythium/shinyproxy-1-click/blob/master/digitalocean/setup.md#setting-up-webhook) on the server, where the
 deployment script is executed.
-The script updates configurations, pulls the new Docker images, and restarts the services.
+The script can update configurations or pull the new Docker images.
+There is a [ShinyProxy 1-Click App](https://marketplace.digitalocean.com/apps/shinyproxy) in the DigitalOcean marketplace that is set up with
+these webhooks. The [documentation](https://github.com/analythium/shinyproxy-1-click/blob/master/digitalocean/secure.md) explains how to set up HTTPS with ShinyProxy and webhooks.
 
@@
 LS \& NT wrote the section on Bioconductor.
-PS wrote the paragraphs about GitLab CI/CD pipelines with Shinyproxy and OpenFaaS templates.
+PS wrote the paragraphs about CI/CD pipelines with Shinyproxy 1-Click app and OpenFaaS templates.
 TS \& JW wrote the section on CyVerse.
@@
 NR is supported in part by the Chan-Zuckerberg Initiative Essential Open Source Software for Science program.
-We would like to thank Celeste R. Brennecka from the Science Writing Support Service of the University of Münster for her editorial support.
+We would like to thank Celeste R. Brennecka from the Scientific Editing Service of the University of Münster for her editorial support.
 
-\bibliography{RJreferences}
+\bibliography{rockerverse}
 

### 2020-06-11 README.Rmd
add HTML link of arXiv paper to README

@@
 
-Article about everything related to containers and R, originally based on [this blog post](http://bit.ly/docker-r) and now **available as a preprint on arXiv.org: https://arxiv.org/abs/2001.10641**.
+Article about everything related to containers and R, originally based on [this blog post](http://bit.ly/docker-r) and now **available as a preprint on arXiv.org: https://arxiv.org/abs/2001.10641** ([read as HTML](https://www.arxiv-vanity.com/papers/2001.10641/)).
 

### 2020-06-22 rockerverse.Rmd
add missing space

--- 2020-03-12 rockerverse.Rmd
@@
 CyVerse has a browser-based 'data science workbench' called the [Discovery Environment](https://cyverse.org/discovery-environment) (DE).
-The DE uses a combination of [HTCondor](https://research.cs.wisc.edu/htcondor/) and Kubernetes for orchestrating container-based analysis and integrates with external HPC, i.e., [NSF-XSEDE](https://www.xsede.org/), through [TAPIS](https://www.tacc.utexas.edu/tapis)(TACC-API's).
+The DE uses a combination of [HTCondor](https://research.cs.wisc.edu/htcondor/) and Kubernetes for orchestrating container-based analysis and integrates with external HPC, i.e., [NSF-XSEDE](https://www.xsede.org/), through [TAPIS](https://www.tacc.utexas.edu/tapis) (TACC-API's).
 CyVerse hosts a multi-petabyte Data Store based on [iRODS](https://irods.org/) with shared access by its users.

### 2020-08-17 rockerverse.Rmd
Remove spurious newline

--- 2020-06-09 rockerverse.Rmd
@@
 After the code has passed the checks, Docker images are built and pushed to the container registry.
-The pipeline finishes with triggering a [webhook](https://github.com/analythium/shinyproxy-1-click/blob/master/digitalocean/setup.md#setting-up-webhook) on the server, where the
-deployment script is executed.
+The pipeline finishes with triggering a [webhook](https://github.com/analythium/shinyproxy-1-click/blob/master/digitalocean/setup.md#setting-up-webhook) on the server, where the deployment script is executed.
 The script can update configurations or pull the new Docker images.

### 2020-08-17 rockerverse.tex
fix Celeste's affiliation
rename bibliography, add zipfile snippet

@@
 HPC, i.e., \href{https://www.xsede.org/}{NSF-XSEDE}, through
-\href{https://www.tacc.utexas.edu/tapis}{TAPIS}(TACC-API's). CyVerse
+\href{https://www.tacc.utexas.edu/tapis}{TAPIS} (TACC-API's). CyVerse
 hosts a multi-petabyte Data Store based on
@@
 Source Software for Science program. We would like to thank Celeste R.
-Brennecka from the Science Writing Support Service of the University of
+Brennecka from the Scientific Editing Service of the University of
 Münster for her editorial support.
 
-\bibliography{RJreferences}
+\bibliography{rockerverse}
 

### 2020-08-17 rockerverse.tex
update tex

@@
 Continuous integration and deployment (CI/CD) for Shiny applications
-using Shinyproxy can be achieved, e.g., via
-\href{https://docs.gitlab.com/ee/ci/pipelines.html}{GitLab pipelines}. A
-pipeline can include building and checking R packages and Shiny apps.
-After the code has passed the checks, Docker images are built and pushed
-to GitLab's container registry. The pipeline finishes with triggering a
-webhook on the server, where the deployment script is executed. The
-script updates configurations, pulls the new Docker images, and restarts
-the services.
+using Shinyproxy can be achieved, e.g., via GitLab pipelines or with a
+combination of GitHub and Docker Hub. A pipeline can include building
+and checking R packages and Shiny apps. After the code has passed the
+checks, Docker images are built and pushed to the container registry.
+The pipeline finishes with triggering a
+\href{https://github.com/analythium/shinyproxy-1-click/blob/master/digitalocean/setup.md\#setting-up-webhook}{webhook}
+on the server, where the deployment script is executed. The script can
+update configurations or pull the new Docker images. There is a
+\href{https://marketplace.digitalocean.com/apps/shinyproxy}{ShinyProxy
+1-Click App} in the DigitalOcean marketplace that is set up with these
+webhooks. The
+\href{https://github.com/analythium/shinyproxy-1-click/blob/master/digitalocean/secure.md}{documentation}
+explains how to set up HTTPS with ShinyProxy and webhooks.
 
@@
 \texttt{Dockerfile} to be archived with a compendium on a data
-repository (e.g.~\href{https://zenodo.org/}{Zenodo},
+repository (e.g. \href{https://zenodo.org/}{Zenodo},
 \href{https://dataverse.org/}{Dataverse},
@@
 shared work environments and GPUs. LS \& NT wrote the section on
-Bioconductor. PS wrote the paragraphs about GitLab CI/CD pipelines with
-Shinyproxy and OpenFaaS templates. TS \& JW wrote the section on
-CyVerse. CvP wrote the section on the usage of Docker containers in
-Dodona. CW wrote the sections on Whole Tale and contributed content
-about publication reproducibility audits. NX contributed content on
-\pkg{liftr}. All authors approved the final version. This articles was
-collaboratively written at
+Bioconductor. PS wrote the paragraphs about CI/CD pipelines with
+Shinyproxy 1-Click app and OpenFaaS templates. TS \& JW wrote the
+section on CyVerse. CvP wrote the section on the usage of Docker
+containers in Dodona. CW wrote the sections on Whole Tale and
+contributed content about publication reproducibility audits. NX
+contributed content on \pkg{liftr}. All authors approved the final
+version. This articles was collaboratively written at
 \href{https://github.com/nuest/rockerverse-paper/}{https://github.com/nuest/rockerverse-paper/}.
